### 2014-02-24 main.tex
initial

\documentclass[10pt]{article}
\usepackage{fullpage}

\usepackage[numbers]{natbib}

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\usepackage{mathpartir} %for typing rules
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{mathabx} % for boxes
\usepackage{graphicx} %to include png images
\usepackage{xcolor} %for colors
\usepackage{url}

%----------------------------------------------------

\usepackage{prettyref}
\newcommand{\pref}[1]{\prettyref{#1}}
\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
\newcommand{\ppref}[1]{\vpageref[]{#1}}
\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
\newrefformat{cha}{Challenge~\ref{#1}}
\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}

%----------------------------------------------------

\usepackage{amsthm}
\theoremstyle{newstyle}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

%----------------------------------------------------

\input{definitions}

\title{Automatic Amortized Analysis for Imperative Code}
\date{}
\author{}

\begin{document}
\maketitle

\section{Decreasing to zero}

We start with a programming language with loops counting
downwards to 0.  Statements and expressions are defined
below.  Assignments of complex expressions can be simulated
by sequencing assignments of simple expressions.
%
\begin{mathpar}
S := \code{assert}(\Gamma)
\mid x \leftarrow E
\mid \code{while}(0 < x)~S
\mid S;S

\and

E := x + y \mid x - y % \mid y
\end{mathpar}
%
We define a deductive system to derive judgements of the
form $\Gamma; Q \vdash S \dashv Q'; \Delta$ with the following
intended semantics
$$
\forall H_1\, H_2\, S\, n,
    ( H_1 \models \Gamma \land
      (H_1, S) \Downarrow_n H_2
    ) \Rightarrow
    ( H_2 \models \Delta \land
      \Phi_{H_1}(Q) \ge n + \Phi_{H_2}(Q')
    ).
$$

\begin{mathpar}

\Rule{Assert}
{}
{ \Gamma; Q \vdash \code{assert}(\Delta) \dashv Q; \Gamma, \Delta }

\\ \Rule{AssignP}
{ \Gamma \models \Delta[x/x+y] }
{\Gamma; Q + y\cdot q_x \vdash x \leftarrow x + y \dashv Q; \Delta}

\and\Rule{AssignM}
{  \Gamma \models \Delta[x/x-y]
\\ \Gamma \models c \le y \le x
}
{ \Gamma; Q \vdash x \leftarrow x - y \dashv c \cdot q_x + V; \Delta }

\\ \Rule{While}
{ \Gamma, 0 < x ; Q \vdash S \dashv Q; \Gamma }
{ \Gamma; Q + M_w \vdash \code{while}(0 < x)~S \dashv Q; \Gamma }

\and\Rule{Seq}
{  \Gamma; Q \vdash S_1 \dashv Q'; \Delta
\\ \Delta; Q' \vdash S_2 \dashv Q''; \Gamma'
}
{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma' }

\end{mathpar}

\section{Decreasing differences}
The second iteration is a system more symmetric where the potentials
are linear functions in differences of variables.  This time, the while
condition can be a bit more complex.  In the rest, $k$
denotes a non-negative constant.
$$
S := \code{while}(k < x - y) \mid \dots
$$

\end{document}

### 2014-02-24 go.sh

@@ -1,0 +1,3 @@
+#!/bin/sh
+xpdf -remote viewer main.pdf 2>/dev/null &
+./tw main

### 2014-02-24 texwatch.c

@@ -1,0 +1,167 @@
+/*% cc -Wall -o # %
+ * texmon texname [xpdfremote]
+ *
+ * Monitors tex files and compile
+ * them on the fly.
+ *
+ * This program makes use of
+ * the inotify(7) system to
+ * watch tex files and rebuild
+ * them when they change.
+ */
+#include <errno.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <sys/inotify.h>
+#include <sys/select.h>
+#include <sys/wait.h>
+#include <unistd.h>
+
+#define MAX_PATH 1024
+
+/* Define here the sequence of commands
+ * to compile your tex files. The % will
+ * be expanded with the tex file name with
+ * the extension chopped.
+ */
+char *compile[] = {
+	"pdflatex %",
+#if 0
+	"bibtex %",
+	"pdflatex %",
+	"pdflatex %",
+#endif
+	0
+};
+
+char *texname, texfile[MAX_PATH];
+char *reloadcmd;
+int errfd, notfd;
+
+void
+err(char *s)
+{
+	fprintf(stderr, "%s!\n", s);
+}
+
+int
+runcmd(char *c)
+{
+	pid_t pid;
+	char s[MAX_PATH], *p;
+	char *args[4];
+	int ret;
+
+	args[0]="/bin/sh";
+	args[1]="-c";
+	args[3]=0;
+
+	switch (pid=fork()) {
+	case 0:
+		p=strchr(c, '%');
+		if (p)
+			snprintf(s, MAX_PATH, "%.*s%s%s",
+				(int)(p-c), c, texname, p+1);
+		else
+			snprintf(s, MAX_PATH, "%s", c);
+		args[2]=s;
+		close(0);
+		dup2(errfd, 1);
+		dup2(errfd, 2);
+		execv(args[0], args);
+		perror("execv");
+		exit(1);
+	default:
+		while (wait(&ret)!=pid)
+			continue;
+		return WIFEXITED(ret)?WEXITSTATUS(ret):1;
+	case -1:
+		perror("fork");
+		return -1;
+	}
+}
+
+void
+run(void)
+{
+	fd_set fs;
+	int code;
+	char **c;
+	struct inotify_event ev;
+
+	for (;;) {
+		FD_ZERO(&fs);
+		FD_SET(notfd, &fs);
+		if (select(notfd+1, &fs, 0, 0, 0)==-1) {
[…]

### 2014-02-27 main.tex
try difference rules

@@
 \begin{mathpar}
-S := \code{assert}(\Gamma)
-\mid x \leftarrow E
+S := \code{skip}
+\mid \code{assert}(\Gamma)
+\mid x \gets E
 \mid \code{while}(0 < x)~S
+\mid \code{if}(?)~S~\code{else}~S
 \mid S;S
@@
 We define a deductive system to derive judgements of the
-form $\Gamma; Q \vdash S \dashv Q'; \Delta$ with the following
+form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$ with the following
 intended semantics
 $$
-\forall H_1\, H_2\, S\, n,
-    ( H_1 \models \Gamma \land
-      (H_1, S) \Downarrow_n H_2
+\forall H\, H'\, S\, n,
+    ( H \models \Gamma \land
+      (H, S) \Downarrow_n H'
     ) \Rightarrow
-    ( H_2 \models \Delta \land
-      \Phi_{H_1}(Q) \ge n + \Phi_{H_2}(Q')
+    ( H' \models \Gamma' \land
+      \Phi_{H}(Q) \ge n + \Phi_{H'}(Q')
     ).
@@
 
-\Rule{Assert}
+\Rule{Skip}
+{}
+{ \Gamma; Q \vdash \code{skip} \dashv Q; \Gamma }
+
+\and \Rule{Assert}
 {}
@@
 \\ \Rule{AssignP}
-{ \Gamma \models \Delta[x/x+y] }
-{\Gamma; Q + y\cdot q_x \vdash x \leftarrow x + y \dashv Q; \Delta}
+{ \Gamma \models \Gamma'[x/x+y] }
+{\Gamma; Q + y\cdot q_x \vdash x \gets x + y \dashv Q; \Gamma'}
 
 \and\Rule{AssignM}
-{  \Gamma \models \Delta[x/x-y]
+{  \Gamma \models \Gamma'[x/x-y]
 \\ \Gamma \models c \le y \le x
 }
-{ \Gamma; Q \vdash x \leftarrow x - y \dashv c \cdot q_x + V; \Delta }
+{ \Gamma; Q \vdash x \gets x - y \dashv c \cdot q_x + V; \Gamma' }
 
 \\ \Rule{While}
-{ \Gamma, 0 < x ; Q \vdash S \dashv Q; \Gamma }
-{ \Gamma; Q + M_w \vdash \code{while}(0 < x)~S \dashv Q; \Gamma }
+{ \Gamma, 0 < x ; Q - M_w \vdash S \dashv Q; \Gamma }
+{ \Gamma; Q \vdash \code{while}(0 < x)~S \dashv Q; \Gamma, x \le 0 }
 
 \and\Rule{Seq}
-{  \Gamma; Q \vdash S_1 \dashv Q'; \Delta
-\\ \Delta; Q' \vdash S_2 \dashv Q''; \Gamma'
+{  \Gamma; Q \vdash S_1 \dashv Q'; \Gamma'
+\\ \Gamma'; Q' \vdash S_2 \dashv Q''; \Gamma''
 }
-{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma' }
+{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
 
@@
 The second iteration is a system more symmetric where the potentials
-are linear functions in differences of variables.  This time, the while
-condition can be a bit more complex.  In the rest, $k$
+are linear functions in differences of variables.  From now on $k$
 denotes a non-negative constant.
 $$
-S := \code{while}(k < x - y) \mid \dots
+S := \code{while}(k < x - y)~S \mid \dots
 $$
 
+\begin{mathpar}
+
+\Rule{AssignOpK}
+{  \Gamma \models \Gamma'[x/x \pm y], y = k
+\\ \Gamma \models x \pm y \in [x, u]
+\\ \textstyle q'_y = q_y + \sum_u q^1_{xu} - \sum_v q_{xv}
+\\ q_{xu} = q^1_{xu} + q^2_{xu}
+}
+{ \textstyle
+  \Gamma; Q \vdash x \gets x \pm y \dashv \sum_u k \cdot q^2_{xu} + Q'; \Gamma'
+}
+
+\and \Rule{AssignOp}
+{  \Gamma \models \Gamma'[x/x \pm y]
+\\ \Gamma \models x \pm y \in [x, u]
+\\ \textstyle q'_y = q_y + \sum_u q_{xu} - \sum_v q_{xv}
+}
+{ \textstyle
+  \Gamma; Q \vdash x \gets x \pm y \dashv Q'; \Gamma'
+}
+
+\\ \mbox{(The other rules are unchanged)}
+
+\end{mathpar}
+
 \end{document}

### 2014-03-05 main.tex
improve notations
bug in AssignOpK
add simple cost aware semantics

@@
 \begin{mathpar}
+
 S := \code{skip}
@@
 $$
+%
+In the rules below we take as convention that $q_{x}$ for a variable
+$x$ is in fact $q_{x0}$ where 0 is a special variable fixed to zero
+in all contraint contexts.
 
@@
 \Rule{AssignOpK}
-{  \Gamma \models \Gamma'[x/x \pm y], y = k
-\\ \Gamma \models x \pm y \in [x, u]
-\\ \textstyle q'_y = q_y + \sum_u q^1_{xu} - \sum_v q_{xv}
-\\ q_{xu} = q^1_{xu} + q^2_{xu}
+{  \Gamma \models \Gamma'[x/x \pm y]
+\\ \Gamma \models |y| = k
+\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
+\\ \textstyle q'_y = q_y
+      + \sum_{u \in \mathcal U} p_{xu}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+\\ q_{xu} = p_{xu} + r_{xu}
 }
 { \textstyle
-  \Gamma; Q \vdash x \gets x \pm y \dashv \sum_u k \cdot q^2_{xu} + Q'; \Gamma'
+  \Gamma; Q
+  \vdash x \gets x \pm y \dashv
+  \sum_u k \cdot r_{xu} + Q'; \Gamma'
+  % q_x is not omitted, v_{x} is in fact a shorthand for
+  % v_{x0} where 0 is a variable equal to 0.
 }
@@
 {  \Gamma \models \Gamma'[x/x \pm y]
-\\ \Gamma \models x \pm y \in [x, u]
-\\ \textstyle q'_y = q_y + \sum_u q_{xu} - \sum_v q_{xv}
+\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
+\\ \textstyle q'_y = q_y
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{xv}
 }
@@
 
+\section{Cost aware semantics}
+
+This big step semantics counts the number of iterations of
+while loops in a program as it executes it.  Non-deterministically,
+the {\sc S:Bubble} rule can be called to interrupt a computation.
+It has the same effect as raising an exception.  We symbolize an
+immutable heap with the notation $\circ H$, this kind of heap just
+bubbles up through the semantics. In the rules, $X$ stands for
+either $H$ or $\circ H$ for some $H$.
+
+\begin{mathpar}
+
+\Rule{S:Bubble}
+{}
+{ (H, S) \Downarrow_0 \circ H }
+
+\and \Rule{S:Skip}
+{}
+{ (H, \code{skip}) \Downarrow_0 H }
+
+\and \Rule{S:Assign}
+{}
+{ (H, x \gets e) \Downarrow_0 H[x \gets \llbracket e \rrbracket_H] }
+
+\and \Rule{S:Assert}
+{ \Gamma \models H }
+{ (H, \code{assert}(\Gamma)) \Downarrow_0 H }
+
+\and \Rule{S:WhileStop}
+{ \neg (k < \llbracket x - y \rrbracket_H) }
+{ (H, \code{while}(k < x - y)~S) \Downarrow_0 H }
+
+\and \Rule{S:WhileStep}
+{  k < \llbracket x - y \rrbracket_H
+\\ (H, S; \code{while}(k < x - y)~S) \Downarrow_n X
+}
+{ (H, \code{while}(k < x - y)~S) \Downarrow_{n+1} X }
+
+\and \Rule{S:If}
+{  (H, S_i) \Downarrow_n X
+\\ i \in \{1, 2\}
+}
+{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_n X }
+
+\and \Rule{S:Seq}
+{  (H, S_1) \Downarrow_{n_1} H'
+\\ (H', S_2) \Downarrow_{n_2} H''
+}
+{ (H, S_1; S_2) \Downarrow_{n_1 + n_2} H'' }
+
+\and \Rule{S:SeqBubble}
+{ (H, S_1) \Downarrow_n \circ H' }
+{ (H, S_1; S_2) \Downarrow_n \circ H' }
+
+\end{mathpar}
+
 \end{document}

### 2014-03-06 main.tex
fix bug in semantics
more tolls

@@
 {}
-{ \Gamma; Q \vdash \code{skip} \dashv Q; \Gamma }
+{ \Gamma; Q + M_s \vdash \code{skip} \dashv Q; \Gamma }
 
@@
 {}
-{ \Gamma; Q \vdash \code{assert}(\Delta) \dashv Q; \Gamma, \Delta }
+{ \Gamma; Q + M_a \vdash \code{assert}(\Delta) \dashv Q; \Gamma, \Delta }
 
-\\ \Rule{AssignP}
-{ \Gamma \models \Gamma'[x/x+y] }
-{\Gamma; Q + y\cdot q_x \vdash x \gets x + y \dashv Q; \Gamma'}
+% \\ \Rule{AssignP}
+% { \Gamma \models \Gamma'[x/x+y] }
+% {\Gamma; Q + M_s + y\cdot q_x \vdash x \gets x + y \dashv Q; \Gamma'}
 
-\and\Rule{AssignM}
-{  \Gamma \models \Gamma'[x/x-y]
-\\ \Gamma \models c \le y \le x
-}
-{ \Gamma; Q \vdash x \gets x - y \dashv c \cdot q_x + V; \Gamma' }
+% \and\Rule{AssignM}
+% {  \Gamma \models \Gamma'[x/x-y]
+% \\ \Gamma \models c \le y \le x
+% }
+% { \Gamma; Q + M_s \vdash x \gets x - y \dashv c \cdot q_x + V; \Gamma' }
 
 \\ \Rule{While}
-{ \Gamma, 0 < x ; Q - M_w \vdash S \dashv Q; \Gamma }
-{ \Gamma; Q \vdash \code{while}(0 < x)~S \dashv Q; \Gamma, x \le 0 }
+{ \Gamma, 0 < x ; Q - M_w^2 \vdash S \dashv Q; \Gamma }
+{ \Gamma; Q + M_w^1 \vdash \code{while}(0 < x)~S \dashv Q - M_w^3; \Gamma, x \le 0 }
 
 \and\Rule{Seq}
-{  \Gamma; Q \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma'; Q' \vdash S_2 \dashv Q''; \Gamma''
+{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
+\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
 }
@@
 { \textstyle
-  \Gamma; Q
+  \Gamma; Q + M_s
   \vdash x \gets x \pm y \dashv
@@
 { \textstyle
-  \Gamma; Q \vdash x \gets x \pm y \dashv Q'; \Gamma'
+  \Gamma; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma'
 }
@@
 while loops in a program as it executes it.  Non-deterministically,
-the {\sc S:Bubble} rule can be called to interrupt a computation.
-It has the same effect as raising an exception.  We symbolize an
-immutable heap with the notation $\circ H$, this kind of heap just
+the {\sc S:Abort} rule can be called to interrupt a computation.
+It has the same effect as raising an exception.  We symbolize a
+frozen heap with the notation $\circ H$, this kind of heap just
 bubbles up through the semantics. In the rules, $X$ stands for
-either $H$ or $\circ H$ for some $H$.
+either $\circ$ or $H$.
 
@@
 
-\Rule{S:Bubble}
-{}
-{ (H, S) \Downarrow_0 \circ H }
+\Rule{S:Abort}
+{ }
+{ (H, S) \Downarrow_0 \circ }
 
 \and \Rule{S:Skip}
-{}
-{ (H, \code{skip}) \Downarrow_0 H }
+{ }
+{ (H, \code{skip}) \Downarrow_{M_s} H }
 
 \and \Rule{S:Assign}
-{}
-{ (H, x \gets e) \Downarrow_0 H[x \gets \llbracket e \rrbracket_H] }
+{ }
+{ (H, x \gets e) \Downarrow_{M_s} H[x \mapsto \llbracket e \rrbracket_H] }
 
 \and \Rule{S:Assert}
-{ \Gamma \models H }
-{ (H, \code{assert}(\Gamma)) \Downarrow_0 H }
+{ H \models \Gamma }
+{ (H, \code{assert}(\Gamma)) \Downarrow_{M_a} H }
 
@@
 { \neg (k < \llbracket x - y \rrbracket_H) }
-{ (H, \code{while}(k < x - y)~S) \Downarrow_0 H }
+{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^3} H }
 
@@
 {  k < \llbracket x - y \rrbracket_H
-\\ (H, S; \code{while}(k < x - y)~S) \Downarrow_n X
+\\ (H, S) \Downarrow_{q_1} H'
+\\ (H', \code{while}(k < x - y)~S) \Downarrow_{q_2} X
 }
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{n+1} X }
+{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot q_1 \cdot M_w^2 \cdot q_2} X }
+
+\and \Rule{S:WhileStepAbort}
+{  k < \llbracket x - y \rrbracket_H
+\\ (H, S) \Downarrow_{q_1} \circ
+}
+{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot q_1 } \circ }
 
 \and \Rule{S:If}
-{  (H, S_i) \Downarrow_n X
+{  (H, S_i) \Downarrow_q X
 \\ i \in \{1, 2\}
 }
-{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_n X }
+{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_{M_c^i \cdot q} X }
 
 \and \Rule{S:Seq}
-{  (H, S_1) \Downarrow_{n_1} H'
-\\ (H', S_2) \Downarrow_{n_2} H''
+{  (H, S_1) \Downarrow_{q_1} H'
+\\ (H', S_2) \Downarrow_{q_2} X
 }
-{ (H, S_1; S_2) \Downarrow_{n_1 + n_2} H'' }
+{ (H, S_1; S_2) \Downarrow_{M_s^1 \cdot q_1 \cdot M_s^2 \cdot q_2} X }
 
-\and \Rule{S:SeqBubble}
-{ (H, S_1) \Downarrow_n \circ H' }
-{ (H, S_1; S_2) \Downarrow_n \circ H' }
+\and \Rule{S:SeqAbort}
+{ (H, S_1) \Downarrow_q \circ }
+{ (H, S_1; S_2) \Downarrow_q \circ }
 

### 2014-03-07 main.tex
use explicit pairs for resource monoid
add soundness part 1

@@
 It has the same effect as raising an exception.  We symbolize a
-frozen heap with the notation $\circ H$, this kind of heap just
-bubbles up through the semantics. In the rules, $X$ stands for
-either $\circ$ or $H$.
+frozen computation with the notation $\circ$.  In the rules, $X$
+stands for either $\circ$ or $H$.
 
@@
 {  k < \llbracket x - y \rrbracket_H
-\\ (H, S) \Downarrow_{q_1} H'
-\\ (H', \code{while}(k < x - y)~S) \Downarrow_{q_2} X
+\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
+\\ (H', \code{while}(k < x - y)~S) \Downarrow_{(q_2, q'_2)} X
 }
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot q_1 \cdot M_w^2 \cdot q_2} X }
+{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }
 
@@
 {  k < \llbracket x - y \rrbracket_H
-\\ (H, S) \Downarrow_{q_1} \circ
+\\ (H, S) \Downarrow_{(q_1, q'_1)} \circ
 }
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot q_1 } \circ }
+{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) } \circ }
 
 \and \Rule{S:If}
-{  (H, S_i) \Downarrow_q X
+{  (H, S_i) \Downarrow_{(q, q')} X
 \\ i \in \{1, 2\}
 }
-{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_{M_c^i \cdot q} X }
+{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_{M_c^i \cdot (q, q')} X }
 
 \and \Rule{S:Seq}
-{  (H, S_1) \Downarrow_{q_1} H'
-\\ (H', S_2) \Downarrow_{q_2} X
+{  (H, S_1) \Downarrow_{(q_1, q'_1)} H'
+\\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
 }
-{ (H, S_1; S_2) \Downarrow_{M_s^1 \cdot q_1 \cdot M_s^2 \cdot q_2} X }
+{ (H, S_1; S_2) \Downarrow_{M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)} X }
 
 \and \Rule{S:SeqAbort}
-{ (H, S_1) \Downarrow_q \circ }
-{ (H, S_1; S_2) \Downarrow_q \circ }
+{ (H, S_1) \Downarrow_{(q,q')} \circ }
+{ (H, S_1; S_2) \Downarrow_{(q,q')} \circ }
 
@@
 
+\section{Soundness theorem}
+
+\begin{theorem}[Terminating case]
+For any $S$, $H$, $H'$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$ such
+that $\Gamma; Q \vdash S \dashv Q'; \Gamma'$ and $H \models \Gamma$ and
+$(H,s) \Downarrow_{(q,q')} H'$, we have
+$H' \models \Gamma'$ and
+$q \le \Phi_H(Q)$ and
+$\Phi_H(Q) \ge q' + \Phi_{H'}(Q')$.
+\end{theorem}
+
 \end{document}

### 2014-03-10 main.tex
start soundness proof

@@
 \usepackage{url}
+\usepackage{enumitem}
 
@@
 We define a deductive system to derive judgements of the
-form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$ with the following
-intended semantics
-$$
-\forall H\, H'\, S\, n,
-    ( H \models \Gamma \land
-      (H, S) \Downarrow_n H'
-    ) \Rightarrow
-    ( H' \models \Gamma' \land
-      \Phi_{H}(Q) \ge n + \Phi_{H'}(Q')
-    ).
-$$
+form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.  The semantics
+of this judgement are defined below in the soundness section.
 
@@
 $q \le \Phi_H(Q)$ and
-$\Phi_H(Q) \ge q' + \Phi_{H'}(Q')$.
+$\Phi_H(Q) - \Phi_{H'}(Q') \ge q - q'$.
 \end{theorem}
 
+\begin{proof} By induction on the big step semantics.
+
+\begin{itemize}
+
+\item Case {\sc S:Skip}.
+  We have $(H, \code{skip}) \Downarrow_{M_s} H$.
+  By assumption, $H \models \Gamma$. \\
+  Since $\Phi_H(Q + M_s) = \Phi_H(Q) + M_s$,
+  \begin{itemize}[topsep=0pt]
+  \item
+    if $M_s \ge 0$ we have,
+      $M_s \le \Phi_H(Q + M_s)$ and
+      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge M_s - 0$;
+  \item
+    if $M_s < 0$ we have,
+      $0 \le \Phi_H(Q + M_s)$ and
+      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge 0 - (-M_s)$.
+  \end{itemize}
+
+\item Case {\sc S:Seq}.
+  By transitivity and induction hypothesis, it is
+  easy to get $H'' \models \Gamma''$.  Now we prove
+  the quantitative part of the theorem.  We have
+  $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
+  We assume $\forall i \in \{1, 2\},~M_s^i \ge 0$.
+  \begin{itemize}[topsep=0pt]
+  \item
+    The first case is when $q'_1 > M_s^2 + q_2$ then $q = M_s^1 + q_1$
+    and since $q_1 \le \Phi_H(Q - M_s^1)$ (by induction) we have
+    $q \le \Phi_H(Q)$.
+  \item
+    The second case is when $q'_1 \le M_s^2 + q_2$ then $q = M_s^1 + q_1 + M_s^2 + q_2 - q'_1$. \\
+    Since $\Phi_H(Q - M_s^1) - \Phi_{H'}(Q') \ge q_1 - q'_1$
+    and $q_2 \le \Phi_{H'}(Q' - M_s^2)$ we have:
+    \begin{align*}
+    q &\le M_s^1 + \Phi_H(Q - M_s^1) - \Phi_{H'}(Q') + M_s^2 + q_2 \\
+      &= \Phi_H(Q) + (q_2 - \Phi_{H'}(Q' - M_s^2)) \le \Phi_H(Q).
+   \end{align*}
+  \end{itemize}
+  We obtain $\Phi_H(Q) - \Phi_{H''}(Q'') \ge q - q'$ by simply chaining
+  the two induction hypothesis and using the fact that $q - q' =
+  (q_1 + M_s^1 - q'_1) + (q_2 + M_s^2 - q'_2)$.
+
+\item Case {\sc S:WhileStop}.
+\item Case {\sc S:WhileStep}.
+
+\end{itemize}
+\end{proof}
+
 \end{document}

### 2014-03-10 main.tex
more cases

@@
 \\ \Rule{While}
-{ \Gamma, 0 < x ; Q - M_w^2 \vdash S \dashv Q; \Gamma }
-{ \Gamma; Q + M_w^1 \vdash \code{while}(0 < x)~S \dashv Q - M_w^3; \Gamma, x \le 0 }
+{ \Gamma, 0 < x ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
+{ \Gamma; Q \vdash \code{while}(0 < x)~S \dashv Q - M_w^3; \Gamma, x \le 0 }
 
@@
 
-This big step semantics counts the number of iterations of
-while loops in a program as it executes it.  Non-deterministically,
-the {\sc S:Abort} rule can be called to interrupt a computation.
+This big step semantics is parameterized by a metric that sets
+the cost of each program operation.  Non-deterministically,
+the rule {\sc S:Abort} can be called to interrupt a computation.
 It has the same effect as raising an exception.  We symbolize a
@@
   $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
-  We assume $\forall i \in \{1, 2\},~M_s^i \ge 0$.
+  We assume $\forall i \in \{1, 2\},~M_s^i \ge 0$, the other cases
+  are similar.
   \begin{itemize}[topsep=0pt]
@@
 \item Case {\sc S:WhileStop}.
+  This proof goes just like in the {\sc S:Skip} case.
+
 \item Case {\sc S:WhileStep}.
+  Here again we assume $\forall i \in \{1, 2, 3\},~M_w^i \ge 0$.
+  \begin{itemize}[topsep=0pt]
+  \item
+    If $q'_1 > M_w^2 + q_2$ then $q = M_w^1 + q_1$ and by induction
+    we get $q_1 \le \Phi_H(Q - M_w^1)$, so $q \le \Phi_H(Q)$.
+  \item
+    If $q'_1 \le M_w^2 + q_2$ then $q = M_w^1 + q_1 + M_w^2 + q_2 - q'_1$.
+    And we proceed like in the sequence case.
+  \end{itemize}
+  We obtain $\Phi_H(Q) - \Phi_{H''}(Q - M_w^3) \ge q - q'$ using the identity
+  $q - q' = (M_w^1 + M_w^2 + q_1 - q'_1) + q_2 - q'_2$ and the two
+  induction hypothesis
+  \begin{itemize}[topsep=0pt]
+  \item $\Phi_H(Q - M_w^1) - \Phi_{H'}(Q + M_w^2) \ge q_1 - q'_1$ and
+  \item $\Phi_{H'}(Q) - \Phi_{H''}(Q - M_w^3) \ge q_2 - q'_2.$
+  \end{itemize}
 

### 2014-04-07 main.tex
just a note
started with polynomial annotations

@@
+% Some notes:
+%  1) Use Appel's SHA-256 implementation as an example
+%
+
 \documentclass[10pt]{article}
@@
 
+\section{Polynomial Potential}
+
+\paragraph{Index Sets}
+
+Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
+family that maps two-element sets of variables to natural numbers,
+that is,
+$$
+I = (i_{\{x,y\}})_{\{x,y\} \subseteq V} \; .
+$$
+%
+We identify a family $I$ with the set $\{ (\{x,y\},i_{\{x,y\}})
+\mid \{x,y\} \subseteq V\}$.
+
+Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
+instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
+from the context.
+%
+We assume that allways $0 \in V$ and sometimes write $i_x$ instead of $i_{\{x,0\}}$.
+
+The \emph{degree} $\deg(I)$ of an index $I = (i_{\{x,y\}})_{\{x,y\}
+  \subseteq V}$ is defined as
+$$
+\deg(I) = \sum_{\{x,y\} \in V} i_{\{x,y\}} \;.
+$$
+We define $\ind_k(V) = \{ I \mid I \in \ind(V) \text{ and } \deg(i) \leq k
+\}$ to be the set of indices of degree at most $k$.
+
+\paragraph{Resource Polynomials}
+
+Let $V$ be a set of variables.  An index $I \in \ind(V)$ denotes a
+\emph{base polynomial} $P_I : \states \to \N$ for $V$ that maps a
+program state $H$ to product of binomial coefficients (a natural
+number).  We define
+$$
+P_I(H) = \prod_{{\{x,y\}} \subseteq V} \binom {|H(x){-}H(y)|} {i_{\{x,y\}}} \; .
+$$
+%
+A \emph{resource polynomial} $R$ for the variable set $V$ is a
+non-negative linear combination of the base polynomials for $V$.
+
+\paragraph{Potential Annotations}
+
+A \emph{potential annotation} for the variable set $V$ is a family
+$$Q = (q_I)_{I \in \ind(V)}$$ 
+of non-negative rational numbers.  Such an annotation denotes the
+resource polynomial $R_Q$ that is defined by
+$$
+R_Q(H) = \sum_{I \in \ind(V)} q_I \cdot P_I(H) \; .
+$$
+%
+We say that the potential annotation $Q$ is of degree $k$ if $q_I = 0$
+for $I \in \ind(V)$ with $\deg(I) > k$.
+
+\paragraph{Additive Shifts}
+
+Let $Q$ be a potential annotation for a variable set $V$ and let
+$\{x,y\} \subseteq V$ be a two-element variable set.  The
+\emph{additive shift} with respect to $\{x,y\}$ is a potential
+annotation $\shift_{\{x,y\}}(Q) = (q'_I)_{I \in \ind(V)} $ for $V$
+that is defined through
+$$
+q'_I = q_I + q_{I^{\{x,y\}{+}1}} \; .
+$$
+For an index $I = (i_{\{x,y\}})_{\{x,y\} \subseteq V}$ we use the
+notation $I^{\{x,y\}{+}k}$ to denote the index
+$(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
+$$
+i'_{\{t,u\}} = \left\{
+  \begin{array}{ll}
+    i_{\{t,u\}} + k  & \text{if } \{t,u\} = \{x,y\} \\
+    i_{\{t,u\}} & \text{otherwise}
+  \end{array}
+\right.
+\;.
+$$
+%
+The additive shift for natural numbers reflects the identity 
+\begin{equation}
+\label{eq:shift}
+\sum_{0 {\leq} i \leq {k}} q_i \binom{n+1}{i} = \sum_{0 {\leq} i \leq {k}} (q_i{+}q_{i+1}) \binom{n}{i}
+\end{equation}
+where $q_{k+1} = 0$.  It is used in the effect system if the
+difference $n+1$ between two variables $x,y$ decreases by one.
+
+\begin{lemma}
+  Let $V$ be a set of variables with $x,y \in V$ and let $H$ be a
+  program state. Let $H'(z) = H(z)$ for $z \not\in \{x,y\}$ and let
+  $|H'(x) {-} H'(y)| = |H(x) {-} H(y)| - 1$.
+  %
+  If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
+\end{lemma}
+
+%%
+% Note: Later we need to shift in many directions at once like 
+%   Q' = shift_{x,y} (shift_{x,u} (Q))
+% To do: Give a combined formula for that (concise constraint system).
+%%
+
 \end{document}

### 2014-04-07 main.tex
more shifts

@@
 
-\begin{lemma}
-  Let $V$ be a set of variables with $x,y \in V$ and let $H$ be a
-  program state. Let $H'(z) = H(z)$ for $z \not\in \{x,y\}$ and let
-  $|H'(x) {-} H'(y)| = |H(x) {-} H(y)| - 1$.
+\begin{lemma} Let $V$ be a set of variables with $x,y \in V$ and let
+  $H$ be a program state. Let $|H'(t) {-} H'(u)| = |H(t) {-} H(u)|$
+  for $\{t,u\} \neq \{x,y\}$ and let $|H'(x) {-} H'(y)| = |H(x) {-}
+  H(y)| - 1$.
   %
   If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
+\end{lemma}
+
+We now study the effect of multiple simultaneous shifts.  Let $Q$ be a
+resource annotation for a variable set $V$ and let $U_1,\ldots,U_n
+\subseteq V$ with $|U_i| = 2$ for all $i$ and $U_i \neq U_j$ for $i
+\neq j$ be pairwise distinct two-element variable sets.  The
+simulations additive shift $\shift_{U_1,\ldots,U_n}(Q)$ of $Q$ with
+respect to $U_1,\ldots,U_n$ is defined by
+$$
+\shift_{U_1,\ldots,U_n}(Q) = \shift_{U_1}( \cdots \shift_{U_n}(Q) \cdots ) \; .
+$$
+%
+\begin{proposition}
+  Let $V$ be a set of variables and let $U_1,\ldots,U_n$ be pairwise
+  distinct two-element variable sets.  Let $|H'(x) {-} H'(y)| = |H(x)
+  {-} H(y)|$ for $\{x,y\} \not\in \{U_1,\ldots,U_n\}$ and let $|H'(x)
+  {-} H'(y)| = |H(x) {-} H(y)| - 1$ for $\{x,y\} \in
+  \{U_1,\ldots,U_n\}$.
+  %
+  If $Q' = \shift_{\{U_1,\ldots,U_n\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
+\end{proposition}
+%
+As shown by the following lemma, the order in which the shifts for the
+individual $U_i$ are applied is insignificant.
+%
+\begin{lemma}
+  Let $\sigma : \{1,\ldots,n\} \to \{1,\ldots,n\}$ be a
+  permutation. Then $\shift_{U_1,\ldots,U_n}(Q) =
+  \shift_{U_{\sigma(1)},\ldots,U_{\sigma(n)}}(Q)$.
+\end{lemma}
+%
+For reasons of efficiency in the constraint generation, we give a more
+direct formula for the simultaneous shift.  Let $I \in \ind(V)$ and
+let $U_1,\ldots,U_n$ be pairwise distinct two-element variable sets.
+We define the index $I^{U_1,\ldots,U_n + k}$ as the family $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
+$$
+i'_{\{t,u\}} = \left\{
+  \begin{array}{ll}
+    i_{\{t,u\}} + k  & \text{if } \{t,u\} \in \{U_1,\ldots,U_n\} \\
+    i_{\{t,u\}} & \text{otherwise}
+  \end{array}
+\right.  \;.
+$$
+
+%
+\begin{lemma}
+  Let $V$ be a variable set and let $U_1,\ldots,U_n$ be pairwise
+  distinct two-element variable sets.
+  %
+  Let $Q = (q_I)_{I \in \ind(V)}$ be a resource annotation for
+  $V$ and let $ Q' = (q'_I)_{I \in \ind(V)}$ where
+  $$
+  q'_I = \sum_{\{j_1,\ldots,j_m\} \subseteq \{1,\ldots,n\} } q_{I^{U_{j_1},\ldots,U_{j_m}+1}} \; .
+  $$
+  Then $Q' = \shift_{U_1,\ldots,U_n}(Q)$.
 \end{lemma}

### 2014-05-14 main.tex
typo
merge with 690 report

@@
 \usepackage{amsthm}
-\theoremstyle{newstyle}
 \newtheorem{lemma}{Lemma}
@@
 
-\section{Decreasing to zero}
+\section{Introduction}
 
-We start with a programming language with loops counting
-downwards to 0.  Statements and expressions are defined
-below.  Assignments of complex expressions can be simulated
-by sequencing assignments of simple expressions.
+After functional correctness, quantitative properties are the
+second thing one wants to know when writing a program.
+For example, an asymptotic runtime bound on a
+sorting algorithm tells us what efficiency we can expect
+of it when running on very large arrays.  But, what if we
+always run it on arrays of size between 5 and 10? In that
+case, the asymptotic is not relevant and we need a more
+precise bound on the runtime.  It is then a good
+software engineering practice to profile the algorithm on
+some examples.
+
+In between asymptotic bounds and profiling there are
+precise bounds, for instance, one can prove that Knuth's Quicksort might
+take at most $20N^2$ operations of a MIX machine when
+called on an array of size $N$.  If you have this information,
+no profiling is needed to compare the algorithms and you
+have a formal guarantee on the behavior of the program. But what if you do not
+have a MIX machine, or you tweaked Knuth's algorithm
+a bit?  Then you are good for another long an tedious
+proof.  We advocate that, it is possible to design a practical
+system that is able to infer precise worst case bounds
+without any user interaction.
+
+This work presents a resource oblivious system to get
+linear resource bounds on imperative programs with
+integers.  Despite its apparent simplicity it is able to
+reproduce multiple results from the literature which
+were obtained using methods based on abstract
+interpretation.  In stark contrast with abstract-interpretation
+based methods, our tool does not require any
+fixpoint computations on an abstract domain to obtain
+loop invariants.  The mechanism we designed, which
+is based on amortized analysis, is able to leverage a local
+information to obtain global invariants.  Additionally, the
+approach we advocate is able to scale to full blown
+imperative languages including constructs like functions
+calls, \code{break} and \code{return}.
+
+The analysis is proved sound with respect to
+the operational semantics instrumented to track resource
+consumption.  It is implemented as a prototype
+in OCaml and available on demand.  We test it on multiple
+algorithms from the literature and on custom tests, it
+is also planned to extend it to handle the full C programming
+language.
+
+\section{Overview and Examples}
+
+\subsection{Amortized Analysis}
+The idea that drives the design of our system is amortized
+analysis.  Say that a program $p$ evaluates on a heap $H$
+by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
+The basic idea of amortized analysis is to define a potential function
+$\Phi$ from heaps to natural numbers, such that $\Phi(H) \ge n$.
+
+Since imperative programs can be composed we must enrich
+the judgement $(\_, \_) \Downarrow_{\_}$ to talk about the heap
+resulting from the program execution $H'$.  So we now write
+$(p, H) \Downarrow_n H'$ to say that $H'$ results from $p$'s
+execution.  We now use two potential functions, one
+that applies before the execution, and one that applies
+after.  The two functions must respect the natural relation
+$\Phi(H) \ge n + \Phi'(H')$.  Intuitively speaking, $\Phi$ must
+give enough \emph{potential} to, both, finish the computation
+and pay for $\Phi'$ on the resulting heap. That way, if $(H, p_1) \Downarrow_n H'$
+and $(H', p_2) \Downarrow_m H''$, we get $\Phi(H) \ge
+n + \Phi'(H')$ and $\Phi'(H') \ge m + \Phi''(H'')$, which
+can be composed to give $\Phi(H) \ge (n + m) + \Phi''(H'')$.
+What we have observed is that, if we define $\Phi \vdash p \dashv \Phi'$
+to be
+$$
+\forall H\,n\,H',~(H, p)\Downarrow_n H' \implies \Phi(H) \ge
+n + \Phi'(H'),
+$$
+then we get the following familiar looking rule:
+$$
+\RuleNolabel
+{\Phi \vdash p_1 \dashv \Phi' \\ \Phi' \vdash p_2 \dashv \Phi''}
+{\Phi \vdash p_1; p_2 \dashv \Phi''}
+.
+$$
+Similarly, other language constructs contrain the potential
+functions to respect more rules that look very similar to
+Hoare logic or effect systems rules.  By fixing the shape of
+a potential function $\Phi$, it becomes possible to leverage
+existing tools to help finding a pair of potential functions
+for an arbitrary program.  This is explained in further details
+in the following sections.
+
+We say that the analysis is amortized because the use of a
+potential function $\Phi$ gives the flexibility to amortize some
+operations.  For example, if heaps are simple mappings from
+$\{\code x, \code y\}$ to natural numbers, we have the following
+judgement (provided that $H~\code x \ge H~\code y \ge 0$ initially)
+$$
+(\Phi := \lambda H.~2\cdot(H~\code x) + 10)
+\vdash \code x = \code x - \code y \dashv
+(\Phi' := \lambda H'.~2\cdot(H'~\code x) + 2\cdot(H'~\code y) + 10).
+$$
+This must be understood as a motion of potential from $\code x$
+to $\code y$. If some resource is consumed $\code y$ times afterwards, we
+can now use $\code y$'s potential to pay for it. But interestingly, the
+initial potential function $\Phi$ only talks about~$\code x$: the potential
+function allowed us to impart the cost of the subsequent resource
+consumptions on the variable~$\code x$.
+This flexibility combined with intelligent decisions on potential
+motions allow our tool to discover linear invariants for surprisingly
+convoluted programs.
+
+\subsection{Motivating Examples}
+
+We now present a representative set of examples that can be
+handled by our tool.  The bounds that are given are for the number
+of ``base statements'' executed by the program, that is, the number
+of assertions, skip and assignment statements executed.  The
+bounds are given as literal expressions where a variable name {\tt x}
+represents the initial value of this variable.
+
+\paragraph{Simple loops.} The simplest loop scheme that is in
+use in typical imperative programs is counting upwards or
+downwards to a given bound N.  The two example programs
+are given below, on their right is the bound inferred by our analysis.
+\begin{center}
+\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
+\lstinline|while x > N ( x = x - 1 )| & $|{\tt N} -  {\tt x}|$ \\
+\lstinline|while x < N ( x = x + 1 )| & $|{\tt N} -  {\tt x}|$ \\
+\end{tabular}
+\end{center}
+
+It is also frequent to iterate every $n$ elements in an array, also
+in this simple case our analysis works without any problem.  Note
+however that if we increment using a variable, it might very well be
+possible that the program diverges!  In this situation no worst
+case bound can be given and the analysis correctly fails.
+\begin{center}
+\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
+\lstinline|while x <= -2 ( x = x + 2 )| & $| {\tt x} | / 2$ \\
+\lstinline|while x <= z - y ( x = x + y )| & Infeasible \\
+\lstinline|assert y > 0; while x <= z - y ( x = x + y )| & $1 + |{\tt x} - {\tt z}|$
+\end{tabular}
+\end{center}
+In the last example, the constant 1 is added to the loop bound
+to account for the initial \texttt{assert} that consumes one
+step.  Also, since the bound is for the worst case, it is tight
+only when {\tt y} is 1.
+
+\paragraph{In the literature.}
+The two examples below are taken from Gulwani's paper~\cite{Gulwani-speed}.
+They are both handled by the Speed tool but require inference of a so-called
+disjunctive invariant.  In the abstract interpretation community, these
+invariants are known to be notoriously difficult to handle.
+
+\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}
+while x < n
+  if y < m    y = y + 1
+  else        x = x + 1
+\end{lstlisting}
+$$
+|{\tt m} - {\tt y}| + |{\tt n} - {\tt x}|
+$$
+\end{minipage}
+\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}
+while x < n
+  if z > x    x = x + 1
+  else        z = z + 1
+\end{lstlisting}
+$$
+|{\tt n} - {\tt x}| + |{\tt n} - {\tt z}|
+$$
+\end{minipage}
+
+\hspace{1ex}
+
+While the first
+example is simple to understand, the second one is much more
+tricky and trying to understand how it actually works is quite challenging.
+However, with the amortized analysis in mind, using the potential transfer
+reasonning, it is almost trivial to prove the termination.  While the Speed
+tool has to find a fairly involved invariant for the loop, our tool is simply
+reasoning locally and works without any special trick.
+%
+Our analysis handles without problems all the other examples
+in~\cite{Gulwani-speed} that have a linear bound.  Here is
+a quick review of some of them together with the bounds we
+inferred.
+
+\begin{minipage}[b]{0.30\textwidth}
+\begin{lstlisting}
+while x < n
+  while x < n
+    x = x + 1
+\end{lstlisting}
+$$
+|{\tt x} - {\tt n}|
+$$
+\end{minipage}
+\begin{minipage}[b]{0.30\textwidth}
+\begin{lstlisting}
+while x < n (x = x + 1);
+while x < n (x = x + 1)
+\end{lstlisting}
+$$
+|{\tt x} - {\tt n}|
+$$
+\end{minipage}
+\begin{minipage}[b]{0.30\textwidth}
+\begin{lstlisting}
+while x < n
+  ( while y < m (y = y+1)
+  ; x = x + 1 )
+\end{lstlisting}
+$$
+|{\tt x} - {\tt n}|
++|{\tt y} - {\tt m}|
+$$
+\end{minipage}
+
+\paragraph{Amortized reasonning.}
+The two examples below show the power of amortized reasoning.
+Because they essentially rely on an amortized reasonning we believe
+that our tool is the only one able to deal with them.
+
+\begin{minipage}[b]{0.45\textwidth}
+\begin{lstlisting}
+assert y >= 1;
+while x + y <= N
+  ( x = x + y
+  ; z = y; z = z + 4
+  ; while z > 0 ( z = z - 1 ) )
+\end{lstlisting}
+$$
+1 + 8\,|{\tt N} - {\tt x}|
+$$
+\end{minipage}
+\begin{minipage}[b]{0.45\textwidth}
+\begin{lstlisting}
+while x > 0
+  ( x = x - 1
+  ; if ?
+      while s > 0 ( s = s - 1 )
+    else s = s + 1 )
+\end{lstlisting}
+$$
+3\,|{\tt x}| + |{\tt s}|
+$$
+\end{minipage}
+
+\hspace{1ex}
+
+Maybe surprisingly, the two examples have linear bounds.  In the first one, the
+worst case is when {\tt y} is 1, in this situation the inner loop will perform 5 times
+$|{\tt N} - {\tt x}|$ iterations. Indeed, for each decrement of {\tt x}, {\tt z} is set
+to 5 and iterated over.  The second example is classic in the amortized analysis
+community, it is the original example from Tarjan's paper~\cite{Tarjan-amort} that
+non deterministically pushes one element on a stack or pops many elements from
+that stack.  In the example source code, the stack depth is represented by the
+variable {\tt s} and the number of iterations is ruled by the initial value of {\tt x}.
+
+\section{Syntax}
+
+Let us now start the formal part of this work.  We begin by defining
+a proper syntax for the language we will be using later on.  This is
+a more algebraic version of the surface syntax used in the previous
+examples.
 %
@@
 S := \code{skip}
-\mid \code{assert}(\Gamma)
+\mid \code{assert}(C)
 \mid x \gets E
-\mid \code{while}(0 < x)~S
-\mid \code{if}(?)~S~\code{else}~S
+\mid \code{while}(C)~S
+\mid \code{if}(C)~S~\code{else}~S
 \mid S;S
@@
 
-E := x + y \mid x - y % \mid y
-\end{mathpar}
-%
-We define a deductive system to derive judgements of the
-form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.  The semantics
-of this judgement are defined below in the soundness section.
-
-\begin{mathpar}
-
-\Rule{Skip}
-{}
-{ \Gamma; Q + M_s \vdash \code{skip} \dashv Q; \Gamma }
-
-\and \Rule{Assert}
-{}
-{ \Gamma; Q + M_a \vdash \code{assert}(\Delta) \dashv Q; \Gamma, \Delta }
-
-% \\ \Rule{AssignP}
-% { \Gamma \models \Gamma'[x/x+y] }
-% {\Gamma; Q + M_s + y\cdot q_x \vdash x \gets x + y \dashv Q; \Gamma'}
-
-% \and\Rule{AssignM}
-% {  \Gamma \models \Gamma'[x/x-y]
-% \\ \Gamma \models c \le y \le x
-% }
-% { \Gamma; Q + M_s \vdash x \gets x - y \dashv c \cdot q_x + V; \Gamma' }
-
-\\ \Rule{While}
-{ \Gamma, 0 < x ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
-{ \Gamma; Q \vdash \code{while}(0 < x)~S \dashv Q - M_w^3; \Gamma, x \le 0 }
-
-\and\Rule{Seq}
-{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
-}
-{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
-
+E := x \mid x \pm y
 \end{mathpar}
 
-\section{Decreasing differences}
-The second iteration is a system more symmetric where the potentials
-are linear functions in differences of variables.  From now on $k$
-denotes a non-negative constant.
+Note that expressions are kept extremely simple, this will help
+greatly when designing the analysis.  Moreover, programs
+containing fancier expressions can easily be compiled to this
+restricted form.  For example $a \gets b + c - d$ is semantically
+equivalent to $a \gets b; a \gets a + c; a \gets a - d$ which
+fits in the above syntax.
+
+The syntax of conditions, noted $C$, is left unspecified
+for the moment.  We will describe it later when the practical
+implementation details are provided.  The only semantic tool
+we need to deal with this unknown is an evaluation
+function $\llbracket \_ \rrbracket_{\_}$ such that, for any
+condition $C$ and any heap $H$ we have $\llbracket
+C \rrbracket_H \in \{ \top, \bot \}$.
+
+Another peculiarity is that there are no integer constants in the
+syntax.  Once again this is for the sake of simplicity.  In practice
+the concrete syntax recognizes ``variables'' like {\tt -4} or
+{\tt 60} as read only variables.  Any heap used later in the
+operational semantics will map these variables to their natural
+value.
+
+\section{Cost Semantics}
+
+The semantics for the syntax defined above needs to account
+for resource consumption.  To account for all kinds of resources
+we leave it as a parameter of the semantics to be specified
+by the user.  This parameter, called a metric, is a tuple of rational
+numbers with the following form
 $$
-S := \code{while}(k < x - y)~S \mid \dots
+(
+M_k, M_s, M_a,
+M_w^1, M_w^2, M_w^3,
+M_c^1, M_c^2,
+M_s^1, M_s^2
+).
 $$
-%
-In the rules below we take as convention that $q_{x}$ for a variable
-$x$ is in fact $q_{x0}$ where 0 is a special variable fixed to zero
-in all contraint contexts.
+Each of these rational numbers indicates the amount of resource
+consumed by an operation.  If one of these numbers is negative,
+it means that some resource is released by the program.
 
-\begin{mathpar}
+The general form of an evaluation judgement is
+$(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
+presented in the overview section appart  from the fact that
+two resource numbers are tracked instead of one.  Intuitively
+$(q, q')$ means that $q$ resources units are necessary to execute
+the program and that $q'$ units are available after its execution.
+For example, $\code{alloc}(3)$ consumes 3 heap cells, so
+the pair resulting from its evaluation would be $(3, 0)$; for
+$\code{free}(3)$, the pair would be $(0, 3)$.  When these
+statements are sequenced with \code{alloc} first, then \code{free},
+the evaluation yields $(3, 3)$.  These resource pairs can
+be composed with the following operation
+$$
+(q_1, q'_1) \cdot (q_2, q'_2) =
+\left\{
+\begin{array}{ll}
+(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
+(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
+\end{array}
+\right.
+.
+$$
+This operation can be seen as the sequential composition of
+two resource consumptions, note that it is not a commutative
+operation, the same way resource freeing and allocation do not
+commute.
+With $(0, 0)$ as a neutral element, the above composition defines
+a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
+It is possible to inject $q \in \mathbb Q$ in this monoid with by taking
+$(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.  This injection
+is used implicitely in the semantics.
 
-\Rule{AssignOpK}
-{  \Gamma \models \Gamma'[x/x \pm y]
-\\ \Gamma \models |y| = k
-\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
-\\ \textstyle q'_y = q_y
-      + \sum_{u \in \mathcal U} p_{xu}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-\\ q_{xu} = p_{xu} + r_{xu}
-}
-{ \textstyle
-  \Gamma; Q + M_s
-  \vdash x \gets x \pm y \dashv
-  \sum_u k \cdot r_{xu} + Q'; \Gamma'
-  % q_x is not omitted, v_{x} is in fact a shorthand for
-  % v_{x0} where 0 is a variable equal to 0.
-}
-
-\and \Rule{AssignOp}
-{  \Gamma \models \Gamma'[x/x \pm y]
-\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
-\\ \textstyle q'_y = q_y
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  \Gamma; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma'
-}
-
-\\ \mbox{(The other rules are unchanged)}
-
-\end{mathpar}
-
-\section{Cost aware semantics}
-
-This big step semantics is parameterized by a metric that sets
-the cost of each program operation.  Non-deterministically,
-the rule {\sc S:Abort} can be called to interrupt a computation.
-It has the same effect as raising an exception.  We symbolize a
-frozen computation with the notation $\circ$.  In the rules, $X$
-stands for either $\circ$ or $H$.
-
+\begin{figure}[ht!]
 \begin{mathpar}
@@
 { }
-{ (H, S) \Downarrow_0 \circ }
+{ (H, S) \Downarrow_{(0,0)} \circ }
 
@@
 { }
-{ (H, \code{skip}) \Downarrow_{M_s} H }
+{ (H, \code{skip}) \Downarrow_{M_k} H }
 
@@
 \and \Rule{S:Assert}
-{ H \models \Gamma }
-{ (H, \code{assert}(\Gamma)) \Downarrow_{M_a} H }
+{ H \models C }
+{ (H, \code{assert}(C)) \Downarrow_{M_a} H }
 
 \and \Rule{S:WhileStop}
-{ \neg (k < \llbracket x - y \rrbracket_H) }
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^3} H }
+{ \llbracket C \rrbracket_H = \bot }
+{ (H, \code{while}(C)~S) \Downarrow_{M_w^3} H }
 
-\and \Rule{S:WhileStep}
-{  k < \llbracket x - y \rrbracket_H
-\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
-\\ (H', \code{while}(k < x - y)~S) \Downarrow_{(q_2, q'_2)} X
-}
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }
-
-\and \Rule{S:WhileStepAbort}
+\and \Rule{S:WhileAbort}
 {  k < \llbracket x - y \rrbracket_H
@@
 
-\and \Rule{S:If}
-{  (H, S_i) \Downarrow_{(q, q')} X
-\\ i \in \{1, 2\}
+\and \Rule{S:WhileStep}
+{  \llbracket C \rrbracket_H = \top
+\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
+\\ (H', \code{while}(C)~S) \Downarrow_{(q_2, q'_2)} X
 }
-{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_{M_c^i \cdot (q, q')} X }
+{ (H, \code{while}(C)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }
+
+\and \Rule{S:IfThen}
+{  (H, S_1) \Downarrow_{(q, q')} X
+\\ \llbracket C \rrbracket_H = \top
+}
+{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^1 \cdot (q, q')} X }
+
+\and \Rule{S:IfElse}
+{  (H, S_2) \Downarrow_{(q, q')} X
+\\ \llbracket C \rrbracket_H = \bot
+}
+{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^2 \cdot (q, q')} X }
 
@@
 \end{mathpar}
+\caption{Operational semantics of statements}
+\label{fig:opsem}
+\end{figure}
 
-\section{Soundness theorem}
+Heaps manipulated by the operational semantics judgements are
+simple maps from variable names to integers.  We designate by
+$H[x \mapsto v]$ the heap $H$ updated to map the variable name
+$x$ to the number $v$.
+%
+The evaluation of expressions is trivial using heaps.
+\begin{mathpar}
+\llbracket x \rrbracket_H = H(x)
+\and
+\llbracket x \pm y \rrbracket_H = H(x) \pm H(y)
+\end{mathpar}
+
+In the evaluation judgement $(H, S) \Downarrow_{(q, q')} X$, the result $X$ can
+be either a new heap, which means the execution terminated successfully
+or it can be the \emph{busy} result $\circ$.  This indicates an early
+termination, it can be triggered at any point using the rule {\sc S:Abort}
+and bubbles up through the execution context.
+Sometimes there is no other choice than to use this latter rule, for
+example when an assertion fails.  In that case the side condition of
+{\sc S:Assert} is not met and the only rule that can possibly be applied
+is {\sc S:Abort}.
+
+The busy result is also useful to handle diverging programs with big-step
+judgements.  In case a program does not terminate, there is no heap $H'$
+such that $(H, S)\Downarrow_{(q, q')} H'$.  The only possible way to
+conclude a big-step judgement is to use the {\sc S:Abort} rule at some
+point in the computation.  One alternative to this technique is to use
+coinductive semantics~\cite{Leroy-coind}, however, using the busy
+result is precise enough in our setting of quantitative resource analysis and
+allows to have only one soundness proof for both unsafe and
+non-terminating programs.  Indeed, stating a result for
+$(H, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
+programs, and partial computations at the same time.
+
+\section{Proof System}
+
+We define in this section an abstract version of the proof system that
+we will use to derive bounds in our practical implementation.
+
+\subsection{Linear Potential Functions}
+
+To get a practical and automated implementation of amortized analysis
+as presented in the overview section we need the fix the shape of
+potential functions.  The general form of a potential function is as follows
+$$
+\Phi(H) = q_0 + \sum_{a \prec b} q_{ab}\,|H(a) - H(b)|.
+$$
+Where $a$ and $b$ range over the set of all program variables $V$ and
+$(q_{ab})$ and $q_0$ are non-negative rational numbers.  The relation
+$\prec$ is an arbitrary complete order on $V$ that is used only to
+avoid considering the two equal quantities $|H(a) - H(b)|$ and
+$|H(b) - H(a)|$ in the above sum.
+
+To simplify the references to the $q$ coefficients later we introduce
+an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
+\{(a, b) \mid a \prec b\}$.  Each index $i$ corresponds to a \emph{base
+function} $f_i$ in the potential function: $0$ corresponds to the constant
+function $\_ \mapsto 1$, and $(a,b)$ corresponds to $H \mapsto
+|H(a) - H(b)|$.  Using these notations we can rewrite the above
+equality as
+$$
+\Phi(H) = \sum_{i \in I} q_i f_i(H).
+$$
+From now on, we will write $ab$ to denote the index $(a,b)$.
+
+\subsection{Proof Rules}
+
+\begin{figure}[ht!]
+\begin{mathpar}
+
+\Rule{Q:Skip}
+{}
+{ \Gamma; Q + M_k \vdash \code{skip} \dashv Q; \Gamma }
+
+\and \Rule{Q:Assert}
+{}
+{ \Gamma; Q + M_a \vdash \code{assert}(C) \dashv Q; \Gamma, C }
+
+\and \Rule{Q:While}
+{ \Gamma, C ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
+{ \Gamma; Q \vdash \code{while}(C)~S \dashv Q - M_w^3; \Gamma, \neg C }
+
+\and \Rule{Q:Set}
+{ \forall u,~q_{yu} = q'_{xu} + q'_{yu}
+\\ q'_{xy} \in \mathbb Q_0^+
+}
+{ \Gamma[x/y]; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
+
+\and \Rule{Q:Incr}
+{ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
+\\ \textstyle q'_y = q_y
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  \Gamma[x/x \pm y]; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
+}
+
+\and \Rule{Q:If}
+{ \Gamma, C; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
+\\ \Gamma, \neg C; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
+}
+{ \Gamma; Q \vdash \code{if} (C)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
+
+\and\Rule{Q:Seq}
+{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
+\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
+}
+{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
+
+\and \Rule{Q:Weak}
+{ \Gamma_1 \models \Gamma_2
+\\ Q_1 \succeq_\Gamma Q_2
+\\ \Gamma_2; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
+\\ Q'_2 \succeq_\Gamma Q'_1
+\\ \Gamma'_2 \models \Gamma'_1
+}
+{ \Gamma_1; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
+
+\and \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |x - y| \}
+\\ \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |x - y| \le u_{xy} \}
+%\\ q'_0 = q_0 + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i
+\\ \forall i \in \mathcal U,~ q'_i \ge q_i - p_i
+\\ \forall i \in \mathcal L,~ q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U \cup \mathcal L,~ q'_i \ge q_i
+}
+{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
+
+\end{mathpar}
+\caption{Abstract proof rules of the quantitative analysis}
+\label{fig:proof}
+\end{figure}
+
+The deductive system defined in~\pref{fig:proof} derives
+judgements of the form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
+Its semantics are formally described in the soundness section but we
+will give an intuition of its meaning here.
+
+\paragraph{Quantitative part.}
+The judgement is really two parts, one logical part, and one quantitative
+part.  The quantitative part is, as shown in the overview section, a
+representation of two potential functions $\Phi$ and $\Phi'$ using
+two maps from indices to non-negative rational numbers
+$Q, Q' : I \rightarrow \mathbb Q_0^+$, these maps are called
+\emph{quantitative contexts}.
+We note $q_i$ for $Q(i)$.  Since $Q$ specifies a value for each
+index in $I$ it is clear that it defines a unique linear potential function
+as described in the previous section. We define $H \mapsto \Phi_H(Q)$
+to be the linear potential function canonically associated to $Q$.
+
+The rules of~\pref{fig:proof} are designed to enforce a semantic
+soundness similar to the one exposed in the overview section.  To
+do this, the rules impose contraints on $Q$ and $Q'$.  To make the
+presentation lighter, we use the following  set of conventions.
+First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
+and $\forall i \neq 0,~q'_i = q_i$.  Second, for any index $i$ such that
+the relation between $q_i$ and $q'_i$ is not specified as a side condition
+of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
+all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
+non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
+0 is a program variable symbolizing the constant 0.
+
+\paragraph{Logical part.}
+The second part of judgements maintains an abstract program state.
+This should be seen as a very weak Hoare logic.  It has to be weak because
+we want to automate it trivially and it can be weak because we only use
+it to get very local and simple knowledge.  The specifics of logical assertions
+in $\Gamma$ are left to the implementation section.
+%
+The purpose of this logical state is to help the rules
+make some sound decisions when relating coefficients of quantitative
+contexts.
+
+Syntactically $\Gamma$ is a list of program conditions.  Because of the
+rules {\sc Q:While} and {\sc Q:If}, we require the program conditions
+to be closed by negation.  Semantically we define the validity
+$H \models \Gamma$ of $\Gamma = C_1, \dots, C_n$ on a given
+heap $H$ to be $\bigwedge_i \llbracket C_i \rrbracket_H = \top$.
+%
+The logical state
+is maintained using regular Hoare style rules.
+The semantic entailment on logical states is noted $\Gamma \models \Delta$,
+its meaning is that $\forall H,~ H\models\Gamma \implies H\models\Delta$.
+
+%\paragraph{Explanation of selected rules.}
+
+\section{Soundness Theorems}
+
+The soundness of a judgement relates the cost semantics with the potential
+functions derived using our proof system.  Depending on the outcome of
+the semantics, we give different soundness statements.
+But, before tackling the issue of soundness let us remind and define some
+convenient notations.
+\begin{itemize}
+\item $H \mapsto \Phi_H(Q)$ for $Q$ a quantitative context, is the
+unique linear potential function that has $Q(i)$ as a coefficient for the
+base function $f_i$.
+\item $H \models \Gamma$ is true if and only if all the assertions in
+$\Gamma$ are true on the heap $H$.
+\end{itemize}
+
+Ideally we would
+like our judgements to be meaningful even in the case of a program
+crash.  More precisely, we want the first potential function to be enough
+to pay for the execution of the program up to the crash point.  This is
+formally stated in the following theorem.
+
+\begin{theorem}[Busy case]
+For any $S$, $H$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$
+such that $H \models \Gamma$, $\Gamma; Q \vdash S \dashv Q'; \Gamma'$,
+  and $(H, S) \Downarrow_{(q,q')} \circ$,
+we have $q \le \Phi_H(Q)$.
+\end{theorem}
+\begin{proof}
+Simple application of the terminating case proved afterwards.
+\end{proof}
+
+In order to prove the harder case for terminating programs, need to first
+focus on the only structural rule of the proof system: {\sc Q:Weak}.  This
+rule can be applied anywhere in the derivation, it is not syntax directed.
+To prove the soundness of this rule, we need a lemma that gives a
+semantic explanation of the $\succeq_\Gamma$ relation.
+
+\begin{lemma}[Relax]
+If $Q' \succeq_\Gamma Q$, for any heap $H$ such that $H \models \Gamma$,
+we have $\Phi_H(Q') \ge \Phi_H(Q)$.
+\end{lemma}
+\begin{proof}
+We observe the following inequations.
+\begin{align*}
+\Phi_H(Q') &= \sum_i q'_i f_i(H) \\
+&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
+  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(H) + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
+  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
+ &\ge \Phi_H(Q)
+  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
+  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
+ &\ge \Phi_H(Q).
+\end{align*}
+Since $H \models \Gamma$, using the transitivity of $\models$ and
+the definition of $\mathcal U$ and $\mathcal L$ we get
+$\forall i\in\mathcal L,~ l_i \le f_i(H)$ and
+$\forall i\in\mathcal U,~ f_i(H)\le u_i$.
+These two inequalities justify the last step.
+\end{proof}
+
+Using this lemma, we can tackle the hardest theorem of soundness.  This theorem
+needs a strong statement to make the proof by induction go through and defines
+the complete meaning of a judgement $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
+In a similar fashion than the busy case, the initial quantitative context is enough
+to perform the whole computation: $q \le \Phi_H(Q)$.  But additionally, the
+two quantitative contexts are also related to the resource consumption $q - q'$
+of the piece of program executed.  This second requirement exactly matches
+the semantic condition described in the overview section: it says that the
+initial potential $\Phi_H(Q)$ is enough to both pay for the resource
+consumption of the program $q - q'$ and pay for the potential of the resulting
+heap $\Phi_{H'}(Q')$.
+
+The proof of this result if very technical but basically goes through by induction
+on the big-step evaluation.  Only a representative subset of all the cases
+is present in the following proof.
 
@@
 
-\begin{proof} By induction on the big step semantics.
+\begin{proof} By nested induction on the big step semantics and on the
+judgement derivation (to account for the structural rule {\sc Q:Weak}).
 
@@
 \item Case {\sc S:Skip}.
-  We have $(H, \code{skip}) \Downarrow_{M_s} H$.
+  We have $(H, \code{skip}) \Downarrow_{M_k} H$.
   By assumption, $H \models \Gamma$. \\
-  Since $\Phi_H(Q + M_s) = \Phi_H(Q) + M_s$,
+  Since $\Phi_H(Q + M_k) = \Phi_H(Q) + M_k$,
   \begin{itemize}[topsep=0pt]
   \item
-    if $M_s \ge 0$ we have,
-      $M_s \le \Phi_H(Q + M_s)$ and
-      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge M_s - 0$;
+    if $M_k \ge 0$ we have,
+      $M_k \le \Phi_H(Q + M_k)$ and
+      $\Phi_H(Q + M_k) - \Phi_H(Q) \ge M_k - 0$;
   \item
-    if $M_s < 0$ we have,
-      $0 \le \Phi_H(Q + M_s)$ and
-      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge 0 - (-M_s)$.
+    if $M_k < 0$ we have,
+      $0 \le \Phi_H(Q + M_k)$ and
+      $\Phi_H(Q + M_k) - \Phi_H(Q) \ge 0 - (-M_k)$.
   \end{itemize}
@@
 \end{proof}
+
+\section{Implementation}
+
+\subsection{Logical Annotations Generation}
+
+\paragraph{Program conditions.}
+The first pass in our OCaml implementation generates the
+logical annotations $\Gamma$ around all program statements.
+These annotations are used later on by the quantitative
+analysis to generate contraints.
+%
+The first design choice we had to make is to give a fixed form
+to program conditions.  Because of their generality and because
+they are closed under negation, we decided to consider
+inequalities over linear combinations as program conditions.
+\begin{mathpar}
+C := L < L \mid L > L \mid L \le L \mid L \ge L
+\and
+L := x \mid k \mid k * L \mid L + L \mid L - L
+\end{mathpar}
+A Condition is represented internally as a pair of one constant
+$k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
+Its intended semantics is
+$$
+\llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
+$$
+Note that conditions of this shape are closed by negation, as showed
+by the following equivalence
+$$
+\neg (k + \sum_v m_v \cdot H(v) \le 0)
+\Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
+$$
+We can also note that all the ``questions'' asked to the logic
+assertions during a proof derivation can be described using
+linear combinations.  Conveniently enough, this language of
+assertions is exactly Presburger's arithmetic.  This means that
+we have a sound and complete decision procedure to answer
+all the questions asked during the analysis.
+
+\paragraph{Logical contexts.}
+The second problem related to the logic is how to figure out
+the logical contexts $\Gamma$ for each program statement.
+This is a question that is answered by abstract interpretation
+but, to show that our work really does not need the full power
+of abstract interpretation, we designed and implemented a very
+simple top-down approach: starting from a statement $S$ and
+a pre-condition $\Gamma$ (given as a list of inequalities), we
+generate $\mathcal P(S, \Gamma)$ a postcondition for $S$ using the
+following equations:
+
+\begin{align*}
+\mathcal P(\code{skip}, \Gamma) &:= \Gamma \\
+\mathcal P(x \gets y, \Gamma) &:= \{ e \le 0 \mid (e \le 0) \in \Gamma \land \{x\} \# e \} \\
+\mathcal P(x \gets x + y, \Gamma) &:= \Gamma[x / x - y] \\
+\mathcal P(x \gets x - y, \Gamma) &:= \Gamma[x / x + y] \\
+\mathcal P(\code{assert}(C), \Gamma) &:= \Gamma,C \\
+\mathcal P(S_1; S_2, \Gamma) &:= \mathcal P(S_2, \mathcal P(S_1, \Gamma)) \\
+\mathcal P(\code{if} (C)~S_1~\code{else}~S_2) &:=
+  \mathcal P(S_1, \Gamma,C) \lor \mathcal P(S_2, \Gamma,\neg C) \\
+\mathcal P(\code{while}(C)~S, \Gamma) &:=
+  \Gamma \lor (\neg C, \mathcal P(S, (C, \{ e \le 0 \mid (e \le 0) \in \Gamma \land {\rm set}(S) \# e \})))
+\end{align*}
+We used the notation ${\rm set}(S)$ to designate the set of variables
+assigned to in the statement $S$ and $s\#e$ to mean that variables
+in $s$ do not appear in the linear combination $e$.  Because contexts
+are simply conjunction of conditions we must define $\Gamma \lor \Delta$
+as a function on contexts:
+$$
+\Gamma \lor \Delta := \{ e \le 0 \mid (e \le 0) \in \Gamma \cup \Delta \land \Gamma \models (e \le 0) \land \Delta \models (e \le 0) \}.
+$$
+It is important to note that the $\mathcal P$ function defined above
+is not performing any fixpoint computation.  Despite this flagrant
+lack of sophistication our complete analysis is still able to figure out some
+tricky invariants as exemplified earlier.
+Finally, to generate annotations for the whole program $S$,
+we run $\mathcal P(S,\bullet)$ (where $\bullet$ is the empty context)
+and memorize all the intermediate results.  In practice, these results are stored in
+a table that can be referenced at any time during the analysis to get the
+pre and post-condition of any statement in the program.
+
+\paragraph{Decision procedure.}
+To decide logical entailment $\Gamma \models \Delta$ we use
+Presburger's decision procedure.  But what this procedure gives us
+is in fact only an answer the the question: Is there any heap $H$
+such that $H \models \Gamma$.  While this sentence is existential
+the question $\Gamma \models \Delta$ is universal.  To use the
+decision procedure, we use the well known trick of double negation.
+\begin{align*}
+\Gamma \models C
+&\Leftrightarrow \neg\neg (\Gamma \models C) \\
+&\Leftrightarrow \neg\neg (\forall H, H\models\Gamma \implies H\models C) \\
+&\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\
+&\Leftrightarrow \neg \exists H, H\models\Gamma \land H\models \neg C \\
+&\Leftrightarrow \neg \exists H, H\models (\Gamma, \neg C)
+\end{align*}
+Thus, to know if $\Gamma \models C$ we ask the decision
+procedure if $\Gamma, \neg C$ is satisfiable and negate the result.
+When we want to know about $\Gamma \models \Delta$ we simply ask
+one question for each of the conjuncts of $\Delta$.
+
+\subsection{Linear Contraints Generation}
+
+\paragraph{Generating a linear program.}
+Because figuring out the coefficents that are in $Q$ right away is a hard
+problem, we separate the search of a derivation in two steps.  As a first
+step we go through the whole program and apply inductively the rules
+given in the proof system section.  During this process our tool uses
+symbolic names for the coefficients $(q_i)$.  Every time a side condition
+must be satisfied by these coefficients, it is recorded using the symbolic
+names in a global list.  When
+the list of all constraints that must be satisfied by the coefficients is
+collected, we feed it to an off the shelf LP-solver.  If the solver successfully
+finds a solution to the given problem, we know that a derivation
+exists for the considered program.  We can even ask the solver
+for the initial $Q$ and get a resource bound for the program.
+If the LP-solver fails to find a solution, an error is reported to the
+user.
+
+\paragraph{Making the proof system algorithmic.}
+One problem with the proof system as described is that it is not
+fully syntax directed.  The rule {\sc Q:Weak} can be applied at
+all times.  However, if we apply this weakening rule at every step
+of the derivation, the number of generated constraints and
+the number of queries to the logic system explode.  This blowup
+causes efficiency problems.  To avoid these performance problems
+our implementation only applies the weakening rules at some
+strategic points in the derivation.  More specifically, at sequence
+points, at the beginning of while loops and around conditional
+statements.  So far, this selection of weakening spots has been
+driven by practice but it might be possible to prove that these
+program points are the only ones where weakening is needed.
 
@@
   %
-  If $Q' = \shift_{\{U_1,\ldots,U_n\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
+  If $Q' = \shift_{U_1,\ldots,U_n}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
 \end{proposition}
@@
 
+\bibliographystyle{abbrvnat}
+\bibliography{lit}
+\addcontentsline{toc}{section}{Bibliography}
 \end{document}

### 2014-05-21 paper/main.tex
some notes

--- 2014-05-14 main.tex
@@
 \maketitle
+
+\section{NOTES}
+
+\subsection{Things to say in the intro}
+
+\paragraph{Quote of Knuth}
+
+\begin{quotation}
+  There's even a fundamental gap in the foundations of my main
+  mathematical specialty, the analysis of algorithms. Consider, for
+  example, a computer program that sorts a list of numbers into
+  order. Thanks to the work of Floyd, Hoare, and others, we have
+  formal definitions of semantics, and tools by which we can verify
+  that sorting is indeed always achieved. My job is to go beyond
+  correctness, to an analysis of such things as the program's running
+  time: I write down a recurrence, say, which is supposed to represent
+  the average number of comparisons made by that program on random
+  input data. I'm 100\% sure that my recurrence correctly describes
+  the program's performance, and all of my colleagues agree with me
+  that the recurrence is "obviously" valid. Yet I have no formal tools
+  by which I can prove that my recurrence is right. I don't really
+  understand my reasoning processes at all! My student Lyle Ramshaw
+  began to create suitable foundations in his thesis (1979), but the
+  problem seems inherently difficult. Nevertheless, I don't lose any
+  sleep over this situation.
+\end{quotation}
+http://www.informit.com/articles/article.aspx?p=2213858
+
+\paragraph{Advanteges over previous work}
+
+\begin{itemize}
+\item have formal soundness proof with respect to cost semantics
+\item can deal with resources that \emph{become available} (e.g, freeing memory) in an execution
+\item derive global bounds for the program as a function of the inputs (as opposed to local loop bounds)
+\item can handle advance control flow such as break, return, mutual recursive functions
+\item do not rely on abstract-interpretation based invariant generation
+\item show that constant factors are very precise with respect to the cost semantics
+\item demonstrate that the analysis is applicable in the formal verification of system code
+\end{itemize}
 

### 2014-05-23 paper/main.tex
some more notes

@@
 
+\paragraph{to do}
+
+\begin{itemize}
+\item describe schedulability analysis for two schedulers
+\item include program logic from the stack paper for arbitrary resources
+\item what operational semantics to use? 
+\item verify automatic tool with program logic
+\item derive some crazy bounds (polynomial? logarithmic?) with the logic
+\item logic can also be used by automation: put linear constraints in the precondition
+\end{itemize}
+
 \paragraph{Advanteges over previous work}
@@
 \begin{itemize}
+\item integrates well with hand-derived bounds in a program logic
 \item have formal soundness proof with respect to cost semantics
+\item can attach certificate to the code (proof-carrying code)
 \item can deal with resources that \emph{become available} (e.g, freeing memory) in an execution

### 2014-06-04 paper/main.tex
new title
alternative title (just a comment)
new introduction

@@
 
-\documentclass[10pt]{article}
-\usepackage{fullpage}
+\documentclass[nocopyrightspace,preprint]{sigplanconf}
 
@@
 
-\title{Automatic Amortized Analysis for Imperative Code}
-\date{}
-\author{}
-
 \begin{document}
+
+\conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
+\copyrightyear{2014}
+\copyrightdata{XXX-X-XXXX-XXXX-X/XX/XX}
+
+%----------------------------------------------------
+
+\titlebanner{Draft -- Do not distribute}        % These are ignored unless
+\preprintfooter{Draft}   % 'preprint' option specified.
+
+% \title{Automatic Amortized Analysis for Imperative Code}
+\title{Compositional Quantitative Resource Analysis of C Programs}
+% Compositional Resource-Bound Certification for C Programs
+
+\authorinfo
+{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
+{Yale University}
+
 \maketitle
+
+\begin{abstract}
+
+\end{abstract}
+
+\category{D.2.4}{Software Engineering}
+{Software/Program Verification}
+\category{F.3.1}{Logics and Meanings of Programs}
+{Specifying and Verifying and Reasoning about Programs}
+
+\terms Verification, Reliability
+
+\keywords Formal Verification, Compiler Construction, Program Logics,
+Stack-Space Bounds, Quantitative Verification
+
+\section{Introduction}
+\label{sec:intro}
+
+In software engineering and software verification, we often would like
+to have static information about the quantitative behavior of
+programs.  For example, stack and heap-space bounds are important to
+ensure the reliability of safety-critical systems [XXX (see PLDI)].
+% Clock-cycle bounds are needed to guarantee the safety of real-time
+% systems [XXX wilhelm].  
+Static energy usage information is critical for autonomous systems and
+has applications in cloud computing [XXX?  (cite this guy form the
+us))].  Worst-case resource bounds can be used to prevent side-channel
+attacks through resource padding [XXX?].  Loop and
+recursion-depth bounds are used to ensure the accuracy of programs
+that are executed on unreliable hardware [XXX oopsla] and complexity
+bounds are needed to verify security protocols [XXX].
+% XXX Mention differential privacy here?
+%
+In general, quantitative performance information at design time can
+provide useful feedback for developers.
+
+Static analysis of quantitative properties of imperative programs is
+an active research area and recent years have seen many innovations.
+Notable tools that have been developed include Speed [XXX], KoAT
+[XXX], Pubs [XXX], RANK [XXX], and LOOPUS [XXX].
+%
+While the these tools can derive impressive results for
+realistic software, there still exist several shortcomings that seem
+to hamper the application of the existing techniques in practice.
+\begin{enumerate}
+\item Analysis tools are black boxes that either deliver a result or fail
+  without enabling \emph{user interaction} to manually or
+  semi-automatically derive bounds for challenging parts of the
+  program.
+\item The computed bounds are often \emph{non-compositional} local
+  bounds (for a single (nested) loop) that are difficult to combine to
+  global whole program bounds.
+\item Existing techniques often rely on complex external tools such as
+  abstract interpretation based invariant generation [XXX speed] or
+  translation of the program into a term-rewriting system [costa,
+  tacas] without providing \emph{verifiable certificates} for the
+  correctness of the derived bound.
+\end{enumerate}
+%
+While there has been progress there is still truth in the following
+quote of Don Knuth from a recent interview [xxx].
+
+\begin{quotation}
+  % Consider, for example, a computer program that sorts a list of
+  % numbers into order. 
+  Thanks to the work of Floyd, Hoare, and others,
+  we have formal definitions of semantics, and tools by which we can
+  verify that sorting is indeed always achieved. My job is to go
+  beyond correctness, to an analysis of such things as the program's
+  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
+  describes the program's performance, and all of my colleagues agree
+  with me that the recurrence is "obviously" valid. Yet I have no
+  formal tools by which I can prove that my recurrence is right. I
+  don't really understand my reasoning processes at all!
+\end{quotation}
+%
+%
+In this paper, we present a novel cost analysis system for C programs
+that works automatically, efficiently and precisely for many examples,
+can be used interactively if the automation fails, and produces
+verifiable certificates for the bounds.  Our contributions are as
+follows.
+%
+\begin{itemize}
+\item We define a operational cost semantics for C programs that
+  accounts a constant cost for each statement that is parametric in a
+  cost metric.  The semantics models cost of terminating and
+  non-terminating executions.  A specific \emph{tick} function can be
+  inserted in the source code to further customize the cost
+  model.
+\item We describe an automatic static analysis technique for deriving
+  bounds for the cost that is defined by the operational semantics.
+  It is based on amortized resource analysis for functional programs
+  [XXX] and derives bounds that are non-negative linear combinations
+  of sizes $|[x,y]|$ of intervals between program variables $x,y$.
+\item We show with a publically available prototype implementation and
+  an experimental evaluation that the our novel amortized resource
+  analysis works efficiently for realistic programs and that the
+  derived constant factors in the bounds are close to or identical
+  with the optimal ones.
+\item We define a quantitative Hoare logic for interactively deriving
+  cost bounds for C programs.  The logic is implemented and proved
+  sound in the Coq Proof Assistant with respect to CompCert C
+  light.  We show that our automatic analysis readily produces
+  derivations in the quantitative Hoare logic.
+% \item We prove the completeness of the program logic
+%   with respect to the operational cost semantics.
+\end{itemize}
+%
+Our approach complements existing works since it provides a semantic
+foundation for the computation of bounds while still providing support
+for automation that sometimes goes beyond the capabilities of existing
+techniques.  The quantitative Hoare logic and the automatic amortized
+analysis enable semi-automatic interactive resource bound development
+in which an automatically derived bound can be used in the
+quantitative logic and a hand derived bound can be used by the
+automatic analysis.  Both the automatic amortized analysis and the
+quantitative logic are naturally compositional and describe the
+resource behavior of code fragments without referring to the source
+code.  As a result, we naturally derive global bounds that are
+functions of the input parameters.  To the best of our knowledge, our
+system is the only one that can deal with resource like memory that
+may become available during the execution.
+
+The main innovation that makes this work practical is our new
+automatic amortized analysis for C code.  It targets programs that
+seem to be common in system software, that is, programs whose resource
+consumption can be described as a function of sizes $|[x,y]|$ of
+intervals of integer variables.  A typical bound would be $B(x,y,z) =
+10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ and we have for example
+$B(100,200,43) = 1043$.  
+%
+Despite the apparent simplicity of the new analysis system, it is able
+to reproduce multiple results from the literature which were obtained
+using methods based on abstract interpretation.  In contrast with
+abstract interpretation--based methods, our technique does not require
+any fixpoint computations on an abstract domain to obtain loop
+invariants.  The mechanism we designed is able to leverage local
+assertions such as $x < y$ that we collect along the branching points
+to obtain global invariants.  We achieve this by generating a linear
+constraint system that reflects resource cost and size changes of
+integer variables in the program.  A solution of the linear program is
+a resource usage bound for the C program.  We can naturally, handle
+advanced control flow such as break, return, mutually-recursive functions.
+
+Following the development steps for automatic amortized analysis for
+functional programs [XXX], we deliberately restrict our self to linear
+bounds in the automatic analysis (there a no restrictions for the
+manually-derived though).  More specifically, bounds have the form
+$\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer variables
+or constants.  The reason is mainly clarity of presentation, since
+an extension to multivariate resource polynomials [XXX] would make the
+inference rules considerably more complected and should better be
+described separately.  However, we developed the linear inference
+system so that the extension to polynomial bounds should work
+smoothly.
+
+We have evaluated the quantitative logic and the automatic analysis
+with system code and examples from the literature. XXX
+
+With the intention, of making the material easily accessible we first
+informally describe the new automatic amortized analysis
+(\pref{sec:inform}).  We then formalize the intuition with the
+operational cost semantics an the inference rules for the amortized
+analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
+quantitative Hoare logic and its soundness proof that is formalized in
+Coq.  In \pref{sec:inter} we prove the soundness of the automatic
+amortized analysis by showing that derivation using the inference
+rules can be seen as derivations in the quantitative logic.  We also
+show how to integrate hand-derived bounds with automatically-derived
+bounds.  \pref{sec:exper} contains the results of the experimental
+evaluation and \pref{sec:related} describes related research.
 

### 2014-06-05 paper/main.tex
cleanup in paper

@@
 
-\section{NOTES}
-
-\subsection{Things to say in the intro}
-
-\paragraph{Quote of Knuth}
-
-\begin{quotation}
-  There's even a fundamental gap in the foundations of my main
-  mathematical specialty, the analysis of algorithms. Consider, for
-  example, a computer program that sorts a list of numbers into
-  order. Thanks to the work of Floyd, Hoare, and others, we have
-  formal definitions of semantics, and tools by which we can verify
-  that sorting is indeed always achieved. My job is to go beyond
-  correctness, to an analysis of such things as the program's running
-  time: I write down a recurrence, say, which is supposed to represent
-  the average number of comparisons made by that program on random
-  input data. I'm 100\% sure that my recurrence correctly describes
-  the program's performance, and all of my colleagues agree with me
-  that the recurrence is "obviously" valid. Yet I have no formal tools
-  by which I can prove that my recurrence is right. I don't really
-  understand my reasoning processes at all! My student Lyle Ramshaw
-  began to create suitable foundations in his thesis (1979), but the
-  problem seems inherently difficult. Nevertheless, I don't lose any
-  sleep over this situation.
-\end{quotation}
-http://www.informit.com/articles/article.aspx?p=2213858
-
-\paragraph{to do}
-
-\begin{itemize}
-\item describe schedulability analysis for two schedulers
-\item include program logic from the stack paper for arbitrary resources
-\item what operational semantics to use? 
-\item verify automatic tool with program logic
-\item derive some crazy bounds (polynomial? logarithmic?) with the logic
-\item logic can also be used by automation: put linear constraints in the precondition
-\end{itemize}
-
-\paragraph{Advanteges over previous work}
-
-\begin{itemize}
-\item integrates well with hand-derived bounds in a program logic
-\item have formal soundness proof with respect to cost semantics
-\item can attach certificate to the code (proof-carrying code)
-\item can deal with resources that \emph{become available} (e.g, freeing memory) in an execution
-\item derive global bounds for the program as a function of the inputs (as opposed to local loop bounds)
-\item can handle advance control flow such as break, return, mutual recursive functions
-\item do not rely on abstract-interpretation based invariant generation
-\item show that constant factors are very precise with respect to the cost semantics
-\item demonstrate that the analysis is applicable in the formal verification of system code
-\end{itemize}
-
-\section{Introduction}
-
-After functional correctness, quantitative properties are the
-second thing one wants to know when writing a program.
-For example, an asymptotic runtime bound on a
-sorting algorithm tells us what efficiency we can expect
-of it when running on very large arrays.  But, what if we
-always run it on arrays of size between 5 and 10? In that
-case, the asymptotic is not relevant and we need a more
-precise bound on the runtime.  It is then a good
-software engineering practice to profile the algorithm on
-some examples.
-
-In between asymptotic bounds and profiling there are
-precise bounds, for instance, one can prove that Knuth's Quicksort might
-take at most $20N^2$ operations of a MIX machine when
-called on an array of size $N$.  If you have this information,
-no profiling is needed to compare the algorithms and you
-have a formal guarantee on the behavior of the program. But what if you do not
-have a MIX machine, or you tweaked Knuth's algorithm
-a bit?  Then you are good for another long an tedious
-proof.  We advocate that, it is possible to design a practical
-system that is able to infer precise worst case bounds
-without any user interaction.
-
-This work presents a resource oblivious system to get
-linear resource bounds on imperative programs with
-integers.  Despite its apparent simplicity it is able to
-reproduce multiple results from the literature which
-were obtained using methods based on abstract
-interpretation.  In stark contrast with abstract-interpretation
-based methods, our tool does not require any
-fixpoint computations on an abstract domain to obtain
-loop invariants.  The mechanism we designed, which
-is based on amortized analysis, is able to leverage a local
-information to obtain global invariants.  Additionally, the
-approach we advocate is able to scale to full blown
-imperative languages including constructs like functions
-calls, \code{break} and \code{return}.
-
-The analysis is proved sound with respect to
-the operational semantics instrumented to track resource
-consumption.  It is implemented as a prototype
-in OCaml and available on demand.  We test it on multiple
-algorithms from the literature and on custom tests, it
-is also planned to extend it to handle the full C programming
-language.
-
-\section{Overview and Examples}
+\section{Amortized Analysis for C Programs}
+\label{sec:inform}
 
@@
 The idea that drives the design of our system is amortized
-analysis.  Say that a program $p$ evaluates on a heap $H$
+analysis [XXX].  Say that a program $p$ evaluates on a heap $H$
 by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
@@
 
+\section{NOTES}
+
+\subsection{Things to say in the intro}
+
+\paragraph{Quote of Knuth}
+
+\begin{quotation}
+  There's even a fundamental gap in the foundations of my main
+  mathematical specialty, the analysis of algorithms. Consider, for
+  example, a computer program that sorts a list of numbers into
+  order. Thanks to the work of Floyd, Hoare, and others, we have
+  formal definitions of semantics, and tools by which we can verify
+  that sorting is indeed always achieved. My job is to go beyond
+  correctness, to an analysis of such things as the program's running
+  time: I write down a recurrence, say, which is supposed to represent
+  the average number of comparisons made by that program on random
+  input data. I'm 100\% sure that my recurrence correctly describes
+  the program's performance, and all of my colleagues agree with me
+  that the recurrence is "obviously" valid. Yet I have no formal tools
+  by which I can prove that my recurrence is right. I don't really
+  understand my reasoning processes at all! My student Lyle Ramshaw
+  began to create suitable foundations in his thesis (1979), but the
+  problem seems inherently difficult. Nevertheless, I don't lose any
+  sleep over this situation.
+\end{quotation}
+http://www.informit.com/articles/article.aspx?p=2213858
+
+\paragraph{to do}
+
+\begin{itemize}
+\item describe schedulability analysis for two schedulers
+\item include program logic from the stack paper for arbitrary resources
+\item what operational semantics to use? 
+\item verify automatic tool with program logic
+\item derive some crazy bounds (polynomial? logarithmic?) with the logic
+\item logic can also be used by automation: put linear constraints in the precondition
+\end{itemize}
+
+\paragraph{Advanteges over previous work}
+
+\begin{itemize}
+\item integrates well with hand-derived bounds in a program logic
+\item have formal soundness proof with respect to cost semantics
+\item can attach certificate to the code (proof-carrying code)
+\item can deal with resources that \emph{become available} (e.g, freeing memory) in an execution
+\item derive global bounds for the program as a function of the inputs (as opposed to local loop bounds)
+\item can handle advance control flow such as break, return, mutual recursive functions
+\item do not rely on abstract-interpretation based invariant generation
+\item show that constant factors are very precise with respect to the cost semantics
+\item demonstrate that the analysis is applicable in the formal verification of system code
+\end{itemize}
+
 \bibliographystyle{abbrvnat}

### 2014-06-07 paper/main.tex
catalog of examples in the paper

@@
 %
-While the these tools can derive impressive results for
-realistic software, there still exist several shortcomings that seem
-to hamper the application of the existing techniques in practice.
+While the these tools can derive impressive results for realistic
+software, there still exist shortcomings that hamper the application
+of the existing techniques in practice.
 \begin{enumerate}
@@
 %
-While there has been progress there is still truth in the following
-quote of Don Knuth from a recent interview [xxx].
-
-\begin{quotation}
+While there has been lot of progress in static quantitative analysis,
+Don Knuth correctly points out in a recent interview [XXX] that the
+state-of-the-art in formal quantitative methods still falls short
+in comparison with semantic techniques [xxx].
+%
+\begin{quote}
   % Consider, for example, a computer program that sorts a list of
@@
   don't really understand my reasoning processes at all!
-\end{quotation}
+  \vspace{-3ex}
+  \begin{flushright}
+    -- Donald Knuth, 2014
+  \end{flushright}
+\end{quote}
 %
@@
 \item We define a operational cost semantics for C programs that
-  accounts a constant cost for each statement that is parametric in a
+  accounts a constant cost for each statement which is parametric in a
   cost metric.  The semantics models cost of terminating and
@@
 code.  As a result, we naturally derive global bounds that are
-functions of the input parameters.  To the best of our knowledge, our
-system is the only one that can deal with resource like memory that
-may become available during the execution.
+functions of the input parameters of the program.  To the best of our
+knowledge, our system is the only one that can deal with resource like
+memory that may become available during the execution.
 
 The main innovation that makes this work practical is our new
-automatic amortized analysis for C code.  It targets programs that
-seem to be common in system software, that is, programs whose resource
+automatic amortized analysis for C code.  It targets programs that are
+common in system software, that is, programs whose resource
 consumption can be described as a function of sizes $|[x,y]|$ of
@@
 a resource usage bound for the C program.  We can naturally, handle
-advanced control flow such as break, return, mutually-recursive functions.
-
-Following the development steps for automatic amortized analysis for
+advanced control flow such as break, return, iteration over negative
+counter variables, and mutually-recursive functions.
+
+Following the development steps of automatic amortized analysis for
 functional programs [XXX], we deliberately restrict our self to linear
 bounds in the automatic analysis (there a no restrictions for the
-manually-derived though).  More specifically, bounds have the form
-$\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer variables
-or constants.  The reason is mainly clarity of presentation, since
-an extension to multivariate resource polynomials [XXX] would make the
-inference rules considerably more complected and should better be
-described separately.  However, we developed the linear inference
-system so that the extension to polynomial bounds should work
-smoothly.
+manually-derived bounds in the logic).  More specifically, bounds have
+the form $\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer
+variables or constants.  The reason is mainly clarity of presentation,
+since an extension to multivariate resource polynomials [XXX] would
+make the inference rules considerably more complected and should
+better be described separately.  However, we developed the linear
+inference system so that the extension to polynomial bounds shall
+work smoothly.
+
+We implemented our analysis for CompCert Clight, a subset of C in
+which loops can only be exited with a \code{break} statement and
+expressions are free of side effects.  C programs can for instance be
+compiled to Clight programs using the verified CompCert C compiler.
+
+Our new quantitative Hoare logic for CompCert Clight is a
+generalization of a program logic that we recently developed to
+formally very stack-space bounds in Coq [PLDI].  In a nutshell, we
+replace the boolean predicates (pre and post conditions) of Hoare
+logic with numeric predicates that map a program state to a
+non-negative number or infinity.  The intuitive meaning of a
+quantitative Hoare triple $\htriple{P}{C}{Q}$ that $C$ consumes $P(S)$
+resources when executed in starting state $S$, and that $Q(S')$
+resources are left if $C$ terminates in state $S'$.  Infinity
+corresponds to \emph{false} in classic Hoare logic.  Our quantitative
+logic is proved sound with respect to a cost semantic for Clight that
+accounts a constant cost for each evaluation step.  This constant cost
+is parametric in a resource metric that can be defined by the user.
 
@@
 \subsection{Amortized Analysis}
-The idea that drives the design of our system is amortized
-analysis [XXX].  Say that a program $p$ evaluates on a heap $H$
-by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
-The basic idea of amortized analysis is to define a potential function
-$\Phi$ from heaps to natural numbers, such that $\Phi(H) \ge n$.
+The idea that drives the design of our system is amortized analysis
+[XXX].  Say that a program $p$ evaluates on a program state $H$ by
+consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.  The
+basic idea of amortized analysis is to define a potential function
+$\Phi$ that maps program states to non-negative numbers, such that
+$\Phi(H) \ge n$.
 
@@
 
+% \acks
+
+% This research is based on work supported in part by NSF grants 1319671
+% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
+% Any opinions, findings, and conclusions contained in this document are
+% those of the authors and do not reflect the views of these agencies.
+
+\appendix
+
+\section{Catalog of Automatically Analyzed Programs}
+
+In this appendix we provide a non-exhaustive catalog of classes of
+programs that can be automatically analyzed by our system.  For
+simplicity, we use cost metric that counts cost $1$ for comparisons
+and assignments and cost $0$ for all other operations.  Of course, the
+examples can also be analyzed with any other cost metric.
+
+We assume that the variables in the code snippets are the inputs to
+the program. Some of the examples contain constants on which the
+computed bound depends.  These constants are randomly chosen to
+present an example but the analysis works for other constants as well.
+Note however that it is sometimes crucial that constants are positive
+(or negative) and that the same constant at different places.
+
+\begin{figure*}[t!]
+\newlength{\progwidth}
+\setlength{\progwidth}{.19\linewidth}
+  \centering
+
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x>0) {
+    x=x-1;
+    y=y+2;
+  }
+  while (y>0) {
+    y=y-1;
+  }
+  while (y>0) {
+    y=y+1;
+  }
+   \end{lstlisting}
+
+$4|[0,x]| + [|0,y|]$
+\\[.7\baselineskip]
+      {\bf t07}
+    \end{center}
+  \end{minipage}%
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (z-y>0) {
+    y = y+1;
+  }
+  while (y>9) {
+    y=y-10;
+  }
+   \end{lstlisting}
+
+$1.1|[y,z]| + 0.1[|0,y|]$
+\\[.7\baselineskip] 
+      {\bf t08}
+    \end{center}
+  \end{minipage}
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x-y>0) {
+    if (*)
+      y=y+1;
+    else
+      x=x-1;
+  }
+   \end{lstlisting}
+
+$|[y,x]|$
+\\[.7\baselineskip]
+      {\bf t10}
+    \end{center}
+  \end{minipage}
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x>0) {
+    x=x-1;
+    if (*) 
+      y=y+1;
+    else {
+      while (y>0)
+	y=y-1;
+    }
+  }
+   \end{lstlisting}
+
+$3|[0,x]| + |[0,y]|$
+\\[.7\baselineskip]
+      {\bf t13}
+    \end{center}
+  \end{minipage}
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n<0) {
+  n=n+1;
+  y=y+1000;
+  while (y>=100){
+    y=y-100;
+  }
+}
+   \end{lstlisting}
+
+$12|[n,0]| + 0.1|[0,y]|$
+\\[.7\baselineskip]
+      {\bf t27}
+    \end{center}
+  \end{minipage}
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x>y) {
+    x=x-1;
+    x=x+1000;
+    y=y+1000;
+  }
+  while (y>0) {
+    y=y-1;
+  }
+  while (x<0) {
+    x=x+1;
+  }
+   \end{lstlisting}
+
+$1004|[y,x]|+|[x,0]|+|[0,y]|$
+\\[.7\baselineskip]
+      {\bf t28}
+    \end{center}
+  \end{minipage}
+
+   \caption{Amortization and Compositionality}
+  \label{fig:cat1}
+\end{figure*}
+
+\paragraph{Amortization and Compositionality}
+
+\pref{fig:cat1} shows code snippets that need amortization are
+compositionality to obtain a whole program bound.
+
+Example \emph{t07} demonstrates two different features of the
+analysis.  For one thing it shows that we can precisely track size
+changes inside loops.  In the first loop, we increment $y$ by $2$ in
+each of the $|[0,x]|$ iterations.  For another thing it shows that we
+automatically recognize dead code if we find conflicting assertions on
+a branching path: After the second loop we know $y \leq 0$ and as a
+result can assign arbitrary potential inside the third loop where we
+know that $y>0$.  As a result, we obtain a tight bound.
+
+Example \emph{t08} shows the ability of the analysis to handle
+negative and non-negative numbers.  Note that there are no
+restrictions on the signs of $z$ and $y$.  We also see again that we
+accurately track the size change of $y$ in the first loop.
+Furthermore, \emph{t08} shows that we do not handle the constants $1$
+or $0$ in any special way.  In all examples you could replace $0$ and
+$1$ with other constants like we did in the second loop and still
+derive a tight bound.  The only information, that the analyzer needs
+is $y \geq c$ before assigning $y = y - c$.
+
+In example \emph{t10} we also do not restrict the inputs $x$ and $y$.
+They can be negative, positive, or zero.  The star {\tt *} in the
+conditional, stands for an arbitrary assertion.  In each branch of the
+conditional we can obtain a constant potential $1$ since the interval
+size $|[y,x]|$ is decreasing.
+
+Example \emph{t13} shows how amortization can be used to handle tricky
+nested loops.  The outer loop is iterated $[|0,x|]$ times.  In the
+conditional, we either (the branching condition is again arbitrary)
+increment the variable $y$ or we execute an inner loop in which $y$ is
+counted back to $0$.  Then analysis computes a tight linear bound for
+this program.  Again, the constants $0$ and $1$ in the inner loop can
+as well be replace by something more interesting, say $9$ and $10$
+like in example \emph{t08}.  Then we still obtain a tight linear
+bound.
+
+Example \emph{t27} is similar to example \emph{t13}.  Instead of
+decrements the variable $x$ in the outer loop we this time increment
+the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
+we increment the variable $y$ by $1000$.  We then execute an inner
+loop that increments $y$ by $100$ until $y=0$.  The analysis can
+derive that only the first execution of the inner loop depends on the
+initial value of $y$.  We again derive a tight bound.
+
+Example \emph{t28} is particularly interesting.  In the first loop we
+decrement the size $|[y,x]|$.  However, we also shift the interval
+$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
+that this does not change the size of the interval and computes the
+tight loop bound $3|[y,x]|$.  The additional to loops are in the
+program to show that the size tracking in the first loop works
+accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
+in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
+the worst case (if $x$ and $y$ are negative).
+
 \bibliographystyle{abbrvnat}
 \bibliography{lit}
-\addcontentsline{toc}{section}{Bibliography}
+
 \end{document}

### 2014-06-07 paper/main.tex
more work on the catalog

@@
     \end{center}
+  \end{minipage}%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  assert (y>=0);
+  while (x-y>0) {
+    x=x-1;
+    x=x-y;
+    z=y;
+    while (z>0) {
+      z=z-1;
+    }
+  }
+   \end{lstlisting}
+
+$1+3|[0,x]|$
+\\[.7\baselineskip]
+      {\bf t15}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  assert (y>=0);
+  while (x-y>0) {
+    x=x-1;
+    x=x-y;
+    z=y;
+    z=z+y;
+    z=z+100;
+    while (z>0) {
+      z=z-1;
+    }
+  }
+   \end{lstlisting}
+
+$1+2|[0,x]|+103|[y,x]|$
+\\[.7\baselineskip]
+      {\bf t16}
+    \end{center}
   \end{minipage}
@@
 
+Sometimes we need some assumptions on the inputs in order to derive a
+bound.  Example \emph{t15} is such a case.  We assume here that the
+input variable $y$ is non-negative and write \code{assert(y>=0)}.  The
+semantic of \code{assert} is that it has no effect if the assertion is
+true and that the program is terminated without further cost
+otherwise.  The cost of an assert is $1$ resource unit in our cost
+metric.  If we enter the loop then we know that $x>0$ and we can
+obtain constant potential from the assignment $x=x-1$.  After the
+assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
+can share the potential $3[0,x]$ before the assignment $x=x-y$ between
+$3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we derive
+a tight linear bound.
+
 \bibliographystyle{abbrvnat}

### 2014-06-08 paper/main.tex
finished the description of t16

@@
 
+\newlength{\progwidth}
+
 \begin{figure*}[t!]
-\newlength{\progwidth}
-\setlength{\progwidth}{.19\linewidth}
+\setlength{\progwidth}{.24\linewidth}
   \centering
 
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x>0) {
-    x=x-1;
-    y=y+2;
-  }
-  while (y>0) {
-    y=y-1;
-  }
-  while (y>0) {
-    y=y+1;
-  }
-   \end{lstlisting}
-
-$4|[0,x]| + [|0,y|]$
-\\[.7\baselineskip]
-      {\bf t07}
-    \end{center}
-  \end{minipage}%
-%
-%
-%
   \begin{minipage}[b]{\progwidth}
@@
   \end{minipage}
+   \caption{Amortization and Compositionality (a)}
+  \label{fig:cat1a}
+\end{figure*}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
+
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x>0) {
+    x=x-1;
+    y=y+2;
+  }
+  while (y>0) {
+    y=y-1;
+  }
+  while (y>0) {
+    y=y+1;
+  }
+   \end{lstlisting}
+
+$4|[0,x]| + [|0,y|]$
+\\[.7\baselineskip]
+      {\bf t07}
+    \end{center}
+  \end{minipage}%
 %
@@
 
-   \caption{Amortization and Compositionality}
-  \label{fig:cat1}
+   \caption{Amortization and Compositionality (b)}
+  \label{fig:cat1b}
 \end{figure*}
@@
 
+Example \emph{t16} is an extension of example \emph{t15}. We again
+assume that $y$ is non-negative and use the same mechanism to iterate
+the outer loop as in \emph{t15}.  In the inner loop, we also count the
+variable $z$ down to zero and perform $|[0,z]|$ interations.  However,
+insead of assigning $z=y$, we assign $z=2y+100$.  The analysis
+computes the tight linear bound $1+2|[0,x]|+103|[y,x]|$.  The
+assignment of potential to the size interval $|[y,x]|$ in instead of
+$|[0,x]|$ is a random choice of the LP solver.  Other optimal
+solutions of the linear program that the analyzer gernerates for
+\emph{t15} would result in bounds such as $1+105|[0,x]|$ or
+$1+6|[0,x]|+99|[y,x]|$.  In fact, the $105$ potential units can be
+shared arbitrarily between the interval sizes $|[y,x]|$ and $|[0,x]|$
+as long as more than $1$ unit is assigned to $|[0,x]|$.  Since we know
+that $y\geq 0$ the choice of the LP solver is optimal but this seems
+to be a coincidence.
+
 \bibliographystyle{abbrvnat}

### 2014-06-10 src/ctest/t07.c
translation of examples to C and more examples from literature

@@ -1,0 +1,19 @@
+
+int main(int argc, char **argv)
+{
+  int x,y;
+
+  while (x>0) {
+    x=x-1;
+    y=y+2;
+  }
+
+  while (y>0) {
+    y=y-1;
+  }
+
+  while (y>0) {
+    y=y+1;
+  }
+
+}

### 2014-06-10 src/ctest/t08.c

@@ -1,0 +1,13 @@
+
+int main(int argc, char **argv)
+{
+  int y,z;
+
+  while (z-y > 0) {
+    y = y + 1;
+  }
+  while (y > 9) {
+    y = y - 10;
+  }
+
+}

### 2014-06-10 src/ctest/t10.c

@@ -1,0 +1,12 @@
+
+int main(int argc, char **argv)
+{
+  int x,y;
+  while (x-y>0) {
+    if (1)
+      y=y+1;
+    else
+      x=x-1;
+  }
+
+}

### 2014-06-10 src/ctest/t11.c

@@ -1,0 +1,12 @@
+
+int main(int argc, char **argv)
+{
+  int x,y,n,m;
+
+  while (n>x) {
+    if (m>y) 
+      y = y+1;
+    else
+      x = x+1;
+  }
+}

### 2014-06-10 src/ctest/t13.c

@@ -1,0 +1,16 @@
+
+int main(int argc, char **argv)
+{
+  int x,y;
+
+  while (x > 0) {
+    x = x -1;
+    if (1) 
+      y = y +1;
+    else {
+      while (y > 0)
+	y = y -1;
+    }
+  }
+
+}

### 2014-06-10 src/ctest/t15.c

@@ -1,0 +1,15 @@
+
+int main(int argc, char **argv)
+{
+  int x,y,z;
+
+  //assert (y>=0);
+  while (x-y>0) {
+    x=x-1;
+    x=x-y;
+    z=y;
+    while (z>0) {
+      z=z-1;
+    }
+  }
+}

### 2014-06-10 src/ctest/t16.c

@@ -1,0 +1,17 @@
+
+int main(int argc, char **argv)
+{
+  int x,y,z;
+
+  //assert (y>=0);
+  while (x-y>0) {
+    x=x-1;
+    x=x-y;
+    z=y;
+    z=z+y;
+    z=z+100;
+    while (z>0) {
+      z=z-1;
+    }
+  }
+}

### 2014-06-10 src/ctest/t19.c

@@ -1,0 +1,13 @@
+
+int main(int argc, char **argv)
+{
+  int i,k;
+
+  while (i>100) {
+    i--;
+  }
+  i=i+k+50;
+  while (i>=0) {
+    i--;
+  }
+}

### 2014-06-10 src/ctest/t20.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int x,y;
+
+  while (x<y) {
+    x=x+1;
+  }
+
+  while (y<x) {
+    y=y+1;
+  }
+
+}

### 2014-06-10 src/ctest/t27.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int n,y;
+
+  while (n<0) {
+    n=n+1;
+    y=y+1000;
+    while (y>=100) {
+      y=y-100;
+    }
+  }
+
+}

### 2014-06-10 src/ctest/t28.c

@@ -1,0 +1,20 @@
+
+int main(int argc, char **argv)
+{
+  int x,y;
+
+  while (x>y) {
+    x=x-1;
+    x=x+1000;
+    y=y+1000;
+  }
+
+  while (y>0) {
+    y=y-1;
+  }
+
+  while (x<0) {
+    x=x+1;
+  }
+
+}

### 2014-06-10 src/ctest/t30.c

@@ -1,0 +1,12 @@
+
+int main(int argc, char **argv)
+{
+  int x,y,t;
+
+  while (x>0) {
+    x=x-1;
+    t=x;
+    x=y;
+    y=t;
+  }
+}

### 2014-06-10 paper/main.tex
more examples in paper

@@
 programs that can be automatically analyzed by our system.  For
-simplicity, we use cost metric that counts cost $1$ for comparisons
-and assignments and cost $0$ for all other operations.  Of course, the
-examples can also be analyzed with any other cost metric.
+simplicity, we use cost metric that counts cost $1$ for assignments
+and cost $0$ for all other operations.  Of course, the examples can
+also be analyzed with any other cost metric.
 
@@
   \label{fig:cat1b}
+\end{figure*}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
+
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (i>100) {
+    i--;
+  }
+  i=i+k+50;
+  while (i>=0) {
+    i--;
+  }
+   \end{lstlisting}
+
+$52 + |[-1,i]| + |[0,k]|$
+\\[.7\baselineskip]
+      {\bf t19}
+    \end{center}
+  \end{minipage}%
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x<y) {
+    x=x+1;
+  }
+  while (y<x) {
+    y=y+1;
+  }
+   \end{lstlisting}
+
+$|[x,y]|+|[y,x]|$
+\\[.7\baselineskip]
+      {\bf t20}
+    \end{center}
+  \end{minipage}%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x>0) {
+    x=x-1;
+    t=x;
+    x=y;
+    y=t;
+  }
+   \end{lstlisting}
+
+$4|[0,x]|+4|[0,y]|$
+\\[.7\baselineskip]
+      {\bf t30}
+    \end{center}
+  \end{minipage}
+
+   \caption{Amortization and Compositionality (c)}
+  \label{fig:cat1c}
 \end{figure*}
@@
 
+Example \emph{t19} demonstrates another nice property that demonstates
+the compositionaliy of the analysis.  The program consists of two
+loops that decrement a variable $i$.  In the first loop, $i$ is
+decremented down to 100 and in the second loop $i$ is increment
+further down to $-1$.  However, between the loops we assign
+$i=i+k+50$.  So in total the program performs $52 + |[-1,i]| +
+|[0,k]|$ increments.  Our analysis finds this tight bound because our
+amortized analysis naturally takes into account the relation between
+the two loops.  Other techniques might derive a more conservative
+bound such as $52 + |[-1,i]| + |[0,k]| + |[100,i]|$.
+
+Example \emph{t20} shows how we can handle programs in which bounds
+contain absolute values like $|x-y|$.  The first loop body is only
+executed if $x<y$ and the second loop body is only executed if $y<x$.
+The analyzers finds a tight bound.
+
+At first sight, example \emph{t30} appears to be a simple loop that
+decrements the variable $x$ down to zero.  However, a closer look
+reveals that the loop actually decrements both input variables $x$ and
+$y$ down to zero before terminating.  In the loop body, first $x$ is
+decremented by one.  Then the values of the variables $x$ and $y$ are
+switched using the local variable $t$ as a buffer.  Out analysis
+infers the tight bound $4|[0,x]|+4|[0,y]|$.
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (n>x) {
+    if (m>y) 
+      y = y+1;
+    else
+      x = x+1;
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.7\baselineskip]
+      {\bf t11}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+ XXX
+   \end{lstlisting}
+
+$XXX$
+\\[.7\baselineskip]
+      {\bf t12}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+ XXX
+   \end{lstlisting}
+
+$XXX$
+\\[.7\baselineskip]
+      {\bf t22}
+    \end{center}
+  \end{minipage}
+
+   \caption{Examples from \cite{GulwaniMC09}}
+  \label{fig:cat2a}
+\end{figure*}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
+
+   \caption{Examples from [GulwaniPLDI09] and [GulwaniPLDI10]}
+  \label{fig:cat2b}
+\end{figure*}
+
+\paragraph{From  the Literature}
+
+Example \emph{t11} has been given in~\cite{GulwaniMC09}.  In the loop,
+we increment
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+ XXX
+   \end{lstlisting}
+
+$XXX$
+\\[.7\baselineskip]
+      {\bf Knuth-Morris-Pratt}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+ XXX
+   \end{lstlisting}
+
+$XXX$
+\\[.7\baselineskip]
+      {\bf GCD}
+    \end{center}
+  \end{minipage}
+
+   \caption{Well-Known Algorithms}
+  \label{fig:cat3}
+\end{figure*}
+
+\paragraph{Well-Known Algorithms}
+
 \bibliographystyle{abbrvnat}

### 2014-06-12 src/t/run.sh
free skip statements in the default metric

@@ -2,21 +2,43 @@ APP="./app"
 TDIR="t"
 TESTS="lex parse lannot eval q"
 
+if [ "$1" = "update" ]; then
+	until
+		sh $0
+	do
+		T=$?
+		echo "Update the test results? [no]"
+		read YN
+		case "$YN" in
+			("y"|"Y"|"yes") sh $0 `printf "t%02d" $T` ;;
+			(*) exit 1 ;;
+		esac
+	done
+	exit
+fi
+
 if [ -f "$TDIR/$1.while" ]; then
 	echo "generating test outputs for $TDIR/$1.while"
 	for TEST in $TESTS; do
 		echo "-- $TEST"
-		$APP -t$TEST < $TDIR/$1.while |& tee $TDIR/$TEST/$1.while
+		$APP -t$TEST < $TDIR/$1.while 2>&1 \
+			| tee $TDIR/$TEST/$1.while
 		echo ""
 	done
 	exit 0
 fi
 
-for TFILE in $TDIR/*.while; do
+for TFILE in $TDIR/t*.while; do
 	printf "testing $TFILE "
+	TNUM=${TFILE#$TDIR/t}; TNUM=${TNUM%.while}
 	for TEST in $TESTS; do
-		$APP -t$TEST < $TFILE |& diff - $TDIR/$TEST/`basename $TFILE` > /tmp/testdiff
-		[ $? -ne 0 ] && { echo " [$TEST failure] (see diff in /tmp/testdiff)"; exit 1; }
+		$APP -t$TEST < $TFILE 2>&1 \
+			| diff - $TDIR/$TEST/`basename $TFILE` > /tmp/testdiff
+		[ $? -ne 0 ] && {
+			echo " [$TEST failure]";
+			echo "*** diff"; cat /tmp/testdiff; echo;
+			exit $TNUM;
+		}
 		printf "[$TEST ok] "
 	done
 	echo ""

### 2014-06-12 paper/main.tex
fixed typo
more c versions of examples

@@
 
-$1.1|[y,z]| + 0.1[|0,y|]$
+$1.1|[y,z]| + 0.1|[0,y]|$
 \\[.7\baselineskip] 
@@
 
-$12|[n,0]| + 0.1|[0,y]|$
+$12|[n,0]| + 0.01|[0,y]|$
 \\[.7\baselineskip]
@@
 
-$4|[0,x]| + [|0,y|]$
+$4|[0,x]| + |[0,y]|$
 \\[.7\baselineskip]
@@
   \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  flag=1;
+  while (flag>0) {
+    if (n>0 && *) {
+      n=n-1;
+      flag=1;
+    } else
+      flag=0;
+  }
+   \end{lstlisting}
+
+$2 + 2|[0, n]|$
+\\[.7\baselineskip]
+      {\bf t47}
+    \end{center}
+  \end{minipage}
 
@@
 Example \emph{t13} shows how amortization can be used to handle tricky
-nested loops.  The outer loop is iterated $[|0,x|]$ times.  In the
+nested loops.  The outer loop is iterated $|[0,x]|$ times.  In the
 conditional, we either (the branching condition is again arbitrary)
@@
 
+Example \emph{t47} demonstrates how we can use integers as Booleans to
+amortize the cost of loops that depend on boolean flags.  The outer
+loop is executed as long as the variable flag is ``true'', that is,
+\code{flag>0}.  Inside the loop, there is a conditional that either
+(if $n>0$) decrements $n$ and assigns \code{flag=1}, or (if $n\geq0$)
+leaves n unchanged and assigns \code{flag=0}.  The analyzer computes
+the tight bound $2 + 2|[0, n]|$.  The potential in the loop invariant
+is $|[0,\mathit{flag}]| + 2|[0, n]|$.  In the \emph{then} branch of
+the conditional, we use the potential $2|[0, n]|$ and the fact that
+$n>0$.  In the \emph{else} branch, we use the potential
+$|[0,\mathit{flag}]|$ and the fact that $\mathit{flag}=1$.
+
 \begin{figure*}[t!]
@@
 \\[.7\baselineskip]
-      {\bf t11}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
- XXX
-   \end{lstlisting}
-
-$XXX$
-\\[.7\baselineskip]
-      {\bf t12}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
- XXX
-   \end{lstlisting}
-
-$XXX$
-\\[.7\baselineskip]
-      {\bf t22}
-    \end{center}
-  \end{minipage}
-
-   \caption{Examples from \cite{GulwaniMC09}}
+      {\bf fig2\_1}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x<n) {
+    if (z>x)
+      x=x+1;
+    else
+      z=z+1;
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[z, n]|$
+\\[.7\baselineskip]
+      {\bf fig2\_2}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (x<n) {
+    while (y<m) {
+      if (*) break;
+      y=y+1;
+    }
+    x=x+1;
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.7\baselineskip]
+      {\bf nested\_multiple}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  x=0;
+  while (x<n) {
+    x=x+1;
+    while (x<n) {
+      if (*) break;
+      x=x+1;
+    }
+  }
+   \end{lstlisting}
+
+$1 + |[0, n]|$
+\\[.7\baselineskip]
+      {\bf nested\_single}
+    \end{center}
+  \end{minipage}
+
+   \caption{Examples from Gulwani et al's SPEED (a)~\cite{GulwaniMC09}}
   \label{fig:cat2a}
@@
   \centering
-
-   \caption{Examples from [GulwaniPLDI09] and [GulwaniPLDI10]}
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  x=0;
+  while (x<n) {
+    if (*) break;
+    x=x+1;
+  }
+  while (x<n)
+    x=x+1;
+   \end{lstlisting}
+
+$1 + |[0,n]|$
+\\[.7\baselineskip]
+      {\bf sequential\_single}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  x=0; y=0;
+  while (x<n) {
+    if (y<m)
+      y=y+1;
+    else
+      x=x+1;
+  }
+   \end{lstlisting}
+$2 + |[0, m]| + |[0, n]|$
+\\[.7\baselineskip]
+      {\bf simple\_multiple}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  x=0;
+  while (x<n) {
+    if (*)
+      x=x+1;
+    else 
+      x=x+1;
+   \end{lstlisting}
+
+$1 + |[0,n]|$
+\\[.7\baselineskip]
+      {\bf simple\_single}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  x=0; y=0;
+  while (*) {
+    if (x<N) {
+      x=x+1; y=y+1;
+    } else if (y<M ) {
+      x=x+1; y=y+1;
+    } else
+      break;
+  }
+   \end{lstlisting}
+
+$2 + 2 |[0, M]| + 2 |[0, N]|$
+
+\\[.7\baselineskip]
+      {\bf simple\_single\_2}
+    \end{center}
+  \end{minipage}
+
+   \caption{Examples from Gulwani et al's SPEED (b)~\cite{GulwaniMC09}}
   \label{fig:cat2b}
 \end{figure*}
-
-\paragraph{From  the Literature}
-
-Example \emph{t11} has been given in~\cite{GulwaniMC09}.  In the loop,
-we increment
 
@@
   \centering
+
+   \caption{Examples from [GulwaniPLDI09] and [GulwaniPLDI10]}
+  \label{fig:cat2c}
+\end{figure*}
+
+\paragraph{From  the Literature}
+
+Our analyzer can derive almost all bounds for programs that have been
+described as challenges in the literature on bound generation.  There
+are only two types of programs that we were not able to handle:
+programs that require \emph{super-linear bounds} and program that
+require \emph{path-sensitive reasoning} to derive a bound.  In the
+following we list all linear examples from the literature that we
+found.  Two of the require path-sensitive reasoning.
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
 %

### 2014-06-12 src/examples/speed/popl10/fig2_1.c

@@ -1,0 +1,12 @@
+
+int main(int argc, char **argv)
+{
+  int n,m,x,y;
+
+  while (n>x) {
+    if (m>y)
+      y = y+1;
+    else
+      x = x+1;
+  }
+}

### 2014-06-12 src/examples/speed/popl10/fig2_2.c

@@ -1,0 +1,12 @@
+
+int main(int argc, char **argv)
+{
+  int x,z,n;
+
+  while (x<n) {
+    if (z>x)
+      x=x+1;
+    else
+      z=z+1;
+  }
+}

### 2014-06-12 src/examples/speed/popl10/nested_multiple.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int x,n,y,m;
+
+  while (x<n) {
+    while (y<m) {
+      if (0) break;
+      y=y+1;
+    }
+    x=x+1;
+  }
+
+}

### 2014-06-12 src/examples/speed/popl10/nested_single.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int x,n,y,m;
+
+  x=0;
+  while (x<n) {
+    x=x+1;
+    while (x<n) {
+      if (0) break;
+      x=x+1;
+    }
+  }
+}

### 2014-06-12 src/examples/speed/popl10/sequential_single.c

@@ -1,0 +1,13 @@
+
+int main(int argc, char **argv)
+{
+  int x,n,y,m;
+
+  x=0;
+  while (x<n) {
+    if (0) break;
+    x=x+1;
+  }
+  while (x<n)
+    x=x+1;
+}

### 2014-06-12 src/examples/speed/popl10/simple_multiple.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int x,n,y,m;
+
+  x=0; y=0;
+
+  while (x<n) {
+    if (y<m)
+      y=y+1;
+    else
+      x=x+1;
+  }
+}

### 2014-06-12 src/examples/speed/popl10/simple_single.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int x,n,y,m;
+
+  x=0;
+  while (x<n) {
+    if (0)
+      x=x+1;
+    else 
+      x=x+1;
+  }
+
+}

### 2014-06-12 src/examples/speed/popl10/simple_single_2.c

@@ -1,0 +1,19 @@
+
+int main(int argc, char **argv)
+{
+  int x,N,y,M;
+
+  x=0;
+  y=0;
+  while (1) {
+    if (x<N) {
+      x=x+1; y=y+1;
+    }
+    else if (y<M ) {
+      x=x+1; y=y+1;
+    }
+  else
+    break;
+  }
+
+}

### 2014-06-12 src/ctest/t15.c
restore useful assertion in C test 15

@@ -3,7 +3,7 @@ int main(int argc, char **argv)
 {
   int x,y,z;
 
-  //assert (y>=0);
+  assert (y>=0);
   while (x-y>0) {
     x=x-1;
     x=x-y;

### 2014-06-12 src/examples/speed/pldi10/ex1.c
more c versions of examples

@@ -1,0 +1,20 @@
+
+int main(int argc, char **argv)
+{
+  int i,j,n;
+
+  i=0;
+  while (i<n) {
+    j=i;
+    j=j+1;
+  }
+  while (j<n) {
+    if (0) {
+      _tick(1);
+      j=j-1;
+      n=n-1;
+    }
+    j=j+1;
+  }
+  i=i+1;
+}

### 2014-06-12 src/examples/speed/pldi10/ex3.c

@@ -1,0 +1,13 @@
+int main(int argc, char **argv)
+{
+  int t,x,n;
+
+  while (n>0) {
+    t = x;
+    n=n-1;
+    while (n>0) {
+      if (0) break;
+      n=n-1;
+    }
+  }
+}

### 2014-06-12 src/examples/speed/pldi10/ex4.c

@@ -1,0 +1,14 @@
+int main(int argc, char **argv)
+{
+  int flag,n;
+
+  flag=1;
+
+  while (flag>0) {
+    flag=0;
+    while (n>0) {
+      n=n-1;       
+      flag=1;
+    }
+  }
+}

### 2014-06-12 paper/main.tex
more work on the catalog

@@
 $2 + 2 |[0, M]| + 2 |[0, N]|$
-
 \\[.7\baselineskip]
@@
   \centering
-
-   \caption{Examples from [GulwaniPLDI09] and [GulwaniPLDI10]}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  assert n>0;
+  assert m>0;
+  va = n; vb = 0;
+  while (va>0 && *) {
+    if (vb<m) { 
+      vb=vb+1; 
+      va=va-1;
+    } else {
+      vb=vb-1;
+      vb=0;
+    }
+  }
+   \end{lstlisting}
+$4 + 4|[0, n]|$
+\\[.7\baselineskip]
+      {\bf fig\_4\_2}
+    \end{center}
+  \end{minipage}
+%
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  assert (0<m);
+  i = n;
+  while (i>0 && *) {
+    if (i<m)
+      i=i-1;
+    else
+      i=i-m;
+  }
+   \end{lstlisting}
+$1 + 1|[0, n]|$
+\\[.7\baselineskip]
+      {\bf fig\_4\_4}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  assert (0 < m < n);
+  i=m;
+  while (0<i<n) {
+    if (dir==fwd) i++;
+    else i--;
+  }
+   \end{lstlisting}
+$---$
+\\[.7\baselineskip]
+      {\bf fig\_4\_5}
+    \end{center}
+  \end{minipage}
+
+   \caption{Examples from [GulwaniPLDI09]}
   \label{fig:cat2c}
+\end{figure*}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.24\linewidth}
+  \centering
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  i=0;
+  while (i<n) {
+    j=i; j=j+1;
+  }
+  while (j<n) {
+    if (0) {
+      _tick(1);
+      j=j-1; n=n-1;
+    }
+    j=j+1;
+  }
+  i=i+1;
+   \end{lstlisting}
+$1 + |[0, n]|$
+\\[.7\baselineskip]
+      {\bf ex1}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (n>0 && m>0) {
+    n--; m--;
+    while (nondet()) {
+      n--; m++;
+    }
+  }
+   \end{lstlisting}
+$---$
+\\[.7\baselineskip]
+      {\bf ex2}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (n>0) {
+    t = x;
+    n=n-1;
+    while (n>0) {
+      if (0) break;
+      n=n-1;
+    }
+  }
+   \end{lstlisting}
+$2 |[0, n]|$
+\\[.7\baselineskip]
+      {\bf ex3}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  flag=1;
+  while (flag>0) {
+    flag=0;
+    while (n>0) {
+      n=n-1;       
+      flag=1;
+    }
+  }
+   \end{lstlisting}
+$2 + 3|[0, n]|$
+\\[.7\baselineskip]
+      {\bf ex4}
+    \end{center}
+  \end{minipage}
+
+   \caption{Examples from [GulwaniPLDI10]}
+  \label{fig:cat3a}
 \end{figure*}

### 2014-06-12 src/ctest/t37.c
more c versions of examples

@@ -1,0 +1,27 @@
+
+void count_down (x) {
+  int a = x;
+  if (a>0) {
+    a = a-1;
+    count_down(a);
+  }
+}
+int copy (x,y) {
+  if (x>0) {
+    y = y+1;
+    x = x-1;
+    y=copy(x,y);
+  };
+  return y;
+}
+void _main (x,y) {
+  y = copy (x,y);
+  count_down(y);
+}
+
+
+
+int main(int argc, char **argv)
+{
+  int x,y;
+}

### 2014-06-12 src/ctest/t39.c

@@ -1,0 +1,30 @@
+
+void count_up (int,int);
+void count_down (int,int);
+
+void count_down (int x,int y)
+{ int a = x;
+  if (a>y) {
+    a = a-1;
+    count_up(a,y);
+  }
+}
+
+void count_up (int x, int y)
+{ int a = y;
+  if (a+1<x) {
+    a = a+2;
+    count_down(x,a);
+  }
+}
+
+void _main (int y, int z) {
+  count_down(y,z);
+}
+
+int main(int argc, char **argv)
+{
+  int x,y;
+}
+
+

### 2014-06-12 src/ctest/t46.c

@@ -1,0 +1,25 @@
+int x,y;
+
+void _tick (int x) {}
+
+void produce () {
+  while (x>0) {
+    _tick(-1); x=x-1; y=y+1;
+  }
+}
+void consume () {
+  while (y>0) {
+    y=y-1; x=x+1; _tick(1);
+  }
+}
+void _main (int y, int z) {
+  consume(); produce(); consume();
+}
+
+
+int main(int argc, char **argv)
+{
+  int x,y;
+}
+
+

### 2014-06-12 src/examples/speed/pldi09/fig1.c
more c versions of examples

@@ -1,0 +1,16 @@
+int main(int argc, char **argv)
+{
+  int x,y;
+
+  x=0;y=0;
+
+  while (1>0) {
+  if x < n (
+    y=y+1;
+    x=x+1
+  ) else if y>0
+    y=y-1
+  else
+    break
+)
+

### 2014-06-12 src/examples/speed/pldi09/fig4_2.c

@@ -1,0 +1,20 @@
+int main(int argc, char **argv)
+{
+  int va,vb,n,m;
+
+/* assert n>0; */
+/* assert m>0; */
+  va = n;
+  vb = 0;
+
+  while (va>0) {
+    if (vb<m) { 
+      vb=vb+1; 
+      va=va-1;
+    }
+    else {
+      vb=vb-1;
+      vb=0;
+    }
+  }
+}

### 2014-06-12 src/examples/speed/pldi09/fig4_4.c

@@ -1,0 +1,15 @@
+
+int main(int argc, char **argv)
+{
+  int i,n,m;
+
+  /* assert 0<m; */
+  i = n;
+
+  while (i>0) {
+    if (i<m)
+      i=i-1;
+    else
+      i=i-m;
+  }
+}

### 2014-06-12 src/examples/speed/pldi09/fig4_5.c

@@ -1,0 +1,14 @@
+
+int main(int argc, char **argv)
+{
+  int i,n,m,dir,fwd;
+
+  /* assert (0 < m < n);  */
+  i=m;
+
+  while (0<i<n) {
+    if (dir==fwd) i++;
+    else i--;
+  }
+
+}

### 2014-06-12 paper/main.tex
more work on the catalog

@@
    \end{lstlisting}
-$1 + |[0, n]|$
+$1 + |[0, n]| \text{ ticks}$
 \\[.7\baselineskip]
@@
 \begin{figure*}[t!]
+ \setlength{\progwidth}{.28\linewidth}
+  \centering
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+void count_down (int x) {
+  int a = x;
+  if (a>0) {
+    a = a-1;
+    count_down(a);
+  }
+}
+int copy (int x, int y) {
+  if (x>0) {
+    x = x-1;
+    y = y+1;
+    y=copy(x,y);
+  };
+  return y;
+}
+void _main (int x,int y) {
+  y = copy (x,y);
+  count_down(y);
+}
+   \end{lstlisting}
+
+$1 + 4|[0, x]| + 2|[0, y]|$
+\\[.7\baselineskip]
+      {\bf t37}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+void count_down (int x,int y)
+{ int a = x;
+  if (a>y) {
+    a = a-1;
+    count_up(a,y);
+  }
+}
+
+void count_up (int x, int y)
+{ int a = y;
+  if (a+1<x) {
+    a = a+2;
+    count_down(x,a);
+  }
+}
+
+void _main (int y, int z) {
+  count_down(y,z);
+}
+   \end{lstlisting}
+
+$1.67 + 1.33 |[z,y]|$
+\\[.7\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+void produce () {
+  while (x>0) {
+    _tick(-1); x=x-1; y=y+1;
+  }
+}
+void consume () {
+  while (y>0) {
+    y=y-1; x=x+1; _tick(1);
+  }
+}
+void _main (int y, int z) {
+  consume(); produce(); consume();
+}
+   \end{lstlisting}
+
+$|[0, y]|$ ticks
+\\[.7\baselineskip]
+      {\bf t46}
+    \end{center}
+  \end{minipage}
+
+   \caption{Programs with (recursive) functions}
+  \label{fig:cat3}
+\end{figure*}
+
+\paragraph{Recursive Functions}
+
+\begin{figure*}[t!]
  \setlength{\progwidth}{.24\linewidth}

### 2014-06-13 paper/main.tex
done with the catalog for now

@@
 simplicity, we use cost metric that counts cost $1$ for assignments
-and cost $0$ for all other operations.  Of course, the examples can
-also be analyzed with any other cost metric.
-
-We assume that the variables in the code snippets are the inputs to
+and cost $0$ for all other operations.  Sometimes we also use the
+\emph{ticks metric} if we want to discuss features like returning
+resources.  Of course, the examples can also be analyzed with any other
+cost metric.
+
+We assume that free variables in the code snippets are the inputs to
 the program. Some of the examples contain constants on which the
@@
 
-   \caption{Examples from Gulwani et al's SPEED (a)~\cite{GulwaniMC09}}
+   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09}} (a)
   \label{fig:cat2a}
@@
 
-   \caption{Examples from Gulwani et al's SPEED (b)~\cite{GulwaniMC09}}
+   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b)}
   \label{fig:cat2b}
@@
    \end{lstlisting}
-$4 + 4|[0, n]|$
+$2 + 4|[0, n]|$
 \\[.7\baselineskip]
@@
    \end{lstlisting}
-$1 + 1|[0, n]|$
+$|[0, n]|$
 \\[.7\baselineskip]
@@
   while (i<n) {
-    j=i; j=j+1;
-  }
+    j=i+1;
     while (j<n) {
-    if (0) {
+      if (*) {
 	_tick(1);
@@
     i=i+1;
-   \end{lstlisting}
-$1 + |[0, n]| \text{ ticks}$
+  }
+   \end{lstlisting}
+$|[0, n]| \text{ ticks}$
 \\[.7\baselineskip]
@@
     while (n>0) {
-      if (0) break;
+      if (*) break;
       n=n-1;
@@
 
-Our analyzer can derive almost all bounds for programs that have been
-described as challenges in the literature on bound generation.  There
-are only two types of programs that we were not able to handle:
-programs that require \emph{super-linear bounds} and program that
-require \emph{path-sensitive reasoning} to derive a bound.  In the
-following we list all linear examples from the literature that we
-found.  Two of the require path-sensitive reasoning.
+Our analyzer can derive almost all linear bounds for programs that
+have been described as challenges in the literature on bound
+generation.  We found only three programs with a linear bound for
+which our analyzer could not find a tight bound.  The first one
+(\emph{fig\_4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
+  reasoning} to derive a bound.  The other two, \emph{ex2} and
+\emph{ex5} (see [GulwaniPLDI10]), from [GulwaniPLDI10] were described
+to have a linear bound but seem to be non-terminating.
+
+Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani' et
+al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
+require inference of a \emph{disjunctive invariant}.  In the abstract
+interpretation community, these invariants are known to be notoriously
+difficult to handle.
+%
+In example \emph{fig2\_1} we have one loop that first increments
+variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
+derive the tight bound $|[x, n]| + |[y, m]|$.
+%
+Example \emph{fig2\_2} is more tricky and trying to understand how it
+works may be challenging.  However, with the amortized analysis in
+mind, using the potential transfer reasoning, it is almost trivial to
+prove the termination.  While the SPEED tool has to find a fairly
+involved invariant for the loop, our tool is simply reasoning locally
+and works without any clever tricks. We obtain the tight bound $|[x,
+n]| + |[z, n]|$.
+
+Example \emph{nested\_multiple} is similar to example \emph{fig2\_1}.
+Instead of incrementing variable $y$ in the outer loop, $y$ is here
+potentially incremented multiple times in each iteration of the outer
+loop.  The idea of example \emph{nested\_single} is similar.  However,
+instead of incrementing variable $y$ in the inner loop, we increment
+$x$, the counter variable of the outer loop. Our analyzer derives a
+tight bound for both programs.  Note that a star \emph{*} in a
+branching condition denotes an arbitrary boolean condition that might
+of course change while iterating (non-deterministic choice).
+
+Example \emph{sequential\_single} is like example
+\emph{nested\_single}.  The only difference is that the inner loop of
+\emph{nested\_single} is now evaluated after the outer loop.  Example
+\emph{simple\_multiple} is a variant of example \emph{fig2\_1} and
+\emph{simple\_single} is a simple variant of \emph{nested\_single}.
+We derive tight bounds for all aforementioned programs.
+
+Example \emph{simple\_single\_2} uses conditionals and a \code{break}
+statement to control loop iterations.  If $x<N$ then variables $x$ and
+$y$ are incremented.  Otherwise, if $y<M$ then the same increment is
+executed.  If $y\geq M$ and $x\geq N$ then the loop is terminated with
+a break.  Our tool computes the bound $2 + 2 |[0, M]| + 2 |[0, N]|$.
+This bound is tight in the sense that there are inputs (such as $M =
+-100$ and $N = 100$) for which the bound precisely describes the
+execution cost.  However, SPEED can compute the more precise bound
+$\max(N,M)$.  We currently cannot express this bound in our system.
+
+Example \emph{fig\_4\_2} from [GulwaniPLDI09] is quite involved.
+Amortized reasoning helps to understand how we derive the bound $4 +
+4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
+constant potential to pay for the two first assignments.  We use the
+remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
+establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
+loop invariant.  In the \emph{if branch} of the conditional, we use
+the constant potential $4$ of the invariant to pay for the two
+assignments ($2$) and the potential of $|0,vb|$ ($2$).  Since we also
+know that $|[0, n]|>0$ we obtain constant potential $4$ and establish
+the loop invariant again.  In the \emph{else branch}, we use the
+potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$ potential
+units to pay for the assignment.  Since we have $4$ constant potential
+units left after the loop, we can subtract them from the initial
+potential $6 + 4|[0, n]|$.
+
+In example \emph{fig\_4\_4} it is essential that $m$ is positive.
+That ensures that we can obtain constant potential for the interval
+size $|[0,i]|$ in the else branch of the conditional since $|[0,i]|$
+decreases.  Example \emph{fig\_4\_5} is one of the three examples that
+we can not handle automatically.  The reasoning is based on the fact
+that the boolean value of the test \code{dir==fwd} does not change
+during the iteration of the loop.  As a result, the variable $i$ is
+either counted down to $0$ or up to $n$.  Our tool cannot handle
+example \emph{fig\_4\_5} because we don't do path sensitive reasoning.
+Note however that it would be more efficient to move the test
+\code{dir==fwd} outside of the loop (this would be also done by an
+optimizing compiler).  The resulting program can be analyzer by our
+tool.
+
+Example \emph{ex1} from [GulwaniPLDI10] specifically focusses on the
+code in \emph{if} statement.  So we use the \emph{tick metric} and
+insert \code{_tick(1)} inside the if statement to derive a bound on
+the number of times the code in the if statement is executed.  Note
+that we cannot derive a bound for the whole program since the outer
+loop is executed a quadratic number of times.  Nevertheless it is
+straightforward to derive a bound on the number of ticks using the
+amortized approach: In the if statement we know that $n>0$ and assign
+$n=n-1$.  So we can use the potential of the interval size $|[0,n]|$
+to pay for the tick.
+
+We cannot analyze example \emph{ex2} for which previous work reported
+a linear bound [GulwaniPLDI10].  The reason is that this example is
+not terminating since the inner loop is not bounded.  The same is true
+for example \emph{ex5} from the same paper, which is not listed here.
+Example \emph{ex2} is similar to example \emph{nested\_single} and
+\emph{ex3} is a variant of example \emph{t47}.
 
@@
 
+Our approach can naturally deal with mutually-recursive functions.
+The recursion patterns can be exactly the same that are used in
+iterations of loops.  In the following, we present three simple
+examples that illustrate the analysis of functions.
+
+Example \emph{t37} illustrates that the analyzer is able to perform
+inter-procedural size tracking.  The function \code{copy} adds the
+argument $x$ to the argument $y$ if $x$ is positive.  However, this
+addition is done in steps of $1$ in each recursive call.  The function
+\code{count\_down} recursively decrements its argument down to $0$.
+The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
+\code{_main} in which we first add $x$ to $y$ using the function
+\code{copy} and then count down the variable $y$ using the function
+\code{count\_down}.  The derived bound is tight.
+
+Example \emph{t39} uses mutual recursion.  The function
+\code{count\_down} is similar to the function with the same name in
+example \emph{t37}.  However, we do not count down to $0$ but to a
+variable $y$ that is passed as an argument and we call the function
+\code{count\_up} afterwards.  The function \code{count\_up} is dual to
+\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
+\code{count\_down}.  For the function $\_main$ that calls
+\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
+1.33 |[z,y]|$.
+
+Example \emph{t46} shows a program that uses and returns resources.
+Again, we use the \emph{tick metric} and the function \code{\_tick} to
+describe the resource usage.  The function \code{produce} produces
+$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
+receives one resource unit.  Similarly, the function \code{consume}
+consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
+$|[0,y]|$ for the function \code{\_main}.  This is only possible since
+amortized analysis naturally tracks the size changes to the variables
+$x$ and $y$, and the interaction between \code{consume} and
+\code{produce}.
+
 \begin{figure*}[t!]

### 2014-06-13 paper/main.tex
more work on intro

@@
 \newrefformat{fig}{Figure~\ref{#1}}
+\newrefformat{app}{Appendix~\ref{#1}}
 \newrefformat{tab}{Table~\ref{#1}}
@@
 \item We show with a publically available prototype implementation and
-  an experimental evaluation that the our novel amortized resource
+  an experimental evaluation that our novel amortized resource
   analysis works efficiently for realistic programs and that the
@@
 %
-Our approach complements existing works since it provides a semantic
+Our approach complements existing work since it provides a semantic
 foundation for the computation of bounds while still providing support
@@
 intervals of integer variables.  A typical bound would be $B(x,y,z) =
-10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ and we have for example
-$B(100,200,43) = 1043$.
+10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
+1043$).
 %
 Despite the apparent simplicity of the new analysis system, it is able
-to reproduce multiple results from the literature which were obtained
-using methods based on abstract interpretation.  In contrast with
-abstract interpretation--based methods, our technique does not require
-any fixpoint computations on an abstract domain to obtain loop
-invariants.  The mechanism we designed is able to leverage local
-assertions such as $x < y$ that we collect along the branching points
-to obtain global invariants.  We achieve this by generating a linear
+to reproduce results from the literature which were obtained using
+methods based on abstract interpretation.  In contrast with abstract
+interpretation--based methods, our technique does not require any
+fixpoint computations on an abstract domain to obtain loop invariants.
+The mechanism we designed is able to leverage local assertions such as
+$x < y$ that we collect along the branching points of the program to
+obtain global invariants.  We achieve this by generating a linear
 constraint system that reflects resource cost and size changes of
@@
 advanced control flow such as break, return, iteration over negative
-counter variables, and mutually-recursive functions.
+integer variables, and mutually-recursive functions.
 
@@
 since an extension to multivariate resource polynomials [XXX] would
-make the inference rules considerably more complected and should
+make the inference rules considerably more involved and should
 better be described separately.  However, we developed the linear
@@
 is parametric in a resource metric that can be defined by the user.
+Our shallow embedding in Coq enables the derivation of both, bounds
+that are parametric in the cost metric or bounds for one specific
+metric.
 
 We have evaluated the quantitative logic and the automatic analysis
-with system code and examples from the literature. XXX
+with system code and examples from the literature. \pref{app:cat}
+contains a $50$ (XXX) challenging loop and recursion patterns that we
+collected from open source software and the literature.  Our analysis
+can find asymptotically tight bounds for $48$ of these patterns and in
+most cases the derived constant factors are tight.  We compared our
+automatic analyzer to other resource-analysis systems.  However, our
+analyzer and Rank are currently the only publically available tools
+that can directly operate on C programs.  We selected $5$ example for
+which we derived linear bounds and translated them to term-rewriting
+systems by hand in order to run them with the tools KoAT and Pubs.
+Both tools and Rank were not able to find linear bounds for the
+examples.  We contacted the authors of LOOPUS and SPEED but were not
+able to obtain a tool to perform a comparison.
+
+To show the practicability of our approach we developed a case study
+with different real-time schedulers implemented in C.  To perform a
+schedulability analysis that takes into account the overhead of the
+scheduler we derived time bounds for context switching and
+initialization of the scheduler.  We then computed time bounds for a
+set of given user task and took account the previously computed
+scheduler overhead to determine whether the scheduling requirements of
+all user tasks can be met.
 
@@
 \section{Catalog of Automatically Analyzed Programs}
+\lable{app:cat}
 

### 2014-06-15 paper/main.tex
work on section 2

@@
 
-\section{Amortized Analysis for C Programs}
+\section{Informal Account}
 \label{sec:inform}
 
-\subsection{Amortized Analysis}
+In this section, we informally introduce the quantiatative program
+logic and the automatic amortized analysis for for Clight programs.
+The purpose of this section is to build some intuition that should
+make the following sections, which contain more details, easy to read.
+
+\subsection{Quantitative Hoare Logic}
+
 The idea that drives the design of our system is amortized analysis
-[XXX].  Say that a program $p$ evaluates on a program state $H$ by
-consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.  The
-basic idea of amortized analysis is to define a potential function
-$\Phi$ that maps program states to non-negative numbers, such that
-$\Phi(H) \ge n$.
-
-Since imperative programs can be composed we must enrich
-the judgement $(\_, \_) \Downarrow_{\_}$ to talk about the heap
-resulting from the program execution $H'$.  So we now write
-$(p, H) \Downarrow_n H'$ to say that $H'$ results from $p$'s
-execution.  We now use two potential functions, one
-that applies before the execution, and one that applies
-after.  The two functions must respect the natural relation
-$\Phi(H) \ge n + \Phi'(H')$.  Intuitively speaking, $\Phi$ must
-give enough \emph{potential} to, both, finish the computation
-and pay for $\Phi'$ on the resulting heap. That way, if $(H, p_1) \Downarrow_n H'$
-and $(H', p_2) \Downarrow_m H''$, we get $\Phi(H) \ge
-n + \Phi'(H')$ and $\Phi'(H') \ge m + \Phi''(H'')$, which
-can be composed to give $\Phi(H) \ge (n + m) + \Phi''(H'')$.
-What we have observed is that, if we define $\Phi \vdash p \dashv \Phi'$
-to be
-$$
-\forall H\,n\,H',~(H, p)\Downarrow_n H' \implies \Phi(H) \ge
-n + \Phi'(H'),
-$$
-then we get the following familiar looking rule:
+[XXX].  Assume that a program $p$ executes on a stating state
+$\state_0$ and consumes $n$ resource units of some quantity we are
+interested in.  We denote that by writing $(p, \state_0)
+\Downarrow_n$.  The basic idea of amortized analysis is to define a
+potential function $\Phi$ that maps program states to non-negative
+numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
+program state such that such that $(p, \state) \Downarrow_n$.
+
+Since programs can be composed we must enrich the evaluation judgement
+$(p, \state) \Downarrow_{n}$ to talk about the state resulting from
+the program execution.  We write $(p, \state) \Downarrow_n \state'$ to
+say that $\state'$ is the program state after the execution of $p$ in
+starting state $\sigma$.  We now use two potential functions, one that
+applies before the execution, and one that applies after.  The two
+functions must respect the relation $\Phi(\state) \ge n +
+\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(p,
+\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
+provide enough \emph{potential} for, both, paying for resource cost of
+the computation and paying for the potential $\Phi'(\state')$ on the
+resulting state $\state'$. That way, if $(\state, p_1) \Downarrow_n
+\state'$ and $(\state', p_2) \Downarrow_m \state''$, we get
+$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
+\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
++ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
+provides an upper bound on the resource consumption of the whole
+program.  What we have observed is that, if we define $\htriple{\Phi}
+{p}{\Phi'}$ to mean
+$$
+\forall \state, n, \state' : (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
+n + \Phi'(\state')
+$$
+then we get the following familiar looking rule.
 $$
 \RuleNolabel
-{\Phi \vdash p_1 \dashv \Phi' \\ \Phi' \vdash p_2 \dashv \Phi''}
-{\Phi \vdash p_1; p_2 \dashv \Phi''}
-.
-$$
-Similarly, other language constructs contrain the potential
-functions to respect more rules that look very similar to
-Hoare logic or effect systems rules.  By fixing the shape of
-a potential function $\Phi$, it becomes possible to leverage
-existing tools to help finding a pair of potential functions
-for an arbitrary program.  This is explained in further details
-in the following sections.
-
-We say that the analysis is amortized because the use of a
-potential function $\Phi$ gives the flexibility to amortize some
-operations.  For example, if heaps are simple mappings from
-$\{\code x, \code y\}$ to natural numbers, we have the following
-judgement (provided that $H~\code x \ge H~\code y \ge 0$ initially)
-$$
-(\Phi := \lambda H.~2\cdot(H~\code x) + 10)
-\vdash \code x = \code x - \code y \dashv
-(\Phi' := \lambda H'.~2\cdot(H'~\code x) + 2\cdot(H'~\code y) + 10).
-$$
-This must be understood as a motion of potential from $\code x$
-to $\code y$. If some resource is consumed $\code y$ times afterwards, we
-can now use $\code y$'s potential to pay for it. But interestingly, the
-initial potential function $\Phi$ only talks about~$\code x$: the potential
-function allowed us to impart the cost of the subsequent resource
-consumptions on the variable~$\code x$.
-This flexibility combined with intelligent decisions on potential
-motions allow our tool to discover linear invariants for surprisingly
-convoluted programs.
+{\htriple{\Phi}{p_1}{\Phi'} \\ \htriple{\Phi'}{p_2}{\Phi''}}
+{\htriple{\Phi} {p_1; p_2}  {\Phi''}}
+$$
+%
+Similarly, other language constructs lead to rules for the potential
+functions that look very similar to Hoare logic or effect systems
+rules.  This rules enable to interactively reason about resource usage
+in a flexible and compositional way, which, as a side effect, produces
+a certificate for the derived resource bound.
+
+\paragraph{Automatic Amortized Analysis}
+
+By fixing the shape of a potential function $\Phi$, it becomes
+possible to leverage existing tools to help finding a pair of
+potential functions for an arbitrary program.  This is explained in
+further details in the following sections.
+
+We say that the analysis is amortized because the use of a potential
+function $\Phi$ gives the flexibility to amortize some operations.
+For example, if program states are simple mappings from variables to
+natural numbers, we have the following judgement (provided that
+initially $\state(\code x) \ge \state(\code y) \ge 0$).
+$$
+\htriple
+{\lambda \state . \, 2 {\cdot} \state (\code x) + 10}
+{\code{ x = x -  y}}
+{\lambda \state . \, 2 {\cdot} \state(\code x) + 2 {\cdot} \state(\code y) + 10}
+$$
+This must be understood as a motion of potential from the
+variable~$\code x$ to the variable~$\code y$. If some resource is
+consumed $\code y$ times afterwards, we can then use $\code y$'s
+potential to pay for it. However, the initial potential function
+$\Phi$ only refers to the variable~$\code x$: the potential function
+allows us to pass on the cost of the subsequent resource consumption
+to~$\code x$.  This flexibility combined with intelligent decisions on
+potential motions allow our tool to discover linear invariants for
+surprisingly convoluted programs.
 
@@
 \section{Catalog of Automatically Analyzed Programs}
-\lable{app:cat}
+\label{app:cat}
 
@@
 \end{document}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% mode: flyspell
+%%% TeX-master: t
+%%% End:

### 2014-06-15 paper/main.tex
more work on section 2

@@
 
-\paragraph{Automatic Amortized Analysis}
-
-By fixing the shape of a potential function $\Phi$, it becomes
-possible to leverage existing tools to help finding a pair of
-potential functions for an arbitrary program.  This is explained in
-further details in the following sections.
+We have implemented and proved sound such a quantitative Hoare for
+CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
+embedding in Coq, every function that can be expressed in Coq can in
+principle be used as a resource bound.  We can use the logic to prove
+a concrete bound for a given resource metric, or a bound that is
+parametric in set of resource metrics.  It is also possible to
+incorporate boolean preconditions into the bounds to express that the
+bound is only valid for a certain class of inputs.  To this end, we
+allow the potential function $\Phi$ in the pre and post conditions to
+take non-negative number or $\infty$ (infinity) as values.  Infinity,
+plays the same role as $\bot$ in Hoare logic.  Boolean assertions can
+be embedded into potential functions by mapping $\bot$ to $\infty$ and
+$\top$ to $0$.
+
+It is for instance possible to derive the following quantitative
+triple that holds for all stack metrics $M$.  A stack metric assigns
+cost $0$ to all evaluation steps except function calls.  Before a call
+$f(x)$, we consume $M(f) > 0$ resources and after the call we return
+$-M(f)$ resources.  So $M(f)$ corresponds to the stack-frame size of
+the function $f$.
+$$
+XXX
+$$
+XXX explanation of triple XXX
 
@@
 to~$\code x$.  This flexibility combined with intelligent decisions on
-potential motions allow our tool to discover linear invariants for
-surprisingly convoluted programs.
+potential motions allows us to design an automatic analysis to
+discover linear invariants for surprisingly convoluted programs.
+
+\subsection{Automatic Amortized Analysis}
+
+A program logic provides a principled foundation for statically
+analyzing programs.  However, program logics need to be supported by
+automatic methods to be useful practice.  That is why we developed a
+new automatic amortized analysis for deriving quantitative Hoare
+triples for common C programs.
+
+To this end, we fix the shape of the potential functions $\Phi$ so
+that it becomes possible to use \emph{linear programming} to compute a
+derivation in the logic.  If we assume that a program state $\state$
+simply maps variable to integers then we require
+$$
+\Phi(\state) = \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
+$$
+where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.
 

### 2014-06-16 paper/main.tex
more work on section 2

@@
 To this end, we fix the shape of the potential functions $\Phi$ so
-that it becomes possible to use \emph{linear programming} to compute a
-derivation in the logic.  If we assume that a program state $\state$
-simply maps variable to integers then we require
-$$
-\Phi(\state) = \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
-$$
-where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.
-
-\subsection{Motivating Examples}
-
-We now present a representative set of examples that can be
-handled by our tool.  The bounds that are given are for the number
-of ``base statements'' executed by the program, that is, the number
-of assertions, skip and assignment statements executed.  The
-bounds are given as literal expressions where a variable name {\tt x}
-represents the initial value of this variable.
-
-\paragraph{Simple loops.} The simplest loop scheme that is in
-use in typical imperative programs is counting upwards or
-downwards to a given bound N.  The two example programs
-are given below, on their right is the bound inferred by our analysis.
-\begin{center}
-\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
-\lstinline|while x > N ( x = x - 1 )| & $|{\tt N} -  {\tt x}|$ \\
-\lstinline|while x < N ( x = x + 1 )| & $|{\tt N} -  {\tt x}|$ \\
-\end{tabular}
-\end{center}
-
-It is also frequent to iterate every $n$ elements in an array, also
-in this simple case our analysis works without any problem.  Note
-however that if we increment using a variable, it might very well be
-possible that the program diverges!  In this situation no worst
-case bound can be given and the analysis correctly fails.
-\begin{center}
-\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
-\lstinline|while x <= -2 ( x = x + 2 )| & $| {\tt x} | / 2$ \\
-\lstinline|while x <= z - y ( x = x + y )| & Infeasible \\
-\lstinline|assert y > 0; while x <= z - y ( x = x + y )| & $1 + |{\tt x} - {\tt z}|$
-\end{tabular}
-\end{center}
-In the last example, the constant 1 is added to the loop bound
-to account for the initial \texttt{assert} that consumes one
-step.  Also, since the bound is for the worst case, it is tight
-only when {\tt y} is 1.
-
-\paragraph{In the literature.}
-The two examples below are taken from Gulwani's paper~\cite{Gulwani-speed}.
-They are both handled by the Speed tool but require inference of a so-called
-disjunctive invariant.  In the abstract interpretation community, these
-invariants are known to be notoriously difficult to handle.
-
-\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}
-while x < n
-  if y < m    y = y + 1
-  else        x = x + 1
-\end{lstlisting}
-$$
-|{\tt m} - {\tt y}| + |{\tt n} - {\tt x}|
-$$
-\end{minipage}
-\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}
-while x < n
-  if z > x    x = x + 1
-  else        z = z + 1
-\end{lstlisting}
-$$
-|{\tt n} - {\tt x}| + |{\tt n} - {\tt z}|
-$$
-\end{minipage}
-
-\hspace{1ex}
-
-While the first
-example is simple to understand, the second one is much more
-tricky and trying to understand how it actually works is quite challenging.
-However, with the amortized analysis in mind, using the potential transfer
-reasonning, it is almost trivial to prove the termination.  While the Speed
-tool has to find a fairly involved invariant for the loop, our tool is simply
-reasoning locally and works without any special trick.
-%
-Our analysis handles without problems all the other examples
-in~\cite{Gulwani-speed} that have a linear bound.  Here is
-a quick review of some of them together with the bounds we
-inferred.
-
-\begin{minipage}[b]{0.30\textwidth}
-\begin{lstlisting}
-while x < n
-  while x < n
-    x = x + 1
-\end{lstlisting}
-$$
-|{\tt x} - {\tt n}|
-$$
-\end{minipage}
-\begin{minipage}[b]{0.30\textwidth}
-\begin{lstlisting}
-while x < n (x = x + 1);
-while x < n (x = x + 1)
-\end{lstlisting}
-$$
-|{\tt x} - {\tt n}|
-$$
-\end{minipage}
-\begin{minipage}[b]{0.30\textwidth}
-\begin{lstlisting}
-while x < n
-  ( while y < m (y = y+1)
-  ; x = x + 1 )
-\end{lstlisting}
-$$
-|{\tt x} - {\tt n}|
-+|{\tt y} - {\tt m}|
-$$
-\end{minipage}
-
-\paragraph{Amortized reasonning.}
-The two examples below show the power of amortized reasoning.
-Because they essentially rely on an amortized reasonning we believe
-that our tool is the only one able to deal with them.
-
-\begin{minipage}[b]{0.45\textwidth}
-\begin{lstlisting}
-assert y >= 1;
-while x + y <= N
-  ( x = x + y
-  ; z = y; z = z + 4
-  ; while z > 0 ( z = z - 1 ) )
-\end{lstlisting}
-$$
-1 + 8\,|{\tt N} - {\tt x}|
-$$
-\end{minipage}
-\begin{minipage}[b]{0.45\textwidth}
-\begin{lstlisting}
-while x > 0
-  ( x = x - 1
-  ; if ?
-      while s > 0 ( s = s - 1 )
-    else s = s + 1 )
-\end{lstlisting}
-$$
-3\,|{\tt x}| + |{\tt s}|
-$$
-\end{minipage}
-
-\hspace{1ex}
-
-Maybe surprisingly, the two examples have linear bounds.  In the first one, the
-worst case is when {\tt y} is 1, in this situation the inner loop will perform 5 times
-$|{\tt N} - {\tt x}|$ iterations. Indeed, for each decrement of {\tt x}, {\tt z} is set
-to 5 and iterated over.  The second example is classic in the amortized analysis
-community, it is the original example from Tarjan's paper~\cite{Tarjan-amort} that
-non deterministically pushes one element on a stack or pops many elements from
-that stack.  In the example source code, the stack depth is represented by the
-variable {\tt s} and the number of iterations is ruled by the initial value of {\tt x}.
+that it becomes possible to use \emph{linear programming} (LP) to
+compute a derivation in the logic.  If we assume that a program state
+$\state$ simply maps variable to integers then we require
+$$
+\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
+$$
+where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.  Note that
+$q_0$ is the constant potential and that we treat constants as program
+variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
+component of $\Phi(\state)$ if $x \in \dom{\state}$.
+
+We then develop an inference rule for each syntactic construct that
+derives sound triples $\htriple{\Phi}p{\Phi'}$ in the quantitative
+logic but enables inference using and LP solver.  This idea is best
+explained by example.  Assume for the remainder of this section that
+we are interested in bounding the number of \emph{ticks} in a program.
+In other words, we use a resource metric that assigns cost $n$ to the
+function call \code{tick(n)} and cost $0$ to all other operations.
+The following example can then be bounded by $|[x,y]|$.
+\begin{lstlisting}
+  while (x<y) {x=x+1; tick(1);}
+\end{lstlisting}
+To derive this bound in our automatic amortized analysis, we start
+with the initial potential $\Phi_0 = |[x,y]|$ (short for
+$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
+loop invariant.  For the loop body we then (like in Hoare logic) have
+to derive a triple like
+$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
+only do so if we exploit the fact that $x<y$ at the beginning of the
+loop body.  The reasoning then works as follows.  We start with the
+potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
+assignment.  If we denote the update version of $x$ after the
+assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
+the potential before and after the assignment \code{x=x+1} holds.
+This means that we have the potential $|[x,y]| + 1$ before the
+statement $tick(1)$.  Since \code{tick(1)} consumes one resource unit,
+we end up with potential $|[x,y]|$ after the loop body and have
+established the loop invariant again.
+
+More formally, we develop (in section XXX) a judgement 
+$$
+\Gamma; Q \vdash S \dashv Q'; \Gamma'
+$$
+such that $\Gamma$ contains logical assertions like $x<y$ that we
+collect along the conditional branches of the program and $Q$ is a
+family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
+indexed by the variables that are currently in scope.  The logical
+assertions in $\Gamma$ are of the form XXXX.  They are only used to
+determine at the assignment if we can extract constant potential to
+pay for future cost.  It is important to note that the reasoning the
+assertions in $\Gamma$ are very basic.  We do not perform any fixpoint
+computations or invariants.
+
+Note that our notion of a potential function is a refinement of the
+concept of a ranking function.  A potential function can be used like
+a ranking function if we use the tick metric and add \code{tick(1)} to
+every back edge (loops and function calls) of the program.  However, a
+potential function is more flexible.  For example, we can use a
+potential function to prove that the following program does not
+consume any resources in the tick metric.
+\begin{lstlisting}
+  while (x<y) {tick(-1); x=x+1; tick(1);}
+\end{lstlisting}
+Similarly we can prove that the following program can be bounded by
+$10|[x,y]|$.  In both cases, we reason exactly like in the first
+version of the while loop to prove the bound.  Of course, such loops
+with different notations can be seamlessly combined in a larger
+program.
+\begin{lstlisting}
+  while (x<y) {x=x+1; tick(10);}
+\end{lstlisting}
+
+To understand how the analysis works, it is instructive to look at some
+more involved examples.
 
@@
 
+\cleardoublepage
 \appendix
@@
 programs that can be automatically analyzed by our system.  For
-simplicity, we use cost metric that counts cost $1$ for assignments
+simplicity, we use a cost metric that counts cost $1$ for assignments
 and cost $0$ for all other operations.  Sometimes we also use the
@@
   y=y+1000;
-  while (y>=100){
+  while (y>=100 && *){
     y=y-100;
@@
 
-$1+3|[0,x]|$
+$3|[0,x]|$
 \\[.7\baselineskip]
@@
 
-\pref{fig:cat1} shows code snippets that need amortization are
-compositionality to obtain a whole program bound.
+Figures~\ref{fig:cat1a}, \ref{fig:cat1b}, and~\ref{fig:cat1c} show
+code snippets that need amortization are compositionality to obtain a
+whole program bound.
 
@@
 changes inside loops.  In the first loop, we increment $y$ by $2$ in
-each of the $|[0,x]|$ iterations.  For another thing it shows that we
-automatically recognize dead code if we find conflicting assertions on
-a branching path: After the second loop we know $y \leq 0$ and as a
-result can assign arbitrary potential inside the third loop where we
-know that $y>0$.  As a result, we obtain a tight bound.
+each of the $|[0,x]|$ iterations.  An in the second loop, we decrement
+$y$.  For another thing it shows that we automatically recognize dead
+code if we find conflicting assertions on a branching path: After the
+second loop we know $y \leq 0$ and as a result can assign arbitrary
+potential inside the third loop where we know that $y>0$.  As a
+result, we obtain a tight bound.
 
@@
 conditional, stands for an arbitrary assertion.  In each branch of the
-conditional we can obtain a constant potential $1$ since the interval
+conditional we can obtain the constant potential $1$ since the interval
 size $|[y,x]|$ is decreasing.
@@
 increment the variable $y$ or we execute an inner loop in which $y$ is
-counted back to $0$.  Then analysis computes a tight linear bound for
+counted back to $0$.  The analysis computes a tight linear bound for
 this program.  Again, the constants $0$ and $1$ in the inner loop can
@@
 Example \emph{t27} is similar to example \emph{t13}.  Instead of
-decrements the variable $x$ in the outer loop we this time increment
+decrementing the variable $x$ in the outer loop we this time increment
 the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
@@
 that this does not change the size of the interval and computes the
-tight loop bound $3|[y,x]|$.  The additional to loops are in the
+tight loop bound $3|[y,x]|$.  The additional two loops are in the
 program to show that the size tracking in the first loop works
@@
 true and that the program is terminated without further cost
-otherwise.  The cost of an assert is $1$ resource unit in our cost
-metric.  If we enter the loop then we know that $x>0$ and we can
-obtain constant potential from the assignment $x=x-1$.  After the
+otherwise.  If we enter the loop then we know that $x>0$ and we can
+obtain constant potential from the assignment \code{x=x-1}.  After the
 assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
-can share the potential $3[0,x]$ before the assignment $x=x-y$ between
-$3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we derive
-a tight linear bound.
+can share the potential $3|[0,x]|$ before the assignment \code{x=x-y}
+between $3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we
+derive a tight linear bound.
 
@@
 variable $z$ down to zero and perform $|[0,z]|$ interations.  However,
-insead of assigning $z=y$, we assign $z=2y+100$.  The analysis
+instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
 computes the tight linear bound $1+2|[0,x]|+103|[y,x]|$.  The
@@
 as long as more than $1$ unit is assigned to $|[0,x]|$.  Since we know
-that $y\geq 0$ the choice of the LP solver is optimal but this seems
+that, $y\geq 0$ the choice of the LP solver is optimal but this seems
 to be a coincidence.
@@
 executed if $x<y$ and the second loop body is only executed if $y<x$.
-The analyzers finds a tight bound.
+The analyzer finds a tight bound.
 
@@
 decremented by one.  Then the values of the variables $x$ and $y$ are
-switched using the local variable $t$ as a buffer.  Out analysis
+switched using the local variable $t$ as a buffer.  Our analysis
 infers the tight bound $4|[0,x]|+4|[0,y]|$.
@@
 
-Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani' et
+Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
 al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
@@
 mind, using the potential transfer reasoning, it is almost trivial to
-prove the termination.  While the SPEED tool has to find a fairly
+prove a bound.  While the SPEED tool has to find a fairly
 involved invariant for the loop, our tool is simply reasoning locally

### 2014-06-16 paper/main.tex
more work on the catalog

@@
 the constant potential $4$ of the invariant to pay for the two
-assignments ($2$) and the potential of $|0,vb|$ ($2$).  Since we also
-know that $|[0, n]|>0$ we obtain constant potential $4$ and establish
-the loop invariant again.  In the \emph{else branch}, we use the
-potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$ potential
-units to pay for the assignment.  Since we have $4$ constant potential
-units left after the loop, we can subtract them from the initial
-potential $6 + 4|[0, n]|$.
+assignments ($2$ units) and the potential of $|[0,vb]|$ ($2$ units).
+Since we also know that $|[0, n]|>0$ we obtain constant potential $4$
+and establish the loop invariant again.  In the \emph{else} branch, we
+use the potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$
+potential units to pay for the assignment.  Since we have $4$ constant
+potential units left after the loop, we can subtract them from the
+initial potential $6 + 4|[0, n]|$.
 
@@
 That ensures that we can obtain constant potential for the interval
-size $|[0,i]|$ in the else branch of the conditional since $|[0,i]|$
-decreases.  Example \emph{fig\_4\_5} is one of the three examples that
-we can not handle automatically.  The reasoning is based on the fact
-that the boolean value of the test \code{dir==fwd} does not change
-during the iteration of the loop.  As a result, the variable $i$ is
-either counted down to $0$ or up to $n$.  Our tool cannot handle
-example \emph{fig\_4\_5} because we don't do path sensitive reasoning.
-Note however that it would be more efficient to move the test
-\code{dir==fwd} outside of the loop (this would be also done by an
-optimizing compiler).  The resulting program can be analyzer by our
+size $|[0,i]|$ in the \emph{else} branch of the conditional since
+$|[0,i]|$ decreases.  Example \emph{fig\_4\_5} is one of the three
+examples that we can not handle automatically.  The execution is
+bounded because the boolean value of the test \code{dir==fwd} does not
+change during the iteration of the loop.  As a result, the variable
+$i$ is either counted down to $0$ or up to $n$.  Our tool cannot
+handle example \emph{fig\_4\_5} because we don't do path sensitive
+reasoning.  Note however that it would be more efficient to move the
+test \code{dir==fwd} outside of the loop (this would be also done by
+an optimizing compiler).  The resulting program could be analyzer by our
 tool.
@@
 Example \emph{ex1} from [GulwaniPLDI10] specifically focusses on the
-code in \emph{if} statement.  So we use the \emph{tick metric} and
+code in the \emph{if} statement.  So we use the \emph{tick metric} and
 insert \code{_tick(1)} inside the if statement to derive a bound on
@@
 amortized approach: In the if statement we know that $n>0$ and assign
-$n=n-1$.  So we can use the potential of the interval size $|[0,n]|$
+\code{n=n-1}.  So we can use the potential of the interval size $|[0,n]|$
 to pay for the tick.
@@
 for example \emph{ex5} from the same paper, which is not listed here.
-Example \emph{ex2} is similar to example \emph{nested\_single} and
-\emph{ex3} is a variant of example \emph{t47}.
+Finally, example \emph{ex2} is similar to example
+\emph{nested\_single}, and \emph{ex3} is a variant of example
+\emph{t47}.
 
@@
 The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
-\code{_main} in which we first add $x$ to $y$ using the function
+\code{\_main} in which we first add $x$ to $y$ using the function
 \code{copy} and then count down the variable $y$ using the function
@@
 \code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
-\code{count\_down}.  For the function $\_main$ that calls
+\code{count\_down}.  For the function $\_main$, which calls
 \code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
@@
 \\[.7\baselineskip]
-      {\bf GCD}
+      {\bf Greatest Common Divisor}
+    \end{center}
+  \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+ XXX
+   \end{lstlisting}
+
+$XXX$
+\\[.7\baselineskip]
+      {\bf Quick Sort}
     \end{center}

### 2014-06-16 paper/main.tex
work on syntax and semantics

@@
 
-Let us now start the formal part of this work.  We begin by defining
-a proper syntax for the language we will be using later on.  This is
-a more algebraic version of the surface syntax used in the previous
-examples.
-%
-\begin{mathpar}
-
-S := \code{skip}
-\mid \code{assert}(C)
-\mid x \gets E
-\mid \code{while}(C)~S
-\mid \code{if}(C)~S~\code{else}~S
+For the sake of simplicity, our quantitative program logic
+applies to the Clight language.  This language is the most
+abstract intermediate language used by CompCert.  It is very
+similar to vanilla C but the expressions are free of
+side-effects and there is only one looping construct.
+%
+\begin{align*}
+S &:= \code{assert}~E
+\mid \code{skip}
+\mid \code{break}
+\mid \code{return}~E
+\mid E \gets E
+\mid E \gets f(E^*)
+\\
+& \mid \code{loop}~S
+\mid \code{if}(E)~S~\code{else}~S
 \mid S;S
-
-\and
-
-E := x \mid x \pm y
-\end{mathpar}
-
-Note that expressions are kept extremely simple, this will help
-greatly when designing the analysis.  Moreover, programs
-containing fancier expressions can easily be compiled to this
-restricted form.  For example $a \gets b + c - d$ is semantically
-equivalent to $a \gets b; a \gets a + c; a \gets a - d$ which
-fits in the above syntax.
-
-The syntax of conditions, noted $C$, is left unspecified
-for the moment.  We will describe it later when the practical
-implementation details are provided.  The only semantic tool
-we need to deal with this unknown is an evaluation
-function $\llbracket \_ \rrbracket_{\_}$ such that, for any
-condition $C$ and any heap $H$ we have $\llbracket
-C \rrbracket_H \in \{ \top, \bot \}$.
-
-Another peculiarity is that there are no integer constants in the
-syntax.  Once again this is for the sake of simplicity.  In practice
-the concrete syntax recognizes ``variables'' like {\tt -4} or
-{\tt 60} as read only variables.  Any heap used later in the
-operational semantics will map these variables to their natural
-value.
-
-\section{Cost Semantics}
-
-The semantics for the syntax defined above needs to account
-for resource consumption.  To account for all kinds of resources
+\end{align*}
+%
+We added the builtin primitive $\code{assert}$ that checks
+if a condition is true before continuing the program
+execution.  This will be used to explicit invariants to the
+automatic analysis later.
+
+The expressions are left abstract in this presentation.  To
+present both the semantics and the program logic, we simply
+need an evaluation function that maps any $e \in E$ to
+a value $\llbracket e \rrbracket_H$ using $H$ as current
+heap.
+
+\section{Semantics}
+
+Values computed by evaluating expressions can be integers, or
+memory locations. In our semantics, integers have infinite
+precision.  While the program logic would perform as well on
+machine integers, our automatic analysis would be unsound.
+%
+\begin{align*}
+V &:= \code{int}~i \mid \code{ptr}~b
+\end{align*}
+
+To formally describe the cost of operation of a given
+program, our semantics tracks resource consumption while
+executing.  To account for all kinds of resources
 we leave it as a parameter of the semantics to be specified
@@
 (
-M_k, M_s, M_a,
-M_w^1, M_w^2, M_w^3,
+M_a, M_s, M_b, M_r, M_u, M_f,
+M_l,
 M_c^1, M_c^2,
-M_s^1, M_s^2
-).
+M_q
+)
 $$
@@
 consumed by an operation.  If one of these numbers is negative,
-it means that some resource is released by the program.
+it means that resources are released by the program.
 
@@
 $(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
-presented in the overview section appart  from the fact that
+presented in the introducing section appart from the fact that
 two resource numbers are tracked instead of one.  Intuitively
@@
 a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
-It is possible to inject $q \in \mathbb Q$ in this monoid with by taking
-$(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.  This injection
-is used implicitely in the semantics.
+It is possible to inject $q \in \mathbb Q$ in this monoid by
+turning it into $(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.
+This injection is used implicitely in the semantics.
 
@@
 
+\and \Rule{S:Assert}
+{ {\rm istrue}~\llbracket e \rrbracket_H}
+{ (H, \code{assert}~e) \Downarrow_{M_a} \code{seq}~H }
+
 \and \Rule{S:Skip}
 { }
-{ (H, \code{skip}) \Downarrow_{M_k} H }
-
-\and \Rule{S:Assign}
+{ (H, \code{skip}) \Downarrow_{M_s} \code{seq}~H }
+
+\and \Rule{S:Break}
 { }
-{ (H, x \gets e) \Downarrow_{M_s} H[x \mapsto \llbracket e \rrbracket_H] }
-
-\and \Rule{S:Assert}
-{ H \models C }
-{ (H, \code{assert}(C)) \Downarrow_{M_a} H }
-
-\and \Rule{S:WhileStop}
-{ \llbracket C \rrbracket_H = \bot }
-{ (H, \code{while}(C)~S) \Downarrow_{M_w^3} H }
-
-\and \Rule{S:WhileAbort}
-{  k < \llbracket x - y \rrbracket_H
-\\ (H, S) \Downarrow_{(q_1, q'_1)} \circ
-}
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) } \circ }
-
-\and \Rule{S:WhileStep}
-{  \llbracket C \rrbracket_H = \top
-\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
-\\ (H', \code{while}(C)~S) \Downarrow_{(q_2, q'_2)} X
-}
-{ (H, \code{while}(C)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }
+{ (H, \code{break}) \Downarrow_{M_b} \code{brk}~H }
+
+\and \Rule{S:Return}
+{ }
+{ (H, \code{return}~e) \Downarrow_{M_r} \code{ret}~H~\llbracket e \rrbracket_H }
+
+\and \Rule{S:Update}
+{ \llbracket e \rrbracket_H = \code{ptr}~b }
+{ (H, e \gets e') \Downarrow_{M_u} \code{seq}~H[b \mapsto \llbracket e' \rrbracket_H] }
+
+\and \Rule{S:FunCall}
+{ \Delta~f = ((a_1, \dots, a_n), S)
+\\ \llbracket e \rrbracket_H = \code{ptr}~b
+\\ (H +\!\!\!+ (a_i \mapsto \llbracket e_i \rrbracket_H)_i, S) \Downarrow_{(q,q')} \code{ret}~H'~v
+}
+{ (H, e \gets f(e_1, \dots, e_n))
+\Downarrow^r_{M_f\cdot(q,q')} \code{seq}~(H'\,\backslash\,(a_i)_i)[b \mapsto v] }
+
+\and \Rule{S:LoopSeq}
+{ (H, S) \Downarrow_{(q_1, q_1')} \code{seq}~H'
+\\ (H', \code{loop}~S) \Downarrow_{(q_2, q_2')} X }
+{ (H, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} X }
+
+\and \Rule{S:LoopBrk}
+{ (H, S) \Downarrow_{(q, q')} \code{brk}~H' }
+{ (H, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~H' }
+
+\and \Rule{S:LoopAbort}
+{ (H, S) \Downarrow_{(q, q')} X
+\\ \forall H.~X \neq \code{seq}~H \land X \neq \code{brk}~H }
+{ (H, \code{loop}~S) \Downarrow_{(q, q')} X }
 
@@
 {  (H, S_1) \Downarrow_{(q, q')} X
-\\ \llbracket C \rrbracket_H = \top
+\\ {\sf istrue}~\llbracket C \rrbracket_H
 }
@@
 {  (H, S_2) \Downarrow_{(q, q')} X
-\\ \llbracket C \rrbracket_H = \bot
+\\ {\sf isfalse}~\llbracket C \rrbracket_H
 }
@@
 \and \Rule{S:Seq}
-{  (H, S_1) \Downarrow_{(q_1, q'_1)} H'
+{  (H, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~H'
 \\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
 }
-{ (H, S_1; S_2) \Downarrow_{M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)} X }
+{ (H, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} X }
 
 \and \Rule{S:SeqAbort}
-{ (H, S_1) \Downarrow_{(q,q')} \circ }
-{ (H, S_1; S_2) \Downarrow_{(q,q')} \circ }
+{ (H, S_1) \Downarrow_{(q,q')} X
+\\ \forall H.~X \neq \code{seq}~H}
+{ (H, S_1; S_2) \Downarrow_{(q,q')} X }
 
@@
     flag=0;
-    while (n>0) {
+    while (n>0 && *) {
       n=n-1;       

### 2014-06-17 paper/main.tex
change notations
more work on section 3

@@
 \mid \code{return}~E
-\mid E \gets E
-\mid E \gets f(E^*)
+\mid x \gets E
+\mid x \gets f(E^*)
 \\
@@
 need an evaluation function that maps any $e \in E$ to
-a value $\llbracket e \rrbracket_H$ using $H$ as current
-heap.
+a value $\llbracket e \rrbracket_\state$ using $\state$ as current
+state.
 
@@
 
-Values computed by evaluating expressions can be integers, or
-memory locations. In our semantics, integers have infinite
-precision.  While the program logic would perform as well on
-machine integers, our automatic analysis would be unsound.
-%
-\begin{align*}
-V &:= \code{int}~i \mid \code{ptr}~b
-\end{align*}
+The operational semantics uses a finite map $\state$ from
+variable names to integers.  This map is represented as
+an associative list to model the semantics of the stack of
+our language. It can be extended with new bindings
+(hiding older bindings with the same name), shrinked by
+forgetting a list of bindings, and updated by changing one
+binding.  We note $(\state, x_1 \mapsto v_1, \dots, x_n \mapsto v_n)$
+an extended heap, $\state~\backslash~(x_1, \dots, x_n)$ a
+state where bindings of $x_1, \dots, x_n$ were removed and
+$\state[x \mapsto v]$ an updated state.
 
@@
 program, our semantics tracks resource consumption while
-executing.  To account for all kinds of resources
-we leave it as a parameter of the semantics to be specified
-by the user.  This parameter, called a metric, is a tuple of rational
-numbers with the following form
+executing.  The semantics is oblivious to the resource of
+interest for the user of our system.  We acheive this
+independence by parameterizing the semantics by a resource
+\emph{metric}, it is a tuple of rational numbers with the
+following form
 $$
@@
 M_q
-)
+).
 $$
@@
 The general form of an evaluation judgement is
-$(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
+$(\state, S) \Downarrow_{(q,q')} O$.  It is similar to the judgement
 presented in the introducing section appart from the fact that
@@
 { }
-{ (H, S) \Downarrow_{(0,0)} \circ }
+{ (\state, S) \Downarrow_{(0,0)} \circ }
 
 \and \Rule{S:Assert}
-{ {\rm istrue}~\llbracket e \rrbracket_H}
-{ (H, \code{assert}~e) \Downarrow_{M_a} \code{seq}~H }
+{ {\rm istrue}~\llbracket e \rrbracket_\state}
+{ (\state, \code{assert}~e) \Downarrow_{M_a} \code{seq}~\state }
 
@@
 { }
-{ (H, \code{skip}) \Downarrow_{M_s} \code{seq}~H }
+{ (\state, \code{skip}) \Downarrow_{M_s} \code{seq}~\state }
 
@@
 { }
-{ (H, \code{break}) \Downarrow_{M_b} \code{brk}~H }
+{ (\state, \code{break}) \Downarrow_{M_b} \code{brk}~\state }
 
@@
 { }
-{ (H, \code{return}~e) \Downarrow_{M_r} \code{ret}~H~\llbracket e \rrbracket_H }
+{ (\state, \code{return}~e) \Downarrow_{M_r}
+  \code{ret}~\state~\llbracket e \rrbracket_\state
+}
 
 \and \Rule{S:Update}
-{ \llbracket e \rrbracket_H = \code{ptr}~b }
-{ (H, e \gets e') \Downarrow_{M_u} \code{seq}~H[b \mapsto \llbracket e' \rrbracket_H] }
+{ }
+{ (\state, x \gets e) \Downarrow_{M_u}
+  \code{seq}~\state[x \mapsto \llbracket e \rrbracket_\state]
+}
 
 \and \Rule{S:FunCall}
-{ \Delta~f = ((a_1, \dots, a_n), S)
-\\ \llbracket e \rrbracket_H = \code{ptr}~b
-\\ (H +\!\!\!+ (a_i \mapsto \llbracket e_i \rrbracket_H)_i, S) \Downarrow_{(q,q')} \code{ret}~H'~v
-}
-{ (H, e \gets f(e_1, \dots, e_n))
-\Downarrow^r_{M_f\cdot(q,q')} \code{seq}~(H'\,\backslash\,(a_i)_i)[b \mapsto v] }
+{ \Delta~f = (x_1, \dots, x_n, S) \\
+  ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
+    \Downarrow_{(q,q')} \code{ret}~\state'~v
+}
+{ (\state, x \gets f(e_1, \dots, e_n)) \Downarrow^r_{M_f\cdot(q,q')}
+  \code{seq}~(\state'~\backslash~(x_i))[x \mapsto v]
+}
 
 \and \Rule{S:LoopSeq}
-{ (H, S) \Downarrow_{(q_1, q_1')} \code{seq}~H'
-\\ (H', \code{loop}~S) \Downarrow_{(q_2, q_2')} X }
-{ (H, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} X }
+{ (\state, S) \Downarrow_{(q_1, q_1')} \code{seq}~\state'
+\\ (\state', \code{loop}~S) \Downarrow_{(q_2, q_2')} O }
+{ (\state, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} O }
 
 \and \Rule{S:LoopBrk}
-{ (H, S) \Downarrow_{(q, q')} \code{brk}~H' }
-{ (H, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~H' }
+{ (\state, S) \Downarrow_{(q, q')} \code{brk}~\state' }
+{ (\state, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~\state' }
 
 \and \Rule{S:LoopAbort}
-{ (H, S) \Downarrow_{(q, q')} X
-\\ \forall H.~X \neq \code{seq}~H \land X \neq \code{brk}~H }
-{ (H, \code{loop}~S) \Downarrow_{(q, q')} X }
+{ (\state, S) \Downarrow_{(q, q')} O
+\\ \forall \state.~O \neq \code{seq}~\state \land O \neq \code{brk}~\state }
+{ (\state, \code{loop}~S) \Downarrow_{(q, q')} O }
 
 \and \Rule{S:IfThen}
-{  (H, S_1) \Downarrow_{(q, q')} X
-\\ {\sf istrue}~\llbracket C \rrbracket_H
-}
-{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^1 \cdot (q, q')} X }
+{  (\state, S_1) \Downarrow_{(q, q')} O
+\\ {\sf istrue}~\llbracket C \rrbracket_\state
+}
+{ (\state, \code{if}(C)~S_1~\code{else}~S_2)
+  \Downarrow_{M_c^1 \cdot (q, q')} O
+}
 
 \and \Rule{S:IfElse}
-{  (H, S_2) \Downarrow_{(q, q')} X
-\\ {\sf isfalse}~\llbracket C \rrbracket_H
-}
-{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^2 \cdot (q, q')} X }
+{  (\state, S_2) \Downarrow_{(q, q')} O
+\\ {\sf isfalse}~\llbracket C \rrbracket_\state
+}
+{ (\state, \code{if}(C)~S_1~\code{else}~S_2)
+  \Downarrow_{M_c^2 \cdot (q, q')} O
+}
 
 \and \Rule{S:Seq}
-{  (H, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~H'
-\\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
-}
-{ (H, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} X }
+{  (\state, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~\state'
+\\ (\state', S_2) \Downarrow_{(q_2, q'_2)} O
+}
+{ (\state, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} O }
 
 \and \Rule{S:SeqAbort}
-{ (H, S_1) \Downarrow_{(q,q')} X
-\\ \forall H.~X \neq \code{seq}~H}
-{ (H, S_1; S_2) \Downarrow_{(q,q')} X }
+{ (\state, S_1) \Downarrow_{(q,q')} O
+\\ \forall \state.~O \neq \code{seq}~\state}
+{ (\state, S_1; S_2) \Downarrow_{(q,q')} O }
 
@@
 
-Heaps manipulated by the operational semantics judgements are
-simple maps from variable names to integers.  We designate by
-$H[x \mapsto v]$ the heap $H$ updated to map the variable name
-$x$ to the number $v$.
-%
-The evaluation of expressions is trivial using heaps.
-\begin{mathpar}
-\llbracket x \rrbracket_H = H(x)
-\and
-\llbracket x \pm y \rrbracket_H = H(x) \pm H(y)
-\end{mathpar}
-
-In the evaluation judgement $(H, S) \Downarrow_{(q, q')} X$, the result $X$ can
-be either a new heap, which means the execution terminated successfully
-or it can be the \emph{busy} result $\circ$.  This indicates an early
+In the evaluation judgement $(\state, S) \Downarrow_{(q, q')} O$,
+the outcome $O$ specifies how the statement $S$ terminated.
+$$
+O := \circ
+\mid \code{seq}~\sigma
+\mid \code{brk}~\sigma
+\mid \code{ret}~\sigma~v
+$$
+The \code{seq} outcome is the simple fallthrough outcome, for example
+a \code{skip} statement will terminate with this outcome without changing
+the heap as pictured in the rule {\sc S:Skip}.  The \code{brk} and
+\code{ret} outcomes describe a statement that terminated with a break or
+return statement.  In the latter case, a return value is also part of the
+outcome.  The use of these different outcomes can be explained by considering
+the loop statement rules:  If the inner statement breaks
+then the loop exits using the rule {\sc S:LoopBrk}, if the inner statement
+falls through another iteration is run using the rule {\sc S:LoopSeq},
+and all other cases are handled in the rule {\sc S:LoopAbort}.
+
+The special outcome $\circ$ indicates an early
 termination, it can be triggered at any point using the rule {\sc S:Abort}

### 2014-06-17 paper/main.tex
uniformize notations
update discussion on \circ

@@
 {  (\state, S_1) \Downarrow_{(q, q')} O
-\\ {\sf istrue}~\llbracket C \rrbracket_\state
-}
-{ (\state, \code{if}(C)~S_1~\code{else}~S_2)
+\\ {\sf istrue}~\llbracket e \rrbracket_\state
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
   \Downarrow_{M_c^1 \cdot (q, q')} O
@@
 {  (\state, S_2) \Downarrow_{(q, q')} O
-\\ {\sf isfalse}~\llbracket C \rrbracket_\state
-}
-{ (\state, \code{if}(C)~S_1~\code{else}~S_2)
+\\ {\sf isfalse}~\llbracket e \rrbracket_\state
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
   \Downarrow_{M_c^2 \cdot (q, q')} O
@@
 The busy result is also useful to handle diverging programs with big-step
-judgements.  In case a program does not terminate, there is no heap $H'$
-such that $(H, S)\Downarrow_{(q, q')} H'$.  The only possible way to
-conclude a big-step judgement is to use the {\sc S:Abort} rule at some
-point in the computation.  One alternative to this technique is to use
-coinductive semantics~\cite{Leroy-coind}, however, using the busy
-result is precise enough in our setting of quantitative resource analysis and
-allows to have only one soundness proof for both unsafe and
-non-terminating programs.  Indeed, stating a result for
-$(H, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
+judgements.  In case a program does not terminate, it is impossible to
+derive $(\state, S)\Downarrow_{(q, q')} O$ with $O \neq \circ$.  The only
+possible way to conclude a big-step judgement is to use the {\sc S:Abort}
+rule at some point in the computation.  One alternative to this technique
+is to use coinductive semantics~\cite{Leroy-coind}, however, using the busy
+result prevents duplication and allows to have only one soundness proof for
+both unsafe and non-terminating programs.  Indeed, stating a result for
+$(\sigma, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
 programs, and partial computations at the same time.

### 2014-06-17 paper/main.tex
start working on the logic
generalize assert rule
more rules
more work on section 2

@@
 analyzing programs.  However, program logics need to be supported by
-automatic methods to be useful practice.  That is why we developed a
+automatic methods to be useful in practice.  That is why we developed a
 new automatic amortized analysis for deriving quantitative Hoare
@@
 We then develop an inference rule for each syntactic construct that
-derives sound triples $\htriple{\Phi}p{\Phi'}$ in the quantitative
-logic but enables inference using and LP solver.  This idea is best
+derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
+logic but enables inference by using an LP solver.  This idea is best
 explained by example.  Assume for the remainder of this section that
@@
 function call \code{tick(n)} and cost $0$ to all other operations.
-The following example can then be bounded by $|[x,y]|$.
-\begin{lstlisting}
-  while (x<y) {x=x+1; tick(1);}
+The cost of the following example can then be bounded by $|[x,y]|$.
+\begin{lstlisting}
+while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
 \end{lstlisting}
@@
 potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
-assignment.  If we denote the update version of $x$ after the
+assignment.  If we denote the updated version of $x$ after the
 assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
@@
 This means that we have the potential $|[x,y]| + 1$ before the
-statement $tick(1)$.  Since \code{tick(1)} consumes one resource unit,
+statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
 we end up with potential $|[x,y]|$ after the loop body and have
@@
 
+\begin{figure}[t]
+  \centering
+\begin{lstlisting}[mathescape]
+${\color{blue}%
+  \cdot; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+while (x<y) {
+  ${\color{blue}%
+  x{<}y; \; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  x=x+1; 
+  ${\color{blue}%
+  \dashv x{\leq}y; \, 1 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv x{\leq}y; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+}
+${\color{blue}%
+ \dashv x{\geq}y; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+\end{lstlisting}
+  \caption{Derivation of a bound on the number of ticks for Example~1.}
+  \label{fig:ex1}
+\end{figure}
 More formally, we develop (in section XXX) a judgement 
@@
 indexed by the variables that are currently in scope.  The logical
-assertions in $\Gamma$ are of the form XXXX.  They are only used to
-determine at the assignment if we can extract constant potential to
-pay for future cost.  It is important to note that the reasoning the
-assertions in $\Gamma$ are very basic.  We do not perform any fixpoint
-computations or invariants.
+context $\Gamma$ is simply a conjunction of inequalities between.  It
+is used to determine at the assignment if we can extract constant
+potential to pay for future cost.  It is important to note that the
+reasoning the assertions in $\Gamma$ are very basic.  We do not
+perform any fixpoint computations or invariants.  \pref{fig:ex1} shows
+the derivation for Example 1.  Note that the logical assertions in
+$\Gamma$ are only used in the assignment \code{x=x+1}.
+
+It would of course be possible to keep track more sophisticated
+assertions in the logical context $\Gamma$.  However, this simple form
+is sufficient for the examples we considered and we can efficiently
+decide queries such as $\Gamma \implies x<y \; ?$, which we are
+interested in.  Nevertheless, this logical part of our analysis system
+is orthogonal to the quantitative part and can be easily extended if
+necessary.
 
@@
 concept of a ranking function.  A potential function can be used like
-a ranking function if we use the tick metric and add \code{tick(1)} to
-every back edge (loops and function calls) of the program.  However, a
-potential function is more flexible.  For example, we can use a
-potential function to prove that the following program does not
+a ranking function if we use the tick metric and add the statement
+\code{tick(1)} to every back edge of the program (loops and function
+calls).  However, a potential function is more flexible.  For example,
+we can use a potential function to prove that Example~2 does not
 consume any resources in the tick metric.
 \begin{lstlisting}
-  while (x<y) {tick(-1); x=x+1; tick(1);}
-\end{lstlisting}
-Similarly we can prove that the following program can be bounded by
-$10|[x,y]|$.  In both cases, we reason exactly like in the first
-version of the while loop to prove the bound.  Of course, such loops
-with different notations can be seamlessly combined in a larger
-program.
-\begin{lstlisting}
-  while (x<y) {x=x+1; tick(10);}
-\end{lstlisting}
-
+  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
+\end{lstlisting}
+Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
+In both cases, we reason exactly like in the first version of the
+while loop to prove the bound.  Of course, such loops with different
+notations can be seamlessly combined in a larger program.
+\begin{lstlisting}
+  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
+\end{lstlisting}
+%
 To understand how the analysis works, it is instructive to look at some
 more involved examples.
+
+XXX add up and down XXX
+
+\newlength{\progwidth}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.22\linewidth}
+  \centering
+  \begin{minipage}[b]{.20\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n>x) {
+  ${\color{blue}%
+  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+  if (m>y) 
+    ${\color{blue}%
+    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
+    y = y+1;
+    ${\color{blue}%
+    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  else
+    ${\color{blue}%
+    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+    x = x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
+}
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.7\baselineskip]
+      {\bf speed\_1}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+while (x<n) {
+  ${\color{blue}%
+  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+  if (z>x)
+    ${\color{blue}%
+    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  else
+    ${\color{blue}%
+    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    z=z+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[z, n]|$
+\\[.7\baselineskip]
+      {\bf speed\_2}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  while (z-y>0) {
+    y = y+1;
+    tick(3);
+  }
+  while (y>9) {
+    y=y-10;
+    tick(1);
+  }
+   \end{lstlisting}
+
+$3.1|[y,z]| + 0.1|[0,y]|$
+\\[.7\baselineskip] 
+      {\bf t08}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n<0) {
+  n=n+1;
+  y=y+1000;
+  while (y>=100 && *){
+    y=y-100;
+    tick(5);
+  }
+  tick(9);
+}
+   \end{lstlisting}
+
+$59|[n,0]| + 0.05|[0,y]|$
+\\[.7\baselineskip]
+      {\bf t27}
+    \end{center}
+  \end{minipage}
+
+  \caption{Derivations of bounds on the number of ticks for
+    challenging examples.  In the potential functions, we only mention
+    the non-zero terms and in the logical context $\Gamma$ we only
+    mention assertions that we actually use in the reasoning.}
+  \label{fig:ex_list}
+\end{figure*}
 
@@
 
-\and \Rule{S:FunCall}
-{ \Delta~f = (x_1, \dots, x_n, S) \\
+\and \Rule{S:Call}
+{ \delta~f = (x_1, \dots, x_n, S) \\
   ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
@@
 \section{Proof System}
+
+\begin{figure*}[t]
+\begin{mathpar}
+
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q \implies \code{istrue}~\sem{e}_\state }
+  { \code{assert}~e }
+  { Q }
+}
+
+\and
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_s + Q }{ \code{skip} }{ Q }
+}
+
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
+}
+
+\and
+\Rule{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+
+\and
+\Rule{L:Call}
+{
+  \Delta f = \forall z.(P_f\,z, Q_f\,z) \\
+  P \models P_f \, y \, (\sem{e_1}_\state,\dots,\sem{e_n}_\state) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[x \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { x \gets f(e_1, \dots, e_n) }
+  { Q }
+}
+
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+
+\and
+\Rule{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+
+\and
+\Rule{L:Weaken}
+{
+  P \models P' \\
+  \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \\
+  Q' \models Q \\
+  B' \models B \\
+  \forall v.\, (R'\,v \models R\,v)
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \mathbb Q_0^+
+}
+{ \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x }
+}
+
+\and
+\Rule{L:Extend}
+{ \forall f\,P_f\,Q_f.\,
+  \Delta'f = \forall z.(P_f\,z, Q_f\,z)
+  \rightarrow
+  \forall y.\,
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+    \htriple{ P_f\,y }{ \delta f }{ \bot })
+\\
+  \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+
+\end{mathpar}
+
+\caption{Rules of the Quantitative Hoare Logic}
+\end{figure*}
 
@@
 
-\cleardoublepage
+\clearpage
 \appendix
@@
 (or negative) and that the same constant at different places.
-
-\newlength{\progwidth}
 

### 2014-06-18 paper/main.tex
small fixes in semantics/logic
more work on section 2

@@
 %
-To understand how the analysis works, it is instructive to look at some
-more involved examples.
+To understand how the analysis works, it is instructive to look at
+some more involved examples.  Examples \emph{speed\_1} and
+\emph{speed\_2} are taken from Gulwani et al~\cite{Gulwani-speed}.
+They are both handled by the SPEED tool~\cite{Gulwani-speed} but
+require inference of a \emph{disjunctive invariant}.  In the abstract
+interpretation community, these invariants are known to be notoriously
+difficult to handle.
+%
+In example \emph{speed\_1} we have one loop that first increments
+variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
+derive the tight bound $|[x, n]| + |[y, m]|$.
+%
+Example \emph{speed\_2} is more tricky and trying to understand how it
+works may be challenging.  However, with the amortized analysis in
+mind, using the potential transfer reasoning, it is almost trivial to
+prove a bound.  While the SPEED tool has to find a fairly
+involved invariant for the loop, our tool is simply reasoning locally
+and works without any clever tricks. We obtain the tight bound $|[x,
+n]| + |[z, n]|$.
 
@@
   \centering
-  \begin{minipage}[b]{.20\linewidth}
-    \begin{center}
-   \begin{lstlisting}
+\hspace{-0.4cm}
+  \begin{minipage}[b]{.18\linewidth}
+    \begin{center}
+   \begin{lstlisting}[]
 while (n>x) {
@@
 while (z-y>0) {
+  ${\color{blue}%
+  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
   y = y+1;
+  ${\color{blue}%
+  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
   tick(3);
-  }
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+${\color{blue}%
+\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
 while (y>9) {
+  ${\color{blue}%
+  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
   y=y-10;
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
   tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
 }
@@
 while (n<0) {
+  ${\color{blue}%
+  n{<}0; \,  P(n,y) \vdash }$
   n=n+1;
+  ${\color{blue}%
+  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
   y=y+1000;
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
   while (y>=100 && *){
+    ${\color{blue}%
+    y{>}99; \, 9 {+} P(n,y) \vdash }$
     y=y-100;
+    ${\color{blue}%
+    \dashv y{>}99; \, 14 {+} P(n,y) \vdash }$
     tick(5);
-  }
+    ${\color{blue}%
+    \dashv \cdot; \, 9 {+} P(n,y) }$
+  }
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
   tick(9);
-}
-   \end{lstlisting}
-
-$59|[n,0]| + 0.05|[0,y]|$
+  ${\color{blue}%
+  \dashv \cdot; \, P(n,y) \vdash }$
+}
+   \end{lstlisting}
+$59|[n,0]| {+} 0.05|[0,y]|$
 \\[.7\baselineskip]
@@
     the non-zero terms and in the logical context $\Gamma$ we only
-    mention assertions that we actually use in the reasoning.}
+    mention assertions that we actually use in the reasoning. In
+    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
+    {+} 0.05|[0,y]|$.}
   \label{fig:ex_list}
 \end{figure*}
+
+\begin{figure}[t]
+  \centering
+    \begin{minipage}[b]{\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+void count_down (int x,int y)
+{ int a = x;
+  if (a>y) {a=a-1; tick(1); count_up(a,y); }
+}
+void count_up (int x, int y)
+{ int a = y;
+  if (a+1<x) {a = a+2; tick(1); count_down(x,a); }
+}
+   \end{lstlisting}
+
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
+\\[.7\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+  \caption{Two mutually-recursive functions with the computed tick bounds.}
+  \label{fig:ex_rec}
+\end{figure}
+
+\subsection{Implementation}
+
+We do not yet generate certificates in the program logic that are
+checkable in Coq.  This is substantial additional engineering effort
+but is conceptually straightforward as we describe in detail in
+Section XXX.
 
@@
 \mid x \gets E
-\mid x \gets f(E^*)
+\mid x \gets f(x^*)
 \\
@@
 \and \Rule{S:Call}
-{ \delta~f = (x_1, \dots, x_n, S) \\
-  ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
+{ \delta~f = (\vec y, S) \\
+  ((\state, y_i \mapsto \state(x_i)), S)
     \Downarrow_{(q,q')} \code{ret}~\state'~v
 }
-{ (\state, x \gets f(e_1, \dots, e_n)) \Downarrow^r_{M_f\cdot(q,q')}
-  \code{seq}~(\state'~\backslash~(x_i))[x \mapsto v]
+{ (\state, r \gets f(\vec x)) \Downarrow_{M_f\cdot(q,q')}
+  \code{seq}~(\state'~\backslash~(x_i))[r \mapsto v]
 }
@@
 {
-  \Delta f = \forall z.(P_f\,z, Q_f\,z) \\
-  P \models P_f \, y \, (\sem{e_1}_\state,\dots,\sem{e_n}_\state) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[x \mapsto v])
+  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
+  P \models P_f \, y \, (\sigma(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[r \mapsto v])
 }
@@
   { M_f + P }
-  { x \gets f(e_1, \dots, e_n) }
+  { r \gets f(\vec x) }
   { Q }
@@
 { \forall f\,P_f\,Q_f.\,
-  \Delta'f = \forall z.(P_f\,z, Q_f\,z)
+  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
   \rightarrow
-  \forall y.\,
+  \forall y\,\vec v.\,
   (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y }{ \delta f }{ \bot })
+    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
 \\
@@
 \\[.7\baselineskip]
-      {\bf fig\_4\_2}
+      {\bf fig4\_2}
     \end{center}
@@
 \\[.7\baselineskip]
-      {\bf fig\_4\_4}
+      {\bf fig4\_4}
     \end{center}
@@
 \\[.7\baselineskip]
-      {\bf fig\_4\_5}
+      {\bf fig4\_5}
     \end{center}
@@
 which our analyzer could not find a tight bound.  The first one
-(\emph{fig\_4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
+(\emph{fig4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
   reasoning} to derive a bound.  The other two, \emph{ex2} and
@@
 
-Example \emph{fig\_4\_2} from [GulwaniPLDI09] is quite involved.
+Example \emph{fig4\_2} from [GulwaniPLDI09] is quite involved.
 Amortized reasoning helps to understand how we derive the bound $4 +
@@
 
-In example \emph{fig\_4\_4} it is essential that $m$ is positive.
+In example \emph{fig4\_4} it is essential that $m$ is positive.
 That ensures that we can obtain constant potential for the interval
 size $|[0,i]|$ in the \emph{else} branch of the conditional since
-$|[0,i]|$ decreases.  Example \emph{fig\_4\_5} is one of the three
+$|[0,i]|$ decreases.  Example \emph{fig4\_5} is one of the three
 examples that we can not handle automatically.  The execution is
@@
 $i$ is either counted down to $0$ or up to $n$.  Our tool cannot
-handle example \emph{fig\_4\_5} because we don't do path sensitive
+handle example \emph{fig4\_5} because we don't do path sensitive
 reasoning.  Note however that it would be more efficient to move the

### 2014-06-18 paper/main.tex
more work on section 2

@@
 
-XXX add up and down XXX
+Example \emph{t08} shows the ability of the analysis to handle
+negative and non-negative numbers.  Note that there are no
+restrictions on the signs of $z$ and $y$.  We also see that we can
+accurately track the size change of $y$ in the first loop.
+Furthermore, \emph{t08} shows that we do not handle the constants $1$
+or $0$ in any special way.  In all examples you could replace $0$ and
+$1$ with other constants like we did in the second loop and still
+derive a tight bound.  The only information, that the analyzer needs
+is $y \geq c$ before assigning $y = y - c$.
+
+Example \emph{t27} shows how amortization can be used to handle tricky
+nested loops.  The outer loop is iterated $|[n,0]|$ times.  In the
+outer loop, we increment the variable $n$ till $n = 0$.  In each of
+the $|[n,0]|$ iterations, we increment the variable $y$ by $1000$.  We
+then non-deterministically execute an inner loop that decrements $y$
+by $100$ until $y<100$.  The analysis can derive that only the first
+execution of the inner loop depends on the initial value of $y$.  We
+again derive a tight bound.
+
+As mentioned, the analysis naturally handles advanced control flow
+like \code{break} and \code{return} statements, and mutual recursion.
+\pref{fig:ex_rec} contains two mutually-recursive functions with their
+automatically derived bounds.  The function \code{count\_down}
+decrements its first argument $x$ until it reaches the second argument
+$y$. It then recursively calls the function \code{count\_up}.  The
+function \code{count\_up} is dual to \code{count\_down}.  Here, we
+count up $y$ by $2$ and recursively call \code{count\_down}.  Our
+analysis system can compute a tight bound.  This is comparable to the
+determination of the point at which to trains meet if they approach
+each other with different speeds.
+
+Please not that we can only show a very small selections of the
+programs that we can handle automatically here.  In \pref{app:cat} you
+find a list of more than $30$ classes of challenging programs that we
+can automatically analyze.  We hope that this gives you an impression
+of the capabilities of the analysis but we also point out that this
+list is far from being exhaustive.
 
@@
 
+We implemented the new automatic amortized analysis using CIL [XXX].
+CIL is a subset of C that is very similar to CompCert Clight but
+provides tool support that facilitated the implementation.  The
+implementation is publically available at XXX.  More details can be
+found in section XXX.
+
 We do not yet generate certificates in the program logic that are
@@
 \mid \code{break}
-\mid \code{return}~E
+\mid \code{return}~x
 \mid x \gets E
@@
 
-The operational semantics uses a finite map $\state$ from
-variable names to integers.  This map is represented as
-an associative list to model the semantics of the stack of
-our language. It can be extended with new bindings
-(hiding older bindings with the same name), shrinked by
-forgetting a list of bindings, and updated by changing one
-binding.  We note $(\state, x_1 \mapsto v_1, \dots, x_n \mapsto v_n)$
-an extended heap, $\state~\backslash~(x_1, \dots, x_n)$ a
-state where bindings of $x_1, \dots, x_n$ were removed and
-$\state[x \mapsto v]$ an updated state.
+The operational semantics uses a mutable state $\state$
+composed of two maps from variable names to integers. The
+first map $\theta$ represents the local variable of a
+function and the second map $\gamma$ gives values to global
+variables of the program.
+
+For simplicity, we assume that local variables and global
+variables are always disjoint sets, this simplifies both
+the semantics and the description of our analysis.  The
+mutable state $\state$ can be accessed and modified, the
+access is written functionally and is defined by
+$$
+(\theta, \gamma)(x) =
+\left\{
+\begin{array}{lr}
+\theta(x) & \mbox{ if } x \in \dom \theta \\
+\gamma(x) & \mbox{ if } x \in \dom \gamma
+\end{array}
+\right..
+$$
+We note $\state[x \mapsto v]$ the state that results from
+updating $\state$ to make $x$ map to $v$.  Similarly to
+the state access, the formal definition of this operation
+is a simple case disjunction.  Thanks to the assumption we
+made above we can be sure that the different cases never
+overlap.
 
@@
 executing.  The semantics is oblivious to the resource of
-interest for the user of our system.  We acheive this
-independence by parameterizing the semantics by a resource
-\emph{metric}, it is a tuple of rational numbers with the
-following form
+interest for the user of our system.  We achieve this
+independence by parameterizing by a resource
+\emph{metric}, it is a tuple of nine rational numbers and
+one map from expressions to rational numbers
 $$
 (
-M_a, M_s, M_b, M_r, M_u, M_f,
+M_a,
+M_b,
+M_r,
+M_u,
+M_f,
 M_l,
-M_c^1, M_c^2,
-M_q
+M_c^1,
+M_c^2,
+M_q,
+M_e
 ).
@@
 Each of these rational numbers indicates the amount of resource
-consumed by an operation.  If one of these numbers is negative,
+consumed by one operation.  If one of these numbers is negative,
 it means that resources are released by the program.
 
-The general form of an evaluation judgement is
-$(\state, S) \Downarrow_{(q,q')} O$.  It is similar to the judgement
-presented in the introducing section appart from the fact that
-two resource numbers are tracked instead of one.  Intuitively
-$(q, q')$ means that $q$ resources units are necessary to execute
-the program and that $q'$ units are available after its execution.
-For example, $\code{alloc}(3)$ consumes 3 heap cells, so
-the pair resulting from its evaluation would be $(3, 0)$; for
-$\code{free}(3)$, the pair would be $(0, 3)$.  When these
-statements are sequenced with \code{alloc} first, then \code{free},
-the evaluation yields $(3, 3)$.  These resource pairs can
-be composed with the following operation
-$$
-(q_1, q'_1) \cdot (q_2, q'_2) =
-\left\{
-\begin{array}{ll}
-(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
-(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
-\end{array}
-\right.
-.
-$$
-This operation can be seen as the sequential composition of
-two resource consumptions, note that it is not a commutative
-operation, the same way resource freeing and allocation do not
-commute.
-With $(0, 0)$ as a neutral element, the above composition defines
-a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
-It is possible to inject $q \in \mathbb Q$ in this monoid by
-turning it into $(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.
-This injection is used implicitely in the semantics.
-
-\begin{figure}[ht!]
+For technical reasons in our soundness proof and to be in line
+with CompCert's Clight, our language's semantics is defined
+in a small-step fashion.  The \pref{fig:opsem} contains the full
+list of reduction rules.  These rules provide a rewrite
+system for program configurations of the form
+$(\state, S, \cont, \cost)$, where $\state$ is the program state,
+$S$ is the statement being executed, $\cont$ is a continuation
+that describes what remains to be done once the execution of $S$
+is over, and $\cost$ is the non-negative quantity of resources
+available for further execution.  All rules that can decrease
+$\cost$ have the implicit side condition that the resource
+quantity available after the step is non-negative.  Using this
+resource counter we can define the resource consumption of a
+program as the minimal resource counter $\cost_*$ that lets the
+program run to termination without errors.
+
+The small-step semantics uses continuations to represent the
+context of execution of a given statement.
+A continuation can be a sequenced computation, it can also result
+from entering a loop or a function, or it can simply be the
+empty continuation (used to start a program).
+$$
+\cont := \Kstop
+\mid \Kseq S \cont
+\mid \Kloop S \cont
+\mid \Kcall r \theta \cont
+$$
+
+\begin{figure}[t!]
 \begin{mathpar}
-
-\Rule{S:Abort}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop \cont, \cost)
+\smallstep
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~x, \Kseq S \cont, \cost)
+\smallstep
+(\state, \code{return}~x, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~x, \Kloop \cont, \cost)
+\smallstep
+(\state, \code{return}~x, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
+\smallstep
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \delta f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma }
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep
+(\state, S_1, \cont, \cost {-} M_c^1)
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma }
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep
+(\state, S_2, \cont, \cost {-} M_c^2)
+}
+%
+\and
+\Rule{S:Seq}
 { }
-{ (\state, S) \Downarrow_{(0,0)} \circ }
-
-\and \Rule{S:Assert}
-{ {\rm istrue}~\llbracket e \rrbracket_\state}
-{ (\state, \code{assert}~e) \Downarrow_{M_a} \code{seq}~\state }
-
-\and \Rule{S:Skip}
-{ }
-{ (\state, \code{skip}) \Downarrow_{M_s} \code{seq}~\state }
-
-\and \Rule{S:Break}
-{ }
-{ (\state, \code{break}) \Downarrow_{M_b} \code{brk}~\state }
-
-\and \Rule{S:Return}
-{ }
-{ (\state, \code{return}~e) \Downarrow_{M_r}
-  \code{ret}~\state~\llbracket e \rrbracket_\state
-}
-
-\and \Rule{S:Update}
-{ }
-{ (\state, x \gets e) \Downarrow_{M_u}
-  \code{seq}~\state[x \mapsto \llbracket e \rrbracket_\state]
-}
-
-\and \Rule{S:Call}
-{ \delta~f = (\vec y, S) \\
-  ((\state, y_i \mapsto \state(x_i)), S)
-    \Downarrow_{(q,q')} \code{ret}~\state'~v
-}
-{ (\state, r \gets f(\vec x)) \Downarrow_{M_f\cdot(q,q')}
-  \code{seq}~(\state'~\backslash~(x_i))[r \mapsto v]
-}
-
-\and \Rule{S:LoopSeq}
-{ (\state, S) \Downarrow_{(q_1, q_1')} \code{seq}~\state'
-\\ (\state', \code{loop}~S) \Downarrow_{(q_2, q_2')} O }
-{ (\state, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} O }
-
-\and \Rule{S:LoopBrk}
-{ (\state, S) \Downarrow_{(q, q')} \code{brk}~\state' }
-{ (\state, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~\state' }
-
-\and \Rule{S:LoopAbort}
-{ (\state, S) \Downarrow_{(q, q')} O
-\\ \forall \state.~O \neq \code{seq}~\state \land O \neq \code{brk}~\state }
-{ (\state, \code{loop}~S) \Downarrow_{(q, q')} O }
-
-\and \Rule{S:IfThen}
-{  (\state, S_1) \Downarrow_{(q, q')} O
-\\ {\sf istrue}~\llbracket e \rrbracket_\state
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
-  \Downarrow_{M_c^1 \cdot (q, q')} O
-}
-
-\and \Rule{S:IfElse}
-{  (\state, S_2) \Downarrow_{(q, q')} O
-\\ {\sf isfalse}~\llbracket e \rrbracket_\state
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
-  \Downarrow_{M_c^2 \cdot (q, q')} O
-}
-
-\and \Rule{S:Seq}
-{  (\state, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~\state'
-\\ (\state', S_2) \Downarrow_{(q_2, q'_2)} O
-}
-{ (\state, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} O }
-
-\and \Rule{S:SeqAbort}
-{ (\state, S_1) \Downarrow_{(q,q')} O
-\\ \forall \state.~O \neq \code{seq}~\state}
-{ (\state, S_1; S_2) \Downarrow_{(q,q')} O }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep
+(\state, S, \cont, \cost {-} M_q)
+}
+%
 
@@
 
-In the evaluation judgement $(\state, S) \Downarrow_{(q, q')} O$,
-the outcome $O$ specifies how the statement $S$ terminated.
-$$
-O := \circ
-\mid \code{seq}~\sigma
-\mid \code{brk}~\sigma
-\mid \code{ret}~\sigma~v
-$$
-The \code{seq} outcome is the simple fallthrough outcome, for example
-a \code{skip} statement will terminate with this outcome without changing
-the heap as pictured in the rule {\sc S:Skip}.  The \code{brk} and
-\code{ret} outcomes describe a statement that terminated with a break or
-return statement.  In the latter case, a return value is also part of the
-outcome.  The use of these different outcomes can be explained by considering
-the loop statement rules:  If the inner statement breaks
-then the loop exits using the rule {\sc S:LoopBrk}, if the inner statement
-falls through another iteration is run using the rule {\sc S:LoopSeq},
-and all other cases are handled in the rule {\sc S:LoopAbort}.
-
-The special outcome $\circ$ indicates an early
-termination, it can be triggered at any point using the rule {\sc S:Abort}
-and bubbles up through the execution context.
-Sometimes there is no other choice than to use this latter rule, for
-example when an assertion fails.  In that case the side condition of
-{\sc S:Assert} is not met and the only rule that can possibly be applied
-is {\sc S:Abort}.
-
-The busy result is also useful to handle diverging programs with big-step
-judgements.  In case a program does not terminate, it is impossible to
-derive $(\state, S)\Downarrow_{(q, q')} O$ with $O \neq \circ$.  The only
-possible way to conclude a big-step judgement is to use the {\sc S:Abort}
-rule at some point in the computation.  One alternative to this technique
-is to use coinductive semantics~\cite{Leroy-coind}, however, using the busy
-result prevents duplication and allows to have only one soundness proof for
-both unsafe and non-terminating programs.  Indeed, stating a result for
-$(\sigma, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
-programs, and partial computations at the same time.
-
 \section{Proof System}
@@
 \begin{mathpar}
-
+%
 \Rule{L:Assert}
@@
 }
-
+%
 \and
@@
   \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
-  P \models P_f \, y \, (\sigma(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[r \mapsto v])
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
 }
@@
   (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
+    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
 \\
@@
 }
-
+%
 \end{mathpar}

### 2014-06-18 paper/main.tex
switch to smallstep

--- 2014-06-18 paper/main.tex
@@
 (
-M_a, M_s, M_b, M_r, M_u, M_f,
+M_a, M_b, M_r, M_u, M_f,
 M_l,
 M_c^1, M_c^2,
-M_q
+M_q, XXX
 ).
@@
 \begin{mathpar}
-
-\Rule{S:Abort}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop \cont, \cost)
+\smallstep
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~e, \Kseq S \cont, \cost)
+\smallstep
+(\state, \code{return}~e, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~e, \Kloop \cont, \cost)
+\smallstep
+(\state, \code{return}~e, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~e, \Kcall r \theta \cont, \cost)
+\smallstep
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \delta f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep
+(\state, S, \Kloop S \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma }
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep
+(\state, S_1, \cont, \cost {-} M_c^1)
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma }
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep
+(\state, S_2, \cont, \cost {-} M_c^2)
+}
+%
+\and
+\Rule{S:Seq}
 { }
-{ (\state, S) \Downarrow_{(0,0)} \circ }
-
-\and \Rule{S:Assert}
-{ {\rm istrue}~\llbracket e \rrbracket_\state}
-{ (\state, \code{assert}~e) \Downarrow_{M_a} \code{seq}~\state }
-
-\and \Rule{S:Skip}
-{ }
-{ (\state, \code{skip}) \Downarrow_{M_s} \code{seq}~\state }
-
-\and \Rule{S:Break}
-{ }
-{ (\state, \code{break}) \Downarrow_{M_b} \code{brk}~\state }
-
-\and \Rule{S:Return}
-{ }
-{ (\state, \code{return}~e) \Downarrow_{M_r}
-  \code{ret}~\state~\llbracket e \rrbracket_\state
-}
-
-\and \Rule{S:Update}
-{ }
-{ (\state, x \gets e) \Downarrow_{M_u}
-  \code{seq}~\state[x \mapsto \llbracket e \rrbracket_\state]
-}
-
-\and \Rule{S:Call}
-{ \delta~f = (\vec y, S) \\
-  ((\state, y_i \mapsto \state(x_i)), S)
-    \Downarrow_{(q,q')} \code{ret}~\state'~v
-}
-{ (\state, r \gets f(\vec x)) \Downarrow_{M_f\cdot(q,q')}
-  \code{seq}~(\state'~\backslash~(x_i))[r \mapsto v]
-}
-
-\and \Rule{S:LoopSeq}
-{ (\state, S) \Downarrow_{(q_1, q_1')} \code{seq}~\state'
-\\ (\state', \code{loop}~S) \Downarrow_{(q_2, q_2')} O }
-{ (\state, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} O }
-
-\and \Rule{S:LoopBrk}
-{ (\state, S) \Downarrow_{(q, q')} \code{brk}~\state' }
-{ (\state, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~\state' }
-
-\and \Rule{S:LoopAbort}
-{ (\state, S) \Downarrow_{(q, q')} O
-\\ \forall \state.~O \neq \code{seq}~\state \land O \neq \code{brk}~\state }
-{ (\state, \code{loop}~S) \Downarrow_{(q, q')} O }
-
-\and \Rule{S:IfThen}
-{  (\state, S_1) \Downarrow_{(q, q')} O
-\\ {\sf istrue}~\llbracket e \rrbracket_\state
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
-  \Downarrow_{M_c^1 \cdot (q, q')} O
-}
-
-\and \Rule{S:IfElse}
-{  (\state, S_2) \Downarrow_{(q, q')} O
-\\ {\sf isfalse}~\llbracket e \rrbracket_\state
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
-  \Downarrow_{M_c^2 \cdot (q, q')} O
-}
-
-\and \Rule{S:Seq}
-{  (\state, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~\state'
-\\ (\state', S_2) \Downarrow_{(q_2, q'_2)} O
-}
-{ (\state, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} O }
-
-\and \Rule{S:SeqAbort}
-{ (\state, S_1) \Downarrow_{(q,q')} O
-\\ \forall \state.~O \neq \code{seq}~\state}
-{ (\state, S_1; S_2) \Downarrow_{(q,q')} O }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep
+(\state, S, \cont, \cost {-} M_q)
+}
+%
 
@@
 O := \circ
-\mid \code{seq}~\sigma
-\mid \code{brk}~\sigma
-\mid \code{ret}~\sigma~v
+\mid \code{seq}~\state
+\mid \code{brk}~\state
+\mid \code{ret}~\state~v
 $$
@@
 both unsafe and non-terminating programs.  Indeed, stating a result for
-$(\sigma, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
+$(\state, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
 programs, and partial computations at the same time.

### 2014-06-18 paper/main.tex
update verbiage on semantics

--- 2014-06-18 paper/main.tex
@@
 
-Example \emph{t08} shows the ability of the analysis to handle
-negative and non-negative numbers.  Note that there are no
-restrictions on the signs of $z$ and $y$.  We also see that we can
-accurately track the size change of $y$ in the first loop.
-Furthermore, \emph{t08} shows that we do not handle the constants $1$
-or $0$ in any special way.  In all examples you could replace $0$ and
-$1$ with other constants like we did in the second loop and still
-derive a tight bound.  The only information, that the analyzer needs
-is $y \geq c$ before assigning $y = y - c$.
-
-Example \emph{t27} shows how amortization can be used to handle tricky
-nested loops.  The outer loop is iterated $|[n,0]|$ times.  In the
-outer loop, we increment the variable $n$ till $n = 0$.  In each of
-the $|[n,0]|$ iterations, we increment the variable $y$ by $1000$.  We
-then non-deterministically execute an inner loop that decrements $y$
-by $100$ until $y<100$.  The analysis can derive that only the first
-execution of the inner loop depends on the initial value of $y$.  We
-again derive a tight bound.
-
-As mentioned, the analysis naturally handles advanced control flow
-like \code{break} and \code{return} statements, and mutual recursion.
-\pref{fig:ex_rec} contains two mutually-recursive functions with their
-automatically derived bounds.  The function \code{count\_down}
-decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{count\_up}.  The
-function \code{count\_up} is dual to \code{count\_down}.  Here, we
-count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis system can compute a tight bound.  This is comparable to the
-determination of the point at which to trains meet if they approach
-each other with different speeds.
-
-Please not that we can only show a very small selections of the
-programs that we can handle automatically here.  In \pref{app:cat} you
-find a list of more than $30$ classes of challenging programs that we
-can automatically analyze.  We hope that this gives you an impression
-of the capabilities of the analysis but we also point out that this
-list is far from being exhaustive.
+XXX add up and down XXX
 
@@
 \subsection{Implementation}
-
-We implemented the new automatic amortized analysis using CIL [XXX].
-CIL is a subset of C that is very similar to CompCert Clight but
-provides tool support that facilitated the implementation.  The
-implementation is publically available at XXX.  More details can be
-found in section XXX.
 
@@
   (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
+    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
 \\

### 2014-06-19 paper/main.tex
change delta f to S_f
typos

--- 2014-06-18 paper/main.tex
@@
 count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis system can compute a tight bound.  This is comparable to the
-determination of the point at which to trains meet if they approach
+analysis system can compute a tight bound.  On this example, it
+amounts to computing the meeting point of two trains that approach
 each other with different speeds.
-
-Please not that we can only show a very small selections of the
+% This is comparable to the
+% determination of the point at which to trains meet if they approach
+% each other with different speeds.
+
+Please note that we can only show a very small selections of the
 programs that we can handle automatically here.  In \pref{app:cat} you
@@
    \begin{lstlisting}
-void count_down (int x,int y)
-{ int a = x;
-  if (a>y) {a=a-1; tick(1); count_up(a,y); }
-}
-void count_up (int x, int y)
-{ int a = y;
-  if (a+1<x) {a = a+2; tick(1); count_down(x,a); }
+void count_down (int x,int y) {
+  if (x>y) { tick(1); count_up(x-1,y); }
+}
+void count_up (int x, int y) {
+  if (y+1<x) { tick(1); count_down(x,y+2); }
 }
@@
 \htriple
-  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
+  { M_r + R\,x }{ \code{return}~x }{ Q }
 }
@@
       x=x+1;
+  }
    \end{lstlisting}

### 2014-06-20 paper/main.tex
minor changes

@@
 In this paper, we present a novel cost analysis system for C programs
-that works automatically, efficiently and precisely for many examples,
-can be used interactively if the automation fails, and produces
+that is based on a solid semantic foundation. It works automatically,
+efficiently and precisely for many examples, can be used interactively
+for manual bound derivation if the automation fails, and produces
 verifiable certificates for the bounds.  Our contributions are as
@@
 techniques.  The quantitative Hoare logic and the automatic amortized
-analysis enable semi-automatic interactive resource bound development
-in which an automatically derived bound can be used in the
-quantitative logic and a hand derived bound can be used by the
-automatic analysis.  Both the automatic amortized analysis and the
-quantitative logic are naturally compositional and describe the
-resource behavior of code fragments without referring to the source
-code.  As a result, we naturally derive global bounds that are
-functions of the input parameters of the program.  To the best of our
-knowledge, our system is the only one that can deal with resource like
-memory that may become available during the execution.
+analysis enable semi-automatic interactive resource-bound development.
+An automatically derived bound can be used in the quantitative logic
+and a hand derived bound can be used by the automatic analysis.  Both
+the automatic amortized analysis and the quantitative logic are
+naturally compositional and describe the resource behavior of code
+fragments without referring to the source code.  As a result, we
+naturally derive global bounds that are functions of the input
+parameters of the program.  To the best of our knowledge, our system
+is the only one that can deal with resource like memory that may
+become available during the execution.
 
@@
 with system code and examples from the literature. \pref{app:cat}
-contains a $50$ (XXX) challenging loop and recursion patterns that we
-collected from open source software and the literature.  Our analysis
-can find asymptotically tight bounds for $48$ of these patterns and in
-most cases the derived constant factors are tight.  We compared our
-automatic analyzer to other resource-analysis systems.  However, our
-analyzer and Rank are currently the only publically available tools
-that can directly operate on C programs.  We selected $5$ example for
-which we derived linear bounds and translated them to term-rewriting
-systems by hand in order to run them with the tools KoAT and Pubs.
-Both tools and Rank were not able to find linear bounds for the
-examples.  We contacted the authors of LOOPUS and SPEED but were not
-able to obtain a tool to perform a comparison.
+contains more than $30$ (XXX) challenging loop and recursion patterns
+that we collected from open source software and the literature.  Our
+analysis can find asymptotically tight bounds for $30$ all but 3 of
+these patterns and in most cases the derived constant factors are
+tight.  We compared our automatic analyzer to other resource-analysis
+systems.  We selected $5$ examples (see Section XXX) for which we
+derived linear bounds and translated them to term-rewriting systems by
+hand in order to run them with the tools KoAT and Pubs.  Both tools
+and Rank were not able to find linear bounds for the examples.  LOOPUS
+is able to derive a bound on XXX of the examples.  But in contrast to
+our analysis, the constant factors are not tight.  We also contacted
+the authors of SPEED but were not able to obtain a tool to perform a
+comparison.
 
@@
 \htriple
-  { Q \implies \code{istrue}~\sem{e}_\state }
+  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
   { \code{assert}~e }
@@
 \caption{Rules of the Quantitative Hoare Logic}
+\label{fig:logic}
 \end{figure*}
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue} \{ A(0, N) + N \} }$
+i=1; m=a[0];
+${\color{blue}%
+  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-i) \} }$
+while (i<N) {
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
+  if (a[i] > m)
+    m = a[i];
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
+  i = i+1;
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
+}
+${\color{blue} \{ 0 \} }$
+\end{lstlisting}
+\caption{Example derivation where we noted $A(i, N)$
+  for $\#\{ k \mid k \ge i \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
+\label{fig:exemple}
+\end{figure}
 

### 2014-06-20 paper/main.tex
add example derivation
more work on the intro

@@
 programs.  For example, stack and heap-space bounds are important to
-ensure the reliability of safety-critical systems [XXX (see PLDI)].
+ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
 % Clock-cycle bounds are needed to guarantee the safety of real-time
@@
 Static energy usage information is critical for autonomous systems and
-has applications in cloud computing [XXX?  (cite this guy form the
-us))].  Worst-case resource bounds can be used to prevent side-channel
-attacks through resource padding [XXX?].  Loop and
-recursion-depth bounds are used to ensure the accuracy of programs
-that are executed on unreliable hardware [XXX oopsla] and complexity
-bounds are needed to verify security protocols [XXX].
+has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
+Worst-case time bounds can help to create constant-time
+implementations that prevent side-channel
+attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
+bounds are used to ensure the accuracy of programs that are executed
+on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
+needed to verify cryptography protocols~\cite{BartheGB09}.
 % XXX Mention differential privacy here?
@@
 some more involved examples.  Examples \emph{speed\_1} and
-\emph{speed\_2} are taken from Gulwani et al~\cite{Gulwani-speed}.
-They are both handled by the SPEED tool~\cite{Gulwani-speed} but
+\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.
+They are both handled by the SPEED tool~\cite{GulwaniMC09} but
 require inference of a \emph{disjunctive invariant}.  In the abstract
@@
 ${\color{blue}%
-  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-i) \} }$
+  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-1) \} }$
 while (i<N) {
@@
 \caption{Example derivation where we noted $A(i, N)$
-  for $\#\{ k \mid k \ge i \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
+  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
 \label{fig:exemple}
@@
 Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
-al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
+al~\cite{GulwaniMC09}.  They are both handled by the SPEED tool but
 require inference of a \emph{disjunctive invariant}.  In the abstract

### 2014-06-21 paper/main.tex
add binary counter increment in the logic

@@
   \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-1) \} }$
-while (i<N) {
+while (i < N) {
   ${\color{blue}%
@@
   if (a[i] > m)
-    m = a[i];
+    m=a[i];
   ${\color{blue}%
     \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
-  i = i+1;
+  i=i+1;
   ${\color{blue}%
     \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
-}
-${\color{blue} \{ 0 \} }$
+} ${\color{blue} \{ 0 \} }$
 \end{lstlisting}
 \caption{Example derivation where we noted $A(i, N)$
-  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
-\label{fig:exemple}
+  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
+  the metric used here assigns a cost of 1 to every assignment
+  and 0 to all other operations.
+  }
+\label{fig:xmplmax}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}%
+  \{ \#_1(a) + 2k \}}$
+while (k > 0) {
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+  while (a[x] == 1 && x < N) {
+    ${\color{blue}%
+      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
+    a[x]=0;
+    ${\color{blue}%
+      \{ (a[x] = 0) + \#_1(a) + 2k + 1 \}}$
+    tick(1);
+    ${\color{blue}%
+      \{ \#_1(a) + 2k \}}$
+    x++; }
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+  if (x < N) {
+    ${\color{blue}%
+      \{ \#_1(a) + 1 + 2(k-1) + 1 \}}$
+    a[x]=1;
+    ${\color{blue}%
+      \{ \#_1(a) + 2(k-1) + 1 \}}$
+    tick(1); }
+  ${\color{blue}%
+    \{ \#_1(a) + 2(k-1) \}}$
+  k--;
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+} ${\color{blue} \{ 0 \}}$
+
+\end{lstlisting}
+\caption{Exemple derivation using amortized reasoning, we noted
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
+  tick metric that assigns cost $n$ to \lstinline{tick(n)} statments and
+  0 to other operations.
+  }
+\label{fig:xmplinc}
 \end{figure}

### 2014-06-21 paper/main.tex
adding more references

@@
 an active research area and recent years have seen many innovations.
-Notable tools that have been developed include Speed [XXX], KoAT
-[XXX], Pubs [XXX], RANK [XXX], and LOOPUS [XXX].
+Notable tools that have been developed include
+SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
+PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
 %
@@
 of the existing techniques in practice.
-\begin{enumerate}
+\begin{itemize}
 \item Analysis tools are black boxes that either deliver a result or fail
@@
 \item Existing techniques often rely on complex external tools such as
-  abstract interpretation based invariant generation [XXX speed] or
-  translation of the program into a term-rewriting system [costa,
-  tacas] without providing \emph{verifiable certificates} for the
-  correctness of the derived bound.
-\end{enumerate}
+  abstract interpretation based invariant
+  generation~\cite{GulwaniMC09} or translation of the program into a
+  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
+  providing \emph{verifiable certificates} for the correctness of the
+  derived bound.
+\end{itemize}
 %
 While there has been lot of progress in static quantitative analysis,
-Don Knuth correctly points out in a recent interview [XXX] that the
-state-of-the-art in formal quantitative methods still falls short
-in comparison with semantic techniques [xxx].
+Knuth correctly points out in a recent interview~\cite{KnuthInter}
+that the state-of-the-art in formal quantitative methods still falls
+short in comparison with semantic techniques.
 %
@@
   \begin{flushright}
-    -- Donald Knuth, 2014
+    -- Donald E. Knuth, 2014
   \end{flushright}
@@
 %
-\begin{itemize}
+\begin{enumerate}
 \item We define a operational cost semantics for C programs that
@@
 %   with respect to the operational cost semantics.
-\end{itemize}
+\end{enumerate}
 %

### 2014-06-21 paper/main.tex
some updates to the new logical derivation

@@
 $$
-\forall \state, n, \state' : (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
+\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
 n + \Phi'(\state')
@@
 \begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
 \begin{lstlisting}
@@
     \{ \#_1(a) + 2k \}}$
-  while (a[x] == 1 && x < N) {
+  while (x < N && a[x] == 1) {
     ${\color{blue}%
@@
     ${\color{blue}%
-      \{ (a[x] = 0) + \#_1(a) + 2k + 1 \}}$
+      \{ \#_1(a) + 2k + 1 \}}$
     tick(1);
@@
   ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
+    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
   if (x < N) {
     ${\color{blue}%
-      \{ \#_1(a) + 1 + 2(k-1) + 1 \}}$
+      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
     a[x]=1;
@@
     \{ \#_1(a) + 2k \}}$
-} ${\color{blue} \{ 0 \}}$
-
-\end{lstlisting}
+} ${\color{blue} \{ \#_1(a) \}}$
+
+\end{lstlisting}
+\end{minipage}
+
 \caption{Exemple derivation using amortized reasoning, we noted
   $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
-  tick metric that assigns cost $n$ to \lstinline{tick(n)} statments and
+  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
   0 to other operations.
@@
 \label{fig:xmplinc}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}%
+  \{ (Z = \log_2(h-l)) + Z \} }$
+bsearch(x,l,h) {
+  if (h-l > 1) {
+    ${\color{blue}%
+      \{ (h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
+    m = h + (h-l)/2;
+    ${\color{blue}%
+      \{ (m = \frac{h+l}{2} \land h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
+    if (a[m]>x) h=m; else l=m;
+    ${\color{blue}%
+      \{ (Z - 1 = \log_2(h-l)) + (Z-1) + 1 \} }$
+    l = bsearch(x,l,h); }
+  ${\color{blue}%
+    \{ Z-1 \} }$
+  return l;
+} ${\color{blue} \{ Z \} }$
+\end{lstlisting}
+\caption{Exemple derivation of a stack usage bound for a binary
+  search program, the metric used assigns a cost of 1 to function
+  calls and $-1$ to return statements, we noted $Z$ an arbitrary
+  logical variable.
+  }
 \end{figure}
@@
 \and \Rule{Q:Set}
-{ \forall u,~q_{yu} = q'_{xu} + q'_{yu}
+{ \forall u.\, q_{yu} = q'_{xu} + q'_{yu}
 \\ q'_{xy} \in \mathbb Q_0^+
@@
 %\\ q'_0 = q_0 + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i
-\\ \forall i \in \mathcal U,~ q'_i \ge q_i - p_i
-\\ \forall i \in \mathcal L,~ q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U \cup \mathcal L,~ q'_i \ge q_i
+\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
 }
@@
 First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
-and $\forall i \neq 0,~q'_i = q_i$.  Second, for any index $i$ such that
+and $\forall i \neq 0.\, q'_i = q_i$.  Second, for any index $i$ such that
 the relation between $q_i$ and $q'_i$ is not specified as a side condition
@@
 The semantic entailment on logical states is noted $\Gamma \models \Delta$,
-its meaning is that $\forall H,~ H\models\Gamma \implies H\models\Delta$.
+its meaning is that $\forall H.\, H\models\Gamma \implies H\models\Delta$.
 
@@
 the definition of $\mathcal U$ and $\mathcal L$ we get
-$\forall i\in\mathcal L,~ l_i \le f_i(H)$ and
-$\forall i\in\mathcal U,~ f_i(H)\le u_i$.
+$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
+$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
 These two inequalities justify the last step.
@@
   $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
-  We assume $\forall i \in \{1, 2\},~M_s^i \ge 0$, the other cases
+  We assume $\forall i \in \{1, 2\}.\, M_s^i \ge 0$, the other cases
   are similar.
@@
 \item Case {\sc S:WhileStep}.
-  Here again we assume $\forall i \in \{1, 2, 3\},~M_w^i \ge 0$.
+  Here again we assume $\forall i \in \{1, 2, 3\}.\, M_w^i \ge 0$.
   \begin{itemize}[topsep=0pt]
@@
 &\Leftrightarrow \neg\neg (\Gamma \models C) \\
-&\Leftrightarrow \neg\neg (\forall H, H\models\Gamma \implies H\models C) \\
+&\Leftrightarrow \neg\neg (\forall H.\, H\models\Gamma \implies H\models C) \\
 &\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\

### 2014-06-21 paper/main.tex
uniformize forall notations

@@
 \begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
 \begin{lstlisting}
@@
     \{ \#_1(a) + 2k \}}$
-  while (x < N && a[x] == 1) {
+  while (a[x] == 1 && x < N) {
     ${\color{blue}%
@@
     ${\color{blue}%
-      \{ \#_1(a) + 2k + 1 \}}$
+      \{ (a[x] = 0) + \#_1(a) + 2k + 1 \}}$
     tick(1);
@@
   ${\color{blue}%
-    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
+    \{ \#_1(a) + 2k \}}$
   if (x < N) {
     ${\color{blue}%
-      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
+      \{ \#_1(a) + 1 + 2(k-1) + 1 \}}$
     a[x]=1;
@@
     \{ \#_1(a) + 2k \}}$
-} ${\color{blue} \{ \#_1(a) \}}$
-
-\end{lstlisting}
-\end{minipage}
-
+} ${\color{blue} \{ 0 \}}$
+
+\end{lstlisting}
 \caption{Exemple derivation using amortized reasoning, we noted
   $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
-  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
+  tick metric that assigns cost $n$ to \lstinline{tick(n)} statments and
   0 to other operations.
@@
 \label{fig:xmplinc}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}%
-  \{ (Z = \log_2(h-l)) + Z \} }$
-bsearch(x,l,h) {
-  if (h-l > 1) {
-    ${\color{blue}%
-      \{ (h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
-    m = h + (h-l)/2;
-    ${\color{blue}%
-      \{ (m = \frac{h+l}{2} \land h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
-    if (a[m]>x) h=m; else l=m;
-    ${\color{blue}%
-      \{ (Z - 1 = \log_2(h-l)) + (Z-1) + 1 \} }$
-    l = bsearch(x,l,h); }
-  ${\color{blue}%
-    \{ Z-1 \} }$
-  return l;
-} ${\color{blue} \{ Z \} }$
-\end{lstlisting}
-\caption{Exemple derivation of a stack usage bound for a binary
-  search program, the metric used assigns a cost of 1 to function
-  calls and $-1$ to return statements, we noted $Z$ an arbitrary
-  logical variable.
-  }
 \end{figure}

### 2014-06-21 paper/main.tex
new example for the stack metric
more work on the informal explanation

--- 2014-06-21 paper/main.tex
@@
 \item We define a operational cost semantics for C programs that
-  accounts a constant cost for each statement which is parametric in a
-  cost metric.  The semantics models cost of terminating and
+  accounts a constant cost for each statement.  The cost is parametric
+  in a cost metric and the semantics models cost of terminating and
   non-terminating executions.  A specific \emph{tick} function can be
-  inserted in the source code to further customize the cost
-  model.
+  inserted in the source code to further customize the cost model.
+
 \item We describe an automatic static analysis technique for deriving
   bounds for the cost that is defined by the operational semantics.
-  It is based on amortized resource analysis for functional programs
-  [XXX] and derives bounds that are non-negative linear combinations
-  of sizes $|[x,y]|$ of intervals between program variables $x,y$.
+  It is based on amortized resource analysis for functional
+  programs~\cite{Jost03,HoffmannAH12} and derives bounds that are
+  non-negative linear combinations of sizes $|[x,y]|$ of intervals
+  between program variables $x,y$.
 \item We show with a publically available prototype implementation and
@@
 Following the development steps of automatic amortized analysis for
-functional programs [XXX], we deliberately restrict our self to linear
-bounds in the automatic analysis (there a no restrictions for the
-manually-derived bounds in the logic).  More specifically, bounds have
-the form $\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer
-variables or constants.  The reason is mainly clarity of presentation,
-since an extension to multivariate resource polynomials [XXX] would
-make the inference rules considerably more involved and should
-better be described separately.  However, we developed the linear
-inference system so that the extension to polynomial bounds shall
-work smoothly.
+functional programs~\cite{Jost03,HoffmannH10}, we deliberately
+restrict our self to linear bounds in the automatic analysis (there a
+no restrictions for the manually-derived bounds in the logic).  More
+specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
+where $a$ and $b$ are integer variables or constants.  The reason is
+mainly clarity of presentation, since an extension to multivariate
+resource polynomials~\cite{HoffmannAH11} would make the inference rules considerably
+more involved and should better be described separately.  However, we
+developed the linear inference system so that the extension to
+polynomial bounds shall work smoothly.
 
@@
 Our new quantitative Hoare logic for CompCert Clight is a
-generalization of a program logic that we recently developed to
-formally very stack-space bounds in Coq [PLDI].  In a nutshell, we
-replace the boolean predicates (pre and post conditions) of Hoare
-logic with numeric predicates that map a program state to a
+generalization of a program logic that we developed to formally very
+stack-space bounds for C programs in Coq~\cite{veristack14}.  In a
+nutshell, we replace the boolean predicates (pre and post conditions)
+of Hoare logic with numeric predicates that map a program state to a
 non-negative number or infinity.  The intuitive meaning of a
-quantitative Hoare triple $\htriple{P}{C}{Q}$ that $C$ consumes $P(S)$
-resources when executed in starting state $S$, and that $Q(S')$
-resources are left if $C$ terminates in state $S'$.  Infinity
+quantitative Hoare triple $\htriple{P}{S}{Q}$ that $C$ consumes $P(\state)$
+resources when executed in starting state $\state$, and that $Q(\state')$
+resources are left if $C$ terminates in state $\state'$.  Infinity
 corresponds to \emph{false} in classic Hoare logic.  Our quantitative
@@
 
-The idea that drives the design of our system is amortized analysis
-[XXX].  Assume that a program $p$ executes on a stating state
-$\state_0$ and consumes $n$ resource units of some quantity we are
-interested in.  We denote that by writing $(p, \state_0)
-\Downarrow_n$.  The basic idea of amortized analysis is to define a
-potential function $\Phi$ that maps program states to non-negative
-numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
-program state such that such that $(p, \state) \Downarrow_n$.
+The idea that drives the design of our system is amortized
+analysis~\cite{Tarjan-amort}.  Assume that a program $S$
+executes on a starting state $\state_0$ and consumes $n$ resource units
+of some quantity we are interested in.  We denote that by writing $(S,
+\state_0) \Downarrow_n$.  The basic idea of amortized analysis is to
+define a potential function $\Phi$ that maps program states to
+non-negative numbers and to show that $\Phi(\state) \geq n$ if
+$\state$ is a program state such that such that $(S, \state)
+\Downarrow_n$.
 
 Since programs can be composed we must enrich the evaluation judgement
-$(p, \state) \Downarrow_{n}$ to talk about the state resulting from
-the program execution.  We write $(p, \state) \Downarrow_n \state'$ to
-say that $\state'$ is the program state after the execution of $p$ in
+$(S, \state) \Downarrow_{n}$ to talk about the state resulting from
+the program execution.  We write $(S, \state) \Downarrow_n \state'$ to
+say that $\state'$ is the program state after the execution of $S$ in
 starting state $\sigma$.  We now use two potential functions, one that
@@
 functions must respect the relation $\Phi(\state) \ge n +
-\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(p,
+\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(S,
 \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
-provide enough \emph{potential} for, both, paying for resource cost of
+provide enough \emph{potential} for both, paying for resource cost of
 the computation and paying for the potential $\Phi'(\state')$ on the
-resulting state $\state'$. That way, if $(\state, p_1) \Downarrow_n
-\state'$ and $(\state', p_2) \Downarrow_m \state''$, we get
+resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
+\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get
 $\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
@@
 program.  What we have observed is that, if we define $\htriple{\Phi}
-{p}{\Phi'}$ to mean
+{S}{\Phi'}$ to mean
 $$
@@
 \RuleNolabel
-{\htriple{\Phi}{p_1}{\Phi'} \\ \htriple{\Phi'}{p_2}{\Phi''}}
-{\htriple{\Phi} {p_1; p_2}  {\Phi''}}
+{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
+{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
 $$
@@
 functions that look very similar to Hoare logic or effect systems
-rules.  This rules enable to interactively reason about resource usage
+rules.  These rules enable interactive reasoning about resource usage
 in a flexible and compositional way, which, as a side effect, produces
@@
 a concrete bound for a given resource metric, or a bound that is
-parametric in set of resource metrics.  It is also possible to
+parametric in a set of resource metrics.  It is also possible to
 incorporate boolean preconditions into the bounds to express that the
@@
 $\top$ to $0$.
+
+\begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}
+${\color{blue}%
+  \{ \#_1(a) + 2k \}}$
+while (k > 0) {
+  x=0;
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+  while (x < N && a[x] == 1) {
+    ${\color{blue}%
+      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
+    a[x]=0;
+    ${\color{blue}%
+      \{ \#_1(a) + 2k + 1 \}}$
+    tick(1);
+    ${\color{blue}%
+      \{ \#_1(a) + 2k \}}$
+    x++; }
+  ${\color{blue}%
+    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
+  if (x < N) {
+    ${\color{blue}%
+      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
+    a[x]=1;
+    ${\color{blue}%
+      \{ \#_1(a) + 2(k-1) + 1 \}}$
+    tick(1); }
+  ${\color{blue}%
+    \{ \#_1(a) + 2(k-1) \}}$
+  k--;
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+} ${\color{blue} \{ \#_1(a) \}}$
+
+\end{lstlisting}
+\end{minipage}
+
+\caption{Exemple derivation using amortized reasoning, we noted
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
+  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
+  0 to other operations.
+  }
+\label{fig:xmplinc}
+\end{figure}
 
@@
 \label{fig:xmplmax}
-\end{figure}
-
-\begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}
-${\color{blue}%
-  \{ \#_1(a) + 2k \}}$
-while (k > 0) {
-  ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
-  while (x < N && a[x] == 1) {
-    ${\color{blue}%
-      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
-    a[x]=0;
-    ${\color{blue}%
-      \{ \#_1(a) + 2k + 1 \}}$
-    tick(1);
-    ${\color{blue}%
-      \{ \#_1(a) + 2k \}}$
-    x++; }
-  ${\color{blue}%
-    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
-  if (x < N) {
-    ${\color{blue}%
-      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
-    a[x]=1;
-    ${\color{blue}%
-      \{ \#_1(a) + 2(k-1) + 1 \}}$
-    tick(1); }
-  ${\color{blue}%
-    \{ \#_1(a) + 2(k-1) \}}$
-  k--;
-  ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
-} ${\color{blue} \{ \#_1(a) \}}$
-
-\end{lstlisting}
-\end{minipage}
-
-\caption{Exemple derivation using amortized reasoning, we noted
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
-  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
-  0 to other operations.
-  }
-\label{fig:xmplinc}
 \end{figure}

### 2014-06-21 paper/main.tex
typos

@@
 
-\caption{Exemple derivation using amortized reasoning, we noted
+\caption{Example derivation using amortized reasoning, we noted
   $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
@@
 \end{lstlisting}
-\caption{Exemple derivation of a stack usage bound for a binary
+\caption{Example derivation of a stack usage bound for a binary
   search program, the metric used assigns a cost of 1 to function

### 2014-06-22 paper/main.tex
more work on section 2

@@
 on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
-needed to verify cryptography protocols~\cite{BartheGB09}.
+needed to verify cryptographic protocols~\cite{BartheGB09}.
 % XXX Mention differential privacy here?
@@
 derived linear bounds and translated them to term-rewriting systems by
-hand in order to run them with the tools KoAT and Pubs.  Both tools
-and Rank were not able to find linear bounds for the examples.  LOOPUS
-is able to derive a bound on XXX of the examples.  But in contrast to
-our analysis, the constant factors are not tight.  We also contacted
-the authors of SPEED but were not able to obtain a tool to perform a
+hand in order to run them with the tools KoAT, PUBS, and Rank.  The
+tools were not able to find linear bounds for the examples.  LOOPUS is
+able to derive a bound on XXX of the examples.  But in contrast to our
+analysis, the constant factors are not tight.  We also contacted the
+authors of SPEED but were not able to obtain a tool to perform a
 comparison.
@@
 
-\caption{Example derivation using amortized reasoning, we noted
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
+\caption{Example derivation using amortized reasoning.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
   tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
@@
 It is for instance possible to derive the following quantitative
-triple that holds for all stack metrics $M$.  A stack metric assigns
-cost $0$ to all evaluation steps except function calls.  Before a call
-$f(x)$, we consume $M(f) > 0$ resources and after the call we return
-$-M(f)$ resources.  So $M(f)$ corresponds to the stack-frame size of
-the function $f$.
-$$
-XXX
-$$
-XXX explanation of triple XXX
+triple for a binary-search that holds for all stack metrics $M$.  A
+stack metric assigns cost $0$ to all evaluation steps except function
+calls.  Before a call $f(x)$, we consume $M_f > 0$ resources and
+after the call we return $M_f$ resources that can be used in
+subsequent function calls.  So $M_f$ corresponds to the stack-frame
+size of the function $f$.
+%
+\begin{center}
+${ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
+\code{bsearch(x,l,h)}
+${ \{ Z{\cdot}M_\code{bsearch} \} }$
+
+\end{center}
+%
+In the pre and post condition, the logical variable $Z$ is used to
+relate the size of the input with the potential that is left after the
+function call.  The logical part $Z = \log_2(h-l)$ of the precondition
+is $0$ if $Z = \log_2(\state_0(h)-\state_0(l))$ in the initial state
+$\state_0$ and $\infty$ otherwise.
+
+Assume for the remainder of this section that we are interested in
+bounding the number of \emph{ticks} in a program.  In other words, we
+use a resource metric that assigns cost $n$ to the function call
+\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
+is an integer and a negative number means that $-n$ resources are
+returned.
 
 We say that the analysis is amortized because the use of a potential
-function $\Phi$ gives the flexibility to amortize some operations.
-For example, if program states are simple mappings from variables to
-natural numbers, we have the following judgement (provided that
-initially $\state(\code x) \ge \state(\code y) \ge 0$).
-$$
-\htriple
-{\lambda \state . \, 2 {\cdot} \state (\code x) + 10}
-{\code{ x = x -  y}}
-{\lambda \state . \, 2 {\cdot} \state(\code x) + 2 {\cdot} \state(\code y) + 10}
-$$
-This must be understood as a motion of potential from the
-variable~$\code x$ to the variable~$\code y$. If some resource is
-consumed $\code y$ times afterwards, we can then use $\code y$'s
-potential to pay for it. However, the initial potential function
-$\Phi$ only refers to the variable~$\code x$: the potential function
-allows us to pass on the cost of the subsequent resource consumption
-to~$\code x$.  This flexibility combined with intelligent decisions on
-potential motions allows us to design an automatic analysis to
-discover linear invariants for surprisingly convoluted programs.
+function $\Phi$ enables us to amortize the cost of operations like in
+typical textbook examples that use the potential method of Tarjan's
+amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
+shows a program that repeatably increments a binary counter that is
+implemented with a boolean-valued ($0$ or $1$) array.  The derived
+bound in the precondition states that the number of ticks that are
+executed by the program are bounded by $\#_1(a) + 2k$.  Here, $k$ is
+the number of increments to the counter that are performed by the
+program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
+the number of $1$'s in the array $a$.  We have inserted the statement
+\code{tick(1)} into the program whenever we update the array $a$.
+
+To derive the bound for the program in~\pref{fig:xmplinc} in the
+quantitative Hoare logic, we follow the textbook
+reasoning~\cite{Algorithms}.  When we assign \code{a[x]=0} in the
+inner loop, we know that $a[x]=1$ before the assignment.  As a result,
+we have $\#_1(a) = \#_1(a') + 1$ if $a'$ is the array $a$ after the
+assignment.  So we obtain one resource unit in addition to the
+invariant (i.e., $\Phi = \#_1(a) + 2k + 1$) that we can use to pay for the
+cost of the statement \code{tick(1)}.  To pay for the tick statement
+in the conditional, we use the fact that $k>0$ to write $2k$ as $1 +
+2(k-1) +1$.  Dual to the first assignment, we have $\#_1(a) +1 =
+\#_1(a')$ if $a$ is the array before and $a'$ is the array after the
+assignment \code{a[x]=1}.  After paying one potential unit for the
+statement $\code{tick(1)}$, we are left with the potential $\#_1(a) +
+2(k - 1)$.  But since we increment $k$ at the end of the outer loop,
+we can establish again our loop invariant $\#_1(a) + 2k$.
+
+See Section XXX for more a formal definition of the logic and
+additional example derivation.
+
+% For example we have the following judgement (provided that initially
+% $\state_0(\code x) \ge \state_0(\code y) \ge 0$).
+% $$
+% \htriple
+% {2{\cdot}x + 10}
+% {\code{ x = x -  y}}
+% {2{\cdot}x + 2{\cdot}y + 10}
+% $$
+% This must be understood as a motion of potential from the
+% variable~$\code x$ to the variable~$\code y$. If some resource is
+% consumed $\code y$ times afterwards, we can then use $\code y$'s
+% potential to pay for it. However, the initial potential function
+% $\Phi$ only refers to the variable~$\code x$: the potential function
+% allows us to pass on the cost of the subsequent resource consumption
+% to~$\code x$.  This flexibility combined with intelligent decisions on
+% potential motions allows us to design an automatic analysis to
+% discover linear invariants for surprisingly convoluted programs.
 
@@
 component of $\Phi(\state)$ if $x \in \dom{\state}$.
-
+%
 We then develop an inference rule for each syntactic construct that
 derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
-logic but enables inference by using an LP solver.  This idea is best
-explained by example.  Assume for the remainder of this section that
-we are interested in bounding the number of \emph{ticks} in a program.
-In other words, we use a resource metric that assigns cost $n$ to the
-function call \code{tick(n)} and cost $0$ to all other operations.
-The cost of the following example can then be bounded by $|[x,y]|$.
+logic but enables inference by using an LP solver.
+
+This idea is best explained by example.  If we again use the
+\emph{ticks} metric that assigns cost $n$ to the function call
+\code{tick(n)} and cost $0$ to all other operations then the cost of
+the following example can be bounded by $|[x,y]|$.
 \begin{lstlisting}
@@
 $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
-only do so if we exploit the fact that $x<y$ at the beginning of the
+only do so if we utilize the fact that $x<y$ at the beginning of the
 loop body.  The reasoning then works as follows.  We start with the
@@
 ${\color{blue}%
-  \cdot; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-while (x<y) {
-  ${\color{blue}%
-  x{<}y; \; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  x=x+1; 
-  ${\color{blue}%
-  \dashv x{\leq}y; \, 1 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv x{\leq}y; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
-}
-${\color{blue}%
- \dashv x{\geq}y; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
-\end{lstlisting}
-  \caption{Derivation of a bound on the number of ticks for Example~1.}
+  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+while (y-x>K) {
+  ${\color{blue}%
+  x{<}y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  x=x+K; 
+  ${\color{blue}%
+  \dashv x{\leq}y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  tick(T);
+  ${\color{blue}%
+  \dashv x{\leq}y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+}
+${\color{blue}%
+ \dashv x{\geq}y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+\end{lstlisting}
+  \caption{Automatic derivation of a tight bound on the number of
+    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
+    $T>0$ are not program variables but denote a concrete constants in
+    the program.  The reasoning works uniformly for all such $K$ and
+    $T$.}
   \label{fig:ex1}
@@
 indexed by the variables that are currently in scope.  The logical
-context $\Gamma$ is simply a conjunction of inequalities between.  It
-is used to determine at the assignment if we can extract constant
-potential to pay for future cost.  It is important to note that the
-reasoning the assertions in $\Gamma$ are very basic.  We do not
-perform any fixpoint computations or invariants.  \pref{fig:ex1} shows
-the derivation for Example 1.  Note that the logical assertions in
-$\Gamma$ are only used in the assignment \code{x=x+1}.
-
-It would of course be possible to keep track more sophisticated
-assertions in the logical context $\Gamma$.  However, this simple form
-is sufficient for the examples we considered and we can efficiently
-decide queries such as $\Gamma \implies x<y \; ?$, which we are
-interested in.  Nevertheless, this logical part of our analysis system
-is orthogonal to the quantitative part and can be easily extended if
-necessary.
+context $\Gamma$ is simply a conjunction of inequalities between
+linear combinations variables.  It is used to determine at the
+assignment if we can extract constant potential to pay for future
+cost.  It is important to note that the reasoning about assertions in
+$\Gamma$ are very basic.  We do not perform any fixpoint computations
+and only derive trivial loop invariants about variables that are
+unchanged in the loop body.
 
@@
 while loop to prove the bound.  Of course, such loops with different
-notations can be seamlessly combined in a larger program.
+tick annotations can be seamlessly combined in a larger program.
 \begin{lstlisting}
@@
 %
-To understand how the analysis works, it is instructive to look at
-some more involved examples.  Examples \emph{speed\_1} and
-\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.
-They are both handled by the SPEED tool~\cite{GulwaniMC09} but
-require inference of a \emph{disjunctive invariant}.  In the abstract
+\pref{fig:ex1} shows a derivation of the bound
+$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
+version of Example~1 in which increment $x$ be a constant $K>0$ and
+consume $T>0$ resources in each iteration.  The reasoning is similar
+to the description of Example~1 except that we obtain the potential
+$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
+assertions in $\Gamma$ are only used in rule for the assignment
+\code{x=x+1}.  To the best of our knowledge, no other tool is
+currently capable of deriving a tight bound on the cost of such a
+loop.  For $T=1$ (many systems focus on the number of loop iterations
+without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
+computes the bound XXX, PUBS~\cite{AlbertAGPZ12} computes the bound
+XXX, Rank~\cite{AliasDFG10} computes the bound XXX, and
+LOOPUS~\cite{SinnZV14} computes the bound XXX.
+
+It might be surprising that we only track simple linear relations in
+the logical context $\Gamma$.  It would of course be possible to keep
+track more sophisticated assertions.  However, this simple form is
+sufficient for the examples we considered and we can efficiently
+decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
+assignment rule.  Nevertheless, this logical part of our analysis
+system is orthogonal to the quantitative part and can be easily
+extended if necessary.
+
+To understand the capabilities of the automatic analysis, it is instructive to
+look at some more involved examples.  Examples \emph{speed\_1} and
+\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.  They
+are both handled by the SPEED tool~\cite{GulwaniMC09} but require
+inference of a \emph{disjunctive invariant}.  In the abstract
 interpretation community, these invariants are known to be notoriously
@@
 \pref{fig:ex_rec} contains two mutually-recursive functions with their
-automatically derived bounds.  The function \code{count\_down}
+automatically derived tick bounds.  The function \code{count\_down}
 decrements its first argument $x$ until it reaches the second argument

### 2014-06-23 paper/main.tex
small bits here and there

@@
   \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-while (y-x>K) {
-  ${\color{blue}%
-  x{<}y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+while (x+K<=y) {
+  ${\color{blue}%
+  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
   x=x+K; 
   ${\color{blue}%
-  \dashv x{\leq}y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
   tick(T);
   ${\color{blue}%
-  \dashv x{\leq}y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
-}
-${\color{blue}%
- \dashv x{\geq}y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+}
+${\color{blue}%
+ \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
 \end{lstlisting}
@@
 without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
-computes the bound XXX, PUBS~\cite{AlbertAGPZ12} computes the bound
-XXX, Rank~\cite{AliasDFG10} computes the bound XXX, and
-LOOPUS~\cite{SinnZV14} computes the bound XXX.
+computes the bound $x + y + 10$, PUBS~\cite{AlbertAGPZ12} computes
+the tight bound $0.1(y-x)$, Rank~\cite{AliasDFG10} computes the bound XXX, and
+LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.
 

### 2014-06-23 paper/main.tex
abstract

@@
 \begin{abstract}
+  The goal of quantitative resource analysis is to provide developers
+  with quantitative information about the runtime behavior of software
+  at development time.  Recent years have seen tremendous progress in
+  automatically deriving worst-case resource bounds.  Yet, many
+  challenges in specifying, interactively deriving, formally
+  certifying, and composing resource bounds remain unsolved.
+
+  This paper describes a novel quantitative resource analysis
+  framework that tackles these challenges for C programs.  First, the
+  runtime resource cost of programs is formalized with a parametric
+  cost semantics for CompCert Clight.  Second, a quantitative Hoare
+  logic that can be used for interactive bound development is
+  implement and proved sound with respect to the cost semantics in the
+  Coq Proof Assistant.  Third, an automatic amortized resource
+  analysis for C programs that can compute derivations in the
+  quantitative Hoare logic is developed and implemented.  Both, the
+  quantitative logic and the automatic amortized analysis are
+  naturally compositional and can be combined to semi-automatically
+  derive global resource bounds.
+
+  The practicability of the analysis framework is demonstrated with an
+  experimental evaluation.  The logic is used to derive XXX A comparison
+  with other automatic tools shows that the constant factors in the
+  bounds that are derived by the automatic amortized analysis are
+  often more precise. XXX
 
@@
 assertions in $\Gamma$ are only used in rule for the assignment
-\code{x=x+1}.  To the best of our knowledge, no other tool is
+\code{x=x+K}.  To the best of our knowledge, no other tool is
 currently capable of deriving a tight bound on the cost of such a
@@
 
-To understand the capabilities of the automatic analysis, it is instructive to
-look at some more involved examples.  Examples \emph{speed\_1} and
-\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.  They
-are both handled by the SPEED tool~\cite{GulwaniMC09} but require
-inference of a \emph{disjunctive invariant}.  In the abstract
+To understand the capabilities of the automatic analysis, it is
+instructive to study more involved examples.  \pref{fig:ex_list}
+contains Examples \emph{speed\_1} and \emph{speed\_2} that are taken
+from previous work on bound analysis~\cite{GulwaniMC09}.  The SPEED
+tool~\cite{GulwaniMC09} derives the bounds as our analysis but
+requires inference of a \emph{disjunctive invariant}.  In the abstract
 interpretation community, these invariants are known to be notoriously
@@
 Example \emph{t27} shows how amortization can be used to handle tricky
-nested loops.  The outer loop is iterated $|[n,0]|$ times.  In the
-outer loop, we increment the variable $n$ till $n = 0$.  In each of
-the $|[n,0]|$ iterations, we increment the variable $y$ by $1000$.  We
-then non-deterministically execute an inner loop that decrements $y$
-by $100$ until $y<100$.  The analysis can derive that only the first
-execution of the inner loop depends on the initial value of $y$.  We
-again derive a tight bound.
+nested loops.  The outer loop we increment the variable $n$ till $n =
+0$.  In each of the $|[n,0]|$ iterations, we increment the variable
+$y$ by $1000$.  We then non-deterministically execute an inner loop
+that decrements $y$ by $100$ until $y<100$.  The analysis can derive
+that only the first execution of the inner loop depends on the initial
+value of $y$.  We again derive tight constant factors.
 
@@
 count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis system can compute a tight bound.  On this example, it
-amounts to computing the meeting point of two trains that approach
-each other with different speeds.
+analysis system can compute a tight bound.  On this example, the
+analysis amounts to computing the meeting point of two trains that
+approach each other with different speeds.
 % This is comparable to the
@@
 
-Please note that we can only show a very small selections of the
-programs that we can handle automatically here.  In \pref{app:cat} you
-find a list of more than $30$ classes of challenging programs that we
-can automatically analyze.  We hope that this gives you an impression
-of the capabilities of the analysis but we also point out that this
-list is far from being exhaustive.
+We only show a very small selections of the programs that we can
+handle automatically here.  In \pref{app:cat} you find a list of more
+than $30$ classes of challenging programs that we can automatically
+analyze.  We hope that this gives you an impression of the
+capabilities of the analysis but we also point out that this list is
+far from being exhaustive.
 
@@
 
-\begin{figure}[ht!]
+\begin{figure*}[t]
 \begin{mathpar}
-
+%
 \Rule{Q:Skip}
 {}
-{ \Gamma; Q + M_k \vdash \code{skip} \dashv Q; \Gamma }
-
+{ \Gamma; B; R; Q \vdash \code{skip} \dashv Q; \Gamma }
+%
+\and \Rule{Q:Break}
+{}
+{ \Gamma; B; R; B + M_b \vdash \code{break} \dashv Q'; \Gamma' }
+%
+\and \Rule{Q:Return}
+{}
+{ \Gamma; B; R; R \, x + M_r \vdash \code{return}~x \dashv Q'; \Gamma' }
+%
 \and \Rule{Q:Assert}
 {}
-{ \Gamma; Q + M_a \vdash \code{assert}(C) \dashv Q; \Gamma, C }
-
-\and \Rule{Q:While}
-{ \Gamma, C ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
-{ \Gamma; Q \vdash \code{while}(C)~S \dashv Q - M_w^3; \Gamma, \neg C }
-
+{ \Gamma; B; R; Q + M_a \vdash \code{assert}~e \dashv Q; \Gamma, e }
+%
+\and \Rule{Q:Loop}
+{ \Gamma; Q'; R; Q \vdash S \dashv M_l + Q; \Gamma }
+{ \Gamma; B; R; Q \vdash \code{loop}~S \dashv Q'; \code{inv}_S~\Gamma }
+%
 \and \Rule{Q:Set}
-{ \forall u.\, q_{yu} = q'_{xu} + q'_{yu}
-\\ q'_{xy} \in \mathbb Q_0^+
-}
-{ \Gamma[x/y]; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
-
+{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
+\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ \Gamma[x/y]; B; R; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
+%
+\and \Rule{Q:IncrPos}
+{ \Gamma \models 0 \le y
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
+\\\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx}
+}
+{ \textstyle
+  \Gamma[x/x + y]; B; R; Q + M_s \vdash x \gets x + y \dashv Q'; \Gamma
+}
+%
+\and \Rule{Q:IncrNeg}
+{ \Gamma \models 0 \le y
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  \Gamma[x/x - y]; B; R; Q + M_s \vdash x \gets x - y \dashv Q'; \Gamma
+}
+%
 \and \Rule{Q:Incr}
-{ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
-\\ \textstyle q'_y = q_y
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{xv}
+{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
 }
 { \textstyle
-  \Gamma[x/x \pm y]; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
-}
-
+  \Gamma[x/x \pm y]; B; R; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
+}
+%
 \and \Rule{Q:If}
-{ \Gamma, C; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma, \neg C; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
-}
-{ \Gamma; Q \vdash \code{if} (C)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
-
+{ \Gamma, e; B; R; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
+\\ \Gamma, \neg e; B; R; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
+}
+{ \Gamma; B; R; Q \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
+%
 \and\Rule{Q:Seq}
-{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
-}
-{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
-
+{  \Gamma; B; R; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
+\\ \Gamma'; B; R; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
+}
+{ \Gamma; B; R; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
+%
+\and\Rule{Q:Call}
+{ (\vec y, s, \Gamma_f, Q_f, Q_f', \Gamma_f') \in \Delta(f)
+\\ P = Q_f \Uparrow_{[\vec y = \vec x]}
+\\ P' = Q'_f \Uparrow_{[r = s]}
+\\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
+\\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
+\\ c \in \mathbb Q_0^+
+}
+{ \Gamma_f[\vec y / \vec x] \land \Gamma_\ell; B; R; Q + c
+  \vdash r \gets f(\vec x) \dashv
+  Q' + c; \Gamma_f'[s/r] \land \Gamma_\ell
+}
+%
 \and \Rule{Q:Weak}
@@
 \\ Q_1 \succeq_\Gamma Q_2
-\\ \Gamma_2; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
+\\ \Gamma_2; B; R; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
 \\ Q'_2 \succeq_\Gamma Q'_1
@@
 }
-{ \Gamma_1; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
-
+{ \Gamma_1; B; R; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
+%
 \and \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |x - y| \}
-\\ \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |x - y| \le u_{xy} \}
-%\\ q'_0 = q_0 + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i
+{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
+\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
 \\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
@@
 { Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
-
+%
 \end{mathpar}
@@
 \label{fig:proof}
-\end{figure}
+\end{figure*}
 

### 2014-06-23 paper/main.tex
new analysis rules
first try at a call rule

@@
 \begin{abstract}
-  The goal of quantitative resource analysis is to provide developers
-  with quantitative information about the runtime behavior of software
-  at development time.  Recent years have seen tremendous progress in
-  automatically deriving worst-case resource bounds.  Yet, many
-  challenges in specifying, interactively deriving, formally
-  certifying, and composing resource bounds remain unsolved.
-
-  This paper describes a novel quantitative resource analysis
-  framework that tackles these challenges for C programs.  First, the
-  runtime resource cost of programs is formalized with a parametric
-  cost semantics for CompCert Clight.  Second, a quantitative Hoare
-  logic that can be used for interactive bound development is
-  implement and proved sound with respect to the cost semantics in the
-  Coq Proof Assistant.  Third, an automatic amortized resource
-  analysis for C programs that can compute derivations in the
-  quantitative Hoare logic is developed and implemented.  Both, the
-  quantitative logic and the automatic amortized analysis are
-  naturally compositional and can be combined to semi-automatically
-  derive global resource bounds.
-
-  The practicability of the analysis framework is demonstrated with an
-  experimental evaluation.  The logic is used to derive XXX A comparison
-  with other automatic tools shows that the constant factors in the
-  bounds that are derived by the automatic amortized analysis are
-  often more precise. XXX
 
@@
 assertions in $\Gamma$ are only used in rule for the assignment
-\code{x=x+K}.  To the best of our knowledge, no other tool is
+\code{x=x+1}.  To the best of our knowledge, no other tool is
 currently capable of deriving a tight bound on the cost of such a
@@
 
-To understand the capabilities of the automatic analysis, it is
-instructive to study more involved examples.  \pref{fig:ex_list}
-contains Examples \emph{speed\_1} and \emph{speed\_2} that are taken
-from previous work on bound analysis~\cite{GulwaniMC09}.  The SPEED
-tool~\cite{GulwaniMC09} derives the bounds as our analysis but
-requires inference of a \emph{disjunctive invariant}.  In the abstract
+To understand the capabilities of the automatic analysis, it is instructive to
+look at some more involved examples.  Examples \emph{speed\_1} and
+\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.  They
+are both handled by the SPEED tool~\cite{GulwaniMC09} but require
+inference of a \emph{disjunctive invariant}.  In the abstract
 interpretation community, these invariants are known to be notoriously
@@
 Example \emph{t27} shows how amortization can be used to handle tricky
-nested loops.  The outer loop we increment the variable $n$ till $n =
-0$.  In each of the $|[n,0]|$ iterations, we increment the variable
-$y$ by $1000$.  We then non-deterministically execute an inner loop
-that decrements $y$ by $100$ until $y<100$.  The analysis can derive
-that only the first execution of the inner loop depends on the initial
-value of $y$.  We again derive tight constant factors.
+nested loops.  The outer loop is iterated $|[n,0]|$ times.  In the
+outer loop, we increment the variable $n$ till $n = 0$.  In each of
+the $|[n,0]|$ iterations, we increment the variable $y$ by $1000$.  We
+then non-deterministically execute an inner loop that decrements $y$
+by $100$ until $y<100$.  The analysis can derive that only the first
+execution of the inner loop depends on the initial value of $y$.  We
+again derive a tight bound.
 
@@
 count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis system can compute a tight bound.  On this example, the
-analysis amounts to computing the meeting point of two trains that
-approach each other with different speeds.
+analysis system can compute a tight bound.  On this example, it
+amounts to computing the meeting point of two trains that approach
+each other with different speeds.
 % This is comparable to the
@@
 
-We only show a very small selections of the programs that we can
-handle automatically here.  In \pref{app:cat} you find a list of more
-than $30$ classes of challenging programs that we can automatically
-analyze.  We hope that this gives you an impression of the
-capabilities of the analysis but we also point out that this list is
-far from being exhaustive.
+Please note that we can only show a very small selections of the
+programs that we can handle automatically here.  In \pref{app:cat} you
+find a list of more than $30$ classes of challenging programs that we
+can automatically analyze.  We hope that this gives you an impression
+of the capabilities of the analysis but we also point out that this
+list is far from being exhaustive.
 

### 2014-06-24 paper/main.tex
syntax and semantics

--- 2014-06-23 paper/main.tex
@@
   Coq Proof Assistant.  Third, an automatic amortized resource
-  analysis for C programs that can compute derivations in the
+  analysis for C programs that computes derivations in the
   quantitative Hoare logic is developed and implemented.  Both, the
@@
 10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
-1043$).
+1043$).  The use of interval sizes in potential functions is an
+important novelty.
 %
@@
 
-\section{Syntax}
-
-For the sake of simplicity, our quantitative program logic
-applies to the Clight language.  This language is the most
-abstract intermediate language used by CompCert.  It is very
-similar to vanilla C but the expressions are free of
-side-effects and there is only one looping construct.
+\section{Syntax and Semantics}
+
+We implemented our cost semantics and the quantitative Hoare logic in
+Coq for \emph{CompCert Clight}.  Clight is the most abstract
+intermediate language used by CompCert.  Mainly, it is a subset of C
+in which loops can only be exited with a \code{break} statement and
+expressions are free of side effects.
+
+\paragraph{Syntax}
+
+In this article, we describe our system for a subset of Clight to
+discuss general ideas instead of technical details.  This subset is
+given by the following grammar.
 %
@@
 \mid S;S
+\mid \code{tick}(n)
 \end{align*}
 %
-We added the builtin primitive $\code{assert}$ that checks
-if a condition is true before continuing the program
-execution.  This will be used to explicit invariants to the
-automatic analysis later.
-
-The expressions are left abstract in this presentation.  To
-present both the semantics and the program logic, we simply
-need an evaluation function that maps any $e \in E$ to
-a value $\llbracket e \rrbracket_\state$ using $\state$ as current
-state.
-
-\section{Semantics}
-
-The operational semantics uses a mutable state $\state$
-composed of two maps from variable names to integers. The
-first map $\theta$ represents the local variable of a
-function and the second map $\gamma$ gives values to global
-variables of the program.
+Expressions are $E$ are left abstract in our presentation.  For our
+analysis framework, it is only important that they are side-effect
+free.
+%
+The most notable difference to full Clight is that we can only assign
+to variables and thus do not consider operations that update the heap.
+Moreover, function arguments and return values are assumed to be
+variables.  In the implementation we can deal with heap updates and
+general function calls and returns.  However, we have not implemented
+function pointers, goto statements, continue statements, and switch
+statements.
+
+We include the built-in primitive $\code{assert}~E$ that terminates
+the program if the argument $E$ evaluates to false and has no effect
+otherwise.  This is useful to express assumptions on the inputs of a
+program for the automatic analysis.  We also add the built-in function
+$\code{tick}(n)$ that can be called with a constant integer $n$ as a
+flexible way to model resource consumption or release (if $n$ is
+negative).
+
+\paragraph{Semantics}
+
+CompCert Clight's operational semantics is based on small-step
+transitions and continuations.  Expressions---which do not have side
+effects---are evaluated in a big-step fashion.  Here, we describe a
+simplified version of Clight's semantics that is sufficient for the
+subset that we are describing.
+
+A program state $\state = (\theta, \gamma)$ composed of two maps from
+variable names to integers. The first map $\theta : \text{Locals} \to
+\Z$ assigns integers to local variables of a function and the second
+map $\gamma : \text{Globals} \to \Z$ gives values to global variables
+of the program.
+%
+The evaluation function \evalE{\cdot}{} that maps an expression $e \in
+E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.
 
 For simplicity, we assume that local variables and global
-variables are always disjoint sets, this simplifies both
-the semantics and the description of our analysis.  The
-mutable state $\state$ can be accessed and modified, the
-access is written functionally and is defined by
+variables are always disjoint sets. % This simplifies both
+% the semantics and the description of our analysis.  
+The mutable state $\state$ can be accessed and modified, the access is
+written functionally and is defined by
 $$
@@
 $$
-We note $\state[x \mapsto v]$ the state that results from
-updating $\state$ to make $x$ map to $v$.  Similarly to
-the state access, the formal definition of this operation
-is a simple case disjunction.  Thanks to the assumption we
-made above we can be sure that the different cases never
-overlap.
-
-To formally describe the cost of operation of a given
-program, our semantics tracks resource consumption while
-executing.  The semantics is oblivious to the resource of
-interest for the user of our system.  We achieve this
-independence by parameterizing by a resource
-\emph{metric}, it is a tuple of nine rational numbers and
-one map from expressions to rational numbers
-$$
-(
+For a program state $\state$, we write $\state[x \mapsto v]$ for the
+program state that maps $x$ to $v$ and behaves as $\state$ for all
+other variables, regardless whether $x \in \dom \theta$ or $x \in \dom
+\gamma$.  Similarly, we just write $\state(x)$ to obtain the value of
+$x$ in program state $\state$.
+
+The small-step semantics is standard, except that it tracks the
+resource consumption of a program.  The semantics is oblivious to the
+resource of interest for the user of our system.  We achieve this
+independence by parameterizing evaluations with a resource
+\emph{metric} $M$; a tuple of nine rational numbers and one map $M_e :
+\text{Expr} \to \Q$ from expressions to rational numbers.
+$$
+M = (
 M_a,
@@
 M_q,
-M_e
-).
+M_e(E)
+)
 $$
 Each of these rational numbers indicates the amount of resource
-consumed by one operation.  If one of these numbers is negative,
-it means that resources are released by the program.
-
-For technical reasons in our soundness proof and to be in line
-with CompCert's Clight, our language's semantics is defined
-in a small-step fashion.  The \pref{fig:opsem} contains the full
-list of reduction rules.  These rules provide a rewrite
-system for program configurations of the form
-$(\state, S, \cont, \cost)$, where $\state$ is the program state,
-$S$ is the statement being executed, $\cont$ is a continuation
-that describes what remains to be done once the execution of $S$
-is over, and $\cost$ is the non-negative quantity of resources
-available for further execution.  All rules that can decrease
-$\cost$ have the implicit side condition that the resource
-quantity available after the step is non-negative.  Using this
-resource counter we can define the resource consumption of a
-program as the minimal resource counter $\cost_*$ that lets the
-program run to termination without errors.
-
-The small-step semantics uses continuations to represent the
-context of execution of a given statement.
-A continuation can be a sequenced computation, it can also result
-from entering a loop or a function, or it can simply be the
-empty continuation (used to start a program).
+consumed by a corresponding operation.  If one of these numbers is
+negative, it means that resources are released by the program.  The
+metric that we use in the implementation can also depend on other
+information that is statically available such as the name of the called
+function or the number of arguments.
+
+\pref{fig:opsem} contains the reduction rules of the semantics.
+The rules define a rewrite system for program configurations of the
+form $(\state, S, \cont, \cost)$, where $\state$ is the program state,
+$S$ is the statement being executed, $\cont$ is a continuation that
+describes what remains to be done after the execution of $S$, and
+$\cost$ is the non-negative number of resources available for further
+execution.  All rules that can decrease $\cost$ have the implicit side
+condition that the resource quantity available after the step is
+non-negative.
+
+A continuation $\cont$ represents the context of the execution of a
+given statement.  A continuation can be the empty continuation
+$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
+\cont$, a loop continuation $\Kloop S \cont$, or the continuation
+$\Kcall r \theta \cont$ of a function call.
 $$
@@
 $$
+%
+During the execution we assume a fixed function context $\fenv$ that
+maps function names to a list of variables and the statement that
+defines the function body.  It is used in the rule \textsc{S:Call}.
+
+The intuitive meaning of an evaluation judgement $(\state, S, \cont,
+\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the following.
+If the statement $S$ is executed in program state $\state$, with
+continuation $\cont$, and with $\cost$ resources available
+then---after a finite number of steps---the evaluation will reach the
+new machine state $(\state', S', \cont')$ and there are $\cost'$
+resources available.  The resource consumption of the evaluation up to
+this point is then $\cost - \cost'$.  If this difference is negative
+then resource became available during the execution.
+
+\pref{lem:sem1} summarizes the main properties of the resource
+counter of the semantics.  It can be proved by induction on the number
+of steps.
+\begin{lemma}
+\label{lem:sem1}
+  \begin{enumerate}
+  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
+    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
+    \smallstep{M}^* (\state', S', \cont', \cost' {+} n)$.
+  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
+    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^* (\state', S',
+    \cont', d')$ then $\cost - \cost' = d - d'$.
+  \end{enumerate}
+\end{lemma}
+%
+If we assume that we have a \emph{positive resource metric}, that is,
+a metric that assigns a positive cost to each evaluation step then we
+can link bounds on the resource consumption under the metric to
+termination.  This is formalized by \pref{lem:sem_term}.
+%
+\begin{lemma}
+ \label{lem:sem_term}
+ Let $M$ be a positive metric and assume that there exists $b \in
+ \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
+ (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.  Then
+ $(\state, S, \cont, d) \smallstep{M}^* (\state'', \code{skip}, \Kstop,
+ d')$ for some $d$, $d'$ and $\state''$.
+\end{lemma}
 
@@
 { (\state, \code{assert}~e, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{skip}, \cont, \cost {-} M_a)
@@
 { (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{break}, \cont, \cost)
@@
 {}
-{ (\state, \code{break}, \Kloop \cont, \cost)
-\smallstep
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
 (\state, \code{skip}, \cont, \cost {-} M_b)
@@
 { (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{return}~x, \cont, \cost)
@@
 {}
-{ (\state, \code{return}~x, \Kloop \cont, \cost)
-\smallstep
+{ (\state, \code{return}~x, \Kloop S \cont, \cost)
+\smallstep{M}
 (\state, \code{return}~x, \cont, \cost)
@@
 { (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state', \code{skip}, \cont, \cost {-} M_r)
@@
 { (\state, x \gets e, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
@@
 \Rule{S:Call}
-{ \delta f = (\vec x, S_f) \\
+{ \fenv f = (\vec x, S_f) \\
   \state = (\theta, \gamma) \\
@@
 { (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
@@
 { (\state, \code{loop}~S, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, S, \Kloop S \cont, \cost)
@@
 { (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{loop}~S, \cont, \cost {-} M_l)
@@
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, S_1, \cont, \cost {-} M_c^1)
@@
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, S_2, \cont, \cost {-} M_c^2)
@@
 { (\state, S_1; S_2, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, S_1, \Kseq {S_2} \cont, \cost)
@@
 { (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, S, \cont, \cost {-} M_q)

### 2014-06-24 paper/main.tex
some fixes and started with the description of the logic

@@
 take non-negative number or $\infty$ (infinity) as values.  Infinity,
-plays the same role as $\bot$ in Hoare logic.  Boolean assertions can
-be embedded into potential functions by mapping $\bot$ to $\infty$ and
-$\top$ to $0$.
+plays the same role as $\mathit{false}$ in Hoare logic.  Boolean assertions can
+be embedded into potential functions by mapping $\mathit{false}$ to $\infty$ and
+$\mathit{true}$ to $0$.
 
@@
 independence by parameterizing evaluations with a resource
-\emph{metric} $M$; a tuple of nine rational numbers and one map $M_e :
-\text{Expr} \to \Q$ from expressions to rational numbers.
+\emph{metric} $M$; a tuple of nine rational numbers, one map $M_e :
+\text{Expr} \to \Q$ from expressions to rational numbers, and one map
+$M_t : \Z \to \Q$ from integers to rational numbers.
 $$
@@
 M_q,
-M_e(E)
+M_e(E),
+M_t(n)
 )
@@
 function or the number of arguments.
+
+\begin{figure}[t!]
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~x, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~x, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \fenv f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep{M}
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma }
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \cont, \cost {-} M_c^1)
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma }
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_2, \cont, \cost {-} M_c^2)
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_q)
+}
+%
+\and
+\Rule{S:Tick}
+{ }
+{ (\state, \code{tick}(n), \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
+}
+
+\end{mathpar}
+\caption{Operational semantics of statements}
+\label{fig:opsem}
+\end{figure}
 
@@
 
-\begin{figure}[t!]
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:RetSeq}
-{}
-{ (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-\and
-%
-\and
-\Rule{S:RetLoop}
-{}
-{ (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetCall}
-{ \state = (\_, \gamma) \\
-  \state' = (\theta, \gamma)[r \mapsto \state(x)]
-}
-{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_r)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Call}
-{ \fenv f = (\vec x, S_f) \\
-  \state = (\theta, \gamma) \\
-  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
-}
-{ (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep{M}
-(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma }
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \cont, \cost {-} M_c^1)
-}
-%
-\and
-\Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma }
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_2, \cont, \cost {-} M_c^2)
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_q)
-}
-%
-
-\end{mathpar}
-\caption{Operational semantics of statements}
-\label{fig:opsem}
-\end{figure}
-
-\section{Proof System}
+\section{Quantitative Hoare Logic}
+
+In this section we describe a simplified version of the quantitative
+Hoare logic that we use in Coq to interactively prove resource bounds.
+
+We generalize classic Hoare logic to express not only classical
+boolean-valued assertions but also assertions that talk about the
+future resource usage.  Instead of the usual assertions $P : \State
+\to \mathit{bool}$ of Hoare logic we use assertions
+$$
+P : \State \to \Q \cup \{ \infty \} \; .
+$$
+This can be understood as a refinement of boolean assertions where
+$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Q$.
+We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
+$\lambda \state . \, \infty$.  To use Coq's support for propositional
+reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
+in the implementation.  For a given $\state \in \State$, such an
+assertion can be seen as a set $B \subseteq \Q$ of valid bounds.
+However, we find the presentation in this article easier to read.
+
+% The continuation-based semantics of a Clight requires that we
+% distinguish pre- and postconditions in the logic to account for
+% different possible ways to exit a block of code.  This is approach is
+% standard in Hoare logics and followed for instance in Appel's
+% separation logic for Clight~\cite{AppelLogic}.  Our postconditions
+% $$
+% Q = (Q^s, Q^b, Q^r) : \Assn \times \Assn \times (\Val \to \Assn)
+% $$
+% provide one assertion $Q^s$ for the case in which the block is exited by
+% fall through, one assertion $Q^b$ if the block is exited by a \code{break}, and a
+% function $Q^r$ from values to assertions in case the block is exited by a
+% \code{return}.  The function argument in the last case represents the return
+% value and the intended meaning is that the resulting assertion is
+% guaranteed to hold for every return value.
+
+% Since we have to deal with recursive functions, we also need a
+% function context
+% $$\Gamma {:} \funs {\to} ((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$ 
+% that maps function names to their specifications, that is, pre- and
+% postconditions.  The precondition depends on the value that is passed
+% to the function by the caller and the memory.  The postcondition
+% depends on the return value and the memory.  For simplicity, we assume
+% that a function has only one argument in this article.  In the Coq
+% implementation, an arbitrary number of function arguments is allowed.
+
+% In summary, a quantitative Hoare triple has the form
+% $$
+% \Gamma \vdash \htriple{P}{S}{Q}
+% $$
+% where $\Gamma$ is a function context, $P : \Assn$ is a precondition,
+% $Q : \Assn \times \Assn \times (\Val \to \Assn)$ is a postcondition,
+% and $S$ is a statement.
+
+% Intuitively, an assertion can be seen as a \emph{potential function}
+% that maps a program state to a non-negative potential.  The potential
+% of the precondition $P$ must be sufficient to cover the cost of the
+% execution of the statement $S$ and the potential $Q$ after the execution
+% of $S$ (as in amortized resource analysis~\cite{HoffmannAH12}).
 
@@
 \end{figure*}
+
+\section{Automatic Amortized Analysis}
 

### 2014-06-24 paper/main.tex
fixed some typos typos

@@
 %
-In this paper, we present a novel cost analysis system for C programs
+In this paper, we present a novel resource analysis framework for C programs
 that is based on a solid semantic foundation. It works automatically,
@@
 parameters of the program.  To the best of our knowledge, our system
-is the only one that can deal with resource like memory that may
-become available during the execution.
+is the only one that can automatically handle resources like memory
+that may become available during the execution.
 

### 2014-06-24 paper/main.tex
minor
Q:Call is wrong

@@
   This paper describes a novel quantitative resource analysis
-  framework that tackles these challenges for C programs.  First, the
-  runtime resource cost of programs is formalized with a parametric
-  cost semantics for CompCert Clight.  Second, a quantitative Hoare
-  logic that can be used for interactive bound development is
-  implement and proved sound with respect to the cost semantics in the
-  Coq Proof Assistant.  Third, an automatic amortized resource
-  analysis for C programs that computes derivations in the
-  quantitative Hoare logic is developed and implemented.  Both, the
-  quantitative logic and the automatic amortized analysis are
-  naturally compositional and can be combined to semi-automatically
-  derive global resource bounds.
+  framework that tackles these challenges for C programs.  The
+  analysis framework consists of three parts.  First, the runtime
+  resource cost of programs is formalized with a parametric cost
+  semantics for CompCert Clight.  Second, a quantitative Hoare logic
+  that can be used for interactive bound development is implement and
+  proved sound with respect to the cost semantics in the Coq Proof
+  Assistant.  Third, an automatic amortized resource analysis for C
+  programs that computes derivations in the quantitative Hoare logic
+  is developed and implemented.  Both, the quantitative logic and the
+  automatic amortized analysis are naturally compositional and can be
+  combined to semi-automatically derive global resource bounds.
 
@@
 non-negative number or infinity.  The intuitive meaning of a
-quantitative Hoare triple $\htriple{P}{S}{Q}$ that $C$ consumes $P(\state)$
-resources when executed in starting state $\state$, and that $Q(\state')$
-resources are left if $C$ terminates in state $\state'$.  Infinity
-corresponds to \emph{false} in classic Hoare logic.  Our quantitative
-logic is proved sound with respect to a cost semantic for Clight that
-accounts a constant cost for each evaluation step.  This constant cost
-is parametric in a resource metric that can be defined by the user.
-Our shallow embedding in Coq enables the derivation of both, bounds
-that are parametric in the cost metric or bounds for one specific
-metric.
+quantitative Hoare triple $\htriple{P}{S}{Q}$ is that $S$ consumes
+less then $P(\state)$ resources when executed in starting state
+$\state$, and that $Q(\state')$ resources are left if $S$ terminates
+in state $\state'$.  Infinity corresponds to \emph{false} in classic
+Hoare logic.  Our quantitative logic is proved sound with respect to a
+cost semantic for Clight that accounts a constant cost for each
+evaluation step.  This constant cost is parametric in a resource
+metric that can be defined by the user.  Our shallow embedding in Coq
+enables the derivation of both, bounds that are parametric in the cost
+metric or bounds for one specific metric.
 
@@
 map $\gamma : \text{Globals} \to \Z$ gives values to global variables
-of the program.
+of the program.  In this article, we assume that all values are
+integers but in the implementation we support all data types of
+Clight.
 %
@@
 
+Due to \code{break} and \code{return} statements of Clight, there are
+different possible ways to exit a block of code.  We also have to keep
+track of the resource specifications of functions.  To account for
+this in the logic, our quantitative Hoare triples have the form
+$$
+\Delta; B; R \vdash
+\htriple
+  { Q }
+  { S }
+  { Q' } \; .
+$$
+The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
+two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
+Hoare logic and the intuitive meaning is as follows.  If $S$ is
+executed with starting state $\state$, the empty continuation
+$\Kstop$, and at least $P(\state)$ resources available then the
+evaluation does not run out of resources and there are at least
+$Q(\state')$ resources left if the evaluation terminates in state
+$\state'$.  However, the assertion $B : \Assn$ provides the
+postcondition for the case in which the code block $S$ is exited by a
+\code{break} statement.  So if the execution is terminated in state
+$\state'$ with a \code{break} then $B(\state')$ resources are
+available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
+case in which the code block $S$ is exited by a \code{return x}
+statement.  The integer argument of $R$ is the return value.
+Finally, the function context 
+$$\Delta {:} \text{Funs} \to \Z^*
+\\
+((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$
+
 % The continuation-based semantics of a Clight requires that we
-% distinguish pre- and postconditions in the logic to account for
-% different possible ways to exit a block of code.  This is approach is
+% distinguish pre- and postconditions in the   This is approach is
 % standard in Hoare logics and followed for instance in Appel's
@@
 \and\Rule{Q:Call}
-{ (\vec y, s, \Gamma_f, Q_f, Q_f', \Gamma_f') \in \Delta(f)
-\\ P = Q_f \Uparrow_{[\vec y = \vec x]}
-\\ P' = Q'_f \Uparrow_{[r = s]}
-\\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
-\\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
-\\ c \in \mathbb Q_0^+
+{ (\vec y, s, \Gamma_f, P, P', \Gamma_f') \in \Delta(f)
+\\ \ell = \text{Locals}(Q)
+\\ R = P[\vec y / \vec x]
+\\ \forall i \in \dom{R} . \, q_i = r_i
+\\ R' = P'[s/r]
+\\ \forall i \in \dom{R'} .\, q'_i = r'_i
+\\ \forall i \not\in \dom{R'} .\, q'_i = 0
+\\ \forall i \in \ell . \, q_i = q_i'
+% \\ P' = Q'_f \Uparrow_{[r = s]}
+% \\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
+% \\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
+% \\ c \in \mathbb Q_0^+
 }

### 2014-06-24 paper/main.tex
fix logic, add explanations

@@
 \begin{center}
-${ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
+${ \{ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
 \code{bsearch(x,l,h)}
@@
 $Q(\state')$ resources left if the evaluation terminates in state
-$\state'$.  However, the assertion $B : \Assn$ provides the
+$\state'$.  The assertion $B : \Assn$ provides the
 postcondition for the case in which the code block $S$ is exited by a
@@
 available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
-case in which the code block $S$ is exited by a \code{return x}
+case in which the code block $S$ is exited by a $\code{return}~x$
 statement.  The integer argument of $R$ is the return value.
-Finally, the function context 
-$$\Delta {:} \text{Funs} \to \Z^*
-\\
-((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$
+Finally, the function context of judgements that we write $\Delta$ is
+a mapping from function names to specification of the form
+$$
+  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
+$$
+The assertion $P_f \, z \, \vec v$ is the precondition of the
+function $f$ and the assertion $Q_f \, z \, v$ is its postcondition.
+They are both parameterized by an arbitary logical variable $z$
+(which can be a tuple) that helps relating them.  The precondition
+is allowed to depend on $\vec v$, the value of all the arguments at
+the function invocation.  Similarly, the postcondition depends on
+the return value $v$ of the function.
 
@@
 \htriple
-  { M_s + Q }{ \code{skip} }{ Q }
+  { Q }{ \code{skip} }{ Q }
 }
@@
 \htriple
-  { M_r + R\,x }{ \code{return}~x }{ Q }
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
 }
@@
 {
-  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
   P \models P_f \, y \, (\state(\vec x)) \land A \\
@@
   { r \gets f(\vec x) }
-  { Q }
+  { Q - M_r }
 }
@@
 \Rule{L:Extend}
-{ \forall f\,P_f\,Q_f.\,
-  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
+{ \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+\\
+  \forall f\,P_f\,Q_f.\,
+  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
   \rightarrow
@@
     \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-\\
-  \Delta \cup \Delta'; B; R \vdash
-  \htriple{ P }{ S }{ Q }
 }

### 2014-06-24 paper/main.tex
Q:Call
some changes to the description of the logic

@@
 $$
-The assertion $P_f \, z \, \vec v$ is the precondition of the
-function $f$ and the assertion $Q_f \, z \, v$ is its postcondition.
-They are both parameterized by an arbitary logical variable $z$
-(which can be a tuple) that helps relating them.  The precondition
-is allowed to depend on $\vec v$, the value of all the arguments at
-the function invocation.  Similarly, the postcondition depends on
-the return value $v$ of the function.
-
-% The continuation-based semantics of a Clight requires that we
-% distinguish pre- and postconditions in the   This is approach is
-% standard in Hoare logics and followed for instance in Appel's
-% separation logic for Clight~\cite{AppelLogic}.  Our postconditions
-% $$
-% Q = (Q^s, Q^b, Q^r) : \Assn \times \Assn \times (\Val \to \Assn)
-% $$
-% provide one assertion $Q^s$ for the case in which the block is exited by
-% fall through, one assertion $Q^b$ if the block is exited by a \code{break}, and a
-% function $Q^r$ from values to assertions in case the block is exited by a
-% \code{return}.  The function argument in the last case represents the return
-% value and the intended meaning is that the resulting assertion is
-% guaranteed to hold for every return value.
-
-% Since we have to deal with recursive functions, we also need a
-% function context
-% $$\Gamma {:} \funs {\to} ((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$ 
-% that maps function names to their specifications, that is, pre- and
-% postconditions.  The precondition depends on the value that is passed
-% to the function by the caller and the memory.  The postcondition
-% depends on the return value and the memory.  For simplicity, we assume
-% that a function has only one argument in this article.  In the Coq
-% implementation, an arbitrary number of function arguments is allowed.
-
-% In summary, a quantitative Hoare triple has the form
-% $$
-% \Gamma \vdash \htriple{P}{S}{Q}
-% $$
-% where $\Gamma$ is a function context, $P : \Assn$ is a precondition,
-% $Q : \Assn \times \Assn \times (\Val \to \Assn)$ is a postcondition,
-% and $S$ is a statement.
-
-% Intuitively, an assertion can be seen as a \emph{potential function}
-% that maps a program state to a non-negative potential.  The potential
-% of the precondition $P$ must be sufficient to cover the cost of the
-% execution of the statement $S$ and the potential $Q$ after the execution
-% of $S$ (as in amortized resource analysis~\cite{HoffmannAH12}).
+The assertion $P_f \, z \, \vec v$ is the precondition of the function
+$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
+both parameterized by an arbitary logical variable $z$ (which can be a
+tuple) that relates the function arguments with the return value.  The
+precondition also depends on $\vec v$, the value of all the arguments
+at the function invocation.  Similarly, the postcondition depends on
+the return value $v$ of the function.  The use of logical variables to
+express relations between different states of an execution is a
+standard technique of Hoare logic.
 
@@
 \and\Rule{Q:Call}
-{ (\vec y, s, \Gamma_f, P, P', \Gamma_f') \in \Delta(f)
-\\ \ell = \text{Locals}(Q)
-\\ R = P[\vec y / \vec x]
-\\ \forall i \in \dom{R} . \, q_i = r_i
-\\ R' = P'[s/r]
-\\ \forall i \in \dom{R'} .\, q'_i = r'_i
-\\ \forall i \not\in \dom{R'} .\, q'_i = 0
-\\ \forall i \in \ell . \, q_i = q_i'
-% \\ P' = Q'_f \Uparrow_{[r = s]}
-% \\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
-% \\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
-% \\ c \in \mathbb Q_0^+
-}
-{ \Gamma_f[\vec y / \vec x] \land \Gamma_\ell; B; R; Q + c
+{ (\vec y, s, \Gamma_f, Q_f, Q_f', \Gamma_f') \in \Delta(f)
+\\ \text{Loc} = \text{Locals}(Q)
+\\ \forall i \neq j . \, x_i \neq x_j
+\\ c \in \mathbb Q_0^+
+\\\\ Q \ge P + S
+\\ P' + S \ge Q'
+\\ R = Q_f[\vec y / \vec x]
+\\ R' = Q'_f[s/r]
+\\ \forall i \in \dom{R} . \, p_i = r_i
+\\ \forall i \in \dom{R'} .\, p'_i = r'_i 
+\\ \forall i \not\in \dom{R'} .\, p'_i = 0
+\\ \forall i \not\in \text{Loc}
+  .\, s_i = 0
+}
+{ \Gamma_f[\vec y / \vec x] \land \Gamma_{\text{Loc}}; B; R; Q + c
   \vdash r \gets f(\vec x) \dashv
-  Q' + c; \Gamma_f'[s/r] \land \Gamma_\ell
+  Q' + c; \Gamma_f'[s/r] \land \Gamma_{\text{Loc}}
 }

### 2014-06-24 paper/main.tex
update to the semantics

@@
 
-For simplicity, we assume that local variables and global
-variables are always disjoint sets. % This simplifies both
-% the semantics and the description of our analysis.  
-The mutable state $\state$ can be accessed and modified, the access is
-written functionally and is defined by
+For simplicity, we assume that local variables and global variables
+are always disjoint.  We just write $\state(x)$ to obtain the value of
+$x$ in program state $\state$.  Such a lookup is defined as
 $$
@@
 $$
-For a program state $\state$, we write $\state[x \mapsto v]$ for the
-program state that maps $x$ to $v$ and behaves as $\state$ for all
-other variables, regardless whether $x \in \dom \theta$ or $x \in \dom
-\gamma$.  Similarly, we just write $\state(x)$ to obtain the value of
-$x$ in program state $\state$.
+For a program state $\state$, we similarly write $\state[x \mapsto v]$
+for the program state that maps $x$ to $v$ and behaves as $\state$ for
+all other variables, regardless whether $x \in \dom \theta$ or $x \in
+\dom \gamma$.
 
@@
 Each of these rational numbers indicates the amount of resource
-consumed by a corresponding operation.  If one of these numbers is
-negative, it means that resources are released by the program.  The
-metric that we use in the implementation can also depend on other
-information that is statically available such as the name of the called
-function or the number of arguments.
+consumed by a corresponding operation.  If the assigned resource cost
+is negative then it means that resources are released.  The metric
+that we use in the implementation can also depend on other information
+that is statically available such as the name of the called function
+or the number of arguments.
 
@@
 
-\pref{fig:opsem} contains the reduction rules of the semantics.
-The rules define a rewrite system for program configurations of the
-form $(\state, S, \cont, \cost)$, where $\state$ is the program state,
-$S$ is the statement being executed, $\cont$ is a continuation that
+\pref{fig:opsem} contains the reduction rules of the semantics.  The
+rules define a rewrite system for program configurations of the form
+$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
+is the statement being executed, $\cont$ is a continuation that
 describes what remains to be done after the execution of $S$, and
-$\cost$ is the non-negative number of resources available for further
-execution.  All rules that can decrease $\cost$ have the implicit side
-condition that the resource quantity available after the step is
-non-negative.
+$\cost \in \Q$ is the non-negative number of resources available for
+further execution.  All rules that can decrease $\cost$ have the
+implicit side condition that the resource quantity available
+\emph{before} the step is non-negative.  This means that we allow
+$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
+transition relation $\smallstep{M}^*$ to indicate that the execution
+ran out of reassesses.  However, every execution that reaches such a
+state is stuck.
 
@@
 The intuitive meaning of an evaluation judgement $(\state, S, \cont,
-\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the following.
-If the statement $S$ is executed in program state $\state$, with
-continuation $\cont$, and with $\cost$ resources available
-then---after a finite number of steps---the evaluation will reach the
-new machine state $(\state', S', \cont')$ and there are $\cost'$
-resources available.  The resource consumption of the evaluation up to
-this point is then $\cost - \cost'$.  If this difference is negative
-then resource became available during the execution.
+\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
+following.  If the statement $S$ is executed in program state
+$\state$, with continuation $\cont$, and with $\cost$ resources
+available then---after a finite number of steps---the evaluation will
+reach the new machine state $(\state', S', \cont',\cost')$ and there
+are $\cost'$ resources available.  If $\cost' \ge 0$ then the
+execution did not run out of resources and the resource consumption up
+to this point is $\cost - \cost'$.  If this difference is negative
+then resource became available during the execution.  If however
+$\cost' < 0$ then the execution ran out of resources and is stuck.
+The cost of the execution is then $\cost \ge 0$.
 
@@
   \begin{enumerate}
-  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
+  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
     \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
-    \smallstep{M}^* (\state', S', \cont', \cost' {+} n)$.
-  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
-    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^* (\state', S',
+    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
+  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
     \cont', d')$ then $\cost - \cost' = d - d'$.
@@
  \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
- (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.  Then
- $(\state, S, \cont, d) \smallstep{M}^* (\state'', \code{skip}, \Kstop,
- d')$ for some $d$, $d'$ and $\state''$.
+ (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
+ Then there exists $B \in \N$ such that $(\state, S, \cont, d)
+ \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
+ and $\state''$ implies $n \le B$.
 \end{lemma}

### 2014-06-24 paper/main.tex
describe meaning of triples

@@
 \end{figure*}
+
+\paragraph{Semantic meaning of triples.}
+
+We already gave an intuition of the meaning of judgements
+derived in the logic, to make it formal we define the
+\emph{resource safety} $\safe n P S \cont$ of an assertion
+and a program configuration as
+\begin{align*}
+  \forall \state \, \cost \, m \, \cost' .\,&
+  (m \le n \land P(\state) \le \cost \land
+    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
+  &\implies 0 \le \cost'.
+\end{align*}
+This predicate is step indexed by an integer $n$ that will
+be used in the soundness proof for the function call and
+loop cases.  The constraint $0 \le \cost'$ imposed by the
+definition makes sure that the program does not stop on a
+resource error.  However, it does not rule out memory safety
+errors or assertion failures, this is because our logic does
+not prove any safety or correctness theorems but only
+focuses on resource usage.
+
+The safety of a continuation $\cont$ is defined using three
+assertions, one for each of the possible outcome of a program
+statement.  We denote it by $\safeK n B R Q \cont$ and define
+it to be
+\begin{align*}
+& \safe n B {\code{break}} \cont \\
+\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
+\land& \safe n Q {\code{skip}} \cont.
+\end{align*}
+
+We can now define the \emph{semantic validity} of a function-context
+free logic judgement $B; R \vdash \htriple P S Q$ as follows
+\begin{align*}
+\valid n B R P S Q := \\
+\forall m\, \cont .\,& (m \le n \land \safeK m B R Q \cont) \\
+&\implies \safe m P S \cont.
+\end{align*}
+We use this validity definition to give a meaning to function
+specifications $\forall z \, \vec v \, v.(P_f \, z \, \vec v,
+Q_f \, z \, v)$ as follows
+$$
+  \forall z \, \vec v \, n .\,
+  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
+$$
+where $S_f$ is the body of the function $f$.
 

### 2014-06-24 paper/main.tex
small fixes to the call rule
new inference rules

@@
 {}
-{ \Gamma; B; R; Q \vdash \code{skip} \dashv Q; \Gamma }
+{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
 %
@@
 {}
-{ \Gamma; B; R; B + M_b \vdash \code{break} \dashv Q'; \Gamma' }
+{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
 %
 \and \Rule{Q:Return}
-{}
-{ \Gamma; B; R; R \, x + M_r \vdash \code{return}~x \dashv Q'; \Gamma' }
+{ P = Q_R[\Vret/x]  
+\\ \forall i \in \dom{P} . \, p_i = q_i
+}
+{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
 %
@@
 {}
-{ \Gamma; B; R; Q + M_a \vdash \code{assert}~e \dashv Q; \Gamma, e }
+{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
 %
 \and \Rule{Q:Loop}
-{ \Gamma; Q'; R; Q \vdash S \dashv M_l + Q; \Gamma }
-{ \Gamma; B; R; Q \vdash \code{loop}~S \dashv Q'; \code{inv}_S~\Gamma }
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
 %
@@
 }
-{ \Gamma[x/y]; B; R; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
+{ B; R; (\Gamma[x/y], Q {+} M_s) \vdash x \gets y \dashv (\Gamma, Q') }
 %
@@
 { \textstyle
-  \Gamma[x/x + y]; B; R; Q + M_s \vdash x \gets x + y \dashv Q'; \Gamma
-}
-%
-\and \Rule{Q:IncrNeg}
+  B; R; (\Gamma[x/x {+} y], Q {+} M_s) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule[rightskip=1cm]{Q:IncrNeg}
 { \Gamma \models 0 \le y
@@
 { \textstyle
-  \Gamma[x/x - y]; B; R; Q + M_s \vdash x \gets x - y \dashv Q'; \Gamma
+  B; R; (\Gamma[x/x {-} y],Q {+} M_s) \vdash x \gets x - y \dashv (\Gamma, Q')
 }
@@
 { \textstyle
-  \Gamma[x/x \pm y]; B; R; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
+  B; R; (\Gamma[x/x \pm y], Q {+} M_s) \vdash x \gets x \pm y \dashv (\Gamma, Q')
 }
@@
 \and \Rule{Q:If}
-{ \Gamma, e; B; R; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma, \neg e; B; R; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
-}
-{ \Gamma; B; R; Q \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
+{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
+\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+}
+{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
 %
 \and\Rule{Q:Seq}
-{  \Gamma; B; R; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma'; B; R; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
-}
-{ \Gamma; B; R; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
+{  B; R; (\Gamma, Q {-} M_s^1) \vdash S_1 \dashv (\Gamma', Q')
+\\ B; R; (\Gamma', Q' {-} M_s^2) \vdash S_2 \dashv (\Gamma'',Q'')
+}
+{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
 %
 \and\Rule{Q:Call}
-{ (\vec y, s, \Gamma_f, Q_f, Q_f', \Gamma_f') \in \Delta(f)
-\\ \text{Loc} = \text{Locals}(Q)
-\\ \forall i \neq j . \, x_i \neq x_j
-\\ c \in \mathbb Q_0^+
-\\\\ Q \ge P + S
-\\ P' + S \ge Q'
-\\ R = Q_f[\vec y / \vec x]
-\\ R' = Q'_f[s/r]
-\\ \forall i \in \dom{R} . \, p_i = r_i
-\\ \forall i \in \dom{R'} .\, p'_i = r'_i 
-\\ \forall i \not\in \dom{R'} .\, p'_i = 0
-\\ \forall i \not\in \text{Loc}
+{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
+\and \text{Loc} = \text{Locals}(Q)
+\and \forall i \neq j . \, x_i \neq x_j
+\and c \in \mathbb Q_0^+
+\\ Q = P + S
+\and Q' = P' + S
+\and U = Q_f[\Vargs / \vec x]
+\and U' = Q'_f[\Vret/r]
+\\ \forall i \in \dom{U} . \, p_i = u_i
+\and \forall i \in \dom{U'} .\, p'_i = u'_i 
+\and \forall i \not\in \dom{U'} .\, p'_i = 0
+\and \forall i \not\in \text{Loc}
   .\, s_i = 0
 }
-{ \Gamma_f[\vec y / \vec x] \land \Gamma_{\text{Loc}}; B; R; Q + c
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
   \vdash r \gets f(\vec x) \dashv
-  Q' + c; \Gamma_f'[s/r] \land \Gamma_{\text{Loc}}
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
 }
@@
 { \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_\Gamma Q_2
-\\ \Gamma_2; B; R; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
-\\ Q'_2 \succeq_\Gamma Q'_1
+\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
 \\ \Gamma'_2 \models \Gamma'_1
 }
-{ \Gamma_1; B; R; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
+{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
 %
@@
 %
+\and \Rule{Q:Abst}
+{ \fenv f = (\vec y, S_f)
+\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
+}
+{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+%
 \end{mathpar}
-\caption{Abstract proof rules of the quantitative analysis}
+\caption{Inference rules of the quantitative analysis}
 \label{fig:proof}

### 2014-06-24 paper/main.tex
more on the soundness

@@
 
-We can now define the \emph{semantic validity} of a function-context
-free logic judgement $B; R \vdash \htriple P S Q$ as follows
+We can now define the \emph{semantic validity} of a logic judgement
+$B; R \vdash \htriple P S Q$ without function context as follows
 \begin{align*}
-\valid n B R P S Q := \\
-\forall m\, \cont .\,& (m \le n \land \safeK m B R Q \cont) \\
-&\implies \safe m P S \cont.
+&\valid n B R P S Q := \\
+&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
+&\qquad \implies \safe m {P+x} S \cont.
 \end{align*}
-We use this validity definition to give a meaning to function
-specifications $\forall z \, \vec v \, v.(P_f \, z \, \vec v,
-Q_f \, z \, v)$ as follows
-$$
-  \forall z \, \vec v \, n .\,
+Note how the validity of a triple embeds the frame rule of
+our logic, this refinement is necessary to have a stronger
+induction hypothesis available during the proof.
+
+Using this semantic validity  we can define the validity of
+a function context $\Delta$ that we write $\validC n \Delta$ as
+\begin{align*}
+  &\forall f .\, \Delta(f) =
+    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
+    \implies \\
+  &\qquad \forall z \, \vec v .\,
   \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
-$$
+\end{align*}
 where $S_f$ is the body of the function $f$.
+
+\begin{theorem}[Soundness of the logic]
+  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
+  $$
+    \forall n .\, \validC n \Delta
+      \implies \forall n .\, \valid n B R P S Q.
+  $$
+\end{theorem}
 

### 2014-06-25 paper/main.tex
description for the rules

@@
 %
-While there has been lot of progress in static quantitative analysis,
+While there has been much progress in static quantitative analysis,
 Knuth correctly points out in a recent interview~\cite{KnuthInter}
@@
 \and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q {+} M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+%
 \Rule{L:Weaken}
@@
 
-\paragraph{Semantic meaning of triples.}
+\paragraph{Rules of the Quantitative Logic}
+
+\pref{fig:logic} shows the inference rules of the quantitative logic.
+The rules are slightly simplified in comparison to the implemented
+rules in Coq.  The main difference is that the presented version do
+not formalize an auxiliary state that is needed to relate pre and
+postconditions (see the following paragraph for more details).
+
+In the rule {\sc L:Skip}, we do not have to account for any resource
+consumption.  As a result, the precondition $Q$ can be any (potential)
+function and we only have to make sure that we do not end up with more
+potential.  Since the execution of \code{skip} leaves the program
+state unchanged, we can simply use the precondition as postcondition.
+The potential functions $B$ for the \code{break} and $R$ for the
+\code{return} part of the postcondition are not reachable and can
+therefore be arbitrary.
+
+The rule {\sc L:Assert} is strange XXX.
+
+The rules {\sc L:Break} and {\sc L:Return} are similar to the rule
+{\sc L:Skip}.  The only difference is that we replace requirement of
+the \code{skip} part of the postcondition with analogue requirements
+on the \code{break} and \code{return} part, respectively.  The
+aforementioned rules not trigger any stack-space consumption nor
+release of stack-space.  This might seem strange in the \code{return}
+case but will become clear in the description of the {\sc L:Call} rule
+later on.
+
+The rules {\sc L:AssignL} and {\sc L:AssignC} for local and global
+assignment follow the traditional backward style of Hoare logic. It is
+ensured that the precondition takes the same value on the initial
+state as the \code{skip} part of the postcondition on the state
+modified by assignment.
+
+Despite its seemingly simplicity, the {\sc L:Seq} rule must not be
+overlooked to understand how the quantitative Hoare logic works.  We
+have to define it in such a way that it accounts for early
+exits in statements. For instance, if $S_1$ contains a \code{break}
+statement then $S_2$ will never be executed so we must ensure in the
+\code{break} part of $S_1$'s postcondition that the \code{break} part
+of $S_1;S_2$ holds. For the same reason, the \code{return} part of
+$S_1$'s postcondition is special.
+
+The {\sc L:Loop} rule uses the same principles as the {\sc Seq} rule
+to tweak the final postcondition. In the case of {\sc L:Loop}, we
+simply ensure that the \code{break} part of the inner statement
+becomes the \code{skip} part of the overall statement. We use $\bot$
+as the \code{break} part of the $\code{loop}\, S$ statement since its
+operational semantics prevent it from terminating differently than by
+a \code{skip} or a \code{return}.
+
+The {\sc L:Call} rule accounts for the actual stack-space usage of
+programs. It enforces that enough stack space is available to call the
+function $f$ by adding $M(f)$ to the precondition and the
+postcondition. The pre- and postconditions are taken from the context
+$\Gamma$. This context is extended using the {\sc L:Abstract} rule
+described below. The assertions in the context are parametric with
+respect to both the function argument value and the return value. This
+allows to specify a bound for a function whose recursion depth depends
+on an input parameter.  The argument parameter is instantiated by the
+call rule using the result of the evaluation of the argument
+expression in the current state. Note the symmetry of the rule {\sc
+  L:Call}: $M(f)$ is added on both sides to account for the stack
+space that becomes available after the call.  This justifies that the
+{\sc L:Return} rule does not account for stack-space release.
+
+Finally, we describe the rules which are not syntax directed. The rule
+{\sc L:Abstract} allows to make a proof on any statement with an extra
+hypothesis in the derivation context provided that we have a proof
+that this hypothesis is true. We can see as the quantitative
+interpretation of the usual logical Modus Ponens.
+
+There are two weakening rules available in the quantitative Hoare
+logic.  The framing rule {\sc L:Frame} is designed to weaken a
+statement by stating that if $S$ needs $P$ bytes to run and leaves $Q$
+bytes free at its end, then it can very well run with $P + c$ bytes
+and return $Q + c$ bytes. It comes very handy when we want to prove
+tight bounds using the $\max$ function as demonstrated
+in~\pref{fig:example_derivation}. The consequence {\sc L:Conseq} rule
+is directly imported from classical Hoare logics except that instead
+of using the logical implication $\Rightarrow$ we use the quantitative
+$\ge$. This rule indeed weakens the statement since it requires more
+resource to run the statement and yields less than what has been
+proved to be available after its termination.
+
+\paragraph{Auxiliary State and Stack Framing}
+
+% As mentioned, our account of the logic in this article is slightly
+% simplified compared to the Coq development to improve readability.
+% % However, the described logic is very close to the implemented version.
+
+% The main difference between the implemented logic and the logic
+% described here is that the latter does not have an
+% \emph{auxiliary state}.
+% %
+% Auxiliary state is a classic extension of Hoare logic (see for
+% example~\cite{Nipkow02}).
+%  It is used to share information
+% between the pre- and postcondition of a triple.  In a logic without
+% auxiliary state (or similar techniques) it is not possible to relate
+% program states before and after a statement.  For example, you cannot
+% specify that the function {\tt\ int\ twice\ ()\ $\{$i = i+i;$\}$\ }
+% doubles the value of the variable $i$.  With an auxiliary variable $Z$
+% it is possible specify this fact in Hoare logic using the triple $\htriple{i =
+%   Z}{\code{twice()}}{ i = 2 \cdot Z}$.
+
+% In the Coq implementation,
+% assertions have the type $\State \times \mathit{Aux} \to \N \cup
+% \{\infty\}$.  The auxiliary state is an arbitrary type in Coq and can
+% be instantiated by the user.  Most of the rules of the logic remain
+% unchanged in the presence of auxiliary state.  The
+% only exception is the consequence rule \textsc{L:Conseq} that reads as
+% follows.
+% {  \footnotesize
+% \begin{mathpar}
+%   \RuleNolabel
+%  {  \htriple{P'}{S}{R}
+%  \\  
+%  {\!\!\!\! \begin{array}{r}
+% \forall \sigma \, a. P(\sigma,a) {\in} \N \! \implies  \! \exists a'.  
+%      P(\sigma,a) {\geq} P'(\sigma,a') \hspace{3.3em} \\ \land (\forall \sigma'\, i.  R^i(\sigma',a') {\geq} Q^i(\sigma',a))
+%  \end{array}}\!\!\!\!
+%  }
+%  { \htriple{P}{S}{Q}
+%  }
+% \end{mathpar}}
+% This is the quantitative version of a consequence rule that has been
+% introduced in the context of Hoare logic and is attributed to Martin
+% Hofmann~\cite{Nipkow02}.  Its typical use case is when we apply the rule {\sc L:Call}
+% to a recursive call.  In this case, the Hoare triple for the function
+% call is proved by an assumption from the derivation context with
+% a slightly different auxiliary state.  In the example derivation
+% in~\pref{fig:bsearch_deriv} this different state is $Z - 1$.  Adapting
+% the derivation hypothesis to prove the recursive call is enabled in
+% our logic by the extended consequence rule introduced above.
+
+% \paragraph{Stack Framing}
+
+% A minor difference in the function application rule is that we only
+% present the rule for function calls with a single argument and without
+% \emph{framing of stack assertions}.  The latter is necessary in the
+% code of a caller to carry over information on the local environment
+% from the precondition of the function call to the postcondition of the
+% function call.  This is a general problem in Hoare logic and stack
+% framing is a well known solution.  For instance, we would like to prove
+% something like $\htriple{y}{ f(x) }{y}$ where $y$ is a local variable.
+% This is not possible with the rule \textsc{L:Call} but it is with the more
+% general rule in the Coq implementation~\iffullversion{\cite{coqfiles}}{\cite{anoncoq}}.
+
+\paragraph{Soundness of Quantitative Hoare Triples.}
 
@@
 }
+%
+\Rule{Q:Tick}
+{}
+{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
 %

### 2014-06-25 paper/main.tex
rule tuning
explanation of logic rules

@@
 M_c^2,
-M_q,
+M_s,
 M_e(E),
@@
 \smallstep{M}
-(\state, S, \cont, \cost {-} M_q)
+(\state, S, \cont, \cost {-} M_s)
 }
@@
 \htriple
-  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
   { \code{assert}~e }
@@
 \htriple
-  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
   { S_1 }{ Q }
@@
 \htriple
-  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
   { S_2 }{ Q }
@@
 \and
-%
 \Rule{L:Weaken}
@@
 
-The rule {\sc L:Assert} is strange XXX.
-
-The rules {\sc L:Break} and {\sc L:Return} are similar to the rule
-{\sc L:Skip}.  The only difference is that we replace requirement of
-the \code{skip} part of the postcondition with analogue requirements
-on the \code{break} and \code{return} part, respectively.  The
-aforementioned rules not trigger any stack-space consumption nor
-release of stack-space.  This might seem strange in the \code{return}
-case but will become clear in the description of the {\sc L:Call} rule
-later on.
-
-The rules {\sc L:AssignL} and {\sc L:AssignC} for local and global
-assignment follow the traditional backward style of Hoare logic. It is
-ensured that the precondition takes the same value on the initial
-state as the \code{skip} part of the postcondition on the state
-modified by assignment.
-
-Despite its seemingly simplicity, the {\sc L:Seq} rule must not be
-overlooked to understand how the quantitative Hoare logic works.  We
-have to define it in such a way that it accounts for early
-exits in statements. For instance, if $S_1$ contains a \code{break}
-statement then $S_2$ will never be executed so we must ensure in the
-\code{break} part of $S_1$'s postcondition that the \code{break} part
-of $S_1;S_2$ holds. For the same reason, the \code{return} part of
-$S_1$'s postcondition is special.
-
-The {\sc L:Loop} rule uses the same principles as the {\sc Seq} rule
-to tweak the final postcondition. In the case of {\sc L:Loop}, we
-simply ensure that the \code{break} part of the inner statement
-becomes the \code{skip} part of the overall statement. We use $\bot$
-as the \code{break} part of the $\code{loop}\, S$ statement since its
-operational semantics prevent it from terminating differently than by
-a \code{skip} or a \code{return}.
-
-The {\sc L:Call} rule accounts for the actual stack-space usage of
+In the rule {\sc L:Assert}, we use the notation
+$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
+require potential $Q + M_a$ in the precondition if $e$ evaluates to
+$\mathit{true}$ in the current program state.  If $e$ evaluates to
+$\mathit{false}$ then the potential in the precondition can be
+arbitrary since the program is terminated.
+
+In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
+be arbitrary since it is unreachable.  Instead, we have to justify the
+potential functions $B$ and $R$ that hold after a \code{return} and a
+\code{break}, respectively.  In {\sc L:Break}, we require to have
+potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
+cost of \code{break} and $B$ to pay for the potential after the
+\code{break}.  In ${\sc L:Return}$ we only require to have potential
+$R$ in the precondition to pay for the potential after the
+\code{return}.  The reason is that we found it to be more convenient
+to account for the execution cost $M_r$ of the return in rule {\sc
+  L:Call} for function calls.
+
+The rule {\sc L:Update} is the standard assignment rule of Hoare
+logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
+the precondition we ensure that $Q$ evaluates to the same number as in
+the postcondition.  We also require that have we the constant
+potential $M_u$ available in the precondition to pay for the cost of
+the update.
+
+Despite its seemingly simplicity, the {\sc L:Seq} rule is important to
+understand how the quantitative Hoare logic works.  We have to define
+it in such a way that it accounts for early exits in statements. For
+instance, if $S_1$ contains a \code{break} statement then $S_2$ will
+never be executed so we must ensure in the \code{break} part $B$ of
+$S_1$'s judgement that the \code{break} part $B$ of of $S_1;S_2$
+holds. The same is true for the return part $R$ of the judgements for
+$S_1$ and $S_2$.  The interaction between the actual pre- and
+postconditions is analogous to standard Hoare logic.  In the
+postcondition of $S_1$ we account for the cost $M_s$ of the execution
+of the sequence.
+
+In the rule {\sc L:Loop}, we ensure that the \code{break} part of the
+loop body $S$ becomes the postcondition of the loop statement. We use
+an arbitrary $B$ as the \code{break} part of the judgement for
+$\code{loop}\, S$ since its operational semantics prevents it from
+terminating differently than by a \code{skip} or a \code{return}.  The
+precondition $I$ of the loop is the loop invariant.  That is why we
+require to have potential $I$ available in the precondition of the
+loop body $S$.  In the post condition of $S$, the potential must be
+sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
+iteration.
+
+The {\sc L:If} is similar to the rule for the conditional in classic
+Hoare logic.  In the preconditions of the judgments for the two
+branches $S_1$ and $S_2$ we lift the boolean assertions
+$\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
+quantitative assertions.  In the precondition of the rule {\sc L:Tick}
+we account for the execution cost $M_t(n)$ of the \code{tick}
+statement that depends on the integer $n$.
+
+In the rule {\sc L:Call}, accounts for the cost usage of
 programs. It enforces that enough stack space is available to call the
@@
 
-\paragraph{Auxiliary State and Stack Framing}
+\paragraph{Auxiliary State and Shallow Embedding}
 
@@
 %
-\and \Rule{Q:Set}
+\and \Rule{Q:Update}
 { q'_{xy}, q'_{yx} \in \mathbb Q_0^+
@@
 }
-{ B; R; (\Gamma[x/y], Q {+} M_s) \vdash x \gets y \dashv (\Gamma, Q') }
-%
-\and \Rule{Q:IncrPos}
+{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
+%
+\and \Rule[leftskip=1.5cm]{Q:IncP}
 { \Gamma \models 0 \le y
@@
       - \sum_{v \not\in \mathcal U} q_{vx}
+\\ M = M_u + M_e(x {+} y)
 }
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_s) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule[rightskip=1cm]{Q:IncrNeg}
+  B; R; (\Gamma[x/x {+} y], Q {+} M) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule[rightskip=1.5cm]{Q:IncM}
 { \Gamma \models 0 \le y
@@
       - \sum_{v \not\in \mathcal U} q_{xv}
+\\ M = M_u + M_e(x {-} y)
 }
 { \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M_s) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:Incr}
+  B; R; (\Gamma[x/x {-} y],Q {+} M) \vdash x \gets x - y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:Inc}
 { \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
 \\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+\\\\ M = M_u + M_e(x {\pm} y)
 }
 { \textstyle
-  B; R; (\Gamma[x/x \pm y], Q {+} M_s) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
 }
@@
 { B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
-\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
 }
@@
 \and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q {-} M_s^1) \vdash S_1 \dashv (\Gamma', Q')
-\\ B; R; (\Gamma', Q' {-} M_s^2) \vdash S_2 \dashv (\Gamma'',Q'')
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
 }
 { B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+%
+\and \Rule{Q:Weak}
+{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\\\ \Gamma_1 \models \Gamma_2
+\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
+\\ \Gamma'_2 \models \Gamma'_1
+}
+{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
 %
@@
 }
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
   \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
-}
-%
-\Rule{Q:Tick}
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
+}
+%
+\and \Rule{Q:Tick}
 {}
 { B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
-%
-\and \Rule{Q:Weak}
-{ \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-\\ \Gamma'_2 \models \Gamma'_1
-}
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
 %
@@
 %
-\and \Rule{Q:Abst}
+\and \Rule{Extend}
 { \fenv f = (\vec y, S_f)

### 2014-06-25 paper/main.tex
more layout
ramblings on logical variables

@@
 The rules are slightly simplified in comparison to the implemented
-rules in Coq.  The main difference is that the presented version do
-not formalize an auxiliary state that is needed to relate pre and
-postconditions (see the following paragraph for more details).
+rules in Coq.  The main difference is that the presented version does
+not formalize the heap operations.
 
@@
 $\mathit{false}$ then the potential in the precondition can be
-arbitrary since the program is terminated.
+arbitrary since the program will be terminated.
 
@@
 cost of \code{break} and $B$ to pay for the potential after the
-\code{break}.  In ${\sc L:Return}$ we only require to have potential
+\code{break}.  In {\sc L:Return} we only require to have potential
 $R$ in the precondition to pay for the potential after the
@@
 is directly imported from classical Hoare logics except that instead
-of using the logical implication $\Rightarrow$ we use the quantitative
+of using the logical implication $\implies$ we use the quantitative
 $\ge$. This rule indeed weakens the statement since it requires more
@@
 
-\paragraph{Auxiliary State and Shallow Embedding}
+\paragraph{Logical Variables and Shallow Embedding}
+
+It is often necessary, when specifying a program, to relate certain
+invariants in the pre- and postcondition.  A standard solution of
+Hoare logics to this problem is \emph{logical variables} (reference XXX).  These
+additional variables also mentioned in the literature as auxiliary
+state are constant across the derivation.  For example, if we note
+$Z$ such a logical variable, we can specify the function
+$\code{double}()$ which doubles the global variable $x$ as
+$$
+  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
+$$
+
+When formalizing Hoare logics in a proof assistant one can either
+fully specify the syntax and semantics of assertions and hence get a
+deep embedding, or use the assertions of the host theory to get a
+shallow embedding.  For simplicity, we used the latter approach in our
+development.  This choice makes it possible to have logical variables
+almost for free: we can simply use the variable and binding mechanisms
+of Coq, our host theory.  When an logical variable is needed we
+introduce it using a universal quantifier of Coq before deriving triples
+in the logic, so the Coq theorem for the above example would look like
+%
+\begin{lstlisting}
+Theorem dbl_triple: forall Z,
+  htriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
+\end{lstlisting}
+
+However, this trick alone is often not sufficient when working with a
+recursive function $f$.  In that case we apply the {\sc L:Extend} rule
+of the logic to assume the specification of $f$ then proceed to prove
+the function body with this induction hypothesis.  But in this process
+it can be the case that the induction hypothesis needs to be used with
+a different value of a logical variable.  To cope with this problem,
+assumptions in the function context $\Delta$ are universally quantified
+over logical variables and the {\sc L:Extend} rule once again leverages
+the host proof assistant to require that the triple on $f$'s body is
+proved for every possible logical variable $y$.
 
@@
 %
-\and \Rule{Q:Loop}
+\and \Rule{Q:Tick}
+{}
+{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
+%
+\\ \Rule{Q:Loop}
 { (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
@@
 %
-\and \Rule[leftskip=1.5cm]{Q:IncP}
+\and \Rule{Q:IncP}
 { \Gamma \models 0 \le y
 \\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
-\\\\ \textstyle q'_{0y} = q_{0y}
+\\ \textstyle q'_{0y} = q_{0y}
       + \sum_{u \in \mathcal U} q_{xu}
       - \sum_{v \not\in \mathcal U} q_{vx}
-\\ M = M_u + M_e(x {+} y)
 }
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule[rightskip=1.5cm]{Q:IncM}
+  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:IncM}
 { \Gamma \models 0 \le y
 \\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\\\ \textstyle q'_{y0} = q_{y0}
+\\ \textstyle q'_{y0} = q_{y0}
       + \sum_{u \in \mathcal U} q_{ux}
       - \sum_{v \not\in \mathcal U} q_{xv}
-\\ M = M_u + M_e(x {-} y)
 }
 { \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M) \vdash x \gets x - y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
 }
@@
 }
-%
-\and \Rule{Q:Tick}
-{}
-{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
 %

### 2014-06-25 paper/main.tex
more rule explanations

@@
 standard technique of Hoare logic.
+%
+To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
+the local variables on the call stack, that is, $\forall z \, \vec v
+\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
+(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.
+
+For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
+program state $\state$ $P(\state) \ge Q(\state)$.
 
@@
 
-In the rule {\sc L:Call}, accounts for the cost usage of
-programs. It enforces that enough stack space is available to call the
-function $f$ by adding $M(f)$ to the precondition and the
-postcondition. The pre- and postconditions are taken from the context
-$\Gamma$. This context is extended using the {\sc L:Abstract} rule
-described below. The assertions in the context are parametric with
-respect to both the function argument value and the return value. This
-allows to specify a bound for a function whose recursion depth depends
-on an input parameter.  The argument parameter is instantiated by the
-call rule using the result of the evaluation of the argument
-expression in the current state. Note the symmetry of the rule {\sc
-  L:Call}: $M(f)$ is added on both sides to account for the stack
-space that becomes available after the call.  This justifies that the
-{\sc L:Return} rule does not account for stack-space release.
-
-Finally, we describe the rules which are not syntax directed. The rule
-{\sc L:Abstract} allows to make a proof on any statement with an extra
-hypothesis in the derivation context provided that we have a proof
-that this hypothesis is true. We can see as the quantitative
-interpretation of the usual logical Modus Ponens.
-
-There are two weakening rules available in the quantitative Hoare
-logic.  The framing rule {\sc L:Frame} is designed to weaken a
-statement by stating that if $S$ needs $P$ bytes to run and leaves $Q$
-bytes free at its end, then it can very well run with $P + c$ bytes
-and return $Q + c$ bytes. It comes very handy when we want to prove
-tight bounds using the $\max$ function as demonstrated
-in~\pref{fig:example_derivation}. The consequence {\sc L:Conseq} rule
-is directly imported from classical Hoare logics except that instead
-of using the logical implication $\implies$ we use the quantitative
-$\ge$. This rule indeed weakens the statement since it requires more
-resource to run the statement and yields less than what has been
-proved to be available after its termination.
+In the rule {\sc L:Call}, accounts for the execution cost of both
+$M_f$ for function calls and $M_r$ for \code{return} statements.  This
+justifies that the {\sc L:Return} rule does not account for resource
+consumption.  The pre- and postconditions $P_f$ and $Q_f$ are taken
+from the function context $\Delta$.  The assertions in the context are
+parametric with respect to both the function argument value and the
+return value. This allows to specify a bound for a function whose
+resource consumption depends on an input parameter.  The argument
+parameter is instantiated by the call rule using the result of the
+evaluation of the argument variables in the current state.  Recall
+that we require that $P_f$ and $Q_f$ do not depend on the local
+variables on the call stack, that is, $\forall z \, \vec v \, \theta
+\, \theta' \, \gamma . \, P_f \, z \, \vec v \, (\theta,\gamma) = P_f
+\, z \, \vec v \, (\theta',\gamma)$.  To transfer potential that
+depends on local variables of the callee from the precondition $P$ to
+the postcondition $Q$, we use an assertion $A : \Assn$ that is
+independent of global variables, that is, $\forall \theta \, \gamma \,
+\gamma' . \, A (\theta,\gamma) = A(\theta,\gamma')$.  It is still
+possible to express relations between global and local variables using
+logical variables (see the following paragraph).
+
+Finally, we describe the rules which are not syntax directed.  There
+are two weakening rules available in the quantitative Hoare logic.
+The framing rule {\sc L:Frame} is designed to weaken a statement by
+stating that if $S$ needs $P$ resources to run and leaves $Q$
+resources available after its execution, then it can very well run
+with $P + c$ resources and return $Q + c$ resources.  The consequence
+{\sc L:Conseq} rule is directly imported from classical Hoare logic
+except that instead of using the logical implication $\Rightarrow$ we
+use the quantitative $\models$ that point-wise applies $\ge$.  This
+rule indeed weakens the statement since it requires more resource to
+run the statement and yields less than what has been proved to be
+available after its termination.
 

### 2014-06-25 paper/main.tex
different changes

@@
   programs that computes derivations in the quantitative Hoare logic
-  is developed and implemented.  Both, the quantitative logic and the
+  is developed and implemented.  It is the first automatic amortized
+  analysis that can derive bounds that depend on negative integers and
+  differences between numbers.  Both, the quantitative logic and the
   automatic amortized analysis are naturally compositional and can be
@@
   % numbers into order. 
-  Thanks to the work of Floyd, Hoare, and others,
+  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
   we have formal definitions of semantics, and tools by which we can
@@
 %
-%
-In this paper, we present a novel resource analysis framework for C programs
-that is based on a solid semantic foundation. It works automatically,
-efficiently and precisely for many examples, can be used interactively
-for manual bound derivation if the automation fails, and produces
-verifiable certificates for the bounds.  Our contributions are as
-follows.
+The goal of this work is to develop a resource analysis framework for
+C programs that is based on a solid semantic foundation.  The choice
+of C is primarily motivated by our ongoing work on the formally
+verified hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is
+mainly developed in C and is supposed to provide verified guaranties
+on timing and memory usage.  Moreover, C is also a natural choice
+because it is still the most widely used language for system
+development; in particular in safety-critical embedded and real-time
+systems where resource bound analyses are often required by regulatory
+authorities~\cite{May2013}.
+
+In a recent work, we formally verified a bound on the stack usage of
+CertiKOS
+
+In this article, we present the first resource analysis framework for
+C that makes it possible to combine automatically and interactively
+derived resource bounds.  It works automatically, efficiently and
+precisely for many examples, can be used interactively for manual
+bound derivation if the automation fails, and produces verifiable
+certificates for the bounds.  Our contributions are as follows.
 %
@@
 
-It is often necessary, when specifying a program, to relate certain
-invariants in the pre- and postcondition.  A standard solution of
-Hoare logics to this problem is \emph{logical variables} (reference XXX).  These
-additional variables also mentioned in the literature as auxiliary
-state are constant across the derivation.  For example, if we note
-$Z$ such a logical variable, we can specify the function
-$\code{double}()$ which doubles the global variable $x$ as
+It is often necessary to relate certain invariants in the pre- and
+postcondition when specifying a program.  A standard solution of Hoare
+logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
+additional variables (also called auxiliary state in the literature)
+are constant across the derivation.  For example, if $Z$ is such
+a logical variable, we can specify the function $\code{double}()$
+which doubles the global variable $x$ as
 $$
@@
 $$
-
+%
 When formalizing Hoare logics in a proof assistant one can either
@@
 of Coq, our host theory.  When an logical variable is needed we
-introduce it using a universal quantifier of Coq before deriving triples
-in the logic, so the Coq theorem for the above example would look like
+introduce it using a universal quantifier of Coq before deriving
+triples in the logic, so the Coq theorem for the above example would
+look as follows.
 %
@@
 \end{lstlisting}
-
+%
 However, this trick alone is often not sufficient when working with a
 recursive function $f$.  In that case we apply the {\sc L:Extend} rule
-of the logic to assume the specification of $f$ then proceed to prove
-the function body with this induction hypothesis.  But in this process
-it can be the case that the induction hypothesis needs to be used with
-a different value of a logical variable.  To cope with this problem,
-assumptions in the function context $\Delta$ are universally quantified
-over logical variables and the {\sc L:Extend} rule once again leverages
-the host proof assistant to require that the triple on $f$'s body is
-proved for every possible logical variable $y$.
+of the logic.  First we add a specification $\forall z\,\vec
+v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
+$\Delta$.  Then we proceed to prove the function $S_f$ body with this
+induction hypothesis.  In this process it can be the case that the
+induction hypothesis needs to be used with a different value of a
+logical variable.  To cope with this problem, assumptions in the
+function context $\Delta$ are universally quantified over logical
+variables.  The {\sc L:Extend} rule uses the host proof assistant to
+require that the triple on $f$'s body is proved for every possible
+logical variable $y$.
 
@@
 
-\section{Polynomial Potential}
-
-\paragraph{Index Sets}
-
-Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
-family that maps two-element sets of variables to natural numbers,
-that is,
-$$
-I = (i_{\{x,y\}})_{\{x,y\} \subseteq V} \; .
-$$
-%
-We identify a family $I$ with the set $\{ (\{x,y\},i_{\{x,y\}})
-\mid \{x,y\} \subseteq V\}$.
-
-Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
-instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
-from the context.
-%
-We assume that allways $0 \in V$ and sometimes write $i_x$ instead of $i_{\{x,0\}}$.
-
-The \emph{degree} $\deg(I)$ of an index $I = (i_{\{x,y\}})_{\{x,y\}
-  \subseteq V}$ is defined as
-$$
-\deg(I) = \sum_{\{x,y\} \in V} i_{\{x,y\}} \;.
-$$
-We define $\ind_k(V) = \{ I \mid I \in \ind(V) \text{ and } \deg(i) \leq k
-\}$ to be the set of indices of degree at most $k$.
-
-\paragraph{Resource Polynomials}
-
-Let $V$ be a set of variables.  An index $I \in \ind(V)$ denotes a
-\emph{base polynomial} $P_I : \states \to \N$ for $V$ that maps a
-program state $H$ to product of binomial coefficients (a natural
-number).  We define
-$$
-P_I(H) = \prod_{{\{x,y\}} \subseteq V} \binom {|H(x){-}H(y)|} {i_{\{x,y\}}} \; .
-$$
-%
-A \emph{resource polynomial} $R$ for the variable set $V$ is a
-non-negative linear combination of the base polynomials for $V$.
-
-\paragraph{Potential Annotations}
-
-A \emph{potential annotation} for the variable set $V$ is a family
-$$Q = (q_I)_{I \in \ind(V)}$$ 
-of non-negative rational numbers.  Such an annotation denotes the
-resource polynomial $R_Q$ that is defined by
-$$
-R_Q(H) = \sum_{I \in \ind(V)} q_I \cdot P_I(H) \; .
-$$
-%
-We say that the potential annotation $Q$ is of degree $k$ if $q_I = 0$
-for $I \in \ind(V)$ with $\deg(I) > k$.
-
-\paragraph{Additive Shifts}
-
-Let $Q$ be a potential annotation for a variable set $V$ and let
-$\{x,y\} \subseteq V$ be a two-element variable set.  The
-\emph{additive shift} with respect to $\{x,y\}$ is a potential
-annotation $\shift_{\{x,y\}}(Q) = (q'_I)_{I \in \ind(V)} $ for $V$
-that is defined through
-$$
-q'_I = q_I + q_{I^{\{x,y\}{+}1}} \; .
-$$
-For an index $I = (i_{\{x,y\}})_{\{x,y\} \subseteq V}$ we use the
-notation $I^{\{x,y\}{+}k}$ to denote the index
-$(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
-$$
-i'_{\{t,u\}} = \left\{
-  \begin{array}{ll}
-    i_{\{t,u\}} + k  & \text{if } \{t,u\} = \{x,y\} \\
-    i_{\{t,u\}} & \text{otherwise}
-  \end{array}
-\right.
-\;.
-$$
-%
-The additive shift for natural numbers reflects the identity 
-\begin{equation}
-\label{eq:shift}
-\sum_{0 {\leq} i \leq {k}} q_i \binom{n+1}{i} = \sum_{0 {\leq} i \leq {k}} (q_i{+}q_{i+1}) \binom{n}{i}
-\end{equation}
-where $q_{k+1} = 0$.  It is used in the effect system if the
-difference $n+1$ between two variables $x,y$ decreases by one.
-
-\begin{lemma} Let $V$ be a set of variables with $x,y \in V$ and let
-  $H$ be a program state. Let $|H'(t) {-} H'(u)| = |H(t) {-} H(u)|$
-  for $\{t,u\} \neq \{x,y\}$ and let $|H'(x) {-} H'(y)| = |H(x) {-}
-  H(y)| - 1$.
-  %
-  If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
-\end{lemma}
-
-We now study the effect of multiple simultaneous shifts.  Let $Q$ be a
-resource annotation for a variable set $V$ and let $U_1,\ldots,U_n
-\subseteq V$ with $|U_i| = 2$ for all $i$ and $U_i \neq U_j$ for $i
-\neq j$ be pairwise distinct two-element variable sets.  The
-simulations additive shift $\shift_{U_1,\ldots,U_n}(Q)$ of $Q$ with
-respect to $U_1,\ldots,U_n$ is defined by
-$$
-\shift_{U_1,\ldots,U_n}(Q) = \shift_{U_1}( \cdots \shift_{U_n}(Q) \cdots ) \; .
-$$
-%
-\begin{proposition}
-  Let $V$ be a set of variables and let $U_1,\ldots,U_n$ be pairwise
-  distinct two-element variable sets.  Let $|H'(x) {-} H'(y)| = |H(x)
-  {-} H(y)|$ for $\{x,y\} \not\in \{U_1,\ldots,U_n\}$ and let $|H'(x)
-  {-} H'(y)| = |H(x) {-} H(y)| - 1$ for $\{x,y\} \in
-  \{U_1,\ldots,U_n\}$.
-  %
-  If $Q' = \shift_{U_1,\ldots,U_n}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
-\end{proposition}
-%
-As shown by the following lemma, the order in which the shifts for the
-individual $U_i$ are applied is insignificant.
-%
-\begin{lemma}
-  Let $\sigma : \{1,\ldots,n\} \to \{1,\ldots,n\}$ be a
-  permutation. Then $\shift_{U_1,\ldots,U_n}(Q) =
-  \shift_{U_{\sigma(1)},\ldots,U_{\sigma(n)}}(Q)$.
-\end{lemma}
-%
-For reasons of efficiency in the constraint generation, we give a more
-direct formula for the simultaneous shift.  Let $I \in \ind(V)$ and
-let $U_1,\ldots,U_n$ be pairwise distinct two-element variable sets.
-We define the index $I^{U_1,\ldots,U_n + k}$ as the family $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
-$$
-i'_{\{t,u\}} = \left\{
-  \begin{array}{ll}
-    i_{\{t,u\}} + k  & \text{if } \{t,u\} \in \{U_1,\ldots,U_n\} \\
-    i_{\{t,u\}} & \text{otherwise}
-  \end{array}
-\right.  \;.
-$$
-
-%
-\begin{lemma}
-  Let $V$ be a variable set and let $U_1,\ldots,U_n$ be pairwise
-  distinct two-element variable sets.
-  %
-  Let $Q = (q_I)_{I \in \ind(V)}$ be a resource annotation for
-  $V$ and let $ Q' = (q'_I)_{I \in \ind(V)}$ where
-  $$
-  q'_I = \sum_{\{j_1,\ldots,j_m\} \subseteq \{1,\ldots,n\} } q_{I^{U_{j_1},\ldots,U_{j_m}+1}} \; .
-  $$
-  Then $Q' = \shift_{U_1,\ldots,U_n}(Q)$.
-\end{lemma}
+% \section{Polynomial Potential}
+
+% \paragraph{Index Sets}
+
+% Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
+% family that maps two-element sets of variables to natural numbers,
+% that is,
+% $$
+% I = (i_{\{x,y\}})_{\{x,y\} \subseteq V} \; .
+% $$
+% %
+% We identify a family $I$ with the set $\{ (\{x,y\},i_{\{x,y\}})
+% \mid \{x,y\} \subseteq V\}$.
+
+% Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
+% instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
+% from the context.
+% %
+% We assume that allways $0 \in V$ and sometimes write $i_x$ instead of $i_{\{x,0\}}$.
+
+% The \emph{degree} $\deg(I)$ of an index $I = (i_{\{x,y\}})_{\{x,y\}
+%   \subseteq V}$ is defined as
+% $$
+% \deg(I) = \sum_{\{x,y\} \in V} i_{\{x,y\}} \;.
+% $$
+% We define $\ind_k(V) = \{ I \mid I \in \ind(V) \text{ and } \deg(i) \leq k
+% \}$ to be the set of indices of degree at most $k$.
+
+% \paragraph{Resource Polynomials}
+
+% Let $V$ be a set of variables.  An index $I \in \ind(V)$ denotes a
+% \emph{base polynomial} $P_I : \states \to \N$ for $V$ that maps a
+% program state $H$ to product of binomial coefficients (a natural
+% number).  We define
+% $$
+% P_I(H) = \prod_{{\{x,y\}} \subseteq V} \binom {|H(x){-}H(y)|} {i_{\{x,y\}}} \; .
+% $$
+% %
+% A \emph{resource polynomial} $R$ for the variable set $V$ is a
+% non-negative linear combination of the base polynomials for $V$.
+
+% \paragraph{Potential Annotations}
+
+% A \emph{potential annotation} for the variable set $V$ is a family
+% $$Q = (q_I)_{I \in \ind(V)}$$ 
+% of non-negative rational numbers.  Such an annotation denotes the
+% resource polynomial $R_Q$ that is defined by
+% $$
+% R_Q(H) = \sum_{I \in \ind(V)} q_I \cdot P_I(H) \; .
+% $$
+% %
+% We say that the potential annotation $Q$ is of degree $k$ if $q_I = 0$
+% for $I \in \ind(V)$ with $\deg(I) > k$.
+
+% \paragraph{Additive Shifts}
+
+% Let $Q$ be a potential annotation for a variable set $V$ and let
+% $\{x,y\} \subseteq V$ be a two-element variable set.  The
+% \emph{additive shift} with respect to $\{x,y\}$ is a potential
+% annotation $\shift_{\{x,y\}}(Q) = (q'_I)_{I \in \ind(V)} $ for $V$
+% that is defined through
+% $$
+% q'_I = q_I + q_{I^{\{x,y\}{+}1}} \; .
+% $$
+% For an index $I = (i_{\{x,y\}})_{\{x,y\} \subseteq V}$ we use the
+% notation $I^{\{x,y\}{+}k}$ to denote the index
+% $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
+% $$
+% i'_{\{t,u\}} = \left\{
+%   \begin{array}{ll}
+%     i_{\{t,u\}} + k  & \text{if } \{t,u\} = \{x,y\} \\
+%     i_{\{t,u\}} & \text{otherwise}
+%   \end{array}
+% \right.
+% \;.
+% $$
+% %
+% The additive shift for natural numbers reflects the identity 
+% \begin{equation}
+% \label{eq:shift}
+% \sum_{0 {\leq} i \leq {k}} q_i \binom{n+1}{i} = \sum_{0 {\leq} i \leq {k}} (q_i{+}q_{i+1}) \binom{n}{i}
+% \end{equation}
+% where $q_{k+1} = 0$.  It is used in the effect system if the
+% difference $n+1$ between two variables $x,y$ decreases by one.
+
+% \begin{lemma} Let $V$ be a set of variables with $x,y \in V$ and let
+%   $H$ be a program state. Let $|H'(t) {-} H'(u)| = |H(t) {-} H(u)|$
+%   for $\{t,u\} \neq \{x,y\}$ and let $|H'(x) {-} H'(y)| = |H(x) {-}
+%   H(y)| - 1$.
+%   %
+%   If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
+% \end{lemma}
+
+% We now study the effect of multiple simultaneous shifts.  Let $Q$ be a
+% resource annotation for a variable set $V$ and let $U_1,\ldots,U_n
+% \subseteq V$ with $|U_i| = 2$ for all $i$ and $U_i \neq U_j$ for $i
+% \neq j$ be pairwise distinct two-element variable sets.  The
+% simulations additive shift $\shift_{U_1,\ldots,U_n}(Q)$ of $Q$ with
+% respect to $U_1,\ldots,U_n$ is defined by
+% $$
+% \shift_{U_1,\ldots,U_n}(Q) = \shift_{U_1}( \cdots \shift_{U_n}(Q) \cdots ) \; .
+% $$
+% %
+% \begin{proposition}
+%   Let $V$ be a set of variables and let $U_1,\ldots,U_n$ be pairwise
+%   distinct two-element variable sets.  Let $|H'(x) {-} H'(y)| = |H(x)
+%   {-} H(y)|$ for $\{x,y\} \not\in \{U_1,\ldots,U_n\}$ and let $|H'(x)
+%   {-} H'(y)| = |H(x) {-} H(y)| - 1$ for $\{x,y\} \in
+%   \{U_1,\ldots,U_n\}$.
+%   %
+%   If $Q' = \shift_{U_1,\ldots,U_n}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
+% \end{proposition}
+% %
+% As shown by the following lemma, the order in which the shifts for the
+% individual $U_i$ are applied is insignificant.
+% %
+% \begin{lemma}
+%   Let $\sigma : \{1,\ldots,n\} \to \{1,\ldots,n\}$ be a
+%   permutation. Then $\shift_{U_1,\ldots,U_n}(Q) =
+%   \shift_{U_{\sigma(1)},\ldots,U_{\sigma(n)}}(Q)$.
+% \end{lemma}
+% %
+% For reasons of efficiency in the constraint generation, we give a more
+% direct formula for the simultaneous shift.  Let $I \in \ind(V)$ and
+% let $U_1,\ldots,U_n$ be pairwise distinct two-element variable sets.
+% We define the index $I^{U_1,\ldots,U_n + k}$ as the family $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
+% $$
+% i'_{\{t,u\}} = \left\{
+%   \begin{array}{ll}
+%     i_{\{t,u\}} + k  & \text{if } \{t,u\} \in \{U_1,\ldots,U_n\} \\
+%     i_{\{t,u\}} & \text{otherwise}
+%   \end{array}
+% \right.  \;.
+% $$
+
+% %
+% \begin{lemma}
+%   Let $V$ be a variable set and let $U_1,\ldots,U_n$ be pairwise
+%   distinct two-element variable sets.
+%   %
+%   Let $Q = (q_I)_{I \in \ind(V)}$ be a resource annotation for
+%   $V$ and let $ Q' = (q'_I)_{I \in \ind(V)}$ where
+%   $$
+%   q'_I = \sum_{\{j_1,\ldots,j_m\} \subseteq \{1,\ldots,n\} } q_{I^{U_{j_1},\ldots,U_{j_m}+1}} \; .
+%   $$
+%   Then $Q' = \shift_{U_1,\ldots,U_n}(Q)$.
+% \end{lemma}
 
@@
 %%
-
-\section{NOTES}
-
-\subsection{Things to say in the intro}
-
-\paragraph{Quote of Knuth}
-
-\begin{quotation}
-  There's even a fundamental gap in the foundations of my main
-  mathematical specialty, the analysis of algorithms. Consider, for
-  example, a computer program that sorts a list of numbers into
-  order. Thanks to the work of Floyd, Hoare, and others, we have
-  formal definitions of semantics, and tools by which we can verify
-  that sorting is indeed always achieved. My job is to go beyond
-  correctness, to an analysis of such things as the program's running
-  time: I write down a recurrence, say, which is supposed to represent
-  the average number of comparisons made by that program on random
-  input data. I'm 100\% sure that my recurrence correctly describes
-  the program's performance, and all of my colleagues agree with me
-  that the recurrence is "obviously" valid. Yet I have no formal tools
-  by which I can prove that my recurrence is right. I don't really
-  understand my reasoning processes at all! My student Lyle Ramshaw
-  began to create suitable foundations in his thesis (1979), but the
-  problem seems inherently difficult. Nevertheless, I don't lose any
-  sleep over this situation.
-\end{quotation}
-http://www.informit.com/articles/article.aspx?p=2213858
-
-\paragraph{to do}
-
-\begin{itemize}
-\item describe schedulability analysis for two schedulers
-\item include program logic from the stack paper for arbitrary resources
-\item what operational semantics to use? 
-\item verify automatic tool with program logic
-\item derive some crazy bounds (polynomial? logarithmic?) with the logic
-\item logic can also be used by automation: put linear constraints in the precondition
-\end{itemize}
-
-\paragraph{Advanteges over previous work}
-
-\begin{itemize}
-\item integrates well with hand-derived bounds in a program logic
-\item have formal soundness proof with respect to cost semantics
-\item can attach certificate to the code (proof-carrying code)
-\item can deal with resources that \emph{become available} (e.g, freeing memory) in an execution
-\item derive global bounds for the program as a function of the inputs (as opposed to local loop bounds)
-\item can handle advance control flow such as break, return, mutual recursive functions
-\item do not rely on abstract-interpretation based invariant generation
-\item show that constant factors are very precise with respect to the cost semantics
-\item demonstrate that the analysis is applicable in the formal verification of system code
-\end{itemize}
 

### 2014-06-25 paper/main.tex
add examples explanations

@@
 Theorem dbl_triple: forall Z,
-  htriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
+  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
 \end{lstlisting}
@@
 
+\paragraph{Using the Quantitative Logic.}
+
+Our first example on~\pref{fig:xmplmax} uses the expressive power of
+the logic to get a very precise runtime bound on a program searching
+a maximal element in an array.  The metric used simply counts the
+assignments performed by the program.  Hence, the resource consumption
+is closely related to the number $A$ of times the test \lstinline{a[i] > m}
+is true during the execution.
+%
+It is clear that $A+1$ is the count of ``maximums so far in the
+array \lstinline{a}'':
+$$
+  \# \{ k \mid 0 \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \},
+$$
+where we write $\# S$ for the cardinal of the set $S$.  This number is
+obviously bounded by $N$, the size of the array.  So any automated tool
+available would derive the linear bound $2 \cdot N$ for that program.
+But using the expressivity of our logic we can get a finer bound
+that depends on the contents of the array before the program execution.
+
+The only non-trivial part of this derivation is finding the loop
+invariant that is used in the while loop. To this end
+it is necessary to define the number $A(i)$ of ``maximums so far
+starting from $i$ in the array \lstinline{a}.''  When the condition
+\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
+so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
+thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
+assignment.  In the other case, no maximum so far is encountered so
+$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
+have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
+of the same program; it can be done either by adapting the proof or by
+applying the {\sc L:Frame} rule one top of the triple already derived.
+More generally, it is clear that without function calls, the {\sc L:Frame}
+is admissible in our system.
+
+The second example on~\pref{fig:xmplbs} shows a use case for
+logical variables as well as a metric for stack consumption.
+We are interested in showing that a binary search function
+$\code{bsearch}$ has logarithmic stack consumption.  We use
+a logical variable $Z$ in the function specification
+$\htriple{(Z = \log_2(h-l)) + Z \cdot F}{}{Z \cdot F}$
+to express that the stack required by the function is fully
+returned afterwards.
+
+The critical step in this proof is the application of the
+{\sc L:Call} rule to prove the recursive call.  At this point
+the context $\Delta$ contains the following specification
+for the function \code{bsearch}
+$$
+  \forall y \, (x,l,h) \, \_.
+  ( (\lambda y \, (\_, l, h) .\, (y {=} \log_2(h {-} l)) + y {\cdot} F)
+  , (\lambda y \, \_ .\, y {\cdot} F)
+  ).
+$$
+Using the {\sc L:Call} rule it is possible to instantiate
+$y$ with $Z - 1$, and because $M_f = F$ and $M_r = -F$ in
+the stack resource metric, the appropriate triple can be
+derived on the call site.  The rest of the proof does not
+involve any resource manipulation and is just bookkeeping
+of logical assertions.
+
 \paragraph{Soundness of Quantitative Hoare Triples.}
@@
 \begin{lstlisting}
-${\color{blue} \{ A(0, N) + N \} }$
+${\color{blue} \{ A(0) + N \} }$
 i=1; m=a[0];
 ${\color{blue}%
-  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-1) \} }$
+  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
 while (i < N) {
   ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
   if (a[i] > m)
@@
   ${\color{blue}%
-    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
+    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
   i=i+1;
   ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
 } ${\color{blue} \{ 0 \} }$
 \end{lstlisting}
-\caption{Example derivation where we noted $A(i, N)$
+\caption{Example derivation where we wrote $A(i)$
   for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
@@
 ${\color{blue}%
-  \{ (Z = \log_2(h-l)) + Z \} }$
+  \{ (Z = \log_2(h-l)) + Z \cdot F \} }$
 bsearch(x,l,h) {
@@
     ${\color{blue}%
-      \{ (h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
+      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
     m = h + (h-l)/2;
     ${\color{blue}%
-      \{ (m = \frac{h+l}{2} \land h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
+      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
     if (a[m]>x) h=m; else l=m;
     ${\color{blue}%
-      \{ (Z - 1 = \log_2(h-l)) + (Z-1) + 1 \} }$
+      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot F + F \} }$
     l = bsearch(x,l,h); }
   ${\color{blue}%
-    \{ Z-1 \} }$
+    \{ (Z-1) \cdot F - (-F) \} }$
   return l;
-} ${\color{blue} \{ Z \} }$
+} ${\color{blue} \{ Z \cdot F \} }$
 \end{lstlisting}
 \caption{Example derivation of a stack usage bound for a binary
-  search program, the metric used assigns a cost of 1 to function
-  calls and $-1$ to return statements, we noted $Z$ an arbitrary
+  search program.  The metric used sets a cost of $F$ before
+  to function call and $-F$ after the call, $F$ is the stack
+  frame size of the function $\code{bsearch}$.  We use $Z$ as
   logical variable.
   }
+\label{fig:xmplbs}
 \end{figure}

### 2014-06-25 paper/main.tex
trying to improve the introduction

@@
 %
-The goal of this work is to develop a resource analysis framework for
-C programs that is based on a solid semantic foundation.  The choice
-of C is primarily motivated by our ongoing work on the formally
-verified hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is
-mainly developed in C and is supposed to provide verified guaranties
-on timing and memory usage.  Moreover, C is also a natural choice
-because it is still the most widely used language for system
-development; in particular in safety-critical embedded and real-time
-systems where resource bound analyses are often required by regulatory
-authorities~\cite{May2013}.
-
-In a recent work, we formally verified a bound on the stack usage of
-CertiKOS
+In this work, we develop a resource analysis framework for C programs
+that is based on a solid semantic foundation in which we start with a
+formally defined cost model of C programs and derive resource bounds
+together with easily verifiable certificates of their soundness with
+respect to the cost model. The choice of C is primarily motivated by
+our ongoing work on the formally verified hypervisor kernel
+CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly developed in C and is
+supposed to provide verified guaranties on timing and memory usage.
+Moreover, C is also a natural choice because it is still the most
+widely used language for system development; in particular in
+safety-critical embedded and real-time systems where resource bound
+analyses are often required by regulatory authorities~\cite{May2013}.
+
+XXXXXXXXXX
+
+Our starting point is a recent work~\cite{veristack14}, in which we
+have formally verified bounds on the stack usage of C programs.  A key
+part of the stack-bound verification is based on a quantitative Hoare
+logic for deriving abstract stack bounds for CompCert Clight programs
+that depend on the depth of (recursive) function calls.  For
+non-recursive programs it is possible to compute derivations for such
+stack bounds in the quantitative Hoare logic automatically.  By
+compiling these bounds together with the C program, we have been able
+to automatically derive stack bounds C programs without recursion such
+as CertiKOS.
+
+Our \emph{first contribution} is a generalization of the quantitative
+Hoare logic for CompCert Clight that enables us to derive resource
+bounds that are parametric in the resource of interest.  The
+quantitative logic is implemented in the Coq Proof Assistant and
+proved sound with respect to a small-step operational cost semantics
+for Clight that counts a constant resource consumption for each
+statement.  The resource consumption is defined by a cost metric and can be
+negative to express the release of resources (e.g, as needed to reason
+about memory usage).  The semantics models cost of terminating and
+non-terminating executions and a specific \emph{tick} function can be
+inserted in the source code to further customize the cost model.
+
+- Problem: very tedious to reason about large programs
+- automatic amortized analysis was inspiration for logic
+- but hard to apply to C programs
+- main problem: negative integers and precise bounds for \emph{for loops} like in python
+  and control C-like control-flow with break and return
+
+Our \emph{second contribution} is
 
@@
 \begin{enumerate}
-\item We define a operational cost semantics for C programs that
-  accounts a constant cost for each statement.  The cost is parametric
-  in a cost metric and the semantics models cost of terminating and
-  non-terminating executions.  A specific \emph{tick} function can be
-  inserted in the source code to further customize the cost model.
+\item
 
@@
 \end{enumerate}
-%
+
+XXXXX
+
 Our approach complements existing work since it provides a semantic

### 2014-06-26 paper/main.tex
more work on the intro

--- 2014-06-25 paper/main.tex
@@
 %
-The goal of this work is to develop a resource analysis framework for
-C programs that is based on a solid semantic foundation.  The choice
-of C is primarily motivated by our ongoing work on the formally
-verified hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is
-mainly developed in C and is supposed to provide verified guaranties
-on timing and memory usage.  Moreover, C is also a natural choice
-because it is still the most widely used language for system
-development; in particular in safety-critical embedded and real-time
-systems where resource bound analyses are often required by regulatory
-authorities~\cite{May2013}.
-
-In a recent work, we formally verified a bound on the stack usage of
-CertiKOS
+In this work, we develop a resource analysis framework for C programs
+that is based on a solid semantic foundation in which we start with a
+formally defined cost model of C programs and derive resource bounds
+together with easily verifiable certificates of their soundness with
+respect to the cost model. The choice of C is primarily motivated by
+our ongoing work on the formally verified hypervisor kernel
+CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly developed in C and is
+supposed to provide verified guaranties on timing and memory usage.
+Moreover, C is a natural choice because it is still the most widely
+used language for system development; in particular in safety-critical
+embedded and real-time systems where resource bound analyses are often
+required by regulatory authorities~\cite{May2013}.
+
+Our starting point is a recent work~\cite{veristack14} in which we
+have formally verified bounds on the stack usage of C programs.  A key
+part of this stack-bound verification is based on a quantitative Hoare
+logic for deriving abstract stack bounds for CompCert Clight programs
+that depend on the depth of (recursive) function calls.  For
+non-recursive programs it is possible to compute derivations for such
+stack bounds in the quantitative Hoare logic automatically.  By
+compiling these bounds together with the C program, we have been able
+to automatically derive verified stack bounds for C programs without
+recursion such as CertiKOS.
+
+Our \emph{first contribution} is a generalization of the quantitative
+Hoare logic for CompCert Clight that enables us to derive resource
+bounds that are parametric in the resource of interest.  The
+quantitative logic is implemented in the Coq Proof Assistant and
+proved sound with respect to a small-step operational cost semantics
+for Clight that counts a constant resource consumption for each atomic
+execution step.  The semantics models the cost of terminating and
+non-terminating executions.  The constant resource consumption at each
+step is parametric in user-definable cost metric and can be negative
+to express the release of resources (as needed, e.g., to reason about
+memory usage).  Furthermore, a specific \emph{tick} function can be
+inserted in the source code to further customize the cost model.
+
+The idea of the quantitative Hoare logic is that we replace the
+boolean predicates (in the pre- and postconditions) of Hoare logic
+with numeric predicates that map a program state to a non-negative
+number or infinity.  The intuitive meaning of a quantitative Hoare
+triple $\htriple{P}{S}{Q}$ is that the program $S$ consumes less then
+$P(\state)$ resources when executed in starting state $\state$, and
+that at least $Q(\state')$ resources are left if $S$ terminates in
+state $\state'$.  We call $P$ the \emph{potential} before the
+execution of $S$ and $Q$ the potential after the execution of $S$.
+Infinity corresponds to \emph{false} in classic Hoare logic and
+boolean assertions can be embedded in the quantitative assertions to
+express that a bound is only valid for a subset of inputs.  In the
+implementation, we use a shallow embedding in Coq which makes the
+logic very flexible.  There is practically no limitation on the format
+of the derived resource bounds since they can be any function that is
+definable in Coq.  The derived bounds can also be parametric in a set
+of cost metrics or specifically apply to one fixed cost metric.
+
+The precision and flexibility of the quantitative Hoare logic provide
+a ample foundation for reasoning about resource consumption of Clight
+programs.  The logic is naturally compositional and the quantitative
+triples can be used to specify resource cost of library functions or
+to describe resource contracts between different functions.  However,
+reasoning about quantitative properties can often be more tedious then
+reasoning about intentional properties.  Consequently, automation is
+inevitable to derive resource bounds for large code bases like
+CertiKOS.  Such an automation was easily achievable in the case of
+constant stack bounds for a code base that does not contain recursive
+functions.  In general however, resource bounds have to be parametric
+in the arguments of a function and depend on the number of loop
+iterations and (recursive) function calls that are performed by the
+function.
+
+For functional programs, there exists resource analysis systems that
+are based on type-based amortized resource analysis~\cite{Jost03,
+  HoffmannAH12} and can automatically derive complex polynomial bounds
+with tight constant factors.  Type-based amortized resource analysis
+has been our main inspiration in the design of the quantitative Hoare
+logic and is therefore a natural candidate to automate the reasoning.
+However, it is a well-known problem that type-based amortized resource
+analysis works well for functional programs with pattern matching but
+not for C-like programs with control flow that depends integer
+arithmetic.  For example, it is a long-time open problem how to use
+type-base amortized analysis to derive bounds for functions whose
+resource consumption depends on (possibly negative) integers and
+non-sequential control flow as introduced by \code{break} and
+\code{return} statements.  Unfortunately, these are the functions with
+non-constant resource consumption that are most common in CertiKOS and
+other system software.
+
+Our \emph{second contribution} is an automatic amortized resource
+analysis for CompCert Clight that computes derivations in the
+quantitative Hoare logic.  It is the first automatic amortized
+resource analysis that can derive bounds that depend on negative
+integers and handles programs with mutually recursive as well as
+\code{break} and \code{return} statements.
+
+XXXX
+
+The main innovation that makes this work practical is our new
+automatic amortized analysis for C code.  It targets programs that are
+common in system software, that is, programs whose resource
+consumption can be described as a function of sizes $|[x,y]|$ of
+intervals of integer variables.  A typical bound would be $B(x,y,z) =
+10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
+1043$).  The use of interval sizes in potential functions is an
+important novelty.
+%
+Despite the apparent simplicity of the new analysis system, it is able
+to reproduce results from the literature which were obtained using
+methods based on abstract interpretation.  In contrast with abstract
+interpretation--based methods, our technique does not require any
+fixpoint computations on an abstract domain to obtain loop invariants.
+The mechanism we designed is able to leverage local assertions such as
+$x < y$ that we collect along the branching points of the program to
+obtain global invariants.  We achieve this by generating a linear
+constraint system that reflects resource cost and size changes of
+integer variables in the program.  A solution of the linear program is
+a resource usage bound for the C program.  We can naturally, handle
+advanced control flow such as break, return, iteration over negative
+integer variables, and mutually-recursive functions.
+
+Following the development steps of automatic amortized analysis for
+functional programs~\cite{Jost03,HoffmannH10}, we deliberately
+restrict our self to linear bounds in the automatic analysis (there a
+no restrictions for the manually-derived bounds in the logic).  More
+specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
+where $a$ and $b$ are integer variables or constants.  The reason is
+mainly clarity of presentation, since an extension to multivariate
+resource polynomials~\cite{HoffmannAH11} would make the inference rules considerably
+more involved and should better be described separately.  However, we
+developed the linear inference system so that the extension to
+polynomial bounds shall work smoothly.
+
+We implemented our analysis for CompCert Clight, a subset of C in
+which loops can only be exited with a \code{break} statement and
+expressions are free of side effects.  C programs can for instance be
+compiled to Clight programs using the verified CompCert C compiler.
+
+XXXXXXXXX
 
@@
 bound derivation if the automation fails, and produces verifiable
-certificates for the bounds.  Our contributions are as follows.
+certificates for the bounds.
+
+XXXXXXX
+
+In summary our contributions are as follows.
 %
 \begin{enumerate}
-\item We define a operational cost semantics for C programs that
-  accounts a constant cost for each statement.  The cost is parametric
-  in a cost metric and the semantics models cost of terminating and
-  non-terminating executions.  A specific \emph{tick} function can be
-  inserted in the source code to further customize the cost model.
+\item
 
@@
 
-The main innovation that makes this work practical is our new
-automatic amortized analysis for C code.  It targets programs that are
-common in system software, that is, programs whose resource
-consumption can be described as a function of sizes $|[x,y]|$ of
-intervals of integer variables.  A typical bound would be $B(x,y,z) =
-10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
-1043$).  The use of interval sizes in potential functions is an
-important novelty.
-%
-Despite the apparent simplicity of the new analysis system, it is able
-to reproduce results from the literature which were obtained using
-methods based on abstract interpretation.  In contrast with abstract
-interpretation--based methods, our technique does not require any
-fixpoint computations on an abstract domain to obtain loop invariants.
-The mechanism we designed is able to leverage local assertions such as
-$x < y$ that we collect along the branching points of the program to
-obtain global invariants.  We achieve this by generating a linear
-constraint system that reflects resource cost and size changes of
-integer variables in the program.  A solution of the linear program is
-a resource usage bound for the C program.  We can naturally, handle
-advanced control flow such as break, return, iteration over negative
-integer variables, and mutually-recursive functions.
-
-Following the development steps of automatic amortized analysis for
-functional programs~\cite{Jost03,HoffmannH10}, we deliberately
-restrict our self to linear bounds in the automatic analysis (there a
-no restrictions for the manually-derived bounds in the logic).  More
-specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
-where $a$ and $b$ are integer variables or constants.  The reason is
-mainly clarity of presentation, since an extension to multivariate
-resource polynomials~\cite{HoffmannAH11} would make the inference rules considerably
-more involved and should better be described separately.  However, we
-developed the linear inference system so that the extension to
-polynomial bounds shall work smoothly.
-
-We implemented our analysis for CompCert Clight, a subset of C in
-which loops can only be exited with a \code{break} statement and
-expressions are free of side effects.  C programs can for instance be
-compiled to Clight programs using the verified CompCert C compiler.
-
-Our new quantitative Hoare logic for CompCert Clight is a
-generalization of a program logic that we developed to formally very
-stack-space bounds for C programs in Coq~\cite{veristack14}.  In a
-nutshell, we replace the boolean predicates (pre and post conditions)
-of Hoare logic with numeric predicates that map a program state to a
-non-negative number or infinity.  The intuitive meaning of a
-quantitative Hoare triple $\htriple{P}{S}{Q}$ is that $S$ consumes
-less then $P(\state)$ resources when executed in starting state
-$\state$, and that $Q(\state')$ resources are left if $S$ terminates
-in state $\state'$.  Infinity corresponds to \emph{false} in classic
-Hoare logic.  Our quantitative logic is proved sound with respect to a
-cost semantic for Clight that accounts a constant cost for each
-evaluation step.  This constant cost is parametric in a resource
-metric that can be defined by the user.  Our shallow embedding in Coq
-enables the derivation of both, bounds that are parametric in the cost
-metric or bounds for one specific metric.
+XXXXX
 

### 2014-06-26 paper/main.tex
adjustments in the example descriptions

@@
 
-It is often necessary to relate certain invariants in the pre- and
-postcondition when specifying a program.  A standard solution of Hoare
+When specifying a program, it is often necessary to relate certain
+invariants in the pre- and postcondition.  A standard solution of Hoare
 logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
@@
 of Coq, our host theory.  When an logical variable is needed we
-introduce it using a universal quantifier of Coq before deriving
-triples in the logic, so the Coq theorem for the above example would
-look as follows.
+introduce it using a universal quantifier of Coq before using the
+logic to derive triples, so the Coq theorem for the above example
+would look as follows.
 %
@@
 v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
-$\Delta$.  Then we proceed to prove the function $S_f$ body with this
-induction hypothesis.  In this process it can be the case that the
-induction hypothesis needs to be used with a different value of a
+$\Delta$.  Then we proceed to prove the function body $S_f$ with this
+induction hypothesis.  In this process it can be the case that we
+have to use the induction hypothesis with a different value of a
 logical variable.  To cope with this problem, assumptions in the
@@
 
-% As mentioned, our account of the logic in this article is slightly
-% simplified compared to the Coq development to improve readability.
-% % However, the described logic is very close to the implemented version.
-
-% The main difference between the implemented logic and the logic
-% described here is that the latter does not have an
-% \emph{auxiliary state}.
-% %
-% Auxiliary state is a classic extension of Hoare logic (see for
-% example~\cite{Nipkow02}).
-%  It is used to share information
-% between the pre- and postcondition of a triple.  In a logic without
-% auxiliary state (or similar techniques) it is not possible to relate
-% program states before and after a statement.  For example, you cannot
-% specify that the function {\tt\ int\ twice\ ()\ $\{$i = i+i;$\}$\ }
-% doubles the value of the variable $i$.  With an auxiliary variable $Z$
-% it is possible specify this fact in Hoare logic using the triple $\htriple{i =
-%   Z}{\code{twice()}}{ i = 2 \cdot Z}$.
-
-% In the Coq implementation,
-% assertions have the type $\State \times \mathit{Aux} \to \N \cup
-% \{\infty\}$.  The auxiliary state is an arbitrary type in Coq and can
-% be instantiated by the user.  Most of the rules of the logic remain
-% unchanged in the presence of auxiliary state.  The
-% only exception is the consequence rule \textsc{L:Conseq} that reads as
-% follows.
-% {  \footnotesize
-% \begin{mathpar}
-%   \RuleNolabel
-%  {  \htriple{P'}{S}{R}
-%  \\  
-%  {\!\!\!\! \begin{array}{r}
-% \forall \sigma \, a. P(\sigma,a) {\in} \N \! \implies  \! \exists a'.  
-%      P(\sigma,a) {\geq} P'(\sigma,a') \hspace{3.3em} \\ \land (\forall \sigma'\, i.  R^i(\sigma',a') {\geq} Q^i(\sigma',a))
-%  \end{array}}\!\!\!\!
-%  }
-%  { \htriple{P}{S}{Q}
-%  }
-% \end{mathpar}}
-% This is the quantitative version of a consequence rule that has been
-% introduced in the context of Hoare logic and is attributed to Martin
-% Hofmann~\cite{Nipkow02}.  Its typical use case is when we apply the rule {\sc L:Call}
-% to a recursive call.  In this case, the Hoare triple for the function
-% call is proved by an assumption from the derivation context with
-% a slightly different auxiliary state.  In the example derivation
-% in~\pref{fig:bsearch_deriv} this different state is $Z - 1$.  Adapting
-% the derivation hypothesis to prove the recursive call is enabled in
-% our logic by the extended consequence rule introduced above.
-
-% \paragraph{Stack Framing}
-
-% A minor difference in the function application rule is that we only
-% present the rule for function calls with a single argument and without
-% \emph{framing of stack assertions}.  The latter is necessary in the
-% code of a caller to carry over information on the local environment
-% from the precondition of the function call to the postcondition of the
-% function call.  This is a general problem in Hoare logic and stack
-% framing is a well known solution.  For instance, we would like to prove
-% something like $\htriple{y}{ f(x) }{y}$ where $y$ is a local variable.
-% This is not possible with the rule \textsc{L:Call} but it is with the more
-% general rule in the Coq implementation~\iffullversion{\cite{coqfiles}}{\cite{anoncoq}}.
-
 \paragraph{Using the Quantitative Logic.}
 
-Our first example on~\pref{fig:xmplmax} uses the expressive power of
-the logic to get a very precise runtime bound on a program searching
-a maximal element in an array.  The metric used simply counts the
-assignments performed by the program.  Hence, the resource consumption
-is closely related to the number $A$ of times the test \lstinline{a[i] > m}
-is true during the execution.
+Our first example on~\pref{fig:xmplmax} derives a precise runtime bound
+on a program that searches a maximal element in an array.
+The metric used simply counts the assignments performed by the program.
+Hence, the resource consumption is closely related to the number $A$ of
+times the test \lstinline{a[i] > m} is true during the execution.
 %
@@
 available would derive the linear bound $2 \cdot N$ for that program.
-But using the expressivity of our logic we can get a finer bound
-that depends on the contents of the array before the program execution.
-
-The only non-trivial part of this derivation is finding the loop
-invariant that is used in the while loop. To this end
+But with the expressivity of our logic it is possible to use the
+previous set cardinal directly and precisely tie the bound to the
+initial contents of the array.
+
+The non-trivial part of this derivation is finding the loop
+invariant that for the while loop. To this end
 it is necessary to define the number $A(i)$ of ``maximums so far
@@
 have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
-of the same program; it can be done either by adapting the proof or by
+of the same program.  It can be done either by adapting the proof or by
 applying the {\sc L:Frame} rule one top of the triple already derived.
 More generally, it is clear that without function calls, the {\sc L:Frame}
-is admissible in our system.
+rule is admissible in our system.
 

### 2014-06-26 paper/main.tex
add some explanations in the soundness

@@
 where $S_f$ is the body of the function $f$.
-
+%
 \begin{theorem}[Soundness of the logic]
@@
     \forall n .\, \validC n \Delta
-      \implies \forall n .\, \valid n B R P S Q.
+      \implies \valid {n {+} 1} B R P S Q.
   $$
 \end{theorem}
+%
+\noindent
+If $S$ is a complete program and $\Delta$ is empty by
+expanding the definitions and remarking that $\validC n \Delta$
+and $\safeK n B R Q \Kstop$ are true for every $n$ we get that 
+$$
+  \forall n .\, \safe n P S \Kstop.
+$$
+This means that for any starting state $\state$, $P(\state)$
+provides enough resource for any run of the program $S$.  This
+setting is actually the main use case of the previous theorem
+which is stated as a stronger result to allow a proof by
+induction.
 

### 2014-06-26 paper/main.tex
typo
revamp automatic analysis a bit

@@
 $$
-where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.  Note that
+where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
 $q_0$ is the constant potential and that we treat constants as program
@@
  Let $M$ be a positive metric and assume that there exists $b \in
- \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
+ \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
@@
 \htriple{ P }{ S }{ Q } \\
-  x \in \mathbb Q_0^+
+  x \in \Qplusz
 }
@@
 The non-trivial part of this derivation is finding the loop
-invariant that for the while loop. To this end
+invariant for the while loop. To this end
 it is necessary to define the number $A(i)$ of ``maximums so far
@@
 of logical assertions.
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue} \{ A(0) + N \} }$
+i=1; m=a[0];
+${\color{blue}%
+  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
+while (i < N) {
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
+  if (a[i] > m)
+    m=a[i];
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
+  i=i+1;
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
+} ${\color{blue} \{ 0 \} }$
+\end{lstlisting}
+\caption{Example derivation where we wrote $A(i)$
+  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
+  the metric used here assigns a cost of 1 to every assignment
+  and 0 to all other operations.
+  }
+\label{fig:xmplmax}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}%
+  \{ (Z = \log_2(h-l)) + Z \cdot F \} }$
+bsearch(x,l,h) {
+  if (h-l > 1) {
+    ${\color{blue}%
+      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
+    m = h + (h-l)/2;
+    ${\color{blue}%
+      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
+    if (a[m]>x) h=m; else l=m;
+    ${\color{blue}%
+      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot F + F \} }$
+    l = bsearch(x,l,h); }
+  ${\color{blue}%
+    \{ (Z-1) \cdot F - (-F) \} }$
+  return l;
+} ${\color{blue} \{ Z \cdot F \} }$
+\end{lstlisting}
+\caption{Example derivation of a stack usage bound for a binary
+  search program.  The metric used sets a cost of $F$ before
+  to function call and $-F$ after the call, $F$ is the stack
+  frame size of the function $\code{bsearch}$.  We use $Z$ as
+  logical variable.
+  }
+\label{fig:xmplbs}
+\end{figure}
 
@@
 
-\section{Automatic Amortized Analysis}
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue} \{ A(0) + N \} }$
-i=1; m=a[0];
-${\color{blue}%
-  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
-while (i < N) {
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
-  if (a[i] > m)
-    m=a[i];
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
-  i=i+1;
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
-} ${\color{blue} \{ 0 \} }$
-\end{lstlisting}
-\caption{Example derivation where we wrote $A(i)$
-  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
-  the metric used here assigns a cost of 1 to every assignment
-  and 0 to all other operations.
-  }
-\label{fig:xmplmax}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}%
-  \{ (Z = \log_2(h-l)) + Z \cdot F \} }$
-bsearch(x,l,h) {
-  if (h-l > 1) {
-    ${\color{blue}%
-      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
-    m = h + (h-l)/2;
-    ${\color{blue}%
-      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
-    if (a[m]>x) h=m; else l=m;
-    ${\color{blue}%
-      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot F + F \} }$
-    l = bsearch(x,l,h); }
-  ${\color{blue}%
-    \{ (Z-1) \cdot F - (-F) \} }$
-  return l;
-} ${\color{blue} \{ Z \cdot F \} }$
-\end{lstlisting}
-\caption{Example derivation of a stack usage bound for a binary
-  search program.  The metric used sets a cost of $F$ before
-  to function call and $-F$ after the call, $F$ is the stack
-  frame size of the function $\code{bsearch}$.  We use $Z$ as
-  logical variable.
-  }
-\label{fig:xmplbs}
-\end{figure}
-
-We define in this section an abstract version of the proof system that
-we will use to derive bounds in our practical implementation.
-
-\subsection{Linear Potential Functions}
-
-To get a practical and automated implementation of amortized analysis
-as presented in the overview section we need the fix the shape of
-potential functions.  The general form of a potential function is as follows
-$$
-\Phi(H) = q_0 + \sum_{a \prec b} q_{ab}\,|H(a) - H(b)|.
-$$
-Where $a$ and $b$ range over the set of all program variables $V$ and
-$(q_{ab})$ and $q_0$ are non-negative rational numbers.  The relation
-$\prec$ is an arbitrary complete order on $V$ that is used only to
-avoid considering the two equal quantities $|H(a) - H(b)|$ and
-$|H(b) - H(a)|$ in the above sum.
-
-To simplify the references to the $q$ coefficients later we introduce
+\section{Automatic Analysis}
+
+In this section we present the deductive system used in our practical
+implementation to derive linear bounds on C programs.
+
+\paragraph{Linear Potential Functions.}
+
+To find derivations in our Quantitative Logic automatically, the main
+insight is to consider only assertions that match a certain shape.
+The general form of a potential function that we consider is
+$$
+  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
+    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}|.
+$$
+
+To simplify the references to the linear coefficients $q$, we introduce
 an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
-\{(a, b) \mid a \prec b\}$.  Each index $i$ corresponds to a \emph{base
+\{(x, y) \mid x, y \in \Var \}$.  Each index $i$ corresponds to a \emph{base
 function} $f_i$ in the potential function: $0$ corresponds to the constant
-function $\_ \mapsto 1$, and $(a,b)$ corresponds to $H \mapsto
-|H(a) - H(b)|$.  Using these notations we can rewrite the above
+function $\_ \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
+|\inter {\state(x)} {\state(y)}|$.  Using these notations we can rewrite the above
 equality as
 $$
-\Phi(H) = \sum_{i \in I} q_i f_i(H).
-$$
-From now on, we will write $ab$ to denote the index $(a,b)$.
-
-\subsection{Proof Rules}
-
-\begin{figure*}[t]
-\begin{mathpar}
-%
-\Rule{Q:Skip}
-{}
-{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
-%
-\and \Rule{Q:Break}
-{}
-{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
-%
-\and \Rule{Q:Return}
-{ P = Q_R[\Vret/x]  
-\\ \forall i \in \dom{P} . \, p_i = q_i
-}
-{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
-%
-\and \Rule{Q:Assert}
-{}
-{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and \Rule{Q:Tick}
-{}
-{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
-%
-\\ \Rule{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
-{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
-%
-\and \Rule{Q:Update}
-{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
-\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
-%
-\and \Rule{Q:IncP}
-{ \Gamma \models 0 \le y
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
-\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:IncM}
-{ \Gamma \models 0 \le y
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:Inc}
-{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-\\\\ M = M_u + M_e(x {\pm} y)
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
-%
-\and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
-}
-{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
-%
-\and \Rule{Q:Weak}
-{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
-\\\\ \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-\\ \Gamma'_2 \models \Gamma'_1
-}
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
-%
-\and\Rule{Q:Call}
-{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
-\and \text{Loc} = \text{Locals}(Q)
-\and \forall i \neq j . \, x_i \neq x_j
-\and c \in \mathbb Q_0^+
-\\ Q = P + S
-\and Q' = P' + S
-\and U = Q_f[\Vargs / \vec x]
-\and U' = Q'_f[\Vret/r]
-\\ \forall i \in \dom{U} . \, p_i = u_i
-\and \forall i \in \dom{U'} .\, p'_i = u'_i 
-\and \forall i \not\in \dom{U'} .\, p'_i = 0
-\and \forall i \not\in \text{Loc}
-  .\, s_i = 0
-}
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
-  \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
-}
-%
-\and \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
-\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
-\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
-\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
-}
-{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
-%
-\and \Rule{Extend}
-{ \fenv f = (\vec y, S_f)
-\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
-}
-{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
-%
-\end{mathpar}
-\caption{Inference rules of the quantitative analysis}
-\label{fig:proof}
-\end{figure*}
-
-The deductive system defined in~\pref{fig:proof} derives
+  \Phi = \sum_{i \in I} q_i f_i.
+$$
+We will often write $xy$ to denote the index $(x,y)$.  The family
+$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense)
+for potential functions, this allows us to represent uniquely any
+linear potential function $\Phi$ as a \emph{quantitative annotation}
+$Q = (q_i)_{i \in I}$, that is a family of rational numbers where
+only a finite number of elements are not zero.
+%
+For soundness, we require that potential functions always give
+non-negative results.  This is acheived by restricting coefficients
+in a quantitative annotation to be always non-negative.
+
+\paragraph{Logical State.}
+
+Quantitative annotations alone are not enough to make our system
+functional.  A minimal logical state needs to be maintained to
+justify certain operations on quantitative annotations.  For
+example when analyzing the code $x \gets x + y$, it is helpful
+to know the sign of $y$ to determine which intervals will increase
+or decrease.  The knowledge needed by our rules can often be
+inferred by local reasoning within usual theories (e.g. Presburger
+arithmetic or bit vectors).
+
+Our implementation uses conjunctions of linear inequalities as
+logical state.  We never compute fixpoints and take $\top$
+as pre- and postconditions for functions. It has proved to be
+sufficient for the variety of examples we designed and found
+in the literature.
+
+In contrast to the quantitative annotations, logical contexts in
+the presented system are left abstract.  This allows for simpler
+rules and leaves room for numerous improvements on top of our
+minimal implementation.
+
+\paragraph{Deduction Rules.}
+
+The deductive system defined in~\pref{fig:auto} derives
 judgements of the form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
@@
 two maps from indices to non-negative rational numbers
-$Q, Q' : I \rightarrow \mathbb Q_0^+$, these maps are called
+$Q, Q' : I \rightarrow \Qplusz$, these maps are called
 \emph{quantitative contexts}.
@@
 
-The rules of~\pref{fig:proof} are designed to enforce a semantic
+The rules of~\pref{fig:auto} are designed to enforce a semantic
 soundness similar to the one exposed in the overview section.  To
@@
 
-\paragraph{Logical part.}
-The second part of judgements maintains an abstract program state.
-This should be seen as a very weak Hoare logic.  It has to be weak because
-we want to automate it trivially and it can be weak because we only use
-it to get very local and simple knowledge.  The specifics of logical assertions
-in $\Gamma$ are left to the implementation section.
-%
-The purpose of this logical state is to help the rules
-make some sound decisions when relating coefficients of quantitative
-contexts.
-
-Syntactically $\Gamma$ is a list of program conditions.  Because of the
-rules {\sc Q:While} and {\sc Q:If}, we require the program conditions
-to be closed by negation.  Semantically we define the validity
-$H \models \Gamma$ of $\Gamma = C_1, \dots, C_n$ on a given
-heap $H$ to be $\bigwedge_i \llbracket C_i \rrbracket_H = \top$.
-%
-The logical state
-is maintained using regular Hoare style rules.
-The semantic entailment on logical states is noted $\Gamma \models \Delta$,
-its meaning is that $\forall H.\, H\models\Gamma \implies H\models\Delta$.
-
-%\paragraph{Explanation of selected rules.}
+\begin{figure*}[t]
+\begin{mathpar}
+%
+\Rule{Q:Skip}
+{}
+{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
+%
+\and \Rule{Q:Break}
+{}
+{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
+%
+\and \Rule{Q:Return}
+{ P = Q_R[\Vret/x]  
+\\ \forall i \in \dom{P} . \, p_i = q_i
+}
+{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
+%
+\and \Rule{Q:Assert}
+{}
+{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
+%
+\and \Rule{Q:Tick}
+{}
+{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
+%
+\\ \Rule{Q:Loop}
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+%
+\and \Rule{Q:Update}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
+%
+\and \Rule{Q:IncP}
+{ \Gamma \models 0 \le y
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
+\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:IncM}
+{ \Gamma \models 0 \le y
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:Inc}
+{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+\\\\ M = M_u + M_e(x {\pm} y)
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:If}
+{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+}
+{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+%
+\and\Rule{Q:Seq}
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+}
+{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+%
+\and \Rule{Q:Weak}
+{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\\\ \Gamma_1 \models \Gamma_2
+\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
+\\ \Gamma'_2 \models \Gamma'_1
+}
+{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
+%
+\and\Rule{Q:Call}
+{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
+\and \text{Loc} = \text{Locals}(Q)
+\and \forall i \neq j . \, x_i \neq x_j
+\and c \in \Qplusz
+\\ Q = P + S
+\and Q' = P' + S
+\and U = Q_f[\Vargs / \vec x]
+\and U' = Q'_f[\Vret/r]
+\\ \forall i \in \dom{U} . \, p_i = u_i
+\and \forall i \in \dom{U'} .\, p'_i = u'_i 
+\and \forall i \not\in \dom{U'} .\, p'_i = 0
+\and \forall i \not\in \text{Loc}
+  .\, s_i = 0
+}
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
+  \vdash r \gets f(\vec x) \dashv
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
+}
+%
+\and \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
+\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
+\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
+}
+{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
+%
+\and \Rule{Extend}
+{ \fenv f = (\vec y, S_f)
+\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
+}
+{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+%
+\end{mathpar}
+\caption{Inference rules of the quantitative analysis}
+\label{fig:auto}
+\end{figure*}
 

### 2014-06-26 paper/main.tex
more explanations of rules

@@
 \Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma }
+{ \code{istrue}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
+}
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
 \smallstep{M}
-(\state, S_1, \cont, \cost {-} M_c^1)
+(\state, S_1, \cont, \cost')
 }
@@
 \Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma }
+{ \code{isfalse}~\sem{e}_\sigma 
+\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
+}
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
 \smallstep{M}
-(\state, S_2, \cont, \cost {-} M_c^2)
+(\state, S_2, \cont, \cost')
 }
@@
 \htriple
-  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
 }
@@
 \htriple
-  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
 }
@@
 \htriple
-  { Q {+} M_t(n) }
+  { Q + M_t(n) }
   { \code{tick}(n) }
@@
 the postcondition.  We also require that have we the constant
-potential $M_u$ available in the precondition to pay for the cost of
-the update.
+potential $M_u + M_e(e)$ available in the precondition to pay for the
+cost of the evaluation of $e$ and the update.
 
@@
 The deductive system defined in~\pref{fig:auto} derives
-judgements of the form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
-Its semantics are formally described in the soundness section but we
-will give an intuition of its meaning here.
-
-\paragraph{Quantitative part.}
-The judgement is really two parts, one logical part, and one quantitative
-part.  The quantitative part is, as shown in the overview section, a
-representation of two potential functions $\Phi$ and $\Phi'$ using
-two maps from indices to non-negative rational numbers
-$Q, Q' : I \rightarrow \Qplusz$, these maps are called
-\emph{quantitative contexts}.
-We note $q_i$ for $Q(i)$.  Since $Q$ specifies a value for each
-index in $I$ it is clear that it defines a unique linear potential function
-as described in the previous section. We define $H \mapsto \Phi_H(Q)$
-to be the linear potential function canonically associated to $Q$.
-
-The rules of~\pref{fig:auto} are designed to enforce a semantic
-soundness similar to the one exposed in the overview section.  To
-do this, the rules impose contraints on $Q$ and $Q'$.  To make the
-presentation lighter, we use the following  set of conventions.
-First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
-and $\forall i \neq 0.\, q'_i = q_i$.  Second, for any index $i$ such that
-the relation between $q_i$ and $q'_i$ is not specified as a side condition
-of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
-all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
-non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
-0 is a program variable symbolizing the constant 0.
+judgements of the form
+$$
+  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
+    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
+$$
+These judgements are analogous to the logic judgements
+$\Delta; B; R \vdash \htriple P S Q$ where each assertion
+splits in two parts, a logical part $\Gamma$ and a quantitative
+part $Q$.  We left the function context $\Delta$ appart in
+the presentation to avoid cluttering notations, but it is
+implicit on all judgements.  We call precondition the pair
+$(\Gamma, Q)$ and postcondition the pair $(\Gamma', Q')$.
+The correspondence between judgements of the automation
+and the ones of the quantitative logic is made formal by our
+soundness proof of the analysis.
+
+Since the different rules of~\pref{fig:auto} manage the
+logical state in exactly the same way as the Quantitative
+Logic, we will focus on the quantitative annotations in the
+sequel.
+%
+As a convention, if $Q$ is a quantitative annotation we
+will write $q_i$ for its $i$th coordinate.  The notation
+$Q \pm n$ used in many rules defines a new context $Q'$
+such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\,
+q'_i = q_i$.  Finally, if a rule mentions $Q$ and $Q'$
+and leaves the latter undefined at some index $i$ we
+will implicitely assume that $q'_i = q_i$.
+
+%
+% Next paragraph is useless shit.
+%
+
+The {\sc Q:Skip} rule simply reuses its postcondition as
+precondition.  This is a sound choice since evaluating
+a $\code{skip}$ statement never costs any resource and
+does not change the program state.  A similar rule is
+the {\sc Q:Tick} rule which enforces that the quantitative
+annotation of the precondition is at least $M_t(n)$ to
+pay for the resource consumption.  Since the program
+state is also unchanged in this case, the logical part
+$\Gamma$ of the precondition can soundly be reused in
+the postcondition.  A slightly different rule is
+{\sc Q:Break}, in that case the regular control flow is
+broken, so anything that follows \code{break} in the
+program will not be executed.  Thus, the postcondition
+$(\Gamma', Q')$ can be arbitrary, and the precondition
+is taken from the break part of the judgement similarly
+to what is done in the logic.
+
+From these three examples alone, one can see that our
+automation follows very closely the rules of the logic.
+The case of the update statement is more interesting
+and specific to linear potential functions.
+In the {\sc Q:IncP} rule the program updates a variable
+$x$ to add it $y \ge 0$.  Since $x$ is changed, the
+quantitative annotation must be updated to reflect the
+change of the program state.  We write $x'$ the value
+of $x$ after the assignment ($x + y$). Since $x$ is
+the only variable changed, only intervals of the form
+$\inter u x$ and $\inter x u$ will be resized.
+Note that for any $u$, $\inter x u$ will get smaller
+with the update,  and if $x' \in \inter x u$ 
+we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.
+But $|\inter x {x'}| = |\inter 0 y|$, this means that
+the potential $q'_{0y}$ in the postcondition can
+be increased by $q_{xu}$ under the guard that
+$x' \in \inter x u$.  Dually, the interval
+$\inter v x$ can get bigger with the update.
+And we know that $|\inter v {x'}| \ge y + |\inter v x|$,
+so we decrease the potential of $\inter 0 y$
+by $q_{vx}$.  Remark that the rule forces it only for
+$v \not\in \mathcal U$, because if not, we know
+that $x \le v$ so $|\inter v x| = 0$.
 
@@
 \and \Rule{Q:Inc}
-{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
 \\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-\\\\ M = M_u + M_e(x {\pm} y)
 }
@@
 }
-{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
 %
@@
 \\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ \Gamma'_2 \models \Gamma'_1
 \\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-\\ \Gamma'_2 \models \Gamma'_1
 }

### 2014-06-26 paper/main.tex
(even) better into

@@
 In this work, we develop a resource analysis framework for C programs
-that is based on a solid semantic foundation in which we start with a
-formally defined cost model of C programs and derive resource bounds
-together with easily verifiable certificates of their soundness with
-respect to the cost model. The choice of C is primarily motivated by
-our ongoing work on the formally verified hypervisor kernel
-CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly developed in C and is
-supposed to provide verified guaranties on timing and memory usage.
-Moreover, C is a natural choice because it is still the most widely
-used language for system development; in particular in safety-critical
-embedded and real-time systems where resource bound analyses are often
-required by regulatory authorities~\cite{May2013}.
-
+that is based on a solid semantic foundation.  The choice of C is
+primarily motivated by our ongoing work on the formally verified
+hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
+developed in C and is supposed to provide verified guaranties on
+timing and memory usage.  Moreover, C is a natural choice because it
+is still the most widely used language for system development; in
+particular in safety-critical embedded and real-time systems where
+resource bound analyses are often required by regulatory
+authorities~\cite{May2013}.  Our \emph{contributions} are as follows.
+%
+\begin{enumerate}
+\item We define a operational cost semantics for CompCert Clight that
+  defines the resource consumption of terminating and diverging
+  executions.  The cost is parametric in a user-definable cost metric
+  and can be negative if resources are released during an execution.
+\item We develop a quantitative Hoare logic for interactively deriving
+  resource bounds for Clight programs.  The logic is implemented and
+  proved sound in the Coq Proof Assistant with respect to the cost
+  semantics of Clight.
+\item We describe an automatic static analysis that computes bounds
+  together with derivations in the quantitative Hoare logic.  The
+  analysis is based on type-based amortized resource analysis for
+  functional programs~\cite{Jost03,HoffmannAH12} and computes
+  derivations by generating simple linear constraints that can be
+  solved by an off-the-shelf LP solver.
+\item We show with a publically available prototype implementation and
+  an experimental evaluation that our novel amortized resource
+  analysis works precisely and efficiently for common loop and
+  recursion patterns and that the derived constant factors in the
+  bounds are close to or identical with the optimal ones.
+% \item We prove the completeness of the program logic
+%   with respect to the operational cost semantics.
+\end{enumerate}
+%
 Our starting point is a recent work~\cite{veristack14} in which we
@@
 logic for deriving abstract stack bounds for CompCert Clight programs
-that depend on the depth of (recursive) function calls.  For
-non-recursive programs it is possible to compute derivations for such
-stack bounds in the quantitative Hoare logic automatically.  By
-compiling these bounds together with the C program, we have been able
-to automatically derive verified stack bounds for C programs without
-recursion such as CertiKOS.
+that depend on the depth of (recursive) function calls.  % For
+% non-recursive programs it is possible to compute derivations for such
+% stack bounds in the quantitative Hoare logic automatically.  
+By compiling these bounds together with the C program, we derive
+verified stack bounds for x86 assembly code.
 
@@
 bounds that are parametric in the resource of interest.  The
-quantitative logic is implemented in the Coq Proof Assistant and
-proved sound with respect to a small-step operational cost semantics
-for Clight that counts a constant resource consumption for each atomic
-execution step.  The semantics models the cost of terminating and
-non-terminating executions.  The constant resource consumption at each
-step is parametric in user-definable cost metric and can be negative
-to express the release of resources (as needed, e.g., to reason about
-memory usage).  Furthermore, a specific \emph{tick} function can be
-inserted in the source code to further customize the cost model.
-
-The idea of the quantitative Hoare logic is that we replace the
-boolean predicates (in the pre- and postconditions) of Hoare logic
-with numeric predicates that map a program state to a non-negative
-number or infinity.  The intuitive meaning of a quantitative Hoare
-triple $\htriple{P}{S}{Q}$ is that the program $S$ consumes less then
-$P(\state)$ resources when executed in starting state $\state$, and
-that at least $Q(\state')$ resources are left if $S$ terminates in
-state $\state'$.  We call $P$ the \emph{potential} before the
-execution of $S$ and $Q$ the potential after the execution of $S$.
-Infinity corresponds to \emph{false} in classic Hoare logic and
-boolean assertions can be embedded in the quantitative assertions to
-express that a bound is only valid for a subset of inputs.  In the
-implementation, we use a shallow embedding in Coq which makes the
-logic very flexible.  There is practically no limitation on the format
-of the derived resource bounds since they can be any function that is
-definable in Coq.  The derived bounds can also be parametric in a set
-of cost metrics or specifically apply to one fixed cost metric.
+quantitative logic is implemented in Coq and proved sound with respect
+to a small-step operational cost semantics for Clight that counts a
+constant resource consumption for each atomic execution step.  The
+constant resource consumption at each step is parametric in
+user-definable cost metric and can be negative to express the release
+of resources (as needed, e.g., to reason about memory usage).
+Furthermore, a specific \emph{tick} function can be inserted in the
+source code to further customize the cost model.  In the
+implementation of the logic, we use a shallow embedding in Coq which
+makes the logic very flexible.  There is practically no limitation on
+the format of the derived resource bounds since they can be any
+function that is definable in Coq.  The derived bounds can also be
+parametric in a set of cost metrics or specifically apply to one fixed
+cost metric.
 
@@
 programs.  The logic is naturally compositional and the quantitative
-triples can be used to specify resource cost of library functions or
-to describe resource contracts between different functions.  However,
-reasoning about quantitative properties can often be more tedious then
-reasoning about intentional properties.  Consequently, automation is
-inevitable to derive resource bounds for large code bases like
-CertiKOS.  Such an automation was easily achievable in the case of
-constant stack bounds for a code base that does not contain recursive
-functions.  In general however, resource bounds have to be parametric
-in the arguments of a function and depend on the number of loop
-iterations and (recursive) function calls that are performed by the
-function.
-
-For functional programs, there exists resource analysis systems that
+Hoare triples can be used to specify resource cost of library
+functions or to describe resource contracts between different
+functions.  However, reasoning about quantitative properties can often
+be more tedious then reasoning about intentional properties.
+Consequently, automation is inevitable to derive resource bounds for
+large code bases like CertiKOS.  Such an automation was easily
+achievable in the case of constant stack bounds for a code base that
+does not contain recursive functions.  In general however, resource
+bounds have to be parametric in the arguments of a function and depend
+on the number of loop iterations and (recursive) function calls that
+are performed by the function.
+
+For functional programs, there exist resource analysis systems that
 are based on type-based amortized resource analysis~\cite{Jost03,
@@
 arithmetic.  For example, it is a long-time open problem how to use
-type-base amortized analysis to derive bounds for functions whose
+type-based amortized analysis to derive bounds for functions whose
 resource consumption depends on (possibly negative) integers and
@@
 integers and handles programs with mutually recursive as well as
-\code{break} and \code{return} statements.
-
-XXXX
-
-The main innovation that makes this work practical is our new
-automatic amortized analysis for C code.  It targets programs that are
-common in system software, that is, programs whose resource
-consumption can be described as a function of sizes $|[x,y]|$ of
-intervals of integer variables.  A typical bound would be $B(x,y,z) =
-10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
-1043$).  The use of interval sizes in potential functions is an
-important novelty.
+\code{break} and \code{return} statements.  It targets programs that
+are common in system software, that is, programs whose resource
+consumption can be described as a function of sizes $|[x,y]| =
+\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
+instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
+  i=i+K)} where the step-size $K>0$ is a constant.
+
+The use of interval sizes in potential functions (an therefore in
+resource bounds) is the main innovation that makes the analysis
+practical for C.
 %
 Despite the apparent simplicity of the new analysis system, it is able
-to reproduce results from the literature which were obtained using
-methods based on abstract interpretation.  In contrast with abstract
+to reproduce results from the literature (e.g,
+SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
+abstract--interpretation base methods.  In contrast with abstract
 interpretation--based methods, our technique does not require any
-fixpoint computations on an abstract domain to obtain loop invariants.
-The mechanism we designed is able to leverage local assertions such as
-$x < y$ that we collect along the branching points of the program to
-obtain global invariants.  We achieve this by generating a linear
+fixpoint computations to obtain loop invariants.  The mechanism we
+designed is able to leverage local assertions such as $x < y$ that we
+collect along the branching points of the program to obtain global
+resource invariants.  We achieve this by generating a linear
 constraint system that reflects resource cost and size changes of
 integer variables in the program.  A solution of the linear program is
-a resource usage bound for the C program.  We can naturally, handle
-advanced control flow such as break, return, iteration over negative
-integer variables, and mutually-recursive functions.
+a resource usage bound for the C program.  % We can naturally, handle
+% advanced control flow such as break, return, iteration over negative
+% integer variables, and mutually-recursive functions.
 
@@
 
-We implemented our analysis for CompCert Clight, a subset of C in
-which loops can only be exited with a \code{break} statement and
-expressions are free of side effects.  C programs can for instance be
-compiled to Clight programs using the verified CompCert C compiler.
-
-XXXXXXXXX
-
-In this article, we present the first resource analysis framework for
-C that makes it possible to combine automatically and interactively
-derived resource bounds.  It works automatically, efficiently and
-precisely for many examples, can be used interactively for manual
-bound derivation if the automation fails, and produces verifiable
-certificates for the bounds.
-
-XXXXXXX
-
-In summary our contributions are as follows.
-%
-\begin{enumerate}
-\item
-
-\item We describe an automatic static analysis technique for deriving
-  bounds for the cost that is defined by the operational semantics.
-  It is based on amortized resource analysis for functional
-  programs~\cite{Jost03,HoffmannAH12} and derives bounds that are
-  non-negative linear combinations of sizes $|[x,y]|$ of intervals
-  between program variables $x,y$.
-\item We show with a publically available prototype implementation and
-  an experimental evaluation that our novel amortized resource
-  analysis works efficiently for realistic programs and that the
-  derived constant factors in the bounds are close to or identical
-  with the optimal ones.
-\item We define a quantitative Hoare logic for interactively deriving
-  cost bounds for C programs.  The logic is implemented and proved
-  sound in the Coq Proof Assistant with respect to CompCert C
-  light.  We show that our automatic analysis readily produces
-  derivations in the quantitative Hoare logic.
-% \item We prove the completeness of the program logic
-%   with respect to the operational cost semantics.
-\end{enumerate}
-
-XXXXX
-
+To the best of our knowledge, this article presents the first resource
+analysis framework for C that makes it possible to combine non-trivial
+automatically derived bounds with interactively derived bounds in a
+prove system that produces verifiable certificates for the bounds.
 Our approach complements existing work since it provides a semantic
@@
 for automation that sometimes goes beyond the capabilities of existing
-techniques.  The quantitative Hoare logic and the automatic amortized
-analysis enable semi-automatic interactive resource-bound development.
-An automatically derived bound can be used in the quantitative logic
-and a hand derived bound can be used by the automatic analysis.  Both
-the automatic amortized analysis and the quantitative logic are
-naturally compositional and describe the resource behavior of code
-fragments without referring to the source code.  As a result, we
-naturally derive global bounds that are functions of the input
-parameters of the program.  To the best of our knowledge, our system
-is the only one that can automatically handle resources like memory
-that may become available during the execution.
-
-XXXXX
+techniques.  Both the automatic amortized analysis and the
+quantitative logic are naturally compositional and describe the
+resource behavior of code fragments without referring to the source
+code.  As a result, we naturally derive global bounds that are
+functions of the input parameters of the program.  Another unique
+feature of our system is, that it can naturally handle resources like
+memory that may become available during the execution.
 
@@
 
-In this section, we informally introduce the quantiatative program
+In this section, we informally introduce the quantitative program
 logic and the automatic amortized analysis for for Clight programs.

### 2014-06-26 paper/main.tex
more descriptions of the automatic analysis

@@
 that $x \le v$ so $|\inter v x| = 0$.
+
+Another very technical rule is {\sc Q:Call}.  It needs to
+account for the stack ``switching'' caused by the function
+call, the argument/return value passing, and the preservation
+of local variables.  We can sum up the requirements for
+this rule as follows,
+\begin{itemize}
+\item
+  any potential in the function pre/postcondition is equalized
+  to its matching potential in the pre/precondition,
+\item
+  the potential of intervals $|\inter x y|$ is preserved
+  if $x$ and $y$ are local variables,
+\item
+  and the unknown potentials after the call (e.g.
+  $|\inter x g|$, with $x$ local and $g$ global)
+  are zeroed out in the postcondition.
+\end{itemize}
+If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
+called, the rule takes care to split the potential of
+$|\inter x y|$ in two parts, one part to perform the
+computation in the function $f$ and one part to keep for
+later use after the function call.  This splitting is
+materialized in the equation $Q = P + S$.  Arguments in
+the function precondition $(\Gamma_f, Q_f)$ are named
+using a fixed vector $\Vargs$ of names different from
+all program variables.  This prevents name conflicts
+from happening and ensures that the substitution
+$[\Vargs/\vec x]$ is meaningful.  Symmetrically,
+we use the unique name $\Vret$ to represent the
+return value in the function postcondition
+$(\Gamma'_f, Q'_f)$.
 

### 2014-06-26 paper/main.tex
some minor updates in the logic section

@@
 deep embedding, or use the assertions of the host theory to get a
-shallow embedding.  For simplicity, we used the latter approach in our
-development.  This choice makes it possible to have logical variables
-almost for free: we can simply use the variable and binding mechanisms
-of Coq, our host theory.  When an logical variable is needed we
-introduce it using a universal quantifier of Coq before using the
-logic to derive triples, so the Coq theorem for the above example
-would look as follows.
+shallow embedding.  Because of its flexibility, we used the latter
+approach in our development.  This choice makes it possible to have
+logical variables almost for free: we can simply use the variable and
+binding mechanisms of Coq, our host theory.  When an logical variable
+is needed we introduce it using a universal quantifier of Coq before
+using the logic to derive triples.  For example, the Coq theorem for
+the above example would look as follows.
 %
@@
 $\Delta$.  Then we proceed to prove the function body $S_f$ with this
-induction hypothesis.  In this process it can be the case that we
-have to use the induction hypothesis with a different value of a
-logical variable.  To cope with this problem, assumptions in the
-function context $\Delta$ are universally quantified over logical
-variables.  The {\sc L:Extend} rule uses the host proof assistant to
-require that the triple on $f$'s body is proved for every possible
-logical variable $y$.
+induction hypothesis.  In this process it can be the case that we have
+to use the induction hypothesis with a different value of a logical
+variable (e.g., because the values of the arguments in the recursive
+call differ from the values of the arguments of the callee).  To cope
+with this problem, assumptions in the function context $\Delta$ are
+universally quantified over logical variables.  The {\sc L:Extend}
+rule uses the host proof assistant to require that the triple on $f$'s
+body is proved for every possible logical variable $y$.
 
@@
 
-Our first example on~\pref{fig:xmplmax} derives a precise runtime bound
-on a program that searches a maximal element in an array.
-The metric used simply counts the assignments performed by the program.
-Hence, the resource consumption is closely related to the number $A$ of
-times the test \lstinline{a[i] > m} is true during the execution.
-%
-It is clear that $A+1$ is the count of ``maximums so far in the
-array \lstinline{a}'':
-$$
-  \# \{ k \mid 0 \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \},
-$$
-where we write $\# S$ for the cardinal of the set $S$.  This number is
+In the following we demonstrate the use of the logic with two example
+derivations.
+
+In the example in~\pref{fig:xmplmax} we derive a precise runtime
+bound on a program that searches a maximal element in an array.  The
+cost metric that we use simply counts the assignments performed by the
+program.  Hence, the resource consumption is closely related to the
+number $A(0)$ of times the test \lstinline{a[i] > m} is true during the
+execution.
+%
+We observe that $A(0)+1$ is the count of ``maximum candidates'' in the
+array \lstinline{a} seen by the algorithm. We define
+$$
+ A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
+$$
+where we write $\# S$ for the cardinal of the set $S$.  $A(0)$ is
 obviously bounded by $N$, the size of the array.  So any automated tool
 available would derive the linear bound $2 \cdot N$ for that program.
-But with the expressivity of our logic it is possible to use the
+But with the expressivity of our  it is possible to use the
 previous set cardinal directly and precisely tie the bound to the
@@
 invariant for the while loop. To this end
-it is necessary to define the number $A(i)$ of ``maximums so far
+it is necessary to define the number $A(i)$ of ``maximum candidates so far
 starting from $i$ in the array \lstinline{a}.''  When the condition
@@
 assignment.  In the other case, no maximum so far is encountered so
-$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
-have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
-of the same program.  It can be done either by adapting the proof or by
-applying the {\sc L:Frame} rule one top of the triple already derived.
-More generally, it is clear that without function calls, the {\sc L:Frame}
-rule is admissible in our system.
-
-The second example on~\pref{fig:xmplbs} shows a use case for
-logical variables as well as a metric for stack consumption.
-We are interested in showing that a binary search function
-$\code{bsearch}$ has logarithmic stack consumption.  We use
-a logical variable $Z$ in the function specification
-$\htriple{(Z = \log_2(h-l)) + Z \cdot F}{}{Z \cdot F}$
-to express that the stack required by the function is fully
-returned afterwards.
+$A(i) = A(i+1)$.  % As a side remark, if $K \ge 0$, it is possible to
+% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
+% of the same program.  It can be done either by adapting the proof or by
+% applying the {\sc L:Frame} rule one top of the triple already derived.
+% More generally, without function calls, the {\sc L:Frame}
+% rule is admissible in our system.
+
+The example in~\pref{fig:xmplbs} shows a use case for logical
+variables as well as a metric for stack consumption. In a stack
+metric, we account a constant cost for a function call ($M_f>0$) that
+is returned after the call ($M_r = -M_f <0$).  All other resource cost
+are $0$.  We are interested in showing that a binary search function
+$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
+variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
+  + Z \cdot F}{}{Z \cdot F}$ to express that the stack required by the
+function is fully returned afterwards.
 
@@
 We already gave an intuition of the meaning of judgements
-derived in the logic, to make it formal we define the
+derived in the logic.  To make it formal, we define the
 \emph{resource safety} $\safe n P S \cont$ of an assertion
@@
     (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
-  &\implies 0 \le \cost'.
+  &\implies \cost' \ge 0.
 \end{align*}
-This predicate is step indexed by an integer $n$ that will
-be used in the soundness proof for the function call and
-loop cases.  The constraint $0 \le \cost'$ imposed by the
-definition makes sure that the program does not stop on a
-resource error.  However, it does not rule out memory safety
-errors or assertion failures, this is because our logic does
-not prove any safety or correctness theorems but only
-focuses on resource usage.
-
-The safety of a continuation $\cont$ is defined using three
+This predicate is step indexed by an integer $n$ that is used in the
+soundness proof for the function call and loop cases.  The constraint
+$\cost' \geq 0$ imposed by the definition ensures that the program does
+not stop because of a resource error (recall that a negative resource
+counter on the right-hand side is a resource failure).  However, it
+does not rule out memory safety errors or assertion failures, this is
+because our logic does not prove any safety or correctness theorems
+but only focuses on resource usage.
+
+The resource safety of a continuation $\cont$ is defined using three
 assertions, one for each of the possible outcome of a program
@@
 \end{align*}
-
-We can now define the \emph{semantic validity} of a logic judgement
-$B; R \vdash \htriple P S Q$ without function context as follows
+%
+We can now define the \emph{semantic validity} of a judgement $\cdot; B; R
+\vdash \htriple P S Q$ of the quantitative Hoare logic with
+empty function context as follows
 \begin{align*}
@@
 where $S_f$ is the body of the function $f$.
+
+XXXXX  Some more explanations.  Where is the difficulty? XXX
+
 %
@@
 \end{theorem}
+
+XXXXX  Some more explanations.  Why n+1?XXX
+
 %

### 2014-06-27 paper/main.tex
typos + remarks on logic soundness

@@
     ${\color{blue}%
-    \dashv y{>}99; \, 14 {+} P(n,y) \vdash }$
+    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
     tick(5);
@@
   ${\color{blue}%
-  \dashv \cdot; \, P(n,y) \vdash }$
+  \dashv \cdot; \, P(n,y) }$
 }
@@
 program.  Hence, the resource consumption is closely related to the
-number $A(0)$ of times the test \lstinline{a[i] > m} is true during the
+number $A(1)$ of times the test \lstinline{a[i] > m} is true during the
 execution.
 %
-We observe that $A(0)+1$ is the count of ``maximum candidates'' in the
+We observe that $A(1)+1$ is the count of ``maximum candidates'' in the
 array \lstinline{a} seen by the algorithm. We define
@@
 $$
-where we write $\# S$ for the cardinal of the set $S$.  $A(0)$ is
+where we write $\# S$ for the cardinal of the set $S$.  $A(1)$ is
 obviously bounded by $N$, the size of the array.  So any automated tool
 available would derive the linear bound $2 \cdot N$ for that program.
-But with the expressivity of our  it is possible to use the
+But with the expressivity of our logic it is possible to use the
 previous set cardinal directly and precisely tie the bound to the
@@
 %
-We can now define the \emph{semantic validity} of a judgement $\cdot; B; R
-\vdash \htriple P S Q$ of the quantitative Hoare logic with
-empty function context as follows
+We can now define the \emph{semantic validity} of a judgement $B; R
+\vdash \htriple P S Q$ of the quantitative logic without function
+context as follows
 \begin{align*}
@@
 
-Using this semantic validity  we can define the validity of
-a function context $\Delta$ that we write $\validC n \Delta$ as
+Using the semantic validity of triples we can define the validity
+of a function context $\Delta$, written $\validC n \Delta$, as
 \begin{align*}
@@
 \end{align*}
-where $S_f$ is the body of the function $f$.
-
-XXXXX  Some more explanations.  Where is the difficulty? XXX
-
+where $S_f$ is the body of the function $f$.  A full
+judgement that mentions a non-empty function context
+$\Delta$ is in fact a guarded statement: it makes
+assumptions on some functions' behavior.  The predicate
+$\validC n \Delta$ gives the precise meaning of the
+assumptions made.  It is also step-indexed to prove the
+soundness of the {\sc L:Extend} rule by induction.
 %
@@
 \end{theorem}
-
-XXXXX  Some more explanations.  Why n+1?XXX
-
 %
 \noindent
-If $S$ is a complete program and $\Delta$ is empty by
-expanding the definitions and remarking that $\validC n \Delta$
-and $\safeK n B R Q \Kstop$ are true for every $n$ we get that 
+The difference $\delta = 1$ between the index in the triple
+validity and the one in context validity is constrained by
+the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
+{\sc L:Call}, the language semantics makes one step and
+proceeds with the function body, so we must have
+$\delta \le 1$ to use the assumptions in $\Delta$.
+For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
+is a valid context for $n$ steps.  The induction hypothesis
+in that case says that if $\Delta \cup \Delta'$ is valid
+for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
+So if we want to solve this goal by induction, it is
+necessary that $\delta \ge 1$.  These two contraints force
+$\delta$ to be exactly one in the theorem statement.
+
+If $S$ is a complete program and $\Delta$ is empty, by
+expanding the definitions we see that $\Delta$ is valid
+for every $n$ and that $\Kstop$ is safe for every $n$, so
+we get
 $$
@@
 $$
-This means that for any starting state $\state$, $P(\state)$
+This means that from any starting state $\state$, $P(\state)$
 provides enough resource for any run of the program $S$.  This
@@
 
-To find derivations in our Quantitative Logic automatically, the main
+To find derivations in our quantitative logic automatically, the main
 insight is to consider only assertions that match a certain shape.
@@
 Since the different rules of~\pref{fig:auto} manage the
-logical state in exactly the same way as the Quantitative
-Logic, we will focus on the quantitative annotations in the
+logical state in exactly the same way as the quantitative
+logic, we will focus on the quantitative annotations in the
 sequel.
@@
 %
-% Next paragraph is useless shit.
+% Next paragraph is shit.
 %
@@
 $\inter v x$ can get bigger with the update.
-And we know that $|\inter v {x'}| \ge y + |\inter v x|$,
+And we know that $|\inter v {x'}| \le y + |\inter v x|$,
 so we decrease the potential of $\inter 0 y$
-by $q_{vx}$.  Remark that the rule forces it only for
-$v \not\in \mathcal U$, because if not, we know
-that $x \le v$ so $|\inter v x| = 0$.
+by $q_{vx}$ to pay for this change.  Remark that
+the rule forces it only for $v \not\in \mathcal U$,
+because if not, we know that $x \le v$ so
+$|\inter v x| = 0$.
 
@@
   any potential in the function pre/postcondition is equalized
-  to its matching potential in the pre/precondition,
+  to its matching potential in the call pre/postcondition,
 \item
@@
 later use after the function call.  This splitting is
-materialized in the equation $Q = P + S$.  Arguments in
-the function precondition $(\Gamma_f, Q_f)$ are named
-using a fixed vector $\Vargs$ of names different from
-all program variables.  This prevents name conflicts
-from happening and ensures that the substitution
-$[\Vargs/\vec x]$ is meaningful.  Symmetrically,
-we use the unique name $\Vret$ to represent the
-return value in the function postcondition
+realized by the equations $Q = P + S$ and $Q' =
+P' + S'$.  Arguments in the function precondition
+$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
+of names different from all program variables.  This
+prevents name conflicts from happening and ensures that
+the substitution $[\Vargs/\vec x]$ is meaningful.
+Symmetrically, we use the unique name $\Vret$ to represent
+the return value in the function postcondition
 $(\Gamma'_f, Q'_f)$.

### 2014-06-27 paper/main.tex
start soundness of the analysis

@@
 $(\Gamma'_f, Q'_f)$.
+
+\paragraph{Soundness Proof.}
+
+The soundness of the analysis leverages the quantitative logic.
+We propose to translate judgements of~\pref{fig:auto} into
+quantitative triples.  We define a translation function
+$\tr$, such that if a judgement $J$ in the automatic analysis
+is derivable, $\tr(J)$ is derivable in the quantitative
+logic.  Using this translated judgement and the soundness
+of our logic it is possible to get a certified resource bound
+for the analyzed program.
+
+The translation of an assertion $(\Gamma, Q)$ in the automatic
+analysis is defined by
+$$
+  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
+$$
+where we write $\Phi_Q$ for the unique linear potential function
+defined by the quantitative annotation $Q$.  We also need to
+translate the assumptions in function contexts of the automatic
+analysis, this is done as follows
+%
+\begin{align*}
+\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
+&\forall z \, \vec v \, v . (
+  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs \mapsto \vec v]), \\
+&\phantom{\forall z \, \vec v \, v . (}
+  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret \mapsto v])
+  ).
+\end{align*}
+%
+These definitions let us translate the judgement $J = B; R; P \vdash
+S \dashv P'$ in the context $\Delta$ by
+$$
+  \tr(J) :=
+  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
+    \htriple{\tr(P)}{S}{\tr(P')}.
+$$
+The soundness of the automatic analysis can now be stated
+formally with the following theorem.
+%
+\begin{theorem}[Soundness of the analysis]
+  If $J$ is a judgement derived with the rules of
+  \pref{fig:auto}, then $\tr(J)$ is a quantitative
+  Hoare triple derivable with the rules of
+  \pref{fig:logic}.
+\end{theorem}
+%
+\noindent
+The proof of this theorem is constructive and basically
+maps each rule of the automatic analysis directly to
+its counterpart in the quantitative logic.  With the
+computational content of the proof, we have an
+effective algorithm to construct proof objects for
+upper bounds derived automatically.
 

### 2014-06-30 paper/main.tex
title?
updates to intro and abstract

@@
 
-% \title{Automatic Amortized Analysis for Imperative Code}
-\title{Compositional Quantitative Resource Analysis of C Programs}
+\title{Compositional Certified Resource Bounds}
+% Certified Resource Bounds for C
+% Compositional Certified Resource Bounds for C
+% Compositional Quantitative Resource Analysis of C Programs
 % Compositional Resource-Bound Certification for C Programs
@@
   analysis framework consists of three parts.  First, the runtime
-  resource cost of programs is formalized with a parametric cost
-  semantics for CompCert Clight.  Second, a quantitative Hoare logic
-  that can be used for interactive bound development is implement and
-  proved sound with respect to the cost semantics in the Coq Proof
-  Assistant.  Third, an automatic amortized resource analysis for C
-  programs that computes derivations in the quantitative Hoare logic
-  is developed and implemented.  It is the first automatic amortized
-  analysis that can derive bounds that depend on negative integers and
-  differences between numbers.  Both, the quantitative logic and the
+  resource consumption of programs is formalized with a parametric
+  cost semantics for CompCert Clight.  Second, a quantitative Hoare
+  logic that can be used for interactive bound development is
+  implement and proved sound with respect to the cost semantics in the
+  Coq Proof Assistant.  The logic is very expressive and a shallow
+  embedding enables a derived bound to be any function that is
+  definable in Coq.  Third, an automatic amortized resource analysis
+  for C programs that computes derivations in the quantitative Hoare
+  logic is developed and implemented.  It is the first automatic
+  amortized analysis that can derive bounds that depend on negative
+  integers and differences between numbers, which is crucial to handle
+  typical systems code.  Both, the quantitative logic and the
   automatic amortized analysis are naturally compositional and can be
@@
   The practicability of the analysis framework is demonstrated with an
-  experimental evaluation.  The logic is used to derive XXX A comparison
-  with other automatic tools shows that the constant factors in the
-  bounds that are derived by the automatic amortized analysis are
-  often more precise. XXX
-
+  experimental evaluation.  The expressivity of the logic is
+  demonstrated by manually deriving customized bounds that are
+  tailored to specific algorithms.  A comparison of the automatic
+  amortized analysis with other automatic tools on 30 challenging
+  examples from the literature and open-source software shows that the
+  derived constant factors in our bounds are often more precise.
 \end{abstract}
@@
   together with derivations in the quantitative Hoare logic.  The
-  analysis is based on type-based amortized resource analysis for
-  functional programs~\cite{Jost03,HoffmannAH12} and computes
+  analysis is inspired by type-based amortized resource analysis for
+  functional programs~\cite{Jost03,HoffmannAH12}.  It computes
   derivations by generating simple linear constraints that can be
-  solved by an off-the-shelf LP solver.
+  solved by an off-the-shelf LP solver and does not require any
+  fixpoint computations to obtain loop invariants.
 \item We show with a publically available prototype implementation and
@@
   analysis works precisely and efficiently for common loop and
-  recursion patterns and that the derived constant factors in the
+  recursion patterns, and that the derived constant factors in the
   bounds are close to or identical with the optimal ones.
@@
 % stack bounds in the quantitative Hoare logic automatically.  
-By compiling these bounds together with the C program, we derive
+By compiling these bounds together with the C program, we have derived
 verified stack bounds for x86 assembly code.
@@
 constant resource consumption for each atomic execution step.  The
-constant resource consumption at each step is parametric in
-user-definable cost metric and can be negative to express the release
-of resources (as needed, e.g., to reason about memory usage).
-Furthermore, a specific \emph{tick} function can be inserted in the
-source code to further customize the cost model.  In the
-implementation of the logic, we use a shallow embedding in Coq which
-makes the logic very flexible.  There is practically no limitation on
-the format of the derived resource bounds since they can be any
-function that is definable in Coq.  The derived bounds can also be
-parametric in a set of cost metrics or specifically apply to one fixed
-cost metric.
-
-The precision and flexibility of the quantitative Hoare logic provide
-a ample foundation for reasoning about resource consumption of Clight
-programs.  The logic is naturally compositional and the quantitative
-Hoare triples can be used to specify resource cost of library
-functions or to describe resource contracts between different
-functions.  However, reasoning about quantitative properties can often
-be more tedious then reasoning about intentional properties.
-Consequently, automation is inevitable to derive resource bounds for
-large code bases like CertiKOS.  Such an automation was easily
-achievable in the case of constant stack bounds for a code base that
-does not contain recursive functions.  In general however, resource
-bounds have to be parametric in the arguments of a function and depend
-on the number of loop iterations and (recursive) function calls that
-are performed by the function.
-
-For functional programs, there exist resource analysis systems that
-are based on type-based amortized resource analysis~\cite{Jost03,
-  HoffmannAH12} and can automatically derive complex polynomial bounds
-with tight constant factors.  Type-based amortized resource analysis
-has been our main inspiration in the design of the quantitative Hoare
-logic and is therefore a natural candidate to automate the reasoning.
-However, it is a well-known problem that type-based amortized resource
-analysis works well for functional programs with pattern matching but
-not for C-like programs with control flow that depends integer
-arithmetic.  For example, it is a long-time open problem how to use
-type-based amortized analysis to derive bounds for functions whose
-resource consumption depends on (possibly negative) integers and
-non-sequential control flow as introduced by \code{break} and
-\code{return} statements.  Unfortunately, these are the functions with
-non-constant resource consumption that are most common in CertiKOS and
-other system software.
+resource consumption at each step is parametric in a user-definable cost
+metric and can be negative to express the release of resources (as
+needed, e.g., to reason about memory usage).  Furthermore, a specific
+\emph{tick} function can be inserted in the source code to further
+customize the cost model.  In the implementation of the logic, we use
+a shallow embedding in Coq which makes the logic very flexible.  There
+is practically no limitation on the format of the derived resource
+bounds since they can be any function that is definable in Coq.  The
+derived bounds can also be parametric in a set of cost metrics or
+specifically apply to one fixed cost metric.
+
+The precision and expressivity of the quantitative Hoare logic provide
+an ample foundation for reasoning about resource consumption of Clight
+programs.  The logic is naturally compositional and the derived
+\emph{quantitative Hoare triples} can be used to specify the resource
+consumption of library functions or to describe resource contracts
+between different functions.  However, reasoning about quantitative
+properties can often be more tedious then reasoning about intentional
+properties.  Consequently, automation is inevitable to derive resource
+bounds for large code bases like CertiKOS.  Such an automation was
+easily achievable in the case of constant stack bounds for a code base
+that does not contain recursive functions~\cite{veristack14}.  In
+general however, resource bounds have to be parametric in the
+arguments of a function and depend on the number of loop iterations
+and (recursive) function calls that are performed by the function.
+
+For functional programs, there exist resource analysis systems---based
+on type-based amortized resource analysis~\cite{Jost03,
+  HoffmannAH12}---that can automatically derive complex polynomial
+bounds with tight constant factors.  This type-based amortized
+resource analysis has been our main inspiration in the design of the
+quantitative Hoare logic and is therefore a natural candidate to
+automate the reasoning.  However, it is a well-known problem that
+type-based amortized resource analysis works well for functional
+programs with pattern matching but not for C-like programs with
+control flow that depends integer arithmetic.  For example, it is a
+long-time open problem how to use type-based amortized analysis to
+derive bounds for functions whose resource consumption depends on
+(possibly negative) integers and non-sequential control flow as
+introduced by \code{break} and \code{return} statements.
+Unfortunately, these are the functions with non-constant resource
+consumption that are most common in CertiKOS and other system
+software.
 
@@
   i=i+K)} where the step-size $K>0$ is a constant.
-
+%
 The use of interval sizes in potential functions (an therefore in
@@
 practical for C.
-%
+
 Despite the apparent simplicity of the new analysis system, it is able
@@
 constraint system that reflects resource cost and size changes of
-integer variables in the program.  A solution of the linear program is
-a resource usage bound for the C program.  % We can naturally, handle
+integer variables in the program.  A solution of the linear program
+immediately yields a resource usage bound for the C
+program.  % We can naturally, handle
 % advanced control flow such as break, return, iteration over negative
@@
 functional programs~\cite{Jost03,HoffmannH10}, we deliberately
-restrict our self to linear bounds in the automatic analysis (there a
-no restrictions for the manually-derived bounds in the logic).  More
-specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
-where $a$ and $b$ are integer variables or constants.  The reason is
-mainly clarity of presentation, since an extension to multivariate
-resource polynomials~\cite{HoffmannAH11} would make the inference rules considerably
-more involved and should better be described separately.  However, we
-developed the linear inference system so that the extension to
-polynomial bounds shall work smoothly.
+restrict our self to linear bounds in the automatic analysis (there
+are no restrictions for the manually-derived bounds in the logic).
+More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
+where $a$ and $b$ are integer variables or constants.  The reason for
+our focus on linear bounds is mainly clarity of presentation.  We
+already experimented with an extension to multivariate resource
+polynomials~\cite{HoffmannAH11} but this would make the inference
+rules considerably more involved and should better be described
+separately.  However, we developed the linear inference system so that
+the extension to polynomial bounds shall work smoothly.
 
@@
 automatically derived bounds with interactively derived bounds in a
-prove system that produces verifiable certificates for the bounds.
+proof system that produces verifiable certificates for the bounds.
 Our approach complements existing work since it provides a semantic
@@
 
-We have evaluated the quantitative logic and the automatic analysis
-with system code and examples from the literature. \pref{app:cat}
-contains more than $30$ (XXX) challenging loop and recursion patterns
-that we collected from open source software and the literature.  Our
-analysis can find asymptotically tight bounds for $30$ all but 3 of
-these patterns and in most cases the derived constant factors are
-tight.  We compared our automatic analyzer to other resource-analysis
-systems.  We selected $5$ examples (see Section XXX) for which we
-derived linear bounds and translated them to term-rewriting systems by
-hand in order to run them with the tools KoAT, PUBS, and Rank.  The
-tools were not able to find linear bounds for the examples.  LOOPUS is
-able to derive a bound on XXX of the examples.  But in contrast to our
-analysis, the constant factors are not tight.  We also contacted the
-authors of SPEED but were not able to obtain a tool to perform a
-comparison.
-
-To show the practicability of our approach we developed a case study
-with different real-time schedulers implemented in C.  To perform a
-schedulability analysis that takes into account the overhead of the
-scheduler we derived time bounds for context switching and
-initialization of the scheduler.  We then computed time bounds for a
-set of given user task and took account the previously computed
-scheduler overhead to determine whether the scheduling requirements of
-all user tasks can be met.
-
-With the intention, of making the material easily accessible we first
-informally describe the new automatic amortized analysis
-(\pref{sec:inform}).  We then formalize the intuition with the
-operational cost semantics an the inference rules for the amortized
-analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
-quantitative Hoare logic and its soundness proof that is formalized in
-Coq.  In \pref{sec:inter} we prove the soundness of the automatic
-amortized analysis by showing that derivation using the inference
-rules can be seen as derivations in the quantitative logic.  We also
-show how to integrate hand-derived bounds with automatically-derived
-bounds.  \pref{sec:exper} contains the results of the experimental
-evaluation and \pref{sec:related} describes related research.
+We have evaluated the automatic analysis with system code and examples
+from the literature.  \pref{app:cat} contains more than $30$
+challenging loop and recursion patterns that we collected from open
+source software and the literature.  Our analysis can find
+asymptotically tight bounds for $30$ all but 3 of these patterns and
+in most cases the derived constant factors are tight.  To compare our
+automatic analyzer with existing techniques, we tested our examples
+with the tools KoAT~\cite{BrockschmidtEFFG14},
+PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and
+LOOPUS~\cite{SinnZV14}.  We also contacted the authors of
+SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
+comparison.  Our experiments shows that the bounds that we derive are
+often more precise than those derived by existing tools.  Only
+LOOPUS~\cite{SinnZV14}, which also uses amortization techniques, is
+able to achieve a similar precision.
+
+Several micro benchmarks demonstrate the practicability and
+expressiveness of the quantitative Hoare logic.  For example, we
+derive a logarithmic bound for a binary search function, a bound that
+depends on the contents of an array to describe the exact cost of a
+function that finds the maximal element in an array, and a linear
+bound that amortizes the cost of $n$ successive increments to a binary
+counter.  XXX In a larger case study, we show how the quantitative
+logic can be used to derive the amortized cost for data-structure
+operations for a table that expands or shrinks depending on the load
+factor.
+
+% To show the practicability of our approach we developed a case study
+% with different real-time schedulers implemented in C.  To perform a
+% schedulability analysis that takes into account the overhead of the
+% scheduler we derived time bounds for context switching and
+% initialization of the scheduler.  We then computed time bounds for a
+% set of given user task and took account the previously computed
+% scheduler overhead to determine whether the scheduling requirements of
+% all user tasks can be met.
+
+% With the intention, of making the material easily accessible we first
+% informally describe the new automatic amortized analysis
+% (\pref{sec:inform}).  We then formalize the intuition with the
+% operational cost semantics an the inference rules for the amortized
+% analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
+% quantitative Hoare logic and its soundness proof that is formalized in
+% Coq.  In \pref{sec:inter} we prove the soundness of the automatic
+% amortized analysis by showing that derivation using the inference
+% rules can be seen as derivations in the quantitative logic.  We also
+% show how to integrate hand-derived bounds with automatically-derived
+% bounds.  \pref{sec:exper} contains the results of the experimental
+% evaluation and \pref{sec:related} describes related research.
 

### 2014-06-30 paper/main.tex
minor update to abstract

@@
   implement and proved sound with respect to the cost semantics in the
-  Coq Proof Assistant.  The logic is very expressive and a shallow
+  Coq Proof Assistant.  The logic is very expressive and shallow
   embedding enables a derived bound to be any function that is
@@
   amortized analysis that can derive bounds that depend on negative
-  integers and differences between numbers, which is crucial to handle
+  integers or differences between numbers, which is crucial to handle
   typical systems code.  Both, the quantitative logic and the
@@
   examples from the literature and open-source software shows that the
-  derived constant factors in our bounds are often more precise.
+  derived bounds are often more precise.
 \end{abstract}

### 2014-06-30 src/ctest/qsort.c
dirty hack for functions in splicing

@@ -1,0 +1,24 @@
+/* Quicksort */
+
+int nondet(void);
+
+void qsort(int *a, int lo, int hi) {
+  int m1, m2, n;
+
+  if (hi - lo < 1)
+    return;
+
+  n = nondet();   /* partition the array */
+  assert( n > 0 );
+  assert( lo + n <= hi );
+
+  m1 = n + lo;
+  m2 = m1 - 1;
+
+  qsort(a, m1, hi);
+  qsort(a, lo, m2);
+}
+
+void main(int *a, int len) {
+  qsort(a, 0, len);
+}

### 2014-06-30 paper/main.tex
add last examples

@@
 \begin{figure*}[t!]
- \setlength{\progwidth}{.24\linewidth}
+ \setlength{\progwidth}{.32\linewidth}
   \centering
@@
    \begin{lstlisting}
- XXX
-   \end{lstlisting}
-
-$XXX$
+int srch(
+  int t[], int n,
+  int p[], int m,
+  int b[]
+) {
+  int i=0, j=0, k=-1;
+
+  while (i < n) {
+    while (j >= 0 && t[i]!=p[j]) {
+      k = b[j];
+      assert(k > 0);
+      assert(k <= j + 1);
+      j -= k;
+    }
+    i++, j++;
+    if (j == m)
+      break;
+  }
+  return i;
+}
+   \end{lstlisting}
+
+$1 + 2|\inter 0 n|$
 \\[.7\baselineskip]
@@
    \begin{lstlisting}
- XXX
-   \end{lstlisting}
-
-$XXX$
+int gcd(int x, int y) {
+  if (x <= 0) return y;
+  if (y <= 0) return x;
+
+  for (;;) {
+    if (x>y) x -= y;
+    else if (y>x) y -= x;
+    else return x;
+  }
+}
+   \end{lstlisting}
+
+$|\inter 0 x| + |\inter 0 y|$
 \\[.7\baselineskip]
@@
    \begin{lstlisting}
- XXX
-   \end{lstlisting}
-
-$XXX$
+void qsort(int a[], int lo, int hi) {
+  int m1, m2, n;
+
+  if (hi - lo < 1) return;
+
+  n = nondet(); /* partition the array */
+  assert( n > 0 );
+  assert( lo + n <= hi );
+
+  m1 = n + lo;
+  m2 = m1 - 1;
+
+  qsort(a, m1, hi);
+  qsort(a, lo, m2);
+}
+
+void main(int a[], int len) {
+  qsort(a, 0, len);
+}
+   \end{lstlisting}
+
+$1 + 2 |\inter 0 {\code{len}}|$
 \\[.7\baselineskip]

### 2014-06-30 src/ctest/gcd.c

@@ -1,0 +1,9 @@
+int gcd(int x, int y) {
+  if (x <= 0) return y;
+  if (y <= 0) return x;
+  for (;;) {
+         if (x>y) x -= y;
+    else if (y>x) y -= x;
+    else return x;
+  }
+}

### 2014-06-30 src/ctest/kmp.c

@@ -1,0 +1,20 @@
+/* Knuth-Morris-Pratt string searching */
+
+int srch(int t[], int n, int p[], int m, int b[])
+{
+  int i = 0, j = 0, k = -1;
+
+  while (i < n)
+  {
+    while (j >= 0 && t[i] != p[j]) {
+       k = b[j];
+       assert(k > 0);
+       assert(k <= j + 1);
+       j -= k;
+    }
+    i++, j++;
+    if (j == m)
+        break;
+  }
+  return i;
+}

### 2014-06-30 paper/main.tex
typos

@@
 between different functions.  However, reasoning about quantitative
-properties can often be more tedious then reasoning about intentional
+properties can often be more tedious than reasoning about intentional
 properties.  Consequently, automation is inevitable to derive resource
@@
 resource analysis that can derive bounds that depend on negative
-integers and handles programs with mutually recursive as well as
+integers and handles programs with mutually recursive functions as well as
 \code{break} and \code{return} statements.  It targets programs that
@@
 SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
-abstract--interpretation base methods.  In contrast with abstract
-interpretation--based methods, our technique does not require any
+abstract-interpretation based methods.  In contrast with abstract-interpretation
+based methods, our technique does not require any
 fixpoint computations to obtain loop invariants.  The mechanism we
@@
 functional programs~\cite{Jost03,HoffmannH10}, we deliberately
-restrict our self to linear bounds in the automatic analysis (there
+restrict ourselves to linear bounds in the automatic analysis (there
 are no restrictions for the manually-derived bounds in the logic).

### 2014-07-01 paper/main.tex
updates to the disussion of the soundness of the logic

@@
 This predicate is step indexed by an integer $n$ that is used in the
-soundness proof for the function call and loop cases.  The constraint
+soundness proof for the function-call and loop cases.  The constraint
 $\cost' \geq 0$ imposed by the definition ensures that the program does
@@
 counter on the right-hand side is a resource failure).  However, it
-does not rule out memory safety errors or assertion failures, this is
+does not rule out memory safety errors or assertion failures. This is
 because our logic does not prove any safety or correctness theorems
 but only focuses on resource usage.
+%
+An interesting detail in the definition is the natural number $m$.  We
+simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
+\cont $.  This would not be the case if we replaced all occurrences of
+$m$ by $n$ in the definition of validity.
 
 The resource safety of a continuation $\cont$ is defined using three
-assertions, one for each of the possible outcome of a program
-statement.  We denote it by $\safeK n B R Q \cont$ and define
-it to be
+assertions, one for each of the possible outcomes of a program
+statement.  We define it as follows.
 \begin{align*}
-& \safe n B {\code{break}} \cont \\
-\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
-\land& \safe n Q {\code{skip}} \cont.
+\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
+& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
+& & \land& \safe n Q {\code{skip}} \cont
 \end{align*}
@@
 \vdash \htriple P S Q$ of the quantitative logic without function
-context as follows
+context as $\valid n B R P S Q :=$
 \begin{align*}
-&\valid n B R P S Q := \\
-&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
-&\qquad \implies \safe m {P+x} S \cont.
+& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
+& \implies \safe m {P+x} S \cont \; .
 \end{align*}
 Note how the validity of a triple embeds the frame rule of
-our logic, this refinement is necessary to have a stronger
+our logic. This refinement is necessary to have a stronger
 induction hypothesis available during the proof.
+%
+We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
+S Q$ implies $\valid n B R P S Q$.
 
@@
 The difference $\delta = 1$ between the index in the triple
-validity and the one in context validity is constrained by
+validity and the one in context validity arises from
 the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
@@
 $$
-  \forall n .\, \safe n P S \Kstop.
+\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
 $$
 This means that from any starting state $\state$, $P(\state)$
-provides enough resource for any run of the program $S$.  This
+provides enough resources for any run of the program $S$.  This
 setting is actually the main use case of the previous theorem

### 2014-07-01 paper/main.tex
some updates in section 5

@@
 \section{Syntax and Semantics}
+\label{sec:sem}
 
@@
 
-\paragraph{Syntax}
+\paragraph{Syntax.}
 
@@
 
-\paragraph{Semantics}
+\paragraph{Semantics.}
 
@@
 We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
-$\lambda \state . \, \infty$.  To use Coq's support for propositional
+$\lambda \state . \, \infty$.  We sometimes call assertions
+\emph{potential functions}.  To use Coq's support for propositional
 reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
@@
 
-\paragraph{Rules of the Quantitative Logic}
+\paragraph{Rules of the Quantitative Logic.}
 
@@
 
-\paragraph{Logical Variables and Shallow Embedding}
+\paragraph{Logical Variables and Shallow Embedding.}
 
@@
 
-In this section we present the deductive system used in our practical
-implementation to derive linear bounds on C programs.
+In this section we present the automatic amortized analysis that we
+use in our implementation to derive linear resource bounds for C
+programs.
 
@@
 
-To find derivations in our quantitative logic automatically, the main
-insight is to consider only assertions that match a certain shape.
-The general form of a potential function that we consider is
+To find derivations in the quantitative Hoare logic automatically, we
+have to focus on bounds that have a certain shape.  The general form
+of \emph{potential functions} (or assertions) that we consider is
 $$
   \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
-    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}|.
-$$
-
-To simplify the references to the linear coefficients $q$, we introduce
-an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
-\{(x, y) \mid x, y \in \Var \}$.  Each index $i$ corresponds to a \emph{base
-function} $f_i$ in the potential function: $0$ corresponds to the constant
-function $\_ \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
-|\inter {\state(x)} {\state(y)}|$.  Using these notations we can rewrite the above
-equality as
+    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
+$$
+%
+Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
+is again a simplified program state as introduced in \pref{sec:sem},
+$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
+references to the linear coefficients $q_i$, we introduce an \emph{index
+  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
+\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
+$f_i$ in the potential function: $0$ corresponds to the constant
+function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
+|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
+rewrite the above equality as
 $$
@@
 $$
-We will often write $xy$ to denote the index $(x,y)$.  The family
-$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense)
-for potential functions, this allows us to represent uniquely any
-linear potential function $\Phi$ as a \emph{quantitative annotation}
-$Q = (q_i)_{i \in I}$, that is a family of rational numbers where
-only a finite number of elements are not zero.
+We often write $xy$ to denote the index $(x,y)$.  The family
+$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense) for
+potential functions.  This allows us to uniquely represent any linear
+potential function $\Phi$ as a \emph{quantitative annotation} $Q =
+(q_i)_{i \in I}$, that is, a family of rational numbers where only a
+finite number of elements are not zero.
 %
@@
 
+In the potential functions, we treat constants as global variables
+that can not be assigned to.  For example, if the program contains the
+constant $8128$ then we have a variable $c_{8128}$ and
+$\state(c_{8128}) = 8128$.  We assume that every program state
+includes the constant $c_0$.
+
 \paragraph{Logical State.}
 
-Quantitative annotations alone are not enough to make our system
-functional.  A minimal logical state needs to be maintained to
-justify certain operations on quantitative annotations.  For
-example when analyzing the code $x \gets x + y$, it is helpful
-to know the sign of $y$ to determine which intervals will increase
-or decrease.  The knowledge needed by our rules can often be
-inferred by local reasoning within usual theories (e.g. Presburger
-arithmetic or bit vectors).
-
+In addition to the quantitative annotations our automatic amortized
+analysis needs to maintain a minimal logical state to justify certain
+operations on quantitative annotations.  For example when analyzing
+the code $x \gets x + y$, it is helpful to know the sign of $y$ to
+determine which intervals will increase or decrease.  The knowledge
+needed by our rules can often be inferred by local reasoning (i.e., in
+basic blocks without recursion and loops) within usual theories
+(e.g. Presburger arithmetic or bit vectors).
+
+In contrast to the quantitative annotations, logical contexts $\Gamma$
+in the presented inference system are left abstract.  This allows for
+simpler rules and leaves room for future improvements.
+%
 Our implementation uses conjunctions of linear inequalities as
@@
 
-In contrast to the quantitative annotations, logical contexts in
-the presented system are left abstract.  This allows for simpler
-rules and leaves room for numerous improvements on top of our
-minimal implementation.
-
-\paragraph{Deduction Rules.}
+\paragraph{Inference Rules.}
 

### 2014-07-01 src/ceval/speed_pldi09_fig1.c
prepare for the evaluation

@@ -1,0 +1,14 @@
+void main(int n)
+{
+  int x=0, y=0;
+
+  for (;;) {
+    if (x < n) {
+      y=y+1;
+      x=x+1;
+    } else if (y > 0)
+      y=y-1;
+    else
+      break;
+  }
+}

### 2014-07-01 src/ceval/speed_pldi09_fig4_2.c

@@ -1,0 +1,18 @@
+void main(int n, int m)
+{
+  int va, vb;
+/* assert(n > 0); */
+  assert(m > 0);
+  va = n;
+  vb = 0;
+
+  while (va > 0) {
+    if (vb < m) { 
+      vb=vb+1; 
+      va=va-1;
+    } else {
+      vb=vb-1;
+      vb=0;
+    }
+  }
+}

### 2014-07-01 src/ceval/speed_pldi09_fig4_4.c

@@ -1,0 +1,13 @@
+void main(int n, int m)
+{
+  int i;
+  assert(0 < m);
+  i = n;
+
+  while (i > 0) {
+    if (i < m)
+      i=i-1;
+    else
+      i=i-m;
+  }
+}

### 2014-07-01 src/ceval/speed_pldi09_fig4_5.c

@@ -1,0 +1,16 @@
+void main(int n, int m, int dir)
+{
+  int i;
+
+  assert(0 < m);
+  assert(m < n);
+
+  i = m;
+
+  while (0 < i && i < n) {
+    if (dir == 1)
+      i++;
+    else
+      i--;
+  }
+}

### 2014-07-01 src/ceval/speed_pldi10_ex1.c

@@ -1,0 +1,20 @@
+int nondet(void);
+
+void main(int n)
+{
+  int i,j;
+
+  i=0;
+  while (i<n) {
+    j=i+1;
+    while (j<n) {
+      if (nondet()) {
+        tick(1);
+        j=j-1;
+        n=n-1;
+      }
+      j=j+1;
+    }
+    i=i+1;
+  }
+}

### 2014-07-01 src/ceval/speed_pldi10_ex3.c

@@ -1,0 +1,12 @@
+int nondet(void);
+
+void main(int n)
+{
+  while (n>0) {
+    n=n-1;
+    while (n>0) {
+      if (nondet()) break;
+      n=n-1;
+    }
+  }
+}

### 2014-07-01 src/ceval/speed_pldi10_ex4.c

@@ -1,0 +1,12 @@
+void main(int n)
+{
+  int flag=1;
+
+  while (flag>0) {
+    flag=0;
+    while (n>0) {
+      n=n-1;       
+      flag=1;
+    }
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_fig2_1.c

@@ -1,0 +1,9 @@
+void main(int x, int y, int n, int m)
+{
+  while (n>x) {
+    if (m>y)
+      y = y+1;
+    else
+      x = x+1;
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_fig2_2.c

@@ -1,0 +1,9 @@
+void main(int x, int z, int n)
+{
+  while (x<n) {
+    if (z>x)
+      x=x+1;
+    else
+      z=z+1;
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_nested_multiple.c

@@ -1,0 +1,13 @@
+int nondet(void);
+
+void main(int x, int n, int y, int m)
+{
+  while (x<n) {
+    while (y<m) {
+      if (nondet())
+        break;
+      y=y+1;
+    }
+    x=x+1;
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_nested_single.c

@@ -1,0 +1,15 @@
+int nondet(void);
+
+void main(int n)
+{
+  int x=0;
+
+  while (x<n) {
+    x=x+1;
+    while (x<n) {
+      if (nondet())
+        break;
+      x=x+1;
+    }
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_sequential_single.c

@@ -1,0 +1,13 @@
+int nondet(void);
+
+void main(int n)
+{
+  int x=0;
+
+  while (x<n) {
+    if (nondet()) break;
+    x=x+1;
+  }
+  while (x<n)
+    x=x+1;
+}

### 2014-07-01 src/ceval/speed_popl10_simple_multiple.c

@@ -1,0 +1,11 @@
+void main(int n, int m)
+{
+  int x=0, y=0;
+
+  while (x<n) {
+    if (y<m)
+      y=y+1;
+    else
+      x=x+1;
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_simple_single.c

@@ -1,0 +1,13 @@
+int nondet(void);
+
+void main(int n)
+{
+  int x=0;
+
+  while (x<n) {
+    if (nondet())
+      x=x+1;
+    else 
+      x=x+1;
+  }
+}

### 2014-07-01 src/ceval/speed_popl10_simple_single_2.c

@@ -1,0 +1,15 @@
+void main(int n, int m)
+{
+  int x=0, y=0;
+
+  for (;;) {
+    if (x<n) {
+      x=x+1; y=y+1;
+    }
+    else if (y<m) {
+      x=x+1; y=y+1;
+    }
+    else
+      break;
+  }
+}

### 2014-07-01 src/ceval/t07.c

@@ -1,0 +1,16 @@
+void fish() { return; }
+
+void main(int x, int y)
+{
+  while (x>0) {
+    x=x-1;
+    y=y+2;
+  }
+  while (y>0) {
+    y=y-1;
+  }
+  fish();
+  while (y>0) {
+    y=y-1;
+  }
+}

### 2014-07-01 src/ceval/t08.c

@@ -1,0 +1,9 @@
+void main(int y, int z)
+{
+  while (z > y) {
+    y = y + 1;
+  }
+  while (y > 9) {
+    y = y - 10;
+  }
+}

### 2014-07-01 src/ceval/t10.c

@@ -1,0 +1,11 @@
+int nondet(void);
+
+void main(int x, int y)
+{
+  while (x>y) {
+    if (nondet())
+      y=y+1;
+    else
+      x=x-1;
+  }
+}

### 2014-07-01 src/ceval/t11.c

@@ -1,0 +1,9 @@
+void main(int x, int y, int n, int m)
+{
+  while (n>x) {
+    if (m>y) 
+      y = y+1;
+    else
+      x = x+1;
+  }
+}

### 2014-07-01 src/ceval/t13.c

@@ -1,0 +1,13 @@
+int nondet(void);
+
+void main(int x, int y)
+{
+  while (x > 0) {
+    x = x-1;
+    if (nondet()) 
+      y = y+1;
+    else
+      while (y > 0)
+        y = y-1;
+  }
+}

### 2014-07-01 src/ceval/t15.c

@@ -1,0 +1,9 @@
+void main(int x, int y, int z)
+{
+  assert(y >= 0);
+  while (x>y) {
+    x-=y+1;
+    for (z=y; z>0; z--)
+      /* nothing */;
+  }
+}

### 2014-07-01 src/ceval/t16.c

@@ -1,0 +1,10 @@
+void main(int x, int y, int z)
+{
+  assert(y >= 0);
+  while (x > y) {
+    x -= y+1;
+    z = 100 + 2*y;
+    while (z > 0)
+      z--;
+  }
+}

### 2014-07-01 src/ceval/t19.c

@@ -1,0 +1,8 @@
+void main(int i, int k)
+{
+  while (i > 100)
+    i--;
+  i=i+k+50;
+  while (i >= 0)
+    i--;
+}

### 2014-07-01 src/ceval/t20.c

@@ -1,0 +1,7 @@
+void main(int x, int y)
+{
+  while (x < y)
+    x++;
+  while (y < x)
+    y++;
+}

### 2014-07-01 src/ceval/t27.c

@@ -1,0 +1,9 @@
+void main(int n, int y)
+{
+  while (n<0) {
+    n++;
+    y+=1000;
+    while (y>=100)
+      y-=100;
+  }
+}

### 2014-07-01 src/ceval/t28.c

@@ -1,0 +1,13 @@
+void main(int x, int y)
+{
+  while (x > y) {
+    x--;
+    x += 1000;
+    y += 1000;
+  }
+
+  while (y>0)
+    y--;
+  while (x<0)
+    x++;
+}

### 2014-07-01 src/ceval/t30.c

@@ -1,0 +1,9 @@
+void main(int x, int y)
+{
+  int t;
+
+  while (x>0) {
+    x--;
+    t=x, x=y, y=t;
+  }
+}

### 2014-07-01 src/ceval/t37.c

@@ -1,0 +1,20 @@
+void count_down(int x) {
+  if (x>0) {
+    x--;
+    count_down(x);
+  }
+}
+
+int copy(int x, int y) {
+  if (x>0) {
+    y++;
+    x--;
+    y=copy(x, y);
+  }
+  return y;
+}
+
+void main(int x, int y) {
+  y = copy(x, y);
+  count_down(y);
+}

### 2014-07-01 src/ceval/t39.c

@@ -1,0 +1,21 @@
+void count_up(int, int);
+
+void count_down(int x,int y)
+{
+  if (x>y) {
+    x--;
+    count_up(x, y);
+  }
+}
+
+void count_up(int x, int y)
+{
+  if (y+1<x) {
+    y += 2;
+    count_down(x, y);
+  }
+}
+
+void main(int y, int z) {
+  count_down(y, z);
+}

### 2014-07-01 src/ceval/t46.c

@@ -1,0 +1,17 @@
+int x, y;
+
+void produce() {
+  while (x > 0) {
+    tick(-1); x--; y++;
+  }
+}
+
+void consume() {
+  while (y > 0) {
+    y--; x++; tick(1);
+  }
+}
+
+void main(int y, int z) {
+  consume(); produce(); consume();
+}

### 2014-07-01 src/ceval/t47.c

@@ -1,0 +1,12 @@
+void main(int n)
+{
+  int flag = 1;
+
+  while (flag>0) {
+    if (n>0) {
+      n--;
+      flag=1;
+    } else
+      flag=0;
+  }
+}

### 2014-07-01 src/ctest/t08.c

@@ -1,13 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int y,z;
-
-  while (z-y > 0) {
-    y = y + 1;
-  }
-  while (y > 9) {
-    y = y - 10;
-  }
-
-}

### 2014-07-01 src/ctest/t10.c

@@ -1,12 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,y;
-  while (x-y>0) {
-    if (1)
-      y=y+1;
-    else
-      x=x-1;
-  }
-
-}

### 2014-07-01 src/ctest/t11.c

@@ -1,12 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,y,n,m;
-
-  while (n>x) {
-    if (m>y) 
-      y = y+1;
-    else
-      x = x+1;
-  }
-}

### 2014-07-01 src/ctest/t13.c

@@ -1,16 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,y;
-
-  while (x > 0) {
-    x = x -1;
-    if (1) 
-      y = y +1;
-    else {
-      while (y > 0)
-	y = y -1;
-    }
-  }
-
-}

### 2014-07-01 src/ctest/t16.c

@@ -1,17 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,y,z;
-
-  //assert (y>=0);
-  while (x-y>0) {
-    x=x-1;
-    x=x-y;
-    z=y;
-    z=z+y;
-    z=z+100;
-    while (z>0) {
-      z=z-1;
-    }
-  }
-}

### 2014-07-01 src/ctest/t20.c

@@ -1,14 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,y;
-
-  while (x<y) {
-    x=x+1;
-  }
-
-  while (y<x) {
-    y=y+1;
-  }
-
-}

### 2014-07-01 src/ctest/t27.c

@@ -1,14 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int n,y;
-
-  while (n<0) {
-    n=n+1;
-    y=y+1000;
-    while (y>=100) {
-      y=y-100;
-    }
-  }
-
-}

### 2014-07-01 src/ctest/t30.c

@@ -1,12 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,y,t;
-
-  while (x>0) {
-    x=x-1;
-    t=x;
-    x=y;
-    y=t;
-  }
-}

### 2014-07-01 src/ctest/t39.c

@@ -1,30 +1,0 @@
-
-void count_up (int,int);
-void count_down (int,int);
-
-void count_down (int x,int y)
-{ int a = x;
-  if (a>y) {
-    a = a-1;
-    count_up(a,y);
-  }
-}
-
-void count_up (int x, int y)
-{ int a = y;
-  if (a+1<x) {
-    a = a+2;
-    count_down(x,a);
-  }
-}
-
-void _main (int y, int z) {
-  count_down(y,z);
-}
-
-int main(int argc, char **argv)
-{
-  int x,y;
-}
-
-

### 2014-07-01 src/ctest/t47.c

@@ -1,17 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int n,flag;
-
-  flag=1;
-
-  while (flag>0) {
-    if (n>0) {
-      n=n-1;
-      flag=1;
-    }
-    else
-      flag=0;
-  }
-
-}

### 2014-07-01 src/examples/speed/pldi09/fig1.c

@@ -1,16 +1,0 @@
-int main(int argc, char **argv)
-{
-  int x,y;
-
-  x=0;y=0;
-
-  while (1>0) {
-  if x < n (
-    y=y+1;
-    x=x+1
-  ) else if y>0
-    y=y-1
-  else
-    break
-)
-

### 2014-07-01 src/examples/speed/pldi09/fig4_2.c

@@ -1,20 +1,0 @@
-int main(int argc, char **argv)
-{
-  int va,vb,n,m;
-
-/* assert n>0; */
-/* assert m>0; */
-  va = n;
-  vb = 0;
-
-  while (va>0) {
-    if (vb<m) { 
-      vb=vb+1; 
-      va=va-1;
-    }
-    else {
-      vb=vb-1;
-      vb=0;
-    }
-  }
-}

### 2014-07-01 src/examples/speed/pldi09/fig4_4.c

@@ -1,15 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int i,n,m;
-
-  /* assert 0<m; */
-  i = n;
-
-  while (i>0) {
-    if (i<m)
-      i=i-1;
-    else
-      i=i-m;
-  }
-}

### 2014-07-01 src/examples/speed/pldi09/fig4_5.c

@@ -1,14 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int i,n,m,dir,fwd;
-
-  /* assert (0 < m < n);  */
-  i=m;
-
-  while (0<i<n) {
-    if (dir==fwd) i++;
-    else i--;
-  }
-
-}

### 2014-07-01 src/examples/speed/pldi10/ex1.c

@@ -1,19 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int i,j,n;
-
-  i=0;
-  while (i<n) {
-    j=i+1;
-    while (j<n) {
-      if (0) {
-	_tick(1);
-	j=j-1;
-	n=n-1;
-      }
-      j=j+1;
-    }
-    i=i+1;
-  }
-}

### 2014-07-01 src/examples/speed/pldi10/ex3.c

@@ -1,13 +1,0 @@
-int main(int argc, char **argv)
-{
-  int t,x,n;
-
-  while (n>0) {
-    t = x;
-    n=n-1;
-    while (n>0) {
-      if (0) break;
-      n=n-1;
-    }
-  }
-}

### 2014-07-01 src/examples/speed/popl10/fig2_1.c

@@ -1,12 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int n,m,x,y;
-
-  while (n>x) {
-    if (m>y)
-      y = y+1;
-    else
-      x = x+1;
-  }
-}

### 2014-07-01 src/examples/speed/popl10/fig2_2.c

@@ -1,12 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,z,n;
-
-  while (x<n) {
-    if (z>x)
-      x=x+1;
-    else
-      z=z+1;
-  }
-}

### 2014-07-01 src/examples/speed/popl10/simple_single.c

@@ -1,14 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,n,y,m;
-
-  x=0;
-  while (x<n) {
-    if (0)
-      x=x+1;
-    else 
-      x=x+1;
-  }
-
-}

### 2014-07-01 src/examples/speed/popl10/simple_single_2.c

@@ -1,19 +1,0 @@
-
-int main(int argc, char **argv)
-{
-  int x,N,y,M;
-
-  x=0;
-  y=0;
-  while (1) {
-    if (x<N) {
-      x=x+1; y=y+1;
-    }
-    else if (y<M ) {
-      x=x+1; y=y+1;
-    }
-  else
-    break;
-  }
-
-}

### 2014-07-01 src/ceval/t15.c
differenciate locals and formals in t1[56]

@@ -1,5 +1,7 @@
-void main(int x, int y, int z)
+void main(int x, int y)
 {
+  int z;
+
   assert(y >= 0);
   while (x>y) {
     x-=y+1;

### 2014-07-01 src/ceval/t16.c

@@ -1,5 +1,7 @@
-void main(int x, int y, int z)
+void main(int x, int y)
 {
+  int z;
+
   assert(y >= 0);
   while (x > y) {
     x -= y+1;

### 2014-07-01 src/ceval/qsort.c
normalize tests for rank

@@ -1,6 +1,6 @@
 /* Quicksort */
 
-int nondet(void);
+int nondet();
 
 void qsort(int *a, int lo, int hi) {
   int m1, m2, n;

### 2014-07-01 src/ceval/speed_pldi10_ex1.c

@@ -1,8 +1,9 @@
-int nondet(void);
+int nondet();
 
 void start(int n)
 {
-  int i,j;
+  int i;
+  int j;
 
   i=0;
   while (i<n) {

### 2014-07-01 src/ceval/speed_pldi10_ex3.c

@@ -1,4 +1,4 @@
-int nondet(void);
+int nondet();
 
 void start(int n)
 {

### 2014-07-01 src/ceval/speed_popl10_simple_multiple.c

@@ -1,6 +1,7 @@
 void start(int n, int m)
 {
-  int x=0, y=0;
+  int x=0;
+  int y=0;
 
   while (x<n) {
     if (y<m)

### 2014-07-01 src/ceval/speed_popl10_simple_single_2.c

@@ -1,6 +1,7 @@
 void start(int n, int m)
 {
-  int x=0, y=0;
+  int x=0;
+  int y=0;
 
   for (;;) {
     if (x<n) {

### 2014-07-01 src/ceval/t10.c

@@ -1,4 +1,4 @@
-int nondet(void);
+int nondet();
 
 void start(int x, int y)
 {

### 2014-07-01 src/ceval/t13.c

@@ -1,13 +1,13 @@
-int nondet(void);
+int nondet();
 
 void start(int x, int y)
 {
   while (x > 0) {
-    x = x-1;
+    x=x-1;
     if (nondet()) 
-      y = y+1;
+      y=y+1;
     else
       while (y > 0)
-        y = y-1;
+        y=y-1;
   }
 }

### 2014-07-01 paper/main.tex
some fixes in the rules
add experiments table
this and that

@@
 
-  The practicability of the analysis framework is demonstrated with an
-  experimental evaluation.  The expressivity of the logic is
-  demonstrated by manually deriving customized bounds that are
-  tailored to specific algorithms.  A comparison of the automatic
-  amortized analysis with other automatic tools on 30 challenging
-  examples from the literature and open-source software shows that the
-  derived bounds are often more precise.
+  An experimental evaluation demonstrates the practicability of the
+  analysis framework.  The expressivity of the logic is shown by
+  manually deriving customized bounds that are tailored to specific
+  algorithms.  A comparison of the automatic amortized analysis with
+  other automatic tools on 30 challenging examples from the literature
+  and open-source software shows that the bounds derived by the
+  automatic amortized analysis are often more precise.
 \end{abstract}
@@
 $$
-  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
+  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
     q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
@@
 
-\paragraph{Inference Rules.}
-
-The deductive system defined in~\pref{fig:auto} derives
-judgements of the form
+\paragraph{Judgements of the Automatic Analysis.}
+
+The inference system for the automatic amortized analysis is defined
+in~\pref{fig:auto}.  The inference rules derive judgements of the form
 $$
@@
 $$
-These judgements are analogous to the logic judgements
-$\Delta; B; R \vdash \htriple P S Q$ where each assertion
-splits in two parts, a logical part $\Gamma$ and a quantitative
-part $Q$.  We left the function context $\Delta$ appart in
-the presentation to avoid cluttering notations, but it is
-implicit on all judgements.  We call precondition the pair
-$(\Gamma, Q)$ and postcondition the pair $(\Gamma', Q')$.
-The correspondence between judgements of the automation
-and the ones of the quantitative logic is made formal by our
-soundness proof of the analysis.
-
-Since the different rules of~\pref{fig:auto} manage the
-logical state in exactly the same way as the quantitative
-logic, we will focus on the quantitative annotations in the
-sequel.
-%
-As a convention, if $Q$ is a quantitative annotation we
-will write $q_i$ for its $i$th coordinate.  The notation
-$Q \pm n$ used in many rules defines a new context $Q'$
-such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\,
-q'_i = q_i$.  Finally, if a rule mentions $Q$ and $Q'$
-and leaves the latter undefined at some index $i$ we
-will implicitely assume that $q'_i = q_i$.
-
-%
-% Next paragraph is shit.
-%
-
-The {\sc Q:Skip} rule simply reuses its postcondition as
-precondition.  This is a sound choice since evaluating
-a $\code{skip}$ statement never costs any resource and
-does not change the program state.  A similar rule is
-the {\sc Q:Tick} rule which enforces that the quantitative
-annotation of the precondition is at least $M_t(n)$ to
-pay for the resource consumption.  Since the program
-state is also unchanged in this case, the logical part
-$\Gamma$ of the precondition can soundly be reused in
-the postcondition.  A slightly different rule is
-{\sc Q:Break}, in that case the regular control flow is
-broken, so anything that follows \code{break} in the
-program will not be executed.  Thus, the postcondition
-$(\Gamma', Q')$ can be arbitrary, and the precondition
-is taken from the break part of the judgement similarly
-to what is done in the logic.
-
-From these three examples alone, one can see that our
-automation follows very closely the rules of the logic.
-The case of the update statement is more interesting
-and specific to linear potential functions.
-In the {\sc Q:IncP} rule the program updates a variable
-$x$ to add it $y \ge 0$.  Since $x$ is changed, the
-quantitative annotation must be updated to reflect the
-change of the program state.  We write $x'$ the value
-of $x$ after the assignment ($x + y$). Since $x$ is
-the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized.
-Note that for any $u$, $\inter x u$ will get smaller
-with the update,  and if $x' \in \inter x u$ 
-we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.
-But $|\inter x {x'}| = |\inter 0 y|$, this means that
-the potential $q'_{0y}$ in the postcondition can
-be increased by $q_{xu}$ under the guard that
-$x' \in \inter x u$.  Dually, the interval
-$\inter v x$ can get bigger with the update.
-And we know that $|\inter v {x'}| \le y + |\inter v x|$,
-so we decrease the potential of $\inter 0 y$
-by $q_{vx}$ to pay for this change.  Remark that
-the rule forces it only for $v \not\in \mathcal U$,
-because if not, we know that $x \le v$ so
-$|\inter v x| = 0$.
+These judgements correspond to the logic judgements $\Delta; B; R
+\vdash \htriple P S Q$ where each assertion splits in two parts, a
+logical part $\Gamma$ and a quantitative part $Q$.  That means that
+$(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
+$(\Gamma_R, Q_R)$ is the postcondition for return statements, and
+$(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
+simple Hoare triple.  We leave out the function context $\Delta$ in
+the presentation to avoid cluttering notations.  Instead we assume a
+fixed function context $\Delta$ that is implicit on all judgements.
+Function contexts are treated exactly as in the logic.  We call
+precondition the pair $(\Gamma, Q)$ and postcondition the pair
+$(\Gamma', Q')$.  The correspondence between judgements of the
+automation and the ones of the quantitative logic is made formal by
+our soundness proof of the automatic amortized analysis at the end of
+this section.
+
+As a convention, if $Q$ and $Q'$ are quantitative annotations then we
+assume that $Q = (q_i)_{i\in I}$ is a family with elements $q_i$, $Q'
+= (q'_i)_{i \in I}$, etc.  The notation $Q \pm n$ used in many rules
+defines a new context $Q'$ such that $q'_0 = q_0 \pm n$ and $\forall i
+\neq 0 .\, q'_i = q_i$.  We have the implicit side condition that all
+rational coefficients are non-negative.  Finally, if a rule mentions
+$Q$ and $Q'$ and leaves the latter undefined at some index $i$ we will
+implicitly assume that $q'_i = q_i$.  As mentioned, we leave the
+logical contexts $\Gamma$ abstract.
+
+We describe the automatic amortized analysis for a subset of
+expressions of C light.  Assignments must have the form \code{x \gets
+  y} or \code{x \gets x \pm y}.  In the implementation, a Clight
+program is converted into this form prior to analysis without changing
+the resource cost.  This achieved by using a series of \emph{cost-free
+  assignments} that we do not result in additional cost in the
+analysis.  Non-linear operations such as \code{x \gets z*y} are simply
+handled by assigning zero potential to coefficients like $q_{xa}$ and
+$q_{ax}$ that contain $x$.
+
+\paragraph{Inference Rules.}
+
+Most of the rules correspond exactly to the respective rules of the
+quantitative Hoare logic.  For example, the rule {\sc Q:Skip} simply
+reuses its postcondition as precondition.  This is sound since
+evaluating a $\code{skip}$ statement never costs any resources and
+does not change the program state.  Similarly, the rule {\sc Q:Tick}
+enforces that the quantitative annotation of the precondition is at
+least $M_t(n)$ to pay for the resource consumption.  Since the program
+state is also unchanged in this case, the logical part $\Gamma$ of the
+precondition can soundly be reused in the postcondition.  In the rule
+{\sc Q:Break}, where the linear control flow is broken, the
+postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
+taken from the break part of the judgement similarly as it is done in
+the quantitative logic.  The rule {\sc Q:Return} exhibits a slight
+difference to the quantitative logic.  While we have a function $\Z
+\to \Assn$ to represent return values in the logic, we assume that a
+special variable $\ret$ is part of the index set of $Q_R$.  The
+substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
+$q_{\ret y} = q'_{x y}$ and $q_{y \ret} = q'_{y x}$ for all $y$ and
+$q_i'=q_i$ otherwise.
+
+The most interesting rules are {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
+  Q:Inc}.
+
+The case of the update statement is more interesting and specific to
+linear potential functions.  In the {\sc Q:IncP} rule the program
+updates a variable $x$ to add it $y \ge 0$.  Since $x$ is changed, the
+quantitative annotation must be updated to reflect the change of the
+program state.  We write $x'$ the value of $x$ after the assignment
+($x + y$). Since $x$ is the only variable changed, only intervals of
+the form $\inter u x$ and $\inter x u$ will be resized.  Note that for
+any $u$, $\inter x u$ will get smaller with the update, and if $x' \in
+\inter x u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'}
+u|$.  But $|\inter x {x'}| = |\inter 0 y|$, this means that the
+potential $q'_{0y}$ in the postcondition can be increased by $q_{xu}$
+under the guard that $x' \in \inter x u$.  Dually, the interval
+$\inter v x$ can get bigger with the update.  And we know that
+$|\inter v {x'}| \le y + |\inter v x|$, so we decrease the potential
+of $\inter 0 y$ by $q_{vx}$ to pay for this change.  Remark that the
+rule forces it only for $v \not\in \mathcal U$, because if not, we
+know that $x \le v$ so $|\inter v x| = 0$.
 
@@
 
+All rules except of {\sc Q:Weak} and {\sc Q:Relax} are syntax
+directed.
+
 \paragraph{Soundness Proof.}
@@
 }
-{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
+{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
 %
@@
 %
-\and \Rule{Q:IncM}
+\and \Rule{Q:DecP}
 { \Gamma \models 0 \le y
@@
 \and \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
-\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
-\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
+{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
+\and \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
 \\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
-}
-{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
-%
-\and \Rule{Extend}
+\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
+\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
+}
+{ Q'   \succeq_\Gamma Q }
+%
+\and \Rule{Q:Extend}
 { \fenv f = (\vec y, S_f)
-\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
+\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
 }
@@
 
+\begin{figure*}
+\tiny
+
+\begin{tabular}{r|p{3cm}c|p{3cm}c|c}
+
+File &
+\multicolumn{2}{|c|}{Rank} &
+\multicolumn{2}{|c|}{Loopus} &
+Amortized Automation
+\\
+\hline
+
+%--------------------------------------------
+\texttt{gcd.c} &
+
+$((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
+$O(n)$ &
+
+Fail &
+&
+
+$|\inter 0 x| + |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{kmp.c} &
+
+$((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
+$O(n^2)$ &
+
+$\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
+$O(n)$ &
+
+$1 + 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{qsort.c} &
+
+Fail &
+&
+
+Fail &
+&
+
+$1 + 2 |\inter 0 {\code{len}}|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi09\_fig1.c} &
+
+Fail &
+&
+
+$\max(n, 0) + \max(n, 0)$ &
+$O(n)$ &
+
+$1 + 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi09\_fig4\_2.c} &
+
+$((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
+$O(n)$ &
+
+Fail &
+&
+
+$1 + 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi09\_fig4\_4.c} &
+
+$((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
+$O(n)$ &
+
+Fail &
+&
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi09\_fig4\_5.c} &
+
+$((((1+1)+(-1+n))+m)+1)+1$ &
+$O(n)$ &
+
+Fail &
+&
+
+Fail
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi10\_ex1.c} &
+
+$(((1+1)+n)+((-1+((1/2){\cdot}n))+((1/2){\cdot}(n{\cdot}n))))+1$ &
+$O(n^2)$ &
+
+$\max(n, 0) + \max(0,  \max(n, 0) {\cdot} \max(0, (n + -1)))$ &
+$O(n^2)$ &
+
+Fail
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi10\_ex3.c} &
+
+$((((1+1)+(-1+n))+(-1+n))+1)+1$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(n, 0)$ &
+$O(n)$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_pldi10\_ex4.c} &
+
+Fail &
+&
+
+Fail &
+&
+
+$1 + 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_fig2\_1.c} &
+
+$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$O(n)$ &
+
+$\max(0, (n - x)) + \max(0, (m - y))$ &
+$O(n)$ &
+
+$|\inter x n| + |\inter y m|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_fig2\_2.c} &
+
+$(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
+$O(n)$ &
+
+$\max(0, (\max(0, (x + 1 + -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
+$O(n)$ &
+
+$|\inter x n| + |\inter z n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_nested\_multiple.c} &
+
+$(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
+$O(n^2)$ &
+
+$\max(0, (m - y)) + \max(0, (n - x))$ &
+$O(n)$ &
+
+$|\inter x n| + |\inter y m|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_nested\_single.c} &
+
+$((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
+$O(n)$ &
+
+$\max(0, min(n, (n + -1))) + \max(0, min(n, (n + -1)))$ &
+$O(n)$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_sequential\_single.c} &
+
+$(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(n, 0)$ &
+$O(n)$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_simple\_multiple.c} &
+
+$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$|\inter 0 m| + |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_simple\_single\_2.c} &
+
+Fail &
+&
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$|\inter 0 n| + |\inter 0 m|$
+\\
+
+%--------------------------------------------
+\texttt{speed\_popl10\_simple\_single.c} &
+
+$(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$O(n)$ &
+
+$\max(n, 0)$ &
+$O(n)$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\texttt{t07.c} &
+
+$(((1+1)+-1)+x)+1$ &
+$O(n)$ &
+
+$\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
+$O(n)$ &
+
+$1 + 3 |\inter 0 x| + |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{t08.c} &
+
+$(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
+$O(n)$ &
+
+$\max(0, (\max(0, (y + -9)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+$O(n)$ &
+
+$1.1 |\inter y z| + 0.1 |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{t10.c} &
+
+$(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
+$O(n)$ &
+
+$\max(0, (x - y))$ &
+$O(n)$ &
+
+$|\inter y x|$
+\\
+
+%--------------------------------------------
+\texttt{t11.c} &
+
+$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$O(n)$ &
+
+$\max(0, (n - x)) + \max(0, (m - y))$ &
+$O(n)$ &
+
+$|\inter x n| + |\inter y m|$
+\\
+
+%--------------------------------------------
+\texttt{t13.c} &
+
+$((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
+$O(n^2)$ &
+
+$\max(x, 0) + \max(0, (\max(x, 0) + \max(y, 0)))$ &
+$O(n)$ &
+
+$2 |\inter 0 x| + |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{t15.c} &
+
+$(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
+$O(n)$ &
+
+Fail &
+&
+
+$|\inter 0 x|$
+\\
+
+%--------------------------------------------
+\texttt{t16.c} &
+
+$(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
+$O(n)$ &
+
+Fail &
+&
+
+$101 |\inter 0 x|$
+\\
+
+%--------------------------------------------
+\texttt{t19.c} &
+
+$(((1+1)+(151+k))+(-100+i))+1$ &
+$O(n)$ &
+
+$\max(0, (i + -100)) + \max(0, (k + i + 51))$ &
+$O(n)$ &
+
+$50 + |\inter {-1} i| + |\inter 0 k|$
+\\
+
+%--------------------------------------------
+\texttt{t20.c} &
+
+$((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
+$O(n)$ &
+
+$\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
+$O(n)$ &
+
+$|\inter x y| + |\inter y x|$
+\\
+
+%--------------------------------------------
+\texttt{t27.c} &
+
+Fail &
+&
+
+$\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
+$O(n)$ &
+
+$0.1|\inter n y| + 11 |\inter n 0|$
+\\
+
+%--------------------------------------------
+\texttt{t28.c} &
+
+$(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
+$O(n)$ &
+
+$\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
+$O(n)$ &
+
+$|\inter x 0| + |\inter 0 y| + 1002 |\inter y x|$
+\\
+
+%--------------------------------------------
+\texttt{t30.c} &
+
+$((1+1)+x)+1$ &
+$O(n)$ &
+
+$\max(0, (1 + \max(x, y, 0)))$ &
+$O(n)$ &
+
+$|\inter 0 x| + |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{t37.c} &
+
+Fail &
+&
+
+Fail &
+&
+
+$3 + 2 |\inter 0 x| + |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{t39.c} &
+
+Fail &
+&
+
+Fail &
+&
+
+$1.33 + 0.67 |\inter z y|$
+\\
+
+%--------------------------------------------
+\texttt{t46.c} &
+
+Fail &
+&
+
+Fail &
+&
+
+$|\inter 0 y|$
+\\
+
+%--------------------------------------------
+\texttt{t47.c} &
+
+$(((1+1)+n)+1)+1$ &
+$O(n)$ &
+
+$1 + \max(n, 0)$ &
+$O(n)$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\end{tabular}
+\caption{That big ugly table}
+\label{fig:eval}
+\end{figure*}
+
 \bibliographystyle{abbrvnat}

### 2014-07-02 paper/main.tex
bugfix in the experiments
add speed results
more meaningless results
more work on section 5

@@
 
-\section{Automatic Analysis}
+\section{Automatic Amortized Analysis}
 
@@
 
-The most interesting rules are {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
-  Q:Inc}.
-
-The case of the update statement is more interesting and specific to
-linear potential functions.  In the {\sc Q:IncP} rule the program
-updates a variable $x$ to add it $y \ge 0$.  Since $x$ is changed, the
+Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
+  Q:Inc} for increments and decrements, which describe how the
+potential is distributed after a size change of the variables.  The
+rule {\sc Q:IncP} is for increments $x \gets x + y$ in which we can
+deduce from the logical context $\Gamma$ that $y$ is not negative.
+Similarly, {\sc Q:DecP} is for increments $x \gets x - y$ in which we
+can deduce that $y \ge 0$.  Of course, we have symmetrical rules {\sc
+  Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
+The rules are equivalent in the case where $y=0$.  Finally, the {\sc
+  Q:Inc} can be applied if we cannot deduce any information about $y$.
+In the implementation, the rules for increments and decrements are
+combined into one syntax directed rule where the specialized rules
+take priority over {\sc Q:Inc}.
+
+To explain how rules for increment and decrement work, it is
+sufficient to understand the rule {\sc Q:IncP}.  The other follow the
+same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
+variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
 quantitative annotation must be updated to reflect the change of the
-program state.  We write $x'$ the value of $x$ after the assignment
-($x + y$). Since $x$ is the only variable changed, only intervals of
-the form $\inter u x$ and $\inter x u$ will be resized.  Note that for
-any $u$, $\inter x u$ will get smaller with the update, and if $x' \in
-\inter x u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'}
-u|$.  But $|\inter x {x'}| = |\inter 0 y|$, this means that the
-potential $q'_{0y}$ in the postcondition can be increased by $q_{xu}$
-under the guard that $x' \in \inter x u$.  Dually, the interval
-$\inter v x$ can get bigger with the update.  And we know that
-$|\inter v {x'}| \le y + |\inter v x|$, so we decrease the potential
-of $\inter 0 y$ by $q_{vx}$ to pay for this change.  Remark that the
-rule forces it only for $v \not\in \mathcal U$, because if not, we
-know that $x \le v$ so $|\inter v x| = 0$.
-
-Another very technical rule is {\sc Q:Call}.  It needs to
-account for the stack ``switching'' caused by the function
-call, the argument/return value passing, and the preservation
-of local variables.  We can sum up the requirements for
-this rule as follows,
+program state.  We write $x'$ the value of $x$ after the assignment.
+Since $x$ is the only variable changed, only intervals of the form
+$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
+$\inter x u$ will get smaller with the update, and if $x' \in \inter x
+u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
+$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
+$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
+guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
+get bigger with the update.  We know that $|\inter v {x'}| \le y +
+|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
+$q_{vx}$ to pay for this change.  Remark that the rule insures this
+only for $v \not\in \mathcal U$ because we know that $x \le v$
+otherwise, and thus $|\inter v x| = 0$.
+
+Another interesting rule is {\sc Q:Call}.  It needs to account for the
+changes to the stack caused by the function call, the argument/return
+value passing, and the preservation of local variables.  We can sum up
+the main ideas of the rule as follows.
 \begin{itemize}
+\item The potential in the pre- and postconditions of the function
+  specification is equalized to its matching potential in the callee's
+  pre/postcondition.
 \item
-  any potential in the function pre/postcondition is equalized
-  to its matching potential in the call pre/postcondition,
-\item
-  the potential of intervals $|\inter x y|$ is preserved
+  The potential of intervals $|\inter x y|$ is preserved
   if $x$ and $y$ are local variables,
 \item
-  and the unknown potentials after the call (e.g.
+  The unknown potentials after the call (e.g.
   $|\inter x g|$, with $x$ local and $g$ global)
-  are zeroed out in the postcondition.
+  are set to zero in the postcondition.
 \end{itemize}
@@
 
-All rules except of {\sc Q:Weak} and {\sc Q:Relax} are syntax
-directed.
+The rule {\sc Q:Weak} is not syntax directed.  In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition.  We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the
+analysis efficient.  The high-level idea of {\sc Q:Weak} is the same
+as in the rule {\sc L:Weak} of the quantitative logic: If we have a
+sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept
+of \emph{more potential} is formalized by the relation $Q'
+\succeq_\Gamma Q$ that is defined in the rule {\sc Relax}.  Here, $Q$
+and $Q'$ are potential annotations and $\Gamma$ is a logical context.
+The relax rule deals also with the important task of transferring
+constant potential (represented by $q_0$) to interval sizes and vise
+versa.  If we can deduce from the logical context that the interval
+size $|\inter x y| \geq l$ is larger then a constant $l$ then we can
+transfer potential $q_{xy} {\cdot} |\inter x y|$ form the interval to
+constant potential $l{\cdot}q_{xy}$ and guarantee that we do not gain
+potential.  Conversely, if $|\inter x y| \leq u$ for a constant $u$
+then we can transfer constant potential $u{\cdot}q_{xy}$ to the
+interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining
+potential.
+
+\paragraph{Automatic Inference via LP Solving.}
+
+We separate the search of a derivation into two steps.  As a first
+step we go through the whole program and apply inductively the
+inference rules of the automatic amortized analysis.  During this
+process our tool uses symbolic names for the coefficients $(q_i)$.
+Every time a linear constraint must be satisfied by these
+coefficients, it is recorded using the symbolic names in a global
+list.  When the list of all constraints that must be satisfied by the
+coefficients is collected, we feed it to an off the shelf
+LP-solver~\footnote{We currently use Coin-Or CLP.}.  If the solver
+successfully finds a solution to the given problem, we know that a
+derivation exists for the considered program.  We can then extract the
+initial $Q$ from the solver and get a resource bound for the program.
+To get a full derivation we simply extract the complete solution from
+the solver, and apply it to the symbolic names $(q_i)$ of the
+coefficients in the derivation.  If the LP-solver fails to find a
+solution, an error is reported to the user.
 
@@
 
-The soundness of the analysis leverages the quantitative logic.
-We propose to translate judgements of~\pref{fig:auto} into
-quantitative triples.  We define a translation function
-$\tr$, such that if a judgement $J$ in the automatic analysis
-is derivable, $\tr(J)$ is derivable in the quantitative
-logic.  Using this translated judgement and the soundness
-of our logic it is possible to get a certified resource bound
-for the analyzed program.
+The soundness of the analysis builds on the quantitative logic.  We
+translate judgements of the automatic amortized analysis as defined in
+\pref{fig:auto} into quantitative Hoare triples.  We define a
+translation function $\tr$, such that if a judgement $J$ in the
+automatic analysis is derivable, $\tr(J)$ is derivable in the
+quantitative logic.  By using $\tr$ to translate derivations of the
+automatic analysis to derivations in the quantitative logic we can
+automatically obtain a certified resource bound for the analyzed
+program.
 
@@
 translate the assumptions in function contexts of the automatic
-analysis, this is done as follows
+analysis.  This is done as follows.
 %
@@
   \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret \mapsto v])
-  ).
+  )
 \end{align*}
@@
 \noindent
-The proof of this theorem is constructive and basically
-maps each rule of the automatic analysis directly to
-its counterpart in the quantitative logic.  With the
-computational content of the proof, we have an
-effective algorithm to construct proof objects for
-upper bounds derived automatically.
+The proof of this theorem is constructive and basically maps each rule
+of the automatic analysis directly to its counterpart in the
+quantitative logic.  The most tricky parts are the translations of the
+rules for increments and decrements and the rule {\sc Q:Weak} for
+weakening because we have to show that the preconditions of the rules
+{\sc L:Weak} or {\sc L:Update}, respectively, are met.  For instance
+we prove the following lemma to show the soundness of the translation
+of {\sc Q:Weak}.
+
+\begin{lemma}[Relax]
+  If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
+  $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
+\end{lemma}
+\begin{proof}
+We observe the following inequations.
+\begin{align*}
+\Phi_{Q'}(\state) &= \sum_i q'_i f_i(H) \\
+&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
+  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(bH)  \\ & \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
+  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
+ &\ge \Phi_Q(\state)
+  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
+  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
+ &\ge \Phi_Q(\state).
+\end{align*}
+Since $H \models \Gamma$, using the transitivity of $\models$ and
+the definition of $\mathcal U$ and $\mathcal L$ we get
+$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
+$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
+These two inequalities justify the last step.
+\end{proof}
+%
+\noindent
+We prove similar lemmas for the rules that deal with assignment.  With
+the computational content of the proof, we have an effective algorithm
+to construct certificates for upper bounds derived automatically.
 
@@
 \and \Rule{Q:IncP}
-{ \Gamma \models 0 \le y
+{ \Gamma \models y \ge 0
 \\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
@@
 \and \Rule{Q:DecP}
-{ \Gamma \models 0 \le y
+{ \Gamma \models y \ge 0
 \\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
@@
 %
+\and \Rule{Q:Extend}
+{ \fenv f = (\vec y, S_f)
+\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
+}
+{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+%
 \and \Rule{Relax}
@@
 { Q'   \succeq_\Gamma Q }
-%
-\and \Rule{Q:Extend}
-{ \fenv f = (\vec y, S_f)
-\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
-}
-{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
-%
+
 \end{mathpar}
@@
 
-\section{Soundness Theorems}
-
-The soundness of a judgement relates the cost semantics with the potential
-functions derived using our proof system.  Depending on the outcome of
-the semantics, we give different soundness statements.
-But, before tackling the issue of soundness let us remind and define some
-convenient notations.
-\begin{itemize}
-\item $H \mapsto \Phi_H(Q)$ for $Q$ a quantitative context, is the
-unique linear potential function that has $Q(i)$ as a coefficient for the
-base function $f_i$.
-\item $H \models \Gamma$ is true if and only if all the assertions in
-$\Gamma$ are true on the heap $H$.
-\end{itemize}
-
-Ideally we would
-like our judgements to be meaningful even in the case of a program
-crash.  More precisely, we want the first potential function to be enough
-to pay for the execution of the program up to the crash point.  This is
-formally stated in the following theorem.
-
-\begin{theorem}[Busy case]
-For any $S$, $H$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$
-such that $H \models \Gamma$, $\Gamma; Q \vdash S \dashv Q'; \Gamma'$,
-  and $(H, S) \Downarrow_{(q,q')} \circ$,
-we have $q \le \Phi_H(Q)$.
-\end{theorem}
-\begin{proof}
-Simple application of the terminating case proved afterwards.
-\end{proof}
-
-In order to prove the harder case for terminating programs, need to first
-focus on the only structural rule of the proof system: {\sc Q:Weak}.  This
-rule can be applied anywhere in the derivation, it is not syntax directed.
-To prove the soundness of this rule, we need a lemma that gives a
-semantic explanation of the $\succeq_\Gamma$ relation.
-
-\begin{lemma}[Relax]
-If $Q' \succeq_\Gamma Q$, for any heap $H$ such that $H \models \Gamma$,
-we have $\Phi_H(Q') \ge \Phi_H(Q)$.
-\end{lemma}
-\begin{proof}
-We observe the following inequations.
-\begin{align*}
-\Phi_H(Q') &= \sum_i q'_i f_i(H) \\
-&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
-  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(H) + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
-  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
- &\ge \Phi_H(Q)
-  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
-  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
- &\ge \Phi_H(Q).
-\end{align*}
-Since $H \models \Gamma$, using the transitivity of $\models$ and
-the definition of $\mathcal U$ and $\mathcal L$ we get
-$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
-$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
-These two inequalities justify the last step.
-\end{proof}
-
-Using this lemma, we can tackle the hardest theorem of soundness.  This theorem
-needs a strong statement to make the proof by induction go through and defines
-the complete meaning of a judgement $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
-In a similar fashion than the busy case, the initial quantitative context is enough
-to perform the whole computation: $q \le \Phi_H(Q)$.  But additionally, the
-two quantitative contexts are also related to the resource consumption $q - q'$
-of the piece of program executed.  This second requirement exactly matches
-the semantic condition described in the overview section: it says that the
-initial potential $\Phi_H(Q)$ is enough to both pay for the resource
-consumption of the program $q - q'$ and pay for the potential of the resulting
-heap $\Phi_{H'}(Q')$.
-
-The proof of this result if very technical but basically goes through by induction
-on the big-step evaluation.  Only a representative subset of all the cases
-is present in the following proof.
-
-\begin{theorem}[Terminating case]
-For any $S$, $H$, $H'$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$ such
-that $\Gamma; Q \vdash S \dashv Q'; \Gamma'$ and $H \models \Gamma$ and
-$(H,s) \Downarrow_{(q,q')} H'$, we have
-$H' \models \Gamma'$ and
-$q \le \Phi_H(Q)$ and
-$\Phi_H(Q) - \Phi_{H'}(Q') \ge q - q'$.
-\end{theorem}
-
-\begin{proof} By nested induction on the big step semantics and on the
-judgement derivation (to account for the structural rule {\sc Q:Weak}).
-
-\begin{itemize}
-
-\item Case {\sc S:Skip}.
-  We have $(H, \code{skip}) \Downarrow_{M_k} H$.
-  By assumption, $H \models \Gamma$. \\
-  Since $\Phi_H(Q + M_k) = \Phi_H(Q) + M_k$,
-  \begin{itemize}[topsep=0pt]
-  \item
-    if $M_k \ge 0$ we have,
-      $M_k \le \Phi_H(Q + M_k)$ and
-      $\Phi_H(Q + M_k) - \Phi_H(Q) \ge M_k - 0$;
-  \item
-    if $M_k < 0$ we have,
-      $0 \le \Phi_H(Q + M_k)$ and
-      $\Phi_H(Q + M_k) - \Phi_H(Q) \ge 0 - (-M_k)$.
-  \end{itemize}
-
-\item Case {\sc S:Seq}.
-  By transitivity and induction hypothesis, it is
-  easy to get $H'' \models \Gamma''$.  Now we prove
-  the quantitative part of the theorem.  We have
-  $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
-  We assume $\forall i \in \{1, 2\}.\, M_s^i \ge 0$, the other cases
-  are similar.
-  \begin{itemize}[topsep=0pt]
-  \item
-    The first case is when $q'_1 > M_s^2 + q_2$ then $q = M_s^1 + q_1$
-    and since $q_1 \le \Phi_H(Q - M_s^1)$ (by induction) we have
-    $q \le \Phi_H(Q)$.
-  \item
-    The second case is when $q'_1 \le M_s^2 + q_2$ then $q = M_s^1 + q_1 + M_s^2 + q_2 - q'_1$. \\
-    Since $\Phi_H(Q - M_s^1) - \Phi_{H'}(Q') \ge q_1 - q'_1$
-    and $q_2 \le \Phi_{H'}(Q' - M_s^2)$ we have:
-    \begin{align*}
-    q &\le M_s^1 + \Phi_H(Q - M_s^1) - \Phi_{H'}(Q') + M_s^2 + q_2 \\
-      &= \Phi_H(Q) + (q_2 - \Phi_{H'}(Q' - M_s^2)) \le \Phi_H(Q).
-   \end{align*}
-  \end{itemize}
-  We obtain $\Phi_H(Q) - \Phi_{H''}(Q'') \ge q - q'$ by simply chaining
-  the two induction hypothesis and using the fact that $q - q' =
-  (q_1 + M_s^1 - q'_1) + (q_2 + M_s^2 - q'_2)$.
-
-\item Case {\sc S:WhileStop}.
-  This proof goes just like in the {\sc S:Skip} case.
-
-\item Case {\sc S:WhileStep}.
-  Here again we assume $\forall i \in \{1, 2, 3\}.\, M_w^i \ge 0$.
-  \begin{itemize}[topsep=0pt]
-  \item
-    If $q'_1 > M_w^2 + q_2$ then $q = M_w^1 + q_1$ and by induction
-    we get $q_1 \le \Phi_H(Q - M_w^1)$, so $q \le \Phi_H(Q)$.
-  \item
-    If $q'_1 \le M_w^2 + q_2$ then $q = M_w^1 + q_1 + M_w^2 + q_2 - q'_1$.
-    And we proceed like in the sequence case.
-  \end{itemize}
-  We obtain $\Phi_H(Q) - \Phi_{H''}(Q - M_w^3) \ge q - q'$ using the identity
-  $q - q' = (M_w^1 + M_w^2 + q_1 - q'_1) + q_2 - q'_2$ and the two
-  induction hypothesis
-  \begin{itemize}[topsep=0pt]
-  \item $\Phi_H(Q - M_w^1) - \Phi_{H'}(Q + M_w^2) \ge q_1 - q'_1$ and
-  \item $\Phi_{H'}(Q) - \Phi_{H''}(Q - M_w^3) \ge q_2 - q'_2.$
-  \end{itemize}
-
-\end{itemize}
-\end{proof}
-
-\section{Implementation}
-
-\subsection{Logical Annotations Generation}
-
-\paragraph{Program conditions.}
-The first pass in our OCaml implementation generates the
-logical annotations $\Gamma$ around all program statements.
-These annotations are used later on by the quantitative
-analysis to generate contraints.
-%
-The first design choice we had to make is to give a fixed form
-to program conditions.  Because of their generality and because
-they are closed under negation, we decided to consider
-inequalities over linear combinations as program conditions.
-\begin{mathpar}
-C := L < L \mid L > L \mid L \le L \mid L \ge L
-\and
-L := x \mid k \mid k * L \mid L + L \mid L - L
-\end{mathpar}
-A Condition is represented internally as a pair of one constant
-$k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
-Its intended semantics is
-$$
-\llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
-$$
-Note that conditions of this shape are closed by negation, as showed
-by the following equivalence
-$$
-\neg (k + \sum_v m_v \cdot H(v) \le 0)
-\Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
-$$
-We can also note that all the ``questions'' asked to the logic
-assertions during a proof derivation can be described using
-linear combinations.  Conveniently enough, this language of
-assertions is exactly Presburger's arithmetic.  This means that
-we have a sound and complete decision procedure to answer
-all the questions asked during the analysis.
-
-\paragraph{Logical contexts.}
-The second problem related to the logic is how to figure out
-the logical contexts $\Gamma$ for each program statement.
-This is a question that is answered by abstract interpretation
-but, to show that our work really does not need the full power
-of abstract interpretation, we designed and implemented a very
-simple top-down approach: starting from a statement $S$ and
-a pre-condition $\Gamma$ (given as a list of inequalities), we
-generate $\mathcal P(S, \Gamma)$ a postcondition for $S$ using the
-following equations:
-
-\begin{align*}
-\mathcal P(\code{skip}, \Gamma) &:= \Gamma \\
-\mathcal P(x \gets y, \Gamma) &:= \{ e \le 0 \mid (e \le 0) \in \Gamma \land \{x\} \# e \} \\
-\mathcal P(x \gets x + y, \Gamma) &:= \Gamma[x / x - y] \\
-\mathcal P(x \gets x - y, \Gamma) &:= \Gamma[x / x + y] \\
-\mathcal P(\code{assert}(C), \Gamma) &:= \Gamma,C \\
-\mathcal P(S_1; S_2, \Gamma) &:= \mathcal P(S_2, \mathcal P(S_1, \Gamma)) \\
-\mathcal P(\code{if} (C)~S_1~\code{else}~S_2) &:=
-  \mathcal P(S_1, \Gamma,C) \lor \mathcal P(S_2, \Gamma,\neg C) \\
-\mathcal P(\code{while}(C)~S, \Gamma) &:=
-  \Gamma \lor (\neg C, \mathcal P(S, (C, \{ e \le 0 \mid (e \le 0) \in \Gamma \land {\rm set}(S) \# e \})))
-\end{align*}
-We used the notation ${\rm set}(S)$ to designate the set of variables
-assigned to in the statement $S$ and $s\#e$ to mean that variables
-in $s$ do not appear in the linear combination $e$.  Because contexts
-are simply conjunction of conditions we must define $\Gamma \lor \Delta$
-as a function on contexts:
-$$
-\Gamma \lor \Delta := \{ e \le 0 \mid (e \le 0) \in \Gamma \cup \Delta \land \Gamma \models (e \le 0) \land \Delta \models (e \le 0) \}.
-$$
-It is important to note that the $\mathcal P$ function defined above
-is not performing any fixpoint computation.  Despite this flagrant
-lack of sophistication our complete analysis is still able to figure out some
-tricky invariants as exemplified earlier.
-Finally, to generate annotations for the whole program $S$,
-we run $\mathcal P(S,\bullet)$ (where $\bullet$ is the empty context)
-and memorize all the intermediate results.  In practice, these results are stored in
-a table that can be referenced at any time during the analysis to get the
-pre and post-condition of any statement in the program.
-
-\paragraph{Decision procedure.}
-To decide logical entailment $\Gamma \models \Delta$ we use
-Presburger's decision procedure.  But what this procedure gives us
-is in fact only an answer the the question: Is there any heap $H$
-such that $H \models \Gamma$.  While this sentence is existential
-the question $\Gamma \models \Delta$ is universal.  To use the
-decision procedure, we use the well known trick of double negation.
-\begin{align*}
-\Gamma \models C
-&\Leftrightarrow \neg\neg (\Gamma \models C) \\
-&\Leftrightarrow \neg\neg (\forall H.\, H\models\Gamma \implies H\models C) \\
-&\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\
-&\Leftrightarrow \neg \exists H, H\models\Gamma \land H\models \neg C \\
-&\Leftrightarrow \neg \exists H, H\models (\Gamma, \neg C)
-\end{align*}
-Thus, to know if $\Gamma \models C$ we ask the decision
-procedure if $\Gamma, \neg C$ is satisfiable and negate the result.
-When we want to know about $\Gamma \models \Delta$ we simply ask
-one question for each of the conjuncts of $\Delta$.
-
-\subsection{Linear Contraints Generation}
-
-\paragraph{Generating a linear program.}
-Because figuring out the coefficents that are in $Q$ right away is a hard
-problem, we separate the search of a derivation in two steps.  As a first
-step we go through the whole program and apply inductively the rules
-given in the proof system section.  During this process our tool uses
-symbolic names for the coefficients $(q_i)$.  Every time a side condition
-must be satisfied by these coefficients, it is recorded using the symbolic
-names in a global list.  When
-the list of all constraints that must be satisfied by the coefficients is
-collected, we feed it to an off the shelf LP-solver.  If the solver successfully
-finds a solution to the given problem, we know that a derivation
-exists for the considered program.  We can even ask the solver
-for the initial $Q$ and get a resource bound for the program.
-If the LP-solver fails to find a solution, an error is reported to the
-user.
-
-\paragraph{Making the proof system algorithmic.}
-One problem with the proof system as described is that it is not
-fully syntax directed.  The rule {\sc Q:Weak} can be applied at
-all times.  However, if we apply this weakening rule at every step
-of the derivation, the number of generated constraints and
-the number of queries to the logic system explode.  This blowup
-causes efficiency problems.  To avoid these performance problems
-our implementation only applies the weakening rules at some
-strategic points in the derivation.  More specifically, at sequence
-points, at the beginning of while loops and around conditional
-statements.  So far, this selection of weakening spots has been
-driven by practice but it might be possible to prove that these
-program points are the only ones where weakening is needed.
+% \section{Implementation}
+
+% \subsection{Logical Annotations Generation}
+
+% \paragraph{Program conditions.}
+% The first pass in our OCaml implementation generates the
+% logical annotations $\Gamma$ around all program statements.
+% These annotations are used later on by the quantitative
+% analysis to generate contraints.
+% %
+% The first design choice we had to make is to give a fixed form
+% to program conditions.  Because of their generality and because
+% they are closed under negation, we decided to consider
+% inequalities over linear combinations as program conditions.
+% \begin{mathpar}
+% C := L < L \mid L > L \mid L \le L \mid L \ge L
+% \and
+% L := x \mid k \mid k * L \mid L + L \mid L - L
+% \end{mathpar}
+% A Condition is represented internally as a pair of one constant
+% $k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
+% Its intended semantics is
+% $$
+% \llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
+% $$
+% Note that conditions of this shape are closed by negation, as showed
+% by the following equivalence
+% $$
+% \neg (k + \sum_v m_v \cdot H(v) \le 0)
+% \Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
+% $$
+% We can also note that all the ``questions'' asked to the logic
+% assertions during a proof derivation can be described using
+% linear combinations.  Conveniently enough, this language of
+% assertions is exactly Presburger's arithmetic.  This means that
+% we have a sound and complete decision procedure to answer
+% all the questions asked during the analysis.
+
+% \paragraph{Logical contexts.}
+% The second problem related to the logic is how to figure out
+% the logical contexts $\Gamma$ for each program statement.
+% This is a question that is answered by abstract interpretation
+% but, to show that our work really does not need the full power
+% of abstract interpretation, we designed and implemented a very
+% simple top-down approach: starting from a statement $S$ and
+% a pre-condition $\Gamma$ (given as a list of inequalities), we
+% generate $\mathcal P(S, \Gamma)$ a postcondition for $S$ using the
+% following equations:
+
+% \begin{align*}
+% \mathcal P(\code{skip}, \Gamma) &:= \Gamma \\
+% \mathcal P(x \gets y, \Gamma) &:= \{ e \le 0 \mid (e \le 0) \in \Gamma \land \{x\} \# e \} \\
+% \mathcal P(x \gets x + y, \Gamma) &:= \Gamma[x / x - y] \\
+% \mathcal P(x \gets x - y, \Gamma) &:= \Gamma[x / x + y] \\
+% \mathcal P(\code{assert}(C), \Gamma) &:= \Gamma,C \\
+% \mathcal P(S_1; S_2, \Gamma) &:= \mathcal P(S_2, \mathcal P(S_1, \Gamma)) \\
+% \mathcal P(\code{if} (C)~S_1~\code{else}~S_2) &:=
+%   \mathcal P(S_1, \Gamma,C) \lor \mathcal P(S_2, \Gamma,\neg C) \\
+% \mathcal P(\code{while}(C)~S, \Gamma) &:=
+%   \Gamma \lor (\neg C, \mathcal P(S, (C, \{ e \le 0 \mid (e \le 0) \in \Gamma \land {\rm set}(S) \# e \})))
+% \end{align*}
+% We used the notation ${\rm set}(S)$ to designate the set of variables
+% assigned to in the statement $S$ and $s\#e$ to mean that variables
+% in $s$ do not appear in the linear combination $e$.  Because contexts
+% are simply conjunction of conditions we must define $\Gamma \lor \Delta$
+% as a function on contexts:
+% $$
+% \Gamma \lor \Delta := \{ e \le 0 \mid (e \le 0) \in \Gamma \cup \Delta \land \Gamma \models (e \le 0) \land \Delta \models (e \le 0) \}.
+% $$
+% It is important to note that the $\mathcal P$ function defined above
+% is not performing any fixpoint computation.  Despite this flagrant
+% lack of sophistication our complete analysis is still able to figure out some
+% tricky invariants as exemplified earlier.
+% Finally, to generate annotations for the whole program $S$,
+% we run $\mathcal P(S,\bullet)$ (where $\bullet$ is the empty context)
+% and memorize all the intermediate results.  In practice, these results are stored in
+% a table that can be referenced at any time during the analysis to get the
+% pre and post-condition of any statement in the program.
+
+% \paragraph{Decision procedure.}
+% To decide logical entailment $\Gamma \models \Delta$ we use
+% Presburger's decision procedure.  But what this procedure gives us
+% is in fact only an answer the the question: Is there any heap $H$
+% such that $H \models \Gamma$.  While this sentence is existential
+% the question $\Gamma \models \Delta$ is universal.  To use the
+% decision procedure, we use the well known trick of double negation.
+% \begin{align*}
+% \Gamma \models C
+% &\Leftrightarrow \neg\neg (\Gamma \models C) \\
+% &\Leftrightarrow \neg\neg (\forall H.\, H\models\Gamma \implies H\models C) \\
+% &\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\
+% &\Leftrightarrow \neg \exists H, H\models\Gamma \land H\models \neg C \\
+% &\Leftrightarrow \neg \exists H, H\models (\Gamma, \neg C)
+% \end{align*}
+% Thus, to know if $\Gamma \models C$ we ask the decision
+% procedure if $\Gamma, \neg C$ is satisfiable and negate the result.
+% When we want to know about $\Gamma \models \Delta$ we simply ask
+% one question for each of the conjuncts of $\Delta$.
+
+% \subsection{Linear Contraints Generation}
+
+% \paragraph{Generating a linear program.}
+% Because figuring out the coefficents that are in $Q$ right away is a hard
+% problem, we separate the search of a derivation in two steps.  As a first
+% step we go through the whole program and apply inductively the rules
+% given in the proof system section.  During this process our tool uses
+% symbolic names for the coefficients $(q_i)$.  Every time a side condition
+% must be satisfied by these coefficients, it is recorded using the symbolic
+% names in a global list.  When
+% the list of all constraints that must be satisfied by the coefficients is
+% collected, we feed it to an off the shelf LP-solver.  If the solver successfully
+% finds a solution to the given problem, we know that a derivation
+% exists for the considered program.  We can even ask the solver
+% for the initial $Q$ and get a resource bound for the program.
+% If the LP-solver fails to find a solution, an error is reported to the
+% user.
+
+% \paragraph{Making the proof system algorithmic.}
+% One problem with the proof system as described is that it is not
+% fully syntax directed.  The rule {\sc Q:Weak} can be applied at
+% all times.  However, if we apply this weakening rule at every step
+% of the derivation, the number of generated constraints and
+% the number of queries to the logic system explode.  This blowup
+% causes efficiency problems.  To avoid these performance problems
+% our implementation only applies the weakening rules at some
+% strategic points in the derivation.  More specifically, at sequence
+% points, at the beginning of while loops and around conditional
+% statements.  So far, this selection of weakening spots has been
+% driven by practice but it might be possible to prove that these
+% program points are the only ones where weakening is needed.
 
@@
 
+\section{Experimental Evaluation}
+
+\section{Related Work}
+
+\section{Conclusion}
+
 % \acks
@@
 
-\begin{tabular}{r|p{3cm}c|p{3cm}c|c}
+\begin{tabular}{r|p{2cm}c|p{2cm}c|p{2cm}c|p{1.8cm}|c}
 
 File &
+\multicolumn{2}{|c|}{KoAT} &
 \multicolumn{2}{|c|}{Rank} &
 \multicolumn{2}{|c|}{Loopus} &
+Speed &
 Amortized Automation
@@
 %--------------------------------------------
-\texttt{gcd.c} &
+\hline \texttt{gcd.c} &
+
+? &
+&
 
@@
 &
+
+? &
 
@@
 %--------------------------------------------
-\texttt{kmp.c} &
+\hline \texttt{kmp.c} &
+
+? &
+&
 
@@
 
+? &
+
 $1 + 2 |\inter 0 n|$
@@
 %--------------------------------------------
-\texttt{qsort.c} &
+\hline \texttt{qsort.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{speed\_pldi09\_fig1.c} &
+\hline \texttt{speed\_pldi09\_fig1.c} &
+
+? &
+&
 
@@
 $O(n)$ &
+
+God damn it &
 
@@
 %--------------------------------------------
-\texttt{speed\_pldi09\_fig4\_2.c} &
+\hline \texttt{speed\_pldi09\_fig4\_2.c} &
+
+Fail &
+&
 
@@
 &
+
+$\frac n m + n$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_pldi09\_fig4\_4.c} &
+\hline \texttt{speed\_pldi09\_fig4\_4.c} &
+
+Fail &
+&
 
@@
 %--------------------------------------------
-\texttt{speed\_pldi09\_fig4\_5.c} &
+\hline \texttt{speed\_pldi09\_fig4\_5.c} &
+
+$28d + 7g + 27$ &
+$O(n)$ &
 
@@
 &
+
+$\max(n, n-m)$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_pldi10\_ex1.c} &
-
-$(((1+1)+n)+((-1+((1/2){\cdot}n))+((1/2){\cdot}(n{\cdot}n))))+1$ &
-$O(n^2)$ &
-
-$\max(n, 0) + \max(0,  \max(n, 0) {\cdot} \max(0, (n + -1)))$ &
-$O(n^2)$ &
-
-Fail
-\\
-
-%--------------------------------------------
-\texttt{speed\_pldi10\_ex3.c} &
-
-$((((1+1)+(-1+n))+(-1+n))+1)+1$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(n, 0)$ &
-$O(n)$ &
+\hline \texttt{speed\_pldi10\_ex1.c} &
+
+Fail &
+&
+
+Fail &
+&
+
+Fail &
+&
+
+$n$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_pldi10\_ex4.c} &
+\hline \texttt{speed\_pldi10\_ex3.c} &
 
@@
 
+$((((1+1)+(-1+n))+(-1+n))+1)+1$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \texttt{speed\_pldi10\_ex4.c} &
+
+$110 A + 33$ &
+$O(n)$ &
+
 Fail &
 &
+
+Fail &
+&
+
+$n + 1$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_fig2\_1.c} &
+\hline \texttt{speed\_popl10\_fig2\_1.c} &
+
+$9a + 9b + 10c + 10d + 7$ &
+$O(n)$ &
 
@@
 $O(n)$ &
+
+$\max(0, n-x) + \max(0, m-y)$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_fig2\_2.c} &
+\hline \texttt{speed\_popl10\_fig2\_2.c} &
+
+$6a + 9b + 3c + 5$ &
+$O(n)$ &
 
@@
 $O(n)$ &
+
+$\max(0, n-x) + \max(0, n-z)$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_nested\_multiple.c} &
+\hline \texttt{speed\_popl10\_nested\_multiple.c} &
+
+Fail &
+&
 
@@
 
+$\max(0, m-y) + \max(0, n-x)$ &
+
 $|\inter x n| + |\inter y m|$
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_nested\_single.c} &
+\hline \texttt{speed\_popl10\_nested\_single.c} &
+
+$48 B + 16$ &
+$O(n)$ &
 
@@
 $O(n)$ &
+
+$n$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_sequential\_single.c} &
+\hline \texttt{speed\_popl10\_sequential\_single.c} &
+
+$21 B + 6$ &
+$O(n)$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_simple\_multiple.c} &
+\hline \texttt{speed\_popl10\_simple\_multiple.c} &
+
+$9 C + 10 D + 7$ &
+$O(n)$ &
 
@@
 $O(n)$ &
+
+$n + m$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_simple\_single\_2.c} &
+\hline \texttt{speed\_popl10\_simple\_single\_2.c} &
+
+$20 D + 12 C + 17$ &
+$O(n)$ &
 
@@
 %--------------------------------------------
-\texttt{speed\_popl10\_simple\_single.c} &
+\hline \texttt{speed\_popl10\_simple\_single.c} &
+
+$4 B + 6$ &
+$O(n)$ &
 
@@
 %--------------------------------------------
-\texttt{t07.c} &
+\hline \texttt{t07.c} &
+
+? &
+&
 
@@
 $O(n)$ &
+
+? &
 
@@
 %--------------------------------------------
-\texttt{t08.c} &
+\hline \texttt{t08.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t10.c} &
+\hline \texttt{t10.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t11.c} &
+\hline \texttt{t11.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t13.c} &
+\hline \texttt{t13.c} &
+
+? &
+&
 
@@
 
+? &
+
 $2 |\inter 0 x| + |\inter 0 y|$
@@
 %--------------------------------------------
-\texttt{t15.c} &
+\hline \texttt{t15.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t16.c} &
+\hline \texttt{t16.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t19.c} &
+\hline \texttt{t19.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t20.c} &
+\hline \texttt{t20.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t27.c} &
+\hline \texttt{t27.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t28.c} &
+\hline \texttt{t28.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t30.c} &
+\hline \texttt{t30.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t37.c} &
+\hline \texttt{t37.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t39.c} &
+\hline \texttt{t39.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t46.c} &
+\hline \texttt{t46.c} &
+
+? &
+&
 
@@
 %--------------------------------------------
-\texttt{t47.c} &
+\hline \texttt{t47.c} &
+
+? &
+&
 
@@
 
-$|\inter 0 n|$
+? &
+
+$1 + |\inter 0 n|$
 \\

### 2014-07-02 paper/fig/run.sh
add material to rebuild figure

@@ -1,0 +1,12 @@
+APP=../../src/app
+N=100
+I=5
+
+for y in `seq -$N $I $N`; do
+  for z in `seq -$N $I $N`; do
+    printf "%03.1d %03.1d " $y $z
+    (echo "y = $y; z = $z;"; cat prog.while) \
+      | $APP -teval \
+      | sed -ne '/cost/{s/^.*: \(.*\)/\1/;p}'
+  done
+done

### 2014-07-02 paper/main.tex
add experimentation examples
related work

@@
   analysis is inspired by type-based amortized resource analysis for
-  functional programs~\cite{Jost03,HoffmannAH12}.  It computes
+  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
   derivations by generating simple linear constraints that can be
@@
 
+\begin{figure*}
+\centering
+\begin{tabular}{c|c|c|c}
+%
+\begin{lstlisting}
+while (z > y)
+  y++;
+while (y > 2)
+  y -= 3;
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (i > 100)
+  i--;
+i += k+50;
+while (i >= 0)
+  i--;
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (x > 0) {
+  x--;
+  t=x, x=y, y=t;
+}
+\end{lstlisting}
+&
+\begin{lstlisting}
+assert(y >= 0);
+while (x>y) {
+  x -= y+1;
+  for (z=y; z > 0; z--)
+    /* skip */;
+}
+\end{lstlisting}
+\\
+%
+\hline
+t08 (slightly modified) & t19 & t30 & t15
+%
+\end{tabular}
+\caption{
+  Comparison of the bounds given by different tools on
+  linear examples.
+}
+\label{fig:compar}
+\end{figure*}
+
 \section{Related Work}
 
+Most closely related to this article is our previous work on
+end-to-end stack-bound verification of Clight
+programs~\cite{veristack14}.  In that work, we have implemented and
+verified a quantitative logic to reason about stack-space usage.  In
+this article, we present a more general quantitative Hoare logic that
+is parametric in the resource of interest.  The main innovation over
+our previous work~\cite{veristack14} is the novel automatic amortized
+analysis for Clight programs that competes logical derivations for
+non-trivial bounds for programs with loops and recursion.
+
+Our work has been inspired by type-based amortized resource analysis
+for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
+are three major improvements over previous work in the current paper,
+which presents the first automatic amortized resource analysis for C.
+First, we solved the long-standing open problem of extending automatic
+amortized resource analysis to compute bounds for programs that loop
+on (possibly negative) integers.  Second, we extended the analysis
+system to deal with non-linear control flow that is introduced by
+\code{break} and \code{return} statements.  Third, for the first time,
+we combined an automatic amortized analysis with a system for
+interactively deriving bounds.
+
+In the development of our quantitative Hoare logic we have drawn
+inspiration from mechanically verified Hoare logics.
+Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
+logics in Isabelle/HOL has been helpful to understand the interaction
+of auxiliary variables with the consequence rule.  The consequence
+rule we use in our Coq implementation is a quantitative version of a
+consequence rule that has been attributed to Martin Hofmann by
+Nipkow~\cite{Nipkow02}.  Appel's separation logic for CompCert
+Clight~\cite{AppelLogic} has been a blueprint for the general
+structure of the quantitative logic.  Since we do not deal with memory
+safety, our logic is much simpler and it would be possible to
+integrate it with Appel's logic.  The continuation passing style that
+we use in the quantitative logic is not only used by
+Appel~\cite{AppelLogic} but also in Hoare logics for low-level
+code~\cite{NiS06,JensenBK13}.
+
+There exist quantitative logics that are integrated into separation
+logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
+quantitative logic.  However, the purpose of these logics is slightly
+different since they focus on the verification of bounds that depend
+on the shape of heap data structures and they are not implemented for
+C.  Also closely related to our logic is a VDM-style logic for
+reasoning about resource usage of an abstract fragment of JVM byte
+code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
+Hoare-style, does not target C code, and has been mainly designed to
+produce certificates for bounds derived for high-level functional
+programs.
+
+There exist many tools that can automatically derive loops and
+recursion bounds for imperative programs such as
+SPEED~\cite{GulwaniMC09,GulwaniZ10}, KoAT~\cite{BrockschmidtEFFG14},
+PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
+ABC~\cite{BlancHHK10} and LOOPUS~\cite{Zuleger11,SinnZV14}.  These
+tools are based on abstract interpretation--based invariant generation
+and/or term rewriting techniques, and they derive impressive results
+on realistic software.  The importance of amortization to derive tight
+bounds is well known in the resource analysis
+community~\cite{AlonsoG12,Moser14,SinnZV14}.  Currently, the only
+other available tools that can be directly applied to C code are Rank
+and LOOPUS.  Our analysis framework does not aim to set a new standard
+for automatic bound analysis.  Our contribution is rather a principled
+approach that produces certificates that are proved sound with respect
+to a formal cost semantics.  Moreover, we have a system that enables
+both automatic and interactive bound derivation, a formal soundness
+proof in Coq, and a method that can handle resources that can be
+released (e.g., memory).  However, as we have shown
+in~\pref{sec:exper}, our automatic amortized analysis matches that the
+state of the art in automatic bound analysis for linear bounds and
+sometimes even derives better constant factors than the other tools.
+Since Rank and Loopus do not handle recursion, our automatic amortized
+analysis is also the only available tool that can derive bounds for
+recursive C programs.
+
+We are only aware of two verified quantitative analysis systems.
+Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
+automatically verify previously inferred loop invariants, size
+relations, and ranking functions for Java Card programs.  However,
+they do not have a formal cost semantics and do not prove the bounds
+correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
+have verified a loop bound analysis for CompCert's RTL intermediate
+language.  However, there is no way to interactively derive bounds or
+to deal with resources like memory usage.  Also Blazy et al.'s
+automatic bound analysis does not compute symbolic bounds.
+
 \section{Conclusion}
+
+cite flops
 
@@
 int srch(
-  int t[], int n,
-  int p[], int m,
+  int t[], int n,  /* haystack */
+  int p[], int m,  /* needle */
   int b[]
@@
 
-$((1+1)+x)+1$ &
-$O(n)$ &
+Fail &
+&
 

### 2014-07-02 paper/main.tex
conclusion

@@
 
+There are techniques~\cite{Braberman08} that can compute the memory
+requirements of object oriented programs with region based garbage
+collection.  These systems infer invariants and use external tools
+that count the number of integer points in the corresponding polytopes
+to obtain bonds.  The described technique can handle loops but not
+recursive or composed functions.
+
 We are only aware of two verified quantitative analysis systems.
@@
 
-cite flops
+We have introduced an novel quantitative analysis system for CompCert
+Clight programs.  To the best of our knowledge, this article presents
+the first resource analysis framework for C that makes it possible to
+combine non-trivial automatically derived bounds with interactively
+derived bounds in a proof system that produces verifiable certificates
+for the bounds.  The main technical innovations are a quantitative
+Hoare logic for reasoning about user-defined resource cost and an
+automatic amortized analysis that derives bounds for programs whose
+resource consumption depends on (possibly negative) integers and
+non-sequential control flow as introduced by \code{break} and
+\code{return} statements.
+
+We will continue to improve our framework with the goal of reasoning
+about resource usage of system software such as the hypervisor kernel
+CertiKOS~\cite{GuVFSC11}, which we are currently developing and
+verifying.  For on thing, we will improve the automation to derive
+\emph{polynomial bounds} and to handle non-linear size changes, as
+already developed for functional programs~\cite{HoffmannS13}.  For
+another thing, we will build on previous work~\cite{HoffmannMZ13}
+the generalize the quantitative Hoare logic to handle \emph{concurrent
+  programs} with locks and lock-free data structures.
 

### 2014-07-03 paper/main.tex
fixed some typos

@@
 already experimented with an extension to multivariate resource
-polynomials~\cite{HoffmannAH11} but this would make the inference
+polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
 rules considerably more involved and should better be described
@@
 functions of the input parameters of the program.  Another unique
-feature of our system is, that it can naturally handle resources like
+feature of our system is, that it can handle resources like
 memory that may become available during the execution.
@@
 SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
-comparison.  Our experiments shows that the bounds that we derive are
+comparison.  Our experiments show that the bounds that we derive are
 often more precise than those derived by existing tools.  Only

### 2014-07-03 paper/main.tex
formatting

@@
 \usepackage{enumitem}
+\usepackage{array} %for stupid tables
 
@@
 \begin{figure*}
-\tiny
-
-\begin{tabular}{r|p{2cm}c|p{2cm}c|p{2cm}c|p{1.8cm}|c}
+\small
+
+\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
+\newcommand{\file}[2][c]{%
+  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
+\renewcommand{\max}[0]{{\rm mx}}
+\renewcommand{\min}[0]{{\rm mn}}
+
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\centering
+\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
 
@@
 
-$((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
-$O(n)$ &
-
-Fail &
+% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
+$(((2{+}1)\dots$ &
+$O(n)$ &
+
+--- &
 &
@@
 
-$|\inter 0 x| + |\inter 0 y|$
+$|\inter 0 x| {+} |\inter 0 y|$
 \\
@@
 
-$((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
+% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
+$(((2{+}(n{+} \dots$ &
 $O(n^2)$ &
 
-$\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
+% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
+$\max(n, 0) \dots$ &
 $O(n)$ &
@@
 
-$1 + 2 |\inter 0 n|$
+$1 {+} 2 |\inter 0 n|$
 \\
@@
 
-Fail &
-&
-
-Fail &
+--- &
+&
+
+--- &
 &
@@
 
-$1 + 2 |\inter 0 {\code{len}}|$
+$1 {+} 2 |\inter 0 {\code{len}}|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_pldi09\_fig1.c} &
-
-? &
-&
-
-Fail &
-&
-
-$\max(n, 0) + \max(n, 0)$ &
-$O(n)$ &
-
-God damn it &
-
-$1 + 2 |\inter 0 n|$
+% \hline \file{\dir{speed\_pldi09}{fig1.c}} &
+% 
+% ? &
+% &
+% 
+% --- &
+% &
+% 
+% $2 \max(n, 0)$ &
+% $O(n)$ &
+% 
+% God damn it &
+% 
+% $1 + 2 |\inter 0 n|$
+% \\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_2.c}} &
+
+--- &
+&
+
+% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}n)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + n$ &
+
+$1 {+} 2 |\inter 0 n|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_pldi09\_fig4\_2.c} &
-
-Fail &
-&
-
-$((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
-$O(n)$ &
-
-Fail &
-&
-
-$\frac n m + n$ &
-
-$1 + 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \texttt{speed\_pldi09\_fig4\_4.c} &
-
-Fail &
-&
-
-$((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
-$O(n)$ &
-
-Fail &
+\hline \file{\dir{speed\_pldi09}{fig4\_4.c}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
 &
@@
 %--------------------------------------------
-\hline \texttt{speed\_pldi09\_fig4\_5.c} &
+\hline \file{\dir{speed\_pldi09}{fig4\_5.c}} &
 
@@
 
-$((((1+1)+(-1+n))+m)+1)+1$ &
-$O(n)$ &
-
-Fail &
+% $((((1+1)+(-1+n))+m)+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
 &
@@
 
-Fail
+---
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_pldi10\_ex1.c} &
-
-Fail &
-&
-
-Fail &
-&
-
-Fail &
+\hline \file{\dir{speed\_pldi10}{ex1.c}} &
+
+--- &
+&
+
+--- &
+&
+
+--- &
 &
@@
 %--------------------------------------------
-\hline \texttt{speed\_pldi10\_ex3.c} &
-
-Fail &
-&
-
-$((((1+1)+(-1+n))+(-1+n))+1)+1$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(n, 0)$ &
+\hline \file{\dir{speed\_pldi10}{ex3.c}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
 $O(n)$ &
@@
 %--------------------------------------------
-\hline \texttt{speed\_pldi10\_ex4.c} &
-
-$110 A + 33$ &
-$O(n)$ &
-
-Fail &
-&
-
-Fail &
+\hline \file{\dir{speed\_pldi10}{ex4.c}} &
+
+$110 a + 33$ &
+$O(n)$ &
+
+--- &
+&
+
+--- &
 &
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_fig2\_1.c} &
-
-$9a + 9b + 10c + 10d + 7$ &
-$O(n)$ &
-
-$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$O(n)$ &
-
-$\max(0, (n - x)) + \max(0, (m - y))$ &
-$O(n)$ &
-
-$\max(0, n-x) + \max(0, m-y)$ &
-
-$|\inter x n| + |\inter y m|$
+\hline \file{\dir{speed\_popl10}{fig2\_1.c}} &
+
+$9a + 9b + \dots$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{+}((-y\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, m{-}y)$ &
+
+$|\inter x n| {+} |\inter y m|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_fig2\_2.c} &
+\hline \file{\dir{speed\_popl10}{fig2\_2.c}} &
 
@@
 
-$(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
-$O(n)$ &
-
-$\max(0, (\max(0, (x + 1 + -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
-$O(n)$ &
-
-$\max(0, n-x) + \max(0, n-z)$ &
-
-$|\inter x n| + |\inter z n|$
+% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
+$((2{-}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
+$\max(0, (x + 1 {-}z)\dots$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, n{-}z)$ &
+
+$|\inter x n| {+} |\inter z n|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_nested\_multiple.c} &
-
-Fail &
-&
-
-$(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
+\hline \file{\dir{speed\_popl10}{nstd\_multiple.c}} &
+
+--- &
+&
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
+$((2{-}x{+}n\dots$ &
 $O(n^2)$ &
 
-$\max(0, (m - y)) + \max(0, (n - x))$ &
-$O(n)$ &
-
-$\max(0, m-y) + \max(0, n-x)$ &
-
-$|\inter x n| + |\inter y m|$
+$\max(0, m {-} y) + \max(0, n {-} x)$ &
+$O(n)$ &
+
+$\max(0, m{-}y) + \max(0, n{-}x)$ &
+
+$|\inter x n| {+} |\inter y m|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_nested\_single.c} &
-
-$48 B + 16$ &
-$O(n)$ &
-
-$((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
-$O(n)$ &
-
-$\max(0, min(n, (n + -1))) + \max(0, min(n, (n + -1)))$ &
+\hline \file{\dir{speed\_popl10}{nstd\_single.c}} &
+
+$48 b + 16$ &
+$O(n)$ &
+
+% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
+$(((1{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(0, \! n {-} 1)\dots$ &
 $O(n)$ &
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_sequential\_single.c} &
-
-$21 B + 6$ &
-$O(n)$ &
-
-$(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(n, 0)$ &
+\hline \file{\dir{speed\_popl10}{sqntl\_single.c}} &
+
+$21 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{}-x{+}n\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
 $O(n)$ &
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_simple\_multiple.c} &
-
-$9 C + 10 D + 7$ &
-$O(n)$ &
-
-$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+\hline \file{\dir{speed\_popl10}{smpl\_multiple.c}} &
+
+$9 c + 10 d + 7$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
 $O(n)$ &
@@
 
-$|\inter 0 m| + |\inter 0 n|$
+$|\inter 0 m| {+} |\inter 0 n|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_simple\_single\_2.c} &
-
-$20 D + 12 C + 17$ &
-$O(n)$ &
-
-Fail &
+\hline \file{\dir{speed\_popl10}{smpl\_single2.c}} &
+
+$20 d + 12 c + 17$ &
+$O(n)$ &
+
+--- &
 &
@@
 
-$|\inter 0 n| + |\inter 0 m|$
+$|\inter 0 n| {+} |\inter 0 m|$
 \\
@@
 %--------------------------------------------
-\hline \texttt{speed\_popl10\_simple\_single.c} &
-
-$4 B + 6$ &
-$O(n)$ &
-
-$(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+\hline \file{\dir{speed\_popl10}{smpl\_single.c}} &
+
+$4 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{-}x{+}n\dots$ &
 $O(n)$ &
@@
 %--------------------------------------------
-\hline \texttt{t07.c} &
+\hline \file{\texttt{t07.c}} &
 
@@
 
-$(((1+1)+-1)+x)+1$ &
-$O(n)$ &
-
-$\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
+$2+x$ &
+$O(n)$ &
+
+% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
+$\max(x, 0)\dots$ &
 $O(n)$ &
@@
 
-$1 + 3 |\inter 0 x| + |\inter 0 y|$
+$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
 \\
@@
 
-$(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
-$O(n)$ &
-
-$\max(0, (\max(0, (y + -9)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
+$((2{+}z{-}y\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y + -9)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+$\max(0, \! y{-}9)\dots$ &
 $O(n)$ &
@@
 
-$1.1 |\inter y z| + 0.1 |\inter 0 y|$
+$1.1 |\inter y z| {+} 0.1 |\inter 0 y|$
 \\
@@
 
-$(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
-$O(n)$ &
-
-$\max(0, (x - y))$ &
+% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
+$((2{-}y{+}x\dots$ &
+$O(n)$ &
+
+$\max(0, x {-} y)$ &
 $O(n)$ &
@@
 
-$(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$O(n)$ &
-
-$\max(0, (n - x)) + \max(0, (m - y))$ &
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x)) + \max(0, m {-} y)$ &
 $O(n)$ &
@@
 
-$|\inter x n| + |\inter y m|$
+$|\inter x n| {+} |\inter y m|$
 \\
@@
 
-$((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
+% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
+$(((1{+}y^2/2\dots$ &
 $O(n^2)$ &
 
-$\max(x, 0) + \max(0, (\max(x, 0) + \max(y, 0)))$ &
+$2{\cdot}\max(x, 0) + \max(y, 0)$ &
 $O(n)$ &
@@
 
-$2 |\inter 0 x| + |\inter 0 y|$
+$2 |\inter 0 x| {+} |\inter 0 y|$
 \\
@@
 
-$(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
-$O(n)$ &
-
-Fail &
+% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
+$((1{+}x\dots$ &
+$O(n)$ &
+
+--- &
 &
@@
 
-$(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
-$O(n)$ &
-
-Fail &
+% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
+$((-99{\cdot}y\dots$ &
+$O(n)$ &
+
+--- &
 &
@@
 
-$(((1+1)+(151+k))+(-100+i))+1$ &
-$O(n)$ &
-
-$\max(0, (i + -100)) + \max(0, (k + i + 51))$ &
+% $(((1+1)+(151+k))+(-100+i))+1$ &
+$((153{+}k\dots$ &
+$O(n)$ &
+
+$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
 $O(n)$ &
@@
 
-$50 + |\inter {-1} i| + |\inter 0 k|$
+$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
 \\
@@
 
-$((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
-$O(n)$ &
-
-$\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
+% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
+$(2{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
+$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
 $O(n)$ &
@@
 
-$|\inter x y| + |\inter y x|$
+$|\inter x y| {+} |\inter y x|$
 \\
@@
 
-Fail &
-&
-
-$\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
+--- &
+&
+
+% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
+$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
 $O(n)$ &
@@
 
-$0.1|\inter n y| + 11 |\inter n 0|$
+$0.1|\inter n y| {+} 11 |\inter n 0|$
 \\
@@
 
-$(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
-$O(n)$ &
-
-$\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
+% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
+$((1{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
+$10^3 \,\max(0, x - y)\dots$ &
 $O(n)$ &
@@
 
-$|\inter x 0| + |\inter 0 y| + 1002 |\inter y x|$
+$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
 \\
@@
 
-Fail &
-&
-
-$\max(0, (1 + \max(x, y, 0)))$ &
+--- &
+&
+
+$1 + \max(x, y)$ &
 $O(n)$ &
@@
 
-$3 + 2 |\inter 0 x| + |\inter 0 y|$
+$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
 \\
@@
 
-$1.33 + 0.67 |\inter z y|$
+$1.33 {+} 0.67 |\inter z y|$
 \\
@@
 
-$(((1+1)+n)+1)+1$ &
+$4+n$ &
 $O(n)$ &
@@
 
-$1 + |\inter 0 n|$
+$1 {+} |\inter 0 n|$
 \\
@@
 \end{tabular}
-\caption{That big ugly table}
+\caption{
+  Experimental evaluation comparing the bounds generated KoAT, Rank,
+  Loopus, Speed, and our automatic amortized analysis on several
+  challenging linear examples. Results for KoAT and Speed were extracted
+  from older publications because KoAT cannot take C programs as 
+  input in its current version and Speed is not publically available.
+  Entries marked with ? signal that we cannot test the tool. Entries
+  marked with --- mean that the tool failed to produce a result.
+}
 \label{fig:eval}

### 2014-07-03 paper/main.tex
add cool figure
minor changes

@@
 \begin{lstlisting}
-while (z > y)
-  y++;
-while (y > 2)
-  y -= 3;
+while (y > x) {
+  x++; tick(1) }
+while (x > 2) {
+  x -= 3; tick(1) }
 \end{lstlisting}
@@
 \hline
-t08 (slightly modified) & t19 & t30 & t15
+t08a (slightly modified t08) & t19 & t30 & t15
 %
@@
 \end{figure*}
+
+\begin{figure}
+\includegraphics[width=\linewidth]{fig/bound3d}
+\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
+  (blue lines) and the measured runtime cost (red crosses) for example
+  \emph{t8a}. For $x\ge 0$ the derived bound is tight.}
+\label{fig:3d}
+\end{figure}
 
@@
 logics in Isabelle/HOL has been helpful to understand the interaction
-of auxiliary variables with the consequence rule.  The consequence
-rule we use in our Coq implementation is a quantitative version of a
-consequence rule that has been attributed to Martin Hofmann by
-Nipkow~\cite{Nipkow02}.  Appel's separation logic for CompCert
-Clight~\cite{AppelLogic} has been a blueprint for the general
-structure of the quantitative logic.  Since we do not deal with memory
-safety, our logic is much simpler and it would be possible to
-integrate it with Appel's logic.  The continuation passing style that
-we use in the quantitative logic is not only used by
+of auxiliary variables with the consequence rule.  Appel's separation
+logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
+the general structure of the quantitative logic.  Since we do not deal
+with memory safety, our logic is much simpler and it would be possible
+to integrate it with Appel's logic.  The continuation passing style
+that we use in the quantitative logic is not only used by
 Appel~\cite{AppelLogic} but also in Hoare logics for low-level

### 2014-07-03 paper/main.tex
minor

@@
 \centering
-\begin{tabular}{c|c|c|c}
-%
+\begin{tabular}{t|c|c|c|c}
+%
+&
 \begin{lstlisting}
@@
 \hline
-t08a (slightly modified t08) & t19 & t30 & t15
-%
+& t08a (slightly modified t08) & t19 & t30 & t15
+%
+\\
+\hline
+LOOPUS \\
+Rank \\
+AAA
 \end{tabular}
-\caption{
-  Comparison of the bounds given by different tools on
-  linear examples.
-}
+\caption{ Comparison of resource bounds derived by different tools on
+  several examples with linear bounds.  AAA stands for our automatic
+  amortized analysis for Clight.  The output of Rank has been manually
+  simplified to fit the table.  }
 \label{fig:compar}
@@
   (blue lines) and the measured runtime cost (red crosses) for example
-  \emph{t8a}. For $x\ge 0$ the derived bound is tight.}
+  \emph{t08a}. For $x\ge 0$ the derived bound is tight.}
 \label{fig:3d}

### 2014-07-03 paper/main.tex
typo
bounds in comparison table

@@
 \end{figure}
-
-\subsection{Implementation}
-
-We implemented the new automatic amortized analysis using CIL [XXX].
-CIL is a subset of C that is very similar to CompCert Clight but
-provides tool support that facilitated the implementation.  The
-implementation is publically available at XXX.  More details can be
-found in section XXX.
-
-We do not yet generate certificates in the program logic that are
-checkable in Coq.  This is substantial additional engineering effort
-but is conceptually straightforward as we describe in detail in
-Section XXX.
 
@@
 \centering
-\begin{tabular}{t|c|c|c|c}
+\begin{tabular}{r|c|c|c|c}
 %
@@
 while (y > x) {
-  x++; tick(1) }
+  x++; tick(1); }
 while (x > 2) {
-  x -= 3; tick(1) }
-\end{lstlisting}
-&
-\begin{lstlisting}
-while (i > 100)
-  i--;
+  x -= 3; tick(1); }
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (i > 100) {
+  i--; tick(1) }
 i += k+50;
-while (i >= 0)
-  i--;
+while (i >= 0) {
+  i--; tick(1); }
 \end{lstlisting}
@@
   t=x, x=y, y=t;
+  tick(1);
 }
@@
   for (z=y; z > 0; z--)
-    /* skip */;
+    tick(1);
+  tick(1);
 }
@@
 \hline
-LOOPUS \\
-Rank \\
+LOOPUS &
+{\small XXX} &
+{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
+{\small % $1 + \max(x, y)$ (???)
+---
+} &
+---
+\\
+%
+%
+%
+%
+Rank &
+{\small $2 {+} y {-} x (???)$ } &
+{\small $54+k+i$ } &
+--- &
+{\small $2+2x-y$ }
+\\
+%
+%
 AAA 
+  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
+  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
+  & $|\inter 0 x| {+} |\inter 0 y|$
+  & $|\inter 0 x|$
 \end{tabular}

### 2014-07-03 eval/t08.c
change t08 to use 3 instead of 10

@@ -3,7 +3,7 @@ void start(int y, int z)
   while (z>y) {
     y=y+1;
   }
-  while (y>9) {
-    y=y-10;
+  while (y>2) {
+    y=y-3;
   }
 }

### 2014-07-03 paper/main.tex
fixed t08

@@
 
-Example \emph{t08} shows the ability of the analysis to handle
+Example \emph{t08a} shows the ability of the analysis to handle
 negative and non-negative numbers.  Note that there are no
@@
 accurately track the size change of $y$ in the first loop.
-Furthermore, \emph{t08} shows that we do not handle the constants $1$
+Furthermore, \emph{t08a} shows that we do not handle the constants $1$
 or $0$ in any special way.  In all examples you could replace $0$ and
@@
 \\[.7\baselineskip] 
-      {\bf t08}
+      {\bf t08a}
     \end{center}
@@
 LOOPUS &
-{\small XXX} &
+{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
 {\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
@@
   (blue lines) and the measured runtime cost (red crosses) for example
-  \emph{t08a}. For $x\ge 0$ the derived bound is tight.}
+  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
 \label{fig:3d}
@@
    \begin{lstlisting}
-  while (z-y>0) {
-    y = y+1;
-  }
-  while (y>9) {
-    y=y-10;
-  }
-   \end{lstlisting}
-
-$1.1|[y,z]| + 0.1|[0,y]|$
+  while (y-x>0) {
+    x = x+1;
+  }
+  while (x>2) {
+    x=x-3;
+  }
+   \end{lstlisting}
+
+$1.33|[x,y]| + 0.33|[0,x]|$
 \\[.7\baselineskip] 
@@
 negative and non-negative numbers.  Note that there are no
-restrictions on the signs of $z$ and $y$.  We also see again that we
-accurately track the size change of $y$ in the first loop.
+restrictions on the signs of $y$ and $x$.  We also see again that we
+accurately track the size change of $x$ in the first loop.
 Furthermore, \emph{t08} shows that we do not handle the constants $1$
@@
 derive a tight bound.  The only information, that the analyzer needs
-is $y \geq c$ before assigning $y = y - c$.
+is $x \geq c$ before assigning $x = x - c$.
 
@@
 
-% $\max(0, (\max(0, (y + -9)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
-$\max(0, \! y{-}9)\dots$ &
+% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+$\max(0, \! y{-}2)\dots$ &
 $O(n)$ &

### 2014-07-03 paper/main.tex
started comparison table
figure caption

@@
 
+\begin{figure*}
+\centering
+\begin{tabular}{r|c|c|c|c|c}
+& KoAT & Rank & LOOPUS & SPEED & AAA \\
+\hline
+\#bounds &
+\\
+\#lin.\ bounds &
+\\
+\#best bounds &
+
+\end{tabular}
+\caption{Comparison of our automatic amortized analysis with other
+  automatic tools.  We have not been able to obtain a version of
+  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
+  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
+  does currently not work on C programs and we use the bounds that
+  have been reported in the author's experimental evaluation.}
+\label{fig:compar}
+\end{figure*}
+
 \section{Related Work}
@@
 Speed &
-Amortized Automation
+AAA
 \\

### 2014-07-03 paper/main.tex
add comparison results
explaining the experiments

@@
 
+We have experimentally evaluated the practicability of our automatic
+amortized analysis with more than 30 challenging loop and recursion
+patterns from open-source code and the
+literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
+examples together with the derived bounds is given in \pref{app:cat}.
+
+\pref{fig:compar} shows four representative examples from the
+evaluation.  Example \emph{t08} is a slightly modified version of
+\emph{t08a} which is described in the introduction.  Example
+\emph{t19} demonstrates the compositionaliy of the analysis.  The
+program consists of two loops that decrement a variable $i$.  In the
+first loop, $i$ is decremented down to 100 and in the second loop $i$
+is increment further down to $-1$.  However, between the loops we
+assign $i=i+k+50$.  So in total the program performs $52 + |[-1,i]| +
+|[0,k]|$ increments.  Our analysis finds this tight bound because our
+amortized analysis naturally takes into account the relation between
+the two loops.  Other techniques derive a more conservative bound.
+
+At first sight, example \emph{t30} appears to be a simple loop that
+decrements the variable $x$ down to zero.  However, a closer look
+reveals that the loop actually decrements both input variables $x$ and
+$y$ down to zero before terminating.  In the loop body, first $x$ is
+decremented by one.  Then the values of the variables $x$ and $y$ are
+switched using the local variable $t$ as a buffer.  Our analysis
+infers the tight bound $|[0,x]|+|[0,y]|$.  Sometimes we need some
+assumptions on the inputs in order to derive a bound.  Example
+\emph{t15} is such a case.  We assume here that the input variable $y$
+is non-negative and write \code{assert(y>=0)}.  If we enter the loop
+then we know that $x>0$ and we can obtain constant potential from the
+assignment \code{x=x-1}.  After the assignment we know that $x\geq y$
+and $y\geq 0$.  As a consequence, we can share the potential
+$|[0,x]|$ before the assignment \code{x=x-y} between $|[0,x]|$ and
+$|0,y|$ after the assignment.  In this way, we derive a tight linear
+bound.
+
+To compare our tool with existing work, we focused on loop bounds and
+use a simple metric that counts the number of back edges (i.e., number
+of loop iterations) that are follows in the execution of the program
+because most other tools only bound this specific cost. In
+\pref{fig:compar}, we show the bounds we derived (AAA) together with
+the bounds derived by LOOPUS~\cite{SinnZV14} and
+Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
+have not been able to obtain this tool.
+KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
+cannot operate on C code and the examples would need to be manually
+translated into a term-rewriting to be analyzed by these tools.  For
+Rank it is not totally clear how the computed bound relates to the C
+program since the computed bound is for transitions in an automaton
+that is derived from the C code.  For instance, the bound $2+y-x$ that
+is derived for \emph{t08}.
+
 \begin{figure*}
@@
 \hline
-& t08a (slightly modified t08) & t19 & t30 & t15
+& t08 (slightly modified t08a) & t19 & t30 & t15
 %
@@
 Rank &
-{\small $2 {+} y {-} x (???)$ } &
+{\small $2 {+} y {-} x (?)$ } &
 {\small $54+k+i$ } &
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
@@
 \hline
-\#bounds &
+\#bounds & 9 & 24 & 20 & 14 & 32
 \\
-\#lin.\ bounds &
+\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
 \\
-\#best bounds &
-
+\#best bounds & 0 & 0 & 11 & 14 & 29
+\\
+\#tested & 14 & 33 & 33 & 14 & 33
 \end{tabular}
@@
 \label{fig:compar}
-\end{figure*}
+\end{figure}
 
@@
 \multicolumn{2}{|c|}{Rank} &
-\multicolumn{2}{|c|}{Loopus} &
-Speed &
+\multicolumn{2}{|c|}{LOOPUS} &
+SPEED &
 AAA
@@
 
-$\frac n m + n$ &
+$\frac n m + m$ &
 
@@
 
-$1 + \max(x, y)$ &
-$O(n)$ &
+--- &
+&
 

### 2014-07-03 paper/main.tex
experiments

@@
 that is derived from the C code.  For instance, the bound $2+y-x$ that
-is derived for \emph{t08}.
+is derived for \emph{t08} only applies to the first loop in the
+program.
+
+\pref{tab:compar} summarizes the results of our experiments.  It shows
+for each tool the number of derived bounds (\#bounds), the number of
+asymptotically tight bounds (\#lin. bounds), the number of bounds with
+the best constant factors in comparison with the other tools (\#best
+bounds), and the number of examples that we were able to test with the
+tool (\#tested).  Since we were not able to run the exponents for KoAT
+and SPEED, we simply used the bounds the have been reported by the
+authors of the respective tools.  The results show that our automatic
+amortized analysis outperforms the existing tools on our example
+programs.  However, this experimental evaluation has to be taken with
+a grain of salt.  Our goal in this work is not to set a new standard
+for automatic bound analysis but only to show that our approach has
+advantages on certain types of examples.  Overall the existing tools
+are more powerful since they can derive polynomial bounds and support
+more features of C.  We were particularly impressed by LOOPUS which is
+very robust, works on large C files, and derives very precise bounds.
+We did not include the running times of the tools in the table since
+all the tested tools work very efficiently and need less then a second
+on all tested examples.
+
+Since we have a formal cost semantics, we can run our examples with
+this semantics for different inputs and measure the cost to compare it
+to our derived bound.  \pref{fig:3d} shows such a comparison for
+example \emph{t08}.  One can see the derived constant factors are the
+best possible if the input variable $x$ is non-negative.  If $x$ is
+negative then the bound is only slightly off.
 
@@
 
-\begin{figure}
+\begin{table}
 \centering
@@
   have been reported in the author's experimental evaluation.}
-\label{fig:compar}
-\end{figure}
+\label{tab:compar}
+\end{table}
 
@@
 
-\begin{figure*}
+\begin{table*}
 \small
@@
 \label{fig:eval}
-\end{figure*}
+\end{table*}
 

### 2014-07-03 paper/main.tex
fix Relax proof

@@
 \begin{align*}
-\Phi_{Q'}(\state) &= \sum_i q'_i f_i(H) \\
+\Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
 &\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
-  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(bH)  \\ & \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
+  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
+  \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
   + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
  &\ge \Phi_Q(\state)
-  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
-  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
+  + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
+  + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
  &\ge \Phi_Q(\state).
 \end{align*}
-Since $H \models \Gamma$, using the transitivity of $\models$ and
+Since $\state \models \Gamma$, using the transitivity of $\models$ and
 the definition of $\mathcal U$ and $\mathcal L$ we get
-$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
-$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
+$\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
+$\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
 These two inequalities justify the last step.

### 2014-07-03 paper/main.tex
updates to intro

@@
 
-\category{D.2.4}{Software Engineering}
-{Software/Program Verification}
-\category{F.3.1}{Logics and Meanings of Programs}
-{Specifying and Verifying and Reasoning about Programs}
-
-\terms Verification, Reliability
-
-\keywords Formal Verification, Compiler Construction, Program Logics,
-Stack-Space Bounds, Quantitative Verification
+% \category{D.2.4}{Software Engineering}
+% {Software/Program Verification}
+% \category{F.3.1}{Logics and Meanings of Programs}
+% {Specifying and Verifying and Reasoning about Programs}
+
+% \terms Verification, Reliability
+
+% \keywords Formal Verification, Compiler Construction, Program Logics,
+% Stack-Space Bounds, Quantitative Verification
 
@@
 \item Existing techniques often rely on complex external tools such as
-  abstract interpretation based invariant
+  abstract interpretation--based invariant
   generation~\cite{GulwaniMC09} or translation of the program into a
@@
   an experimental evaluation that our novel amortized resource
-  analysis works precisely and efficiently for common loop and
+  analysis works precisely and efficiently for challenging loop and
   recursion patterns, and that the derived constant factors in the
@@
 %
+To the best of our knowledge, this article presents the first resource
+analysis framework for C that makes it possible to combine non-trivial
+automatically derived bounds with interactively derived bounds in a
+proof system that produces verifiable certificates for the bounds.
+Our approach complements existing work since it provides a semantic
+foundation for the computation of bounds while still providing support
+for automation that sometimes goes beyond the capabilities of existing
+techniques.  Both the automatic amortized analysis and the
+quantitative logic are naturally compositional and describe the
+resource behavior of code fragments without referring to the source
+code.  As a result, we directly derive global bounds that are
+functions of the input parameters of the program.  Another unique
+feature of our system is, that it can handle resources like memory
+that may become available during the execution.
+
 Our starting point is a recent work~\cite{veristack14} in which we
@@
 % stack bounds in the quantitative Hoare logic automatically.  
-By compiling these bounds together with the C program, we have derived
-verified stack bounds for x86 assembly code.
-
+% By compiling these bounds together with the C program, we have derived
+% verified stack bounds for x86 assembly code.
+%
 Our \emph{first contribution} is a generalization of the quantitative
 Hoare logic for CompCert Clight that enables us to derive resource
-bounds that are parametric in the resource of interest.  The
-quantitative logic is implemented in Coq and proved sound with respect
-to a small-step operational cost semantics for Clight that counts a
-constant resource consumption for each atomic execution step.  The
-resource consumption at each step is parametric in a user-definable cost
-metric and can be negative to express the release of resources (as
-needed, e.g., to reason about memory usage).  Furthermore, a specific
-\emph{tick} function can be inserted in the source code to further
-customize the cost model.  In the implementation of the logic, we use
+bounds that are parametric in the resource of interest.  % The
+% quantitative logic is implemented in Coq and proved sound with respect
+% to a small-step operational cost semantics for Clight.
+% that counts a
+% constant resource consumption for each atomic execution step.  The
+% resource consumption at each step is parametric in a user-definable cost
+% metric and can be negative to express the release of resources (as
+% needed, e.g., to reason about memory usage).  Furthermore, a specific
+% \emph{tick} function can be inserted in the source code to further
+% customize the cost model.
+In the implementation of the logic, we use
 a shallow embedding in Coq which makes the logic very flexible.  There
@@
 an ample foundation for reasoning about resource consumption of Clight
-programs.  The logic is naturally compositional and the derived
-\emph{quantitative Hoare triples} can be used to specify the resource
-consumption of library functions or to describe resource contracts
-between different functions.  However, reasoning about quantitative
+programs.  % The logic is naturally compositional and the derived
+% \emph{quantitative Hoare triples} can be used to specify the resource
+% consumption of library functions or to describe resource contracts
+% between different functions. 
+However, reasoning about quantitative
 properties can often be more tedious than reasoning about intentional
@@
 bounds with tight constant factors.  This type-based amortized
-resource analysis has been our main inspiration in the design of the
+resource analysis has been an inspiration in the design of the
 quantitative Hoare logic and is therefore a natural candidate to
@@
 
-Our \emph{second contribution} is an automatic amortized resource
+Our \emph{main contribution} is an automatic amortized resource
 analysis for CompCert Clight that computes derivations in the
@@
 resource analysis that can derive bounds that depend on negative
-integers and handles programs with mutually recursive functions as well as
-\code{break} and \code{return} statements.  It targets programs that
-are common in system software, that is, programs whose resource
+integers and handles programs with mutually recursive functions as
+well as \code{break} and \code{return} statements.  Using
+potential-based approach of amortized analysis allows us to
+automatically handle programs that are need special treatment in other
+techniques such as
+\begin{itemize}
+\item interaction between \emph{sequential loops or function calls} through
+  size changes of variables,
+\item \emph{nested loops} that influence each others number of iterations,
+\item and \emph{tricky iteration patterns} as found for instance in
+  the Knuth-Morris-Pratt algorithm for string search.
+\end{itemize}
+%
+The main innovation that makes the analysis practical for C is the use
+of interval sizes in potential functions instead of the sizes of
+variables.  This leads to precise bounds of programs whose resource
 consumption can be described as a function of sizes $|[x,y]| =
@@
 %
-The use of interval sizes in potential functions (an therefore in
-resource bounds) is the main innovation that makes the analysis
-practical for C.
 
@@
 SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
-abstract-interpretation based methods.  In contrast with abstract-interpretation
-based methods, our technique does not require any
+abstract interpretation--based methods such as disjunctive invariant
+generation~\cite{GulwaniMC09} and symbolic backward
+execution~\cite{GulwaniZ10}.  In contrast with abstract
+interpretation--based methods, our technique does not require any
 fixpoint computations to obtain loop invariants.  The mechanism we
@@
 the extension to polynomial bounds shall work smoothly.
-
-To the best of our knowledge, this article presents the first resource
-analysis framework for C that makes it possible to combine non-trivial
-automatically derived bounds with interactively derived bounds in a
-proof system that produces verifiable certificates for the bounds.
-Our approach complements existing work since it provides a semantic
-foundation for the computation of bounds while still providing support
-for automation that sometimes goes beyond the capabilities of existing
-techniques.  Both the automatic amortized analysis and the
-quantitative logic are naturally compositional and describe the
-resource behavior of code fragments without referring to the source
-code.  As a result, we naturally derive global bounds that are
-functions of the input parameters of the program.  Another unique
-feature of our system is, that it can handle resources like
-memory that may become available during the execution.
 

### 2014-07-03 paper/main.tex
changes in section 2

@@
 
-\section{Informal Account}
+\section{Overview and Examples}
 \label{sec:inform}
@@
 The purpose of this section is to build some intuition that should
-make the following sections, which contain more details, easy to read.
+make the following sections easy to read.
 
@@
 
-The idea that drives the design of our system is amortized
-analysis~\cite{Tarjan-amort}.  Assume that a program $S$
-executes on a starting state $\state_0$ and consumes $n$ resource units
-of some quantity we are interested in.  We denote that by writing $(S,
-\state_0) \Downarrow_n$.  The basic idea of amortized analysis is to
-define a potential function $\Phi$ that maps program states to
-non-negative numbers and to show that $\Phi(\state) \geq n$ if
-$\state$ is a program state such that such that $(S, \state)
-\Downarrow_n$.
-
-Since programs can be composed we must enrich the evaluation judgement
-$(S, \state) \Downarrow_{n}$ to talk about the state resulting from
-the program execution.  We write $(S, \state) \Downarrow_n \state'$ to
-say that $\state'$ is the program state after the execution of $S$ in
-starting state $\sigma$.  We now use two potential functions, one that
-applies before the execution, and one that applies after.  The two
-functions must respect the relation $\Phi(\state) \ge n +
-\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(S,
-\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
-provide enough \emph{potential} for both, paying for resource cost of
-the computation and paying for the potential $\Phi'(\state')$ on the
-resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
-\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get
-$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
-\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
-+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
-provides an upper bound on the resource consumption of the whole
-program.  What we have observed is that, if we define $\htriple{\Phi}
-{S}{\Phi'}$ to mean
+The idea that drives the design of our framework is amortized
+analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
+starting state $\state$ and consumes $n$ resource units of some
+quantity we are interested in.  We denote that by writing $(S, \state)
+\Downarrow_n \state'$ where $\state'$ the program state after the
+execution.  The basic idea of amortized analysis is to define a
+\emph{potential function} $\Phi$ that maps program states to non-negative
+numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
+program state such that such that $(S, \state) \Downarrow_n \state'$.
+
+To obtain a compositional analysis we also have to take into account
+the state resulting from the program execution.  We thus use two
+potential functions, one that applies before the execution, and one
+that applies after.  The two functions must respect the relation
+$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
+$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
+$\Phi(\state)$ must provide enough \emph{potential} for both, paying
+for resource cost of the computation and paying for the potential
+$\Phi'(\state')$ on the resulting state $\state'$. That way, if
+$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
+\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
+$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed to
+$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
+potential function $\Phi$ provides an upper bound on the resource
+consumption of the whole program.  What we have observed is that, if
+we define $\htriple{\Phi} {S}{\Phi'}$ to mean
 $$
@@
 Similarly, other language constructs lead to rules for the potential
-functions that look very similar to Hoare logic or effect systems
+functions that look very similar to Hoare logic or effect system
 rules.  These rules enable interactive reasoning about resource usage
@@
 
-We have implemented and proved sound such a quantitative Hoare for
-CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
-embedding in Coq, every function that can be expressed in Coq can in
-principle be used as a resource bound.  We can use the logic to prove
-a concrete bound for a given resource metric, or a bound that is
-parametric in a set of resource metrics.  It is also possible to
-incorporate boolean preconditions into the bounds to express that the
-bound is only valid for a certain class of inputs.  To this end, we
-allow the potential function $\Phi$ in the pre and post conditions to
-take non-negative number or $\infty$ (infinity) as values.  Infinity,
-plays the same role as $\mathit{false}$ in Hoare logic.  Boolean assertions can
-be embedded into potential functions by mapping $\mathit{false}$ to $\infty$ and
+% We have implemented and proved sound such a quantitative Hoare for
+% CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
+% embedding in Coq, every function that can be expressed in Coq can in
+% principle be used as a resource bound.
+
+It is also possible to incorporate boolean preconditions into the
+bounds to express that the bound is only valid for a certain class of
+inputs.  To this end, we allow the potential function $\Phi$ in the
+pre and post conditions to take non-negative number or $\infty$
+(infinity) as values.  Infinity, plays the same role as
+$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
+into potential functions by mapping $\mathit{false}$ to $\infty$ and
 $\mathit{true}$ to $0$.
@@
   $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
-  0 to other operations.
+  tick metric that assigns cost $n$ to the statement \lstinline{tick(n)} and
+  0 to all other operations.
   }
@@
 
-It is for instance possible to derive the following quantitative
-triple for a binary-search that holds for all stack metrics $M$.  A
-stack metric assigns cost $0$ to all evaluation steps except function
-calls.  Before a call $f(x)$, we consume $M_f > 0$ resources and
-after the call we return $M_f$ resources that can be used in
-subsequent function calls.  So $M_f$ corresponds to the stack-frame
-size of the function $f$.
+We can use the logic to prove a concrete bound for a given resource
+metric, or a bound that is parametric in a set of resource metrics.
+For example, it is possible to derive the following quantitative
+triple for a binary-search function \code{bsearch} that holds for all
+stack metrics $M$.  
 %
@@
 ${ \{ Z{\cdot}M_\code{bsearch} \} }$
-
 \end{center}
 %
-In the pre and post condition, the logical variable $Z$ is used to
-relate the size of the input with the potential that is left after the
-function call.  The logical part $Z = \log_2(h-l)$ of the precondition
-is $0$ if $Z = \log_2(\state_0(h)-\state_0(l))$ in the initial state
-$\state_0$ and $\infty$ otherwise.
-
-Assume for the remainder of this section that we are interested in
-bounding the number of \emph{ticks} in a program.  In other words, we
-use a resource metric that assigns cost $n$ to the function call
-\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
-is an integer and a negative number means that $-n$ resources are
-returned.
+A stack metric assigns cost $0$ to all evaluation steps except
+function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
+and after the call we return $M_f$ resources that can be used in
+subsequent function calls.  So $M_f$ corresponds to the stack-frame
+size of the function $f$.  In the pre- and postcondition, the logical
+variable $Z$ is used to relate the size of the input with the
+potential that is left after the function call.  The logical part $Z =
+\log_2(h-l)$ of the precondition is $0$ if $Z =
+\log_2(\state_0(h)-\state_0(l))$ in the initial state $\state_0$ and
+$\infty$ otherwise.
+
+For simplicity, assume for the remainder of this section that we are
+interested in bounding the number of \emph{ticks} in a program.  In
+other words, we use a resource metric that assigns cost $n$ to the
+function call \code{tick(n)} and cost $0$ to all other operations.
+The argument $n$ is an integer and a negative number means that $-n$
+resources are returned.
 
@@
 the number of $1$'s in the array $a$.  We have inserted the statement
-\code{tick(1)} into the program whenever we update the array $a$.
+\code{tick(1)} into the program whenever we update the array $a$.  A
+naive analysis of the algorithm would lead to a quadratic bound since
+there can be a linear number of updates in the inner loop.
 
@@
 
-See Section XXX for more a formal definition of the logic and
+See \pref{sec:logic} for more a formal definition of the logic and
 additional example derivation.
@@
 that it becomes possible to use \emph{linear programming} (LP) to
-compute a derivation in the logic.  If we assume that a program state
-$\state$ simply maps variable to integers then we require
+compute a derivation in the logic.  If we assume for now that a
+program state $\state$ simply maps variable to integers then we
+require
 $$
@@
 only do so if we utilize the fact that $x<y$ at the beginning of the
-loop body.  The reasoning then works as follows.  We start with the
-potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
-assignment.  If we denote the updated version of $x$ after the
+loop body.  % The reasoning then works as follows. We start with the
+% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
+% assignment.  
+If we denote the updated version of $x$ after the
 assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
@@
 established the loop invariant again.
+
+Note that our notion of a potential function is a generalization of the
+concept of a ranking function.  A potential function can be used like
+a ranking function if we use the tick metric and add the statement
+\code{tick(1)} to every back edge of the program (loops and function
+calls).  However, a potential function is more flexible.  For example,
+we can use a potential function to prove that Example~2 does not
+consume any resources in the tick metric.
+\begin{lstlisting}
+  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
+\end{lstlisting}
+Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
+In both cases, we reason exactly like in the first version of the
+while loop to prove the bound.  Of course, such loops with different
+tick annotations can be seamlessly combined in a larger program.
+\begin{lstlisting}
+  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
+\end{lstlisting}
+%
+Also note that there are no restrictions on the signs of the input
+variables in the examples.
 
@@
 \end{figure}
-More formally, we develop (in section XXX) a judgement 
+%
+\newlength{\progwidth}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.22\linewidth}
+  \centering
+\hspace{-0.4cm}
+  \begin{minipage}[b]{.18\linewidth}
+    \begin{center}
+   \begin{lstlisting}[]
+while (n>x) {
+  ${\color{blue}%
+  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+  if (m>y) 
+    ${\color{blue}%
+    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
+    y = y+1;
+    ${\color{blue}%
+    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  else
+    ${\color{blue}%
+    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+    x = x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
+}
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.7\baselineskip]
+      {\bf speed\_1}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+while (x<n) {
+  ${\color{blue}%
+  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+  if (z>x)
+    ${\color{blue}%
+    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  else
+    ${\color{blue}%
+    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    z=z+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[z, n]|$
+\\[.7\baselineskip]
+      {\bf speed\_2}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (z-y>0) {
+  ${\color{blue}%
+  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y = y+1;
+  ${\color{blue}%
+  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(3);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+${\color{blue}%
+\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+while (y>9) {
+  ${\color{blue}%
+  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y=y-10;
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+   \end{lstlisting}
+
+$3.1|[y,z]| + 0.1|[0,y]|$
+\\[.7\baselineskip] 
+      {\bf t08a}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n<0) {
+  ${\color{blue}%
+  n{<}0; \,  P(n,y) \vdash }$
+  n=n+1;
+  ${\color{blue}%
+  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
+  y=y+1000;
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  while (y>=100 && *){
+    ${\color{blue}%
+    y{>}99; \, 9 {+} P(n,y) \vdash }$
+    y=y-100;
+    ${\color{blue}%
+    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
+    tick(5);
+    ${\color{blue}%
+    \dashv \cdot; \, 9 {+} P(n,y) }$
+  }
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  tick(9);
+  ${\color{blue}%
+  \dashv \cdot; \, P(n,y) }$
+}
+   \end{lstlisting}
+$59|[n,0]| {+} 0.05|[0,y]|$
+\\[.7\baselineskip]
+      {\bf t27}
+    \end{center}
+  \end{minipage}
+
+  \caption{Derivations of bounds on the number of ticks for
+    challenging examples.  In the potential functions, we only mention
+    the non-zero terms and in the logical context $\Gamma$ we only
+    mention assertions that we actually use in the reasoning. In
+    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
+    {+} 0.05|[0,y]|$.}
+  \label{fig:ex_list}
+\end{figure*}
+%
+%
+More formally, we develop (\pref{sec:AAA}) a judgement 
 $$
@@
 cost.  It is important to note that the reasoning about assertions in
-$\Gamma$ are very basic.  We do not perform any fixpoint computations
+$\Gamma$ is very basic.  We do not perform any fixpoint computations
 and only derive trivial loop invariants about variables that are
@@
 
-Note that our notion of a potential function is a refinement of the
-concept of a ranking function.  A potential function can be used like
-a ranking function if we use the tick metric and add the statement
-\code{tick(1)} to every back edge of the program (loops and function
-calls).  However, a potential function is more flexible.  For example,
-we can use a potential function to prove that Example~2 does not
-consume any resources in the tick metric.
-\begin{lstlisting}
-  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
-\end{lstlisting}
-Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
-In both cases, we reason exactly like in the first version of the
-while loop to prove the bound.  Of course, such loops with different
-tick annotations can be seamlessly combined in a larger program.
-\begin{lstlisting}
-  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
-\end{lstlisting}
-%
 \pref{fig:ex1} shows a derivation of the bound
@@
 assertions in $\Gamma$ are only used in rule for the assignment
-\code{x=x+K}.  To the best of our knowledge, no other tool is
+\code{x=x+K}.  To the best of our knowledge, no other tool for C is
 currently capable of deriving a tight bound on the cost of such a
@@
 without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
-computes the bound $x + y + 10$, PUBS~\cite{AlbertAGPZ12} computes
-the tight bound $0.1(y-x)$, Rank~\cite{AliasDFG10} computes the bound XXX, and
-LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.
+computes the bound $x + y + 10$, Rank~\cite{AliasDFG10} computes the
+bound $y-x-7$, and LOOPUS~\cite{SinnZV14} computes the bound
+$y-(x+9)$.  Only PUBS~\cite{AlbertAGPZ12} computes the tight bound
+$0.1(y-x)$ if we translate the program into a term-rewriting system by
+hand.
 
@@
 
-To understand the capabilities of the automatic analysis, it is
-instructive to study more involved examples.  \pref{fig:ex_list}
-contains Examples \emph{speed\_1} and \emph{speed\_2} that are taken
-from previous work on bound analysis~\cite{GulwaniMC09}.  The SPEED
-tool~\cite{GulwaniMC09} derives the bounds as our analysis but
-requires inference of a \emph{disjunctive invariant}.  In the abstract
+As mentioned, the automatic analysis can handle challenging example
+programs without special tricks or techniques.  Examples
+\emph{speed\_1} and \emph{speed\_2} that are taken from previous work
+on bound analysis~\cite{GulwaniMC09} and demonstrate the our method
+can handle \emph{tricky iteration patterns}.  The SPEED
+tool~\cite{GulwaniMC09} derives the same bounds as our analysis but
+requires inference of \emph{disjunctive invariants}.  In the abstract
 interpretation community, these invariants are known to be notoriously
@@
 %
-Example \emph{speed\_2} is more tricky and trying to understand how it
-works may be challenging.  However, with the amortized analysis in
-mind, using the potential transfer reasoning, it is almost trivial to
-prove a bound.  While the SPEED tool has to find a fairly
-involved invariant for the loop, our tool is simply reasoning locally
-and works without any clever tricks. We obtain the tight bound $|[x,
-n]| + |[z, n]|$.
-
-Example \emph{t08a} shows the ability of the analysis to handle
-negative and non-negative numbers.  Note that there are no
-restrictions on the signs of $z$ and $y$.  We also see that we can
-accurately track the size change of $y$ in the first loop.
-Furthermore, \emph{t08a} shows that we do not handle the constants $1$
-or $0$ in any special way.  In all examples you could replace $0$ and
-$1$ with other constants like we did in the second loop and still
-derive a tight bound.  The only information, that the analyzer needs
-is $y \geq c$ before assigning $y = y - c$.
-
-Example \emph{t27} shows how amortization can be used to handle tricky
-nested loops.  The outer loop we increment the variable $n$ till $n =
-0$.  In each of the $|[n,0]|$ iterations, we increment the variable
-$y$ by $1000$.  We then non-deterministically execute an inner loop
-that decrements $y$ by $100$ until $y<100$.  The analysis can derive
-that only the first execution of the inner loop depends on the initial
-value of $y$.  We again derive tight constant factors.
-
-As mentioned, the analysis naturally handles advanced control flow
-like \code{break} and \code{return} statements, and mutual recursion.
+Example \emph{speed\_2} is more tricky and we found it hard to
+manually find a bound.  However, with the amortized analysis in mind,
+using the potential transfer reasoning, it is almost trivial to prove
+the tight bound $|[x, n]| + |[z, n]|$.
+
+Example \emph{t08a} shows the ability of the analysis to discover
+interaction between \emph{sequential loops or function calls} through
+size changes of variables.  We accurately track the size change of $y$
+in the first loop by transferring the potential $0.1$ form $|[y,z]| to
+|[0,y]|$.  Furthermore, \emph{t08a} shows that we do not handle the
+constants $1$ or $0$ in any special way.  In all examples you could
+replace $0$ and $1$ with other constants like we did in the second
+loop and still derive a tight bound.  The only information, that the
+analyzer needs is $y \geq c$ before assigning $y = y - c$.
+
+Example \emph{t27} shows how amortization can be used to handle
+\emph{interacting nested loops}.  The outer loop we increment the variable $n$
+till $n = 0$.  In each of the $|[n,0]|$ iterations, we increment the
+variable $y$ by $1000$.  We then non-deterministically execute an
+inner loop that decrements $y$ by $100$ until $y<100$.  The analysis
+can derive that only the first execution of the inner loop depends on
+the initial value of $y$.  We again derive tight constant factors.
+%
+
+As mentioned, the analysis also handles advanced control flow like
+\code{break} and \code{return} statements, and mutual recursion.
 \pref{fig:ex_rec} contains two mutually-recursive functions with their
@@
 count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis system can compute a tight bound.  On this example, the
-analysis amounts to computing the meeting point of two trains that
-approach each other with different speeds.
+analysis system is the only available system that can compute a tight
+bound on this example.  The analysis amounts to computing the meeting
+point of two trains that approach each other with different speeds.
 % This is comparable to the
@@
 
-We only show a very small selections of the programs that we can
-handle automatically here.  In \pref{app:cat} you find a list of more
-than $30$ classes of challenging programs that we can automatically
-analyze.  We hope that this gives you an impression of the
-capabilities of the analysis but we also point out that this list is
-far from being exhaustive.
-
-\newlength{\progwidth}
-
-\begin{figure*}[t!]
- \setlength{\progwidth}{.22\linewidth}
-  \centering
-\hspace{-0.4cm}
-  \begin{minipage}[b]{.18\linewidth}
-    \begin{center}
-   \begin{lstlisting}[]
-while (n>x) {
-  ${\color{blue}%
-  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-  if (m>y) 
-    ${\color{blue}%
-    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
-    y = y+1;
-    ${\color{blue}%
-    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  else
-    ${\color{blue}%
-    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-    x = x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
-}
-   \end{lstlisting}
-
-$|[x, n]| + |[y, m]|$
-\\[.7\baselineskip]
-      {\bf speed\_1}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-
-while (x<n) {
-  ${\color{blue}%
-  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-  if (z>x)
-    ${\color{blue}%
-    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    x=x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  else
-    ${\color{blue}%
-    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    z=z+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[z, n]|$
-\\[.7\baselineskip]
-      {\bf speed\_2}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (z-y>0) {
-  ${\color{blue}%
-  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y = y+1;
-  ${\color{blue}%
-  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(3);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-${\color{blue}%
-\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-while (y>9) {
-  ${\color{blue}%
-  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y=y-10;
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-   \end{lstlisting}
-
-$3.1|[y,z]| + 0.1|[0,y]|$
-\\[.7\baselineskip] 
-      {\bf t08a}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (n<0) {
-  ${\color{blue}%
-  n{<}0; \,  P(n,y) \vdash }$
-  n=n+1;
-  ${\color{blue}%
-  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
-  y=y+1000;
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  while (y>=100 && *){
-    ${\color{blue}%
-    y{>}99; \, 9 {+} P(n,y) \vdash }$
-    y=y-100;
-    ${\color{blue}%
-    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
-    tick(5);
-    ${\color{blue}%
-    \dashv \cdot; \, 9 {+} P(n,y) }$
-  }
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  tick(9);
-  ${\color{blue}%
-  \dashv \cdot; \, P(n,y) }$
-}
-   \end{lstlisting}
-$59|[n,0]| {+} 0.05|[0,y]|$
-\\[.7\baselineskip]
-      {\bf t27}
-    \end{center}
-  \end{minipage}
-
-  \caption{Derivations of bounds on the number of ticks for
-    challenging examples.  In the potential functions, we only mention
-    the non-zero terms and in the logical context $\Gamma$ we only
-    mention assertions that we actually use in the reasoning. In
-    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
-    {+} 0.05|[0,y]|$.}
-  \label{fig:ex_list}
-\end{figure*}
+We only show a small selections of the programs that we can handle
+automatically here.  In \pref{app:cat} is a list of more than $30$
+classes of challenging programs that we can automatically analyze.
+\pref{sec:exper} contains a more detailed comparison with other tools
+for automatic bound derivation.
 
@@
 \section{Quantitative Hoare Logic}
+\label{sec:logic}
 
@@
 \section{Automatic Amortized Analysis}
+\label{sec:AAA}
 
@@
 \section{Experimental Evaluation}
+\label{sec:exper}
 

### 2014-07-04 paper/main.tex
explanations for "well-known algorithms"

@@
-% Some notes:
-%  1) Use Appel's SHA-256 implementation as an example
-%
-
 \documentclass[nocopyrightspace,preprint]{sigplanconf}
@@
 
+\pref{fig:cat3} shows well-known algorithms that can be automatically
+analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
+the search function of the Knuth-Morris-Pratt algorithm for string
+search.  To derive a bound for this algorithm we have add an assertion
+that indicates the bounds of the values that are stored in the array
+\code{b}.  Using this information we derive a tight linear bound.
+
+Example \emph{Greatest Common Divisor} is the usual implementation of
+the GCD algorithm.  We automatically derive a linear bound.  Note that
+the two tests at the beginning that check if the inputs are positive
+are essential.  Finally, example \emph{Quick Sort} shows a skeleton of
+the quick sort sorting algorithm.  Since we can only derive linear
+bounds we left out the inner loop that swaps the array elements and
+determines the position \code{n+lo} of the pivot.  We only assert that
+\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
+
 \begin{table*}

### 2014-07-04 paper/main.tex
started preparing popl submission

@@
-\documentclass[nocopyrightspace,preprint]{sigplanconf}
+%Comment for anonymization
+%\def\fullversion{}
+
+\documentclass[nocopyrightspace]{sigplanconf}
+
+\newcommand{\iffull}[2]{\ifx\fullversion\undefined{#2}\else{#1}\fi}
+\newcommand{\ifshort}[2]{\ifx\fullversion\undefined{#1}\else{#2}\fi}
+
+\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
+\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
+\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
+\newcommand{\paraskip}[0]{\ifshort{\vspace{-2pt}}{}}
+\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
+\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
+\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}
+
+ % \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
+ %  \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
+ %  \captionsetup[figure]{format=myformat}
+ % \renewcommand{\captionlabelfont}{\bf}
 
@@
 
+% \abovedisplayskip=4pt
+% \belowdisplayskip=4pt
+% \abovedisplayshortskip=4pt
+% \belowdisplayshortskip=4pt
+
 \conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
@@
 
-\title{Compositional Certified Resource Bounds}
+\title{Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}}
 % Certified Resource Bounds for C
@@
 \authorinfo
-{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
-{Yale University}
+{\iffull{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}{\vspace{-1cm}}}
+{\iffull{Yale University}{\vspace{-1cm}}}
 
@@
   combined to semi-automatically derive global resource bounds.
-
+  %
   An experimental evaluation demonstrates the practicability of the
@@
 
+\sectskip
 \section{Introduction}
 \label{sec:intro}
+\aftersectskip
 
@@
 software, there still exist shortcomings that hamper the application
-of the existing techniques in practice.
+of existing techniques in practice.
+\itemskip
 \begin{itemize}
-\item Analysis tools are black boxes that either deliver a result or fail
+\item Analysis tools are black boxes that \iffull{either}{} deliver a result or fail
   without enabling \emph{user interaction} to manually or
@@
   program.
+\itemskipIn
 \item The computed bounds are often \emph{non-compositional} local
@@
   global whole program bounds.
+\itemskipIn
 \item Existing techniques often rely on complex external tools such as
@@
 \end{itemize}
+\itemskip
 %
@@
 %
+\itemskip
 \begin{quote}
@@
   don't really understand my reasoning processes at all!
-  \vspace{-3ex}
+  \vspace{-4ex}
   \begin{flushright}
@@
 \end{quote}
+\itemskip
 %
@@
 that is based on a solid semantic foundation.  The choice of C is
-primarily motivated by our ongoing work on the formally verified
+primarily motivated by \iffull{our}{the} ongoing work on the formally verified
 hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
@@
 %
+\itemskip
 \begin{enumerate}
@@
   and can be negative if resources are released during an execution.
+
 \item We develop a quantitative Hoare logic for interactively deriving
@@
   semantics of Clight.
+\itemskipIn
 \item We describe an automatic static analysis that computes bounds
@@
   fixpoint computations to obtain loop invariants.
+\itemskipIn
 \item We show with a publically available prototype implementation and
@@
 \end{enumerate}
+\itemskip
 %
@@
 
-Our starting point is a recent work~\cite{veristack14} in which we
-have formally verified bounds on the stack usage of C programs.  A key
-part of this stack-bound verification is based on a quantitative Hoare
-logic for deriving abstract stack bounds for CompCert Clight programs
-that depend on the depth of (recursive) function calls.  % For
+Our starting point is a recent work~\cite{veristack14} \iffull{in
+  which we have}{that has} formally verified bounds on the stack usage
+of C programs.  A key part of this stack-bound verification is based
+on a quantitative Hoare logic for deriving abstract \emph{stack
+  bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
+  depth of (recursive) function calls.}  % For
 % non-recursive programs it is possible to compute derivations for such
@@
 introduced by \code{break} and \code{return} statements.
-Unfortunately, these are the functions with non-constant resource
-consumption that are most common in CertiKOS and other system
+Unfortunately, these are the functions \iffull{with non-constant resource
+consumption}{} that are most common in CertiKOS and other system
 software.
@@
 resource analysis that can derive bounds that depend on negative
-integers and handles programs with mutually recursive functions as
-well as \code{break} and \code{return} statements.  Using
+integers. It also handles programs with mutually recursive functions as
+well as \code{break} and \code{return} statements.  Using the
 potential-based approach of amortized analysis allows us to
-automatically handle programs that are need special treatment in other
-techniques such as
+use a single mechanism to analyze programs that need special treatment in other
+techniques.  For instance, we can naturally handle
+\itemskip
 \begin{itemize}
@@
   size changes of variables,
+\itemskipIn
 \item \emph{nested loops} that influence each others number of iterations,
+\itemskipIn
 \item and \emph{tricky iteration patterns} as found for instance in
@@
 
+\sectskip
 \section{Overview and Examples}
 \label{sec:inform}
+\aftersectskip
 
@@
 
+\sectskip
 \section{Syntax and Semantics}
 \label{sec:sem}
+\aftersectskip
 
@@
 \label{lem:sem1}
+\itemskip
   \begin{enumerate}
@@
     \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
+\itemskipIn
   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
@@
   \end{enumerate}
+\itemskip
 \end{lemma}
@@
 
+\sectskip
 \section{Quantitative Hoare Logic}
 \label{sec:logic}
+\aftersectskip
 
@@
 
+\sectskip
 \section{Automatic Amortized Analysis}
 \label{sec:AAA}
+\aftersectskip
 
@@
 the main ideas of the rule as follows.
+\itemskip
 \begin{itemize}
@@
   pre/postcondition.
+\itemskipIn
 \item
@@
   if $x$ and $y$ are local variables,
+\itemskipIn
 \item
@@
 \end{itemize}
+\itemskip
 If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
@@
 
+% \sectskip
 % \section{Implementation}
@@
 
+% \sectskip
 % \section{Polynomial Potential}
@@
 
+\sectskip
 \section{Experimental Evaluation}
 \label{sec:exper}
+\aftersectskip
 
@@
 
+\sectskip
 \section{Related Work}
+\label{sec:related}
+\aftersectskip
 
@@
 
+\sectskip
 \section{Conclusion}
+\label{sec:concl}
+\aftersectskip
 
@@
 
+\sectskip
 \bibliographystyle{abbrvnat}

### 2014-07-04 paper/main.tex
more updates to the intro

@@
 quantitative Hoare logic and is therefore a natural candidate to
-automate the reasoning.  However, it is a well-known problem that
-type-based amortized resource analysis works well for functional
-programs with pattern matching but not for C-like programs with
-control flow that depends integer arithmetic.  For example, it is a
-long-time open problem how to use type-based amortized analysis to
-derive bounds for functions whose resource consumption depends on
-(possibly negative) integers and non-sequential control flow as
-introduced by \code{break} and \code{return} statements.
-Unfortunately, these are the functions \iffull{with non-constant resource
-consumption}{} that are most common in CertiKOS and other system
-software.
+automate the reasoning.  Type-based amortized resource analysis works
+well for functional programs with pattern matching but it has been a
+long-time open problem to apply it to C-like programs with control
+flow that depends integer arithmetic, negative numbers, and non-linear
+control flow.
+% example, it is a
+% long-time open problem how to use type-based amortized analysis to
+% derive bounds for functions whose resource consumption depends on
+% (possibly negative) integers and non-sequential control flow as
+% introduced by \code{break} and \code{return} statements.
+% Unfortunately, these are the functions \iffull{with non-constant resource
+% consumption}{} that are most common in CertiKOS and other system
+% software.
 
@@
 resource analysis that can derive bounds that depend on negative
-integers. It also handles programs with mutually recursive functions as
-well as \code{break} and \code{return} statements.  Using the
-potential-based approach of amortized analysis allows us to
-use a single mechanism to analyze programs that need special treatment in other
-techniques.  For instance, we can naturally handle
+integers. It also handles programs with mutually recursive functions
+as well as \code{break} and \code{return} statements.  The
+potential-based approach of amortized analysis provides a single
+mechanism for analyzing programs that need special treatment in other
+techniques; including
 \itemskip
@@
 We have evaluated the automatic analysis with system code and examples
-from the literature.  \pref{app:cat} contains more than $30$
+from the literature.  \pref{app:cat} \ifshort{in the extended version
+  of this article\cite{anon_extended}}{} contains more than $30$
 challenging loop and recursion patterns that we collected from open
 source software and the literature.  Our analysis can find
-asymptotically tight bounds for $30$ all but 3 of these patterns and
-in most cases the derived constant factors are tight.  To compare our
+asymptotically tight bounds for all but 3 of these patterns and in
+most cases the derived constant factors are tight.  To compare our
 automatic analyzer with existing techniques, we tested our examples
-with the tools KoAT~\cite{BrockschmidtEFFG14},
-PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and
-LOOPUS~\cite{SinnZV14}.  We also contacted the authors of
-SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
-comparison.  Our experiments show that the bounds that we derive are
-often more precise than those derived by existing tools.  Only
-LOOPUS~\cite{SinnZV14}, which also uses amortization techniques, is
-able to achieve a similar precision.
-
+with the tools such as KoAT~\cite{BrockschmidtEFFG14},
+Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
+show that the bounds that we derive are often more precise than those
+derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
+uses amortization techniques, is able to achieve a similar precision.
+%
 Several micro benchmarks demonstrate the practicability and
@@
 bound that amortizes the cost of $n$ successive increments to a binary
-counter.  XXX In a larger case study, we show how the quantitative
-logic can be used to derive the amortized cost for data-structure
-operations for a table that expands or shrinks depending on the load
-factor.
+counter.
 

### 2014-07-04 paper/main.tex
more work on POPL version

@@
-%Comment for anonymization
+% Comment for anonymization
+% Also look for "SHORT" comments
+
 %\def\fullversion{}
 
-\documentclass[nocopyrightspace]{sigplanconf}
+\documentclass[nocopyrightspace,preprint]{sigplanconf}
 
@@
 
- % \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
- %  \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
- %  \captionsetup[figure]{format=myformat}
- % \renewcommand{\captionlabelfont}{\bf}
+%SHORT
+\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
+\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
+\captionsetup[figure]{format=myformat}
+\renewcommand{\captionlabelfont}{\bf}
+%SHORT
 
@@
 \input{definitions}
-
 \begin{document}
 
-% \abovedisplayskip=4pt
-% \belowdisplayskip=4pt
-% \abovedisplayshortskip=4pt
-% \belowdisplayshortskip=4pt
+%SHORT
+\lstset{basicstyle=\tt\scriptsize}
+\abovedisplayskip=3pt
+\belowdisplayskip=3pt
+\abovedisplayshortskip=3pt
+\belowdisplayshortskip=3pt
+%SHORT
 
@@
 logic and the automatic amortized analysis for for Clight programs.
-The purpose of this section is to build some intuition that should
-make the following sections easy to read.
+% The purpose of this section is to build some intuition that should
+% make the following sections easy to read.
 
@@
 To obtain a compositional analysis we also have to take into account
-the state resulting from the program execution.  We thus use two
+the state resulting from a program's execution.  We thus use two
 potential functions, one that applies before the execution, and one
@@
 $\mathit{true}$ to $0$.
-
+%
 \begin{figure}
@@
 \end{figure}
-
+%
+\iffull{
+}{}
 We can use the logic to prove a concrete bound for a given resource
@@
 \begin{center}
-${ \{ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
+${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
 \code{bsearch(x,l,h)}
@@
 potential that is left after the function call.  The logical part $Z =
-\log_2(h-l)$ of the precondition is $0$ if $Z =
-\log_2(\state_0(h)-\state_0(l))$ in the initial state $\state_0$ and
+\log_2(h-\ell)$ of the precondition is $0$ if $Z =
+\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
 $\infty$ otherwise.
@@
 For simplicity, assume for the remainder of this section that we are
-interested in bounding the number of \emph{ticks} in a program.  In
-other words, we use a resource metric that assigns cost $n$ to the
-function call \code{tick(n)} and cost $0$ to all other operations.
-The argument $n$ is an integer and a negative number means that $-n$
-resources are returned.
+interested in bounding the number of \emph{ticks} in a program.  Thus,
+we use a resource metric that assigns cost $n$ to the function call
+\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
+is an integer and a negative number means that $-n$ resources are
+returned.
 
@@
 
+\begin{figure}[t]
+  \centering
+\begin{lstlisting}[mathescape]
+${\color{blue}%
+  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+while (x+K<=y) {
+  ${\color{blue}%
+  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  x=x+K; 
+  ${\color{blue}%
+  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  tick(T);
+  ${\color{blue}%
+  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+}
+${\color{blue}%
+ \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+\end{lstlisting}
+  \caption{Automatic derivation of a tight bound on the number of
+    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
+    $T>0$ are not program variables but denote a concrete constants in
+    the program.  The reasoning works uniformly for all such $K$ and
+    $T$.}
+  \label{fig:ex1}
+\end{figure}
+
+%
+\newlength{\progwidth}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.22\linewidth}
+  \centering
+\hspace{-0.4cm}
+  \begin{minipage}[b]{.18\linewidth}
+    \begin{center}
+   \begin{lstlisting}[]
+while (n>x) {
+  ${\color{blue}%
+  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+  if (m>y) 
+    ${\color{blue}%
+    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
+    y = y+1;
+    ${\color{blue}%
+    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  else
+    ${\color{blue}%
+    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+    x = x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
+}
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.7\baselineskip]
+      {\bf speed\_1}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+while (x<n) {
+  ${\color{blue}%
+  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+  if (z>x)
+    ${\color{blue}%
+    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  else
+    ${\color{blue}%
+    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    z=z+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[z, n]|$
+\\[.7\baselineskip]
+      {\bf speed\_2}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (z-y>0) {
+  ${\color{blue}%
+  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y = y+1;
+  ${\color{blue}%
+  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(3);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+${\color{blue}%
+\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+while (y>9) {
+  ${\color{blue}%
+  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y=y-10;
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+   \end{lstlisting}
+
+$3.1|[y,z]| + 0.1|[0,y]|$
+\\[.7\baselineskip] 
+      {\bf t08a}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n<0) {
+  ${\color{blue}%
+  n{<}0; \,  P(n,y) \vdash }$
+  n=n+1;
+  ${\color{blue}%
+  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
+  y=y+1000;
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  while (y>=100 && *){
+    ${\color{blue}%
+    y{>}99; \, 9 {+} P(n,y) \vdash }$
+    y=y-100;
+    ${\color{blue}%
+    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
+    tick(5);
+    ${\color{blue}%
+    \dashv \cdot; \, 9 {+} P(n,y) }$
+  }
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  tick(9);
+  ${\color{blue}%
+  \dashv \cdot; \, P(n,y) }$
+}
+   \end{lstlisting}
+$59|[n,0]| {+} 0.05|[0,y]|$
+\\[.7\baselineskip]
+      {\bf t27}
+    \end{center}
+  \end{minipage}
+
+  \caption{Derivations of bounds on the number of ticks for
+    challenging examples.  In the potential functions, we only mention
+    the non-zero terms and in the logical context $\Gamma$ we only
+    mention assertions that we actually use in the reasoning. In
+    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
+    {+} 0.05|[0,y]|$.}
+  \label{fig:ex_list}
+\end{figure*}
+%
+%
+
 \subsection{Automatic Amortized Analysis}
@@
 
-\begin{figure}[t]
-  \centering
-\begin{lstlisting}[mathescape]
-${\color{blue}%
-  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-while (x+K<=y) {
-  ${\color{blue}%
-  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  x=x+K; 
-  ${\color{blue}%
-  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  tick(T);
-  ${\color{blue}%
-  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
-}
-${\color{blue}%
- \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
-\end{lstlisting}
-  \caption{Automatic derivation of a tight bound on the number of
-    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
-    $T>0$ are not program variables but denote a concrete constants in
-    the program.  The reasoning works uniformly for all such $K$ and
-    $T$.}
-  \label{fig:ex1}
-\end{figure}
-%
-\newlength{\progwidth}
-
-\begin{figure*}[t!]
- \setlength{\progwidth}{.22\linewidth}
-  \centering
-\hspace{-0.4cm}
-  \begin{minipage}[b]{.18\linewidth}
-    \begin{center}
-   \begin{lstlisting}[]
-while (n>x) {
-  ${\color{blue}%
-  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-  if (m>y) 
-    ${\color{blue}%
-    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
-    y = y+1;
-    ${\color{blue}%
-    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  else
-    ${\color{blue}%
-    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-    x = x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
-}
-   \end{lstlisting}
-
-$|[x, n]| + |[y, m]|$
-\\[.7\baselineskip]
-      {\bf speed\_1}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-
-while (x<n) {
-  ${\color{blue}%
-  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-  if (z>x)
-    ${\color{blue}%
-    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    x=x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  else
-    ${\color{blue}%
-    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    z=z+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[z, n]|$
-\\[.7\baselineskip]
-      {\bf speed\_2}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (z-y>0) {
-  ${\color{blue}%
-  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y = y+1;
-  ${\color{blue}%
-  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(3);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-${\color{blue}%
-\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-while (y>9) {
-  ${\color{blue}%
-  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y=y-10;
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-   \end{lstlisting}
-
-$3.1|[y,z]| + 0.1|[0,y]|$
-\\[.7\baselineskip] 
-      {\bf t08a}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (n<0) {
-  ${\color{blue}%
-  n{<}0; \,  P(n,y) \vdash }$
-  n=n+1;
-  ${\color{blue}%
-  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
-  y=y+1000;
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  while (y>=100 && *){
-    ${\color{blue}%
-    y{>}99; \, 9 {+} P(n,y) \vdash }$
-    y=y-100;
-    ${\color{blue}%
-    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
-    tick(5);
-    ${\color{blue}%
-    \dashv \cdot; \, 9 {+} P(n,y) }$
-  }
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  tick(9);
-  ${\color{blue}%
-  \dashv \cdot; \, P(n,y) }$
-}
-   \end{lstlisting}
-$59|[n,0]| {+} 0.05|[0,y]|$
-\\[.7\baselineskip]
-      {\bf t27}
-    \end{center}
-  \end{minipage}
-
-  \caption{Derivations of bounds on the number of ticks for
-    challenging examples.  In the potential functions, we only mention
-    the non-zero terms and in the logical context $\Gamma$ we only
-    mention assertions that we actually use in the reasoning. In
-    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
-    {+} 0.05|[0,y]|$.}
-  \label{fig:ex_list}
-\end{figure*}
-%
-%
 More formally, we develop (\pref{sec:AAA}) a judgement 

### 2014-07-04 paper/main.tex
more work torwards popl deadline

@@
 $|[x, n]| + |[y, m]|$
-\\[.7\baselineskip]
+\\[.4\baselineskip]
       {\bf speed\_1}
@@
 $|[x, n]| + |[z, n]|$
-\\[.7\baselineskip]
+\\[.4\baselineskip]
       {\bf speed\_2}
@@
 $3.1|[y,z]| + 0.1|[0,y]|$
-\\[.7\baselineskip] 
+\\[.4\baselineskip] 
       {\bf t08a}
@@
 $59|[n,0]| {+} 0.05|[0,y]|$
-\\[.7\baselineskip]
+\\[.4\baselineskip]
       {\bf t27}
@@
   \end{minipage}
-
-  \caption{Derivations of bounds on the number of ticks for
-    challenging examples.  In the potential functions, we only mention
-    the non-zero terms and in the logical context $\Gamma$ we only
-    mention assertions that we actually use in the reasoning. In
-    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
-    {+} 0.05|[0,y]|$.}
+\vspace{1ex}
+\caption{Derivations of bounds on the number of ticks for challenging
+  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
+  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
+  \emph{t08a} contains \emph{sequential loops} so that the iterations
+  of the second loop depend on the first, and \emph{t27} contains
+  interacting \emph{nested loops}. In the potential functions, we only
+  mention the non-zero terms and in the logical context $\Gamma$ we
+  only mention assertions that we actually use in the reasoning. In
+  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
+  0.05|[0,y]|$.}
   \label{fig:ex_list}
@@
 
-To this end, we fix the shape of the potential functions $\Phi$ so
-that it becomes possible to use \emph{linear programming} (LP) to
-compute a derivation in the logic.  If we assume for now that a
-program state $\state$ simply maps variable to integers then we
+To enable automation, we fix the shape of the potential functions
+$\Phi$ so that it becomes possible to use \emph{linear programming}
+(LP) to compute a derivation in the logic.  If we assume for now that
+a program state $\state$ simply maps variable to integers then we
 require
@@
 \emph{ticks} metric that assigns cost $n$ to the function call
-\code{tick(n)} and cost $0$ to all other operations then the cost of
-the following example can be bounded by $|[x,y]|$.
-\begin{lstlisting}
+\code{tick(n)} \iffull{and cost $0$ to all other operations}{} then
+the cost of the following example can be bounded by $|[x,y]|$. Note
+that there are no restrictions on the signs of the input variables in
+the examples.
+\begin{lstlisting}[basicstyle=\tt\small]
 while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
@@
 consume any resources in the tick metric.
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\small]
   while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
@@
 tick annotations can be seamlessly combined in a larger program.
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\small]
   while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
@@
 %
-Also note that there are no restrictions on the signs of the input
-variables in the examples.
-
 More formally, we develop (\pref{sec:AAA}) a judgement 
@@
 assertions in $\Gamma$ are only used in rule for the assignment
-\code{x=x+K}.  To the best of our knowledge, no other tool for C is
-currently capable of deriving a tight bound on the cost of such a
-loop.  For $T=1$ (many systems focus on the number of loop iterations
-without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
-computes the bound $x + y + 10$, Rank~\cite{AliasDFG10} computes the
-bound $y-x-7$, and LOOPUS~\cite{SinnZV14} computes the bound
-$y-(x+9)$.  Only PUBS~\cite{AlbertAGPZ12} computes the tight bound
-$0.1(y-x)$ if we translate the program into a term-rewriting system by
-hand.
-
-It might be surprising that we only track simple linear relations in
+\code{x=x+K}.  To the best of our knowledge, no other implemented tool
+for C is currently capable of deriving a tight bound on the cost of
+such a loop.  For $T=1$ (many systems focus on the number of loop
+iterations without a cost model) and $K=10$,
+KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
+Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
+LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
+PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
+translate the program into a term-rewriting system by hand.
+
+To automate the reasoning, we first introduce an unknown rational
+variables for each factor in the potential functions.  We then use our
+inference rules (see \pref{sec:AAA}) to emit linear inequalities on
+these variables that enforce that the potential annotations are sound.
+For instance if $K = 1$ in \pref{fig:ex1} then we would have the
+annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
+|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
+p_{(y,x)} {\cdot} |[y,x]|$ after the assignment where $q_i$ and $p_i$
+are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
+$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.
+
+\iffull{It might be surprising that we}{We} only track simple linear relations in
 the logical context $\Gamma$.  It would of course be possible to keep
@@
 programs without special tricks or techniques.  Examples
-\emph{speed\_1} and \emph{speed\_2} that are taken from previous work
-on bound analysis~\cite{GulwaniMC09} and demonstrate the our method
-can handle \emph{tricky iteration patterns}.  The SPEED
-tool~\cite{GulwaniMC09} derives the same bounds as our analysis but
-requires inference of \emph{disjunctive invariants}.  In the abstract
-interpretation community, these invariants are known to be notoriously
-difficult to handle.
+\emph{speed\_1} and \emph{speed\_2} that are taken from previous
+work~\cite{GulwaniMC09} and demonstrate the our method can handle
+\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
+derives the same bounds as our analysis but requires inference of
+\emph{disjunctive invariants}.  In the abstract interpretation
+community, these invariants are known to be notoriously difficult to
+handle.
 %
@@
 Example \emph{speed\_2} is more tricky and we found it hard to
-manually find a bound.  However, with the amortized analysis in mind,
-using the potential transfer reasoning, it is almost trivial to prove
-the tight bound $|[x, n]| + |[z, n]|$.
+manually find a bound.  However, using potential transfer reasoning of
+amortized analysis, it is straightforward to prove the tight bound
+$|[x, n]| + |[z, n]|$.
 
 Example \emph{t08a} shows the ability of the analysis to discover
-interaction between \emph{sequential loops or function calls} through
-size changes of variables.  We accurately track the size change of $y$
-in the first loop by transferring the potential $0.1$ form $|[y,z]| to
-|[0,y]|$.  Furthermore, \emph{t08a} shows that we do not handle the
-constants $1$ or $0$ in any special way.  In all examples you could
-replace $0$ and $1$ with other constants like we did in the second
-loop and still derive a tight bound.  The only information, that the
-analyzer needs is $y \geq c$ before assigning $y = y - c$.
-
+interaction between \emph{sequential loops} through size changes of
+variables.  We accurately track the size change of $y$ in the first
+loop by transferring the potential $0.1$ form $|[y,z]|$ to $|[0,y]|$.
+Furthermore, \emph{t08a} shows again that we do not handle the
+constants $1$ or $0$ in any special way.  In all examples we could
+replace $0$ and $1$ with other constants like in the second loop and
+still derive a tight bound.  \iffull{The only information, that the analyzer
+needs is $y \geq c$ before assigning $y = y - c$.}{}
+%
 Example \emph{t27} shows how amortization can be used to handle
-\emph{interacting nested loops}.  The outer loop we increment the variable $n$
-till $n = 0$.  In each of the $|[n,0]|$ iterations, we increment the
-variable $y$ by $1000$.  We then non-deterministically execute an
-inner loop that decrements $y$ by $100$ until $y<100$.  The analysis
-can derive that only the first execution of the inner loop depends on
-the initial value of $y$.  We again derive tight constant factors.
+\emph{interacting nested loops}.  In the outer loop we increment the
+variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations, we
+increment the variable $y$ by $1000$.  We then non-deterministically
+(expressed by \code{*}) execute an inner loop that decrements $y$ by
+$100$ until $y<100$.  The analysis discovers that only the first
+execution of the inner loop depends on the initial value of $y$.  We
+again derive tight constant factors.
 %
@@
 count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis system is the only available system that can compute a tight
-bound on this example.  The analysis amounts to computing the meeting
-point of two trains that approach each other with different speeds.
+analysis is the only available system that computes a tight
+bound on this example.  \iffull{The analysis amounts to computing the meeting
+point of two trains that approach each other with different speeds.}{}
 % This is comparable to the
@@
 
-We only show a small selections of the programs that we can handle
-automatically here.  In \pref{app:cat} is a list of more than $30$
+\iffull{We only show a small selections of the programs that we can handle
+automatically here.}{}  In \pref{app:cat} is a list of more than $30$
 classes of challenging programs that we can automatically analyze.
 \pref{sec:exper} contains a more detailed comparison with other tools
-for automatic bound derivation.
+for \iffull{automatic}{} bound derivation.
 
@@
 void count_down (int x,int y) {
-  if (x>y) { tick(1); count_up(x-1,y); }
-}
+  if (x>y) { tick(1); count_up(x-1,y); } }
+
 void count_up (int x, int y) {
-  if (y+1<x) { tick(1); count_down(x,y+2); }
-}
+  if (y+1<x) { tick(1); count_down(x,y+2); } }
+
    \end{lstlisting}
@@
 $\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
-\\[.7\baselineskip]
+\\[.4\baselineskip]
       {\bf t39}
@@
   \end{minipage}
-  \caption{Two mutually-recursive functions with the computed tick bounds.}
+\vspace{-1.5ex}
+
+  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
   \label{fig:ex_rec}
@@
 
+\begin{figure}[t!]
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~x, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~x, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \fenv f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep{M}
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \cont, \cost')
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma 
+\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_2, \cont, \cost')
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
+}
+%
+\and
+\Rule{S:Tick}
+{ }
+{ (\state, \code{tick}(n), \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
+}
+
+\end{mathpar}
+\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
+\label{fig:opsem}
+\end{figure}
+
 \paragraph{Semantics.}
@@
 %
-The evaluation function \evalE{\cdot}{} that maps an expression $e \in
+The evaluation function \evalE{\cdot}{} maps an expression $e \in
 E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.
@@
 are always disjoint.  We just write $\state(x)$ to obtain the value of
-$x$ in program state $\state$.  Such a lookup is defined as
+$x$ in program state $\state$. \iffull{Such a lookup is defined as
 $$
@@
 \right..
-$$
-For a program state $\state$, we similarly write $\state[x \mapsto v]$
+$$}{}
+\iffull{For a program state $\state$, we}{We} similarly write $\state[x \mapsto v]$
 for the program state that maps $x$ to $v$ and behaves as $\state$ for
@@
 
-\begin{figure}[t!]
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:RetSeq}
-{}
-{ (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-\and
-%
-\and
-\Rule{S:RetLoop}
-{}
-{ (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetCall}
-{ \state = (\_, \gamma) \\
-  \state' = (\theta, \gamma)[r \mapsto \state(x)]
-}
-{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_r)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Call}
-{ \fenv f = (\vec x, S_f) \\
-  \state = (\theta, \gamma) \\
-  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
-}
-{ (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep{M}
-(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \cont, \cost')
-}
-%
-\and
-\Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma 
-\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_2, \cont, \cost')
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
-}
-%
-\and
-\Rule{S:Tick}
-{ }
-{ (\state, \code{tick}(n), \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_t(n))
-}
-
-\end{mathpar}
-\caption{Operational semantics of statements}
-\label{fig:opsem}
-\end{figure}
-
-\pref{fig:opsem} contains the reduction rules of the semantics.  The
+\pref{fig:opsem} contains \ifshort{selected}{} the reduction rules of the semantics.  The
 rules define a rewrite system for program configurations of the form
@@
 
+\iffull{
 A continuation $\cont$ represents the context of the execution of a
@@
 defines the function body.  It is used in the rule \textsc{S:Call}.
-
-The intuitive meaning of an evaluation judgement $(\state, S, \cont,
-\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
+}{}
+
+The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
+S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
 following.  If the statement $S$ is executed in program state
@@
 
+\iffull{
 \pref{lem:sem1} summarizes the main properties of the resource
@@
 \end{lemma}
+}{}
 
@@
 Hoare logic that we use in Coq to interactively prove resource bounds.
-
+%
 We generalize classic Hoare logic to express not only classical

### 2014-07-04 paper/main.tex
more formatting and shortening

@@
 We have evaluated the automatic analysis with system code and examples
-from the literature.  \pref{app:cat} \ifshort{in the extended version
-  of this article\cite{anon_extended}}{} contains more than $30$
+from the literature.  \iffull{\pref{app:cat}}{The extended version
+  of this article~\cite{anon_extended}}{} contains more than $30$
 challenging loop and recursion patterns that we collected from open
@@
 %
-\and
+\iffull{\and
 \Rule{S:RetSeq}
@@
 (\state', \code{skip}, \cont, \cost {-} M_r)
-}
+}}{}
 %
@@
 %
-\and
+\iffull{\and
 \Rule{S:Call}
@@
 (\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}
+}}{}
 %
@@
 %
-\and
+\iffull{\and
 \Rule{S:IfTrue}
@@
 (\state, S_2, \cont, \cost')
-}
+}}{}
 %
@@
 %
-\and
+\iffull{\and
 \Rule{S:Tick}
@@
 (\state, \code{skip}, \cont, \cost {-} M_t(n))
-}
+}}{}
 
@@
 \end{lemma}
-}{}
+} 
+{
+The extended version of this article~\cite{anon_extended} contains all rules lemmas 
+that state the main properties of the cost semantics.
+}
 
@@
 tuple) that relates the function arguments with the return value.  The
-precondition also depends on $\vec v$, the value of all the arguments
+precondition also depends on $\vec v$, the values of the arguments
 at the function invocation.  Similarly, the postcondition depends on
@@
 \begin{figure*}[t]
+%SHORT
+\small
+%SHORT
 \begin{mathpar}
 %
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q }{ \code{skip} }{ Q }
+}
+%
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+%
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Call}
+{
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
 \Rule{L:Assert}
@@
 \and
-\Rule{L:Skip}
+\Rule{L:Tick}
 { }
@@
 \htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
-\htriple
-  { M_f + P }
-  { r \gets f(\vec x) }
-  { Q - M_r }
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
 }
@@
   { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
 }
@@
 \and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
 \Rule{L:Weaken}
 {
-  P \models P' \\
-  \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \\
-  Q' \models Q \\
-  B' \models B \\
-  \forall v.\, (R'\,v \models R\,v)
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
 }
@@
 }
-{ \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x }
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
 }
@@
 to account for the execution cost $M_r$ of the return in rule {\sc
-  L:Call} for function calls.
+  L:Call}\iffull{ for function calls.}{.}
 
@@
 cost of the evaluation of $e$ and the update.
-
-Despite its seemingly simplicity, the {\sc L:Seq} rule is important to
-understand how the quantitative Hoare logic works.  We have to define
-it in such a way that it accounts for early exits in statements. For
-instance, if $S_1$ contains a \code{break} statement then $S_2$ will
-never be executed so we must ensure in the \code{break} part $B$ of
-$S_1$'s judgement that the \code{break} part $B$ of of $S_1;S_2$
-holds. The same is true for the return part $R$ of the judgements for
-$S_1$ and $S_2$.  The interaction between the actual pre- and
-postconditions is analogous to standard Hoare logic.  In the
-postcondition of $S_1$ we account for the cost $M_s$ of the execution
-of the sequence.
-
-In the rule {\sc L:Loop}, we ensure that the \code{break} part of the
+\iffull{
+}{}
+The rule {\sc L:Seq} rule is crucial to understand how the
+quantitative Hoare logic works.  To account for early exits of
+statements, we must ensure in the \code{break} part $B$ of $S_1$'s
+judgement that the \code{break} part $B$ of of $S_1;S_2$ holds. The
+same is true for the return part $R$ of the judgements for $S_1$ and
+$S_2$.  The interaction between the actual pre- and postconditions is
+analogous to standard Hoare logic.  In the postcondition of $S_1$ we
+account for the cost $M_s$ of the execution of the sequence.
+
+In the rule {\sc L:Loop}, the \code{break} part of the
 loop body $S$ becomes the postcondition of the loop statement. We use
 an arbitrary $B$ as the \code{break} part of the judgement for
-$\code{loop}\, S$ since its operational semantics prevents it from
-terminating differently than by a \code{skip} or a \code{return}.  The
+$\code{loop}\, S$ since its operational semantics ensures that it can
+only terminate with a \code{skip} or a \code{return}.  The
 precondition $I$ of the loop is the loop invariant.  That is why we
 require to have potential $I$ available in the precondition of the
-loop body $S$.  In the post condition of $S$, the potential must be
+loop body $S$.  In the postcondition of $S$, the potential must be
 sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
@@
 
+\iffull{
 The {\sc L:If} is similar to the rule for the conditional in classic
@@
 statement that depends on the integer $n$.
+}{}
 
@@
 parameter is instantiated by the call rule using the result of the
-evaluation of the argument variables in the current state.  Recall
-that we require that $P_f$ and $Q_f$ do not depend on the local
-variables on the call stack, that is, $\forall z \, \vec v \, \theta
-\, \theta' \, \gamma . \, P_f \, z \, \vec v \, (\theta,\gamma) = P_f
-\, z \, \vec v \, (\theta',\gamma)$.  To transfer potential that
-depends on local variables of the callee from the precondition $P$ to
-the postcondition $Q$, we use an assertion $A : \Assn$ that is
-independent of global variables, that is, $\forall \theta \, \gamma \,
-\gamma' . \, A (\theta,\gamma) = A(\theta,\gamma')$.  It is still
-possible to express relations between global and local variables using
-logical variables (see the following paragraph).
+evaluation of the argument variables in the current state.
+\iffull{Recall that we require that $P_f$ and $Q_f$ do not depend on
+  the local variables on the call stack, that is, $\forall z \, \vec v
+  \, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
+  (\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.}{} To
+transfer potential that depends on local variables of the callee from
+the precondition $P$ to the postcondition $Q$, we use an assertion $A
+: \Assn$ that is independent of global variables, that is, $\forall
+\theta \, \gamma \, \gamma' . \, A (\theta,\gamma) =
+A(\theta,\gamma')$.  It is still possible to express relations between
+global and local variables using logical variables (see the
+\iffull{following paragraph}{extended version} for details).
 
@@
 
+\iffull{
 \paragraph{Logical Variables and Shallow Embedding.}
@@
 body is proved for every possible logical variable $y$.
+}{}
 
@@
 where we write $\# S$ for the cardinal of the set $S$.  $A(1)$ is
-obviously bounded by $N$, the size of the array.  So any automated tool
+bounded by $N$, the size of the array.  So any automated tool
 available would derive the linear bound $2 \cdot N$ for that program.
@@
 in the literature.
+
+\begin{figure*}[t]
+%SHORT
+\small
+%SHORT
+\begin{mathpar}
+%
+\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+{ }
+{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!!}
+%
+\and
+\Rule{Q:Break}
+{ }
+{\!\!  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') \!\!}
+%
+\and
+\Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
+{ }
+{ \!\! B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) \!\!}
+%
+\and \Rule{Q:Return}
+{ P = Q_R[\Vret/x]  
+\\ \forall i \in \dom{P} . \, p_i = q_i
+}
+{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
+%
+\and \Rule{Q:Update}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
+%
+\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+%
+\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\ \textstyle q'_{0y} = q_{0y} 
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:Inc}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:If}
+{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+}
+{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+%
+\and\Rule{Q:Seq}
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+}
+{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+%
+\and\Rule{Q:Call}
+{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
+\\ \text{Loc} = \text{Locals}(Q) \!
+\\ \forall i \neq j . \, x_i \neq x_j \!
+\\ c \in \Qplusz \!
+\\ Q = P + S \!
+\\ Q' = P' + S
+\\ U = Q_f[\Vargs / \vec x]
+\\ U' = Q'_f[\Vret/r]
+\\ \forall i \in \dom{U} . \, p_i = u_i
+\\ \forall i \in \dom{U'} .\, p'_i = u'_i 
+\\ \forall i \not\in \dom{U'} .\, p'_i = 0
+\\ \forall i \not\in \text{Loc}
+  .\, s_i = 0
+}
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
+  \vdash r \gets f(\vec x) \dashv
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
+}
+%
+\and 
+\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
+{ }
+{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
+%
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ \fenv f = (\vec y, S_f)
+\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
+}
+{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+%
+\and \Rule{Q:Weak}
+{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\ Q_1 \succeq_{\Gamma_1} Q_2 
+\\ \Gamma'_2 \models \Gamma'_1 
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1 
+}
+{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
+%
+\and \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
+\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
+\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
+}
+{ Q' \succeq_\Gamma Q }
+
+\end{mathpar}
+\caption{Inference rules of the quantitative analysis.}
+\label{fig:auto}
+\end{figure*}
 
@@
 weakening because we have to show that the preconditions of the rules
-{\sc L:Weak} or {\sc L:Update}, respectively, are met.  For instance
+{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
 we prove the following lemma to show the soundness of the translation
@@
 to construct certificates for upper bounds derived automatically.
-
-\begin{figure*}[t]
-\begin{mathpar}
-%
-\Rule{Q:Skip}
-{}
-{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
-%
-\and \Rule{Q:Break}
-{}
-{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
-%
-\and \Rule{Q:Return}
-{ P = Q_R[\Vret/x]  
-\\ \forall i \in \dom{P} . \, p_i = q_i
-}
-{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
-%
-\and \Rule{Q:Assert}
-{}
-{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and \Rule{Q:Tick}
-{}
-{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
-%
-\\ \Rule{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
-{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
-%
-\and \Rule{Q:Update}
-{ q'_{xy}, q'_{yx} \in \Qplusz
-\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
-%
-\and \Rule{Q:IncP}
-{ \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
-\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:DecP}
-{ \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
-%
-\and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
-}
-{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
-%
-\and \Rule{Q:Weak}
-{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
-\\\\ \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ \Gamma'_2 \models \Gamma'_1
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-}
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
-%
-\and\Rule{Q:Call}
-{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
-\and \text{Loc} = \text{Locals}(Q)
-\and \forall i \neq j . \, x_i \neq x_j
-\and c \in \Qplusz
-\\ Q = P + S
-\and Q' = P' + S
-\and U = Q_f[\Vargs / \vec x]
-\and U' = Q'_f[\Vret/r]
-\\ \forall i \in \dom{U} . \, p_i = u_i
-\and \forall i \in \dom{U'} .\, p'_i = u'_i 
-\and \forall i \not\in \dom{U'} .\, p'_i = 0
-\and \forall i \not\in \text{Loc}
-  .\, s_i = 0
-}
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
-  \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
-}
-%
-\and \Rule{Q:Extend}
-{ \fenv f = (\vec y, S_f)
-\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
-}
-{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
-%
-\and \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
-\and \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
-\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
-\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
-}
-{ Q'   \succeq_\Gamma Q }
-
-\end{mathpar}
-\caption{Inference rules of the quantitative analysis}
-\label{fig:auto}
-\end{figure*}
+}{}
 
@@
 
-Most closely related to this article is our previous work on
-end-to-end stack-bound verification of Clight
-programs~\cite{veristack14}.  In that work, we have implemented and
-verified a quantitative logic to reason about stack-space usage.  In
-this article, we present a more general quantitative Hoare logic that
-is parametric in the resource of interest.  The main innovation over
-our previous work~\cite{veristack14} is the novel automatic amortized
-analysis for Clight programs that competes logical derivations for
-non-trivial bounds for programs with loops and recursion.
+Most closely related to this article is \iffull{our}{a} previous work
+on end-to-end stack-bound verification of Clight
+programs~\cite{veristack14}.  In that work, \iffull{we}{the authors}
+have implemented and verified a quantitative logic to reason about
+stack-space usage.  In this article, we present a more general
+quantitative Hoare logic that is parametric in the resource of
+interest.  The main innovation is the novel automatic amortized analysis for
+Clight programs that competes logical derivations for non-trivial
+bounds for programs with loops and recursion.
 
@@
 
+\iffull{
 \clearpage
@@
 \end{table*}
+}{}
 

### 2014-07-06 paper/main.tex
more shortening for popl

@@
 cost metric that we use simply counts the assignments performed by the
-program.  Hence, the resource consumption is closely related to the
-number $A(1)$ of times the test \lstinline{a[i] > m} is true during the
+program.  Hence, the resource cost is closely related to the
+number of times the test \lstinline{a[i] > m} is true during the
 execution.
 %
-We observe that $A(1)+1$ is the count of ``maximum candidates'' in the
-array \lstinline{a} seen by the algorithm. We define
+If we define
 $$
@@
 $$
-where we write $\# S$ for the cardinal of the set $S$.  $A(1)$ is
-bounded by $N$, the size of the array.  So any automated tool
-available would derive the linear bound $2 \cdot N$ for that program.
-But with the expressivity of our logic it is possible to use the
-previous set cardinal directly and precisely tie the bound to the
-initial contents of the array.
-
-The non-trivial part of this derivation is finding the loop
-invariant for the while loop. To this end
-it is necessary to define the number $A(i)$ of ``maximum candidates so far
-starting from $i$ in the array \lstinline{a}.''  When the condition
-\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
-so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
-thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
-assignment.  In the other case, no maximum so far is encountered so
-$A(i) = A(i+1)$.  % As a side remark, if $K \ge 0$, it is possible to
+where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
+the number of ``maximum candidates'' in the array \lstinline{a} seen
+by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
+So any automated tool would at best derive the linear bound $2 \cdot
+N$ for that program.  But with the expressivity of our logic it is
+possible to use the previous set cardinal directly and precisely tie
+the bound to the initial contents of the array.
+%
+The non-trivial part of this derivation is finding the loop invariant
+$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
+When the condition \lstinline{a[i] > m} is true, we know that we
+encountered a ``maximum so far'' because $m$ is a maximal element of
+\lstinline{a[$0 \dots i$]}, thus $A(i) = 1 + A(i+1)$ and we get one
+potential unit to pay for the assignment.  In the other case, no
+maximum so far is encountered so $A(i) = A(i+1)$.
+% As a side remark, if $K \ge 0$, it is possible to
 % have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
@@
 variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
-  + Z \cdot F}{}{Z \cdot F}$ to express that the stack required by the
-function is fully returned afterwards.
-
-The critical step in this proof is the application of the
-{\sc L:Call} rule to prove the recursive call.  At this point
-the context $\Delta$ contains the following specification
-for the function \code{bsearch}
-$$
+  + Z \cdot M_\code{bsearch}}{}{Z \cdot M_\code{bsearch}}$ to express that the stack required by the
+function is returned after the call.
+%
+The critical step in the proof is the application of the
+{\sc L:Call} rule to the recursive call.  At this point
+the context $\Delta$ contains the specification
+$
   \forall y \, (x,l,h) \, \_.
-  ( (\lambda y \, (\_, l, h) .\, (y {=} \log_2(h {-} l)) + y {\cdot} F)
-  , (\lambda y \, \_ .\, y {\cdot} F)
-  ).
-$$
-Using the {\sc L:Call} rule it is possible to instantiate
-$y$ with $Z - 1$, and because $M_f = F$ and $M_r = -F$ in
-the stack resource metric, the appropriate triple can be
-derived on the call site.  The rest of the proof does not
-involve any resource manipulation and is just bookkeeping
-of logical assertions.
+  ( (\lambda y \, (\_, l, h) .\, (y = \log_2(h {-} l)) {+} y {\cdot} M_\code{bsearch})
+  , (\lambda y \, \_ .\, y {\cdot} M_\code{bsearch})
+  )
+$.
+%
+Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
+- 1$, and because $M_f = M_\code{bsearch}$ and $M_r =
+-M_\code{bsearch}$ in the stack metric.  The rest of the proof does
+not involve any resource manipulation and is just bookkeeping of
+logical assertions.
 
@@
 ${\color{blue}%
-  \{ (Z = \log_2(h-l)) + Z \cdot F \} }$
+  \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
 bsearch(x,l,h) {
@@
     ${\color{blue}%
-      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
+      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
     m = h + (h-l)/2;
     ${\color{blue}%
-      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
+      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
     if (a[m]>x) h=m; else l=m;
     ${\color{blue}%
-      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot F + F \} }$
+      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
     l = bsearch(x,l,h); }
   ${\color{blue}%
-    \{ (Z-1) \cdot F - (-F) \} }$
+    \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
   return l;
-} ${\color{blue} \{ Z \cdot F \} }$
+} ${\color{blue} \{ Z \cdot M_\code{bsearch} \} }$
 \end{lstlisting}
 \caption{Example derivation of a stack usage bound for a binary
-  search program.  The metric used sets a cost of $F$ before
-  to function call and $-F$ after the call, $F$ is the stack
-  frame size of the function $\code{bsearch}$.  We use $Z$ as
+  search program.  The used resource metric defines the cost $M_\code{bsearch}$ before
+  the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
+  frame size of the function $\code{bsearch}$.  $Z$ is a
   logical variable.
@@
 \emph{resource safety} $\safe n P S \cont$ of an assertion
-and a program configuration as
+and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
 \begin{align*}
-  \forall \state \, \cost \, m \, \cost' .\,&
-  (m \le n \land P(\state) \le \cost \land
-    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
-  &\implies \cost' \ge 0.
+  (m {\le} n \land P(\state) {\le} \cost \land
+    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
+  \implies \cost' {\ge} 0.
 \end{align*}
-This predicate is step indexed by an integer $n$ that is used in the
-soundness proof for the function-call and loop cases.  The constraint
-$\cost' \geq 0$ imposed by the definition ensures that the program does
-not stop because of a resource error (recall that a negative resource
-counter on the right-hand side is a resource failure).  However, it
-does not rule out memory safety errors or assertion failures. This is
-because our logic does not prove any safety or correctness theorems
-but only focuses on resource usage.
+This predicate is step indexed by an integer $n$ that is used for
+induction in the soundness proof for the function-call and loop cases.
+The constraint $\cost' \geq 0$ imposed by the definition ensures that
+the program does not stop because of a resource error (recall that a
+negative resource counter on the right-hand side is a resource
+failure).  However, it does not rule out memory safety errors or
+assertion failures. This is because our logic does not prove any
+safety or correctness theorems but only focuses on resource usage.
 %
@@
 \cont $.  This would not be the case if we replaced all occurrences of
-$m$ by $n$ in the definition of validity.
+$m$ by $n$ in the definition\iffull{ of validity.}{.}
 
@@
 assertions, one for each of the possible outcomes of a program
-statement.  We define it as follows.
+statement.  \iffull{We define it as follows.
 \begin{align*}
@@
 & & \land& \safe n Q {\code{skip}} \cont
-\end{align*}
+\end{align*}}
+{
+We define $\safeK n B R Q \cont \mathop{:=}
+\safe n B {\code{break}} \cont \, \land \, 
+\safe n Q {\code{skip}} \cont \, \land \, 
+\safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont$.
+
+}
 %
@@
 
-Using the semantic validity of triples we can define the validity
+Using the semantic validity of triples we define the validity
 of a function context $\Delta$, written $\validC n \Delta$, as
@@
 \end{align*}
-where $S_f$ is the body of the function $f$.  A full
+where $S_f$ is the body of the function $f$. \iffull{A full
 judgement that mentions a non-empty function context
 $\Delta$ is in fact a guarded statement: it makes
-assumptions on some functions' behavior.  The predicate
+assumptions on some functions' behavior.}{}  The predicate
 $\validC n \Delta$ gives the precise meaning of the
@@
 %
+We are now able to state the soundness of the quantitative logic.
+%
 \begin{theorem}[Soundness of the logic]
-  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
-  $$
+  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
+  $
     \forall n .\, \validC n \Delta
       \implies \valid {n {+} 1} B R P S Q.
-  $$
+  $
 \end{theorem}
@@
 
-If $S$ is a complete program and $\Delta$ is empty, by
+Assume that $S$ is a complete program and $\Delta$ is empty.  By
 expanding the definitions we see that $\Delta$ is valid
-for every $n$ and that $\Kstop$ is safe for every $n$, so
-we get
+for every $n$ and that $\Kstop$ is safe for every $n$. So
+we derive
 $$
@@
 This means that from any starting state $\state$, $P(\state)$
-provides enough resources for any run of the program $S$.  This
+provides enough resources for any run of the program $S$.  \iffull{This
 setting is actually the main use case of the previous theorem
 which is stated as a stronger result to allow a proof by
-induction.
+induction.}{}
 
@@
 In this section we present the automatic amortized analysis that we
-use in our implementation to derive linear resource bounds for C
+use in our implementation to derive resource bounds for C
 programs.
@@
 
+\bibliographystyle{abbrvnat}
+\iffull{
+\bibliography{lit}
+}
+{
 \sectskip
-\bibliographystyle{abbrvnat}
+\footnotesize
 \bibliography{lit}
+}
 

### 2014-07-06 paper/main.tex
more shortening

@@
 \captionsetup[figure]{format=myformat}
+\captionsetup[table]{format=myformat}
 \renewcommand{\captionlabelfont}{\bf}
@@
 
-\iffull{We only show a small selections of the programs that we can handle
-automatically here.}{}  In \pref{app:cat} is a list of more than $30$
+\iffull{We only show a small selections of the programs that we can
+  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
+  extended version~\cite{anon_extended}} is a list of more than $30$
 classes of challenging programs that we can automatically analyze.
-\pref{sec:exper} contains a more detailed comparison with other tools
-for \iffull{automatic}{} bound derivation.
+\pref{sec:exper} contains a more detailed comparison with other
+tools\iffull{for automatic bound derivation.}{.}
 
@@
 |\inter {\state(x)} {\state(y)}|$.  Using these notations we can
-rewrite the above equality as
+rewrite the above equality as\iffull{
 $$
   \Phi = \sum_{i \in I} q_i f_i.
-$$
-We often write $xy$ to denote the index $(x,y)$.  The family
-$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense) for
-potential functions.  This allows us to uniquely represent any linear
-potential function $\Phi$ as a \emph{quantitative annotation} $Q =
-(q_i)_{i \in I}$, that is, a family of rational numbers where only a
-finite number of elements are not zero.
-%
-For soundness, we require that potential functions always give
-non-negative results.  This is acheived by restricting coefficients
-in a quantitative annotation to be always non-negative.
+$$}
+{ $\Phi = \sum_{i \in I} q_i f_i$.}  We often write $xy$ to denote the
+index $(x,y)$.  The family $(f_i)_{i\in I}$ is actually a basis (in
+the linear-algebra sense) for potential functions.  This allows us to
+uniquely represent any linear potential function
+$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
+is, a family of \ifshort{non-negative}{} rational numbers where only
+a finite number of elements are not zero. \iffull{ For soundness, we
+  require that potential functions always give non-negative results.
+  This is acheived by restricting coefficients in a quantitative
+  annotation to be always non-negative.}{}
 
@@
 determine which intervals will increase or decrease.  The knowledge
-needed by our rules can often be inferred by local reasoning (i.e., in
+needed by our rules can be inferred by local reasoning (i.e., in
 basic blocks without recursion and loops) within usual theories
@@
 as pre- and postconditions for functions. It has proved to be
-sufficient for the variety of examples we designed and found
+sufficient for the variety of examples we are interested in or found 
 in the literature.
@@
 \begin{mathpar}
+\iffull{
 %
@@
 %
-\and \Rule{Q:Return}
+\and }{}
+\Rule{Q:Return}
 { P = Q_R[\Vret/x]  
@@
 %
-\and 
+\iffull{\and 
 \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
@@
 { B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+}{}
+%
+\and
+\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
 { \fenv f = (\vec y, S_f)
-\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
+\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
 }
@@
 \\ \!\! \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2 
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2 
 \\ \Gamma'_2 \models \Gamma'_1 
@@
 \end{mathpar}
-\caption{Inference rules of the quantitative analysis.}
+\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
 \label{fig:auto}
@@
 fixed function context $\Delta$ that is implicit on all judgements.
-Function contexts are treated exactly as in the logic.  We call
-precondition the pair $(\Gamma, Q)$ and postcondition the pair
-$(\Gamma', Q')$.  The correspondence between judgements of the
+Function contexts are treated exactly as in the logic.  % We call
+% precondition the pair $(\Gamma, Q)$ and postcondition the pair
+% $(\Gamma', Q')$.  
+The correspondence between judgements of the
 automation and the ones of the quantitative logic is made formal by
@@
 the resource cost.  This achieved by using a series of \emph{cost-free
-  assignments} that we do not result in additional cost in the
+  assignments} that do not result in additional cost in the
 analysis.  Non-linear operations such as \code{x \gets z*y} are simply
 handled by assigning zero potential to coefficients like $q_{xa}$ and
-$q_{ax}$ that contain $x$.
+$q_{ax}$ that contain $x$ after the assignment.
+
+\begin{figure*}
+\centering
+\begin{tabular}{r|c|c|c|c}
+%
+& t08 (slightly modified t08a) & t19 & t30 & t15\\
+%
+\hline
+&
+\begin{lstlisting}
+while (y > x) {
+  x++; tick(1); }
+while (x > 2) {
+  x -= 3; tick(1); }
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (i > 100) {
+  i--; tick(1) }
+i += k+50;
+while (i >= 0) {
+  i--; tick(1); }
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (x > 0) {
+  x--;
+  t=x, x=y, y=t;
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}
+assert(y >= 0);
+while (x>y) {
+  x -= y+1;
+  for (z=y; z > 0; z--)
+    tick(1);
+  tick(1);
+}
+\end{lstlisting}
+%
+\\
+\hline
+LOOPUS &
+{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
+{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
+{\small % $1 + \max(x, y)$ (???)
+---
+} &
+---
+\\
+%
+%
+%
+%
+Rank &
+{\small $2 {+} y {-} x (?)$ } &
+{\small $54+k+i$ } &
+--- &
+{\small $2+2x-y$ }
+\\
+%
+%
+AAA 
+  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
+  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
+  & $|\inter 0 x| {+} |\inter 0 y|$
+  & $|\inter 0 x|$
+\end{tabular}
+\vspace{.1cm}
+\caption{ Comparison of resource bounds derived by different tools on
+  several examples with linear bounds.  AAA stands for our automatic
+  amortized analysis for Clight.  The output of Rank has been manually
+  simplified to fit the table.  }
+\label{fig:compar}
+\end{figure*}
 
@@
 Most of the rules correspond exactly to the respective rules of the
-quantitative Hoare logic.  For example, the rule {\sc Q:Skip} simply
-reuses its postcondition as precondition.  This is sound since
-evaluating a $\code{skip}$ statement never costs any resources and
-does not change the program state.  Similarly, the rule {\sc Q:Tick}
-enforces that the quantitative annotation of the precondition is at
-least $M_t(n)$ to pay for the resource consumption.  Since the program
-state is also unchanged in this case, the logical part $\Gamma$ of the
-precondition can soundly be reused in the postcondition.  In the rule
-{\sc Q:Break}, where the linear control flow is broken, the
+quantitative Hoare logic.  
+\iffull{
+  For example, the rule {\sc Q:Skip} simply reuses its postcondition
+  as precondition.  This is sound since evaluating a $\code{skip}$
+  statement never costs any resources and does not change the program
+  state.  Similarly, the rule {\sc Q:Tick} enforces that the
+  quantitative annotation of the precondition is at least $M_t(n)$ to
+  pay for the resource consumption.  Since the program state is also
+  unchanged in this case, the logical part $\Gamma$ of the
+  precondition can soundly be reused in the postcondition.  In the
+  rule {\sc Q:Break}, where the linear control flow is broken, the
   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
-taken from the break part of the judgement similarly as it is done in
-the quantitative logic.  The rule {\sc Q:Return} exhibits a slight
+  taken from the break part of the judgement similarly as it is done
+  in the quantitative logic.
+}
+%
+{ Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
+  {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
+  rules can be found in the extended version.
+}
+%
+The rule {\sc Q:Return} exhibits a slight
 difference to the quantitative logic.  While we have a function $\Z
@@
   Q:Inc} for increments and decrements, which describe how the
-potential is distributed after a size change of the variables.  The
-rule {\sc Q:IncP} is for increments $x \gets x + y$ in which we can
-deduce from the logical context $\Gamma$ that $y$ is not negative.
+potential is distributed after a size change of a variable.  The rule
+{\sc Q:IncP} is for increments $x \gets x + y$ in which we can deduce
+from the logical context $\Gamma$ that $y$ is not negative.
 Similarly, {\sc Q:DecP} is for increments $x \gets x - y$ in which we
@@
   Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
-The rules are equivalent in the case where $y=0$.  Finally, the {\sc
-  Q:Inc} can be applied if we cannot deduce any information about $y$.
-In the implementation, the rules for increments and decrements are
-combined into one syntax directed rule where the specialized rules
+The rules are equivalent in the case where $y=0$.  Finally, the rule
+{\sc Q:Inc} can be applied if we cannot deduce any information about
+$y$.  In the implementation, the rules for increments and decrements
+are combined into one syntax directed rule where the specialized rules
 take priority over {\sc Q:Inc}.
@@
 quantitative annotation must be updated to reflect the change of the
-program state.  We write $x'$ the value of $x$ after the assignment.
+program state.  We write $x'$ for the value of $x$ after the assignment.
 Since $x$ is the only variable changed, only intervals of the form
@@
 |\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
-$q_{vx}$ to pay for this change.  Remark that the rule insures this
+$q_{vx}$ to pay for this change.  The rule insures this
 only for $v \not\in \mathcal U$ because we know that $x \le v$
@@
   specification is equalized to its matching potential in the callee's
-  pre/postcondition.
+  pre- and postcondition.
 \itemskipIn
 \item
-  The potential of intervals $|\inter x y|$ is preserved
-  if $x$ and $y$ are local variables,
-\itemskipIn
+  $\!$The potential of intervals $|\inter x y|$ is preserved
+  if $x$ and $y$ are local\iffull{ variables,}{,}
 \item
@@
 If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
-called, the rule takes care to split the potential of
+called, {\sc Q:Call} splits the potential of
 $|\inter x y|$ in two parts, one part to perform the
@@
 later use after the function call.  This splitting is
-realized by the equations $Q = P + S$ and $Q' =
-P' + S'$.  Arguments in the function precondition
+realized by the equations $Q = P {+} S$ and $Q' =
+P' {+} S'$.  Arguments in the function precondition
 $(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
@@
 Symmetrically, we use the unique name $\Vret$ to represent
-the return value in the function postcondition
+the return value in the function's postcondition
 $(\Gamma'_f, Q'_f)$.
@@
 we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition.  We could integrate weakening into every
+sequential composition.  \iffull{We could integrate weakening into every
 syntax directed rule but this simple heuristic helps to make the
-analysis efficient.  The high-level idea of {\sc Q:Weak} is the same
+analysis efficient.}{}  The high-level idea of {\sc Q:Weak} is the same
 as in the rule {\sc L:Weak} of the quantitative logic: If we have a
@@
 and $Q'$ are potential annotations and $\Gamma$ is a logical context.
-The relax rule deals also with the important task of transferring
+The rule {\sc Q:Relax} deals also with the important task of transferring
 constant potential (represented by $q_0$) to interval sizes and vise
@@
 inference rules of the automatic amortized analysis.  During this
-process our tool uses symbolic names for the coefficients $(q_i)$.
-Every time a linear constraint must be satisfied by these
-coefficients, it is recorded using the symbolic names in a global
-list.  When the list of all constraints that must be satisfied by the
-coefficients is collected, we feed it to an off the shelf
-LP-solver~\footnote{We currently use Coin-Or CLP.}.  If the solver
-successfully finds a solution to the given problem, we know that a
-derivation exists for the considered program.  We can then extract the
-initial $Q$ from the solver and get a resource bound for the program.
-To get a full derivation we simply extract the complete solution from
-the solver, and apply it to the symbolic names $(q_i)$ of the
-coefficients in the derivation.  If the LP-solver fails to find a
-solution, an error is reported to the user.
+process our tool uses symbolic names for the rational coefficients
+$(q_i)$ in the rules.  Every time a linear constraint must be
+satisfied by these coefficients, it is recorded in a global list using
+the symbolic names.  When then feed the collected constraints to an
+off the shelf LP-solver~\footnote{We currently use Coin-Or's CLP.}.
+If the solver successfully finds a solution, we know that a derivation
+exists for the considered program.  We can then extract the initial
+$Q$ from the solver and get a resource bound for the program.  To get
+a full derivation we simply extract the complete solution from the
+solver, and apply it to the symbolic names $(q_i)$ of the coefficients
+in the derivation.  If the LP-solver fails to find a solution, an
+error is reported to the user.
 
@@
 translate the assumptions in function contexts of the automatic
-analysis.  This is done as follows.
-%
-\begin{align*}
-\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
-&\forall z \, \vec v \, v . (
-  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs \mapsto \vec v]), \\
-&\phantom{\forall z \, \vec v \, v . (}
-  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret \mapsto v])
+analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) := \forall z \, \vec v \, v . $
+%
+$$
+(
+  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
+  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
   )
-\end{align*}
+$$
 %
@@
 %
-\begin{theorem}[Soundness of the analysis]
+\begin{theorem}[Soundness of the automatic analysis]
   If $J$ is a judgement derived with the rules of
@@
 literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
-examples together with the derived bounds is given in \pref{app:cat}.
+examples \iffull{together with the derived bounds is given in
+  \pref{app:cat}.}{is given in the extended
+  version~\cite{anon_extended}.}
 
@@
 program.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{r|c|c|c|c|c}
+& KoAT & Rank & LOOPUS & SPEED & AAA \\
+\hline
+\#bounds & 9 & 24 & 20 & 14 & 32
+\\
+\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
+\\
+\#best bounds & 0 & 0 & 11 & 14 & 29
+\\
+\#tested & 14 & 33 & 33 & 14 & 33
+\end{tabular}
+\vspace{.1cm}
+\caption{Comparison of our automatic amortized analysis with other
+  automatic tools.  We have not been able to obtain a version of
+  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
+  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
+  does currently not work on C programs and we use the bounds that
+  have been reported in the author's experimental evaluation.}
+\label{tab:compar}
+\end{table}
 
@@
 
-\begin{figure*}
-\centering
-\begin{tabular}{r|c|c|c|c}
-%
-&
-\begin{lstlisting}
-while (y > x) {
-  x++; tick(1); }
-while (x > 2) {
-  x -= 3; tick(1); }
-\end{lstlisting}
-&
-\begin{lstlisting}
-while (i > 100) {
-  i--; tick(1) }
-i += k+50;
-while (i >= 0) {
-  i--; tick(1); }
-\end{lstlisting}
-&
-\begin{lstlisting}
-while (x > 0) {
-  x--;
-  t=x, x=y, y=t;
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}
-assert(y >= 0);
-while (x>y) {
-  x -= y+1;
-  for (z=y; z > 0; z--)
-    tick(1);
-  tick(1);
-}
-\end{lstlisting}
-\\
-%
-\hline
-& t08 (slightly modified t08a) & t19 & t30 & t15
-%
-\\
-\hline
-LOOPUS &
-{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
-{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
-{\small % $1 + \max(x, y)$ (???)
----
-} &
----
-\\
-%
-%
-%
-%
-Rank &
-{\small $2 {+} y {-} x (?)$ } &
-{\small $54+k+i$ } &
---- &
-{\small $2+2x-y$ }
-\\
-%
-%
-AAA 
-  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
-  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
-  & $|\inter 0 x| {+} |\inter 0 y|$
-  & $|\inter 0 x|$
-\end{tabular}
-\caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  AAA stands for our automatic
-  amortized analysis for Clight.  The output of Rank has been manually
-  simplified to fit the table.  }
-\label{fig:compar}
-\end{figure*}
-
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/bound3d}
-\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for example
-  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
-\label{fig:3d}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{r|c|c|c|c|c}
-& KoAT & Rank & LOOPUS & SPEED & AAA \\
-\hline
-\#bounds & 9 & 24 & 20 & 14 & 32
-\\
-\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
-\\
-\#best bounds & 0 & 0 & 11 & 14 & 29
-\\
-\#tested & 14 & 33 & 33 & 14 & 33
-\end{tabular}
-\caption{Comparison of our automatic amortized analysis with other
-  automatic tools.  We have not been able to obtain a version of
-  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
-  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
-  does currently not work on C programs and we use the bounds that
-  have been reported in the author's experimental evaluation.}
-\label{tab:compar}
-\end{table}
-
 \sectskip
@@
 programs.
+
+\begin{figure}[t]
+\includegraphics[width=\linewidth]{fig/bound3d}
+\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
+  (blue lines) and the measured runtime cost (red crosses) for example
+  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
+\label{fig:3d}
+\end{figure}
 
@@
 \sectskip
-\footnotesize
+\scriptsize
 \bibliography{lit}

### 2014-07-06 paper/main.tex
some fixes to make the full version compile
more cleanup for full version
_tick -> tick _main -> main

@@
 % Also look for "SHORT" comments
-
-%\def\fullversion{}
+\def\fullversion{}
 
@@
 
+\iffull{
 \begin{figure}[t!]
@@
 %
-\iffull{\and
+\and
 \Rule{S:RetSeq}
@@
 (\state', \code{skip}, \cont, \cost {-} M_r)
-}}{}
+}
 %
@@
 %
-\iffull{\and
+\and
 \Rule{S:Call}
@@
 (\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}}{}
+}
 %
@@
 %
-\iffull{\and
+\and
 \Rule{S:IfTrue}
@@
 (\state, S_2, \cont, \cost')
-}}{}
+}
 %
@@
 %
-\iffull{\and
+\and
 \Rule{S:Tick}
@@
 (\state, \code{skip}, \cont, \cost {-} M_t(n))
-}}{}
+}
 
@@
 \end{figure}
+}{}
+
+\ifshort{
+\begin{figure}[t!]
+\small
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
+}
+\end{mathpar}
+\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
+\label{fig:opsem}
+\end{figure}
+}{}
 
@@
 
+\iffull{
 \begin{figure*}[t]
-%SHORT
-\small
-%SHORT
 \begin{mathpar}
@@
 \end{figure*}
+}{}
+
+\ifshort{
+\begin{figure*}[t]
+\small
+\begin{mathpar}
+%
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q }{ \code{skip} }{ Q }
+}
+%
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+%
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Call}
+{
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
+  { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
+\and
+\Rule{L:Extend}
+{ \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+\\
+  \forall f\,P_f\,Q_f.\,
+  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
+  \rightarrow
+  \forall y\,\vec v.\,
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\end{mathpar}
+
+\caption{Rules of the Quantitative Hoare Logic}
+\label{fig:logic}
+\end{figure*}
+}{}
 
@@
 
-\iffull{
+\ifx\fullversion\undefined{}\else{
 \paragraph{Logical Variables and Shallow Embedding.}
@@
 body is proved for every possible logical variable $y$.
-}{}
+}\fi
 
@@
 
+\iffull{
 \begin{figure*}[t]
-%SHORT
-\small
-%SHORT
 \begin{mathpar}
-\iffull{
 %
@@
 %
-\and }{}
+\and
 \Rule{Q:Return}
@@
 %
-\iffull{\and 
+\and 
 \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
@@
 { B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-}{}
 %
@@
 \end{figure*}
+}{}
+
+\ifshort{
+\begin{figure*}[t]
+\small
+\begin{mathpar}
+\iffull{
+%
+\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+{ }
+{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!!}
+%
+\and
+\Rule{Q:Break}
+{ }
+{\!\!  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') \!\!}
+%
+\and
+\Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
+{ }
+{ \!\! B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) \!\!}
+%
+\and }{}
+\Rule{Q:Return}
+{ P = Q_R[\Vret/x]  
+\\ \forall i \in \dom{P} . \, p_i = q_i
+}
+{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
+%
+\and \Rule{Q:Update}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
+%
+\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+%
+\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\ \textstyle q'_{0y} = q_{0y} 
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:Inc}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:If}
+{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+}
+{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+%
+\and\Rule{Q:Seq}
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+}
+{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+%
+\and\Rule{Q:Call}
+{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
+\\ \text{Loc} = \text{Locals}(Q) \!
+\\ \forall i \neq j . \, x_i \neq x_j \!
+\\ c \in \Qplusz \!
+\\ Q = P + S \!
+\\ Q' = P' + S
+\\ U = Q_f[\Vargs / \vec x]
+\\ U' = Q'_f[\Vret/r]
+\\ \forall i \in \dom{U} . \, p_i = u_i
+\\ \forall i \in \dom{U'} .\, p'_i = u'_i 
+\\ \forall i \not\in \dom{U'} .\, p'_i = 0
+\\ \forall i \not\in \text{Loc}
+  .\, s_i = 0
+}
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
+  \vdash r \gets f(\vec x) \dashv
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
+}
+%
+\iffull{\and 
+\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
+{ }
+{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
+}{}
+%
+\and
+\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ \fenv f = (\vec y, S_f)
+\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
+}
+{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+%
+\and \Rule{Q:Weak}
+{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2 
+\\ \Gamma'_2 \models \Gamma'_1 
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1 
+}
+{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
+%
+\and \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
+\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
+\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
+}
+{ Q' \succeq_\Gamma Q }
+
+\end{mathpar}
+\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
+\label{fig:auto}
+\end{figure*}
+}{}
 
@@
 while (y > x) {
-  x++; tick(1); }
+  x++; tick(1); 
+}
 while (x > 2) {
-  x -= 3; tick(1); }
+  x -= 3; tick(1); 
+}
 \end{lstlisting}
@@
 while (i > 100) {
-  i--; tick(1) }
+  i--; tick(1) 
+}
 i += k+50;
@@
 
-\begin{table}[h]
+\begin{table}[t]
 \centering
@@
 
-\iffull{
-\clearpage
+\ifx\fullversion\undefined{}\else{
+%\clearpage
 \appendix
@@
       if (*) {
-        _tick(1);
+        tick(1);
         j=j-1; n=n-1;
@@
 code in the \emph{if} statement.  So we use the \emph{tick metric} and
-insert \code{_tick(1)} inside the if statement to derive a bound on
+insert \code{tick(1)} inside the if statement to derive a bound on
 the number of times the code in the if statement is executed.  Note
@@
 }
-void _main (int x,int y) {
+void main (int x,int y) {
   y = copy (x,y);
@@
 
-void _main (int y, int z) {
+void main (int y, int z) {
   count_down(y,z);
@@
   while (x>0) {
-    _tick(-1); x=x-1; y=y+1;
+    tick(-1); x=x-1; y=y+1;
   }
@@
   while (y>0) {
-    y=y-1; x=x+1; _tick(1);
-  }
-}
-void _main (int y, int z) {
+    y=y-1; x=x+1; tick(1);
+  }
+}
+void main (int y, int z) {
   consume(); produce(); consume();
@@
 The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
-\code{\_main} in which we first add $x$ to $y$ using the function
+\code{\main} in which we first add $x$ to $y$ using the function
 \code{copy} and then count down the variable $y$ using the function
@@
 \code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
-\code{count\_down}.  For the function $\_main$, which calls
+\code{count\_down}.  For the function $\main$, which calls
 \code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
@@
 Example \emph{t46} shows a program that uses and returns resources.
-Again, we use the \emph{tick metric} and the function \code{\_tick} to
+Again, we use the \emph{tick metric} and the function \code{\tick} to
 describe the resource usage.  The function \code{produce} produces
@@
 consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
-$|[0,y]|$ for the function \code{\_main}.  This is only possible since
+$|[0,y]|$ for the function \code{\main}.  This is only possible since
 amortized analysis naturally tracks the size changes to the variables
@@
 \end{table*}
-}{}
+}\fi
 

### 2014-07-06 paper/main.tex
more shortening

@@
 % Also look for "SHORT" comments
-\def\fullversion{}
+% \def\fullversion{}
 
@@
 \begin{align*}
-& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
+& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
 & \implies \safe m {P+x} S \cont \; .
@@
 \centering
-\begin{tabular}{r|c|c|c|c}
-%
-& t08 (slightly modified t08a) & t19 & t30 & t15\\
+\begin{tabular}{r|c|c|c|c|c}
+%
+& t08 (modified t08a) & t19 & t30 & t15 & t13\\
 %
@@
 \end{lstlisting}
+&
+\begin{lstlisting}
+  while (x>0) {
+    x=x-1;
+    if (*) y=y+1;
+    else
+      while (y>0) {
+        y=y-1; tick(1); }
+    tick(1); }
+\end{lstlisting}
 %
@@
 \hline
-LOOPUS &
-{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
-{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
-{\small % $1 + \max(x, y)$ (???)
----
-} &
----
+%
+%
+%
+AAA 
+  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
+  & {\small $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ }
+  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
+  & {\small $|\inter 0 x|$}
+  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
 \\
-%
-%
-%
 %
@@
 {\small $54+k+i$ } &
---- &
-{\small $2+2x-y$ }
+{\small ---} &
+{\small $2+2x-y$ } &
+{\small $0.5{\cdot}y^2 {+} yx +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
 \\
 %
-%
-AAA 
-  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
-  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
-  & $|\inter 0 x| {+} |\inter 0 y|$
-  & $|\inter 0 x|$
+LOOPUS &
+{\small 
+  \begin{tabular}{c}
+    $\max(0, x{-}2)$\\ $+ 2\max(0, y {-} x)$
+  \end{tabular}
+} &
+{\small 
+  \begin{tabular}{c}
+    $\max(0, i{-}100)$ \\
+    $+ \max(0, k {+} i {+} 51)$
+  \end{tabular}
+ } &
+{\small ---}% $1 + \max(x, y)$ (???) 
+&
+{\small ---}
+&
+{\small $2\max(x, 0) {+} \max(y, 0)$}
 \end{tabular}
@@
 
-\pref{fig:compar} shows four representative examples from the
+\pref{fig:compar} shows five representative loop patterns from the
 evaluation.  Example \emph{t08} is a slightly modified version of
@@
 is increment further down to $-1$.  However, between the loops we
-assign $i=i+k+50$.  So in total the program performs $52 + |[-1,i]| +
+assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
 |[0,k]|$ increments.  Our analysis finds this tight bound because our
 amortized analysis naturally takes into account the relation between
-the two loops.  Other techniques derive a more conservative bound.
+the two loops.
 
@@
 
+Example \emph{t13} shows how amortization can be used to find linear
+bounds for nested loops.  The outer loop is iterated $|[0,x]|$ times.
+In the conditional, we either (the branching condition is arbitrary)
+increment the variable $y$ or we execute an inner loop in which $y$ is
+counted back to $0$.  The analysis computes a tight bound.
+\iffull{Again, the constants $0$ and $1$ in the
+  inner loop can as well be replace by something more interesting, say
+  $9$ and $10$ like in example \emph{t08}.  Then we still obtain a
+  tight linear bound.}{}
+
 To compare our tool with existing work, we focused on loop bounds and
@@
 cannot operate on C code and the examples would need to be manually
-translated into a term-rewriting to be analyzed by these tools.  For
+translated into a term-rewriting system to be analyzed by these tools.  For
 Rank it is not totally clear how the computed bound relates to the C
@@
 bounds), and the number of examples that we were able to test with the
-tool (\#tested).  Since we were not able to run the exponents for KoAT
+tool (\#tested).  Since we were not able to run the experiments for KoAT
 and SPEED, we simply used the bounds the have been reported by the
@@
 for automatic bound analysis but only to show that our approach has
-advantages on certain types of examples.  Overall the existing tools
+advantages on examples with linear bounds.  Overall the existing tools
 are more powerful since they can derive polynomial bounds and support
@@
 We did not include the running times of the tools in the table since
-all the tested tools work very efficiently and need less then a second
-on all tested examples.
+all tested tools work very efficiently and need less then a second
+on every tested example.
 
@@
 interest.  The main innovation is the novel automatic amortized analysis for
-Clight programs that competes logical derivations for non-trivial
+Clight programs that computes logical derivations for non-trivial
 bounds for programs with loops and recursion.
@@
 \code{break} and \code{return} statements.  Third, for the first time,
-we combined an automatic amortized analysis with a system for
+we have combined an automatic amortized analysis with a system for
 interactively deriving bounds.
@@
 logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
-the general structure of the quantitative logic.  Since we do not deal
+the general structure of the quantitative logic.  \iffull{Since we do not deal
 with memory safety, our logic is much simpler and it would be possible
-to integrate it with Appel's logic.  The continuation passing style
+to integrate it with Appel's logic.}{}  The continuation passing style
 that we use in the quantitative logic is not only used by
@@
 code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
-Hoare-style, does not target C code, and has been mainly designed to
-produce certificates for bounds derived for high-level functional
-programs.
+Hoare-style, does not target C code, and is not designed for
+interactive bound development but to produce certificates for bounds
+derived for high-level functional programs.
 
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{fig/bound3d}
+\center
+\includegraphics[width=.9\linewidth]{fig/bound3d}
 \caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
@@
 language.  However, there is no way to interactively derive bounds or
-to deal with resources like memory usage.  Also Blazy et al.'s
-automatic bound analysis does not compute symbolic bounds.
+to deal with resources like memory usage.  Blazy et al.'s automatic
+bound analysis does also not compute symbolic bounds.
 
@@
 
-We will continue to improve our framework with the goal of reasoning
+We will continue to improve our framework to reasoning more precisely
 about resource usage of system software such as the hypervisor kernel
-CertiKOS~\cite{GuVFSC11}, which we are currently developing and
-verifying.  For on thing, we will improve the automation to derive
-\emph{polynomial bounds} and to handle non-linear size changes, as
-already developed for functional programs~\cite{HoffmannS13}.  For
-another thing, we will build on previous work~\cite{HoffmannMZ13}
-the generalize the quantitative Hoare logic to handle \emph{concurrent
-  programs} with locks and lock-free data structures.
+CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
+  and verifying}{is currently developed and verified}.  For on thing,
+we will improve the automation to derive \emph{polynomial bounds} and
+to handle non-linear size changes, as already developed for functional
+programs~\cite{HoffmannS13}.  For another thing, we will build on
+previous work~\cite{HoffmannMZ13} to generalize the quantitative
+Hoare logic to handle \emph{concurrent programs}\iffull{ with locks and
+lock-free data structures.}{.}
 

### 2014-07-07 paper/main.tex
typo
some typos

@@
 In this section, we informally introduce the quantitative program
-logic and the automatic amortized analysis for for Clight programs.
+logic and the automatic amortized analysis for Clight programs.
 % The purpose of this section is to build some intuition that should
@@
 quantity we are interested in.  We denote that by writing $(S, \state)
-\Downarrow_n \state'$ where $\state'$ the program state after the
+\Downarrow_n \state'$ where $\state'$ is the program state after the
 execution.  The basic idea of amortized analysis is to define a
@@
 numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
-program state such that such that $(S, \state) \Downarrow_n \state'$.
-
-To obtain a compositional analysis we also have to take into account
+program state such that $(S, \state) \Downarrow_n \state'$.
+
+To obtain a compositional reasnoning we also have to take into account
 the state resulting from a program's execution.  We thus use two
@@
 $\Phi(\state)$ must provide enough \emph{potential} for both, paying
-for resource cost of the computation and paying for the potential
+for the resource cost of the computation and paying for the potential
 $\Phi'(\state')$ on the resulting state $\state'$. That way, if
@@
 \state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
-$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed to
+$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
 $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
@@
 \forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
-n + \Phi'(\state')
+n + \Phi'(\state') \; ,
 $$
@@
 inputs.  To this end, we allow the potential function $\Phi$ in the
-pre and post conditions to take non-negative number or $\infty$
+pre- and postconditions to take non-negative number or $\infty$
 (infinity) as values.  Infinity, plays the same role as
@@
   $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \lstinline{tick(n)} and
+  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
   0 to all other operations.
@@
 For example, it is possible to derive the following quantitative
-triple for a binary-search function \code{bsearch} that holds for all
-stack metrics $M$.  
+triple for a binary-search function \code{bsearch}.% that holds for all
+%stack metrics $M$.  
 %
@@
 %
+It holds for all stack metrics $M$.
 A stack metric assigns cost $0$ to all evaluation steps except
@@
 For simplicity, assume for the remainder of this section that we are
-interested in bounding the number of \emph{ticks} in a program.  Thus,
-we use a resource metric that assigns cost $n$ to the function call
+interested in bounding the number of \emph{ticks} in a program.  That
+is, we use a resource metric that assigns cost $n$ to the function call
 \code{tick(n)} and cost $0$ to all other operations.  The argument $n$
@@
 amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
-shows a program that repeatably increments a binary counter that is
+shows a program that repeatedly increments a binary counter
 implemented with a boolean-valued ($0$ or $1$) array.  The derived
-bound in the precondition states that the number of ticks that are
-executed by the program are bounded by $\#_1(a) + 2k$.  Here, $k$ is
+bound in the precondition states that the number of ticks
+executed by the program is bounded by $\#_1(a) + 2k$.  Here, $k$ is
 the number of increments to the counter that are performed by the
@@
 See \pref{sec:logic} for more a formal definition of the logic and
-additional example derivation.
+additional example derivations.
 
@@
 (LP) to compute a derivation in the logic.  If we assume for now that
-a program state $\state$ simply maps variable to integers then we
+a program state $\state$ simply maps variables to integers then we
 require
@@
 We then develop an inference rule for each syntactic construct that
-derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
-logic but enables inference by using an LP solver.
+derives a sound triple $\htriple{\Phi}P{\Phi'}$ in the quantitative
+logic but enables inference using an LP solver.
 
@@
 \emph{ticks} metric that assigns cost $n$ to the function call
-\code{tick(n)} \iffull{and cost $0$ to all other operations}{} then
+\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
 the cost of the following example can be bounded by $|[x,y]|$. Note
@@
 \begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
+while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
 \end{lstlisting}
@@
 to derive a triple like
-$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
+$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
 only do so if we utilize the fact that $x<y$ at the beginning of the
@@
 \begin{lstlisting}[basicstyle=\tt\small]
-  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
+  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
 \end{lstlisting}
@@
 \begin{lstlisting}[basicstyle=\tt\small]
-  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
+  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
 \end{lstlisting}
@@
 family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
-indexed by the variables that are currently in scope.  The logical
+indexed by the variables currently in scope.  The logical
 context $\Gamma$ is simply a conjunction of inequalities between
-linear combinations variables.  It is used to determine at the
-assignment if we can extract constant potential to pay for future
+linear combinations of variables.  It is used, for example, to determine
+at the assignment if we can extract constant potential to pay for future
 cost.  It is important to note that the reasoning about assertions in
@@
 version of Example~1 in which increment $x$ be a constant $K>0$ and
-consume $T>0$ resources in each iteration.  The reasoning is similar
-to the description of Example~1 except that we obtain the potential
+that consumes $T>0$ resources in each iteration.  The reasoning is similar
+to the one of Example~1 except that we obtain the potential
 $K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
@@
 To automate the reasoning, we first introduce an unknown rational
-variables for each factor in the potential functions.  We then use our
+variable for each factor in the potential functions.  We then use our
 inference rules (see \pref{sec:AAA}) to emit linear inequalities on
@@
 the logical context $\Gamma$.  It would of course be possible to keep
-track more sophisticated assertions.  However, this simple form is
+of track more sophisticated assertions.  However, this simple form is
 sufficient for the examples we considered and we can efficiently
@@
 program.  Hence, the resource cost is closely related to the
-number of times the test \lstinline{a[i] > m} is true during the
+number of times the test \code{a[i] > m} is true during the
 execution.
@@
 where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
-the number of ``maximum candidates'' in the array \lstinline{a} seen
+the number of ``maximum candidates'' in the array \code{a} seen
 by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
@@
 $(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
-When the condition \lstinline{a[i] > m} is true, we know that we
+When the condition \code{a[i] > m} is true, we know that we
 encountered a ``maximum so far'' because $m$ is a maximal element of
-\lstinline{a[$0 \dots i$]}, thus $A(i) = 1 + A(i+1)$ and we get one
+\code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
 potential unit to pay for the assignment.  In the other case, no
@@
 to our derived bound.  \pref{fig:3d} shows such a comparison for
-example \emph{t08}.  One can see the derived constant factors are the
+example \emph{t08}.  One can see that the derived constant factors are the
 best possible if the input variable $x$ is non-negative.  If $x$ is

### 2014-07-07 paper/main.tex
more typos

@@
 the logical context $\Gamma$.  It would of course be possible to keep
-of track more sophisticated assertions.  However, this simple form is
+track of more sophisticated assertions.  However, this simple form is
 sufficient for the examples we considered and we can efficiently
@@
 \emph{speed\_1} and \emph{speed\_2} that are taken from previous
-work~\cite{GulwaniMC09} and demonstrate the our method can handle
+work~\cite{GulwaniMC09} and demonstrate that our method can handle
 \emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
-derives the same bounds as our analysis but requires inference of
-\emph{disjunctive invariants}.  In the abstract interpretation
+derives the same bounds as our analysis but requires heuristics for
+its counter instrumentation.  It can also be handled with inference of
+\emph{disjunctive invariants}, but in the abstract interpretation
 community, these invariants are known to be notoriously difficult to
@@
 Example \emph{speed\_2} is more tricky and we found it hard to
-manually find a bound.  However, using potential transfer reasoning of
-amortized analysis, it is straightforward to prove the tight bound
+manually find a bound.  However, using potential transfer reasoning as
+in amortized analysis, it is easy to prove the tight bound
 $|[x, n]| + |[z, n]|$.
@@
 Example \emph{t08a} shows the ability of the analysis to discover
-interaction between \emph{sequential loops} through size changes of
+interaction between \emph{sequenced loops} through size change of
 variables.  We accurately track the size change of $y$ in the first
@@
 \emph{interacting nested loops}.  In the outer loop we increment the
-variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations, we
-increment the variable $y$ by $1000$.  We then non-deterministically
+variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
+increment the variable $y$ by $1000$.  Then we non-deterministically
 (expressed by \code{*}) execute an inner loop that decrements $y$ by
@@
 %
-Expressions are $E$ are left abstract in our presentation.  For our
+Expressions $E$ are left abstract in our presentation.  For our
 analysis framework, it is only important that they are side-effect
@@
 general function calls and returns.  However, we have not implemented
-function pointers, goto statements, continue statements, and switch
-statements.
+function pointers, \code{goto} statements, \code{continue} statements,
+and \code{switch} statements.
 
@@
 effects---are evaluated in a big-step fashion.  Here, we describe a
-simplified version of Clight's semantics that is sufficient for the
-subset that we are describing.
+simplified version of Clight's semantics for the subset we are
+describing.
 
@@
 The evaluation function \evalE{\cdot}{} maps an expression $e \in
-E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.
+E$ to a value $\evalE{e}{\state} \in \Z$ in the program state
+$\state$.
 
 For simplicity, we assume that local variables and global variables
-are always disjoint.  We just write $\state(x)$ to obtain the value of
+are always different.  We just write $\state(x)$ to obtain the value of
 $x$ in program state $\state$. \iffull{Such a lookup is defined as

### 2014-07-07 paper/main.tex
improved abstract

@@
   at development time.  Recent years have seen tremendous progress in
-  automatically deriving worst-case resource bounds.  Yet, many
+  automatically deriving worst-case resource bounds, yet many
   challenges in specifying, interactively deriving, formally
@@
   framework that tackles these challenges for C programs.  The
-  analysis framework consists of three parts.  First, the runtime
-  resource consumption of programs is formalized with a parametric
-  cost semantics for CompCert Clight.  Second, a quantitative Hoare
-  logic that can be used for interactive bound development is
-  implement and proved sound with respect to the cost semantics in the
-  Coq Proof Assistant.  The logic is very expressive and shallow
-  embedding enables a derived bound to be any function that is
+  analysis framework consists of three parts: First, a parametric cost
+  semantics formalizes the resource consumption of CompCert Clight
+  programs.  Second, a quantitative Hoare logic enables users to
+  interactively develop resource bounds in the Coq Proof Assistant.
+  The logic is proved sound with respect to the cost semantics, and
+  shallow embedding enables a derived bound to be any function that is
   definable in Coq.  Third, an automatic amortized resource analysis
-  for C programs that computes derivations in the quantitative Hoare
-  logic is developed and implemented.  It is the first automatic
-  amortized analysis that can derive bounds that depend on negative
-  integers or differences between numbers, which is crucial to handle
-  typical systems code.  Both, the quantitative logic and the
-  automatic amortized analysis are naturally compositional and can be
-  combined to semi-automatically derive global resource bounds.
+  for C programs computes derivations in the quantitative Hoare logic.
+  It is the first automatic amortized analysis that can derive bounds
+  that depend on negative integers or differences between numbers,
+  which is crucial to handle typical systems code.  Both, the
+  quantitative logic and the automatic amortized analysis are
+  naturally compositional and can be combined to semi-automatically
+  derive global resource bounds.
   %
@@
   algorithms.  A comparison of the automatic amortized analysis with
-  other automatic tools on 30 challenging examples from the literature
-  and open-source software shows that the bounds derived by the
-  automatic amortized analysis are often more precise.
+  other automatic tools on 30 challenging loop and recursion patterns
+  from the literature and open-source software shows that the bounds
+  derived by the automatic amortized analysis are often more precise.
 \end{abstract}

### 2014-07-07 paper/main.tex
change in abstract
Joshua's remarks for the intro

@@
   definable in Coq.  Third, an automatic amortized resource analysis
-  for C programs computes derivations in the quantitative Hoare logic.
+  for Clight computes derivations in the quantitative Hoare logic.
   It is the first automatic amortized analysis that can derive bounds
@@
 software, there still exist shortcomings that hamper the application
-of existing techniques in practice.
+of existing techniques in practice:
 \itemskip
@@
 resource bound analyses are often required by regulatory
-authorities~\cite{May2013}.  Our \emph{contributions} are as follows.
+authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
 %
@@
   derivations by generating simple linear constraints that can be
-  solved by an off-the-shelf LP solver and does not require any
+  solved by an off-the-shelf LP solver, and does not require any
   fixpoint computations to obtain loop invariants.
 \itemskipIn
-\item We show with a publically available prototype implementation and
-  an experimental evaluation that our novel amortized resource
+\item We show with a publically available prototype implementation, and
+  an experimental evaluation, that our novel amortized resource
   analysis works precisely and efficiently for challenging loop and
@@
 for automation that sometimes goes beyond the capabilities of existing
-techniques.  Both the automatic amortized analysis and the
+techniques.  Furthermore, the automatic amortized analysis and the
 quantitative logic are naturally compositional and describe the
@@
 feature of our system is, that it can handle resources like memory
-that may become available during the execution.
+that may become available during execution.
 
@@
 constraint system that reflects resource cost and size changes of
-integer variables in the program.  A solution of the linear program
+integer variables in the program; a solution of the linear program
 immediately yields a resource usage bound for the C
@@
 source software and the literature.  Our analysis can find
-asymptotically tight bounds for all but 3 of these patterns and in
+asymptotically tight bounds for all but 3 of these patterns, and in
 most cases the derived constant factors are tight.  To compare our

### 2014-07-07 paper/main.tex
Joshua's remakrs for section 2

@@
 starting state $\state$ and consumes $n$ resource units of some
-quantity we are interested in.  We denote that by writing $(S, \state)
+user-defined quantity.  We denote that by writing $(S, \state)
 \Downarrow_n \state'$ where $\state'$ is the program state after the
@@
 numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
-program state such that $(S, \state) \Downarrow_n \state'$.
+program state such that $(S, \state) \Downarrow_n \state'$ to prove 
+a bound on the resource usage.
 
@@
 }{}
-We can use the logic to prove a concrete bound for a given resource
+We can then use the logic to prove a concrete bound for a given resource
 metric, or a bound that is parametric in a set of resource metrics.
-For example, it is possible to derive the following quantitative
-triple for a binary-search function \code{bsearch}.% that holds for all
-%stack metrics $M$.  
+For example, we can derive the following quantitative
+triple for a binary-search function \code{bsearch} that holds for all
+stack metrics $M$.  
 %
@@
 %
-It holds for all stack metrics $M$.
 A stack metric assigns cost $0$ to all evaluation steps except
@@
 analyzing programs.  However, program logics need to be supported by
-automatic methods to be useful in practice.  That is why we developed a
-new automatic amortized analysis for deriving quantitative Hoare
-triples for common C programs.
+automatic methods to be useful in practice.  % That is why we developed a
+% new automatic amortized analysis for deriving quantitative Hoare
+% triples for common C programs.
 
@@
 \code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
-the cost of the following example can be bounded by $|[x,y]|$. Note
+the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
 that there are no restrictions on the signs of the input variables in
-the examples.
+the examples.}
 \begin{lstlisting}[basicstyle=\tt\small]
@@
 linear combinations of variables.  It is used, for example, to determine
-at the assignment if we can extract constant potential to pay for future
+at variable assignments if we can extract constant potential to pay for future
 cost.  It is important to note that the reasoning about assertions in
@@
 $\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
-version of Example~1 in which increment $x$ be a constant $K>0$ and
-that consumes $T>0$ resources in each iteration.  The reasoning is similar
+version of Example~1 in which we increment $x$ by a constant $K>0$ and
+consume $T>0$ resources in each iteration.  The reasoning is similar
 to the one of Example~1 except that we obtain the potential
@@
 variable for each factor in the potential functions.  We then use our
-inference rules (see \pref{sec:AAA}) to emit linear inequalities on
-these variables that enforce that the potential annotations are sound.
-For instance if $K = 1$ in \pref{fig:ex1} then we would have the
+inference rules (see \pref{sec:AAA}) to emit linear constraints on
+these variables that enforce that variable assignments that respect
+the constraints correspond to sound potential annotations.  For
+instance if $K = 1$ in \pref{fig:ex1} then we would have the
 annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
 |[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
-p_{(y,x)} {\cdot} |[y,x]|$ after the assignment where $q_i$ and $p_i$
+p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
 are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
@@
 \iffull{It might be surprising that we}{We} only track simple linear relations in
-the logical context $\Gamma$.  It would of course be possible to keep
-track of more sophisticated assertions.  However, this simple form is
+the logical context $\Gamma$.  While it would of course be possible to keep
+track of more sophisticated assertions, this simple form is
 sufficient for the examples we considered and we can efficiently
@@
 
-As mentioned, the automatic analysis can handle challenging example
+As mentioned earlier, the automatic analysis can handle challenging example
 programs without special tricks or techniques.  Examples
-\emph{speed\_1} and \emph{speed\_2} that are taken from previous
-work~\cite{GulwaniMC09} and demonstrate that our method can handle
+\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
+work~\cite{GulwaniMC09},  demonstrate that our method can handle
 \emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
@@
 %
-Example \emph{speed\_2} is more tricky and we found it hard to
-manually find a bound.  However, using potential transfer reasoning as
+Example \emph{speed\_2} is even trickier, and we found it hard to
+find a bound manually.  However, using potential transfer reasoning as
 in amortized analysis, it is easy to prove the tight bound
@@
 variables.  We accurately track the size change of $y$ in the first
-loop by transferring the potential $0.1$ form $|[y,z]|$ to $|[0,y]|$.
+loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
 Furthermore, \emph{t08a} shows again that we do not handle the
@@
 decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{count\_up}.  The
-function \code{count\_up} is dual to \code{count\_down}.  Here, we
+$y$. It then recursively calls the function \code{count\_up},
+which is dual to \code{count\_down}.  Here, we
 count up $y$ by $2$ and recursively call \code{count\_down}.  Our
@@
 So if we want to solve this goal by induction, it is
-necessary that $\delta \ge 1$.  These two contraints force
+necessary that $\delta \ge 1$.  These two constraints force
 $\delta$ to be exactly one in the theorem statement.
@@
 % These annotations are used later on by the quantitative
-% analysis to generate contraints.
+% analysis to generate constraints.
 % %
@@
 
-% \subsection{Linear Contraints Generation}
+% \subsection{Linear Constraints Generation}
 

### 2014-07-07 paper/main.tex
Joshua's remarks for section 3

@@
 
-In this article, we describe our system for a subset of Clight to
-discuss general ideas instead of technical details.  This subset is
-given by the following grammar.
+In this article, we describe our system for a subset of Clight that is
+sufficient to discuss the general ideas.  This subset is given by the
+following grammar.
 %
@@
 Moreover, function arguments and return values are assumed to be
-variables.  In the implementation we can deal with heap updates and
-general function calls and returns.  However, we have not implemented
-function pointers, \code{goto} statements, \code{continue} statements,
-and \code{switch} statements.
+variables.  This is only for simplifying the presentation; in the
+implementation we can deal with heap updates and general function
+calls and returns.  However, we have not implemented our framework for function
+pointers, \code{goto} statements, \code{continue} statements, and
+\code{switch} statements.
 
@@
 A program state $\state = (\theta, \gamma)$ composed of two maps from
-variable names to integers. The first map $\theta : \text{Locals} \to
-\Z$ assigns integers to local variables of a function and the second
-map $\gamma : \text{Globals} \to \Z$ gives values to global variables
+variable names to integers. The first map, $\theta : \text{Locals} \to
+\Z$, assigns integers to local variables of a function, and the second
+map, $\gamma : \text{Globals} \to \Z$, gives values to global variables
 of the program.  In this article, we assume that all values are
@@
 $$}{}
-\iffull{For a program state $\state$, we}{We} similarly write $\state[x \mapsto v]$
+\iffull{For a program state $\state$, we}{Similarly} we write $\state[x \mapsto v]$
 for the program state that maps $x$ to $v$ and behaves as $\state$ for
@@
 The small-step semantics is standard, except that it tracks the
-resource consumption of a program.  The semantics is oblivious to the
+resource consumption of a program.  The semantics is parametric in the
 resource of interest for the user of our system.  We achieve this
@@
 \text{Expr} \to \Q$ from expressions to rational numbers, and one map
-$M_t : \Z \to \Q$ from integers to rational numbers.
-$$
+$M_t : \Z \to \Q$ from integers to rational numbers
+%
+\iffull{.$$
 M = (
@@
 )
-$$
+$$}{$M_a$, $M_b$, $M_r$, $\ldots\;$.
+
+}
 Each of these rational numbers indicates the amount of resource

### 2014-07-07 paper/main.tex
shortened section 4

@@
 
-In the rule {\sc L:Skip}, we do not have to account for any resource
+\iffull{In the rule {\sc L:Skip}, we do not have to account for any resource
 consumption.  As a result, the precondition $Q$ can be any (potential)
@@
 \code{return} part of the postcondition are not reachable and can
-therefore be arbitrary.
+therefore be arbitrary.}{}
 
@@
 
-The rule {\sc L:Update} is the standard assignment rule of Hoare
+\iffull{The rule {\sc L:Update} is the standard assignment rule of Hoare
 logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
@@
 cost of the evaluation of $e$ and the update.
-\iffull{
-}{}
+
 The rule {\sc L:Seq} rule is crucial to understand how the
@@
 
+}
+{
+The rules {\sc L:Update} and {\sc L:Seq} correspond to the respective 
+rules of standard Hoare logic.  
+}
 In the rule {\sc L:Loop}, the \code{break} part of the
@@
 only terminate with a \code{skip} or a \code{return}.  The
-precondition $I$ of the loop is the loop invariant.  That is why we
+precondition $I$ of the loop is the loop invariant.  \iffull{That is why we
 require to have potential $I$ available in the precondition of the
-loop body $S$.  In the postcondition of $S$, the potential must be
+loop body $S$.}{}  In the postcondition of $S$, the potential must be
 sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
@@
 In the rule {\sc L:Call}, accounts for the execution cost of both
-$M_f$ for function calls and $M_r$ for \code{return} statements.  This
+$M_f$ for function calls and $M_r$ for \code{return} statements.  \iffull{This
 justifies that the {\sc L:Return} rule does not account for resource
-consumption.  The pre- and postconditions $P_f$ and $Q_f$ are taken
+consumption.}{}  The pre- and postconditions $P_f$ and $Q_f$ are taken
 from the function context $\Delta$.  The assertions in the context are
@@
 A(\theta,\gamma')$.  It is still possible to express relations between
-global and local variables using logical variables (see the
-\iffull{following paragraph}{extended version} for details).
+global and local variables using logical variables\iffull{ (see the
+following paragraph for details).}{.}
 
@@
 except that instead of using the logical implication $\Rightarrow$ we
-use the quantitative $\models$ that point-wise applies $\ge$.  This
+use the quantitative $\models$ that point-wise applies $\ge$.  \iffull{This
 rule indeed weakens the statement since it requires more resource to
 run the statement and yields less than what has been proved to be
-available after its termination.
+available after its termination.}{}
 
@@
 $q_{ax}$ that contain $x$ after the assignment.
-
-\begin{figure*}
-\centering
-\begin{tabular}{r|c|c|c|c|c}
-%
-& t08 (modified t08a) & t19 & t30 & t15 & t13\\
-%
-\hline
-&
-\begin{lstlisting}
-while (y > x) {
-  x++; tick(1); 
-}
-while (x > 2) {
-  x -= 3; tick(1); 
-}
-\end{lstlisting}
-&
-\begin{lstlisting}
-while (i > 100) {
-  i--; tick(1) 
-}
-i += k+50;
-while (i >= 0) {
-  i--; tick(1); }
-\end{lstlisting}
-&
-\begin{lstlisting}
-while (x > 0) {
-  x--;
-  t=x, x=y, y=t;
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}
-assert(y >= 0);
-while (x>y) {
-  x -= y+1;
-  for (z=y; z > 0; z--)
-    tick(1);
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}
-  while (x>0) {
-    x=x-1;
-    if (*) y=y+1;
-    else
-      while (y>0) {
-        y=y-1; tick(1); }
-    tick(1); }
-\end{lstlisting}
-%
-\\
-\hline
-%
-%
-%
-AAA 
-  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
-  & {\small $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ }
-  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
-  & {\small $|\inter 0 x|$}
-  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
-\\
-%
-Rank &
-{\small $2 {+} y {-} x (?)$ } &
-{\small $54+k+i$ } &
-{\small ---} &
-{\small $2+2x-y$ } &
-{\small $0.5{\cdot}y^2 {+} yx +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
-\\
-%
-LOOPUS &
-{\small 
-  \begin{tabular}{c}
-    $\max(0, x{-}2)$\\ $+ 2\max(0, y {-} x)$
-  \end{tabular}
-} &
-{\small 
-  \begin{tabular}{c}
-    $\max(0, i{-}100)$ \\
-    $+ \max(0, k {+} i {+} 51)$
-  \end{tabular}
- } &
-{\small ---}% $1 + \max(x, y)$ (???) 
-&
-{\small ---}
-&
-{\small $2\max(x, 0) {+} \max(y, 0)$}
-\end{tabular}
-\vspace{.1cm}
-\caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  AAA stands for our automatic
-  amortized analysis for Clight.  The output of Rank has been manually
-  simplified to fit the table.  }
-\label{fig:compar}
-\end{figure*}
 
@@
 error is reported to the user.
+
+\begin{figure*}
+\centering
+\begin{tabular}{r|c|c|c|c|c}
+%
+& t08 (modified t08a) & t19 & t30 & t15 & t13\\
+%
+\hline
+&
+\begin{lstlisting}
+while (y > x) {
+  x++; tick(1); 
+}
+while (x > 2) {
+  x -= 3; tick(1); 
+}
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (i > 100) {
+  i--; tick(1) 
+}
+i += k+50;
+while (i >= 0) {
+  i--; tick(1); }
+\end{lstlisting}
+&
+\begin{lstlisting}
+while (x > 0) {
+  x--;
+  t=x, x=y, y=t;
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}
+assert(y >= 0);
+while (x>y) {
+  x -= y+1;
+  for (z=y; z > 0; z--)
+    tick(1);
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}
+  while (x>0) {
+    x=x-1;
+    if (*) y=y+1;
+    else
+      while (y>0) {
+        y=y-1; tick(1); }
+    tick(1); }
+\end{lstlisting}
+%
+\\
+\hline
+%
+%
+%
+AAA 
+  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
+  & {\small $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ }
+  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
+  & {\small $|\inter 0 x|$}
+  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
+\\
+%
+Rank &
+{\small $2 {+} y {-} x (?)$ } &
+{\small $54+k+i$ } &
+{\small ---} &
+{\small $2+2x-y$ } &
+{\small $0.5{\cdot}y^2 {+} yx +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
+\\
+%
+LOOPUS &
+{\small 
+  \begin{tabular}{c}
+    $\max(0, x{-}2)$\\ $+ 2\max(0, y {-} x)$
+  \end{tabular}
+} &
+{\small 
+  \begin{tabular}{c}
+    $\max(0, i{-}100)$ \\
+    $+ \max(0, k {+} i {+} 51)$
+  \end{tabular}
+ } &
+{\small ---}% $1 + \max(x, y)$ (???) 
+&
+{\small ---}
+&
+{\small $2\max(x, 0) {+} \max(y, 0)$}
+\end{tabular}
+\vspace{.1cm}
+\caption{ Comparison of resource bounds derived by different tools on
+  several examples with linear bounds.  AAA stands for our automatic
+  amortized analysis for Clight.  The output of Rank has been manually
+  simplified to fit the table.  }
+\label{fig:compar}
+\end{figure*}
 

### 2014-07-07 paper/main.tex
some restructuring to make figures appear at the right places

@@
 \end{figure}
+}{}
+
+\ifshort{
+\begin{figure*}[t]
+\small
+\begin{mathpar}
+%
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q }{ \code{skip} }{ Q }
+}
+%
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+%
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Call}
+{
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
+  { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
+\and
+\Rule{L:Extend}
+{ \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+\\
+  \forall f\,P_f\,Q_f.\,
+  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
+  \rightarrow
+  \forall y\,\vec v.\,
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\end{mathpar}
+
+\caption{Rules of the Quantitative Hoare Logic}
+\label{fig:logic}
+\end{figure*}
 }{}
@@
 \begin{figure*}[t]
-\begin{mathpar}
-%
-\Rule{L:Skip}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
-\htriple
-  { M_f + P }
-  { r \gets f(\vec x) }
-  { Q - M_r }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\and
-\Rule{L:Extend}
-{ \Delta \cup \Delta'; B; R \vdash
-  \htriple{ P }{ S }{ Q }
-\\
-  \forall f\,P_f\,Q_f.\,
-  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
-  \rightarrow
-  \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\end{mathpar}
-
-\caption{Rules of the Quantitative Hoare Logic}
-\label{fig:logic}
-\end{figure*}
-}{}
-
-\ifshort{
-\begin{figure*}[t]
-\small
 \begin{mathpar}
@@
 program.
+\begin{figure}[t]
+  \centering
+    \begin{minipage}[b]{.8\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+int srch(
+  int t[], int n,  /* haystack */
+  int p[], int m,  /* needle */
+  int b[]
+) { int i=0, j=0, k=-1;
+    while (i < n) {
+      while (j >= 0 && t[i]!=p[j]) {
+        k = b[j];
+        assert(k > 0 && k <= j + 1);
+        j -= k; tick(1)
+      }
+      i++, j++;
+      if (j == m) break;
+      tick(1);
+    }
+    return i;
+  }
+   \end{lstlisting}
+\vspace{-1.5ex}
+$1 + 2|\inter 0 n|$
+    \end{center}
+\vspace{.1ex}
+  \end{minipage}
+  \caption{The Knuth-Morris-Pratt algorithm for string search.}
+  \label{fig:KMP}
+\end{figure}
 

### 2014-07-07 paper/main.tex
added knuth-morris-pratt

@@
 
+\begin{figure}[t]
+  \centering
+    \begin{minipage}[b]{.9\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+int srch(
+  int t[], int n,  /* haystack */
+  int p[], int m,  /* needle */
+  int b[]
+) { int i=0, j=0, k=-1;
+    while (i < n) {
+      while (j >= 0 && t[i]!=p[j]) {
+        k = b[j];
+        assert(k > 0 && k <= j + 1);
+        j -= k; tick(1)
+      }
+      i++, j++;
+      if (j == m) break;
+      tick(1);
+    }
+    return i;
+  }
+   \end{lstlisting}
+\vspace{-1.2cm}
+\hspace{3.8cm}
+\footnotesize
+\begin{tabular}{|rl|}
+\hline
+AAA & $1 {+} 2|\inter 0 n|$ \\
+Rank & $O(n^2)$ \\
+LOOPUS & $2 {+} 3\max(n, 0)$\\
+\hline
+\end{tabular}
+    \end{center}
+\vspace{-0.1cm}
+  \end{minipage}
+  \caption{The Knuth-Morris-Pratt algorithm for string search.}
+  \label{fig:KMP}
+\end{figure}
+
+Finally, \pref{fig:KMP} contains the search function of the
+Knuth-Morris-Pratt algorithm for string search.  Our automatic
+amortized analysis finds the tight linear bound $1+2|[0,n]|$.  We need
+to assert that the elements $b[j]$ of the failure table $b$ are in the
+interval $[1,j+1]$.  This is guaranteed by
+construction of the table in the initialization procedure of the
+algorithm, which we can also analyze automatically.  We need the
+assertion since we do not infer any logical assertion on the contents
+of the heap.  Rank derives a complex quadratic bound and LOOPUS
+derives a linear bound.
+
 To compare our tool with existing work, we focused on loop bounds and
@@
 program.
-\begin{figure}[t]
-  \centering
-    \begin{minipage}[b]{.8\linewidth}
-    \begin{center}
-   \begin{lstlisting}
-int srch(
-  int t[], int n,  /* haystack */
-  int p[], int m,  /* needle */
-  int b[]
-) { int i=0, j=0, k=-1;
-    while (i < n) {
-      while (j >= 0 && t[i]!=p[j]) {
-        k = b[j];
-        assert(k > 0 && k <= j + 1);
-        j -= k; tick(1)
-      }
-      i++, j++;
-      if (j == m) break;
-      tick(1);
-    }
-    return i;
-  }
-   \end{lstlisting}
-\vspace{-1.5ex}
-$1 + 2|\inter 0 n|$
-    \end{center}
-\vspace{.1ex}
-  \end{minipage}
-  \caption{The Knuth-Morris-Pratt algorithm for string search.}
-  \label{fig:KMP}
-\end{figure}
 

### 2014-07-07 paper/main.tex
typo

@@
 independence by parameterizing evaluations with a resource
-\emph{metric} $M$; a tuple of nine rational numbers, one map $M_e :
+\emph{metric} $M$; a tuple of nine rational numbers \ifshort{$M_a$, $M_b$, $M_r$, $\ldots$}{}, one map $M_e :
 \text{Expr} \to \Q$ from expressions to rational numbers, and one map
-$M_t : \Z \to \Q$ from integers to rational numbers
-%
-\iffull{.$$
+$M_t : \Z \to \Q$ from integers to rational numbers.
+%
+\iffull{$$
 M = (
@@
 )
-$$}{$M_a$, $M_b$, $M_r$, $\ldots\;$.
+$$}{
 

### 2014-07-07 paper/main.tex
some updates to the extended version

@@
 }
+
+\iffull{
+\begin{figure*}[t]
+\begin{mathpar}
+%
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q }{ \code{skip} }{ Q }
+}
+%
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+%
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Call}
+{
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
+  { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
+\and
+\Rule{L:Extend}
+{ \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+\\
+  \forall f\,P_f\,Q_f.\,
+  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
+  \rightarrow
+  \forall y\,\vec v.\,
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\end{mathpar}
+
+\caption{Rules of the Quantitative Hoare Logic}
+\label{fig:logic}
+\end{figure*}
+}{}
 
@@
 
-\iffull{
-\begin{figure*}[t]
-\begin{mathpar}
-%
-\Rule{L:Skip}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
-\htriple
-  { M_f + P }
-  { r \gets f(\vec x) }
-  { Q - M_r }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\and
-\Rule{L:Extend}
-{ \Delta \cup \Delta'; B; R \vdash
-  \htriple{ P }{ S }{ Q }
-\\
-  \forall f\,P_f\,Q_f.\,
-  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
-  \rightarrow
-  \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\end{mathpar}
-
-\caption{Rules of the Quantitative Hoare Logic}
-\label{fig:logic}
-\end{figure*}
-}{}
-
 \paragraph{Rules of the Quantitative Logic.}
@@
 {  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
 }
@@
 &
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
 while (y > x) {
@@
 &
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
 while (i > 100) {
@@
 &
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
 while (x > 0) {
@@
 &
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
 assert(y >= 0);
@@
 &
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
   while (x>0) {
@@
 
-\paragraph{Recursive Functions}
-
-Our approach can naturally deal with mutually-recursive functions.
-The recursion patterns can be exactly the same that are used in
-iterations of loops.  In the following, we present three simple
-examples that illustrate the analysis of functions.
-
-Example \emph{t37} illustrates that the analyzer is able to perform
-inter-procedural size tracking.  The function \code{copy} adds the
-argument $x$ to the argument $y$ if $x$ is positive.  However, this
-addition is done in steps of $1$ in each recursive call.  The function
-\code{count\_down} recursively decrements its argument down to $0$.
-The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
-\code{\main} in which we first add $x$ to $y$ using the function
-\code{copy} and then count down the variable $y$ using the function
-\code{count\_down}.  The derived bound is tight.
-
-Example \emph{t39} uses mutual recursion.  The function
-\code{count\_down} is similar to the function with the same name in
-example \emph{t37}.  However, we do not count down to $0$ but to a
-variable $y$ that is passed as an argument and we call the function
-\code{count\_up} afterwards.  The function \code{count\_up} is dual to
-\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
-\code{count\_down}.  For the function $\main$, which calls
-\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
-1.33 |[z,y]|$.
-
-Example \emph{t46} shows a program that uses and returns resources.
-Again, we use the \emph{tick metric} and the function \code{\tick} to
-describe the resource usage.  The function \code{produce} produces
-$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
-receives one resource unit.  Similarly, the function \code{consume}
-consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
-$|[0,y]|$ for the function \code{\main}.  This is only possible since
-amortized analysis naturally tracks the size changes to the variables
-$x$ and $y$, and the interaction between \code{consume} and
-\code{produce}.
-
 \begin{figure*}[t!]
@@
 
-\paragraph{Well-Known Algorithms}
-
-\pref{fig:cat3} shows well-known algorithms that can be automatically
-analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
-the search function of the Knuth-Morris-Pratt algorithm for string
-search.  To derive a bound for this algorithm we have add an assertion
-that indicates the bounds of the values that are stored in the array
-\code{b}.  Using this information we derive a tight linear bound.
-
-Example \emph{Greatest Common Divisor} is the usual implementation of
-the GCD algorithm.  We automatically derive a linear bound.  Note that
-the two tests at the beginning that check if the inputs are positive
-are essential.  Finally, example \emph{Quick Sort} shows a skeleton of
-the quick sort sorting algorithm.  Since we can only derive linear
-bounds we left out the inner loop that swaps the array elements and
-determines the position \code{n+lo} of the pivot.  We only assert that
-\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
+\paragraph{Recursive Functions}
+
+Our approach can naturally deal with mutually-recursive functions.
+The recursion patterns can be exactly the same that are used in
+iterations of loops.  In the following, we present three simple
+examples that illustrate the analysis of functions.
+
+Example \emph{t37} illustrates that the analyzer is able to perform
+inter-procedural size tracking.  The function \code{copy} adds the
+argument $x$ to the argument $y$ if $x$ is positive.  However, this
+addition is done in steps of $1$ in each recursive call.  The function
+\code{count\_down} recursively decrements its argument down to $0$.
+The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
+\code{main} in which we first add $x$ to $y$ using the function
+\code{copy} and then count down the variable $y$ using the function
+\code{count\_down}.  The derived bound is tight.
+
+Example \emph{t39} uses mutual recursion.  The function
+\code{count\_down} is similar to the function with the same name in
+example \emph{t37}.  However, we do not count down to $0$ but to a
+variable $y$ that is passed as an argument and we call the function
+\code{count\_up} afterwards.  The function \code{count\_up} is dual to
+\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
+\code{count\_down}.  For the function \code{main}, which calls
+\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
+1.33 |[z,y]|$.
+
+Example \emph{t46} shows a program that uses and returns resources.
+Again, we use the \emph{tick metric} and the function \code{tick} to
+describe the resource usage.  The function \code{produce} produces
+$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
+receives one resource unit.  Similarly, the function \code{consume}
+consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
+$|[0,y]|$ for the function \code{main}.  This is only possible since
+amortized analysis naturally tracks the size changes to the variables
+$x$ and $y$, and the interaction between \code{consume} and
+\code{produce}.
 
@@
 \end{table*}
+
+\paragraph{Well-Known Algorithms}
+
+\pref{fig:cat3} shows well-known algorithms that can be automatically
+analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
+the search function of the Knuth-Morris-Pratt algorithm for string
+search.  To derive a bound for this algorithm we have add an assertion
+that indicates the bounds of the values that are stored in the array
+\code{b}.  Using this information we derive a tight linear bound.
+
+Example \emph{Greatest Common Divisor} is the usual implementation of
+the GCD algorithm.  We automatically derive a linear bound.  Note that
+the two tests at the beginning that check if the inputs are positive
+are essential.  Finally, example \emph{Quick Sort} shows a skeleton of
+the quick sort sorting algorithm.  Since we can only derive linear
+bounds we left out the inner loop that swaps the array elements and
+determines the position \code{n+lo} of the pivot.  We only assert that
+\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
+
 }\fi

### 2014-07-08 paper/main.tex
unifying bounds in the figures and the big table of the appendix

@@
 % Also look for "SHORT" comments
-% \def\fullversion{}
+\def\fullversion{}
 
@@
 %SHORT
-\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
-\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
-\captionsetup[figure]{format=myformat}
-\captionsetup[table]{format=myformat}
-\renewcommand{\captionlabelfont}{\bf}
+% \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
+% \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
+% \captionsetup[figure]{format=myformat}
+% \captionsetup[table]{format=myformat}
+% \renewcommand{\captionlabelfont}{\bf}
 %SHORT
@@
 %SHORT
-\lstset{basicstyle=\tt\scriptsize}
-\abovedisplayskip=3pt
-\belowdisplayskip=3pt
-\abovedisplayshortskip=3pt
-\belowdisplayshortskip=3pt
+% \lstset{basicstyle=\tt\scriptsize}
+% \abovedisplayskip=3pt
+% \belowdisplayskip=3pt
+% \abovedisplayshortskip=3pt
+% \belowdisplayshortskip=3pt
 %SHORT
@@
 
-\titlebanner{Draft -- Do not distribute}        % These are ignored unless
-\preprintfooter{Draft}   % 'preprint' option specified.
+\titlebanner{\iffull{Extended Version}{}}%{Draft -- Do not distribute}        % These are ignored unless
+\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{}}%{Draft}   % 'preprint' option specified.
 
@@
 source software and the literature.  Our analysis can find
-asymptotically tight bounds for all but 3 of these patterns, and in
+asymptotically tight bounds for all but 1 of these patterns, and in
 most cases the derived constant factors are tight.  To compare our
@@
 programs that can be automatically analyzed by our system.  For
-simplicity, we use a cost metric that counts cost $1$ for assignments
-and cost $0$ for all other operations.  Sometimes we also use the
-\emph{ticks metric} if we want to discuss features like returning
-resources.  Of course, the examples can also be analyzed with any other
-cost metric.
-
-We assume that free variables in the code snippets are the inputs to
+simplicity, and to compare our analysis with existing tools, we use a
+cost metric that counts the number of loop iterations and function
+calls.  This is the only cost metric supported by tools like
+KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
+we also use the \emph{tick metric} if we want to discuss features such as
+resource restitution.  The tick metric assigns cost $n$ to the
+statement \code{tick(n)} and cost $0$ to all other statements. Of
+course, the examples can also be analyzed with any other cost metric
+in our system.
+
+We assume that free variables in the code snippets are the inputs of
 the program. Some of the examples contain constants on which the
@@
 Note however that it is sometimes crucial that constants are positive
-(or negative) and that the same constant at different places.
+(or negative) or that other relations hold.
+
+\pref{tab:eval} contains the details of our comparative evaluation.
 
@@
 
-$3|[0,x]| + |[0,y]|$
+$2|[0,x]| + |[0,y]|$
 \\[.7\baselineskip]
@@
 
-$12|[n,0]| + 0.01|[0,y]|$
+$11|[n,0]| + 0.01|[0,y]|$
 \\[.7\baselineskip]
@@
   \end{minipage}
-   \caption{Amortization and Compositionality (a)}
+   \caption{Amortization and Compositionality (a).}
   \label{fig:cat1a}
@@
 
-$4|[0,x]| + |[0,y]|$
+$1 + 3|[0,x]| + |[0,y]|$
 \\[.7\baselineskip]
@@
 
-$1004|[y,x]|+|[x,0]|+|[0,y]|$
+$1002|[y,x]|+|[x,0]|+|[0,y]|$
 \\[.7\baselineskip]
@@
 
-$3|[0,x]|$
+$|[0,x]|$
 \\[.7\baselineskip]
@@
 
-$1+2|[0,x]|+103|[y,x]|$
+$101|[y,x]|$
 \\[.7\baselineskip]
@@
 
-   \caption{Amortization and Compositionality (b)}
+   \caption{Amortization and Compositionality (b).}
   \label{fig:cat1b}
@@
 
-$52 + |[-1,i]| + |[0,k]|$
+$50 + |[-1,i]| + |[0,k]|$
 \\[.7\baselineskip]
@@
 
-$4|[0,x]|+4|[0,y]|$
+$|[0,x]|+|[0,y]|$
 \\[.7\baselineskip]
@@
 
-$2 + 2|[0, n]|$
+$1 + |[0, n]|$
 \\[.7\baselineskip]
@@
 
-   \caption{Amortization and Compositionality (c)}
+   \caption{Amortization and Compositionality (c).}
   \label{fig:cat1c}
@@
 Figures~\ref{fig:cat1a}, \ref{fig:cat1b}, and~\ref{fig:cat1c} show
-code snippets that need amortization are compositionality to obtain a
+code snippets that need amortization and compositionality to obtain a
 whole program bound.
@@
 that this does not change the size of the interval and computes the
-tight loop bound $3|[y,x]|$.  The additional two loops are in the
+tight loop bound $|[y,x]|$.  The additional two loops are in the
 program to show that the size tracking in the first loop works
@@
 assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
-can share the potential $3|[0,x]|$ before the assignment \code{x=x-y}
-between $3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we
+can share the potential $|[0,x]|$ before the assignment \code{x=x-y}
+between $|[0,x]|$ and $|[0,y]|$ after the assignment.  In this way, we
 derive a tight linear bound.
@@
 the outer loop as in \emph{t15}.  In the inner loop, we also count the
-variable $z$ down to zero and perform $|[0,z]|$ interations.  However,
+variable $z$ down to zero and perform $|[0,z]|$ iterations.  However,
 instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
-computes the tight linear bound $1+2|[0,x]|+103|[y,x]|$.  The
-assignment of potential to the size interval $|[y,x]|$ in instead of
-$|[0,x]|$ is a random choice of the LP solver.  Other optimal
-solutions of the linear program that the analyzer gernerates for
-\emph{t15} would result in bounds such as $1+105|[0,x]|$ or
-$1+6|[0,x]|+99|[y,x]|$.  In fact, the $105$ potential units can be
-shared arbitrarily between the interval sizes $|[y,x]|$ and $|[0,x]|$
-as long as more than $1$ unit is assigned to $|[0,x]|$.  Since we know
-that, $y\geq 0$ the choice of the LP solver is optimal but this seems
-to be a coincidence.
-
-Example \emph{t19} demonstrates another nice property that demonstates
-the compositionaliy of the analysis.  The program consists of two
-loops that decrement a variable $i$.  In the first loop, $i$ is
-decremented down to 100 and in the second loop $i$ is increment
-further down to $-1$.  However, between the loops we assign
-$i=i+k+50$.  So in total the program performs $52 + |[-1,i]| +
-|[0,k]|$ increments.  Our analysis finds this tight bound because our
-amortized analysis naturally takes into account the relation between
-the two loops.  Other techniques might derive a more conservative
-bound such as $52 + |[-1,i]| + |[0,k]| + |[100,i]|$.
+computes the linear bound $101|[0,x]|$.  The
+assignment of potential to the size interval $|[0,x]|$ in instead of
+$|[y,x]|$ is a random choice of the LP solver.
+
+Example \emph{t19} consists of two loops that decrement a variable
+$i$.  In the first loop, $i$ is decremented down to 100 and in the
+second loop $i$ is increment further down to $-1$.  However, between
+the loops we assign $i=i+k+50$.  So in total the program performs $50
++ |[-1,i]| + |[0,k]|$ iterations.  Our analysis finds this tight bound
+because our amortized analysis naturally takes into account the
+relation between the two loops.  Techniques that do not use
+amortization derive a more conservative bound such as $50 + |[-1,i]| +
+|[0,k]| + |[100,i]|$.
 
@@
 switched using the local variable $t$ as a buffer.  Our analysis
-infers the tight bound $4|[0,x]|+4|[0,y]|$.
+infers the tight bound $|[0,x]|+|[0,y]|$.
 
@@
 leaves n unchanged and assigns \code{flag=0}.  The analyzer computes
-the tight bound $2 + 2|[0, n]|$.  The potential in the loop invariant
-is $|[0,\mathit{flag}]| + 2|[0, n]|$.  In the \emph{then} branch of
-the conditional, we use the potential $2|[0, n]|$ and the fact that
+the tight bound $1 + |[0, n]|$.  The potential in the loop invariant
+is $|[0,\mathit{flag}]| + |[0, n]|$.  In the \emph{then} branch of
+the conditional, we use the potential $|[0, n]|$ and the fact that
 $n>0$.  In the \emph{else} branch, we use the potential
@@
 
-$1 + |[0, n]|$
+$|[0, n]|$
 \\[.7\baselineskip]
@@
 
-   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09}} (a)
+   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (a).}
   \label{fig:cat2a}
@@
 
-$1 + |[0,n]|$
+$|[0,n]|$
 \\[.7\baselineskip]
@@
    \end{lstlisting}
-$2 + |[0, m]| + |[0, n]|$
+$|[0, m]| + |[0, n]|$
 \\[.7\baselineskip]
@@
 
-$2 + 2 |[0, M]| + 2 |[0, N]|$
+$|[0, M]| + |[0, N]|$
 \\[.7\baselineskip]
@@
 
-   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b)}
+   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b).}
   \label{fig:cat2b}
@@
    \end{lstlisting}
-$2 + 4|[0, n]|$
+$1 + 2|[0, n]|$
 \\[.7\baselineskip]
@@
 
-   \caption{Examples from [GulwaniPLDI09]}
+   \caption{Examples from~\cite{GulwaniJK09}.}
   \label{fig:cat2c}
@@
       n--; m++;
-    }
+    };
+    tick(1);
   }
    \end{lstlisting}
-$---$
+$|[0,n]|$ ticks
 \\[.7\baselineskip]
@@
    \end{lstlisting}
-$2 |[0, n]|$
+$|[0, n]|$
 \\[.7\baselineskip]
@@
    \end{lstlisting}
-$2 + 3|[0, n]|$
+$1 + 2|[0, n]|$
 \\[.7\baselineskip]
@@
 
-   \caption{Examples from [GulwaniPLDI10]}
+   \caption{Examples from~\cite{GulwaniZ10}.}
   \label{fig:cat3a}
@@
 have been described as challenges in the literature on bound
-generation.  We found only three programs with a linear bound for
-which our analyzer could not find a tight bound.  The first one
-(\emph{fig4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
-  reasoning} to derive a bound.  The other two, \emph{ex2} and
-\emph{ex5} (see [GulwaniPLDI10]), from [GulwaniPLDI10] were described
-to have a linear bound but seem to be non-terminating.
+generation.  We found only one program with a linear bound for
+which our analyzer could not find a tight bound:  Example
+(\emph{fig4\_5}) from~~\cite{GulwaniJK09} requires \emph{path-sensitive
+  reasoning} to derive a bound.
 
@@
 executed.  If $y\geq M$ and $x\geq N$ then the loop is terminated with
-a break.  Our tool computes the bound $2 + 2 |[0, M]| + 2 |[0, N]|$.
+a break.  Our tool computes the bound $|[0, M]| + |[0, N]|$.
 This bound is tight in the sense that there are inputs (such as $M =
@@
 
-Example \emph{fig4\_2} from [GulwaniPLDI09] is quite involved.
-Amortized reasoning helps to understand how we derive the bound $4 +
-4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
-constant potential to pay for the two first assignments.  We use the
-remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
-establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
-loop invariant.  In the \emph{if branch} of the conditional, we use
-the constant potential $4$ of the invariant to pay for the two
-assignments ($2$ units) and the potential of $|[0,vb]|$ ($2$ units).
-Since we also know that $|[0, n]|>0$ we obtain constant potential $4$
-and establish the loop invariant again.  In the \emph{else} branch, we
-use the potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$
-potential units to pay for the assignment.  Since we have $4$ constant
-potential units left after the loop, we can subtract them from the
-initial potential $6 + 4|[0, n]|$.
-
-In example \emph{fig4\_4} it is essential that $m$ is positive.
-That ensures that we can obtain constant potential for the interval
-size $|[0,i]|$ in the \emph{else} branch of the conditional since
-$|[0,i]|$ decreases.  Example \emph{fig4\_5} is one of the three
-examples that we can not handle automatically.  The execution is
-bounded because the boolean value of the test \code{dir==fwd} does not
-change during the iteration of the loop.  As a result, the variable
-$i$ is either counted down to $0$ or up to $n$.  Our tool cannot
-handle example \emph{fig4\_5} because we don't do path sensitive
-reasoning.  Note however that it would be more efficient to move the
-test \code{dir==fwd} outside of the loop (this would be also done by
-an optimizing compiler).  The resulting program could be analyzer by our
-tool.
-
-Example \emph{ex1} from [GulwaniPLDI10] specifically focusses on the
+Example \emph{fig4\_2} from~\cite{GulwaniJK09} is quite involved.
+Amortized reasoning helps to understand how we derive the bound $1 +
+2|[0, n]|$.  We start with potential $1 + 2|[0, n]|$ and use the fact
+that $vb=0$ to establish the potential $1 + 2|[0, n]| + |[0,vb]|$ that
+serves as a loop invariant.  In the \emph{if branch} of the
+conditional, we use the constant potential $1$ of the invariant to pay
+for the potential of $|[0,vb]|$.  Since we also know that $|[0, n]|>0$
+we obtain constant potential $2$ that we use to pay for the loop
+iteration ($1$ unit) and to establish the loop invariant again ($1$
+unit).  In the \emph{else} branch, we use the potential $|[0,vb]|$
+and the fact $vb>0$ to obtain $1$ potential units to pay for the
+loop iteration.
+
+In example \emph{fig4\_4} it is essential that $m$ is positive.  That
+ensures that we can obtain constant potential for the interval size
+$|[0,i]|$ in the \emph{else} branch of the conditional since $|[0,i]|$
+decreases.  Example \emph{fig4\_5} is an examples that we can not
+handle automatically.  The execution is bounded because the boolean
+value of the test \code{dir==fwd} does not change during the iteration
+of the loop.  As a result, the variable $i$ is either counted down to
+$0$ or up to $n$.  Our tool cannot handle example \emph{fig4\_5}
+because we don't do path sensitive reasoning.  Note however that it
+would be more efficient to move the test \code{dir==fwd} outside of
+the loop (this would be also done by an optimizing compiler).  The
+resulting program can then be analyzed by our tool.
+
+Example \emph{ex1} from~\cite{GulwaniZ10} specifically focuses on the
 code in the \emph{if} statement.  So we use the \emph{tick metric} and
@@
 
-We cannot analyze example \emph{ex2} for which previous work reported
-a linear bound [GulwaniPLDI10].  The reason is that this example is
-not terminating since the inner loop is not bounded.  The same is true
-for example \emph{ex5} from the same paper, which is not listed here.
+Similar to example \emph{ex1}, example \emph{ex2}
+form~\cite{GulwaniZ10} focuses on the number of iterations of the
+\emph{outer loop}.  While the whole program is not terminating, the
+number of iterations of the outer loop is bounded by $|[0,n]|$.
 Finally, example \emph{ex2} is similar to example
@@
 
-$1 + 4|[0, x]| + 2|[0, y]|$
+$3 + 2|[0, x]| + |[0, y]|$
 \\[.7\baselineskip]
@@
 
-$1.67 + 1.33 |[z,y]|$
+$1.33 + 0.67 |[z,y]|$
 \\[.7\baselineskip]
@@
 \code{count\_down} recursively decrements its argument down to $0$.
-The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
+The derived bound $3 + 2|[0, x]| +2|[0, y]|$ is for the function
 \code{main} in which we first add $x$ to $y$ using the function
@@
 \code{count\_down}.  For the function \code{main}, which calls
-\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
-1.33 |[z,y]|$.
+\code{count\_down(y,z)}, the analyzer computes the tight bound $1.33 +
+0.67 |[z,y]|$.
 
@@
 
-$1.1 |\inter y z| {+} 0.1 |\inter 0 y|$
+$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
 \\
@@
 
-$0.1|\inter n y| {+} 11 |\inter n 0|$
+$0.01|\inter n y| {+} 11 |\inter n 0|$
 \\
@@
   Experimental evaluation comparing the bounds generated KoAT, Rank,
-  Loopus, Speed, and our automatic amortized analysis on several
-  challenging linear examples. Results for KoAT and Speed were extracted
-  from older publications because KoAT cannot take C programs as 
-  input in its current version and Speed is not publically available.
-  Entries marked with ? signal that we cannot test the tool. Entries
-  marked with --- mean that the tool failed to produce a result.
-}
-\label{fig:eval}
+  LOOPUS, SPEED, and our automatic amortized analysis on several
+  challenging linear examples. Results for KoAT and SPEED were extracted
+  from previous publications because KoAT cannot take C programs as 
+  input in its current version and SPEED is not available.
+  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
+  marked with --- indicate that the tool failed to produce a result.
+}
+\label{tab:eval}
 \end{table*}

### 2014-07-08 paper/main.tex
back to the short version
typos

@@
 %SHORT
-% \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
-% \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
-% \captionsetup[figure]{format=myformat}
-% \captionsetup[table]{format=myformat}
-% \renewcommand{\captionlabelfont}{\bf}
+\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
+\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
+\captionsetup[figure]{format=myformat}
+\captionsetup[table]{format=myformat}
+\renewcommand{\captionlabelfont}{\bf}
 %SHORT
@@
 %SHORT
-% \lstset{basicstyle=\tt\scriptsize}
-% \abovedisplayskip=3pt
-% \belowdisplayskip=3pt
-% \abovedisplayshortskip=3pt
-% \belowdisplayshortskip=3pt
+\lstset{basicstyle=\tt\scriptsize}
+\abovedisplayskip=3pt
+\belowdisplayskip=3pt
+\abovedisplayshortskip=3pt
+\belowdisplayshortskip=3pt
 %SHORT
@@
   %
-  An experimental evaluation demonstrates the practicability of the
+  An experimental evaluation demonstrates the practicality of the
   analysis framework.  The expressivity of the logic is shown by
@@
 \itemskipIn
-\item We show with a publically available prototype implementation, and
+\item We show with a publicly available prototype implementation, and
   an experimental evaluation, that our novel amortized resource
@@
 long-time open problem to apply it to C-like programs with control
-flow that depends integer arithmetic, negative numbers, and non-linear
+flow that depends on integer arithmetic, negative numbers, and non-linear
 control flow.
@@
 resource invariants.  We achieve this by generating a linear
-constraint system that reflects resource cost and size changes of
+constraint system that reflects the resource cost and size changes of
 integer variables in the program; a solution of the linear program
@@
 automatic analyzer with existing techniques, we tested our examples
-with the tools such as KoAT~\cite{BrockschmidtEFFG14},
+with tools such as KoAT~\cite{BrockschmidtEFFG14},
 Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
@@
 %
-Several micro benchmarks demonstrate the practicability and
+Several micro benchmarks demonstrate the practicality and
 expressiveness of the quantitative Hoare logic.  For example, we
@@
 depends on the contents of an array to describe the exact cost of a
-function that finds the maximal element in an array, and a linear
-bound that amortizes the cost of $n$ successive increments to a binary
+function that finds a maximal element in an array, and a linear
+bound that amortizes the cost of $k$ successive increments to a binary
 counter.
 
-% To show the practicability of our approach we developed a case study
+% To show the practicality of our approach we developed a case study
 % with different real-time schedulers implemented in C.  To perform a
@@
 $$
-\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
+\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
 n + \Phi'(\state') \; ,
@@
 
-It is also possible to incorporate boolean preconditions into the
-bounds to express that the bound is only valid for a certain class of
+It is also possible to incorporate boolean conditions into a
+bound to express that the bound is only valid for a certain class of
 inputs.  To this end, we allow the potential function $\Phi$ in the
-pre- and postconditions to take non-negative number or $\infty$
+pre- and postconditions to take non-negative numbers or $\infty$
 (infinity) as values.  Infinity, plays the same role as
@@
 interested in bounding the number of \emph{ticks} in a program.  That
-is, we use a resource metric that assigns cost $n$ to the function call
+is, we use the \emph{tick metric} that assigns cost $n$ to the function call
 \code{tick(n)} and cost $0$ to all other operations.  The argument $n$
@@
     ticks for a standard \emph{for loop}.  The parameters $K>0$ and
-    $T>0$ are not program variables but denote a concrete constants in
+    $T>0$ are not program variables but denote concrete constants in
     the program.  The reasoning works uniformly for all such $K$ and
@@
     m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
-    y = y+1;
+    y=y+1;
     ${\color{blue}%
@@
     n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-    x = x+1;
+    x=x+1;
     ${\color{blue}%
@@
   y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y = y+1;
+  y=y+1;
   ${\color{blue}%
@@
 This idea is best explained by example.  If we again use the
-\emph{ticks} metric that assigns cost $n$ to the function call
+tick metric that assigns cost $n$ to the function call
 \code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
@@
 $K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions in $\Gamma$ are only used in rule for the assignment
+assertions in $\Gamma$ are only used in the rule for the assignment
 \code{x=x+K}.  To the best of our knowledge, no other implemented tool
@@
 iterations without a cost model) and $K=10$,
-KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
+KoAT~\cite{BrockschmidtEFFG14} computes the bound $|x| + |y| + 10$,
 Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
-LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
+LOOPUS~\cite{SinnZV14} computes the bound $y-x-9$.  Only
 PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
@@
 derives the same bounds as our analysis but requires heuristics for
-its counter instrumentation.  It can also be handled with inference of
+its counter instrumentation.  These loops can also be handled with inference of
 \emph{disjunctive invariants}, but in the abstract interpretation
 community, these invariants are known to be notoriously difficult to
-handle.
+generate.
 %
@@
 
-We include the built-in primitive $\code{assert}~E$ that terminates
-the program if the argument $E$ evaluates to false and has no effect
+We include the built-in primitive $\code{assert}~e$ that terminates
+the program if the argument $e$ evaluates to false and has no effect
 otherwise.  This is useful to express assumptions on the inputs of a
 program for the automatic analysis.  We also add the built-in function
-$\code{tick}(n)$ that can be called with a constant integer $n$ as a
+\code{tick(n)} that can be called with a constant integer $n$ as a
 flexible way to model resource consumption or release (if $n$ is
@@
 effects---are evaluated in a big-step fashion.  Here, we describe a
-simplified version of Clight's semantics for the subset we are
-describing.
-
-A program state $\state = (\theta, \gamma)$ composed of two maps from
+simplified version of Clight's semantics for the subset we consider.
+
+A program state $\state = (\theta, \gamma)$ is composed of two maps from
 variable names to integers. The first map, $\theta : \text{Locals} \to
@@
 
-\pref{fig:opsem} contains \ifshort{selected}{} the reduction rules of the semantics.  The
+\pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
 rules define a rewrite system for program configurations of the form
@@
 transition relation $\smallstep{M}^*$ to indicate that the execution
-ran out of reassesses.  However, every execution that reaches such a
+ran out of resources.  However, every execution that reaches such a
 state is stuck.
@@
 {
-The extended version of this article~\cite{anon_extended} contains all rules lemmas 
-that state the main properties of the cost semantics.
+The extended version of this article~\cite{anon_extended} contains all rules and
+lemmas that state the main properties of the cost semantics.
 }
@@
 $$
-P : \State \to \Q \cup \{ \infty \} \; .
+P : \State \to \Qplusz \cup \{ \infty \} \; .
 $$
 This can be understood as a refinement of boolean assertions where
-$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Q$.
-We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
+$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
+We write $\Assn$ for $\State \to \Qplusz \cup \{ \infty \}$ and $\bot$ for
 $\lambda \state . \, \infty$.  We sometimes call assertions
 \emph{potential functions}.  To use Coq's support for propositional
-reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
+reasoning, assertions have the type $\State \to \Qplusz \to \mathrm{Prop}$
 in the implementation.  For a given $\state \in \State$, such an
@@
 Finally, the function context of judgements that we write $\Delta$ is
-a mapping from function names to specification of the form
+a mapping from function names to specifications of the form
 $$
@@
 To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
-the local variables on the call stack, that is, $\forall z \, \vec v
+the local variables on the stack, that is, $\forall z \, \vec v
 \, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
@@
 
-In the rule {\sc L:Call}, accounts for the execution cost of both
+The rule {\sc L:Call} accounts for the execution cost of both
 $M_f$ for function calls and $M_r$ for \code{return} statements.  \iffull{This
 justifies that the {\sc L:Return} rule does not account for resource
-consumption.}{}  The pre- and postconditions $P_f$ and $Q_f$ are taken
+consumption.}{}  The pre- and postcondition $P_f$ and $Q_f$ are taken
 from the function context $\Delta$.  The assertions in the context are
-parametric with respect to both the function argument value and the
+parametric with respect to both the function arguments' value and the
 return value. This allows to specify a bound for a function whose
-resource consumption depends on an input parameter.  The argument
-parameter is instantiated by the call rule using the result of the
+resource consumption depends on input parameters.  The argument
+parameters are instantiated by the call rule using the result of the
 evaluation of the argument variables in the current state.
@@
   set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
-\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
+\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
 $f_i$ in the potential function: $0$ corresponds to the constant
@@
 In the potential functions, we treat constants as global variables
-that can not be assigned to.  For example, if the program contains the
+that cannot be assigned to.  For example, if the program contains the
 constant $8128$ then we have a variable $c_{8128}$ and
@@
 $Q$ and $Q'$ and leaves the latter undefined at some index $i$ we will
-implicitly assume that $q'_i = q_i$.  As mentioned, we leave the
-logical contexts $\Gamma$ abstract.
+implicitly assume that $q'_i = q_i$.
 
 We describe the automatic amortized analysis for a subset of
-expressions of C light.  Assignments must have the form \code{x \gets
-  y} or \code{x \gets x \pm y}.  In the implementation, a Clight
+expressions of Clight.  Assignments must have the form $x \gets
+y$ or $x \gets x \pm y$.  In the implementation, a Clight
 program is converted into this form prior to analysis without changing
@@
   assignments} that do not result in additional cost in the
-analysis.  Non-linear operations such as \code{x \gets z*y} are simply
+analysis.  Non-linear operations such as $x \gets z*y$ are simply
 handled by assigning zero potential to coefficients like $q_{xa}$ and
@@
 substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
-$q_{\ret y} = q'_{x y}$ and $q_{y \ret} = q'_{y x}$ for all $y$ and
+$q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
 $q_i'=q_i$ otherwise.
@@
 potential is distributed after a size change of a variable.  The rule
-{\sc Q:IncP} is for increments $x \gets x + y$ in which we can deduce
-from the logical context $\Gamma$ that $y$ is not negative.
-Similarly, {\sc Q:DecP} is for increments $x \gets x - y$ in which we
-can deduce that $y \ge 0$.  Of course, we have symmetrical rules {\sc
+{\sc Q:IncP} is for increments $x \gets x + y$ and {\sc Q:DecP} is
+for decrements $x \gets x - y$, they both apply only when we can deduce
+from the logical context $\Gamma$ that $y \ge 0$.
+Of course, we have symmetrical rules {\sc
   Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
@@
 To explain how rules for increment and decrement work, it is
-sufficient to understand the rule {\sc Q:IncP}.  The other follow the
+sufficient to understand the rule {\sc Q:IncP}.  The others follow the
 same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
@@
 |\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
-$q_{vx}$ to pay for this change.  The rule insures this
+$q_{vx}$ to pay for this change.  The rule ensures this
 only for $v \not\in \mathcal U$ because we know that $x \le v$
@@
 Another interesting rule is {\sc Q:Call}.  It needs to account for the
-changes to the stack caused by the function call, the argument/return
+changes to the stack caused by the function call, the arguments/return
 value passing, and the preservation of local variables.  We can sum up
@@
 \begin{itemize}
-\item The potential in the pre- and postconditions of the function
+\item The potential in the pre- and postcondition of the function
   specification is equalized to its matching potential in the callee's
@@
 \itemskip
-If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
+If $x$ and $y$ are local variables and $f(x,y)$ is
 called, {\sc Q:Call} splits the potential of
@@
 and $Q'$ are potential annotations and $\Gamma$ is a logical context.
-The rule {\sc Q:Relax} deals also with the important task of transferring
-constant potential (represented by $q_0$) to interval sizes and vise
+The rule {\sc Relax} deals also with the important task of transferring
+constant potential (represented by $q_0$) to interval sizes and vice
 versa.  If we can deduce from the logical context that the interval
@@
 satisfied by these coefficients, it is recorded in a global list using
-the symbolic names.  When then feed the collected constraints to an
-off the shelf LP-solver~\footnote{We currently use Coin-Or's CLP.}.
+the symbolic names.  We then feed the collected constraints to an
+off the shelf LP solver~\footnote{We currently use Coin-Or's CLP.}.
 If the solver successfully finds a solution, we know that a derivation
@@
 solver, and apply it to the symbolic names $(q_i)$ of the coefficients
-in the derivation.  If the LP-solver fails to find a solution, an
+in the derivation.  If the LP solver fails to find a solution, an
 error is reported to the user.
@@
 while (i > 100) {
-  i--; tick(1) 
-}
+  i--; tick(1) }
 i += k+50;
 while (i >= 0) {
-  i--; tick(1); }
+  i--; tick(1);
+}
 \end{lstlisting}
@@
 assert(y >= 0);
-while (x>y) {
+while (x > y) {
   x -= y+1;
@@
 \begin{lstlisting}[basicstyle=\tt\scriptsize]
-  while (x>0) {
-    x=x-1;
-    if (*) y=y+1;
+  while (x > 0) {
+    x--;
+    if (*) y++;
     else
-      while (y>0) {
-        y=y-1; tick(1); }
+      while (y > 0) {
+        y--; tick(1); }
     tick(1); }
@@
 
-We have experimentally evaluated the practicability of our automatic
+We have experimentally evaluated the practicality of our automatic
 amortized analysis with more than 30 challenging loop and recursion
@@
 first loop, $i$ is decremented down to 100 and in the second loop $i$
-is increment further down to $-1$.  However, between the loops we
+is decremented further down to $-1$.  However, between the loops we
 assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
@@
 $|[0,x]|$ before the assignment \code{x=x-y} between $|[0,x]|$ and
-$|0,y|$ after the assignment.  In this way, we derive a tight linear
+$|[0,y]|$ after the assignment.  In this way, we derive a tight linear
 bound.
@@
 use a simple metric that counts the number of back edges (i.e., number
-of loop iterations) that are follows in the execution of the program
+of loop iterations) that are followed in the execution of the program
 because most other tools only bound this specific cost. In
@@
 released (e.g., memory).  However, as we have shown
-in~\pref{sec:exper}, our automatic amortized analysis matches that the
+in~\pref{sec:exper}, our automatic amortized analysis matches the
 state of the art in automatic bound analysis for linear bounds and
@@
 
-We will continue to improve our framework to reasoning more precisely
+We will continue to improve our framework to reason more precisely
 about resource usage of system software such as the hypervisor kernel
 CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
-  and verifying}{is currently developed and verified}.  For on thing,
+  and verifying}{is currently developed and verified}.  For one thing,
 we will improve the automation to derive \emph{polynomial bounds} and

### 2014-07-08 paper/main.tex
minor update

@@
 
-The rule {\sc L:Call} accounts for the execution cost of both
-$M_f$ for function calls and $M_r$ for \code{return} statements.  \iffull{This
-justifies that the {\sc L:Return} rule does not account for resource
-consumption.}{}  The pre- and postcondition $P_f$ and $Q_f$ are taken
-from the function context $\Delta$.  The assertions in the context are
-parametric with respect to both the function arguments' value and the
-return value. This allows to specify a bound for a function whose
-resource consumption depends on input parameters.  The argument
-parameters are instantiated by the call rule using the result of the
-evaluation of the argument variables in the current state.
-\iffull{Recall that we require that $P_f$ and $Q_f$ do not depend on
-  the local variables on the call stack, that is, $\forall z \, \vec v
-  \, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
-  (\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.}{} To
-transfer potential that depends on local variables of the callee from
-the precondition $P$ to the postcondition $Q$, we use an assertion $A
-: \Assn$ that is independent of global variables, that is, $\forall
-\theta \, \gamma \, \gamma' . \, A (\theta,\gamma) =
-A(\theta,\gamma')$.  It is still possible to express relations between
-global and local variables using logical variables\iffull{ (see the
-following paragraph for details).}{.}
+The rule {\sc L:Call} accounts for the execution cost of both $M_f$
+for function calls and $M_r$ for \code{return} statements.
+\iffull{This justifies that the {\sc L:Return} rule does not account
+  for resource consumption.}{} The pre- and postcondition $P_f$ and
+$Q_f$ are taken from the function context $\Delta$.  The assertions in
+the context are parametric with respect to both the values of the
+function arguments and the return value. This allows us to specify a
+bound for a function whose resource consumption depends on its
+arguments.  The arguments are instantiated by the call rule
+using the result of the evaluation of the argument variables in the
+current state.  \iffull{Recall that we require that $P_f$ and $Q_f$ do
+  not depend on the local variables on the call stack, that is,
+  $\forall z \, \vec v \, \theta \, \theta' \, \gamma . \, P_f \, z \,
+  \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
+  (\theta',\gamma)$.}{} To transfer potential that depends on local
+variables of the callee from the precondition $P$ to the postcondition
+$Q$, we use an assertion $A : \Assn$ that is independent of global
+variables, that is, $\forall \theta \, \gamma \, \gamma' . \, A
+(\theta,\gamma) = A(\theta,\gamma')$.  It is still possible to express
+relations between global and local variables using logical
+variables\iffull{ (see the following paragraph for details).}{.}
 
@@
 { $\Phi = \sum_{i \in I} q_i f_i$.}  We often write $xy$ to denote the
-index $(x,y)$.  The family $(f_i)_{i\in I}$ is actually a basis (in
-the linear-algebra sense) for potential functions.  This allows us to
+index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
+the linear-algebra sense) for potential functions.}{}  This allows us to
 uniquely represent any linear potential function

### 2014-07-08 paper/main.tex
minor change
fixed a typo
fixed rule Q:Return
removed preprint option
typo

@@
 % Comment for anonymization
+% Add preprint option for the extended version
 % Also look for "SHORT" comments
@@
 
-\documentclass[nocopyrightspace,preprint]{sigplanconf}
+\documentclass[nocopyrightspace]{sigplanconf}
 
@@
 \begin{itemize}
-\item Analysis tools are black boxes that \iffull{either}{} deliver a result or fail
+\item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
   without enabling \emph{user interaction} to manually or
@@
 \begin{enumerate}
-\item We define a operational cost semantics for CompCert Clight that
+\item We define an operational cost semantics for CompCert Clight that
   defines the resource consumption of terminating and diverging
@@
 { P = Q_R[\Vret/x]
+\\ \Gamma = \Gamma_R[\Vret/x]
 \\ \forall i \in \dom{P} . \, p_i = q_i
 }
-{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
+{ B; (\Gamma_R,Q_R); (\Gamma, Q) \vdash \code{return}~x \dashv (\Gamma',Q') }
 %
@@
 \and }{}
-\Rule{Q:Return}
+\Rule[leftskip=.2cm,rightskip=.1cm]{Q:Return}
 { P = Q_R[\Vret/x]  
@@
 }
-{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
-%
-\and \Rule{Q:Update}
+{ B; (\Gamma_R,Q_R); (\Gamma_R[\Vret/x], Q) \vdash \code{return}~x \dashv (\Gamma',Q') }
+%
+\and \Rule[leftskip=.1cm,rightskip=.2cm]{Q:Update}
 { q'_{xy}, q'_{yx} \in \Qplusz
@@
 language.  However, there is no way to interactively derive bounds or
-to deal with resources like memory usage.  Blazy et al.'s automatic
-bound analysis does also not compute symbolic bounds.
+to deal with resources like memory usage.  Furthermore, Blazy et al.'s
+automatic bound analysis does not compute symbolic bounds.
 

### 2014-07-08 paper/main.tex
typo

@@
 Our approach complements existing work since it provides a semantic
-foundation for the computation of bounds while still providing support
-for automation that sometimes goes beyond the capabilities of existing
+foundation for the computation of bounds while still supporting
+automation that sometimes goes beyond the capabilities of existing
 techniques.  Furthermore, the automatic amortized analysis and the
@@
 integer variables in the program; a solution of the linear program
-immediately yields a resource usage bound for the C
+immediately yields a resource-usage bound for the C
 program.  % We can naturally, handle
@@
 %
-More formally, we develop (\pref{sec:AAA}) a judgement 
+More formally, we develop (in \pref{sec:AAA}) a judgement 
 $$
@@
 evaluation.  Example \emph{t08} is a slightly modified version of
-\emph{t08a} which is described in the introduction.  Example
+\emph{t08a}, which is described in~\pref{sec:inform}.  Example
 \emph{t19} demonstrates the compositionaliy of the analysis.  The

### 2014-07-08 paper/main.tex
typo

@@
 in the implementation.  For a given $\state \in \State$, such an
-assertion can be seen as a set $B \subseteq \Q$ of valid bounds.
+assertion can be seen as a set $B \subseteq \Qplusz$ of valid bounds.
 However, we find the presentation in this article easier to read.
@@
 algorithm, which we can also analyze automatically.  We need the
-assertion since we do not infer any logical assertion on the contents
+assertion since we do not infer any logical assertions on the contents
 of the heap.  Rank derives a complex quadratic bound and LOOPUS
@@
 tool (\#tested).  Since we were not able to run the experiments for KoAT
-and SPEED, we simply used the bounds the have been reported by the
+and SPEED, we simply used the bounds that have been reported by the
 authors of the respective tools.  The results show that our automatic

### 2014-07-08 paper/main.tex
typo

@@
 
-There exist many tools that can automatically derive loops and
+There exist many tools that can automatically derive loop and
 recursion bounds for imperative programs such as

### 2014-07-08 paper/main.tex
typo

@@
   $\!$The potential of intervals $|\inter x y|$ is preserved
-  if $x$ and $y$ are local\iffull{ variables,}{,}
+  if $x$ and $y$ are local\iffull{ variables.}{.}
 \item
@@
 There are techniques~\cite{Braberman08} that can compute the memory
-requirements of object oriented programs with region based garbage
+requirements of object oriented programs with region-based garbage
 collection.  These systems infer invariants and use external tools

### 2014-07-15 paper/main.tex
switched to tr version
short version with names for distribution

@@
 
-\documentclass[nocopyrightspace]{sigplanconf}
+\documentclass[nocopyrightspace,preprint]{sigplanconf}
 
@@
 \titlebanner{\iffull{Extended Version}{}}%{Draft -- Do not distribute}        % These are ignored unless
-\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{}}%{Draft}   % 'preprint' option specified.
+\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.
 
@@
 \authorinfo
-{\iffull{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}{\vspace{-1cm}}}
-{\iffull{Yale University}{\vspace{-1cm}}}
+{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
+{Yale University \vspace{-1cm}}
 

### 2014-09-09 paper/main.tex
example derivation

@@
 
+\section{Example Derivation in the Automatic System}
+\label{app:derivation}
+
+\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
+
+\begin{figure*}[t]
+  \centering
+  \footnotesize
+  \begin{mathpar}
+  \inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
+  { \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
+    { 
+      \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1.5cm]
+      { 
+        \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
+        {
+\inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
+{
+  \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
+  { }
+  {(x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
+   \dashv (\cdot, P^\text{de})
+  }
+}
+{(x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
+\dashv (\cdot, P^\text{we})
+}
+\and
+\inferrule*[right={\scriptsize (Q:Tick)},leftskip=0cm,rightskip=0cm]
+{ }
+{(x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
+ \dashv (\cdot, P^\text{ti})
+}
+} {
+          (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq}) 
+        }
+      }
+      { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if}) }
+      \and
+      \inferrule*[right={\scriptsize (Q:Weak)},leftskip=4.5cm,rightskip=4cm]
+      { \inferrule*[right={\scriptsize (Q:Break)},leftskip=0cm,rightskip=0cm]
+        { }
+        {(x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash \code{break} \dashv (\bot,P^\text{br})}
+      }
+      {
+        (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash \code{break} \dashv (\cdot,P^\text{el})
+      }
+    }
+    { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm} 
+    }
+  }
+  {
+   (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P) 
+  }
+\end{mathpar}
+\flushleft Constraints:
+$$
+\begin{array}{llll}
+P = B^\text{lo} & Q  = Q^\text{lo} & Q^\text{lo} = P^\text{lo} \\
+B^\text{el} = B^\text{if} = B^\text{lo} & Q^\text{el} = Q^\text{if} = Q^\text{lo} & P^\text{el} = P^\text{if} = P^\text{lo} \\
+B^\text{el} = B^\text{br}  & Q^\text{el} \succeq_{(x{<}10)} Q^\text{br}  & P^\text{br} \succeq_{(\cdot)} P^\text{el} \\
+B^\text{br} = Q^\text{br}\\
+B^\text{if} = B^\text{sq}  & Q^\text{if} \succeq_{(x{<}10)} Q^\text{sq}  & P^\text{sq} \succeq_{(\cdot)} P^\text{if} \\
+B^\text{sq} = B^\text{we} = B^\text{ti} & Q^\text{sq} = Q^\text{we} & P^\text{we} = Q^\text{ti} & P^\text{ti} = P^\text{sq} \\
+Q^\text{ti} = P^\text{ti} + 5 \\
+B^\text{we} = B^\text{de}  & Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}  & P^\text{de} \succeq_{(\cdot)} P^\text{we} \\
+p^\text{de}_{0,10} = q^\text{de}_{0,10} + q^\text{de}_{0,x}  & p^\text{de}_{0} = q^\text{de}_{0}
+& \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{de}_{\alpha,\beta} = q^\text{de}_{\alpha,\beta}
+\end{array}
+$$
+\flushleft Linear Objective Function:
+$$
+q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x}
+$$
+\flushleft Constant Objective Function:
+$$
+q_0 + 11{\cdot}q_{0,10}  
+$$
+\caption{An example derivation as produced by the automatic tool.}
+\label{fig:derivation}
+\end{figure*}
+
+\pref{fig:derivation} contains an example derivation as it is produced
+by our automatic analyzer.  The upper case letters (with optional
+superscript) such as $Q^\text{de}$ stand for a family of rational
+variables that are later part of the constraint system that is passed
+to the LP solver.  For example $Q^\text{de}$ stands for the potential
+function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
+q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
+q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
+variables such as $q^\text{de}_{x,10}$ are yet unknown and later
+instantiated by the LP solver.  The derivation with the instantiated
+variables is then a sound derivation in the verified program logic.
+
+The rules are syntax directed and applied inductively.  That means
+that we follow the inductive definition of the syntax.  For example,
+the outermost expression is a loop, so we use the rule \text{Q:Loop}
+at the root of the derivation tree.  At this point, we don't know yet
+whether a loop invariant exists.  But we produce the constraints
+$Q^\text{lo} = P^\text{lo}$ which is short for the following
+constraint set.
+$$
+\begin{array}{lll}
+q^\text{lo}_0 = p^\text{lo}_0 &
+q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
+q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
+q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
+q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
+q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
+\end{array}
+$$
+%
+This constraint set expresses the fact that the potential before and
+after the loop body is equal and thus constitutes an invariant.
+
+In general, the weakening rule is applied after every syntax directed
+rule.  Applying weakening everywhere does not lead to performance
+problems. However, it can be left out in practice at some places to
+increase the efficiency of the tool.
+
+The weakening operation $\succeq_{\Gamma}$ is defined in the rule
+\textsc{Relax}.  For example, we have
+$$
+\begin{array}{lll}
+P^\text{de} \succeq_{(\cdot)} P^\text{we}
+& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
+& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
+& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+\end{array}
+$$
+
+After the constraint generation, the LP solver is provided with a
+so-called objective function that it will minimize.  We wish to
+minimize the initial potential, which is the resource bound on the
+whole program.  Here it is given by $Q$.  Moreover, we would like to
+express that minimization of linear potential such as
+$$
+q_{10,x} |[10,x]|
+$$
+takes priority over minimization of constant potential such as
+$$
+q_{0,10} |[0,10]| \; .
+$$
+
+To this end, modern LP solvers support efficient resolving of
+constraint sets with additional constraints: First we take our initial
+constraint set as given in \pref{fig:derivation}.  We then as the
+solver to find a solution that satisfies the constraint and minimizes
+the linear expression 
+$$
+q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} \; .
+$$
+Note the linear penalties (this is a random choice) that we give to
+some of the factors.  We use it to express that the interval $[10,x]$
+is smaller than the interval $[0,x]$ and, similarly, that the interval
+$[x,0]$ is smaller than the interval $[x,10]$.
+
+The LP solver now returns a solution of the constraint set and an
+objective value, that is, a mapping from variables to floating-point
+numbers and the value of the objective function with this
+instantiation.  The solver also memorizes the optimization path that
+lead to the optimal solution.  In this case, the objective value would
+be
+$$
+5000
+$$
+since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
+We now add the constraint
+$$
+q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} = 5000
+$$
+to our constraint set and ask the solver to optimize the objective
+function
+$$
+q_0 + 11{\cdot}q_{0,10} \; .
+$$
+This happens in almost no time in practice.
+
+The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
+the derived bound is $0.5 |[0,x]|$.
+
+Another interesting note is that the constraints that we generate have
+a particularly simple form that is know as \emph{network problem}.
+Such problems can be solve in linear time in practice.
+
 \section{Catalog of Automatically Analyzed Programs}

### 2014-09-09 paper/main.tex

@@
 
-\sectskip
-\section{Conclusion}
-\label{sec:concl}
-\aftersectskip
-
-We have introduced an novel quantitative analysis system for CompCert
-Clight programs.  To the best of our knowledge, this article presents
-the first resource analysis framework for C that makes it possible to
-combine non-trivial automatically derived bounds with interactively
-derived bounds in a proof system that produces verifiable certificates
-for the bounds.  The main technical innovations are a quantitative
-Hoare logic for reasoning about user-defined resource cost and an
-automatic amortized analysis that derives bounds for programs whose
-resource consumption depends on (possibly negative) integers and
-non-sequential control flow as introduced by \code{break} and
-\code{return} statements.
-
-We will continue to improve our framework to reason more precisely
-about resource usage of system software such as the hypervisor kernel
-CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
-  and verifying}{is currently developed and verified}.  For one thing,
-we will improve the automation to derive \emph{polynomial bounds} and
-to handle non-linear size changes, as already developed for functional
-programs~\cite{HoffmannS13}.  For another thing, we will build on
-previous work~\cite{HoffmannMZ13} to generalize the quantitative
-Hoare logic to handle \emph{concurrent programs}\iffull{ with locks and
-lock-free data structures.}{.}
-
-% \acks
-
-% This research is based on work supported in part by NSF grants 1319671
-% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
-% Any opinions, findings, and conclusions contained in this document are
-% those of the authors and do not reflect the views of these agencies.
-
-\ifx\fullversion\undefined{}\else{
-%\clearpage
-\appendix
-
-\section{Example Derivation in the Automatic System}
-\label{app:derivation}
-
 \newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
@@
 $$
-\flushleft Linear Objective Function:
-$$
+\vspace{1.5ex}
+
+\begin{tabular}{l@{\hspace{5em}}l}
+ Linear Objective Function: &  Constant Objective Function: \\
+\hspace{3em} $
 q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x}
-$$
-\flushleft Constant Objective Function:
-$$
+$
+&
+\hspace{3em}
+$
 q_0 + 11{\cdot}q_{0,10}  
-$$
+$
+\end{tabular}
+
 \caption{An example derivation as produced by the automatic tool.}
@@
 \end{figure*}
+\sectskip
+\section{Conclusion}
+\label{sec:concl}
+\aftersectskip
+
+We have introduced an novel quantitative analysis system for CompCert
+Clight programs.  To the best of our knowledge, this article presents
+the first resource analysis framework for C that makes it possible to
+combine non-trivial automatically derived bounds with interactively
+derived bounds in a proof system that produces verifiable certificates
+for the bounds.  The main technical innovations are a quantitative
+Hoare logic for reasoning about user-defined resource cost and an
+automatic amortized analysis that derives bounds for programs whose
+resource consumption depends on (possibly negative) integers and
+non-sequential control flow as introduced by \code{break} and
+\code{return} statements.
+
+We will continue to improve our framework to reason more precisely
+about resource usage of system software such as the hypervisor kernel
+CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
+  and verifying}{is currently developed and verified}.  For one thing,
+we will improve the automation to derive \emph{polynomial bounds} and
+to handle non-linear size changes, as already developed for functional
+programs~\cite{HoffmannS13}.  For another thing, we will build on
+previous work~\cite{HoffmannMZ13} to generalize the quantitative
+Hoare logic to handle \emph{concurrent programs}\iffull{ with locks and
+lock-free data structures.}{.}
+
+% \acks
+
+% This research is based on work supported in part by NSF grants 1319671
+% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
+% Any opinions, findings, and conclusions contained in this document are
+% those of the authors and do not reflect the views of these agencies.
+
+\ifx\fullversion\undefined{}\else{
+%\clearpage
+\appendix
+
+\newpage
+
+\section{Example Derivation in the Automatic System}
+\label{app:derivation}
 
@@
 be
-$$
+$
 5000
-$$
+$
 since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
@@
 Such problems can be solve in linear time in practice.
+
+\newpage
 

### 2014-09-09 paper/main.tex
fixed some typos

@@
 instantiation.  The solver also memorizes the optimization path that
-lead to the optimal solution.  In this case, the objective value would
+led to the optimal solution.  In this case, the objective value would
 be
@@
 Another interesting note is that the constraints that we generate have
-a particularly simple form that is know as \emph{network problem}.
-Such problems can be solve in linear time in practice.
+a particularly simple form that is known as \emph{network problem}.
+Such problems can be solved in linear time in practice.
 

### 2014-10-22 paper/main.tex
start working on pldi intro

@@
 
-In software engineering and software verification, we often would like
-to have static information about the quantitative behavior of
-programs.  For example, stack and heap-space bounds are important to
-ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
-% Clock-cycle bounds are needed to guarantee the safety of real-time
-% systems [XXX wilhelm].  
-Static energy usage information is critical for autonomous systems and
-has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
-Worst-case time bounds can help to create constant-time
-implementations that prevent side-channel
-attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
-bounds are used to ensure the accuracy of programs that are executed
-on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
-needed to verify cryptographic protocols~\cite{BartheGB09}.
-% XXX Mention differential privacy here?
-%
-In general, quantitative performance information at design time can
-provide useful feedback for developers.
-
-Static analysis of quantitative properties of imperative programs is
-an active research area and recent years have seen many innovations.
-Notable tools that have been developed include
-SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
-PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
-%
-While the these tools can derive impressive results for realistic
-software, there still exist shortcomings that hamper the application
-of existing techniques in practice:
-\itemskip
+In software engineering and software verification,
+we often would like to have static information
+about the quantitative behavior of programs.
+For example, stack and heap-space bounds
+are important to ensure the reliability of
+safety-critical systems~\cite{veristack14,Regehr05}.
+Static energy usage information is critical
+for autonomous systems and has applications in
+cloud computing~\cite{CohenZSL12,CarrollH10}.
+Worst-case time bounds can help to create
+constant-time implementations that prevent
+side-channel attacks~\cite{KasperS09,BartheBCLP14}.
+Loop and recursion-depth bounds are used to
+ensure the accuracy of programs that are executed
+on unreliable hardware~\cite{CarbinMR13} and
+complexity bounds are needed to verify cryptographic
+protocols~\cite{BartheGB09}.  In general, quantitative
+performance information at design time can provide useful
+feedback for developers.
+
+Static analysis of quantitative properties of imperative
+programs is an active research area and recent years
+have seen many innovations.  Notable tools that
+have been developed include SPEED~\cite{GulwaniMC09},
+KoAT~\cite{BrockschmidtEFFG14}, PUBS~\cite{AlbertAGPZ12},
+Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
+
+These tools can derive impressive results on realistic
+system software.  However, there are still theoritical
+shortcomings that remain open problems to this day:
+
 \begin{itemize}
-\item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
-  without enabling \emph{user interaction} to manually or
-  semi-automatically derive bounds for challenging parts of the
-  program.
-\itemskipIn
-\item The computed bounds are often \emph{non-compositional} local
-  bounds (for a single (nested) loop) that are difficult to combine to
+\item The computed bounds are often
+  \emph{non-compositional} local bounds (for a single
+  (nested) loop) that are difficult to combine to
   global whole program bounds.
 \itemskipIn
-\item Existing techniques often rely on complex external tools such as
-  abstract interpretation--based invariant
-  generation~\cite{GulwaniMC09} or translation of the program into a
-  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
-  providing \emph{verifiable certificates} for the correctness of the
-  derived bound.
+\item When an automatic tool fails on a loop of the
+  program, no user interaction is possible to derive
+  a bound on the full program.
+\itemskipIn
+\item Existing tools work on a highly abstracted
+  version of the program, thus, are unable to provide
+  a simple syntactic proof of the correctness of their
+  bound.
+\itemskipIn
+\item Maybe we can mention resource freeing                    XXX
 \end{itemize}
-\itemskip
-%
-While there has been much progress in static quantitative analysis,
-Knuth correctly points out in a recent interview~\cite{KnuthInter}
-that the state-of-the-art in formal quantitative methods still falls
-short in comparison with semantic techniques.
-%
-\itemskip
-\begin{quote}
-  % Consider, for example, a computer program that sorts a list of
-  % numbers into order. 
-  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
-  we have formal definitions of semantics, and tools by which we can
-  verify that sorting is indeed always achieved. My job is to go
-  beyond correctness, to an analysis of such things as the program's
-  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
-  describes the program's performance, and all of my colleagues agree
-  with me that the recurrence is "obviously" valid. Yet I have no
-  formal tools by which I can prove that my recurrence is right. I
-  don't really understand my reasoning processes at all!
-  \vspace{-4ex}
-  \begin{flushright}
-    -- Donald E. Knuth, 2014
-  \end{flushright}
-\end{quote}
-\itemskip
-%
-In this work, we develop a resource analysis framework for C programs
-that is based on a solid semantic foundation.  The choice of C is
-primarily motivated by \iffull{our}{the} ongoing work on the formally verified
-hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
-developed in C and is supposed to provide verified guaranties on
-timing and memory usage.  Moreover, C is a natural choice because it
-is still the most widely used language for system development; in
-particular in safety-critical embedded and real-time systems where
-resource bound analyses are often required by regulatory
-authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
-%
-\itemskip
-\begin{enumerate}
-\item We define an operational cost semantics for CompCert Clight that
-  defines the resource consumption of terminating and diverging
-  executions.  The cost is parametric in a user-definable cost metric
-  and can be negative if resources are released during an execution.
-
-\item We develop a quantitative Hoare logic for interactively deriving
-  resource bounds for Clight programs.  The logic is implemented and
-  proved sound in the Coq Proof Assistant with respect to the cost
-  semantics of Clight.
+
+This paper presents a new framework to derive worst-case
+resource bounds for C programs.  Our choice of C is
+mainly motivated by its strong presence in system and
+embedded software.  These applications often have to meet
+strict time and memory constraints.  To ensure that these
+requirements are met, profiling is not enough because non
+exhaustive.  In contrast, static analysis tools provide
+stronger and sound guarantees.
+
+Our compositional framework is built of three parts: 1-
+a clear cost model for C programs based on instrumented
+semantics; 2- a quantitative Hoare logic allowing
+compositional reasoning on the resource consumption of
+C programs at a syntactic level; 3- an automated tool
+that derives bounds in the previous logic.
+
+Compared to existing work, our framework is a jump in
+compositionality and correctness: the analysis is no
+longer centered on single loops but on complete programs,
+a mix of automation and interaction can be used to derive
+global bounds, and bounds are computed and proved within
+a clear syntactic deductive system.  Compositionality is
+due first to the logic that soundly allows mixes of hand
+derived bounds and automatically generated bounds.  It is
+also due to the automated tool that keeps track of loop
+bounds and size changes of variables in a unified system.
+
+The cost model and the quantitative Hoare logic
+presented are based on a recent work~\cite{veristack14}
+that formally verifies bounds on the stack usage of
+C programs.  We generalize this work by allowing to
+reason on any resource.  The automation, which is the
+keystone of this paper, is based on amortized analysis
+techniques~\cite{jan-something}.  It makes the analysis
+of large programs tractable by automatically generating
+resource bounds in the previous logic on a large class
+of programs.
+
+Our automated analysis is able to infer linear
+resource bounds on C programs with integer loops.
+To our knowledge this is the first tool based on
+amortized analysis able to deal with negative numbers and
+differences of variables.  It is also the first resource
+analysis tool for C that deals naturally with recursive
+functions, resources that can be freed (e.g. memory),
+and sequenced loops.  Automated tools for functional
+programs already existed~\cite{Jost03,HoffmannAH12}
+but it has been a long-time open problem to apply it to
+C-like programming languages with control flow depending
+on integer arithmetic.
+
+Compared to more classical approaches based on ranking
+functions, tools that are based on amortized analysis
+only need one mechanism to provide features that are
+usually treated non uniformly:
+
+\begin{itemize}
+\item interaction between \emph{sequential loops or
+  function calls} through size changes of variables,
 \itemskipIn
-\item We describe an automatic static analysis that computes bounds
-  together with derivations in the quantitative Hoare logic.  The
-  analysis is inspired by type-based amortized resource analysis for
-  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
-  derivations by generating simple linear constraints that can be
-  solved by an off-the-shelf LP solver, and does not require any
-  fixpoint computations to obtain loop invariants.
-\itemskipIn
-\item We show with a publicly available prototype implementation, and
-  an experimental evaluation, that our novel amortized resource
-  analysis works precisely and efficiently for challenging loop and
-  recursion patterns, and that the derived constant factors in the
-  bounds are close to or identical with the optimal ones.
-% \item We prove the completeness of the program logic
-%   with respect to the operational cost semantics.
-\end{enumerate}
-\itemskip
-%
-To the best of our knowledge, this article presents the first resource
-analysis framework for C that makes it possible to combine non-trivial
-automatically derived bounds with interactively derived bounds in a
-proof system that produces verifiable certificates for the bounds.
-Our approach complements existing work since it provides a semantic
-foundation for the computation of bounds while still supporting
-automation that sometimes goes beyond the capabilities of existing
-techniques.  Furthermore, the automatic amortized analysis and the
-quantitative logic are naturally compositional and describe the
-resource behavior of code fragments without referring to the source
-code.  As a result, we directly derive global bounds that are
-functions of the input parameters of the program.  Another unique
-feature of our system is, that it can handle resources like memory
-that may become available during execution.
-
-Our starting point is a recent work~\cite{veristack14} \iffull{in
-  which we have}{that has} formally verified bounds on the stack usage
-of C programs.  A key part of this stack-bound verification is based
-on a quantitative Hoare logic for deriving abstract \emph{stack
-  bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
-  depth of (recursive) function calls.}  % For
-% non-recursive programs it is possible to compute derivations for such
-% stack bounds in the quantitative Hoare logic automatically.  
-% By compiling these bounds together with the C program, we have derived
-% verified stack bounds for x86 assembly code.
-%
-Our \emph{first contribution} is a generalization of the quantitative
-Hoare logic for CompCert Clight that enables us to derive resource
-bounds that are parametric in the resource of interest.  % The
-% quantitative logic is implemented in Coq and proved sound with respect
-% to a small-step operational cost semantics for Clight.
-% that counts a
-% constant resource consumption for each atomic execution step.  The
-% resource consumption at each step is parametric in a user-definable cost
-% metric and can be negative to express the release of resources (as
-% needed, e.g., to reason about memory usage).  Furthermore, a specific
-% \emph{tick} function can be inserted in the source code to further
-% customize the cost model.
-In the implementation of the logic, we use
-a shallow embedding in Coq which makes the logic very flexible.  There
-is practically no limitation on the format of the derived resource
-bounds since they can be any function that is definable in Coq.  The
-derived bounds can also be parametric in a set of cost metrics or
-specifically apply to one fixed cost metric.
-
-The precision and expressivity of the quantitative Hoare logic provide
-an ample foundation for reasoning about resource consumption of Clight
-programs.  % The logic is naturally compositional and the derived
-% \emph{quantitative Hoare triples} can be used to specify the resource
-% consumption of library functions or to describe resource contracts
-% between different functions. 
-However, reasoning about quantitative
-properties can often be more tedious than reasoning about intentional
-properties.  Consequently, automation is inevitable to derive resource
-bounds for large code bases like CertiKOS.  Such an automation was
-easily achievable in the case of constant stack bounds for a code base
-that does not contain recursive functions~\cite{veristack14}.  In
-general however, resource bounds have to be parametric in the
-arguments of a function and depend on the number of loop iterations
-and (recursive) function calls that are performed by the function.
-
-For functional programs, there exist resource analysis systems---based
-on type-based amortized resource analysis~\cite{Jost03,
-  HoffmannAH12}---that can automatically derive complex polynomial
-bounds with tight constant factors.  This type-based amortized
-resource analysis has been an inspiration in the design of the
-quantitative Hoare logic and is therefore a natural candidate to
-automate the reasoning.  Type-based amortized resource analysis works
-well for functional programs with pattern matching but it has been a
-long-time open problem to apply it to C-like programs with control
-flow that depends on integer arithmetic, negative numbers, and non-linear
-control flow.
-% example, it is a
-% long-time open problem how to use type-based amortized analysis to
-% derive bounds for functions whose resource consumption depends on
-% (possibly negative) integers and non-sequential control flow as
-% introduced by \code{break} and \code{return} statements.
-% Unfortunately, these are the functions \iffull{with non-constant resource
-% consumption}{} that are most common in CertiKOS and other system
-% software.
-
-Our \emph{main contribution} is an automatic amortized resource
-analysis for CompCert Clight that computes derivations in the
-quantitative Hoare logic.  It is the first automatic amortized
-resource analysis that can derive bounds that depend on negative
-integers. It also handles programs with mutually recursive functions
-as well as \code{break} and \code{return} statements.  The
-potential-based approach of amortized analysis provides a single
-mechanism for analyzing programs that need special treatment in other
-techniques; including
-\itemskip
-\begin{itemize}
-\item interaction between \emph{sequential loops or function calls} through
-  size changes of variables,
-\itemskipIn
-\item \emph{nested loops} that influence each others number of iterations,
-\itemskipIn
-\item and \emph{tricky iteration patterns} as found for instance in
-  the Knuth-Morris-Pratt algorithm for string search.
+\item \emph{nested loops} that influence each other
+  with the same set of modified variables,
+\item and \emph{amortized bounds} as found
+  in the Knuth-Morris-Pratt algorithm for string
+  search.
 \end{itemize}
-\itemskip
-%
-The main innovation that makes the analysis practical for C is the use
-of interval sizes in potential functions instead of the sizes of
-variables.  This leads to precise bounds of programs whose resource
-consumption can be described as a function of sizes $|[x,y]| =
-\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
-instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
-  i=i+K)} where the step-size $K>0$ is a constant.
-%
-
-Despite the apparent simplicity of the new analysis system, it is able
-to reproduce results from the literature (e.g,
-SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
-abstract interpretation--based methods such as disjunctive invariant
-generation~\cite{GulwaniMC09} and symbolic backward
-execution~\cite{GulwaniZ10}.  In contrast with abstract
-interpretation--based methods, our technique does not require any
-fixpoint computations to obtain loop invariants.  The mechanism we
-designed is able to leverage local assertions such as $x < y$ that we
-collect along the branching points of the program to obtain global
-resource invariants.  We achieve this by generating a linear
-constraint system that reflects the resource cost and size changes of
-integer variables in the program; a solution of the linear program
-immediately yields a resource-usage bound for the C
-program.  % We can naturally, handle
-% advanced control flow such as break, return, iteration over negative
-% integer variables, and mutually-recursive functions.
-
-Following the development steps of automatic amortized analysis for
-functional programs~\cite{Jost03,HoffmannH10}, we deliberately
-restrict ourselves to linear bounds in the automatic analysis (there
-are no restrictions for the manually-derived bounds in the logic).
-More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
-where $a$ and $b$ are integer variables or constants.  The reason for
-our focus on linear bounds is mainly clarity of presentation.  We
-already experimented with an extension to multivariate resource
-polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
-rules considerably more involved and should better be described
-separately.  However, we developed the linear inference system so that
-the extension to polynomial bounds shall work smoothly.
-
-We have evaluated the automatic analysis with system code and examples
-from the literature.  \iffull{\pref{app:cat}}{The extended version
-  of this article~\cite{anon_extended}}{} contains more than $30$
-challenging loop and recursion patterns that we collected from open
-source software and the literature.  Our analysis can find
-asymptotically tight bounds for all but 1 of these patterns, and in
-most cases the derived constant factors are tight.  To compare our
-automatic analyzer with existing techniques, we tested our examples
-with tools such as KoAT~\cite{BrockschmidtEFFG14},
-Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
-show that the bounds that we derive are often more precise than those
-derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
-uses amortization techniques, is able to achieve a similar precision.
-%
-Several micro benchmarks demonstrate the practicality and
-expressiveness of the quantitative Hoare logic.  For example, we
-derive a logarithmic bound for a binary search function, a bound that
-depends on the contents of an array to describe the exact cost of a
-function that finds a maximal element in an array, and a linear
-bound that amortizes the cost of $k$ successive increments to a binary
-counter.
-
-% To show the practicality of our approach we developed a case study
-% with different real-time schedulers implemented in C.  To perform a
-% schedulability analysis that takes into account the overhead of the
-% scheduler we derived time bounds for context switching and
-% initialization of the scheduler.  We then computed time bounds for a
-% set of given user task and took account the previously computed
-% scheduler overhead to determine whether the scheduling requirements of
-% all user tasks can be met.
-
-% With the intention, of making the material easily accessible we first
-% informally describe the new automatic amortized analysis
-% (\pref{sec:inform}).  We then formalize the intuition with the
-% operational cost semantics an the inference rules for the amortized
-% analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
-% quantitative Hoare logic and its soundness proof that is formalized in
-% Coq.  In \pref{sec:inter} we prove the soundness of the automatic
-% amortized analysis by showing that derivation using the inference
-% rules can be seen as derivations in the quantitative logic.  We also
-% show how to integrate hand-derived bounds with automatically-derived
-% bounds.  \pref{sec:exper} contains the results of the experimental
-% evaluation and \pref{sec:related} describes related research.
+
+This paper presents a compositional framework for
+deriving certified resource bounds on C programs. We
+emphasize the following contributions: 1- an automatic
+analysis that generates linear compositional and
+certified resource bounds; 2- an infrastructure verified
+in Coq that lets a programmer derive complex free-form
+resource bounds on complete C programs.
+
+Give a description of the sections here.                    XXX
+
+% In software engineering and software verification, we often would like
+% to have static information about the quantitative behavior of
+% programs.  For example, stack and heap-space bounds are important to
+% ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
+% % Clock-cycle bounds are needed to guarantee the safety of real-time
+% % systems [XXX wilhelm].
+% Static energy usage information is critical for autonomous systems and
+% has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
+% Worst-case time bounds can help to create constant-time
+% implementations that prevent side-channel
+% attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
+% bounds are used to ensure the accuracy of programs that are executed
+% on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
+% needed to verify cryptographic protocols~\cite{BartheGB09}.
+% % XXX Mention differential privacy here?
+% %
+% In general, quantitative performance information at design time can
+% provide useful feedback for developers.
+%
+% Static analysis of quantitative properties of imperative programs is
+% an active research area and recent years have seen many innovations.
+% Notable tools that have been developed include
+% SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
+% PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
+% %
+% While the these tools can derive impressive results for realistic
+% software, there still exist shortcomings that hamper the application
+% of existing techniques in practice:
+% \itemskip
+% \begin{itemize}
+% \item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
+%   without enabling \emph{user interaction} to manually or
+%   semi-automatically derive bounds for challenging parts of the
+%   program.
+% \itemskipIn
+% \item The computed bounds are often \emph{non-compositional} local
+%   bounds (for a single (nested) loop) that are difficult to combine to
+%   global whole program bounds.
+% \itemskipIn
+% \item Existing techniques often rely on complex external tools such as
+%   abstract interpretation--based invariant
+%   generation~\cite{GulwaniMC09} or translation of the program into a
+%   term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
+%   providing \emph{verifiable certificates} for the correctness of the
+%   derived bound.
+% \end{itemize}
+% \itemskip
+% %
+% While there has been much progress in static quantitative analysis,
+% Knuth correctly points out in a recent interview~\cite{KnuthInter}
+% that the state-of-the-art in formal quantitative methods still falls
+% short in comparison with semantic techniques.
+% %
+% \itemskip
+% \begin{quote}
+%   % Consider, for example, a computer program that sorts a list of
+%   % numbers into order.
+%   $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
+%   we have formal definitions of semantics, and tools by which we can
+%   verify that sorting is indeed always achieved. My job is to go
+%   beyond correctness, to an analysis of such things as the program's
+%   running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
+%   describes the program's performance, and all of my colleagues agree
+%   with me that the recurrence is "obviously" valid. Yet I have no
+%   formal tools by which I can prove that my recurrence is right. I
+%   don't really understand my reasoning processes at all!
+%   \vspace{-4ex}
+%   \begin{flushright}
+%     -- Donald E. Knuth, 2014
+%   \end{flushright}
+% \end{quote}
+% \itemskip
+% %
+% In this work, we develop a resource analysis framework for C programs
+% that is based on a solid semantic foundation.  The choice of C is
+% primarily motivated by \iffull{our}{the} ongoing work on the formally verified
+% hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
+% developed in C and is supposed to provide verified guaranties on
+% timing and memory usage.  Moreover, C is a natural choice because it
+% is still the most widely used language for system development; in
+% particular in safety-critical embedded and real-time systems where
+% resource bound analyses are often required by regulatory
+% authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
+% %
+% \itemskip
+% \begin{enumerate}
+% \item We define an operational cost semantics for CompCert Clight that
+%   defines the resource consumption of terminating and diverging
+%   executions.  The cost is parametric in a user-definable cost metric
+%   and can be negative if resources are released during an execution.
+%
+% \item We develop a quantitative Hoare logic for interactively deriving
+%   resource bounds for Clight programs.  The logic is implemented and
+%   proved sound in the Coq Proof Assistant with respect to the cost
+%   semantics of Clight.
+% \itemskipIn
+% \item We describe an automatic static analysis that computes bounds
+%   together with derivations in the quantitative Hoare logic.  The
+%   analysis is inspired by type-based amortized resource analysis for
+%   functional programs~\cite{Jost03,HoffmannAH11}.  It computes
+%   derivations by generating simple linear constraints that can be
+%   solved by an off-the-shelf LP solver, and does not require any
+%   fixpoint computations to obtain loop invariants.
+% \itemskipIn
+% \item We show with a publicly available prototype implementation, and
+%   an experimental evaluation, that our novel amortized resource
+%   analysis works precisely and efficiently for challenging loop and
+%   recursion patterns, and that the derived constant factors in the
+%   bounds are close to or identical with the optimal ones.
+% % \item We prove the completeness of the program logic
+% %   with respect to the operational cost semantics.
+% \end{enumerate}
+% \itemskip
+% %
+% To the best of our knowledge, this article presents the first resource
+% analysis framework for C that makes it possible to combine non-trivial
+% automatically derived bounds with interactively derived bounds in a
+% proof system that produces verifiable certificates for the bounds.
+% Our approach complements existing work since it provides a semantic
+% foundation for the computation of bounds while still supporting
+% automation that sometimes goes beyond the capabilities of existing
+% techniques.  Furthermore, the automatic amortized analysis and the
+% quantitative logic are naturally compositional and describe the
+% resource behavior of code fragments without referring to the source
+% code.  As a result, we directly derive global bounds that are
+% functions of the input parameters of the program.  Another unique
+% feature of our system is, that it can handle resources like memory
+% that may become available during execution.
+%
+% Our starting point is a recent work~\cite{veristack14} \iffull{in
+%   which we have}{that has} formally verified bounds on the stack usage
+% of C programs.  A key part of this stack-bound verification is based
+% on a quantitative Hoare logic for deriving abstract \emph{stack
+%   bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
+%   depth of (recursive) function calls.}  % For
+% % non-recursive programs it is possible to compute derivations for such
+% % stack bounds in the quantitative Hoare logic automatically.
+% % By compiling these bounds together with the C program, we have derived
+% % verified stack bounds for x86 assembly code.
+% %
+% Our \emph{first contribution} is a generalization of the quantitative
+% Hoare logic for CompCert Clight that enables us to derive resource
+% bounds that are parametric in the resource of interest.  % The
+% % quantitative logic is implemented in Coq and proved sound with respect
+% % to a small-step operational cost semantics for Clight.
+% % that counts a
+% % constant resource consumption for each atomic execution step.  The
+% % resource consumption at each step is parametric in a user-definable cost
+% % metric and can be negative to express the release of resources (as
+% % needed, e.g., to reason about memory usage).  Furthermore, a specific
+% % \emph{tick} function can be inserted in the source code to further
+% % customize the cost model.
+% In the implementation of the logic, we use
+% a shallow embedding in Coq which makes the logic very flexible.  There
+% is practically no limitation on the format of the derived resource
+% bounds since they can be any function that is definable in Coq.  The
+% derived bounds can also be parametric in a set of cost metrics or
+% specifically apply to one fixed cost metric.
+%
+% The precision and expressivity of the quantitative Hoare logic provide
+% an ample foundation for reasoning about resource consumption of Clight
+% programs.  % The logic is naturally compositional and the derived
+% % \emph{quantitative Hoare triples} can be used to specify the resource
+% % consumption of library functions or to describe resource contracts
+% % between different functions.
+% However, reasoning about quantitative
+% properties can often be more tedious than reasoning about intentional
+% properties.  Consequently, automation is inevitable to derive resource
+% bounds for large code bases like CertiKOS.  Such an automation was
+% easily achievable in the case of constant stack bounds for a code base
+% that does not contain recursive functions~\cite{veristack14}.  In
+% general however, resource bounds have to be parametric in the
+% arguments of a function and depend on the number of loop iterations
+% and (recursive) function calls that are performed by the function.
+%
+% For functional programs, there exist resource analysis systems---based
+% on type-based amortized resource analysis~\cite{Jost03,
+%   HoffmannAH12}---that can automatically derive complex polynomial
+% bounds with tight constant factors.  This type-based amortized
+% resource analysis has been an inspiration in the design of the
+% quantitative Hoare logic and is therefore a natural candidate to
+% automate the reasoning.  Type-based amortized resource analysis works
+% well for functional programs with pattern matching but it has been a
+% long-time open problem to apply it to C-like programs with control
+% flow that depends on integer arithmetic, negative numbers, and non-linear
+% control flow.
+% % example, it is a
+% % long-time open problem how to use type-based amortized analysis to
+% % derive bounds for functions whose resource consumption depends on
+% % (possibly negative) integers and non-sequential control flow as
+% % introduced by \code{break} and \code{return} statements.
+% % Unfortunately, these are the functions \iffull{with non-constant resource
+% % consumption}{} that are most common in CertiKOS and other system
+% % software.
+%
+% Our \emph{main contribution} is an automatic amortized resource
+% analysis for CompCert Clight that computes derivations in the
+% quantitative Hoare logic.  It is the first automatic amortized
+% resource analysis that can derive bounds that depend on negative
+% integers. It also handles programs with mutually recursive functions
+% as well as \code{break} and \code{return} statements.  The
+% potential-based approach of amortized analysis provides a single
+% mechanism for analyzing programs that need special treatment in other
+% techniques; including
+% \itemskip
+% \begin{itemize}
+% \item interaction between \emph{sequential loops or function calls} through
+%   size changes of variables,
+% \itemskipIn
+% \item \emph{nested loops} that influence each others number of iterations,
+% \itemskipIn
+% \item and \emph{tricky iteration patterns} as found for instance in
+%   the Knuth-Morris-Pratt algorithm for string search.
+% \end{itemize}
+% \itemskip
+% %
+% The main innovation that makes the analysis practical for C is the use
+% of interval sizes in potential functions instead of the sizes of
+% variables.  This leads to precise bounds of programs whose resource
+% consumption can be described as a function of sizes $|[x,y]| =
+% \max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
+% instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
+%   i=i+K)} where the step-size $K>0$ is a constant.
+% %
+%
+% Despite the apparent simplicity of the new analysis system, it is able
+% to reproduce results from the literature (e.g,
+% SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
+% abstract interpretation--based methods such as disjunctive invariant
+% generation~\cite{GulwaniMC09} and symbolic backward
+% execution~\cite{GulwaniZ10}.  In contrast with abstract
+% interpretation--based methods, our technique does not require any
+% fixpoint computations to obtain loop invariants.  The mechanism we
+% designed is able to leverage local assertions such as $x < y$ that we
+% collect along the branching points of the program to obtain global
+% resource invariants.  We achieve this by generating a linear
+% constraint system that reflects the resource cost and size changes of
+% integer variables in the program; a solution of the linear program
+% immediately yields a resource-usage bound for the C
+% program.  % We can naturally, handle
+% % advanced control flow such as break, return, iteration over negative
+% % integer variables, and mutually-recursive functions.
+%
+% Following the development steps of automatic amortized analysis for
+% functional programs~\cite{Jost03,HoffmannH10}, we deliberately
+% restrict ourselves to linear bounds in the automatic analysis (there
+% are no restrictions for the manually-derived bounds in the logic).
+% More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
+% where $a$ and $b$ are integer variables or constants.  The reason for
+% our focus on linear bounds is mainly clarity of presentation.  We
+% already experimented with an extension to multivariate resource
+% polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
+% rules considerably more involved and should better be described
+% separately.  However, we developed the linear inference system so that
+% the extension to polynomial bounds shall work smoothly.
+%
+% We have evaluated the automatic analysis with system code and examples
+% from the literature.  \iffull{\pref{app:cat}}{The extended version
+%   of this article~\cite{anon_extended}}{} contains more than $30$
+% challenging loop and recursion patterns that we collected from open
+% source software and the literature.  Our analysis can find
+% asymptotically tight bounds for all but 1 of these patterns, and in
+% most cases the derived constant factors are tight.  To compare our
+% automatic analyzer with existing techniques, we tested our examples
+% with tools such as KoAT~\cite{BrockschmidtEFFG14},
+% Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
+% show that the bounds that we derive are often more precise than those
+% derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
+% uses amortization techniques, is able to achieve a similar precision.
+% %
+% Several micro benchmarks demonstrate the practicality and
+% expressiveness of the quantitative Hoare logic.  For example, we
+% derive a logarithmic bound for a binary search function, a bound that
+% depends on the contents of an array to describe the exact cost of a
+% function that finds a maximal element in an array, and a linear
+% bound that amortizes the cost of $k$ successive increments to a binary
+% counter.
+%
+% % To show the practicality of our approach we developed a case study
+% % with different real-time schedulers implemented in C.  To perform a
+% % schedulability analysis that takes into account the overhead of the
+% % scheduler we derived time bounds for context switching and
+% % initialization of the scheduler.  We then computed time bounds for a
+% % set of given user task and took account the previously computed
+% % scheduler overhead to determine whether the scheduling requirements of
+% % all user tasks can be met.
+%
+% % With the intention, of making the material easily accessible we first
+% % informally describe the new automatic amortized analysis
+% % (\pref{sec:inform}).  We then formalize the intuition with the
+% % operational cost semantics an the inference rules for the amortized
+% % analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
+% % quantitative Hoare logic and its soundness proof that is formalized in
+% % Coq.  In \pref{sec:inter} we prove the soundness of the automatic
+% % amortized analysis by showing that derivation using the inference
+% % rules can be seen as derivations in the quantitative logic.  We also
+% % show how to integrate hand-derived bounds with automatically-derived
+% % bounds.  \pref{sec:exper} contains the results of the experimental
+% % evaluation and \pref{sec:related} describes related research.
 

### 2014-10-23 paper/main.tex
work on the intro

@@
 Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
-
 These tools can derive impressive results on realistic
@@
 \itemskipIn
-\item Maybe we can mention resource freeing                    XXX
+\item Resources that can be freed like memory cannot
+  be handled with the usual techniques.
 \end{itemize}
@@
 embedded software.  These applications often have to meet
-strict time and memory constraints.  To ensure that these
+strict time and memory constraints.
+
+XXX [this has nothing to do here]    To ensure that these
 requirements are met, profiling is not enough because non
@@
 that formally verifies bounds on the stack usage of
-C programs.  We generalize this work by allowing to
-reason on any resource.  The automation, which is the
+C programs.  We generalize this work by allowing
+to reason on any resource (e.g. evaluation steps,
+memory, or stack usage).  The automation, which is the
 keystone of this paper, is based on amortized analysis
@@
 of large programs tractable by automatically generating
-resource bounds in the previous logic on a large class
-of programs.
+resource bounds that are certified in the previous logic
+on a large class of programs.
 
@@
 \end{itemize}
+
+The main innovation that makes amortized analysis
+work on imperative languages is to track multivariate
+quantites instead of simple functions of program
+variables.  This leads to precise bounds expressed as
+functions of sizes $|[x,y]| = \max(0,y-x)$ of intervals.
 

### 2014-10-27 paper/main.tex
work on the new speach

--- 2014-10-22 paper/main.tex
@@
 
-Static analysis of quantitative properties of imperative
-programs is an active research area and recent years
-have seen many innovations.  Notable tools that
-have been developed include SPEED~\cite{GulwaniMC09},
-KoAT~\cite{BrockschmidtEFFG14}, PUBS~\cite{AlbertAGPZ12},
-Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
-
-These tools can derive impressive results on realistic
-system software.  However, there are still theoritical
-shortcomings that remain open problems to this day:
-
-\begin{itemize}
-\item The computed bounds are often
-  \emph{non-compositional} local bounds (for a single
-  (nested) loop) that are difficult to combine to
-  global whole program bounds.
-\itemskipIn
-\item When an automatic tool fails on a loop of the
-  program, no user interaction is possible to derive
-  a bound on the full program.
-\itemskipIn
-\item Existing tools work on a highly abstracted
-  version of the program, thus, are unable to provide
-  a simple syntactic proof of the correctness of their
-  bound.
-\itemskipIn
-\item Maybe we can mention resource freeing                    XXX
-\end{itemize}
-
-This paper presents a new framework to derive worst-case
-resource bounds for C programs.  Our choice of C is
-mainly motivated by its strong presence in system and
-embedded software.  These applications often have to meet
-strict time and memory constraints.  To ensure that these
-requirements are met, profiling is not enough because non
-exhaustive.  In contrast, static analysis tools provide
-stronger and sound guarantees.
-
-Our compositional framework is built of three parts: 1-
-a clear cost model for C programs based on instrumented
-semantics; 2- a quantitative Hoare logic allowing
-compositional reasoning on the resource consumption of
-C programs at a syntactic level; 3- an automated tool
-that derives bounds in the previous logic.
-
-Compared to existing work, our framework is a jump in
-compositionality and correctness: the analysis is no
-longer centered on single loops but on complete programs,
-a mix of automation and interaction can be used to derive
-global bounds, and bounds are computed and proved within
-a clear syntactic deductive system.  Compositionality is
-due first to the logic that soundly allows mixes of hand
-derived bounds and automatically generated bounds.  It is
-also due to the automated tool that keeps track of loop
-bounds and size changes of variables in a unified system.
-
-The cost model and the quantitative Hoare logic
-presented are based on a recent work~\cite{veristack14}
-that formally verifies bounds on the stack usage of
-C programs.  We generalize this work by allowing to
-reason on any resource.  The automation, which is the
-keystone of this paper, is based on amortized analysis
-techniques~\cite{jan-something}.  It makes the analysis
-of large programs tractable by automatically generating
-resource bounds in the previous logic on a large class
-of programs.
+% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW
+
+Available techniques split in two categories.  In the
+first one, tools derive impressive bounds for
+imperative programs, but are not fully compositional.
+Notable recent work like
+SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
+PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
+and LOOPUS~\cite{SinnZV14} falls in this category.
+They lack compositionality in at least three ways.
+First, they all base their analysis on some form of
+\emph{ranking-function}.
+Since this
+is often a local analysis, the loop bounds are expressed
+as arithmetic expressions in terms of variables as they
+are just before the loop.  This makes it hard to give a
+resource bound on sequenced loops or function calls in
+terms of the input parameters of a procedure.
+XXX MENTION GULWANI PAPER [1] XXX
+Second, while all imperative programming languages
+provide a function or procedure abstraction, all
+available tools are not able to abstract their resource
+behavior and have to inline the procedure body to perform
+any kind of analysis.  Third, when such a tool fails on
+given program, the only way out to get a resource bound
+is to rewrite the program.  There is no possibility
+of sound interaction with the tool, and manual proofs
+cannot be composed with automatically inferred bounds.
+One of the biggest challenge in the way of interaction
+with these automated tools is that they work on a highly
+abstract version of the program that has a lot more
+behaviors than the original program.
+
+In the second category, tools also derive very tight
+bounds on functional programs~\cite{Jost03,HoffmannAH12}.
+These tools are highly compositional, scale for large
+programs, and work directly on the syntax.  Despite all
+these theoretical and practical advantages, it has been
+a long-time open problem how to make them work in the
+presence of negative integers, mutation and non-linear
+control flow.  For example, the first challenge of
+imperative programs is the absence of useful type system
+to hook the analysis on.  Another issue is that these
+tools associate a \emph{potential} directly to program
+variables, doing this in imperative programs would
+fail for loops as simple as \lstinline{for(i=x;i<y;i++)}
+where $y - i$ decreases, but not $|i|$.
+Mutation is yet another challenge, while functional
+programs relate multiple fixed inputs (the typing
+context) with one output (the expression typed),
+imperative programs can change the value of variables
+in scope, so the potential of these variables has to be
+updated after mutations or unsound bounds can be derived.
+
+This paper presents a new framework for automatically
+deriving resource bounds on C programs.  This new
+approach is an attempt to unify the two previous
+categories: it solves the compositionality issues of
+imperative world by adapting amortized-analysis based
+techniques from the functional world.
 
@@
 To our knowledge this is the first tool based on
-amortized analysis able to deal with negative numbers and
-differences of variables.  It is also the first resource
-analysis tool for C that deals naturally with recursive
-functions, resources that can be freed (e.g. memory),
-and sequenced loops.  Automated tools for functional
-programs already existed~\cite{Jost03,HoffmannAH12}
-but it has been a long-time open problem to apply it to
-C-like programming languages with control flow depending
-on integer arithmetic.
-
-Compared to more classical approaches based on ranking
-functions, tools that are based on amortized analysis
-only need one mechanism to provide features that are
-usually treated non uniformly:
+amortized analysis able to deal with negative numbers
+and differences of variables.  It is also the first
+resource analysis tool for C that deals naturally with
+recursive functions, resources that can be freed (e.g.
+memory), and sequenced loops.  Compared to more classical
+apporaches based on ranking functions, our tool inherits
+the benefits of amortized reasoning and can handle using
+only one simple mechanism:
 
 \begin{itemize}
-\item interaction between \emph{sequential loops or
+\item interactions between \emph{sequential loops or
   function calls} through size changes of variables,
@@
 \item \emph{nested loops} that influence each other
-  with the same set of modified variables,
-\item and \emph{amortized bounds} as found
-  in the Knuth-Morris-Pratt algorithm for string
-  search.
+  with the same set of modified variables
+\itemskipIn
+\item and \emph{amortized bounds} as found, for example, in
+  the Knuth-Morris-Pratt algorithm for string search.
 \end{itemize}
 
-This paper presents a compositional framework for
-deriving certified resource bounds on C programs. We
-emphasize the following contributions: 1- an automatic
-analysis that generates linear compositional and
-certified resource bounds; 2- an infrastructure verified
-in Coq that lets a programmer derive complex free-form
-resource bounds on complete C programs.
+The main innovations that make amortized analysis work
+on imperative languages is to base the analysis on a
+Hoare-like logic and track multivariate quantities instead
+of program variables directly.  This leads to precise bounds
+expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
+intervals.
+
+% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW
 

### 2014-10-29 paper/main.tex
improvements

@@
 Second, while all imperative programming languages
-provide a function or procedure abstraction, all
-available tools are not able to abstract their resource
-behavior and have to inline the procedure body to perform
-any kind of analysis.  Third, when such a tool fails on
-given program, the only way out to get a resource bound
-is to rewrite the program.  There is no possibility
-of sound interaction with the tool, and manual proofs
+provide a function or procedure abstraction, available
+tools are not able to abstract their resource behavior
+and have to inline the procedure body to perform
+their analysis.  Third, when such a tool fails on
+program, the only way out to get a resource bound is to
+rewrite the program.  There is no opportunity for sound
+interaction with the tool, for example, manual proofs
 cannot be composed with automatically inferred bounds.
-One of the biggest challenge in the way of interaction
-with these automated tools is that they work on a highly
-abstract version of the program that has a lot more
-behaviors than the original program.
+One of the toughest challenges in the way of interaction
+is that available tools work on a highly abstract version
+of the source program. This abstracted program has a lot
+more behaviors than the original one and is usually not
+able to express memory related operations.
 

### 2014-10-29 paper/main.tex
improvements

@@
 terms of the input parameters of a procedure.
-XXX MENTION GULWANI PAPER [1] XXX
+XXX MENTION GULWANI AND KOAT XXX
 Second, while all imperative programming languages

### 2014-11-03 paper/main.tex
improvements

@@
 
-Available techniques split in two categories.  In the
-first one, tools derive impressive bounds for
-imperative programs, but are not fully compositional.
+Available techniques split in two categories.  Tools
+from the first one derive impressive bounds for
+imperative programs, but they are not fully compositional.
 Notable recent work like
@@
 Since this
-is often a local analysis, the loop bounds are expressed
-as arithmetic expressions in terms of variables as they
-are just before the loop.  This makes it hard to give a
+is often a local analysis, the loop bounds are
+arithmetic expressions that depend on the value of
+variables  just before the loop.
+This makes it hard to give a
 resource bound on sequenced loops or function calls in
 terms of the input parameters of a procedure.
-XXX MENTION GULWANI AND KOAT XXX
 Second, while all imperative programming languages
@@
 tools are not able to abstract their resource behavior
-and have to inline the procedure body to perform
-their analysis.  Third, when such a tool fails on
-program, the only way out to get a resource bound is to
-rewrite the program.  There is no opportunity for sound
-interaction with the tool, for example, manual proofs
-cannot be composed with automatically inferred bounds.
-One of the toughest challenges in the way of interaction
-is that available tools work on a highly abstract version
-of the source program. This abstracted program has a lot
-more behaviors than the original one and is usually not
-able to express memory related operations.
-
-In the second category, tools also derive very tight
+and have to inline the procedure body to perform their
+analysis.  Third, when a tool fails to find a resource
+bound for a program, the program has to be rewritten.
+There is no opportunity for sound interaction with the
+tool, for example, manual proofs cannot be composed
+with automatically inferred bounds.  One reason for this
+impossibility to interact is that available tools work
+on a highly abstract version of the source program. This
+abstracted program has a lot more behaviors than the
+original one and is usually expressed in a language that
+is not able to express memory operations.
+
+In the second category, tools derive very tight
 bounds on functional programs~\cite{Jost03,HoffmannAH12}.
-These tools are highly compositional, scale for large
+They are highly compositional, scale for large
 programs, and work directly on the syntax.  Despite all
@@
 in scope, so the potential of these variables has to be
-updated after mutations or unsound bounds can be derived.
+updated after mutations or, unsound bounds can be derived.
 
@@
 approach is an attempt to unify the two previous
-categories: it solves the compositionality issues of
+categories: it solves the compositionality issues of the
 imperative world by adapting amortized-analysis based
@@
 apporaches based on ranking functions, our tool inherits
-the benefits of amortized reasoning and can handle using
+the benefits of amortized reasoning.  It handles, using
 only one simple mechanism:
@@
 \item \emph{nested loops} that influence each other
-  with the same set of modified variables
+  with the same set of modified variables,
 \itemskipIn

### 2014-11-04 paper/main.tex
start working on section 2

@@
 tool, for example, manual proofs cannot be composed
-with automatically inferred bounds.  One reason for this
-impossibility to interact is that available tools work
-on a highly abstract version of the source program. This
+with automatically inferred bounds.  One reason for
+this impossibility to interact is that available tools
+work on an abstract version of the source program. This
 abstracted program has a lot more behaviors than the
 original one and is usually expressed in a language that
-is not able to express memory operations.
+does not feature memory operations.
 
@@
 apporaches based on ranking functions, our tool inherits
-the benefits of amortized reasoning.  It handles, using
-only one simple mechanism:
+the benefits of amortized reasoning.  Using only one
+simple mechanism, it handles:
 
@@
 \section{Overview and Examples}
+\aftersectskip
+
+In this section, we informally introduce the automatic amortized
+analysis (AAA) for Clight programs.  Some example bounds are derived
+step by step following our algorithm.
+
+\subsection{Bounding resource usage with potential functions}
+
+The idea that underlies the design of our framework is amortized
+analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes
+on a starting state $\state$ and consumes $n$ resource units of
+some user-defined quantity.  We denote that by writing $(S, \state)
+\Downarrow_n \state'$ where $\state'$ is the program state after
+the execution.  The basic idea of amortized analysis is to define a
+\emph{potential function} $\Phi$ that maps program states to non-negative
+numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a program
+state such that $(S, \state) \Downarrow_n \state'$.  This gives that
+$\Phi(\state)$ is a valid resource bound.
+
+To obtain a compositional reasoning we also have to take into account the
+state resulting from a program's execution.  We thus use two potential
+functions, one that applies before the execution, and one that applies
+after.  The two functions must respect the relation $\Phi(\state)
+\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
+that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
+must provide enough \emph{potential} for both, paying for the resource
+cost of the computation and paying for the potential $\Phi'(\state')$ on
+the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
+\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
+\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
+This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
+Note that the initial potential function $\Phi$ provides an upper bound
+on the resource consumption of the whole program.  What we have observed
+is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
+$$
+\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
+n + \Phi'(\state') \; ,
+$$
+then we get the following familiar looking rule.
+$$
+\RuleNolabel
+{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
+{\htriple{\Phi} {S_1; S_2}  {\Phi''}}.
+$$
+%
+Similarly, other language constructs lead to rules for the potential
+functions that look very similar to Hoare logic or effect system
+rules.  These rules enable interactive reasoning about resource usage
+in a flexible and compositional way, which, as a side effect, produces
+a certificate for the derived resource bound.
+
+The considerations above already show a departure from classical
+techniques for imperative programs.  Indeed, reasoning with two
+potential functions is more compositional than using
+ranking-functions because it forces to consider the sequencing of program
+operations.  In the previous rule, $\Phi$ gives a bound for
+$S_1; S_2$ through the intermediate potential $\Phi'$, even though
+it was derived on $S_1$ only.
+
+% Potential functions can access the whole program state, so this
+% framework can readily be used to derive bounds that depend on the
+% memory contents.
+
+\subsection{Automatic Amortized Analysis}
+
+The reasoning presented above is very general and flexible.  But in
+practice, to make the reasoning tractable on large code bases, we would
+like to infer automatically the potential functions.  The first step
+towards this goal is to reduce our search space by fixing the shape
+of potential functions $\Phi$.  We settle on a form that allows
+\emph{linear programming} (LP) to help inference:
+$$
+\Phi(\state) = q_0 + \sum_{x,y \in \dom(\state)}
+  q_{(x,y)} \cdot |[\state(x),\state(y)]|
+$$
+where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  We assumed
+that the program state is a simple map from program variables to integers
+and we treat constants as program variables (that cannot be changed
+by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
+the sum.  We then develop an inference rule for each syntactic construct
+that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
+the previous section.
+
+This idea is best explained by example.  If we use the tick metric that
+assigns cost $n$ to the function call \code{tick(n)} and cost $0$ to all
+other operations then the cost of the following example can be bounded
+by $|[x,y]|$.\footnote{Note that there are no restrictions on the signs
+of the input variables in the examples.}
+\begin{lstlisting}[basicstyle=\tt\small]
+while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
+\end{lstlisting}
+To derive this bound in our automatic amortized analysis, we start with
+the initial potential $\Phi_0 = |[x,y]|$ (short for $\Phi_0(\state)
+= |[\state(x),\state(y)]|$) which we also use as the loop invariant.
+For the loop body we have (like in Hoare logic) to derive a triple
+like $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do
+so if we utilize the fact that $x<y$ at the beginning of the loop body.
+The reasoning then works as follows. We start with the potential $|[x,y]|$
+and the fact that $|[x,y]| > 0$ before the assignment.  If we denote the
+updated version of $x$ after the assignment by $x'$ then the relation
+$|[x,y]| = |[x',y]| + 1$ between the potential before and after the
+assignment \code{x=x+1} holds.  This means that we have the potential
+$|[x,y]| + 1$ before the statement \code{tick(1)}.  Since \code{tick(1)}
+consumes one resource unit, we end up with potential $|[x,y]|$ after
+the loop body and have established the loop invariant again.
+
+\subsection{Compositionality in practice}
+
+The following example is typical in cryptographic primitives
+implementations.  These primitives work by block, so when data of
+arbitrary length needs to be processed, it is munched in blocks and
+the leftovers are stored in a buffer for future use when more data
+is available.  This looping pattern is particularly well handled by
+our tool because the LP solver is leveraged to generate a tight bound.
+The optimization problem it is given encodes that constant bounds are
+preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
+AAA is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
++ \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
+counted as constant.
+%
+\begin{lstlisting}
+for (; l >= 8; l -= 8)
+    // process one block
+    tick(N);
+for (; l > 0; l--)
+    // save leftovers
+    tick(1);
+\end{lstlisting}
+%
+It is worth taking some time to explain the constant contribution
+to the resource bound when $N<8$.  Remark first that the cost of
+the second loop is $|[0,l]|$, and after the first one, we still have
+$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
+to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
+we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
+loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
+potential units instead.  This explains the somewhat intriguing value
+of the constant term in the bound.  All this reasoning is done within
+our tool automatically and simply relies on general rules presented
+in further sections.
+
+This level of precision and compositionality is only achieved by
+our work, no other available tool will attain the above tight bounds
+on this common kind of looping code.  Thanks to the modularity of
+amortized analysis the intricate interactions between the bounds of
+the two loops can be modeled as one linear program, and because LP
+solvers can optimize the quality of the final bound we can keep
+it close to the actual resource consumption.
+
+\sectskip
+\section{Overview and Examples}
 \label{sec:inform}

### 2014-11-05 paper/main.tex
add the quicksort example

@@
 
+With two concrete examples from open-source projects we will try to
+show the reader that compositionality is indeed useful in the analysis
+of common code.
+
 The following example is typical in cryptographic primitives
@@
 it close to the actual resource consumption.
+
+A clean theoretical framework for compositional bounds sometimes
+even allow to get better asymptotic bounds.  During our experiments
+we found an example of this situation in some code from the cBench
+benchmark suite.
+%
+\begin{lstlisting}
+for (;;) {
+    do { l++; tick(1); } while (l < h && nondet());
+    do { h--; tick(1); } while (h > l && nondet());
+    if (h <= l) break;
+    tick(1); /* swap two elements */ }
+\end{lstlisting}
+%
+The above snippet is the inner loop of a Quicksort implementation, more
+precisely, it is the partitioning part of the algorithm.  This partition
+loop has linear complexity, and feeding into our analysis gives the
+correct and tight worst-case bound $2+3|[l,h]|$.  However, all available
+tools for C will derive a quadratic bound instead.  With the amortized
+analysis approach in mind, it is clear that the three loops get their
+potential from the same interval $[l,h]$.  But on the other hand, current
+techniques will try to find one linear ranking-function for each loop
+and combine them multiplicatively or additively, resulting in a loose
+quadratic bound.
 

### 2014-11-06 paper/main.tex
pass through intro

@@
 %----------------------------------------------------
+
+\newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}
 
@@
 
-Available techniques split in two categories.  Tools
-from the first one derive impressive bounds for
+Available techniques for deriving worst-case resource bounds
+split in two categories.  Techniques
+in the first category derive impressive bounds for
 imperative programs, but they are not fully compositional.
-Notable recent work like
+This is problematic if one needs to derive global whole-program bounds.
+In the second category, techniques derive tight whole-program
+bounds on functional programs.
+They are highly compositional, scale for large
+programs, and work directly on the syntax.  However, despite all
+these theoretical and practical advantages, it has been
+a long-time open problem to make them work in the
+presence of negative integers, mutation, and non-linear
+control flow. As a result, they are not suitable for
+typical imperative code.
+%
+\jan{I changed the order here a bit. Good style: First sentence is a
+  summary of the following paragraph.}
+
+Notable tools in the first category include
 SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
 PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
-and LOOPUS~\cite{SinnZV14} falls in this category.
+and LOOPUS~\cite{SinnZV14}.
 They lack compositionality in at least three ways.
@@
 terms of the input parameters of a procedure.
-Second, while all imperative programming languages
+Second, while all popular imperative programming languages
 provide a function or procedure abstraction, available
@@
 bound for a program, the program has to be rewritten.
-There is no opportunity for sound interaction with the
-tool, for example, manual proofs cannot be composed
-with automatically inferred bounds.  One reason for
-this impossibility to interact is that available tools
+There is no possibility for sound user interaction with the
+tool. For example, there is no concept of manual proofs for
+resource bounds and no framework that could support composition
+of manually derived bounds with automatically inferred bounds. 
+One reason for this lack of interaction is that available tools
 work on an abstract version of the source program. This
-abstracted program has a lot more behaviors than the
+abstracted program has more behaviors than the
 original one and is usually expressed in a language that
 does not feature memory operations. 
-
-In the second category, tools derive very tight
-bounds on functional programs~\cite{Jost03,HoffmannAH12}.
-They are highly compositional, scale for large
-programs, and work directly on the syntax.  Despite all
-these theoretical and practical advantages, it has been
-a long-time open problem how to make them work in the
-presence of negative integers, mutation and non-linear
-control flow.  For example, the first challenge of
-imperative programs is the absence of useful type system
-to hook the analysis on.  Another issue is that these
-tools associate a \emph{potential} directly to program
-variables, doing this in imperative programs would
-fail for loops as simple as \lstinline{for(i=x;i<y;i++)}
-where $y - i$ decreases, but not $|i|$.
-Mutation is yet another challenge, while functional
-programs relate multiple fixed inputs (the typing
-context) with one output (the expression typed),
-imperative programs can change the value of variables
-in scope, so the potential of these variables has to be
+%
+\jan{@Quentin: What do you mean by ``more behaviors''?}
+
+In the second category, are based on type systems for functional
+programs and the potential method of amortized
+analysis~\cite{Jost03,HoffmannAH12}.  The first challenge of
+imperative programs is already the absence of a fine-grained type system to hook
+the analysis on.  Another issue is that the techniques associate
+\emph{potential} directly to program variables.  In
+imperative programs, this would fail for loops as simple as
+\lstinline{for(i=x;i<y;i++)} where $y - i$ decreases, but not $|i|$.
+Mutation is yet another challenge.  While functional programs relate
+multiple fixed inputs (the typing context) with one output (the
+expression typed), imperative programs can change the value of
+variables in scope, so the potential of these variables has to be
 updated after mutations or, unsound bounds can be derived.
+%
+\jan{@Quentin: There is a mismatch between the previous two paragraphs.
+  In the first one we talk about things that category one techniques
+  can't do. In the second one we talk about ``challenges'' for the
+  category two. We instead should just say what the state-of-the-art
+  can't do. We also have to mention Hofmann/Jost ESOP'06 and Atkey ESOP'10.}
 
@@
 Our automated analysis is able to infer linear
-resource bounds on C programs with integer loops.
-To our knowledge this is the first tool based on
-amortized analysis able to deal with negative numbers
+resource bounds on C programs with mutually-recursive functions and integer loops.
+To our knowledge this is the first technique based on
+amortized analysis that is able to derive bounds that depend on negative numbers
 and differences of variables.  It is also the first
-resource analysis tool for C that deals naturally with
-recursive functions, resources that can be freed (e.g.
+resource analysis technique for C that deals naturally with
+recursive functions, resources that can become available during execution (e.g., when freeing
 memory), and sequenced loops.  Compared to more classical
@@
 
+\jan{Here we need a description of the implementation and the
+  experiments.  Message: You can immediately see the great benefits of
+  the unification of the two categories in practice. It scales to
+  larger programs than existing tools (for global bounds) and is as
+  good on local bounds (for tricky loop and recursion patterns).}
+
 % NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW
@@
 Give a description of the sections here.                    XXX
+
+\jan{We also mention before that existing imperative tools don't allow
+  for user interaction. We should pick this up again and say that we
+  can do it. But then we have to mention the PLDI paper. My suggestion:
+  \begin{itemize}
+  \item mention PLDI'14 in the paragraph about the second category and
+    say it has great flexibility for manual bound derivation but
+    almost no support for automation
+  \item just add another bullet to say that we have the same
+    flexibility for manual bound derivation as in PLDI'14 and that
+    manual and automatic bounds and be composed
+  \end{itemize}
+}
 

### 2014-11-06 paper/main.tex
pass through section 2

@@
 \end{itemize}
+\jan{We can say that our technique is the only one that reduces the problem to
+  LP solving. I used that as a selling point in the DARPA-STAC proposal.}
 
@@
     flexibility for manual bound derivation as in PLDI'14 and that
-    manual and automatic bounds and be composed
+    manual and automatic bounds can be composed
   \end{itemize}
@@
 In this section, we informally introduce the automatic amortized
-analysis (AAA) for Clight programs.  Some example bounds are derived
-step by step following our algorithm.
-
-\subsection{Bounding resource usage with potential functions}
+analysis (AAA) for Clight programs and demonstrate the technique with
+example bounds that we derive step-by-step following our analysis
+system.
+
+\jan{Watch out with the headings. Either upper case or lower case, but
+  don't mix.}
+
+\subsection{Bounding Resource Usage with Potential Functions}
 
@@
 practice, to make the reasoning tractable on large code bases, we would
-like to infer automatically the potential functions.  The first step
+like to automatically infer the potential functions.  The first step
 towards this goal is to reduce our search space by fixing the shape
@@
 \emph{linear programming} (LP) to help inference:
+%
+\jan{I wouldn't say that LP is the reason we picked this format. We
+  rather picked it because it covers the most common cases.}
+%
 $$
@@
 $$
-where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  We assumed
+where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Here, we assume
 that the program state is a simple map from program variables to integers
@@
 
-\subsection{Compositionality in practice}
-
-With two concrete examples from open-source projects we will try to
-show the reader that compositionality is indeed useful in the analysis
-of common code.
+\jan{I would add another subsection here to explain how we automate
+  the analysis with LP solving.  Message: Efficient reduction to LP
+  solving and LP solvers are great and super fast. Maybe mention
+  'network problems'.}
+
+\subsection{Compositionality in Practice}
+
+With two concrete examples from open-source projects we demonstrate
+that the compositionality of our method is indeed needed in the
+analysis of common code.
+
+\jan{I don't like to talk about ``the reader''. Often you can leave it
+  out. It is fine to address the reader with ``you''. See BUGS in
+  Writing.}
 
@@
 %
+\jan{N is unbound above. The explanation must be much slower
+  here. Explain what the code does and how the ticks account for the
+  cost. Then: What is the resource bound?. Then: What do we derive?
+  Btw: Why do you use tick(1) and not tick(M)?}
+%
 \begin{lstlisting}
@@
 quadratic bound.
+%
+\jan{Did you test the last example with other tools? What are the
+  bounds they get? State the bounds here.}
 

### 2014-11-07 paper/main.tex
a comment for the structure
update to structure comment
minor corrections

@@
 \documentclass[nocopyrightspace,preprint]{sigplanconf}
+
+% Structure:
+%
+% 1) Intro
+% 2) Overview and Examples
+% 3) Cost Semantics for Clight
+% 4) Compositional Resource-Bound Analysis
+% 5) Automatic Inference via LP Solving
+% 6) Auxiliary Variables and User Interaction
+% 7) Quantitative Hoare Logic and Soundness Proof
+% 8) Experimental Evaluation
+% 9) Related Work
+% 10) Conclusion
 
@@
 
-In the second category, are based on type systems for functional
-programs and the potential method of amortized
+Tools in the second category are based on type systems for functional
+programs and on the potential method of amortized
 analysis~\cite{Jost03,HoffmannAH12}.  The first challenge of

### 2014-11-07 paper/main.tex
new abstract

@@
 \begin{abstract}
-  The goal of quantitative resource analysis is to provide developers
-  with quantitative information about the runtime behavior of software
-  at development time.  Recent years have seen tremendous progress in
-  automatically deriving worst-case resource bounds, yet many
-  challenges in specifying, interactively deriving, formally
-  certifying, and composing resource bounds remain unsolved.
-
-  This paper describes a novel quantitative resource analysis
-  framework that tackles these challenges for C programs.  The
-  analysis framework consists of three parts: First, a parametric cost
-  semantics formalizes the resource consumption of CompCert Clight
-  programs.  Second, a quantitative Hoare logic enables users to
-  interactively develop resource bounds in the Coq Proof Assistant.
-  The logic is proved sound with respect to the cost semantics, and
-  shallow embedding enables a derived bound to be any function that is
-  definable in Coq.  Third, an automatic amortized resource analysis
-  for Clight computes derivations in the quantitative Hoare logic.
-  It is the first automatic amortized analysis that can derive bounds
-  that depend on negative integers or differences between numbers,
-  which is crucial to handle typical systems code.  Both, the
-  quantitative logic and the automatic amortized analysis are
-  naturally compositional and can be combined to semi-automatically
-  derive global resource bounds.
-  %
-  An experimental evaluation demonstrates the practicality of the
-  analysis framework.  The expressivity of the logic is shown by
-  manually deriving customized bounds that are tailored to specific
-  algorithms.  A comparison of the automatic amortized analysis with
-  other automatic tools on 30 challenging loop and recursion patterns
-  from the literature and open-source software shows that the bounds
-  derived by the automatic amortized analysis are often more precise.
+This paper presents a new approach to quantitative
+analysis for C programs.  The technique described uses
+ideas from amortized analysis and abstract interpretation
+to derive tight linear worst-case bounds that come with a
+proof certificate.  Contrary to most available tools for
+C, ours has a compositional design that provides the
+programmer with bounds expressed in terms of meaningful
+variables.  The compositionality allows to handle uniformly
+nested sequenced loops and recursive sequenced function
+calls.  The automatic analysis is implemented
+as a tool that leverages state-of-the-art SMT solvers
+LP solvers to automatically infer resource bounds on
+a wide range of C programs.  It is the first automatic
+amortized analysis that can derive bounds that depend
+on negative integers and differences between numbers.
+To prove its soundness and give a concrete shape to proof
+certificates, we developed a quantitative Hoare logic
+that we proved sound using the Coq proof assistant.  This
+logic allows manual proofs of arbitrarily complex resource
+bounds on C programs and interaction with the automation.
+To prove the scalability and precision of this new tool,
+we run it on multiple code snippets from the literature,
+reproduce results from previous work, and analyze a
+consequent part of a modern benchmark suite.
+
+%   The goal of quantitative resource analysis is to provide developers
+%   with quantitative information about the runtime behavior of software
+%   at development time.  Recent years have seen tremendous progress in
+%   automatically deriving worst-case resource bounds, yet many
+%   challenges in specifying, interactively deriving, formally
+%   certifying, and composing resource bounds remain unsolved.
+%
+%   This paper describes a novel quantitative resource analysis
+%   framework that tackles these challenges for C programs.  The
+%   analysis framework consists of three parts: First, a parametric cost
+%   semantics formalizes the resource consumption of CompCert Clight
+%   programs.  Second, a quantitative Hoare logic enables users to
+%   interactively develop resource bounds in the Coq Proof Assistant.
+%   The logic is proved sound with respect to the cost semantics, and
+%   shallow embedding enables a derived bound to be any function that is
+%   definable in Coq.  Third, an automatic amortized resource analysis
+%   for Clight computes derivations in the quantitative Hoare logic.
+%   It is the first automatic amortized analysis that can derive bounds
+%   that depend on negative integers or differences between numbers,
+%   which is crucial to handle typical systems code.  Both, the
+%   quantitative logic and the automatic amortized analysis are
+%   naturally compositional and can be combined to semi-automatically
+%   derive global resource bounds.
+%   %
+%   An experimental evaluation demonstrates the practicality of the
+%   analysis framework.  The expressivity of the logic is shown by
+%   manually deriving customized bounds that are tailored to specific
+%   algorithms.  A comparison of the automatic amortized analysis with
+%   other automatic tools on 30 challenging loop and recursion patterns
+%   from the literature and open-source software shows that the bounds
+%   derived by the automatic amortized analysis are often more precise.
 \end{abstract}

### 2014-11-07 paper/main.tex
experiments, they look ridiculous

@@
 
+\begin{table}[t]
+\centering
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+\end{tabular}
+\vspace{.1cm}
+\caption{Evaluation of our automatic amortized analysis on larger
+C code from the cBench benchmark suite. The function analyzed might
+be calling helper functions which do not appear in the list but have
+to be analyzed.  These functions are accounted for in the LoC column
+where not only the lines count of the analyzed function is called
+but also the line count of all the funtions it calls. The line count
+is computed on pre-processed C with all unnecessary declarations
+removed.}
+\label{tab:compar}
+\end{table}
+
 \sectskip

### 2014-11-08 paper/main.tex
updated the intro

@@
 Available techniques for deriving worst-case resource bounds
-split in two categories.  Techniques
+fall into two categories.  Techniques
 in the first category derive impressive bounds for
-imperative programs, but they are not fully compositional.
+numerical imperative programs, but they are not fully compositional.
 This is problematic if one needs to derive global whole-program bounds.
 In the second category, techniques derive tight whole-program
-bounds on functional programs.
+bounds on functional programs that operate on inductive data structures.
 They are highly compositional, scale for large
@@
 a long-time open problem to make them work in the
-presence of negative integers, mutation, and non-linear
+presence of signed integers, mutation, and non-linear
 control flow. As a result, they are not suitable for
 typical imperative code.
-%
-\jan{I changed the order here a bit. Good style: First sentence is a
-  summary of the following paragraph.}
 
@@
 First, they all base their analysis on some form of
-\emph{ranking-function}.
+\emph{ranking function}.
 Since this
-is often a local analysis, the loop bounds are
+is usually a local analysis and loop bounds are
 arithmetic expressions that depend on the value of
@@
 This makes it hard to give a
-resource bound on sequenced loops or function calls in
+resource bound on a sequence of loops and procedure calls in
 terms of the input parameters of a procedure.
@@
 work on an abstract version of the source program. This
-abstracted program has more behaviors than the
-original one and is usually expressed in a language that
-does not feature memory operations.
-%
-\jan{@Quentin: What do you mean by ``more behaviors''?}
-
-Tools in the second category are based on type systems for functional
-programs and on the potential method of amortized
-analysis~\cite{Jost03,HoffmannAH12}.  The first challenge of
-imperative programs is already the absence of a fine-grained type system to hook
-the analysis on.  Another issue is that the techniques associate
-\emph{potential} directly to program variables.  In
-imperative programs, this would fail for loops as simple as
-\lstinline{for(i=x;i<y;i++)} where $y - i$ decreases, but not $|i|$.
-Mutation is yet another challenge.  While functional programs relate
-multiple fixed inputs (the typing context) with one output (the
-expression typed), imperative programs can change the value of
-variables in scope, so the potential of these variables has to be
-updated after mutations or, unsound bounds can be derived.
-%
-\jan{@Quentin: There is a mismatch between the previous two paragraphs.
-  In the first one we talk about things that category one techniques
-  can't do. In the second one we talk about ``challenges'' for the
-  category two. We instead should just say what the state-of-the-art
-  can't do. We also have to mention Hofmann/Jost ESOP'06 and Atkey ESOP'10.}
+abstracted program has sometimes more behaviors than the
+original one (introduced by non-determinism) and is usually
+expressed in a language that does not feature memory operations.
+
+Tools in the second category originate form the potential method of
+amortized analysis and type systems for functional
+programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
+definitions of object oriented programs~\cite{Jost06} and
+data-structure predicates of separation logic~\cite{Atkey10} can play
+the role of the type system in imperative programs. However, these
+techniques fail for numerical imperative programs because they lack
+fine-grained data-structure information to hook the analysis on.
+Another issue is that the techniques associate \emph{potential}
+directly with single program variables.  In imperative programs, this
+fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
+$y - i$ decreases, but not $|i|$.  Mutation is yet another problem.
+While functional programs relate multiple fixed inputs (the typing
+context) with one output (the expression typed), imperative programs
+can change the value of variables in scope, so the potential of these
+variables has to be updated after mutations to ensure soundness.
 
@@
 approach is an attempt to unify the two previous
-categories: it solves the compositionality issues of the
-imperative world by adapting amortized-analysis based
+categories: It solves the compositionality issues of techniques
+for numerical imperative code by adapting amortized-analysis based
 techniques from the functional world.
-
+%
 Our automated analysis is able to infer linear
 resource bounds on C programs with mutually-recursive functions and integer loops.
+%
+The resource behavior of functions can be summarized
+in a function specification that can be used a every
+call-side without access to the function body.
+%
 To our knowledge this is the first technique based on
@@
 \end{itemize}
-\jan{We can say that our technique is the only one that reduces the problem to
-  LP solving. I used that as a selling point in the DARPA-STAC proposal.}
-
-The main innovations that make amortized analysis work
+
+The main innovation that makes amortized analysis work
 on imperative languages is to base the analysis on a
@@
 expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
-intervals.
+intervals. A distinctive feature of our analysis system is that
+it reduces bound derivation to a linear optimization problem
+that can solved by off-the-shelf LP solvers.  This enables the
+efficient inference of global bounds for larger programs.
+
+The use of the potential method of amortized analysis makes user
+interaction possible in different ways. For one thing, we can directly
+combine the new automatic analysis with manually derived bounds in a
+previously-developed quantitative Hoare logic~\cite{veristack14} (see
+Section XXX).  For another thing, we describe a new mechanism that
+allows the separation of quantitative and qualitative verification
+(see Section XXX).  Using this mechanism, the user can guide the
+analysis by using auxiliary variables and logical assertions that can
+be verified by existing qualitative tools such as Hoare logic or
+abstract interpretation.  In this way, we can benefit from existing
+automation techniques and provide a middle-ground between fully
+automatic and fully manual verification to derive. This enables the
+semi-automatic inference of non-linear bounds, such as polynomial,
+logarithmic, and exponential bounds.
 
@@
 
-% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW
-
-Give a description of the sections here.                    XXX
-
-\jan{We also mention before that existing imperative tools don't allow
-  for user interaction. We should pick this up again and say that we
-  can do it. But then we have to mention the PLDI paper. My suggestion:
+% Give a description of the sections here.                    XXX
+% Jan: We don't need that. Usually, people don't read it.
+
+In summary, we make the following \emph{contributions}.
+
 \begin{itemize}
-  \item mention PLDI'14 in the paragraph about the second category and
-    say it has great flexibility for manual bound derivation but
-    almost no support for automation
-  \item just add another bullet to say that we have the same
-    flexibility for manual bound derivation as in PLDI'14 and that
-    manual and automatic bounds can be composed
+\item We develop the first automatic amortized analysis for C
+  programs. It is naturally compositional, tracks size changes of
+  variables to derive global bounds, can handle mutually-recursive
+  functions, generates resource specifications for functions, derives
+  easily verifiable certificates, and handles resources that can
+  become available during an execution.
+\item We show how to reduce the inference of \emph{linear resource
+    bounds} to efficient LP solving.
+\item We describe a new method of harnessing existing qualitative
+  verification techniques to guide the automatic amortized analysis
+  to derive \emph{non-linear} resource bounds.
+\item We prove the soundness of the analysis with respect to a
+  parametric cost semantics for C programs. The cost model can be
+  further customized with function calls (\code{tick(n)}) that
+  indicate resource usage. This mechanism is also covered by the
+  soundness proof.
+\item We present experiments with a prototype implementation on more
+  than XXX lines of C code. A detailed comparison shows that our
+  prototype is the only tool that can derive global bounds for larger
+  C programs while being as powerful as existing tools when deriving
+  linear local bounds for tricky loop and recursion patterns.
 \end{itemize}
-}
 

### 2014-11-08 paper/main.tex
add logical state examples

@@
 }{}
+\begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}
+${\color{blue}%
+  \mbox{logical state invariant } \{ \code{na} = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1 && na > 0) {
+    a[x]=0;
+    na--; tick(1);
+    x++; }
+  if (x < N) {
+    a[x]=1;
+    na++; tick(1); }
+  k--;
+}
+
+\end{lstlisting}
+\end{minipage}
+
+\caption{Example of assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
+  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
+  0 to all other operations.  The bound derived automatically is
+  $2|[0,\code{k}]| + |[0,\code{na}]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+%
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ \code{lg} \ge \log_2(h-l) \cdot M_\code{bsearch} \} }$
+bsearch(x,l,h,lg) {
+  if (h-l > 1 && lg > 0) {
+    m = h + (h-l)/2;
+    lg--; if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h,lg);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We use the tick metric to model the stack consumption of the program.
+  Before a function call to \code{bsearch}, a tick statement pays for
+  the space used by stack frame of the function.  The space is recoverd
+  after the call.
+  The bound derived automatically is $|[0,\code{lg}]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
+%
+\iffull{
+}{}
 We can then use the logic to prove a concrete bound for a given resource

### 2014-11-08 paper/main.tex
start working on section 6

@@
 \begin{document}
+
+\lstset{moredelim=[is][\underbar]{_}{_},keepspaces=true}
 
@@
 \sectskip
+\section{Logical State and User Interaction}
+\label{sec:anno}
+\aftersectskip
+
+\begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ \code{na} = \#_1(\code{a})} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1 _&& na > 0_) {
+    a[x]=0; _na--;_
+    tick(1);
+    x++; }
+  if (x < N) {
+    a[x]=1; _na++;_
+    tick(1); }
+  k--;
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Example of assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
+  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
+  0 to all other operations.  The bound derived automatically is
+  $2|[0,\code{k}]| + |[0,\code{na}]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ \code{lg} \ge \log_2(h-l) \cdot M_\code{bsearch} \} }$
+bsearch(x,l,h_,lg_) {
+  if (h-l > 1 && lg > 0) {
+    m = h + (h-l)/2;
+    _lg--;_ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h_,lg_);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We use the tick metric to model the stack consumption of the program.
+  Before a function call to \code{bsearch}, a tick statement pays for
+  the space used by stack frame of the function.  The space is recoverd
+  after the call.
+  The bound derived automatically is $|[0,\code{lg}]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
+
+While complete automation is often desirable, it cannot possibly
+be universal since the problem is undecidable.  In this section
+we present two techniques to derive complex resource bounds by
+leveraging our automation.
+
+When the resource bound of a program depends on the contents of the heap,
+or is non-linear (e.g. logarithmic, exponential), we propose to
+introduce \emph{logical state}.
+This technique is inspired by XXX (speed) and consists in introducing
+extra variables to the program that do not change its behavior, but help
+the automatic tool to figure out a bound.
+
+The systematic technique consists in the following steps. First,
+a program $P$ that fails to be analyzed automatically is enriched
+by extra variables $\vec x$ called logical variables to form a
+program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
+for the logical variables is shown to be such that:
+$$
+\forall n\, \sigma\, \sigma'.\,
+  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
+    {\implies}
+  \exists n', n' \le n \land
+  (\sigma, P) \Downarrow_{n'} \sigma'.
+$$
+Since the annotated program and the original one are often
+syntactically close, the proof of this result goes
+by simple induction on the resource-aware evaluation judgement.
+Third, using existing automation tools, a bound $B(\vec x)$ for
+$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
+$\vec X$, gives the final resource bound for the program $P$.
+
+This idea is illustrated by the program on \pref{fig:xmplincaux}.
+Parts of the code underlined are annotations that were added
+to the original program text.  The top-level loop increments
+a binary counter \code{k} times.  A naive analysis of
+the algorithm yields the quadratic bound $\code{k}\cdot\code{N}$.
+However, the algorithm is in fact linear and its cost is bounded
+by $2\code{k} + \#_1(\code{a})$ where $\#_1(\code{a})$ denotes
+the number of one entries in the array \code{a}.  Since this
+number depends on the heap contents, no tool available
+for C is able to derive the linear bound.  However, it can be inferred
+by our automated tool if a logical variable \code{na} is introduced.
+This logical variable is a reification, in the program, of the
+number $\#_1(\code{a})$.  For example, on line 5 of the example
+we are setting \code{a[x]} to 0 and because of the condition we
+know that this array entry was 1.  To reflect this change on
+$\#_1(\code{a})$, the logical variable \code{na} is decremented.
+Similarly, on line 9, an array entry which was 0 becomes 1, so
+\code{na} is incremented.  To complete the step 2 of the systematic
+procedure described above, we must show that the extra test
+\code{na > 0} at line 4 does not influence the loop behavior.  It
+is done by proving inductively that $\code{na} = \#_1(\code{a})$
+and remarking that since $\code{a[x]} = 1$ is true, we must have
+$\#_1(\code{a}) > 0$, thus \code{na > 0} in the test. %  So the
+% test is in fact semantically equivalent to \code{true}.
+
+Another simple example is given on \pref{fig:xmplbsaux} where
+a logarithmic bound on the stack consumption of a binary search
+program is proved using logical variable annotations.  Once again,
+annotations are underlined in the program text.  In this example,
+to ease the proof of equivalence between the annotated program
+and the original one, we use the inequality $\code{lg} \ge
+\log_2(h-l)\cdot M_\code{bsearch}$ as invariant.  This trick
+allows a simpler proof because, when working with integer
+arithmetic, it is not always the case that $\log_2(x/2) =
+\log_2(x)-1$.
+
+\sectskip
 \section{Overview and Examples}
@@
 } ${\color{blue} \{ \#_1(a) \}}$
-
 \end{lstlisting}
@@
 \label{fig:xmplinc}
-\end{figure}
-%
-\iffull{
-}{}
-\begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}
-${\color{blue}%
-  \mbox{logical state invariant } \{ \code{na} = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1 && na > 0) {
-    a[x]=0;
-    na--; tick(1);
-    x++; }
-  if (x < N) {
-    a[x]=1;
-    na++; tick(1); }
-  k--;
-}
-
-\end{lstlisting}
-\end{minipage}
-
-\caption{Example of assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
-  0 to all other operations.  The bound derived automatically is
-  $2|[0,\code{k}]| + |[0,\code{na}]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-%
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ \code{lg} \ge \log_2(h-l) \cdot M_\code{bsearch} \} }$
-bsearch(x,l,h,lg) {
-  if (h-l > 1 && lg > 0) {
-    m = h + (h-l)/2;
-    lg--; if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h,lg);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We use the tick metric to model the stack consumption of the program.
-  Before a function call to \code{bsearch}, a tick statement pays for
-  the space used by stack frame of the function.  The space is recoverd
-  after the call.
-  The bound derived automatically is $|[0,\code{lg}]|$.
-  }
-\label{fig:xmplbsaux}
 \end{figure}

### 2014-11-09 paper/main.tex
add paragraph about interaction w/ logic

@@
   \mbox{logical state invariant }
-  \{ \code{lg} \ge \log_2(h-l) \cdot M_\code{bsearch} \} }$
+  \{ \code{lg} > \log_2(\code h-\code l) \} }$
 bsearch(x,l,h_,lg_) {
-  if (h-l > 1 && lg > 0) {
-    m = h + (h-l)/2;
+  if (h-l > 1 _&& lg > 0_) {
+    m = l + (h-l)/2;
     _lg--;_ if (a[m]>x) h=m; else l=m;
@@
   the space used by stack frame of the function.  The space is recoverd
-  after the call.
+  after the call.  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
   The bound derived automatically is $|[0,\code{lg}]|$.
@@
 to ease the proof of equivalence between the annotated program
-and the original one, we use the inequality $\code{lg} \ge
-\log_2(h-l)\cdot M_\code{bsearch}$ as invariant.  This trick
-allows a simpler proof because, when working with integer
-arithmetic, it is not always the case that $\log_2(x/2) =
-\log_2(x)-1$.
+and the original one, we use the inequality $\code{lg} >
+\log_2(\code h-\code l)$ as invariant.  This trick allows a simpler
+proof because, when working with integer arithmetic, it is not always
+the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
+
+On the other hand, it might be the case that even with annotations, a
+resource bound cannot be found automatically.  With existing tools, this
+situation requires a full rewrite of the program.  However, in our
+framework, it is possible to write a manual proof and combine it with
+bounds derived by the automatic analysis.  This interaction is made
+formal using our Quantitative Hoare Logic.  This resource aware program
+logic is an extension of the one presented in~\cite{veristack14} and
+was originally designed to prove the soundness of our automatic
+analysis.  To make it suitable for this end, we made it independent
+of the resource used (e.g. stack usage, heap usage, clock ticks) and
+implemented a smoother handling of the auxiliary state through a shallow
+embedding in Coq.  Despite being first a theoritical tool, the logic can
+be used directly as in~\cite{veristack14}.  Its judgements are inspired
+by typical Hoare logics and have the shape
+$$
+\Delta \vdash \htriple P S Q
+$$
+where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
+are quantitative assertions, i.e. functions from the program states
+to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
+of the logic are described in further details in the extended version of
+this paper~\cite{anon_extended}.  Because our automated tool produces
+proofs of the resource bounds in the Quantitative Hoare Logic, it can
+be used as an assistant when deriving a complex bound by hand in the
+logic.  The interaction is guaranteed to be sound by the soundness
+theorem of our Quantitative Hoare Logic.
 

### 2014-11-09 paper/main.tex
try to uniformize notations

@@
 loop has linear complexity, and feeding into our analysis gives the
-correct and tight worst-case bound $2+3|[l,h]|$.  However, all available
-tools for C will derive a quadratic bound instead.  With the amortized
+correct and tight worst-case bound $2+3|[l,h]|$.  With the amortized
 analysis approach in mind, it is clear that the three loops get their
-potential from the same interval $[l,h]$.  But on the other hand, current
-techniques will try to find one linear ranking-function for each loop
-and combine them multiplicatively or additively, resulting in a loose
-quadratic bound.
-%
-\jan{Did you test the last example with other tools? What are the
-  bounds they get? State the bounds here.}
+potential from the same interval $[l,h]$.  However, KoAT fails to
+prove the termination of the loop and LOOPUS derives the quadratic bound
+$(h-l-1)^2$.  Following the classic technique, these tools try to find one
+linear ranking-function for each loop and combine them multiplicatively
+or additively, resulting in a loose quadratic bound or no bound at all.
 
@@
 ${\color{blue}%
-  \mbox{logical state invariant } \{ \code{na} = \#_1(\code{a})} \}$
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
 while (k > 0) {
@@
   0 to all other operations.  The bound derived automatically is
-  $2|[0,\code{k}]| + |[0,\code{na}]|$.
+  $2|[0,k]| + |[0,na]|$.
   }
@@
   \mbox{logical state invariant }
-  \{ \code{lg} > \log_2(\code h-\code l) \} }$
+  \{ lg > \log_2(h - l) \} }$
 bsearch(x,l,h_,lg_) {
@@
   logarithm of $x$ in base 2.
-  The bound derived automatically is $|[0,\code{lg}]|$.
+  The bound derived automatically is $|[0,lg]|$.
   }
@@
 to the original program text.  The top-level loop increments
-a binary counter \code{k} times.  A naive analysis of
-the algorithm yields the quadratic bound $\code{k}\cdot\code{N}$.
+a binary counter $k$ times.  A naive analysis of
+the algorithm yields the quadratic bound $k \cdot N$.
 However, the algorithm is in fact linear and its cost is bounded
-by $2\code{k} + \#_1(\code{a})$ where $\#_1(\code{a})$ denotes
-the number of one entries in the array \code{a}.  Since this
+by $2k + \#_1(a)$ where $\#_1(a)$ denotes
+the number of one entries in the array $a$.  Since this
 number depends on the heap contents, no tool available
 for C is able to derive the linear bound.  However, it can be inferred
-by our automated tool if a logical variable \code{na} is introduced.
+by our automated tool if a logical variable $na$ is introduced.
 This logical variable is a reification, in the program, of the
-number $\#_1(\code{a})$.  For example, on line 5 of the example
-we are setting \code{a[x]} to 0 and because of the condition we
+number $\#_1(a)$.  For example, on line 5 of the example
+we are setting \lstinline{a[x]} to 0 and because of the condition we
 know that this array entry was 1.  To reflect this change on
-$\#_1(\code{a})$, the logical variable \code{na} is decremented.
+$\#_1(a)$, the logical variable $na$ is decremented.
 Similarly, on line 9, an array entry which was 0 becomes 1, so
-\code{na} is incremented.  To complete the step 2 of the systematic
+$na$ is incremented.  To complete the step 2 of the systematic
 procedure described above, we must show that the extra test
-\code{na > 0} at line 4 does not influence the loop behavior.  It
-is done by proving inductively that $\code{na} = \#_1(\code{a})$
-and remarking that since $\code{a[x]} = 1$ is true, we must have
-$\#_1(\code{a}) > 0$, thus \code{na > 0} in the test. %  So the
-% test is in fact semantically equivalent to \code{true}.
+\lstinline{na > 0} at line 4 does not influence the loop behavior.  It
+is done by proving inductively that $na = \#_1(a)$
+and remarking that since \lstinline{a[x] == 1} is true, we must have
+$\#_1(a) > 0$, thus \lstinline{na > 0} is true in the test. %  So the
+% test is in fact semantically equivalent to \lstinline{true}.
 
@@
 to ease the proof of equivalence between the annotated program
-and the original one, we use the inequality $\code{lg} >
-\log_2(\code h-\code l)$ as invariant.  This trick allows a simpler
+and the original one, we use the inequality $lg >
+\log_2(h-l)$ as invariant.  This trick allows a simpler
 proof because, when working with integer arithmetic, it is not always

### 2014-11-10 paper/main.tex
another pass through the intro

@@
 protocols~\cite{BartheGB09}.  In general, quantitative
-performance information at design time can provide useful
+resource information at design time can provide useful
 feedback for developers.
@@
 This is problematic if one needs to derive global whole-program bounds.
-In the second category, techniques derive tight whole-program
+Techniques in the second category derive tight whole-program
 bounds on functional programs that operate on inductive data structures.
 They are highly compositional, scale for large
-programs, and work directly on the syntax.  However, despite all
-these theoretical and practical advantages, it has been
+programs, and work directly on the syntax.  However, % despite all
+% these theoretical and practical advantages, 
+it has been
 a long-time open problem to make them work in the
@@
 is usually a local analysis and loop bounds are
-arithmetic expressions that depend on the value of
+arithmetic expressions that depend on the values of
 variables just before the loop.
@@
 provide a function or procedure abstraction, available
-tools are not able to abstract their resource behavior
+tools are not able to abstract resource behavior
 and have to inline the procedure body to perform their
 analysis.  Third, when a tool fails to find a resource
-bound for a program, the program has to be rewritten.
-There is no possibility for sound user interaction with the
-tool. For example, there is no concept of manual proofs for
+bound for a program, there is no possibility for sound
+user interaction with the tool other than to the rewrite
+the program. For example, there is no concept of manual proofs of
 resource bounds and no framework that could support composition
@@
 the role of the type system in imperative programs. However, these
-techniques fail for numerical imperative programs because they lack
+techniques fail for numerical imperative programs because these programs lack
 fine-grained data-structure information to hook the analysis on.
-Another issue is that the techniques associate \emph{potential}
+Another issue is that existing techniques associate \emph{potential}
 directly with single program variables.  In imperative programs, this
@@
 The resource behavior of functions can be summarized
-in a function specification that can be used a every
-call-side without access to the function body.
+in a function specification that can be used at every
+call-side without accessing the function body.
 %
@@
 resource analysis technique for C that deals naturally with
-recursive functions, resources that can become available during execution (e.g., when freeing
-memory), and sequenced loops.  Compared to more classical
+recursive functions and sequenced loops, and can handle resources that can
+become available during execution (e.g., when freeing
+memory).  Compared to more classical
 apporaches based on ranking functions, our tool inherits
@@
 on imperative languages is to base the analysis on a
-Hoare-like logic and track multivariate quantities instead
-of program variables directly.  This leads to precise bounds
+Hoare-like logic and to track multivariate quantities instead
+of program variables.  This leads to precise bounds
 expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
-intervals. A distinctive feature of our analysis system is that
+intervals. 
+%
+A distinctive feature of our analysis system is that
 it reduces bound derivation to a linear optimization problem
@@
 automation techniques and provide a middle-ground between fully
-automatic and fully manual verification to derive. This enables the
-semi-automatic inference of non-linear bounds, such as polynomial,
-logarithmic, and exponential bounds.
-
-\jan{Here we need a description of the implementation and the
-  experiments.  Message: You can immediately see the great benefits of
-  the unification of the two categories in practice. It scales to
-  larger programs than existing tools (for global bounds) and is as
-  good on local bounds (for tricky loop and recursion patterns).}
+automatic and fully manual verification for bound derivation. This
+enables the semi-automatic inference of non-linear bounds, such as
+polynomial, logarithmic, and exponential bounds.
+
+We have evaluated the automatic analysis with system code and examples
+from the literature.  Our tool automatically derived global resource
+bounds for more than XXX lines of C code from the XXX benchmark suite.
+%
+\iffull{\pref{app:cat}}{The extended version
+  of this article~\cite{anon_extended}}{} contains more than $30$
+challenging loop and recursion patterns that we collected from open
+source software and the literature.  Our analysis can find
+asymptotically tight bounds for all but 1 of these patterns, and in
+most cases the derived constant factors are tight.  To compare our
+automatic analyzer with existing techniques, we tested our examples
+with tools such as KoAT~\cite{BrockschmidtEFFG14},
+Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
+show that the bounds that we derive are often more precise than those
+derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
+uses amortization techniques, is able to achieve a similar precision.
+%
+Several micro benchmarks demonstrate the practicality and
+expressiveness of the user guided bound inference.  For example, we
+derive a logarithmic bound for a binary search function, a bound that
+depends on the contents of an array to describe the exact cost of a
+function that finds a maximal element in an array, and a linear
+bound that amortizes the cost of $k$ successive increments to a binary
+counter.
 
@@
   become available during an execution.
-\item We show how to reduce the inference of \emph{linear resource
-    bounds} to efficient LP solving.
+\item We show how to reduce the inference of \emph{linear} resource
+    bounds to efficient LP solving.
 \item We describe a new method of harnessing existing qualitative
   verification techniques to guide the automatic amortized analysis
-  to derive \emph{non-linear} resource bounds.
+  to derive \emph{non-linear} resource bounds with LP solving.
 \item We prove the soundness of the analysis with respect to a
@@
   further customized with function calls (\code{tick(n)}) that
-  indicate resource usage. This mechanism is also covered by the
-  soundness proof.
+  indicate resource usage. % This mechanism is also covered by the
+  % soundness proof.
 \item We present experiments with a prototype implementation on more

### 2014-11-10 paper/main.tex
use assertions for logic variables

@@
   x=0;
-  while (x < N && a[x] == 1 _&& na > 0_) {
+  while (x < N && a[x] == 1) {
+    _assert(na > 0);_
     a[x]=0; _na--;_
@@
 bsearch(x,l,h_,lg_) {
-  if (h-l > 1 _&& lg > 0_) {
+  if (h-l > 1) {
     m = l + (h-l)/2;
+    _assert(lg > 0);_
     _lg--;_ if (a[m]>x) h=m; else l=m;
@@
 This logical variable is a reification, in the program, of the
-number $\#_1(a)$.  For example, on line 5 of the example
+number $\#_1(a)$.  For example, on line 6 of the example
 we are setting \lstinline{a[x]} to 0 and because of the condition we
@@
 $\#_1(a)$, the logical variable $na$ is decremented.
-Similarly, on line 9, an array entry which was 0 becomes 1, so
+Similarly, on line 10, an array entry which was 0 becomes 1, so
 $na$ is incremented.  To complete the step 2 of the systematic
-procedure described above, we must show that the extra test
-\lstinline{na > 0} at line 4 does not influence the loop behavior.  It
+procedure described above, we must show that the extra assertion
+\lstinline{na > 0} on line 5 is always true.  It
 is done by proving inductively that $na = \#_1(a)$
 and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) > 0$, thus \lstinline{na > 0} is true in the test. %  So the
+$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
 % test is in fact semantically equivalent to \lstinline{true}.

### 2014-11-10 paper/main.tex
new class
shuffle some sections, drop old intro

@@
 
-\documentclass[nocopyrightspace,preprint]{sigplanconf}
+\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}
 
@@
 %SHORT
-
-\usepackage[numbers]{natbib}
 
@@
 
-\lstset{moredelim=[is][\underbar]{_}{_},keepspaces=true}
+\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
 
@@
 
-% In software engineering and software verification, we often would like
-% to have static information about the quantitative behavior of
-% programs.  For example, stack and heap-space bounds are important to
-% ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
-% % Clock-cycle bounds are needed to guarantee the safety of real-time
-% % systems [XXX wilhelm].
-% Static energy usage information is critical for autonomous systems and
-% has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
-% Worst-case time bounds can help to create constant-time
-% implementations that prevent side-channel
-% attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
-% bounds are used to ensure the accuracy of programs that are executed
-% on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
-% needed to verify cryptographic protocols~\cite{BartheGB09}.
-% % XXX Mention differential privacy here?
-% %
-% In general, quantitative performance information at design time can
-% provide useful feedback for developers.
-%
-% Static analysis of quantitative properties of imperative programs is
-% an active research area and recent years have seen many innovations.
-% Notable tools that have been developed include
-% SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
-% PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
-% %
-% While the these tools can derive impressive results for realistic
-% software, there still exist shortcomings that hamper the application
-% of existing techniques in practice:
-% \itemskip
-% \begin{itemize}
-% \item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
-%   without enabling \emph{user interaction} to manually or
-%   semi-automatically derive bounds for challenging parts of the
-%   program.
-% \itemskipIn
-% \item The computed bounds are often \emph{non-compositional} local
-%   bounds (for a single (nested) loop) that are difficult to combine to
-%   global whole program bounds.
-% \itemskipIn
-% \item Existing techniques often rely on complex external tools such as
-%   abstract interpretation--based invariant
-%   generation~\cite{GulwaniMC09} or translation of the program into a
-%   term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
-%   providing \emph{verifiable certificates} for the correctness of the
-%   derived bound.
-% \end{itemize}
-% \itemskip
-% %
-% While there has been much progress in static quantitative analysis,
-% Knuth correctly points out in a recent interview~\cite{KnuthInter}
-% that the state-of-the-art in formal quantitative methods still falls
-% short in comparison with semantic techniques.
-% %
-% \itemskip
-% \begin{quote}
-%   % Consider, for example, a computer program that sorts a list of
-%   % numbers into order.
-%   $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
-%   we have formal definitions of semantics, and tools by which we can
-%   verify that sorting is indeed always achieved. My job is to go
-%   beyond correctness, to an analysis of such things as the program's
-%   running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
-%   describes the program's performance, and all of my colleagues agree
-%   with me that the recurrence is "obviously" valid. Yet I have no
-%   formal tools by which I can prove that my recurrence is right. I
-%   don't really understand my reasoning processes at all!
-%   \vspace{-4ex}
-%   \begin{flushright}
-%     -- Donald E. Knuth, 2014
-%   \end{flushright}
-% \end{quote}
-% \itemskip
-% %
-% In this work, we develop a resource analysis framework for C programs
-% that is based on a solid semantic foundation.  The choice of C is
-% primarily motivated by \iffull{our}{the} ongoing work on the formally verified
-% hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
-% developed in C and is supposed to provide verified guaranties on
-% timing and memory usage.  Moreover, C is a natural choice because it
-% is still the most widely used language for system development; in
-% particular in safety-critical embedded and real-time systems where
-% resource bound analyses are often required by regulatory
-% authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
-% %
-% \itemskip
-% \begin{enumerate}
-% \item We define an operational cost semantics for CompCert Clight that
-%   defines the resource consumption of terminating and diverging
-%   executions.  The cost is parametric in a user-definable cost metric
-%   and can be negative if resources are released during an execution.
-%
-% \item We develop a quantitative Hoare logic for interactively deriving
-%   resource bounds for Clight programs.  The logic is implemented and
-%   proved sound in the Coq Proof Assistant with respect to the cost
-%   semantics of Clight.
-% \itemskipIn
-% \item We describe an automatic static analysis that computes bounds
-%   together with derivations in the quantitative Hoare logic.  The
-%   analysis is inspired by type-based amortized resource analysis for
-%   functional programs~\cite{Jost03,HoffmannAH11}.  It computes
-%   derivations by generating simple linear constraints that can be
-%   solved by an off-the-shelf LP solver, and does not require any
-%   fixpoint computations to obtain loop invariants.
-% \itemskipIn
-% \item We show with a publicly available prototype implementation, and
-%   an experimental evaluation, that our novel amortized resource
-%   analysis works precisely and efficiently for challenging loop and
-%   recursion patterns, and that the derived constant factors in the
-%   bounds are close to or identical with the optimal ones.
-% % \item We prove the completeness of the program logic
-% %   with respect to the operational cost semantics.
-% \end{enumerate}
-% \itemskip
-% %
-% To the best of our knowledge, this article presents the first resource
-% analysis framework for C that makes it possible to combine non-trivial
-% automatically derived bounds with interactively derived bounds in a
-% proof system that produces verifiable certificates for the bounds.
-% Our approach complements existing work since it provides a semantic
-% foundation for the computation of bounds while still supporting
-% automation that sometimes goes beyond the capabilities of existing
-% techniques.  Furthermore, the automatic amortized analysis and the
-% quantitative logic are naturally compositional and describe the
-% resource behavior of code fragments without referring to the source
-% code.  As a result, we directly derive global bounds that are
-% functions of the input parameters of the program.  Another unique
-% feature of our system is, that it can handle resources like memory
-% that may become available during execution.
-%
-% Our starting point is a recent work~\cite{veristack14} \iffull{in
-%   which we have}{that has} formally verified bounds on the stack usage
-% of C programs.  A key part of this stack-bound verification is based
-% on a quantitative Hoare logic for deriving abstract \emph{stack
-%   bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
-%   depth of (recursive) function calls.}  % For
-% % non-recursive programs it is possible to compute derivations for such
-% % stack bounds in the quantitative Hoare logic automatically.
-% % By compiling these bounds together with the C program, we have derived
-% % verified stack bounds for x86 assembly code.
-% %
-% Our \emph{first contribution} is a generalization of the quantitative
-% Hoare logic for CompCert Clight that enables us to derive resource
-% bounds that are parametric in the resource of interest.  % The
-% % quantitative logic is implemented in Coq and proved sound with respect
-% % to a small-step operational cost semantics for Clight.
-% % that counts a
-% % constant resource consumption for each atomic execution step.  The
-% % resource consumption at each step is parametric in a user-definable cost
-% % metric and can be negative to express the release of resources (as
-% % needed, e.g., to reason about memory usage).  Furthermore, a specific
-% % \emph{tick} function can be inserted in the source code to further
-% % customize the cost model.
-% In the implementation of the logic, we use
-% a shallow embedding in Coq which makes the logic very flexible.  There
-% is practically no limitation on the format of the derived resource
-% bounds since they can be any function that is definable in Coq.  The
-% derived bounds can also be parametric in a set of cost metrics or
-% specifically apply to one fixed cost metric.
-%
-% The precision and expressivity of the quantitative Hoare logic provide
-% an ample foundation for reasoning about resource consumption of Clight
-% programs.  % The logic is naturally compositional and the derived
-% % \emph{quantitative Hoare triples} can be used to specify the resource
-% % consumption of library functions or to describe resource contracts
-% % between different functions.
-% However, reasoning about quantitative
-% properties can often be more tedious than reasoning about intentional
-% properties.  Consequently, automation is inevitable to derive resource
-% bounds for large code bases like CertiKOS.  Such an automation was
-% easily achievable in the case of constant stack bounds for a code base
-% that does not contain recursive functions~\cite{veristack14}.  In
-% general however, resource bounds have to be parametric in the
-% arguments of a function and depend on the number of loop iterations
-% and (recursive) function calls that are performed by the function.
-%
-% For functional programs, there exist resource analysis systems---based
-% on type-based amortized resource analysis~\cite{Jost03,
-%   HoffmannAH12}---that can automatically derive complex polynomial
-% bounds with tight constant factors.  This type-based amortized
-% resource analysis has been an inspiration in the design of the
-% quantitative Hoare logic and is therefore a natural candidate to
-% automate the reasoning.  Type-based amortized resource analysis works
-% well for functional programs with pattern matching but it has been a
-% long-time open problem to apply it to C-like programs with control
-% flow that depends on integer arithmetic, negative numbers, and non-linear
-% control flow.
-% % example, it is a
-% % long-time open problem how to use type-based amortized analysis to
-% % derive bounds for functions whose resource consumption depends on
-% % (possibly negative) integers and non-sequential control flow as
-% % introduced by \code{break} and \code{return} statements.
-% % Unfortunately, these are the functions \iffull{with non-constant resource
-% % consumption}{} that are most common in CertiKOS and other system
-% % software.
-%
-% Our \emph{main contribution} is an automatic amortized resource
-% analysis for CompCert Clight that computes derivations in the
-% quantitative Hoare logic.  It is the first automatic amortized
-% resource analysis that can derive bounds that depend on negative
-% integers. It also handles programs with mutually recursive functions
-% as well as \code{break} and \code{return} statements.  The
-% potential-based approach of amortized analysis provides a single
-% mechanism for analyzing programs that need special treatment in other
-% techniques; including
-% \itemskip
-% \begin{itemize}
-% \item interaction between \emph{sequential loops or function calls} through
-%   size changes of variables,
-% \itemskipIn
-% \item \emph{nested loops} that influence each others number of iterations,
-% \itemskipIn
-% \item and \emph{tricky iteration patterns} as found for instance in
-%   the Knuth-Morris-Pratt algorithm for string search.
-% \end{itemize}
-% \itemskip
-% %
-% The main innovation that makes the analysis practical for C is the use
-% of interval sizes in potential functions instead of the sizes of
-% variables.  This leads to precise bounds of programs whose resource
-% consumption can be described as a function of sizes $|[x,y]| =
-% \max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
-% instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
-%   i=i+K)} where the step-size $K>0$ is a constant.
-% %
-%
-% Despite the apparent simplicity of the new analysis system, it is able
-% to reproduce results from the literature (e.g,
-% SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
-% abstract interpretation--based methods such as disjunctive invariant
-% generation~\cite{GulwaniMC09} and symbolic backward
-% execution~\cite{GulwaniZ10}.  In contrast with abstract
-% interpretation--based methods, our technique does not require any
-% fixpoint computations to obtain loop invariants.  The mechanism we
-% designed is able to leverage local assertions such as $x < y$ that we
-% collect along the branching points of the program to obtain global
-% resource invariants.  We achieve this by generating a linear
-% constraint system that reflects the resource cost and size changes of
-% integer variables in the program; a solution of the linear program
-% immediately yields a resource-usage bound for the C
-% program.  % We can naturally, handle
-% % advanced control flow such as break, return, iteration over negative
-% % integer variables, and mutually-recursive functions.
-%
-% Following the development steps of automatic amortized analysis for
-% functional programs~\cite{Jost03,HoffmannH10}, we deliberately
-% restrict ourselves to linear bounds in the automatic analysis (there
-% are no restrictions for the manually-derived bounds in the logic).
-% More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
-% where $a$ and $b$ are integer variables or constants.  The reason for
-% our focus on linear bounds is mainly clarity of presentation.  We
-% already experimented with an extension to multivariate resource
-% polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
-% rules considerably more involved and should better be described
-% separately.  However, we developed the linear inference system so that
-% the extension to polynomial bounds shall work smoothly.
-%
-% We have evaluated the automatic analysis with system code and examples
-% from the literature.  \iffull{\pref{app:cat}}{The extended version
-%   of this article~\cite{anon_extended}}{} contains more than $30$
-% challenging loop and recursion patterns that we collected from open
-% source software and the literature.  Our analysis can find
-% asymptotically tight bounds for all but 1 of these patterns, and in
-% most cases the derived constant factors are tight.  To compare our
-% automatic analyzer with existing techniques, we tested our examples
-% with tools such as KoAT~\cite{BrockschmidtEFFG14},
-% Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
-% show that the bounds that we derive are often more precise than those
-% derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
-% uses amortization techniques, is able to achieve a similar precision.
-% %
-% Several micro benchmarks demonstrate the practicality and
-% expressiveness of the quantitative Hoare logic.  For example, we
-% derive a logarithmic bound for a binary search function, a bound that
-% depends on the contents of an array to describe the exact cost of a
-% function that finds a maximal element in an array, and a linear
-% bound that amortizes the cost of $k$ successive increments to a binary
-% counter.
-%
-% % To show the practicality of our approach we developed a case study
-% % with different real-time schedulers implemented in C.  To perform a
-% % schedulability analysis that takes into account the overhead of the
-% % scheduler we derived time bounds for context switching and
-% % initialization of the scheduler.  We then computed time bounds for a
-% % set of given user task and took account the previously computed
-% % scheduler overhead to determine whether the scheduling requirements of
-% % all user tasks can be met.
-%
-% % With the intention, of making the material easily accessible we first
-% % informally describe the new automatic amortized analysis
-% % (\pref{sec:inform}).  We then formalize the intuition with the
-% % operational cost semantics an the inference rules for the amortized
-% % analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
-% % quantitative Hoare logic and its soundness proof that is formalized in
-% % Coq.  In \pref{sec:inter} we prove the soundness of the automatic
-% % amortized analysis by showing that derivation using the inference
-% % rules can be seen as derivations in the quantitative logic.  We also
-% % show how to integrate hand-derived bounds with automatically-derived
-% % bounds.  \pref{sec:exper} contains the results of the experimental
-% % evaluation and \pref{sec:related} describes related research.
-
 \sectskip
@@
 or additively, resulting in a loose quadratic bound or no bound at all.
-
-\sectskip
-\section{Logical State and User Interaction}
-\label{sec:anno}
-\aftersectskip
-
-\begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    _assert(na > 0);_
-    a[x]=0; _na--;_
-    tick(1);
-    x++; }
-  if (x < N) {
-    a[x]=1; _na++;_
-    tick(1); }
-  k--;
-}
-\end{lstlisting}
-\end{minipage}
-\caption{Example of assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
-  0 to all other operations.  The bound derived automatically is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h_,lg_) {
-  if (h-l > 1) {
-    m = l + (h-l)/2;
-    _assert(lg > 0);_
-    _lg--;_ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h_,lg_);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We use the tick metric to model the stack consumption of the program.
-  Before a function call to \code{bsearch}, a tick statement pays for
-  the space used by stack frame of the function.  The space is recoverd
-  after the call.  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The bound derived automatically is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
-
-While complete automation is often desirable, it cannot possibly
-be universal since the problem is undecidable.  In this section
-we present two techniques to derive complex resource bounds by
-leveraging our automation.
-
-When the resource bound of a program depends on the contents of the heap,
-or is non-linear (e.g. logarithmic, exponential), we propose to
-introduce \emph{logical state}.
-This technique is inspired by XXX (speed) and consists in introducing
-extra variables to the program that do not change its behavior, but help
-the automatic tool to figure out a bound.
-
-The systematic technique consists in the following steps. First,
-a program $P$ that fails to be analyzed automatically is enriched
-by extra variables $\vec x$ called logical variables to form a
-program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
-for the logical variables is shown to be such that:
-$$
-\forall n\, \sigma\, \sigma'.\,
-  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
-    {\implies}
-  \exists n', n' \le n \land
-  (\sigma, P) \Downarrow_{n'} \sigma'.
-$$
-Since the annotated program and the original one are often
-syntactically close, the proof of this result goes
-by simple induction on the resource-aware evaluation judgement.
-Third, using existing automation tools, a bound $B(\vec x)$ for
-$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
-$\vec X$, gives the final resource bound for the program $P$.
-
-This idea is illustrated by the program on \pref{fig:xmplincaux}.
-Parts of the code underlined are annotations that were added
-to the original program text.  The top-level loop increments
-a binary counter $k$ times.  A naive analysis of
-the algorithm yields the quadratic bound $k \cdot N$.
-However, the algorithm is in fact linear and its cost is bounded
-by $2k + \#_1(a)$ where $\#_1(a)$ denotes
-the number of one entries in the array $a$.  Since this
-number depends on the heap contents, no tool available
-for C is able to derive the linear bound.  However, it can be inferred
-by our automated tool if a logical variable $na$ is introduced.
-This logical variable is a reification, in the program, of the
-number $\#_1(a)$.  For example, on line 6 of the example
-we are setting \lstinline{a[x]} to 0 and because of the condition we
-know that this array entry was 1.  To reflect this change on
-$\#_1(a)$, the logical variable $na$ is decremented.
-Similarly, on line 10, an array entry which was 0 becomes 1, so
-$na$ is incremented.  To complete the step 2 of the systematic
-procedure described above, we must show that the extra assertion
-\lstinline{na > 0} on line 5 is always true.  It
-is done by proving inductively that $na = \#_1(a)$
-and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
-% test is in fact semantically equivalent to \lstinline{true}.
-
-Another simple example is given on \pref{fig:xmplbsaux} where
-a logarithmic bound on the stack consumption of a binary search
-program is proved using logical variable annotations.  Once again,
-annotations are underlined in the program text.  In this example,
-to ease the proof of equivalence between the annotated program
-and the original one, we use the inequality $lg >
-\log_2(h-l)$ as invariant.  This trick allows a simpler
-proof because, when working with integer arithmetic, it is not always
-the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
-
-On the other hand, it might be the case that even with annotations, a
-resource bound cannot be found automatically.  With existing tools, this
-situation requires a full rewrite of the program.  However, in our
-framework, it is possible to write a manual proof and combine it with
-bounds derived by the automatic analysis.  This interaction is made
-formal using our Quantitative Hoare Logic.  This resource aware program
-logic is an extension of the one presented in~\cite{veristack14} and
-was originally designed to prove the soundness of our automatic
-analysis.  To make it suitable for this end, we made it independent
-of the resource used (e.g. stack usage, heap usage, clock ticks) and
-implemented a smoother handling of the auxiliary state through a shallow
-embedding in Coq.  Despite being first a theoritical tool, the logic can
-be used directly as in~\cite{veristack14}.  Its judgements are inspired
-by typical Hoare logics and have the shape
-$$
-\Delta \vdash \htriple P S Q
-$$
-where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
-are quantitative assertions, i.e. functions from the program states
-to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
-of the logic are described in further details in the extended version of
-this paper~\cite{anon_extended}.  Because our automated tool produces
-proofs of the resource bounds in the Quantitative Hoare Logic, it can
-be used as an assistant when deriving a complex bound by hand in the
-logic.  The interaction is guaranteed to be sound by the soundness
-theorem of our Quantitative Hoare Logic.
-
-\sectskip
-\section{Overview and Examples}
-\label{sec:inform}
-\aftersectskip
-
-In this section, we informally introduce the quantitative program
-logic and the automatic amortized analysis for Clight programs.
-% The purpose of this section is to build some intuition that should
-% make the following sections easy to read.
-
-\subsection{Quantitative Hoare Logic}
-
-The idea that drives the design of our framework is amortized
-analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
-starting state $\state$ and consumes $n$ resource units of some
-user-defined quantity.  We denote that by writing $(S, \state)
-\Downarrow_n \state'$ where $\state'$ is the program state after the
-execution.  The basic idea of amortized analysis is to define a
-\emph{potential function} $\Phi$ that maps program states to non-negative
-numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
-program state such that $(S, \state) \Downarrow_n \state'$ to prove
-a bound on the resource usage.
-
-To obtain a compositional reasnoning we also have to take into account
-the state resulting from a program's execution.  We thus use two
-potential functions, one that applies before the execution, and one
-that applies after.  The two functions must respect the relation
-$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
-$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
-$\Phi(\state)$ must provide enough \emph{potential} for both, paying
-for the resource cost of the computation and paying for the potential
-$\Phi'(\state')$ on the resulting state $\state'$. That way, if
-$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
-\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
-$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
-$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
-potential function $\Phi$ provides an upper bound on the resource
-consumption of the whole program.  What we have observed is that, if
-we define $\htriple{\Phi} {S}{\Phi'}$ to mean
-$$
-\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
-n + \Phi'(\state') \; ,
-$$
-then we get the following familiar looking rule.
-$$
-\RuleNolabel
-{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
-{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
-$$
-%
-Similarly, other language constructs lead to rules for the potential
-functions that look very similar to Hoare logic or effect system
-rules.  These rules enable interactive reasoning about resource usage
-in a flexible and compositional way, which, as a side effect, produces
-a certificate for the derived resource bound.
-
-% We have implemented and proved sound such a quantitative Hoare for
-% CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
-% embedding in Coq, every function that can be expressed in Coq can in
-% principle be used as a resource bound.
-
-It is also possible to incorporate boolean conditions into a
-bound to express that the bound is only valid for a certain class of
-inputs.  To this end, we allow the potential function $\Phi$ in the
-pre- and postconditions to take non-negative numbers or $\infty$
-(infinity) as values.  Infinity, plays the same role as
-$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
-into potential functions by mapping $\mathit{false}$ to $\infty$ and
-$\mathit{true}$ to $0$.
-%
-\begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}
-${\color{blue}%
-  \{ \#_1(a) + 2k \}}$
-while (k > 0) {
-  x=0;
-  ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
-  while (x < N && a[x] == 1) {
-    ${\color{blue}%
-      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
-    a[x]=0;
-    ${\color{blue}%
-      \{ \#_1(a) + 2k + 1 \}}$
-    tick(1);
-    ${\color{blue}%
-      \{ \#_1(a) + 2k \}}$
-    x++; }
-  ${\color{blue}%
-    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
-  if (x < N) {
-    ${\color{blue}%
-      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
-    a[x]=1;
-    ${\color{blue}%
-      \{ \#_1(a) + 2(k-1) + 1 \}}$
-    tick(1); }
-  ${\color{blue}%
-    \{ \#_1(a) + 2(k-1) \}}$
-  k--;
-  ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
-} ${\color{blue} \{ \#_1(a) \}}$
-\end{lstlisting}
-\end{minipage}
-
-\caption{Example derivation using amortized reasoning.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
-  0 to all other operations.
-  }
-\label{fig:xmplinc}
-\end{figure}
-%
-\iffull{
-}{}
-We can then use the logic to prove a concrete bound for a given resource
-metric, or a bound that is parametric in a set of resource metrics.
-For example, we can derive the following quantitative
-triple for a binary-search function \code{bsearch} that holds for all
-stack metrics $M$.
-%
-\begin{center}
-${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
-\code{bsearch(x,l,h)}
-${ \{ Z{\cdot}M_\code{bsearch} \} }$
-\end{center}
-%
-A stack metric assigns cost $0$ to all evaluation steps except
-function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
-and after the call we return $M_f$ resources that can be used in
-subsequent function calls.  So $M_f$ corresponds to the stack-frame
-size of the function $f$.  In the pre- and postcondition, the logical
-variable $Z$ is used to relate the size of the input with the
-potential that is left after the function call.  The logical part $Z =
-\log_2(h-\ell)$ of the precondition is $0$ if $Z =
-\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
-$\infty$ otherwise.
-
-For simplicity, assume for the remainder of this section that we are
-interested in bounding the number of \emph{ticks} in a program.  That
-is, we use the \emph{tick metric} that assigns cost $n$ to the function call
-\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
-is an integer and a negative number means that $-n$ resources are
-returned.
-
-We say that the analysis is amortized because the use of a potential
-function $\Phi$ enables us to amortize the cost of operations like in
-typical textbook examples that use the potential method of Tarjan's
-amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
-shows a program that repeatedly increments a binary counter
-implemented with a boolean-valued ($0$ or $1$) array.  The derived
-bound in the precondition states that the number of ticks
-executed by the program is bounded by $\#_1(a) + 2k$.  Here, $k$ is
-the number of increments to the counter that are performed by the
-program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
-the number of $1$'s in the array $a$.  We have inserted the statement
-\code{tick(1)} into the program whenever we update the array $a$.  A
-naive analysis of the algorithm would lead to a quadratic bound since
-there can be a linear number of updates in the inner loop.
-
-To derive the bound for the program in~\pref{fig:xmplinc} in the
-quantitative Hoare logic, we follow the textbook
-reasoning~\cite{Algorithms}.  When we assign \code{a[x]=0} in the
-inner loop, we know that $a[x]=1$ before the assignment.  As a result,
-we have $\#_1(a) = \#_1(a') + 1$ if $a'$ is the array $a$ after the
-assignment.  So we obtain one resource unit in addition to the
-invariant (i.e., $\Phi = \#_1(a) + 2k + 1$) that we can use to pay for the
-cost of the statement \code{tick(1)}.  To pay for the tick statement
-in the conditional, we use the fact that $k>0$ to write $2k$ as $1 +
-2(k-1) +1$.  Dual to the first assignment, we have $\#_1(a) +1 =
-\#_1(a')$ if $a$ is the array before and $a'$ is the array after the
-assignment \code{a[x]=1}.  After paying one potential unit for the
-statement $\code{tick(1)}$, we are left with the potential $\#_1(a) +
-2(k - 1)$.  But since we increment $k$ at the end of the outer loop,
-we can establish again our loop invariant $\#_1(a) + 2k$.
-
-See \pref{sec:logic} for more a formal definition of the logic and
-additional example derivations.
-
-% For example we have the following judgement (provided that initially
-% $\state_0(\code x) \ge \state_0(\code y) \ge 0$).
-% $$
-% \htriple
-% {2{\cdot}x + 10}
-% {\code{ x = x -  y}}
-% {2{\cdot}x + 2{\cdot}y + 10}
-% $$
-% This must be understood as a motion of potential from the
-% variable~$\code x$ to the variable~$\code y$. If some resource is
-% consumed $\code y$ times afterwards, we can then use $\code y$'s
-% potential to pay for it. However, the initial potential function
-% $\Phi$ only refers to the variable~$\code x$: the potential function
-% allows us to pass on the cost of the subsequent resource consumption
-% to~$\code x$.  This flexibility combined with intelligent decisions on
-% potential motions allows us to design an automatic analysis to
-% discover linear invariants for surprisingly convoluted programs.
-
-\begin{figure}[t]
-  \centering
-\begin{lstlisting}[mathescape]
-${\color{blue}%
-  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-while (x+K<=y) {
-  ${\color{blue}%
-  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  x=x+K;
-  ${\color{blue}%
-  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  tick(T);
-  ${\color{blue}%
-  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
-}
-${\color{blue}%
- \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
-\end{lstlisting}
-  \caption{Automatic derivation of a tight bound on the number of
-    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
-    $T>0$ are not program variables but denote concrete constants in
-    the program.  The reasoning works uniformly for all such $K$ and
-    $T$.}
-  \label{fig:ex1}
-\end{figure}
-
-%
-\newlength{\progwidth}
-
-\begin{figure*}[t!]
- \setlength{\progwidth}{.22\linewidth}
-  \centering
-\hspace{-0.4cm}
-  \begin{minipage}[b]{.18\linewidth}
-    \begin{center}
-   \begin{lstlisting}[]
-while (n>x) {
-  ${\color{blue}%
-  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-  if (m>y)
-    ${\color{blue}%
-    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
-    y=y+1;
-    ${\color{blue}%
-    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  else
-    ${\color{blue}%
-    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-    x=x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
-}
-   \end{lstlisting}
-
-$|[x, n]| + |[y, m]|$
-\\[.4\baselineskip]
-      {\bf speed\_1}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-
-while (x<n) {
-  ${\color{blue}%
-  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-  if (z>x)
-    ${\color{blue}%
-    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    x=x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  else
-    ${\color{blue}%
-    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    z=z+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[z, n]|$
-\\[.4\baselineskip]
-      {\bf speed\_2}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (z-y>0) {
-  ${\color{blue}%
-  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y=y+1;
-  ${\color{blue}%
-  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(3);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-${\color{blue}%
-\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-while (y>9) {
-  ${\color{blue}%
-  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y=y-10;
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-   \end{lstlisting}
-
-$3.1|[y,z]| + 0.1|[0,y]|$
-\\[.4\baselineskip]
-      {\bf t08a}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (n<0) {
-  ${\color{blue}%
-  n{<}0; \,  P(n,y) \vdash }$
-  n=n+1;
-  ${\color{blue}%
-  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
-  y=y+1000;
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  while (y>=100 && *){
-    ${\color{blue}%
-    y{>}99; \, 9 {+} P(n,y) \vdash }$
-    y=y-100;
-    ${\color{blue}%
-    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
-    tick(5);
-    ${\color{blue}%
-    \dashv \cdot; \, 9 {+} P(n,y) }$
-  }
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  tick(9);
-  ${\color{blue}%
-  \dashv \cdot; \, P(n,y) }$
-}
-   \end{lstlisting}
-$59|[n,0]| {+} 0.05|[0,y]|$
-\\[.4\baselineskip]
-      {\bf t27}
-    \end{center}
-  \end{minipage}
-\vspace{1ex}
-\caption{Derivations of bounds on the number of ticks for challenging
-  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
-  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
-  \emph{t08a} contains \emph{sequential loops} so that the iterations
-  of the second loop depend on the first, and \emph{t27} contains
-  interacting \emph{nested loops}. In the potential functions, we only
-  mention the non-zero terms and in the logical context $\Gamma$ we
-  only mention assertions that we actually use in the reasoning. In
-  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
-  0.05|[0,y]|$.}
-  \label{fig:ex_list}
-\end{figure*}
-%
-%
-
-\subsection{Automatic Amortized Analysis}
-
-A program logic provides a principled foundation for statically
-analyzing programs.  However, program logics need to be supported by
-automatic methods to be useful in practice.  % That is why we developed a
-% new automatic amortized analysis for deriving quantitative Hoare
-% triples for common C programs.
-
-To enable automation, we fix the shape of the potential functions
-$\Phi$ so that it becomes possible to use \emph{linear programming}
-(LP) to compute a derivation in the logic.  If we assume for now that
-a program state $\state$ simply maps variables to integers then we
-require
-$$
-\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
-$$
-where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
-$q_0$ is the constant potential and that we treat constants as program
-variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
-component of $\Phi(\state)$ if $x \in \dom{\state}$.
-%
-We then develop an inference rule for each syntactic construct that
-derives a sound triple $\htriple{\Phi}P{\Phi'}$ in the quantitative
-logic but enables inference using an LP solver.
-
-This idea is best explained by example.  If we again use the
-tick metric that assigns cost $n$ to the function call
-\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
-the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
-that there are no restrictions on the signs of the input variables in
-the examples.}
-\begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
-\end{lstlisting}
-To derive this bound in our automatic amortized analysis, we start
-with the initial potential $\Phi_0 = |[x,y]|$ (short for
-$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
-loop invariant.  For the loop body we then (like in Hoare logic) have
-to derive a triple like
-$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
-only do so if we utilize the fact that $x<y$ at the beginning of the
-loop body.  % The reasoning then works as follows. We start with the
-% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
-% assignment.
-If we denote the updated version of $x$ after the
-assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
-the potential before and after the assignment \code{x=x+1} holds.
-This means that we have the potential $|[x,y]| + 1$ before the
-statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
-we end up with potential $|[x,y]|$ after the loop body and have
-established the loop invariant again.
-
-Note that our notion of a potential function is a generalization of the
-concept of a ranking function.  A potential function can be used like
-a ranking function if we use the tick metric and add the statement
-\code{tick(1)} to every back edge of the program (loops and function
-calls).  However, a potential function is more flexible.  For example,
-we can use a potential function to prove that Example~2 does not
-consume any resources in the tick metric.
-\begin{lstlisting}[basicstyle=\tt\small]
-  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
-\end{lstlisting}
-Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
-In both cases, we reason exactly like in the first version of the
-while loop to prove the bound.  Of course, such loops with different
-tick annotations can be seamlessly combined in a larger program.
-\begin{lstlisting}[basicstyle=\tt\small]
-  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
-\end{lstlisting}
-%
-More formally, we develop (in \pref{sec:AAA}) a judgement
-$$
-\Gamma; Q \vdash S \dashv Q'; \Gamma'
-$$
-such that $\Gamma$ contains logical assertions like $x<y$ that we
-collect along the conditional branches of the program and $Q$ is a
-family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
-indexed by the variables currently in scope.  The logical
-context $\Gamma$ is simply a conjunction of inequalities between
-linear combinations of variables.  It is used, for example, to determine
-at variable assignments if we can extract constant potential to pay for future
-cost.  It is important to note that the reasoning about assertions in
-$\Gamma$ is very basic.  We do not perform any fixpoint computations
-and only derive trivial loop invariants about variables that are
-unchanged in the loop body.
-
-\pref{fig:ex1} shows a derivation of the bound
-$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
-version of Example~1 in which we increment $x$ by a constant $K>0$ and
-consume $T>0$ resources in each iteration.  The reasoning is similar
-to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions in $\Gamma$ are only used in the rule for the assignment
-\code{x=x+K}.  To the best of our knowledge, no other implemented tool
-for C is currently capable of deriving a tight bound on the cost of
-such a loop.  For $T=1$ (many systems focus on the number of loop
-iterations without a cost model) and $K=10$,
-KoAT~\cite{BrockschmidtEFFG14} computes the bound $|x| + |y| + 10$,
-Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
-LOOPUS~\cite{SinnZV14} computes the bound $y-x-9$.  Only
-PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
-translate the program into a term-rewriting system by hand.
-
-To automate the reasoning, we first introduce an unknown rational
-variable for each factor in the potential functions.  We then use our
-inference rules (see \pref{sec:AAA}) to emit linear constraints on
-these variables that enforce that variable assignments that respect
-the constraints correspond to sound potential annotations.  For
-instance if $K = 1$ in \pref{fig:ex1} then we would have the
-annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
-|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
-p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
-are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
-$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.
-
-\iffull{It might be surprising that we}{We} only track simple linear relations in
-the logical context $\Gamma$.  While it would of course be possible to keep
-track of more sophisticated assertions, this simple form is
-sufficient for the examples we considered and we can efficiently
-decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
-assignment rule.  Nevertheless, this logical part of our analysis
-system is orthogonal to the quantitative part and can be easily
-extended if necessary.
-
-As mentioned earlier, the automatic analysis can handle challenging example
-programs without special tricks or techniques.  Examples
-\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
-work~\cite{GulwaniMC09},  demonstrate that our method can handle
-\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
-derives the same bounds as our analysis but requires heuristics for
-its counter instrumentation.  These loops can also be handled with inference of
-\emph{disjunctive invariants}, but in the abstract interpretation
-community, these invariants are known to be notoriously difficult to
-generate.
-%
-In example \emph{speed\_1} we have one loop that first increments
-variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
-derive the tight bound $|[x, n]| + |[y, m]|$.
-%
-Example \emph{speed\_2} is even trickier, and we found it hard to
-find a bound manually.  However, using potential transfer reasoning as
-in amortized analysis, it is easy to prove the tight bound
-$|[x, n]| + |[z, n]|$.
-
-Example \emph{t08a} shows the ability of the analysis to discover
-interaction between \emph{sequenced loops} through size change of
-variables.  We accurately track the size change of $y$ in the first
-loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
-Furthermore, \emph{t08a} shows again that we do not handle the
-constants $1$ or $0$ in any special way.  In all examples we could
-replace $0$ and $1$ with other constants like in the second loop and
-still derive a tight bound.  \iffull{The only information, that the analyzer
-needs is $y \geq c$ before assigning $y = y - c$.}{}
-%
-Example \emph{t27} shows how amortization can be used to handle
-\emph{interacting nested loops}.  In the outer loop we increment the
-variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
-increment the variable $y$ by $1000$.  Then we non-deterministically
-(expressed by \code{*}) execute an inner loop that decrements $y$ by
-$100$ until $y<100$.  The analysis discovers that only the first
-execution of the inner loop depends on the initial value of $y$.  We
-again derive tight constant factors.
-%
-
-As mentioned, the analysis also handles advanced control flow like
-\code{break} and \code{return} statements, and mutual recursion.
-\pref{fig:ex_rec} contains two mutually-recursive functions with their
-automatically derived tick bounds.  The function \code{count\_down}
-decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{count\_up},
-which is dual to \code{count\_down}.  Here, we
-count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis is the only available system that computes a tight
-bound on this example.  \iffull{The analysis amounts to computing the meeting
-point of two trains that approach each other with different speeds.}{}
-% This is comparable to the
-% determination of the point at which to trains meet if they approach
-% each other with different speeds.
-
-\iffull{We only show a small selections of the programs that we can
-  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
-  extended version~\cite{anon_extended}} is a list of more than $30$
-classes of challenging programs that we can automatically analyze.
-\pref{sec:exper} contains a more detailed comparison with other
-tools\iffull{for automatic bound derivation.}{.}
-
-\begin{figure}[t]
-  \centering
-    \begin{minipage}[b]{\linewidth}
-    \begin{center}
-   \begin{lstlisting}
-void count_down (int x,int y) {
-  if (x>y) { tick(1); count_up(x-1,y); } }
-
-void count_up (int x, int y) {
-  if (y+1<x) { tick(1); count_down(x,y+2); } }
-
-   \end{lstlisting}
-
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
-\\[.4\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-\vspace{-1.5ex}
-
-  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
-  \label{fig:ex_rec}
-\end{figure}
-
-\sectskip
-\section{Syntax and Semantics}
-\label{sec:sem}
-\aftersectskip
-
-We implemented our cost semantics and the quantitative Hoare logic in
-Coq for \emph{CompCert Clight}.  Clight is the most abstract
-intermediate language used by CompCert.  Mainly, it is a subset of C
-in which loops can only be exited with a \code{break} statement and
-expressions are free of side effects.
-
-\paragraph{Syntax.}
-
-In this article, we describe our system for a subset of Clight that is
-sufficient to discuss the general ideas.  This subset is given by the
-following grammar.
-%
-\begin{align*}
-S &:= \code{assert}~E
-\mid \code{skip}
-\mid \code{break}
-\mid \code{return}~x
-\mid x \gets E
-\mid x \gets f(x^*)
-\\
-& \mid \code{loop}~S
-\mid \code{if}(E)~S~\code{else}~S
-\mid S;S
-\mid \code{tick}(n)
-\end{align*}
-%
-Expressions $E$ are left abstract in our presentation.  For our
-analysis framework, it is only important that they are side-effect
-free.
-%
-The most notable difference to full Clight is that we can only assign
-to variables and thus do not consider operations that update the heap.
-Moreover, function arguments and return values are assumed to be
-variables.  This is only for simplifying the presentation; in the
-implementation we can deal with heap updates and general function
-calls and returns.  However, we have not implemented our framework for function
-pointers, \code{goto} statements, \code{continue} statements, and
-\code{switch} statements.
-
-We include the built-in primitive $\code{assert}~e$ that terminates
-the program if the argument $e$ evaluates to false and has no effect
-otherwise.  This is useful to express assumptions on the inputs of a
-program for the automatic analysis.  We also add the built-in function
-\code{tick(n)} that can be called with a constant integer $n$ as a
-flexible way to model resource consumption or release (if $n$ is
-negative).
-
-\iffull{
-\begin{figure}[t!]
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:RetSeq}
-{}
-{ (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-\and
-%
-\and
-\Rule{S:RetLoop}
-{}
-{ (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetCall}
-{ \state = (\_, \gamma) \\
-  \state' = (\theta, \gamma)[r \mapsto \state(x)]
-}
-{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_r)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Call}
-{ \fenv f = (\vec x, S_f) \\
-  \state = (\theta, \gamma) \\
-  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
-}
-{ (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep{M}
-(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \cont, \cost')
-}
-%
-\and
-\Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_2, \cont, \cost')
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
-}
-%
-\and
-\Rule{S:Tick}
-{ }
-{ (\state, \code{tick}(n), \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_t(n))
-}
-
-\end{mathpar}
-\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
-\label{fig:opsem}
-\end{figure}
-}{}
-
-\ifshort{
-\begin{figure}[t!]
-\small
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
-}
-\end{mathpar}
-\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
-\label{fig:opsem}
-\end{figure}
-}{}
-
-\ifshort{
-\begin{figure*}[t]
-\small
-\begin{mathpar}
-%
-\Rule{L:Skip}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
-\htriple
-  { M_f + P }
-  { r \gets f(\vec x) }
-  { Q - M_r }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\and
-\Rule{L:Extend}
-{ \Delta \cup \Delta'; B; R \vdash
-  \htriple{ P }{ S }{ Q }
-\\
-  \forall f\,P_f\,Q_f.\,
-  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
-  \rightarrow
-  \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\end{mathpar}
-
-\caption{Rules of the Quantitative Hoare Logic}
-\label{fig:logic}
-\end{figure*}
-}{}
-
-\paragraph{Semantics.}
-
-CompCert Clight's operational semantics is based on small-step
-transitions and continuations.  Expressions---which do not have side
-effects---are evaluated in a big-step fashion.  Here, we describe a
-simplified version of Clight's semantics for the subset we consider.
-
-A program state $\state = (\theta, \gamma)$ is composed of two maps from
-variable names to integers. The first map, $\theta : \text{Locals} \to
-\Z$, assigns integers to local variables of a function, and the second
-map, $\gamma : \text{Globals} \to \Z$, gives values to global variables
-of the program.  In this article, we assume that all values are
-integers but in the implementation we support all data types of
-Clight.
-%
-The evaluation function \evalE{\cdot}{} maps an expression $e \in
-E$ to a value $\evalE{e}{\state} \in \Z$ in the program state
-$\state$.
-
-For simplicity, we assume that local variables and global variables
-are always different.  We just write $\state(x)$ to obtain the value of
-$x$ in program state $\state$. \iffull{Such a lookup is defined as
-$$
-(\theta, \gamma)(x) =
-\left\{
-\begin{array}{lr}
-\theta(x) & \mbox{ if } x \in \dom \theta \\
-\gamma(x) & \mbox{ if } x \in \dom \gamma
-\end{array}
-\right..
-$$}{}
-\iffull{For a program state $\state$, we}{Similarly} we write $\state[x \mapsto v]$
-for the program state that maps $x$ to $v$ and behaves as $\state$ for
-all other variables, regardless whether $x \in \dom \theta$ or $x \in
-\dom \gamma$.
-
-The small-step semantics is standard, except that it tracks the
-resource consumption of a program.  The semantics is parametric in the
-resource of interest for the user of our system.  We achieve this
-independence by parameterizing evaluations with a resource
-\emph{metric} $M$; a tuple of nine rational numbers \ifshort{$M_a$, $M_b$, $M_r$, $\ldots$}{}, one map $M_e :
-\text{Expr} \to \Q$ from expressions to rational numbers, and one map
-$M_t : \Z \to \Q$ from integers to rational numbers.
-%
-\iffull{$$
-M = (
-M_a,
-M_b,
-M_r,
-M_u,
-M_f,
-M_l,
-M_c^1,
-M_c^2,
-M_s,
-M_e(E),
-M_t(n)
-)
-$$}{
-
-}
-Each of these rational numbers indicates the amount of resource
-consumed by a corresponding operation.  If the assigned resource cost
-is negative then it means that resources are released.  The metric
-that we use in the implementation can also depend on other information
-that is statically available such as the name of the called function
-or the number of arguments.
-
-\pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
-rules define a rewrite system for program configurations of the form
-$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
-is the statement being executed, $\cont$ is a continuation that
-describes what remains to be done after the execution of $S$, and
-$\cost \in \Q$ is the non-negative number of resources available for
-further execution.  All rules that can decrease $\cost$ have the
-implicit side condition that the resource quantity available
-\emph{before} the step is non-negative.  This means that we allow
-$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
-transition relation $\smallstep{M}^*$ to indicate that the execution
-ran out of resources.  However, every execution that reaches such a
-state is stuck.
-
-\iffull{
-A continuation $\cont$ represents the context of the execution of a
-given statement.  A continuation can be the empty continuation
-$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
-\cont$, a loop continuation $\Kloop S \cont$, or the continuation
-$\Kcall r \theta \cont$ of a function call.
-$$
-\cont := \Kstop
-\mid \Kseq S \cont
-\mid \Kloop S \cont
-\mid \Kcall r \theta \cont
-$$
-%
-During the execution we assume a fixed function context $\fenv$ that
-maps function names to a list of variables and the statement that
-defines the function body.  It is used in the rule \textsc{S:Call}.
-}{}
-
-The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
-S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
-following.  If the statement $S$ is executed in program state
-$\state$, with continuation $\cont$, and with $\cost$ resources
-available then---after a finite number of steps---the evaluation will
-reach the new machine state $(\state', S', \cont',\cost')$ and there
-are $\cost'$ resources available.  If $\cost' \ge 0$ then the
-execution did not run out of resources and the resource consumption up
-to this point is $\cost - \cost'$.  If this difference is negative
-then resource became available during the execution.  If however
-$\cost' < 0$ then the execution ran out of resources and is stuck.
-The cost of the execution is then $\cost \ge 0$.
-
-\iffull{
-\pref{lem:sem1} summarizes the main properties of the resource
-counter of the semantics.  It can be proved by induction on the number
-of steps.
-\begin{lemma}
-\label{lem:sem1}
-\itemskip
-  \begin{enumerate}
-  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
-    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
-    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
-\itemskipIn
-  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
-    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
-    \cont', d')$ then $\cost - \cost' = d - d'$.
-  \end{enumerate}
-\itemskip
-\end{lemma}
-%
-If we assume that we have a \emph{positive resource metric}, that is,
-a metric that assigns a positive cost to each evaluation step then we
-can link bounds on the resource consumption under the metric to
-termination.  This is formalized by \pref{lem:sem_term}.
-%
-\begin{lemma}
- \label{lem:sem_term}
- Let $M$ be a positive metric and assume that there exists $b \in
- \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
- (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
- Then there exists $B \in \N$ such that $(\state, S, \cont, d)
- \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
- and $\state''$ implies $n \le B$.
-\end{lemma}
-}
-{
-The extended version of this article~\cite{anon_extended} contains all rules and
-lemmas that state the main properties of the cost semantics.
-}
-
-\iffull{
-\begin{figure*}[t]
-\begin{mathpar}
-%
-\Rule{L:Skip}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
-\htriple
-  { M_f + P }
-  { r \gets f(\vec x) }
-  { Q - M_r }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\and
-\Rule{L:Extend}
-{ \Delta \cup \Delta'; B; R \vdash
-  \htriple{ P }{ S }{ Q }
-\\
-  \forall f\,P_f\,Q_f.\,
-  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
-  \rightarrow
-  \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\end{mathpar}
-
-\caption{Rules of the Quantitative Hoare Logic}
-\label{fig:logic}
-\end{figure*}
-}{}
-
-\sectskip
-\section{Quantitative Hoare Logic}
-\label{sec:logic}
-\aftersectskip
-
-In this section we describe a simplified version of the quantitative
-Hoare logic that we use in Coq to interactively prove resource bounds.
-%
-We generalize classic Hoare logic to express not only classical
-boolean-valued assertions but also assertions that talk about the
-future resource usage.  Instead of the usual assertions $P : \State
-\to \mathit{bool}$ of Hoare logic we use assertions
-$$
-P : \State \to \Qplusz \cup \{ \infty \} \; .
-$$
-This can be understood as a refinement of boolean assertions where
-$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
-We write $\Assn$ for $\State \to \Qplusz \cup \{ \infty \}$ and $\bot$ for
-$\lambda \state . \, \infty$.  We sometimes call assertions
-\emph{potential functions}.  To use Coq's support for propositional
-reasoning, assertions have the type $\State \to \Qplusz \to \mathrm{Prop}$
-in the implementation.  For a given $\state \in \State$, such an
-assertion can be seen as a set $B \subseteq \Qplusz$ of valid bounds.
-However, we find the presentation in this article easier to read.
-
-Due to \code{break} and \code{return} statements of Clight, there are
-different possible ways to exit a block of code.  We also have to keep
-track of the resource specifications of functions.  To account for
-this in the logic, our quantitative Hoare triples have the form
-$$
-\Delta; B; R \vdash
-\htriple
-  { Q }
-  { S }
-  { Q' } \; .
-$$
-The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
-two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
-Hoare logic and the intuitive meaning is as follows.  If $S$ is
-executed with starting state $\state$, the empty continuation
-$\Kstop$, and at least $P(\state)$ resources available then the
-evaluation does not run out of resources and there are at least
-$Q(\state')$ resources left if the evaluation terminates in state
-$\state'$.  The assertion $B : \Assn$ provides the
-postcondition for the case in which the code block $S$ is exited by a
-\code{break} statement.  So if the execution is terminated in state
-$\state'$ with a \code{break} then $B(\state')$ resources are
-available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
-case in which the code block $S$ is exited by a $\code{return}~x$
-statement.  The integer argument of $R$ is the return value.
-Finally, the function context of judgements that we write $\Delta$ is
-a mapping from function names to specifications of the form
-$$
-  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
-$$
-The assertion $P_f \, z \, \vec v$ is the precondition of the function
-$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
-both parameterized by an arbitary logical variable $z$ (which can be a
-tuple) that relates the function arguments with the return value.  The
-precondition also depends on $\vec v$, the values of the arguments
-at the function invocation.  Similarly, the postcondition depends on
-the return value $v$ of the function.  The use of logical variables to
-express relations between different states of an execution is a
-standard technique of Hoare logic.
-%
-To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
-the local variables on the stack, that is, $\forall z \, \vec v
-\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
-(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.
-
-For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
-program state $\state$ $P(\state) \ge Q(\state)$.
-
-\paragraph{Rules of the Quantitative Logic.}
-
-\pref{fig:logic} shows the inference rules of the quantitative logic.
-The rules are slightly simplified in comparison to the implemented
-rules in Coq.  The main difference is that the presented version does
-not formalize the heap operations.
-
-\iffull{In the rule {\sc L:Skip}, we do not have to account for any resource
-consumption.  As a result, the precondition $Q$ can be any (potential)
-function and we only have to make sure that we do not end up with more
-potential.  Since the execution of \code{skip} leaves the program
-state unchanged, we can simply use the precondition as postcondition.
-The potential functions $B$ for the \code{break} and $R$ for the
-\code{return} part of the postcondition are not reachable and can
-therefore be arbitrary.}{}
-
-In the rule {\sc L:Assert}, we use the notation
-$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
-require potential $Q + M_a$ in the precondition if $e$ evaluates to
-$\mathit{true}$ in the current program state.  If $e$ evaluates to
-$\mathit{false}$ then the potential in the precondition can be
-arbitrary since the program will be terminated.
-
-In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
-be arbitrary since it is unreachable.  Instead, we have to justify the
-potential functions $B$ and $R$ that hold after a \code{return} and a
-\code{break}, respectively.  In {\sc L:Break}, we require to have
-potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
-cost of \code{break} and $B$ to pay for the potential after the
-\code{break}.  In {\sc L:Return} we only require to have potential
-$R$ in the precondition to pay for the potential after the
-\code{return}.  The reason is that we found it to be more convenient
-to account for the execution cost $M_r$ of the return in rule {\sc
-  L:Call}\iffull{ for function calls.}{.}
-
-\iffull{The rule {\sc L:Update} is the standard assignment rule of Hoare
-logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
-the precondition we ensure that $Q$ evaluates to the same number as in
-the postcondition.  We also require that have we the constant
-potential $M_u + M_e(e)$ available in the precondition to pay for the
-cost of the evaluation of $e$ and the update.
-
-The rule {\sc L:Seq} rule is crucial to understand how the
-quantitative Hoare logic works.  To account for early exits of
-statements, we must ensure in the \code{break} part $B$ of $S_1$'s
-judgement that the \code{break} part $B$ of of $S_1;S_2$ holds. The
-same is true for the return part $R$ of the judgements for $S_1$ and
-$S_2$.  The interaction between the actual pre- and postconditions is
-analogous to standard Hoare logic.  In the postcondition of $S_1$ we
-account for the cost $M_s$ of the execution of the sequence.
-
-}
-{
-The rules {\sc L:Update} and {\sc L:Seq} correspond to the respective
-rules of standard Hoare logic.
-}
-In the rule {\sc L:Loop}, the \code{break} part of the
-loop body $S$ becomes the postcondition of the loop statement. We use
-an arbitrary $B$ as the \code{break} part of the judgement for
-$\code{loop}\, S$ since its operational semantics ensures that it can
-only terminate with a \code{skip} or a \code{return}.  The
-precondition $I$ of the loop is the loop invariant.  \iffull{That is why we
-require to have potential $I$ available in the precondition of the
-loop body $S$.}{}  In the postcondition of $S$, the potential must be
-sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
-iteration.
-
-\iffull{
-The {\sc L:If} is similar to the rule for the conditional in classic
-Hoare logic.  In the preconditions of the judgments for the two
-branches $S_1$ and $S_2$ we lift the boolean assertions
-$\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
-quantitative assertions.  In the precondition of the rule {\sc L:Tick}
-we account for the execution cost $M_t(n)$ of the \code{tick}
-statement that depends on the integer $n$.
-}{}
-
-The rule {\sc L:Call} accounts for the execution cost of both $M_f$
-for function calls and $M_r$ for \code{return} statements.
-\iffull{This justifies that the {\sc L:Return} rule does not account
-  for resource consumption.}{} The pre- and postcondition $P_f$ and
-$Q_f$ are taken from the function context $\Delta$.  The assertions in
-the context are parametric with respect to both the values of the
-function arguments and the return value. This allows us to specify a
-bound for a function whose resource consumption depends on its
-arguments.  The arguments are instantiated by the call rule
-using the result of the evaluation of the argument variables in the
-current state.  \iffull{Recall that we require that $P_f$ and $Q_f$ do
-  not depend on the local variables on the call stack, that is,
-  $\forall z \, \vec v \, \theta \, \theta' \, \gamma . \, P_f \, z \,
-  \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
-  (\theta',\gamma)$.}{} To transfer potential that depends on local
-variables of the callee from the precondition $P$ to the postcondition
-$Q$, we use an assertion $A : \Assn$ that is independent of global
-variables, that is, $\forall \theta \, \gamma \, \gamma' . \, A
-(\theta,\gamma) = A(\theta,\gamma')$.  It is still possible to express
-relations between global and local variables using logical
-variables\iffull{ (see the following paragraph for details).}{.}
-
-Finally, we describe the rules which are not syntax directed.  There
-are two weakening rules available in the quantitative Hoare logic.
-The framing rule {\sc L:Frame} is designed to weaken a statement by
-stating that if $S$ needs $P$ resources to run and leaves $Q$
-resources available after its execution, then it can very well run
-with $P + c$ resources and return $Q + c$ resources.  The consequence
-{\sc L:Conseq} rule is directly imported from classical Hoare logic
-except that instead of using the logical implication $\Rightarrow$ we
-use the quantitative $\models$ that point-wise applies $\ge$.  \iffull{This
-rule indeed weakens the statement since it requires more resource to
-run the statement and yields less than what has been proved to be
-available after its termination.}{}
-
-\ifx\fullversion\undefined{}\else{
-\paragraph{Logical Variables and Shallow Embedding.}
-
-When specifying a program, it is often necessary to relate certain
-invariants in the pre- and postcondition.  A standard solution of Hoare
-logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
-additional variables (also called auxiliary state in the literature)
-are constant across the derivation.  For example, if $Z$ is such
-a logical variable, we can specify the function $\code{double}()$
-which doubles the global variable $x$ as
-$$
-  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
-$$
-%
-When formalizing Hoare logics in a proof assistant one can either
-fully specify the syntax and semantics of assertions and hence get a
-deep embedding, or use the assertions of the host theory to get a
-shallow embedding.  Because of its flexibility, we used the latter
-approach in our development.  This choice makes it possible to have
-logical variables almost for free: we can simply use the variable and
-binding mechanisms of Coq, our host theory.  When an logical variable
-is needed we introduce it using a universal quantifier of Coq before
-using the logic to derive triples.  For example, the Coq theorem for
-the above example would look as follows.
-%
-\begin{lstlisting}
-Theorem dbl_triple: forall Z,
-  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
-\end{lstlisting}
-%
-However, this trick alone is often not sufficient when working with a
-recursive function $f$.  In that case we apply the {\sc L:Extend} rule
-of the logic.  First we add a specification $\forall z\,\vec
-v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
-$\Delta$.  Then we proceed to prove the function body $S_f$ with this
-induction hypothesis.  In this process it can be the case that we have
-to use the induction hypothesis with a different value of a logical
-variable (e.g., because the values of the arguments in the recursive
-call differ from the values of the arguments of the callee).  To cope
-with this problem, assumptions in the function context $\Delta$ are
-universally quantified over logical variables.  The {\sc L:Extend}
-rule uses the host proof assistant to require that the triple on $f$'s
-body is proved for every possible logical variable $y$.
-}\fi
-
-\paragraph{Using the Quantitative Logic.}
-
-In the following we demonstrate the use of the logic with two example
-derivations.
-
-In the example in~\pref{fig:xmplmax} we derive a precise runtime
-bound on a program that searches a maximal element in an array.  The
-cost metric that we use simply counts the assignments performed by the
-program.  Hence, the resource cost is closely related to the
-number of times the test \code{a[i] > m} is true during the
-execution.
-%
-If we define
-$$
- A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
-$$
-where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
-the number of ``maximum candidates'' in the array \code{a} seen
-by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
-So any automated tool would at best derive the linear bound $2 \cdot
-N$ for that program.  But with the expressivity of our logic it is
-possible to use the previous set cardinal directly and precisely tie
-the bound to the initial contents of the array.
-%
-The non-trivial part of this derivation is finding the loop invariant
-$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
-When the condition \code{a[i] > m} is true, we know that we
-encountered a ``maximum so far'' because $m$ is a maximal element of
-\code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
-potential unit to pay for the assignment.  In the other case, no
-maximum so far is encountered so $A(i) = A(i+1)$.
-% As a side remark, if $K \ge 0$, it is possible to
-% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
-% of the same program.  It can be done either by adapting the proof or by
-% applying the {\sc L:Frame} rule one top of the triple already derived.
-% More generally, without function calls, the {\sc L:Frame}
-% rule is admissible in our system.
-
-The example in~\pref{fig:xmplbs} shows a use case for logical
-variables as well as a metric for stack consumption. In a stack
-metric, we account a constant cost for a function call ($M_f>0$) that
-is returned after the call ($M_r = -M_f <0$).  All other resource cost
-are $0$.  We are interested in showing that a binary search function
-$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
-variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
-  + Z \cdot M_\code{bsearch}}{}{Z \cdot M_\code{bsearch}}$ to express that the stack required by the
-function is returned after the call.
-%
-The critical step in the proof is the application of the
-{\sc L:Call} rule to the recursive call.  At this point
-the context $\Delta$ contains the specification
-$
-  \forall y \, (x,l,h) \, \_.
-  ( (\lambda y \, (\_, l, h) .\, (y = \log_2(h {-} l)) {+} y {\cdot} M_\code{bsearch})
-  , (\lambda y \, \_ .\, y {\cdot} M_\code{bsearch})
-  )
-$.
-%
-Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
-- 1$, and because $M_f = M_\code{bsearch}$ and $M_r =
--M_\code{bsearch}$ in the stack metric.  The rest of the proof does
-not involve any resource manipulation and is just bookkeeping of
-logical assertions.
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue} \{ A(0) + N \} }$
-i=1; m=a[0];
-${\color{blue}%
-  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
-while (i < N) {
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
-  if (a[i] > m)
-    m=a[i];
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
-  i=i+1;
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
-} ${\color{blue} \{ 0 \} }$
-\end{lstlisting}
-\caption{Example derivation where we wrote $A(i)$
-  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
-  the metric used here assigns a cost of 1 to every assignment
-  and 0 to all other operations.
-  }
-\label{fig:xmplmax}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}%
-  \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
-bsearch(x,l,h) {
-  if (h-l > 1) {
-    ${\color{blue}%
-      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
-    m = h + (h-l)/2;
-    ${\color{blue}%
-      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
-    if (a[m]>x) h=m; else l=m;
-    ${\color{blue}%
-      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
-    l = bsearch(x,l,h); }
-  ${\color{blue}%
-    \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
-  return l;
-} ${\color{blue} \{ Z \cdot M_\code{bsearch} \} }$
-\end{lstlisting}
-\caption{Example derivation of a stack usage bound for a binary
-  search program.  The used resource metric defines the cost $M_\code{bsearch}$ before
-  the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
-  frame size of the function $\code{bsearch}$.  $Z$ is a
-  logical variable.
-  }
-\label{fig:xmplbs}
-\end{figure}
-
-\paragraph{Soundness of Quantitative Hoare Triples.}
-
-We already gave an intuition of the meaning of judgements
-derived in the logic.  To make it formal, we define the
-\emph{resource safety} $\safe n P S \cont$ of an assertion
-and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
-\begin{align*}
-  (m {\le} n \land P(\state) {\le} \cost \land
-    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
-  \implies \cost' {\ge} 0.
-\end{align*}
-This predicate is step indexed by an integer $n$ that is used for
-induction in the soundness proof for the function-call and loop cases.
-The constraint $\cost' \geq 0$ imposed by the definition ensures that
-the program does not stop because of a resource error (recall that a
-negative resource counter on the right-hand side is a resource
-failure).  However, it does not rule out memory safety errors or
-assertion failures. This is because our logic does not prove any
-safety or correctness theorems but only focuses on resource usage.
-%
-An interesting detail in the definition is the natural number $m$.  We
-simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
-\cont $.  This would not be the case if we replaced all occurrences of
-$m$ by $n$ in the definition\iffull{ of validity.}{.}
-
-The resource safety of a continuation $\cont$ is defined using three
-assertions, one for each of the possible outcomes of a program
-statement.  \iffull{We define it as follows.
-\begin{align*}
-\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
-& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
-& & \land& \safe n Q {\code{skip}} \cont
-\end{align*}}
-{
-We define $\safeK n B R Q \cont \mathop{:=}
-\safe n B {\code{break}} \cont \, \land \,
-\safe n Q {\code{skip}} \cont \, \land \,
-\safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont$.
-
-}
-%
-We can now define the \emph{semantic validity} of a judgement $B; R
-\vdash \htriple P S Q$ of the quantitative logic without function
-context as $\valid n B R P S Q :=$
-\begin{align*}
-& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
-& \implies \safe m {P+x} S \cont \; .
-\end{align*}
-Note how the validity of a triple embeds the frame rule of
-our logic. This refinement is necessary to have a stronger
-induction hypothesis available during the proof.
-%
-We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
-S Q$ implies $\valid n B R P S Q$.
-
-Using the semantic validity of triples we define the validity
-of a function context $\Delta$, written $\validC n \Delta$, as
-\begin{align*}
-  &\forall f .\, \Delta(f) =
-    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
-    \implies \\
-  &\qquad \forall z \, \vec v .\,
-  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
-\end{align*}
-where $S_f$ is the body of the function $f$. \iffull{A full
-judgement that mentions a non-empty function context
-$\Delta$ is in fact a guarded statement: it makes
-assumptions on some functions' behavior.}{}  The predicate
-$\validC n \Delta$ gives the precise meaning of the
-assumptions made.  It is also step-indexed to prove the
-soundness of the {\sc L:Extend} rule by induction.
-%
-We are now able to state the soundness of the quantitative logic.
-%
-\begin{theorem}[Soundness of the logic]
-  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
-  $
-    \forall n .\, \validC n \Delta
-      \implies \valid {n {+} 1} B R P S Q.
-  $
-\end{theorem}
-%
-\noindent
-The difference $\delta = 1$ between the index in the triple
-validity and the one in context validity arises from
-the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
-{\sc L:Call}, the language semantics makes one step and
-proceeds with the function body, so we must have
-$\delta \le 1$ to use the assumptions in $\Delta$.
-For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
-is a valid context for $n$ steps.  The induction hypothesis
-in that case says that if $\Delta \cup \Delta'$ is valid
-for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
-So if we want to solve this goal by induction, it is
-necessary that $\delta \ge 1$.  These two constraints force
-$\delta$ to be exactly one in the theorem statement.
-
-Assume that $S$ is a complete program and $\Delta$ is empty.  By
-expanding the definitions we see that $\Delta$ is valid
-for every $n$ and that $\Kstop$ is safe for every $n$. So
-we derive
-$$
-\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
-$$
-This means that from any starting state $\state$, $P(\state)$
-provides enough resources for any run of the program $S$.  \iffull{This
-setting is actually the main use case of the previous theorem
-which is stated as a stronger result to allow a proof by
-induction.}{}
 
@@
 \sectskip
+\section{Logical State and User Interaction}
+\label{sec:anno}
+\aftersectskip
+
+\begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1) {
+    @assert(na > 0);@
+    a[x]=0; @na--;@
+    tick(1);
+    x++; }
+  if (x < N) {
+    a[x]=1; @na++;@
+    tick(1); }
+  k--;
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Example of assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
+  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
+  0 to all other operations.  The bound derived automatically is
+  $2|[0,k]| + |[0,na]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ lg > \log_2(h - l) \} }$
+bsearch(x,l,h@,lg@) {
+  if (h-l > 1) {
+    @assert(lg > 0);@
+    m = l + (h-l)/2;
+    @lg--;@ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h@,lg@);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We use the tick metric to model the stack consumption of the program.
+  Before a function call to \code{bsearch}, a tick statement pays for
+  the space used by stack frame of the function.  The space is recoverd
+  after the call.  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
+  The bound derived automatically is $|[0,lg]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
+
+While complete automation is often desirable, it cannot possibly
+be universal since the problem is undecidable.  In this section
+we present two techniques to derive complex resource bounds by
+leveraging our automation.
+
+When the resource bound of a program depends on the contents of the heap,
+or is non-linear (e.g. logarithmic, exponential), we propose to
+introduce \emph{logical state}.
+This technique is inspired by XXX (speed) and consists in introducing
+extra variables to the program that do not change its behavior, but help
+the automatic tool to figure out a bound.
+
+The systematic technique consists in the following steps. First,
+a program $P$ that fails to be analyzed automatically is enriched
+by extra variables $\vec x$ called logical variables to form a
+program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
+for the logical variables is shown to be such that:
+$$
+\forall n\, \sigma\, \sigma'.\,
+  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
+    {\implies}
+  \exists n' {\le} n.\,
+    (\sigma, P) \Downarrow_{n'} \sigma'.
+$$
+Since the annotated program and the original one are often
+syntactically close, the proof of this result goes
+by simple induction on the resource-aware evaluation judgement.
+Third, using existing automation tools, a bound $B(\vec x)$ for
+$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
+$\vec X$, gives the final resource bound for the program $P$.
+
+This idea is illustrated by the program on \pref{fig:xmplincaux}.
+Parts of the code in blue are annotations that were added
+to the original program text.  The top-level loop increments
+a binary counter $k$ times.  A naive analysis of
+the algorithm yields the quadratic bound $k \cdot N$.
+However, the algorithm is in fact linear and its cost is bounded
+by $2k + \#_1(a)$ where $\#_1(a)$ denotes
+the number of one entries in the array $a$.  Since this
+number depends on the heap contents, no tool available
+for C is able to derive the linear bound.  However, it can be inferred
+by our automated tool if a logical variable $na$ is introduced.
+This logical variable is a reification, in the program, of the
+number $\#_1(a)$.  For example, on line 6 of the example
+we are setting \lstinline{a[x]} to 0 and because of the condition we
+know that this array entry was 1.  To reflect this change on
+$\#_1(a)$, the logical variable $na$ is decremented.
+Similarly, on line 10, an array entry which was 0 becomes 1, so
+$na$ is incremented.  To complete the step 2 of the systematic
+procedure described above, we must show that the extra assertion
+\lstinline{na > 0} on line 5 is always true.  It
+is done by proving inductively that $na = \#_1(a)$
+and remarking that since \lstinline{a[x] == 1} is true, we must have
+$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
+% test is in fact semantically equivalent to \lstinline{true}.
+
+Another simple example is given on \pref{fig:xmplbsaux} where
+a logarithmic bound on the stack consumption of a binary search
+program is proved using logical variable annotations.  Once again,
+annotations are in blue in the program text.  In this example,
+to ease the proof of equivalence between the annotated program
+and the original one, we use the inequality $lg >
+\log_2(h-l)$ as invariant.  This trick allows a simpler
+proof because, when working with integer arithmetic, it is not always
+the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
+
+On the other hand, it might be the case that even with annotations, a
+resource bound cannot be found automatically.  With existing tools, this
+situation requires a full rewrite of the program.  However, in our
+framework, it is possible to write a manual proof and combine it with
+bounds derived by the automatic analysis.  This interaction is made
+formal using our Quantitative Hoare Logic.  This resource aware program
+logic is an extension of the one presented in~\cite{veristack14} and
+was originally designed to prove the soundness of our automatic
+analysis.  To make it suitable for this end, we made it independent
+of the resource used (e.g. stack usage, heap usage, clock ticks) and
+implemented a smoother handling of the auxiliary state through a shallow
+embedding in Coq.  Despite being first a theoritical tool, the logic can
+be used directly as in~\cite{veristack14}.  Its judgements are inspired
+by typical Hoare logics and have the shape
+$$
+\Delta \vdash \htriple P S Q
+$$
+where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
+are quantitative assertions, i.e. functions from the program states
+to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
+of the logic are described in further details in the extended version of
+this paper~\cite{anon_extended}.  Because our automated tool produces
+proofs of the resource bounds in the Quantitative Hoare Logic, it can
+be used as an assistant when deriving a complex bound by hand in the
+logic.  The interaction is guaranteed to be sound by the soundness
+theorem of our Quantitative Hoare Logic.
+
+\sectskip
+\section{Overview and Examples}
+\label{sec:inform}
+\aftersectskip
+
+In this section, we informally introduce the quantitative program
+logic and the automatic amortized analysis for Clight programs.
+% The purpose of this section is to build some intuition that should
+% make the following sections easy to read.
+
+\subsection{Quantitative Hoare Logic}
+
+The idea that drives the design of our framework is amortized
+analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
+starting state $\state$ and consumes $n$ resource units of some
+user-defined quantity.  We denote that by writing $(S, \state)
+\Downarrow_n \state'$ where $\state'$ is the program state after the
+execution.  The basic idea of amortized analysis is to define a
+\emph{potential function} $\Phi$ that maps program states to non-negative
+numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
+program state such that $(S, \state) \Downarrow_n \state'$ to prove
+a bound on the resource usage.
+
+To obtain a compositional reasnoning we also have to take into account
+the state resulting from a program's execution.  We thus use two
+potential functions, one that applies before the execution, and one
+that applies after.  The two functions must respect the relation
+$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
+$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
+$\Phi(\state)$ must provide enough \emph{potential} for both, paying
+for the resource cost of the computation and paying for the potential
+$\Phi'(\state')$ on the resulting state $\state'$. That way, if
+$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
+\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
+$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
+$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
+potential function $\Phi$ provides an upper bound on the resource
+consumption of the whole program.  What we have observed is that, if
+we define $\htriple{\Phi} {S}{\Phi'}$ to mean
+$$
+\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
+n + \Phi'(\state') \; ,
+$$
+then we get the following familiar looking rule.
+$$
+\RuleNolabel
+{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
+{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
+$$
+%
+Similarly, other language constructs lead to rules for the potential
+functions that look very similar to Hoare logic or effect system
+rules.  These rules enable interactive reasoning about resource usage
+in a flexible and compositional way, which, as a side effect, produces
+a certificate for the derived resource bound.
+
+% We have implemented and proved sound such a quantitative Hoare for
+% CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
+% embedding in Coq, every function that can be expressed in Coq can in
+% principle be used as a resource bound.
+
+It is also possible to incorporate boolean conditions into a
+bound to express that the bound is only valid for a certain class of
+inputs.  To this end, we allow the potential function $\Phi$ in the
+pre- and postconditions to take non-negative numbers or $\infty$
+(infinity) as values.  Infinity, plays the same role as
+$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
+into potential functions by mapping $\mathit{false}$ to $\infty$ and
+$\mathit{true}$ to $0$.
+%
+\begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}
+${\color{blue}%
+  \{ \#_1(a) + 2k \}}$
+while (k > 0) {
+  x=0;
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+  while (x < N && a[x] == 1) {
+    ${\color{blue}%
+      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
+    a[x]=0;
+    ${\color{blue}%
+      \{ \#_1(a) + 2k + 1 \}}$
+    tick(1);
+    ${\color{blue}%
+      \{ \#_1(a) + 2k \}}$
+    x++; }
+  ${\color{blue}%
+    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
+  if (x < N) {
+    ${\color{blue}%
+      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
+    a[x]=1;
+    ${\color{blue}%
+      \{ \#_1(a) + 2(k-1) + 1 \}}$
+    tick(1); }
+  ${\color{blue}%
+    \{ \#_1(a) + 2(k-1) \}}$
+  k--;
+  ${\color{blue}%
+    \{ \#_1(a) + 2k \}}$
+} ${\color{blue} \{ \#_1(a) \}}$
+\end{lstlisting}
+\end{minipage}
+
+\caption{Example derivation using amortized reasoning.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
+  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
+  0 to all other operations.
+  }
+\label{fig:xmplinc}
+\end{figure}
+%
+\iffull{
+}{}
+We can then use the logic to prove a concrete bound for a given resource
+metric, or a bound that is parametric in a set of resource metrics.
+For example, we can derive the following quantitative
+triple for a binary-search function \code{bsearch} that holds for all
+stack metrics $M$.
+%
+\begin{center}
+${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
+\code{bsearch(x,l,h)}
+${ \{ Z{\cdot}M_\code{bsearch} \} }$
+\end{center}
+%
+A stack metric assigns cost $0$ to all evaluation steps except
+function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
+and after the call we return $M_f$ resources that can be used in
+subsequent function calls.  So $M_f$ corresponds to the stack-frame
+size of the function $f$.  In the pre- and postcondition, the logical
+variable $Z$ is used to relate the size of the input with the
+potential that is left after the function call.  The logical part $Z =
+\log_2(h-\ell)$ of the precondition is $0$ if $Z =
+\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
+$\infty$ otherwise.
+
+For simplicity, assume for the remainder of this section that we are
+interested in bounding the number of \emph{ticks} in a program.  That
+is, we use the \emph{tick metric} that assigns cost $n$ to the function call
+\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
+is an integer and a negative number means that $-n$ resources are
+returned.
+
+We say that the analysis is amortized because the use of a potential
+function $\Phi$ enables us to amortize the cost of operations like in
+typical textbook examples that use the potential method of Tarjan's
+amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
+shows a program that repeatedly increments a binary counter
+implemented with a boolean-valued ($0$ or $1$) array.  The derived
+bound in the precondition states that the number of ticks
+executed by the program is bounded by $\#_1(a) + 2k$.  Here, $k$ is
+the number of increments to the counter that are performed by the
+program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
+the number of $1$'s in the array $a$.  We have inserted the statement
+\code{tick(1)} into the program whenever we update the array $a$.  A
+naive analysis of the algorithm would lead to a quadratic bound since
+there can be a linear number of updates in the inner loop.
+
+To derive the bound for the program in~\pref{fig:xmplinc} in the
+quantitative Hoare logic, we follow the textbook
+reasoning~\cite{Algorithms}.  When we assign \code{a[x]=0} in the
+inner loop, we know that $a[x]=1$ before the assignment.  As a result,
+we have $\#_1(a) = \#_1(a') + 1$ if $a'$ is the array $a$ after the
+assignment.  So we obtain one resource unit in addition to the
+invariant (i.e., $\Phi = \#_1(a) + 2k + 1$) that we can use to pay for the
+cost of the statement \code{tick(1)}.  To pay for the tick statement
+in the conditional, we use the fact that $k>0$ to write $2k$ as $1 +
+2(k-1) +1$.  Dual to the first assignment, we have $\#_1(a) +1 =
+\#_1(a')$ if $a$ is the array before and $a'$ is the array after the
+assignment \code{a[x]=1}.  After paying one potential unit for the
+statement $\code{tick(1)}$, we are left with the potential $\#_1(a) +
+2(k - 1)$.  But since we increment $k$ at the end of the outer loop,
+we can establish again our loop invariant $\#_1(a) + 2k$.
+
+See \pref{sec:logic} for more a formal definition of the logic and
+additional example derivations.
+
+% For example we have the following judgement (provided that initially
+% $\state_0(\code x) \ge \state_0(\code y) \ge 0$).
+% $$
+% \htriple
+% {2{\cdot}x + 10}
+% {\code{ x = x -  y}}
+% {2{\cdot}x + 2{\cdot}y + 10}
+% $$
+% This must be understood as a motion of potential from the
+% variable~$\code x$ to the variable~$\code y$. If some resource is
+% consumed $\code y$ times afterwards, we can then use $\code y$'s
+% potential to pay for it. However, the initial potential function
+% $\Phi$ only refers to the variable~$\code x$: the potential function
+% allows us to pass on the cost of the subsequent resource consumption
+% to~$\code x$.  This flexibility combined with intelligent decisions on
+% potential motions allows us to design an automatic analysis to
+% discover linear invariants for surprisingly convoluted programs.
+
+\begin{figure}[t]
+  \centering
+\begin{lstlisting}[mathescape]
+${\color{blue}%
+  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+while (x+K<=y) {
+  ${\color{blue}%
+  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  x=x+K;
+  ${\color{blue}%
+  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  tick(T);
+  ${\color{blue}%
+  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+}
+${\color{blue}%
+ \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+\end{lstlisting}
+  \caption{Automatic derivation of a tight bound on the number of
+    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
+    $T>0$ are not program variables but denote concrete constants in
+    the program.  The reasoning works uniformly for all such $K$ and
+    $T$.}
+  \label{fig:ex1}
+\end{figure}
+
+%
+\newlength{\progwidth}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.22\linewidth}
+  \centering
+\hspace{-0.4cm}
+  \begin{minipage}[b]{.18\linewidth}
+    \begin{center}
+   \begin{lstlisting}[]
+while (n>x) {
+  ${\color{blue}%
+  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+  if (m>y)
+    ${\color{blue}%
+    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
+    y=y+1;
+    ${\color{blue}%
+    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  else
+    ${\color{blue}%
+    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
+}
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.4\baselineskip]
+      {\bf speed\_1}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+while (x<n) {
+  ${\color{blue}%
+  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+  if (z>x)
+    ${\color{blue}%
+    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  else
+    ${\color{blue}%
+    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    z=z+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[z, n]|$
+\\[.4\baselineskip]
+      {\bf speed\_2}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (z-y>0) {
+  ${\color{blue}%
+  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y=y+1;
+  ${\color{blue}%
+  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(3);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+${\color{blue}%
+\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+while (y>9) {
+  ${\color{blue}%
+  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y=y-10;
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+   \end{lstlisting}
+
+$3.1|[y,z]| + 0.1|[0,y]|$
+\\[.4\baselineskip]
+      {\bf t08a}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n<0) {
+  ${\color{blue}%
+  n{<}0; \,  P(n,y) \vdash }$
+  n=n+1;
+  ${\color{blue}%
+  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
+  y=y+1000;
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  while (y>=100 && *){
+    ${\color{blue}%
+    y{>}99; \, 9 {+} P(n,y) \vdash }$
+    y=y-100;
+    ${\color{blue}%
+    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
+    tick(5);
+    ${\color{blue}%
+    \dashv \cdot; \, 9 {+} P(n,y) }$
+  }
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  tick(9);
+  ${\color{blue}%
+  \dashv \cdot; \, P(n,y) }$
+}
+   \end{lstlisting}
+$59|[n,0]| {+} 0.05|[0,y]|$
+\\[.4\baselineskip]
+      {\bf t27}
+    \end{center}
+  \end{minipage}
+\vspace{1ex}
+\caption{Derivations of bounds on the number of ticks for challenging
+  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
+  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
+  \emph{t08a} contains \emph{sequential loops} so that the iterations
+  of the second loop depend on the first, and \emph{t27} contains
+  interacting \emph{nested loops}. In the potential functions, we only
+  mention the non-zero terms and in the logical context $\Gamma$ we
+  only mention assertions that we actually use in the reasoning. In
+  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
+  0.05|[0,y]|$.}
+  \label{fig:ex_list}
+\end{figure*}
+%
+%
+
+\subsection{Automatic Amortized Analysis}
+
+A program logic provides a principled foundation for statically
+analyzing programs.  However, program logics need to be supported by
+automatic methods to be useful in practice.  % That is why we developed a
+% new automatic amortized analysis for deriving quantitative Hoare
+% triples for common C programs.
+
+To enable automation, we fix the shape of the potential functions
+$\Phi$ so that it becomes possible to use \emph{linear programming}
+(LP) to compute a derivation in the logic.  If we assume for now that
+a program state $\state$ simply maps variables to integers then we
+require
+$$
+\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
+$$
+where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
+$q_0$ is the constant potential and that we treat constants as program
+variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
+component of $\Phi(\state)$ if $x \in \dom{\state}$.
+%
+We then develop an inference rule for each syntactic construct that
+derives a sound triple $\htriple{\Phi}P{\Phi'}$ in the quantitative
+logic but enables inference using an LP solver.
+
+This idea is best explained by example.  If we again use the
+tick metric that assigns cost $n$ to the function call
+\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
+the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
+that there are no restrictions on the signs of the input variables in
+the examples.}
+\begin{lstlisting}[basicstyle=\tt\small]
+while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
+\end{lstlisting}
+To derive this bound in our automatic amortized analysis, we start
+with the initial potential $\Phi_0 = |[x,y]|$ (short for
+$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
+loop invariant.  For the loop body we then (like in Hoare logic) have
+to derive a triple like
+$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
+only do so if we utilize the fact that $x<y$ at the beginning of the
+loop body.  % The reasoning then works as follows. We start with the
+% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
+% assignment.
+If we denote the updated version of $x$ after the
+assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
+the potential before and after the assignment \code{x=x+1} holds.
+This means that we have the potential $|[x,y]| + 1$ before the
+statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
+we end up with potential $|[x,y]|$ after the loop body and have
+established the loop invariant again.
+
+Note that our notion of a potential function is a generalization of the
+concept of a ranking function.  A potential function can be used like
+a ranking function if we use the tick metric and add the statement
+\code{tick(1)} to every back edge of the program (loops and function
+calls).  However, a potential function is more flexible.  For example,
+we can use a potential function to prove that Example~2 does not
+consume any resources in the tick metric.
+\begin{lstlisting}[basicstyle=\tt\small]
+  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
+\end{lstlisting}
+Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
+In both cases, we reason exactly like in the first version of the
+while loop to prove the bound.  Of course, such loops with different
+tick annotations can be seamlessly combined in a larger program.
+\begin{lstlisting}[basicstyle=\tt\small]
+  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
+\end{lstlisting}
+%
+More formally, we develop (in \pref{sec:AAA}) a judgement
+$$
+\Gamma; Q \vdash S \dashv Q'; \Gamma'
+$$
+such that $\Gamma$ contains logical assertions like $x<y$ that we
+collect along the conditional branches of the program and $Q$ is a
+family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
+indexed by the variables currently in scope.  The logical
+context $\Gamma$ is simply a conjunction of inequalities between
+linear combinations of variables.  It is used, for example, to determine
+at variable assignments if we can extract constant potential to pay for future
+cost.  It is important to note that the reasoning about assertions in
+$\Gamma$ is very basic.  We do not perform any fixpoint computations
+and only derive trivial loop invariants about variables that are
+unchanged in the loop body.
+
+\pref{fig:ex1} shows a derivation of the bound
+$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
+version of Example~1 in which we increment $x$ by a constant $K>0$ and
+consume $T>0$ resources in each iteration.  The reasoning is similar
+to the one of Example~1 except that we obtain the potential
+$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
+assertions in $\Gamma$ are only used in the rule for the assignment
+\code{x=x+K}.  To the best of our knowledge, no other implemented tool
+for C is currently capable of deriving a tight bound on the cost of
+such a loop.  For $T=1$ (many systems focus on the number of loop
+iterations without a cost model) and $K=10$,
+KoAT~\cite{BrockschmidtEFFG14} computes the bound $|x| + |y| + 10$,
+Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
+LOOPUS~\cite{SinnZV14} computes the bound $y-x-9$.  Only
+PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
+translate the program into a term-rewriting system by hand.
+
+To automate the reasoning, we first introduce an unknown rational
+variable for each factor in the potential functions.  We then use our
+inference rules (see \pref{sec:AAA}) to emit linear constraints on
+these variables that enforce that variable assignments that respect
+the constraints correspond to sound potential annotations.  For
+instance if $K = 1$ in \pref{fig:ex1} then we would have the
+annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
+|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
+p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
+are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
+$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.
+
+\iffull{It might be surprising that we}{We} only track simple linear relations in
+the logical context $\Gamma$.  While it would of course be possible to keep
+track of more sophisticated assertions, this simple form is
+sufficient for the examples we considered and we can efficiently
+decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
+assignment rule.  Nevertheless, this logical part of our analysis
+system is orthogonal to the quantitative part and can be easily
+extended if necessary.
+
+As mentioned earlier, the automatic analysis can handle challenging example
+programs without special tricks or techniques.  Examples
+\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
+work~\cite{GulwaniMC09},  demonstrate that our method can handle
+\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
+derives the same bounds as our analysis but requires heuristics for
+its counter instrumentation.  These loops can also be handled with inference of
+\emph{disjunctive invariants}, but in the abstract interpretation
+community, these invariants are known to be notoriously difficult to
+generate.
+%
+In example \emph{speed\_1} we have one loop that first increments
+variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
+derive the tight bound $|[x, n]| + |[y, m]|$.
+%
+Example \emph{speed\_2} is even trickier, and we found it hard to
+find a bound manually.  However, using potential transfer reasoning as
+in amortized analysis, it is easy to prove the tight bound
+$|[x, n]| + |[z, n]|$.
+
+Example \emph{t08a} shows the ability of the analysis to discover
+interaction between \emph{sequenced loops} through size change of
+variables.  We accurately track the size change of $y$ in the first
+loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
+Furthermore, \emph{t08a} shows again that we do not handle the
+constants $1$ or $0$ in any special way.  In all examples we could
+replace $0$ and $1$ with other constants like in the second loop and
+still derive a tight bound.  \iffull{The only information, that the analyzer
+needs is $y \geq c$ before assigning $y = y - c$.}{}
+%
+Example \emph{t27} shows how amortization can be used to handle
+\emph{interacting nested loops}.  In the outer loop we increment the
+variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
+increment the variable $y$ by $1000$.  Then we non-deterministically
+(expressed by \code{*}) execute an inner loop that decrements $y$ by
+$100$ until $y<100$.  The analysis discovers that only the first
+execution of the inner loop depends on the initial value of $y$.  We
+again derive tight constant factors.
+%
+
+As mentioned, the analysis also handles advanced control flow like
+\code{break} and \code{return} statements, and mutual recursion.
+\pref{fig:ex_rec} contains two mutually-recursive functions with their
+automatically derived tick bounds.  The function \code{count\_down}
+decrements its first argument $x$ until it reaches the second argument
+$y$. It then recursively calls the function \code{count\_up},
+which is dual to \code{count\_down}.  Here, we
+count up $y$ by $2$ and recursively call \code{count\_down}.  Our
+analysis is the only available system that computes a tight
+bound on this example.  \iffull{The analysis amounts to computing the meeting
+point of two trains that approach each other with different speeds.}{}
+% This is comparable to the
+% determination of the point at which to trains meet if they approach
+% each other with different speeds.
+
+\iffull{We only show a small selections of the programs that we can
+  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
+  extended version~\cite{anon_extended}} is a list of more than $30$
+classes of challenging programs that we can automatically analyze.
+\pref{sec:exper} contains a more detailed comparison with other
+tools\iffull{for automatic bound derivation.}{.}
+
+\begin{figure}[t]
+  \centering
+    \begin{minipage}[b]{\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+void count_down (int x,int y) {
+  if (x>y) { tick(1); count_up(x-1,y); } }
+
+void count_up (int x, int y) {
+  if (y+1<x) { tick(1); count_down(x,y+2); } }
+
+   \end{lstlisting}
+
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
+\\[.4\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+\vspace{-1.5ex}
+
+  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
+  \label{fig:ex_rec}
+\end{figure}
+
+\sectskip
+\section{Syntax and Semantics}
+\label{sec:sem}
+\aftersectskip
+
+We implemented our cost semantics and the quantitative Hoare logic in
+Coq for \emph{CompCert Clight}.  Clight is the most abstract
+intermediate language used by CompCert.  Mainly, it is a subset of C
+in which loops can only be exited with a \code{break} statement and
+expressions are free of side effects.
+
+\paragraph{Syntax.}
+
+In this article, we describe our system for a subset of Clight that is
+sufficient to discuss the general ideas.  This subset is given by the
+following grammar.
+%
+\begin{align*}
+S &:= \code{assert}~E
+\mid \code{skip}
+\mid \code{break}
+\mid \code{return}~x
+\mid x \gets E
+\mid x \gets f(x^*)
+\\
+& \mid \code{loop}~S
+\mid \code{if}(E)~S~\code{else}~S
+\mid S;S
+\mid \code{tick}(n)
+\end{align*}
+%
+Expressions $E$ are left abstract in our presentation.  For our
+analysis framework, it is only important that they are side-effect
+free.
+%
+The most notable difference to full Clight is that we can only assign
+to variables and thus do not consider operations that update the heap.
+Moreover, function arguments and return values are assumed to be
+variables.  This is only for simplifying the presentation; in the
+implementation we can deal with heap updates and general function
+calls and returns.  However, we have not implemented our framework for function
+pointers, \code{goto} statements, \code{continue} statements, and
+\code{switch} statements.
+
+We include the built-in primitive $\code{assert}~e$ that terminates
+the program if the argument $e$ evaluates to false and has no effect
+otherwise.  This is useful to express assumptions on the inputs of a
+program for the automatic analysis.  We also add the built-in function
+\code{tick(n)} that can be called with a constant integer $n$ as a
+flexible way to model resource consumption or release (if $n$ is
+negative).
+
+\iffull{
+\begin{figure}[t!]
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~x, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~x, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \fenv f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep{M}
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \cont, \cost')
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_2, \cont, \cost')
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
+}
+%
+\and
+\Rule{S:Tick}
+{ }
+{ (\state, \code{tick}(n), \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
+}
+
+\end{mathpar}
+\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
+\label{fig:opsem}
+\end{figure}
+}{}
+
+\ifshort{
+\begin{figure}[t!]
+\small
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
+}
+\end{mathpar}
+\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
+\label{fig:opsem}
+\end{figure}
+}{}
+
+\ifshort{
+\begin{figure*}[t]
+\small
+\begin{mathpar}
+%
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q }{ \code{skip} }{ Q }
+}
+%
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+%
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Call}
+{
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
+  { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
+\and
+\Rule{L:Extend}
+{ \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+\\
+  \forall f\,P_f\,Q_f.\,
+  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
+  \rightarrow
+  \forall y\,\vec v.\,
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\end{mathpar}
+
+\caption{Rules of the Quantitative Hoare Logic}
+\label{fig:logic}
+\end{figure*}
+}{}
+
+\paragraph{Semantics.}
+
+CompCert Clight's operational semantics is based on small-step
+transitions and continuations.  Expressions---which do not have side
+effects---are evaluated in a big-step fashion.  Here, we describe a
+simplified version of Clight's semantics for the subset we consider.
+
+A program state $\state = (\theta, \gamma)$ is composed of two maps from
+variable names to integers. The first map, $\theta : \text{Locals} \to
+\Z$, assigns integers to local variables of a function, and the second
+map, $\gamma : \text{Globals} \to \Z$, gives values to global variables
+of the program.  In this article, we assume that all values are
+integers but in the implementation we support all data types of
+Clight.
+%
+The evaluation function \evalE{\cdot}{} maps an expression $e \in
+E$ to a value $\evalE{e}{\state} \in \Z$ in the program state
+$\state$.
+
+For simplicity, we assume that local variables and global variables
+are always different.  We just write $\state(x)$ to obtain the value of
+$x$ in program state $\state$. \iffull{Such a lookup is defined as
+$$
+(\theta, \gamma)(x) =
+\left\{
+\begin{array}{lr}
+\theta(x) & \mbox{ if } x \in \dom \theta \\
+\gamma(x) & \mbox{ if } x \in \dom \gamma
+\end{array}
+\right..
+$$}{}
+\iffull{For a program state $\state$, we}{Similarly} we write $\state[x \mapsto v]$
+for the program state that maps $x$ to $v$ and behaves as $\state$ for
+all other variables, regardless whether $x \in \dom \theta$ or $x \in
+\dom \gamma$.
+
+The small-step semantics is standard, except that it tracks the
+resource consumption of a program.  The semantics is parametric in the
+resource of interest for the user of our system.  We achieve this
+independence by parameterizing evaluations with a resource
+\emph{metric} $M$; a tuple of nine rational numbers \ifshort{$M_a$, $M_b$, $M_r$, $\ldots$}{}, one map $M_e :
+\text{Expr} \to \Q$ from expressions to rational numbers, and one map
+$M_t : \Z \to \Q$ from integers to rational numbers.
+%
+\iffull{$$
+M = (
+M_a,
+M_b,
+M_r,
+M_u,
+M_f,
+M_l,
+M_c^1,
+M_c^2,
+M_s,
+M_e(E),
+M_t(n)
+)
+$$}{
+
+}
+Each of these rational numbers indicates the amount of resource
+consumed by a corresponding operation.  If the assigned resource cost
+is negative then it means that resources are released.  The metric
+that we use in the implementation can also depend on other information
+that is statically available such as the name of the called function
+or the number of arguments.
+
+\pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
+rules define a rewrite system for program configurations of the form
+$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
+is the statement being executed, $\cont$ is a continuation that
+describes what remains to be done after the execution of $S$, and
+$\cost \in \Q$ is the non-negative number of resources available for
+further execution.  All rules that can decrease $\cost$ have the
+implicit side condition that the resource quantity available
+\emph{before} the step is non-negative.  This means that we allow
+$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
+transition relation $\smallstep{M}^*$ to indicate that the execution
+ran out of resources.  However, every execution that reaches such a
+state is stuck.
+
+\iffull{
+A continuation $\cont$ represents the context of the execution of a
+given statement.  A continuation can be the empty continuation
+$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
+\cont$, a loop continuation $\Kloop S \cont$, or the continuation
+$\Kcall r \theta \cont$ of a function call.
+$$
+\cont := \Kstop
+\mid \Kseq S \cont
+\mid \Kloop S \cont
+\mid \Kcall r \theta \cont
+$$
+%
+During the execution we assume a fixed function context $\fenv$ that
+maps function names to a list of variables and the statement that
+defines the function body.  It is used in the rule \textsc{S:Call}.
+}{}
+
+The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
+S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
+following.  If the statement $S$ is executed in program state
+$\state$, with continuation $\cont$, and with $\cost$ resources
+available then---after a finite number of steps---the evaluation will
+reach the new machine state $(\state', S', \cont',\cost')$ and there
+are $\cost'$ resources available.  If $\cost' \ge 0$ then the
+execution did not run out of resources and the resource consumption up
+to this point is $\cost - \cost'$.  If this difference is negative
+then resource became available during the execution.  If however
+$\cost' < 0$ then the execution ran out of resources and is stuck.
+The cost of the execution is then $\cost \ge 0$.
+
+\iffull{
+\pref{lem:sem1} summarizes the main properties of the resource
+counter of the semantics.  It can be proved by induction on the number
+of steps.
+\begin{lemma}
+\label{lem:sem1}
+\itemskip
+  \begin{enumerate}
+  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
+    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
+\itemskipIn
+  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
+    \cont', d')$ then $\cost - \cost' = d - d'$.
+  \end{enumerate}
+\itemskip
+\end{lemma}
+%
+If we assume that we have a \emph{positive resource metric}, that is,
+a metric that assigns a positive cost to each evaluation step then we
+can link bounds on the resource consumption under the metric to
+termination.  This is formalized by \pref{lem:sem_term}.
+%
+\begin{lemma}
+ \label{lem:sem_term}
+ Let $M$ be a positive metric and assume that there exists $b \in
+ \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
+ (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
+ Then there exists $B \in \N$ such that $(\state, S, \cont, d)
+ \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
+ and $\state''$ implies $n \le B$.
+\end{lemma}
+}
+{
+The extended version of this article~\cite{anon_extended} contains all rules and
+lemmas that state the main properties of the cost semantics.
+}
+
+\iffull{
+\begin{figure*}[t]
+\begin{mathpar}
+%
+\Rule{L:Skip}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q }{ \code{skip} }{ Q }
+}
+%
+\and
+\Rule{L:Break}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { M_b + B }{ \code{break} }{ Q }
+}
+%
+\and
+\Rule{L:Return}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { R\,(\state(x)) }{ \code{return}~x }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Call}
+{
+  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
+\htriple
+  { M_f + P }
+  { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
+  { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
+\and
+\Rule{L:Extend}
+{ \Delta \cup \Delta'; B; R \vdash
+  \htriple{ P }{ S }{ Q }
+\\
+  \forall f\,P_f\,Q_f.\,
+  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
+  \rightarrow
+  \forall y\,\vec v.\,
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\end{mathpar}
+
+\caption{Rules of the Quantitative Hoare Logic}
+\label{fig:logic}
+\end{figure*}
+}{}
+
+\sectskip
+\section{Quantitative Hoare Logic}
+\label{sec:logic}
+\aftersectskip
+
+In this section we describe a simplified version of the quantitative
+Hoare logic that we use in Coq to interactively prove resource bounds.
+%
+We generalize classic Hoare logic to express not only classical
+boolean-valued assertions but also assertions that talk about the
+future resource usage.  Instead of the usual assertions $P : \State
+\to \mathit{bool}$ of Hoare logic we use assertions
+$$
+P : \State \to \Qplusz \cup \{ \infty \} \; .
+$$
+This can be understood as a refinement of boolean assertions where
+$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
+We write $\Assn$ for $\State \to \Qplusz \cup \{ \infty \}$ and $\bot$ for
+$\lambda \state . \, \infty$.  We sometimes call assertions
+\emph{potential functions}.  To use Coq's support for propositional
+reasoning, assertions have the type $\State \to \Qplusz \to \mathrm{Prop}$
+in the implementation.  For a given $\state \in \State$, such an
+assertion can be seen as a set $B \subseteq \Qplusz$ of valid bounds.
+However, we find the presentation in this article easier to read.
+
+Due to \code{break} and \code{return} statements of Clight, there are
+different possible ways to exit a block of code.  We also have to keep
+track of the resource specifications of functions.  To account for
+this in the logic, our quantitative Hoare triples have the form
+$$
+\Delta; B; R \vdash
+\htriple
+  { Q }
+  { S }
+  { Q' } \; .
+$$
+The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
+two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
+Hoare logic and the intuitive meaning is as follows.  If $S$ is
+executed with starting state $\state$, the empty continuation
+$\Kstop$, and at least $P(\state)$ resources available then the
+evaluation does not run out of resources and there are at least
+$Q(\state')$ resources left if the evaluation terminates in state
+$\state'$.  The assertion $B : \Assn$ provides the
+postcondition for the case in which the code block $S$ is exited by a
+\code{break} statement.  So if the execution is terminated in state
+$\state'$ with a \code{break} then $B(\state')$ resources are
+available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
+case in which the code block $S$ is exited by a $\code{return}~x$
+statement.  The integer argument of $R$ is the return value.
+Finally, the function context of judgements that we write $\Delta$ is
+a mapping from function names to specifications of the form
+$$
+  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
+$$
+The assertion $P_f \, z \, \vec v$ is the precondition of the function
+$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
+both parameterized by an arbitary logical variable $z$ (which can be a
+tuple) that relates the function arguments with the return value.  The
+precondition also depends on $\vec v$, the values of the arguments
+at the function invocation.  Similarly, the postcondition depends on
+the return value $v$ of the function.  The use of logical variables to
+express relations between different states of an execution is a
+standard technique of Hoare logic.
+%
+To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
+the local variables on the stack, that is, $\forall z \, \vec v
+\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
+(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.
+
+For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
+program state $\state$ $P(\state) \ge Q(\state)$.
+
+\paragraph{Rules of the Quantitative Logic.}
+
+\pref{fig:logic} shows the inference rules of the quantitative logic.
+The rules are slightly simplified in comparison to the implemented
+rules in Coq.  The main difference is that the presented version does
+not formalize the heap operations.
+
+\iffull{In the rule {\sc L:Skip}, we do not have to account for any resource
+consumption.  As a result, the precondition $Q$ can be any (potential)
+function and we only have to make sure that we do not end up with more
+potential.  Since the execution of \code{skip} leaves the program
+state unchanged, we can simply use the precondition as postcondition.
+The potential functions $B$ for the \code{break} and $R$ for the
+\code{return} part of the postcondition are not reachable and can
+therefore be arbitrary.}{}
+
+In the rule {\sc L:Assert}, we use the notation
+$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
+require potential $Q + M_a$ in the precondition if $e$ evaluates to
+$\mathit{true}$ in the current program state.  If $e$ evaluates to
+$\mathit{false}$ then the potential in the precondition can be
+arbitrary since the program will be terminated.
+
+In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
+be arbitrary since it is unreachable.  Instead, we have to justify the
+potential functions $B$ and $R$ that hold after a \code{return} and a
+\code{break}, respectively.  In {\sc L:Break}, we require to have
+potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
+cost of \code{break} and $B$ to pay for the potential after the
+\code{break}.  In {\sc L:Return} we only require to have potential
+$R$ in the precondition to pay for the potential after the
+\code{return}.  The reason is that we found it to be more convenient
+to account for the execution cost $M_r$ of the return in rule {\sc
+  L:Call}\iffull{ for function calls.}{.}
+
+\iffull{The rule {\sc L:Update} is the standard assignment rule of Hoare
+logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
+the precondition we ensure that $Q$ evaluates to the same number as in
+the postcondition.  We also require that have we the constant
+potential $M_u + M_e(e)$ available in the precondition to pay for the
+cost of the evaluation of $e$ and the update.
+
+The rule {\sc L:Seq} rule is crucial to understand how the
+quantitative Hoare logic works.  To account for early exits of
+statements, we must ensure in the \code{break} part $B$ of $S_1$'s
+judgement that the \code{break} part $B$ of of $S_1;S_2$ holds. The
+same is true for the return part $R$ of the judgements for $S_1$ and
+$S_2$.  The interaction between the actual pre- and postconditions is
+analogous to standard Hoare logic.  In the postcondition of $S_1$ we
+account for the cost $M_s$ of the execution of the sequence.
+
+}
+{
+The rules {\sc L:Update} and {\sc L:Seq} correspond to the respective
+rules of standard Hoare logic.
+}
+In the rule {\sc L:Loop}, the \code{break} part of the
+loop body $S$ becomes the postcondition of the loop statement. We use
+an arbitrary $B$ as the \code{break} part of the judgement for
+$\code{loop}\, S$ since its operational semantics ensures that it can
+only terminate with a \code{skip} or a \code{return}.  The
+precondition $I$ of the loop is the loop invariant.  \iffull{That is why we
+require to have potential $I$ available in the precondition of the
+loop body $S$.}{}  In the postcondition of $S$, the potential must be
+sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
+iteration.
+
+\iffull{
+The {\sc L:If} is similar to the rule for the conditional in classic
+Hoare logic.  In the preconditions of the judgments for the two
+branches $S_1$ and $S_2$ we lift the boolean assertions
+$\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
+quantitative assertions.  In the precondition of the rule {\sc L:Tick}
+we account for the execution cost $M_t(n)$ of the \code{tick}
+statement that depends on the integer $n$.
+}{}
+
+The rule {\sc L:Call} accounts for the execution cost of both $M_f$
+for function calls and $M_r$ for \code{return} statements.
+\iffull{This justifies that the {\sc L:Return} rule does not account
+  for resource consumption.}{} The pre- and postcondition $P_f$ and
+$Q_f$ are taken from the function context $\Delta$.  The assertions in
+the context are parametric with respect to both the values of the
+function arguments and the return value. This allows us to specify a
+bound for a function whose resource consumption depends on its
+arguments.  The arguments are instantiated by the call rule
+using the result of the evaluation of the argument variables in the
+current state.  \iffull{Recall that we require that $P_f$ and $Q_f$ do
+  not depend on the local variables on the call stack, that is,
+  $\forall z \, \vec v \, \theta \, \theta' \, \gamma . \, P_f \, z \,
+  \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
+  (\theta',\gamma)$.}{} To transfer potential that depends on local
+variables of the callee from the precondition $P$ to the postcondition
+$Q$, we use an assertion $A : \Assn$ that is independent of global
+variables, that is, $\forall \theta \, \gamma \, \gamma' . \, A
+(\theta,\gamma) = A(\theta,\gamma')$.  It is still possible to express
+relations between global and local variables using logical
+variables\iffull{ (see the following paragraph for details).}{.}
+
+Finally, we describe the rules which are not syntax directed.  There
+are two weakening rules available in the quantitative Hoare logic.
+The framing rule {\sc L:Frame} is designed to weaken a statement by
+stating that if $S$ needs $P$ resources to run and leaves $Q$
+resources available after its execution, then it can very well run
+with $P + c$ resources and return $Q + c$ resources.  The consequence
+{\sc L:Conseq} rule is directly imported from classical Hoare logic
+except that instead of using the logical implication $\Rightarrow$ we
+use the quantitative $\models$ that point-wise applies $\ge$.  \iffull{This
+rule indeed weakens the statement since it requires more resource to
+run the statement and yields less than what has been proved to be
+available after its termination.}{}
+
+\ifx\fullversion\undefined{}\else{
+\paragraph{Logical Variables and Shallow Embedding.}
+
+When specifying a program, it is often necessary to relate certain
+invariants in the pre- and postcondition.  A standard solution of Hoare
+logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
+additional variables (also called auxiliary state in the literature)
+are constant across the derivation.  For example, if $Z$ is such
+a logical variable, we can specify the function $\code{double}()$
+which doubles the global variable $x$ as
+$$
+  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
+$$
+%
+When formalizing Hoare logics in a proof assistant one can either
+fully specify the syntax and semantics of assertions and hence get a
+deep embedding, or use the assertions of the host theory to get a
+shallow embedding.  Because of its flexibility, we used the latter
+approach in our development.  This choice makes it possible to have
+logical variables almost for free: we can simply use the variable and
+binding mechanisms of Coq, our host theory.  When an logical variable
+is needed we introduce it using a universal quantifier of Coq before
+using the logic to derive triples.  For example, the Coq theorem for
+the above example would look as follows.
+%
+\begin{lstlisting}
+Theorem dbl_triple: forall Z,
+  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
+\end{lstlisting}
+%
+However, this trick alone is often not sufficient when working with a
+recursive function $f$.  In that case we apply the {\sc L:Extend} rule
+of the logic.  First we add a specification $\forall z\,\vec
+v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
+$\Delta$.  Then we proceed to prove the function body $S_f$ with this
+induction hypothesis.  In this process it can be the case that we have
+to use the induction hypothesis with a different value of a logical
+variable (e.g., because the values of the arguments in the recursive
+call differ from the values of the arguments of the callee).  To cope
+with this problem, assumptions in the function context $\Delta$ are
+universally quantified over logical variables.  The {\sc L:Extend}
+rule uses the host proof assistant to require that the triple on $f$'s
+body is proved for every possible logical variable $y$.
+}\fi
+
+\paragraph{Using the Quantitative Logic.}
+
+In the following we demonstrate the use of the logic with two example
+derivations.
+
+In the example in~\pref{fig:xmplmax} we derive a precise runtime
+bound on a program that searches a maximal element in an array.  The
+cost metric that we use simply counts the assignments performed by the
+program.  Hence, the resource cost is closely related to the
+number of times the test \code{a[i] > m} is true during the
+execution.
+%
+If we define
+$$
+ A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
+$$
+where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
+the number of ``maximum candidates'' in the array \code{a} seen
+by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
+So any automated tool would at best derive the linear bound $2 \cdot
+N$ for that program.  But with the expressivity of our logic it is
+possible to use the previous set cardinal directly and precisely tie
+the bound to the initial contents of the array.
+%
+The non-trivial part of this derivation is finding the loop invariant
+$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
+When the condition \code{a[i] > m} is true, we know that we
+encountered a ``maximum so far'' because $m$ is a maximal element of
+\code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
+potential unit to pay for the assignment.  In the other case, no
+maximum so far is encountered so $A(i) = A(i+1)$.
+% As a side remark, if $K \ge 0$, it is possible to
+% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
+% of the same program.  It can be done either by adapting the proof or by
+% applying the {\sc L:Frame} rule one top of the triple already derived.
+% More generally, without function calls, the {\sc L:Frame}
+% rule is admissible in our system.
+
+The example in~\pref{fig:xmplbs} shows a use case for logical
+variables as well as a metric for stack consumption. In a stack
+metric, we account a constant cost for a function call ($M_f>0$) that
+is returned after the call ($M_r = -M_f <0$).  All other resource cost
+are $0$.  We are interested in showing that a binary search function
+$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
+variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
+  + Z \cdot M_\code{bsearch}}{}{Z \cdot M_\code{bsearch}}$ to express that the stack required by the
+function is returned after the call.
+%
+The critical step in the proof is the application of the
+{\sc L:Call} rule to the recursive call.  At this point
+the context $\Delta$ contains the specification
+$
+  \forall y \, (x,l,h) \, \_.
+  ( (\lambda y \, (\_, l, h) .\, (y = \log_2(h {-} l)) {+} y {\cdot} M_\code{bsearch})
+  , (\lambda y \, \_ .\, y {\cdot} M_\code{bsearch})
+  )
+$.
+%
+Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
+- 1$, and because $M_f = M_\code{bsearch}$ and $M_r =
+-M_\code{bsearch}$ in the stack metric.  The rest of the proof does
+not involve any resource manipulation and is just bookkeeping of
+logical assertions.
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue} \{ A(0) + N \} }$
+i=1; m=a[0];
+${\color{blue}%
+  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
+while (i < N) {
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
+  if (a[i] > m)
+    m=a[i];
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
+  i=i+1;
+  ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
+} ${\color{blue} \{ 0 \} }$
+\end{lstlisting}
+\caption{Example derivation where we wrote $A(i)$
+  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
+  the metric used here assigns a cost of 1 to every assignment
+  and 0 to all other operations.
+  }
+\label{fig:xmplmax}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}%
+  \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
+bsearch(x,l,h) {
+  if (h-l > 1) {
+    ${\color{blue}%
+      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
+    m = h + (h-l)/2;
+    ${\color{blue}%
+      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
+    if (a[m]>x) h=m; else l=m;
+    ${\color{blue}%
+      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
+    l = bsearch(x,l,h); }
+  ${\color{blue}%
+    \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
+  return l;
+} ${\color{blue} \{ Z \cdot M_\code{bsearch} \} }$
+\end{lstlisting}
+\caption{Example derivation of a stack usage bound for a binary
+  search program.  The used resource metric defines the cost $M_\code{bsearch}$ before
+  the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
+  frame size of the function $\code{bsearch}$.  $Z$ is a
+  logical variable.
+  }
+\label{fig:xmplbs}
+\end{figure}
+
+\paragraph{Soundness of Quantitative Hoare Triples.}
+
+We already gave an intuition of the meaning of judgements
+derived in the logic.  To make it formal, we define the
+\emph{resource safety} $\safe n P S \cont$ of an assertion
+and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
+\begin{align*}
+  (m {\le} n \land P(\state) {\le} \cost \land
+    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
+  \implies \cost' {\ge} 0.
+\end{align*}
+This predicate is step indexed by an integer $n$ that is used for
+induction in the soundness proof for the function-call and loop cases.
+The constraint $\cost' \geq 0$ imposed by the definition ensures that
+the program does not stop because of a resource error (recall that a
+negative resource counter on the right-hand side is a resource
+failure).  However, it does not rule out memory safety errors or
+assertion failures. This is because our logic does not prove any
+safety or correctness theorems but only focuses on resource usage.
+%
+An interesting detail in the definition is the natural number $m$.  We
+simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
+\cont $.  This would not be the case if we replaced all occurrences of
+$m$ by $n$ in the definition\iffull{ of validity.}{.}
+
+The resource safety of a continuation $\cont$ is defined using three
+assertions, one for each of the possible outcomes of a program
+statement.  \iffull{We define it as follows.
+\begin{align*}
+\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
+& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
+& & \land& \safe n Q {\code{skip}} \cont
+\end{align*}}
+{
+We define $\safeK n B R Q \cont \mathop{:=}
+\safe n B {\code{break}} \cont \, \land \,
+\safe n Q {\code{skip}} \cont \, \land \,
+\safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont$.
+
+}
+%
+We can now define the \emph{semantic validity} of a judgement $B; R
+\vdash \htriple P S Q$ of the quantitative logic without function
+context as $\valid n B R P S Q :=$
+\begin{align*}
+& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
+& \implies \safe m {P+x} S \cont \; .
+\end{align*}
+Note how the validity of a triple embeds the frame rule of
+our logic. This refinement is necessary to have a stronger
+induction hypothesis available during the proof.
+%
+We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
+S Q$ implies $\valid n B R P S Q$.
+
+Using the semantic validity of triples we define the validity
+of a function context $\Delta$, written $\validC n \Delta$, as
+\begin{align*}
+  &\forall f .\, \Delta(f) =
+    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
+    \implies \\
+  &\qquad \forall z \, \vec v .\,
+  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
+\end{align*}
+where $S_f$ is the body of the function $f$. \iffull{A full
+judgement that mentions a non-empty function context
+$\Delta$ is in fact a guarded statement: it makes
+assumptions on some functions' behavior.}{}  The predicate
+$\validC n \Delta$ gives the precise meaning of the
+assumptions made.  It is also step-indexed to prove the
+soundness of the {\sc L:Extend} rule by induction.
+%
+We are now able to state the soundness of the quantitative logic.
+%
+\begin{theorem}[Soundness of the logic]
+  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
+  $
+    \forall n .\, \validC n \Delta
+      \implies \valid {n {+} 1} B R P S Q.
+  $
+\end{theorem}
+%
+\noindent
+The difference $\delta = 1$ between the index in the triple
+validity and the one in context validity arises from
+the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
+{\sc L:Call}, the language semantics makes one step and
+proceeds with the function body, so we must have
+$\delta \le 1$ to use the assumptions in $\Delta$.
+For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
+is a valid context for $n$ steps.  The induction hypothesis
+in that case says that if $\Delta \cup \Delta'$ is valid
+for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
+So if we want to solve this goal by induction, it is
+necessary that $\delta \ge 1$.  These two constraints force
+$\delta$ to be exactly one in the theorem statement.
+
+Assume that $S$ is a complete program and $\Delta$ is empty.  By
+expanding the definitions we see that $\Delta$ is valid
+for every $n$ and that $\Kstop$ is safe for every $n$. So
+we derive
+$$
+\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
+$$
+This means that from any starting state $\state$, $P(\state)$
+provides enough resources for any run of the program $S$.  \iffull{This
+setting is actually the main use case of the previous theorem
+which is stated as a stronger result to allow a proof by
+induction.}{}
+
+\sectskip
 \section{Experimental Evaluation}

### 2014-11-10 paper/main.tex
minor updates at the first part

@@
 
-We have evaluated the automatic analysis with system code and examples
-from the literature.  Our tool automatically derived global resource
-bounds for more than XXX lines of C code from the XXX benchmark suite.
+We have implemented the automatic analysis in the tool
+\toolname and experimentally evaluated its effectiveness by
+analyzing system code and examples from the literature.
+\toolname has automatically derived global resource bounds for
+more than XXX lines of C code from the XXX benchmark suite.
 %
@@
 In this section, we informally introduce the automatic amortized
-analysis (AAA) for Clight programs and demonstrate the technique with
+analysis for Clight programs and demonstrate the technique with
 example bounds that we derive step-by-step following our analysis
 system.
-
-\jan{Watch out with the headings. Either upper case or lower case, but
-  don't mix.}
 
@@
 numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a program
-state such that $(S, \state) \Downarrow_n \state'$.  This gives that
+state such that $(S, \state) \Downarrow_n \state'$.  Then
 $\Phi(\state)$ is a valid resource bound.
@@
 preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
-AAA is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
+\toolname is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
 + \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
@@
 \hline
-AAA & $1 {+} 2|\inter 0 n|$ \\
+\toolname & $1 {+} 2|\inter 0 n|$ \\
 Rank & $O(n^2)$ \\
@@
 because most other tools only bound this specific cost. In
-\pref{fig:compar}, we show the bounds we derived (AAA) together with
+\pref{fig:compar}, we show the bounds we derived (\toolname) together with
 the bounds derived by LOOPUS~\cite{SinnZV14} and
@@
 \begin{tabular}{r|c|c|c|c|c}
-& KoAT & Rank & LOOPUS & SPEED & AAA \\
+& KoAT & Rank & LOOPUS & SPEED & \toolname \\
 \hline
@@
 SPEED &
-AAA
+\toolname
 \\

### 2014-11-10 paper/main.tex
trying to clean up section 2

@@
 in the first category derive impressive bounds for
-numerical imperative programs, but they are not fully compositional.
+numerical imperative programs, but are not fully compositional.
 This is problematic if one needs to derive global whole-program bounds.
@@
 %
+This rule already shows a departure from classical techniques that are
+based on ranking functions.  Reasoning with two potential functions
+promotes compositional reasoning by considering the sequencing of
+program operations.  In the previous rule, $\Phi$ gives a bound for
+$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
+was derived on $S_1$ only.
+%
 Similarly, other language constructs lead to rules for the potential
@@
 
-The considerations above already show a departure from classical
-techniques for imperative programs.  Indeed, reasoning with two
-potential functions is more compositional than using
-ranking-functions because it forces to consider the sequencing of program
-operations.  In the previous rule, $\Phi$ gives a bound for
-$S_1; S_2$ through the intermediate potential $\Phi'$, even though
-it was derived on $S_1$ only.
-
 % Potential functions can access the whole program state, so this
@@
 
-\subsection{Automatic Amortized Analysis}
-
-The reasoning presented above is very general and flexible.  But in
-practice, to make the reasoning tractable on large code bases, we would
+\subsection{Automating  Amortized Analysis}
+
+To make amortized reasoning feasible on large code bases, we would
 like to automatically infer the potential functions.  The first step
-towards this goal is to reduce our search space by fixing the shape
-of potential functions $\Phi$.  We settle on a form that allows
-\emph{linear programming} (LP) to help inference:
-%
-\jan{I wouldn't say that LP is the reason we picked this format. We
-  rather picked it because it covers the most common cases.}
+towards this goal is to reduce our search space by fixing the shape of
+potential functions $\Phi$.  In this paper, we focus on a form that
+can express tight bounds for many typical programs and allows
+inference view \emph{linear programming}:
 %
 $$
-\Phi(\state) = q_0 + \sum_{x,y \in \dom(\state)}
+\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}}
   q_{(x,y)} \cdot |[\state(x),\state(y)]|
 $$
-where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Here, we assume
-that the program state is a simple map from program variables to integers
+where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  For now, we assume
+that a program state is a simple map from program variables to integers
 and we treat constants as program variables (that cannot be changed
 by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
-the sum.  We then develop an inference rule for each syntactic construct
-that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
-the previous section.
-
-This idea is best explained by example.  If we use the tick metric that
-assigns cost $n$ to the function call \code{tick(n)} and cost $0$ to all
-other operations then the cost of the following example can be bounded
-by $|[x,y]|$.\footnote{Note that there are no restrictions on the signs
-of the input variables in the examples.}
+the sum.  % We then develop an inference rule for each syntactic construct
+% that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
+% the previous section.
+
+\begin{figure}[t]
+  \centering
+\begin{lstlisting}[mathescape]
+${\color{blue}%
+  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+while (x+K<=y) {
+  ${\color{blue}%
+  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  x=x+K;
+  ${\color{blue}%
+  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  tick(T);
+  ${\color{blue}%
+  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+}
+${\color{blue}%
+ \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+\end{lstlisting}
+  \caption{Automatic derivation of a tight bound on the number of
+    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
+    $T>0$ are not program variables but denote concrete constants in
+    the program.  The reasoning works uniformly for all such $K$ and
+    $T$.}
+  \label{fig:ex1}
+\end{figure}
+
+\newlength{\progwidth}
+
+\begin{figure*}[t!]
+ \setlength{\progwidth}{.22\linewidth}
+  \centering
+\hspace{-0.4cm}
+  \begin{minipage}[b]{.18\linewidth}
+    \begin{center}
+   \begin{lstlisting}[]
+while (n>x) {
+  ${\color{blue}%
+  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+  if (m>y)
+    ${\color{blue}%
+    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
+    y=y+1;
+    ${\color{blue}%
+    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  else
+    ${\color{blue}%
+    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
+}
+   \end{lstlisting}
+
+$|[x, n]| + |[y, m]|$
+\\[.4\baselineskip]
+      {\bf speed\_1}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+while (x<n) {
+  ${\color{blue}%
+  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+  if (z>x)
+    ${\color{blue}%
+    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    x=x+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  else
+    ${\color{blue}%
+    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    z=z+1;
+    ${\color{blue}%
+    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
+  }
+   \end{lstlisting}
+
+$|[x, n]| + |[z, n]|$
+\\[.4\baselineskip]
+      {\bf speed\_2}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (z-y>0) {
+  ${\color{blue}%
+  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y=y+1;
+  ${\color{blue}%
+  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(3);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+${\color{blue}%
+\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+while (y>9) {
+  ${\color{blue}%
+  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  y=y-10;
+  ${\color{blue}%
+  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  tick(1);
+  ${\color{blue}%
+  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+}
+   \end{lstlisting}
+
+$3.1|[y,z]| + 0.1|[0,y]|$
+\\[.4\baselineskip]
+      {\bf t08a}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+while (n<0) {
+  ${\color{blue}%
+  n{<}0; \,  P(n,y) \vdash }$
+  n=n+1;
+  ${\color{blue}%
+  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
+  y=y+1000;
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  while (y>=100 && *){
+    ${\color{blue}%
+    y{>}99; \, 9 {+} P(n,y) \vdash }$
+    y=y-100;
+    ${\color{blue}%
+    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
+    tick(5);
+    ${\color{blue}%
+    \dashv \cdot; \, 9 {+} P(n,y) }$
+  }
+  ${\color{blue}%
+  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  tick(9);
+  ${\color{blue}%
+  \dashv \cdot; \, P(n,y) }$
+}
+   \end{lstlisting}
+$59|[n,0]| {+} 0.05|[0,y]|$
+\\[.4\baselineskip]
+      {\bf t27}
+    \end{center}
+  \end{minipage}
+\vspace{1ex}
+\caption{Derivations of bounds on the number of ticks for challenging
+  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
+  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
+  \emph{t08a} contains \emph{sequential loops} so that the iterations
+  of the second loop depend on the first, and \emph{t27} contains
+  interacting \emph{nested loops}. In the potential functions, we only
+  mention the non-zero terms and in the logical context $\Gamma$ we
+  only mention assertions that we actually use in the reasoning. In
+  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
+  0.05|[0,y]|$.}
+  \label{fig:ex_list}
+\end{figure*}
+
+The automatic derivation of a resource bound using these potential
+functions is best explained by example.  If we use the tick metric
+that assigns cost $n$ to the function call \code{tick(n)} and cost $0$
+to all other operations then the cost of the following example can be
+bounded by $|[x,y]|$.% \footnote{Note that there are no restrictions on
+  % the signs of the input variables in the examples.}
 \begin{lstlisting}[basicstyle=\tt\small]
@@
 \end{lstlisting}
-To derive this bound in our automatic amortized analysis, we start with
+To derive this bound in our amortized analysis, we start with
 the initial potential $\Phi_0 = |[x,y]|$ (short for $\Phi_0(\state)
@@
 
-\jan{I would add another subsection here to explain how we automate
-  the analysis with LP solving.  Message: Efficient reduction to LP
-  solving and LP solvers are great and super fast. Maybe mention
-  'network problems'.}
-
-\subsection{Compositionality in Practice}
+\pref{fig:ex1} shows a derivation of the bound
+$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
+version of Example~1 in which we increment $x$ by a constant $K>0$ and
+consume $T>0$ resources in each iteration.  The reasoning is similar
+to the one of Example~1 except that we obtain the potential
+$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
+assertions in $\Gamma$ are only used in the rule for the assignment
+\code{x=x+K}.  To the best of our knowledge, no other implemented tool
+for C is currently capable of deriving a tight bound on the cost of
+such a loop.  For $T=1$ (many systems focus on the number of loop
+iterations without a cost model) and $K=10$, KoAT computes the bound
+$|x| + |y| + 10$, Rank computes the bound $y-x-7$, and LOOPUS computes
+the bound $y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if
+we translate the program into a term-rewriting system by hand.
+
+To automate the reasoning, we first introduce an unknown rational
+variable for each factor in the potential functions.  We then use our
+inference rules (see \pref{sec:AAA}) to emit linear constraints on
+these variables that enforce that variable assignments that respect
+the constraints correspond to sound potential annotations.  For
+instance if $K = 1$ in \pref{fig:ex1} then we would have the
+annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
+|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
+p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
+are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
+$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
+q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
+the coefficients of the initial potential function. LP solvers are
+among the most mature constraint solving tools. As a result, the bound
+derivation works very efficiently in practice.
+
+\iffull{It might be surprising that we}{We} only track simple linear relations in
+the logical context $\Gamma$.  While it would of course be possible to keep
+track of more sophisticated assertions, this simple form is
+sufficient for the examples we considered and we can efficiently
+decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
+assignment rule.  Nevertheless, this logical part of our analysis
+system is orthogonal to the quantitative part and can be easily
+extended if necessary.
+
+\ifx\fullversion\undefined{}\else{
+Note that our notion of a potential function is a generalization of the
+concept of a ranking function.  A potential function can be used like
+a ranking function if we use the tick metric and add the statement
+\code{tick(1)} to every back edge of the program (loops and function
+calls).  However, a potential function is more flexible.  For example,
+we can use a potential function to prove that Example~2 does not
+consume any resources in the tick metric.
+\begin{lstlisting}[basicstyle=\tt\small]
+#\!\!\!\!\!# while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
+\end{lstlisting}
+Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
+In both cases, we reason exactly like in the first version of the
+while loop to prove the bound.  Of course, such loops with different
+tick annotations can be seamlessly combined in a larger program.
+\begin{lstlisting}[basicstyle=\tt\small]
+while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
+\end{lstlisting}
+}\fi
+
+\subsection{Advantages of the Potential Method }
+
+The potential based approach addresses several challenges in resource
+bound analysis. This includes amortized bounds in which individual
+loop iterations have different resource consumption, sequential loops
+or function calls that require the tracking of size changes of
+variables, and nested loops that interact by modifying the same set of
+variables.
+
+\paragraph{Challenging Loop Patterns} As mentioned earlier, the
+automatic analysis can handle challenging example programs without
+special tricks or techniques.  Examples \emph{speed\_1} and
+\emph{speed\_2}, that are taken from previous work~\cite{GulwaniMC09},
+demonstrate that our method can handle \emph{tricky iteration
+  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
+bounds as our analysis but requires heuristics for its counter
+instrumentation.  These loops can also be handled with inference of
+\emph{disjunctive invariants}, but in the abstract interpretation
+community, these invariants are known to be notoriously difficult to
+generate.
+%
+In example \emph{speed\_1} we have one loop that first increments
+variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
+derive the tight bound $|[x, n]| + |[y, m]|$.
+%
+Example \emph{speed\_2} is even trickier, and we found it hard to
+find a bound manually.  However, using potential transfer reasoning as
+in amortized analysis, it is easy to prove the tight bound
+$|[x, n]| + |[z, n]|$.
+
+Example \emph{t08a} shows the ability of the analysis to discover
+interaction between \emph{sequenced loops} through size change of
+variables.  We accurately track the size change of $y$ in the first
+loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
+Furthermore, \emph{t08a} shows again that we do not handle the
+constants $1$ or $0$ in any special way.  In all examples we could
+replace $0$ and $1$ with other constants like in the second loop and
+still derive a tight bound.  \iffull{The only information, that the analyzer
+needs is $y \geq c$ before assigning $y = y - c$.}{}
+%
+Example \emph{t27} shows how amortization can be used to handle
+\emph{interacting nested loops}.  In the outer loop we increment the
+variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
+increment the variable $y$ by $1000$.  Then we non-deterministically
+(expressed by \code{*}) execute an inner loop that decrements $y$ by
+$100$ until $y<100$.  The analysis discovers that only the first
+execution of the inner loop depends on the initial value of $y$.  We
+again derive tight constant factors.
+%
+
+As mentioned, the analysis also handles advanced control flow like
+\code{break} and \code{return} statements, and mutual recursion.
+\pref{fig:ex_rec} contains two mutually-recursive functions with their
+automatically derived tick bounds.  The function \code{count\_down}
+decrements its first argument $x$ until it reaches the second argument
+$y$. It then recursively calls the function \code{count\_up},
+which is dual to \code{count\_down}.  Here, we
+count up $y$ by $2$ and recursively call \code{count\_down}.  Our
+analysis is the only available system that computes a tight
+bound on this example.  \iffull{The analysis amounts to computing the meeting
+point of two trains that approach each other with different speeds.}{}
+% This is comparable to the
+% determination of the point at which to trains meet if they approach
+% each other with different speeds.
+
+\iffull{We only show a small selections of the programs that we can
+  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
+  extended version~\cite{anon_extended}} is a list of more than $30$
+classes of challenging programs that we can automatically analyze.
+\pref{sec:exper} contains a more detailed comparison with other
+tools\iffull{for automatic bound derivation.}{.}
+
+\begin{figure}[t]
+  \centering
+    \begin{minipage}[b]{\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+void count_down (int x,int y) {
+  if (x>y) { tick(1); count_up(x-1,y); } }
+
+void count_up (int x, int y) {
+  if (y+1<x) { tick(1); count_down(x,y+2); } }
+
+   \end{lstlisting}
+
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
+\\[.4\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+\vspace{-1.5ex}
+
+  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
+  \label{fig:ex_rec}
+\end{figure}
+
+\paragraph{Compositionality}
 
@@
 or additively, resulting in a loose quadratic bound or no bound at all.
+
+\paragraph{User Guidance for Non-Linear Bounds}
 
@@
 \sectskip
-\section{Overview and Examples}
-\label{sec:inform}
-\aftersectskip
-
-In this section, we informally introduce the quantitative program
-logic and the automatic amortized analysis for Clight programs.
-% The purpose of this section is to build some intuition that should
-% make the following sections easy to read.
-
-\subsection{Quantitative Hoare Logic}
-
-The idea that drives the design of our framework is amortized
-analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
-starting state $\state$ and consumes $n$ resource units of some
-user-defined quantity.  We denote that by writing $(S, \state)
-\Downarrow_n \state'$ where $\state'$ is the program state after the
-execution.  The basic idea of amortized analysis is to define a
-\emph{potential function} $\Phi$ that maps program states to non-negative
-numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
-program state such that $(S, \state) \Downarrow_n \state'$ to prove
-a bound on the resource usage.
-
-To obtain a compositional reasnoning we also have to take into account
-the state resulting from a program's execution.  We thus use two
-potential functions, one that applies before the execution, and one
-that applies after.  The two functions must respect the relation
-$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
-$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
-$\Phi(\state)$ must provide enough \emph{potential} for both, paying
-for the resource cost of the computation and paying for the potential
-$\Phi'(\state')$ on the resulting state $\state'$. That way, if
-$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
-\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
-$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
-$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
-potential function $\Phi$ provides an upper bound on the resource
-consumption of the whole program.  What we have observed is that, if
-we define $\htriple{\Phi} {S}{\Phi'}$ to mean
-$$
-\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
-n + \Phi'(\state') \; ,
-$$
-then we get the following familiar looking rule.
-$$
-\RuleNolabel
-{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
-{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
-$$
-%
-Similarly, other language constructs lead to rules for the potential
-functions that look very similar to Hoare logic or effect system
-rules.  These rules enable interactive reasoning about resource usage
-in a flexible and compositional way, which, as a side effect, produces
-a certificate for the derived resource bound.
-
-% We have implemented and proved sound such a quantitative Hoare for
-% CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
-% embedding in Coq, every function that can be expressed in Coq can in
-% principle be used as a resource bound.
-
-It is also possible to incorporate boolean conditions into a
-bound to express that the bound is only valid for a certain class of
-inputs.  To this end, we allow the potential function $\Phi$ in the
-pre- and postconditions to take non-negative numbers or $\infty$
-(infinity) as values.  Infinity, plays the same role as
-$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
-into potential functions by mapping $\mathit{false}$ to $\infty$ and
-$\mathit{true}$ to $0$.
-%
-\begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}
-${\color{blue}%
-  \{ \#_1(a) + 2k \}}$
-while (k > 0) {
-  x=0;
-  ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
-  while (x < N && a[x] == 1) {
-    ${\color{blue}%
-      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
-    a[x]=0;
-    ${\color{blue}%
-      \{ \#_1(a) + 2k + 1 \}}$
-    tick(1);
-    ${\color{blue}%
-      \{ \#_1(a) + 2k \}}$
-    x++; }
-  ${\color{blue}%
-    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
-  if (x < N) {
-    ${\color{blue}%
-      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
-    a[x]=1;
-    ${\color{blue}%
-      \{ \#_1(a) + 2(k-1) + 1 \}}$
-    tick(1); }
-  ${\color{blue}%
-    \{ \#_1(a) + 2(k-1) \}}$
-  k--;
-  ${\color{blue}%
-    \{ \#_1(a) + 2k \}}$
-} ${\color{blue} \{ \#_1(a) \}}$
-\end{lstlisting}
-\end{minipage}
-
-\caption{Example derivation using amortized reasoning.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
-  0 to all other operations.
-  }
-\label{fig:xmplinc}
-\end{figure}
-%
-\iffull{
-}{}
-We can then use the logic to prove a concrete bound for a given resource
-metric, or a bound that is parametric in a set of resource metrics.
-For example, we can derive the following quantitative
-triple for a binary-search function \code{bsearch} that holds for all
-stack metrics $M$.
-%
-\begin{center}
-${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
-\code{bsearch(x,l,h)}
-${ \{ Z{\cdot}M_\code{bsearch} \} }$
-\end{center}
-%
-A stack metric assigns cost $0$ to all evaluation steps except
-function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
-and after the call we return $M_f$ resources that can be used in
-subsequent function calls.  So $M_f$ corresponds to the stack-frame
-size of the function $f$.  In the pre- and postcondition, the logical
-variable $Z$ is used to relate the size of the input with the
-potential that is left after the function call.  The logical part $Z =
-\log_2(h-\ell)$ of the precondition is $0$ if $Z =
-\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
-$\infty$ otherwise.
-
-For simplicity, assume for the remainder of this section that we are
-interested in bounding the number of \emph{ticks} in a program.  That
-is, we use the \emph{tick metric} that assigns cost $n$ to the function call
-\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
-is an integer and a negative number means that $-n$ resources are
-returned.
-
-We say that the analysis is amortized because the use of a potential
-function $\Phi$ enables us to amortize the cost of operations like in
-typical textbook examples that use the potential method of Tarjan's
-amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
-shows a program that repeatedly increments a binary counter
-implemented with a boolean-valued ($0$ or $1$) array.  The derived
-bound in the precondition states that the number of ticks
-executed by the program is bounded by $\#_1(a) + 2k$.  Here, $k$ is
-the number of increments to the counter that are performed by the
-program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
-the number of $1$'s in the array $a$.  We have inserted the statement
-\code{tick(1)} into the program whenever we update the array $a$.  A
-naive analysis of the algorithm would lead to a quadratic bound since
-there can be a linear number of updates in the inner loop.
-
-To derive the bound for the program in~\pref{fig:xmplinc} in the
-quantitative Hoare logic, we follow the textbook
-reasoning~\cite{Algorithms}.  When we assign \code{a[x]=0} in the
-inner loop, we know that $a[x]=1$ before the assignment.  As a result,
-we have $\#_1(a) = \#_1(a') + 1$ if $a'$ is the array $a$ after the
-assignment.  So we obtain one resource unit in addition to the
-invariant (i.e., $\Phi = \#_1(a) + 2k + 1$) that we can use to pay for the
-cost of the statement \code{tick(1)}.  To pay for the tick statement
-in the conditional, we use the fact that $k>0$ to write $2k$ as $1 +
-2(k-1) +1$.  Dual to the first assignment, we have $\#_1(a) +1 =
-\#_1(a')$ if $a$ is the array before and $a'$ is the array after the
-assignment \code{a[x]=1}.  After paying one potential unit for the
-statement $\code{tick(1)}$, we are left with the potential $\#_1(a) +
-2(k - 1)$.  But since we increment $k$ at the end of the outer loop,
-we can establish again our loop invariant $\#_1(a) + 2k$.
-
-See \pref{sec:logic} for more a formal definition of the logic and
-additional example derivations.
-
-% For example we have the following judgement (provided that initially
-% $\state_0(\code x) \ge \state_0(\code y) \ge 0$).
-% $$
-% \htriple
-% {2{\cdot}x + 10}
-% {\code{ x = x -  y}}
-% {2{\cdot}x + 2{\cdot}y + 10}
-% $$
-% This must be understood as a motion of potential from the
-% variable~$\code x$ to the variable~$\code y$. If some resource is
-% consumed $\code y$ times afterwards, we can then use $\code y$'s
-% potential to pay for it. However, the initial potential function
-% $\Phi$ only refers to the variable~$\code x$: the potential function
-% allows us to pass on the cost of the subsequent resource consumption
-% to~$\code x$.  This flexibility combined with intelligent decisions on
-% potential motions allows us to design an automatic analysis to
-% discover linear invariants for surprisingly convoluted programs.
-
-\begin{figure}[t]
-  \centering
-\begin{lstlisting}[mathescape]
-${\color{blue}%
-  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-while (x+K<=y) {
-  ${\color{blue}%
-  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  x=x+K;
-  ${\color{blue}%
-  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
-  tick(T);
-  ${\color{blue}%
-  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
-}
-${\color{blue}%
- \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
-\end{lstlisting}
-  \caption{Automatic derivation of a tight bound on the number of
-    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
-    $T>0$ are not program variables but denote concrete constants in
-    the program.  The reasoning works uniformly for all such $K$ and
-    $T$.}
-  \label{fig:ex1}
-\end{figure}
-
-%
-\newlength{\progwidth}
-
-\begin{figure*}[t!]
- \setlength{\progwidth}{.22\linewidth}
-  \centering
-\hspace{-0.4cm}
-  \begin{minipage}[b]{.18\linewidth}
-    \begin{center}
-   \begin{lstlisting}[]
-while (n>x) {
-  ${\color{blue}%
-  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-  if (m>y)
-    ${\color{blue}%
-    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
-    y=y+1;
-    ${\color{blue}%
-    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  else
-    ${\color{blue}%
-    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
-    x=x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
-}
-   \end{lstlisting}
-
-$|[x, n]| + |[y, m]|$
-\\[.4\baselineskip]
-      {\bf speed\_1}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-
-while (x<n) {
-  ${\color{blue}%
-  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-  if (z>x)
-    ${\color{blue}%
-    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    x=x+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  else
-    ${\color{blue}%
-    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
-    z=z+1;
-    ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[z, n]|$
-\\[.4\baselineskip]
-      {\bf speed\_2}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (z-y>0) {
-  ${\color{blue}%
-  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y=y+1;
-  ${\color{blue}%
-  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(3);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-${\color{blue}%
-\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-while (y>9) {
-  ${\color{blue}%
-  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  y=y-10;
-  ${\color{blue}%
-  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
-  tick(1);
-  ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
-   \end{lstlisting}
-
-$3.1|[y,z]| + 0.1|[0,y]|$
-\\[.4\baselineskip]
-      {\bf t08a}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (n<0) {
-  ${\color{blue}%
-  n{<}0; \,  P(n,y) \vdash }$
-  n=n+1;
-  ${\color{blue}%
-  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
-  y=y+1000;
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  while (y>=100 && *){
-    ${\color{blue}%
-    y{>}99; \, 9 {+} P(n,y) \vdash }$
-    y=y-100;
-    ${\color{blue}%
-    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
-    tick(5);
-    ${\color{blue}%
-    \dashv \cdot; \, 9 {+} P(n,y) }$
-  }
-  ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
-  tick(9);
-  ${\color{blue}%
-  \dashv \cdot; \, P(n,y) }$
-}
-   \end{lstlisting}
-$59|[n,0]| {+} 0.05|[0,y]|$
-\\[.4\baselineskip]
-      {\bf t27}
-    \end{center}
-  \end{minipage}
-\vspace{1ex}
-\caption{Derivations of bounds on the number of ticks for challenging
-  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
-  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
-  \emph{t08a} contains \emph{sequential loops} so that the iterations
-  of the second loop depend on the first, and \emph{t27} contains
-  interacting \emph{nested loops}. In the potential functions, we only
-  mention the non-zero terms and in the logical context $\Gamma$ we
-  only mention assertions that we actually use in the reasoning. In
-  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
-  0.05|[0,y]|$.}
-  \label{fig:ex_list}
-\end{figure*}
-%
-%
-
-\subsection{Automatic Amortized Analysis}
-
-A program logic provides a principled foundation for statically
-analyzing programs.  However, program logics need to be supported by
-automatic methods to be useful in practice.  % That is why we developed a
-% new automatic amortized analysis for deriving quantitative Hoare
-% triples for common C programs.
-
-To enable automation, we fix the shape of the potential functions
-$\Phi$ so that it becomes possible to use \emph{linear programming}
-(LP) to compute a derivation in the logic.  If we assume for now that
-a program state $\state$ simply maps variables to integers then we
-require
-$$
-\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
-$$
-where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
-$q_0$ is the constant potential and that we treat constants as program
-variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
-component of $\Phi(\state)$ if $x \in \dom{\state}$.
-%
-We then develop an inference rule for each syntactic construct that
-derives a sound triple $\htriple{\Phi}P{\Phi'}$ in the quantitative
-logic but enables inference using an LP solver.
-
-This idea is best explained by example.  If we again use the
-tick metric that assigns cost $n$ to the function call
-\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
-the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
-that there are no restrictions on the signs of the input variables in
-the examples.}
-\begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
-\end{lstlisting}
-To derive this bound in our automatic amortized analysis, we start
-with the initial potential $\Phi_0 = |[x,y]|$ (short for
-$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
-loop invariant.  For the loop body we then (like in Hoare logic) have
-to derive a triple like
-$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
-only do so if we utilize the fact that $x<y$ at the beginning of the
-loop body.  % The reasoning then works as follows. We start with the
-% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
-% assignment.
-If we denote the updated version of $x$ after the
-assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
-the potential before and after the assignment \code{x=x+1} holds.
-This means that we have the potential $|[x,y]| + 1$ before the
-statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
-we end up with potential $|[x,y]|$ after the loop body and have
-established the loop invariant again.
-
-Note that our notion of a potential function is a generalization of the
-concept of a ranking function.  A potential function can be used like
-a ranking function if we use the tick metric and add the statement
-\code{tick(1)} to every back edge of the program (loops and function
-calls).  However, a potential function is more flexible.  For example,
-we can use a potential function to prove that Example~2 does not
-consume any resources in the tick metric.
-\begin{lstlisting}[basicstyle=\tt\small]
-  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
-\end{lstlisting}
-Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
-In both cases, we reason exactly like in the first version of the
-while loop to prove the bound.  Of course, such loops with different
-tick annotations can be seamlessly combined in a larger program.
-\begin{lstlisting}[basicstyle=\tt\small]
-  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
-\end{lstlisting}
-%
-More formally, we develop (in \pref{sec:AAA}) a judgement
-$$
-\Gamma; Q \vdash S \dashv Q'; \Gamma'
-$$
-such that $\Gamma$ contains logical assertions like $x<y$ that we
-collect along the conditional branches of the program and $Q$ is a
-family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
-indexed by the variables currently in scope.  The logical
-context $\Gamma$ is simply a conjunction of inequalities between
-linear combinations of variables.  It is used, for example, to determine
-at variable assignments if we can extract constant potential to pay for future
-cost.  It is important to note that the reasoning about assertions in
-$\Gamma$ is very basic.  We do not perform any fixpoint computations
-and only derive trivial loop invariants about variables that are
-unchanged in the loop body.
-
-\pref{fig:ex1} shows a derivation of the bound
-$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
-version of Example~1 in which we increment $x$ by a constant $K>0$ and
-consume $T>0$ resources in each iteration.  The reasoning is similar
-to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions in $\Gamma$ are only used in the rule for the assignment
-\code{x=x+K}.  To the best of our knowledge, no other implemented tool
-for C is currently capable of deriving a tight bound on the cost of
-such a loop.  For $T=1$ (many systems focus on the number of loop
-iterations without a cost model) and $K=10$,
-KoAT~\cite{BrockschmidtEFFG14} computes the bound $|x| + |y| + 10$,
-Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
-LOOPUS~\cite{SinnZV14} computes the bound $y-x-9$.  Only
-PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
-translate the program into a term-rewriting system by hand.
-
-To automate the reasoning, we first introduce an unknown rational
-variable for each factor in the potential functions.  We then use our
-inference rules (see \pref{sec:AAA}) to emit linear constraints on
-these variables that enforce that variable assignments that respect
-the constraints correspond to sound potential annotations.  For
-instance if $K = 1$ in \pref{fig:ex1} then we would have the
-annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
-|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
-p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
-are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
-$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.
-
-\iffull{It might be surprising that we}{We} only track simple linear relations in
-the logical context $\Gamma$.  While it would of course be possible to keep
-track of more sophisticated assertions, this simple form is
-sufficient for the examples we considered and we can efficiently
-decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
-assignment rule.  Nevertheless, this logical part of our analysis
-system is orthogonal to the quantitative part and can be easily
-extended if necessary.
-
-As mentioned earlier, the automatic analysis can handle challenging example
-programs without special tricks or techniques.  Examples
-\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
-work~\cite{GulwaniMC09},  demonstrate that our method can handle
-\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
-derives the same bounds as our analysis but requires heuristics for
-its counter instrumentation.  These loops can also be handled with inference of
-\emph{disjunctive invariants}, but in the abstract interpretation
-community, these invariants are known to be notoriously difficult to
-generate.
-%
-In example \emph{speed\_1} we have one loop that first increments
-variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
-derive the tight bound $|[x, n]| + |[y, m]|$.
-%
-Example \emph{speed\_2} is even trickier, and we found it hard to
-find a bound manually.  However, using potential transfer reasoning as
-in amortized analysis, it is easy to prove the tight bound
-$|[x, n]| + |[z, n]|$.
-
-Example \emph{t08a} shows the ability of the analysis to discover
-interaction between \emph{sequenced loops} through size change of
-variables.  We accurately track the size change of $y$ in the first
-loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
-Furthermore, \emph{t08a} shows again that we do not handle the
-constants $1$ or $0$ in any special way.  In all examples we could
-replace $0$ and $1$ with other constants like in the second loop and
-still derive a tight bound.  \iffull{The only information, that the analyzer
-needs is $y \geq c$ before assigning $y = y - c$.}{}
-%
-Example \emph{t27} shows how amortization can be used to handle
-\emph{interacting nested loops}.  In the outer loop we increment the
-variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
-increment the variable $y$ by $1000$.  Then we non-deterministically
-(expressed by \code{*}) execute an inner loop that decrements $y$ by
-$100$ until $y<100$.  The analysis discovers that only the first
-execution of the inner loop depends on the initial value of $y$.  We
-again derive tight constant factors.
-%
-
-As mentioned, the analysis also handles advanced control flow like
-\code{break} and \code{return} statements, and mutual recursion.
-\pref{fig:ex_rec} contains two mutually-recursive functions with their
-automatically derived tick bounds.  The function \code{count\_down}
-decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{count\_up},
-which is dual to \code{count\_down}.  Here, we
-count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis is the only available system that computes a tight
-bound on this example.  \iffull{The analysis amounts to computing the meeting
-point of two trains that approach each other with different speeds.}{}
-% This is comparable to the
-% determination of the point at which to trains meet if they approach
-% each other with different speeds.
-
-\iffull{We only show a small selections of the programs that we can
-  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
-  extended version~\cite{anon_extended}} is a list of more than $30$
-classes of challenging programs that we can automatically analyze.
-\pref{sec:exper} contains a more detailed comparison with other
-tools\iffull{for automatic bound derivation.}{.}
-
-\begin{figure}[t]
-  \centering
-    \begin{minipage}[b]{\linewidth}
-    \begin{center}
-   \begin{lstlisting}
-void count_down (int x,int y) {
-  if (x>y) { tick(1); count_up(x-1,y); } }
-
-void count_up (int x, int y) {
-  if (y+1<x) { tick(1); count_down(x,y+2); } }
-
-   \end{lstlisting}
-
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
-\\[.4\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-\vspace{-1.5ex}
-
-  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
-  \label{fig:ex_rec}
-\end{figure}
-
-\sectskip
 \section{Syntax and Semantics}

### 2014-11-10 paper/main.tex
freshen section 3, update plan

@@
 %
-% 1) Intro
-% 2) Overview and Examples
-% 3) Cost Semantics for Clight
-% 4) Compositional Resource-Bound Analysis
-% 5) Automatic Inference via LP Solving
-% 6) Auxiliary Variables and User Interaction
-% 7) Quantitative Hoare Logic and Soundness Proof
-% 8) Experimental Evaluation
-% 9) Related Work
-% 10) Conclusion
+%    Abstract                              1c
+% 1) Intro                                 3c
+% 2) The Potential Method                  1c
+% 3) Compositional Resource-Bound Analysis 4c
+%    - Gulwani/ examples
+% 4) Derivation Rules                      2c
+% 5) Automatic Inference via LP Solving    2c
+% 6) Logical State and User Interaction    2c
+% 7) Soundness Proof                       2c
+%    1) Cost Semantics for Clight
+%    2) Quantitative Hoare Logic
+% 8) Experimental Evaluation               2c
+% 9) Related Work                          1c
+% a) Conclusion                            1c
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \sectskip
-\section{Automatic Amortized Analysis}
+\section{Automated Reasoning with \toolname}
 \label{sec:AAA}
@@
 
-In this section we present the automatic amortized analysis that we
-use in our implementation to derive resource bounds for C
-programs.
+In this section we present formally the analysis that we
+use in our implementation to derive resource bounds for C code.
 
@@
 
-To find derivations in the quantitative Hoare logic automatically, we
-have to focus on bounds that have a certain shape.  The general form
-of \emph{potential functions} (or assertions) that we consider is
+To find resource bounds automatically, we first need to restrict our
+search space.  This is done by focusing on bounds that have a certain
+shape.  The general form of \emph{potential functions} that we consider is
 $$
@@
 Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
-is again a simplified program state as introduced in \pref{sec:sem},
-$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
+is a simplified program state that maps variable names to integers,
+$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
 references to the linear coefficients $q_i$, we introduce an \emph{index
@@
 rewrite the above equality as\iffull{
-$$
-  \Phi = \sum_{i \in I} q_i f_i.
-$$}
-{ $\Phi = \sum_{i \in I} q_i f_i$.}  We often write $xy$ to denote the
+$$ \Phi = \sum_{i \in I} q_i f_i.$$}{
+$\Phi = \sum_{i \in I} q_i f_i$.}%
+We often write $xy$ to denote the
 index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
@@
 
-\paragraph{Logical State.}
+\paragraph{Abstract Program State.}
 
 In addition to the quantitative annotations our automatic amortized
-analysis needs to maintain a minimal logical state to justify certain
+analysis needs to maintain a minimal abstract state to justify certain
 operations on quantitative annotations.  For example when analyzing
@@
 determine which intervals will increase or decrease.  The knowledge
-needed by our rules can be inferred by local reasoning (i.e., in
+needed by our rules can be inferred by local reasoning (i.e. in
 basic blocks without recursion and loops) within usual theories
@@
 simpler rules and leaves room for future improvements.
-%
-Our implementation uses conjunctions of linear inequalities as
-logical state.  We never compute fixpoints and take $\top$
-as pre- and postconditions for functions. It has proved to be
-sufficient for the variety of examples we are interested in or found
-in the literature.
+Our implementation finds these logical contexts using abstract
+interpretation using the domain of linear inequalities.  We observed that
+the rules of the analysis often requires only minimal local knowledge.
+This means that it is not necessary for us to compute precise loop
+invariants and only a rough fixpoint (e.g. keeping only inequalities on
+variables unchanged by the loop) is sufficient to obtain good bounds.
 
@@
 { }
-{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!!}
+{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
 %
@@
 { Q' \succeq_\Gamma Q }
-
 \end{mathpar}
-\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
+\caption{Inference rules of the quantitative analysis.}
 \label{fig:auto}
@@
 \end{mathpar}
-\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
+\caption{Selected inference rules of the quantitative analysis.}
 \label{fig:auto}
@@
 $$
-These judgements correspond to the logic judgements $\Delta; B; R
-\vdash \htriple P S Q$ where each assertion splits in two parts, a
-logical part $\Gamma$ and a quantitative part $Q$.  That means that
-$(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
-$(\Gamma_R, Q_R)$ is the postcondition for return statements, and
-$(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
-simple Hoare triple.  We leave out the function context $\Delta$ in
-the presentation to avoid cluttering notations.  Instead we assume a
-fixed function context $\Delta$ that is implicit on all judgements.
-Function contexts are treated exactly as in the logic.  % We call
-% precondition the pair $(\Gamma, Q)$ and postcondition the pair
-% $(\Gamma', Q')$.
-The correspondence between judgements of the
-automation and the ones of the quantitative logic is made formal by
-our soundness proof of the automatic amortized analysis at the end of
-this section.
-
-As a convention, if $Q$ and $Q'$ are quantitative annotations then we
-assume that $Q = (q_i)_{i\in I}$ is a family with elements $q_i$, $Q'
-= (q'_i)_{i \in I}$, etc.  The notation $Q \pm n$ used in many rules
-defines a new context $Q'$ such that $q'_0 = q_0 \pm n$ and $\forall i
-\neq 0 .\, q'_i = q_i$.  We have the implicit side condition that all
-rational coefficients are non-negative.  Finally, if a rule mentions
-$Q$ and $Q'$ and leaves the latter undefined at some index $i$ we will
-implicitly assume that $q'_i = q_i$.
+
+These judgements are similar to classic Hoare triples in the presence
+of the early exit statements \code{break} and \code{return}.  The part
+$(\Gamma_B, Q_B)$ is the postcondition used when breaking out of
+a loop.  For example, it is set upon entry of looping statements by
+the rule {\sc Q:Loop} and used before \code{break} statements by the
+rule {\sc Q:Break}.  Similarly $(\Gamma_R, Q_R)$ is the postcondition
+used before returning from a function call.  The part $(\Gamma, Q) \vdash
+S \dashv (\Gamma', Q')$ of the judgement can be seen as a regular
+Hoare triple.  All assertions are split into two parts, the logical
+part and the quantitative part.  The quantitative part $Q$ represents a
+potential function as a collection of non-negative numbers $q_i$ indexed
+by the index set $I$.  The logical part $\Gamma$ is left abstract but
+is enforced by our derivation system to respect classic Hoare logic
+constraints.
+
+% These judgements correspond to the logic judgements $\Delta; B; R
+% \vdash \htriple P S Q$ where each assertion splits in two parts, a
+% logical part $\Gamma$ and a quantitative part $Q$.  That means that
+% $(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
+% $(\Gamma_R, Q_R)$ is the postcondition for return statements, and
+% $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
+% simple Hoare triple.  We leave out the function context $\Delta$ in
+% the presentation to avoid cluttering notations.  Instead we assume a
+% fixed function context $\Delta$ that is implicit on all judgements.
+% Function contexts are treated exactly as in the logic.  % We call
+% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
+% % $(\Gamma', Q')$.
+% The correspondence between judgements of the
+% automation and the ones of the quantitative logic is made formal by
+% our soundness proof of the automatic amortized analysis at the end of
+% this section.
+
+As a convention, if $Q$ and $Q'$ are quantitative annotations
+we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
+The notation $Q \pm n$ used in many rules defines a new context $Q'$
+such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
+In all the rules, we have the implicit side condition that all rational
+coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
+and leaves the latter undefined at some index $i$ we will implicitly
+assume that $q'_i = q_i$.
 
 We describe the automatic amortized analysis for a subset of
-expressions of Clight.  Assignments must have the form $x \gets
+expressions of Clight.  Assignments are restricted to the form $x \gets
 y$ or $x \gets x \pm y$.  In the implementation, a Clight
 program is converted into this form prior to analysis without changing
-the resource cost.  This achieved by using a series of \emph{cost-free
+the resource cost.  This is achieved by using a series of \emph{cost-free
   assignments} that do not result in additional cost in the
-analysis.  Non-linear operations such as $x \gets z*y$ are simply
-handled by assigning zero potential to coefficients like $q_{xa}$ and
-$q_{ax}$ that contain $x$ after the assignment.
+semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
+a[y]$ are simply handled by assigning zero potential to coefficients
+like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
 
@@
 
-Most of the rules correspond exactly to the respective rules of the
-quantitative Hoare logic.
-\iffull{
-  For example, the rule {\sc Q:Skip} simply reuses its postcondition
-  as precondition.  This is sound since evaluating a $\code{skip}$
-  statement never costs any resources and does not change the program
-  state.  Similarly, the rule {\sc Q:Tick} enforces that the
-  quantitative annotation of the precondition is at least $M_t(n)$ to
-  pay for the resource consumption.  Since the program state is also
-  unchanged in this case, the logical part $\Gamma$ of the
-  precondition can soundly be reused in the postcondition.  In the
-  rule {\sc Q:Break}, where the linear control flow is broken, the
-  postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
-  taken from the break part of the judgement similarly as it is done
-  in the quantitative logic.
-}
-%
-{ Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
-  {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
-  rules can be found in the extended version.
-}
-%
-The rule {\sc Q:Return} exhibits a slight
-difference to the quantitative logic.  While we have a function $\Z
-\to \Assn$ to represent return values in the logic, we assume that a
-special variable $\ret$ is part of the index set of $Q_R$.  The
-substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
-$q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
-$q_i'=q_i$ otherwise.
-
-Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
-  Q:Inc} for increments and decrements, which describe how the
-potential is distributed after a size change of a variable.  The rule
-{\sc Q:IncP} is for increments $x \gets x + y$ and {\sc Q:DecP} is
-for decrements $x \gets x - y$, they both apply only when we can deduce
-from the logical context $\Gamma$ that $y \ge 0$.
-Of course, we have symmetrical rules {\sc
-  Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
-The rules are equivalent in the case where $y=0$.  Finally, the rule
-{\sc Q:Inc} can be applied if we cannot deduce any information about
-$y$.  In the implementation, the rules for increments and decrements
-are combined into one syntax directed rule where the specialized rules
-take priority over {\sc Q:Inc}.
+The rules of our derivation system must serve two purposes.  They must
+\emph{attach} potential to certain program variable intervals and use
+this potential, when it is allowed, to \emph{pay} for resource consuming
+operations.  These two purposes are easily illustrated on the {\sc Q:Skip}
+rule.  This rule reuses its precondition as postcondition.  This is
+explained by two facts: first, no resource is consumed by the skip
+operation, thus no potential needs to be taken out of the precondition
+to pay for the evaluation; second, the program state is not changed by
+the execution of a skip statement, thus all the potential available
+before the execution of the skip statement is still available after.
+These two observations would also justify a rule with the side-condition
+$Q' \le Q$, because some potential is always safely lost.  For the sake
+of orthogonality, it is not how we present the rule and this form of
+weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}.
+
+% Most of the rules correspond exactly to the respective rules of the
+% quantitative Hoare logic.
+% \iffull{
+%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
+%   as precondition.  This is sound since evaluating a $\code{skip}$
+%   statement never costs any resources and does not change the program
+%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
+%   quantitative annotation of the precondition is at least $M_t(n)$ to
+%   pay for the resource consumption.  Since the program state is also
+%   unchanged in this case, the logical part $\Gamma$ of the
+%   precondition can soundly be reused in the postcondition.  In the
+%   rule {\sc Q:Break}, where the linear control flow is broken, the
+%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
+%   taken from the break part of the judgement similarly as it is done
+%   in the quantitative logic.
+% }
+% %
+% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
+%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
+%   rules can be found in the extended version.
+% }
+% %
+% The rule {\sc Q:Return} exhibits a slight
+% difference to the quantitative logic.  While we have a function $\Z
+% \to \Assn$ to represent return values in the logic, we assume that a
+% special variable $\ret$ is part of the index set of $Q_R$.  The
+% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
+% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
+% $q_i'=q_i$ otherwise.
+
+Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
+which describe how the potential is distributed after a size change of
+a variable.  The rule {\sc Q:IncP} is for increments $x \gets x + y$
+and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
+when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
+Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
+that can be applied if $y$ is not positive.  The rules are equivalent
+in the case where $y=0$.  Finally, the rule {\sc Q:Inc} can be applied
+if we cannot deduce any information about $y$.  In the implementation,
+the rules for increments and decrements are combined into one syntax
+directed rule where the specialized rules take priority over {\sc Q:Inc}.
 
@@
 sequential composition.  \iffull{We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the
-analysis efficient.}{}  The high-level idea of {\sc Q:Weak} is the same
-as in the rule {\sc L:Weak} of the quantitative logic: If we have a
-sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept
-of \emph{more potential} is formalized by the relation $Q'
-\succeq_\Gamma Q$ that is defined in the rule {\sc Relax}.  Here, $Q$
-and $Q'$ are potential annotations and $\Gamma$ is a logical context.
-The rule {\sc Relax} deals also with the important task of transferring
-constant potential (represented by $q_0$) to interval sizes and vice
-versa.  If we can deduce from the logical context that the interval
-size $|\inter x y| \geq l$ is larger then a constant $l$ then we can
-transfer potential $q_{xy} {\cdot} |\inter x y|$ form the interval to
-constant potential $l{\cdot}q_{xy}$ and guarantee that we do not gain
-potential.  Conversely, if $|\inter x y| \leq u$ for a constant $u$
-then we can transfer constant potential $u{\cdot}q_{xy}$ to the
-interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining
-potential.
-
-\paragraph{Automatic Inference via LP Solving.}
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
+annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
+deals also with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq l$ is
+larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
+without gaining potential.
+
+\sectskip
+\section{Automatic Inference via LP Solving.}
+\label{sec:lp}
+\aftersectskip
 
@@
 
-\paragraph{Soundness Proof.}
+\sectskip
+\section{Soundness Proof}
+\label{sec:soundness}
+\aftersectskip
 

### 2014-11-11 paper/main.tex
start work on soundness proof

@@
 
-The soundness of the analysis builds on the quantitative logic.  We
-translate judgements of the automatic amortized analysis as defined in
-\pref{fig:auto} into quantitative Hoare triples.  We define a
-translation function $\tr$, such that if a judgement $J$ in the
-automatic analysis is derivable, $\tr(J)$ is derivable in the
-quantitative logic.  By using $\tr$ to translate derivations of the
-automatic analysis to derivations in the quantitative logic we can
-automatically obtain a certified resource bound for the analyzed
-program.
+The soundness of the analysis builds on a new cost semantics for Clight
+and an extended Quantitative Logic.  Using these two tools, the soundness
+proof of \toolname is done by a exhibiting a translation morphism.
+
+\subsection{Cost Semantics for Clight}
+
+To base the soundness proof on a formal ground, we start by defining a
+new cost-aware operational semantics for Clight.  CompCert Clight's
+operational semantics is based on small-step transitions and
+continuations.  Expressions---which do not have side effects---are
+evaluated in a big-step fashion.
+
+A program state $\state = (\theta, \gamma)$ is composed of two maps
+from variable names to integers. The first map, $\theta : \text{Locals}
+\to \Z$, assigns integers to local variables of a function, and the
+second map, $\gamma : \text{Globals} \to \Z$, gives values to global
+variables of the program.  In this article, we assume that all values are
+integers but in the implementation we support all data types of Clight.
+The evaluation function \evalE{\cdot}{} maps an expression $e \in E$
+to a value $\evalE{e}{\state} \in \Z$ in the program state $\state$.
+% For simplicity, we assume that local variables and global variables
+% are disjoint.
+We write $\state(x)$ to obtain the value of $x$ in program state
+$\state$.
+% \iffull{Such a lookup is defined as
+% $$
+% (\theta, \gamma)(x) =
+% \left\{
+% \begin{array}{lr}
+% \theta(x) & \mbox{ if } x \in \dom \theta \\
+% \gamma(x) & \mbox{ if } x \in \dom \gamma
+% \end{array}
+% \right..
+% $$}{}
+Similarly, we write $\state[x \mapsto v]$ for the state based on $\state$
+where the value of $x$ is updated to $v$.
+
+The small-step semantics we define is standard, except that it tracks
+the resource consumption of a program.  The semantics is parametric in
+the resource of interest for the user of our system.  We achieve this
+independence by parameterizing evaluations with a resource \emph{metric}
+$M$; a tuple of nine rational numbers and two maps.  Each of these
+parameters indicates the amount of resource consumed by a corresponding
+step in the semantics.  Resources can be released by assigning a negative
+cost. %  The metric used in the implementation can also
+% depend on other information that is statically available such as the
+% name of the called function or the number of arguments.
+Two rules for update and tick follow.
+\begin{mathpar}
+\Rule[rightskip=0.3cm]{U}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\rightarrow\\\\
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+\and
+\Rule[leftskip=0.3cm,rightskip=0.2cm]{T}
+{ }
+{ (\state, \code{tick}(n), \cont, \cost)
+\rightarrow\\\\
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
+}
+\end{mathpar}
+The rules keep track of the resource cost $\cost$ at all times, and
+all rules have as implicit side condition that $\cost$ in the start
+configuration is non-negative.  It makes it possible
+to detect a resource crash as a stuck configuration where $\cost<0$.
+
+% \pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
+% rules define a rewrite system for program configurations of the form
+% $(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
+% is the statement being executed, $\cont$ is a continuation that
+% describes what remains to be done after the execution of $S$, and
+% $\cost \in \Q$ is the non-negative number of resources available for
+% further execution.  All rules that can decrease $\cost$ have the
+% implicit side condition that the resource quantity available
+% \emph{before} the step is non-negative.  This means that we allow
+% $(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
+% transition relation $\smallstep{M}^*$ to indicate that the execution
+% ran out of resources.  However, every execution that reaches such a
+% state is stuck.
+%
+% \iffull{
+% A continuation $\cont$ represents the context of the execution of a
+% given statement.  A continuation can be the empty continuation
+% $\Kstop$ (used to start a program), a sequenced computation $\Kseq S
+% \cont$, a loop continuation $\Kloop S \cont$, or the continuation
+% $\Kcall r \theta \cont$ of a function call.
+% $$
+% \cont := \Kstop
+% \mid \Kseq S \cont
+% \mid \Kloop S \cont
+% \mid \Kcall r \theta \cont
+% $$
+% %
+% During the execution we assume a fixed function context $\fenv$ that
+% maps function names to a list of variables and the statement that
+% defines the function body.  It is used in the rule \textsc{S:Call}.
+% }{}
+%
+% The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
+% S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
+% following.  If the statement $S$ is executed in program state
+% $\state$, with continuation $\cont$, and with $\cost$ resources
+% available then---after a finite number of steps---the evaluation will
+% reach the new machine state $(\state', S', \cont',\cost')$ and there
+% are $\cost'$ resources available.  If $\cost' \ge 0$ then the
+% execution did not run out of resources and the resource consumption up
+% to this point is $\cost - \cost'$.  If this difference is negative
+% then resource became available during the execution.  If however
+% $\cost' < 0$ then the execution ran out of resources and is stuck.
+% The cost of the execution is then $\cost \ge 0$.
+%
+% \iffull{
+% \pref{lem:sem1} summarizes the main properties of the resource
+% counter of the semantics.  It can be proved by induction on the number
+% of steps.
+% \begin{lemma}
+% \label{lem:sem1}
+% \itemskip
+%   \begin{enumerate}
+%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+%     \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
+%     \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
+% \itemskipIn
+%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+%     \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
+%     \cont', d')$ then $\cost - \cost' = d - d'$.
+%   \end{enumerate}
+% \itemskip
+% \end{lemma}
+% %
+% If we assume that we have a \emph{positive resource metric}, that is,
+% a metric that assigns a positive cost to each evaluation step then we
+% can link bounds on the resource consumption under the metric to
+% termination.  This is formalized by \pref{lem:sem_term}.
+% %
+% \begin{lemma}
+%  \label{lem:sem_term}
+%  Let $M$ be a positive metric and assume that there exists $b \in
+%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
+%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
+%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
+%  \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
+%  and $\state''$ implies $n \le B$.
+% \end{lemma}
+% }
+
+\subsection{Quantitative Hoare Logic}
+
+To prove the soundness of \toolname we found it useful to go through
+an intermediate step using a Quantitative Hoare Logic.  This logic
+is at the same time a convenient semantic tool and a clean way to
+interface manual proofs with our automation, as explained in XXX.
+The logic is an extension of the one presented in \cite{veristack14},
+it adds support for arbitrary resources and simplifies the
+handling of auxiliary state.
+
+Following the previous work, we define quantitative Hoare triples
+as $\Delta;B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
+maps from program states to $\Qplusz\cup\{\infty\}$ in order to represent
+an amount of resources available.  The assertions $B$ and $R$ are
+postconditions for the case in which the block $S$ exits by a \code{break}
+or \code{return} statement.  Additionally, $R$ depends on an extra value
+that is returned by the function.  The meaning of the triple $\htriple Q S
+{Q'}$ is as follows: If $S$ is executed with starting state $\state$,
+the empty continuation \Kstop{} and at least $Q(\state)$ resource units
+available then the evaluation does not run out of resources and there
+are at least $Q(\state')$ resources left if the evaluation terminates
+in state $\state'$.  The logic rules are similar to the ones in previous
+work and are simply extended to account for all the costs found in
+our cost-aware semantics.
+
+\subsection{The Soundness Theorem}
+
+We translate judgements of the automatic amortized analysis as defined in
+\pref{fig:auto} into quantitative Hoare triples.  We define a translation
+function $\tr$, such that if a judgement $J$ in the automatic analysis
+is derivable, $\tr(J)$ is derivable in the quantitative logic.  By using
+$\tr$ to translate derivations of the automatic analysis to derivations
+in the quantitative logic we can automatically obtain a certified resource
+bound for the analyzed program.
 
@@
 (
-  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
-  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
+  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
+  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
   )
@@
 These definitions let us translate the judgement $J = B; R; P \vdash
-S \dashv P'$ in the context $\Delta$ by
+S \dashv P'$ in the context $\Delta$ by $\tr(J) :=$
 $$
-  \tr(J) :=
   (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
@@
 weakening because we have to show that the preconditions of the rules
-{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
-we prove the following lemma to show the soundness of the translation
-of {\sc Q:Weak}.
-
-\begin{lemma}[Relax]
-  If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
-  $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
-\end{lemma}
-\begin{proof}
-We observe the following inequations.
-\begin{align*}
-\Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
-&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
-  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
-  \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
-  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
- &\ge \Phi_Q(\state)
-  + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
-  + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
- &\ge \Phi_Q(\state).
-\end{align*}
-Since $\state \models \Gamma$, using the transitivity of $\models$ and
-the definition of $\mathcal U$ and $\mathcal L$ we get
-$\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
-$\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
-These two inequalities justify the last step.
-\end{proof}
-%
-\noindent
-We prove similar lemmas for the rules that deal with assignment.  With
-the computational content of the proof, we have an effective algorithm
-to construct certificates for upper bounds derived automatically.
-}{}
+{\sc L:Weak} or {\sc L:Update}, respectively, are met. %  \iffull{ For instance
+% we prove the following lemma to show the soundness of the translation
+% of {\sc Q:Weak}.
+%
+% \begin{lemma}[Relax]
+%   If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
+%   $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
+% \end{lemma}
+% \begin{proof}
+% We observe the following inequations.
+% \begin{align*}
+% \Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
+% &\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
+%   + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
+%   \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
+%   + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
+%  &\ge \Phi_Q(\state)
+%   + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
+%   + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
+%  &\ge \Phi_Q(\state).
+% \end{align*}
+% Since $\state \models \Gamma$, using the transitivity of $\models$ and
+% the definition of $\mathcal U$ and $\mathcal L$ we get
+% $\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
+% $\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
+% These two inequalities justify the last step.
+% \end{proof}
+% %
+% \noindent
+% We prove similar lemmas for the rules that deal with assignment.  With
+% the computational content of the proof, we have an effective algorithm
+% to construct certificates for upper bounds derived automatically.
+% }{}
 
@@
 
-\paragraph{Semantics.}
-
-CompCert Clight's operational semantics is based on small-step
-transitions and continuations.  Expressions---which do not have side
-effects---are evaluated in a big-step fashion.  Here, we describe a
-simplified version of Clight's semantics for the subset we consider.
-
-A program state $\state = (\theta, \gamma)$ is composed of two maps from
-variable names to integers. The first map, $\theta : \text{Locals} \to
-\Z$, assigns integers to local variables of a function, and the second
-map, $\gamma : \text{Globals} \to \Z$, gives values to global variables
-of the program.  In this article, we assume that all values are
-integers but in the implementation we support all data types of
-Clight.
-%
-The evaluation function \evalE{\cdot}{} maps an expression $e \in
-E$ to a value $\evalE{e}{\state} \in \Z$ in the program state
-$\state$.
-
-For simplicity, we assume that local variables and global variables
-are always different.  We just write $\state(x)$ to obtain the value of
-$x$ in program state $\state$. \iffull{Such a lookup is defined as
-$$
-(\theta, \gamma)(x) =
-\left\{
-\begin{array}{lr}
-\theta(x) & \mbox{ if } x \in \dom \theta \\
-\gamma(x) & \mbox{ if } x \in \dom \gamma
-\end{array}
-\right..
-$$}{}
-\iffull{For a program state $\state$, we}{Similarly} we write $\state[x \mapsto v]$
-for the program state that maps $x$ to $v$ and behaves as $\state$ for
-all other variables, regardless whether $x \in \dom \theta$ or $x \in
-\dom \gamma$.
-
-The small-step semantics is standard, except that it tracks the
-resource consumption of a program.  The semantics is parametric in the
-resource of interest for the user of our system.  We achieve this
-independence by parameterizing evaluations with a resource
-\emph{metric} $M$; a tuple of nine rational numbers \ifshort{$M_a$, $M_b$, $M_r$, $\ldots$}{}, one map $M_e :
-\text{Expr} \to \Q$ from expressions to rational numbers, and one map
-$M_t : \Z \to \Q$ from integers to rational numbers.
-%
-\iffull{$$
-M = (
-M_a,
-M_b,
-M_r,
-M_u,
-M_f,
-M_l,
-M_c^1,
-M_c^2,
-M_s,
-M_e(E),
-M_t(n)
-)
-$$}{
-
-}
-Each of these rational numbers indicates the amount of resource
-consumed by a corresponding operation.  If the assigned resource cost
-is negative then it means that resources are released.  The metric
-that we use in the implementation can also depend on other information
-that is statically available such as the name of the called function
-or the number of arguments.
-
-\pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
-rules define a rewrite system for program configurations of the form
-$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
-is the statement being executed, $\cont$ is a continuation that
-describes what remains to be done after the execution of $S$, and
-$\cost \in \Q$ is the non-negative number of resources available for
-further execution.  All rules that can decrease $\cost$ have the
-implicit side condition that the resource quantity available
-\emph{before} the step is non-negative.  This means that we allow
-$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
-transition relation $\smallstep{M}^*$ to indicate that the execution
-ran out of resources.  However, every execution that reaches such a
-state is stuck.
-
-\iffull{
-A continuation $\cont$ represents the context of the execution of a
-given statement.  A continuation can be the empty continuation
-$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
-\cont$, a loop continuation $\Kloop S \cont$, or the continuation
-$\Kcall r \theta \cont$ of a function call.
-$$
-\cont := \Kstop
-\mid \Kseq S \cont
-\mid \Kloop S \cont
-\mid \Kcall r \theta \cont
-$$
-%
-During the execution we assume a fixed function context $\fenv$ that
-maps function names to a list of variables and the statement that
-defines the function body.  It is used in the rule \textsc{S:Call}.
-}{}
-
-The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
-S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
-following.  If the statement $S$ is executed in program state
-$\state$, with continuation $\cont$, and with $\cost$ resources
-available then---after a finite number of steps---the evaluation will
-reach the new machine state $(\state', S', \cont',\cost')$ and there
-are $\cost'$ resources available.  If $\cost' \ge 0$ then the
-execution did not run out of resources and the resource consumption up
-to this point is $\cost - \cost'$.  If this difference is negative
-then resource became available during the execution.  If however
-$\cost' < 0$ then the execution ran out of resources and is stuck.
-The cost of the execution is then $\cost \ge 0$.
-
-\iffull{
-\pref{lem:sem1} summarizes the main properties of the resource
-counter of the semantics.  It can be proved by induction on the number
-of steps.
-\begin{lemma}
-\label{lem:sem1}
-\itemskip
-  \begin{enumerate}
-  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
-    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
-    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
-\itemskipIn
-  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
-    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
-    \cont', d')$ then $\cost - \cost' = d - d'$.
-  \end{enumerate}
-\itemskip
-\end{lemma}
-%
-If we assume that we have a \emph{positive resource metric}, that is,
-a metric that assigns a positive cost to each evaluation step then we
-can link bounds on the resource consumption under the metric to
-termination.  This is formalized by \pref{lem:sem_term}.
-%
-\begin{lemma}
- \label{lem:sem_term}
- Let $M$ be a positive metric and assume that there exists $b \in
- \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
- (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
- Then there exists $B \in \N$ such that $(\state, S, \cont, d)
- \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
- and $\state''$ implies $n \le B$.
-\end{lemma}
-}
-{
-The extended version of this article~\cite{anon_extended} contains all rules and
-lemmas that state the main properties of the cost semantics.
-}
-
 \iffull{

### 2014-11-11 paper/main.tex
restructuring of the first quater or so

@@
 \sectskip
-\section{Overview and Examples}
+\section{The Potential Method}
 \aftersectskip
-
-In this section, we informally introduce the automatic amortized
-analysis for Clight programs and demonstrate the technique with
-example bounds that we derive step-by-step following our analysis
-system.
-
-\subsection{Bounding Resource Usage with Potential Functions}
 
@@
 
-% Potential functions can access the whole program state, so this
-% framework can readily be used to derive bounds that depend on the
-% memory contents.
-
-\subsection{Automating  Amortized Analysis}
-
-To make amortized reasoning feasible on large code bases, we would
-like to automatically infer the potential functions.  The first step
-towards this goal is to reduce our search space by fixing the shape of
-potential functions $\Phi$.  In this paper, we focus on a form that
-can express tight bounds for many typical programs and allows
-inference view \emph{linear programming}:
-%
-$$
-\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}}
-  q_{(x,y)} \cdot |[\state(x),\state(y)]|
-$$
-where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  For now, we assume
-that a program state is a simple map from program variables to integers
-and we treat constants as program variables (that cannot be changed
-by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
-the sum.  % We then develop an inference rule for each syntactic construct
-% that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
-% the previous section.
+% \subsection{Automating  Amortized Analysis}
+
+% To make amortized reasoning feasible on large code bases, we would
+% like to automatically infer the potential functions.  The first step
+% towards this goal is to reduce our search space by fixing the shape of
+% potential functions $\Phi$.  In this paper, we focus on a form that
+% can express tight bounds for many typical programs and allows
+% inference view \emph{linear programming}:
+% %
+% $$
+% \Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}}
+%   q_{(x,y)} \cdot |[\state(x),\state(y)]|
+% $$
+% where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  For now, we assume
+% that a program state is a simple map from program variables to integers
+% and we treat constants as program variables (that cannot be changed
+% by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
+% the sum.  % We then develop an inference rule for each syntactic construct
+% % that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
+% % the previous section.
 
@@
 \end{lstlisting}
-  \caption{Automatic derivation of a tight bound on the number of
+  \caption{Derivation of a tight bound on the number of
     ticks for a standard \emph{for loop}.  The parameters $K>0$ and
-    $T>0$ are not program variables but denote concrete constants in
-    the program.  The reasoning works uniformly for all such $K$ and
-    $T$.}
+    $T>0$ are not program variables but denote concrete constants. 
+    % The reasoning works uniformly for all such $K$ and
+    % $T$.
+  }
   \label{fig:ex1}
@@
 
-The automatic derivation of a resource bound using these potential
-functions is best explained by example.  If we use the tick metric
-that assigns cost $n$ to the function call \code{tick(n)} and cost $0$
-to all other operations then the cost of the following example can be
-bounded by $|[x,y]|$.% \footnote{Note that there are no restrictions on
+The derivation of a resource bound using potential functions is best
+explained by example.  If we use the tick metric that assigns cost $n$
+to the function call \code{tick(n)} and cost $0$ to all other
+operations then the cost of the following example can be bounded by
+$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
   % the signs of the input variables in the examples.}
@@
 \end{lstlisting}
-To derive this bound in our amortized analysis, we start with
-the initial potential $\Phi_0 = |[x,y]|$ (short for $\Phi_0(\state)
-= |[\state(x),\state(y)]|$) which we also use as the loop invariant.
-For the loop body we have (like in Hoare logic) to derive a triple
-like $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do
-so if we utilize the fact that $x<y$ at the beginning of the loop body.
-The reasoning then works as follows. We start with the potential $|[x,y]|$
-and the fact that $|[x,y]| > 0$ before the assignment.  If we denote the
-updated version of $x$ after the assignment by $x'$ then the relation
-$|[x,y]| = |[x',y]| + 1$ between the potential before and after the
-assignment \code{x=x+1} holds.  This means that we have the potential
-$|[x,y]| + 1$ before the statement \code{tick(1)}.  Since \code{tick(1)}
-consumes one resource unit, we end up with potential $|[x,y]|$ after
-the loop body and have established the loop invariant again.
+To derive this bound, we start with the initial potential $\Phi_0 =
+|[x,y]|$, which
+we also use as the loop invariant.  For the loop body we have (like in
+Hoare logic) to derive a triple like
+$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
+if we utilize the fact that $x<y$ at the beginning of the loop body.
+The reasoning then works as follows. We start with the potential
+$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
+we denote the updated version of $x$ after the assignment by $x'$ then
+the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
+after the assignment \code{x=x+1} holds.  This means that we have the
+potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
+\code{tick(1)} consumes one resource unit, we end up with potential
+$|[x,y]|$ after the loop body and have established the loop invariant
+again.
 
@@
 assertions in $\Gamma$ are only used in the rule for the assignment
-\code{x=x+K}.  To the best of our knowledge, no other implemented tool
-for C is currently capable of deriving a tight bound on the cost of
-such a loop.  For $T=1$ (many systems focus on the number of loop
-iterations without a cost model) and $K=10$, KoAT computes the bound
-$|x| + |y| + 10$, Rank computes the bound $y-x-7$, and LOOPUS computes
-the bound $y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if
-we translate the program into a term-rewriting system by hand.
-
-To automate the reasoning, we first introduce an unknown rational
-variable for each factor in the potential functions.  We then use our
-inference rules (see \pref{sec:AAA}) to emit linear constraints on
-these variables that enforce that variable assignments that respect
-the constraints correspond to sound potential annotations.  For
-instance if $K = 1$ in \pref{fig:ex1} then we would have the
-annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
-|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
-p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
-are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
-$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
-q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
-the coefficients of the initial potential function. LP solvers are
-among the most mature constraint solving tools. As a result, the bound
-derivation works very efficiently in practice.
-
-\iffull{It might be surprising that we}{We} only track simple linear relations in
-the logical context $\Gamma$.  While it would of course be possible to keep
-track of more sophisticated assertions, this simple form is
-sufficient for the examples we considered and we can efficiently
-decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
-assignment rule.  Nevertheless, this logical part of our analysis
-system is orthogonal to the quantitative part and can be easily
-extended if necessary.
-
-\ifx\fullversion\undefined{}\else{
+\code{x=x+K}.
+
+To the best of our knowledge, no other implemented tool for C is
+currently capable of deriving a tight bound on the cost of such a
+loop.  For $T=1$ (many systems focus on the number of loop iterations
+without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
+10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
+$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
+translate the program into a term-rewriting system by
+hand. Nevertheless, we will show in \jan{Next section} that the
+potential method makes automatic bound derivation straightforward.
+
+% To automate the reasoning, we first introduce an unknown rational
+% variable for each factor in the potential functions.  We then use our
+% inference rules (see \pref{sec:AAA}) to emit linear constraints on
+% these variables that enforce that variable assignments that respect
+% the constraints correspond to sound potential annotations.  For
+% instance if $K = 1$ in \pref{fig:ex1} then we would have the
+% annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
+% |[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
+% p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
+% are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
+% $p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
+% q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
+% the coefficients of the initial potential function. LP solvers are
+% among the most mature constraint solving tools. As a result, the bound
+% derivation works very efficiently in practice.
+
+% \iffull{It might be surprising that we}{We} only track simple linear relations in
+% the logical context $\Gamma$.  While it would of course be possible to keep
+% track of more sophisticated assertions, this simple form is
+% sufficient for the examples we considered and we can efficiently
+% decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
+% assignment rule.  Nevertheless, this logical part of our analysis
+% system is orthogonal to the quantitative part and can be easily
+% extended if necessary.
+
 Note that our notion of a potential function is a generalization of the
@@
 \end{lstlisting}
-}\fi
-
-\subsection{Advantages of the Potential Method }
-
-The potential based approach addresses several challenges in resource
-bound analysis. This includes amortized bounds in which individual
-loop iterations have different resource consumption, sequential loops
-or function calls that require the tracking of size changes of
-variables, and nested loops that interact by modifying the same set of
-variables.
-
-\paragraph{Challenging Loop Patterns} As mentioned earlier, the
-automatic analysis can handle challenging example programs without
-special tricks or techniques.  Examples \emph{speed\_1} and
+
+In general, the potential based approach addresses several challenges
+in resource bound analysis. This includes amortized bounds in which
+individual loop iterations have different resource consumption,
+sequential loops or function calls that require the tracking of size
+changes of variables, and nested loops that interact by modifying the
+same set of variables.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\sectskip
+\section{Compositional Resource-Bound Analysis}
+\label{sec:AAA}
+\aftersectskip
+
+In this section we describe the high-level design of the automatic
+amortized analysis that we implemented in $\toolname$. We then explain
+and motivate our design decisions with examples.
+
+\paragraph{Linear Potential Functions.}
+
+To find resource bounds automatically, we first need to restrict our
+search space.  In this work, we focus on a form that can express tight
+bounds for many typical programs and allows inference view
+\emph{linear programming}.  The general form of \emph{potential
+  functions} that we consider is
+$$
+  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
+    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
+$$
+%
+Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
+is a simplified program state that maps variable names to integers,
+$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
+references to the linear coefficients $q_i$, we introduce an \emph{index
+  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
+\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
+$f_i$ in the potential function: $0$ corresponds to the constant
+function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
+|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
+rewrite the above equality as\iffull{
+$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$}{
+$\Phi(\state) = \sum_{i \in I} q_i f_i(\state)$.}%
+We often write $xy$ to denote the
+index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
+the linear-algebra sense) for potential functions.}{}  This allows us to
+uniquely represent any linear potential function
+$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
+is, a family of \ifshort{non-negative}{} rational numbers where only
+a finite number of elements are not zero. \iffull{ For soundness, we
+  require that potential functions always give non-negative results.
+  This is acheived by restricting coefficients in a quantitative
+  annotation to be always non-negative.}{}
+
+In the potential functions, we treat constants as global variables
+that cannot be assigned to.  For example, if the program contains the
+constant $8128$ then we have a variable $c_{8128}$ and
+$\state(c_{8128}) = 8128$.  We assume that every program state
+includes the constant $c_0$.
+
+\paragraph{Abstract Program State.}
+
+In addition to the quantitative annotations our automatic amortized
+analysis needs to maintain a minimal abstract state to justify certain
+operations on quantitative annotations.  For example when analyzing
+the code $x \gets x + y$, it is helpful to know the sign of $y$ to
+determine which intervals will increase or decrease.  The knowledge
+needed by our rules can be inferred by local reasoning (i.e. in
+basic blocks without recursion and loops) within usual theories
+(e.g. Presburger arithmetic or bit vectors).
+
+In contrast to the quantitative annotations, logical contexts $\Gamma$
+in the presented inference system are left abstract.  This allows for
+simpler rules and leaves room for future improvements.
+Our implementation finds these logical contexts using abstract
+interpretation using the domain of linear inequalities.  We observed that
+the rules of the analysis often requires only minimal local knowledge.
+This means that it is not necessary for us to compute precise loop
+invariants and only a rough fixpoint (e.g. keeping only inequalities on
+variables unchanged by the loop) is sufficient to obtain good bounds.
+
+\paragraph{Examples: Challenging Loop Patterns.}
+
+One might think that our set of potential functions is too simplistic
+to be able to express and prove bounds for realistic
+programs. Nevertheless, we can handle challenging example programs
+without special tricks or techniques.  Examples \emph{speed\_1} and
 \emph{speed\_2}, that are taken from previous work~\cite{GulwaniMC09},
@@
 
-\paragraph{Compositionality}
+\paragraph{Examples: Compositionality.}
 
@@
 analysis of common code.
-
-\jan{I don't like to talk about ``the reader''. Often you can leave it
-  out. It is fine to address the reader with ``you''. See BUGS in
-  Writing.}
 
@@
 or additively, resulting in a loose quadratic bound or no bound at all.
-
-\paragraph{User Guidance for Non-Linear Bounds}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\sectskip
-\section{Automated Reasoning with \toolname}
-\label{sec:AAA}
-\aftersectskip
-
-In this section we present formally the analysis that we
-use in our implementation to derive resource bounds for C code.
-
-\paragraph{Linear Potential Functions.}
-
-To find resource bounds automatically, we first need to restrict our
-search space.  This is done by focusing on bounds that have a certain
-shape.  The general form of \emph{potential functions} that we consider is
-$$
-  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
-    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
-$$
-%
-Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
-is a simplified program state that maps variable names to integers,
-$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
-references to the linear coefficients $q_i$, we introduce an \emph{index
-  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
-\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
-$f_i$ in the potential function: $0$ corresponds to the constant
-function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
-|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
-rewrite the above equality as\iffull{
-$$ \Phi = \sum_{i \in I} q_i f_i.$$}{
-$\Phi = \sum_{i \in I} q_i f_i$.}%
-We often write $xy$ to denote the
-index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
-the linear-algebra sense) for potential functions.}{}  This allows us to
-uniquely represent any linear potential function
-$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
-is, a family of \ifshort{non-negative}{} rational numbers where only
-a finite number of elements are not zero. \iffull{ For soundness, we
-  require that potential functions always give non-negative results.
-  This is acheived by restricting coefficients in a quantitative
-  annotation to be always non-negative.}{}
-
-In the potential functions, we treat constants as global variables
-that cannot be assigned to.  For example, if the program contains the
-constant $8128$ then we have a variable $c_{8128}$ and
-$\state(c_{8128}) = 8128$.  We assume that every program state
-includes the constant $c_0$.
-
-\paragraph{Abstract Program State.}
-
-In addition to the quantitative annotations our automatic amortized
-analysis needs to maintain a minimal abstract state to justify certain
-operations on quantitative annotations.  For example when analyzing
-the code $x \gets x + y$, it is helpful to know the sign of $y$ to
-determine which intervals will increase or decrease.  The knowledge
-needed by our rules can be inferred by local reasoning (i.e. in
-basic blocks without recursion and loops) within usual theories
-(e.g. Presburger arithmetic or bit vectors).
-
-In contrast to the quantitative annotations, logical contexts $\Gamma$
-in the presented inference system are left abstract.  This allows for
-simpler rules and leaves room for future improvements.
-Our implementation finds these logical contexts using abstract
-interpretation using the domain of linear inequalities.  We observed that
-the rules of the analysis often requires only minimal local knowledge.
-This means that it is not necessary for us to compute precise loop
-invariants and only a rough fixpoint (e.g. keeping only inequalities on
-variables unchanged by the loop) is sufficient to obtain good bounds.
 
@@
 
-\paragraph{Judgements of the Automatic Analysis.}
-
-The inference system for the automatic amortized analysis is defined
-in~\pref{fig:auto}.  The inference rules derive judgements of the form
+\section{Derivation System}
+
+In the following we describe the local and compositional derivation
+rules of the automatic amortized analysis.
+
+\paragraph{Judgements.}
+
+The derivation system for the automatic amortized analysis is defined
+in~\pref{fig:auto}.  The derivation rules derive judgements of the form
 $$
@@
 
-\paragraph{Inference Rules.}
+\paragraph{Derivation Rules.}
 
@@
 step we go through the whole program and apply inductively the
-inference rules of the automatic amortized analysis.  During this
+derivation rules of the automatic amortized analysis.  During this
 process our tool uses symbolic names for the rational coefficients
@@
 
+\jan{ the following should go somewhere here:
+
+Note that potential functions can access the whole program state, so
+this framework can readily be used to derive bounds that depend on the
+memory contents.
+}
+
 \begin{figure}

### 2014-11-11 paper/main.tex
minor
new abstract

@@
 \begin{abstract}
-This paper presents a new approach to quantitative
-analysis for C programs.  The technique described uses
-ideas from amortized analysis and abstract interpretation
-to derive tight linear worst-case bounds that come with a
-proof certificate.  Contrary to most available tools for
-C, ours has a compositional design that provides the
-programmer with bounds expressed in terms of meaningful
-variables.  The compositionality allows to handle uniformly
-nested sequenced loops and recursive sequenced function
-calls.  The automatic analysis is implemented
-as a tool that leverages state-of-the-art SMT solvers
-LP solvers to automatically infer resource bounds on
-a wide range of C programs.  It is the first automatic
-amortized analysis that can derive bounds that depend
-on negative integers and differences between numbers.
-To prove its soundness and give a concrete shape to proof
-certificates, we developed a quantitative Hoare logic
-that we proved sound using the Coq proof assistant.  This
-logic allows manual proofs of arbitrarily complex resource
-bounds on C programs and interaction with the automation.
-To prove the scalability and precision of this new tool,
-we run it on multiple code snippets from the literature,
-reproduce results from previous work, and analyze a
-consequent part of a modern benchmark suite.
+  This paper presents a new approach for automatically deriving
+  worst-case resource bounds for C programs.
+  %
+  The described technique combines ideas from amortized analysis and
+  abstract interpretation in a unified framework to address four
+  challenges for state-of-the-art techniques: compositionality, user
+  interaction, generation of proof certificates, and scalability.
+  %
+  \emph{Compositionality} is achieved by incorporating the potential
+  method of amortized analysis.  It enables the derivation of global
+  whole-program bounds with local derivation rules by naturally
+  tracking size changes of variables in sequenced loops and function
+  calls.  The resource consumption of functions is described abstractly
+  and a function call can be analyzed without access to the function
+  body.
+  %\
+  \emph{User interaction} is supported with a new mechanism that
+  clearly separates qualitative and quantitative verification. A user
+  can guide the analysis to derive complex non-linear bounds by using
+  auxiliary variables and assertions.  The assertions are
+  separately proved using established qualitative techniques such as
+  abstract interpretation or Hoare logic.
+  %
+  \emph{Proof certificates} are automatically generated from the local
+  derivation rules. A soundness proof of the derivation system with
+  respect to a formal cost semantics guarantees the validity of the
+  certificates.
+  %
+  \emph{Scalability} is attained by an efficient reduction of
+  bound inference to a linear optimization problem that can be solved
+  by off-the-shelf LP solvers.
+  %
+  The analysis framework is implemented in the publically-available
+  tool $\toolname$. An experimental evaluation demonstrates the
+  advantages of the new technique with a comparison of $\toolname$ with
+  existing tools on challenging micro benchmarks and the analysis of
+  more than XXX lines of C code from the XXX benchmark suite.
+
+% Contrary to most available tools for
+% C, ours has a compositional design that provides the
+% programmer with bounds expressed in terms of meaningful
+% variables.  The compositionality allows to handle uniformly
+% nested sequenced loops and recursive sequenced function
+% calls.  The automatic analysis is implemented
+% as a tool that leverages state-of-the-art SMT solvers
+% LP solvers to automatically infer resource bounds on
+% a wide range of C programs.  It is the first automatic
+% amortized analysis that can derive bounds that depend
+% on negative integers and differences between numbers.
+% To prove its soundness and give a concrete shape to proof
+% certificates, we developed a quantitative Hoare logic
+% that we proved sound using the Coq proof assistant.  This
+% logic allows manual proofs of arbitrarily complex resource
+% bounds on C programs and interaction with the automation.
+% To prove the scalability and precision of this new tool,
+% we run it on multiple code snippets from the literature,
+% reproduce results from previous work, and analyze a
+% consequent part of a modern benchmark suite.
 
@@
     tick(5);
-    ${\color{blue}%
-    \dashv \cdot; \, 9 {+} P(n,y) }$
-  }
-  ${\color{blue}%
+  } ${\color{blue}%
   \dashv \cdot; \, 9 {+} P(n,y) \vdash }$

### 2014-11-11 paper/main.tex
minor modifications

@@
 \pref{sec:exper} contains a more detailed comparison with other
-tools\iffull{for automatic bound derivation.}{.}
+tools\iffull{ for automatic bound derivation.}{.}
 
@@
 this potential, when it is allowed, to \emph{pay} for resource consuming
-operations.  These two purposes are easily illustrated on the {\sc Q:Skip}
-rule.  This rule reuses its precondition as postcondition.  This is
+operations.  These two purposes are illustrated on the {\sc Q:Skip}
+rule.  This rule reuses its precondition as postcondition, it is
 explained by two facts: first, no resource is consumed by the skip
-operation, thus no potential needs to be taken out of the precondition
-to pay for the evaluation; second, the program state is not changed by
+operation, thus no potential has to be used to pay for the evaluation;
+second, the program state is not changed by
 the execution of a skip statement, thus all the potential available
@@
 These two observations would also justify a rule with the side-condition
-$Q' \le Q$, because some potential is always safely lost.  For the sake
+$Q \ge Q'$, because some potential is always safely lost.  For the sake
 of orthogonality, it is not how we present the rule and this form of
-weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}.
+weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
+described later.
 
@@
 implemented a smoother handling of the auxiliary state through a shallow
-embedding in Coq.  Despite being first a theoritical tool, the logic can
+embedding in Coq.  Despite being first a theoretical tool, the logic can
 be used directly as in~\cite{veristack14}.  Its judgements are inspired

### 2014-11-11 paper/main.tex
section shuffling
integration of the example derivation

@@
 
-\begin{figure*}[t!]
+\begin{figure*}
  \setlength{\progwidth}{.22\linewidth}
@@
 \iffull{
-\begin{figure*}[t]
+\begin{figure*}
 \begin{mathpar}
@@
 \ifshort{
-\begin{figure*}[t]
+\begin{figure*}
 \small
@@
 \sectskip
-\section{Automatic Inference via LP Solving.}
+\section{Automatic Inference via LP Solving}
 \label{sec:lp}
@@
 the symbolic names.  We then feed the collected constraints to an
-off the shelf LP solver~\footnote{We currently use Coin-Or's CLP.}.
+off the shelf LP solver. (We currently use Coin-Or's CLP.)
 If the solver successfully finds a solution, we know that a derivation
@@
 error is reported to the user.
+
+\pref{fig:derivation} contains an example derivation as it is produced
+by our automatic analyzer.  The upper case letters (with optional
+superscript) such as $Q^\text{de}$ stand for a family of rational
+variables that are later part of the constraint system that is passed
+to the LP solver.  For example $Q^\text{de}$ stands for the potential
+function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
+q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
+q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
+variables such as $q^\text{de}_{x,10}$ are yet unknown and later
+instantiated by the LP solver.  The derivation with the instantiated
+variables is then a sound derivation in the verified program logic.
+
+The rules are syntax directed and applied inductively.  That means
+that we follow the inductive definition of the syntax.  For example,
+the outermost expression is a loop, so we use the rule \text{Q:Loop}
+at the root of the derivation tree.  At this point, we don't know yet
+whether a loop invariant exists.  But we produce the constraints
+$Q^\text{lo} = P^\text{lo}$ which is short for the following
+constraint set.
+$$
+\begin{array}{lll}
+q^\text{lo}_0 = p^\text{lo}_0 &
+q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
+q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
+q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
+q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
+q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
+\end{array}
+$$
+%
+This constraint set expresses the fact that the potential before and
+after the loop body is equal and thus constitutes an invariant.
+
+In general, the weakening rule is applied after every syntax directed
+rule.  Applying weakening everywhere does not lead to performance
+problems. However, it can be left out in practice at some places to
+increase the efficiency of the tool.
+
+The weakening operation $\succeq_{\Gamma}$ is defined in the rule
+\textsc{Relax}.  For example, we have
+$$
+\begin{array}{lll}
+P^\text{de} \succeq_{(\cdot)} P^\text{we}
+& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
+& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
+& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+\end{array}
+$$
+
+After the constraint generation, the LP solver is provided with a
+so-called objective function that it will minimize.  We wish to
+minimize the initial potential, which is the resource bound on the
+whole program.  Here it is given by $Q$.  Moreover, we would like to
+express that minimization of linear potential such as
+$$
+q_{10,x} |[10,x]|
+$$
+takes priority over minimization of constant potential such as
+$$
+q_{0,10} |[0,10]| \; .
+$$
+
+To this end, modern LP solvers support efficient resolving of
+constraint sets with additional constraints: First we take our initial
+constraint set as given in \pref{fig:derivation}.  We then as the
+solver to find a solution that satisfies the constraint and minimizes
+the linear expression
+$$
+q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} \; .
+$$
+Note the linear penalties (this is a random choice) that we give to
+some of the factors.  We use it to express that the interval $[10,x]$
+is smaller than the interval $[0,x]$ and, similarly, that the interval
+$[x,0]$ is smaller than the interval $[x,10]$.
+
+The LP solver now returns a solution of the constraint set and an
+objective value, that is, a mapping from variables to floating-point
+numbers and the value of the objective function with this
+instantiation.  The solver also memorizes the optimization path that
+led to the optimal solution.  In this case, the objective value would
+be
+$
+5000
+$
+since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
+We now add the constraint
+$$
+q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} = 5000
+$$
+to our constraint set and ask the solver to optimize the objective
+function
+$$
+q_0 + 11{\cdot}q_{0,10} \; .
+$$
+This happens in almost no time in practice.
+
+The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
+the derived bound is $0.5 |[0,x]|$.
+
+Another interesting note is that the constraints that we generate have
+a particularly simple form that is known as \emph{network problem}.
+Such problems can be solved in linear time in practice.
+
+\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
+
+\begin{figure*}
+\centering
+\footnotesize
+\begin{mathpar}
+\inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
+{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
+  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1cm]
+    { \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
+      { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
+        { \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
+          {
+          }
+          { (x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
+            \dashv (\cdot, P^\text{de})
+          }
+        }
+        { (x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
+          \dashv (\cdot, P^\text{we})
+        }
+        \and
+        \inferrule*[right={\scriptsize (Q:Tick)},leftskip=0cm,rightskip=0cm]
+        {
+        }
+        { (x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
+          \dashv (\cdot, P^\text{ti})
+        }
+      }
+      { (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
+        \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq})
+      }
+    }
+    { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
+      \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if})
+    }
+    \and
+    \inferrule*[right={\scriptsize (Q:Weak)},leftskip=4.5cm,rightskip=4cm]
+    { \inferrule*[right={\scriptsize (Q:Break)},leftskip=0cm,rightskip=0cm]
+      { }
+      { (x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash
+        \code{break} \dashv (\bot,P^\text{br})
+      }
+    }
+    { (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
+      \code{break} \dashv (\cdot,P^\text{el})
+    }
+  }
+  { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
+    \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm}
+  }
+}
+{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P)
+}
+\end{mathpar}
+\flushleft Constraints:
+$$
+\begin{array}{lll}
+P {=} B^\text{lo} \land Q {=} Q^\text{lo} {=} P^\text{lo}
+& B^\text{el} {=} B^\text{if} {=} B^\text{lo}
+    \land Q^\text{el} {=} Q^\text{if} {=} Q^\text{lo}
+    \land P^\text{el} {=} P^\text{if} {=} P^\text{lo}
+& B^\text{el} {=} B^\text{br} \land Q^\text{el} \succeq_{(x{<}10)} Q^\text{br}
+    \land P^\text{br} \succeq_{(\cdot)} P^\text{el}
+\\
+B^\text{br} {=} Q^\text{br}
+& B^\text{if} {=} B^\text{sq} \land Q^\text{if} \succeq_{(x{<}10)} Q^\text{sq}
+    \land P^\text{sq} \succeq_{(\cdot)} P^\text{if}
+& B^\text{sq} {=} B^\text{we} {=} B^\text{ti} \land Q^\text{sq} {=} Q^\text{we}
+    \land P^\text{we} {=} Q^\text{ti} \land P^\text{ti} {=} P^\text{sq}
+\\
+Q^\text{ti} {=} P^\text{ti} + 5
+& B^\text{we} {=} B^\text{de} \land Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}
+    \land P^\text{de} \succeq_{(\cdot)} P^\text{we}
+& p^\text{de}_{0,10} {=} q^\text{de}_{0,10} + q^\text{de}_{0,x}
+    \land p^\text{de}_{0} {=} q^\text{de}_{0}
+    \land \forall (\alpha, \beta) \neq (0,10)  . \,
+              p^\text{de}_{\alpha,\beta} {=} q^\text{de}_{\alpha,\beta}
+\end{array}
+$$
+\vspace{1.5ex}
+
+\begin{tabular}{l@{\hspace{5em}}l}
+ Linear Objective Function: &  Constant Objective Function: \\
+\hspace{3em} $
+q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x}
+$
+&
+\hspace{3em}
+$
+q_0 + 11{\cdot}q_{0,10}
+$
+\end{tabular}
+
+\caption{An example derivation as produced by the automatic tool.}
+\label{fig:derivation}
+\end{figure*}
 
@@
 \end{figure*}
+
+\sectskip
+\section{Logical State and User Interaction}
+\label{sec:anno}
+\aftersectskip
+
+\jan{ the following should go somewhere here:
+
+Note that potential functions can access the whole program state, so
+this framework can readily be used to derive bounds that depend on the
+memory contents.
+}
+
+\begin{figure}
+\center
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1) {
+    @assert(na > 0);@
+    a[x]=0; @na--;@
+    tick(1);
+    x++; }
+  if (x < N) {
+    a[x]=1; @na++;@
+    tick(1); }
+  k--;
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Example of assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
+  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
+  0 to all other operations.  The bound derived automatically is
+  $2|[0,k]| + |[0,na]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ lg > \log_2(h - l) \} }$
+bsearch(x,l,h@,lg@) {
+  if (h-l > 1) {
+    @assert(lg > 0);@
+    m = l + (h-l)/2;
+    @lg--;@ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h@,lg@);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We use the tick metric to model the stack consumption of the program.
+  Before a function call to \code{bsearch}, a tick statement pays for
+  the space used by stack frame of the function.  The space is recoverd
+  after the call.  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
+  The bound derived automatically is $|[0,lg]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
+
+While complete automation is often desirable, it cannot possibly
+be universal since the problem is undecidable.  In this section
+we present two techniques to derive complex resource bounds by
+leveraging our automation.
+
+When the resource bound of a program depends on the contents of the heap,
+or is non-linear (e.g. logarithmic, exponential), we propose to
+introduce \emph{logical state}.
+This technique is inspired by XXX (speed) and consists in introducing
+extra variables to the program that do not change its behavior, but help
+the automatic tool to figure out a bound.
+
+The systematic technique consists in the following steps. First,
+a program $P$ that fails to be analyzed automatically is enriched
+by extra variables $\vec x$ called logical variables to form a
+program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
+for the logical variables is shown to be such that:
+$$
+\forall n\, \sigma\, \sigma'.\,
+  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
+    {\implies}
+  \exists n' {\le} n.\,
+    (\sigma, P) \Downarrow_{n'} \sigma'.
+$$
+Since the annotated program and the original one are often
+syntactically close, the proof of this result goes
+by simple induction on the resource-aware evaluation judgement.
+Third, using existing automation tools, a bound $B(\vec x)$ for
+$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
+$\vec X$, gives the final resource bound for the program $P$.
+
+This idea is illustrated by the program on \pref{fig:xmplincaux}.
+Parts of the code in blue are annotations that were added
+to the original program text.  The top-level loop increments
+a binary counter $k$ times.  A naive analysis of
+the algorithm yields the quadratic bound $k \cdot N$.
+However, the algorithm is in fact linear and its cost is bounded
+by $2k + \#_1(a)$ where $\#_1(a)$ denotes
+the number of one entries in the array $a$.  Since this
+number depends on the heap contents, no tool available
+for C is able to derive the linear bound.  However, it can be inferred
+by our automated tool if a logical variable $na$ is introduced.
+This logical variable is a reification, in the program, of the
+number $\#_1(a)$.  For example, on line 6 of the example
+we are setting \lstinline{a[x]} to 0 and because of the condition we
+know that this array entry was 1.  To reflect this change on
+$\#_1(a)$, the logical variable $na$ is decremented.
+Similarly, on line 10, an array entry which was 0 becomes 1, so
+$na$ is incremented.  To complete the step 2 of the systematic
+procedure described above, we must show that the extra assertion
+\lstinline{na > 0} on line 5 is always true.  It
+is done by proving inductively that $na = \#_1(a)$
+and remarking that since \lstinline{a[x] == 1} is true, we must have
+$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
+% test is in fact semantically equivalent to \lstinline{true}.
+
+Another simple example is given on \pref{fig:xmplbsaux} where
+a logarithmic bound on the stack consumption of a binary search
+program is proved using logical variable annotations.  Once again,
+annotations are in blue in the program text.  In this example,
+to ease the proof of equivalence between the annotated program
+and the original one, we use the inequality $lg >
+\log_2(h-l)$ as invariant.  This trick allows a simpler
+proof because, when working with integer arithmetic, it is not always
+the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
+
+On the other hand, it might be the case that even with annotations, a
+resource bound cannot be found automatically.  With existing tools, this
+situation requires a full rewrite of the program.  However, in our
+framework, it is possible to write a manual proof and combine it with
+bounds derived by the automatic analysis.  This interaction is made
+formal using our Quantitative Hoare Logic.  This resource aware program
+logic is an extension of the one presented in~\cite{veristack14} and
+was originally designed to prove the soundness of our automatic
+analysis.  To make it suitable for this end, we made it independent
+of the resource used (e.g. stack usage, heap usage, clock ticks) and
+implemented a smoother handling of the auxiliary state through a shallow
+embedding in Coq.  Despite being first a theoretical tool, the logic can
+be used directly as in~\cite{veristack14}.  Its judgements are inspired
+by typical Hoare logics and have the shape
+$$
+\Delta \vdash \htriple P S Q
+$$
+where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
+are quantitative assertions, i.e. functions from the program states
+to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
+of the logic are described in further details in the extended version of
+this paper~\cite{anon_extended}.  Because our automated tool produces
+proofs of the resource bounds in the Quantitative Hoare Logic, it can
+be used as an assistant when deriving a complex bound by hand in the
+logic.  The interaction is guaranteed to be sound by the soundness
+theorem of our Quantitative Hoare Logic.
 
@@
 \sectskip
-\section{Logical State and User Interaction}
-\label{sec:anno}
-\aftersectskip
-
-\jan{ the following should go somewhere here:
-
-Note that potential functions can access the whole program state, so
-this framework can readily be used to derive bounds that depend on the
-memory contents.
-}
-
-\begin{figure}
-\center
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    @assert(na > 0);@
-    a[x]=0; @na--;@
-    tick(1);
-    x++; }
-  if (x < N) {
-    a[x]=1; @na++;@
-    tick(1); }
-  k--;
-}
-\end{lstlisting}
-\end{minipage}
-\caption{Example of assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
-  0 to all other operations.  The bound derived automatically is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h@,lg@) {
-  if (h-l > 1) {
-    @assert(lg > 0);@
-    m = l + (h-l)/2;
-    @lg--;@ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h@,lg@);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We use the tick metric to model the stack consumption of the program.
-  Before a function call to \code{bsearch}, a tick statement pays for
-  the space used by stack frame of the function.  The space is recoverd
-  after the call.  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The bound derived automatically is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
-
-While complete automation is often desirable, it cannot possibly
-be universal since the problem is undecidable.  In this section
-we present two techniques to derive complex resource bounds by
-leveraging our automation.
-
-When the resource bound of a program depends on the contents of the heap,
-or is non-linear (e.g. logarithmic, exponential), we propose to
-introduce \emph{logical state}.
-This technique is inspired by XXX (speed) and consists in introducing
-extra variables to the program that do not change its behavior, but help
-the automatic tool to figure out a bound.
-
-The systematic technique consists in the following steps. First,
-a program $P$ that fails to be analyzed automatically is enriched
-by extra variables $\vec x$ called logical variables to form a
-program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
-for the logical variables is shown to be such that:
-$$
-\forall n\, \sigma\, \sigma'.\,
-  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
-    {\implies}
-  \exists n' {\le} n.\,
-    (\sigma, P) \Downarrow_{n'} \sigma'.
-$$
-Since the annotated program and the original one are often
-syntactically close, the proof of this result goes
-by simple induction on the resource-aware evaluation judgement.
-Third, using existing automation tools, a bound $B(\vec x)$ for
-$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
-$\vec X$, gives the final resource bound for the program $P$.
-
-This idea is illustrated by the program on \pref{fig:xmplincaux}.
-Parts of the code in blue are annotations that were added
-to the original program text.  The top-level loop increments
-a binary counter $k$ times.  A naive analysis of
-the algorithm yields the quadratic bound $k \cdot N$.
-However, the algorithm is in fact linear and its cost is bounded
-by $2k + \#_1(a)$ where $\#_1(a)$ denotes
-the number of one entries in the array $a$.  Since this
-number depends on the heap contents, no tool available
-for C is able to derive the linear bound.  However, it can be inferred
-by our automated tool if a logical variable $na$ is introduced.
-This logical variable is a reification, in the program, of the
-number $\#_1(a)$.  For example, on line 6 of the example
-we are setting \lstinline{a[x]} to 0 and because of the condition we
-know that this array entry was 1.  To reflect this change on
-$\#_1(a)$, the logical variable $na$ is decremented.
-Similarly, on line 10, an array entry which was 0 becomes 1, so
-$na$ is incremented.  To complete the step 2 of the systematic
-procedure described above, we must show that the extra assertion
-\lstinline{na > 0} on line 5 is always true.  It
-is done by proving inductively that $na = \#_1(a)$
-and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
-% test is in fact semantically equivalent to \lstinline{true}.
-
-Another simple example is given on \pref{fig:xmplbsaux} where
-a logarithmic bound on the stack consumption of a binary search
-program is proved using logical variable annotations.  Once again,
-annotations are in blue in the program text.  In this example,
-to ease the proof of equivalence between the annotated program
-and the original one, we use the inequality $lg >
-\log_2(h-l)$ as invariant.  This trick allows a simpler
-proof because, when working with integer arithmetic, it is not always
-the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
-
-On the other hand, it might be the case that even with annotations, a
-resource bound cannot be found automatically.  With existing tools, this
-situation requires a full rewrite of the program.  However, in our
-framework, it is possible to write a manual proof and combine it with
-bounds derived by the automatic analysis.  This interaction is made
-formal using our Quantitative Hoare Logic.  This resource aware program
-logic is an extension of the one presented in~\cite{veristack14} and
-was originally designed to prove the soundness of our automatic
-analysis.  To make it suitable for this end, we made it independent
-of the resource used (e.g. stack usage, heap usage, clock ticks) and
-implemented a smoother handling of the auxiliary state through a shallow
-embedding in Coq.  Despite being first a theoretical tool, the logic can
-be used directly as in~\cite{veristack14}.  Its judgements are inspired
-by typical Hoare logics and have the shape
-$$
-\Delta \vdash \htriple P S Q
-$$
-where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
-are quantitative assertions, i.e. functions from the program states
-to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
-of the logic are described in further details in the extended version of
-this paper~\cite{anon_extended}.  Because our automated tool produces
-proofs of the resource bounds in the Quantitative Hoare Logic, it can
-be used as an assistant when deriving a complex bound by hand in the
-logic.  The interaction is guaranteed to be sound by the soundness
-theorem of our Quantitative Hoare Logic.
-
-\sectskip
 \section{Syntax and Semantics}
@@
 
-\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
-
-\begin{figure*}[t]
-  \centering
-  \footnotesize
-  \begin{mathpar}
-  \inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
-  { \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
-    {
-      \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1.5cm]
-      {
-        \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
-        {
-\inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
-{
-  \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
-  { }
-  {(x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
-   \dashv (\cdot, P^\text{de})
-  }
-}
-{(x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
-\dashv (\cdot, P^\text{we})
-}
-\and
-\inferrule*[right={\scriptsize (Q:Tick)},leftskip=0cm,rightskip=0cm]
-{ }
-{(x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
- \dashv (\cdot, P^\text{ti})
-}
-} {
-          (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq})
-        }
-      }
-      { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if}) }
-      \and
-      \inferrule*[right={\scriptsize (Q:Weak)},leftskip=4.5cm,rightskip=4cm]
-      { \inferrule*[right={\scriptsize (Q:Break)},leftskip=0cm,rightskip=0cm]
-        { }
-        {(x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash \code{break} \dashv (\bot,P^\text{br})}
-      }
-      {
-        (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash \code{break} \dashv (\cdot,P^\text{el})
-      }
-    }
-    { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm}
-    }
-  }
-  {
-   (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P)
-  }
-\end{mathpar}
-\flushleft Constraints:
-$$
-\begin{array}{llll}
-P = B^\text{lo} & Q  = Q^\text{lo} & Q^\text{lo} = P^\text{lo} \\
-B^\text{el} = B^\text{if} = B^\text{lo} & Q^\text{el} = Q^\text{if} = Q^\text{lo} & P^\text{el} = P^\text{if} = P^\text{lo} \\
-B^\text{el} = B^\text{br}  & Q^\text{el} \succeq_{(x{<}10)} Q^\text{br}  & P^\text{br} \succeq_{(\cdot)} P^\text{el} \\
-B^\text{br} = Q^\text{br}\\
-B^\text{if} = B^\text{sq}  & Q^\text{if} \succeq_{(x{<}10)} Q^\text{sq}  & P^\text{sq} \succeq_{(\cdot)} P^\text{if} \\
-B^\text{sq} = B^\text{we} = B^\text{ti} & Q^\text{sq} = Q^\text{we} & P^\text{we} = Q^\text{ti} & P^\text{ti} = P^\text{sq} \\
-Q^\text{ti} = P^\text{ti} + 5 \\
-B^\text{we} = B^\text{de}  & Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}  & P^\text{de} \succeq_{(\cdot)} P^\text{we} \\
-p^\text{de}_{0,10} = q^\text{de}_{0,10} + q^\text{de}_{0,x}  & p^\text{de}_{0} = q^\text{de}_{0}
-& \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{de}_{\alpha,\beta} = q^\text{de}_{\alpha,\beta}
-\end{array}
-$$
-\vspace{1.5ex}
-
-\begin{tabular}{l@{\hspace{5em}}l}
- Linear Objective Function: &  Constant Objective Function: \\
-\hspace{3em} $
-q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x}
-$
-&
-\hspace{3em}
-$
-q_0 + 11{\cdot}q_{0,10}
-$
-\end{tabular}
-
-\caption{An example derivation as produced by the automatic tool.}
-\label{fig:derivation}
-\end{figure*}
 \sectskip
@@
 
-\section{Example Derivation in the Automatic System}
-\label{app:derivation}
-
-\pref{fig:derivation} contains an example derivation as it is produced
-by our automatic analyzer.  The upper case letters (with optional
-superscript) such as $Q^\text{de}$ stand for a family of rational
-variables that are later part of the constraint system that is passed
-to the LP solver.  For example $Q^\text{de}$ stands for the potential
-function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
-q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
-q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
-variables such as $q^\text{de}_{x,10}$ are yet unknown and later
-instantiated by the LP solver.  The derivation with the instantiated
-variables is then a sound derivation in the verified program logic.
-
-The rules are syntax directed and applied inductively.  That means
-that we follow the inductive definition of the syntax.  For example,
-the outermost expression is a loop, so we use the rule \text{Q:Loop}
-at the root of the derivation tree.  At this point, we don't know yet
-whether a loop invariant exists.  But we produce the constraints
-$Q^\text{lo} = P^\text{lo}$ which is short for the following
-constraint set.
-$$
-\begin{array}{lll}
-q^\text{lo}_0 = p^\text{lo}_0 &
-q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
-q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
-q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
-q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
-q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
-\end{array}
-$$
-%
-This constraint set expresses the fact that the potential before and
-after the loop body is equal and thus constitutes an invariant.
-
-In general, the weakening rule is applied after every syntax directed
-rule.  Applying weakening everywhere does not lead to performance
-problems. However, it can be left out in practice at some places to
-increase the efficiency of the tool.
-
-The weakening operation $\succeq_{\Gamma}$ is defined in the rule
-\textsc{Relax}.  For example, we have
-$$
-\begin{array}{lll}
-P^\text{de} \succeq_{(\cdot)} P^\text{we}
-& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
-& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
-& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
-\end{array}
-$$
-
-After the constraint generation, the LP solver is provided with a
-so-called objective function that it will minimize.  We wish to
-minimize the initial potential, which is the resource bound on the
-whole program.  Here it is given by $Q$.  Moreover, we would like to
-express that minimization of linear potential such as
-$$
-q_{10,x} |[10,x]|
-$$
-takes priority over minimization of constant potential such as
-$$
-q_{0,10} |[0,10]| \; .
-$$
-
-To this end, modern LP solvers support efficient resolving of
-constraint sets with additional constraints: First we take our initial
-constraint set as given in \pref{fig:derivation}.  We then as the
-solver to find a solution that satisfies the constraint and minimizes
-the linear expression
-$$
-q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} \; .
-$$
-Note the linear penalties (this is a random choice) that we give to
-some of the factors.  We use it to express that the interval $[10,x]$
-is smaller than the interval $[0,x]$ and, similarly, that the interval
-$[x,0]$ is smaller than the interval $[x,10]$.
-
-The LP solver now returns a solution of the constraint set and an
-objective value, that is, a mapping from variables to floating-point
-numbers and the value of the objective function with this
-instantiation.  The solver also memorizes the optimization path that
-led to the optimal solution.  In this case, the objective value would
-be
-$
-5000
-$
-since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
-We now add the constraint
-$$
-q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} = 5000
-$$
-to our constraint set and ask the solver to optimize the objective
-function
-$$
-q_0 + 11{\cdot}q_{0,10} \; .
-$$
-This happens in almost no time in practice.
-
-The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
-the derived bound is $0.5 |[0,x]|$.
-
-Another interesting note is that the constraints that we generate have
-a particularly simple form that is known as \emph{network problem}.
-Such problems can be solved in linear time in practice.
-
-\newpage
-
 \section{Catalog of Automatically Analyzed Programs}
@@
 
-\begin{figure*}[t!]
+\begin{figure*}
 \setlength{\progwidth}{.24\linewidth}
@@
 
-\begin{figure*}[t!]
+\begin{figure*}
  \setlength{\progwidth}{.24\linewidth}
@@
 
-\begin{figure*}[t!]
+\begin{figure*}
  \setlength{\progwidth}{.28\linewidth}
@@
 
-\begin{figure*}[t!]
+\begin{figure*}
  \setlength{\progwidth}{.32\linewidth}

### 2014-11-11 paper/main.tex
more work on section 5

@@
 
-\sectskip
-\section{Automatic Inference via LP Solving}
-\label{sec:lp}
-\aftersectskip
-
-We separate the search of a derivation into two steps.  As a first
-step we go through the whole program and apply inductively the
-derivation rules of the automatic amortized analysis.  During this
-process our tool uses symbolic names for the rational coefficients
-$(q_i)$ in the rules.  Every time a linear constraint must be
-satisfied by these coefficients, it is recorded in a global list using
-the symbolic names.  We then feed the collected constraints to an
-off the shelf LP solver. (We currently use Coin-Or's CLP.)
-If the solver successfully finds a solution, we know that a derivation
-exists for the considered program.  We can then extract the initial
-$Q$ from the solver and get a resource bound for the program.  To get
-a full derivation we simply extract the complete solution from the
-solver, and apply it to the symbolic names $(q_i)$ of the coefficients
-in the derivation.  If the LP solver fails to find a solution, an
-error is reported to the user.
-
-\pref{fig:derivation} contains an example derivation as it is produced
-by our automatic analyzer.  The upper case letters (with optional
-superscript) such as $Q^\text{de}$ stand for a family of rational
-variables that are later part of the constraint system that is passed
-to the LP solver.  For example $Q^\text{de}$ stands for the potential
-function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
-q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
-q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
-variables such as $q^\text{de}_{x,10}$ are yet unknown and later
-instantiated by the LP solver.  The derivation with the instantiated
-variables is then a sound derivation in the verified program logic.
-
-The rules are syntax directed and applied inductively.  That means
-that we follow the inductive definition of the syntax.  For example,
-the outermost expression is a loop, so we use the rule \text{Q:Loop}
-at the root of the derivation tree.  At this point, we don't know yet
-whether a loop invariant exists.  But we produce the constraints
-$Q^\text{lo} = P^\text{lo}$ which is short for the following
-constraint set.
-$$
-\begin{array}{lll}
-q^\text{lo}_0 = p^\text{lo}_0 &
-q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
-q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
-q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
-q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
-q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
-\end{array}
-$$
-%
-This constraint set expresses the fact that the potential before and
-after the loop body is equal and thus constitutes an invariant.
-
-In general, the weakening rule is applied after every syntax directed
-rule.  Applying weakening everywhere does not lead to performance
-problems. However, it can be left out in practice at some places to
-increase the efficiency of the tool.
-
-The weakening operation $\succeq_{\Gamma}$ is defined in the rule
-\textsc{Relax}.  For example, we have
-$$
-\begin{array}{lll}
-P^\text{de} \succeq_{(\cdot)} P^\text{we}
-& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
-& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
-& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
-\end{array}
-$$
-
-After the constraint generation, the LP solver is provided with a
-so-called objective function that it will minimize.  We wish to
-minimize the initial potential, which is the resource bound on the
-whole program.  Here it is given by $Q$.  Moreover, we would like to
-express that minimization of linear potential such as
-$$
-q_{10,x} |[10,x]|
-$$
-takes priority over minimization of constant potential such as
-$$
-q_{0,10} |[0,10]| \; .
-$$
-
-To this end, modern LP solvers support efficient resolving of
-constraint sets with additional constraints: First we take our initial
-constraint set as given in \pref{fig:derivation}.  We then as the
-solver to find a solution that satisfies the constraint and minimizes
-the linear expression
-$$
-q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} \; .
-$$
-Note the linear penalties (this is a random choice) that we give to
-some of the factors.  We use it to express that the interval $[10,x]$
-is smaller than the interval $[0,x]$ and, similarly, that the interval
-$[x,0]$ is smaller than the interval $[x,10]$.
-
-The LP solver now returns a solution of the constraint set and an
-objective value, that is, a mapping from variables to floating-point
-numbers and the value of the objective function with this
-instantiation.  The solver also memorizes the optimization path that
-led to the optimal solution.  In this case, the objective value would
-be
-$
-5000
-$
-since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
-We now add the constraint
-$$
-q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} = 5000
-$$
-to our constraint set and ask the solver to optimize the objective
-function
-$$
-q_0 + 11{\cdot}q_{0,10} \; .
-$$
-This happens in almost no time in practice.
-
-The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
-the derived bound is $0.5 |[0,x]|$.
-
-Another interesting note is that the constraints that we generate have
-a particularly simple form that is known as \emph{network problem}.
-Such problems can be solved in linear time in practice.
-
 \newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
-
 \begin{figure*}
@@
 \end{mathpar}
-\flushleft Constraints:
+\flushleft
+\begin{tabular}{l}Constraints:\end{tabular}
 $$
@@
 $$
-\vspace{1.5ex}
-
 \begin{tabular}{l@{\hspace{5em}}l}
- Linear Objective Function: &  Constant Objective Function: \\
-\hspace{3em} $
-q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x}
+  Linear Objective Function: $
+    1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
   $
-&
-\hspace{3em}
-$
-q_0 + 11{\cdot}q_{0,10}
+& Constant Objective Function: $
+    1{\cdot}q_0 + 11{\cdot}q_{0,10}
   $
 \end{tabular}
-
 \caption{An example derivation as produced by the automatic tool.}
@@
 \end{figure*}
+
+\sectskip
+\section{Automatic Inference via LP Solving}
+\label{sec:lp}
+\aftersectskip
+
+We separate the search of a derivation into two steps.  As a first
+step we go through the whole program and apply inductively the
+derivation rules of the automatic amortized analysis.  During this
+process our tool uses symbolic names for the rational coefficients
+$(q_i)$ in the rules.  Every time a linear constraint must be
+satisfied by these coefficients, it is recorded in a global list using
+the symbolic names.  We then feed the collected constraints to an
+off the shelf LP solver. (We currently use Coin-Or's CLP.)
+If the solver successfully finds a solution, we know that a derivation
+exists for the considered program.  We can then extract the initial
+$Q$ from the solver and get a resource bound for the program.  To get
+a full derivation we simply extract the complete solution from the
+solver, and apply it to the symbolic names $(q_i)$ of the coefficients
+in the derivation.  If the LP solver fails to find a solution, an
+error is reported to the user.
+
+\pref{fig:derivation} contains an example derivation as it is produced
+by our automatic analyzer.  The upper case letters (with optional
+superscript) such as $Q^\text{de}$ are families of
+variables that are later part of the constraint system that is passed
+to the LP solver.  For example $Q^\text{de}$ stands for the potential
+function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
+q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
+q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
+variables such as $q^\text{de}_{x,10}$ are yet unknown and later
+instantiated by the LP solver.
+
+The rules are syntax directed and applied inductively.  For example,
+the outermost expression is a loop, so we use the rule \text{Q:Loop}
+at the root of the derivation tree.  At this point, we don't know yet
+whether a loop invariant exists.  But we produce the constraints
+$Q^\text{lo} = P^\text{lo}$\iffull{ which is short for the following
+constraint set.
+$$
+\begin{array}{lll}
+q^\text{lo}_0 = p^\text{lo}_0 &
+q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
+q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
+q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
+q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
+q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
+\end{array}
+$$}{.}
+%
+This constraint expresses the fact that the potential before and
+after the loop body is equal and thus constitutes an invariant.
+
+In general, the weakening rule can be applied after every syntax directed
+rule.  However, it can be left out in practice at some places to
+increase the efficiency of the tool.
+The weakening operation $\succeq_{\Gamma}$ is defined by the rule
+\textsc{Relax}, it is parameterized by a logical context that is
+used to gather information on interval sizes.  For example, we have
+$$
+\begin{array}{lll}
+P^\text{de} \succeq_{(\cdot)} P^\text{we}
+& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
+& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
+& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+\end{array}
+$$
+
+After the constraint generation, the LP solver is provided with a
+so-called objective function to minimize.  We wish to
+minimize the initial potential, which constitutes the resource bound on the
+whole program.  Here it is given by $Q$.  Moreover, we would like to
+express that minimization of linear potential such as
+$q_{10,x} |[10,x]|$
+takes priority over minimization of constant potential such as
+$q_{0,10} |[0,10]|$.
+
+To get a tight bound, we use modern LP solvers that allow constraint
+solving and minimization at the same time:  First we consider our initial
+constraint set as given in \pref{fig:derivation} and ask the
+solver to find a solution that satisfies the constraints and minimizes
+the linear expression
+$$
+1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x} \; .
+$$
+The penalties given to certain factors are used
+to prioritize certain intervals.  For example, a bound with
+$[10,x]$ will be prefered to another with $[0,x]$ because $|[10,x]|
+\le |[0,x]|$.
+%
+The LP solver now returns a solution of the constraint set and an
+objective value\iffull{, that is, a mapping from variables to floating-point
+numbers and the value of the objective function with this
+instantiation}{}.  The solver also memorizes the optimization path that
+led to the optimal solution.  In this case, the objective value would
+be 5000
+since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
+We now add the constraint
+$$
+1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
+ \le 5000
+$$
+to our constraint set and ask the solver to optimize the objective
+function
+$$
+q_0 + 11{\cdot}q_{0,10} \; .
+$$
+This happens in almost no time in practice.
+The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
+the derived bound is $0.5 |[0,x]|$.
+
+Another interesting note is that the constraints that we generate have
+a particularly simple form that is known as \emph{network problem}.
+Such problems can be solved in linear time in practice.
 

### 2014-11-11 paper/main.tex
done with RC of abstract and intro

@@
 
-\title{Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}}
+\title{Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}}
 % Certified Resource Bounds for C
@@
   existing tools on challenging micro benchmarks and the analysis of
-  more than XXX lines of C code from the XXX benchmark suite.
+  more than 2000 lines of C code from the cBench benchmark suite.
 
@@
 
-Available techniques for deriving worst-case resource bounds
+Available techniques for automatically deriving worst-case resource bounds
 fall into two categories.  Techniques
@@
 Techniques in the second category derive tight whole-program
-bounds on functional programs that operate on inductive data structures.
+bounds for programs with regular loop and recursion patterns that
+decrease the size of an individual variable or data structure.
 They are highly compositional, scale for large
@@
 it has been
-a long-time open problem to make them work in the
-presence of signed integers, mutation, and non-linear
-control flow. As a result, they are not suitable for
-typical imperative code.
-
-Notable tools in the first category include
+a long-time open problem to make them work for typical imperative code
+with non-regular iteration patterns, signed integers, mutation, and non-linear
+control flow. % As a result, they are not suitable for
+% typical imperative code.
+
+Tools in the first category include
 SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
@@
 and LOOPUS~\cite{SinnZV14}.
-They lack compositionality in at least three ways.
+They lack compositionality in at least two ways.
 First, they all base their analysis on some form of
-\emph{ranking function}.
-Since this
-is usually a local analysis and loop bounds are
+\emph{ranking function} or \emph{counter instrumentation}
+that is linked to a local analysis.  As a result, loop bounds are
 arithmetic expressions that depend on the values of
@@
 This makes it hard to give a
-resource bound on a sequence of loops and procedure calls in
-terms of the input parameters of a procedure.
+resource bound on a sequence of loops and function calls in
+terms of the input parameters of a function.
 Second, while all popular imperative programming languages
@@
 and have to inline the procedure body to perform their
-analysis.  Third, when a tool fails to find a resource
-bound for a program, there is no possibility for sound
-user interaction with the tool other than to the rewrite
-the program. For example, there is no concept of manual proofs of
-resource bounds and no framework that could support composition
-of manually derived bounds with automatically inferred bounds.
-One reason for this lack of interaction is that available tools
-work on an abstract version of the source program. This
-abstracted program has sometimes more behaviors than the
-original one (introduced by non-determinism) and is usually
-expressed in a language that does not feature memory operations.
+analysis.
 
@@
 fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
-$y - i$ decreases, but not $|i|$.  Mutation is yet another problem.
-While functional programs relate multiple fixed inputs (the typing
-context) with one output (the expression typed), imperative programs
-can change the value of variables in scope, so the potential of these
-variables has to be updated after mutations to ensure soundness.
+$y - i$ decreases, but not $|i|$. %  Mutation is yet another problem.
+% While functional programs relate multiple fixed inputs (the typing
+% context) with one output (the expression typed), imperative programs
+% can change the value of variables in scope, so the potential of these
+% variables has to be updated after mutations to ensure soundness.
+
+A general problem with existing tools is user interaction. When a tool
+fails to find a resource bound for a program, there is no possibility
+for sound user interaction to guide the tool during during bound
+derivation.
+For example, there is no concept of manual proofs of
+resource bounds and no framework that could support composition
+of manually derived bounds with automatically inferred bounds.
+
+% Finally, the tools often operate on an abstract version of the and use
+% complex external tools from term rewriting and abstract
+% interpretation. This makes it hard to generate a proof certificate for
+% the bounds or even to prove the soundness of the tools with respect to
+% a formal cost semantics.
 
@@
 %
-Our automated analysis is able to infer linear
+Our automated analysis is able to infer
 resource bounds on C programs with mutually-recursive functions and integer loops.
@@
 simple mechanism, it handles:
-
+\itemskip
 \begin{itemize}
@@
 \end{itemize}
-
+\itemskip
 The main innovation that makes amortized analysis work
@@
 A distinctive feature of our analysis system is that
-it reduces bound derivation to a linear optimization problem
+it reduces linear bound inference  to a linear optimization problem
 that can solved by off-the-shelf LP solvers.  This enables the
 efficient inference of global bounds for larger programs.
+%
+Moreover, our local inference rules automatically generate
+proof certificates that can be easily checked in linear time.
 
@@
 previously-developed quantitative Hoare logic~\cite{veristack14} (see
-Section XXX).  For another thing, we describe a new mechanism that
+\pref{sec:soundness}).  For another thing, we describe a new mechanism that
 allows the separation of quantitative and qualitative verification
-(see Section XXX).  Using this mechanism, the user can guide the
+(see \pref{sec:anno}).  Using this mechanism, the user can guide the
 analysis by using auxiliary variables and logical assertions that can
@@
 
-We have implemented the automatic analysis in the tool
+We have implemented the analysis system in the tool
 \toolname and experimentally evaluated its effectiveness by
@@
 \toolname has automatically derived global resource bounds for
-more than XXX lines of C code from the XXX benchmark suite.
+more than 2000 lines of C code from the cBench benchmark suite.
 %
@@
 asymptotically tight bounds for all but 1 of these patterns, and in
-most cases the derived constant factors are tight.  To compare our
-automatic analyzer with existing techniques, we tested our examples
+most cases the derived constant factors are tight.  To compare $\toolname$ 
+with existing techniques, we tested our examples
 with tools such as KoAT~\cite{BrockschmidtEFFG14},
@@
 uses amortization techniques, is able to achieve a similar precision.
-%
+
 Several micro benchmarks demonstrate the practicality and
@@
 depends on the contents of an array to describe the exact cost of a
-function that finds a maximal element in an array, and a linear
-bound that amortizes the cost of $k$ successive increments to a binary
-counter.
-
-% Give a description of the sections here.                    XXX
-% Jan: We don't need that. Usually, people don't read it.
+function that finds a maximal element in an array, and a bound 
+that amortizes the cost of $k$ successive increments to a binary
+counter (see \pref{sec:anno}).
 
 In summary, we make the following \emph{contributions}.
-
+\itemskip
 \begin{itemize}
@@
   variables to derive global bounds, can handle mutually-recursive
-  functions, generates resource specifications for functions, derives
-  easily verifiable certificates, and handles resources that can
-  become available during an execution.
-\item We show how to reduce the inference of \emph{linear} resource
+  functions, generates resource abstractions for functions, derives
+  proof certificates, and handles resources that can
+  become available during execution.
+\item We show how to automatically reduce the inference of \emph{linear} resource
     bounds to efficient LP solving.
@@
 \item We present experiments with a prototype implementation on more
-  than XXX lines of C code. A detailed comparison shows that our
+  than 2000 lines of C code. A detailed comparison shows that our
   prototype is the only tool that can derive global bounds for larger
@@
 \end{itemize}
+\itemskip
+%
+{\footnotesize \sl
+[To the reviewers: The auxiliary material contains an anonymized
+version of the extended article, and the source code of $\toolname$
+and all examples.] }
 

### 2014-11-11 paper/main.tex
minor

@@
 %
+\iffull{}
 {\footnotesize \sl

### 2014-11-11 paper/main.tex
shortening of section 2

@@
 is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
+\itemskip
+\itemskip
 $$
@@
 n + \Phi'(\state') \; ,
+\vspace{-1.5ex}
 $$
-then we get the following familiar looking rule.
+then we get the following familiar looking rule
+\itemskip
 $$
@@
 {\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
-{\htriple{\Phi} {S_1; S_2}  {\Phi''}}.
+{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
+\vspace{-1.5ex}
 $$
@@
 
+The derivation of a resource bound using potential functions is best
+explained by example.  If we use the tick metric that assigns cost $n$
+to the function call \code{tick(n)} and cost $0$ to all other
+operations then the cost of the following example can be bounded by
+$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
+  % the signs of the input variables in the examples.}
+\begin{lstlisting}[basicstyle=\tt\small]
+while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
+\end{lstlisting}
+To derive this bound, we start with the initial potential $\Phi_0 =
+|[x,y]|$, which
+we also use as the loop invariant.  For the loop body we have (like in
+Hoare logic) to derive a triple like
+$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
+if we utilize the fact that $x<y$ at the beginning of the loop body.
+The reasoning then works as follows. We start with the potential
+$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
+we denote the updated version of $x$ after the assignment by $x'$ then
+the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
+after the assignment \code{x=x+1} holds.  This means that we have the
+potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
+\code{tick(1)} consumes one resource unit, we end up with potential
+$|[x,y]|$ after the loop body and have established the loop invariant
+again.
+
 \begin{figure}[t]
   \centering
+\vspace{-1.5ex}
 \begin{lstlisting}[mathescape]
@@
 \end{figure}
+
+\pref{fig:ex1} shows a derivation of the bound
+$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
+version of Example~1 in which we increment $x$ by a constant $K>0$ and
+consume $T>0$ resources in each iteration.  The reasoning is similar
+to the one of Example~1 except that we obtain the potential
+$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
+assertions in $\Gamma$ are only used in the rule for the assignment
+\code{x=x+K}.
+
+To the best of our knowledge, no other implemented tool for C is
+currently capable of deriving a tight bound on the cost of such a
+loop.  For $T=1$ (many systems focus on the number of loop iterations
+without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
+10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
+$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
+translate the program into a term-rewriting system by
+hand. Nevertheless, we will show in \pref{sec:AAA} that the
+potential method makes automatic bound derivation straightforward.
+
+% To automate the reasoning, we first introduce an unknown rational
+% variable for each factor in the potential functions.  We then use our
+% inference rules (see \pref{sec:AAA}) to emit linear constraints on
+% these variables that enforce that variable assignments that respect
+% the constraints correspond to sound potential annotations.  For
+% instance if $K = 1$ in \pref{fig:ex1} then we would have the
+% annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
+% |[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
+% p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
+% are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
+% $p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
+% q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
+% the coefficients of the initial potential function. LP solvers are
+% among the most mature constraint solving tools. As a result, the bound
+% derivation works very efficiently in practice.
+
+% \iffull{It might be surprising that we}{We} only track simple linear relations in
+% the logical context $\Gamma$.  While it would of course be possible to keep
+% track of more sophisticated assertions, this simple form is
+% sufficient for the examples we considered and we can efficiently
+% decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
+% assignment rule.  Nevertheless, this logical part of our analysis
+% system is orthogonal to the quantitative part and can be easily
+% extended if necessary.
+
+The concept of a potential function is a generalization of the
+concept of a ranking function.  A potential function can be used like
+a ranking function if we use the tick metric and add the statement
+\code{tick(1)} to every back edge of the program (loops and function
+calls).  However, a potential function is more flexible.  For example,
+we can use a potential function to prove that Example~2 does not
+consume any resources in the tick metric.
+\begin{lstlisting}[basicstyle=\tt\small]
+#\!\!\!\!\!# while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
+\end{lstlisting}
+Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
+In both cases, we reason exactly like in the first version of the
+while loop to prove the bound.  Of course, such loops with different
+tick annotations can be seamlessly combined in a larger program.
+\begin{lstlisting}[basicstyle=\tt\small]
+while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
+\end{lstlisting}
+
+\iffull{In general, the potential based approach addresses several challenges
+in resource bound analysis. This includes amortized bounds in which
+individual loop iterations have different resource consumption,
+sequential loops or function calls that require the tracking of size
+changes of variables, and nested loops that interact by modifying the
+same set of variables.}{}
 
@@
 \end{figure*}
-
-The derivation of a resource bound using potential functions is best
-explained by example.  If we use the tick metric that assigns cost $n$
-to the function call \code{tick(n)} and cost $0$ to all other
-operations then the cost of the following example can be bounded by
-$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
-  % the signs of the input variables in the examples.}
-\begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
-\end{lstlisting}
-To derive this bound, we start with the initial potential $\Phi_0 =
-|[x,y]|$, which
-we also use as the loop invariant.  For the loop body we have (like in
-Hoare logic) to derive a triple like
-$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
-if we utilize the fact that $x<y$ at the beginning of the loop body.
-The reasoning then works as follows. We start with the potential
-$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
-we denote the updated version of $x$ after the assignment by $x'$ then
-the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
-after the assignment \code{x=x+1} holds.  This means that we have the
-potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
-\code{tick(1)} consumes one resource unit, we end up with potential
-$|[x,y]|$ after the loop body and have established the loop invariant
-again.
-
-\pref{fig:ex1} shows a derivation of the bound
-$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
-version of Example~1 in which we increment $x$ by a constant $K>0$ and
-consume $T>0$ resources in each iteration.  The reasoning is similar
-to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions in $\Gamma$ are only used in the rule for the assignment
-\code{x=x+K}.
-
-To the best of our knowledge, no other implemented tool for C is
-currently capable of deriving a tight bound on the cost of such a
-loop.  For $T=1$ (many systems focus on the number of loop iterations
-without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
-10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
-$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
-translate the program into a term-rewriting system by
-hand. Nevertheless, we will show in \jan{Next section} that the
-potential method makes automatic bound derivation straightforward.
-
-% To automate the reasoning, we first introduce an unknown rational
-% variable for each factor in the potential functions.  We then use our
-% inference rules (see \pref{sec:AAA}) to emit linear constraints on
-% these variables that enforce that variable assignments that respect
-% the constraints correspond to sound potential annotations.  For
-% instance if $K = 1$ in \pref{fig:ex1} then we would have the
-% annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
-% |[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
-% p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
-% are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
-% $p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
-% q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
-% the coefficients of the initial potential function. LP solvers are
-% among the most mature constraint solving tools. As a result, the bound
-% derivation works very efficiently in practice.
-
-% \iffull{It might be surprising that we}{We} only track simple linear relations in
-% the logical context $\Gamma$.  While it would of course be possible to keep
-% track of more sophisticated assertions, this simple form is
-% sufficient for the examples we considered and we can efficiently
-% decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
-% assignment rule.  Nevertheless, this logical part of our analysis
-% system is orthogonal to the quantitative part and can be easily
-% extended if necessary.
-
-Note that our notion of a potential function is a generalization of the
-concept of a ranking function.  A potential function can be used like
-a ranking function if we use the tick metric and add the statement
-\code{tick(1)} to every back edge of the program (loops and function
-calls).  However, a potential function is more flexible.  For example,
-we can use a potential function to prove that Example~2 does not
-consume any resources in the tick metric.
-\begin{lstlisting}[basicstyle=\tt\small]
-#\!\!\!\!\!# while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
-\end{lstlisting}
-Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
-In both cases, we reason exactly like in the first version of the
-while loop to prove the bound.  Of course, such loops with different
-tick annotations can be seamlessly combined in a larger program.
-\begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
-\end{lstlisting}
-
-In general, the potential based approach addresses several challenges
-in resource bound analysis. This includes amortized bounds in which
-individual loop iterations have different resource consumption,
-sequential loops or function calls that require the tracking of size
-changes of variables, and nested loops that interact by modifying the
-same set of variables.
 

### 2014-11-11 paper/main.tex
compress the rules
cosmetics in the rules figure

@@
 \begin{mathpar}
-\iffull{
-%
-\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
-{ }
-{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!!}
-%
-\and
-\Rule{Q:Break}
-{ }
-{\!\!  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') \!\!}
-%
-\and
-\Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
-{ }
-{ \!\! B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) \!\!}
-%
-\and }{}
-\Rule[leftskip=.2cm,rightskip=.1cm]{Q:Return}
+\Rule[leftskip=1.2cm,rightskip=.5cm]{Q:Ret}
 { P = Q_R[\Vret/x]
@@
 %
-\and \Rule[leftskip=.1cm,rightskip=.2cm]{Q:Update}
+\and \Rule[leftskip=.5cm,rightskip=1.2cm]{Q:Update}
 { q'_{xy}, q'_{yx} \in \Qplusz
@@
 \\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\ \textstyle q'_{0y} = q_{0y}
+\\\\ \textstyle q'_{0y} = q_{0y}
       + \sum_{u \in \mathcal U} q_{xu}
@@
 %
-\and \Rule{Q:DecP}
+\and \Rule[leftskip=1cm,rightskip=.5cm]{Q:DecP}
 { \Gamma \models y \ge 0
@@
 %
-\and \Rule{Q:Inc}
+\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
 { M = M_u + M_e(x {\pm} y)
@@
 %
-\and\Rule{Q:Seq}
+\and \Rule{Q:Seq}
 {  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
@@
 }
-%
-\iffull{\and
-\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
-{ }
-{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-}{}
 %

### 2014-11-11 paper/main.tex
move logic and syntax to the appendix

@@
 
-\sectskip
-\section{Syntax and Semantics}
-\label{sec:sem}
-\aftersectskip
-
-We implemented our cost semantics and the quantitative Hoare logic in
-Coq for \emph{CompCert Clight}.  Clight is the most abstract
-intermediate language used by CompCert.  Mainly, it is a subset of C
-in which loops can only be exited with a \code{break} statement and
-expressions are free of side effects.
-
-\paragraph{Syntax.}
-
-In this article, we describe our system for a subset of Clight that is
-sufficient to discuss the general ideas.  This subset is given by the
-following grammar.
-%
-\begin{align*}
-S &:= \code{assert}~E
-\mid \code{skip}
-\mid \code{break}
-\mid \code{return}~x
-\mid x \gets E
-\mid x \gets f(x^*)
-\\
-& \mid \code{loop}~S
-\mid \code{if}(E)~S~\code{else}~S
-\mid S;S
-\mid \code{tick}(n)
-\end{align*}
-%
-Expressions $E$ are left abstract in our presentation.  For our
-analysis framework, it is only important that they are side-effect
-free.
-%
-The most notable difference to full Clight is that we can only assign
-to variables and thus do not consider operations that update the heap.
-Moreover, function arguments and return values are assumed to be
-variables.  This is only for simplifying the presentation; in the
-implementation we can deal with heap updates and general function
-calls and returns.  However, we have not implemented our framework for function
-pointers, \code{goto} statements, \code{continue} statements, and
-\code{switch} statements.
-
-We include the built-in primitive $\code{assert}~e$ that terminates
-the program if the argument $e$ evaluates to false and has no effect
-otherwise.  This is useful to express assumptions on the inputs of a
-program for the automatic analysis.  We also add the built-in function
-\code{tick(n)} that can be called with a constant integer $n$ as a
-flexible way to model resource consumption or release (if $n$ is
-negative).
-
-\iffull{
-\begin{figure}[t!]
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:RetSeq}
-{}
-{ (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-\and
-%
-\and
-\Rule{S:RetLoop}
-{}
-{ (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetCall}
-{ \state = (\_, \gamma) \\
-  \state' = (\theta, \gamma)[r \mapsto \state(x)]
-}
-{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_r)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Call}
-{ \fenv f = (\vec x, S_f) \\
-  \state = (\theta, \gamma) \\
-  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
-}
-{ (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep{M}
-(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \cont, \cost')
-}
-%
-\and
-\Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_2, \cont, \cost')
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
-}
-%
-\and
-\Rule{S:Tick}
-{ }
-{ (\state, \code{tick}(n), \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_t(n))
-}
-
-\end{mathpar}
-\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
-\label{fig:opsem}
-\end{figure}
-}{}
-
-\ifshort{
-\begin{figure}[t!]
-\small
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
-}
-\end{mathpar}
-\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
-\label{fig:opsem}
-\end{figure}
-}{}
-
 \ifshort{
@@
 
-\iffull{
+\sectskip
+\section{Experimental Evaluation}
+\label{sec:exper}
+\aftersectskip
+
+We have experimentally evaluated the practicality of our automatic
+amortized analysis with more than 30 challenging loop and recursion
+patterns from open-source code and the
+literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
+examples \iffull{together with the derived bounds is given in
+  \pref{app:cat}.}{is given in the extended
+  version~\cite{anon_extended}.}
+
+\pref{fig:compar} shows five representative loop patterns from the
+evaluation.  Example \emph{t08} is a slightly modified version of
+\emph{t08a}, which is described in~\pref{sec:inform}.  Example
+\emph{t19} demonstrates the compositionaliy of the analysis.  The
+program consists of two loops that decrement a variable $i$.  In the
+first loop, $i$ is decremented down to 100 and in the second loop $i$
+is decremented further down to $-1$.  However, between the loops we
+assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
+|[0,k]|$ increments.  Our analysis finds this tight bound because our
+amortized analysis naturally takes into account the relation between
+the two loops.
+
+At first sight, example \emph{t30} appears to be a simple loop that
+decrements the variable $x$ down to zero.  However, a closer look
+reveals that the loop actually decrements both input variables $x$ and
+$y$ down to zero before terminating.  In the loop body, first $x$ is
+decremented by one.  Then the values of the variables $x$ and $y$ are
+switched using the local variable $t$ as a buffer.  Our analysis
+infers the tight bound $|[0,x]|+|[0,y]|$.  Sometimes we need some
+assumptions on the inputs in order to derive a bound.  Example
+\emph{t15} is such a case.  We assume here that the input variable $y$
+is non-negative and write \code{assert(y>=0)}.  If we enter the loop
+then we know that $x>0$ and we can obtain constant potential from the
+assignment \code{x=x-1}.  After the assignment we know that $x\geq y$
+and $y\geq 0$.  As a consequence, we can share the potential
+$|[0,x]|$ before the assignment \code{x=x-y} between $|[0,x]|$ and
+$|[0,y]|$ after the assignment.  In this way, we derive a tight linear
+bound.
+
+Example \emph{t13} shows how amortization can be used to find linear
+bounds for nested loops.  The outer loop is iterated $|[0,x]|$ times.
+In the conditional, we either (the branching condition is arbitrary)
+increment the variable $y$ or we execute an inner loop in which $y$ is
+counted back to $0$.  The analysis computes a tight bound.
+\iffull{Again, the constants $0$ and $1$ in the
+  inner loop can as well be replace by something more interesting, say
+  $9$ and $10$ like in example \emph{t08}.  Then we still obtain a
+  tight linear bound.}{}
+
+\begin{figure}[t]
+  \centering
+    \begin{minipage}[b]{.9\linewidth}
+    \begin{center}
+   \begin{lstlisting}
+int srch(
+  int t[], int n,  /* haystack */
+  int p[], int m,  /* needle */
+  int b[]
+) { int i=0, j=0, k=-1;
+    while (i < n) {
+      while (j >= 0 && t[i]!=p[j]) {
+        k = b[j];
+        assert(k > 0 && k <= j + 1);
+        j -= k; tick(1)
+      }
+      i++, j++;
+      if (j == m) break;
+      tick(1);
+    }
+    return i;
+  }
+   \end{lstlisting}
+\vspace{-1.2cm}
+\hspace{3.8cm}
+\footnotesize
+\begin{tabular}{|rl|}
+\hline
+\toolname & $1 {+} 2|\inter 0 n|$ \\
+Rank & $O(n^2)$ \\
+LOOPUS & $2 {+} 3\max(n, 0)$\\
+\hline
+\end{tabular}
+    \end{center}
+\vspace{-0.1cm}
+  \end{minipage}
+  \caption{The Knuth-Morris-Pratt algorithm for string search.}
+  \label{fig:KMP}
+\end{figure}
+
+Finally, \pref{fig:KMP} contains the search function of the
+Knuth-Morris-Pratt algorithm for string search.  Our automatic
+amortized analysis finds the tight linear bound $1+2|[0,n]|$.  We need
+to assert that the elements $b[j]$ of the failure table $b$ are in the
+interval $[1,j+1]$.  This is guaranteed by
+construction of the table in the initialization procedure of the
+algorithm, which we can also analyze automatically.  We need the
+assertion since we do not infer any logical assertions on the contents
+of the heap.  Rank derives a complex quadratic bound and LOOPUS
+derives a linear bound.
+
+To compare our tool with existing work, we focused on loop bounds and
+use a simple metric that counts the number of back edges (i.e., number
+of loop iterations) that are followed in the execution of the program
+because most other tools only bound this specific cost. In
+\pref{fig:compar}, we show the bounds we derived (\toolname) together with
+the bounds derived by LOOPUS~\cite{SinnZV14} and
+Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
+have not been able to obtain this tool.
+KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
+cannot operate on C code and the examples would need to be manually
+translated into a term-rewriting system to be analyzed by these tools.  For
+Rank it is not totally clear how the computed bound relates to the C
+program since the computed bound is for transitions in an automaton
+that is derived from the C code.  For instance, the bound $2+y-x$ that
+is derived for \emph{t08} only applies to the first loop in the
+program.
+
+\begin{table}[t]
+\centering
+\begin{tabular}{r|c|c|c|c|c}
+& KoAT & Rank & LOOPUS & SPEED & \toolname \\
+\hline
+\#bounds & 9 & 24 & 20 & 14 & 32
+\\
+\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
+\\
+\#best bounds & 0 & 0 & 11 & 14 & 29
+\\
+\#tested & 14 & 33 & 33 & 14 & 33
+\end{tabular}
+\vspace{.1cm}
+\caption{Comparison of our automatic amortized analysis with other
+  automatic tools.  We have not been able to obtain a version of
+  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
+  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
+  does currently not work on C programs and we use the bounds that
+  have been reported in the author's experimental evaluation.}
+\label{tab:compar}
+\end{table}
+
+\pref{tab:compar} summarizes the results of our experiments.  It shows
+for each tool the number of derived bounds (\#bounds), the number of
+asymptotically tight bounds (\#lin. bounds), the number of bounds with
+the best constant factors in comparison with the other tools (\#best
+bounds), and the number of examples that we were able to test with the
+tool (\#tested).  Since we were not able to run the experiments for KoAT
+and SPEED, we simply used the bounds that have been reported by the
+authors of the respective tools.  The results show that our automatic
+amortized analysis outperforms the existing tools on our example
+programs.  However, this experimental evaluation has to be taken with
+a grain of salt.  Our goal in this work is not to set a new standard
+for automatic bound analysis but only to show that our approach has
+advantages on examples with linear bounds.  Overall the existing tools
+are more powerful since they can derive polynomial bounds and support
+more features of C.  We were particularly impressed by LOOPUS which is
+very robust, works on large C files, and derives very precise bounds.
+We did not include the running times of the tools in the table since
+all tested tools work very efficiently and need less then a second
+on every tested example.
+
+Since we have a formal cost semantics, we can run our examples with
+this semantics for different inputs and measure the cost to compare it
+to our derived bound.  \pref{fig:3d} shows such a comparison for
+example \emph{t08}.  One can see that the derived constant factors are the
+best possible if the input variable $x$ is non-negative.  If $x$ is
+negative then the bound is only slightly off.
+
+\begin{table}[t]
+\centering
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+\end{tabular}
+\vspace{.1cm}
+\caption{Evaluation of our automatic amortized analysis on larger
+C code from the cBench benchmark suite. The function analyzed might
+be calling helper functions which do not appear in the list but have
+to be analyzed.  These functions are accounted for in the LoC column
+where not only the lines count of the analyzed function is called
+but also the line count of all the funtions it calls. The line count
+is computed on pre-processed C with all unnecessary declarations
+removed.}
+\label{tab:compar}
+\end{table}
+
+\sectskip
+\section{Related Work}
+\label{sec:related}
+\aftersectskip
+
+Most closely related to this article is \iffull{our}{a} previous work
+on end-to-end stack-bound verification of Clight
+programs~\cite{veristack14}.  In that work, \iffull{we}{the authors}
+have implemented and verified a quantitative logic to reason about
+stack-space usage.  In this article, we present a more general
+quantitative Hoare logic that is parametric in the resource of
+interest.  The main innovation is the novel automatic amortized analysis for
+Clight programs that computes logical derivations for non-trivial
+bounds for programs with loops and recursion.
+
+Our work has been inspired by type-based amortized resource analysis
+for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
+are three major improvements over previous work in the current paper,
+which presents the first automatic amortized resource analysis for C.
+First, we solved the long-standing open problem of extending automatic
+amortized resource analysis to compute bounds for programs that loop
+on (possibly negative) integers.  Second, we extended the analysis
+system to deal with non-linear control flow that is introduced by
+\code{break} and \code{return} statements.  Third, for the first time,
+we have combined an automatic amortized analysis with a system for
+interactively deriving bounds.
+
+In the development of our quantitative Hoare logic we have drawn
+inspiration from mechanically verified Hoare logics.
+Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
+logics in Isabelle/HOL has been helpful to understand the interaction
+of auxiliary variables with the consequence rule.  Appel's separation
+logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
+the general structure of the quantitative logic.  \iffull{Since we do not deal
+with memory safety, our logic is much simpler and it would be possible
+to integrate it with Appel's logic.}{}  The continuation passing style
+that we use in the quantitative logic is not only used by
+Appel~\cite{AppelLogic} but also in Hoare logics for low-level
+code~\cite{NiS06,JensenBK13}.
+
+There exist quantitative logics that are integrated into separation
+logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
+quantitative logic.  However, the purpose of these logics is slightly
+different since they focus on the verification of bounds that depend
+on the shape of heap data structures and they are not implemented for
+C.  Also closely related to our logic is a VDM-style logic for
+reasoning about resource usage of an abstract fragment of JVM byte
+code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
+Hoare-style, does not target C code, and is not designed for
+interactive bound development but to produce certificates for bounds
+derived for high-level functional programs.
+
+\begin{figure}[t]
+\center
+\includegraphics[width=.9\linewidth]{fig/bound3d}
+\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
+  (blue lines) and the measured runtime cost (red crosses) for example
+  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
+\label{fig:3d}
+\end{figure}
+
+There exist many tools that can automatically derive loop and
+recursion bounds for imperative programs such as
+SPEED~\cite{GulwaniMC09,GulwaniZ10}, KoAT~\cite{BrockschmidtEFFG14},
+PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
+ABC~\cite{BlancHHK10} and LOOPUS~\cite{Zuleger11,SinnZV14}.  These
+tools are based on abstract interpretation--based invariant generation
+and/or term rewriting techniques, and they derive impressive results
+on realistic software.  The importance of amortization to derive tight
+bounds is well known in the resource analysis
+community~\cite{AlonsoG12,Moser14,SinnZV14}.  Currently, the only
+other available tools that can be directly applied to C code are Rank
+and LOOPUS.  Our analysis framework does not aim to set a new standard
+for automatic bound analysis.  Our contribution is rather a principled
+approach that produces certificates that are proved sound with respect
+to a formal cost semantics.  Moreover, we have a system that enables
+both automatic and interactive bound derivation, a formal soundness
+proof in Coq, and a method that can handle resources that can be
+released (e.g., memory).  However, as we have shown
+in~\pref{sec:exper}, our automatic amortized analysis matches the
+state of the art in automatic bound analysis for linear bounds and
+sometimes even derives better constant factors than the other tools.
+Since Rank and Loopus do not handle recursion, our automatic amortized
+analysis is also the only available tool that can derive bounds for
+recursive C programs.
+
+There are techniques~\cite{Braberman08} that can compute the memory
+requirements of object oriented programs with region-based garbage
+collection.  These systems infer invariants and use external tools
+that count the number of integer points in the corresponding polytopes
+to obtain bonds.  The described technique can handle loops but not
+recursive or composed functions.
+
+We are only aware of two verified quantitative analysis systems.
+Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
+automatically verify previously inferred loop invariants, size
+relations, and ranking functions for Java Card programs.  However,
+they do not have a formal cost semantics and do not prove the bounds
+correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
+have verified a loop bound analysis for CompCert's RTL intermediate
+language.  However, there is no way to interactively derive bounds or
+to deal with resources like memory usage.  Furthermore, Blazy et al.'s
+automatic bound analysis does not compute symbolic bounds.
+
+\sectskip
+\section{Conclusion}
+\label{sec:concl}
+\aftersectskip
+
+We have introduced an novel quantitative analysis system for CompCert
+Clight programs.  To the best of our knowledge, this article presents
+the first resource analysis framework for C that makes it possible to
+combine non-trivial automatically derived bounds with interactively
+derived bounds in a proof system that produces verifiable certificates
+for the bounds.  The main technical innovations are a quantitative
+Hoare logic for reasoning about user-defined resource cost and an
+automatic amortized analysis that derives bounds for programs whose
+resource consumption depends on (possibly negative) integers and
+non-sequential control flow as introduced by \code{break} and
+\code{return} statements.
+
+We will continue to improve our framework to reason more precisely
+about resource usage of system software such as the hypervisor kernel
+CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
+  and verifying}{is currently developed and verified}.  For one thing,
+we will improve the automation to derive \emph{polynomial bounds} and
+to handle non-linear size changes, as already developed for functional
+programs~\cite{HoffmannS13}.  For another thing, we will build on
+previous work~\cite{HoffmannMZ13} to generalize the quantitative
+Hoare logic to handle \emph{concurrent programs}\iffull{ with locks and
+lock-free data structures.}{.}
+
+% \acks
+
+% This research is based on work supported in part by NSF grants 1319671
+% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
+% Any opinions, findings, and conclusions contained in this document are
+% those of the authors and do not reflect the views of these agencies.
+
+\ifx\fullversion\undefined{}\else{
+%\clearpage
+\appendix
+
+\newpage
+
+\section{Quantitative Hoare Logic}
+\label{app:logic}
+
 \begin{figure*}
@@
 \end{figure*}
-}{}
-
-\sectskip
-\section{Quantitative Hoare Logic}
-\label{sec:logic}
-\aftersectskip
 
@@
 
-\iffull{In the rule {\sc L:Skip}, we do not have to account for any resource
+In the rule {\sc L:Skip}, we do not have to account for any resource
 consumption.  As a result, the precondition $Q$ can be any (potential)
@@
 \code{return} part of the postcondition are not reachable and can
-therefore be arbitrary.}{}
+therefore be arbitrary.
 
@@
 to account for the execution cost $M_r$ of the return in rule {\sc
-  L:Call}\iffull{ for function calls.}{.}
-
-\iffull{The rule {\sc L:Update} is the standard assignment rule of Hoare
+  L:Call} for function calls.
+
+The rule {\sc L:Update} is the standard assignment rule of Hoare
 logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
@@
 
-}
-{
-The rules {\sc L:Update} and {\sc L:Seq} correspond to the respective
-rules of standard Hoare logic.
-}
 In the rule {\sc L:Loop}, the \code{break} part of the
@@
 only terminate with a \code{skip} or a \code{return}.  The
-precondition $I$ of the loop is the loop invariant.  \iffull{That is why we
+precondition $I$ of the loop is the loop invariant.  That is why we
 require to have potential $I$ available in the precondition of the
-loop body $S$.}{}  In the postcondition of $S$, the potential must be
+loop body $S$.  In the postcondition of $S$, the potential must be
 sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
@@
 
-\iffull{
 The {\sc L:If} is similar to the rule for the conditional in classic
@@
 statement that depends on the integer $n$.
-}{}
 
@@
 for function calls and $M_r$ for \code{return} statements.
-\iffull{This justifies that the {\sc L:Return} rule does not account
-  for resource consumption.}{} The pre- and postcondition $P_f$ and
+This justifies that the {\sc L:Return} rule does not account
+  for resource consumption. The pre- and postcondition $P_f$ and
 $Q_f$ are taken from the function context $\Delta$.  The assertions in
@@
 using the result of the evaluation of the argument variables in the
-current state.  \iffull{Recall that we require that $P_f$ and $Q_f$ do
+current state.  Recall that we require that $P_f$ and $Q_f$ do
   not depend on the local variables on the call stack, that is,
@@
   \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
-  (\theta',\gamma)$.}{} To transfer potential that depends on local
+  (\theta',\gamma)$. To transfer potential that depends on local
 variables of the callee from the precondition $P$ to the postcondition
@@
 relations between global and local variables using logical
-variables\iffull{ (see the following paragraph for details).}{.}
+variables (see the following paragraph for details).
 
@@
 except that instead of using the logical implication $\Rightarrow$ we
-use the quantitative $\models$ that point-wise applies $\ge$.  \iffull{This
+use the quantitative $\models$ that point-wise applies $\ge$.  This
 rule indeed weakens the statement since it requires more resource to
 run the statement and yields less than what has been proved to be
-available after its termination.}{}
-
-\ifx\fullversion\undefined{}\else{
+available after its termination.
+
 \paragraph{Logical Variables and Shallow Embedding.}
@@
 body is proved for every possible logical variable $y$.
-}\fi
 
@@
 \cont $.  This would not be the case if we replaced all occurrences of
-$m$ by $n$ in the definition\iffull{ of validity.}{.}
+$m$ by $n$ in the definition of validity.
 
@@
 assertions, one for each of the possible outcomes of a program
-statement.  \iffull{We define it as follows.
+statement.  We define it as follows.
 \begin{align*}
@@
 & & \land& \safe n Q {\code{skip}} \cont
-\end{align*}}
-{
-We define $\safeK n B R Q \cont \mathop{:=}
-\safe n B {\code{break}} \cont \, \land \,
-\safe n Q {\code{skip}} \cont \, \land \,
-\safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont$.
-
-}
+\end{align*}
 %
@@
 \end{align*}
-where $S_f$ is the body of the function $f$. \iffull{A full
+where $S_f$ is the body of the function $f$. A full
 judgement that mentions a non-empty function context
 $\Delta$ is in fact a guarded statement: it makes
-assumptions on some functions' behavior.}{}  The predicate
+assumptions on some functions' behavior.  The predicate
 $\validC n \Delta$ gives the precise meaning of the
@@
 This means that from any starting state $\state$, $P(\state)$
-provides enough resources for any run of the program $S$.  \iffull{This
+provides enough resources for any run of the program $S$.  This
 setting is actually the main use case of the previous theorem
 which is stated as a stronger result to allow a proof by
-induction.}{}
-
-\sectskip
-\section{Experimental Evaluation}
-\label{sec:exper}
-\aftersectskip
-
-We have experimentally evaluated the practicality of our automatic
-amortized analysis with more than 30 challenging loop and recursion
-patterns from open-source code and the
-literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
-examples \iffull{together with the derived bounds is given in
-  \pref{app:cat}.}{is given in the extended
-  version~\cite{anon_extended}.}
-
-\pref{fig:compar} shows five representative loop patterns from the
-evaluation.  Example \emph{t08} is a slightly modified version of
-\emph{t08a}, which is described in~\pref{sec:inform}.  Example
-\emph{t19} demonstrates the compositionaliy of the analysis.  The
-program consists of two loops that decrement a variable $i$.  In the
-first loop, $i$ is decremented down to 100 and in the second loop $i$
-is decremented further down to $-1$.  However, between the loops we
-assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
-|[0,k]|$ increments.  Our analysis finds this tight bound because our
-amortized analysis naturally takes into account the relation between
-the two loops.
-
-At first sight, example \emph{t30} appears to be a simple loop that
-decrements the variable $x$ down to zero.  However, a closer look
-reveals that the loop actually decrements both input variables $x$ and
-$y$ down to zero before terminating.  In the loop body, first $x$ is
-decremented by one.  Then the values of the variables $x$ and $y$ are
-switched using the local variable $t$ as a buffer.  Our analysis
-infers the tight bound $|[0,x]|+|[0,y]|$.  Sometimes we need some
-assumptions on the inputs in order to derive a bound.  Example
-\emph{t15} is such a case.  We assume here that the input variable $y$
-is non-negative and write \code{assert(y>=0)}.  If we enter the loop
-then we know that $x>0$ and we can obtain constant potential from the
-assignment \code{x=x-1}.  After the assignment we know that $x\geq y$
-and $y\geq 0$.  As a consequence, we can share the potential
-$|[0,x]|$ before the assignment \code{x=x-y} between $|[0,x]|$ and
-$|[0,y]|$ after the assignment.  In this way, we derive a tight linear
-bound.
-
-Example \emph{t13} shows how amortization can be used to find linear
-bounds for nested loops.  The outer loop is iterated $|[0,x]|$ times.
-In the conditional, we either (the branching condition is arbitrary)
-increment the variable $y$ or we execute an inner loop in which $y$ is
-counted back to $0$.  The analysis computes a tight bound.
-\iffull{Again, the constants $0$ and $1$ in the
-  inner loop can as well be replace by something more interesting, say
-  $9$ and $10$ like in example \emph{t08}.  Then we still obtain a
-  tight linear bound.}{}
-
-\begin{figure}[t]
-  \centering
-    \begin{minipage}[b]{.9\linewidth}
-    \begin{center}
-   \begin{lstlisting}
-int srch(
-  int t[], int n,  /* haystack */
-  int p[], int m,  /* needle */
-  int b[]
-) { int i=0, j=0, k=-1;
-    while (i < n) {
-      while (j >= 0 && t[i]!=p[j]) {
-        k = b[j];
-        assert(k > 0 && k <= j + 1);
-        j -= k; tick(1)
-      }
-      i++, j++;
-      if (j == m) break;
-      tick(1);
-    }
-    return i;
-  }
-   \end{lstlisting}
-\vspace{-1.2cm}
-\hspace{3.8cm}
-\footnotesize
-\begin{tabular}{|rl|}
-\hline
-\toolname & $1 {+} 2|\inter 0 n|$ \\
-Rank & $O(n^2)$ \\
-LOOPUS & $2 {+} 3\max(n, 0)$\\
-\hline
-\end{tabular}
-    \end{center}
-\vspace{-0.1cm}
-  \end{minipage}
-  \caption{The Knuth-Morris-Pratt algorithm for string search.}
-  \label{fig:KMP}
+induction.
+
+\section{Syntax and Semantics}
+\label{app:sem}
+
+We implemented our cost semantics and the quantitative Hoare logic in
+Coq for \emph{CompCert Clight}.  Clight is the most abstract
+intermediate language used by CompCert.  Mainly, it is a subset of C
+in which loops can only be exited with a \code{break} statement and
+expressions are free of side effects.
+
+\paragraph{Syntax.}
+
+In this article, we describe our system for a subset of Clight that is
+sufficient to discuss the general ideas.  This subset is given by the
+following grammar.
+%
+\begin{align*}
+S &:= \code{assert}~E
+\mid \code{skip}
+\mid \code{break}
+\mid \code{return}~x
+\mid x \gets E
+\mid x \gets f(x^*)
+\\
+& \mid \code{loop}~S
+\mid \code{if}(E)~S~\code{else}~S
+\mid S;S
+\mid \code{tick}(n)
+\end{align*}
+%
+Expressions $E$ are left abstract in our presentation.  For our
+analysis framework, it is only important that they are side-effect
+free.
+%
+The most notable difference to full Clight is that we can only assign
+to variables and thus do not consider operations that update the heap.
+Moreover, function arguments and return values are assumed to be
+variables.  This is only for simplifying the presentation; in the
+implementation we can deal with heap updates and general function
+calls and returns.  However, we have not implemented our framework for function
+pointers, \code{goto} statements, \code{continue} statements, and
+\code{switch} statements.
+
+We include the built-in primitive $\code{assert}~e$ that terminates
+the program if the argument $e$ evaluates to false and has no effect
+otherwise.  This is useful to express assumptions on the inputs of a
+program for the automatic analysis.  We also add the built-in function
+\code{tick(n)} that can be called with a constant integer $n$ as a
+flexible way to model resource consumption or release (if $n$ is
+negative).
+
+\begin{figure}[t!]
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~x, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~x, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \fenv f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep{M}
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \cont, \cost')
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_2, \cont, \cost')
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
+}
+%
+\and
+\Rule{S:Tick}
+{ }
+{ (\state, \code{tick}(n), \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
+}
+
+\end{mathpar}
+\caption{Rules of the operational semantics of statements.}
+\label{fig:opsem}
 \end{figure}
-
-Finally, \pref{fig:KMP} contains the search function of the
-Knuth-Morris-Pratt algorithm for string search.  Our automatic
-amortized analysis finds the tight linear bound $1+2|[0,n]|$.  We need
-to assert that the elements $b[j]$ of the failure table $b$ are in the
-interval $[1,j+1]$.  This is guaranteed by
-construction of the table in the initialization procedure of the
-algorithm, which we can also analyze automatically.  We need the
-assertion since we do not infer any logical assertions on the contents
-of the heap.  Rank derives a complex quadratic bound and LOOPUS
-derives a linear bound.
-
-To compare our tool with existing work, we focused on loop bounds and
-use a simple metric that counts the number of back edges (i.e., number
-of loop iterations) that are followed in the execution of the program
-because most other tools only bound this specific cost. In
-\pref{fig:compar}, we show the bounds we derived (\toolname) together with
-the bounds derived by LOOPUS~\cite{SinnZV14} and
-Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
-have not been able to obtain this tool.
-KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
-cannot operate on C code and the examples would need to be manually
-translated into a term-rewriting system to be analyzed by these tools.  For
-Rank it is not totally clear how the computed bound relates to the C
-program since the computed bound is for transitions in an automaton
-that is derived from the C code.  For instance, the bound $2+y-x$ that
-is derived for \emph{t08} only applies to the first loop in the
-program.
-
-\begin{table}[t]
-\centering
-\begin{tabular}{r|c|c|c|c|c}
-& KoAT & Rank & LOOPUS & SPEED & \toolname \\
-\hline
-\#bounds & 9 & 24 & 20 & 14 & 32
-\\
-\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
-\\
-\#best bounds & 0 & 0 & 11 & 14 & 29
-\\
-\#tested & 14 & 33 & 33 & 14 & 33
-\end{tabular}
-\vspace{.1cm}
-\caption{Comparison of our automatic amortized analysis with other
-  automatic tools.  We have not been able to obtain a version of
-  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
-  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
-  does currently not work on C programs and we use the bounds that
-  have been reported in the author's experimental evaluation.}
-\label{tab:compar}
-\end{table}
-
-\pref{tab:compar} summarizes the results of our experiments.  It shows
-for each tool the number of derived bounds (\#bounds), the number of
-asymptotically tight bounds (\#lin. bounds), the number of bounds with
-the best constant factors in comparison with the other tools (\#best
-bounds), and the number of examples that we were able to test with the
-tool (\#tested).  Since we were not able to run the experiments for KoAT
-and SPEED, we simply used the bounds that have been reported by the
-authors of the respective tools.  The results show that our automatic
-amortized analysis outperforms the existing tools on our example
-programs.  However, this experimental evaluation has to be taken with
-a grain of salt.  Our goal in this work is not to set a new standard
-for automatic bound analysis but only to show that our approach has
-advantages on examples with linear bounds.  Overall the existing tools
-are more powerful since they can derive polynomial bounds and support
-more features of C.  We were particularly impressed by LOOPUS which is
-very robust, works on large C files, and derives very precise bounds.
-We did not include the running times of the tools in the table since
-all tested tools work very efficiently and need less then a second
-on every tested example.
-
-Since we have a formal cost semantics, we can run our examples with
-this semantics for different inputs and measure the cost to compare it
-to our derived bound.  \pref{fig:3d} shows such a comparison for
-example \emph{t08}.  One can see that the derived constant factors are the
-best possible if the input variable $x$ is non-negative.  If $x$ is
-negative then the bound is only slightly off.
-
-\begin{table}[t]
-\centering
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-\end{tabular}
-\vspace{.1cm}
-\caption{Evaluation of our automatic amortized analysis on larger
-C code from the cBench benchmark suite. The function analyzed might
-be calling helper functions which do not appear in the list but have
-to be analyzed.  These functions are accounted for in the LoC column
-where not only the lines count of the analyzed function is called
-but also the line count of all the funtions it calls. The line count
-is computed on pre-processed C with all unnecessary declarations
-removed.}
-\label{tab:compar}
-\end{table}
-
-\sectskip
-\section{Related Work}
-\label{sec:related}
-\aftersectskip
-
-Most closely related to this article is \iffull{our}{a} previous work
-on end-to-end stack-bound verification of Clight
-programs~\cite{veristack14}.  In that work, \iffull{we}{the authors}
-have implemented and verified a quantitative logic to reason about
-stack-space usage.  In this article, we present a more general
-quantitative Hoare logic that is parametric in the resource of
-interest.  The main innovation is the novel automatic amortized analysis for
-Clight programs that computes logical derivations for non-trivial
-bounds for programs with loops and recursion.
-
-Our work has been inspired by type-based amortized resource analysis
-for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
-are three major improvements over previous work in the current paper,
-which presents the first automatic amortized resource analysis for C.
-First, we solved the long-standing open problem of extending automatic
-amortized resource analysis to compute bounds for programs that loop
-on (possibly negative) integers.  Second, we extended the analysis
-system to deal with non-linear control flow that is introduced by
-\code{break} and \code{return} statements.  Third, for the first time,
-we have combined an automatic amortized analysis with a system for
-interactively deriving bounds.
-
-In the development of our quantitative Hoare logic we have drawn
-inspiration from mechanically verified Hoare logics.
-Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
-logics in Isabelle/HOL has been helpful to understand the interaction
-of auxiliary variables with the consequence rule.  Appel's separation
-logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
-the general structure of the quantitative logic.  \iffull{Since we do not deal
-with memory safety, our logic is much simpler and it would be possible
-to integrate it with Appel's logic.}{}  The continuation passing style
-that we use in the quantitative logic is not only used by
-Appel~\cite{AppelLogic} but also in Hoare logics for low-level
-code~\cite{NiS06,JensenBK13}.
-
-There exist quantitative logics that are integrated into separation
-logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
-quantitative logic.  However, the purpose of these logics is slightly
-different since they focus on the verification of bounds that depend
-on the shape of heap data structures and they are not implemented for
-C.  Also closely related to our logic is a VDM-style logic for
-reasoning about resource usage of an abstract fragment of JVM byte
-code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
-Hoare-style, does not target C code, and is not designed for
-interactive bound development but to produce certificates for bounds
-derived for high-level functional programs.
-
-\begin{figure}[t]
-\center
-\includegraphics[width=.9\linewidth]{fig/bound3d}
-\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for example
-  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
-\label{fig:3d}
-\end{figure}
-
-There exist many tools that can automatically derive loop and
-recursion bounds for imperative programs such as
-SPEED~\cite{GulwaniMC09,GulwaniZ10}, KoAT~\cite{BrockschmidtEFFG14},
-PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
-ABC~\cite{BlancHHK10} and LOOPUS~\cite{Zuleger11,SinnZV14}.  These
-tools are based on abstract interpretation--based invariant generation
-and/or term rewriting techniques, and they derive impressive results
-on realistic software.  The importance of amortization to derive tight
-bounds is well known in the resource analysis
-community~\cite{AlonsoG12,Moser14,SinnZV14}.  Currently, the only
-other available tools that can be directly applied to C code are Rank
-and LOOPUS.  Our analysis framework does not aim to set a new standard
-for automatic bound analysis.  Our contribution is rather a principled
-approach that produces certificates that are proved sound with respect
-to a formal cost semantics.  Moreover, we have a system that enables
-both automatic and interactive bound derivation, a formal soundness
-proof in Coq, and a method that can handle resources that can be
-released (e.g., memory).  However, as we have shown
-in~\pref{sec:exper}, our automatic amortized analysis matches the
-state of the art in automatic bound analysis for linear bounds and
-sometimes even derives better constant factors than the other tools.
-Since Rank and Loopus do not handle recursion, our automatic amortized
-analysis is also the only available tool that can derive bounds for
-recursive C programs.
-
-There are techniques~\cite{Braberman08} that can compute the memory
-requirements of object oriented programs with region-based garbage
-collection.  These systems infer invariants and use external tools
-that count the number of integer points in the corresponding polytopes
-to obtain bonds.  The described technique can handle loops but not
-recursive or composed functions.
-
-We are only aware of two verified quantitative analysis systems.
-Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
-automatically verify previously inferred loop invariants, size
-relations, and ranking functions for Java Card programs.  However,
-they do not have a formal cost semantics and do not prove the bounds
-correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
-have verified a loop bound analysis for CompCert's RTL intermediate
-language.  However, there is no way to interactively derive bounds or
-to deal with resources like memory usage.  Furthermore, Blazy et al.'s
-automatic bound analysis does not compute symbolic bounds.
-
-\sectskip
-\section{Conclusion}
-\label{sec:concl}
-\aftersectskip
-
-We have introduced an novel quantitative analysis system for CompCert
-Clight programs.  To the best of our knowledge, this article presents
-the first resource analysis framework for C that makes it possible to
-combine non-trivial automatically derived bounds with interactively
-derived bounds in a proof system that produces verifiable certificates
-for the bounds.  The main technical innovations are a quantitative
-Hoare logic for reasoning about user-defined resource cost and an
-automatic amortized analysis that derives bounds for programs whose
-resource consumption depends on (possibly negative) integers and
-non-sequential control flow as introduced by \code{break} and
-\code{return} statements.
-
-We will continue to improve our framework to reason more precisely
-about resource usage of system software such as the hypervisor kernel
-CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
-  and verifying}{is currently developed and verified}.  For one thing,
-we will improve the automation to derive \emph{polynomial bounds} and
-to handle non-linear size changes, as already developed for functional
-programs~\cite{HoffmannS13}.  For another thing, we will build on
-previous work~\cite{HoffmannMZ13} to generalize the quantitative
-Hoare logic to handle \emph{concurrent programs}\iffull{ with locks and
-lock-free data structures.}{.}
-
-% \acks
-
-% This research is based on work supported in part by NSF grants 1319671
-% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
-% Any opinions, findings, and conclusions contained in this document are
-% those of the authors and do not reflect the views of these agencies.
-
-\ifx\fullversion\undefined{}\else{
-%\clearpage
-\appendix
-
-\newpage
 

### 2014-11-11 paper/main.tex
surgery to trim fat on Derivation System

@@
 for (;;) {
-    do { l++; tick(1); } while (l < h && nondet());
-    do { h--; tick(1); } while (h > l && nondet());
+    do { l++; tick(1); } while (l < h && *);
+    do { h--; tick(1); } while (h > l && *);
     if (h <= l) break;
@@
 \begin{mathpar}
-\Rule[leftskip=1.2cm,rightskip=.5cm]{Q:Ret}
-{ P = Q_R[\Vret/x]
-\\ \forall i \in \dom{P} . \, p_i = q_i
-}
-{ B; (\Gamma_R,Q_R); (\Gamma_R[\Vret/x], Q) \vdash \code{return}~x \dashv (\Gamma',Q') }
-%
-\and \Rule[leftskip=.5cm,rightskip=1.2cm]{Q:Update}
+\Rule{Q:Update}
 { q'_{xy}, q'_{yx} \in \Qplusz
@@
 %
-\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
 { (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
@@
 %
-\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
+\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
 { \Gamma \models y \ge 0 \!
@@
 %
-\and \Rule[leftskip=1cm,rightskip=.5cm]{Q:DecP}
-{ \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
-%
 \and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
@@
 %
-\and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
-%
-\and \Rule{Q:Seq}
+\\ \Rule{Q:Seq}
 {  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
@@
 { B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
-%
-\and\Rule{Q:Call}
-{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
-\\ \text{Loc} = \text{Locals}(Q) \!
-\\ \forall i \neq j . \, x_i \neq x_j \!
-\\ c \in \Qplusz \!
-\\ Q = P + S \!
-\\ Q' = P' + S
-\\ U = Q_f[\Vargs / \vec x]
-\\ U' = Q'_f[\Vret/r]
-\\ \forall i \in \dom{U} . \, p_i = u_i
-\\ \forall i \in \dom{U'} .\, p'_i = u'_i
-\\ \forall i \not\in \dom{U'} .\, p'_i = 0
-\\ \forall i \not\in \text{Loc}
-  .\, s_i = 0
-}
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
-  \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
-}
-%
-\and
-\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
-{ \fenv f = (\vec y, S_f)
-\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
-}
-{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
 %
@@
 %
-\and \Rule{Relax}
+\\ \Rule{Relax}
 { \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
@@
 
-Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
+Most interesting are the rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
 which describe how the potential is distributed after a size change of
-a variable.  The rule {\sc Q:IncP} is for increments $x \gets x + y$
+a variable.
+\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
 and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
@@
 Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-that can be applied if $y$ is not positive.  The rules are equivalent
-in the case where $y=0$.  Finally, the rule {\sc Q:Inc} can be applied
+that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
+$x \gets x+y$, it applies when we can deduce from the logical context
+$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
+can be applied for decrements and when $y$ is not positive.}
+The rules are all equivalent in the case where $y=0$.
+Finally, the rule {\sc Q:Inc} can be applied
 if we cannot deduce any information about $y$.  In the implementation,
@@
 
-To explain how rules for increment and decrement work, it is
+\iffull{To explain how rules for increment and decrement work, it is
 sufficient to understand the rule {\sc Q:IncP}.  The others follow the
-same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
+same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
 variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
@@
 
+\iffull{
 Another interesting rule is {\sc Q:Call}.  It needs to account for the
@@
 $(\Gamma'_f, Q'_f)$.
+}{}
 

### 2014-11-11 paper/main.tex
show a lot less rules for the logic

@@
 weakening because we have to show that the preconditions of the rules
-{\sc L:Weak} or {\sc L:Update}, respectively, are met. %  \iffull{ For instance
-% we prove the following lemma to show the soundness of the translation
-% of {\sc Q:Weak}.
-%
-% \begin{lemma}[Relax]
-%   If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
-%   $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
-% \end{lemma}
-% \begin{proof}
-% We observe the following inequations.
-% \begin{align*}
-% \Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
-% &\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
-%   + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
-%   \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
-%   + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
-%  &\ge \Phi_Q(\state)
-%   + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
-%   + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
-%  &\ge \Phi_Q(\state).
-% \end{align*}
-% Since $\state \models \Gamma$, using the transitivity of $\models$ and
-% the definition of $\mathcal U$ and $\mathcal L$ we get
-% $\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
-% $\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
-% These two inequalities justify the last step.
-% \end{proof}
-% %
-% \noindent
-% We prove similar lemmas for the rules that deal with assignment.  With
-% the computational content of the proof, we have an effective algorithm
-% to construct certificates for upper bounds derived automatically.
-% }{}
-
-% \sectskip
-% \section{Implementation}
-
-% \subsection{Logical Annotations Generation}
-
-% \paragraph{Program conditions.}
-% The first pass in our OCaml implementation generates the
-% logical annotations $\Gamma$ around all program statements.
-% These annotations are used later on by the quantitative
-% analysis to generate constraints.
-% %
-% The first design choice we had to make is to give a fixed form
-% to program conditions.  Because of their generality and because
-% they are closed under negation, we decided to consider
-% inequalities over linear combinations as program conditions.
-% \begin{mathpar}
-% C := L < L \mid L > L \mid L \le L \mid L \ge L
-% \and
-% L := x \mid k \mid k * L \mid L + L \mid L - L
-% \end{mathpar}
-% A Condition is represented internally as a pair of one constant
-% $k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
-% Its intended semantics is
-% $$
-% \llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
-% $$
-% Note that conditions of this shape are closed by negation, as showed
-% by the following equivalence
-% $$
-% \neg (k + \sum_v m_v \cdot H(v) \le 0)
-% \Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
-% $$
-% We can also note that all the ``questions'' asked to the logic
-% assertions during a proof derivation can be described using
-% linear combinations.  Conveniently enough, this language of
-% assertions is exactly Presburger's arithmetic.  This means that
-% we have a sound and complete decision procedure to answer
-% all the questions asked during the analysis.
-
-% \paragraph{Logical contexts.}
-% The second problem related to the logic is how to figure out
-% the logical contexts $\Gamma$ for each program statement.
-% This is a question that is answered by abstract interpretation
-% but, to show that our work really does not need the full power
-% of abstract interpretation, we designed and implemented a very
-% simple top-down approach: starting from a statement $S$ and
-% a pre-condition $\Gamma$ (given as a list of inequalities), we
-% generate $\mathcal P(S, \Gamma)$ a postcondition for $S$ using the
-% following equations:
-
-% \begin{align*}
-% \mathcal P(\code{skip}, \Gamma) &:= \Gamma \\
-% \mathcal P(x \gets y, \Gamma) &:= \{ e \le 0 \mid (e \le 0) \in \Gamma \land \{x\} \# e \} \\
-% \mathcal P(x \gets x + y, \Gamma) &:= \Gamma[x / x - y] \\
-% \mathcal P(x \gets x - y, \Gamma) &:= \Gamma[x / x + y] \\
-% \mathcal P(\code{assert}(C), \Gamma) &:= \Gamma,C \\
-% \mathcal P(S_1; S_2, \Gamma) &:= \mathcal P(S_2, \mathcal P(S_1, \Gamma)) \\
-% \mathcal P(\code{if} (C)~S_1~\code{else}~S_2) &:=
-%   \mathcal P(S_1, \Gamma,C) \lor \mathcal P(S_2, \Gamma,\neg C) \\
-% \mathcal P(\code{while}(C)~S, \Gamma) &:=
-%   \Gamma \lor (\neg C, \mathcal P(S, (C, \{ e \le 0 \mid (e \le 0) \in \Gamma \land {\rm set}(S) \# e \})))
-% \end{align*}
-% We used the notation ${\rm set}(S)$ to designate the set of variables
-% assigned to in the statement $S$ and $s\#e$ to mean that variables
-% in $s$ do not appear in the linear combination $e$.  Because contexts
-% are simply conjunction of conditions we must define $\Gamma \lor \Delta$
-% as a function on contexts:
-% $$
-% \Gamma \lor \Delta := \{ e \le 0 \mid (e \le 0) \in \Gamma \cup \Delta \land \Gamma \models (e \le 0) \land \Delta \models (e \le 0) \}.
-% $$
-% It is important to note that the $\mathcal P$ function defined above
-% is not performing any fixpoint computation.  Despite this flagrant
-% lack of sophistication our complete analysis is still able to figure out some
-% tricky invariants as exemplified earlier.
-% Finally, to generate annotations for the whole program $S$,
-% we run $\mathcal P(S,\bullet)$ (where $\bullet$ is the empty context)
-% and memorize all the intermediate results.  In practice, these results are stored in
-% a table that can be referenced at any time during the analysis to get the
-% pre and post-condition of any statement in the program.
-
-% \paragraph{Decision procedure.}
-% To decide logical entailment $\Gamma \models \Delta$ we use
-% Presburger's decision procedure.  But what this procedure gives us
-% is in fact only an answer the the question: Is there any heap $H$
-% such that $H \models \Gamma$.  While this sentence is existential
-% the question $\Gamma \models \Delta$ is universal.  To use the
-% decision procedure, we use the well known trick of double negation.
-% \begin{align*}
-% \Gamma \models C
-% &\Leftrightarrow \neg\neg (\Gamma \models C) \\
-% &\Leftrightarrow \neg\neg (\forall H.\, H\models\Gamma \implies H\models C) \\
-% &\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\
-% &\Leftrightarrow \neg \exists H, H\models\Gamma \land H\models \neg C \\
-% &\Leftrightarrow \neg \exists H, H\models (\Gamma, \neg C)
-% \end{align*}
-% Thus, to know if $\Gamma \models C$ we ask the decision
-% procedure if $\Gamma, \neg C$ is satisfiable and negate the result.
-% When we want to know about $\Gamma \models \Delta$ we simply ask
-% one question for each of the conjuncts of $\Delta$.
-
-% \subsection{Linear Constraints Generation}
-
-% \paragraph{Generating a linear program.}
-% Because figuring out the coefficents that are in $Q$ right away is a hard
-% problem, we separate the search of a derivation in two steps.  As a first
-% step we go through the whole program and apply inductively the rules
-% given in the proof system section.  During this process our tool uses
-% symbolic names for the coefficients $(q_i)$.  Every time a side condition
-% must be satisfied by these coefficients, it is recorded using the symbolic
-% names in a global list.  When
-% the list of all constraints that must be satisfied by the coefficients is
-% collected, we feed it to an off the shelf LP-solver.  If the solver successfully
-% finds a solution to the given problem, we know that a derivation
-% exists for the considered program.  We can even ask the solver
-% for the initial $Q$ and get a resource bound for the program.
-% If the LP-solver fails to find a solution, an error is reported to the
-% user.
-
-% \paragraph{Making the proof system algorithmic.}
-% One problem with the proof system as described is that it is not
-% fully syntax directed.  The rule {\sc Q:Weak} can be applied at
-% all times.  However, if we apply this weakening rule at every step
-% of the derivation, the number of generated constraints and
-% the number of queries to the logic system explode.  This blowup
-% causes efficiency problems.  To avoid these performance problems
-% our implementation only applies the weakening rules at some
-% strategic points in the derivation.  More specifically, at sequence
-% points, at the beginning of while loops and around conditional
-% statements.  So far, this selection of weakening spots has been
-% driven by practice but it might be possible to prove that these
-% program points are the only ones where weakening is needed.
-
-% \sectskip
-% \section{Polynomial Potential}
-
-% \paragraph{Index Sets}
-
-% Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
-% family that maps two-element sets of variables to natural numbers,
-% that is,
-% $$
-% I = (i_{\{x,y\}})_{\{x,y\} \subseteq V} \; .
-% $$
-% %
-% We identify a family $I$ with the set $\{ (\{x,y\},i_{\{x,y\}})
-% \mid \{x,y\} \subseteq V\}$.
-
-% Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
-% instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
-% from the context.
-% %
-% We assume that allways $0 \in V$ and sometimes write $i_x$ instead of $i_{\{x,0\}}$.
-
-% The \emph{degree} $\deg(I)$ of an index $I = (i_{\{x,y\}})_{\{x,y\}
-%   \subseteq V}$ is defined as
-% $$
-% \deg(I) = \sum_{\{x,y\} \in V} i_{\{x,y\}} \;.
-% $$
-% We define $\ind_k(V) = \{ I \mid I \in \ind(V) \text{ and } \deg(i) \leq k
-% \}$ to be the set of indices of degree at most $k$.
-
-% \paragraph{Resource Polynomials}
-
-% Let $V$ be a set of variables.  An index $I \in \ind(V)$ denotes a
-% \emph{base polynomial} $P_I : \states \to \N$ for $V$ that maps a
-% program state $H$ to product of binomial coefficients (a natural
-% number).  We define
-% $$
-% P_I(H) = \prod_{{\{x,y\}} \subseteq V} \binom {|H(x){-}H(y)|} {i_{\{x,y\}}} \; .
-% $$
-% %
-% A \emph{resource polynomial} $R$ for the variable set $V$ is a
-% non-negative linear combination of the base polynomials for $V$.
-
-% \paragraph{Potential Annotations}
-
-% A \emph{potential annotation} for the variable set $V$ is a family
-% $$Q = (q_I)_{I \in \ind(V)}$$
-% of non-negative rational numbers.  Such an annotation denotes the
-% resource polynomial $R_Q$ that is defined by
-% $$
-% R_Q(H) = \sum_{I \in \ind(V)} q_I \cdot P_I(H) \; .
-% $$
-% %
-% We say that the potential annotation $Q$ is of degree $k$ if $q_I = 0$
-% for $I \in \ind(V)$ with $\deg(I) > k$.
-
-% \paragraph{Additive Shifts}
-
-% Let $Q$ be a potential annotation for a variable set $V$ and let
-% $\{x,y\} \subseteq V$ be a two-element variable set.  The
-% \emph{additive shift} with respect to $\{x,y\}$ is a potential
-% annotation $\shift_{\{x,y\}}(Q) = (q'_I)_{I \in \ind(V)} $ for $V$
-% that is defined through
-% $$
-% q'_I = q_I + q_{I^{\{x,y\}{+}1}} \; .
-% $$
-% For an index $I = (i_{\{x,y\}})_{\{x,y\} \subseteq V}$ we use the
-% notation $I^{\{x,y\}{+}k}$ to denote the index
-% $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
-% $$
-% i'_{\{t,u\}} = \left\{
-%   \begin{array}{ll}
-%     i_{\{t,u\}} + k  & \text{if } \{t,u\} = \{x,y\} \\
-%     i_{\{t,u\}} & \text{otherwise}
-%   \end{array}
-% \right.
-% \;.
-% $$
-% %
-% The additive shift for natural numbers reflects the identity
-% \begin{equation}
-% \label{eq:shift}
-% \sum_{0 {\leq} i \leq {k}} q_i \binom{n+1}{i} = \sum_{0 {\leq} i \leq {k}} (q_i{+}q_{i+1}) \binom{n}{i}
-% \end{equation}
-% where $q_{k+1} = 0$.  It is used in the effect system if the
-% difference $n+1$ between two variables $x,y$ decreases by one.
-
-% \begin{lemma} Let $V$ be a set of variables with $x,y \in V$ and let
-%   $H$ be a program state. Let $|H'(t) {-} H'(u)| = |H(t) {-} H(u)|$
-%   for $\{t,u\} \neq \{x,y\}$ and let $|H'(x) {-} H'(y)| = |H(x) {-}
-%   H(y)| - 1$.
-%   %
-%   If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
-% \end{lemma}
-
-% We now study the effect of multiple simultaneous shifts.  Let $Q$ be a
-% resource annotation for a variable set $V$ and let $U_1,\ldots,U_n
-% \subseteq V$ with $|U_i| = 2$ for all $i$ and $U_i \neq U_j$ for $i
-% \neq j$ be pairwise distinct two-element variable sets.  The
-% simulations additive shift $\shift_{U_1,\ldots,U_n}(Q)$ of $Q$ with
-% respect to $U_1,\ldots,U_n$ is defined by
-% $$
-% \shift_{U_1,\ldots,U_n}(Q) = \shift_{U_1}( \cdots \shift_{U_n}(Q) \cdots ) \; .
-% $$
-% %
-% \begin{proposition}
-%   Let $V$ be a set of variables and let $U_1,\ldots,U_n$ be pairwise
-%   distinct two-element variable sets.  Let $|H'(x) {-} H'(y)| = |H(x)
-%   {-} H(y)|$ for $\{x,y\} \not\in \{U_1,\ldots,U_n\}$ and let $|H'(x)
-%   {-} H'(y)| = |H(x) {-} H(y)| - 1$ for $\{x,y\} \in
-%   \{U_1,\ldots,U_n\}$.
-%   %
-%   If $Q' = \shift_{U_1,\ldots,U_n}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
-% \end{proposition}
-% %
-% As shown by the following lemma, the order in which the shifts for the
-% individual $U_i$ are applied is insignificant.
-% %
-% \begin{lemma}
-%   Let $\sigma : \{1,\ldots,n\} \to \{1,\ldots,n\}$ be a
-%   permutation. Then $\shift_{U_1,\ldots,U_n}(Q) =
-%   \shift_{U_{\sigma(1)},\ldots,U_{\sigma(n)}}(Q)$.
-% \end{lemma}
-% %
-% For reasons of efficiency in the constraint generation, we give a more
-% direct formula for the simultaneous shift.  Let $I \in \ind(V)$ and
-% let $U_1,\ldots,U_n$ be pairwise distinct two-element variable sets.
-% We define the index $I^{U_1,\ldots,U_n + k}$ as the family $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
-% $$
-% i'_{\{t,u\}} = \left\{
-%   \begin{array}{ll}
-%     i_{\{t,u\}} + k  & \text{if } \{t,u\} \in \{U_1,\ldots,U_n\} \\
-%     i_{\{t,u\}} & \text{otherwise}
-%   \end{array}
-% \right.  \;.
-% $$
-
-% %
-% \begin{lemma}
-%   Let $V$ be a variable set and let $U_1,\ldots,U_n$ be pairwise
-%   distinct two-element variable sets.
-%   %
-%   Let $Q = (q_I)_{I \in \ind(V)}$ be a resource annotation for
-%   $V$ and let $ Q' = (q'_I)_{I \in \ind(V)}$ where
-%   $$
-%   q'_I = \sum_{\{j_1,\ldots,j_m\} \subseteq \{1,\ldots,n\} } q_{I^{U_{j_1},\ldots,U_{j_m}+1}} \; .
-%   $$
-%   Then $Q' = \shift_{U_1,\ldots,U_n}(Q)$.
-% \end{lemma}
-
-%%
-% Note: Later we need to shift in many directions at once like
-%   Q' = shift_{x,y} (shift_{x,u} (Q))
-% To do: Give a combined formula for that (concise constraint system).
-%%
-
-\ifshort{
+{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
+we prove the following lemma to show the soundness of the translation
+of {\sc Q:Weak}.
+
+\begin{lemma}[Relax]
+  If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
+  $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
+\end{lemma}
+\begin{proof}
+We observe the following inequations.
+\begin{align*}
+\Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
+&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
+  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
+  \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
+  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
+ &\ge \Phi_Q(\state)
+  + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
+  + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
+ &\ge \Phi_Q(\state).
+\end{align*}
+Since $\state \models \Gamma$, using the transitivity of $\models$ and
+the definition of $\mathcal U$ and $\mathcal L$ we get
+$\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
+$\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
+These two inequalities justify the last step.
+\end{proof}
+%
+\noindent
+We prove similar lemmas for the rules that deal with assignment.  With
+the computational content of the proof, we have an effective algorithm
+to construct certificates for upper bounds derived automatically.
+}{}
+
 \begin{figure*}
@@
 %
-\Rule{L:Skip}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+\Rule{L:Update}
 { }
@@
   { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
-\htriple
-  { M_f + P }
-  { r \gets f(\vec x) }
-  { Q - M_r }
 }
@@
 }
-%
-\and
-\Rule{L:If}
+\and \Rule{L:Seq}
 { \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
+\htriple{ P }{ S_1 }{ Q' + M_s }
 \\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
 }
 { \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+\htriple{ P }{ S_1;S_2 }{ Q }
 }
@@
 {
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
   \!\! P \models P' \!\! \\
@@
   \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+  \!\! Q' \models Q \!\!
 }
@@
 %
-\and
-\Rule{L:Extend}
-{ \Delta \cup \Delta'; B; R \vdash
-  \htriple{ P }{ S }{ Q }
-\\
-  \forall f\,P_f\,Q_f.\,
-  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
-  \rightarrow
-  \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
 \end{mathpar}
 
-\caption{Rules of the Quantitative Hoare Logic}
+\caption{Selected rules of the Quantitative Hoare Logic}
 \label{fig:logic}
 \end{figure*}
-}{}
 
@@
 \caption{Rules of the Quantitative Hoare Logic}
-\label{fig:logic}
+\label{fig:logicapp}
 \end{figure*}
@@
 
-\pref{fig:logic} shows the inference rules of the quantitative logic.
+\pref{fig:logicapp} shows the inference rules of the quantitative logic.
 The rules are slightly simplified in comparison to the implemented

### 2014-11-11 paper/main.tex
working on section 3

--- 2014-11-11 paper/main.tex
@@
    \end{lstlisting}
-
+\vspace{-2.5ex}
 $|[x, n]| + |[y, m]|$
@@
    \end{lstlisting}
-
+\vspace{-2.5ex}
 $|[x, n]| + |[z, n]|$
@@
    \end{lstlisting}
-
+\vspace{-2.5ex}
 $3.1|[y,z]| + 0.1|[0,y]|$
@@
    \end{lstlisting}
+\vspace{-2.5ex}
 $59|[n,0]| {+} 0.05|[0,y]|$
@@
   mention the non-zero terms and in the logical context $\Gamma$ we
-  only mention assertions that we actually use in the reasoning. In
+  only mention assertions that we use. In
   Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
@@
 In this section we describe the high-level design of the automatic
-amortized analysis that we implemented in $\toolname$. We then explain
-and motivate our design decisions with examples.
-
+amortized analysis that we implemented in $\toolname$. Examples explain
+and motivate our design decisions.
+
+\paraskip
 \paragraph{Linear Potential Functions.}
@@
 To find resource bounds automatically, we first need to restrict our
-search space.  In this work, we focus on a form that can express tight
-bounds for many typical programs and allows inference view
-\emph{linear programming}.  The general form of \emph{potential
-  functions} that we consider is
+search space.  In this work, we focus on a the following form of
+potential functions, which can express tight bounds for many typical
+programs and allows inference view \emph{linear programming}.
 $$
@@
 
+\paraskip
 \paragraph{Abstract Program State.}
@@
 
-\paragraph{Examples: Challenging Loop Patterns.}
+\begin{figure*}
+ \setlength{\progwidth}{.30\linewidth}
+  \centering
+\hspace{-0.4cm}
+  \begin{minipage}[b]{.4\linewidth}
+    \begin{center}
+   \begin{lstlisting}[]
+void count_down (int x,int y) {
+  if (x>y) { tick(1); count_up(x-1,y);}
+}
+void count_up (int x, int y) {
+  if (y+1<x) { tick(1); count_down(x,y+2);}
+}
+   \end{lstlisting}
+\vspace{-2.5ex}
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
+\\[.4\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+   \end{lstlisting}
+\vspace{-2.5ex}
+$XXX$
+\\[.4\baselineskip]
+      {\bf XXX}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+
+   \end{lstlisting}
+\vspace{-2.5ex}
+$XXX$
+\\[.4\baselineskip]
+      {\bf XXX}
+    \end{center}
+  \end{minipage}
+\vspace{1ex}
+\caption{Exampel \emph{t39} shows two mutually-recursive functions
+  with the computed tick bounds.  The derived constant factors are
+  tight.}
+  \label{fig:ex_list_2}
+\end{figure*}
+
+\paraskip
+\paragraph{Examples: Challenging Loops.}
 
@@
 \code{break} and \code{return} statements, and mutual recursion.
-\pref{fig:ex_rec} contains two mutually-recursive functions with their
+Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
 automatically derived tick bounds.  The function \code{count\_down}
@@
 which is dual to \code{count\_down}.  Here, we
-count up $y$ by $2$ and recursively call \code{count\_down}.  Our
-analysis is the only available system that computes a tight
-bound on this example.  \iffull{The analysis amounts to computing the meeting
+count up $y$ by $2$ and recursively call \code{count\_down}.  $\toolname$ is 
+the only available system that computes a tight
+bound.  \iffull{The analysis amounts to computing the meeting
 point of two trains that approach each other with different speeds.}{}
@@
 
-\begin{figure}[t]
-  \centering
-    \begin{minipage}[b]{\linewidth}
-    \begin{center}
-   \begin{lstlisting}
-void count_down (int x,int y) {
-  if (x>y) { tick(1); count_up(x-1,y); } }
-
-void count_up (int x, int y) {
-  if (y+1<x) { tick(1); count_down(x,y+2); } }
-
-   \end{lstlisting}
-
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
-\\[.4\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-\vspace{-1.5ex}
-
-  \caption{Two mutually-recursive functions with the computed tick bounds.  The derived constant factors are tight.}
-  \label{fig:ex_rec}
-\end{figure}
-
+\paraskip
 \paragraph{Examples: Compositionality.}
@@
 With two concrete examples from open-source projects we demonstrate
-that the compositionality of our method is indeed needed in the
-analysis of common code.
+that the compositionality of our method is indeed crucial in practice.
 

### 2014-11-11 paper/main.tex
still working on section 3

@@
 \begin{figure*}
- \setlength{\progwidth}{.30\linewidth}
   \centering
 \hspace{-0.4cm}
-  \begin{minipage}[b]{.4\linewidth}
+  \begin{minipage}[b]{7cm}
     \begin{center}
    \begin{lstlisting}[]
-void count_down (int x,int y) {
-  if (x>y) { tick(1); count_up(x-1,y);}
-}
-void count_up (int x, int y) {
-  if (y+1<x) { tick(1); count_down(x,y+2);}
+void c_down (int x,int y) {
+  if (x>y) { tick(1); c_up(x-1,y); }
+}
+void c_up (int x, int y) {
+  if (y+1<x) {tick(1); c_down(x,y+2);}
 }
@@
 \vspace{-2.5ex}
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
 \\[.4\baselineskip]
@@
 %
-  \begin{minipage}[b]{\progwidth}
+  \begin{minipage}[b]{5cm}
     \begin{center}
    \begin{lstlisting}
-
+for (; l >= 8; l -= 8)
+  // process one block
+  tick(N);
+for (; l > 0; l--)
+  // save leftovers
+  tick(1);
    \end{lstlisting}
@@
 \\[.4\baselineskip]
-      {\bf XXX}
+      {\bf t61}
     \end{center}
@@
 %
-  \begin{minipage}[b]{\progwidth}
+  \begin{minipage}[b]{5.5cm}
     \begin{center}
@@
 \vspace{-2.5ex}
+\begin{lstlisting}
+for (;;) {
+  do { l++; tick(1); } 
+    while (l < h && *);
+  do { h--; tick(1); } 
+    while (h > l && *);
+  if (h <= l) break;
+  tick(1); /* swap elems. */ }
+\end{lstlisting}
 $XXX$
 \\[.4\baselineskip]
-      {\bf XXX}
+      {\bf t62}
     \end{center}
@@
 Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
-automatically derived tick bounds.  The function \code{count\_down}
+automatically derived tick bounds.  The function \code{c\_down}
 decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{count\_up},
-which is dual to \code{count\_down}.  Here, we
-count up $y$ by $2$ and recursively call \code{count\_down}.  $\toolname$ is 
+$y$. It then recursively calls the function \code{c\_up},
+which is dual to \code{c\_down}.  Here, we
+c up $y$ by $2$ and recursively call \code{c\_down}.  $\toolname$ is 
 the only available system that computes a tight

### 2014-11-12 paper/main.tex
done with section 3

@@
 \vspace{-2.5ex}
-$XXX$
+$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
+$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
 \\[.4\baselineskip]
@@
 \end{lstlisting}
-$XXX$
+$2+3|[l,h]|$
 \\[.4\baselineskip]
@@
 \vspace{1ex}
-\caption{Exampel \emph{t39} shows two mutually-recursive functions
-  with the computed tick bounds.  The derived constant factors are
-  tight.}
+\caption{Example \emph{t39} shows two mutually-recursive functions
+  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
+  demonstrate the unique compositionality of our system. In \emph{61},
+  $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
+  are tight.}
   \label{fig:ex_list_2}
@@
 \paraskip
-\paragraph{Examples: Challenging Loops.}
+\paragraph{Challenging Loops.}
 
@@
 
+\paraskip
+\paragraph{Nested and Sequenced Loops.}
+
 Example \emph{t08a} shows the ability of the analysis to discover
@@
 %
+
+\paraskip
+\paragraph{Mutual Recursive Functions.}
 
@@
 
+\paraskip
+\paragraph{Compositionality.}
+
+With two concrete examples from open-source projects we demonstrate
+that the compositionality of our method is indeed crucial in practice.
+\jan{what projects?}
+
+Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations of block-based
+cryptographic primitives: Data of arbitrary length, is consumed in
+blocks and the leftover is stored in a buffer for future use when more
+data is available.  This looping pattern is particularly well handled
+by our method. If $N \geq 8$, $\toolname$ infers the bound
+$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8}
++ \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the
+cost in the second loop (\code{tick(1)}) are random choices and
+$\toolname$ would also derive tight bound for other values.
+
+To understand the resource bound for the case $N<8$, first note that the cost of
+the second loop is $|[0,l]|$.  After the first loop, we still have
+$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
+to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
+we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
+loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
+potential units instead.  
+%
+This level of precision and compositionality is only achieved by
+our novel analysis, no other available tool derives the aforementioned
+ tight bounds.
+
+Example \emph{t62} (\pref{fig:ex_list_2}) 
+is the inner loop of a quick sort implementation in cBench.  More
+precisely, it is the partitioning part of the algorithm.  This partition
+loop has linear complexity, and feeding into our analysis gives the
+correct and tight worst-case bound $2+3|[l,h]|.$  With the amortized
+analysis approach in mind, it is clear that the three loops get their
+potential from the same interval $[l,h]$.  However, KoAT fails to
+find a bound and LOOPUS derives the quadratic bound
+$(h-l-1)^2$.  Following the classical technique, these tools try to find one
+linear ranking-function for each loop and combine them multiplicatively
+or additively.
+
 \iffull{We only show a small selections of the programs that we can
@@
 tools\iffull{ for automatic bound derivation.}{.}
-
-\paraskip
-\paragraph{Examples: Compositionality.}
-
-With two concrete examples from open-source projects we demonstrate
-that the compositionality of our method is indeed crucial in practice.
-
-The following example is typical in cryptographic primitives
-implementations.  These primitives work by block, so when data of
-arbitrary length needs to be processed, it is munched in blocks and
-the leftovers are stored in a buffer for future use when more data
-is available.  This looping pattern is particularly well handled by
-our tool because the LP solver is leveraged to generate a tight bound.
-The optimization problem it is given encodes that constant bounds are
-preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
-\toolname is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
-+ \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
-counted as constant.
-%
-\jan{N is unbound above. The explanation must be much slower
-  here. Explain what the code does and how the ticks account for the
-  cost. Then: What is the resource bound?. Then: What do we derive?
-  Btw: Why do you use tick(1) and not tick(M)?}
-%
-\begin{lstlisting}
-for (; l >= 8; l -= 8)
-    // process one block
-    tick(N);
-for (; l > 0; l--)
-    // save leftovers
-    tick(1);
-\end{lstlisting}
-%
-It is worth taking some time to explain the constant contribution
-to the resource bound when $N<8$.  Remark first that the cost of
-the second loop is $|[0,l]|$, and after the first one, we still have
-$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
-to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
-we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
-loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
-potential units instead.  This explains the somewhat intriguing value
-of the constant term in the bound.  All this reasoning is done within
-our tool automatically and simply relies on general rules presented
-in further sections.
-
-This level of precision and compositionality is only achieved by
-our work, no other available tool will attain the above tight bounds
-on this common kind of looping code.  Thanks to the modularity of
-amortized analysis the intricate interactions between the bounds of
-the two loops can be modeled as one linear program, and because LP
-solvers can optimize the quality of the final bound we can keep
-it close to the actual resource consumption.
-
-A clean theoretical framework for compositional bounds sometimes
-even allow to get better asymptotic bounds.  During our experiments
-we found an example of this situation in some code from the cBench
-benchmark suite.
-%
-\begin{lstlisting}
-for (;;) {
-    do { l++; tick(1); } while (l < h && *);
-    do { h--; tick(1); } while (h > l && *);
-    if (h <= l) break;
-    tick(1); /* swap two elements */ }
-\end{lstlisting}
-%
-The above snippet is the inner loop of a Quicksort implementation, more
-precisely, it is the partitioning part of the algorithm.  This partition
-loop has linear complexity, and feeding into our analysis gives the
-correct and tight worst-case bound $2+3|[l,h]|$.  With the amortized
-analysis approach in mind, it is clear that the three loops get their
-potential from the same interval $[l,h]$.  However, KoAT fails to
-prove the termination of the loop and LOOPUS derives the quadratic bound
-$(h-l-1)^2$.  Following the classic technique, these tools try to find one
-linear ranking-function for each loop and combine them multiplicatively
-or additively, resulting in a loose quadratic bound or no bound at all.
 
@@
 
+\paraskip
 \paragraph{Judgements.}

### 2014-11-12 paper/main.tex
add example in full version
uniformize code

@@
 void c_down (int x,int y) {
-  if (x>y) { tick(1); c_up(x-1,y); }
+  if (x>y) {tick(1); c_up(x-1,y);}
 }
@@
    \begin{lstlisting}
-for (; l >= 8; l -= 8)
-  // process one block
+for (; l>=8; l-=8)
+  /* process one block */
   tick(N);
-for (; l > 0; l--)
-  // save leftovers
+for (; l>0; l--)
+  /* save leftovers */
   tick(1);
@@
   do { l++; tick(1); }
-    while (l < h && *);
+    while (l<h && *);
   do { h--; tick(1); }
-    while (h > l && *);
-  if (h <= l) break;
+    while (h>l && *);
+  if (h<=l) break;
   tick(1); /* swap elems. */ }
@@
 \end{figure}
+
+\ifdefined\fullversion
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}
+  \mbox{logical state invariant }
+  \{ nm = A(0) \} }$
+i=1; m=a[0];
+while (i < N) {
+  if (a[i] > m) {
+    @assert(nm > 0); nm--;@
+    m = a[i], tick(1);
+  } i++, tick(1);
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We write $A(i)$ for
+  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
+  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
+  }
+\label{fig:xmplmax}
+\end{figure}
+\fi
 
@@
 \begin{table*}
-\small
+\footnotesize
 

### 2014-11-12 paper/main.tex
work on lp section and analysis soundness

@@
 solver to find a solution that satisfies the constraints and minimizes
-the linear expression
+the linear expression \ifshort{$L := 1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x}
++ 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}$.}{
 $$
-1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x} \; .
-$$
+L := 1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x} \; .
+$$}
 The penalties given to certain factors are used
@@
 since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
-We now add the constraint
+We now add the constraint \ifshort{$L \le 5000$}{
 $$
@@
  \le 5000
-$$
+$$}
 to our constraint set and ask the solver to optimize the objective
-function
+function \ifshort{$q_0 + 11{\cdot}q_{0,10}$.}{
 $$
 q_0 + 11{\cdot}q_{0,10} \; .
-$$
+$$}
 This happens in almost no time in practice.
-The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
-the derived bound is $0.5 |[0,x]|$.
+The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise, thus
+the bound is $0.5 |[0,x]|$.
 
@@
 
+\begin{figure*}
+\small
+\begin{mathpar}
+%
+\Rule{L:Update}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
+  { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+\and \Rule{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
+\end{mathpar}
+\caption{Selected rules of the Quantitative Hoare Logic}
+\label{fig:logic}
+\end{figure*}
+
 To prove the soundness of \toolname we found it useful to go through
@@
 work and are simply extended to account for all the costs found in
-our cost-aware semantics.
+our cost-aware semantics.  Some rules are presented in \pref{fig:logic}.
+
+Finally, still following previous work, we define the a continuation based
+soundness of triples and prove the soundness of all rules in Coq.  The
+full version of this paper \cite{anon_extended}, gives more explanations
+on the rules and provides a thorough overview of our soundness proof.
 
@@
 
-We translate judgements of the automatic amortized analysis as defined in
-\pref{fig:auto} into quantitative Hoare triples.  We define a translation
+We use the
+logic described in the previous paragraph as the target of a translation
+function for the automatic derivation system.  This reveals two orthogonal
+aspects of our proof that relies, on one side, on amortized reasoning
+(as encoded in the quantitative logic), and on the other side, on
+combinatorial properties of the linear potential functions that we
+are using (as encoded in the automatic analysis).
+
+We define a translation
 function $\tr$, such that if a judgement $J$ in the automatic analysis
@@
 }{}
-
-\begin{figure*}
-\small
-\begin{mathpar}
-%
-\Rule{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-\and \Rule{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\end{mathpar}
-
-\caption{Selected rules of the Quantitative Hoare Logic}
-\label{fig:logic}
-\end{figure*}
 

### 2014-11-12 paper/main.tex
formatting sections 4 and 5

@@
 %SHORT
-% \lstset{basicstyle=\tt\scriptsize}
-% \abovedisplayskip=3pt
-% \belowdisplayskip=3pt
-% \abovedisplayshortskip=3pt
-% \belowdisplayshortskip=3pt
+\ifx\fullversion\undefined
+\abovedisplayskip=3pt
+\belowdisplayskip=3pt
+\abovedisplayshortskip=3pt
+\belowdisplayshortskip=3pt
+\fi
 %SHORT
@@
 is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
-\itemskip
-\itemskip
 $$
@@
 n + \Phi'(\state') \; ,
-\vspace{-1.5ex}
 $$
 then we get the following familiar looking rule
-\itemskip
 $$
@@
 {\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
-\vspace{-1.5ex}
 $$
@@
   \centering
+\vspace{-.6cm}
 \hspace{-0.4cm}
@@
   \centering
+\vspace{-1.8cm}
 \hspace{-0.4cm}
@@
 \small
+\vspace{-.3cm}
 \begin{mathpar}
@@
 
+\sectskip
 \section{Derivation System}
@@
 $$
-
-These judgements are similar to classic Hoare triples in the presence
-of the early exit statements \code{break} and \code{return}.  The part
-$(\Gamma_B, Q_B)$ is the postcondition used when breaking out of
-a loop.  For example, it is set upon entry of looping statements by
-the rule {\sc Q:Loop} and used before \code{break} statements by the
-rule {\sc Q:Break}.  Similarly $(\Gamma_R, Q_R)$ is the postcondition
-used before returning from a function call.  The part $(\Gamma, Q) \vdash
-S \dashv (\Gamma', Q')$ of the judgement can be seen as a regular
-Hoare triple.  All assertions are split into two parts, the logical
-part and the quantitative part.  The quantitative part $Q$ represents a
-potential function as a collection of non-negative numbers $q_i$ indexed
-by the index set $I$.  The logical part $\Gamma$ is left abstract but
-is enforced by our derivation system to respect classic Hoare logic
-constraints.
+%
+The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
+can be seen as a regular Hoare triple.  All assertions are split into
+two parts, the logical part and the quantitative part.  The
+quantitative part $Q$ represents a potential function as a collection
+of non-negative numbers $q_i$ indexed by the index set $I$.  The
+logical part $\Gamma$ is left abstract but is enforced by our
+derivation system to respect classic Hoare logic constraints.
+%
+The meaning of this basic judgment is as follows: If $S$ is executed
+with starting state $\state$, the assertions in $\Gamma$
+hold, and at least $Q(\state)$ resource units are available then the
+evaluation does not run out of resources and---if the execution
+terminates in state $\state'$---there are at least $Q(\state')$
+resources left and $\Gamma'$ holds for $\state'$.
+
+The judgement is a bit more involved since we have to take into
+account the early exit statements \code{break} and \code{return}.
+This is similar to classical Hoare triples the presence of non-linear
+control flow. In the judgement, $(\Gamma_B, Q_B)$ is the postcondition
+that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
+% of looping statements by the rule {\sc Q:Loop} and used before
+% \code{break} statements by the rule {\sc Q:Break}.  
+Similarly, $(\Gamma_R, Q_R)$ is the postcondition that holds when
+returning from a function call with \code{return}. (This becomes clear
+in the rules.)
 
@@
 coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
-and leaves the latter undefined at some index $i$ we will implicitly
-assume that $q'_i = q_i$.
+and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
 
@@
 semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-a[y]$ are simply handled by assigning zero potential to coefficients
+a[y]$ are handled by assigning $0$ to coefficients
 like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
 
+\paraskip
 \paragraph{Derivation Rules.}
@@
 before the execution of the skip statement is still available after.
+\iffull{
 These two observations would also justify a rule with the side-condition
@@
 weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
-described later.
+described later.}{}
 
@@
 
-Most interesting are the rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
-which describe how the potential is distributed after a size change of
+The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
+describe how the potential is distributed after a size change of
 a variable.
@@
 The rules are all equivalent in the case where $y=0$.
-Finally, the rule {\sc Q:Inc} can be applied
+The rule {\sc Q:Inc} can be applied
 if we cannot deduce any information about $y$.  In the implementation,
 the rules for increments and decrements are combined into one syntax
-directed rule where the specialized rules take priority over {\sc Q:Inc}.
+directed rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
 
@@
 $q_{vx}$ to pay for this change.  The rule ensures this
-only for $v \not\in \mathcal U$ because we know that $x \le v$
+only for $v \not\in \mathcal U$ because $x \le v$
 otherwise, and thus $|\inter v x| = 0$.
@@
 x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
-without gaining potential.
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$\iffull{ without gaining potential.}{.}
 
@@
 \begin{figure*}
+\vspace{-.3cm}
 \centering
@@
 }
+\vspace{-.3cm}
 \end{mathpar}
@@
 \begin{tabular}{l}Constraints:\end{tabular}
+\vspace{-.2cm}
 $$
@@
 \end{tabular}
+\vspace{.1cm}
 \caption{An example derivation as produced by the automatic tool.}
@@
 We separate the search of a derivation into two steps.  As a first
-step we go through the whole program and apply inductively the
-derivation rules of the automatic amortized analysis.  During this
+step we go through the functions of the program and apply inductively
+the derivation rules of the automatic amortized analysis.  This is
+done in a bottom-up way, for each strongly connected component of the
+call graph.  During this
 process our tool uses symbolic names for the rational coefficients
@@
 the symbolic names.  We then feed the collected constraints to an
-off the shelf LP solver. (We currently use Coin-Or's CLP.)
+off-the-shelf LP solver. (We currently use Coin-Or's CLP.)
 If the solver successfully finds a solution, we know that a derivation
@@
 in the derivation.  If the LP solver fails to find a solution, an
-error is reported to the user.
+error is reported.
 
 \pref{fig:derivation} contains an example derivation as it is produced
-by our automatic analyzer.  The upper case letters (with optional
+by our $\toolname$.  The upper case letters (with optional
 superscript) such as $Q^\text{de}$ are families of
@@
 $$
-
-After the constraint generation, the LP solver is provided with a
-so-called objective function to minimize.  We wish to
+%
+After the constraint generation, the LP solver is provided with an
+objective function to minimize.  We wish to
 minimize the initial potential, which constitutes the resource bound on the
@@
 solver to find a solution that satisfies the constraints and minimizes
-the linear expression \ifshort{$L := 1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x}
-+ 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}$.}{
+the linear expression
+\iffull{
 $$
-L := 1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x} \; .
+1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x} \; .
 $$}
+{$
+1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
+$.}
+%
 The penalties given to certain factors are used
@@
 since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
-We now add the constraint \ifshort{$L \le 5000$}{
+We now add the constraint
+\iffull{
 $$
@@
  \le 5000
-$$}
+$$}{
+$
+1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
+ \le 5000
+$
+}
 to our constraint set and ask the solver to optimize the objective
-function \ifshort{$q_0 + 11{\cdot}q_{0,10}$.}{
+function
+\iffull{
 $$
@@
 $$}
+{$
+q_0 + 11{\cdot}q_{0,10}$.}
 This happens in almost no time in practice.
@@
 
-Another interesting note is that the constraints that we generate have
+The constraints that we generate have
 a particularly simple form that is known as \emph{network problem}.

### 2014-11-12 paper/main.tex
resolved conflict
work on the soundness theorem

@@
 
-Finally, still following previous work, we define the a continuation based
+Finally, we define a continuation based
 soundness of triples and prove the soundness of all rules in Coq.  The
@@
 We use the
-logic described in the previous paragraph as the target of a translation
+logic from previous paragraph as the target of a translation
 function for the automatic derivation system.  This reveals two orthogonal
-aspects of our proof that relies, on one side, on amortized reasoning
-(as encoded in the quantitative logic), and on the other side, on
-combinatorial properties of the linear potential functions that we
-are using (as encoded in the automatic analysis).
-
-We define a translation
-function $\tr$, such that if a judgement $J$ in the automatic analysis
+aspects of the proof: on one side, it relies on amortized reasoning
+(the quantitative logic rules), and on the other side, it
+uses combinatorial properties of our linear potential functions
+(the automatic analysis rules).
+
+Technically, we define a translation
+function $\tr$ such that if a judgement $J$ in the automatic analysis
 is derivable, $\tr(J)$ is derivable in the quantitative logic.  By using
 $\tr$ to translate derivations of the automatic analysis to derivations
-in the quantitative logic we can automatically obtain a certified resource
+in the quantitative logic we can directly obtain a certified resource
 bound for the analyzed program.
@@
 where we write $\Phi_Q$ for the unique linear potential function
-defined by the quantitative annotation $Q$.  We also need to
+defined by the quantitative annotation $Q$.  The logical context $\Gamma$
+is implicitely lifted to a quantitative assertion by mapping a state
+$\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
+\iffull{We also need to
 translate the assumptions in function contexts of the automatic
-analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) := \forall z \, \vec v \, v . $
+analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
+\forall z \, \vec v \, v . $
 %
@@
   )
-$$
-%
-These definitions let us translate the judgement $J = B; R; P \vdash
-S \dashv P'$ in the context $\Delta$ by $\tr(J) :=$
-$$
+$$}{}
+%
+\iffull{These definitions let}{This lets} us translate the judgement
+$J := B; R; P \vdash S \dashv P'$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
+\iffull{$$
   (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
     \htriple{\tr(P)}{S}{\tr(P')}.
+$$}{%
 $$
+  \tr(J) := \tr(B); \tr(R) \vdash \htriple{\tr(P)} S {\tr(P')}.
+$$}
 The soundness of the automatic analysis can now be stated
@@
 \begin{theorem}[Soundness of the automatic analysis]
-  If $J$ is a judgement derived with the rules of
-  \pref{fig:auto}, then $\tr(J)$ is a quantitative
-  Hoare triple derivable with the rules of
-  \pref{fig:logic}.
+  If $J$ is a judgement derived by the automatic analyis,
+  then $\tr(J)$ is a quantitative Hoare triple derivable
+  in the quantitative logic.
 \end{theorem}
@@
 rules for increments and decrements and the rule {\sc Q:Weak} for
-weakening because we have to show that the preconditions of the rules
-{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
-we prove the following lemma to show the soundness of the translation
-of {\sc Q:Weak}.
+weakening because they make essential use of the algebraic properties
+of our linear potential functions.
+\iffull{ For instance we prove the following lemma to show the soundness
+of the translation of {\sc Q:Weak}.
 

### 2014-11-12 paper/main.tex
some formatting

@@
   of the second loop depend on the first, and \emph{t27} contains
-  interacting \emph{nested loops}. In the potential functions, we only
+  interacting \emph{nested loops}. \iffull{In the potential functions, we only
   mention the non-zero terms and in the logical context $\Gamma$ we
-  only mention assertions that we use. In
+  only mention assertions that we use.}{} In
   Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
@@
 \small
-\vspace{-.3cm}
+\vspace{-.6cm}
+\def \MathparLineskip {\lineskip=0.35cm}
 \begin{mathpar}
@@
 { Q' \succeq_\Gamma Q }
-
+\vspace{-.3cm}
 \end{mathpar}
@@
 like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+
+\paraskip
+\paragraph{Function Specifications.}
+
+\jan{Here we have to say something about function specifications and
+  function calls.}
+\jan{Here we have to say something about function specifications and
+  function calls.}
+\jan{Here we have to say something about function specifications and
+  function calls.}
+\jan{Here we have to say something about function specifications and
+  function calls.}
+\jan{Here we have to say something about function specifications and
+  function calls.}
+\jan{Here we have to say something about function specifications and
+  function calls.}
 
@@
 \begin{figure*}
-\vspace{-.3cm}
+\vspace{-.6cm}
 \centering
@@
 
-\begin{figure*}
-\centering
-\begin{tabular}{r|c|c|c|c|c}
-%
-& t08 (modified t08a) & t19 & t30 & t15 & t13\\
-%
-\hline
-&
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
-while (y > x) {
-  x++; tick(1);
-}
-while (x > 2) {
-  x -= 3; tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
-while (i > 100) {
-  i--; tick(1) }
-i += k+50;
-while (i >= 0) {
-  i--; tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
-while (x > 0) {
-  x--;
-  t=x, x=y, y=t;
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
-assert(y >= 0);
-while (x > y) {
-  x -= y+1;
-  for (z=y; z > 0; z--)
-    tick(1);
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
-  while (x > 0) {
-    x--;
-    if (*) y++;
-    else
-      while (y > 0) {
-        y--; tick(1); }
-    tick(1); }
-\end{lstlisting}
-%
-\\
-\hline
-%
-%
-%
-AAA
-  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
-  & {\small $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ }
-  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
-  & {\small $|\inter 0 x|$}
-  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
-\\
-%
-Rank &
-{\small $2 {+} y {-} x (?)$ } &
-{\small $54+k+i$ } &
-{\small ---} &
-{\small $2+2x-y$ } &
-{\small $0.5{\cdot}y^2 {+} yx +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
-\\
-%
-LOOPUS &
-{\small
-  \begin{tabular}{c}
-    $\max(0, x{-}2)$\\ $+ 2\max(0, y {-} x)$
-  \end{tabular}
-} &
-{\small
-  \begin{tabular}{c}
-    $\max(0, i{-}100)$ \\
-    $+ \max(0, k {+} i {+} 51)$
-  \end{tabular}
- } &
-{\small ---}% $1 + \max(x, y)$ (???)
-&
-{\small ---}
-&
-{\small $2\max(x, 0) {+} \max(y, 0)$}
-\end{tabular}
-\vspace{.1cm}
-\caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  AAA stands for our automatic
-  amortized analysis for Clight.  The output of Rank has been manually
-  simplified to fit the table.  }
-\label{fig:compar}
-\end{figure*}
-
 \sectskip
@@
 \aftersectskip
-
-\jan{ the following should go somewhere here:
-
-Note that potential functions can access the whole program state, so
-this framework can readily be used to derive bounds that depend on the
-memory contents.
-}
 
@@
 theorem of our Quantitative Hoare Logic.
+
+\jan{ the following should go somewhere here:
+
+Note that potential functions can access the whole program state, so
+this framework can readily be used to derive bounds that depend on the
+memory contents.
+}
 
@@
 }{}
+
+\begin{figure*}
+\centering
+\begin{tabular}{r|c|c|c|c|c}
+%
+& t08 (modified t08a) & t19 & t30 & t15 & t13\\
+%
+\hline
+&
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
+while (y > x) {
+  x++; tick(1);
+}
+while (x > 2) {
+  x -= 3; tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
+while (i > 100) {
+  i--; tick(1) }
+i += k+50;
+while (i >= 0) {
+  i--; tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
+while (x > 0) {
+  x--;
+  t=x, x=y, y=t;
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
+assert(y >= 0);
+while (x > y) {
+  x -= y+1;
+  for (z=y; z > 0; z--)
+    tick(1);
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\scriptsize]
+  while (x > 0) {
+    x--;
+    if (*) y++;
+    else
+      while (y > 0) {
+        y--; tick(1); }
+    tick(1); }
+\end{lstlisting}
+%
+\\
+\hline
+%
+%
+%
+AAA
+  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
+  & {\small $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ }
+  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
+  & {\small $|\inter 0 x|$}
+  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
+\\
+%
+Rank &
+{\small $2 {+} y {-} x (?)$ } &
+{\small $54+k+i$ } &
+{\small ---} &
+{\small $2+2x-y$ } &
+{\small $0.5{\cdot}y^2 {+} yx +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
+\\
+%
+LOOPUS &
+{\small
+  \begin{tabular}{c}
+    $\max(0, x{-}2)$\\ $+ 2\max(0, y {-} x)$
+  \end{tabular}
+} &
+{\small
+  \begin{tabular}{c}
+    $\max(0, i{-}100)$ \\
+    $+ \max(0, k {+} i {+} 51)$
+  \end{tabular}
+ } &
+{\small ---}% $1 + \max(x, y)$ (???)
+&
+{\small ---}
+&
+{\small $2\max(x, 0) {+} \max(y, 0)$}
+\end{tabular}
+\vspace{.1cm}
+\caption{ Comparison of resource bounds derived by different tools on
+  several examples with linear bounds.  AAA stands for our automatic
+  amortized analysis for Clight.  The output of Rank has been manually
+  simplified to fit the table.  }
+\label{fig:compar}
+\end{figure*}
 

### 2014-11-12 paper/main.tex
better spacing
compress section 7

@@
 
-\jan{Here we have to say something about function specifications and
-  function calls.}
-\jan{Here we have to say something about function specifications and
-  function calls.}
-\jan{Here we have to say something about function specifications and
-  function calls.}
 \jan{Here we have to say something about function specifications and
@@
 
-The rules are syntax directed and applied inductively.  For example,
+In general, the weakening rule can be applied after every syntax directed
+rule.  However, it can be left out in practice at some places to
+increase the efficiency of the tool.
+The weakening operation $\succeq_{\Gamma}$ is defined by the rule
+\textsc{Relax}, it is parameterized by a logical context that is
+used to gather information on interval sizes.  For example, we have
+$$
+\begin{array}{lll}
+P^\text{de} \succeq_{(\cdot)} P^\text{we}
+& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
+& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
+& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+\end{array}
+$$
+%
+The other rules are syntax directed and applied inductively.  For example,
 the outermost expression is a loop, so we use the rule \text{Q:Loop}
@@
 
-In general, the weakening rule can be applied after every syntax directed
-rule.  However, it can be left out in practice at some places to
-increase the efficiency of the tool.
-The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}, it is parameterized by a logical context that is
-used to gather information on interval sizes.  For example, we have
-$$
-\begin{array}{lll}
-P^\text{de} \succeq_{(\cdot)} P^\text{we}
-& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
-& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
-& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
-\end{array}
-$$
-%
 After the constraint generation, the LP solver is provided with an
@@
 
-\subsection{Cost Semantics for Clight}
-
+\paragraph{Cost Semantics for Clight.}
 To base the soundness proof on a formal ground, we start by defining a
@@
 parameters indicates the amount of resource consumed by a corresponding
-step in the semantics.  Resources can be released by assigning a negative
+step in the semantics.  Resources can be released by using a negative
 cost. %  The metric used in the implementation can also
@@
 % name of the called function or the number of arguments.
-Two rules for update and tick follow.
+Two sample rules for update and tick follow.
 \begin{mathpar}
@@
 
-\subsection{Quantitative Hoare Logic}
-
-\begin{figure*}
-\small
-\begin{mathpar}
-%
-\Rule{L:Update}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-\and \Rule{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
-\\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\end{mathpar}
-\caption{Selected rules of the Quantitative Hoare Logic}
-\label{fig:logic}
-\end{figure*}
-
-To prove the soundness of \toolname we found it useful to go through
-an intermediate step using a Quantitative Hoare Logic.  This logic
-is at the same time a convenient semantic tool and a clean way to
-interface manual proofs with our automation, as explained in XXX.
-The logic is an extension of the one presented in \cite{veristack14},
-it adds support for arbitrary resources and simplifies the
-handling of auxiliary state.
-
-Following the previous work, we define quantitative Hoare triples
-as $\Delta;B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
-maps from program states to $\Qplusz\cup\{\infty\}$ in order to represent
+% \begin{figure*}
+% \small
+% \begin{mathpar}
+% %
+% \Rule{L:Update}
+% { }
+% { \Delta; B; R \vdash
+% \htriple
+%   { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+% }
+% %
+% \and
+% \Rule{L:Assert}
+% { }
+% { \Delta; B; R \vdash
+% \htriple
+%   { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+%   { \code{assert}~e }
+%   { Q }
+% }
+% %
+% \and
+% \Rule{L:Tick}
+% { }
+% { \Delta; B; R \vdash
+% \htriple
+%   { Q + M_t(n) }
+%   { \code{tick}(n) }
+%   { Q }
+% }
+% %
+% \and
+% \Rule{L:Loop}
+% { \Delta; Q; R \vdash
+% \htriple
+%   { I }{ S }{ I + M_l }
+% }
+% { \Delta; B; R \vdash
+% \htriple
+%   { I }{ \code{loop}~S }{ Q }
+% }
+% \and \Rule{L:Seq}
+% { \Delta; B; R \vdash
+% \htriple{ P }{ S_1 }{ Q' + M_s }
+% \\
+% \!\!  \Delta; B; R \vdash
+% \htriple{ Q' }{ S_2 }{ Q }
+% }
+% { \Delta; B; R \vdash
+% \htriple{ P }{ S_1;S_2 }{ Q }
+% }
+% %
+% \and
+% \Rule{L:Weaken}
+% {
+%   \!\! B' \models B \!\! \\
+%   \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
+%   \!\! P \models P' \!\! \\
+%   \!\! \Delta; B'; R' \vdash
+%   \htriple{ P' }{ S }{ Q' } \!\! \\
+%   \!\! Q' \models Q \!\!
+% }
+% { \Delta; B; R \vdash
+% \htriple{ P }{ S }{ Q }
+% }
+% %
+% \and
+% \Rule{L:Frame}
+% { \Delta; B; R \vdash
+% \htriple{ P }{ S }{ Q } \\
+%   x \in \Qplusz
+% }
+% { \!\! \Delta; B + x; R + x \vdash
+% \htriple{ P + x }{ S }{ Q + x } \!\!
+% }
+% %
+% \end{mathpar}
+% \caption{Selected rules of the Quantitative Hoare Logic}
+% \label{fig:logic}
+% \end{figure*}
+
+\paragraph{Quantitative Hoare Logic.}
+To prove the soundness of \toolname we found it useful to go through an
+intermediate step using a Quantitative Hoare Logic.  This logic is at
+the same time a convenient semantic tool and a clean way to interface
+manual proofs with our automation.  We base it on the one presented in
+\cite{veristack14}, we add support for arbitrary resources and simplify
+the handling of auxiliary state.
+
+Following this previous work, we define quantitative Hoare triples
+as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
+maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
 an amount of resources available.  The assertions $B$ and $R$ are
 postconditions for the case in which the block $S$ exits by a \code{break}
-or \code{return} statement.  Additionally, $R$ depends on an extra value
-that is returned by the function.  The meaning of the triple $\htriple Q S
-{Q'}$ is as follows: If $S$ is executed with starting state $\state$,
-the empty continuation \Kstop{} and at least $Q(\state)$ resource units
-available then the evaluation does not run out of resources and there
-are at least $Q(\state')$ resources left if the evaluation terminates
-in state $\state'$.  The logic rules are similar to the ones in previous
-work and are simply extended to account for all the costs found in
-our cost-aware semantics.  Some rules are presented in \pref{fig:logic}.
-
-Finally, we define a continuation based
-soundness of triples and prove the soundness of all rules in Coq.  The
-full version of this paper \cite{anon_extended}, gives more explanations
-on the rules and provides a thorough overview of our soundness proof.
-
-\subsection{The Soundness Theorem}
-
-We use the
-logic from previous paragraph as the target of a translation
+or \code{return} statement.  Additionally, $R$ depends on the return
+value of the current function.  The meaning of the triple $\htriple Q
+S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
+the empty continuation, and at least $Q(\state)$ resource units available
+then the evaluation does not run out of resources and there are at
+least $Q(\state')$ resources left if the evaluation terminates in
+state $\state'$.  The logic rules are similar to the ones in previous
+work and are simply extended to account for all the costs found in our
+cost-aware semantics.
+
+Finally, we define a strong compositional continuation based soundness for
+triples and prove the validity of all the rules in Coq.  \ifshort{The full
+version of this paper \cite{anon_extended}, provides explanations for
+many rules and a thorough overview of our soundness proof.}{Details about
+the logic and its soundness proof are provided in the \pref{app:logic}.}
+
+\paragraph{The Soundness Theorem.}
+We use the logic from previous paragraph as the target of a translation
 function for the automatic derivation system.  This reveals two orthogonal
-aspects of the proof: on one side, it relies on amortized reasoning
-(the quantitative logic rules), and on the other side, it
-uses combinatorial properties of our linear potential functions
-(the automatic analysis rules).
-
-Technically, we define a translation
-function $\tr$ such that if a judgement $J$ in the automatic analysis
-is derivable, $\tr(J)$ is derivable in the quantitative logic.  By using
-$\tr$ to translate derivations of the automatic analysis to derivations
-in the quantitative logic we can directly obtain a certified resource
-bound for the analyzed program.
+aspects of the proof: on one side, it relies on amortized reasoning (the
+quantitative logic rules), and on the other side, it uses combinatorial
+properties of our linear potential functions (the automatic analysis
+rules).
+
+Technically, we define a translation function $\tr$ such that if
+a judgement $J$ in the automatic analysis is derivable, $\tr(J)$
+is derivable in the quantitative logic.  By using $\tr$ to translate
+derivations of the automatic analysis to derivations in the quantitative
+logic we can directly obtain a certified resource bound for the analyzed
+program.
 

### 2014-11-12 paper/main.tex
section 6

@@
     a[x]=0; @na--;@
-    tick(1);
-    x++; }
-  if (x < N) {
-    a[x]=1; @na++;@
-    tick(1); }
+    tick(1); x++; }
+  if (x < N) { a[x]=1; @na++;@ tick(1); }
   k--;
@@
 \end{minipage}
-\caption{Example of assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
-  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
-  0 to all other operations.  The bound derived automatically is
+\caption{Assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
+  tick metric.  The derived bound  is
   $2|[0,k]| + |[0,na]|$.
@@
 \end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We use the tick metric to model the stack consumption of the program.
-  Before a function call to \code{bsearch}, a tick statement pays for
-  the space used by stack frame of the function.  The space is recoverd
-  after the call.  We write $\log_2(x)$ for the integer part of
+\caption{Assisted bound derivation using logical state.
+  % We use the tick metric to model the stack consumption of the program.
+  % Before a function call to \code{bsearch}, a tick statement pays for
+  % the space used by stack frame of the function.  The space is recoverd
+  % after the call.
+  We write $\log_2(x)$ for the integer part of
   logarithm of $x$ in base 2.
-  The bound derived automatically is $|[0,lg]|$.
+  The semi-automatically derived bound is $|[0,lg]|$.
   }
@@
 
-While complete automation is often desirable, it cannot possibly
-be universal since the problem is undecidable.  In this section
-we present two techniques to derive complex resource bounds by
-leveraging our automation.
-
-When the resource bound of a program depends on the contents of the heap,
-or is non-linear (e.g. logarithmic, exponential), we propose to
-introduce \emph{logical state}.
-This technique is inspired by XXX (speed) and consists in introducing
-extra variables to the program that do not change its behavior, but help
-the automatic tool to figure out a bound.
-
-The systematic technique consists in the following steps. First,
-a program $P$ that fails to be analyzed automatically is enriched
-by extra variables $\vec x$ called logical variables to form a
-program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
-for the logical variables is shown to be such that:
+While complete automation is desirable, it is not always possible
+since the problem of bound derivation is undecidable.  In this section
+we present a new technique to derive complex resource bounds
+semi-automatically by leveraging our automation. Our goal is to
+develop an interface between bound derivation and
+established qualitative verification techniques.
+
+When the resource bound of a program depends on the contents of the
+heap, or is non-linear (e.g. logarithmic, exponential), we propose to
+introduce a \emph{logical state} using \emph{auxiliary variables}.
+Auxiliary variables do not change the behavior of the program and
+guide $\toolname$ during bound derivation.
+
+More precisely, the technique consists in the following steps. First,
+a program $P$ that fails to be analyzed automatically is enriched by
+auxiliary variables $\vec x$ to form a program
+$P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
+logical variables is selected so that:
 $$
@@
 $$
-Since the annotated program and the original one are often
+Since the annotated program and the original one are usually
 syntactically close, the proof of this result goes
@@
 
-On the other hand, it might be the case that even with annotations, a
-resource bound cannot be found automatically.  With existing tools, this
-situation requires a full rewrite of the program.  However, in our
-framework, it is possible to write a manual proof and combine it with
-bounds derived by the automatic analysis.  This interaction is made
-formal using our Quantitative Hoare Logic.  This resource aware program
-logic is an extension of the one presented in~\cite{veristack14} and
-was originally designed to prove the soundness of our automatic
-analysis.  To make it suitable for this end, we made it independent
-of the resource used (e.g. stack usage, heap usage, clock ticks) and
-implemented a smoother handling of the auxiliary state through a shallow
-embedding in Coq.  Despite being first a theoretical tool, the logic can
-be used directly as in~\cite{veristack14}.  Its judgements are inspired
-by typical Hoare logics and have the shape
-$$
-\Delta \vdash \htriple P S Q
-$$
-where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
-are quantitative assertions, i.e. functions from the program states
-to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
-of the logic are described in further details in the extended version of
-this paper~\cite{anon_extended}.  Because our automated tool produces
-proofs of the resource bounds in the Quantitative Hoare Logic, it can
-be used as an assistant when deriving a complex bound by hand in the
-logic.  The interaction is guaranteed to be sound by the soundness
-theorem of our Quantitative Hoare Logic.
-
-\jan{ the following should go somewhere here:
-
-Note that potential functions can access the whole program state, so
-this framework can readily be used to derive bounds that depend on the
-memory contents.
-}
+% On the other hand, it might be the case that even with annotations, a
+% resource bound cannot be found automatically.  With existing tools, this
+% situation requires a full rewrite of the program.  However, in our
+% framework, it is possible to write a manual proof and combine it with
+% bounds derived by the automatic analysis.  This interaction is made
+% formal using our Quantitative Hoare Logic.  This resource aware program
+% logic is an extension of the one presented in~\cite{veristack14} and
+% was originally designed to prove the soundness of our automatic
+% analysis.  To make it suitable for this end, we made it independent
+% of the resource used (e.g. stack usage, heap usage, clock ticks) and
+% implemented a smoother handling of the auxiliary state through a shallow
+% embedding in Coq.  Despite being first a theoretical tool, the logic can
+% be used directly as in~\cite{veristack14}.  Its judgements are inspired
+% by typical Hoare logics and have the shape
+% $$
+% \Delta \vdash \htriple P S Q
+% $$
+% where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
+% are quantitative assertions, i.e. functions from the program states
+% to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
+% of the logic are described in further details in the extended version of
+% this paper~\cite{anon_extended}.  Because our automated tool produces
+% proofs of the resource bounds in the Quantitative Hoare Logic, it can
+% be used as an assistant when deriving a complex bound by hand in the
+% logic.  The interaction is guaranteed to be sound by the soundness
+% theorem of our Quantitative Hoare Logic.
 

### 2014-11-12 paper/main.tex
pack figure 8

@@
 \begin{lstlisting}[basicstyle=\tt\scriptsize]
-while (y > x) {
+while (y>x) {
   x++; tick(1);
 }
-while (x > 2) {
+while (x>2) {
   x -= 3; tick(1);
@@
 \begin{lstlisting}[basicstyle=\tt\scriptsize]
-while (i > 100) {
+while (i>100) {
   i--; tick(1) }
 i += k+50;
-while (i >= 0) {
+while (i>=0) {
   i--; tick(1);
@@
 \begin{lstlisting}[basicstyle=\tt\scriptsize]
-while (x > 0) {
+while (x>0) {
   x--;
@@
 \begin{lstlisting}[basicstyle=\tt\scriptsize]
-assert(y >= 0);
+assert(y>=0);
 while (x > y) {
   x -= y+1;
-  for (z=y; z > 0; z--)
+  for (z=y; z>0; z--)
     tick(1);
@@
 \begin{lstlisting}[basicstyle=\tt\scriptsize]
-  while (x > 0) {
+  while (x>0) {
     x--;
@@
     else
-      while (y > 0) {
-        y--; tick(1); }
+      while (y>0)
+        y--, tick(1);
     tick(1); }
@@
 %
-AAA
+{\footnotesize \toolname}
   & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
@@
 %
-Rank &
-{\small $2 {+} y {-} x (?)$ } &
-{\small $54+k+i$ } &
-{\small ---} &
-{\small $2+2x-y$ } &
-{\small $0.5{\cdot}y^2 {+} yx +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
+{\footnotesize Rank}
+  & {\small $2 {+} y {-} x (?)$ }
+  & {\small $54+k+i$ }
+  & {\small ---}
+  & {\small $2+2x-y$ }
+  & {\small $0.5{\cdot}y^2 {+} yx \ldots$}
+           % +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
 \\
 %
-LOOPUS &
+{\footnotesize LOOPUS}
+&
 {\small
@@
   \end{tabular}
-} &
+}
+&
 {\small
@@
   \end{tabular}
- } &
+}
+&
 {\small ---}% $1 + \max(x, y)$ (???)
@@
 &
-{\small $2\max(x, 0) {+} \max(y, 0)$}
+{\small \begin{tabular}{c}
+           $2\max(x, 0)$ \\
+           $+ \max(y, 0)$
+        \end{tabular}}
 \end{tabular}

### 2014-11-12 paper/main.tex
section 8

@@
 The soundness of the analysis builds on a new cost semantics for Clight
-and an extended Quantitative Logic.  Using these two tools, the soundness
-proof of \toolname is done by a exhibiting a translation morphism.
+and an extended quantitative logic.  Using these two tools, the soundness
+of the automatic analysis described in \pref{sec:AAA}
+is proved by a translation morphism to the logic.
+
+\iffull{}{ The main parts of the soundness proved are formalized with
+  Coq and available for download. The full definitions of the cost
+  semantics and the quantitative Hoare logic, and more details on the
+  soundness proof can be found in the extended version of this
+  article.}
 
@@
 independence by parameterizing evaluations with a resource \emph{metric}
-$M$; a tuple of nine rational numbers and two maps.  Each of these
+$M$; a tuple of rational numbers and two maps.  Each of these
 parameters indicates the amount of resource consumed by a corresponding
@@
 \noindent
-The proof of this theorem is constructive and basically maps each rule
+The proof of this theorem is constructive and maps each rule
 of the automatic analysis directly to its counterpart in the

### 2014-11-12 paper/main.tex
experiments

@@
 
+\paraskip
 \paragraph{Cost Semantics for Clight.}
@@
 
+\paraskip
 \paragraph{Quantitative Hoare Logic.}
@@
 \centering
+\vspace{-.5cm}
 \begin{tabular}{r|c|c|c|c|c}
@@
 evaluation.  Example \emph{t08} is a slightly modified version of
-\emph{t08a}, which is described in~\pref{sec:inform}.  Example
+\emph{t08a}, which is described in~\pref{sec:AAA}.  Example
 \emph{t19} demonstrates the compositionaliy of the analysis.  The
@@
   tight linear bound.}{}
-
+%
+\iffull{
 \begin{figure}[t]
@@
 derives a linear bound.
+}
+%
+{The extended version also contains a discussion of the automatic
+  bound derivation for the Knuth-Morris-Pratt algorithm for string
+  search. $\toolname$ finds the tight linear bound $1+2|[0,n]|$.}
 
@@
 translated into a term-rewriting system to be analyzed by these tools.  For
-Rank it is not totally clear how the computed bound relates to the C
+Rank it is not completely clear how the computed bound relates to the C
 program since the computed bound is for transitions in an automaton
@@
 \begin{table}[t]
+\small
 \centering
@@
 \vspace{.1cm}
-\caption{Comparison of our automatic amortized analysis with other
-  automatic tools.  We have not been able to obtain a version of
-  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
-  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
-  does currently not work on C programs and we use the bounds that
-  have been reported in the author's experimental evaluation.}
+\caption{Comparison of $\toolname$ with other
+  automatic tools.}
 \label{tab:compar}
@@
 removed.}
-\label{tab:compar}
+\label{tab:cbench}
 \end{table}
@@
   (blue lines) and the measured runtime cost (red crosses) for example
-  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
+  \emph{t08}. For $x\ge 0$ the bound is tight.}
 \label{fig:3d}
@@
 can share the potential $|[0,x]|$ before the assignment \code{x=x-y}
-between $|[0,x]|$ and $|[0,y]|$ after the assignment.  In this way, we
-derive a tight linear bound.
+between $|[0,x]|$ and $|[0,y]|$ after the assignment.  \iffull{In this way, we
+derive a tight linear bound.}{}
 

### 2014-11-12 paper/main.tex
changes to experimental evaluation

@@
 %
-{\footnotesize \toolname}
+\toolname
   & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
@@
 %
-{\footnotesize Rank}
+Rank
   & {\small $2 {+} y {-} x (?)$ }
@@
 %
-{\footnotesize LOOPUS}
+{\sc loopus}
 &
@@
 \caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  AAA stands for our automatic
-  amortized analysis for Clight.  The output of Rank has been manually
+  several examples with linear bounds.  The output of Rank has been manually
   simplified to fit the table.  }
@@
 evaluation.  Example \emph{t08} is a slightly modified version of
-\emph{t08a}, which is described in~\pref{sec:AAA}.  Example
-\emph{t19} demonstrates the compositionaliy of the analysis.  The
-program consists of two loops that decrement a variable $i$.  In the
-first loop, $i$ is decremented down to 100 and in the second loop $i$
-is decremented further down to $-1$.  However, between the loops we
-assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
-|[0,k]|$ increments.  Our analysis finds this tight bound because our
-amortized analysis naturally takes into account the relation between
-the two loops.
-
-At first sight, example \emph{t30} appears to be a simple loop that
-decrements the variable $x$ down to zero.  However, a closer look
-reveals that the loop actually decrements both input variables $x$ and
-$y$ down to zero before terminating.  In the loop body, first $x$ is
-decremented by one.  Then the values of the variables $x$ and $y$ are
-switched using the local variable $t$ as a buffer.  Our analysis
-infers the tight bound $|[0,x]|+|[0,y]|$.  Sometimes we need some
-assumptions on the inputs in order to derive a bound.  Example
-\emph{t15} is such a case.  We assume here that the input variable $y$
-is non-negative and write \code{assert(y>=0)}.  If we enter the loop
-then we know that $x>0$ and we can obtain constant potential from the
-assignment \code{x=x-1}.  After the assignment we know that $x\geq y$
-and $y\geq 0$.  As a consequence, we can share the potential
-$|[0,x]|$ before the assignment \code{x=x-y} between $|[0,x]|$ and
-$|[0,y]|$ after the assignment.  In this way, we derive a tight linear
-bound.
+\emph{t08a}, which is described in~\pref{sec:AAA}.  Example \emph{t19}
+demonstrates the compositionaliy of the analysis.  The program consists
+of two loops that decrement a variable $i$.  In the first loop, $i$ is
+decremented down to 100 and in the second loop $i$ is decremented further
+down to $-1$.  However, between the loops we assign \code{i=i+k+50}.
+So in total the program performs $52 + |[-1,i]| + |[0,k]|$ increments.
+Our analysis finds this tight bound because our amortized analysis
+naturally takes into account the relation between the two loops.
+Example \emph{t30} decrements both input variables $x$ and $y$ down to
+zero in an unconventional way.  In the loop body, first $x$ is decremented
+by one.  Then the values of the variables $x$ and $y$ are switched using
+the local variable $t$ as a buffer.  Our analysis infers the tight
+bound $|[0,x]|+|[0,y]|$.  Sometimes we need some assumptions on the
+inputs in order to derive a bound.  Example \emph{t15} is such a case.
+We assume here that the input variable $y$ is non-negative and write
+\lstinline{assert(y>=0)}.  The assignment \lstinline{x -= y+1} in the
+loop is split in \lstinline{x--} and \lstinline{x -= y}. If we enter the
+loop then we know that $x>0$, so we can obtain constant potential from
+\lstinline{x--}.  Then we know that $x\geq y\geq 0$, as a consequence
+we can share the potential of $|[0,x]|$ between $|[0,x]|$ and $|[0,y]|$
+after \lstinline{x -= y}.  In this way, we derive a tight linear bound.
 
@@
   tight linear bound.}{}
-%
-\iffull{
+
+\ifdefined\fullversion
 \begin{figure}[t]
@@
 derives a linear bound.
-}
-%
-{The extended version also contains a discussion of the automatic
+\else
+The extended version also contains a discussion of the automatic
 bound derivation for the Knuth-Morris-Pratt algorithm for string
-  search. $\toolname$ finds the tight linear bound $1+2|[0,n]|$.}
+search. $\toolname$ finds the tight linear bound $1+2|[0,n]|$.
+\fi
 

### 2014-11-12 paper/main.tex
add periodic exemple

@@
   \centering
-\vspace{-1.8cm}
-\hspace{-0.4cm}
+\ifshort{\vspace{-1.8cm}
+\hspace{-0.4cm}}{}
   \begin{minipage}[b]{7cm}
@@
 \begin{figure*}
- \setlength{\progwidth}{.24\linewidth}
+ \setlength{\progwidth}{.18\linewidth}
   \centering
@@
   \end{minipage}
+%
+%
+  \begin{minipage}[b]{\progwidth}
+    \begin{center}
+   \begin{lstlisting}
+  i=1;
+  j=0;
+  while (j<x) {
+    j++;
+    if (i>=4) {
+      i=1;
+      tick(40);
+    } else
+       i++;
+    tick(1);
+  }
+   \end{lstlisting}
+
+$11|[0,x]|$
+\\[.7\baselineskip]
+      {\bf t63}
+    \end{center}
+  \end{minipage}
 
@@
 $|[y,x]|$ is a random choice of the LP solver.
+
+Example \emph{t63} shows how amortized reasoning handles periodically
+expensive operations.  The top-level loop is iterated $x$ times in
+a regular fashion, but every 4 iterations of this loop an expensive
+opereation modeled by \code{tick(40)} is performed.  Taking the worst
+case of the loop body would yield the pessimistic bound $41|[0,x]|$,
+however our tool is able to amortize the expensive operation and derives
+instead the tight bound $11|[0,x]|$.
 

### 2014-11-12 paper/main.tex
add function specs paragraph

@@
 
-\jan{Here we have to say something about function specifications and
-  function calls.}
-\jan{Here we have to say something about function specifications and
-  function calls.}
-\jan{Here we have to say something about function specifications and
-  function calls.}
+During the analysis, function specifications are quadruples
+$(\Gamma_f, Q_f, \Gamma'_f, Q'_f)$ where $\Gamma_f, Q_f$ depend on
+$\Vargs$, and $\Gamma'_f, Q'_f$ depend on $\Vret$.  These parameters
+are instantiated by appropriate variables on call sites.  A distinctive
+feature of our analysis is that it respects the function abstraction:
+when deriving a function specification it generates a set of
+constraints and the above quadruple; once done, the constraint
+sets can readily be reused for every call site and the function need
+not be analyzed multiple times.
 

### 2014-11-12 paper/main.tex
related work

@@
 we can share the potential of $|[0,x]|$ between $|[0,x]|$ and $|[0,y]|$
-after \lstinline{x -= y}.  In this way, we derive a tight linear bound.
+after \lstinline{x -= y}.  % In this way, we derive a tight linear bound.
 
@@
   tight linear bound.}{}
-
+%
 \ifdefined\fullversion
@@
 programs.  However, this experimental evaluation has to be taken with
-a grain of salt.  Our goal in this work is not to set a new standard
-for automatic bound analysis but only to show that our approach has
-advantages on examples with linear bounds.  Overall the existing tools
-are more powerful since they can derive polynomial bounds and support
+a grain of salt.  Existing tools complement $\toolname$
+since they can derive polynomial bounds and support
 more features of C.  We were particularly impressed by LOOPUS which is
 very robust, works on large C files, and derives very precise bounds.
-We did not include the running times of the tools in the table since
+\iffull{We did not include the running times of the tools in the table since
 all tested tools work very efficiently and need less then a second
-on every tested example.
+on every tested example.}{}
+
+\pref{tab:cbench} contains a compilation of the results of our
+experiments with the cBench benchmark suite. It shows a representative
+list of automatically derived function bounds. In total we analyzed
+more than 2000 lines of code.
+%
+In the LOC column
+where not only count the LOC of the analyzed function but
+also the LOC of all the function it calls. \iffull{The line count is
+computed on pre-processed C with all unnecessary declarations removed.}{}
+\jan{What is the cost measure?}
 
@@
 example \emph{t08}.  One can see that the derived constant factors are the
-best possible if the input variable $x$ is non-negative.  If $x$ is
-negative then the bound is only slightly off.
-
+best possible if the input variable $x$ is non-negative.  \iffull{If $x$ is
+negative then the bound is only slightly off.}{}
+
+\sectskip
+\section{Related Work}
+\label{sec:related}
+\aftersectskip
+
+A recent project~\cite{veristack14} has implemented and verified a
+quantitative logic to reason about stack-space usage, and modified the
+verified CompCert C compiler to translate C level bound to x86 stack
+bounds.  This quantitative logic is also based on the potential method
+but has very rudimentary support for automation that cannot
+automatically derive symbolic bounds and is not based on efficient LP
+solving.  In this article, we use a more general quantitative Hoare
+logic that is parametric in the resource of interest.  Our main
+contribution is the automatic amortized analysis for Clight that can
+be combined with~\cite{veristack14} to automatically derive parametric
+verified stack bounds for recursive functions.
+
+Our work has been inspired by type-based amortized resource analysis
+for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
+are three major improvements over previous work in the current paper,
+which presents the first automatic amortized resource analysis for C.
+First, we solved the long-standing open problem of extending automatic
+amortized resource analysis to compute bounds for programs that loop
+on (possibly negative) integers.  Second, we extended the analysis
+system to deal with non-linear control flow that is introduced by
+\code{break} and \code{return} statements.  Third, for the first time,
+we have combined an automatic amortized analysis with a system for
+interactively deriving bounds. In particular, recent
+systems~\cite{HoffmannS13} that deal with integers and arrays cannot
+derive bounds that depend on values in mutable locations or possibly
+negative integers, and thus do not apply to most of our example
+programs.
+
+\iffull{
+In the development of our quantitative Hoare logic we have drawn
+inspiration from mechanically verified Hoare logics.
+Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
+logics in Isabelle/HOL has been helpful to understand the interaction
+of auxiliary variables with the consequence rule.  Appel's separation
+logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
+the general structure of the quantitative logic.  \iffull{Since we do not deal
+with memory safety, our logic is much simpler and it would be possible
+to integrate it with Appel's logic.}{}  The continuation passing style
+that we use in the quantitative logic is not only used by
+Appel~\cite{AppelLogic} but also in Hoare logics for low-level
+code~\cite{NiS06,JensenBK13}.
+
+There exist quantitative logics that are integrated into separation
+logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
+quantitative logic.  However, the purpose of these logics is slightly
+different since they focus on the verification of bounds that depend
+on the shape of heap data structures and they are not implemented for
+C.  Also closely related to our logic is a VDM-style logic for
+reasoning about resource usage of an abstract fragment of JVM byte
+code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
+Hoare-style, does not target C code, and is not designed for
+interactive bound development but to produce certificates for bounds
+derived for high-level functional programs.
+}{}
 \begin{table}[t]
@@
 \vspace{.1cm}
-\caption{Evaluation of our automatic amortized analysis on larger
-C code from the cBench benchmark suite. The function analyzed might
-be calling helper functions which do not appear in the list but have
-to be analyzed.  These functions are accounted for in the LoC column
-where not only the lines count of the analyzed function is called
-but also the line count of all the funtions it calls. The line count
-is computed on pre-processed C with all unnecessary declarations
-removed.}
+\caption{Derived bounds for functions from cBench.}
 \label{tab:cbench}
 \end{table}
-
-\sectskip
-\section{Related Work}
-\label{sec:related}
-\aftersectskip
-
-Most closely related to this article is \iffull{our}{a} previous work
-on end-to-end stack-bound verification of Clight
-programs~\cite{veristack14}.  In that work, \iffull{we}{the authors}
-have implemented and verified a quantitative logic to reason about
-stack-space usage.  In this article, we present a more general
-quantitative Hoare logic that is parametric in the resource of
-interest.  The main innovation is the novel automatic amortized analysis for
-Clight programs that computes logical derivations for non-trivial
-bounds for programs with loops and recursion.
-
-Our work has been inspired by type-based amortized resource analysis
-for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
-are three major improvements over previous work in the current paper,
-which presents the first automatic amortized resource analysis for C.
-First, we solved the long-standing open problem of extending automatic
-amortized resource analysis to compute bounds for programs that loop
-on (possibly negative) integers.  Second, we extended the analysis
-system to deal with non-linear control flow that is introduced by
-\code{break} and \code{return} statements.  Third, for the first time,
-we have combined an automatic amortized analysis with a system for
-interactively deriving bounds.
-
-In the development of our quantitative Hoare logic we have drawn
-inspiration from mechanically verified Hoare logics.
-Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
-logics in Isabelle/HOL has been helpful to understand the interaction
-of auxiliary variables with the consequence rule.  Appel's separation
-logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
-the general structure of the quantitative logic.  \iffull{Since we do not deal
-with memory safety, our logic is much simpler and it would be possible
-to integrate it with Appel's logic.}{}  The continuation passing style
-that we use in the quantitative logic is not only used by
-Appel~\cite{AppelLogic} but also in Hoare logics for low-level
-code~\cite{NiS06,JensenBK13}.
-
-There exist quantitative logics that are integrated into separation
-logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
-quantitative logic.  However, the purpose of these logics is slightly
-different since they focus on the verification of bounds that depend
-on the shape of heap data structures and they are not implemented for
-C.  Also closely related to our logic is a VDM-style logic for
-reasoning about resource usage of an abstract fragment of JVM byte
-code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
-Hoare-style, does not target C code, and is not designed for
-interactive bound development but to produce certificates for bounds
-derived for high-level functional programs.
 
@@
 other available tools that can be directly applied to C code are Rank
-and LOOPUS.  Our analysis framework does not aim to set a new standard
-for automatic bound analysis.  Our contribution is rather a principled
-approach that produces certificates that are proved sound with respect
-to a formal cost semantics.  Moreover, we have a system that enables
-both automatic and interactive bound derivation, a formal soundness
-proof in Coq, and a method that can handle resources that can be
-released (e.g., memory).  However, as we have shown
-in~\pref{sec:exper}, our automatic amortized analysis matches the
-state of the art in automatic bound analysis for linear bounds and
-sometimes even derives better constant factors than the other tools.
-Since Rank and Loopus do not handle recursion, our automatic amortized
-analysis is also the only available tool that can derive bounds for
-recursive C programs.
+and LOOPUS.  As demonstrated, $\toolname$ is more compositional than
+the aforementioned tools. Our technique, is the only one that is able
+to generate resource specifications for functions, deal with resources
+like memory that might become available, generates proof certificates
+for the bounds, and has support for user guidance that separates
+qualitative and quantitative reasoning.
 
@@
 recursive or composed functions.
-
+%
 We are only aware of two verified quantitative analysis systems.

### 2014-11-12 paper/main.tex
spacing

@@
 &
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
 while (y>x) {
@@
 &
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
+\hspace{-.5cm}\begin{lstlisting}[basicstyle=\tt\footnotesize]
 while (i>100) {
@@
 }
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
+\end{lstlisting}\hspace{-.5cm}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
 while (x>0) {
@@
 &
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
 assert(y>=0);
@@
 &
-\begin{lstlisting}[basicstyle=\tt\scriptsize]
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
 while (x>0) {
@@
 \toolname
-  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
-  & {\small $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ }
+  & {\small $\!\!\!1.33|[x,y]| {+} 0.33 |[0,x]|\!\!\!$ }
+  & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
   & {\small $|\inter 0 x| {+} |\inter 0 y|$}
@@
 %
-{\sc loopus}
+{\hspace{-.2cm}LOOPUS}
 &
@@
 {\small
-  \begin{tabular}{c}
+  $\!\!\!\!$\begin{tabular}{c}
     $\max(0, i{-}100)$ \\
     $+ \max(0, k {+} i {+} 51)$
-  \end{tabular}
+  \end{tabular}$\!\!\!\!$
 }

### 2014-11-12 paper/main.tex
conclusion

@@
   i--; tick(1);
-}
-\end{lstlisting}\hspace{-.5cm}
+}\end{lstlisting}\hspace{-.5cm}
 &
@@
 
-We have introduced an novel quantitative analysis system for CompCert
-Clight programs.  To the best of our knowledge, this article presents
-the first resource analysis framework for C that makes it possible to
-combine non-trivial automatically derived bounds with interactively
-derived bounds in a proof system that produces verifiable certificates
-for the bounds.  The main technical innovations are a quantitative
-Hoare logic for reasoning about user-defined resource cost and an
-automatic amortized analysis that derives bounds for programs whose
-resource consumption depends on (possibly negative) integers and
-non-sequential control flow as introduced by \code{break} and
-\code{return} statements.
-
-We will continue to improve our framework to reason more precisely
-about resource usage of system software such as the hypervisor kernel
-CertiKOS~\cite{GuVFSC11}, which \iffull{we are currently developing
-  and verifying}{is currently developed and verified}.  For one thing,
-we will improve the automation to derive \emph{polynomial bounds} and
-to handle non-linear size changes, as already developed for functional
-programs~\cite{HoffmannS13}.  For another thing, we will build on
-previous work~\cite{HoffmannMZ13} to generalize the quantitative
-Hoare logic to handle \emph{concurrent programs}\iffull{ with locks and
-lock-free data structures.}{.}
+We have developed a novel analysis framework for compositional and
+certified worst-case resource bound analysis for C programs.  The
+framework combines ideas from existing abstract interpretation--based
+techniques with the potential method of amortized analysis.  The
+technical innovations are implemented in the publically available tool
+$\toolname$. To the best of our knowledge, $\toolname$ is the first
+analysis tool for C programs that automatically reduces the derivation
+of symbolic bounds to efficient LP solving.
+
+We have demonstrated that our approach improves the state-of-the-art
+in resource bound analysis for C programs in three ways. First, our
+technique is naturally compositional, tracks size changes of
+variables, and can abstractly specify the resource cost of functions
+(\pref{sec:AAA}). Second, it is easily combinable with established
+qualitative verification to guide semi-automatic bound derivation
+(\pref{sec:anno}). Third, we have shown that the local inference rules
+of the derivation system automatically produce easily checkable
+certificates for the derived bounds (\pref{sec:soundness}).
+
+Our system is the first amortized resource analysis for C programs.
+It addresses the longstanding open problem of extending automatic
+amortized resource analysis to compute bounds for programs that loop
+on singed integers and to deal with non-linear control flow that is
+introduced by break and return statements.
+
+This work is the starting point for several projects that we plan to
+investigate in the future. Examples include the extension to
+concurrent programs, better integration of low-level features like
+hardware caches, and the extension of the automatic analysis to
+multivariate resource polynomials~\cite{HoffmannAH11}.
 

### 2014-11-12 paper/main.tex
spacing

@@
 The rule {\sc Q:Inc} can be applied
-if we cannot deduce any information about $y$.  In the implementation,
-the rules for increments and decrements are combined into one syntax
-directed rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
+if we cannot deduce any information about $y$.  In $\toolname$,
+the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
 
@@
 \newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
-\begin{figure*}
+\begin{figure*}[t]
 \vspace{-.6cm}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \center

### 2014-11-13 paper/main.tex
some formatting in the extended version
more examples and technical remarks

@@
 that the compositionality of our method is indeed crucial in practice.
-\jan{what projects?}
-
-Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations of block-based
-cryptographic primitives: Data of arbitrary length, is consumed in
-blocks and the leftover is stored in a buffer for future use when more
-data is available.  This looping pattern is particularly well handled
-by our method. If $N \geq 8$, $\toolname$ infers the bound
-$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8}
-+ \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the
-cost in the second loop (\code{tick(1)}) are random choices and
-$\toolname$ would also derive tight bound for other values.
+
+Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
+of block-based cryptographic primitives: Data of arbitrary length, is
+consumed in blocks and the leftover is stored in a buffer for future use
+when more data is available.  It is present in all the block encryption
+routines of PGP.  The same pattern is also used in performance critical
+code to unroll a loop, for example we found it in a bit manipulating
+function of libtiff library and a CRC computation routine of MAD,
+an MPEG audio decoder.  This looping pattern is particularly well
+handled by our method. If $N \geq 8$, $\toolname$ infers the bound
+$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8} +
+\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
+in the second loop (\code{tick(1)}) are random choices and $\toolname$
+would also derive tight bound for other values.
 
@@
 }{}
+
 \begin{table}[t]
@@
 \code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{MD5Final} & 195 & 141 & 0.22 \\
 \code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
+\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\hline
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1
 \end{tabular}
 \vspace{.1cm}
-\caption{Derived bounds for functions from cBench.}
+\caption{Derived bounds for functions from cBench. \jan{FUCKING backedge metric}}
 \label{tab:cbench}
@@
 certificates for the derived bounds (\pref{sec:soundness}).
-
+%
 Our system is the first amortized resource analysis for C programs.
@@
 \newpage
+
+\section{Syntax and Semantics}
+\label{app:sem}
+
+We implemented our cost semantics and the quantitative Hoare logic in
+Coq for \emph{CompCert Clight}.  Clight is the most abstract
+intermediate language used by CompCert.  Mainly, it is a subset of C
+in which loops can only be exited with a \code{break} statement and
+expressions are free of side effects.
+
+\paragraph{Syntax.}
+
+In this article, we describe our system for a subset of Clight that is
+sufficient to discuss the general ideas.  This subset is given by the
+following grammar.
+%
+\begin{align*}
+S &:= \code{assert}~E
+\mid \code{skip}
+\mid \code{break}
+\mid \code{return}~x
+\mid x \gets E
+\mid x \gets f(x^*)
+\\
+& \mid \code{loop}~S
+\mid \code{if}(E)~S~\code{else}~S
+\mid S;S
+\mid \code{tick}(n)
+\end{align*}
+%
+Expressions $E$ are left abstract in our presentation.  For our
+analysis framework, it is only important that they are side-effect
+free.
+%
+The most notable difference to full Clight is that we can only assign
+to variables and thus do not consider operations that update the heap.
+Moreover, function arguments and return values are assumed to be
+variables.  This is only for simplifying the presentation; in the
+implementation we can deal with heap updates and general function
+calls and returns.  However, we have not implemented our framework for function
+pointers, \code{goto} statements, \code{continue} statements, and
+\code{switch} statements.
+
+We include the built-in primitive $\code{assert}~e$ that terminates
+the program if the argument $e$ evaluates to false and has no effect
+otherwise.  This is useful to express assumptions on the inputs of a
+program for the automatic analysis.  We also add the built-in function
+\code{tick(n)} that can be called with a constant integer $n$ as a
+flexible way to model resource consumption or release (if $n$ is
+negative).
+
+\begin{figure*}[t!]
+\begin{mathpar}
+%
+\Rule{S:Assert}
+{ \code{istrue}~\sem{e}_\state }
+{ (\state, \code{assert}~e, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
+}
+%
+\and
+\Rule{S:BrkSeq}
+{}
+{ (\state, \code{break}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{break}, \cont, \cost)
+}
+%
+\and
+\Rule{S:BrkLoop}
+{}
+{ (\state, \code{break}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
+}
+%
+\and
+\Rule{S:RetSeq}
+{}
+{ (\state, \code{return}~x, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+\and
+%
+\and
+\Rule{S:RetLoop}
+{}
+{ (\state, \code{return}~x, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{return}~x, \cont, \cost)
+}
+%
+\and
+\Rule{S:RetCall}
+{ \state = (\_, \gamma) \\
+  \state' = (\theta, \gamma)[r \mapsto \state(x)]
+}
+{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_r)
+}
+%
+\and
+\Rule{S:Update}
+{ \state' = \state[x \mapsto \sem{e}_\state] }
+{ (\state, x \gets e, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+}
+%
+\and
+\Rule{S:Call}
+{ \fenv f = (\vec x, S_f) \\
+  \state = (\theta, \gamma) \\
+  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
+}
+{ (\state, r \gets f(\vec y), \cont, \cost)
+\smallstep{M}
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+}
+%
+\and
+\Rule{S:Loop}
+{}
+{ (\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, S, \Kloop S \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipLoop}
+{}
+{ (\state, \code{skip}, \Kloop S \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
+}
+%
+\and
+\Rule{S:IfTrue}
+{ \code{istrue}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \cont, \cost')
+}
+%
+\and
+\Rule{S:IfFalse}
+{ \code{isfalse}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
+}
+{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_2, \cont, \cost')
+}
+%
+\and
+\Rule{S:Seq}
+{ }
+{ (\state, S_1; S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \Kseq {S_2} \cont, \cost)
+}
+%
+\and
+\Rule{S:SkipSeq}
+{}
+{ (\state, \code{skip}, \Kseq S \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
+}
+%
+\and
+\Rule{S:Tick}
+{ }
+{ (\state, \code{tick}(n), \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
+}
+
+\end{mathpar}
+\caption{Rules of the operational semantics of statements.}
+\label{fig:opsem}
+\end{figure*}
 
@@
 \begin{align*}
-\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
-& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
-& & \land& \safe n Q {\code{skip}} \cont
+\safeK n B R Q \cont & :=  \safe n B {\code{break}} \cont \\
+ \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
+ \land& \safe n Q {\code{skip}} \cont
 \end{align*}
@@
 %
-\begin{theorem}[Soundness of the logic]
+\begin{theorem}[Soundness]
   If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
@@
 
-\section{Syntax and Semantics}
-\label{app:sem}
-
-We implemented our cost semantics and the quantitative Hoare logic in
-Coq for \emph{CompCert Clight}.  Clight is the most abstract
-intermediate language used by CompCert.  Mainly, it is a subset of C
-in which loops can only be exited with a \code{break} statement and
-expressions are free of side effects.
-
-\paragraph{Syntax.}
-
-In this article, we describe our system for a subset of Clight that is
-sufficient to discuss the general ideas.  This subset is given by the
-following grammar.
-%
-\begin{align*}
-S &:= \code{assert}~E
-\mid \code{skip}
-\mid \code{break}
-\mid \code{return}~x
-\mid x \gets E
-\mid x \gets f(x^*)
-\\
-& \mid \code{loop}~S
-\mid \code{if}(E)~S~\code{else}~S
-\mid S;S
-\mid \code{tick}(n)
-\end{align*}
-%
-Expressions $E$ are left abstract in our presentation.  For our
-analysis framework, it is only important that they are side-effect
-free.
-%
-The most notable difference to full Clight is that we can only assign
-to variables and thus do not consider operations that update the heap.
-Moreover, function arguments and return values are assumed to be
-variables.  This is only for simplifying the presentation; in the
-implementation we can deal with heap updates and general function
-calls and returns.  However, we have not implemented our framework for function
-pointers, \code{goto} statements, \code{continue} statements, and
-\code{switch} statements.
-
-We include the built-in primitive $\code{assert}~e$ that terminates
-the program if the argument $e$ evaluates to false and has no effect
-otherwise.  This is useful to express assumptions on the inputs of a
-program for the automatic analysis.  We also add the built-in function
-\code{tick(n)} that can be called with a constant integer $n$ as a
-flexible way to model resource consumption or release (if $n$ is
-negative).
-
-\begin{figure}[t!]
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{}
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{}
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
-}
-%
-\and
-\Rule{S:RetSeq}
-{}
-{ (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-\and
-%
-\and
-\Rule{S:RetLoop}
-{}
-{ (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{return}~x, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetCall}
-{ \state = (\_, \gamma) \\
-  \state' = (\theta, \gamma)[r \mapsto \state(x)]
-}
-{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_r)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
-}
-%
-\and
-\Rule{S:Call}
-{ \fenv f = (\vec x, S_f) \\
-  \state = (\theta, \gamma) \\
-  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
-}
-{ (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep{M}
-(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
-}
-%
-\and
-\Rule{S:Loop}
-{}
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{}
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
-}
-%
-\and
-\Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \cont, \cost')
-}
-%
-\and
-\Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_2, \cont, \cost')
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{}
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
-}
-%
-\and
-\Rule{S:Tick}
-{ }
-{ (\state, \code{tick}(n), \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_t(n))
-}
-
-\end{mathpar}
-\caption{Rules of the operational semantics of statements.}
-\label{fig:opsem}
-\end{figure}
-
 \section{Catalog of Automatically Analyzed Programs}

### 2014-11-13 paper/main.tex
fixing a bunch of typos

@@
   %
-  The analysis framework is implemented in the publically-available
+  The analysis framework is implemented in the publicly-available
   tool $\toolname$. An experimental evaluation demonstrates the
@@
 memory).  Compared to more classical
-apporaches based on ranking functions, our tool inherits
+approaches based on ranking functions, our tool inherits
 the benefits of amortized reasoning.  Using only one
@@
   require that potential functions always give non-negative results.
-  This is acheived by restricting coefficients in a quantitative
+  This is achieved by restricting coefficients in a quantitative
   annotation to be always non-negative.}{}
@@
 to prioritize certain intervals.  For example, a bound with
-$[10,x]$ will be prefered to another with $[0,x]$ because $|[10,x]|
+$[10,x]$ will be preferred to another with $[0,x]$ because $|[10,x]|
 \le |[0,x]|$.
@@
 defined by the quantitative annotation $Q$.  The logical context $\Gamma$
-is implicitely lifted to a quantitative assertion by mapping a state
+is implicitly lifted to a quantitative assertion by mapping a state
 $\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
@@
 \begin{theorem}[Soundness of the automatic analysis]
-  If $J$ is a judgement derived by the automatic analyis,
+  If $J$ is a judgement derived by the automatic analysis,
   then $\tr(J)$ is a quantitative Hoare triple derivable
@@
 \emph{t08a}, which is described in~\pref{sec:AAA}.  Example \emph{t19}
-demonstrates the compositionaliy of the analysis.  The program consists
+demonstrates the compositionally of the analysis.  The program consists
 of two loops that decrement a variable $i$.  In the first loop, $i$ is
@@
 techniques with the potential method of amortized analysis.  The
-technical innovations are implemented in the publically available tool
+technical innovations are implemented in the publicly available tool
 $\toolname$. To the best of our knowledge, $\toolname$ is the first
@@
 $f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
-both parameterized by an arbitary logical variable $z$ (which can be a
+both parameterized by an arbitrary logical variable $z$ (which can be a
 tuple) that relates the function arguments with the return value.  The
@@
 a regular fashion, but every 4 iterations of this loop an expensive
-opereation modeled by \code{tick(40)} is performed.  Taking the worst
+operation modeled by \code{tick(40)} is performed.  Taking the worst
 case of the loop body would yield the pessimistic bound $41|[0,x]|$,

### 2014-11-13 paper/main.tex
notes on the benchmarks

@@
 list of automatically derived function bounds. In total we analyzed
-more than 2000 lines of code.
-%
-In the LOC column
-where not only count the LOC of the analyzed function but
-also the LOC of all the function it calls. \iffull{The line count is
+more than 2000 lines of code.  In the LoC column
+we not only count the lines of the analyzed function but
+also the ones of all the function it calls. \iffull{The line count is
 computed on pre-processed C with all unnecessary declarations removed.}{}
-\jan{What is the cost measure?}
+We analyzed the functions using a metric that assigns a cost 1 to all
+the back-edges in the control flow (loops, and function calls).
+The functions that took the longest to analyze are often unrolled loops
+where many variables are assigned, this stresses our analysis because
+the number of LP variables has a quadratic growth in program variables.
+Even on these stressful examples, the analysis could finish in less
+than 2 seconds.  These times can even be greatly reduced by adding
+manual annotations to indicate $\toolname$ that certain variables do
+not have to be tracked.  We could also envision to implement heuristics.
+
+The trickiest of these benchmarks is the \code{sha\_update} function,
+it is composed of one loop calling two helper functions that in turn
+have 6 and 1 inner loops.  The essence of the SHA algorithm makes
+the compositionality of our analysis essential to get a tight bound
+since loops on the same index are sequenced 4 and 2 times without
+resetting it.  All other tools available have constant factors
+quadratically bigger than ours on this example.
 
@@
 \hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
 \end{tabular}

### 2014-11-13 paper/main.tex
some typos

@@
 are important to ensure the reliability of
-safety-critical systems~\cite{veristack14,Regehr05}.
+safety-critical systems~\cite{Regehr05}.
 Static energy usage information is critical
@@
 cloud computing~\cite{CohenZSL12,CarrollH10}.
-Worst-case time bounds can help to create
+Worst-case time bounds can help create
 constant-time implementations that prevent
@@
 in the first category derive impressive bounds for
-numerical imperative programs, but are not fully compositional.
+numerical imperative programs, but are not compositional.
 This is problematic if one needs to derive global whole-program bounds.
@@
 provide a function or procedure abstraction, available
-tools are not able to abstract resource behavior
-and have to inline the procedure body to perform their
+tools are not able to abstract resource behavior;
+instead, they have to inline the procedure body to perform their
 analysis.
@@
 the role of the type system in imperative programs. However, these
-techniques fail for numerical imperative programs because these programs lack
+techniques fail for numerical imperative programs because they lack
 fine-grained data-structure information to hook the analysis on.
@@
 fails to find a resource bound for a program, there is no possibility
-for sound user interaction to guide the tool during during bound
+for sound user interaction to guide the tool during bound
 derivation.
 For example, there is no concept of manual proofs of
-resource bounds and no framework that could support composition
+resource bounds and no framework can support composition
 of manually derived bounds with automatically inferred bounds.
@@
 deriving resource bounds on C programs.  This new
-approach is an attempt to unify the two previous
+approach is an attempt to unify the two aforementioned
 categories: It solves the compositionality issues of techniques
-for numerical imperative code by adapting amortized-analysis based
+for numerical imperative code by adapting amortized-analysis--based
 techniques from the functional world.
@@
 in a function specification that can be used at every
-call-side without accessing the function body.
+call site without accessing the function body.
 %
@@
 
-Several micro benchmarks demonstrate the practicality and
+Examples from cBench and micro benchmarks demonstrate the practicality and
 expressiveness of the user guided bound inference.  For example, we
-derive a logarithmic bound for a binary search function, a bound that
+derive a logarithmic bound for a binary search function\iffull{, a bound that
 depends on the contents of an array to describe the exact cost of a
-function that finds a maximal element in an array, and a bound
-that amortizes the cost of $k$ successive increments to a binary
+function that finds a maximal element in an array, }{ }and a bound
+that amortizes the cost of $k$\iffull{ successive }{ }increments to a binary
 counter (see \pref{sec:anno}).
@@
   become available during execution.
+\itemskipIn
 \item We show how to automatically reduce the inference of \emph{linear} resource
     bounds to efficient LP solving.
+\itemskipIn
 \item We describe a new method of harnessing existing qualitative
@@
   to derive \emph{non-linear} resource bounds with LP solving.
+\itemskipIn
 \item We prove the soundness of the analysis with respect to a
@@
   % soundness proof.
-\item We present experiments with a prototype implementation on more
+\itemskipIn
+\item We implemented our resource bound analysis in the publically-available tool $\toolname$.
+\itemskipIn
+\item We present experiments with $\toolname$ on more
   than 2000 lines of C code. A detailed comparison shows that our
@@
 based on ranking functions.  Reasoning with two potential functions
-promotes compositional reasoning by considering the sequencing of
-program operations.  In the previous rule, $\Phi$ gives a bound for
+promotes compositional reasoning by focusing on the sequencing of
+programs.  In the previous rule, $\Phi$ gives a bound for
 $S_1; S_2$ through the intermediate potential $\Phi'$, even though it
@@
 functions that look very similar to Hoare logic or effect system
-rules.  These rules enable interactive reasoning about resource usage
+rules.  These rules enable reasoning about resource usage
 in a flexible and compositional way, which, as a side effect, produces
@@
 $K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions in $\Gamma$ are only used in the rule for the assignment
+assertions are only used in the rule for the assignment
 \code{x=x+K}.
@@
 translate the program into a term-rewriting system by
-hand. Nevertheless, we will show in \pref{sec:AAA} that the
+hand. We will show in the following sections that the
 potential method makes automatic bound derivation straightforward.
@@
 \begin{lstlisting}[basicstyle=\tt\small]
-#\!\!\!\!\!# while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
+#\!\!\!\!\!#while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
+
+#\!\!\!\!\!#while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
 \end{lstlisting}
@@
 tick annotations can be seamlessly combined in a larger program.
-\begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
-\end{lstlisting}
 
@@
 potential functions, which can express tight bounds for many typical
-programs and allows inference view \emph{linear programming}.
+programs and allows for inference with \emph{linear programming}.
 $$
@@
 Our implementation finds these logical contexts using abstract
-interpretation using the domain of linear inequalities.  We observed that
+interpretation with the domain of linear inequalities.  We observed that
 the rules of the analysis often requires only minimal local knowledge.
@@
   with the computed tick bounds.  Example \emph{t61} and \emph{t62}
-  demonstrate the unique compositionality of our system. In \emph{61},
+  demonstrate the unique compositionality of our system. In \emph{t61},
   $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
@@
 without special tricks or techniques.  Examples \emph{speed\_1} and
-\emph{speed\_2}, that are taken from previous work~\cite{GulwaniMC09},
+\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
 demonstrate that our method can handle \emph{tricky iteration
@@
 
-Example \emph{t08a} shows the ability of the analysis to discover
+Example \emph{t08a} in Figure~\ref{fig:ex_list} shows the ability of the analysis to discover
 interaction between \emph{sequenced loops} through size change of
@@
 %
-Example \emph{t27} shows how amortization can be used to handle
+Example \emph{t27} in Figure~\ref{fig:ex_list}
+shows how amortization can be used to handle
 \emph{interacting nested loops}.  In the outer loop we increment the
@@
 function of libtiff library and a CRC computation routine of MAD,
-an MPEG audio decoder.  This looping pattern is particularly well
-handled by our method. If $N \geq 8$, $\toolname$ infers the bound
-$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8} +
+an MPEG audio decoder.  This looping pattern is handled particularly well
+by our method. If $N \geq 8$, $\toolname$ infers the bound
+$\frac{N}{8} |[0,l]|$, but if $N<8$, $\toolname$ infers $7\frac{8-N}{8} +
 \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
@@
 The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
-can be seen as a regular Hoare triple.  All assertions are split into
+can be seen as a quantitative Hoare triple.  All assertions are split into
 two parts, the logical part and the quantitative part.  The
@@
 account the early exit statements \code{break} and \code{return}.
-This is similar to classical Hoare triples the presence of non-linear
+This is similar to classical Hoare triples in the presence of non-linear
 control flow. In the judgement, $(\Gamma_B, Q_B)$ is the postcondition
@@
 }{}
-
-The rule {\sc Q:Weak} is not syntax directed.  In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition.  \iffull{We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
-annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
-deals also with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq l$ is
-larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$\iffull{ without gaining potential.}{.}
 
@@
 
+The rule {\sc Q:Weak} is not syntax directed.  In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition.  \iffull{We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
+annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
+deals also with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq l$ is
+larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$\iffull{ without gaining potential.}{.}
+
 \sectskip
@@
 
-More precisely, the technique consists in the following steps. First,
+More precisely, the technique consists of the following steps. First,
 a program $P$ that fails to be analyzed automatically is enriched by
@@
 
-This idea is illustrated by the program on \pref{fig:xmplincaux}.
+This idea is illustrated by the program in \pref{fig:xmplincaux}.
 Parts of the code in blue are annotations that were added
@@
 and the original one, we use the inequality $lg >
-\log_2(h-l)$ as invariant.  This trick allows a simpler
+\log_2(h-l)$ as invariant.  This technique allows a simpler
 proof because, when working with integer arithmetic, it is not always
@@
 \vspace{.1cm}
-\caption{Derived bounds for functions from cBench. \jan{FUCKING backedge metric}}
+\caption{Derived bounds for functions from cBench.}
 \label{tab:cbench}
@@
 amortized resource analysis to compute bounds for programs that loop
-on singed integers and to deal with non-linear control flow that is
+on signed integers and to deal with non-linear control flow that is
 introduced by break and return statements.

### 2014-11-13 paper/main.tex
formatting

@@
 not be analyzed multiple times.
+%
+Therefore, the derivation rules are parametric in a function context $\Delta$
+that we leave implicit in the rules that we present here. More details
+can be found in the extended version.
 
@@
 }{}
+
+\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition. We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax}\iffull{ that
+also deals with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq l$ is
+larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}
 
@@
 
-The rule {\sc Q:Weak} is not syntax directed.  In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition.  \iffull{We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
-annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
-deals also with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq l$ is
-larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$\iffull{ without gaining potential.}{.}
-
 \sectskip
@@
 
-The constraints that we generate have
+\iffull{The constraints that we generate have
 a particularly simple form that is known as \emph{network problem}.
-Such problems can be solved in linear time in practice.
+Such problems can be solved in linear time in practice.}{}
 
@@
 \end{lstlisting}
+\vspace{-.2cm}
 \end{minipage}
@@
 \begin{figure}
+\vspace{-.3cm}
 \begin{lstlisting}[numbers=left]
@@
 \end{lstlisting}
+\vspace{-.2cm}
 \caption{Assisted bound derivation using logical state.
@@
 the back-edges in the control flow (loops, and function calls).
-The functions that took the longest to analyze are often unrolled loops
-where many variables are assigned, this stresses our analysis because
+The bounds for the functions \code{ycc\_rgb\_conv} and \code{uv\_decode} 
+have been inferred with user interaction as described in \pref{sec:anno}.
+The most challenging functions for $\toolname$ have unrolled loops
+where many variables are assigned.  This stresses our analysis because
 the number of LP variables has a quadratic growth in program variables.
 Even on these stressful examples, the analysis could finish in less
-than 2 seconds.  These times can even be greatly reduced by adding
+than 2 seconds.  \iffull{These times can be reduced by adding
 manual annotations to indicate $\toolname$ that certain variables do
-not have to be tracked.  We could also envision to implement heuristics.
-
-The trickiest of these benchmarks is the \code{sha\_update} function,
-it is composed of one loop calling two helper functions that in turn
+not have to be tracked.  We could also envision to implement heuristics.}{}
+%
+For example, the \code{sha\_update} function
+is composed of one loop calling two helper functions that in turn
 have 6 and 1 inner loops.  The essence of the SHA algorithm makes
@@
 since loops on the same index are sequenced 4 and 2 times without
-resetting it.  All other tools available have constant factors
-quadratically bigger than ours on this example.
+resetting it.  All other tools derive much larger constant factors.
 
@@
 \code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{MD5Final} & 195 & 141 & 0.22 \\
+\iffull{\code{MD5Final} & 195 & 141 & 0.22 \\}{}
 \code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
+\iffull{\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\}{}
 \code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
@@
 requirements of object oriented programs with region-based garbage
-collection.  These systems infer invariants and use external tools
+collection.  \iffull{These systems infer invariants and use external tools
 that count the number of integer points in the corresponding polytopes
 to obtain bonds.  The described technique can handle loops but not
-recursive or composed functions.
+recursive or composed functions.}{ These systems can handle loops but not
+recursive or composed functions.}
 %
@@
 have verified a loop bound analysis for CompCert's RTL intermediate
-language.  However, there is no way to interactively derive bounds or
-to deal with resources like memory usage.  Furthermore, Blazy et al.'s
-automatic bound analysis does not compute symbolic bounds.
+language.  However, this automatic bound analysis does not
+compute symbolic bounds. \iffull{Furthermore, there is no way to
+  interactively derive bounds or to deal with resources like memory
+  usage.}
 
@@
 framework combines ideas from existing abstract interpretation--based
-techniques with the potential method of amortized analysis.  The
-technical innovations are implemented in the publicly available tool
+techniques with the potential method of amortized analysis.  It is implemented in the publicly available tool
 $\toolname$. To the best of our knowledge, $\toolname$ is the first
 analysis tool for C programs that automatically reduces the derivation
-of symbolic bounds to efficient LP solving.
+of symbolic bounds to LP solving.
 
@@
 amortized resource analysis to compute bounds for programs that loop
-on signed integers and to deal with non-linear control flow that is
-introduced by break and return statements.
+on signed integers and to deal with non-linear control flow\ifshort{.}{ that is
+introduced by break and return statements.}
 
 This work is the starting point for several projects that we plan to
-investigate in the future. Examples include the extension to
-concurrent programs, better integration of low-level features like
-hardware caches, and the extension of the automatic analysis to
+investigate in the future, such as the extension to
+concurrency, better integration of low-level features like
+memory caches, and the extension of the automatic analysis to
 multivariate resource polynomials~\cite{HoffmannAH11}.

### 2014-11-13 paper/main.tex
some minor edits

--- 2014-11-13 paper/main.tex
@@
 fails to find a resource bound for a program, there is no possibility
-for sound user interaction to guide the tool during during bound
+for sound user interaction to guide the tool during bound
 derivation.
@@
 without special tricks or techniques.  Examples \emph{speed\_1} and
-\emph{speed\_2}, that are taken from previous work~\cite{GulwaniMC09},
+\emph{speed\_2} in Figure~\ref{fig:ex_list_2}, which are taken from previous work~\cite{GulwaniMC09},
 demonstrate that our method can handle \emph{tricky iteration

### 2014-11-13 paper/main.tex
updates in section 5

--- 2014-11-13 paper/main.tex
@@
 \begin{figure*}[t]
-\vspace{-.6cm}
+\vspace{-.75cm}
 \centering
@@
 \vspace{.1cm}
-\caption{An example derivation as produced by the automatic tool.}
+\caption{An example derivation as produced $\toolname$. The constraints are resolved by an off-the-shelf LP solver.}
 \label{fig:derivation}
@@
 the derivation rules of the automatic amortized analysis.  This is
-done in a bottom-up way, for each strongly connected component of the
+done in a bottom-up way, for each strongly connected component (SCC) of the
 call graph.  During this
 process our tool uses symbolic names for the rational coefficients
-$(q_i)$ in the rules.  Every time a linear constraint must be
-satisfied by these coefficients, it is recorded in a global list using
-the symbolic names.  We then feed the collected constraints to an
-off-the-shelf LP solver. (We currently use Coin-Or's CLP.)
+$q_i$ in the rules.  Each time a linear constraint must be
+satisfied by these coefficients, it is recorded in a global list for the SCC using
+the symbolic names.  We instantiate this constraint list for every call of
+a function from this SCC.
+
+We then feed the collected constraints to an
+off-the-shelf LP solver (currently Coin-Or's CLP).
 If the solver successfully finds a solution, we know that a derivation
-exists for the considered program.  We can then extract the initial
-$Q$ from the solver and get a resource bound for the program.  To get
-a full derivation we simply extract the complete solution from the
-solver, and apply it to the symbolic names $(q_i)$ of the coefficients
+exists and extract the values for the initial
+$Q$ from the solver to get a resource bound for the program.  To get
+a full derivation we extract the complete solution from the
+solver, and apply it to the symbolic names $q_i$ of the coefficients
 in the derivation.  If the LP solver fails to find a solution, an
@@
 
-\pref{fig:derivation} contains an example derivation as it is produced
-by our $\toolname$.  The upper case letters (with optional
+\pref{fig:derivation} contains an example derivation as produced
+by $\toolname$.  The upper case letters (with optional
 superscript) such as $Q^\text{de}$ are families of
@@
 The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}, it is parameterized by a logical context that is
+\textsc{Relax}. It is parameterized by a logical context that is
 used to gather information on interval sizes.  For example, we have
@@
 the outermost expression is a loop, so we use the rule \text{Q:Loop}
-at the root of the derivation tree.  At this point, we don't know yet
+at the root of the derivation tree.  At this point, we do not know yet
 whether a loop invariant exists.  But we produce the constraints
@@
 %
-This constraint expresses the fact that the potential before and
-after the loop body is equal and thus constitutes an invariant.
+These constraints express the fact that the potential functions before and
+after the loop body are equal and thus constitute an invariant.
 
 After the constraint generation, the LP solver is provided with an
-objective function to minimize.  We wish to
-minimize the initial potential, which constitutes the resource bound on the
+objective function to be minimized.  We wish to
+minimize the initial potential, which is a resource bound on the
 whole program.  Here it is given by $Q$.  Moreover, we would like to
@@
 This happens in almost no time in practice.
-The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise, thus
-the bound is $0.5 |[0,x]|$.
+The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise. Thus
+the derived bound is $0.5 |[0,x]|$.
 

### 2014-11-13 paper/main.tex
minor

--- 2014-11-13 paper/main.tex
@@
 without special tricks or techniques.  Examples \emph{speed\_1} and
-\emph{speed\_2} in Figure~\ref{fig:ex_list_2}, which are taken from previous work~\cite{GulwaniMC09},
+\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
 demonstrate that our method can handle \emph{tricky iteration

### 2014-11-13 paper/main.tex
section 7

--- 2014-11-13 paper/main.tex
@@
 This idea is illustrated by the program in \pref{fig:xmplincaux}.
-Parts of the code in blue are annotations that were added
+The parts of the code in blue are annotations that were added
 to the original program text.  The top-level loop increments
@@
 by our automated tool if a logical variable $na$ is introduced.
-This logical variable is a reification, in the program, of the
-number $\#_1(a)$.  For example, on line 6 of the example
+This logical variable is a reification of the
+number $\#_1(a)$ in the program.  For example, on line 6 of the example
 we are setting \lstinline{a[x]} to 0 and because of the condition we
@@
 and the original one, we use the inequality $lg >
-\log_2(h-l)$ as invariant.  This technique allows a simpler
+\log_2(h-l)$ as an invariant.  This technique allows a simpler
 proof because, when working with integer arithmetic, it is not always
-the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
+the case that $\log_2(x-x/2) = \log_2(x)-1$.
 
@@
 To base the soundness proof on a formal ground, we start by defining a
-new cost-aware operational semantics for Clight.  CompCert Clight's
+new cost-aware operational semantics for CompCert Clight.  Clight's
 operational semantics is based on small-step transitions and
@@
 
-The small-step semantics we define is standard, except that it tracks
+The small-step semantics is standard, except that it tracks
 the resource consumption of a program.  The semantics is parametric in
@@
 \end{mathpar}
-The rules keep track of the resource cost $\cost$ at all times, and
-all rules have as implicit side condition that $\cost$ in the start
-configuration is non-negative.  It makes it possible
-to detect a resource crash as a stuck configuration where $\cost<0$.
+The rules have as implicit side condition that $\cost$ is
+non-negative.  This makes it possible to detect a resource crash as a
+stuck configuration where $\cost<0$.
 
@@
 \paragraph{Quantitative Hoare Logic.}
-To prove the soundness of \toolname we found it useful to go through an
-intermediate step using a Quantitative Hoare Logic.  This logic is at
-the same time a convenient semantic tool and a clean way to interface
-manual proofs with our automation.  We base it on the one presented in
-\cite{veristack14}, we add support for arbitrary resources and simplify
-the handling of auxiliary state.
-
-Following this previous work, we define quantitative Hoare triples
+To prove the soundness of \toolname we found it useful to go through
+an intermediate step using a quantitative Hoare logic.  This logic is
+at the same time a convenient semantic tool and a clean way to
+interface manual proofs with our automation.  We base it on a logic
+for stack usage~\cite{veristack14}, add support for arbitrary
+resources, and simplify the handling of auxiliary state.
+
+We define quantitative Hoare triples
 as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
@@
 least $Q(\state')$ resources left if the evaluation terminates in
-state $\state'$.  The logic rules are similar to the ones in previous
-work and are simply extended to account for all the costs found in our
+$\state'$.  The logic rules are similar to the ones in previous
+work and generalized to account for the cost introduced by  our
 cost-aware semantics.
@@
 version of this paper \cite{anon_extended}, provides explanations for
-many rules and a thorough overview of our soundness proof.}{Details about
+the rules and a thorough overview of our soundness proof.}{Details about
 the logic and its soundness proof are provided in the \pref{app:logic}.}
@@
 \paragraph{The Soundness Theorem.}
-We use the logic from previous paragraph as the target of a translation
+We use the quantitative logic as the target of a translation
 function for the automatic derivation system.  This reveals two orthogonal

### 2014-11-13 paper/main.tex
moved a figure
spacing

@@
   \centering
-\vspace{-.6cm}
+\vspace{-.8cm}
 \hspace{-0.4cm}
@@
 by our method. If $N \geq 8$, $\toolname$ infers the bound
-$\frac{N}{8} |[0,l]|$, but if $N<8$, $\toolname$ infers $7\frac{8-N}{8} +
+$\frac{N}{8} |[0,l]|$, but if $N<8$, it infers $7\frac{8-N}{8} +
 \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
@@
 %
-The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
+The part $(\Gamma, Q) \vdash \!\! S \!\! \dashv (\Gamma', Q')$ of the judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 with starting state $\state$, the assertions in $\Gamma$
-hold, and at least $Q(\state)$ resource units are available then the
-evaluation does not run out of resources and---if the execution
-terminates in state $\state'$---there are at least $Q(\state')$
-resources left and $\Gamma'$ holds for $\state'$.
+hold, and at least $Q(\state)$ resources are available then the
+evaluation does not run out of resources and, if the execution
+terminates in state $\state'$, there are at least $Q(\state')$
+\mbox{resources left and $\Gamma'$ holds for $\state'$.}
 
@@
 Similarly, $(\Gamma_R, Q_R)$ is the postcondition that holds when
-returning from a function call with \code{return}. (This becomes clear
-in the rules.)
+returning from a function call with \code{return}. \iffull{(This becomes clear
+in the rules.)}{}
 
@@
 otherwise, and thus $|\inter v x| = 0$.
-
-\iffull{
-Another interesting rule is {\sc Q:Call}.  It needs to account for the
-changes to the stack caused by the function call, the arguments/return
-value passing, and the preservation of local variables.  We can sum up
-the main ideas of the rule as follows.
-\itemskip
-\begin{itemize}
-\item The potential in the pre- and postcondition of the function
-  specification is equalized to its matching potential in the callee's
-  pre- and postcondition.
-\itemskipIn
-\item
-  $\!$The potential of intervals $|\inter x y|$ is preserved
-  if $x$ and $y$ are local\iffull{ variables.}{.}
-\item
-  The unknown potentials after the call (e.g.
-  $|\inter x g|$, with $x$ local and $g$ global)
-  are set to zero in the postcondition.
-\end{itemize}
-\itemskip
-If $x$ and $y$ are local variables and $f(x,y)$ is
-called, {\sc Q:Call} splits the potential of
-$|\inter x y|$ in two parts, one part to perform the
-computation in the function $f$ and one part to keep for
-later use after the function call.  This splitting is
-realized by the equations $Q = P {+} S$ and $Q' =
-P' {+} S'$.  Arguments in the function precondition
-$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
-of names different from all program variables.  This
-prevents name conflicts from happening and ensures that
-the substitution $[\Vargs/\vec x]$ is meaningful.
-Symmetrically, we use the unique name $\Vret$ to represent
-the return value in the function's postcondition
-$(\Gamma'_f, Q'_f)$.
-}{}
-
-\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition. We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}\iffull{ that
-also deals with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq l$ is
-larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}
 
@@
 \end{figure*}
+
+\iffull{
+Another interesting rule is {\sc Q:Call}.  It needs to account for the
+changes to the stack caused by the function call, the arguments/return
+value passing, and the preservation of local variables.  We can sum up
+the main ideas of the rule as follows.
+\itemskip
+\begin{itemize}
+\item The potential in the pre- and postcondition of the function
+  specification is equalized to its matching potential in the callee's
+  pre- and postcondition.
+\itemskipIn
+\item
+  $\!$The potential of intervals $|\inter x y|$ is preserved
+  if $x$ and $y$ are local\iffull{ variables.}{.}
+\item
+  The unknown potentials after the call (e.g.
+  $|\inter x g|$, with $x$ local and $g$ global)
+  are set to zero in the postcondition.
+\end{itemize}
+\itemskip
+If $x$ and $y$ are local variables and $f(x,y)$ is
+called, {\sc Q:Call} splits the potential of
+$|\inter x y|$ in two parts, one part to perform the
+computation in the function $f$ and one part to keep for
+later use after the function call.  This splitting is
+realized by the equations $Q = P {+} S$ and $Q' =
+P' {+} S'$.  Arguments in the function precondition
+$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
+of names different from all program variables.  This
+prevents name conflicts from happening and ensures that
+the substitution $[\Vargs/\vec x]$ is meaningful.
+Symmetrically, we use the unique name $\Vret$ to represent
+the return value in the function's postcondition
+$(\Gamma'_f, Q'_f)$.
+}{}
+
+\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition. We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax}\iffull{ that
+also deals with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq l$ is
+larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}
 

### 2014-11-13 paper/main.tex
new example

@@
 %
-& t08 (modified t08a) & t19 & t30 & t15 & t13\\
+& t63 & t19 & t30 & t15 & t13\\
 %
@@
 \begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (y>x) {
-  x++; tick(1);
-}
-while (x>2) {
-  x -= 3; tick(1);
-}
+i=1; j=0;
+while (j<x) {
+  j++;
+  if (i>=4)
+    i=1, tick(40);
+  else i++;
+  tick(1); }
 \end{lstlisting}
@@
 \toolname
-  & {\small $\!\!\!1.33|[x,y]| {+} 0.33 |[0,x]|\!\!\!$ }
+  & {\small $11|[0,x]|$ }
   & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
@@
 Rank
-  & {\small $2 {+} y {-} x (?)$ }
+  & {\small $23{\cdot}x - 14$ }
   & {\small $54+k+i$ }
@@
 {\small
-  \begin{tabular}{c}
-    $\max(0, x{-}2)$\\ $+ 2\max(0, y {-} x)$
-  \end{tabular}
+  $41\max(x,0)$
 }
@@
   $\!\!\!\!$\begin{tabular}{c}
-    $\max(0, i{-}100)$ \\
-    $+ \max(0, k {+} i {+} 51)$
+    $\max(i{-}100, 0)$ \\
+    $+ \max(k {+} i {+} 51, 0)$
   \end{tabular}$\!\!\!\!$
@@
 \pref{fig:compar} shows five representative loop patterns from the
-evaluation.  Example \emph{t08} is a slightly modified version of
-\emph{t08a}, which is described in~\pref{sec:AAA}.  Example \emph{t19}
+evaluation.
+Example \emph{t63} is a loop that performs an expensive operation every
+4 steps.  $\toolname$ is the only tool able to amortize this cost over
+the input parameter $x$.
+Example \emph{t19}
 demonstrates the compositionally of the analysis.  The program consists

### 2014-11-13 paper/main.tex
t63 -> t09

@@
 %
-& t63 & t19 & t30 & t15 & t13\\
+& t09 & t19 & t30 & t15 & t13\\
 %
@@
 evaluation.
-Example \emph{t63} is a loop that performs an expensive operation every
+Example \emph{t09} is a loop that performs an expensive operation every
 4 steps.  $\toolname$ is the only tool able to amortize this cost over
@@
 \\[.7\baselineskip]
-      {\bf t63}
+      {\bf t09}
     \end{center}
@@
 
-Example \emph{t63} shows how amortized reasoning handles periodically
+Example \emph{t09} shows how amortized reasoning handles periodically
 expensive operations.  The top-level loop is iterated $x$ times in

### 2014-11-13 paper/main.tex
some changes to the intro section to explain the weakness of existing potential-based methods.

--- 2014-11-13 paper/main.tex
@@
 protocols~\cite{BartheGB09}.  In general, quantitative
-resource information at design time can provide useful
+resource information % at design time
+can provide useful
 feedback for developers.
@@
 
-Available techniques for automatically deriving worst-case resource bounds
-fall into two categories.  Techniques
-in the first category derive impressive bounds for
-numerical imperative programs, but are not compositional.
-This is problematic if one needs to derive global whole-program bounds.
-Techniques in the second category derive tight whole-program
-bounds for programs with regular loop and recursion patterns that
-decrease the size of an individual variable or data structure.
-They are highly compositional, scale for large
-programs, and work directly on the syntax.  However, % despite all
-% these theoretical and practical advantages,
-it has been
-a long-time open problem to make them work for typical imperative code
+Available techniques for {\em automatically} deriving worst-case
+resource bounds fall into two categories.  Techniques in the first
+category derive impressive bounds for numerical imperative programs,
+but are not compositional.  This is problematic if one needs to derive
+global whole-program bounds.
+%%%%%%
+Techniques in the second category derive
+tight whole-program bounds for programs with regular loop or
+recursion patterns that decrease the size of an individual variable or
+data structure. They are highly compositional, scale for large
+programs, and work directly on the syntax.  However,
+they do not support multivariate interval-based resource bounds
+(e.g., $x-y$) which are common in C programs. Indeed, 
+it has been a long-time open problem to develop compositional
+resource analysis techniques that can work for typical imperative code
 with non-regular iteration patterns, signed integers, mutation, and non-linear
@@
 
-Tools in the second category originate form the potential method of
-amortized analysis and type systems for functional
+Tools in the second category originate form the {\em potential method}
+of amortized analysis and type systems for functional
 programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
-definitions of object oriented programs~\cite{Jost06} and
+definitions of object-oriented programs~\cite{Jost06} and
 data-structure predicates of separation logic~\cite{Atkey10} can play
-the role of the type system in imperative programs. However, these
-techniques fail for numerical imperative programs because they lack
-fine-grained data-structure information to hook the analysis on.
-Another issue is that existing techniques associate \emph{potential}
+the role of the type system in imperative programs.
+However, a fundamental weakness of these potential-based techniques
+is that they can only associate \emph{potential}
 directly with single program variables. In imperative programs, this
@@
 % variables has to be updated after mutations to ensure soundness.
+
+% these
+% techniques fail for numerical imperative programs because they lack
+% fine-grained data-structure information to hook the analysis on.
+% Another issue is that existing techniques associate \emph{potential}
+% directly with single program variables.
 
@@
   \centering
-\vspace{-.8cm}
+\vspace{-.6cm}
 \hspace{-0.4cm}
@@
 by our method. If $N \geq 8$, $\toolname$ infers the bound
-$\frac{N}{8} |[0,l]|$, but if $N<8$, it infers $7\frac{8-N}{8} +
+$\frac{N}{8} |[0,l]|$, but if $N<8$, $\toolname$ infers $7\frac{8-N}{8} +
 \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
@@
 %
-The part $(\Gamma, Q) \vdash \!\! S \!\! \dashv (\Gamma', Q')$ of the judgement
+The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 with starting state $\state$, the assertions in $\Gamma$
-hold, and at least $Q(\state)$ resources are available then the
-evaluation does not run out of resources and, if the execution
-terminates in state $\state'$, there are at least $Q(\state')$
-\mbox{resources left and $\Gamma'$ holds for $\state'$.}
+hold, and at least $Q(\state)$ resource units are available then the
+evaluation does not run out of resources and---if the execution
+terminates in state $\state'$---there are at least $Q(\state')$
+resources left and $\Gamma'$ holds for $\state'$.
 
@@
 Similarly, $(\Gamma_R, Q_R)$ is the postcondition that holds when
-returning from a function call with \code{return}. \iffull{(This becomes clear
-in the rules.)}{}
+returning from a function call with \code{return}. (This becomes clear
+in the rules.)
 

### 2014-11-13 paper/main.tex
minor

@@
 For example, there is no concept of manual proofs of
-resource bounds and no framework can support composition
+resource bounds; and no framework can support composition
 of manually derived bounds with automatically inferred bounds.
@@
 introduce a \emph{logical state} using \emph{auxiliary variables}.
-Auxiliary variables do not change the behavior of the program and
-guide $\toolname$ during bound derivation.
+Auxiliary variables guide $\toolname$ during bound derivation
+but they do not change the behavior of the program and
 

### 2014-11-13 paper/main.tex
minor

@@
 
-This paper presents a new framework for automatically
+This paper presents a new compositional framework for automatically
 deriving resource bounds on C programs.  This new
@@
 \itemskipIn
-\item We implemented our resource bound analysis in the publically-available tool $\toolname$.
+\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
 \itemskipIn

### 2014-11-13 paper/main.tex
some minor updates in section 8

@@
 Example \emph{t19}
-demonstrates the compositionally of the analysis.  The program consists
+demonstrates the compositionality of the analysis.  The program consists
 of two loops that decrement a variable $i$.  In the first loop, $i$ is
@@
 down to $-1$.  However, between the loops we assign \code{i=i+k+50}.
-So in total the program performs $52 + |[-1,i]| + |[0,k]|$ increments.
+So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
 Our analysis finds this tight bound because our amortized analysis
@@
 increment the variable $y$ or we execute an inner loop in which $y$ is
-counted back to $0$.  The analysis computes a tight bound.
+counted back to $0$.  $\toolname$ computes a tight bound.
 \iffull{Again, the constants $0$ and $1$ in the
@@
 because most other tools only bound this specific cost. In
-\pref{fig:compar}, we show the bounds we derived (\toolname) together with
+\pref{fig:compar}, we show the bounds we derived ($\toolname$) together with
 the bounds derived by LOOPUS~\cite{SinnZV14} and
@@
 is composed of one loop calling two helper functions that in turn
-have 6 and 1 inner loops.  The essence of the SHA algorithm makes
-the compositionality of our analysis essential to get a tight bound
+have 6 and 1 inner loops.  In the analysis of the SHA algorithm,
+the compositionality of our analysis is essential to get a tight bound
 since loops on the same index are sequenced 4 and 2 times without
@@
 to our derived bound.  \pref{fig:3d} shows such a comparison for
-example \emph{t08}.  One can see that the derived constant factors are the
-best possible if the input variable $x$ is non-negative.  \iffull{If $x$ is
-negative then the bound is only slightly off.}{}
+example \emph{t08}, a variant of \emph{t08a} from \pref{sec:AAA}.  One
+can see that the derived constant factors are the best possible if the
+input variable $x$ is non-negative.  \iffull{If $x$ is negative then
+  the bound is only slightly off.}{}
 

### 2014-11-13 paper/main.tex
more rewording in intro

--- 2014-11-13 paper/main.tex
@@
 the role of the type system in imperative programs.
-However, a fundamental weakness of these potential-based techniques
+However, a major weakness of existing potential-based techniques
 is that they can only associate \emph{potential}
-directly with single program variables. In imperative programs, this
+with individual program variable or data structure.
+For C programs, this
 fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
-$y - i$ decreases, but not $|i|$. %  Mutation is yet another problem.
+$y-i$ decreases, but not $|i|$. %  Mutation is yet another problem.
 % While functional programs relate multiple fixed inputs (the typing
@@
 
-A general problem with existing tools is user interaction. When a tool
+A general problem with existing tools (in both categories) is user interaction. When a tool
 fails to find a resource bound for a program, there is no possibility

### 2014-11-13 paper/main.tex
bugfix in code for full version
another fix for the full version
moved a figure
spacing
bugfixes in formatting

@@
 \newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
-\newcommand{\paraskip}[0]{\ifshort{\vspace{-2pt}}{}}
+\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
 \newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
@@
 \belowdisplayshortskip=3pt
+\else
 \fi
@@
   \centering
-\vspace{-.8cm}
+\vspace{-.3cm}
 \hspace{-0.4cm}
@@
   tick(1);
-  ${\color{blue}%
+} ${\color{blue}%
   \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
-}
    \end{lstlisting}
@@
   tick(1);
-  ${\color{blue}%
+} ${\color{blue}%
   \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
-  }
    \end{lstlisting}
@@
   tick(1);
-  ${\color{blue}%
+} ${\color{blue}%
   \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
-}
    \end{lstlisting}
@@
   tick(9);
-  ${\color{blue}%
+} ${\color{blue}%
   \dashv \cdot; \, P(n,y) }$
-}
    \end{lstlisting}
@@
   \centering
-\ifshort{\vspace{-1.8cm}
+\ifshort{\vspace{-.3cm}
 \hspace{-0.4cm}}{}
@@
    \end{lstlisting}
-\vspace{-2.5ex}
+\vspace{-2.7ex}
 $0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
@@
    \end{lstlisting}
-\vspace{-2.5ex}
+\vspace{-2.8ex}
 $\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
@@
     \begin{center}
-   \begin{lstlisting}
-
-   \end{lstlisting}
-\vspace{-2.5ex}
 \begin{lstlisting}
@@
 when more data is available.  It is present in all the block encryption
-routines of PGP.  The same pattern is also used in performance critical
-code to unroll a loop, for example we found it in a bit manipulating
+routines of PGP and also used in performance critical
+code to unroll a loop. For example we found it in a bit manipulating
 function of libtiff library and a CRC computation routine of MAD,
-an MPEG audio decoder.  This looping pattern is handled particularly well
+an MPEG decoder.  This looping pattern is handled particularly well
 by our method. If $N \geq 8$, $\toolname$ infers the bound
@@
 \small
-\vspace{-.6cm}
+\vspace{-.4cm}
 \def \MathparLineskip {\lineskip=0.35cm}
@@
 \begin{figure*}[t]
-\vspace{-.75cm}
+\vspace{-.5cm}
 \centering
@@
 \textsc{Relax}. It is parameterized by a logical context that is
-used to gather information on interval sizes.  For example, we have
-$$
-\begin{array}{lll}
-P^\text{de} \succeq_{(\cdot)} P^\text{we}
-& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
-& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
-& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
-\end{array}
-$$
+used to gather information on interval sizes.  For example, $P^\text{de} \succeq_{(\cdot)} P^\text{we} \equiv$
+{\small
+  \begin{align*}
+ p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10} 
+\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
+\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+  \end{align*}
+}
 %
@@
 
+\iffull{
+\begin{table}[t]
+\centering
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{MD5Final} & 195 & 141 & 0.22 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
+\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\hline
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
+\end{tabular}
+\vspace{.1cm}
+\caption{Derived bounds for functions from cBench.}
+\label{tab:cbench}
+\end{table}
+}
+{
+\begin{table}[t]
+\centering
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\hline
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
+\end{tabular}
+\vspace{.1cm}
+\caption{Derived bounds for functions from cBench.}
+\label{tab:cbench}
+\end{table}
+}
+
+\begin{figure}[t]
+\center
+\includegraphics[width=.9\linewidth]{fig/bound3d}
+\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
+  (blue lines) and the measured runtime cost (red crosses) for example
+  \emph{t08}. For $x\ge 0$ the bound is tight.}
+\label{fig:3d}
+\end{figure}
+
 \sectskip
@@
 }{}
-
-\begin{table}[t]
-\centering
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\iffull{\code{MD5Final} & 195 & 141 & 0.22 \\}{}
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-\iffull{\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\}{}
-\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-\hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-\end{tabular}
-\vspace{.1cm}
-\caption{Derived bounds for functions from cBench.}
-\label{tab:cbench}
-\end{table}
-
-\begin{figure}[t]
-\center
-\includegraphics[width=.9\linewidth]{fig/bound3d}
-\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for example
-  \emph{t08}. For $x\ge 0$ the bound is tight.}
-\label{fig:3d}
-\end{figure}
 

### 2014-11-13 paper/main.tex
- updates to related work - formatting

@@
 \end{tabular}
-\vspace{.1cm}
+\vspace{.01cm}
 \caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  The output of Rank has been manually
-  simplified to fit the table.  }
+  several examples with linear bounds.  \iffull{The output of Rank has been manually
+  simplified to fit the table.}{}  }
 \label{fig:compar}
@@
 \end{tabular}
-\vspace{.1cm}
+\vspace{.01cm}
 \caption{Comparison of $\toolname$ with other
@@
 \centering
+\small
 \begin{tabular}{r|c|c|c}
@@
 \center
+\vspace{-.2cm}
 \includegraphics[width=.9\linewidth]{fig/bound3d}
@@
 automatically derive symbolic bounds and is not based on efficient LP
-solving.  In this article, we use a more general quantitative Hoare
-logic that is parametric in the resource of interest.  Our main
-contribution is the automatic amortized analysis for Clight that can
-be combined with~\cite{veristack14} to automatically derive parametric
-verified stack bounds for recursive functions.
+solving.  In contrast, our main
+contribution is an automatic amortized analysis for C that can
+derive parametric bounds for loops
+and recursive functions fully automatically.
+%
+In this article, we also use a more general quantitative Hoare
+logic that is parametric in the resource of interest.
 
 Our work has been inspired by type-based amortized resource analysis
-for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
-are three major improvements over previous work in the current paper,
-which presents the first automatic amortized resource analysis for C.
-First, we solved the long-standing open problem of extending automatic
-amortized resource analysis to compute bounds for programs that loop
-on (possibly negative) integers.  Second, we extended the analysis
-system to deal with non-linear control flow that is introduced by
-\code{break} and \code{return} statements.  Third, for the first time,
-we have combined an automatic amortized analysis with a system for
-interactively deriving bounds. In particular, recent
-systems~\cite{HoffmannS13} that deal with integers and arrays cannot
-derive bounds that depend on values in mutable locations or possibly
-negative integers, and thus do not apply to most of our example
-programs.
+for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  Here,
+we present the first automatic amortized resource analysis for
+C. None of the existing techniques can handle the example programs we
+describe in this work.
+%
+The automatic analysis of realistic C programs is enabled by three
+major improvements over previous work.  First, we solved the
+long-standing open problem of extending automatic amortized resource
+analysis to compute bounds for programs that loop on (possibly
+negative) integers without decreasing one individual number in each
+iteration.  Second, we extended the analysis system to deal with
+non-linear control flow that is introduced by \code{break} and
+\code{return} statements.  Third, for the first time, we have combined
+an automatic amortized analysis with a system for interactively
+deriving bounds. In particular, recent systems~\cite{HoffmannS13} that
+deal with integers and arrays cannot derive bounds that depend on
+values in mutable locations, possibly negative integers, or on
+differences between integers.
 

### 2014-11-13 paper/main.tex
minor
add knuth morris pratt
new total number of lines analyzed

@@
   existing tools on challenging micro benchmarks and the analysis of
-  more than 2000 lines of C code from the cBench benchmark suite.
+  more than $\totalLoc$ lines of C code from the cBench benchmark suite.
 
@@
 \toolname has automatically derived global resource bounds for
-more than 2000 lines of C code from the cBench benchmark suite.
+more than $\totalLoc$ lines of C code from the cBench benchmark suite.
 %
@@
 \item We present experiments with $\toolname$ on more
-  than 2000 lines of C code. A detailed comparison shows that our
+  than $\totalLoc$ lines of C code. A detailed comparison shows that our
   prototype is the only tool that can derive global bounds for larger
@@
 list of automatically derived function bounds. In total we analyzed
-more than 2000 lines of code.  In the LoC column
+more than $\totalLoc$ lines of code.  In the LoC column
 we not only count the lines of the analyzed function but
@@
 
-\iffull{
 \begin{table}[t]
@@
 \code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
 \hline
@@
 \end{table}
-}
-{
-\begin{table}[t]
-\small
-\centering
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-\hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-\end{tabular}
-\vspace{.1cm}
-\caption{Derived bounds for functions from cBench.}
-\label{tab:cbench}
-\end{table}
-}
+
+% \begin{table}[t]
+% \small
+% \centering
+% \begin{tabular}{r|c|c|c}
+% Function & LoC & Bound & Time (s) \\
+% \hline
+% \code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+% \code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+% \code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+% \code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+% \code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+% \code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+% \code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+% \code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+% \code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+% \code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
+% \hline
+% \code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+% \code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
+% \end{tabular}
+% \vspace{.1cm}
+% \caption{Derived bounds for functions from cBench.}
+% \label{tab:cbench}
+% \end{table}
 

### 2014-11-13 paper/main.tex
minor

--- 2014-11-13 paper/main.tex
@@
 
-A recent project~\cite{veristack14} has implemented and verified a
-quantitative logic to reason about stack-space usage, and modified the
-verified CompCert C compiler to translate C level bound to x86 stack
-bounds.  This quantitative logic is also based on the potential method
-but has very rudimentary support for automation that cannot
-automatically derive symbolic bounds and is not based on efficient LP
-solving.  In contrast, our main
-contribution is an automatic amortized analysis for C that can
-derive parametric bounds for loops
-and recursive functions fully automatically.
-%
-In this article, we also use a more general quantitative Hoare
-logic that is parametric in the resource of interest.
-
 Our work has been inspired by type-based amortized resource analysis
@@
 The automatic analysis of realistic C programs is enabled by three
-major improvements over previous work.  First, we solved the
+major improvements over previous work.  First, we extended the
+analysis system to associate potential with not just individual
+program variables but also multivariate intervals
+and, more generally, auxiliary variables.  Second, we solved the
 long-standing open problem of extending automatic amortized resource
@@
 negative) integers without decreasing one individual number in each
-iteration.  Second, we extended the analysis system to deal with
-non-linear control flow that is introduced by \code{break} and
-\code{return} statements.  Third, for the first time, we have combined
-an automatic amortized analysis with a system for interactively
-deriving bounds. In particular, recent systems~\cite{HoffmannS13} that
-deal with integers and arrays cannot derive bounds that depend on
-values in mutable locations, possibly negative integers, or on
-differences between integers.
+iteration.  Third, for the first time, we have combined an automatic
+amortized analysis with a system for interactively deriving bounds. In
+particular, recent systems~\cite{HoffmannS13} that deal with integers
+and arrays cannot derive bounds that depend on values in mutable
+locations, possibly negative integers, or on differences between
+integers.
+
+A recent project~\cite{veristack14} has implemented and verified a
+quantitative logic to reason about stack-space usage, and modified the
+verified CompCert C compiler to translate C level bound to x86 stack
+bounds.  This quantitative logic is also based on the potential method
+but has very rudimentary support for automation. It is not based on
+efficient LP solving so it cannot automatically derive symbolic
+bounds.  In contrast, our main contribution is an automatic amortized
+analysis for C that can derive parametric bounds for loops and
+recursive functions fully automatically.
+%
+In this article, we also use a more general quantitative Hoare
+logic that is parametric in the resource of interest.
 
@@
 Our system is the first amortized resource analysis for C programs.
-It addresses the longstanding open problem of extending automatic
+It addresses the long-standing open problem of extending automatic
 amortized resource analysis to compute bounds for programs that loop

### 2014-11-13 paper/main.tex
minor

@@
 it reduces linear bound inference  to a linear optimization problem
-that can solved by off-the-shelf LP solvers.  This enables the
+that can be solved by off-the-shelf LP solvers.  This enables the
 efficient inference of global bounds for larger programs.

### 2014-11-13 paper/main.tex
exchanged the Q:Update rule with Q:Skip
minor

@@
 \begin{mathpar}
-\Rule{Q:Update}
-{ q'_{xy}, q'_{yx} \in \Qplusz
-\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
+\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+{ }
+{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
 %
@@
 Auxiliary variables guide $\toolname$ during bound derivation
-but they do not change the behavior of the program and
+but they do not change the behavior of the program.
 

### 2014-11-13 paper/main.tex
minor

@@
 %
-In example \emph{speed\_1} we have one loop that first increments
+In Example \emph{speed\_1} we have one loop that first increments
 variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
@@
 
-Another simple example is given on \pref{fig:xmplbsaux} where
+Another simple example is given in \pref{fig:xmplbsaux} where
 a logarithmic bound on the stack consumption of a binary search
@@
   inner loop can as well be replace by something more interesting, say
-  $9$ and $10$ like in example \emph{t08}.  Then we still obtain a
+  $9$ and $10$ like in Example \emph{t08}.  Then we still obtain a
   tight linear bound.}{}
@@
 to our derived bound.  \pref{fig:3d} shows such a comparison for
-example \emph{t08}, a variant of \emph{t08a} from \pref{sec:AAA}.  One
+Example \emph{t08}, a variant of \emph{t08a} from \pref{sec:AAA}.  One
 can see that the derived constant factors are the best possible if the
@@
 \caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for example
+  (blue lines) and the measured runtime cost (red crosses) for Example
   \emph{t08}. For $x\ge 0$ the bound is tight.}
@@
 
-In example \emph{t10} we also do not restrict the inputs $x$ and $y$.
+In Example \emph{t10} we also do not restrict the inputs $x$ and $y$.
 They can be negative, positive, or zero.  The star {\tt *} in the
@@
 as well be replace by something more interesting, say $9$ and $10$
-like in example \emph{t08}.  Then we still obtain a tight linear
+like in Example \emph{t08}.  Then we still obtain a tight linear
 bound.
 
-Example \emph{t27} is similar to example \emph{t13}.  Instead of
+Example \emph{t27} is similar to Example \emph{t13}.  Instead of
 decrementing the variable $x$ in the outer loop we this time increment
@@
 
-Example \emph{t16} is an extension of example \emph{t15}. We again
+Example \emph{t16} is an extension of Example \emph{t15}. We again
 assume that $y$ is non-negative and use the same mechanism to iterate
@@
 
-At first sight, example \emph{t30} appears to be a simple loop that
+At first sight, Example \emph{t30} appears to be a simple loop that
 decrements the variable $x$ down to zero.  However, a closer look
@@
 %
-In example \emph{fig2\_1} we have one loop that first increments
+In Example \emph{fig2\_1} we have one loop that first increments
 variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
@@
 
-Example \emph{nested\_multiple} is similar to example \emph{fig2\_1}.
+Example \emph{nested\_multiple} is similar to Example \emph{fig2\_1}.
 Instead of incrementing variable $y$ in the outer loop, $y$ is here
 potentially incremented multiple times in each iteration of the outer
-loop.  The idea of example \emph{nested\_single} is similar.  However,
+loop.  The idea of Example \emph{nested\_single} is similar.  However,
 instead of incrementing variable $y$ in the inner loop, we increment
@@
 
-Example \emph{sequential\_single} is like example
+Example \emph{sequential\_single} is like Example
 \emph{nested\_single}.  The only difference is that the inner loop of
 \emph{nested\_single} is now evaluated after the outer loop.  Example
-\emph{simple\_multiple} is a variant of example \emph{fig2\_1} and
+\emph{simple\_multiple} is a variant of Example \emph{fig2\_1} and
 \emph{simple\_single} is a simple variant of \emph{nested\_single}.
@@
 
-In example \emph{fig4\_4} it is essential that $m$ is positive.  That
+In Example \emph{fig4\_4} it is essential that $m$ is positive.  That
 ensures that we can obtain constant potential for the interval size
@@
 of the loop.  As a result, the variable $i$ is either counted down to
-$0$ or up to $n$.  Our tool cannot handle example \emph{fig4\_5}
+$0$ or up to $n$.  Our tool cannot handle Example \emph{fig4\_5}
 because we don't do path sensitive reasoning.  Note however that it
@@
 
-Similar to example \emph{ex1}, example \emph{ex2}
+Similar to Example \emph{ex1}, Example \emph{ex2}
 form~\cite{GulwaniZ10} focuses on the number of iterations of the
@@
 number of iterations of the outer loop is bounded by $|[0,n]|$.
-Finally, example \emph{ex2} is similar to example
-\emph{nested\_single}, and \emph{ex3} is a variant of example
+Finally, Example \emph{ex2} is similar to example
+\emph{nested\_single}, and \emph{ex3} is a variant of Example
 \emph{t47}.
@@
 \code{count\_down} is similar to the function with the same name in
-example \emph{t37}.  However, we do not count down to $0$ but to a
+Example \emph{t37}.  However, we do not count down to $0$ but to a
 variable $y$ that is passed as an argument and we call the function
@@
 the two tests at the beginning that check if the inputs are positive
-are essential.  Finally, example \emph{Quick Sort} shows a skeleton of
+are essential.  Finally, Example \emph{Quick Sort} shows a skeleton of
 the quick sort sorting algorithm.  Since we can only derive linear

### 2014-11-13 paper/main.tex
minor
typos

@@
 \itemskip
-The main innovation that makes amortized analysis work
-on imperative languages is to base the analysis on a
+The main innovations that make amortized analysis work
+on imperative languages are to base the analysis on a
 Hoare-like logic and to track multivariate quantities instead
@@
 source software and the literature.  Our analysis can find
-asymptotically tight bounds for all but 1 of these patterns, and in
+asymptotically tight bounds for all but one of these patterns, and in
 most cases the derived constant factors are tight.  To compare $\toolname$
@@
 
-In contrast to the quantitative annotations, logical contexts $\Gamma$
-in the presented inference system are left abstract.  This allows for
-simpler rules and leaves room for future improvements.
+The abstract program state is represented as \emph{logical contexts} in the
+derivation system used by our automated tool.
 Our implementation finds these logical contexts using abstract
@@
 \paraskip
-\paragraph{Mutual Recursive Functions.}
+\paragraph{Mutually Recursive Functions.}
 
@@
 which is dual to \code{c\_down}.  Here, we
-c up $y$ by $2$ and recursively call \code{c\_down}.  $\toolname$ is
+count up $y$ by $2$ and call \code{c\_down}.  $\toolname$ is
 the only available system that computes a tight
@@
 Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
-of block-based cryptographic primitives: Data of arbitrary length, is
+of block-based cryptographic primitives: Data of arbitrary length is
 consumed in blocks and the leftover is stored in a buffer for future use
@@
 code to unroll a loop. For example we found it in a bit manipulating
-function of libtiff library and a CRC computation routine of MAD,
+function of the libtiff library and a CRC computation routine of MAD,
 an MPEG decoder.  This looping pattern is handled particularly well
@@
 loop has linear complexity, and feeding into our analysis gives the
-correct and tight worst-case bound $2+3|[l,h]|.$  With the amortized
-analysis approach in mind, it is clear that the three loops get their
+correct and tight worst-case bound $2+3|[l,h]|.$  With the potential
+method in mind, it is clear that the three loops get their
 potential from the same interval $[l,h]$.  However, KoAT fails to
@@
 
-\iffull{}{ The main parts of the soundness proved are formalized with
+\iffull{}{ The main parts of the soundness proof are formalized with
   Coq and available for download. The full definitions of the cost

### 2014-11-13 paper/main.tex
phrasing

@@
 constraints and the above quadruple; once done, the constraint
-sets can readily be reused for every call site and the function need
+set can readily be reused for every call site and the function need
 not be analyzed multiple times.
@@
 Therefore, the derivation rules are parametric in a function context $\Delta$
-that we leave implicit in the rules that we present here. More details
+that we leave implicit in the rules presented here. More details
 can be found in the extended version.
@@
 the derivation rules of the automatic amortized analysis.  This is
-done in a bottom-up way, for each strongly connected component (SCC) of the
+done in a bottom-up way for each strongly connected component (SCC) of the
 call graph.  During this
@@
 satisfied by these coefficients, it is recorded in a global list for the SCC using
-the symbolic names.  We instantiate this constraint list for every call of
-a function from this SCC.
+the symbolic names.  We reuse the constraint list for
+every call from outside the SCC.
 

### 2014-11-13 paper/main.tex
typos and layout
minor

--- 2014-11-13 paper/main.tex
@@
   \tr(J) := \tr(B); \tr(R) \vdash \htriple{\tr(P)} S {\tr(P')}.
-$$}
-The soundness of the automatic analysis can now be stated
+$$}%
+\noindent{}The soundness of the automatic analysis can now be stated
 formally with the following theorem.

### 2014-11-13 paper/main.tex

--- 2014-11-13 paper/main.tex
@@
 \begin{figure}[t]
-\center
 \begin{minipage}[b]{0.8\linewidth}
@@
 $\#_1(a)$, the logical variable $na$ is decremented.
-Similarly, on line 10, an array entry which was 0 becomes 1, so
+Similarly, on line 8, an array entry which was 0 becomes 1, so
 $na$ is incremented.  To complete the step 2 of the systematic

### 2014-11-13 paper/main.tex
some small updates in related work

@@
 %
-The automatic analysis of realistic C programs is enabled by three
+The automatic analysis of realistic C programs is enabled by two
 major improvements over previous work.  First, we extended the
@@
 program variables but also multivariate intervals
-and, more generally, auxiliary variables.  Second, we solved the
+and, more generally, auxiliary variables.  In this way, we solved the
 long-standing open problem of extending automatic amortized resource
@@
 negative) integers without decreasing one individual number in each
-iteration.  Third, for the first time, we have combined an automatic
+iteration.  Second, for the first time, we have combined an automatic
 amortized analysis with a system for interactively deriving bounds. In
@@
 but has very rudimentary support for automation. It is not based on
-efficient LP solving so it cannot automatically derive symbolic
+efficient LP solving and cannot automatically derive symbolic
 bounds.  In contrast, our main contribution is an automatic amortized

### 2014-11-13 paper/main.tex
Jan does not know C well

@@
 while (i>100) {
-  i--; tick(1) }
-i += k+50;
+  i--; tick(1);
+} i += k+50;
 while (i>=0) {

### 2014-11-13 paper/main.tex
citing the coin-or website
spacing

@@
 We then feed the collected constraints to an
-off-the-shelf LP solver (currently Coin-Or's CLP).
+off-the-shelf LP solver (currently CLP~\cite{clp}).
 If the solver successfully finds a solution, we know that a derivation
@@
 When the resource bound of a program depends on the contents of the
-heap, or is non-linear (e.g. logarithmic, exponential), we propose to
+heap, or is non-linear (e.g. logarithmic, exponential), we
 introduce a \emph{logical state} using \emph{auxiliary variables}.
@@
 \center
-\vspace{-.2cm}
+% \vspace{-.2cm}
 \includegraphics[width=.9\linewidth]{fig/bound3d}

### 2014-11-13 paper/main.tex
smt vs lp paragraph

@@
 used to gather information on interval sizes.  For example, $P^\text{de} \succeq_{(\cdot)} P^\text{we} \equiv$
+\vspace{-.3cm}
 {\small
@@
 }
-%
+\vspace{-.8cm}
+
+\noindent
 The other rules are syntax directed and applied inductively.  For example,
@@
 the derived bound is $0.5 |[0,x]|$.
+
+A notable advantage of the LP-based approach compared to
+SMT-solver--based techniques is that a satisfying assignment is
+a proof certificate instead of a counter example.
+
+\vspace{-.24cm}
 
@@
 
+\begin{table}[t]
+\centering
+\small
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{MD5Final} & 195 & 141 & 0.22 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+% \code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
+\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
+\hline
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
+\end{tabular}
+\vspace{.1cm}
+\caption{Derived bounds for functions from cBench.}
+\label{tab:cbench}
+\end{table}
+
 \pref{tab:compar} summarizes the results of our experiments.  It shows
@@
   the bound is only slightly off.}{}
-
-\begin{table}[t]
-\centering
-\small
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{MD5Final} & 195 & 141 & 0.22 \\
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
-\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
-\hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-\end{tabular}
-\vspace{.1cm}
-\caption{Derived bounds for functions from cBench.}
-\label{tab:cbench}
-\end{table}
 

### 2014-12-09 paper/main.tex
minor
work in progress (integration of the new rules)

@@
 % Also look for "SHORT" comments
-% \def\fullversion{}
+ \def\fullversion{}
 
@@
 is a simplified program state that maps variable names to integers,
-$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
+$|\inter a b| = \max(0,b-a)$, and $q_i \in \Q$.  To simplify the
 references to the linear coefficients $q_i$, we introduce an \emph{index
@@
 $\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
-is, a family of \ifshort{non-negative}{} rational numbers where only
-a finite number of elements are not zero. \iffull{ For soundness, we
-  require that potential functions always give non-negative results.
-  This is achieved by restricting coefficients in a quantitative
-  annotation to be always non-negative.}{}
+is, a family of rational numbers where only
+a finite number of elements are not zero.  \iffull{As a notable difference from
+previous works based on amortized-analysis, we do not require all the coefficients
+of our potential function to be positive.  In practice, we will see using
+a semantic argument that despite this additional freedom, potential functions
+always remain positive.}{}
 
@@
 { }
-{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
+{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
 %
@@
 { }
-{\!\!  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') \!\!}
+{ (\Gamma, Q_B); R; (\Gamma, Q_B) \vdash \code{break} \dashv (\Gamma', Q') }
 %
@@
 \Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
-{ }
-{ \!\! B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) \!\!}
+{ XXX }
+{ B; R; (\Gamma, Q{+}n) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
 %
@@
 \and \Rule{Q:Update}
-{ q'_{xy}, q'_{yx} \in \Qplusz
+{ q'_{xy}, q'_{yx} \in \Q
 \\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
@@
 \\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q) }
 { B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
@@
 \and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
-{ \Gamma \models y \ge 0 \!
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx} \!
+{ \Gamma \models y \ge 0
+\\ \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
 }
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
 }
@@
 { \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
 }
 { \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {-} y],Q) \vdash x \gets x - y \dashv (\Gamma, Q')
 }
@@
 \and \Rule{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
+\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
 }
 { \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {\pm} y], Q) \vdash x \gets x \pm y \dashv (\Gamma, Q')
 }
@@
 \and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+{ B; R; (\Gamma \land e, Q) \vdash S_1 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q) \vdash S_2 \dashv (\Gamma',Q')
 }
@@
 \and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q')
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
 }
@@
 }
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
   \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
 }
@@
 { }
-{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
+{ B; R; (\Gamma, Q) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
 %

### 2014-12-11 paper/main.tex
attempt to explain the new increment rule

@@
 \Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
-{ XXX }
-{ B; R; (\Gamma, Q{+}n) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
+{ n < 0 \implies Q \ge 0 }
+{ B; R; (\Gamma, Q) \vdash \code{tick}(n) \dashv (\Gamma, Q{-}n) }
 %
@@
 %
-\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
 { (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q) }
 { B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
-%
-\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
-{ \Gamma \models y \ge 0
-\\ \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:DecP}
-{ \Gamma \models y \ge 0
-\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {-} y],Q) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
 %
@@
 { \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:Dec}
+{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{xu}, -q_{ux})
+\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{ux}, -q_{xu})
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {-} y], Q) \vdash x \gets x - y \dashv (\Gamma, Q')
 }
@@
 }
-{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
 %
@@
 %
-\and
-\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
 { }
@@
 %
-\and
-\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
 { \fenv f = (\vec y, S_f)
+\\ Q' \ge 0
+\\ Q'_f \ge 0
 \\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
@@
 \and \Rule{Q:Weak}
-{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
-\\ \!\! \Gamma_1 \models \Gamma_2
-\\\\ Q_1 \succeq_{\Gamma_1} Q_2
+{  \!\! \Gamma_1 \models \Gamma_2
+\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
 \\ \Gamma'_2 \models \Gamma'_1
@@
 \\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\ \forall i.\, p_i, r_i \in \Qplusz
 \\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
@@
 \ifshort{
-\begin{figure*}
-\small
-\vspace{-.4cm}
-\def \MathparLineskip {\lineskip=0.35cm}
-\begin{mathpar}
-\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
-{ }
-{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
-%
-\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
-{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
-%
-\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
-{ \Gamma \models y \ge 0 \!
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx} \!
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
-}
-%
-\\ \Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
-}
-{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
-%
-\and \Rule{Q:Weak}
-{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
-\\ \!\! \Gamma_1 \models \Gamma_2
-\\\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ \Gamma'_2 \models \Gamma'_1
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-}
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
-%
-\\ \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
-\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
-\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
-\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
-}
-{ Q' \succeq_\Gamma Q }
-\vspace{-.3cm}
-\end{mathpar}
-\caption{Selected inference rules of the quantitative analysis.}
-\label{fig:auto}
-\end{figure*}
+                    \jan{XXX PLEASE INSERT FIGURE HERE XXX}
 }{}
@@
 The derivation system for the automatic amortized analysis is defined
-in~\pref{fig:auto}.  The derivation rules derive judgements of the form
+in~\pref{fig:auto}.  The derivation rules yield judgements of the form
 $$
@@
 %
-The part $(\Gamma, Q) \vdash \!\! S \!\! \dashv (\Gamma', Q')$ of the judgement
+The part $(\Gamma, Q) \! \vdash \! S \! \dashv \! (\Gamma', Q')$ of the judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 quantitative part $Q$ represents a potential function as a collection
-of non-negative numbers $q_i$ indexed by the index set $I$.  The
+of rational numbers $q_i$ indexed by the index set $I$.  The
 logical part $\Gamma$ is left abstract but is enforced by our
@@
 with starting state $\state$, the assertions in $\Gamma$
-hold, and at least $Q(\state)$ resources are available then the
+hold, $Q' {\ge} 0$, and at least $Q(\state)$ resources are available then the
 evaluation does not run out of resources and, if the execution
-terminates in state $\state'$, there are at least $Q(\state')$
-\mbox{resources left and $\Gamma'$ holds for $\state'$.}
+terminates in state $\state'$, there are at least $Q'(\state')$
+resources left and $\Gamma'$ holds for $\state'$.
 
@@
 such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
-In all the rules, we have the implicit side condition that all rational
-coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
+%
+% THE FOLLOWING IS NO LONGER TRUE
+% In all the rules, we have the implicit side condition that all rational
+% coefficients are non-negative.
+%
+Finally, if a rule mentions $Q$ and $Q'$
 and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
@@
 a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+like $q_{xa}$ and $q_{ax}$ that depend on $x$ after the assignment.
 
@@
 Therefore, the derivation rules are parametric in a function context $\Delta$
-that we leave implicit in the rules presented here. More details
-can be found in the extended version.
+that we leave implicit in the rules presented here.
+% More details can be found in the extended version.
 
@@
 rule.  This rule reuses its precondition as postcondition, it is
-explained by two facts: first, no resource is consumed by the skip
+explained by two observations: first, no resource is consumed by the skip
 operation, thus no potential has to be used to pay for the evaluation;
@@
 
-The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
-describe how the potential is distributed after a size change of
-a variable.
-\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
-and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
-when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
-Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
-$x \gets x+y$, it applies when we can deduce from the logical context
-$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
-can be applied for decrements and when $y$ is not positive.}
-The rules are all equivalent in the case where $y=0$.
-The rule {\sc Q:Inc} can be applied
-if we cannot deduce any information about $y$.  In $\toolname$,
-the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
-
-\iffull{To explain how rules for increment and decrement work, it is
-sufficient to understand the rule {\sc Q:IncP}.  The others follow the
-same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
-variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
+The rules {\sc Q:Inc} and {\sc Q:Dec} describe how the potential is
+distributed after a size change of a variable.
+The rule {\sc Q:Inc} is for incements $x \gets x+y$ and the symmetric
+rule {\sc Q:Dec} is for decrements $x \gets x-y$.
+The two rules are equivalent in the case where $y=0$.
+
+The two previous rules are essentially the same, so we only focus on
+the explanation of {\sc Q:Inc}. In this rule, the program updates a
+variable $x$ with $x+y$.  Since $x$ is changed, the
 quantitative annotation must be updated to reflect the change of the
@@
 Since $x$ is the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
-$\inter x u$ will get smaller with the update, and if $x' \in \inter x
-u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
-$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
-$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
-guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
-get bigger with the update.  We know that $|\inter v {x'}| \le y +
-|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
-$q_{vx}$ to pay for this change.  The rule ensures this
-only for $v \not\in \mathcal U$ because $x \le v$
-otherwise, and thus $|\inter v x| = 0$.
+$\inter u x$ and $\inter x u$ will be resized.  Consider $S(x) = q_{xu}
+|\inter x u| + q_{ux} |\inter u x|$, during the increment, $S$ will
+increase by at most $y \cdot \max(q_{ux}, -q_{xu})$.
+This means that we must decrease the potential $q_{0y}$ of $|\inter 0 y|$
+by $\max(q_{ux}, -q_{xu})$ to pay for the change on
+$|\inter x u|$ and $|\inter u x|$.  A symmetric argument explains the
+change on $q_{y0}$.
 

### 2014-12-18 cbench/macro/test_macro.c
add macro benchmark used in author response

@@ -1,0 +1,624 @@
+void memmove(void *dst, void *src, int n) {
+	char *d = dst, *s = src;
+	while (n > 0) {
+		// *d = *s; // unsupported heap assignment
+		d = d+1;
+		s = s+1;
+		n = n-1;
+	}
+}
+
+void memcpy(void *dst, void *src, int n) {
+	char *d = dst, *s = src;
+	while (n > 0) {
+		// *d = *s; // unsupported heap assignment
+		d = d+1;
+		s = s+1;
+		n = n-1;
+	}
+}
+
+void memset(void *p, int i, unsigned n) {
+	char *c = p;
+	while (n > 0) {
+		// *c = i; // unsupported heap assignment
+		c = c+1;
+		n = n-1;
+	}
+}
+
+typedef unsigned char boolean;
+typedef unsigned char byte;
+typedef byte *byteptr;
+typedef char *string;
+typedef unsigned short word16;
+
+
+
+typedef unsigned long word32;
+
+struct IdeaCfbContext {
+ byte oldcipher[8];
+ byte iv[8];
+ word16 key[(6*8 +4)];
+ int bufleft;
+};
+
+struct IdeaRandContext {
+ byte outbuf[8];
+ word16 key[(6*8 +4)];
+ int bufleft;
+ byte internalbuf[8];
+};
+
+typedef word16 uint16;
+
+ static uint16
+ mulInv(uint16 x)
+{
+    uint16 t0, t1;
+    uint16 q, y;
+    int X;
+
+    if (x <= 1)
+ return x;
+    t1 = 0x10001L / x;
+    y = 0x10001L % x;
+    if (y == 1)
+ return (1 - t1);
+    t0 = 1;
+    X = 32;
+    assert(X>0);
+    do {
+ q = x / y;
+ x = x % y;
+ t0 += q * t1;
+ if (x == 1)
+     return t0;
+ q = y / x;
+ y = y % x;
+ t1 += q * t0;
+ if (X <= 0) break;
+ X--;
+    } while (y != 1);
+    return (1 - t1);
+}
+
+
+
+
+static void ideaExpandKey(byte const *userkey, word16 * EK)
+{
+    int i, j;
+
+    for (j = 0; j < 8; j++) {
+ EK[j] = (userkey[0] << 8) + userkey[1];
+ userkey += 2;
+    }
+    for (i = 0; j < (6*8 +4); j++) {
+ i++;
[…]

### 2014-12-18 paper/main.tex
reflect bug fixes from the code in the rules
continue updating the paper

@@
 \and \Rule{Q:Update}
-{ q'_{xy}, q'_{yx} \in \Q
-\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+{ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
 }
@@
 \and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q) }
+{ Q \ge Q'
+\\ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q)
+}
 { B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
@@
 { \fenv f = (\vec y, S_f)
-\\ Q' \ge 0
+\\ Q_f \ge 0
 \\ Q'_f \ge 0
@@
 \ifshort{
-                    \jan{XXX PLEASE INSERT FIGURE HERE XXX}
+                    \jan{XXX THIS FIGURE NEEDS TO BE UPDATED XXX}
 }{}
@@
 \inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
-{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
-  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1cm]
+{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=1cm,width=20cm]
+  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2.5cm,rightskip=2.5cm, vdots=1cm]
     { \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
       { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
-        { \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
+        { \inferrule*[right={\scriptsize (Q:Dec)},leftskip=0cm,rightskip=0cm]
           {
           }
-          { (x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
-            \dashv (\cdot, P^\text{de})
-          }
-        }
-        { (x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
+          { (\cdot, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
+            \dashv (x {\ge} 0, P^\text{de})
+          }
+        }
+        { (\cdot, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
           \dashv (\cdot, P^\text{we})
@@
         }
-        { (x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
+        { (\cdot, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
           \dashv (\cdot, P^\text{ti})
@@
       }
-      { (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
+      { (\cdot, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
         \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq})
@@
     }
-    { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
+    { (\cdot, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
       \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if})
@@
       { }
-      { (x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash
+      { (\cdot, B^\text{br}); (\cdot,Q^\text{br}) \vdash
         \code{break} \dashv (\bot,P^\text{br})
@@
     }
-    { (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
+    { (\cdot, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
       \code{break} \dashv (\cdot,P^\text{el})
@@
   }
-  { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
+  { \hspace{3cm} (\cdot, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
     \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm}
@@
 }
-{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P)
+{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (\cdot, P)
 }
@@
 & B^\text{we} {=} B^\text{de} \land Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}
-    \land P^\text{de} \succeq_{(\cdot)} P^\text{we}
+    \land P^\text{de} \succeq_{(x{\ge}0)} P^\text{we}
 & p^\text{de}_{0,10} {=} q^\text{de}_{0,10} + q^\text{de}_{0,x}
@@
 The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}. It is parameterized by a logical context that is
-used to gather information on interval sizes.  For example, $P^\text{de} \succeq_{(\cdot)} P^\text{we} \equiv$
-\vspace{-.3cm}
+\textsc{Relax}. It is parameterized by a logical context
+used to gather information on interval sizes.  For example,
+$P^\text{de} \succeq_{(x{\ge}0)} P^\text{we} \equiv$
+% SSS SPACING
+%\vspace{-.3cm}
 {\small
 \begin{align*}
- p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}
-\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
-\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0,10}
+  \land p^\text{we}_{x,0} \leq p^\text{de}_{x,0} + u_{x,0} - v_{0,x} \land \;\;
+  \\
+  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}
+  \bigwedge_{(\alpha, \beta) \neq (0,10), (x,0)} p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
 \end{align*}
 }
-\vspace{-.8cm}
+% SSS SPACING
+%\vspace{-.8cm}
 
@@
 
-\vspace{-.24cm}
+% SSS SPACING
+%\vspace{-.24cm}
 
@@
 
-\iffull{}{ The main parts of the soundness proof are formalized with
+\iffull{}{The main parts of the soundness proof are formalized with
   Coq and available for download. The full definitions of the cost
@@
 the resource of interest for the user of our system.  We achieve this
-independence by parameterizing evaluations with a resource \emph{metric}
-$M$; a tuple of rational numbers and two maps.  Each of these
-parameters indicates the amount of resource consumed by a corresponding
-step in the semantics.  Resources can be released by using a negative
-cost. %  The metric used in the implementation can also
-% depend on other information that is statically available such as the
-% name of the called function or the number of arguments.
-Two sample rules for update and tick follow.
+independence by using a tick instruction that allows the programmer
+to specify how much and when resource is consumed.  If the resource
+usage is bound to the programming language (e.g. stack usage, number
+loop iterations and function calls), the tick statements can
+be inserted automatically by a pre-processing phase.  Resources can be
+released by using a negative tick.
+Two sample rules of the semantics for update and tick follow.
+
 \begin{mathpar}
-\Rule[rightskip=0.3cm]{U}
+\Rule{U}
 { \state' = \state[x \mapsto \sem{e}_\state] }
@@
 \rightarrow\\\\
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+(\state', \code{skip}, \cont, \cost)
 }
@@
 \rightarrow\\\\
-(\state, \code{skip}, \cont, \cost {-} M_t(n))
+(\state, \code{skip}, \cont, \cost {-} n)
 }
@@
 as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
-maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
+maps from program states to an element of $\Q\cup\{\infty\}$ that represents
 an amount of resources available.  The assertions $B$ and $R$ are
@@
 S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
-the empty continuation, and at least $Q(\state)$ resource units available
+the empty continuation, $Q'(\state) \ge 0$ for all state $\sigma$, and at least $Q(\state)$ resource units available
 then the evaluation does not run out of resources and there are at
-least $Q(\state')$ resources left if the evaluation terminates in
+least $Q'(\state')$ resources left if the evaluation terminates in state
 $\state'$.  The logic rules are similar to the ones in previous
-work and generalized to account for the cost introduced by  our
-cost-aware semantics.
+work.
 
@@
 translate the assumptions in function contexts of the automatic
-analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
-\forall z \, \vec v \, v . $
-%
-$$
+analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=$
+%
+\begin{multline*}
+\forall z \, \vec v \, v . \,
 (
-  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
+  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]), \\
   \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
   )
-$$}{}
+\end{multline*}}{}%
 %
@@
 $$}%
-\noindent{}The soundness of the automatic analysis can now be stated
+The soundness of the automatic analysis can now be stated
 formally with the following theorem.

### 2014-12-18 paper/main.tex
update logic rules and restore semantics paragraph

@@
 stuck configuration where $\cost<0$.
-
-% \pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
-% rules define a rewrite system for program configurations of the form
-% $(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
-% is the statement being executed, $\cont$ is a continuation that
-% describes what remains to be done after the execution of $S$, and
-% $\cost \in \Q$ is the non-negative number of resources available for
-% further execution.  All rules that can decrease $\cost$ have the
-% implicit side condition that the resource quantity available
-% \emph{before} the step is non-negative.  This means that we allow
-% $(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
-% transition relation $\smallstep{M}^*$ to indicate that the execution
-% ran out of resources.  However, every execution that reaches such a
-% state is stuck.
-%
-% \iffull{
-% A continuation $\cont$ represents the context of the execution of a
-% given statement.  A continuation can be the empty continuation
-% $\Kstop$ (used to start a program), a sequenced computation $\Kseq S
-% \cont$, a loop continuation $\Kloop S \cont$, or the continuation
-% $\Kcall r \theta \cont$ of a function call.
-% $$
-% \cont := \Kstop
-% \mid \Kseq S \cont
-% \mid \Kloop S \cont
-% \mid \Kcall r \theta \cont
-% $$
-% %
-% During the execution we assume a fixed function context $\fenv$ that
-% maps function names to a list of variables and the statement that
-% defines the function body.  It is used in the rule \textsc{S:Call}.
-% }{}
-%
-% The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
-% S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
-% following.  If the statement $S$ is executed in program state
-% $\state$, with continuation $\cont$, and with $\cost$ resources
-% available then---after a finite number of steps---the evaluation will
-% reach the new machine state $(\state', S', \cont',\cost')$ and there
-% are $\cost'$ resources available.  If $\cost' \ge 0$ then the
-% execution did not run out of resources and the resource consumption up
-% to this point is $\cost - \cost'$.  If this difference is negative
-% then resource became available during the execution.  If however
-% $\cost' < 0$ then the execution ran out of resources and is stuck.
-% The cost of the execution is then $\cost \ge 0$.
-%
-% \iffull{
-% \pref{lem:sem1} summarizes the main properties of the resource
-% counter of the semantics.  It can be proved by induction on the number
-% of steps.
-% \begin{lemma}
-% \label{lem:sem1}
-% \itemskip
-%   \begin{enumerate}
-%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
-%     \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
-%     \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
-% \itemskipIn
-%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
-%     \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
-%     \cont', d')$ then $\cost - \cost' = d - d'$.
-%   \end{enumerate}
-% \itemskip
-% \end{lemma}
-% %
-% If we assume that we have a \emph{positive resource metric}, that is,
-% a metric that assigns a positive cost to each evaluation step then we
-% can link bounds on the resource consumption under the metric to
-% termination.  This is formalized by \pref{lem:sem_term}.
-% %
-% \begin{lemma}
-%  \label{lem:sem_term}
-%  Let $M$ be a positive metric and assume that there exists $b \in
-%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
-%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
-%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
-%  \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
-%  and $\state''$ implies $n \le B$.
-% \end{lemma}
-% }
 
@@
 
-In this article, we describe our system for a subset of Clight that is
-sufficient to discuss the general ideas.  This subset is given by the
-following grammar.
+We describe our system for a subset of Clight that is sufficient to
+discuss the general ideas.  This subset is given by the following
+grammar.
 %
@@
 The most notable difference to full Clight is that we can only assign
-to variables and thus do not consider operations that update the heap.
+to variables and thus do not consider operations that update the heap
+or the stack through a pointer.
 Moreover, function arguments and return values are assumed to be
@@
 pointers, \code{goto} statements, \code{continue} statements, and
-\code{switch} statements.
+stack variables that have their address taken.
 
@@
 program for the automatic analysis.  We also add the built-in function
-\code{tick(n)} that can be called with a constant integer $n$ as a
-flexible way to model resource consumption or release (if $n$ is
+\code{tick(n)} that can be called with a constant rational number $n$
+as a flexible way to model resource consumption or release (if $n$ is
 negative).
@@
 { (\state, \code{assert}~e, \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_a)
+\smallstep
+(\state, \code{skip}, \cont, \cost)
 }
@@
 \Rule{S:BrkSeq}
-{}
+{ }
 { (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep{M}
+\smallstep
 (\state, \code{break}, \cont, \cost)
@@
 \Rule{S:BrkLoop}
-{}
+{ }
 { (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_b)
+\smallstep
+(\state, \code{skip}, \cont, \cost)
 }
@@
 \Rule{S:RetSeq}
-{}
+{ }
 { (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep{M}
+\smallstep
 (\state, \code{return}~x, \cont, \cost)
@@
 \Rule{S:RetLoop}
-{}
+{ }
 { (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep{M}
+\smallstep
 (\state, \code{return}~x, \cont, \cost)
@@
 { (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_r)
+\smallstep
+(\state', \code{skip}, \cont, \cost)
 }
@@
 { (\state, x \gets e, \cont, \cost)
-\smallstep{M}
-(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
+\smallstep
+(\state', \code{skip}, \cont, \cost))
 }
@@
 { (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep{M}
-(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
+\smallstep
+(\state', S_f, \Kcall r \theta \cont, \cost)
 }
@@
 \Rule{S:Loop}
-{}
+{ }
 { (\state, \code{loop}~S, \cont, \cost)
-\smallstep{M}
+\smallstep
 (\state, S, \Kloop S \cont, \cost)
@@
 \Rule{S:SkipLoop}
-{}
+{ }
 { (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep{M}
-(\state, \code{loop}~S, \cont, \cost {-} M_l)
+\smallstep
+(\state, \code{loop}~S, \cont, \cost)
 }
@@
 { \code{istrue}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
 }
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_1, \cont, \cost')
+\smallstep
+(\state, S_1, \cont, \cost)
 }
@@
 { \code{isfalse}~\sem{e}_\sigma
-\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
 }
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep{M}
-(\state, S_2, \cont, \cost')
+\smallstep
+(\state, S_2, \cont, \cost)
 }
@@
 { (\state, S_1; S_2, \cont, \cost)
-\smallstep{M}
+\smallstep
 (\state, S_1, \Kseq {S_2} \cont, \cost)
@@
 \Rule{S:SkipSeq}
-{}
+{ }
 { (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep{M}
-(\state, S, \cont, \cost {-} M_s)
+\smallstep
+(\state, S, \cont, \cost)
 }
@@
 { (\state, \code{tick}(n), \cont, \cost)
-\smallstep{M}
-(\state, \code{skip}, \cont, \cost {-} M_t(n))
+\smallstep
+(\state, \code{skip}, \cont, \cost - n)
 }
@@
 \end{figure*}
+
+\paragraph{Semantics.}
+
+\pref{fig:opsem} contains the reduction rules of the semantics.  The rules
+define a rewrite system for program configurations of the form $(\state,
+S, \cont, \cost)$, where $\state$ is the program state, $S$ is the
+statement being executed, $\cont$ is a continuation that describes what
+remains to be done after the execution of $S$, and $\cost \in \Q$ is
+the amount of resources available for further execution.  All the rules
+have the implicit side condition that the resource quantity available
+\emph{before} the step is non-negative.  This means that we allow
+$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side transition
+relation $\smallstep^*$ to indicate that the execution ran out of
+resources.  However, every execution that reaches such a state is stuck.
+
+A continuation $\cont$ represents the context of the execution of a
+given statement.  A continuation can be the empty continuation
+$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
+\cont$, a loop continuation $\Kloop S \cont$, or the continuation
+$\Kcall r \theta \cont$ of a function call.
+$$
+\cont := \Kstop
+\mid \Kseq S \cont
+\mid \Kloop S \cont
+\mid \Kcall r \theta \cont
+$$
+%
+During the execution we assume a fixed function context $\fenv$ that
+maps function names to a list of variables and the statement that
+defines the function body.  It is used in the rule \textsc{S:Call}.
+
+The intuitive meaning of an evaluation $(\state,
+S, \cont, \cost) \smallstep^* (\state', S', \cont', \cost')$ is the
+following.  If the statement $S$ is executed in program state
+$\state$, with continuation $\cont$, and with $\cost$ resources
+available then---after a finite number of steps---the evaluation will
+reach the new machine state $(\state', S', \cont',\cost')$ and there
+are $\cost'$ resources available.  If $\cost' \ge 0$ then the
+execution did not run out of resources and the resource consumption up
+to this point is $\cost - \cost'$.  If this difference is negative
+then resource became available during the execution.  If however
+$\cost' < 0$ then the execution ran out of resources and is stuck.
+The cost of the execution is then $\cost \ge 0$.
+
+\pref{lem:sem1} summarizes the main properties of the resource
+counter of the semantics.  It can be proved by induction on the number
+of steps.
+\begin{lemma}
+\label{lem:sem1}
+\itemskip
+  \begin{enumerate}
+  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
+    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
+    \smallstep^k (\state', S', \cont', \cost' {+} n)$.
+\itemskipIn
+  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
+    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep^k (\state', S',
+    \cont', d')$ then $\cost - \cost' = d - d'$.
+  \end{enumerate}
+\itemskip
+\end{lemma}
+
+% OOPS this is no longer true... This is an argument agains a presentation
+% with ticks only. XXX
+%
+% If we assume that we have a \emph{positive resource metric}, that is,
+% a metric that assigns a positive cost to each evaluation step then we
+% can link bounds on the resource consumption under the metric to
+% termination.  This is formalized by \pref{lem:sem_term}.
+%
+% \begin{lemma}
+%  \label{lem:sem_term}
+%  Let $M$ be a positive metric and assume that there exists $b \in
+%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep^*
+%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
+%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
+%  \smallstep^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
+%  and $\state''$ implies $n \le B$.
+% \end{lemma}
 
@@
 \htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
+  { B }{ \code{break} }{ Q }
+}
+\and
+\Rule{L:Tick}
+{ n < 0 \implies Q \ge 0}
+{ \Delta; B; R \vdash
+\htriple
+  { Q }
+  { \code{tick}(n) }
+  { Q - n }
+}
+%
 %
@@
 \and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+\Rule{L:Assert}
 { }
@@
 \htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+  { \code{istrue}~\sem{e}_\state \implies Q }
+  { \code{assert}~e }
+  { Q }
 }
@@
 \and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+\Rule{L:Update}
+{ }
 { \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
+\htriple
+  { \lambda \state.\, Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ I \ge Q
+\\ \Delta; Q; R \vdash
+   \htriple { I }{ S }{ I }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P }
+  { S_1 }{ Q }
+\\\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' }
 \\
-\!\!  \Delta; B; R \vdash
+  \Delta; B; R \vdash
 \htriple{ Q' }{ S_2 }{ Q }
@@
   \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+  P \ge P_f \, y \, (\state(\vec x)) + A \\
+  \forall v.\, (Q_f \, y \, v + A \ge \lambda \state.\, Q \, \state[r \mapsto v])
 }
@@
 \htriple
-  { M_f + P }
+  { P }
   { r \gets f(\vec x) }
-  { Q - M_r }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
   { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
 }
@@
 \\
+  P_f \ge 0
+\\\\
   \forall f\,P_f\,Q_f.\,
@@
 %
+\and
+\Rule{L:Weaken}
+{
+  \Delta; B'; R' \vdash \htriple{ P' }{ S }{ Q' } \\\\
+  P \ge P' \!\! \\
+  \!\! Q' \ge Q \!\! \\
+  \!\! B' \ge B \!\! \\
+  \!\! \forall v.\, (R'\,v \ge R\,v)
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Q
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
 \end{mathpar}
@@
 $$
-P : \State \to \Qplusz \cup \{ \infty \} \; .
+P : \State \to \Q \cup \{ \infty \} \; .
 $$
@@
 $\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
-We write $\Assn$ for $\State \to \Qplusz \cup \{ \infty \}$ and $\bot$ for
+We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
 $\lambda \state . \, \infty$.  We sometimes call assertions
 \emph{potential functions}.  To use Coq's support for propositional
-reasoning, assertions have the type $\State \to \Qplusz \to \mathrm{Prop}$
+reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
 in the implementation.  For a given $\state \in \State$, such an
-assertion can be seen as a set $B \subseteq \Qplusz$ of valid bounds.
-However, we find the presentation in this article easier to read.
+assertion can be seen as a set $B \subseteq \Q$.  However, we find
+the presentation in this article easier to read.
 
@@
 Hoare logic and the intuitive meaning is as follows.  If $S$ is
-executed with starting state $\state$, the empty continuation
-$\Kstop$, and at least $P(\state)$ resources available then the
+executed with starting state $\state$, with the empty continuation
+$\Kstop$, $Q' \ge 0$, and at least $Q(\state)$ resources available then the
 evaluation does not run out of resources and there are at least
-$Q(\state')$ resources left if the evaluation terminates in state
+$Q'(\state')$ resources left if the evaluation terminates in state
 $\state'$.  The assertion $B : \Assn$ provides the
@@
 
-For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
-program state $\state$ $P(\state) \ge Q(\state)$.
+For two assertions $P,Q : \Assn$, we write $P \ge Q$ to if for all
+program state $\state$, $P(\state) \ge Q(\state)$.
 
@@
 except that instead of using the logical implication $\Rightarrow$ we
-use the quantitative $\models$ that point-wise applies $\ge$.  This
-rule indeed weakens the statement since it requires more resource to
-run the statement and yields less than what has been proved to be
-available after its termination.
+use the quantitative $\ge$.  This rule indeed weakens the statement
+since it requires more resource to run the statement and yields less
+than what has been proved to be available after its termination.
 
@@
   (m {\le} n \land P(\state) {\le} \cost \land
-    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
+    (\state, S, \cont, \cost) \smallstep^m (\_, \_, \_, \cost'))
   \implies \cost' {\ge} 0.

### 2014-12-22 paper/poly.tex
dig back the polynomial potential section

--- 2014-04-07 main.tex
@@
-% Some notes:
-%  1) Use Appel's SHA-256 implementation as an example
-%
-
-\documentclass[10pt]{article}
-\usepackage{fullpage}
-
-\usepackage[numbers]{natbib}
-
-\usepackage[utf8]{inputenc} %for utf8 input
-\usepackage{amssymb} %for shift symbol
-\usepackage{amsmath}
-\usepackage{listings} %for code
-\usepackage{mathpartir} %for typing rules
-\usepackage{microtype} %better micro typing
-\usepackage{stmaryrd} %for llbracket
-\usepackage{mathabx} % for boxes
-\usepackage{graphicx} %to include png images
-\usepackage{xcolor} %for colors
-\usepackage{url}
-\usepackage{enumitem}
-
-%----------------------------------------------------
-
-\usepackage{prettyref}
-\newcommand{\pref}[1]{\prettyref{#1}}
-\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
-\newcommand{\ppref}[1]{\vpageref[]{#1}}
-\newrefformat{fig}{Figure~\ref{#1}}
-\newrefformat{tab}{Table~\ref{#1}}
-\newrefformat{cha}{Challenge~\ref{#1}}
-\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}
-
-%----------------------------------------------------
-
-\usepackage{amsthm}
-\theoremstyle{newstyle}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}{Proposition}
-
-%----------------------------------------------------
-
-\input{definitions}
-
-\title{Automatic Amortized Analysis for Imperative Code}
-\date{}
-\author{}
-
-\begin{document}
-\maketitle
-
-\section{Decreasing to zero}
-
-We start with a programming language with loops counting
-downwards to 0.  Statements and expressions are defined
-below.  Assignments of complex expressions can be simulated
-by sequencing assignments of simple expressions.
-%
-\begin{mathpar}
-
-S := \code{skip}
-\mid \code{assert}(\Gamma)
-\mid x \gets E
-\mid \code{while}(0 < x)~S
-\mid \code{if}(?)~S~\code{else}~S
-\mid S;S
-
-\and
-
-E := x + y \mid x - y % \mid y
-\end{mathpar}
-%
-We define a deductive system to derive judgements of the
-form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.  The semantics
-of this judgement are defined below in the soundness section.
-
-\begin{mathpar}
-
-\Rule{Skip}
-{}
-{ \Gamma; Q + M_s \vdash \code{skip} \dashv Q; \Gamma }
-
-\and \Rule{Assert}
-{}
-{ \Gamma; Q + M_a \vdash \code{assert}(\Delta) \dashv Q; \Gamma, \Delta }
-
-% \\ \Rule{AssignP}
-% { \Gamma \models \Gamma'[x/x+y] }
-% {\Gamma; Q + M_s + y\cdot q_x \vdash x \gets x + y \dashv Q; \Gamma'}
-
-% \and\Rule{AssignM}
-% {  \Gamma \models \Gamma'[x/x-y]
-% \\ \Gamma \models c \le y \le x
-% }
-% { \Gamma; Q + M_s \vdash x \gets x - y \dashv c \cdot q_x + V; \Gamma' }
-
-\\ \Rule{While}
-{ \Gamma, 0 < x ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
-{ \Gamma; Q \vdash \code{while}(0 < x)~S \dashv Q - M_w^3; \Gamma, x \le 0 }
-
-\and\Rule{Seq}
-{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
-\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
-}
-{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
-
-\end{mathpar}
-
-\section{Decreasing differences}
-The second iteration is a system more symmetric where the potentials
-are linear functions in differences of variables.  From now on $k$
-denotes a non-negative constant.
-$$
-S := \code{while}(k < x - y)~S \mid \dots
-$$
-%
-In the rules below we take as convention that $q_{x}$ for a variable
-$x$ is in fact $q_{x0}$ where 0 is a special variable fixed to zero
-in all contraint contexts.
-
-\begin{mathpar}
-
-\Rule{AssignOpK}
-{  \Gamma \models \Gamma'[x/x \pm y]
-\\ \Gamma \models |y| = k
-\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
-\\ \textstyle q'_y = q_y
-      + \sum_{u \in \mathcal U} p_{xu}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-\\ q_{xu} = p_{xu} + r_{xu}
-}
-{ \textstyle
-  \Gamma; Q + M_s
-  \vdash x \gets x \pm y \dashv
-  \sum_u k \cdot r_{xu} + Q'; \Gamma'
-  % q_x is not omitted, v_{x} is in fact a shorthand for
-  % v_{x0} where 0 is a variable equal to 0.
-}
-
-\and \Rule{AssignOp}
-{  \Gamma \models \Gamma'[x/x \pm y]
-\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
-\\ \textstyle q'_y = q_y
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  \Gamma; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma'
-}
-
-\\ \mbox{(The other rules are unchanged)}
-
-\end{mathpar}
-
-\section{Cost aware semantics}
-
-This big step semantics is parameterized by a metric that sets
-the cost of each program operation.  Non-deterministically,
-the rule {\sc S:Abort} can be called to interrupt a computation.
-It has the same effect as raising an exception.  We symbolize a
-frozen computation with the notation $\circ$.  In the rules, $X$
-stands for either $\circ$ or $H$.
-
-\begin{mathpar}
-
-\Rule{S:Abort}
-{ }
-{ (H, S) \Downarrow_0 \circ }
-
-\and \Rule{S:Skip}
-{ }
-{ (H, \code{skip}) \Downarrow_{M_s} H }
-
-\and \Rule{S:Assign}
-{ }
-{ (H, x \gets e) \Downarrow_{M_s} H[x \mapsto \llbracket e \rrbracket_H] }
-
-\and \Rule{S:Assert}
-{ H \models \Gamma }
-{ (H, \code{assert}(\Gamma)) \Downarrow_{M_a} H }
-
-\and \Rule{S:WhileStop}
-{ \neg (k < \llbracket x - y \rrbracket_H) }
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^3} H }
-
-\and \Rule{S:WhileStep}
-{  k < \llbracket x - y \rrbracket_H
-\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
-\\ (H', \code{while}(k < x - y)~S) \Downarrow_{(q_2, q'_2)} X
-}
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }
-
-\and \Rule{S:WhileStepAbort}
-{  k < \llbracket x - y \rrbracket_H
-\\ (H, S) \Downarrow_{(q_1, q'_1)} \circ
-}
-{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) } \circ }
-
-\and \Rule{S:If}
-{  (H, S_i) \Downarrow_{(q, q')} X
-\\ i \in \{1, 2\}
-}
-{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_{M_c^i \cdot (q, q')} X }
-
-\and \Rule{S:Seq}
-{  (H, S_1) \Downarrow_{(q_1, q'_1)} H'
-\\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
-}
-{ (H, S_1; S_2) \Downarrow_{M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)} X }
-
-\and \Rule{S:SeqAbort}
-{ (H, S_1) \Downarrow_{(q,q')} \circ }
-{ (H, S_1; S_2) \Downarrow_{(q,q')} \circ }
-
-\end{mathpar}
-
-\section{Soundness theorem}
-
-\begin{theorem}[Terminating case]
-For any $S$, $H$, $H'$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$ such
-that $\Gamma; Q \vdash S \dashv Q'; \Gamma'$ and $H \models \Gamma$ and
-$(H,s) \Downarrow_{(q,q')} H'$, we have
-$H' \models \Gamma'$ and
-$q \le \Phi_H(Q)$ and
-$\Phi_H(Q) - \Phi_{H'}(Q') \ge q - q'$.
-\end{theorem}
-
-\begin{proof} By induction on the big step semantics.
-
-\begin{itemize}
-
-\item Case {\sc S:Skip}.
-  We have $(H, \code{skip}) \Downarrow_{M_s} H$.
-  By assumption, $H \models \Gamma$. \\
-  Since $\Phi_H(Q + M_s) = \Phi_H(Q) + M_s$,
-  \begin{itemize}[topsep=0pt]
-  \item
-    if $M_s \ge 0$ we have,
-      $M_s \le \Phi_H(Q + M_s)$ and
-      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge M_s - 0$;
-  \item
-    if $M_s < 0$ we have,
-      $0 \le \Phi_H(Q + M_s)$ and
-      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge 0 - (-M_s)$.
-  \end{itemize}
-
-\item Case {\sc S:Seq}.
-  By transitivity and induction hypothesis, it is
-  easy to get $H'' \models \Gamma''$.  Now we prove
-  the quantitative part of the theorem.  We have
-  $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
-  We assume $\forall i \in \{1, 2\},~M_s^i \ge 0$, the other cases
-  are similar.
-  \begin{itemize}[topsep=0pt]
-  \item
-    The first case is when $q'_1 > M_s^2 + q_2$ then $q = M_s^1 + q_1$
-    and since $q_1 \le \Phi_H(Q - M_s^1)$ (by induction) we have
-    $q \le \Phi_H(Q)$.
-  \item
-    The second case is when $q'_1 \le M_s^2 + q_2$ then $q = M_s^1 + q_1 + M_s^2 + q_2 - q'_1$. \\
-    Since $\Phi_H(Q - M_s^1) - \Phi_{H'}(Q') \ge q_1 - q'_1$
-    and $q_2 \le \Phi_{H'}(Q' - M_s^2)$ we have:
-    \begin{align*}
-    q &\le M_s^1 + \Phi_H(Q - M_s^1) - \Phi_{H'}(Q') + M_s^2 + q_2 \\
-      &= \Phi_H(Q) + (q_2 - \Phi_{H'}(Q' - M_s^2)) \le \Phi_H(Q).
-   \end{align*}
-  \end{itemize}
-  We obtain $\Phi_H(Q) - \Phi_{H''}(Q'') \ge q - q'$ by simply chaining
-  the two induction hypothesis and using the fact that $q - q' =
-  (q_1 + M_s^1 - q'_1) + (q_2 + M_s^2 - q'_2)$.
-
-\item Case {\sc S:WhileStop}.
-  This proof goes just like in the {\sc S:Skip} case.
-
-\item Case {\sc S:WhileStep}.
-  Here again we assume $\forall i \in \{1, 2, 3\},~M_w^i \ge 0$.
-  \begin{itemize}[topsep=0pt]
-  \item
-    If $q'_1 > M_w^2 + q_2$ then $q = M_w^1 + q_1$ and by induction
-    we get $q_1 \le \Phi_H(Q - M_w^1)$, so $q \le \Phi_H(Q)$.
-  \item
-    If $q'_1 \le M_w^2 + q_2$ then $q = M_w^1 + q_1 + M_w^2 + q_2 - q'_1$.
-    And we proceed like in the sequence case.
-  \end{itemize}
-  We obtain $\Phi_H(Q) - \Phi_{H''}(Q - M_w^3) \ge q - q'$ using the identity
-  $q - q' = (M_w^1 + M_w^2 + q_1 - q'_1) + q_2 - q'_2$ and the two
-  induction hypothesis
-  \begin{itemize}[topsep=0pt]
-  \item $\Phi_H(Q - M_w^1) - \Phi_{H'}(Q + M_w^2) \ge q_1 - q'_1$ and
-  \item $\Phi_{H'}(Q) - \Phi_{H''}(Q - M_w^3) \ge q_2 - q'_2.$
-  \end{itemize}
-
-\end{itemize}
-\end{proof}
-
 \section{Polynomial Potential}
@@
 %%
-
-\end{document}

### 2014-12-24 paper/main.tex
fix frame rule and update smallstep definition
finish modifications in tex!

@@
 variables, and can abstractly specify the resource cost of functions
-(\pref{sec:AAA}). Second, it is easily combinable with established
+(\pref{sec:AAA}). Second, it is easily combined with established
 qualitative verification to guide semi-automatic bound derivation
@@
 { }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
@@
 { n < 0 \implies Q \ge 0}
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
@@
 { I \ge Q
-\\ \Delta; Q; R \vdash
+\\ \Delta; Q; R \ldash
    \htriple { I }{ S }{ I }
 }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
@@
 \and
+\Rule{L:Seq}
+{ \Delta; B; R \ldash
+\htriple{ P }{ S_1 }{ Q' }
+\\\\
+  \Delta; B; R \ldash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \ldash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
 \Rule{L:If}
-{ \Delta; B; R \vdash
+{% P \ge 0
+%\\
+  \Delta; B; R \ldash
   \htriple
@@
   { S_1 }{ Q }
-\\\\
-  \Delta; B; R \vdash
+\\
+  \Delta; B; R \ldash
 \htriple
@@
 }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
   { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' }
-\\
-  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
 }
@@
 }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
@@
 \Rule{L:Extend}
-{  \Delta \cup \Delta'; B; R \vdash
+{  \Delta \cup \Delta'; B; R \ldash
   \htriple{ P }{ S }{ Q }
@@
   \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
+  (\Delta \cup \Delta'; \bot; Q_f\,y \ldash
     \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
 }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple{ P }{ S }{ Q }
@@
 {
-  \Delta; B'; R' \vdash \htriple{ P' }{ S }{ Q' } \\\\
+  \Delta; B'; R' \ldash \htriple{ P' }{ S }{ Q' } \\\\
   P \ge P' \!\! \\
@@
 }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple{ P }{ S }{ Q }
@@
 \Rule{L:Frame}
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple{ P }{ S }{ Q } \\
-  x \in \Q
-}
-{ \!\! \Delta; B + x; R + x \vdash
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \ldash
 \htriple{ P + x }{ S }{ Q + x } \!\!
@@
 \end{mathpar}
-
 \caption{Rules of the Quantitative Hoare Logic}
@@
 $$
-\Delta; B; R \vdash
+\Delta; B; R \ldash
 \htriple
@@
 
+The {\sc L:Tick} rule accounts for the cost $n$ of the tick statement.
+This rule is the only one to account directly for a cost on in the
+assertions because calling the \code{tick} function is the only way to
+change the amount of resource available.
+The rule also has a side condition that ensures that the potential
+remains positive during all the program execution.  It is an
+essential semantic property of this system.
+
 In the rule {\sc L:Assert}, we use the notation
-$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
-require potential $Q + M_a$ in the precondition if $e$ evaluates to
+$\code{istrue}~\sem{e}_\state \implies Q$ to express that we
+require potential $Q$ in the precondition if $e$ evaluates to
 $\mathit{true}$ in the current program state.  If $e$ evaluates to
-$\mathit{false}$ then the potential in the precondition can be
-arbitrary since the program will be terminated.
+$\mathit{false}$ then the potential in the precondition can be any
+non-negative number since the program will be terminated.
 
@@
 \code{break}, respectively.  In {\sc L:Break}, we require to have
-potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
-cost of \code{break} and $B$ to pay for the potential after the
+potential $B$ in the precondition to pay for the potential after the
 \code{break}.  In {\sc L:Return} we only require to have potential
 $R$ in the precondition to pay for the potential after the
-\code{return}.  The reason is that we found it to be more convenient
-to account for the execution cost $M_r$ of the return in rule {\sc
-  L:Call} for function calls.
+\code{return}.
 
@@
 the precondition we ensure that $Q$ evaluates to the same number as in
-the postcondition.  We also require that have we the constant
-potential $M_u + M_e(e)$ available in the precondition to pay for the
-cost of the evaluation of $e$ and the update.
+the postcondition.
 
@@
 $S_2$.  The interaction between the actual pre- and postconditions is
-analogous to standard Hoare logic.  In the postcondition of $S_1$ we
-account for the cost $M_s$ of the execution of the sequence.
+analogous to standard Hoare logic.
 
@@
 require to have potential $I$ available in the precondition of the
-loop body $S$.  In the postcondition of $S$, the potential must be
-sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
-iteration.
+loop body $S$.
 
@@
 $\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
-quantitative assertions.  In the precondition of the rule {\sc L:Tick}
-we account for the execution cost $M_t(n)$ of the \code{tick}
-statement that depends on the integer $n$.
-
-The rule {\sc L:Call} accounts for the execution cost of both $M_f$
-for function calls and $M_r$ for \code{return} statements.
-This justifies that the {\sc L:Return} rule does not account
-  for resource consumption. The pre- and postcondition $P_f$ and
-$Q_f$ are taken from the function context $\Delta$.  The assertions in
+quantitative assertions.
+
+The pre- and postcondition $P_f$ and $Q_f$ used in the {\sc L:Call} rule
+are taken from the function context $\Delta$.  The assertions in
 the context are parametric with respect to both the values of the
@@
 logical variables almost for free: we can simply use the variable and
-binding mechanisms of Coq, our host theory.  When an logical variable
+binding mechanisms of Coq, our host theory.  When a logical variable
 is needed we introduce it using a universal quantifier of Coq before
@@
 Theorem dbl_triple: forall Z,
-  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
+  qtriple ($\lambda\sigma$.$\sigma$(x)=Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x)=2*Z).
 \end{lstlisting}
@@
 In the example in~\pref{fig:xmplmax} we derive a precise runtime
-bound on a program that searches a maximal element in an array.  The
-cost metric that we use simply counts the assignments performed by the
-program.  Hence, the resource cost is closely related to the
-number of times the test \code{a[i] > m} is true during the
-execution.
+bound on a program that searches a maximal element in an array.
+Because of the location of the tick statements, the resource cost
+is closely related to the number of times the test \code{a[i] > m}
+is true during the execution.
 %
@@
 \code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
-potential unit to pay for the assignment.  In the other case, no
+potential unit to pay for the tick in the conditional.  In the other case, no
 maximum so far is encountered so $A(i) = A(i+1)$.
-% As a side remark, if $K \ge 0$, it is possible to
-% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
-% of the same program.  It can be done either by adapting the proof or by
-% applying the {\sc L:Frame} rule one top of the triple already derived.
-% More generally, without function calls, the {\sc L:Frame}
-% rule is admissible in our system.
+As a side remark, if $K \ge 0$, it is possible to
+have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
+of the same program.  It can be done either by adapting the proof or by
+applying the {\sc L:Frame} rule one top of the triple already derived.
+More generally, without function calls, the {\sc L:Frame}
+rule is admissible in our system.
 
 The example in~\pref{fig:xmplbs} shows a use case for logical
-variables as well as a metric for stack consumption. In a stack
-metric, we account a constant cost for a function call ($M_f>0$) that
-is returned after the call ($M_r = -M_f <0$).  All other resource cost
-are $0$.  We are interested in showing that a binary search function
+variables. On this example, we model the stack consumption:
+we account a constant cost for a function call ($M_f>0$) that
+is returned after the call ($M_r = -M_f <0$).
+We are interested in showing that a binary search function
 $\code{bsearch}$ has logarithmic stack consumption.  We use a logical
@@
 Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
-- 1$, and because $M_f = M_\code{bsearch}$ and $M_r =
--M_\code{bsearch}$ in the stack metric.  The rest of the proof does
-not involve any resource manipulation and is just bookkeeping of
-logical assertions.
+- 1$.  The rest of the proof is here to make sure that the first tick
+can be paid for using regular algebra on the $\log_2$ function.
 
@@
 ${\color{blue} \{ A(0) + N \} }$
-i=1; m=a[0];
-${\color{blue}%
-  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
+i=1; tick(1); m=a[0]; tick(1);
 while (i < N) {
@@
   if (a[i] > m)
-    m=a[i];
+    m=a[i], tick(1);
   ${\color{blue}%
@@
   ${\color{blue}%
+    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) + 1 \} }$
+  tick(1);
+  ${\color{blue}%
     \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
@@
   for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
-  the metric used here assigns a cost of 1 to every assignment
-  and 0 to all other operations.
+  we use ticks in this example to count the number of assignments performed
+  by the program.
   }
@@
 ${\color{blue}%
-  \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
+  P(Z) := \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
 bsearch(x,l,h) {
@@
       \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
-    l = bsearch(x,l,h); }
-  ${\color{blue}%
+    tick($M_\code{bsearch}$);
+    ${\color{blue}\{ P(Z-1) \}}$ l = bsearch(x,l,h); ${\color{blue}\{ Q(Z-1) \}}$
+    tick($-M_\code{bsearch}$);
+  } ${\color{blue}%
     \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
   return l;
-} ${\color{blue} \{ Z \cdot M_\code{bsearch} \} }$
+} ${\color{blue} Q(Z) := \{ Z \cdot M_\code{bsearch} \} }$
 \end{lstlisting}
 \caption{Example derivation of a stack usage bound for a binary
-  search program.  The used resource metric defines the cost $M_\code{bsearch}$ before
+  search program.  The stack usage is modeled by adding a cost $M_\code{bsearch}$ before
   the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
-  frame size of the function $\code{bsearch}$.  $Z$ is a
-  logical variable.
+  frame size of the function $\code{bsearch}$.  $Z$ is a logical variable.
   }
@@
 We already gave an intuition of the meaning of judgements
-derived in the logic.  To make it formal, we define the
-\emph{resource safety} $\safe n P S \cont$ of an assertion
-and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
-\begin{align*}
-  (m {\le} n \land P(\state) {\le} \cost \land
-    (\state, S, \cont, \cost) \smallstep^m (\_, \_, \_, \cost'))
-  \implies \cost' {\ge} 0.
-\end{align*}
-This predicate is step indexed by an integer $n$ that is used for
-induction in the soundness proof for the function-call and loop cases.
-The constraint $\cost' \geq 0$ imposed by the definition ensures that
-the program does not stop because of a resource error (recall that a
-negative resource counter on the right-hand side is a resource
-failure).  However, it does not rule out memory safety errors or
-assertion failures. This is because our logic does not prove any
-safety or correctness theorems but only focuses on resource usage.
-%
-An interesting detail in the definition is the natural number $m$.  We
-simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
-\cont $.  This would not be the case if we replaced all occurrences of
-$m$ by $n$ in the definition of validity.
+derived in the logic.  We start by defining an indexed
+predicate $\cdot \downarrow_n$ that defines a \emph{resource-safe}
+program configuration:
+\begin{multline*}
+  (\state, S, K, c) \downarrow_n \, := \\
+  \forall m \le n.\, \forall c'.\,
+    (\state, S, K, c) \smallstep^m
+    (\_, \_, \_, c') \implies 0 \le c'.
+\end{multline*}
+This predicate means that a program configuration can execute
+for a certain number of steps without running out of resources.
+Note that nothing is said about the program safety, i.e. a stuck
+configuration $C$ will satisfy $\forall n.\, C\downarrow_n$. This
+is because our logic does not prove any safety or correctness
+theorems but only focuses on resource usage.
+We can now be used to define the resource safety
+of a configuration with respect to a quantitative assertion.
+\begin{multline*}
+  \safe n P S \cont \, := \\
+  \forall \state\, c.\, P(\state) \le c \implies
+    0 \le c \wedge (\state, S, K, c) \downarrow_n.
+\end{multline*}
+Once again, we use an index in the definition of the predicate, it
+is used to perform proofs by induction for the function-call and
+loop rules of the logic.
+%
+An interesting and essential property of the previous predicate is
+that $\safe {n+1} P S \cont \implies \safe n P S \cont$.  To prove
+this weakening property, we use the quantification over all numbers
+smaller than $n$ in the definition of $\cdot \downarrow_n$.
 
@@
 \safeK n B R Q \cont & :=  \safe n B {\code{break}} \cont \\
- \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
+ \land& \forall x. \safe n {\lambda \state.R\,(\state(x))} {\code{return}~x} \cont \\
  \land& \safe n Q {\code{skip}} \cont
@@
 We can now define the \emph{semantic validity} of a judgement $B; R
-\vdash \htriple P S Q$ of the quantitative logic without function
+\ldash \htriple P S Q$ of the quantitative logic without function
 context as $\valid n B R P S Q :=$
-\begin{align*}
-& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
-& \implies \safe m {P+x} S \cont \; .
-\end{align*}
+\begin{multline*}
+\forall m \le n.\, \forall x \geq 0 .\, \forall \cont . \\
+  \safeK m {B{+}x} {R{+}x} {Q{+}x} \cont \\
+  \implies \safe m {P+x} S \cont .
+\end{multline*}
 Note how the validity of a triple embeds the frame rule of
@@
   &\forall f .\, \Delta(f) =
-    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
+    \forall z \, \vec v \, v .\, (P_f \, z \, \vec v, Q_f \, z \, v)
     \implies \\
@@
 \begin{theorem}[Soundness]
-  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
+  \label{thm:sound}
+  If $\Delta; B; R \ldash \htriple P S Q$ is derivable then
   $
     \forall n .\, \validC n \Delta
-      \implies \valid {n {+} 1} B R P S Q.
+      {\implies} \valid {n {+} 1} B R P S Q.
   $
@@
 expanding the definitions we see that $\Delta$ is valid
-for every $n$ and that $\Kstop$ is safe for every $n$. So
-we derive
+for every $n$ and that $\Kstop$ is safe for every $n$ if both
+$Q$, $R$, and $B$ are non-negative on all possible program
+states. So the \pref{thm:sound} becomes
 $$
-\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
+\Delta; B; R \ldash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
 $$

### 2014-12-24 paper/main.tex
final cosmetic modifications

@@
 }
-void main (int x,int y) {
+void start (int x,int y) {
   y = copy (x,y);
@@
 
-void main (int y, int z) {
+void start (int y, int z) {
   count_down(y,z);
@@
 }
-void main (int y, int z) {
+void start (int y, int z) {
   consume(); produce(); consume();
@@
   while (i < n) {
-    while (j >= 0 && t[i]!=p[j]) {
+    while (j >= 0
+        && t[i]!=p[j]) {
       k = b[j];
@@
 int gcd(int x, int y) {
-  if (x <= 0) return y;
-  if (y <= 0) return x;
+  if (x <= 0)
+    return y;
+  if (y <= 0)
+    return x;
 
@@
    \begin{lstlisting}
-void qsort(int a[], int lo, int hi) {
+void qsort(
+  int a[], int lo, int hi
+) {
   int m1, m2, n;
@@
 
-  n = nondet(); /* partition the array */
+  n = nondet(); /* partition */
   assert( n > 0 );
@@
 
-void main(int a[], int len) {
+void start(int a[], int len) {
   qsort(a, 0, len);

### 2015-01-05 paper/main.tex
remove references to resource metrics

@@
 The derivation of a resource bound using potential functions is best
-explained by example.  If we use the tick metric that assigns cost $n$
-to the function call \code{tick(n)} and cost $0$ to all other
-operations then the cost of the following example can be bounded by
+explained by example.  If we use a cost model that assigns cost $n$ to
+the function call \code{tick(n)} and cost $0$ to all other operations
+then the cost of the following example can be bounded by
 $|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
@@
 concept of a ranking function.  A potential function can be used like
-a ranking function if we use the tick metric and add the statement
+a ranking function if we add the statement
 \code{tick(1)} to every back edge of the program (loops and function
@@
 we can use a potential function to prove that Example~2 does not
-consume any resources in the tick metric.
+consume any resources.
 \begin{lstlisting}[basicstyle=\tt\small]
@@
 \caption{Assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
-  tick metric.  The derived bound  is
+  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$.
+  The derived bound is
   $2|[0,k]| + |[0,na]|$.
@@
 \caption{Assisted bound derivation using logical state.
-  % We use the tick metric to model the stack consumption of the program.
+  % We model the stack consumption of the program using tick statements.
   % Before a function call to \code{bsearch}, a tick statement pays for
@@
   $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
-  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
+  The bound derived is $|[0,N]| + |[0,nm]|$.
   }
@@
 To compare our tool with existing work, we focused on loop bounds and
-use a simple metric that counts the number of back edges (i.e., number
+use tick statements to counts the number of back edges (i.e., number
 of loop iterations) that are followed in the execution of the program
@@
 computed on pre-processed C with all unnecessary declarations removed.}{}
-We analyzed the functions using a metric that assigns a cost 1 to all
+We analyzed the functions by giving a cost 1 to all
 the back-edges in the control flow (loops, and function calls).
@@
 programs that can be automatically analyzed by our system.  For
-simplicity, and to compare our analysis with existing tools, we use a
-cost metric that counts the number of loop iterations and function
-calls.  This is the only cost metric supported by tools like
+simplicity, and to compare our analysis with existing tools, we use
+tick statements to count the number of loop iterations and function
+calls.  This is the only resource that can be measured by tools like
 KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
-we also use the \emph{tick metric} if we want to discuss features such as
-resource restitution.  The tick metric assigns cost $n$ to the
-statement \code{tick(n)} and cost $0$ to all other statements. Of
-course, the examples can also be analyzed with any other cost metric
-in our system.
+we also use (negative) tick statements to discuss features such as
+resource restitution.
 
@@
 Example \emph{ex1} from~\cite{GulwaniZ10} specifically focuses on the
-code in the \emph{if} statement.  So we use the \emph{tick metric} and
+code in the \emph{if} statement.  So we
 insert \code{tick(1)} inside the if statement to derive a bound on
@@
 Example \emph{t46} shows a program that uses and returns resources.
-Again, we use the \emph{tick metric} and the function \code{tick} to
+Again, we use the function \code{tick} to
 describe the resource usage.  The function \code{produce} produces

### 2015-01-06 paper/main.tex
minor fixes

@@
   }
-\label{fig:xmplmax}
+% \label{fig:xmplmax2}
 \end{figure}
@@
 programs that can be automatically analyzed by our system.  For
-simplicity, and to compare our analysis with existing tools, we use
-tick statements to count the number of loop iterations and function
-calls.  This is the only resource that can be measured by tools like
+simplicity, and to compare our analysis with existing tools,
+the examples are automatically instrumented with tick statements
+to count the number of loop iterations and function calls.
+This is the only resource that can be measured by tools like
 KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
-we also use (negative) tick statements to discuss features such as
+we use tick statements explicitely to discuss features such as
 resource restitution.
@@
    \caption{Well-Known Algorithms}
-  \label{fig:cat3}
+  \label{fig:wknown}
 \end{figure*}
@@
 
-\pref{fig:cat3} shows well-known algorithms that can be automatically
+\pref{fig:wknown} shows well-known algorithms that can be automatically
 analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows

### 2015-01-20 paper/poly.tex
add a lemma about multiplication of binomial coeffs

@@
+\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}
+
+\usepackage[utf8]{inputenc} %for utf8 input
+\usepackage{amssymb} %for shift symbol
+\usepackage{amsmath}
+\usepackage{listings} %for code
+\usepackage{mathpartir} %for typing rules
+\usepackage{microtype} %better micro typing
+\usepackage{stmaryrd} %for llbracket
+\usepackage{mathabx} % for boxes
+\usepackage{graphicx} %to include png images
+\usepackage{xcolor} %for colors
+\usepackage{url}
+\usepackage{enumitem}
+\usepackage{array} %for stupid tables
+
+%----------------------------------------------------
+
+\usepackage{prettyref}
+\newcommand{\pref}[1]{\prettyref{#1}}
+\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
+\newcommand{\ppref}[1]{\vpageref[]{#1}}
+\newrefformat{fig}{Figure~\ref{#1}}
+\newrefformat{app}{Appendix~\ref{#1}}
+\newrefformat{tab}{Table~\ref{#1}}
+\newrefformat{cha}{Challenge~\ref{#1}}
+\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}
+
+%----------------------------------------------------
+
+\usepackage{amsthm}
+\newtheorem{lemma}{Lemma}
+\newtheorem{corollary}{Corollary}
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}{Proposition}
+
+%----------------------------------------------------
+
+\input{definitions}
+\begin{document}
+
 \section{Polynomial Potential}
@@
 
+\subsection{Binomial basis}
+
+\begin{lemma}
+  There exists a family $(c^{ij}_k)_k$, such that
+  $$
+  \sum_k c^{ij}_k \binom n k = \binom n i \binom n j,
+  $$
+  and $c^{ij}_k = \binom k j \binom j {i+j - k}$ satisfies
+  the previous equation.
+\end{lemma}
+\begin{proof}
+  \begin{align*}
+    \binom n i \binom n j
+      &= \sum_k \binom {n-j} k \binom j {i-k} \binom n j \\
+      &= \sum_k \binom {n-j} {k-j} \binom n j \binom j {i+j-k} \\
+      &= \sum_k \binom n k \binom k j \binom j {i+j-k}.
+  \end{align*}
+  In the first equality, we use Vandermonde's convolution
+  formula on $\binom n i$.  In the second equality we change
+  the summation variable to $k-j$.  And in the third equation
+  we use the identity proved below.
+  \begin{align*}
+    \binom {n-j} {k-j} \binom n j
+      &= \frac {(n-j)!} {(n-k)! (k-j)!} \; \frac {n!} {(n-j)! j!} \\
+      &= \frac {n!} {(n-k)! (k-j)! j!} \\
+      &= \frac {n!} {(n-k)! k!} \; \frac {k!} {(k-j)! j!} \\
+      &= \binom n k \binom k j.
+  \end{align*}
+\end{proof}
+
 %%
@@
 %%
+
+\end{document}

### 2015-03-06 paper/poly.tex
start writing down the polynomial system

@@
-\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}
+\documentclass[nocopyrightspace,preprint]{sigplanconf-pldi15}
 
@@
 
-\section{Polynomial Potential}
+\section{Canonical Polynomial Potential}
+
+The following presentation of amortized analysis for resource
+inference on integer programs permits to obtain precise polynomial
+bounds by naturally extending the previous linear work. Innovations
+of this presentation include
+\begin{enumerate}
+\item the use of canonical polynoms $X^i$ at no cost on precision,
+ \\ (I wish)
+\item the use of negative coefficients in potential annotations,
+\item the support for operations with mixed effects on multivariate potential.
+\end{enumerate}
+
+\paragraph{Index Sets}
+
+Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
+family that maps pairs of variables to non-zero natural numbers,
+$$
+I = (i_{(x,y)})_{(x,y) \in V^2} .
+$$
+%
+We identify a family $I$ with the set of pairs $\{ ((x,y),i_{(x,y)})
+\mid (x,y) \in V^2\}$.
+Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
+instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
+from the context.
+%
+
+The \emph{degree} of an index $I = (i_{(x,y)})$
+is defined by
+$
+\deg(I) = \sum_{(x,y)} i_{(x,y)}.
+$
+We define $\ind_k(V) = \{ I \in \ind(V) \mid \deg(I) \leq k
+\}$ to be the set of indices of degree at most $k$.  Remark
+that with these notations, $\ind_0(V)$ for any $V$ is a
+singleton---it contains only the empty index.
+
+\paragraph{Resource Polynomials}
+
+Let $V$ be a set of variabls.  An index $I \in \ind(V)$ denotes a
+\emph{base polynomial} $P_I : \State \to \Q$ that maps a program
+state $\state$ to a rational number.  The definition of $P_I$ is
+$
+P_I(\state) = \prod_{(x,y)} |[\state(x),\state(y)]|^{i_{(x,y)}}.
+$
+Remark that following the standard convention for $\prod$, the
+empty index represents the constant base polynomial 1.
+
+A \emph{resource polynomial} is a linear combination of
+base polynomials with rational coefficients.  Resource polynomials
+form a vector space over $\Q$.
+
+\paragraph{Potential Annotations}
+
+We call \emph{potential annotations}
+the coordinates of a resource polynomial in the basis of
+all base polynomials.  An annotation $Q$ is readily represented
+by a collection $(q_I)_I$ binding rational numbers to indices.
+For example, the potential annotation $\{\{((x,y),2)\} \mapsto 0.5,
+\emptyset \mapsto 3\}$ specifies the resource polynomial
+$
+  P_Q(\state) = 0.5\cdot|[\state(x),\state(y)]|^2 + 3.
+$
+
+The notion of degree can be generalized to potential annotations:
+the annotation $Q = (q_I)_I$ has degree $\max_{q_I} \deg(I)$.
+Remark that the degree of an annotation is exactly the degree of
+the polynomial it represents.
+
+\paragraph{Potential Method for Resource Analysis}
+
+The previous potential annotations will be used to decorate
+programs and bound their resource consumption.  A program
+$P$ consumes $n$ resource units turning a state $\state$ into $\state'$,
+we note this $\state \rightarrow^n_P \state'$.  The \emph{potential
+method} make us look for two annotations $Q$ and $Q'$ such
+that
+$$
+  \forall n\, \state\, \state'.\, \state \rightarrow^n_P \state'
+    \implies P_Q(\state) \ge n + P_{Q'}(\state').
+$$
+The composition of these annotations through the whole program
+using a quantitative logic permits to derive a resource bound.
+In the sequel we explicit a relation that is sufficient to
+satisfy the previous condition when the program $P$ is an
+increment operation $x \gets x+y$.
+
+\paragraph{Shifts of Resource Polynomials}
+
+By symmetry we can consider the case where $P=x \gets x+y$ with
+$y>0$, the case where $y \le 0$ is only a few signs away.  Let
+$Q$ be a potential annotation, we will describe a $Q'$ and a
+$P^y_.$ map such that if $\state \rightarrow_P \state'$ (with no resource cost),
+$P_Q(\state) = P^y_{Q'}(\state')$.  We call $Q'$ a shift of $Q$ and
+write $Q' = \qsh_x^+ Q$.  (When $y\le0$ we get $\qsh_x^-Q$.)
+
+The shift of $Q$ is defined on an extended set of indices, namely
+$\ind^\square = \ind(V \cup \{\square\})$.  This square is not a real
+variable, but a helper with special semantics.
+We can now pose $\qsh_x^+ Q = (q'_I)_{I\in\ind^\square}$ where,
+
+\begin{align*}
+q'_I =
+q_{I[\square \gets x]}
+\prod_z
+   \binom{i_{(\square,z)} {+} i_{(x,z)}}{i_{(\square,z)}}
+   \binom{i_{(z,\square)} {+} i_{(z,x)}}{i_{(z,\square)}}
+   (-1)^{i_{(z,\square)}}
+   .
+\end{align*}
+
+The substitution operation $I[\square \gets x]$ defines an index $I'$
+in $\ind(V)$ with the same degree as $I$.  This index satisfies
+$$
+i'_{(x,z)} {=} i_{(x,z)} {+} i_{(\square,z)}, \;
+i'_{(z,x)} {=} i_{(z,x)} {+} i_{(z,\square)}, \;
+i'_{(u,v)} {=} i_{(u,v)} \mbox{ o.w.}
+$$
+Notations may seem heavy, but what they really express is that
+replacing $\square$ by $x$ in $x^2\square^3z^2$ yields $x^5z^2$.
+
+\clearpage
+\section{Binomial Polynomial Potential}
+
+This section presents the binomial version of the previous system.
+In practice, negative coefficients could be enough to avoid having to use
+this basis.  I also suspect that there is no significant simplifications
+in the presentation using the binomial basis.
 

### 2015-03-10 paper/poly.tex
complete the shifts section

@@
 Remark that following the standard convention for $\prod$, the
-empty index represents the constant base polynomial 1.
+empty index represents the constant base polynomial 1.  We will
+later use the notation $P_{(x,y)}(\state) = |[\state(x),\state(y)]|$
+to define $P_I$ as $P_I = \prod_{(x,y)} P_{(x,y)}$.
 
@@
 programs and bound their resource consumption.  A program
-$P$ consumes $n$ resource units turning a state $\state$ into $\state'$,
-we note this $\state \rightarrow^n_P \state'$.  The \emph{potential
+$C$ consumes $n$ resource units turning a state $\state$ into $\state'$,
+we note this $\state \rightarrow^n_C \state'$.  The \emph{potential
 method} make us look for two annotations $Q$ and $Q'$ such
@@
 $$
-  \forall n\, \state\, \state'.\, \state \rightarrow^n_P \state'
+  \forall n\, \state\, \state'.\, \state \rightarrow^n_C \state'
     \implies P_Q(\state) \ge n + P_{Q'}(\state').
@@
 In the sequel we explicit a relation that is sufficient to
-satisfy the previous condition when the program $P$ is an
+satisfy the previous condition when the program $C$ is an
 increment operation $x \gets x+y$.
@@
 
-By symmetry we can consider the case where $P=x \gets x+y$ with
+By symmetry we can consider the case where $C=x \gets x+y$ with
 $y>0$, the case where $y \le 0$ is only a few signs away.  Let
 $Q$ be a potential annotation, we will describe a $Q'$ and a
-$P^y_.$ map such that if $\state \rightarrow_P \state'$ (with no resource cost),
+$P^y_.$ map such that if $\state \rightarrow_C \state'$ (with no resource cost),
 $P_Q(\state) = P^y_{Q'}(\state')$.  We call $Q'$ a shift of $Q$ and
@@
 replacing $\square$ by $x$ in $x^2\square^3z^2$ yields $x^5z^2$.
+
+We now define an interpretation $P_.^y$ of indices containing
+$\square$:
+\begin{align*}
+P^y_{(\square,z)}(\state) &=
+  \left| [\sigma(x){+}\sigma(y), \sigma(z)] \,\triangle\,
+         [\sigma(x), \sigma(z)] \right| \\
+P^y_{(z,\square)}(\state) &=
+  \left| [\sigma(z), \sigma(x){+}\sigma(y)] \,\triangle\,
+         [\sigma(z), \sigma(x)] \right| \\
+P^y_{(\square, \square)}(\state) &= 0 \\
+P^y_{(u,v)}(\state) &= |[\sigma(u), \sigma(v)]|,
+\end{align*}
+we use $A \,\triangle\, B$ to denote the symmetric difference
+of $A$ and $B$.
+The quantity $P^y_{(\square,z)}(\state)$ is the size change of
+the interval $[x, z]$.  If $\state$ is the state before the
+assignment and $\state'$ is the state after the assignment, we
+have
+$P_{(x,z)}(\state) = P^y_{(x,z)}(\state') + P^y_{(\square,z)}(\state')$
+and
+$P_{(z,x)}(\state) = P^y_{(z,x)}(\state') - P^y_{(z,\square)}(\state')$.
+With these definitions, we have the following lemma.
+\begin{lemma}
+If $\state \rightarrow_C \state'$ and $\state(y)>0$ then
+$P_Q(\state) = P^y_{\qsh^+_x Q}(\state')$.
+\end{lemma}
 

### 2015-03-11 paper/poly.tex
add clumsy proof
odd spacing
include graphics

@@
          [\sigma(z), \sigma(x)] \right| \\
-P^y_{(\square, \square)}(\state) &= 0 \\
+P^y_{(\square, \square)}(\state) &= P_{(x,\square)}(\state)
+  = P_{(\square,x)}(\state) = 0 \\
 P^y_{(u,v)}(\state) &= |[\sigma(u), \sigma(v)]|,
@@
 \end{lemma}
+\begin{proof}
+\small
+\begin{align*}
+&P^y_{\qsh_x^+ Q}(\state')
+= \sum_I q'_I \prod_{(u,v)} P_{uv}^{uv} \\
+&= \sum_I q_{I[\square\gets x]} P_I \prod_z
+  \binom{\square z {+} xz}{\square z}P_{\square z}^{\square z} P_{xz}^{xz}
+  \binom{z\! \square\! {+} zx}{z \square}(-P_{z \square})^{z \square} P_{zx}^{zx} \\
+&= \sum_J q_J P_J \sum_{I[\square\gets x] = J} \prod_z
+  \binom{\square z {+} xz}{\square z}P_{\square z}^{\square z} P_{xz}^{xz}
+  \binom{z\! \square\! {+} zx}{z \square}(-P_{z \square})^{z \square} P_{zx}^{zx} \\
+&= \sum_J q_J P_J \prod_z
+  \left( \sum_{\square z + xz = j_{xz}}
+    \binom{\square z {+} xz}{\square z}P_{\square z}^{\square z} P_{xz}^{xz} \right)
+  \left( \sum_{z\square {+} zx = j_{zx}} \dots \right) \\
+&= \sum_J q_J P_J \prod_z
+  (P_{xz} + P_{\square z})^{j_{xz}}
+  (P_{zx} - P_{z \square})^{j_{zx}} \\
+&= \sum_J q_J P_J(\state) = P_Q(\state)
+\end{align*}
+\end{proof}
+
+\paragraph{Projection of Resource Polynomials}
+
+\DeclareGraphicsRule{*}{mps}{*}{}
+
+The shift operation perserves all the potential after an increment
+but it introduces a new dummy variable $\square$.  Because annotations
+need to be uniform for compositionality, we present in this paragraph
+a projection that eliminates this extra variable.  The main identity that
+we use is $P_{\square z}(\state) + P_{z \square}(\state) =
+P_{0y}(\state)$.  It can be checked on all the three cases illustrated
+below.
+$$
+\includegraphics{fig/ivals.1}
+$$
+In the first case, $P_{\square z}$ will be 0 and $P_{z \square}$ will
+be non zero; the third case witnesses the opposite situation; and in
+the second case, both are non-zero.
 

### 2015-03-16 paper/poly.tex
more proofs, the notations are clumsy

@@
 empty index represents the constant base polynomial 1.  We will
-later use the notation $P_{(x,y)}(\state) = |[\state(x),\state(y)]|$
-to define $P_I$ as $P_I = \prod_{(x,y)} P_{(x,y)}$.
+later use the notation $P_{xy}(\state) = |[\state(x),\state(y)]|$
+to define $P_I$ as $P_I = \prod_{(x,y)} P_{xy}$.
 
@@
 
-The substitution operation $I[\square \gets x]$ defines an index $I'$
+The substitution operation $I[\square \gets x]$ defines an index $J$
 in $\ind(V)$ with the same degree as $I$.  This index satisfies
 $$
-i'_{(x,z)} {=} i_{(x,z)} {+} i_{(\square,z)}, \;
-i'_{(z,x)} {=} i_{(z,x)} {+} i_{(z,\square)}, \;
-i'_{(u,v)} {=} i_{(u,v)} \mbox{ o.w.}
+j_{(x,z)} {=} i_{(x,z)} {+} i_{(\square,z)}, \;
+j_{(z,x)} {=} i_{(z,x)} {+} i_{(z,\square)}, \;
+j_{(u,v)} {=} i_{(u,v)} \mbox{ o.w.}
 $$
@@
 \begin{align*}
-P^y_{(\square,z)}(\state) &=
+P^y_{\square z}(\state) &=
   \left| [\sigma(x){+}\sigma(y), \sigma(z)] \,\triangle\,
          [\sigma(x), \sigma(z)] \right| \\
-P^y_{(z,\square)}(\state) &=
+P^y_{z \square}(\state) &=
   \left| [\sigma(z), \sigma(x){+}\sigma(y)] \,\triangle\,
          [\sigma(z), \sigma(x)] \right| \\
-P^y_{(\square, \square)}(\state) &= P_{(x,\square)}(\state)
-  = P_{(\square,x)}(\state) = 0 \\
-P^y_{(u,v)}(\state) &= |[\sigma(u), \sigma(v)]|,
+P^y_{\square \square}(\state) &= P_{x \square}(\state)
+  = P_{\square x}(\state) = 0 \\
+P^y_{uv}(\state) &= |[\sigma(u), \sigma(v)]| \mbox{ otherwise},
 \end{align*}
@@
 of $A$ and $B$.
-The quantity $P^y_{(\square,z)}(\state)$ is the size change of
+The quantity $P^y_{\square z}(\state)$ is the size change of
 the interval $[x, z]$.  If $\state$ is the state before the
@@
 have
-$P_{(x,z)}(\state) = P^y_{(x,z)}(\state') + P^y_{(\square,z)}(\state')$
+$P_{xz}(\state) = P^y_{xz}(\state') + P^y_{\square z}(\state')$
 and
-$P_{(z,x)}(\state) = P^y_{(z,x)}(\state') - P^y_{(z,\square)}(\state')$.
+$P_{zx}(\state) = P^y_{zx}(\state') - P^y_{z \square}(\state')$.
 With these definitions, we have the following lemma.
@@
 = \sum_I q'_I \prod_{(u,v)} P_{uv}^{uv} \\
-&= \sum_I q_{I[\square\gets x]} P_I \prod_z
+&= \sum_I q_{I[\square\gets x]} X_I \prod_z
   \binom{\square z {+} xz}{\square z}P_{\square z}^{\square z} P_{xz}^{xz}
   \binom{z\! \square\! {+} zx}{z \square}(-P_{z \square})^{z \square} P_{zx}^{zx} \\
-&= \sum_J q_J P_J \sum_{I[\square\gets x] = J} \prod_z
+&= \sum_J q_J X_J \sum_{I[\square\gets x] = J} \prod_z
   \binom{\square z {+} xz}{\square z}P_{\square z}^{\square z} P_{xz}^{xz}
   \binom{z\! \square\! {+} zx}{z \square}(-P_{z \square})^{z \square} P_{zx}^{zx} \\
-&= \sum_J q_J P_J \prod_z
+&= \sum_J q_J X_J \prod_z
   \left( \sum_{\square z + xz = j_{xz}}
@@
   \left( \sum_{z\square {+} zx = j_{zx}} \dots \right) \\
-&= \sum_J q_J P_J \prod_z
+&= \sum_J q_J X_J \prod_z
   (P_{xz} + P_{\square z})^{j_{xz}}
@@
 the second case, both are non-zero.
+
+Given an index I from $\ind(V\cup \{ \square \})$, we define its projection
+$J = I|_{0y} \in \ind(V)$ over the interval $[0,y]$ as
+$$
+j_{0y} = i_{0y} + \sum_{z \in V} i_{\square z} + i_{z \square}, \quad
+j_{uv} = i_{uv} \mbox{ o.w.}
+$$
+Once again this is a degree preserving transformation.  We use it to
+relate indices containing $P_{0y}^k$ and its possible \emph{developments}
+in terms of $P_{\square z}$ and $P_{z \square}$.  A development gives a
+possible rewrite of indices that contain the base polynomial $
+|[0,y]|^k$ where k is non-zero.  Formally, it is a pair $\delta =
+(\delta_0, \delta_m)$ of a natural number $\delta_0$ and a map $\delta_m$
+from program variables to natural numbers such that $\delta_m(x) = 0$.
+A development $\delta$ has degree $\delta_0 + \sum_z \delta_m(z)$.  We
+will write $\Delta(J)$ for the set of all developments that have degree
+$j_{0y}$.  We will also use the shorter functional notation $\delta(z)$
+to denote $\delta_m(z)$.
+
+To see the action of a development on an index, consider the
+index $J = |[0,y]|^2$, the development $\delta = (2, \_ \mapsto 0)$
+is in $\Delta(J)$ and transforms $J$ into itself.  The development
+$\delta' = (1, \{ z \mapsto 1, \_ \mapsto 0\})$ is also into $\Delta(J)$ and
+transforms $J$ into the resource polynomial
+$|[0,y]|P_{\square z} + |[0,y]|P_{z \square}$.
+
+If $Q$ is a resource polynomial on the index set $\ind(V\cup \{\square\})$ we
+can now define the projection $Q' = Q|_{0y} \in \ind(V)$ as
+$$
+q'_J = \sum_{\delta \in \Delta(J)} q^\delta_J,
+$$
+where the numbers $q^\delta_J$ are constrained to satisfy the following
+family of equations
+$$
+q_I = q^\delta_{I|_{0y}} \prod_z \binom{i_{\square z} + i_{z \square}}{i_{\square z}}.
+$$
+
+\begin{lemma}
+If $Q \in \ind(V \cup \{\square\})$, then for all $\state$,
+$P^y_Q(\state) = P_{Q|_{0y}}(\state)$.
+\end{lemma}
+\begin{proof}
+\begin{align*}
+&P_{Q|_{0y}}(\state)
+  = \sum_J q_J P_J(\state)
+  = \sum_J X_J \sum_{\delta \in \Delta(J)} q_J^\delta P_{0y}^{0y} \\
+&= \sum_J X_J \sum_{\delta \in \Delta(J)}
+    q_J^\delta P_{0y}^{\delta_0}
+    \prod_z (P_{\square z} + P_{z \square})^{\delta(z)} \\
+&= \sum_J X_J \sum_{\delta \in \Delta(J)}
+    q_J^\delta P_{0y}^{\delta_0}
+    \prod_z \sum_{i+j=\delta(z)} \binom{\delta(z)}{i}
+      P_{\square z}^i P_{z \square}^j \\
+&= \sum_J \sum_{I|_{0y} = J}
+    q_J^{\delta_I} X_J P_{0y}^{i_{0y}}
+    \prod_z \binom{i_{\square z} + i_{z \square}}{i_{\square z}}
+      P_{\square z}^{i_{\square z}} P_{z \square}^{i_{z \square}} \\
+&= \sum_I
+    \left( q_{I|_{0y}}^{\delta_I}
+    \prod_z \binom{i_{\square z} + i_{z \square}}{i_{\square z}}
+    \right) P_I^y
+  = \sum_I q_I P_I^y = P^y_Q(\state).
+\end{align*}
+\end{proof}
 

### 2015-04-06 paper/main.tex
transition to final style

--- 2014-12-24 paper/main.tex
@@
 % Also look for "SHORT" comments
- \def\fullversion{}
-
-\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}
+% \def\fullversion{}
+
+\documentclass{sigplanconf}
 
@@
 %SHORT
-\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
+\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captions
 \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
@@
 \usepackage{listings} %for code
+\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
 \usepackage{mathpartir} %for typing rules
@@
 \input{definitions}
+\newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}
+
 \begin{document}
 
-\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{PLDI'15}{, June 13--17, 2015, Portland, OR, USA}
+\copyrightyear{2015}
+\copyrightdata{978-1-4503-3468-6/15/06}
+\doi{nnnnnnn.nnnnnnn}
+
+\titlebanner{\iffull{Extended Version}{}}%{Draft -- Do not distribute}        % These are ignored unless
+\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.
+
+\title{Compositional Certified Resource Bounds}
+% Certified Resource Bounds for C
+% Compositional Certified Resource Bounds for C
+% Compositional Quantitative Resource Analysis of C Programs
+% Compositional Resource-Bound Certification for C Programs
+
+\authorinfo
+{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
+{Yale University \vspace{-1cm}}
+
+\maketitle
 
@@
 
-\conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
-\copyrightyear{2014}
-\copyrightdata{XXX-X-XXXX-XXXX-X/XX/XX}
-
 %----------------------------------------------------
-
-\newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}
-
-\titlebanner{\iffull{Extended Version}{}}%{Draft -- Do not distribute}        % These are ignored unless
-\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.
-
-\title{Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}}
-% Certified Resource Bounds for C
-% Compositional Certified Resource Bounds for C
-% Compositional Quantitative Resource Analysis of C Programs
-% Compositional Resource-Bound Certification for C Programs
-
-\authorinfo
-{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
-{Yale University \vspace{-1cm}}
-
-\maketitle
 
@@
 is a simplified program state that maps variable names to integers,
-$|\inter a b| = \max(0,b-a)$, and $q_i \in \Q$.  To simplify the
+$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
 references to the linear coefficients $q_i$, we introduce an \emph{index
@@
 $\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
-is, a family of rational numbers where only
-a finite number of elements are not zero.  \iffull{As a notable difference from
-previous works based on amortized-analysis, we do not require all the coefficients
-of our potential function to be positive.  In practice, we will see using
-a semantic argument that despite this additional freedom, potential functions
-always remain positive.}{}
+is, a family of \ifshort{non-negative}{} rational numbers where only
+a finite number of elements are not zero. \iffull{ For soundness, we
+  require that potential functions always give non-negative results.
+  This is achieved by restricting coefficients in a quantitative
+  annotation to be always non-negative.}{}
 
@@
 { }
-{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
+{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
 %
@@
 { }
-{ (\Gamma, Q_B); R; (\Gamma, Q_B) \vdash \code{break} \dashv (\Gamma', Q') }
+{\!\!  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') \!\!}
 %
@@
 \Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
-{ n < 0 \implies Q \ge 0 }
-{ B; R; (\Gamma, Q) \vdash \code{tick}(n) \dashv (\Gamma, Q{-}n) }
+{ }
+{ \!\! B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) \!\!}
 %
@@
 \and \Rule{Q:Update}
-{ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
 }
@@
 %
-\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ Q \ge Q'
-\\ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q)
-}
+\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
 { B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
 %
+\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
+}
+%
 \and \Rule{Q:Inc}
-{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
-\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
 }
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:Dec}
-{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{xu}, -q_{ux})
-\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{ux}, -q_{xu})
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {-} y], Q) \vdash x \gets x - y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
 }
@@
 \and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
+}
+{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
 %
 \and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q')
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
+\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
 }
@@
 }
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
   \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
-}
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
+}
+%
+\and
+\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
 { }
-{ B; R; (\Gamma, Q) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
+%
+\and
+\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
 { \fenv f = (\vec y, S_f)
-\\ Q_f \ge 0
-\\ Q'_f \ge 0
 \\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
@@
 \and \Rule{Q:Weak}
-{  \!\! \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2
 \\ \Gamma'_2 \models \Gamma'_1
@@
 \\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\ \forall i.\, p_i, r_i \in \Qplusz
 \\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
@@
 \ifshort{
-                    \jan{XXX THIS FIGURE NEEDS TO BE UPDATED XXX}
+\begin{figure*}
+\small
+\vspace{-.4cm}
+\def \MathparLineskip {\lineskip=0.35cm}
+\begin{mathpar}
+\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+{ }
+{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
+%
+\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+%
+\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+}
+{ \textstyle
+  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+}
+%
+\\ \Rule{Q:Seq}
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+}
+{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+%
+\and \Rule{Q:Weak}
+{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ \Gamma'_2 \models \Gamma'_1
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
+}
+{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
+%
+\\ \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
+\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
+\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
+}
+{ Q' \succeq_\Gamma Q }
+\vspace{-.3cm}
+\end{mathpar}
+\caption{Selected inference rules of the quantitative analysis.}
+\label{fig:auto}
+\end{figure*}
 }{}
@@
 The derivation system for the automatic amortized analysis is defined
-in~\pref{fig:auto}.  The derivation rules yield judgements of the form
+in~\pref{fig:auto}.  The derivation rules derive judgements of the form
 $$
@@
 %
-The part $(\Gamma, Q) \! \vdash \! S \! \dashv \! (\Gamma', Q')$ of the judgement
+The part $(\Gamma, Q) \vdash \!\! S \!\! \dashv (\Gamma', Q')$ of the judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 quantitative part $Q$ represents a potential function as a collection
-of rational numbers $q_i$ indexed by the index set $I$.  The
+of non-negative numbers $q_i$ indexed by the index set $I$.  The
 logical part $\Gamma$ is left abstract but is enforced by our
@@
 with starting state $\state$, the assertions in $\Gamma$
-hold, $Q' {\ge} 0$, and at least $Q(\state)$ resources are available then the
+hold, and at least $Q(\state)$ resources are available then the
 evaluation does not run out of resources and, if the execution
-terminates in state $\state'$, there are at least $Q'(\state')$
-resources left and $\Gamma'$ holds for $\state'$.
+terminates in state $\state'$, there are at least $Q(\state')$
+\mbox{resources left and $\Gamma'$ holds for $\state'$.}
 
@@
 such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
-%
-% THE FOLLOWING IS NO LONGER TRUE
-% In all the rules, we have the implicit side condition that all rational
-% coefficients are non-negative.
-%
-Finally, if a rule mentions $Q$ and $Q'$
+In all the rules, we have the implicit side condition that all rational
+coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
 and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
@@
 a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that depend on $x$ after the assignment.
+like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
 
@@
 Therefore, the derivation rules are parametric in a function context $\Delta$
-that we leave implicit in the rules presented here.
-% More details can be found in the extended version.
+that we leave implicit in the rules presented here. More details
+can be found in the extended version.
 
@@
 rule.  This rule reuses its precondition as postcondition, it is
-explained by two observations: first, no resource is consumed by the skip
+explained by two facts: first, no resource is consumed by the skip
 operation, thus no potential has to be used to pay for the evaluation;
@@
 
-The rules {\sc Q:Inc} and {\sc Q:Dec} describe how the potential is
-distributed after a size change of a variable.
-The rule {\sc Q:Inc} is for incements $x \gets x+y$ and the symmetric
-rule {\sc Q:Dec} is for decrements $x \gets x-y$.
-The two rules are equivalent in the case where $y=0$.
-
-The two previous rules are essentially the same, so we only focus on
-the explanation of {\sc Q:Inc}. In this rule, the program updates a
-variable $x$ with $x+y$.  Since $x$ is changed, the
+The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
+describe how the potential is distributed after a size change of
+a variable.
+\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
+and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
+when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
+Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
+that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
+$x \gets x+y$, it applies when we can deduce from the logical context
+$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
+can be applied for decrements and when $y$ is not positive.}
+The rules are all equivalent in the case where $y=0$.
+The rule {\sc Q:Inc} can be applied
+if we cannot deduce any information about $y$.  In $\toolname$,
+the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
+
+\iffull{To explain how rules for increment and decrement work, it is
+sufficient to understand the rule {\sc Q:IncP}.  The others follow the
+same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
+variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
 quantitative annotation must be updated to reflect the change of the
@@
 Since $x$ is the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized.  Consider $S(x) = q_{xu}
-|\inter x u| + q_{ux} |\inter u x|$, during the increment, $S$ will
-increase by at most $y \cdot \max(q_{ux}, -q_{xu})$.
-This means that we must decrease the potential $q_{0y}$ of $|\inter 0 y|$
-by $\max(q_{ux}, -q_{xu})$ to pay for the change on
-$|\inter x u|$ and $|\inter u x|$.  A symmetric argument explains the
-change on $q_{y0}$.
+$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
+$\inter x u$ will get smaller with the update, and if $x' \in \inter x
+u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
+$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
+$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
+guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
+get bigger with the update.  We know that $|\inter v {x'}| \le y +
+|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
+$q_{vx}$ to pay for this change.  The rule ensures this
+only for $v \not\in \mathcal U$ because $x \le v$
+otherwise, and thus $|\inter v x| = 0$.
 
@@
 \inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
-{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=1cm,width=20cm]
-  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2.5cm,rightskip=2.5cm, vdots=1cm]
+{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
+  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1cm]
     { \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
       { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
-        { \inferrule*[right={\scriptsize (Q:Dec)},leftskip=0cm,rightskip=0cm]
+        { \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
           {
           }
-          { (\cdot, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
-            \dashv (x {\ge} 0, P^\text{de})
-          }
-        }
-        { (\cdot, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
+          { (x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
+            \dashv (\cdot, P^\text{de})
+          }
+        }
+        { (x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
           \dashv (\cdot, P^\text{we})
@@
         }
-        { (\cdot, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
+        { (x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
           \dashv (\cdot, P^\text{ti})
@@
       }
-      { (\cdot, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
+      { (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
         \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq})
@@
     }
-    { (\cdot, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
+    { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
       \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if})
@@
       { }
-      { (\cdot, B^\text{br}); (\cdot,Q^\text{br}) \vdash
+      { (x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash
         \code{break} \dashv (\bot,P^\text{br})
@@
     }
-    { (\cdot, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
+    { (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
       \code{break} \dashv (\cdot,P^\text{el})
@@
   }
-  { \hspace{3cm} (\cdot, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
+  { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
     \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm}
@@
 }
-{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (\cdot, P)
+{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P)
 }
@@
 & B^\text{we} {=} B^\text{de} \land Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}
-    \land P^\text{de} \succeq_{(x{\ge}0)} P^\text{we}
+    \land P^\text{de} \succeq_{(\cdot)} P^\text{we}
 & p^\text{de}_{0,10} {=} q^\text{de}_{0,10} + q^\text{de}_{0,x}
@@
 The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}. It is parameterized by a logical context
-used to gather information on interval sizes.  For example,
-$P^\text{de} \succeq_{(x{\ge}0)} P^\text{we} \equiv$
-% SSS SPACING
-%\vspace{-.3cm}
+\textsc{Relax}. It is parameterized by a logical context that is
+used to gather information on interval sizes.  For example, $P^\text{de} \succeq_{(\cdot)} P^\text{we} \equiv$
+\vspace{-.3cm}
 {\small
   \begin{align*}
-  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0,10}
-  \land p^\text{we}_{x,0} \leq p^\text{de}_{x,0} + u_{x,0} - v_{0,x} \land \;\;
-  \\
-  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}
-  \bigwedge_{(\alpha, \beta) \neq (0,10), (x,0)} p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
+ p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}
+\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
+\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
   \end{align*}
 }
-% SSS SPACING
-%\vspace{-.8cm}
+\vspace{-.8cm}
 
@@
 
-% SSS SPACING
-%\vspace{-.24cm}
+\vspace{-.24cm}
 
@@
 
-\iffull{}{The main parts of the soundness proof are formalized with
+\iffull{}{ The main parts of the soundness proof are formalized with
   Coq and available for download. The full definitions of the cost
@@
 the resource of interest for the user of our system.  We achieve this
-independence by using a tick instruction that allows the programmer
-to specify how much and when resource is consumed.  If the resource
-usage is bound to the programming language (e.g. stack usage, number
-loop iterations and function calls), the tick statements can
-be inserted automatically by a pre-processing phase.  Resources can be
-released by using a negative tick.
-Two sample rules of the semantics for update and tick follow.
-
+independence by parameterizing evaluations with a resource \emph{metric}
+$M$; a tuple of rational numbers and two maps.  Each of these
+parameters indicates the amount of resource consumed by a corresponding
+step in the semantics.  Resources can be released by using a negative
+cost. %  The metric used in the implementation can also
+% depend on other information that is statically available such as the
+% name of the called function or the number of arguments.
+Two sample rules for update and tick follow.
 \begin{mathpar}
-\Rule{U}
+\Rule[rightskip=0.3cm]{U}
 { \state' = \state[x \mapsto \sem{e}_\state] }
@@
 \rightarrow\\\\
-(\state', \code{skip}, \cont, \cost)
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
 }
@@
 \rightarrow\\\\
-(\state, \code{skip}, \cont, \cost {-} n)
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
 }
@@
 stuck configuration where $\cost<0$.
+
+% \pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
+% rules define a rewrite system for program configurations of the form
+% $(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
+% is the statement being executed, $\cont$ is a continuation that
+% describes what remains to be done after the execution of $S$, and
+% $\cost \in \Q$ is the non-negative number of resources available for
+% further execution.  All rules that can decrease $\cost$ have the
+% implicit side condition that the resource quantity available
+% \emph{before} the step is non-negative.  This means that we allow
+% $(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
+% transition relation $\smallstep{M}^*$ to indicate that the execution
+% ran out of resources.  However, every execution that reaches such a
+% state is stuck.
+%
+% \iffull{
+% A continuation $\cont$ represents the context of the execution of a
+% given statement.  A continuation can be the empty continuation
+% $\Kstop$ (used to start a program), a sequenced computation $\Kseq S
+% \cont$, a loop continuation $\Kloop S \cont$, or the continuation
+% $\Kcall r \theta \cont$ of a function call.
+% $$
+% \cont := \Kstop
+% \mid \Kseq S \cont
+% \mid \Kloop S \cont
+% \mid \Kcall r \theta \cont
+% $$
+% %
+% During the execution we assume a fixed function context $\fenv$ that
+% maps function names to a list of variables and the statement that
+% defines the function body.  It is used in the rule \textsc{S:Call}.
+% }{}
+%
+% The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
+% S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
+% following.  If the statement $S$ is executed in program state
+% $\state$, with continuation $\cont$, and with $\cost$ resources
+% available then---after a finite number of steps---the evaluation will
+% reach the new machine state $(\state', S', \cont',\cost')$ and there
+% are $\cost'$ resources available.  If $\cost' \ge 0$ then the
+% execution did not run out of resources and the resource consumption up
+% to this point is $\cost - \cost'$.  If this difference is negative
+% then resource became available during the execution.  If however
+% $\cost' < 0$ then the execution ran out of resources and is stuck.
+% The cost of the execution is then $\cost \ge 0$.
+%
+% \iffull{
+% \pref{lem:sem1} summarizes the main properties of the resource
+% counter of the semantics.  It can be proved by induction on the number
+% of steps.
+% \begin{lemma}
+% \label{lem:sem1}
+% \itemskip
+%   \begin{enumerate}
+%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+%     \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
+%     \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
+% \itemskipIn
+%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
+%     \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
+%     \cont', d')$ then $\cost - \cost' = d - d'$.
+%   \end{enumerate}
+% \itemskip
+% \end{lemma}
+% %
+% If we assume that we have a \emph{positive resource metric}, that is,
+% a metric that assigns a positive cost to each evaluation step then we
+% can link bounds on the resource consumption under the metric to
+% termination.  This is formalized by \pref{lem:sem_term}.
+% %
+% \begin{lemma}
+%  \label{lem:sem_term}
+%  Let $M$ be a positive metric and assume that there exists $b \in
+%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
+%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
+%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
+%  \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
+%  and $\state''$ implies $n \le B$.
+% \end{lemma}
+% }
 
@@
 as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
-maps from program states to an element of $\Q\cup\{\infty\}$ that represents
+maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
 an amount of resources available.  The assertions $B$ and $R$ are
@@
 S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
-the empty continuation, $Q'(\state) \ge 0$ for all state $\sigma$, and at least $Q(\state)$ resource units available
+the empty continuation, and at least $Q(\state)$ resource units available
 then the evaluation does not run out of resources and there are at
-least $Q'(\state')$ resources left if the evaluation terminates in state
+least $Q(\state')$ resources left if the evaluation terminates in
 $\state'$.  The logic rules are similar to the ones in previous
-work.
+work and generalized to account for the cost introduced by  our
+cost-aware semantics.
 
@@
 translate the assumptions in function contexts of the automatic
-analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=$
-%
-\begin{multline*}
-\forall z \, \vec v \, v . \,
+analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
+\forall z \, \vec v \, v . $
+%
+$$
 (
-  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]), \\
+  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
   \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
   )
-\end{multline*}}{}%
+$$}{}
 %
@@
 $$}%
-The soundness of the automatic analysis can now be stated
+\noindent{}The soundness of the automatic analysis can now be stated
 formally with the following theorem.
@@
 variables, and can abstractly specify the resource cost of functions
-(\pref{sec:AAA}). Second, it is easily combined with established
+(\pref{sec:AAA}). Second, it is easily combinable with established
 qualitative verification to guide semi-automatic bound derivation
@@
 
-We describe our system for a subset of Clight that is sufficient to
-discuss the general ideas.  This subset is given by the following
-grammar.
+In this article, we describe our system for a subset of Clight that is
+sufficient to discuss the general ideas.  This subset is given by the
+following grammar.
 %
@@
 The most notable difference to full Clight is that we can only assign
-to variables and thus do not consider operations that update the heap
-or the stack through a pointer.
+to variables and thus do not consider operations that update the heap.
 Moreover, function arguments and return values are assumed to be
@@
 pointers, \code{goto} statements, \code{continue} statements, and
-stack variables that have their address taken.
+\code{switch} statements.
 
@@
 program for the automatic analysis.  We also add the built-in function
-\code{tick(n)} that can be called with a constant rational number $n$
-as a flexible way to model resource consumption or release (if $n$ is
+\code{tick(n)} that can be called with a constant integer $n$ as a
+flexible way to model resource consumption or release (if $n$ is
 negative).
@@
 { (\state, \code{assert}~e, \cont, \cost)
-\smallstep
-(\state, \code{skip}, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_a)
 }
@@
 \Rule{S:BrkSeq}
-{ }
+{}
 { (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{break}, \cont, \cost)
@@
 \Rule{S:BrkLoop}
-{ }
+{}
 { (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep
-(\state, \code{skip}, \cont, \cost)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_b)
 }
@@
 \Rule{S:RetSeq}
-{ }
+{}
 { (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{return}~x, \cont, \cost)
@@
 \Rule{S:RetLoop}
-{ }
+{}
 { (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, \code{return}~x, \cont, \cost)
@@
 { (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep
-(\state', \code{skip}, \cont, \cost)
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_r)
 }
@@
 { (\state, x \gets e, \cont, \cost)
-\smallstep
-(\state', \code{skip}, \cont, \cost))
+\smallstep{M}
+(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
 }
@@
 { (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep
-(\state', S_f, \Kcall r \theta \cont, \cost)
+\smallstep{M}
+(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
 }
@@
 \Rule{S:Loop}
-{ }
+{}
 { (\state, \code{loop}~S, \cont, \cost)
-\smallstep
+\smallstep{M}
 (\state, S, \Kloop S \cont, \cost)
@@
 \Rule{S:SkipLoop}
-{ }
+{}
 { (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep
-(\state, \code{loop}~S, \cont, \cost)
+\smallstep{M}
+(\state, \code{loop}~S, \cont, \cost {-} M_l)
 }
@@
 { \code{istrue}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
 }
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep
-(\state, S_1, \cont, \cost)
+\smallstep{M}
+(\state, S_1, \cont, \cost')
 }
@@
 { \code{isfalse}~\sem{e}_\sigma
+\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
 }
 { (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep
-(\state, S_2, \cont, \cost)
+\smallstep{M}
+(\state, S_2, \cont, \cost')
 }
@@
 \Rule{S:SkipSeq}
-{ }
+{}
 { (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep
-(\state, S, \cont, \cost)
+\smallstep{M}
+(\state, S, \cont, \cost {-} M_s)
 }
@@
 { (\state, \code{tick}(n), \cont, \cost)
-\smallstep
-(\state, \code{skip}, \cont, \cost - n)
+\smallstep{M}
+(\state, \code{skip}, \cont, \cost {-} M_t(n))
 }
@@
 
-\paragraph{Semantics.}
-
-\pref{fig:opsem} contains the reduction rules of the semantics.  The rules
-define a rewrite system for program configurations of the form $(\state,
-S, \cont, \cost)$, where $\state$ is the program state, $S$ is the
-statement being executed, $\cont$ is a continuation that describes what
-remains to be done after the execution of $S$, and $\cost \in \Q$ is
-the amount of resources available for further execution.  All the rules
-have the implicit side condition that the resource quantity available
-\emph{before} the step is non-negative.  This means that we allow
-$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side transition
-relation $\smallstep^*$ to indicate that the execution ran out of
-resources.  However, every execution that reaches such a state is stuck.
-
-A continuation $\cont$ represents the context of the execution of a
-given statement.  A continuation can be the empty continuation
-$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
-\cont$, a loop continuation $\Kloop S \cont$, or the continuation
-$\Kcall r \theta \cont$ of a function call.
-$$
-\cont := \Kstop
-\mid \Kseq S \cont
-\mid \Kloop S \cont
-\mid \Kcall r \theta \cont
-$$
-%
-During the execution we assume a fixed function context $\fenv$ that
-maps function names to a list of variables and the statement that
-defines the function body.  It is used in the rule \textsc{S:Call}.
-
-The intuitive meaning of an evaluation $(\state,
-S, \cont, \cost) \smallstep^* (\state', S', \cont', \cost')$ is the
-following.  If the statement $S$ is executed in program state
-$\state$, with continuation $\cont$, and with $\cost$ resources
-available then---after a finite number of steps---the evaluation will
-reach the new machine state $(\state', S', \cont',\cost')$ and there
-are $\cost'$ resources available.  If $\cost' \ge 0$ then the
-execution did not run out of resources and the resource consumption up
-to this point is $\cost - \cost'$.  If this difference is negative
-then resource became available during the execution.  If however
-$\cost' < 0$ then the execution ran out of resources and is stuck.
-The cost of the execution is then $\cost \ge 0$.
-
-\pref{lem:sem1} summarizes the main properties of the resource
-counter of the semantics.  It can be proved by induction on the number
-of steps.
-\begin{lemma}
-\label{lem:sem1}
-\itemskip
-  \begin{enumerate}
-  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
-    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
-    \smallstep^k (\state', S', \cont', \cost' {+} n)$.
-\itemskipIn
-  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
-    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep^k (\state', S',
-    \cont', d')$ then $\cost - \cost' = d - d'$.
-  \end{enumerate}
-\itemskip
-\end{lemma}
-
-% OOPS this is no longer true... This is an argument agains a presentation
-% with ticks only. XXX
-%
-% If we assume that we have a \emph{positive resource metric}, that is,
-% a metric that assigns a positive cost to each evaluation step then we
-% can link bounds on the resource consumption under the metric to
-% termination.  This is formalized by \pref{lem:sem_term}.
-%
-% \begin{lemma}
-%  \label{lem:sem_term}
-%  Let $M$ be a positive metric and assume that there exists $b \in
-%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep^*
-%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
-%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
-%  \smallstep^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
-%  and $\state''$ implies $n \le B$.
-% \end{lemma}
-
 \section{Quantitative Hoare Logic}
@@
 { }
-{ \Delta; B; R \ldash
+{ \Delta; B; R \vdash
 \htriple
@@
 { }
-{ \Delta; B; R \ldash
+{ \Delta; B; R \vdash
 \htriple
-  { B }{ \code{break} }{ Q }
-}
-\and
-\Rule{L:Tick}
-{ n < 0 \implies Q \ge 0}
-{ \Delta; B; R \ldash
-\htriple
-  { Q }
-  { \code{tick}(n) }
-  { Q - n }
-}
-%
+  { M_b + B }{ \code{break} }{ Q }
+}
 %
@@
 \and
-\Rule{L:Assert}
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
 { }
-{ \Delta; B; R \ldash
+{ \Delta; B; R \vdash
 \htriple
-  { \code{istrue}~\sem{e}_\state \implies Q }
-  { \code{assert}~e }
-  { Q }
+  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
 }
@@
 \and
-\Rule{L:Update}
-{ }
-{ \Delta; B; R \ldash
-\htriple
-  { \lambda \state.\, Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ I \ge Q
-\\ \Delta; Q; R \ldash
-   \htriple { I }{ S }{ I }
-}
-{ \Delta; B; R \ldash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:Seq}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S_1 }{ Q' }
-\\\\
-  \Delta; B; R \ldash
+\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S_1 }{ Q' + M_s }
+\\
+\!\!  \Delta; B; R \vdash
 \htriple{ Q' }{ S_2 }{ Q }
 }
-{ \Delta; B; R \ldash
+{ \Delta; B; R \vdash
 \htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{% P \ge 0
-%\\
-  \Delta; B; R \ldash
-  \htriple
-  { \code{istrue}~\sem{e}_\state + P }
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \ldash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \ldash
-\htriple
-  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
 }
@@
   \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \ge P_f \, y \, (\state(\vec x)) + A \\
-  \forall v.\, (Q_f \, y \, v + A \ge \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \ldash
+  P \models P_f \, y \, (\state(\vec x)) \land A \\
+  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \vdash
 \htriple
-  { P }
+  { M_f + P }
   { r \gets f(\vec x) }
+  { Q - M_r }
+}
+%
+\and
+\Rule{L:Assert}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
+  { \code{assert}~e }
   { Q }
+}
+%
+\and
+\Rule{L:Tick}
+{ }
+{ \Delta; B; R \vdash
+\htriple
+  { Q + M_t(n) }
+  { \code{tick}(n) }
+  { Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ \Delta; Q; R \vdash
+\htriple
+  { I }{ S }{ I + M_l }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{ \Delta; B; R \vdash
+\htriple
+  { \code{istrue}~\sem{e}_\state + P - M_c^1}
+  { S_1 }{ Q }
+\\
+  \Delta; B; R \vdash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \vdash
+\htriple
+  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
+}
+%
+\and
+\Rule{L:Weaken}
+{
+  \!\! P \models P' \!\! \\
+  \!\! \Delta; B'; R' \vdash
+  \htriple{ P' }{ S }{ Q' } \!\! \\
+  \!\! Q' \models Q \!\! \\
+  \!\! B' \models B \!\! \\
+  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
+}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \vdash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \vdash
+\htriple{ P + x }{ S }{ Q + x } \!\!
 }
@@
 \Rule{L:Extend}
-{  \Delta \cup \Delta'; B; R \ldash
+{ \Delta \cup \Delta'; B; R \vdash
   \htriple{ P }{ S }{ Q }
 \\
-  P_f \ge 0
-\\\\
   \forall f\,P_f\,Q_f.\,
@@
   \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \ldash
+  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
     \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
 }
-{ \Delta; B; R \ldash
+{ \Delta; B; R \vdash
 \htriple{ P }{ S }{ Q }
@@
 %
-\and
-\Rule{L:Weaken}
-{
-  \Delta; B'; R' \ldash \htriple{ P' }{ S }{ Q' } \\\\
-  P \ge P' \!\! \\
-  \!\! Q' \ge Q \!\! \\
-  \!\! B' \ge B \!\! \\
-  \!\! \forall v.\, (R'\,v \ge R\,v)
-}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \ldash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
 \end{mathpar}
+
 \caption{Rules of the Quantitative Hoare Logic}
@@
 $$
-P : \State \to \Q \cup \{ \infty \} \; .
+P : \State \to \Qplusz \cup \{ \infty \} \; .
 $$
@@
 $\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
-We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
+We write $\Assn$ for $\State \to \Qplusz \cup \{ \infty \}$ and $\bot$ for
 $\lambda \state . \, \infty$.  We sometimes call assertions
 \emph{potential functions}.  To use Coq's support for propositional
-reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
+reasoning, assertions have the type $\State \to \Qplusz \to \mathrm{Prop}$
 in the implementation.  For a given $\state \in \State$, such an
-assertion can be seen as a set $B \subseteq \Q$.  However, we find
-the presentation in this article easier to read.
+assertion can be seen as a set $B \subseteq \Qplusz$ of valid bounds.
+However, we find the presentation in this article easier to read.
 
@@
 $$
-\Delta; B; R \ldash
+\Delta; B; R \vdash
 \htriple
@@
 Hoare logic and the intuitive meaning is as follows.  If $S$ is
-executed with starting state $\state$, with the empty continuation
-$\Kstop$, $Q' \ge 0$, and at least $Q(\state)$ resources available then the
+executed with starting state $\state$, the empty continuation
+$\Kstop$, and at least $P(\state)$ resources available then the
 evaluation does not run out of resources and there are at least
-$Q'(\state')$ resources left if the evaluation terminates in state
+$Q(\state')$ resources left if the evaluation terminates in state
 $\state'$.  The assertion $B : \Assn$ provides the
@@
 
-For two assertions $P,Q : \Assn$, we write $P \ge Q$ to if for all
-program state $\state$, $P(\state) \ge Q(\state)$.
+For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
+program state $\state$ $P(\state) \ge Q(\state)$.
 
@@
 
-The {\sc L:Tick} rule accounts for the cost $n$ of the tick statement.
-This rule is the only one to account directly for a cost on in the
-assertions because calling the \code{tick} function is the only way to
-change the amount of resource available.
-The rule also has a side condition that ensures that the potential
-remains positive during all the program execution.  It is an
-essential semantic property of this system.
-
 In the rule {\sc L:Assert}, we use the notation
-$\code{istrue}~\sem{e}_\state \implies Q$ to express that we
-require potential $Q$ in the precondition if $e$ evaluates to
+$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
+require potential $Q + M_a$ in the precondition if $e$ evaluates to
 $\mathit{true}$ in the current program state.  If $e$ evaluates to
-$\mathit{false}$ then the potential in the precondition can be any
-non-negative number since the program will be terminated.
+$\mathit{false}$ then the potential in the precondition can be
+arbitrary since the program will be terminated.
 
@@
 \code{break}, respectively.  In {\sc L:Break}, we require to have
-potential $B$ in the precondition to pay for the potential after the
+potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
+cost of \code{break} and $B$ to pay for the potential after the
 \code{break}.  In {\sc L:Return} we only require to have potential
 $R$ in the precondition to pay for the potential after the
-\code{return}.
+\code{return}.  The reason is that we found it to be more convenient
+to account for the execution cost $M_r$ of the return in rule {\sc
+  L:Call} for function calls.
 
@@
 the precondition we ensure that $Q$ evaluates to the same number as in
-the postcondition.
+the postcondition.  We also require that have we the constant
+potential $M_u + M_e(e)$ available in the precondition to pay for the
+cost of the evaluation of $e$ and the update.
 
@@
 $S_2$.  The interaction between the actual pre- and postconditions is
-analogous to standard Hoare logic.
+analogous to standard Hoare logic.  In the postcondition of $S_1$ we
+account for the cost $M_s$ of the execution of the sequence.
 
@@
 require to have potential $I$ available in the precondition of the
-loop body $S$.
+loop body $S$.  In the postcondition of $S$, the potential must be
+sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
+iteration.
 
@@
 $\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
-quantitative assertions.
-
-The pre- and postcondition $P_f$ and $Q_f$ used in the {\sc L:Call} rule
-are taken from the function context $\Delta$.  The assertions in
+quantitative assertions.  In the precondition of the rule {\sc L:Tick}
+we account for the execution cost $M_t(n)$ of the \code{tick}
+statement that depends on the integer $n$.
+
+The rule {\sc L:Call} accounts for the execution cost of both $M_f$
+for function calls and $M_r$ for \code{return} statements.
+This justifies that the {\sc L:Return} rule does not account
+  for resource consumption. The pre- and postcondition $P_f$ and
+$Q_f$ are taken from the function context $\Delta$.  The assertions in
 the context are parametric with respect to both the values of the
@@
 except that instead of using the logical implication $\Rightarrow$ we
-use the quantitative $\ge$.  This rule indeed weakens the statement
-since it requires more resource to run the statement and yields less
-than what has been proved to be available after its termination.
+use the quantitative $\models$ that point-wise applies $\ge$.  This
+rule indeed weakens the statement since it requires more resource to
+run the statement and yields less than what has been proved to be
+available after its termination.
 
@@
 logical variables almost for free: we can simply use the variable and
-binding mechanisms of Coq, our host theory.  When a logical variable
+binding mechanisms of Coq, our host theory.  When an logical variable
 is needed we introduce it using a universal quantifier of Coq before
@@
 Theorem dbl_triple: forall Z,
-  qtriple ($\lambda\sigma$.$\sigma$(x)=Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x)=2*Z).
+  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
 \end{lstlisting}
@@
 In the example in~\pref{fig:xmplmax} we derive a precise runtime
-bound on a program that searches a maximal element in an array.
-Because of the location of the tick statements, the resource cost
-is closely related to the number of times the test \code{a[i] > m}
-is true during the execution.
+bound on a program that searches a maximal element in an array.  The
+cost metric that we use simply counts the assignments performed by the
+program.  Hence, the resource cost is closely related to the
+number of times the test \code{a[i] > m} is true during the
+execution.
 %
@@
 \code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
-potential unit to pay for the tick in the conditional.  In the other case, no
+potential unit to pay for the assignment.  In the other case, no
 maximum so far is encountered so $A(i) = A(i+1)$.
-As a side remark, if $K \ge 0$, it is possible to
-have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
-of the same program.  It can be done either by adapting the proof or by
-applying the {\sc L:Frame} rule one top of the triple already derived.
-More generally, without function calls, the {\sc L:Frame}
-rule is admissible in our system.
+% As a side remark, if $K \ge 0$, it is possible to
+% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
+% of the same program.  It can be done either by adapting the proof or by
+% applying the {\sc L:Frame} rule one top of the triple already derived.
+% More generally, without function calls, the {\sc L:Frame}
+% rule is admissible in our system.
 
 The example in~\pref{fig:xmplbs} shows a use case for logical
-variables. On this example, we model the stack consumption:
-we account a constant cost for a function call ($M_f>0$) that
-is returned after the call ($M_r = -M_f <0$).
-We are interested in showing that a binary search function
+variables as well as a metric for stack consumption. In a stack
+metric, we account a constant cost for a function call ($M_f>0$) that
+is returned after the call ($M_r = -M_f <0$).  All other resource cost
+are $0$.  We are interested in showing that a binary search function
 $\code{bsearch}$ has logarithmic stack consumption.  We use a logical
@@
 Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
-- 1$.  The rest of the proof is here to make sure that the first tick
-can be paid for using regular algebra on the $\log_2$ function.
+- 1$, and because $M_f = M_\code{bsearch}$ and $M_r =
+-M_\code{bsearch}$ in the stack metric.  The rest of the proof does
+not involve any resource manipulation and is just bookkeeping of
+logical assertions.
 
@@
 ${\color{blue} \{ A(0) + N \} }$
-i=1; tick(1); m=a[0]; tick(1);
+i=1; m=a[0];
+${\color{blue}%
+  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
 while (i < N) {
@@
   if (a[i] > m)
-    m=a[i], tick(1);
+    m=a[i];
   ${\color{blue}%
@@
   ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) + 1 \} }$
-  tick(1);
-  ${\color{blue}%
     \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
@@
   for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
-  we use ticks in this example to count the number of assignments performed
-  by the program.
+  the metric used here assigns a cost of 1 to every assignment
+  and 0 to all other operations.
   }
@@
 ${\color{blue}%
-  P(Z) := \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
+  \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
 bsearch(x,l,h) {
@@
       \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
-    tick($M_\code{bsearch}$);
-    ${\color{blue}\{ P(Z-1) \}}$ l = bsearch(x,l,h); ${\color{blue}\{ Q(Z-1) \}}$
-    tick($-M_\code{bsearch}$);
-  } ${\color{blue}%
+    l = bsearch(x,l,h); }
+  ${\color{blue}%
     \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
   return l;
-} ${\color{blue} Q(Z) := \{ Z \cdot M_\code{bsearch} \} }$
+} ${\color{blue} \{ Z \cdot M_\code{bsearch} \} }$
 \end{lstlisting}
 \caption{Example derivation of a stack usage bound for a binary
-  search program.  The stack usage is modeled by adding a cost $M_\code{bsearch}$ before
+  search program.  The used resource metric defines the cost $M_\code{bsearch}$ before
   the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
-  frame size of the function $\code{bsearch}$.  $Z$ is a logical variable.
+  frame size of the function $\code{bsearch}$.  $Z$ is a
+  logical variable.
   }
@@
 We already gave an intuition of the meaning of judgements
-derived in the logic.  We start by defining an indexed
-predicate $\cdot \downarrow_n$ that defines a \emph{resource-safe}
-program configuration:
-\begin{multline*}
-  (\state, S, K, c) \downarrow_n \, := \\
-  \forall m \le n.\, \forall c'.\,
-    (\state, S, K, c) \smallstep^m
-    (\_, \_, \_, c') \implies 0 \le c'.
-\end{multline*}
-This predicate means that a program configuration can execute
-for a certain number of steps without running out of resources.
-Note that nothing is said about the program safety, i.e. a stuck
-configuration $C$ will satisfy $\forall n.\, C\downarrow_n$. This
-is because our logic does not prove any safety or correctness
-theorems but only focuses on resource usage.
-We can now be used to define the resource safety
-of a configuration with respect to a quantitative assertion.
-\begin{multline*}
-  \safe n P S \cont \, := \\
-  \forall \state\, c.\, P(\state) \le c \implies
-    0 \le c \wedge (\state, S, K, c) \downarrow_n.
-\end{multline*}
-Once again, we use an index in the definition of the predicate, it
-is used to perform proofs by induction for the function-call and
-loop rules of the logic.
-%
-An interesting and essential property of the previous predicate is
-that $\safe {n+1} P S \cont \implies \safe n P S \cont$.  To prove
-this weakening property, we use the quantification over all numbers
-smaller than $n$ in the definition of $\cdot \downarrow_n$.
+derived in the logic.  To make it formal, we define the
+\emph{resource safety} $\safe n P S \cont$ of an assertion
+and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
+\begin{align*}
+  (m {\le} n \land P(\state) {\le} \cost \land
+    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
+  \implies \cost' {\ge} 0.
+\end{align*}
+This predicate is step indexed by an integer $n$ that is used for
+induction in the soundness proof for the function-call and loop cases.
+The constraint $\cost' \geq 0$ imposed by the definition ensures that
+the program does not stop because of a resource error (recall that a
+negative resource counter on the right-hand side is a resource
+failure).  However, it does not rule out memory safety errors or
+assertion failures. This is because our logic does not prove any
+safety or correctness theorems but only focuses on resource usage.
+%
+An interesting detail in the definition is the natural number $m$.  We
+simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
+\cont $.  This would not be the case if we replaced all occurrences of
+$m$ by $n$ in the definition of validity.
 
@@
 \safeK n B R Q \cont & :=  \safe n B {\code{break}} \cont \\
- \land& \forall x. \safe n {\lambda \state.R\,(\state(x))} {\code{return}~x} \cont \\
+ \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
  \land& \safe n Q {\code{skip}} \cont
@@
 We can now define the \emph{semantic validity} of a judgement $B; R
-\ldash \htriple P S Q$ of the quantitative logic without function
+\vdash \htriple P S Q$ of the quantitative logic without function
 context as $\valid n B R P S Q :=$
-\begin{multline*}
-\forall m \le n.\, \forall x \geq 0 .\, \forall \cont . \\
-  \safeK m {B{+}x} {R{+}x} {Q{+}x} \cont \\
-  \implies \safe m {P+x} S \cont .
-\end{multline*}
+\begin{align*}
+& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
+& \implies \safe m {P+x} S \cont \; .
+\end{align*}
 Note how the validity of a triple embeds the frame rule of
@@
   &\forall f .\, \Delta(f) =
-    \forall z \, \vec v \, v .\, (P_f \, z \, \vec v, Q_f \, z \, v)
+    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
     \implies \\
@@
 \begin{theorem}[Soundness]
-  \label{thm:sound}
-  If $\Delta; B; R \ldash \htriple P S Q$ is derivable then
+  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
   $
     \forall n .\, \validC n \Delta
-      {\implies} \valid {n {+} 1} B R P S Q.
+      \implies \valid {n {+} 1} B R P S Q.
   $
@@
 expanding the definitions we see that $\Delta$ is valid
-for every $n$ and that $\Kstop$ is safe for every $n$ if both
-$Q$, $R$, and $B$ are non-negative on all possible program
-states. So the \pref{thm:sound} becomes
+for every $n$ and that $\Kstop$ is safe for every $n$. So
+we derive
 $$
-\Delta; B; R \ldash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
+\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
 $$
@@
 }
-void start (int x,int y) {
+void main (int x,int y) {
   y = copy (x,y);
@@
 
-void start (int y, int z) {
+void main (int y, int z) {
   count_down(y,z);
@@
 }
-void start (int y, int z) {
+void main (int y, int z) {
   consume(); produce(); consume();
@@
   while (i < n) {
-    while (j >= 0
-        && t[i]!=p[j]) {
+    while (j >= 0 && t[i]!=p[j]) {
       k = b[j];
@@
 int gcd(int x, int y) {
-  if (x <= 0)
-    return y;
-  if (y <= 0)
-    return x;
+  if (x <= 0) return y;
+  if (y <= 0) return x;
 
@@
    \begin{lstlisting}
-void qsort(
-  int a[], int lo, int hi
-) {
+void qsort(int a[], int lo, int hi) {
   int m1, m2, n;
@@
 
-  n = nondet(); /* partition */
+  n = nondet(); /* partition the array */
   assert( n > 0 );
@@
 
-void start(int a[], int len) {
+void main(int a[], int len) {
   qsort(a, 0, len);

### 2015-04-07 paper/main.tex
syntax changes

@@
 \itemskip
-%
-\iffull{}
-{\footnotesize \sl
-[To the reviewers: The auxiliary material contains an anonymized
-version of the extended article, and the source code of $\toolname$
-and all examples.] }
 
@@
 ${\color{blue}%
-  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
 while (x+K<=y) {
   ${\color{blue}%
-  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
   x=x+K;
   ${\color{blue}%
-  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
+  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
   tick(T);
   ${\color{blue}%
-  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
+  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
 }
 ${\color{blue}%
- \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
+ \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
 \end{lstlisting}
@@
   ${\color{blue}%
-  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+  \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
   if (m>y)
     ${\color{blue}%
-    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
+    \{ m{>}y; \, |[x, n]| {+} |[y, m]| \} }$
     y=y+1;
     ${\color{blue}%
-    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+    \{ \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
   else
     ${\color{blue}%
-    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
+    \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
     x=x+1;
     ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
+    \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
   ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
+  \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
   tick(1);
 } ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
+  \{ \cdot; \, |[x, n]| {+} |[y, m]| \} }$
    \end{lstlisting}
@@
   ${\color{blue}%
-  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+  \{ x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
   if (z>x)
     ${\color{blue}%
-    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    \{ x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
     x=x+1;
     ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+    \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
   else
     ${\color{blue}%
-    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
+    \{ z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
     z=z+1;
     ${\color{blue}%
-    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
+    \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
   ${\color{blue}%
-  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
+  \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
   tick(1);
 } ${\color{blue}%
-  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
+  \{ \cdot; \, |[x, n]| {+} |[z, n]| \} }$
    \end{lstlisting}
@@
   ${\color{blue}%
-  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  \{ y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
   y=y+1;
   ${\color{blue}%
-  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  \{ \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
   tick(3);
   ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+  \{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
 }
 ${\color{blue}%
-\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+\{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
 while (y>9) {
   ${\color{blue}%
-  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  \{ y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
   y=y-10;
   ${\color{blue}%
-  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
+  \{ \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
   tick(1);
 } ${\color{blue}%
-  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
+  \{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
    \end{lstlisting}
@@
   ${\color{blue}%
-  n{<}0; \,  P(n,y) \vdash }$
+  \{ n{<}0; \,  P(n,y) \} }$
   n=n+1;
   ${\color{blue}%
-  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
+  \{ \cdot; \,  59 {+} P(n,y) \} }$
   y=y+1000;
   ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  \{ \cdot; \, 9 {+} P(n,y) \} }$
   while (y>=100 && *){
     ${\color{blue}%
-    y{>}99; \, 9 {+} P(n,y) \vdash }$
+    \{ y{>}99; \, 9 {+} P(n,y) \} }$
     y=y-100;
     ${\color{blue}%
-    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
+    \{ \cdot; \, 14 {+} P(n,y) \} }$
     tick(5);
   } ${\color{blue}%
-  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
+  \{ \cdot; \, 9 {+} P(n,y) \} }$
   tick(9);
 } ${\color{blue}%
-  \dashv \cdot; \, P(n,y) }$
+  \{ \cdot; \, P(n,y) \} }$
    \end{lstlisting}
@@
 { }
-{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
+{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
 %
 \and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
-{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
+{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
 %
@@
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
+  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
 }
@@
 { \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
 }
@@
 \\ \Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
-}
-{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
+\\\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
 %
 \and \Rule{Q:Weak}
-{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
 \\ \!\! \Gamma_1 \models \Gamma_2
@@
 }
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
+{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
 %
@@
 $$
-  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
-    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
+  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
+    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
 $$
 %
-The part $(\Gamma, Q) \vdash \!\! S \!\! \dashv (\Gamma', Q')$ of the judgement
+The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 This is similar to classical Hoare triples in the presence of non-linear
-control flow. In the judgement, $(\Gamma_B, Q_B)$ is the postcondition
+control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
 that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
@@
 % \code{break} statements by the rule {\sc Q:Break}.
-Similarly, $(\Gamma_R, Q_R)$ is the postcondition that holds when
+Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
 returning from a function call with \code{return}. \iffull{(This becomes clear
@@
 % logical part $\Gamma$ and a quantitative part $Q$.  That means that
-% $(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
-% $(\Gamma_R, Q_R)$ is the postcondition for return statements, and
-% $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
+% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
+% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
+% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
 % simple Hoare triple.  We leave out the function context $\Delta$ in
@@
 During the analysis, function specifications are quadruples
-$(\Gamma_f, Q_f, \Gamma'_f, Q'_f)$ where $\Gamma_f, Q_f$ depend on
-$\Vargs$, and $\Gamma'_f, Q'_f$ depend on $\Vret$.  These parameters
+$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
+$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
 are instantiated by appropriate variables on call sites.  A distinctive
@@
 
-\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}
+\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x \geq 10)~(x=x - 10; tick(5))~else~break}}}
 \begin{figure*}[t]
@@
           }
-          { (x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
-            \dashv (\cdot, P^\text{de})
-          }
-        }
-        { (x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
-          \dashv (\cdot, P^\text{we})
+          { (x {<} 10; B^\text{de}) \vdash
+	    \htriple{ x {\geq} 10; Q^\text{de} }{ \code{x = x - 10} }{ \cdot; P^\text{de} }
+          }
+        }
+        { (x {<} 10; B^\text{we}) \vdash
+	  \htriple{ x {\geq} 10; Q^\text{we} }{ \code{x = x - 10} }{ \cdot; P^\text{we} }
         }
@@
         }
-        { (x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
-          \dashv (\cdot, P^\text{ti})
-        }
-      }
-      { (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
-        \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq})
-      }
-    }
-    { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
-      \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if})
+        { (x {<} 10; B^\text{ti}) \vdash
+	  \htriple{ \cdot; Q^\text{ti} }{ \code{tick(5)} }{ \cdot; P^\text{ti} }
+        }
+      }
+      { (x {<} 10; B^\text{sq}) \vdash
+	\htriple{ x {\geq} 10; Q^\text{sq} }{ \code{x=x - 10; tick(5)} }{ \cdot; P^\text{sq} }
+      }
+    }
+    { (x {<} 10; B^\text{if}) \vdash
+      \htriple{ x {\geq} 10; Q^\text{if} }{ \code{x=x - 10; tick(5)} }{ \cdot; P^\text{if} }
     }
@@
       { }
-      { (x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash
-        \code{break} \dashv (\bot,P^\text{br})
-      }
-    }
-    { (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
-      \code{break} \dashv (\cdot,P^\text{el})
-    }
-  }
-  { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
-    \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm}
-  }
-}
-{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P)
+      { (x {<} 10; B^\text{br}) \vdash
+	\htriple{ x {<} 10; Q^\text{br} }{ \code{break} }{ \bot; P^\text{br} }
+      }
+    }
+    { (x {<} 10; B^\text{el}) \vdash
+      \htriple{ x {<} 10; Q^\text{el} }{ \code{break} }{ \cdot; P^\text{el} }
+    }
+  }
+  { \hspace{3cm} (x {<} 10; B^\text{lo}) \vdash
+    \htriple{ \cdot; Q^\text{lo} }{ \loopbody }{ \cdot; P^\text{lo} } \hspace{3cm}
+  }
+}
+{ (\cdot; B) \vdash
+  \htriple{ \cdot; Q) }{ \code{loop}~{\loopbody} }{ x {<} 10; P }
 }

### 2015-04-08 paper/main.tex
drop a few space tricks
reformat some maths

--- 2015-01-06 paper/main.tex
@@
 
-\ifshort{
-                    \jan{XXX THIS FIGURE NEEDS TO BE UPDATED XXX}
-}{}
+\jan{QUENTIN FIX ABOVE FIGURE}
 
@@
 $$
-  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
-    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
+  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
+    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
 $$
 %
-The part $(\Gamma, Q) \! \vdash \! S \! \dashv \! (\Gamma', Q')$ of the judgement
+The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 \inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
-{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=1cm,width=20cm]
-  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2.5cm,rightskip=2.5cm, vdots=1cm]
+{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
+  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1cm]
     { \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
       { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
-        { \inferrule*[right={\scriptsize (Q:Dec)},leftskip=0cm,rightskip=0cm]
+        { \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
           {
           }
-          { (\cdot, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
-            \dashv (x {\ge} 0, P^\text{de})
-          }
-        }
-        { (\cdot, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
-          \dashv (\cdot, P^\text{we})
+          { (x {<} 10; B^\text{de}) \vdash
+	    \htriple{ x {\geq} 10; Q^\text{de} }{ \code{x = x - 10} }{ \cdot; P^\text{de} }
+          }
+        }
+        { (x {<} 10; B^\text{we}) \vdash
+	  \htriple{ x {\geq} 10; Q^\text{we} }{ \code{x = x - 10} }{ \cdot; P^\text{we} }
         }
@@
         }
-        { (\cdot, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
-          \dashv (\cdot, P^\text{ti})
-        }
-      }
-      { (\cdot, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash
-        \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq})
-      }
-    }
-    { (\cdot, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash
-      \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if})
+        { (x {<} 10; B^\text{ti}) \vdash
+	  \htriple{ \cdot; Q^\text{ti} }{ \code{tick(5)} }{ \cdot; P^\text{ti} }
+        }
+      }
+      { (x {<} 10; B^\text{sq}) \vdash
+	\htriple{ x {\geq} 10; Q^\text{sq} }{ \code{x=x - 10; tick(5)} }{ \cdot; P^\text{sq} }
+      }
+    }
+    { (x {<} 10; B^\text{if}) \vdash
+      \htriple{ x {\geq} 10; Q^\text{if} }{ \code{x=x - 10; tick(5)} }{ \cdot; P^\text{if} }
     }
@@
       { }
-      { (\cdot, B^\text{br}); (\cdot,Q^\text{br}) \vdash
-        \code{break} \dashv (\bot,P^\text{br})
-      }
-    }
-    { (\cdot, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash
-      \code{break} \dashv (\cdot,P^\text{el})
-    }
-  }
-  { \hspace{3cm} (\cdot, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash
-    \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm}
-  }
-}
-{ (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (\cdot, P)
+      { (x {<} 10; B^\text{br}) \vdash
+	\htriple{ x {<} 10; Q^\text{br} }{ \code{break} }{ \bot; P^\text{br} }
+      }
+    }
+    { (x {<} 10; B^\text{el}) \vdash
+      \htriple{ x {<} 10; Q^\text{el} }{ \code{break} }{ \cdot; P^\text{el} }
+    }
+  }
+  { \hspace{3cm} (x {<} 10; B^\text{lo}) \vdash
+    \htriple{ \cdot; Q^\text{lo} }{ \loopbody }{ \cdot; P^\text{lo} } \hspace{3cm}
+  }
+}
+{ (\cdot; B) \vdash
+  \htriple{ \cdot; Q) }{ \code{loop}~{\loopbody} }{ x {<} 10; P }
 }
@@
 The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}. It is parameterized by a logical context
+\textsc{Relax}. It is parameterized by a logical context that is
 used to gather information on interval sizes.  For example,
-$P^\text{de} \succeq_{(x{\ge}0)} P^\text{we} \equiv$
-% SSS SPACING
-%\vspace{-.3cm}
-{\small
+{%\small
   \begin{align*}
-  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0,10}
-  \land p^\text{we}_{x,0} \leq p^\text{de}_{x,0} + u_{x,0} - v_{0,x} \land \;\;
-  \\
-  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}
-  \bigwedge_{(\alpha, \beta) \neq (0,10), (x,0)} p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
+  P^\text{de} \succeq_{(\cdot)} P^\text{we} &\equiv
+ p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10} \\
+&\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
+&\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
   \end{align*}
 }
-% SSS SPACING
-%\vspace{-.8cm}
 

### 2015-04-08 paper/main.tex
reformat some maths

--- 2015-04-07 paper/main.tex
@@
 %SHORT
-\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captions
-\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
-\captionsetup[figure]{format=myformat}
-\captionsetup[table]{format=myformat}
-\renewcommand{\captionlabelfont}{\bf}
+%\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captions
+%\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
+%\captionsetup[figure]{format=myformat}
+%\captionsetup[table]{format=myformat}
+%\renewcommand{\captionlabelfont}{\bf}
 %SHORT
@@
 { Q' \succeq_\Gamma Q }
-\vspace{-.3cm}
+%\vspace{-.3cm}
 \end{mathpar}
@@
 \textsc{Relax}. It is parameterized by a logical context that is
-used to gather information on interval sizes.  For example, $P^\text{de} \succeq_{(\cdot)} P^\text{we} \equiv$
-\vspace{-.3cm}
-{\small
+used to gather information on interval sizes.  For example,
+{%\small
   \begin{align*}
- p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}
-\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
-\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
+  P^\text{de} \succeq_{(\cdot)} P^\text{we} &\equiv
+ p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10} \\
+&\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
+&\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
   \end{align*}
 }
-\vspace{-.8cm}
 

### 2015-04-08 paper/main.tex
compactify
rework the t62 explanation

@@
   demonstrate the unique compositionality of our system. In \emph{t61},
-  $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
-  are tight.}
+  $N\geq 0$ is a fixed but arbitrary constant. % The derived constant factors
+  %are tight.
+  }
   \label{fig:ex_list_2}
@@
 precisely, it is the partitioning part of the algorithm.  This partition
-loop has linear complexity, and feeding into our analysis gives the
-correct and tight worst-case bound $2+3|[l,h]|.$  With the potential
-method in mind, it is clear that the three loops get their
-potential from the same interval $[l,h]$.  However, KoAT fails to
+loop has linear complexity, and feeding it to our analysis gives a
+worst-case bound $2+3|[l,h]|$.  The bound is not optimal but it can be
+refined by rewriting the program. % to let our logical inference
+%discover more invariants.
+%With the potential
+%method in mind, it is clear that the three loops get their
+%potential from the same interval $[l,h]$ regardless of their nesting.
+With the potential method in mind, we can reason as follows.
+%
+If $h \geq l$ initially, the cost of the loop is 2.
+Otherwise, the cost of each round (at
+most 3) is payed using the potential of $[l,h]$ by the first
+increment to $l$ because we know that $l<h$.  The two inner loops
+can also use $[l,h]$ to pay for their inner costs.
+%
+KoAT fails to
 find a bound and LOOPUS derives the quadratic bound
 $(h-l-1)^2$.  Following the classical technique, these tools try to find one
-linear ranking-function for each loop and combine them multiplicatively
+ranking function for each loop and combine them multiplicatively
 or additively.
@@
 }
-
-\noindent
 The other rules are syntax directed and applied inductively.  For example,

### 2015-04-08 paper/main.tex
add a citation for CompCert

--- 2015-04-08 paper/main.tex
@@
 We describe the automatic amortized analysis for a subset of
-expressions of Clight.  Assignments are restricted to the form $x \gets
+expressions of Clight.  Clight is the first intermediate
+language of the CompCert compiler~\cite{CompcertCACM}, it
+is a cleaner C with a unified looping construct and
+side-effect free expressions.
+Assignments are restricted to the form $x \gets
 y$ or $x \gets x \pm y$.  In the implementation, a Clight
@@
 To base the soundness proof on a formal ground, we start by defining a
-new cost-aware operational semantics for CompCert Clight.  Clight's
+new cost-aware operational semantics for Clight.  Clight's
 operational semantics is based on small-step transitions and

### 2015-04-09 paper/main.tex
looks like PLDI conferenceinfo was broken
add nice and well deserved aec badge
add a note in the logical state section

--- 2015-04-08 paper/main.tex
@@
 \usepackage{array} %for stupid tables
+\usepackage[firstpage]{draftwatermark}
+\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[width=2.0cm]{fig/aec-badge-pldi.pdf}}}
+\SetWatermarkAngle{0}
 
@@
 
-\conferenceinfo{PLDI'15}{, June 13--17, 2015, Portland, OR, USA}
+\conferenceinfo{PLDI'15}{June 13--17, 2015, Portland, OR, USA}
 \copyrightyear{2015}
@@
 
-\begin{figure}[t]
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    @assert(na > 0);@
-    a[x]=0; @na--;@
-    tick(1); x++; }
-  if (x < N) { a[x]=1; @na++;@ tick(1); }
-  k--;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\end{minipage}
-\caption{Assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
-  tick metric.  The derived bound  is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-
-\begin{figure}
-\vspace{-.3cm}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h@,lg@) {
-  if (h-l > 1) {
-    @assert(lg > 0);@
-    m = l + (h-l)/2;
-    @lg--;@ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h@,lg@);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\caption{Assisted bound derivation using logical state.
-  % We use the tick metric to model the stack consumption of the program.
-  % Before a function call to \code{bsearch}, a tick statement pays for
-  % the space used by stack frame of the function.  The space is recoverd
-  % after the call.
-  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The semi-automatically derived bound is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
-
-\ifdefined\fullversion
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}
-  \mbox{logical state invariant }
-  \{ nm = A(0) \} }$
-i=1; m=a[0];
-while (i < N) {
-  if (a[i] > m) {
-    @assert(nm > 0); nm--;@
-    m = a[i], tick(1);
-  } i++, tick(1);
-}
-\end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We write $A(i)$ for
-  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
-  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
-  }
-\label{fig:xmplmax}
-\end{figure}
-\fi
-
 While complete automation is desirable, it is not always possible
@@
 a program $P$ that fails to be analyzed automatically is enriched by
-auxiliary variables $\vec x$ to form a program
+auxiliary variables $\vec x$ and assertions to form a program
 $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
@@
 and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
-% test is in fact semantically equivalent to \lstinline{true}.
+$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails.
 
@@
 the case that $\log_2(x-x/2) = \log_2(x)-1$.
+
+Generally, we observed that because the instrumented program
+is structurally same as the original one, it is sufficient to prove
+that the added assertions never fail to show that they have identical
+resource consumption.  This can be piggybacked on standard
+static-analysis tools.
 
@@
 % theorem of our Quantitative Hoare Logic.
+
+\begin{figure}[t]
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1) {
+    @assert(na > 0);@
+    a[x]=0; @na--;@
+    tick(1); x++; }
+  if (x < N) { a[x]=1; @na++;@ tick(1); }
+  k--;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\end{minipage}
+\caption{Assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
+  tick metric.  The derived bound  is
+  $2|[0,k]| + |[0,na]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
+\begin{figure}
+\vspace{-.3cm}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ lg > \log_2(h - l) \} }$
+bsearch(x,l,h@,lg@) {
+  if (h-l > 1) {
+    @assert(lg > 0);@
+    m = l + (h-l)/2;
+    @lg--;@ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h@,lg@);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\caption{Assisted bound derivation using logical state.
+  % We use the tick metric to model the stack consumption of the program.
+  % Before a function call to \code{bsearch}, a tick statement pays for
+  % the space used by stack frame of the function.  The space is recoverd
+  % after the call.
+  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
+  The semi-automatically derived bound is $|[0,lg]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
+
+\ifdefined\fullversion
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}
+  \mbox{logical state invariant }
+  \{ nm = A(0) \} }$
+i=1; m=a[0];
+while (i < N) {
+  if (a[i] > m) {
+    @assert(nm > 0); nm--;@
+    m = a[i], tick(1);
+  } i++, tick(1);
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We write $A(i)$ for
+  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
+  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
+  }
+\label{fig:xmplmax}
+\end{figure}
+\fi
 

### 2015-04-09 paper/main.tex
add a limitations section

@@
   handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
-  extended version~\cite{anon_extended}} is a list of more than $30$
+  extended version~\cite{anon_extended}} is a list of more than 30
 classes of challenging programs that we can automatically analyze.
@@
 tools\iffull{ for automatic bound derivation.}{.}
+
+\paraskip
+\paragraph{Limitations.}
+The implementation of our analysis does not currently support all of Clight.
+Function pointers, goto statements, and stack-allocated
+variables that have their address taken are not supported by the
+automatic analysis.  While these limitations concern the current
+implementation, our technique is in principle capable to handle them.
+
+For the sake of simplicity, the automated system described here is
+restricted to finding only linear bounds.  However, the amortized analysis
+technique was shown to work with polynomial bounds~\cite{HoffmannAH11};
+we leave this extension of our system as future work.
+
+Even some linear programs cannot be analyzed automatically by
+$\toolname$, it is usually the case for programs that rely on heap
+invariants (like nul-terminated C strings), for programs making
+use of non-linear operations (like \code{\%} or \code{*}), or for
+programs whose
+termination can only be proved by complex path-sensitive reasoning.
 

### 2015-04-09 paper/main.tex
use scale=1 for the badge
AEC badge will be added later

@@
 \usepackage{array} %for stupid tables
-\usepackage[firstpage]{draftwatermark}
-\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[width=2.0cm]{fig/aec-badge-pldi.pdf}}}
-\SetWatermarkAngle{0}
+%\usepackage[firstpage]{draftwatermark}
+%\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[scale=0.1]{fig/aec-badge-pldi.pdf}}}
+%\SetWatermarkAngle{0}
 
@@
 {Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
-{Yale University \vspace{-1cm}}
+{Yale University}
+{\{first.last\}@yale.edu}
 
@@
 
-\vspace{-.24cm}
-
 \iffull{The constraints that we generate have
@@
 \aftersectskip
+
+\begin{figure}[t]
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1) {
+    @assert(na > 0);@
+    a[x]=0; @na--;@
+    tick(1); x++; }
+  if (x < N) { a[x]=1; @na++;@ tick(1); }
+  k--;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\end{minipage}
+\caption{Assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
+  tick metric.  The derived bound  is
+  $2|[0,k]| + |[0,na]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
+\begin{figure}
+\vspace{-.3cm}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ lg > \log_2(h - l) \} }$
+bsearch(x,l,h@,lg@) {
+  if (h-l > 1) {
+    @assert(lg > 0);@
+    m = l + (h-l)/2;
+    @lg--;@ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h@,lg@);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\caption{Assisted bound derivation using logical state.
+  % We use the tick metric to model the stack consumption of the program.
+  % Before a function call to \code{bsearch}, a tick statement pays for
+  % the space used by stack frame of the function.  The space is recoverd
+  % after the call.
+  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
+  The semi-automatically derived bound is $|[0,lg]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
+
+\ifdefined\fullversion
+\begin{figure}
+\begin{lstlisting}
+${\color{blue}
+  \mbox{logical state invariant }
+  \{ nm = A(0) \} }$
+i=1; m=a[0];
+while (i < N) {
+  if (a[i] > m) {
+    @assert(nm > 0); nm--;@
+    m = a[i], tick(1);
+  } i++, tick(1);
+}
+\end{lstlisting}
+\caption{Example of assisted bound derivation using logical state.
+  We write $A(i)$ for
+  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
+  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
+  }
+\label{fig:xmplmax}
+\end{figure}
+\fi
 
@@
 % theorem of our Quantitative Hoare Logic.
-
-\begin{figure}[t]
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    @assert(na > 0);@
-    a[x]=0; @na--;@
-    tick(1); x++; }
-  if (x < N) { a[x]=1; @na++;@ tick(1); }
-  k--;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\end{minipage}
-\caption{Assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
-  tick metric.  The derived bound  is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-
-\begin{figure}
-\vspace{-.3cm}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h@,lg@) {
-  if (h-l > 1) {
-    @assert(lg > 0);@
-    m = l + (h-l)/2;
-    @lg--;@ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h@,lg@);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\caption{Assisted bound derivation using logical state.
-  % We use the tick metric to model the stack consumption of the program.
-  % Before a function call to \code{bsearch}, a tick statement pays for
-  % the space used by stack frame of the function.  The space is recoverd
-  % after the call.
-  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The semi-automatically derived bound is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
-
-\ifdefined\fullversion
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}
-  \mbox{logical state invariant }
-  \{ nm = A(0) \} }$
-i=1; m=a[0];
-while (i < N) {
-  if (a[i] > m) {
-    @assert(nm > 0); nm--;@
-    m = a[i], tick(1);
-  } i++, tick(1);
-}
-\end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We write $A(i)$ for
-  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
-  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
-  }
-\label{fig:xmplmax}
-\end{figure}
-\fi
 

### 2015-04-09 paper/main.tex
layout work, not so bad now!

@@
 
+\ifshort{
+\begin{figure*}
+\small
+\vspace{-.4cm}
+\def \MathparLineskip {\lineskip=0.35cm}
+\begin{mathpar}
+\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+{ }
+{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
+%
+\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
+{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
+%
+\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
+}
+%
+\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
+}
+%
+\\ \Rule{Q:Seq}
+{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
+\\\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
+%
+\and \Rule{Q:Weak}
+{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ \Gamma'_2 \models \Gamma'_1
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
+}
+{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
+%
+\\ \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
+\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
+\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
+}
+{ Q' \succeq_\Gamma Q }
+%\vspace{-.3cm}
+\end{mathpar}
+\caption{Selected inference rules of the quantitative analysis.}
+\label{fig:auto}
+\end{figure*}
+}{}
+
 \paraskip
@@
 }{}
-
-\ifshort{
-\begin{figure*}
-\small
-\vspace{-.4cm}
-\def \MathparLineskip {\lineskip=0.35cm}
-\begin{mathpar}
-\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
-{ }
-{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
-%
-\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
-{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
-%
-\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
-{ \Gamma \models y \ge 0 \!
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx} \!
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
-}
-%
-\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
-}
-%
-\\ \Rule{Q:Seq}
-{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
-\\\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
-}
-{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
-%
-\and \Rule{Q:Weak}
-{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
-\\ \!\! \Gamma_1 \models \Gamma_2
-\\\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ \Gamma'_2 \models \Gamma'_1
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-}
-{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
-%
-\\ \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
-\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
-\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
-\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
-}
-{ Q' \succeq_\Gamma Q }
-%\vspace{-.3cm}
-\end{mathpar}
-\caption{Selected inference rules of the quantitative analysis.}
-\label{fig:auto}
-\end{figure*}
-}{}
-
-\sectskip
-\section{Derivation System}
-
-In the following we describe the local and compositional derivation
-rules of the automatic amortized analysis.
-
-\paraskip
-\paragraph{Judgements.}
-
-The derivation system for the automatic amortized analysis is defined
-in~\pref{fig:auto}.  The derivation rules derive judgements of the form
-$$
-  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
-    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
-$$
-%
-The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
-can be seen as a quantitative Hoare triple.  All assertions are split into
-two parts, the logical part and the quantitative part.  The
-quantitative part $Q$ represents a potential function as a collection
-of non-negative numbers $q_i$ indexed by the index set $I$.  The
-logical part $\Gamma$ is left abstract but is enforced by our
-derivation system to respect classic Hoare logic constraints.
-%
-The meaning of this basic judgment is as follows: If $S$ is executed
-with starting state $\state$, the assertions in $\Gamma$
-hold, and at least $Q(\state)$ resources are available then the
-evaluation does not run out of resources and, if the execution
-terminates in state $\state'$, there are at least $Q(\state')$
-\mbox{resources left and $\Gamma'$ holds for $\state'$.}
-
-The judgement is a bit more involved since we have to take into
-account the early exit statements \code{break} and \code{return}.
-This is similar to classical Hoare triples in the presence of non-linear
-control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
-that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
-% of looping statements by the rule {\sc Q:Loop} and used before
-% \code{break} statements by the rule {\sc Q:Break}.
-Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
-returning from a function call with \code{return}. \iffull{(This becomes clear
-in the rules.)}{}
-
-% These judgements correspond to the logic judgements $\Delta; B; R
-% \vdash \htriple P S Q$ where each assertion splits in two parts, a
-% logical part $\Gamma$ and a quantitative part $Q$.  That means that
-% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
-% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
-% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
-% simple Hoare triple.  We leave out the function context $\Delta$ in
-% the presentation to avoid cluttering notations.  Instead we assume a
-% fixed function context $\Delta$ that is implicit on all judgements.
-% Function contexts are treated exactly as in the logic.  % We call
-% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
-% % $(\Gamma', Q')$.
-% The correspondence between judgements of the
-% automation and the ones of the quantitative logic is made formal by
-% our soundness proof of the automatic amortized analysis at the end of
-% this section.
-
-As a convention, if $Q$ and $Q'$ are quantitative annotations
-we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
-The notation $Q \pm n$ used in many rules defines a new context $Q'$
-such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
-In all the rules, we have the implicit side condition that all rational
-coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
-and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
-
-We describe the automatic amortized analysis for a subset of
-expressions of Clight.  Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}, it
-is a cleaner C with a unified looping construct and
-side-effect free expressions.
-Assignments are restricted to the form $x \gets
-y$ or $x \gets x \pm y$.  In the implementation, a Clight
-program is converted into this form prior to analysis without changing
-the resource cost.  This is achieved by using a series of \emph{cost-free
-  assignments} that do not result in additional cost in the
-semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
-
-\paraskip
-\paragraph{Function Specifications.}
-
-During the analysis, function specifications are quadruples
-$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
-$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
-are instantiated by appropriate variables on call sites.  A distinctive
-feature of our analysis is that it respects the function abstraction:
-when deriving a function specification it generates a set of
-constraints and the above quadruple; once done, the constraint
-set can readily be reused for every call site and the function need
-not be analyzed multiple times.
-%
-Therefore, the derivation rules are parametric in a function context $\Delta$
-that we leave implicit in the rules presented here. More details
-can be found in the extended version.
-
-\paraskip
-\paragraph{Derivation Rules.}
-
-The rules of our derivation system must serve two purposes.  They must
-\emph{attach} potential to certain program variable intervals and use
-this potential, when it is allowed, to \emph{pay} for resource consuming
-operations.  These two purposes are illustrated on the {\sc Q:Skip}
-rule.  This rule reuses its precondition as postcondition, it is
-explained by two facts: first, no resource is consumed by the skip
-operation, thus no potential has to be used to pay for the evaluation;
-second, the program state is not changed by
-the execution of a skip statement, thus all the potential available
-before the execution of the skip statement is still available after.
-\iffull{
-These two observations would also justify a rule with the side-condition
-$Q \ge Q'$, because some potential is always safely lost.  For the sake
-of orthogonality, it is not how we present the rule and this form of
-weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
-described later.}{}
-
-% Most of the rules correspond exactly to the respective rules of the
-% quantitative Hoare logic.
-% \iffull{
-%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
-%   as precondition.  This is sound since evaluating a $\code{skip}$
-%   statement never costs any resources and does not change the program
-%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
-%   quantitative annotation of the precondition is at least $M_t(n)$ to
-%   pay for the resource consumption.  Since the program state is also
-%   unchanged in this case, the logical part $\Gamma$ of the
-%   precondition can soundly be reused in the postcondition.  In the
-%   rule {\sc Q:Break}, where the linear control flow is broken, the
-%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
-%   taken from the break part of the judgement similarly as it is done
-%   in the quantitative logic.
-% }
-% %
-% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
-%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
-%   rules can be found in the extended version.
-% }
-% %
-% The rule {\sc Q:Return} exhibits a slight
-% difference to the quantitative logic.  While we have a function $\Z
-% \to \Assn$ to represent return values in the logic, we assume that a
-% special variable $\ret$ is part of the index set of $Q_R$.  The
-% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
-% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
-% $q_i'=q_i$ otherwise.
-
-The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
-describe how the potential is distributed after a size change of
-a variable.
-\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
-and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
-when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
-Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
-$x \gets x+y$, it applies when we can deduce from the logical context
-$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
-can be applied for decrements and when $y$ is not positive.}
-The rules are all equivalent in the case where $y=0$.
-The rule {\sc Q:Inc} can be applied
-if we cannot deduce any information about $y$.  In $\toolname$,
-the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
-
-\iffull{To explain how rules for increment and decrement work, it is
-sufficient to understand the rule {\sc Q:IncP}.  The others follow the
-same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
-variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
-quantitative annotation must be updated to reflect the change of the
-program state.  We write $x'$ for the value of $x$ after the assignment.
-Since $x$ is the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
-$\inter x u$ will get smaller with the update, and if $x' \in \inter x
-u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
-$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
-$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
-guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
-get bigger with the update.  We know that $|\inter v {x'}| \le y +
-|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
-$q_{vx}$ to pay for this change.  The rule ensures this
-only for $v \not\in \mathcal U$ because $x \le v$
-otherwise, and thus $|\inter v x| = 0$.
 
@@
 
+\sectskip
+\section{Derivation System}
+
+In the following we describe the local and compositional derivation
+rules of the automatic amortized analysis.
+
+\paraskip
+\paragraph{Judgements.}
+
+The derivation system for the automatic amortized analysis is defined
+in~\pref{fig:auto}.  The derivation rules derive judgements of the form
+$$
+  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
+    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
+$$
+%
+The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
+can be seen as a quantitative Hoare triple.  All assertions are split into
+two parts, the logical part and the quantitative part.  The
+quantitative part $Q$ represents a potential function as a collection
+of non-negative numbers $q_i$ indexed by the index set $I$.  The
+logical part $\Gamma$ is left abstract but is enforced by our
+derivation system to respect classic Hoare logic constraints.
+%
+The meaning of this basic judgment is as follows: If $S$ is executed
+with starting state $\state$, the assertions in $\Gamma$
+hold, and at least $Q(\state)$ resources are available then the
+evaluation does not run out of resources and, if the execution
+terminates in state $\state'$, there are at least $Q(\state')$
+\mbox{resources left and $\Gamma'$ holds for $\state'$.}
+
+The judgement is a bit more involved since we have to take into
+account the early exit statements \code{break} and \code{return}.
+This is similar to classical Hoare triples in the presence of non-linear
+control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
+that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
+% of looping statements by the rule {\sc Q:Loop} and used before
+% \code{break} statements by the rule {\sc Q:Break}.
+Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
+returning from a function call with \code{return}. \iffull{(This becomes clear
+in the rules.)}{}
+
+% These judgements correspond to the logic judgements $\Delta; B; R
+% \vdash \htriple P S Q$ where each assertion splits in two parts, a
+% logical part $\Gamma$ and a quantitative part $Q$.  That means that
+% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
+% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
+% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
+% simple Hoare triple.  We leave out the function context $\Delta$ in
+% the presentation to avoid cluttering notations.  Instead we assume a
+% fixed function context $\Delta$ that is implicit on all judgements.
+% Function contexts are treated exactly as in the logic.  % We call
+% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
+% % $(\Gamma', Q')$.
+% The correspondence between judgements of the
+% automation and the ones of the quantitative logic is made formal by
+% our soundness proof of the automatic amortized analysis at the end of
+% this section.
+
+As a convention, if $Q$ and $Q'$ are quantitative annotations
+we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
+The notation $Q \pm n$ used in many rules defines a new context $Q'$
+such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
+In all the rules, we have the implicit side condition that all rational
+coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
+and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
+
+We describe the automatic amortized analysis for a subset of
+expressions of Clight.  Clight is the first intermediate
+language of the CompCert compiler~\cite{CompcertCACM}, it
+is a cleaner C with a unified looping construct and
+side-effect free expressions.
+Assignments are restricted to the form $x \gets
+y$ or $x \gets x \pm y$.  In the implementation, a Clight
+program is converted into this form prior to analysis without changing
+the resource cost.  This is achieved by using a series of \emph{cost-free
+  assignments} that do not result in additional cost in the
+semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
+a[y]$ are handled by assigning $0$ to coefficients
+like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+
+\paraskip
+\paragraph{Function Specifications.}
+
+During the analysis, function specifications are quadruples
+$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
+$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
+are instantiated by appropriate variables on call sites.  A distinctive
+feature of our analysis is that it respects the function abstraction:
+when deriving a function specification it generates a set of
+constraints and the above quadruple; once done, the constraint
+set can readily be reused for every call site and the function need
+not be analyzed multiple times.
+%
+Therefore, the derivation rules are parametric in a function context $\Delta$
+that we leave implicit in the rules presented here. More details
+can be found in the extended version.
+
+\paraskip
+\paragraph{Derivation Rules.}
+
+The rules of our derivation system must serve two purposes.  They must
+\emph{attach} potential to certain program variable intervals and use
+this potential, when it is allowed, to \emph{pay} for resource consuming
+operations.  These two purposes are illustrated on the {\sc Q:Skip}
+rule.  This rule reuses its precondition as postcondition, it is
+explained by two facts: first, no resource is consumed by the skip
+operation, thus no potential has to be used to pay for the evaluation;
+second, the program state is not changed by
+the execution of a skip statement, thus all the potential available
+before the execution of the skip statement is still available after.
+\iffull{
+These two observations would also justify a rule with the side-condition
+$Q \ge Q'$, because some potential is always safely lost.  For the sake
+of orthogonality, it is not how we present the rule and this form of
+weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
+described later.}{}
+
+% Most of the rules correspond exactly to the respective rules of the
+% quantitative Hoare logic.
+% \iffull{
+%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
+%   as precondition.  This is sound since evaluating a $\code{skip}$
+%   statement never costs any resources and does not change the program
+%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
+%   quantitative annotation of the precondition is at least $M_t(n)$ to
+%   pay for the resource consumption.  Since the program state is also
+%   unchanged in this case, the logical part $\Gamma$ of the
+%   precondition can soundly be reused in the postcondition.  In the
+%   rule {\sc Q:Break}, where the linear control flow is broken, the
+%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
+%   taken from the break part of the judgement similarly as it is done
+%   in the quantitative logic.
+% }
+% %
+% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
+%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
+%   rules can be found in the extended version.
+% }
+% %
+% The rule {\sc Q:Return} exhibits a slight
+% difference to the quantitative logic.  While we have a function $\Z
+% \to \Assn$ to represent return values in the logic, we assume that a
+% special variable $\ret$ is part of the index set of $Q_R$.  The
+% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
+% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
+% $q_i'=q_i$ otherwise.
+
+The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
+describe how the potential is distributed after a size change of
+a variable.
+\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
+and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
+when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
+Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
+that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
+$x \gets x+y$, it applies when we can deduce from the logical context
+$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
+can be applied for decrements and when $y$ is not positive.}
+The rules are all equivalent in the case where $y=0$.
+The rule {\sc Q:Inc} can be applied
+if we cannot deduce any information about $y$.  In $\toolname$,
+the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
+
+\iffull{To explain how rules for increment and decrement work, it is
+sufficient to understand the rule {\sc Q:IncP}.  The others follow the
+same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
+variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
+quantitative annotation must be updated to reflect the change of the
+program state.  We write $x'$ for the value of $x$ after the assignment.
+Since $x$ is the only variable changed, only intervals of the form
+$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
+$\inter x u$ will get smaller with the update, and if $x' \in \inter x
+u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
+$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
+$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
+guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
+get bigger with the update.  We know that $|\inter v {x'}| \le y +
+|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
+$q_{vx}$ to pay for this change.  The rule ensures this
+only for $v \not\in \mathcal U$ because $x \le v$
+otherwise, and thus $|\inter v x| = 0$.
+
 \iffull{
@@
 
+The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
+loop body one needs to find an invariant potential $Q$ that will pay for
+the iterations.  At each iteration, $M_l$ resources are spent to jump back,
+this explains the postcondition $Q{+}M_l$. %  Assuming $M_l>0$, and if $Q$
+%is finite on the initial heap (i.e. the precondition is not $\bot$), the
+%loop can only run for a finite number of iterations.
+To exit the loop,
+the programmer uses the \code{break} statement, this explains why the postcondition
+for the full $\code{loop}~S$ statement $\{ \Gamma'; Q' \}$ is used as break
+postcondition in the derivation for $S$.
+
 \iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
@@
   article.}
+
+\begin{figure*}
+\centering
+\vspace{-.5cm}
+\begin{tabular}{r|c|c|c|c|c}
+%
+& t09 & t19 & t30 & t15 & t13\\
+%
+\hline
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+i=1; j=0;
+while (j<x) {
+  j++;
+  if (i>=4)
+    i=1, tick(40);
+  else i++;
+  tick(1); }
+\end{lstlisting}
+&
+\hspace{-.5cm}\begin{lstlisting}[basicstyle=\tt\footnotesize]
+while (i>100) {
+  i--; tick(1);
+} i += k+50;
+while (i>=0) {
+  i--; tick(1);
+}\end{lstlisting}\hspace{-.5cm}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+while (x>0) {
+  x--;
+  t=x, x=y, y=t;
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+assert(y>=0);
+while (x > y) {
+  x -= y+1;
+  for (z=y; z>0; z--)
+    tick(1);
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+while (x>0) {
+  x--;
+  if (*) y++;
+  else
+    while (y>0)
+      y--, tick(1);
+  tick(1); }
+\end{lstlisting}
+%
+\\
+\hline
+%
+%
+%
+\toolname
+  & {\small $11|[0,x]|$ }
+  & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
+  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
+  & {\small $|\inter 0 x|$}
+  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
+\\
+%
+Rank
+  & {\small $23{\cdot}x - 14$ }
+  & {\small $54+k+i$ }
+  & {\small ---}
+  & {\small $2+2x-y$ }
+  & {\small $0.5{\cdot}y^2 {+} yx \ldots$}
+           % +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
+\\
+%
+{\hspace{-.2cm}LOOPUS}
+&
+{\small
+  $41\max(x,0)$
+}
+&
+{\small
+  $\!\!\!\!$\begin{tabular}{c}
+    $\max(i{-}100, 0)$ \\
+    $+ \max(k {+} i {+} 51, 0)$
+  \end{tabular}$\!\!\!\!$
+}
+&
+{\small ---}% $1 + \max(x, y)$ (???)
+&
+{\small ---}
+&
+{\small \begin{tabular}{c}
+           $2\max(x, 0)$ \\
+           $+ \max(y, 0)$
+        \end{tabular}}
+\end{tabular}
+\vspace{.01cm}
+\caption{ Comparison of resource bounds derived by different tools on
+  several examples with linear bounds.  \iffull{The output of Rank has been manually
+  simplified to fit the table.}{}  }
+\label{fig:compar}
+\end{figure*}
 
@@
 
-\begin{figure*}
-\centering
-\vspace{-.5cm}
-\begin{tabular}{r|c|c|c|c|c}
-%
-& t09 & t19 & t30 & t15 & t13\\
-%
-\hline
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-i=1; j=0;
-while (j<x) {
-  j++;
-  if (i>=4)
-    i=1, tick(40);
-  else i++;
-  tick(1); }
-\end{lstlisting}
-&
-\hspace{-.5cm}\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (i>100) {
-  i--; tick(1);
-} i += k+50;
-while (i>=0) {
-  i--; tick(1);
-}\end{lstlisting}\hspace{-.5cm}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (x>0) {
-  x--;
-  t=x, x=y, y=t;
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-assert(y>=0);
-while (x > y) {
-  x -= y+1;
-  for (z=y; z>0; z--)
-    tick(1);
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (x>0) {
-  x--;
-  if (*) y++;
-  else
-    while (y>0)
-      y--, tick(1);
-  tick(1); }
-\end{lstlisting}
-%
-\\
-\hline
-%
-%
-%
-\toolname
-  & {\small $11|[0,x]|$ }
-  & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
-  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
-  & {\small $|\inter 0 x|$}
-  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
-\\
-%
-Rank
-  & {\small $23{\cdot}x - 14$ }
-  & {\small $54+k+i$ }
-  & {\small ---}
-  & {\small $2+2x-y$ }
-  & {\small $0.5{\cdot}y^2 {+} yx \ldots$}
-           % +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
-\\
-%
-{\hspace{-.2cm}LOOPUS}
-&
-{\small
-  $41\max(x,0)$
-}
-&
-{\small
-  $\!\!\!\!$\begin{tabular}{c}
-    $\max(i{-}100, 0)$ \\
-    $+ \max(k {+} i {+} 51, 0)$
-  \end{tabular}$\!\!\!\!$
-}
-&
-{\small ---}% $1 + \max(x, y)$ (???)
-&
-{\small ---}
-&
-{\small \begin{tabular}{c}
-           $2\max(x, 0)$ \\
-           $+ \max(y, 0)$
-        \end{tabular}}
-\end{tabular}
-\vspace{.01cm}
-\caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  \iffull{The output of Rank has been manually
-  simplified to fit the table.}{}  }
-\label{fig:compar}
-\end{figure*}
-
 \sectskip

### 2015-04-09 paper/main.tex
fix some phrasing

@@
 Generally, we observed that because the instrumented program
-is structurally same as the original one, it is sufficient to prove
-that the added assertions never fail to show that they have identical
-resource consumption.  This can be piggybacked on standard
-static-analysis tools.
+is structurally same as the original one, it is enough to prove
+that the added assertions never fail in order to show the two programs
+have identical resource consumption.  This can usually be piggybacked
+on standard static-analysis tools.
 

### 2015-04-12 paper/main.tex
oh beautiful rules
spacing
removed scriptsize before the bib
added acm terms and categories, and keywords

@@
 
-% \category{D.2.4}{Software Engineering}
-% {Software/Program Verification}
-% \category{F.3.1}{Logics and Meanings of Programs}
-% {Specifying and Verifying and Reasoning about Programs}
-
-% \terms Verification, Reliability
-
-% \keywords Formal Verification, Compiler Construction, Program Logics,
-% Stack-Space Bounds, Quantitative Verification
+\category{D.2.4}{Software Engineering}
+{Software/Program Verification}
+\category{F.3.1}{Logics and Meanings of Programs}
+{Specifying and Verifying and Reasoning about Programs}
+
+\terms Verification, Reliability
+
+\keywords Quantitative Verification, Resource Bound Analysis, Static
+Program Analysis, Amortized Analysis, LP Solving, Program Logic
 
@@
 \begin{mathpar}
-\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+\Rule{Q:Skip}
 { }
@@
 %
-\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+\and \Rule[leftskip=.5cm, rightskip=.5cm]{Q:Assert}
+{ }
+{ B, R \vdash \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
+%
+\and \Rule{Q:Break}
+{ }
+{(\Gamma; Q_B), R \vdash \htriple{ \Gamma; Q_B {+} M_b }{ \code{break} }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Return}
+{ P = Q_R[\Vret/x]
+\\ \Gamma = \Gamma_R[\Vret/x]
+\\ \forall i \in \dom{P} . \, p_i = q_i
+}
+{ B, (\Gamma_R; Q_R) \vdash \htriple{ \Gamma; Q }{ \code{return}~x }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Loop}
 { (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
@@
 %
-\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
+\and \Rule{Q:Seq}
+{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
+\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
+%
+\and \Rule{Q:If}
+{ B, R \vdash \htriple{ \Gamma {\land} e; Q {-} M_c^1) }{ S_1 }{ \Gamma'; Q' }
+\\\\ B, R \vdash \htriple{ \Gamma {\land} \neg e; Q {-} M_c^2 }{ S_2 }{ \Gamma'; Q' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q{+}M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ \Gamma'; Q' } }
+%
+\\ \Rule{Q:IncP}
 { \Gamma \models y \ge 0 \!
@@
 %
-\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
+\and \Rule{Q:Inc}
 { M = M_u + M_e(x {\pm} y)
@@
 %
-\\ \Rule{Q:Seq}
-{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
-\\\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
-}
-{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
+\and \Rule{Q:Set}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ B, R \vdash \htriple{ \Gamma[x/y]; Q {+} M_u {+} M_e(y) }{ x \gets y }{ \Gamma; Q' } }
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {-} y]; Q {+} M_u {+} M_e(x {-} y) }{ x \gets x - y }{ \Gamma; Q' }
+}
+%
+\and\Rule{Q:Call}
+{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
+\\ \text{Loc} = \text{Locals}(Q) \!
+\\ \forall i \neq j . \, x_i \neq x_j \!
+\\ c \in \Qplusz \!
+\\ Q = P + S \!
+\\ Q' = P' + S
+\\ U = Q_f[\Vargs / \vec x]
+\\ U' = Q'_f[\Vret/r]
+\\ \forall i \in \dom{U} . \, p_i = u_i
+\\ \forall i \in \dom{U'} .\, p'_i = u'_i
+\\ \forall i \not\in \dom{U'} .\, p'_i = 0
+\\ \forall i \not\in \text{Loc}
+  .\, s_i = 0
+}
+{ B, R \vdash \htriple
+  { \Gamma_f[\Vargs / \vec x] {\land} \Gamma_{\text{Loc}}; Q {+} c {+} M_f }
+  { r \gets f(\vec x) }
+  { \Gamma_f'[\Vret/r] {\land} \Gamma_{\text{Loc}}; Q' {+} c {-} M_r }
+}
+%
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ \fenv f = (\vec y, S_f)
+\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{ \Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y] }{ S_f }{ \Gamma'; Q' }
+}
+{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
 %
@@
 \sectskip
-\scriptsize
 \bibliography{lit}

### 2015-04-13 paper/main.tex
added acks
minor changes

@@
 {Yale University}
-{\{first.last\}@yale.edu}
+{\{quentin.carbonneaux, jan.hoffmann, zhong.shao\}@yale.edu}
 
@@
 {Specifying and Verifying and Reasoning about Programs}
-
+\vspace{-3pt}
 \terms Verification, Reliability
-
+\vspace{-3pt}
 \keywords Quantitative Verification, Resource Bound Analysis, Static
-Program Analysis, Amortized Analysis, LP Solving, Program Logic
+Analysis, Amortized Analysis, LP Solving, Program Logic
 
@@
 
-% \acks
-
-% This research is based on work supported in part by NSF grants 1319671
-% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
-% Any opinions, findings, and conclusions contained in this document are
-% those of the authors and do not reflect the views of these agencies.
+\acks
+
+This research is based on work supported in part by NSF grants 1319671
+and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
+Any opinions, findings, and conclusions contained in this document are
+those of the authors and do not reflect the views of these agencies.
 

### 2015-04-13 paper/main.tex
move limitations in a section, cosmetics on section 2

@@
 we also use as the loop invariant.  For the loop body we have (like in
-Hoare logic) to derive a triple like
+Hoare logic) to derive a triple
 $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
@@
 to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions are only used in the rule for the assignment
+$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
+logical assertions from potential functions by semicolons. Note that the
+logical assertions are only used in the rule for the assignment
 \code{x=x+K}.
@@
 tools\iffull{ for automatic bound derivation.}{.}
-
-\paraskip
-\paragraph{Limitations.}
-The implementation of our analysis does not currently support all of Clight.
-Function pointers, goto statements, and stack-allocated
-variables that have their address taken are not supported by the
-automatic analysis.  While these limitations concern the current
-implementation, our technique is in principle capable to handle them.
-
-For the sake of simplicity, the automated system described here is
-restricted to finding only linear bounds.  However, the amortized analysis
-technique was shown to work with polynomial bounds~\cite{HoffmannAH11};
-we leave this extension of our system as future work.
-
-Even some linear programs cannot be analyzed automatically by
-$\toolname$, it is usually the case for programs that rely on heap
-invariants (like nul-terminated C strings), for programs making
-use of non-linear operations (like \code{\%} or \code{*}), or for
-programs whose
-termination can only be proved by complex path-sensitive reasoning.
 
@@
 \sectskip
+\section{Limitations}
+\label{sec:limitations}
+\aftersectskip
+
+Our current implementation of $\toolname$ does not currently support
+all of Clight.  Programs with function pointers, goto statements,
+and dereferences of stack-allocated variables cannot be analyzed
+automatically.  While these limitations concern the current
+implementation, our technique is in principle capable to handle them.
+
+For the sake of simplicity, the automated system described here is
+restricted to finding only linear bounds.  However, the amortized analysis
+technique was shown to work with polynomial bounds~\cite{HoffmannAH11};
+we leave this extension of our system as future work.
+
+Even certain linear programs cannot be analyzed automatically by
+$\toolname$, it is usually the case for programs that rely on heap
+invariants (like nul-terminated C strings), for programs making
+use of non-linear operations (like \code{\%} or \code{*}), or for
+programs whose
+termination can only be proved by complex path-sensitive reasoning.
+
+\sectskip
 \section{Related Work}

### 2015-04-13 paper/main.tex
remove + 0.|[y,x]| in figure 1
add a note on non-linear operations

@@
 ${\color{blue}%
-  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
+  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
 while (x+K<=y) {
   ${\color{blue}%
-  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
+  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
   x=x+K;
   ${\color{blue}%
-  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
+  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| \} }$
   tick(T);
   ${\color{blue}%
-  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
+  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
 }
 ${\color{blue}%
- \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
+ \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
 \end{lstlisting}
@@
 that cannot be assigned to.  For example, if the program contains the
-constant $8128$ then we have a variable $c_{8128}$ and
-$\state(c_{8128}) = 8128$.  We assume that every program state
+constant $1988$ then we have a variable $c_{1988}$ and
+$\state(c_{1988}) = 1988$.  We assume that every program state
 includes the constant $c_0$.
@@
 like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+This sound treatment ensures that no further loop bounds
+depend on the result of the non-linear operation.
 

### 2015-04-13 paper/main.tex
more rule descriptions

@@
 
-The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
+The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
 describe how the potential is distributed after a size change of
@@
 The rule {\sc Q:Inc} can be applied
-if we cannot deduce any information about $y$.  In $\toolname$,
-the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}
-
-\iffull{To explain how rules for increment and decrement work, it is
+if we cannot find the sign of $y$.  % In $\toolname$,
+% the rules for increments and decrements are combined into one rule\iffull{ where
+% the specialized rules take priority over {\sc Q:Inc}.}{.}
+
+To explain how rules for increment and decrement work, it is
 sufficient to understand the rule {\sc Q:IncP}.  The others follow the
-same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
+same idea and are symmetrical. In {\sc Q:IncP}, the program updates a
 variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
@@
 
-\iffull{
+The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
+loop body one needs to find an invariant potential $Q$ that will pay for
+the iterations.  At each iteration, $M_l$ resources are spent to jump
+back, this explains the postcondition $Q{+}M_l$.  To exit the loop,
+the programmer uses the \code{break} statement, it is the reason why
+the postcondition for the full $\code{loop}~S$ statement $\{ \Gamma';
+Q' \}$ is used as break postcondition in the derivation for $S$.
+
 Another interesting rule is {\sc Q:Call}.  It needs to account for the
@@
 \item
-  $\!$The potential of intervals $|\inter x y|$ is preserved
+  The potential of intervals $|\inter x y|$ is preserved
+  across a function call
   if $x$ and $y$ are local\iffull{ variables.}{.}
@@
 P' {+} S'$.  Arguments in the function precondition
-$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
+$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
 of names different from all program variables.  This
@@
 the return value in the function's postcondition
-$(\Gamma'_f, Q'_f)$.
-}{}
-
-The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
-loop body one needs to find an invariant potential $Q$ that will pay for
-the iterations.  At each iteration, $M_l$ resources are spent to jump back,
-this explains the postcondition $Q{+}M_l$. %  Assuming $M_l>0$, and if $Q$
-%is finite on the initial heap (i.e. the precondition is not $\bot$), the
-%loop can only run for a finite number of iterations.
-To exit the loop,
-the programmer uses the \code{break} statement, this explains why the postcondition
-for the full $\code{loop}~S$ statement $\{ \Gamma'; Q' \}$ is used as break
-postcondition in the derivation for $S$.
-
-\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
+$(\Gamma'_f; Q'_f)$.
+
+The rule {\sc Q:Weak} is not syntax directed. In the implementation
 we apply {\sc Q:Weak} before loops and between the two statements of a
@@
 syntax directed rule but this simple heuristic helps to make the analysis
-efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
+efficient.  The high-level idea of {\sc Q:Weak} is the following: If
 we have a sound judgement, then it is sound to add more potential to the
@@
 \emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}\iffull{ that
+that is defined in the rule {\sc Relax} that
 also deals with the important task of transferring constant potential
@@
 from the logical context that the interval size $|\inter x y| \geq l$ is
-larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
+larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
 and guarantee that we do not gain potential.  Conversely, if $|\inter
 x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
+without gaining potential.
 

### 2015-04-13 paper/main.tex
fix inconsistencies in section 7

@@
 \and\Rule{Q:Call}
-{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
+{ (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
 \\ \text{Loc} = \text{Locals}(Q) \!
@@
 We define quantitative Hoare triples
-as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
+as $B;R\vdash_L \htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
 maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
@@
 $$
-  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
+  \tr(\Gamma, Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
 $$
@@
 %
-\iffull{These definitions let}{This lets} us translate the judgement
-$J := B; R; P \vdash S \dashv P'$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
+These definitions let us translate the judgement
+$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
 \iffull{$$
@@
 $$
-  \tr(J) := \tr(B); \tr(R) \vdash \htriple{\tr(P)} S {\tr(P')}.
+  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
 $$}%

### 2015-04-13 paper/main.tex
create Cost Aware Clight, add Q:Tick rule

@@
 %
-\and \Rule[leftskip=.5cm, rightskip=.5cm]{Q:Assert}
+\and \Rule{Q:Assert}
 { }
 { B, R \vdash \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
+%
+\and \Rule{Q:Tick}
+{ }
+{ B, R \vdash \htriple{ \Gamma; Q {+} M_t(n) }{ \code{tick}(n) }{ \Gamma; Q } }
 %
@@
 }
-{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
 %
@@
 \end{mathpar}
-\caption{Selected inference rules of the quantitative analysis.}
+\caption{Inference rules of the quantitative analysis.}
 \label{fig:auto}
@@
 \paraskip
+\paragraph{Cost Aware Clight.}
+
+We present the rules for a subset of the
+expressions of Clight.  Clight is the first intermediate
+language of the CompCert compiler~\cite{CompcertCACM}, it
+is a cleaner C with a unified looping construct and
+side-effect free expressions.  We reuse most of CompCert's
+syntax but instrument the semantics with a resource metric
+$M$ that accounts for the cost (an arbitrary rational number)
+of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
+evaluating the expression \code{exp}; the rationals $M_f$
+and $M_r$ account respectively for the cost of a call to the
+function $f$ and the cost of returning from it.
+More details on this cost-aware semantics are provided in \pref{sec:soundness}.
+
+In the rules, assignments are restricted to the form $x \gets
+y$ or $x \gets x \pm y$.  In the implementation, a Clight
+program is converted into this form prior to analysis without changing
+the resource cost.  This is achieved by using a series of \emph{cost-free
+  assignments} that do not result in additional cost in the
+semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
+a[y]$ are handled by assigning $0$ to coefficients
+like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+This sound treatment ensures that no further loop bounds
+depend on the result of the non-linear operation.
+
+\paraskip
 \paragraph{Judgements.}
@@
 terminates in state $\state'$, there are at least $Q(\state')$
-\mbox{resources left and $\Gamma'$ holds for $\state'$.}
+resources left and $\Gamma'$ holds for $\state'$.
 
@@
 Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
-returning from a function call with \code{return}. \iffull{(This becomes clear
+returning from a function call. \iffull{(This becomes clear
 in the rules.)}{}
@@
 and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
-
-We describe the automatic amortized analysis for a subset of
-expressions of Clight.  Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}, it
-is a cleaner C with a unified looping construct and
-side-effect free expressions.
-Assignments are restricted to the form $x \gets
-y$ or $x \gets x \pm y$.  In the implementation, a Clight
-program is converted into this form prior to analysis without changing
-the resource cost.  This is achieved by using a series of \emph{cost-free
-  assignments} that do not result in additional cost in the
-semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
-This sound treatment ensures that no further loop bounds
-depend on the result of the non-linear operation.
 
@@
 all of Clight.  Programs with function pointers, goto statements,
-and dereferences of stack-allocated variables cannot be analyzed
-automatically.  While these limitations concern the current
-implementation, our technique is in principle capable to handle them.
+continue statements, and dereferences of stack-allocated variables
+cannot be analyzed automatically.  While these limitations concern the
+current implementation, our technique is in principle capable to handle
+them.
 

### 2015-04-13 paper/main.tex
add note on LP certificates, move figures

@@
 
+\sectskip
+\section{Derivation System}
+
+In the following we describe the local and compositional derivation
+rules of the automatic amortized analysis.
+
+\paraskip
+\paragraph{Cost Aware Clight.}
+
+We present the rules for a subset of the
+expressions of Clight.  Clight is the first intermediate
+language of the CompCert compiler~\cite{CompcertCACM}, it
+is a cleaner C with a unified looping construct and
+side-effect free expressions.  We reuse most of CompCert's
+syntax but instrument the semantics with a resource metric
+$M$ that accounts for the cost (an arbitrary rational number)
+of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
+evaluating the expression \code{exp}; the rationals $M_f$
+and $M_r$ account respectively for the cost of a call to the
+function $f$ and the cost of returning from it.
+More details are provided in \pref{sec:soundness}.
+
+In the rules, assignments are restricted to the form $x \gets
+y$ or $x \gets x \pm y$.  In the implementation, a Clight
+program is converted into this form prior to analysis without changing
+the resource cost.  This is achieved by using a series of \emph{cost-free
+  assignments} that do not result in additional cost in the
+semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
+a[y]$ are handled by assigning $0$ to coefficients
+like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+This sound treatment ensures that no further loop bounds
+depend on the result of the non-linear operation.
+
+\paraskip
+\paragraph{Judgements.}
+
+The derivation system for the automatic amortized analysis is defined
+in~\pref{fig:auto}.  The derivation rules derive judgements of the form
+$$
+  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
+    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
+$$
+%
+The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
+can be seen as a quantitative Hoare triple.  All assertions are split into
+two parts, the logical part and the quantitative part.  The
+quantitative part $Q$ represents a potential function as a collection
+of non-negative numbers $q_i$ indexed by the index set $I$.  The
+logical part $\Gamma$ is left abstract but is enforced by our
+derivation system to respect classic Hoare logic constraints.
+%
+The meaning of this basic judgment is as follows: If $S$ is executed
+with starting state $\state$, the assertions in $\Gamma$
+hold, and at least $Q(\state)$ resources are available then the
+evaluation does not run out of resources and, if the execution
+terminates in state $\state'$, there are at least $Q(\state')$
+resources left and $\Gamma'$ holds for $\state'$.
+
+The judgement is a bit more involved since we have to take into
+account the early exit statements \code{break} and \code{return}.
+This is similar to classical Hoare triples in the presence of non-linear
+control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
+that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
+% of looping statements by the rule {\sc Q:Loop} and used before
+% \code{break} statements by the rule {\sc Q:Break}.
+Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
+returning from a function call. \iffull{(This becomes clear
+in the rules.)}{}
+
+% These judgements correspond to the logic judgements $\Delta; B; R
+% \vdash \htriple P S Q$ where each assertion splits in two parts, a
+% logical part $\Gamma$ and a quantitative part $Q$.  That means that
+% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
+% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
+% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
+% simple Hoare triple.  We leave out the function context $\Delta$ in
+% the presentation to avoid cluttering notations.  Instead we assume a
+% fixed function context $\Delta$ that is implicit on all judgements.
+% Function contexts are treated exactly as in the logic.  % We call
+% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
+% % $(\Gamma', Q')$.
+% The correspondence between judgements of the
+% automation and the ones of the quantitative logic is made formal by
+% our soundness proof of the automatic amortized analysis at the end of
+% this section.
+
+As a convention, if $Q$ and $Q'$ are quantitative annotations
+we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
+The notation $Q \pm n$ used in many rules defines a new context $Q'$
+such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
+In all the rules, we have the implicit side condition that all rational
+coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
+and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
+
+\paraskip
+\paragraph{Function Specifications.}
+
+During the analysis, function specifications are quadruples
+$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
+$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
+are instantiated by appropriate variables on call sites.  A distinctive
+feature of our analysis is that it respects the function abstraction:
+when deriving a function specification it generates a set of
+constraints and the above quadruple; once done, the constraint
+set can readily be reused for every call site and the function need
+not be analyzed multiple times.
+%
+Therefore, the derivation rules are parametric in a function context $\Delta$
+that we leave implicit in the rules presented here. More details
+can be found in the extended version.
+
+\paraskip
+\paragraph{Derivation Rules.}
+
+The rules of our derivation system must serve two purposes.  They must
+\emph{attach} potential to certain program variable intervals and use
+this potential, when it is allowed, to \emph{pay} for resource consuming
+operations.  These two purposes are illustrated on the {\sc Q:Skip}
+rule.  This rule reuses its precondition as postcondition, it is
+explained by two facts: first, no resource is consumed by the skip
+operation, thus no potential has to be used to pay for the evaluation;
+second, the program state is not changed by
+the execution of a skip statement, thus all the potential available
+before the execution of the skip statement is still available after.
+\iffull{
+These two observations would also justify a rule with the side-condition
+$Q \ge Q'$, because some potential is always safely lost.  For the sake
+of orthogonality, it is not how we present the rule and this form of
+weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
+described later.}{}
+
+% Most of the rules correspond exactly to the respective rules of the
+% quantitative Hoare logic.
+% \iffull{
+%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
+%   as precondition.  This is sound since evaluating a $\code{skip}$
+%   statement never costs any resources and does not change the program
+%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
+%   quantitative annotation of the precondition is at least $M_t(n)$ to
+%   pay for the resource consumption.  Since the program state is also
+%   unchanged in this case, the logical part $\Gamma$ of the
+%   precondition can soundly be reused in the postcondition.  In the
+%   rule {\sc Q:Break}, where the linear control flow is broken, the
+%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
+%   taken from the break part of the judgement similarly as it is done
+%   in the quantitative logic.
+% }
+% %
+% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
+%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
+%   rules can be found in the extended version.
+% }
+% %
+% The rule {\sc Q:Return} exhibits a slight
+% difference to the quantitative logic.  While we have a function $\Z
+% \to \Assn$ to represent return values in the logic, we assume that a
+% special variable $\ret$ is part of the index set of $Q_R$.  The
+% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
+% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
+% $q_i'=q_i$ otherwise.
+
+The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
+describe how the potential is distributed after a size change of
+a variable.
+\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
+and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
+when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
+Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
+that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
+$x \gets x+y$, it applies when we can deduce from the logical context
+$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
+can be applied for decrements and when $y$ is not positive.}
+The rules are all equivalent in the case where $y=0$.
+The rule {\sc Q:Inc} can be applied
+if we cannot find the sign of $y$.  % In $\toolname$,
+% the rules for increments and decrements are combined into one rule\iffull{ where
+% the specialized rules take priority over {\sc Q:Inc}.}{.}
+
+To explain how rules for increment and decrement work, it is
+sufficient to understand the rule {\sc Q:IncP}.  The others follow the
+same idea and are symmetrical. In {\sc Q:IncP}, the program updates a
+variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
+quantitative annotation must be updated to reflect the change of the
+program state.  We write $x'$ for the value of $x$ after the assignment.
+Since $x$ is the only variable changed, only intervals of the form
+$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
+$\inter x u$ will get smaller with the update, and if $x' \in \inter x
+u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
+$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
+$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
+guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
+get bigger with the update.  We know that $|\inter v {x'}| \le y +
+|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
+$q_{vx}$ to pay for this change.  The rule ensures this
+only for $v \not\in \mathcal U$ because $x \le v$
+otherwise, and thus $|\inter v x| = 0$.
+
+The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
+loop body one needs to find an invariant potential $Q$ that will pay for
+the iterations.  At each iteration, $M_l$ resources are spent to jump
+back, this explains the postcondition $Q{+}M_l$.  To exit the loop,
+the programmer uses the \code{break} statement, it is the reason why
+the postcondition for the full $\code{loop}~S$ statement $\{ \Gamma';
+Q' \}$ is used as break postcondition in the derivation for $S$.
+
+Another interesting rule is {\sc Q:Call}.  It needs to account for the
+changes to the stack caused by the function call, the arguments/return
+value passing, and the preservation of local variables.  We can sum up
+the main ideas of the rule as follows.
+\itemskip
+\begin{itemize}
+\item The potential in the pre- and postcondition of the function
+  specification is equalized to its matching potential in the callee's
+  pre- and postcondition.
+\itemskipIn
+\item
+  The potential of intervals $|\inter x y|$ is preserved
+  across a function call
+  if $x$ and $y$ are local\iffull{ variables.}{.}
+\item
+  The unknown potentials after the call (e.g.
+  $|\inter x g|$, with $x$ local and $g$ global)
+  are set to zero in the postcondition.
+\end{itemize}
+\itemskip
+If $x$ and $y$ are local variables and $f(x,y)$ is
+called, {\sc Q:Call} splits the potential of
+$|\inter x y|$ in two parts, one part to perform the
+computation in the function $f$ and one part to keep for
+later use after the function call.  This splitting is
+realized by the equations $Q = P {+} S$ and $Q' =
+P' {+} S'$.  Arguments in the function precondition
+$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
+of names different from all program variables.  This
+prevents name conflicts from happening and ensures that
+the substitution $[\Vargs/\vec x]$ is meaningful.
+Symmetrically, we use the unique name $\Vret$ to represent
+the return value in the function's postcondition
+$(\Gamma'_f; Q'_f)$.
+
+The rule {\sc Q:Weak} is not syntax directed. In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition. We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax} that
+also deals with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq l$ is
+larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
+without gaining potential.
+
 \newcommand{\loopbody}[0]{\ensuremath{\code{if~(x \geq 10)~(x=x - 10; tick(5))~else~break}}}
@@
 \sectskip
-\section{Derivation System}
-
-In the following we describe the local and compositional derivation
-rules of the automatic amortized analysis.
-
-\paraskip
-\paragraph{Cost Aware Clight.}
-
-We present the rules for a subset of the
-expressions of Clight.  Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}, it
-is a cleaner C with a unified looping construct and
-side-effect free expressions.  We reuse most of CompCert's
-syntax but instrument the semantics with a resource metric
-$M$ that accounts for the cost (an arbitrary rational number)
-of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
-evaluating the expression \code{exp}; the rationals $M_f$
-and $M_r$ account respectively for the cost of a call to the
-function $f$ and the cost of returning from it.
-More details on this cost-aware semantics are provided in \pref{sec:soundness}.
-
-In the rules, assignments are restricted to the form $x \gets
-y$ or $x \gets x \pm y$.  In the implementation, a Clight
-program is converted into this form prior to analysis without changing
-the resource cost.  This is achieved by using a series of \emph{cost-free
-  assignments} that do not result in additional cost in the
-semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
-This sound treatment ensures that no further loop bounds
-depend on the result of the non-linear operation.
-
-\paraskip
-\paragraph{Judgements.}
-
-The derivation system for the automatic amortized analysis is defined
-in~\pref{fig:auto}.  The derivation rules derive judgements of the form
-$$
-  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
-    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
-$$
-%
-The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
-can be seen as a quantitative Hoare triple.  All assertions are split into
-two parts, the logical part and the quantitative part.  The
-quantitative part $Q$ represents a potential function as a collection
-of non-negative numbers $q_i$ indexed by the index set $I$.  The
-logical part $\Gamma$ is left abstract but is enforced by our
-derivation system to respect classic Hoare logic constraints.
-%
-The meaning of this basic judgment is as follows: If $S$ is executed
-with starting state $\state$, the assertions in $\Gamma$
-hold, and at least $Q(\state)$ resources are available then the
-evaluation does not run out of resources and, if the execution
-terminates in state $\state'$, there are at least $Q(\state')$
-resources left and $\Gamma'$ holds for $\state'$.
-
-The judgement is a bit more involved since we have to take into
-account the early exit statements \code{break} and \code{return}.
-This is similar to classical Hoare triples in the presence of non-linear
-control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
-that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
-% of looping statements by the rule {\sc Q:Loop} and used before
-% \code{break} statements by the rule {\sc Q:Break}.
-Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
-returning from a function call. \iffull{(This becomes clear
-in the rules.)}{}
-
-% These judgements correspond to the logic judgements $\Delta; B; R
-% \vdash \htriple P S Q$ where each assertion splits in two parts, a
-% logical part $\Gamma$ and a quantitative part $Q$.  That means that
-% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
-% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
-% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
-% simple Hoare triple.  We leave out the function context $\Delta$ in
-% the presentation to avoid cluttering notations.  Instead we assume a
-% fixed function context $\Delta$ that is implicit on all judgements.
-% Function contexts are treated exactly as in the logic.  % We call
-% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
-% % $(\Gamma', Q')$.
-% The correspondence between judgements of the
-% automation and the ones of the quantitative logic is made formal by
-% our soundness proof of the automatic amortized analysis at the end of
-% this section.
-
-As a convention, if $Q$ and $Q'$ are quantitative annotations
-we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
-The notation $Q \pm n$ used in many rules defines a new context $Q'$
-such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
-In all the rules, we have the implicit side condition that all rational
-coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
-and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
-
-\paraskip
-\paragraph{Function Specifications.}
-
-During the analysis, function specifications are quadruples
-$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
-$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
-are instantiated by appropriate variables on call sites.  A distinctive
-feature of our analysis is that it respects the function abstraction:
-when deriving a function specification it generates a set of
-constraints and the above quadruple; once done, the constraint
-set can readily be reused for every call site and the function need
-not be analyzed multiple times.
-%
-Therefore, the derivation rules are parametric in a function context $\Delta$
-that we leave implicit in the rules presented here. More details
-can be found in the extended version.
-
-\paraskip
-\paragraph{Derivation Rules.}
-
-The rules of our derivation system must serve two purposes.  They must
-\emph{attach} potential to certain program variable intervals and use
-this potential, when it is allowed, to \emph{pay} for resource consuming
-operations.  These two purposes are illustrated on the {\sc Q:Skip}
-rule.  This rule reuses its precondition as postcondition, it is
-explained by two facts: first, no resource is consumed by the skip
-operation, thus no potential has to be used to pay for the evaluation;
-second, the program state is not changed by
-the execution of a skip statement, thus all the potential available
-before the execution of the skip statement is still available after.
-\iffull{
-These two observations would also justify a rule with the side-condition
-$Q \ge Q'$, because some potential is always safely lost.  For the sake
-of orthogonality, it is not how we present the rule and this form of
-weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
-described later.}{}
-
-% Most of the rules correspond exactly to the respective rules of the
-% quantitative Hoare logic.
-% \iffull{
-%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
-%   as precondition.  This is sound since evaluating a $\code{skip}$
-%   statement never costs any resources and does not change the program
-%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
-%   quantitative annotation of the precondition is at least $M_t(n)$ to
-%   pay for the resource consumption.  Since the program state is also
-%   unchanged in this case, the logical part $\Gamma$ of the
-%   precondition can soundly be reused in the postcondition.  In the
-%   rule {\sc Q:Break}, where the linear control flow is broken, the
-%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
-%   taken from the break part of the judgement similarly as it is done
-%   in the quantitative logic.
-% }
-% %
-% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
-%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
-%   rules can be found in the extended version.
-% }
-% %
-% The rule {\sc Q:Return} exhibits a slight
-% difference to the quantitative logic.  While we have a function $\Z
-% \to \Assn$ to represent return values in the logic, we assume that a
-% special variable $\ret$ is part of the index set of $Q_R$.  The
-% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
-% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
-% $q_i'=q_i$ otherwise.
-
-The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
-describe how the potential is distributed after a size change of
-a variable.
-\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
-and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
-when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
-Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
-$x \gets x+y$, it applies when we can deduce from the logical context
-$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
-can be applied for decrements and when $y$ is not positive.}
-The rules are all equivalent in the case where $y=0$.
-The rule {\sc Q:Inc} can be applied
-if we cannot find the sign of $y$.  % In $\toolname$,
-% the rules for increments and decrements are combined into one rule\iffull{ where
-% the specialized rules take priority over {\sc Q:Inc}.}{.}
-
-To explain how rules for increment and decrement work, it is
-sufficient to understand the rule {\sc Q:IncP}.  The others follow the
-same idea and are symmetrical. In {\sc Q:IncP}, the program updates a
-variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
-quantitative annotation must be updated to reflect the change of the
-program state.  We write $x'$ for the value of $x$ after the assignment.
-Since $x$ is the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
-$\inter x u$ will get smaller with the update, and if $x' \in \inter x
-u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
-$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
-$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
-guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
-get bigger with the update.  We know that $|\inter v {x'}| \le y +
-|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
-$q_{vx}$ to pay for this change.  The rule ensures this
-only for $v \not\in \mathcal U$ because $x \le v$
-otherwise, and thus $|\inter v x| = 0$.
-
-The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
-loop body one needs to find an invariant potential $Q$ that will pay for
-the iterations.  At each iteration, $M_l$ resources are spent to jump
-back, this explains the postcondition $Q{+}M_l$.  To exit the loop,
-the programmer uses the \code{break} statement, it is the reason why
-the postcondition for the full $\code{loop}~S$ statement $\{ \Gamma';
-Q' \}$ is used as break postcondition in the derivation for $S$.
-
-Another interesting rule is {\sc Q:Call}.  It needs to account for the
-changes to the stack caused by the function call, the arguments/return
-value passing, and the preservation of local variables.  We can sum up
-the main ideas of the rule as follows.
-\itemskip
-\begin{itemize}
-\item The potential in the pre- and postcondition of the function
-  specification is equalized to its matching potential in the callee's
-  pre- and postcondition.
-\itemskipIn
-\item
-  The potential of intervals $|\inter x y|$ is preserved
-  across a function call
-  if $x$ and $y$ are local\iffull{ variables.}{.}
-\item
-  The unknown potentials after the call (e.g.
-  $|\inter x g|$, with $x$ local and $g$ global)
-  are set to zero in the postcondition.
-\end{itemize}
-\itemskip
-If $x$ and $y$ are local variables and $f(x,y)$ is
-called, {\sc Q:Call} splits the potential of
-$|\inter x y|$ in two parts, one part to perform the
-computation in the function $f$ and one part to keep for
-later use after the function call.  This splitting is
-realized by the equations $Q = P {+} S$ and $Q' =
-P' {+} S'$.  Arguments in the function precondition
-$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
-of names different from all program variables.  This
-prevents name conflicts from happening and ensures that
-the substitution $[\Vargs/\vec x]$ is meaningful.
-Symmetrically, we use the unique name $\Vret$ to represent
-the return value in the function's postcondition
-$(\Gamma'_f; Q'_f)$.
-
-The rule {\sc Q:Weak} is not syntax directed. In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition. We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.  The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax} that
-also deals with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq l$ is
-larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
-without gaining potential.
-
-\sectskip
 \section{Automatic Inference via LP Solving}
@@
 SMT-solver--based techniques is that a satisfying assignment is
-a proof certificate instead of a counter example.
+a proof certificate instead of a counter example.  To provide
+high-assurance bounds, this certificate can be checked
+in linear time by a simple validator.
 
@@
 \aftersectskip
-
-\begin{figure}[t]
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    @assert(na > 0);@
-    a[x]=0; @na--;@
-    tick(1); x++; }
-  if (x < N) { a[x]=1; @na++;@ tick(1); }
-  k--;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\end{minipage}
-\caption{Assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
-  tick metric.  The derived bound  is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-
-\begin{figure}
-\vspace{-.3cm}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h@,lg@) {
-  if (h-l > 1) {
-    @assert(lg > 0);@
-    m = l + (h-l)/2;
-    @lg--;@ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h@,lg@);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\caption{Assisted bound derivation using logical state.
-  % We use the tick metric to model the stack consumption of the program.
-  % Before a function call to \code{bsearch}, a tick statement pays for
-  % the space used by stack frame of the function.  The space is recoverd
-  % after the call.
-  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The semi-automatically derived bound is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
 
@@
 $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
-logical variables is selected so that:
+logical variables is selected to satisfy the proposition:
 $$
@@
 
+\begin{figure}[t]
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1) {
+    @assert(na > 0);@
+    a[x]=0; @na--;@
+    tick(1); x++; }
+  if (x < N) { a[x]=1; @na++;@ tick(1); }
+  k--;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\end{minipage}
+\caption{Assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
+  tick metric.  The derived bound  is
+  $2|[0,k]| + |[0,na]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
 Another simple example is given in \pref{fig:xmplbsaux} where
@@
 on standard static-analysis tools.
+
+\begin{figure}
+\vspace{-.3cm}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ lg > \log_2(h - l) \} }$
+bsearch(x,l,h@,lg@) {
+  if (h-l > 1) {
+    @assert(lg > 0);@
+    m = l + (h-l)/2;
+    @lg--;@ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h@,lg@);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\caption{Assisted bound derivation using logical state.
+  % We use the tick metric to model the stack consumption of the program.
+  % Before a function call to \code{bsearch}, a tick statement pays for
+  % the space used by stack frame of the function.  The space is recoverd
+  % after the call.
+  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
+  The semi-automatically derived bound is $|[0,lg]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
 

### 2015-04-13 paper/main.tex
flush last page, add names to TR
avoid repetitions
typos

@@
 \usepackage{array} %for stupid tables
+\usepackage{flushend} %equalize columns length on last page
 %\usepackage[firstpage]{draftwatermark}
@@
 is that they can only associate \emph{potential}
-with individual program variable or data structure.
+with individual program variables or data structures.
 For C programs, this
@@
 $K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
-logical assertions from potential functions by semicolons. Note that the
+logical assertions from potential functions with semicolons. Note that the
 logical assertions are only used in the rule for the assignment
@@
 To find resource bounds automatically, we first need to restrict our
-search space.  In this work, we focus on a the following form of
+search space.  In this work, we focus on the following form of
 potential functions, which can express tight bounds for many typical
@@
 
-In addition to the quantitative annotations our automatic amortized
+In addition to the quantitative annotations, our automatic amortized
 analysis needs to maintain a minimal abstract state to justify certain
@@
 interpretation with the domain of linear inequalities.  We observed that
-the rules of the analysis often requires only minimal local knowledge.
+the rules of the analysis often require only minimal local knowledge.
 This means that it is not necessary for us to compute precise loop
@@
 precisely, it is the partitioning part of the algorithm.  This partition
-loop has linear complexity, and feeding it to our analysis gives a
-worst-case bound $2+3|[l,h]|$.  The bound is not optimal but it can be
+loop has linear complexity, and feeding it to our analysis gives the
+worst-case bound $2+3|[l,h]|$.  This bound is not optimal but it can be
 refined by rewriting the program. % to let our logical inference
@@
 %potential from the same interval $[l,h]$ regardless of their nesting.
-With the potential method in mind, we can reason as follows.
+To understand the bound, we can reason as follows.
 %
@@
 Otherwise, the cost of each round (at
-most 3) is payed using the potential of $[l,h]$ by the first
+most 3) can be payed using the potential of $[l,h]$ by the first
 increment to $l$ because we know that $l<h$.  The two inner loops
@@
 
-We present the rules for a subset of the
-expressions of Clight.  Clight is the first intermediate
+We present the rules for a subset of Clight.
+Clight is the first intermediate
 language of the CompCert compiler~\cite{CompcertCACM}, it
@@
 evaluation does not run out of resources and, if the execution
-terminates in state $\state'$, there are at least $Q(\state')$
+terminates in state $\state'$, there are at least $Q'(\state')$
 resources left and $\Gamma'$ holds for $\state'$.
@@
 a variable.
-\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
+The rule {\sc Q:IncP} is for increments $x \gets x + y$
 and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
@@
 Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
-$x \gets x+y$, it applies when we can deduce from the logical context
-$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
-can be applied for decrements and when $y$ is not positive.}
+that can be applied if $y$ is not positive.
 The rules are all equivalent in the case where $y=0$.
@@
 
-The rule {\sc Q:Weak} is not syntax directed. In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition. We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.  The high-level idea of {\sc Q:Weak} is the following: If
+The rule {\sc Q:Weak} is the only rule not syntax directed. %In the implementation
+%we apply {\sc Q:Weak} before loops and between the two statements of a
+%sequential composition.
+We could integrate weakenings into every syntax directed rule but, for
+the sake of efficiency, the implementation uses a simple heuristic instead.
+The high-level idea of {\sc Q:Weak} is the following: If
 we have a sound judgement, then it is sound to add more potential to the
@@
 \emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax} that
+that is defined in the rule {\sc Relax}. This rule
 also deals with the important task of transferring constant potential
@@
 
-We separate the search of a derivation into two steps.  As a first
+We separate the search of a derivation in two steps.  As a first
 step we go through the functions of the program and apply inductively
@@
 $Q$ from the solver to get a resource bound for the program.  To get
-a full derivation we extract the complete solution from the
-solver, and apply it to the symbolic names $q_i$ of the coefficients
+a full derivation, we extract the complete solution from the
+solver and apply it to the symbolic names $q_i$ of the coefficients
 in the derivation.  If the LP solver fails to find a solution, an
@@
 logical variables is selected to satisfy the proposition:
-$$
+\begin{equation} \tag{*}
 \forall n\, \sigma\, \sigma'.\,
@@
     (\sigma, P) \Downarrow_{n'} \sigma'.
-$$
+\end{equation}
 Since the annotated program and the original one are usually
@@
 that the added assertions never fail in order to show the two programs
-have identical resource consumption.  This can usually be piggybacked
+satisfy the proposition (*).  This can usually be piggybacked
 on standard static-analysis tools.
@@
 then the evaluation does not run out of resources and there are at
-least $Q(\state')$ resources left if the evaluation terminates in
+least $Q'(\state')$ resources left if the evaluation terminates in
 $\state'$.  The logic rules are similar to the ones in previous
@@
 
-Finally, we define a strong compositional continuation based soundness for
+Finally, we define a strong compositional continuation-based soundness for
 triples and prove the validity of all the rules in Coq.  \ifshort{The full
@@
 
-The translation of an assertion $(\Gamma, Q)$ in the automatic
+The translation of an assertion $(\Gamma; Q)$ in the automatic
 analysis is defined by
 $$
-  \tr(\Gamma, Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
+  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
 $$
@@
 decremented down to 100 and in the second loop $i$ is decremented further
-down to $-1$.  However, between the loops we assign \code{i=i+k+50}.
+down to $-1$.  However, between the loops we assign \lstinline{i += k+50}.
 So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
@@
 zero in an unconventional way.  In the loop body, first $x$ is decremented
-by one.  Then the values of the variables $x$ and $y$ are switched using
+by one, then the values of the variables $x$ and $y$ are switched using
 the local variable $t$ as a buffer.  Our analysis infers the tight
@@
 
-Our current implementation of $\toolname$ does not currently support
+Our implementation does not currently support
 all of Clight.  Programs with function pointers, goto statements,

### 2015-04-13 paper/main.tex
read limitations section; clarified the part about non-linear ops

@@
 $\toolname$, it is usually the case for programs that rely on heap
-invariants (like nul-terminated C strings), for programs making
-use of non-linear operations (like \code{\%} or \code{*}), or for
-programs whose
+invariants (like nul-terminated C strings), for programs in which
+resource usage depends on the result of non-linear operations (like
+\code{\%} or \code{*}) in a non-trivial way, or for programs whose
 termination can only be proved by complex path-sensitive reasoning.

### 2015-04-13 paper/main.tex
read through 'Cost Aware Clight'. minor changes

@@
 Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}, it
-is a cleaner C with a unified looping construct and
+language of the CompCert compiler~\cite{CompcertCACM}. It
+is a subset of C with a unified looping construct and
 side-effect free expressions.  We reuse most of CompCert's
@@
 of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
-evaluating the expression \code{exp}; the rationals $M_f$
+evaluating the expression \code{exp}. The rationals $M_f$
 and $M_r$ account respectively for the cost of a call to the

### 2015-04-14 paper/main.tex
minor

@@
 
-This research is based on work supported in part by NSF grants 1319671
-and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
-Any opinions, findings, and conclusions contained in this document are
-those of the authors and do not reflect the views of these agencies.
+We thank members of the FLINT team at Yale and anonymous referees for
+helpful comments and suggestions that improved this paper and the
+implemented tools.  This research is based on work supported in part
+by NSF grants 1319671 and 1065451, DARPA grants FA8750-10-2-0254 and
+FA8750-12-2-0293, and ONR Grant N00014-12-1-0478.  Any opinions,
+findings, and conclusions contained in this document are those of the
+authors and do not reflect the views of these agencies.
 

### 2015-04-14 paper/main.tex
fix overfull hbox in Figure 2
less space tricks
add country to affiliation
typo fix plus more spacing

@@
 
-\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
-\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
-\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
-\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
-\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
-\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
-\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}
+%\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
+%\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
+%\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
+%\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
+%\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
+%\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
+%\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}
+
+\newcommand{\itemskip}[0]{}
+\newcommand{\itemskipIn}[0]{}
+\newcommand{\sectskip}[0]{}
+\newcommand{\paraskip}[0]{}
+\newcommand{\lstskip}[0]{}
+\newcommand{\aftersectskip}[0]{}
+\newcommand{\subsectskip}[0]{}
 
@@
 %SHORT
-\ifx\fullversion\undefined
-\abovedisplayskip=3pt
-\belowdisplayskip=3pt
-\abovedisplayshortskip=3pt
-\belowdisplayshortskip=3pt
-\else
-\fi
+%\ifx\fullversion\undefined
+%\abovedisplayskip=3pt
+%\belowdisplayskip=3pt
+%\abovedisplayshortskip=3pt
+%\belowdisplayshortskip=3pt
+%\else
+%\fi
 %SHORT
@@
 {Specifying and Verifying and Reasoning about Programs}
-\vspace{-3pt}
+%\vspace{-3pt}
 \terms Verification, Reliability
-\vspace{-3pt}
+%\vspace{-3pt}
 \keywords Quantitative Verification, Resource Bound Analysis, Static
@@
 The resource behavior of functions can be summarized
-in a function specification that can be used at every
+in a functional specification that can be used at every
 call site without accessing the function body.
@@
 resource analysis technique for C that deals naturally with
-recursive functions and sequenced loops, and can handle resources that can
+recursive functions and sequenced loops, and can handle resources that may
 become available during execution (e.g., when freeing
@@
   functions, generates resource abstractions for functions, derives
-  proof certificates, and handles resources that can
+  proof certificates, and handles resources that may
   become available during execution.
@@
   \centering
-\vspace{-1.5ex}
+%\vspace{-1.5ex}
 \begin{lstlisting}[mathescape]
@@
 \begin{figure*}
- \setlength{\progwidth}{.22\linewidth}
   \centering
-\vspace{-.3cm}
+%\vspace{-.3cm}
 \hspace{-0.4cm}
-  \begin{minipage}[b]{.18\linewidth}
+  \begin{minipage}[b]{.24\linewidth}
     \begin{center}
@@
    \end{lstlisting}
-\vspace{-2.5ex}
+%\vspace{-2.5ex}
 $|[x, n]| + |[y, m]|$
@@
 %
-  \begin{minipage}[b]{\progwidth}
+  \begin{minipage}[b]{0.27\linewidth}
     \begin{center}
@@
    \end{lstlisting}
-\vspace{-2.5ex}
+%\vspace{-2.5ex}
 $|[x, n]| + |[z, n]|$
@@
 %
-  \begin{minipage}[b]{\progwidth}
+  \begin{minipage}[b]{0.25\linewidth}
     \begin{center}
@@
    \end{lstlisting}
-\vspace{-2.5ex}
+%\vspace{-2.5ex}
 $3.1|[y,z]| + 0.1|[0,y]|$
@@
 %
-  \begin{minipage}[b]{\progwidth}
+  \begin{minipage}[b]{0.22\linewidth}
     \begin{center}
@@
    \end{lstlisting}
-\vspace{-2.5ex}
+%\vspace{-2.5ex}
 $59|[n,0]| {+} 0.05|[0,y]|$
@@
   \centering
-\ifshort{\vspace{-.3cm}
+\ifshort{%\vspace{-.3cm}
 \hspace{-0.4cm}}{}
@@
    \end{lstlisting}
-\vspace{-2.7ex}
+%\vspace{-2.7ex}
 $0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
@@
    \end{lstlisting}
-\vspace{-2.8ex}
+%\vspace{-2.8ex}
 $\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
@@
 \small
-\vspace{-.4cm}
+%\vspace{-.4cm}
 \def \MathparLineskip {\lineskip=0.35cm}
@@
 \begin{figure*}[t]
-\vspace{-.5cm}
+%\vspace{-.5cm}
 \centering
@@
 }
-\vspace{-.3cm}
+%\vspace{-.3cm}
 \end{mathpar}
@@
 \begin{tabular}{l}Constraints:\end{tabular}
-\vspace{-.2cm}
+%\vspace{-.2cm}
 $$
@@
 \end{lstlisting}
-\vspace{-.2cm}
+%\vspace{-.2cm}
 \end{minipage}
@@
 \begin{figure}
-\vspace{-.3cm}
+%\vspace{-.3cm}
 \begin{lstlisting}[numbers=left]
@@
 \end{lstlisting}
-\vspace{-.2cm}
+%\vspace{-.2cm}
 \caption{Assisted bound derivation using logical state.
@@
 \centering
-\vspace{-.5cm}
+%\vspace{-.5cm}
 \begin{tabular}{r|c|c|c|c|c}
@@
    \end{lstlisting}
-\vspace{-1.2cm}
+%\vspace{-1.2cm}
 \hspace{3.8cm}
@@
     \end{center}
-\vspace{-0.1cm}
+%\vspace{-0.1cm}
   \end{minipage}
@@
 
-Since we have a formal cost semantics, we can run our examples with
-this semantics for different inputs and measure the cost to compare it
+With our formal cost semantics, we can run our examples 
+for different inputs and measure the cost to compare it
 to our derived bound.  \pref{fig:3d} shows such a comparison for
@@
 \center
-% \vspace{-.2cm}
+\vspace{-.2cm}
 \includegraphics[width=.9\linewidth]{fig/bound3d}
@@
 analysis for C that can derive parametric bounds for loops and
-recursive functions fully automatically.
-%
-In this article, we also use a more general quantitative Hoare
-logic that is parametric in the resource of interest.
+recursive functions fully automatically. We
+use a more general quantitative Hoare
+logic that is parametric over the resource of interest.
 
@@
 and LOOPUS.  As demonstrated, $\toolname$ is more compositional than
-the aforementioned tools. Our technique, is the only one that is able
-to generate resource specifications for functions, deal with resources
-like memory that might become available, generates proof certificates
-for the bounds, and has support for user guidance that separates
+the aforementioned tools. Our technique, is the only one that can
+generate resource specifications for functions, deal with resources
+like memory that might become available, generate proof certificates
+for the bounds, and  support user guidance that separates
 qualitative and quantitative reasoning.
@@
 framework combines ideas from existing abstract interpretation--based
-techniques with the potential method of amortized analysis.  It is implemented in the publicly available tool
-$\toolname$. To the best of our knowledge, $\toolname$ is the first
-analysis tool for C programs that automatically reduces the derivation
-of symbolic bounds to LP solving.
+techniques with the potential method of amortized analysis.  It is
+implemented in the publicly available tool $\toolname$. To the best of
+our knowledge, $\toolname$ is the first tool for C programs
+that automatically reduces the derivation of symbolic bounds to LP
+solving.
 

### 2015-04-14 paper/main.tex
fix wrong limitation

@@
 {Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
-{Yale University}
+{Yale University, USA}
 {\{quentin.carbonneaux, jan.hoffmann, zhong.shao\}@yale.edu}
@@
 all of Clight.  Programs with function pointers, goto statements,
-continue statements, and dereferences of stack-allocated variables
+continue statements, and pointers to stack-allocated variables
 cannot be analyzed automatically.  While these limitations concern the

### 2015-04-14 paper/main.tex
add large table
mention the large table in section 8
trying to save some space, not enough

@@
 \usepackage{flushend} %equalize columns length on last page
+\usepackage{hhline}
 %\usepackage[firstpage]{draftwatermark}
@@
 procedure described above, we must show that the extra assertion
-\lstinline{na > 0} on line 5 is always true.  It
-is done by proving inductively that $na = \#_1(a)$
+\lstinline{na > 0} on line 5 cannot fail.  We do it
+by proving inductively that $na = \#_1(a)$
 and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails.
+$\#_1(a) {>} 0$, thus the assertion \lstinline{na > 0} never fails.
 
@@
 and the original one, we use the inequality $lg >
-\log_2(h-l)$ as an invariant.  This technique allows a simpler
+\log_2(h-l)$ as invariant.  This allows a simpler
 proof because, when working with integer arithmetic, it is not always
@@
 of the automatic analysis directly to its counterpart in the
-quantitative logic.  The most tricky parts are the translations of the
+quantitative logic.  The trickiest parts are the translations of the
 rules for increments and decrements and the rule {\sc Q:Weak} for
 weakening because they make essential use of the algebraic properties
-of our linear potential functions.
+of the potential functions.
 \iffull{ For instance we prove the following lemma to show the soundness
@@
 
-\pref{tab:compar} summarizes the results of our experiments.  It shows
+\pref{tab:compar} summarizes the results of our experiments presented
+in \pref{tab:eval}.  It shows
 for each tool the number of derived bounds (\#bounds), the number of
@@
 
+\paragraph{Well-Known Algorithms}
+
+\pref{fig:cat3} shows well-known algorithms that can be automatically
+analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
+the search function of the Knuth-Morris-Pratt algorithm for string
+search.  To derive a bound for this algorithm we have add an assertion
+that indicates the bounds of the values that are stored in the array
+\code{b}.  Using this information we derive a tight linear bound.
+
+Example \emph{Greatest Common Divisor} is the usual implementation of
+the GCD algorithm.  We automatically derive a linear bound.  Note that
+the two tests at the beginning that check if the inputs are positive
+are essential.  Finally, Example \emph{Quick Sort} shows a skeleton of
+the quick sort sorting algorithm.  Since we can only derive linear
+bounds we left out the inner loop that swaps the array elements and
+determines the position \code{n+lo} of the pivot.  We only assert that
+\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
+
+}\fi
+
 \begin{table*}
@@
 \renewcommand{\max}[0]{{\rm mx}}
-\renewcommand{\min}[0]{{\rm mn}}
 
@@
 File &
-\multicolumn{2}{|c|}{KoAT} &
-\multicolumn{2}{|c|}{Rank} &
-\multicolumn{2}{|c|}{LOOPUS} &
+\multicolumn{2}{c|}{KoAT} &
+\multicolumn{2}{c|}{Rank} &
+\multicolumn{2}{c|}{LOOPUS} &
 SPEED &
@@
 \\
-\hline
 
 %--------------------------------------------
-\hline \texttt{gcd.c} &
+\hhline{=|==|==|==|=|=} \texttt{gcd.c} &
 
@@
   Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
-  marked with --- indicate that the tool failed to produce a result.
+  marked with --- indicate that the tool failed to produce a result. We write
+  $\max(a,b)$ for the maximum of $a$ and $b$.
 }
@@
 \end{table*}
-
-\paragraph{Well-Known Algorithms}
-
-\pref{fig:cat3} shows well-known algorithms that can be automatically
-analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
-the search function of the Knuth-Morris-Pratt algorithm for string
-search.  To derive a bound for this algorithm we have add an assertion
-that indicates the bounds of the values that are stored in the array
-\code{b}.  Using this information we derive a tight linear bound.
-
-Example \emph{Greatest Common Divisor} is the usual implementation of
-the GCD algorithm.  We automatically derive a linear bound.  Note that
-the two tests at the beginning that check if the inputs are positive
-are essential.  Finally, Example \emph{Quick Sort} shows a skeleton of
-the quick sort sorting algorithm.  Since we can only derive linear
-bounds we left out the inner loop that swaps the array elements and
-determines the position \code{n+lo} of the pivot.  We only assert that
-\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
-
-}\fi
 

### 2015-04-14 paper/main.tex
bad space tricks on page 10, please check
moved the table to the main body
move table captions on top, trade some space tricks

@@
 \begin{table}[t]
+\nocaptionrule \caption{Comparison of $\toolname$ with other
+  automatic tools.}
+\vspace{6pt}
+\label{tab:compar}
 \small
@@
 \end{tabular}
-\vspace{.01cm}
-\caption{Comparison of $\toolname$ with other
-  automatic tools.}
-\label{tab:compar}
-\end{table}
-
-\begin{table}[t]
-\centering
-\small
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{MD5Final} & 195 & 141 & 0.22 \\
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-% \code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
-\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
-\hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-\end{tabular}
-\vspace{.1cm}
-\caption{Derived bounds for functions from cBench.}
-\label{tab:cbench}
+\vspace{-6pt}
 \end{table}
@@
 on every tested example.}{}
+
+\begin{table*}
+\nocaptionrule \caption{
+  Experimental evaluation comparing the bounds generated KoAT, Rank,
+  LOOPUS, SPEED, and our automatic amortized analysis on several
+  challenging linear examples. Results for KoAT and SPEED were extracted
+  from previous publications because KoAT cannot take C programs as
+  input in its current version and SPEED is not available.
+  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
+  marked with --- indicate that the tool failed to produce a result. We write
+  $\max(a,b)$ for the maximum of $a$ and $b$.
+}
+\vspace{6pt}
+\label{tab:eval}
+\footnotesize
+
+\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
+\newcommand{\file}[2][c]{%
+  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
+\renewcommand{\max}[0]{{\rm mx}}
+
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\centering
+\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
+
+File &
+\multicolumn{2}{c|}{KoAT} &
+\multicolumn{2}{c|}{Rank} &
+\multicolumn{2}{c|}{LOOPUS} &
+SPEED &
+\toolname
+\\
+
+%--------------------------------------------
+\hhline{=|==|==|==|=|=} \texttt{gcd.c} &
+
+? &
+&
+
+% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
+$(((2{+}1)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$|\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{kmp.c} &
+
+? &
+&
+
+% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
+$(((2{+}(n{+} \dots$ &
+$O(n^2)$ &
+
+% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
+$\max(n, 0) \dots$ &
+$O(n)$ &
+
+? &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \texttt{qsort.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$1 {+} 2 |\inter 0 {\code{len}}|$
+\\
+
+%--------------------------------------------
+% \hline \file{\dir{speed\_pldi09}{fig1.c}} &
+%
+% ? &
+% &
+%
+% --- &
+% &
+%
+% $2 \max(n, 0)$ &
+% $O(n)$ &
+%
+% God damn it &
+%
+% $1 + 2 |\inter 0 n|$
+% \\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_2.c}} &
+
+--- &
+&
+
+% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}n)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + n$ &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_4.c}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + m$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_5.c}} &
+
+$28d + 7g + 27$ &
+$O(n)$ &
+
+% $((((1+1)+(-1+n))+m)+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\max(n, n-m)$ &
+
+---
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi10}{ex1.c}} &
+
+--- &
+&
+
+--- &
+&
+
+--- &
+&
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi10}{ex3.c}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi10}{ex4.c}} &
+
+$110 a + 33$ &
+$O(n)$ &
+
+--- &
+&
+
+--- &
+&
+
+$n + 1$ &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{fig2\_1.c}} &
+
+$9a + 9b + \dots$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{+}((-y\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, m{-}y)$ &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{fig2\_2.c}} &
+
+$6a + 9b + 3c + 5$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
+$((2{-}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
+$\max(0, (x + 1 {-}z)\dots$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, n{-}z)$ &
+
+$|\inter x n| {+} |\inter z n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{nstd\_multiple.c}} &
+
+--- &
+&
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
+$((2{-}x{+}n\dots$ &
+$O(n^2)$ &
+
+$\max(0, m {-} y) + \max(0, n {-} x)$ &
+$O(n)$ &
+
+$\max(0, m{-}y) + \max(0, n{-}x)$ &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{nstd\_single.c}} &
+
+$48 b + 16$ &
+$O(n)$ &
+
+% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
+$(((1{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(0, \! n {-} 1)\dots$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{sqntl\_single.c}} &
+
+$21 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{}-x{+}n\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{smpl\_multiple.c}} &
+
+$9 c + 10 d + 7$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$n + m$ &
+
+$|\inter 0 m| {+} |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{smpl\_single2.c}} &
+
+$20 d + 12 c + 17$ &
+$O(n)$ &
+
+--- &
+&
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$n + m$ &
+
+$|\inter 0 n| {+} |\inter 0 m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{smpl\_single.c}} &
+
+$4 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\texttt{t07.c}} &
+
+? &
+&
+
+$2+x$ &
+$O(n)$ &
+
+% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
+$\max(x, 0)\dots$ &
+$O(n)$ &
+
+? &
+
+$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t08.c} &
+
+? &
+&
+
+% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
+$((2{+}z{-}y\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+$\max(0, \! y{-}2)\dots$ &
+$O(n)$ &
+
+? &
+
+$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t10.c} &
+
+? &
+&
+
+% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
+$((2{-}y{+}x\dots$ &
+$O(n)$ &
+
+$\max(0, x {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t11.c} &
+
+? &
+&
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x)) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t13.c} &
+
+? &
+&
+
+% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
+$(((1{+}y^2/2\dots$ &
+$O(n^2)$ &
+
+$2{\cdot}\max(x, 0) + \max(y, 0)$ &
+$O(n)$ &
+
+? &
+
+$2 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t15.c} &
+
+? &
+&
+
+% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
+$((1{+}x\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$|\inter 0 x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t16.c} &
+
+? &
+&
+
+% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
+$((-99{\cdot}y\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$101 |\inter 0 x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t19.c} &
+
+? &
+&
+
+% $(((1+1)+(151+k))+(-100+i))+1$ &
+$((153{+}k\dots$ &
+$O(n)$ &
+
+$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
+$O(n)$ &
+
+? &
+
+$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t20.c} &
+
+? &
+&
+
+% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
+$(2{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
+$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter x y| {+} |\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t27.c} &
+
+? &
+&
+
+--- &
+&
+
+% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
+$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
+$O(n)$ &
+
+? &
+
+$0.01|\inter n y| {+} 11 |\inter n 0|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t28.c} &
+
+? &
+&
+
+% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
+$((1{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
+$10^3 \,\max(0, x - y)\dots$ &
+$O(n)$ &
+
+? &
+
+$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t30.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$|\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t37.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t39.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$1.33 {+} 0.67 |\inter z y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t46.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$|\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t47.c} &
+
+? &
+&
+
+$4+n$ &
+$O(n)$ &
+
+$1 + \max(n, 0)$ &
+$O(n)$ &
+
+? &
+
+$1 {+} |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\end{tabular}
+\end{table*}
 
@@
 
-\begin{figure}[t]
-\center
-\vspace{-.2cm}
-\includegraphics[width=.9\linewidth]{fig/bound3d}
-\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for Example
-  \emph{t08}. For $x\ge 0$ the bound is tight.}
-\label{fig:3d}
-\end{figure}
+\begin{table}[t]
+\nocaptionrule \caption{Derived bounds for functions from cBench.}
+\vspace{6pt}
+\label{tab:cbench}
+\centering
+\small
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{MD5Final} & 195 & 141 & 0.22 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+% \code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
+\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
+\hline
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
+\end{tabular}
+\vspace{-4pt}
+\end{table}
 
@@
 integers.
+
+\begin{figure}[t]
+\center
+\vspace{-.2cm}
+\includegraphics[width=.9\linewidth]{fig/bound3d}
+\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
+  (blue lines) and the measured runtime cost (red crosses) for Example
+  \emph{t08}. For $x\ge 0$ the bound is tight.}
+\label{fig:3d}
+\vspace{-10pt}
+\end{figure}
 
@@
 
-\begin{table*}
-\footnotesize
-
-\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
-\newcommand{\file}[2][c]{%
-  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
-\renewcommand{\max}[0]{{\rm mx}}
-
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\centering
-\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
-
-File &
-\multicolumn{2}{c|}{KoAT} &
-\multicolumn{2}{c|}{Rank} &
-\multicolumn{2}{c|}{LOOPUS} &
-SPEED &
-\toolname
-\\
-
-%--------------------------------------------
-\hhline{=|==|==|==|=|=} \texttt{gcd.c} &
-
-? &
-&
-
-% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
-$(((2{+}1)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{kmp.c} &
-
-? &
-&
-
-% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
-$(((2{+}(n{+} \dots$ &
-$O(n^2)$ &
-
-% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
-$\max(n, 0) \dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \texttt{qsort.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1 {+} 2 |\inter 0 {\code{len}}|$
-\\
-
-%--------------------------------------------
-% \hline \file{\dir{speed\_pldi09}{fig1.c}} &
-%
-% ? &
-% &
-%
-% --- &
-% &
-%
-% $2 \max(n, 0)$ &
-% $O(n)$ &
-%
-% God damn it &
-%
-% $1 + 2 |\inter 0 n|$
-% \\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_2.c}} &
-
---- &
-&
-
-% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}n)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + n$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_4.c}} &
-
---- &
-&
-
-% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + m$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_5.c}} &
-
-$28d + 7g + 27$ &
-$O(n)$ &
-
-% $((((1+1)+(-1+n))+m)+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, n-m)$ &
-
----
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex1.c}} &
-
---- &
-&
-
---- &
-&
-
---- &
-&
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex3.c}} &
-
---- &
-&
-
-% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex4.c}} &
-
-$110 a + 33$ &
-$O(n)$ &
-
---- &
-&
-
---- &
-&
-
-$n + 1$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{fig2\_1.c}} &
-
-$9a + 9b + \dots$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{+}((-y\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, m{-}y)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{fig2\_2.c}} &
-
-$6a + 9b + 3c + 5$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
-$((2{-}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
-$\max(0, (x + 1 {-}z)\dots$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, n{-}z)$ &
-
-$|\inter x n| {+} |\inter z n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{nstd\_multiple.c}} &
-
---- &
-&
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n^2)$ &
-
-$\max(0, m {-} y) + \max(0, n {-} x)$ &
-$O(n)$ &
-
-$\max(0, m{-}y) + \max(0, n{-}x)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{nstd\_single.c}} &
-
-$48 b + 16$ &
-$O(n)$ &
-
-% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
-$(((1{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(0, \! n {-} 1)\dots$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{sqntl\_single.c}} &
-
-$21 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{}-x{+}n\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_multiple.c}} &
-
-$9 c + 10 d + 7$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 m| {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_single2.c}} &
-
-$20 d + 12 c + 17$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 n| {+} |\inter 0 m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_single.c}} &
-
-$4 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\texttt{t07.c}} &
-
-? &
-&
-
-$2+x$ &
-$O(n)$ &
-
-% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
-$\max(x, 0)\dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t08.c} &
-
-? &
-&
-
-% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
-$((2{+}z{-}y\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
-$\max(0, \! y{-}2)\dots$ &
-$O(n)$ &
-
-? &
-
-$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t10.c} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
-$((2{-}y{+}x\dots$ &
-$O(n)$ &
-
-$\max(0, x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t11.c} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x)) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t13.c} &
-
-? &
-&
-
-% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
-$(((1{+}y^2/2\dots$ &
-$O(n^2)$ &
-
-$2{\cdot}\max(x, 0) + \max(y, 0)$ &
-$O(n)$ &
-
-? &
-
-$2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t15.c} &
-
-? &
-&
-
-% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
-$((1{+}x\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t16.c} &
-
-? &
-&
-
-% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
-$((-99{\cdot}y\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$101 |\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t19.c} &
-
-? &
-&
-
-% $(((1+1)+(151+k))+(-100+i))+1$ &
-$((153{+}k\dots$ &
-$O(n)$ &
-
-$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
-$O(n)$ &
-
-? &
-
-$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t20.c} &
-
-? &
-&
-
-% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
-$(2{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
-$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x y| {+} |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t27.c} &
-
-? &
-&
-
---- &
-&
-
-% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
-$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
-$O(n)$ &
-
-? &
-
-$0.01|\inter n y| {+} 11 |\inter n 0|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t28.c} &
-
-? &
-&
-
-% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
-$((1{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
-$10^3 \,\max(0, x - y)\dots$ &
-$O(n)$ &
-
-? &
-
-$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t30.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t37.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t39.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1.33 {+} 0.67 |\inter z y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t46.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t47.c} &
-
-? &
-&
-
-$4+n$ &
-$O(n)$ &
-
-$1 + \max(n, 0)$ &
-$O(n)$ &
-
-? &
-
-$1 {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\end{tabular}
-\caption{
-  Experimental evaluation comparing the bounds generated KoAT, Rank,
-  LOOPUS, SPEED, and our automatic amortized analysis on several
-  challenging linear examples. Results for KoAT and SPEED were extracted
-  from previous publications because KoAT cannot take C programs as
-  input in its current version and SPEED is not available.
-  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
-  marked with --- indicate that the tool failed to produce a result. We write
-  $\max(a,b)$ for the maximum of $a$ and $b$.
-}
-\label{tab:eval}
-\end{table*}
-
 \bibliographystyle{abbrvnat}

### 2015-04-15 paper/main.tex
move renewcommand{\max} before Table 2 caption
pass through derivation Rules

@@
 rule.  This rule reuses its precondition as postcondition, it is
-explained by two facts: first, no resource is consumed by the skip
-operation, thus no potential has to be used to pay for the evaluation;
-second, the program state is not changed by
-the execution of a skip statement, thus all the potential available
+explained by two facts: First, no resource is consumed by the skip
+operation, thus no potential has to be used to pay for the evaluation.
+Second, the program state is not changed by
+the execution of a skip statement. Thus all potential available
 before the execution of the skip statement is still available after.
@@
 The rule {\sc Q:IncP} is for increments $x \gets x + y$
-and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
+and {\sc Q:DecP} is for decrements $x \gets x - y$. They both apply only
 when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
-Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-that can be applied if $y$ is not positive.
+Of course, there are symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
+(not presented here) that can be applied if $y$ is negative.
 The rules are all equivalent in the case where $y=0$.
@@
 
-The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
-loop body one needs to find an invariant potential $Q$ that will pay for
-the iterations.  At each iteration, $M_l$ resources are spent to jump
-back, this explains the postcondition $Q{+}M_l$.  To exit the loop,
-the programmer uses the \code{break} statement, it is the reason why
-the postcondition for the full $\code{loop}~S$ statement $\{ \Gamma';
-Q' \}$ is used as break postcondition in the derivation for $S$.
+The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it
+on a loop body, one needs to find an invariant potential $Q$ that will
+pay for the iterations.  At each iteration, $M_l$ resources are spent
+to jump back. This explains the postcondition $Q{+}M_l$. Since the
+loop can only be exited with a \code{break} statement, the
+postcondition $\{ \Gamma'; Q' \}$ for the statement $\code{loop}~S$ is
+used as break postcondition in the derivation for $S$.
 
@@
 called, {\sc Q:Call} splits the potential of
-$|\inter x y|$ in two parts, one part to perform the
+$|\inter x y|$ in two parts. One part to perform the
 computation in the function $f$ and one part to keep for
@@
 of names different from all program variables.  This
-prevents name conflicts from happening and ensures that
+prevents name conflicts and ensures that
 the substitution $[\Vargs/\vec x]$ is meaningful.
@@
 
-The rule {\sc Q:Weak} is the only rule not syntax directed. %In the implementation
+The rule {\sc Q:Weak} is the only rule that is not syntax directed. %In the implementation
 %we apply {\sc Q:Weak} before loops and between the two statements of a
@@
 (represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq l$ is
-larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
+from the logical context that the interval size $|\inter x y| \geq \ell$ is
+larger than a constant $\ell$ then we can turn the potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval into the constant potential $\ell{\cdot}q_{xy}$
 and guarantee that we do not gain potential.  Conversely, if $|\inter
@@
 \begin{table*}
+\renewcommand{\max}[0]{{\rm mx}}
 \nocaptionrule \caption{
@@
   \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
-\renewcommand{\max}[0]{{\rm mx}}
 

### 2015-04-15 paper/main.tex
reordered the figures to fit page limit
edited caption of the big table

@@
 
+\paraskip
+\paragraph{Challenging Loops.}
+
+One might think that our set of potential functions is too simplistic
+to be able to express and prove bounds for realistic
+programs. Nevertheless, we can handle challenging example programs
+without special tricks or techniques.  Examples \emph{speed\_1} and
+\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
+demonstrate that our method can handle \emph{tricky iteration
+  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
+bounds as our analysis but requires heuristics for its counter
+instrumentation.  These loops can also be handled with inference of
+\emph{disjunctive invariants}, but in the abstract interpretation
+community, these invariants are known to be notoriously difficult to
+generate.
+%
+In Example \emph{speed\_1} we have one loop that first increments
+variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
+derive the tight bound $|[x, n]| + |[y, m]|$.
+%
+Example \emph{speed\_2} is even trickier, and we found it hard to
+find a bound manually.  However, using potential transfer reasoning as
+in amortized analysis, it is easy to prove the tight bound
+$|[x, n]| + |[z, n]|$.
+
+\paraskip
+\paragraph{Nested and Sequenced Loops.}
+
+Example \emph{t08a} in Figure~\ref{fig:ex_list} shows the ability of the analysis to discover
+interaction between \emph{sequenced loops} through size change of
+variables.  We accurately track the size change of $y$ in the first
+loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
+Furthermore, \emph{t08a} shows again that we do not handle the
+constants $1$ or $0$ in any special way.  In all examples we could
+replace $0$ and $1$ with other constants like in the second loop and
+still derive a tight bound.  \iffull{The only information, that the analyzer
+needs is $y \geq c$ before assigning $y = y - c$.}{}
+%
+Example \emph{t27} in Figure~\ref{fig:ex_list}
+shows how amortization can be used to handle
+\emph{interacting nested loops}.  In the outer loop we increment the
+variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
+increment the variable $y$ by $1000$.  Then we non-deterministically
+(expressed by \code{*}) execute an inner loop that decrements $y$ by
+$100$ until $y<100$.  The analysis discovers that only the first
+execution of the inner loop depends on the initial value of $y$.  We
+again derive tight constant factors.
+%
+
+\paraskip
+\paragraph{Mutually Recursive Functions.}
+
+As mentioned, the analysis also handles advanced control flow like
+\code{break} and \code{return} statements, and mutual recursion.
+Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
+automatically derived tick bounds.  The function \code{c\_down}
+decrements its first argument $x$ until it reaches the second argument
+$y$. It then recursively calls the function \code{c\_up},
+which is dual to \code{c\_down}.  Here, we
+count up $y$ by $2$ and call \code{c\_down}.  $\toolname$ is
+the only available system that computes a tight
+bound.  \iffull{The analysis amounts to computing the meeting
+point of two trains that approach each other with different speeds.}{}
+% This is comparable to the
+% determination of the point at which to trains meet if they approach
+% each other with different speeds.
+
 \begin{figure*}
@@
 \paraskip
-\paragraph{Challenging Loops.}
-
-One might think that our set of potential functions is too simplistic
-to be able to express and prove bounds for realistic
-programs. Nevertheless, we can handle challenging example programs
-without special tricks or techniques.  Examples \emph{speed\_1} and
-\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
-demonstrate that our method can handle \emph{tricky iteration
-  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
-bounds as our analysis but requires heuristics for its counter
-instrumentation.  These loops can also be handled with inference of
-\emph{disjunctive invariants}, but in the abstract interpretation
-community, these invariants are known to be notoriously difficult to
-generate.
-%
-In Example \emph{speed\_1} we have one loop that first increments
-variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
-derive the tight bound $|[x, n]| + |[y, m]|$.
-%
-Example \emph{speed\_2} is even trickier, and we found it hard to
-find a bound manually.  However, using potential transfer reasoning as
-in amortized analysis, it is easy to prove the tight bound
-$|[x, n]| + |[z, n]|$.
-
-\paraskip
-\paragraph{Nested and Sequenced Loops.}
-
-Example \emph{t08a} in Figure~\ref{fig:ex_list} shows the ability of the analysis to discover
-interaction between \emph{sequenced loops} through size change of
-variables.  We accurately track the size change of $y$ in the first
-loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
-Furthermore, \emph{t08a} shows again that we do not handle the
-constants $1$ or $0$ in any special way.  In all examples we could
-replace $0$ and $1$ with other constants like in the second loop and
-still derive a tight bound.  \iffull{The only information, that the analyzer
-needs is $y \geq c$ before assigning $y = y - c$.}{}
-%
-Example \emph{t27} in Figure~\ref{fig:ex_list}
-shows how amortization can be used to handle
-\emph{interacting nested loops}.  In the outer loop we increment the
-variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
-increment the variable $y$ by $1000$.  Then we non-deterministically
-(expressed by \code{*}) execute an inner loop that decrements $y$ by
-$100$ until $y<100$.  The analysis discovers that only the first
-execution of the inner loop depends on the initial value of $y$.  We
-again derive tight constant factors.
-%
-
-\paraskip
-\paragraph{Mutually Recursive Functions.}
-
-As mentioned, the analysis also handles advanced control flow like
-\code{break} and \code{return} statements, and mutual recursion.
-Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
-automatically derived tick bounds.  The function \code{c\_down}
-decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{c\_up},
-which is dual to \code{c\_down}.  Here, we
-count up $y$ by $2$ and call \code{c\_down}.  $\toolname$ is
-the only available system that computes a tight
-bound.  \iffull{The analysis amounts to computing the meeting
-point of two trains that approach each other with different speeds.}{}
-% This is comparable to the
-% determination of the point at which to trains meet if they approach
-% each other with different speeds.
-
-\ifshort{
-\begin{figure*}
-\small
-%\vspace{-.4cm}
-\def \MathparLineskip {\lineskip=0.35cm}
-\begin{mathpar}
-\Rule{Q:Skip}
-{ }
-{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
-%
-\and \Rule{Q:Assert}
-{ }
-{ B, R \vdash \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
-%
-\and \Rule{Q:Tick}
-{ }
-{ B, R \vdash \htriple{ \Gamma; Q {+} M_t(n) }{ \code{tick}(n) }{ \Gamma; Q } }
-%
-\and \Rule{Q:Break}
-{ }
-{(\Gamma; Q_B), R \vdash \htriple{ \Gamma; Q_B {+} M_b }{ \code{break} }{ \Gamma'; Q' } }
-%
-\and \Rule{Q:Return}
-{ P = Q_R[\Vret/x]
-\\ \Gamma = \Gamma_R[\Vret/x]
-\\ \forall i \in \dom{P} . \, p_i = q_i
-}
-{ B, (\Gamma_R; Q_R) \vdash \htriple{ \Gamma; Q }{ \code{return}~x }{ \Gamma'; Q' } }
-%
-\and \Rule{Q:Loop}
-{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
-{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
-%
-\and \Rule{Q:Seq}
-{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
-\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
-}
-{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
-%
-\and \Rule{Q:If}
-{ B, R \vdash \htriple{ \Gamma {\land} e; Q {-} M_c^1) }{ S_1 }{ \Gamma'; Q' }
-\\\\ B, R \vdash \htriple{ \Gamma {\land} \neg e; Q {-} M_c^2 }{ S_2 }{ \Gamma'; Q' }
-}
-{ B, R \vdash \htriple{ \Gamma; Q{+}M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ \Gamma'; Q' } }
-%
-\\ \Rule{Q:IncP}
-{ \Gamma \models y \ge 0 \!
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx} \!
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
-}
-%
-\and \Rule{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
-}
-%
-\and \Rule{Q:Set}
-{ q'_{xy}, q'_{yx} \in \Qplusz
-\\\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B, R \vdash \htriple{ \Gamma[x/y]; Q {+} M_u {+} M_e(y) }{ x \gets y }{ \Gamma; Q' } }
-%
-\and \Rule{Q:DecP}
-{ \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {-} y]; Q {+} M_u {+} M_e(x {-} y) }{ x \gets x - y }{ \Gamma; Q' }
-}
-%
-\and\Rule{Q:Call}
-{ (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
-\\ \text{Loc} = \text{Locals}(Q) \!
-\\ \forall i \neq j . \, x_i \neq x_j \!
-\\ c \in \Qplusz \!
-\\ Q = P + S \!
-\\ Q' = P' + S
-\\ U = Q_f[\Vargs / \vec x]
-\\ U' = Q'_f[\Vret/r]
-\\ \forall i \in \dom{U} . \, p_i = u_i
-\\ \forall i \in \dom{U'} .\, p'_i = u'_i
-\\ \forall i \not\in \dom{U'} .\, p'_i = 0
-\\ \forall i \not\in \text{Loc}
-  .\, s_i = 0
-}
-{ B, R \vdash \htriple
-  { \Gamma_f[\Vargs / \vec x] {\land} \Gamma_{\text{Loc}}; Q {+} c {+} M_f }
-  { r \gets f(\vec x) }
-  { \Gamma_f'[\Vret/r] {\land} \Gamma_{\text{Loc}}; Q' {+} c {-} M_r }
-}
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
-{ \fenv f = (\vec y, S_f)
-\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{ \Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y] }{ S_f }{ \Gamma'; Q' }
-}
-{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
-%
-\and \Rule{Q:Weak}
-{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
-\\ \!\! \Gamma_1 \models \Gamma_2
-\\\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ \Gamma'_2 \models \Gamma'_1
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-}
-{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
-%
-\\ \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
-\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
-\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
-\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
-}
-{ Q' \succeq_\Gamma Q }
-%\vspace{-.3cm}
-\end{mathpar}
-\caption{Inference rules of the quantitative analysis.}
-\label{fig:auto}
-\end{figure*}
-}{}
-
-\paraskip
 \paragraph{Compositionality.}
@@
 and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
+
+\ifshort{
+\begin{figure*}
+\small
+%\vspace{-.4cm}
+\def \MathparLineskip {\lineskip=0.35cm}
+\begin{mathpar}
+\Rule{Q:Skip}
+{ }
+{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
+%
+\and \Rule{Q:Assert}
+{ }
+{ B, R \vdash \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
+%
+\and \Rule{Q:Tick}
+{ }
+{ B, R \vdash \htriple{ \Gamma; Q {+} M_t(n) }{ \code{tick}(n) }{ \Gamma; Q } }
+%
+\and \Rule{Q:Break}
+{ }
+{(\Gamma; Q_B), R \vdash \htriple{ \Gamma; Q_B {+} M_b }{ \code{break} }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Return}
+{ P = Q_R[\Vret/x]
+\\ \Gamma = \Gamma_R[\Vret/x]
+\\ \forall i \in \dom{P} . \, p_i = q_i
+}
+{ B, (\Gamma_R; Q_R) \vdash \htriple{ \Gamma; Q }{ \code{return}~x }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Loop}
+{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
+{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Seq}
+{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
+\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
+%
+\and \Rule{Q:If}
+{ B, R \vdash \htriple{ \Gamma {\land} e; Q {-} M_c^1) }{ S_1 }{ \Gamma'; Q' }
+\\\\ B, R \vdash \htriple{ \Gamma {\land} \neg e; Q {-} M_c^2 }{ S_2 }{ \Gamma'; Q' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q{+}M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ \Gamma'; Q' } }
+%
+\\ \Rule{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
+}
+%
+\and \Rule{Q:Inc}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
+}
+%
+\and \Rule{Q:Set}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ B, R \vdash \htriple{ \Gamma[x/y]; Q {+} M_u {+} M_e(y) }{ x \gets y }{ \Gamma; Q' } }
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {-} y]; Q {+} M_u {+} M_e(x {-} y) }{ x \gets x - y }{ \Gamma; Q' }
+}
+%
+\and\Rule{Q:Call}
+{ (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
+\\ \text{Loc} = \text{Locals}(Q) \!
+\\ \forall i \neq j . \, x_i \neq x_j \!
+\\ c \in \Qplusz \!
+\\ Q = P + S \!
+\\ Q' = P' + S
+\\ U = Q_f[\Vargs / \vec x]
+\\ U' = Q'_f[\Vret/r]
+\\ \forall i \in \dom{U} . \, p_i = u_i
+\\ \forall i \in \dom{U'} .\, p'_i = u'_i
+\\ \forall i \not\in \dom{U'} .\, p'_i = 0
+\\ \forall i \not\in \text{Loc}
+  .\, s_i = 0
+}
+{ B, R \vdash \htriple
+  { \Gamma_f[\Vargs / \vec x] {\land} \Gamma_{\text{Loc}}; Q {+} c {+} M_f }
+  { r \gets f(\vec x) }
+  { \Gamma_f'[\Vret/r] {\land} \Gamma_{\text{Loc}}; Q' {+} c {-} M_r }
+}
+%
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ \fenv f = (\vec y, S_f)
+\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{ \Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y] }{ S_f }{ \Gamma'; Q' }
+}
+{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
+%
+\and \Rule{Q:Weak}
+{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2
+\\ \Gamma'_2 \models \Gamma'_1
+\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
+}
+{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
+%
+\\ \Rule{Relax}
+{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
+\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
+\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
+\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
+\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
+\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
+}
+{ Q' \succeq_\Gamma Q }
+%\vspace{-.3cm}
+\end{mathpar}
+\caption{Inference rules of the quantitative analysis.}
+\label{fig:auto}
+\end{figure*}
+}{}
 
@@
 \nocaptionrule \caption{
-  Experimental evaluation comparing the bounds generated KoAT, Rank,
-  LOOPUS, SPEED, and our automatic amortized analysis on several
+  Comparison of the bounds generated by KoAT, Rank,
+  LOOPUS, SPEED, and our tool $\toolname$ on several
   challenging linear examples. Results for KoAT and SPEED were extracted

### 2015-04-15 paper/main.tex
spacing fix in table 2
move figure 3 just below figure 2
add the large figure as an appendix

@@
 
+\begin{figure*}
+  \centering
+\ifshort{%\vspace{-.3cm}
+\hspace{-0.4cm}}{}
+  \begin{minipage}[b]{7cm}
+    \begin{center}
+   \begin{lstlisting}[]
+void c_down (int x,int y) {
+  if (x>y) {tick(1); c_up(x-1,y);}
+}
+void c_up (int x, int y) {
+  if (y+1<x) {tick(1); c_down(x,y+2);}
+}
+   \end{lstlisting}
+%\vspace{-2.7ex}
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
+\\[.4\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{5cm}
+    \begin{center}
+   \begin{lstlisting}
+for (; l>=8; l-=8)
+  /* process one block */
+  tick(N);
+for (; l>0; l--)
+  /* save leftovers */
+  tick(1);
+   \end{lstlisting}
+%\vspace{-2.8ex}
+$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
+$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
+\\[.4\baselineskip]
+      {\bf t61}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{5.5cm}
+    \begin{center}
+\begin{lstlisting}
+for (;;) {
+  do { l++; tick(1); }
+    while (l<h && *);
+  do { h--; tick(1); }
+    while (h>l && *);
+  if (h<=l) break;
+  tick(1); /* swap elems. */ }
+\end{lstlisting}
+$2+3|[l,h]|$
+\\[.4\baselineskip]
+      {\bf t62}
+    \end{center}
+  \end{minipage}
+\vspace{1ex}
+\caption{Example \emph{t39} shows two mutually-recursive functions
+  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
+  demonstrate the unique compositionality of our system. In \emph{t61},
+  $N\geq 0$ is a fixed but arbitrary constant. % The derived constant factors
+  %are tight.
+  }
+  \label{fig:ex_list_2}
+\end{figure*}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % each other with different speeds.
-
-\begin{figure*}
-  \centering
-\ifshort{%\vspace{-.3cm}
-\hspace{-0.4cm}}{}
-  \begin{minipage}[b]{7cm}
-    \begin{center}
-   \begin{lstlisting}[]
-void c_down (int x,int y) {
-  if (x>y) {tick(1); c_up(x-1,y);}
-}
-void c_up (int x, int y) {
-  if (y+1<x) {tick(1); c_down(x,y+2);}
-}
-   \end{lstlisting}
-%\vspace{-2.7ex}
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
-\\[.4\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{5cm}
-    \begin{center}
-   \begin{lstlisting}
-for (; l>=8; l-=8)
-  /* process one block */
-  tick(N);
-for (; l>0; l--)
-  /* save leftovers */
-  tick(1);
-   \end{lstlisting}
-%\vspace{-2.8ex}
-$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
-$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
-\\[.4\baselineskip]
-      {\bf t61}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{5.5cm}
-    \begin{center}
-\begin{lstlisting}
-for (;;) {
-  do { l++; tick(1); }
-    while (l<h && *);
-  do { h--; tick(1); }
-    while (h>l && *);
-  if (h<=l) break;
-  tick(1); /* swap elems. */ }
-\end{lstlisting}
-$2+3|[l,h]|$
-\\[.4\baselineskip]
-      {\bf t62}
-    \end{center}
-  \end{minipage}
-\vspace{1ex}
-\caption{Example \emph{t39} shows two mutually-recursive functions
-  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
-  demonstrate the unique compositionality of our system. In \emph{t61},
-  $N\geq 0$ is a fixed but arbitrary constant. % The derived constant factors
-  %are tight.
-  }
-  \label{fig:ex_list_2}
-\end{figure*}
 
@@
 
-\begin{table*}
-\renewcommand{\max}[0]{{\rm mx}}
-\nocaptionrule \caption{
-  Comparison of the bounds generated by KoAT, Rank,
-  LOOPUS, SPEED, and our tool $\toolname$ on several
-  challenging linear examples. Results for KoAT and SPEED were extracted
-  from previous publications because KoAT cannot take C programs as
-  input in its current version and SPEED is not available.
-  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
-  marked with --- indicate that the tool failed to produce a result. We write
-  $\max(a,b)$ for the maximum of $a$ and $b$.
-}
-\vspace{6pt}
-\label{tab:eval}
-\footnotesize
-
-\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
-\newcommand{\file}[2][c]{%
-  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
-
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\centering
-\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
-
-File &
-\multicolumn{2}{c|}{KoAT} &
-\multicolumn{2}{c|}{Rank} &
-\multicolumn{2}{c|}{LOOPUS} &
-SPEED &
-\toolname
-\\
-
-%--------------------------------------------
-\hhline{=|==|==|==|=|=} \texttt{gcd.c} &
-
-? &
-&
-
-% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
-$(((2{+}1)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{kmp.c} &
-
-? &
-&
-
-% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
-$(((2{+}(n{+} \dots$ &
-$O(n^2)$ &
-
-% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
-$\max(n, 0) \dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \texttt{qsort.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1 {+} 2 |\inter 0 {\code{len}}|$
-\\
-
-%--------------------------------------------
-% \hline \file{\dir{speed\_pldi09}{fig1.c}} &
-%
-% ? &
-% &
-%
-% --- &
-% &
-%
-% $2 \max(n, 0)$ &
-% $O(n)$ &
-%
-% God damn it &
-%
-% $1 + 2 |\inter 0 n|$
-% \\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_2.c}} &
-
---- &
-&
-
-% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}n)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + n$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_4.c}} &
-
---- &
-&
-
-% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + m$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_5.c}} &
-
-$28d + 7g + 27$ &
-$O(n)$ &
-
-% $((((1+1)+(-1+n))+m)+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, n-m)$ &
-
----
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex1.c}} &
-
---- &
-&
-
---- &
-&
-
---- &
-&
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex3.c}} &
-
---- &
-&
-
-% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex4.c}} &
-
-$110 a + 33$ &
-$O(n)$ &
-
---- &
-&
-
---- &
-&
-
-$n + 1$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{fig2\_1.c}} &
-
-$9a + 9b + \dots$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{+}((-y\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, m{-}y)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{fig2\_2.c}} &
-
-$6a + 9b + 3c + 5$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
-$((2{-}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
-$\max(0, (x + 1 {-}z)\dots$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, n{-}z)$ &
-
-$|\inter x n| {+} |\inter z n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{nstd\_multiple.c}} &
-
---- &
-&
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n^2)$ &
-
-$\max(0, m {-} y) + \max(0, n {-} x)$ &
-$O(n)$ &
-
-$\max(0, m{-}y) + \max(0, n{-}x)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{nstd\_single.c}} &
-
-$48 b + 16$ &
-$O(n)$ &
-
-% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
-$(((1{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(0, \! n {-} 1)\dots$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{sqntl\_single.c}} &
-
-$21 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{}-x{+}n\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_multiple.c}} &
-
-$9 c + 10 d + 7$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 m| {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_single2.c}} &
-
-$20 d + 12 c + 17$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 n| {+} |\inter 0 m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_single.c}} &
-
-$4 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\texttt{t07.c}} &
-
-? &
-&
-
-$2+x$ &
-$O(n)$ &
-
-% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
-$\max(x, 0)\dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t08.c} &
-
-? &
-&
-
-% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
-$((2{+}z{-}y\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
-$\max(0, \! y{-}2)\dots$ &
-$O(n)$ &
-
-? &
-
-$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t10.c} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
-$((2{-}y{+}x\dots$ &
-$O(n)$ &
-
-$\max(0, x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t11.c} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x)) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t13.c} &
-
-? &
-&
-
-% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
-$(((1{+}y^2/2\dots$ &
-$O(n^2)$ &
-
-$2{\cdot}\max(x, 0) + \max(y, 0)$ &
-$O(n)$ &
-
-? &
-
-$2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t15.c} &
-
-? &
-&
-
-% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
-$((1{+}x\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t16.c} &
-
-? &
-&
-
-% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
-$((-99{\cdot}y\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$101 |\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t19.c} &
-
-? &
-&
-
-% $(((1+1)+(151+k))+(-100+i))+1$ &
-$((153{+}k\dots$ &
-$O(n)$ &
-
-$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
-$O(n)$ &
-
-? &
-
-$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t20.c} &
-
-? &
-&
-
-% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
-$(2{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
-$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x y| {+} |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t27.c} &
-
-? &
-&
-
---- &
-&
-
-% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
-$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
-$O(n)$ &
-
-? &
-
-$0.01|\inter n y| {+} 11 |\inter n 0|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t28.c} &
-
-? &
-&
-
-% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
-$((1{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
-$10^3 \,\max(0, x - y)\dots$ &
-$O(n)$ &
-
-? &
-
-$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t30.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t37.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t39.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1.33 {+} 0.67 |\inter z y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t46.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t47.c} &
-
-? &
-&
-
-$4+n$ &
-$O(n)$ &
-
-$1 + \max(n, 0)$ &
-$O(n)$ &
-
-? &
-
-$1 {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\end{tabular}
-\end{table*}
-
 \pref{tab:cbench} contains a compilation of the results of our
@@
 
+\begin{table*}[h!]
+\appendix
+\section{Complete Experimental Results for the Tool Comparison}
+
+\renewcommand{\max}[0]{{\rm mx}}
+\nocaptionrule \caption{
+  Comparison of the bounds generated by KoAT, Rank,
+  LOOPUS, SPEED, and our tool $\toolname$ on several
+  challenging linear examples. Results for KoAT and SPEED were extracted
+  from previous publications~\cite{GulwaniJK09,GulwaniMC09,GulwaniZ10,SinnZV14} because KoAT cannot take C programs as
+  input in its current version and SPEED is not available.
+  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
+  marked with --- indicate that the tool failed to produce a result. We write
+  $\max(a,b)$ for the maximum of $a$ and $b$.
+}
+\vspace{6pt}
+\label{tab:eval}
+\footnotesize
+
+\newcommand{\dir}[2]{\code{#1} \\ \code{#2}}
+\newcommand{\file}[2][c]{%
+  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
+
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\centering
+\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
+
+Function &
+\multicolumn{2}{c|}{KoAT} &
+\multicolumn{2}{c|}{Rank} &
+\multicolumn{2}{c|}{LOOPUS} &
+SPEED &
+\toolname
+\\
+
+%--------------------------------------------
+\hhline{=|==|==|==|=|=} \code{gcd} &
+
+? &
+&
+
+% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
+$(((2{+}1)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$|\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{kmp} &
+
+? &
+&
+
+% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
+$(((2{+}(n{+} \dots$ &
+$O(n^2)$ &
+
+% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
+$\max(n, 0) \dots$ &
+$O(n)$ &
+
+? &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \code{qsort} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$1 {+} 2 |\inter 0 {\code{len}}|$
+\\
+
+%--------------------------------------------
+% \hline \file{\dir{SPEED\ PLDI09}{fig1}} &
+%
+% ? &
+% &
+%
+% --- &
+% &
+%
+% $2 \max(n, 0)$ &
+% $O(n)$ &
+%
+% God damn it &
+%
+% $1 + 2 |\inter 0 n|$
+% \\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ pldi09}{fig4\ 2}} &
+
+--- &
+&
+
+% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}n)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + n$ &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ pldi09}{fig4\ 4}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + m$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ pldi09}{fig4\ 5}} &
+
+$28d + 7g + 27$ &
+$O(n)$ &
+
+% $((((1+1)+(-1+n))+m)+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\max(n, n-m)$ &
+
+---
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ pldi10}{ex1}} &
+
+--- &
+&
+
+--- &
+&
+
+--- &
+&
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ pldi10}{ex3}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ pldi10}{ex4}} &
+
+$110 a + 33$ &
+$O(n)$ &
+
+--- &
+&
+
+--- &
+&
+
+$n + 1$ &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{fig2\ 1}} &
+
+$9a + 9b + \dots$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{+}((-y\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, m{-}y)$ &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{fig2\ 2}} &
+
+$6a + 9b + 3c + 5$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
+$((2{-}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
+$\max(0, (x + 1 {-}z)\dots$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, n{-}z)$ &
+
+$|\inter x n| {+} |\inter z n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{nested\ multiple}} &
+
+--- &
+&
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
+$((2{-}x{+}n\dots$ &
+$O(n^2)$ &
+
+$\max(0, m {-} y) + \max(0, n {-} x)$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, m{-}y) $ &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{nested\ single}} &
+
+$48 b + 16$ &
+$O(n)$ &
+
+% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
+$(((1{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(0, \! n {-} 1)\dots$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{sequential\ single}} &
+
+$21 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{}-x{+}n\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{simple\ multiple}} &
+
+$9 c + 10 d + 7$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$n + m$ &
+
+$|\inter 0 m| {+} |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{simple\ single2}} &
+
+$20 d + 12 c + 17$ &
+$O(n)$ &
+
+--- &
+&
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$n + m$ &
+
+$|\inter 0 n| {+} |\inter 0 m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\ popl10}{simple\ single}} &
+
+$4 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\code{t07}} &
+
+? &
+&
+
+$2+x$ &
+$O(n)$ &
+
+% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
+$\max(x, 0)\dots$ &
+$O(n)$ &
+
+? &
+
+$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{t08} &
+
+? &
+&
+
+% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
+$((2{+}z{-}y\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+$\max(0, \! y{-}2)\dots$ &
+$O(n)$ &
+
+? &
+
+$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{t10} &
+
+? &
+&
+
+% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
+$((2{-}y{+}x\dots$ &
+$O(n)$ &
+
+$\max(0, x {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \code{t11} &
+
+? &
+&
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x)) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \code{t13} &
+
+? &
+&
+
+% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
+$(((1{+}y^2/2\dots$ &
+$O(n^2)$ &
+
+$2{\cdot}\max(x, 0) + \max(y, 0)$ &
+$O(n)$ &
+
+? &
+
+$2 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{t15} &
+
+? &
+&
+
+% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
+$((1{+}x\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$|\inter 0 x|$
+\\
+
+%--------------------------------------------
+\hline \code{t16} &
+
+? &
+&
+
+% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
+$((-99{\cdot}y\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$101 |\inter 0 x|$
+\\
+
+%--------------------------------------------
+\hline \code{t19} &
+
+? &
+&
+
+% $(((1+1)+(151+k))+(-100+i))+1$ &
+$((153{+}k\dots$ &
+$O(n)$ &
+
+$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
+$O(n)$ &
+
+? &
+
+$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
+\\
+
+%--------------------------------------------
+\hline \code{t20} &
+
+? &
+&
+
+% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
+$(2{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
+$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter x y| {+} |\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \code{t27} &
+
+? &
+&
+
+--- &
+&
+
+% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
+$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
+$O(n)$ &
+
+? &
+
+$0.01|\inter n y| {+} 11 |\inter n 0|$
+\\
+
+%--------------------------------------------
+\hline \code{t28} &
+
+? &
+&
+
+% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
+$((1{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
+$10^3 \,\max(0, x - y)\dots$ &
+$O(n)$ &
+
+? &
+
+$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \code{t30} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$|\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{t37} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{t39} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$1.33 {+} 0.67 |\inter z y|$
+\\
+
+%--------------------------------------------
+\hline \code{t46} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$|\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \code{t47} &
+
+? &
+&
+
+$4+n$ &
+$O(n)$ &
+
+$1 + \max(n, 0)$ &
+$O(n)$ &
+
+? &
+
+$1 {+} |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\end{tabular}
+\end{table*}
+
 \end{document}

### 2015-04-15 paper/main.tex
add the DOI number, add notes in Table 3 caption

@@
 \copyrightdata{978-1-4503-3468-6/15/06}
-\doi{nnnnnnn.nnnnnnn}
+\doi{2737924.2737955}
 
@@
 \pref{tab:compar} summarizes the results of our experiments presented
-in \pref{tab:eval}.  It shows
+in \pref{app:eval}.  It shows
 for each tool the number of derived bounds (\#bounds), the number of
@@
 \section{Complete Experimental Results for the Tool Comparison}
+\label{app:eval}
 
@@
   marked with --- indicate that the tool failed to produce a result. We write
-  $\max(a,b)$ for the maximum of $a$ and $b$.
+  $\max(a,b)$ for the maximum of $a$ and $b$.  Functions with names of the form \code{tXX} are complex tests
+  that we designed during the development of $\toolname$.  The source code for all the functions
+  is available in the extended version~\cite{anon_extended}.
 }

### 2015-04-15 paper/main.tex
some updates to the caption

@@
   marked with --- indicate that the tool failed to produce a result. We write
-  $\max(a,b)$ for the maximum of $a$ and $b$.  Functions with names of the form \code{tXX} are complex tests
-  that we designed during the development of $\toolname$.  The source code for all the functions
+  $\max(a,b)$ for the maximum of $a$ and $b$.  Functions with names of the form \code{tXX} are challenging tests
+  that we designed during the development of $\toolname$.  The source code for all functions
   is available in the extended version~\cite{anon_extended}.

### 2015-07-27 paper/main.tex
add proper copyright notice
start changing notations in tr

--- 2015-04-15 paper/main.tex
@@
 % Also look for "SHORT" comments
-% \def\fullversion{}
-
-\documentclass{sigplanconf}
+ \def\fullversion{}
+
+\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}
 
@@
 
-%\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
-%\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
-%\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
-%\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
-%\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
-%\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
-%\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}
-
-\newcommand{\itemskip}[0]{}
-\newcommand{\itemskipIn}[0]{}
-\newcommand{\sectskip}[0]{}
-\newcommand{\paraskip}[0]{}
-\newcommand{\lstskip}[0]{}
-\newcommand{\aftersectskip}[0]{}
-\newcommand{\subsectskip}[0]{}
+\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
+\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
+\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
+\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
+\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
+\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
+\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}
 
 %SHORT
-%\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captions
-%\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
-%\captionsetup[figure]{format=myformat}
-%\captionsetup[table]{format=myformat}
-%\renewcommand{\captionlabelfont}{\bf}
+\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
+\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
+\captionsetup[figure]{format=myformat}
+\captionsetup[table]{format=myformat}
+\renewcommand{\captionlabelfont}{\bf}
 %SHORT
@@
 \usepackage{listings} %for code
-\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
 \usepackage{mathpartir} %for typing rules
@@
 \usepackage{array} %for stupid tables
-\usepackage{flushend} %equalize columns length on last page
-\usepackage{hhline}
-%\usepackage[firstpage]{draftwatermark}
-%\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[scale=0.1]{fig/aec-badge-pldi.pdf}}}
-%\SetWatermarkAngle{0}
 
@@
 \input{definitions}
+\begin{document}
+
+\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
+
+%SHORT
+\ifx\fullversion\undefined
+\abovedisplayskip=3pt
+\belowdisplayskip=3pt
+\abovedisplayshortskip=3pt
+\belowdisplayshortskip=3pt
+\else
+\fi
+%SHORT
+
+\conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
+\copyrightyear{2014}
+\copyrightdata{XXX-X-XXXX-XXXX-X/XX/XX}
+
+%----------------------------------------------------
+
 \newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{PLDI'15}{June 13--17, 2015, Portland, OR, USA}
-\copyrightyear{2015}
-\copyrightdata{978-1-4503-3468-6/15/06}
-\doi{nnnnnnn.nnnnnnn}
 
@@
 
-\title{Compositional Certified Resource Bounds}
+\title{Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}}
 % Certified Resource Bounds for C
@@
 {Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
-{Yale University, USA}
-{\{quentin.carbonneaux, jan.hoffmann, zhong.shao\}@yale.edu}
+{Yale University \vspace{-1cm}}
 
 \maketitle
-
-%SHORT
-%\ifx\fullversion\undefined
-%\abovedisplayskip=3pt
-%\belowdisplayskip=3pt
-%\abovedisplayshortskip=3pt
-%\belowdisplayshortskip=3pt
-%\else
-%\fi
-%SHORT
-
-%----------------------------------------------------
 
@@
 
-\category{D.2.4}{Software Engineering}
-{Software/Program Verification}
-\category{F.3.1}{Logics and Meanings of Programs}
-{Specifying and Verifying and Reasoning about Programs}
-%\vspace{-3pt}
-\terms Verification, Reliability
-%\vspace{-3pt}
-\keywords Quantitative Verification, Resource Bound Analysis, Static
-Analysis, Amortized Analysis, LP Solving, Program Logic
+% \category{D.2.4}{Software Engineering}
+% {Software/Program Verification}
+% \category{F.3.1}{Logics and Meanings of Programs}
+% {Specifying and Verifying and Reasoning about Programs}
+
+% \terms Verification, Reliability
+
+% \keywords Formal Verification, Compiler Construction, Program Logics,
+% Stack-Space Bounds, Quantitative Verification
 
@@
 is that they can only associate \emph{potential}
-with individual program variables or data structures.
+with individual program variable or data structure.
 For C programs, this
@@
 The resource behavior of functions can be summarized
-in a functional specification that can be used at every
+in a function specification that can be used at every
 call site without accessing the function body.
@@
 resource analysis technique for C that deals naturally with
-recursive functions and sequenced loops, and can handle resources that may
+recursive functions and sequenced loops, and can handle resources that can
 become available during execution (e.g., when freeing
@@
   functions, generates resource abstractions for functions, derives
-  proof certificates, and handles resources that may
+  proof certificates, and handles resources that can
   become available during execution.
@@
 \itemskip
+%
+\iffull{}
+{\footnotesize \sl
+[To the reviewers: The auxiliary material contains an anonymized
+version of the extended article, and the source code of $\toolname$
+and all examples.] }
 
@@
 we also use as the loop invariant.  For the loop body we have (like in
-Hoare logic) to derive a triple
+Hoare logic) to derive a triple like
 $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
@@
   \centering
-%\vspace{-1.5ex}
+\vspace{-1.5ex}
 \begin{lstlisting}[mathescape]
 ${\color{blue}%
-  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
+  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
 while (x+K<=y) {
   ${\color{blue}%
-  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
+  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
   x=x+K;
   ${\color{blue}%
-  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| \} }$
+  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
   tick(T);
   ${\color{blue}%
-  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
+  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
 }
 ${\color{blue}%
- \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
+ \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
 \end{lstlisting}
@@
 to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
-logical assertions from potential functions with semicolons. Note that the
-logical assertions are only used in the rule for the assignment
+$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
+assertions are only used in the rule for the assignment
 \code{x=x+K}.
@@
 \begin{figure*}
+ \setlength{\progwidth}{.22\linewidth}
   \centering
-%\vspace{-.3cm}
+\vspace{-.3cm}
 \hspace{-0.4cm}
-  \begin{minipage}[b]{.24\linewidth}
+  \begin{minipage}[b]{.18\linewidth}
     \begin{center}
@@
    \end{lstlisting}
-%\vspace{-2.5ex}
+\vspace{-2.5ex}
 $|[x, n]| + |[y, m]|$
@@
 %
-  \begin{minipage}[b]{0.27\linewidth}
+  \begin{minipage}[b]{\progwidth}
     \begin{center}
@@
    \end{lstlisting}
-%\vspace{-2.5ex}
+\vspace{-2.5ex}
 $|[x, n]| + |[z, n]|$
@@
 %
-  \begin{minipage}[b]{0.25\linewidth}
+  \begin{minipage}[b]{\progwidth}
     \begin{center}
@@
    \end{lstlisting}
-%\vspace{-2.5ex}
+\vspace{-2.5ex}
 $3.1|[y,z]| + 0.1|[0,y]|$
@@
 %
-  \begin{minipage}[b]{0.22\linewidth}
+  \begin{minipage}[b]{\progwidth}
     \begin{center}
@@
    \end{lstlisting}
-%\vspace{-2.5ex}
+\vspace{-2.5ex}
 $59|[n,0]| {+} 0.05|[0,y]|$
@@
 
-\begin{figure*}
-  \centering
-\ifshort{%\vspace{-.3cm}
-\hspace{-0.4cm}}{}
-  \begin{minipage}[b]{7cm}
-    \begin{center}
-   \begin{lstlisting}[]
-void c_down (int x,int y) {
-  if (x>y) {tick(1); c_up(x-1,y);}
-}
-void c_up (int x, int y) {
-  if (y+1<x) {tick(1); c_down(x,y+2);}
-}
-   \end{lstlisting}
-%\vspace{-2.7ex}
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
-\\[.4\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{5cm}
-    \begin{center}
-   \begin{lstlisting}
-for (; l>=8; l-=8)
-  /* process one block */
-  tick(N);
-for (; l>0; l--)
-  /* save leftovers */
-  tick(1);
-   \end{lstlisting}
-%\vspace{-2.8ex}
-$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
-$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
-\\[.4\baselineskip]
-      {\bf t61}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{5.5cm}
-    \begin{center}
-\begin{lstlisting}
-for (;;) {
-  do { l++; tick(1); }
-    while (l<h && *);
-  do { h--; tick(1); }
-    while (h>l && *);
-  if (h<=l) break;
-  tick(1); /* swap elems. */ }
-\end{lstlisting}
-$2+3|[l,h]|$
-\\[.4\baselineskip]
-      {\bf t62}
-    \end{center}
-  \end{minipage}
-\vspace{1ex}
-\caption{Example \emph{t39} shows two mutually-recursive functions
-  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
-  demonstrate the unique compositionality of our system. In \emph{t61},
-  $N\geq 0$ is a fixed but arbitrary constant. % The derived constant factors
-  %are tight.
-  }
-  \label{fig:ex_list_2}
-\end{figure*}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 To find resource bounds automatically, we first need to restrict our
-search space.  In this work, we focus on the following form of
+search space.  In this work, we focus on a the following form of
 potential functions, which can express tight bounds for many typical
@@
 that cannot be assigned to.  For example, if the program contains the
-constant $1988$ then we have a variable $c_{1988}$ and
-$\state(c_{1988}) = 1988$.  We assume that every program state
+constant $8128$ then we have a variable $c_{8128}$ and
+$\state(c_{8128}) = 8128$.  We assume that every program state
 includes the constant $c_0$.
@@
 
-In addition to the quantitative annotations, our automatic amortized
+In addition to the quantitative annotations our automatic amortized
 analysis needs to maintain a minimal abstract state to justify certain
@@
 interpretation with the domain of linear inequalities.  We observed that
-the rules of the analysis often require only minimal local knowledge.
+the rules of the analysis often requires only minimal local knowledge.
 This means that it is not necessary for us to compute precise loop
@@
 variables unchanged by the loop) is sufficient to obtain good bounds.
+
+\begin{figure*}
+  \centering
+\ifshort{\vspace{-.3cm}
+\hspace{-0.4cm}}{}
+  \begin{minipage}[b]{7cm}
+    \begin{center}
+   \begin{lstlisting}[]
+void c_down (int x,int y) {
+  if (x>y) {tick(1); c_up(x-1,y);}
+}
+void c_up (int x, int y) {
+  if (y+1<x) {tick(1); c_down(x,y+2);}
+}
+   \end{lstlisting}
+\vspace{-2.7ex}
+$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
+$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
+\\[.4\baselineskip]
+      {\bf t39}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{5cm}
+    \begin{center}
+   \begin{lstlisting}
+for (; l>=8; l-=8)
+  /* process one block */
+  tick(N);
+for (; l>0; l--)
+  /* save leftovers */
+  tick(1);
+   \end{lstlisting}
+\vspace{-2.8ex}
+$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
+$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
+\\[.4\baselineskip]
+      {\bf t61}
+    \end{center}
+  \end{minipage}
+%
+\hfill
+%
+  \begin{minipage}[b]{5.5cm}
+    \begin{center}
+\begin{lstlisting}
+for (;;) {
+  do { l++; tick(1); }
+    while (l<h && *);
+  do { h--; tick(1); }
+    while (h>l && *);
+  if (h<=l) break;
+  tick(1); /* swap elems. */ }
+\end{lstlisting}
+$2+3|[l,h]|$
+\\[.4\baselineskip]
+      {\bf t62}
+    \end{center}
+  \end{minipage}
+\vspace{1ex}
+\caption{Example \emph{t39} shows two mutually-recursive functions
+  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
+  demonstrate the unique compositionality of our system. In \emph{t61},
+  $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
+  are tight.}
+  \label{fig:ex_list_2}
+\end{figure*}
 
@@
 precisely, it is the partitioning part of the algorithm.  This partition
-loop has linear complexity, and feeding it to our analysis gives the
-worst-case bound $2+3|[l,h]|$.  This bound is not optimal but it can be
-refined by rewriting the program. % to let our logical inference
-%discover more invariants.
-%With the potential
-%method in mind, it is clear that the three loops get their
-%potential from the same interval $[l,h]$ regardless of their nesting.
-To understand the bound, we can reason as follows.
-%
-If $h \geq l$ initially, the cost of the loop is 2.
-Otherwise, the cost of each round (at
-most 3) can be payed using the potential of $[l,h]$ by the first
-increment to $l$ because we know that $l<h$.  The two inner loops
-can also use $[l,h]$ to pay for their inner costs.
-%
-KoAT fails to
+loop has linear complexity, and feeding into our analysis gives the
+correct and tight worst-case bound $2+3|[l,h]|.$  With the potential
+method in mind, it is clear that the three loops get their
+potential from the same interval $[l,h]$.  However, KoAT fails to
 find a bound and LOOPUS derives the quadratic bound
 $(h-l-1)^2$.  Following the classical technique, these tools try to find one
-ranking function for each loop and combine them multiplicatively
+linear ranking-function for each loop and combine them multiplicatively
 or additively.
@@
   handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
-  extended version~\cite{anon_extended}} is a list of more than 30
+  extended version~\cite{anon_extended}} is a list of more than $30$
 classes of challenging programs that we can automatically analyze.
@@
 \Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
-{ }
-{ \!\! B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) \!\!}
+{ n < 0 \implies Q \ge 0 }
+{ B; R; (\Gamma, Q) \vdash \code{tick}(n) \dashv (\Gamma, Q{-}n) }
 %
@@
 \and \Rule{Q:Update}
-{ q'_{xy}, q'_{yx} \in \Qplusz
-\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+{ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
 }
@@
 %
-\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
+\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ Q \ge Q'
+\\ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q)
+}
 { B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
 %
-\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
-{ \Gamma \models y \ge 0 \!
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx} \!
+\and \Rule{Q:Inc}
+{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
+\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
 }
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:DecP}
-{ \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
+  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
+}
+%
+\and \Rule{Q:Dec}
+{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{xu}, -q_{ux})
+\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{ux}, -q_{xu})
 }
 { \textstyle
-  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-}
-{ \textstyle
-  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
+  B; R; (\Gamma[x/x {-} y], Q) \vdash x \gets x - y \dashv (\Gamma, Q')
 }
@@
 \and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+{ B; R; (\Gamma \land e, Q) \vdash S_1 \dashv (\Gamma',Q')
+\\\\ B; R; (\Gamma \land \neg e, Q) \vdash S_2 \dashv (\Gamma',Q')
+}
+{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
 %
 \and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
-\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
+{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q')
+\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
 }
@@
 }
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
+{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
   \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
-}
-%
-\and
-\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
+  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
+}
+%
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
 { }
-{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and
-\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ B; R; (\Gamma, Q) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
+%
+\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
 { \fenv f = (\vec y, S_f)
+\\ Q_f \ge 0
+\\ Q'_f \ge 0
 \\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
@@
 
+\jan{QUENTIN FIX ABOVE FIGURE}
+
 \sectskip
@@
 \paraskip
-\paragraph{Cost Aware Clight.}
-
-We present the rules for a subset of Clight.
-Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}. It
-is a subset of C with a unified looping construct and
-side-effect free expressions.  We reuse most of CompCert's
-syntax but instrument the semantics with a resource metric
-$M$ that accounts for the cost (an arbitrary rational number)
-of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
-evaluating the expression \code{exp}. The rationals $M_f$
-and $M_r$ account respectively for the cost of a call to the
-function $f$ and the cost of returning from it.
-More details are provided in \pref{sec:soundness}.
-
-In the rules, assignments are restricted to the form $x \gets
-y$ or $x \gets x \pm y$.  In the implementation, a Clight
-program is converted into this form prior to analysis without changing
-the resource cost.  This is achieved by using a series of \emph{cost-free
-  assignments} that do not result in additional cost in the
-semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
-This sound treatment ensures that no further loop bounds
-depend on the result of the non-linear operation.
-
-\paraskip
 \paragraph{Judgements.}
@@
 with starting state $\state$, the assertions in $\Gamma$
-hold, and at least $Q(\state)$ resources are available then the
+hold, $Q' {\ge} 0$, and at least $Q(\state)$ resources are available then the
 evaluation does not run out of resources and, if the execution
@@
 Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
-returning from a function call. \iffull{(This becomes clear
+returning from a function call with \code{return}. \iffull{(This becomes clear
 in the rules.)}{}
@@
 such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
-In all the rules, we have the implicit side condition that all rational
-coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
+%
+% THE FOLLOWING IS NO LONGER TRUE
+% In all the rules, we have the implicit side condition that all rational
+% coefficients are non-negative.
+%
+Finally, if a rule mentions $Q$ and $Q'$
 and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.
 
-\ifshort{
-\begin{figure*}
-\small
-%\vspace{-.4cm}
-\def \MathparLineskip {\lineskip=0.35cm}
-\begin{mathpar}
-\Rule{Q:Skip}
-{ }
-{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
-%
-\and \Rule{Q:Assert}
-{ }
-{ B, R \vdash \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
-%
-\and \Rule{Q:Tick}
-{ }
-{ B, R \vdash \htriple{ \Gamma; Q {+} M_t(n) }{ \code{tick}(n) }{ \Gamma; Q } }
-%
-\and \Rule{Q:Break}
-{ }
-{(\Gamma; Q_B), R \vdash \htriple{ \Gamma; Q_B {+} M_b }{ \code{break} }{ \Gamma'; Q' } }
-%
-\and \Rule{Q:Return}
-{ P = Q_R[\Vret/x]
-\\ \Gamma = \Gamma_R[\Vret/x]
-\\ \forall i \in \dom{P} . \, p_i = q_i
-}
-{ B, (\Gamma_R; Q_R) \vdash \htriple{ \Gamma; Q }{ \code{return}~x }{ \Gamma'; Q' } }
-%
-\and \Rule{Q:Loop}
-{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
-{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
-%
-\and \Rule{Q:Seq}
-{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
-\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
-}
-{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
-%
-\and \Rule{Q:If}
-{ B, R \vdash \htriple{ \Gamma {\land} e; Q {-} M_c^1) }{ S_1 }{ \Gamma'; Q' }
-\\\\ B, R \vdash \htriple{ \Gamma {\land} \neg e; Q {-} M_c^2 }{ S_2 }{ \Gamma'; Q' }
-}
-{ B, R \vdash \htriple{ \Gamma; Q{+}M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ \Gamma'; Q' } }
-%
-\\ \Rule{Q:IncP}
-{ \Gamma \models y \ge 0 \!
-\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
-\\\\ \textstyle q'_{0y} = q_{0y}
-      + \sum_{u \in \mathcal U} q_{xu}
-      - \sum_{v \not\in \mathcal U} q_{vx} \!
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
-}
-%
-\and \Rule{Q:Inc}
-{ M = M_u + M_e(x {\pm} y)
-\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
-\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
-}
-%
-\and \Rule{Q:Set}
-{ q'_{xy}, q'_{yx} \in \Qplusz
-\\\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B, R \vdash \htriple{ \Gamma[x/y]; Q {+} M_u {+} M_e(y) }{ x \gets y }{ \Gamma; Q' } }
-%
-\and \Rule{Q:DecP}
-{ \Gamma \models y \ge 0
-\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
-\\\\ \textstyle q'_{y0} = q_{y0}
-      + \sum_{u \in \mathcal U} q_{ux}
-      - \sum_{v \not\in \mathcal U} q_{xv}
-}
-{ \textstyle
-  B, R \vdash \htriple{ \Gamma[x/x {-} y]; Q {+} M_u {+} M_e(x {-} y) }{ x \gets x - y }{ \Gamma; Q' }
-}
-%
-\and\Rule{Q:Call}
-{ (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
-\\ \text{Loc} = \text{Locals}(Q) \!
-\\ \forall i \neq j . \, x_i \neq x_j \!
-\\ c \in \Qplusz \!
-\\ Q = P + S \!
-\\ Q' = P' + S
-\\ U = Q_f[\Vargs / \vec x]
-\\ U' = Q'_f[\Vret/r]
-\\ \forall i \in \dom{U} . \, p_i = u_i
-\\ \forall i \in \dom{U'} .\, p'_i = u'_i
-\\ \forall i \not\in \dom{U'} .\, p'_i = 0
-\\ \forall i \not\in \text{Loc}
-  .\, s_i = 0
-}
-{ B, R \vdash \htriple
-  { \Gamma_f[\Vargs / \vec x] {\land} \Gamma_{\text{Loc}}; Q {+} c {+} M_f }
-  { r \gets f(\vec x) }
-  { \Gamma_f'[\Vret/r] {\land} \Gamma_{\text{Loc}}; Q' {+} c {-} M_r }
-}
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
-{ \fenv f = (\vec y, S_f)
-\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{ \Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y] }{ S_f }{ \Gamma'; Q' }
-}
-{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
-%
-\and \Rule{Q:Weak}
-{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
-\\ \!\! \Gamma_1 \models \Gamma_2
-\\\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ \Gamma'_2 \models \Gamma'_1
-\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
-}
-{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
-%
-\\ \Rule{Relax}
-{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
-\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
-\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
-\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
-\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
-}
-{ Q' \succeq_\Gamma Q }
-%\vspace{-.3cm}
-\end{mathpar}
-\caption{Inference rules of the quantitative analysis.}
-\label{fig:auto}
-\end{figure*}
-}{}
+We describe the automatic amortized analysis for a subset of
+expressions of Clight.  Assignments are restricted to the form $x \gets
+y$ or $x \gets x \pm y$.  In the implementation, a Clight
+program is converted into this form prior to analysis without changing
+the resource cost.  This is achieved by using a series of \emph{cost-free
+  assignments} that do not result in additional cost in the
+semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
+a[y]$ are handled by assigning $0$ to coefficients
+like $q_{xa}$ and $q_{ax}$ that depend on $x$ after the assignment.
 
@@
 rule.  This rule reuses its precondition as postcondition, it is
-explained by two facts: First, no resource is consumed by the skip
-operation, thus no potential has to be used to pay for the evaluation.
-Second, the program state is not changed by
-the execution of a skip statement. Thus all potential available
+explained by two observations: first, no resource is consumed by the skip
+operation, thus no potential has to be used to pay for the evaluation;
+second, the program state is not changed by
+the execution of a skip statement, thus all the potential available
 before the execution of the skip statement is still available after.
@@
 
-The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
-describe how the potential is distributed after a size change of
-a variable.
-The rule {\sc Q:IncP} is for increments $x \gets x + y$
-and {\sc Q:DecP} is for decrements $x \gets x - y$. They both apply only
-when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
-Of course, there are symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
-(not presented here) that can be applied if $y$ is negative.
-The rules are all equivalent in the case where $y=0$.
-The rule {\sc Q:Inc} can be applied
-if we cannot find the sign of $y$.  % In $\toolname$,
-% the rules for increments and decrements are combined into one rule\iffull{ where
-% the specialized rules take priority over {\sc Q:Inc}.}{.}
-
-To explain how rules for increment and decrement work, it is
-sufficient to understand the rule {\sc Q:IncP}.  The others follow the
-same idea and are symmetrical. In {\sc Q:IncP}, the program updates a
-variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
+The rules {\sc Q:Inc} and {\sc Q:Dec} describe how the potential is
+distributed after a size change of a variable.
+The rule {\sc Q:Inc} is for incements $x \gets x+y$ and the symmetric
+rule {\sc Q:Dec} is for decrements $x \gets x-y$.
+The two rules are equivalent in the case where $y=0$.
+
+The two previous rules are essentially the same, so we only focus on
+the explanation of {\sc Q:Inc}. In this rule, the program updates a
+variable $x$ with $x+y$.  Since $x$ is changed, the
 quantitative annotation must be updated to reflect the change of the
@@
 Since $x$ is the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
-$\inter x u$ will get smaller with the update, and if $x' \in \inter x
-u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
-$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
-$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
-guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
-get bigger with the update.  We know that $|\inter v {x'}| \le y +
-|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
-$q_{vx}$ to pay for this change.  The rule ensures this
-only for $v \not\in \mathcal U$ because $x \le v$
-otherwise, and thus $|\inter v x| = 0$.
-
-The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it
-on a loop body, one needs to find an invariant potential $Q$ that will
-pay for the iterations.  At each iteration, $M_l$ resources are spent
-to jump back. This explains the postcondition $Q{+}M_l$. Since the
-loop can only be exited with a \code{break} statement, the
-postcondition $\{ \Gamma'; Q' \}$ for the statement $\code{loop}~S$ is
-used as break postcondition in the derivation for $S$.
-
-Another interesting rule is {\sc Q:Call}.  It needs to account for the
-changes to the stack caused by the function call, the arguments/return
-value passing, and the preservation of local variables.  We can sum up
-the main ideas of the rule as follows.
-\itemskip
-\begin{itemize}
-\item The potential in the pre- and postcondition of the function
-  specification is equalized to its matching potential in the callee's
-  pre- and postcondition.
-\itemskipIn
-\item
-  The potential of intervals $|\inter x y|$ is preserved
-  across a function call
-  if $x$ and $y$ are local\iffull{ variables.}{.}
-\item
-  The unknown potentials after the call (e.g.
-  $|\inter x g|$, with $x$ local and $g$ global)
-  are set to zero in the postcondition.
-\end{itemize}
-\itemskip
-If $x$ and $y$ are local variables and $f(x,y)$ is
-called, {\sc Q:Call} splits the potential of
-$|\inter x y|$ in two parts. One part to perform the
-computation in the function $f$ and one part to keep for
-later use after the function call.  This splitting is
-realized by the equations $Q = P {+} S$ and $Q' =
-P' {+} S'$.  Arguments in the function precondition
-$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
-of names different from all program variables.  This
-prevents name conflicts and ensures that
-the substitution $[\Vargs/\vec x]$ is meaningful.
-Symmetrically, we use the unique name $\Vret$ to represent
-the return value in the function's postcondition
-$(\Gamma'_f; Q'_f)$.
-
-The rule {\sc Q:Weak} is the only rule that is not syntax directed. %In the implementation
-%we apply {\sc Q:Weak} before loops and between the two statements of a
-%sequential composition.
-We could integrate weakenings into every syntax directed rule but, for
-the sake of efficiency, the implementation uses a simple heuristic instead.
-The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}. This rule
-also deals with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq \ell$ is
-larger than a constant $\ell$ then we can turn the potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval into the constant potential $\ell{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
-without gaining potential.
+$\inter u x$ and $\inter x u$ will be resized.  Consider $S(x) = q_{xu}
+|\inter x u| + q_{ux} |\inter u x|$, during the increment, $S$ will
+increase by at most $y \cdot \max(q_{ux}, -q_{xu})$.
+This means that we must decrease the potential $q_{0y}$ of $|\inter 0 y|$
+by $\max(q_{ux}, -q_{xu})$ to pay for the change on
+$|\inter x u|$ and $|\inter u x|$.  A symmetric argument explains the
+change on $q_{y0}$.
 
@@
 \begin{figure*}[t]
-%\vspace{-.5cm}
+\vspace{-.5cm}
 \centering
@@
 }
-%\vspace{-.3cm}
+\vspace{-.3cm}
 \end{mathpar}
@@
 \begin{tabular}{l}Constraints:\end{tabular}
-%\vspace{-.2cm}
+\vspace{-.2cm}
 $$
@@
 
+\iffull{
+Another interesting rule is {\sc Q:Call}.  It needs to account for the
+changes to the stack caused by the function call, the arguments/return
+value passing, and the preservation of local variables.  We can sum up
+the main ideas of the rule as follows.
+\itemskip
+\begin{itemize}
+\item The potential in the pre- and postcondition of the function
+  specification is equalized to its matching potential in the callee's
+  pre- and postcondition.
+\itemskipIn
+\item
+  $\!$The potential of intervals $|\inter x y|$ is preserved
+  if $x$ and $y$ are local\iffull{ variables.}{.}
+\item
+  The unknown potentials after the call (e.g.
+  $|\inter x g|$, with $x$ local and $g$ global)
+  are set to zero in the postcondition.
+\end{itemize}
+\itemskip
+If $x$ and $y$ are local variables and $f(x,y)$ is
+called, {\sc Q:Call} splits the potential of
+$|\inter x y|$ in two parts, one part to perform the
+computation in the function $f$ and one part to keep for
+later use after the function call.  This splitting is
+realized by the equations $Q = P {+} S$ and $Q' =
+P' {+} S'$.  Arguments in the function precondition
+$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
+of names different from all program variables.  This
+prevents name conflicts from happening and ensures that
+the substitution $[\Vargs/\vec x]$ is meaningful.
+Symmetrically, we use the unique name $\Vret$ to represent
+the return value in the function's postcondition
+$(\Gamma'_f, Q'_f)$.
+}{}
+
+\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition. We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax}\iffull{ that
+also deals with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq l$ is
+larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}
+
 \sectskip
@@
 
-We separate the search of a derivation in two steps.  As a first
+We separate the search of a derivation into two steps.  As a first
 step we go through the functions of the program and apply inductively
@@
 $Q$ from the solver to get a resource bound for the program.  To get
-a full derivation, we extract the complete solution from the
-solver and apply it to the symbolic names $q_i$ of the coefficients
+a full derivation we extract the complete solution from the
+solver, and apply it to the symbolic names $q_i$ of the coefficients
 in the derivation.  If the LP solver fails to find a solution, an
@@
 The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}. It is parameterized by a logical context that is
+\textsc{Relax}. It is parameterized by a logical context
 used to gather information on interval sizes.  For example,
-{%\small
+$P^\text{de} \succeq_{(x{\ge}0)} P^\text{we} \equiv$
+% SSS SPACING
+%\vspace{-.3cm}
+{\small
 \begin{align*}
-  P^\text{de} \succeq_{(\cdot)} P^\text{we} &\equiv
- p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10} \\
-&\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
-&\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
+  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0,10}
+  \land p^\text{we}_{x,0} \leq p^\text{de}_{x,0} + u_{x,0} - v_{0,x} \land \;\;
+  \\
+  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}
+  \bigwedge_{(\alpha, \beta) \neq (0,10), (x,0)} p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
 \end{align*}
 }
+% SSS SPACING
+%\vspace{-.8cm}
+
+\noindent
 The other rules are syntax directed and applied inductively.  For example,
@@
 SMT-solver--based techniques is that a satisfying assignment is
-a proof certificate instead of a counter example.  To provide
-high-assurance bounds, this certificate can be checked
-in linear time by a simple validator.
+a proof certificate instead of a counter example.
+
+% SSS SPACING
+%\vspace{-.24cm}
 
@@
 \aftersectskip
+
+\begin{figure}[t]
+\begin{minipage}[b]{0.8\linewidth}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
+while (k > 0) {
+  x=0;
+  while (x < N && a[x] == 1) {
+    @assert(na > 0);@
+    a[x]=0; @na--;@
+    tick(1); x++; }
+  if (x < N) { a[x]=1; @na++;@ tick(1); }
+  k--;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\end{minipage}
+\caption{Assisted bound derivation using logical state.  We write
+  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$.
+  The derived bound is
+  $2|[0,k]| + |[0,na]|$.
+  }
+\label{fig:xmplincaux}
+\end{figure}
+
+\begin{figure}
+\vspace{-.3cm}
+\begin{lstlisting}[numbers=left]
+${\color{blue}%
+  \mbox{logical state invariant }
+  \{ lg > \log_2(h - l) \} }$
+bsearch(x,l,h@,lg@) {
+  if (h-l > 1) {
+    @assert(lg > 0);@
+    m = l + (h-l)/2;
+    @lg--;@ if (a[m]>x) h=m; else l=m;
+    tick($M_\code{bsearch}$);
+    l = bsearch(x,l,h@,lg@);
+    tick($-M_\code{bsearch}$);
+  } else return l;
+}
+\end{lstlisting}
+\vspace{-.2cm}
+\caption{Assisted bound derivation using logical state.
+  % We model the stack consumption of the program using tick statements.
+  % Before a function call to \code{bsearch}, a tick statement pays for
+  % the space used by stack frame of the function.  The space is recoverd
+  % after the call.
+  We write $\log_2(x)$ for the integer part of
+  logarithm of $x$ in base 2.
+  The semi-automatically derived bound is $|[0,lg]|$.
+  }
+\label{fig:xmplbsaux}
+\end{figure}
 
@@
   $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
-  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
-  }
-\label{fig:xmplmax}
+  The bound derived is $|[0,N]| + |[0,nm]|$.
+  }
+% \label{fig:xmplmax2}
 \end{figure}
@@
 a program $P$ that fails to be analyzed automatically is enriched by
-auxiliary variables $\vec x$ and assertions to form a program
+auxiliary variables $\vec x$ to form a program
 $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
-logical variables is selected to satisfy the proposition:
-\begin{equation} \tag{*}
+logical variables is selected so that:
+$$
 \forall n\, \sigma\, \sigma'.\,
@@
     (\sigma, P) \Downarrow_{n'} \sigma'.
-\end{equation}
+$$
 Since the annotated program and the original one are usually
@@
 procedure described above, we must show that the extra assertion
-\lstinline{na > 0} on line 5 cannot fail.  We do it
-by proving inductively that $na = \#_1(a)$
+\lstinline{na > 0} on line 5 is always true.  It
+is done by proving inductively that $na = \#_1(a)$
 and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) {>} 0$, thus the assertion \lstinline{na > 0} never fails.
-
-\begin{figure}[t]
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    @assert(na > 0);@
-    a[x]=0; @na--;@
-    tick(1); x++; }
-  if (x < N) { a[x]=1; @na++;@ tick(1); }
-  k--;
-}
-\end{lstlisting}
-%\vspace{-.2cm}
-\end{minipage}
-\caption{Assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
-  tick metric.  The derived bound  is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
+$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
+% test is in fact semantically equivalent to \lstinline{true}.
 
@@
 and the original one, we use the inequality $lg >
-\log_2(h-l)$ as invariant.  This allows a simpler
+\log_2(h-l)$ as an invariant.  This technique allows a simpler
 proof because, when working with integer arithmetic, it is not always
 the case that $\log_2(x-x/2) = \log_2(x)-1$.
-
-Generally, we observed that because the instrumented program
-is structurally same as the original one, it is enough to prove
-that the added assertions never fail in order to show the two programs
-satisfy the proposition (*).  This can usually be piggybacked
-on standard static-analysis tools.
-
-\begin{figure}
-%\vspace{-.3cm}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h@,lg@) {
-  if (h-l > 1) {
-    @assert(lg > 0);@
-    m = l + (h-l)/2;
-    @lg--;@ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h@,lg@);
-    tick($-M_\code{bsearch}$);
-  } else return l;
-}
-\end{lstlisting}
-%\vspace{-.2cm}
-\caption{Assisted bound derivation using logical state.
-  % We use the tick metric to model the stack consumption of the program.
-  % Before a function call to \code{bsearch}, a tick statement pays for
-  % the space used by stack frame of the function.  The space is recoverd
-  % after the call.
-  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The semi-automatically derived bound is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
 
@@
 
-\begin{figure*}
-\centering
-%\vspace{-.5cm}
-\begin{tabular}{r|c|c|c|c|c}
-%
-& t09 & t19 & t30 & t15 & t13\\
-%
-\hline
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-i=1; j=0;
-while (j<x) {
-  j++;
-  if (i>=4)
-    i=1, tick(40);
-  else i++;
-  tick(1); }
-\end{lstlisting}
-&
-\hspace{-.5cm}\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (i>100) {
-  i--; tick(1);
-} i += k+50;
-while (i>=0) {
-  i--; tick(1);
-}\end{lstlisting}\hspace{-.5cm}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (x>0) {
-  x--;
-  t=x, x=y, y=t;
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-assert(y>=0);
-while (x > y) {
-  x -= y+1;
-  for (z=y; z>0; z--)
-    tick(1);
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (x>0) {
-  x--;
-  if (*) y++;
-  else
-    while (y>0)
-      y--, tick(1);
-  tick(1); }
-\end{lstlisting}
-%
-\\
-\hline
-%
-%
-%
-\toolname
-  & {\small $11|[0,x]|$ }
-  & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
-  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
-  & {\small $|\inter 0 x|$}
-  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
-\\
-%
-Rank
-  & {\small $23{\cdot}x - 14$ }
-  & {\small $54+k+i$ }
-  & {\small ---}
-  & {\small $2+2x-y$ }
-  & {\small $0.5{\cdot}y^2 {+} yx \ldots$}
-           % +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
-\\
-%
-{\hspace{-.2cm}LOOPUS}
-&
-{\small
-  $41\max(x,0)$
-}
-&
-{\small
-  $\!\!\!\!$\begin{tabular}{c}
-    $\max(i{-}100, 0)$ \\
-    $+ \max(k {+} i {+} 51, 0)$
-  \end{tabular}$\!\!\!\!$
-}
-&
-{\small ---}% $1 + \max(x, y)$ (???)
-&
-{\small ---}
-&
-{\small \begin{tabular}{c}
-           $2\max(x, 0)$ \\
-           $+ \max(y, 0)$
-        \end{tabular}}
-\end{tabular}
-\vspace{.01cm}
-\caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  \iffull{The output of Rank has been manually
-  simplified to fit the table.}{}  }
-\label{fig:compar}
-\end{figure*}
-
 \paraskip
@@
 To base the soundness proof on a formal ground, we start by defining a
-new cost-aware operational semantics for Clight.  Clight's
+new cost-aware operational semantics for CompCert Clight.  Clight's
 operational semantics is based on small-step transitions and
@@
 We define quantitative Hoare triples
-as $B;R\vdash_L \htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
-maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
+as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
+maps from program states to an element of $\Q\cup\{\infty\}$ that represents
 an amount of resources available.  The assertions $B$ and $R$ are
@@
 S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
-the empty continuation, and at least $Q(\state)$ resource units available
+the empty continuation, $Q'(\state) \ge 0$ for all state $\sigma$, and at least $Q(\state)$ resource units available
 then the evaluation does not run out of resources and there are at
-least $Q'(\state')$ resources left if the evaluation terminates in
+least $Q'(\state')$ resources left if the evaluation terminates in state
 $\state'$.  The logic rules are similar to the ones in previous
-work and generalized to account for the cost introduced by  our
-cost-aware semantics.
-
-Finally, we define a strong compositional continuation-based soundness for
+work.
+
+Finally, we define a strong compositional continuation based soundness for
 triples and prove the validity of all the rules in Coq.  \ifshort{The full
@@
 
-The translation of an assertion $(\Gamma; Q)$ in the automatic
+The translation of an assertion $(\Gamma, Q)$ in the automatic
 analysis is defined by
 $$
-  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
+  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
 $$
@@
 translate the assumptions in function contexts of the automatic
-analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
-\forall z \, \vec v \, v . $
-%
-$$
+analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=$
+%
+\begin{multline*}
+\forall z \, \vec v \, v . \,
 (
-  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
+  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]), \\
   \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
   )
-$$}{}
-%
-These definitions let us translate the judgement
-$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
+\end{multline*}}{}%
+%
+\iffull{These definitions let}{This lets} us translate the judgement
+$J := B; R; P \vdash S \dashv P'$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
 \iffull{$$
@@
 $$
-  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
+  \tr(J) := \tr(B); \tr(R) \vdash \htriple{\tr(P)} S {\tr(P')}.
 $$}%
-\noindent{}The soundness of the automatic analysis can now be stated
+The soundness of the automatic analysis can now be stated
 formally with the following theorem.
@@
 of the automatic analysis directly to its counterpart in the
-quantitative logic.  The trickiest parts are the translations of the
+quantitative logic.  The most tricky parts are the translations of the
 rules for increments and decrements and the rule {\sc Q:Weak} for
 weakening because they make essential use of the algebraic properties
-of the potential functions.
+of our linear potential functions.
 \iffull{ For instance we prove the following lemma to show the soundness
@@
 
+\begin{figure*}
+\centering
+\vspace{-.5cm}
+\begin{tabular}{r|c|c|c|c|c}
+%
+& t09 & t19 & t30 & t15 & t13\\
+%
+\hline
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+i=1; j=0;
+while (j<x) {
+  j++;
+  if (i>=4)
+    i=1, tick(40);
+  else i++;
+  tick(1); }
+\end{lstlisting}
+&
+\hspace{-.5cm}\begin{lstlisting}[basicstyle=\tt\footnotesize]
+while (i>100) {
+  i--; tick(1);
+} i += k+50;
+while (i>=0) {
+  i--; tick(1);
+}\end{lstlisting}\hspace{-.5cm}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+while (x>0) {
+  x--;
+  t=x, x=y, y=t;
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+assert(y>=0);
+while (x > y) {
+  x -= y+1;
+  for (z=y; z>0; z--)
+    tick(1);
+  tick(1);
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[basicstyle=\tt\footnotesize]
+while (x>0) {
+  x--;
+  if (*) y++;
+  else
+    while (y>0)
+      y--, tick(1);
+  tick(1); }
+\end{lstlisting}
+%
+\\
+\hline
+%
+%
+%
+\toolname
+  & {\small $11|[0,x]|$ }
+  & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
+  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
+  & {\small $|\inter 0 x|$}
+  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
+\\
+%
+Rank
+  & {\small $23{\cdot}x - 14$ }
+  & {\small $54+k+i$ }
+  & {\small ---}
+  & {\small $2+2x-y$ }
+  & {\small $0.5{\cdot}y^2 {+} yx \ldots$}
+           % +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
+\\
+%
+{\hspace{-.2cm}LOOPUS}
+&
+{\small
+  $41\max(x,0)$
+}
+&
+{\small
+  $\!\!\!\!$\begin{tabular}{c}
+    $\max(i{-}100, 0)$ \\
+    $+ \max(k {+} i {+} 51, 0)$
+  \end{tabular}$\!\!\!\!$
+}
+&
+{\small ---}% $1 + \max(x, y)$ (???)
+&
+{\small ---}
+&
+{\small \begin{tabular}{c}
+           $2\max(x, 0)$ \\
+           $+ \max(y, 0)$
+        \end{tabular}}
+\end{tabular}
+\vspace{.01cm}
+\caption{ Comparison of resource bounds derived by different tools on
+  several examples with linear bounds.  \iffull{The output of Rank has been manually
+  simplified to fit the table.}{}  }
+\label{fig:compar}
+\end{figure*}
+
 \sectskip
@@
 decremented down to 100 and in the second loop $i$ is decremented further
-down to $-1$.  However, between the loops we assign \lstinline{i += k+50}.
+down to $-1$.  However, between the loops we assign \code{i=i+k+50}.
 So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
@@
 zero in an unconventional way.  In the loop body, first $x$ is decremented
-by one, then the values of the variables $x$ and $y$ are switched using
+by one.  Then the values of the variables $x$ and $y$ are switched using
 the local variable $t$ as a buffer.  Our analysis infers the tight
@@
    \end{lstlisting}
-%\vspace{-1.2cm}
+\vspace{-1.2cm}
 \hspace{3.8cm}
@@
     \end{center}
-%\vspace{-0.1cm}
+\vspace{-0.1cm}
   \end{minipage}
@@
 \begin{table}[t]
-\nocaptionrule \caption{Comparison of $\toolname$ with other
-  automatic tools.}
-\vspace{6pt}
-\label{tab:compar}
 \small
@@
 \end{tabular}
-\vspace{-6pt}
+\vspace{.01cm}
+\caption{Comparison of $\toolname$ with other
+  automatic tools.}
+\label{tab:compar}
 \end{table}
 
-\pref{tab:compar} summarizes the results of our experiments presented
-in \pref{tab:eval}.  It shows
+\begin{table}[t]
+\centering
+\small
+\begin{tabular}{r|c|c|c}
+Function & LoC & Bound & Time (s) \\
+\hline
+\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
+\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
+\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
+\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
+\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
+\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
+\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
+\code{MD5Final} & 195 & 141 & 0.22 \\
+\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
+% \code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
+\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
+\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
+\hline
+\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
+\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
+\end{tabular}
+\vspace{.1cm}
+\caption{Derived bounds for functions from cBench.}
+\label{tab:cbench}
+\end{table}
+
+\pref{tab:compar} summarizes the results of our experiments.  It shows
 for each tool the number of derived bounds (\#bounds), the number of
@@
 
-With our formal cost semantics, we can run our examples
-for different inputs and measure the cost to compare it
+Since we have a formal cost semantics, we can run our examples with
+this semantics for different inputs and measure the cost to compare it
 to our derived bound.  \pref{fig:3d} shows such a comparison for
@@
 
-\begin{table}[t]
-\nocaptionrule \caption{Derived bounds for functions from cBench.}
-\vspace{6pt}
-\label{tab:cbench}
-\centering
-\small
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{MD5Final} & 195 & 141 & 0.22 \\
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-% \code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
-\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
-\hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-\end{tabular}
-\vspace{-4pt}
-\end{table}
-
-\sectskip
-\section{Limitations}
-\label{sec:limitations}
-\aftersectskip
-
-Our implementation does not currently support
-all of Clight.  Programs with function pointers, goto statements,
-continue statements, and pointers to stack-allocated variables
-cannot be analyzed automatically.  While these limitations concern the
-current implementation, our technique is in principle capable to handle
-them.
-
-For the sake of simplicity, the automated system described here is
-restricted to finding only linear bounds.  However, the amortized analysis
-technique was shown to work with polynomial bounds~\cite{HoffmannAH11};
-we leave this extension of our system as future work.
-
-Even certain linear programs cannot be analyzed automatically by
-$\toolname$, it is usually the case for programs that rely on heap
-invariants (like nul-terminated C strings), for programs in which
-resource usage depends on the result of non-linear operations (like
-\code{\%} or \code{*}) in a non-trivial way, or for programs whose
-termination can only be proved by complex path-sensitive reasoning.
+\begin{figure}[t]
+\center
+% \vspace{-.2cm}
+\includegraphics[width=.9\linewidth]{fig/bound3d}
+\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
+  (blue lines) and the measured runtime cost (red crosses) for Example
+  \emph{t08}. For $x\ge 0$ the bound is tight.}
+\label{fig:3d}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-\center
-\vspace{-.2cm}
-\includegraphics[width=.9\linewidth]{fig/bound3d}
-\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for Example
-  \emph{t08}. For $x\ge 0$ the bound is tight.}
-\label{fig:3d}
-\vspace{-10pt}
-\end{figure}
-
 A recent project~\cite{veristack14} has implemented and verified a
@@
 analysis for C that can derive parametric bounds for loops and
-recursive functions fully automatically. We
-use a more general quantitative Hoare
-logic that is parametric over the resource of interest.
+recursive functions fully automatically.
+%
+In this article, we also use a more general quantitative Hoare
+logic that is parametric in the resource of interest.
 
@@
 and LOOPUS.  As demonstrated, $\toolname$ is more compositional than
-the aforementioned tools. Our technique, is the only one that can
-generate resource specifications for functions, deal with resources
-like memory that might become available, generate proof certificates
-for the bounds, and  support user guidance that separates
+the aforementioned tools. Our technique, is the only one that is able
+to generate resource specifications for functions, deal with resources
+like memory that might become available, generates proof certificates
+for the bounds, and has support for user guidance that separates
 qualitative and quantitative reasoning.
@@
 framework combines ideas from existing abstract interpretation--based
-techniques with the potential method of amortized analysis.  It is
-implemented in the publicly available tool $\toolname$. To the best of
-our knowledge, $\toolname$ is the first tool for C programs
-that automatically reduces the derivation of symbolic bounds to LP
-solving.
+techniques with the potential method of amortized analysis.  It is implemented in the publicly available tool
+$\toolname$. To the best of our knowledge, $\toolname$ is the first
+analysis tool for C programs that automatically reduces the derivation
+of symbolic bounds to LP solving.
 
@@
 
-\acks
-
-We thank members of the FLINT team at Yale and anonymous referees for
-helpful comments and suggestions that improved this paper and the
-implemented tools.  This research is based on work supported in part
-by NSF grants 1319671 and 1065451, DARPA grants FA8750-10-2-0254 and
-FA8750-12-2-0293, and ONR Grant N00014-12-1-0478.  Any opinions,
-findings, and conclusions contained in this document are those of the
-authors and do not reflect the views of these agencies.
+% \acks
+
+% This research is based on work supported in part by NSF grants 1319671
+% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
+% Any opinions, findings, and conclusions contained in this document are
+% those of the authors and do not reflect the views of these agencies.
 
@@
 
+\paragraph{Semantics.}
+
+\pref{fig:opsem} contains the reduction rules of the semantics.  The rules
+define a rewrite system for program configurations of the form $(\state,
+S, \cont, \cost)$, where $\state$ is the program state, $S$ is the
+statement being executed, $\cont$ is a continuation that describes what
+remains to be done after the execution of $S$, and $\cost \in \Q$ is
+the amount of resources available for further execution.  All the rules
+have the implicit side condition that the resource quantity available
+\emph{before} the step is non-negative.  This means that we allow
+$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side transition
+relation $\smallstep^*$ to indicate that the execution ran out of
+resources.  However, every execution that reaches such a state is stuck.
+
+A continuation $\cont$ represents the context of the execution of a
+given statement.  A continuation can be the empty continuation
+$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
+\cont$, a loop continuation $\Kloop S \cont$, or the continuation
+$\Kcall r \theta \cont$ of a function call.
+$$
+\cont := \Kstop
+\mid \Kseq S \cont
+\mid \Kloop S \cont
+\mid \Kcall r \theta \cont
+$$
+%
+During the execution we assume a fixed function context $\fenv$ that
+maps function names to a list of variables and the statement that
+defines the function body.  It is used in the rule \textsc{S:Call}.
+
+The intuitive meaning of an evaluation $(\state,
+S, \cont, \cost) \smallstep^* (\state', S', \cont', \cost')$ is the
+following.  If the statement $S$ is executed in program state
+$\state$, with continuation $\cont$, and with $\cost$ resources
+available then---after a finite number of steps---the evaluation will
+reach the new machine state $(\state', S', \cont',\cost')$ and there
+are $\cost'$ resources available.  If $\cost' \ge 0$ then the
+execution did not run out of resources and the resource consumption up
+to this point is $\cost - \cost'$.  If this difference is negative
+then resource became available during the execution.  If however
+$\cost' < 0$ then the execution ran out of resources and is stuck.
+The cost of the execution is then $\cost \ge 0$.
+
+\pref{lem:sem1} summarizes the main properties of the resource
+counter of the semantics.  It can be proved by induction on the number
+of steps.
+\begin{lemma}
+\label{lem:sem1}
+\itemskip
+  \begin{enumerate}
+  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
+    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
+    \smallstep^k (\state', S', \cont', \cost' {+} n)$.
+\itemskipIn
+  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
+    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep^k (\state', S',
+    \cont', d')$ then $\cost - \cost' = d - d'$.
+  \end{enumerate}
+\itemskip
+\end{lemma}
+
+% OOPS this is no longer true... This is an argument agains a presentation
+% with ticks only. XXX
+%
+% If we assume that we have a \emph{positive resource metric}, that is,
+% a metric that assigns a positive cost to each evaluation step then we
+% can link bounds on the resource consumption under the metric to
+% termination.  This is formalized by \pref{lem:sem_term}.
+%
+% \begin{lemma}
+%  \label{lem:sem_term}
+%  Let $M$ be a positive metric and assume that there exists $b \in
+%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep^*
+%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
+%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
+%  \smallstep^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
+%  and $\state''$ implies $n \le B$.
+% \end{lemma}
+
 \section{Quantitative Hoare Logic}
@@
 { }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
-  { M_b + B }{ \code{break} }{ Q }
-}
+  { B }{ \code{break} }{ Q }
+}
+\and
+\Rule{L:Tick}
+{ n < 0 \implies Q \ge 0}
+{ \Delta; B; R \ldash
+\htriple
+  { Q }
+  { \code{tick}(n) }
+  { Q - n }
+}
+%
 %
@@
 \and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
+\Rule{L:Assert}
 { }
-{ \Delta; B; R \vdash
+{ \Delta; B; R \ldash
 \htriple
-  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+  { \code{istrue}~\sem{e}_\state \implies Q }
+  { \code{assert}~e }
+  { Q }
 }
@@
 \and
-\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1 }{ Q' + M_s }
+\Rule{L:Update}
+{ }
+{ \Delta; B; R \ldash
+\htriple
+  { \lambda \state.\, Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
+}
+%
+\and
+\Rule{L:Loop}
+{ I \ge Q
+\\ \Delta; Q; R \ldash
+   \htriple { I }{ S }{ I }
+}
+{ \Delta; B; R \ldash
+\htriple
+  { I }{ \code{loop}~S }{ Q }
+}
+%
+\and
+\Rule{L:Seq}
+{ \Delta; B; R \ldash
+\htriple{ P }{ S_1 }{ Q' }
+\\\\
+  \Delta; B; R \ldash
+\htriple{ Q' }{ S_2 }{ Q }
+}
+{ \Delta; B; R \ldash
+\htriple{ P }{ S_1;S_2 }{ Q }
+}
+%
+\and
+\Rule{L:If}
+{% P \ge 0
+%\\
+  \Delta; B; R \ldash
+  \htriple
+  { \code{istrue}~\sem{e}_\state + P }
+  { S_1 }{ Q }
 \\
-\!\!  \Delta; B; R \vdash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S_1;S_2 }{ Q }
+  \Delta; B; R \ldash
+\htriple
+  { \code{isfalse}~\sem{e}_\state + P }
+  { S_2 }{ Q }
+}
+{ \Delta; B; R \ldash
+\htriple
+  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
 }
@@
   \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \models P_f \, y \, (\state(\vec x)) \land A \\
-  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \vdash
+  P \ge P_f \, y \, (\state(\vec x)) + A \\
+  \forall v.\, (Q_f \, y \, v + A \ge \lambda \state.\, Q \, \state[r \mapsto v])
+}
+{ \Delta; B; R \ldash
 \htriple
-  { M_f + P }
+  { P }
   { r \gets f(\vec x) }
-  { Q - M_r }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-  { \code{assert}~e }
   { Q }
-}
-%
-\and
-\Rule{L:Tick}
-{ }
-{ \Delta; B; R \vdash
-\htriple
-  { Q + M_t(n) }
-  { \code{tick}(n) }
-  { Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ \Delta; Q; R \vdash
-\htriple
-  { I }{ S }{ I + M_l }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{ \Delta; B; R \vdash
-\htriple
-  { \code{istrue}~\sem{e}_\state + P - M_c^1}
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \vdash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \vdash
-\htriple
-  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \!\! P \models P' \!\! \\
-  \!\! \Delta; B'; R' \vdash
-  \htriple{ P' }{ S }{ Q' } \!\! \\
-  \!\! Q' \models Q \!\! \\
-  \!\! B' \models B \!\! \\
-  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
-}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \vdash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \vdash
-\htriple{ P + x }{ S }{ Q + x } \!\!
 }
@@
 \Rule{L:Extend}
-{ \Delta \cup \Delta'; B; R \vdash
+{  \Delta \cup \Delta'; B; R \ldash
   \htriple{ P }{ S }{ Q }
 \\
+  P_f \ge 0
+\\\\
   \forall f\,P_f\,Q_f.\,
@@
 %
+\and
+\Rule{L:Weaken}
+{
+  \Delta; B'; R' \ldash \htriple{ P' }{ S }{ Q' } \\\\
+  P \ge P' \!\! \\
+  \!\! Q' \ge Q \!\! \\
+  \!\! B' \ge B \!\! \\
+  \!\! \forall v.\, (R'\,v \ge R\,v)
+}
+{ \Delta; B; R \ldash
+\htriple{ P }{ S }{ Q }
+}
+%
+\and
+\Rule{L:Frame}
+{ \Delta; B; R \ldash
+\htriple{ P }{ S }{ Q } \\
+  x \in \Qplusz
+}
+{ \!\! \Delta; B + x; R + x \ldash
+\htriple{ P + x }{ S }{ Q + x } \!\!
+}
+%
 \end{mathpar}
-
 \caption{Rules of the Quantitative Hoare Logic}
@@
 We already gave an intuition of the meaning of judgements
-derived in the logic.  To make it formal, we define the
-\emph{resource safety} $\safe n P S \cont$ of an assertion
-and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
-\begin{align*}
-  (m {\le} n \land P(\state) {\le} \cost \land
-    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
-  \implies \cost' {\ge} 0.
-\end{align*}
-This predicate is step indexed by an integer $n$ that is used for
-induction in the soundness proof for the function-call and loop cases.
-The constraint $\cost' \geq 0$ imposed by the definition ensures that
-the program does not stop because of a resource error (recall that a
-negative resource counter on the right-hand side is a resource
-failure).  However, it does not rule out memory safety errors or
-assertion failures. This is because our logic does not prove any
-safety or correctness theorems but only focuses on resource usage.
-%
-An interesting detail in the definition is the natural number $m$.  We
-simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
-\cont $.  This would not be the case if we replaced all occurrences of
-$m$ by $n$ in the definition of validity.
+derived in the logic.  We start by defining an indexed
+predicate $\cdot \downarrow_n$ that defines a \emph{resource-safe}
+program configuration:
+\begin{multline*}
+  (\state, S, K, c) \downarrow_n \, := \\
+  \forall m \le n.\, \forall c'.\,
+    (\state, S, K, c) \smallstep^m
+    (\_, \_, \_, c') \implies 0 \le c'.
+\end{multline*}
+This predicate means that a program configuration can execute
+for a certain number of steps without running out of resources.
+Note that nothing is said about the program safety, i.e. a stuck
+configuration $C$ will satisfy $\forall n.\, C\downarrow_n$. This
+is because our logic does not prove any safety or correctness
+theorems but only focuses on resource usage.
+We can now be used to define the resource safety
+of a configuration with respect to a quantitative assertion.
+\begin{multline*}
+  \safe n P S \cont \, := \\
+  \forall \state\, c.\, P(\state) \le c \implies
+    0 \le c \wedge (\state, S, K, c) \downarrow_n.
+\end{multline*}
+Once again, we use an index in the definition of the predicate, it
+is used to perform proofs by induction for the function-call and
+loop rules of the logic.
+%
+An interesting and essential property of the previous predicate is
+that $\safe {n+1} P S \cont \implies \safe n P S \cont$.  To prove
+this weakening property, we use the quantification over all numbers
+smaller than $n$ in the definition of $\cdot \downarrow_n$.
 
@@
 programs that can be automatically analyzed by our system.  For
-simplicity, and to compare our analysis with existing tools, we use a
-cost metric that counts the number of loop iterations and function
-calls.  This is the only cost metric supported by tools like
+simplicity, and to compare our analysis with existing tools,
+the examples are automatically instrumented with tick statements
+to count the number of loop iterations and function calls.
+This is the only resource that can be measured by tools like
 KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
-we also use the \emph{tick metric} if we want to discuss features such as
-resource restitution.  The tick metric assigns cost $n$ to the
-statement \code{tick(n)} and cost $0$ to all other statements. Of
-course, the examples can also be analyzed with any other cost metric
-in our system.
+we use tick statements explicitely to discuss features such as
+resource restitution.
 
@@
 
+\begin{table*}
+\footnotesize
+
+\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
+\newcommand{\file}[2][c]{%
+  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
+\renewcommand{\max}[0]{{\rm mx}}
+\renewcommand{\min}[0]{{\rm mn}}
+
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\centering
+\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
+
+File &
+\multicolumn{2}{|c|}{KoAT} &
+\multicolumn{2}{|c|}{Rank} &
+\multicolumn{2}{|c|}{LOOPUS} &
+SPEED &
+\toolname
+\\
+\hline
+
+%--------------------------------------------
+\hline \texttt{gcd.c} &
+
+? &
+&
+
+% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
+$(((2{+}1)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$|\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{kmp.c} &
+
+? &
+&
+
+% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
+$(((2{+}(n{+} \dots$ &
+$O(n^2)$ &
+
+% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
+$\max(n, 0) \dots$ &
+$O(n)$ &
+
+? &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \texttt{qsort.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$1 {+} 2 |\inter 0 {\code{len}}|$
+\\
+
+%--------------------------------------------
+% \hline \file{\dir{speed\_pldi09}{fig1.c}} &
+%
+% ? &
+% &
+%
+% --- &
+% &
+%
+% $2 \max(n, 0)$ &
+% $O(n)$ &
+%
+% God damn it &
+%
+% $1 + 2 |\inter 0 n|$
+% \\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_2.c}} &
+
+--- &
+&
+
+% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}n)\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + n$ &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_4.c}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\frac n m + m$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi09}{fig4\_5.c}} &
+
+$28d + 7g + 27$ &
+$O(n)$ &
+
+% $((((1+1)+(-1+n))+m)+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+$\max(n, n-m)$ &
+
+---
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi10}{ex1.c}} &
+
+--- &
+&
+
+--- &
+&
+
+--- &
+&
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi10}{ex3.c}} &
+
+--- &
+&
+
+% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
+$(((2{+}(-1\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_pldi10}{ex4.c}} &
+
+$110 a + 33$ &
+$O(n)$ &
+
+--- &
+&
+
+--- &
+&
+
+$n + 1$ &
+
+$1 {+} 2 |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{fig2\_1.c}} &
+
+$9a + 9b + \dots$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{+}((-y\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, m{-}y)$ &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{fig2\_2.c}} &
+
+$6a + 9b + 3c + 5$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
+$((2{-}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
+$\max(0, (x + 1 {-}z)\dots$ &
+$O(n)$ &
+
+$\max(0, n{-}x) + \max(0, n{-}z)$ &
+
+$|\inter x n| {+} |\inter z n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{nstd\_multiple.c}} &
+
+--- &
+&
+
+% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
+$((2{-}x{+}n\dots$ &
+$O(n^2)$ &
+
+$\max(0, m {-} y) + \max(0, n {-} x)$ &
+$O(n)$ &
+
+$\max(0, m{-}y) + \max(0, n{-}x)$ &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{nstd\_single.c}} &
+
+$48 b + 16$ &
+$O(n)$ &
+
+% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
+$(((1{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(0, \! n {-} 1)\dots$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{sqntl\_single.c}} &
+
+$21 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{}-x{+}n\dots$ &
+$O(n)$ &
+
+$2{\cdot}\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{smpl\_multiple.c}} &
+
+$9 c + 10 d + 7$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$n + m$ &
+
+$|\inter 0 m| {+} |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{smpl\_single2.c}} &
+
+$20 d + 12 c + 17$ &
+$O(n)$ &
+
+--- &
+&
+
+$\max(n, 0) + \max(m, 0)$ &
+$O(n)$ &
+
+$n + m$ &
+
+$|\inter 0 n| {+} |\inter 0 m|$
+\\
+
+%--------------------------------------------
+\hline \file{\dir{speed\_popl10}{smpl\_single.c}} &
+
+$4 b + 6$ &
+$O(n)$ &
+
+% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
+$((2{-}x{+}n\dots$ &
+$O(n)$ &
+
+$\max(n, 0)$ &
+$O(n)$ &
+
+$n$ &
+
+$|\inter 0 n|$
+\\
+
+%--------------------------------------------
+\hline \file{\texttt{t07.c}} &
+
+? &
+&
+
+$2+x$ &
+$O(n)$ &
+
+% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
+$\max(x, 0)\dots$ &
+$O(n)$ &
+
+? &
+
+$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t08.c} &
+
+? &
+&
+
+% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
+$((2{+}z{-}y\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
+$\max(0, \! y{-}2)\dots$ &
+$O(n)$ &
+
+? &
+
+$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t10.c} &
+
+? &
+&
+
+% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
+$((2{-}y{+}x\dots$ &
+$O(n)$ &
+
+$\max(0, x {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t11.c} &
+
+? &
+&
+
+% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
+$((2{-}y{+}m\dots$ &
+$O(n)$ &
+
+$\max(0, n {-} x)) + \max(0, m {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter x n| {+} |\inter y m|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t13.c} &
+
+? &
+&
+
+% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
+$(((1{+}y^2/2\dots$ &
+$O(n^2)$ &
+
+$2{\cdot}\max(x, 0) + \max(y, 0)$ &
+$O(n)$ &
+
+? &
+
+$2 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t15.c} &
+
+? &
+&
+
+% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
+$((1{+}x\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$|\inter 0 x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t16.c} &
+
+? &
+&
+
+% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
+$((-99{\cdot}y\dots$ &
+$O(n)$ &
+
+--- &
+&
+
+? &
+
+$101 |\inter 0 x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t19.c} &
+
+? &
+&
+
+% $(((1+1)+(151+k))+(-100+i))+1$ &
+$((153{+}k\dots$ &
+$O(n)$ &
+
+$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
+$O(n)$ &
+
+? &
+
+$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t20.c} &
+
+? &
+&
+
+% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
+$(2{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
+$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
+$O(n)$ &
+
+? &
+
+$|\inter x y| {+} |\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t27.c} &
+
+? &
+&
+
+--- &
+&
+
+% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
+$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
+$O(n)$ &
+
+? &
+
+$0.01|\inter n y| {+} 11 |\inter n 0|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t28.c} &
+
+? &
+&
+
+% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
+$((1{-}y{+}x\dots$ &
+$O(n)$ &
+
+% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
+$10^3 \,\max(0, x - y)\dots$ &
+$O(n)$ &
+
+? &
+
+$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t30.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$|\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t37.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t39.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$1.33 {+} 0.67 |\inter z y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t46.c} &
+
+? &
+&
+
+--- &
+&
+
+--- &
+&
+
+? &
+
+$|\inter 0 y|$
+\\
+
+%--------------------------------------------
+\hline \texttt{t47.c} &
+
+? &
+&
+
+$4+n$ &
+$O(n)$ &
+
+$1 + \max(n, 0)$ &
+$O(n)$ &
+
+? &
+
+$1 {+} |\inter 0 n|$
+\\
+
+%--------------------------------------------
+\end{tabular}
+\caption{
+  Experimental evaluation comparing the bounds generated KoAT, Rank,
+  LOOPUS, SPEED, and our automatic amortized analysis on several
+  challenging linear examples. Results for KoAT and SPEED were extracted
+  from previous publications because KoAT cannot take C programs as
+  input in its current version and SPEED is not available.
+  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
+  marked with --- indicate that the tool failed to produce a result.
+}
+\label{tab:eval}
+\end{table*}
+
 \paragraph{Well-Known Algorithms}
 
-\pref{fig:cat3} shows well-known algorithms that can be automatically
+\pref{fig:wknown} shows well-known algorithms that can be automatically
 analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
@@
 \sectskip
+\scriptsize
 \bibliography{lit}
 }
-
-\begin{table*}[h!]
-\appendix
-\section{Complete Experimental Results for the Tool Comparison}
-
-\renewcommand{\max}[0]{{\rm mx}}
-\nocaptionrule \caption{
-  Comparison of the bounds generated by KoAT, Rank,
-  LOOPUS, SPEED, and our tool $\toolname$ on several
-  challenging linear examples. Results for KoAT and SPEED were extracted
-  from previous publications~\cite{GulwaniJK09,GulwaniMC09,GulwaniZ10,SinnZV14} because KoAT cannot take C programs as
-  input in its current version and SPEED is not available.
-  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
-  marked with --- indicate that the tool failed to produce a result. We write
-  $\max(a,b)$ for the maximum of $a$ and $b$.
-}
-\vspace{6pt}
-\label{tab:eval}
-\footnotesize
-
-\newcommand{\dir}[2]{\code{#1} \\ \code{#2}}
-\newcommand{\file}[2][c]{%
-  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
-
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\centering
-\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
-
-Function &
-\multicolumn{2}{c|}{KoAT} &
-\multicolumn{2}{c|}{Rank} &
-\multicolumn{2}{c|}{LOOPUS} &
-SPEED &
-\toolname
-\\
-
-%--------------------------------------------
-\hhline{=|==|==|==|=|=} \code{gcd} &
-
-? &
-&
-
-% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
-$(((2{+}1)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{kmp} &
-
-? &
-&
-
-% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
-$(((2{+}(n{+} \dots$ &
-$O(n^2)$ &
-
-% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
-$\max(n, 0) \dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \code{qsort} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1 {+} 2 |\inter 0 {\code{len}}|$
-\\
-
-%--------------------------------------------
-% \hline \file{\dir{SPEED\ PLDI09}{fig1}} &
-%
-% ? &
-% &
-%
-% --- &
-% &
-%
-% $2 \max(n, 0)$ &
-% $O(n)$ &
-%
-% God damn it &
-%
-% $1 + 2 |\inter 0 n|$
-% \\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ pldi09}{fig4\ 2}} &
-
---- &
-&
-
-% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}n)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + n$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ pldi09}{fig4\ 4}} &
-
---- &
-&
-
-% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + m$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ pldi09}{fig4\ 5}} &
-
-$28d + 7g + 27$ &
-$O(n)$ &
-
-% $((((1+1)+(-1+n))+m)+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, n-m)$ &
-
----
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ pldi10}{ex1}} &
-
---- &
-&
-
---- &
-&
-
---- &
-&
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ pldi10}{ex3}} &
-
---- &
-&
-
-% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ pldi10}{ex4}} &
-
-$110 a + 33$ &
-$O(n)$ &
-
---- &
-&
-
---- &
-&
-
-$n + 1$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{fig2\ 1}} &
-
-$9a + 9b + \dots$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{+}((-y\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, m{-}y)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{fig2\ 2}} &
-
-$6a + 9b + 3c + 5$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
-$((2{-}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
-$\max(0, (x + 1 {-}z)\dots$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, n{-}z)$ &
-
-$|\inter x n| {+} |\inter z n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{nested\ multiple}} &
-
---- &
-&
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n^2)$ &
-
-$\max(0, m {-} y) + \max(0, n {-} x)$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, m{-}y) $ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{nested\ single}} &
-
-$48 b + 16$ &
-$O(n)$ &
-
-% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
-$(((1{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(0, \! n {-} 1)\dots$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{sequential\ single}} &
-
-$21 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{}-x{+}n\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{simple\ multiple}} &
-
-$9 c + 10 d + 7$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 m| {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{simple\ single2}} &
-
-$20 d + 12 c + 17$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 n| {+} |\inter 0 m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\ popl10}{simple\ single}} &
-
-$4 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\code{t07}} &
-
-? &
-&
-
-$2+x$ &
-$O(n)$ &
-
-% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
-$\max(x, 0)\dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{t08} &
-
-? &
-&
-
-% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
-$((2{+}z{-}y\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
-$\max(0, \! y{-}2)\dots$ &
-$O(n)$ &
-
-? &
-
-$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{t10} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
-$((2{-}y{+}x\dots$ &
-$O(n)$ &
-
-$\max(0, x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \code{t11} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x)) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \code{t13} &
-
-? &
-&
-
-% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
-$(((1{+}y^2/2\dots$ &
-$O(n^2)$ &
-
-$2{\cdot}\max(x, 0) + \max(y, 0)$ &
-$O(n)$ &
-
-? &
-
-$2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{t15} &
-
-? &
-&
-
-% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
-$((1{+}x\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \code{t16} &
-
-? &
-&
-
-% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
-$((-99{\cdot}y\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$101 |\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \code{t19} &
-
-? &
-&
-
-% $(((1+1)+(151+k))+(-100+i))+1$ &
-$((153{+}k\dots$ &
-$O(n)$ &
-
-$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
-$O(n)$ &
-
-? &
-
-$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
-\\
-
-%--------------------------------------------
-\hline \code{t20} &
-
-? &
-&
-
-% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
-$(2{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
-$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x y| {+} |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \code{t27} &
-
-? &
-&
-
---- &
-&
-
-% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
-$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
-$O(n)$ &
-
-? &
-
-$0.01|\inter n y| {+} 11 |\inter n 0|$
-\\
-
-%--------------------------------------------
-\hline \code{t28} &
-
-? &
-&
-
-% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
-$((1{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
-$10^3 \,\max(0, x - y)\dots$ &
-$O(n)$ &
-
-? &
-
-$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \code{t30} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{t37} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{t39} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1.33 {+} 0.67 |\inter z y|$
-\\
-
-%--------------------------------------------
-\hline \code{t46} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \code{t47} &
-
-? &
-&
-
-$4+n$ &
-$O(n)$ &
-
-$1 + \max(n, 0)$ &
-$O(n)$ &
-
-? &
-
-$1 {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\end{tabular}
-\end{table*}
 

### 2015-07-27 paper/main.tex
import rules from pldi branch

@@
 
-\iffull{
 \begin{figure*}
+\small
+\def \MathparLineskip {\lineskip=0.35cm}
 \begin{mathpar}
-%
-\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+\Rule{Q:Skip}
 { }
-{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
-%
-\and
-\Rule{Q:Break}
+{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
+%
+\and \Rule{Q:Assert}
 { }
-{ (\Gamma, Q_B); R; (\Gamma, Q_B) \vdash \code{break} \dashv (\Gamma', Q') }
-%
-\and
-\Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
-{ n < 0 \implies Q \ge 0 }
-{ B; R; (\Gamma, Q) \vdash \code{tick}(n) \dashv (\Gamma, Q{-}n) }
-%
-\and
-\Rule{Q:Return}
+{ B, R \vdash \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
+%
+\and \Rule{Q:Break}
+{ }
+{(\Gamma; Q_B), R \vdash \htriple{ \Gamma; Q_B {+} M_b }{ \code{break} }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Return}
 { P = Q_R[\Vret/x]
@@
 }
-{ B; (\Gamma_R,Q_R); (\Gamma, Q) \vdash \code{return}~x \dashv (\Gamma',Q') }
-%
-\and \Rule{Q:Update}
-{ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
-%
-\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
-{ Q \ge Q'
-\\ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q)
-}
-{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+{ B, (\Gamma_R; Q_R) \vdash \htriple{ \Gamma; Q }{ \code{return}~x }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Loop}
+{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
+{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
+%
+\and \Rule{Q:Seq}
+{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
+\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
+%
+\and \Rule{Q:If}
+{ B, R \vdash \htriple{ \Gamma {\land} e; Q {-} M_c^1) }{ S_1 }{ \Gamma'; Q' }
+\\\\ B, R \vdash \htriple{ \Gamma {\land} \neg e; Q {-} M_c^2 }{ S_2 }{ \Gamma'; Q' }
+}
+{ B, R \vdash \htriple{ \Gamma; Q{+}M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ \Gamma'; Q' } }
+%
+\\ \Rule{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
+}
 %
 \and \Rule{Q:Inc}
-{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
-\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
 }
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:Dec}
-{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{xu}, -q_{ux})
-\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{ux}, -q_{xu})
+  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
+}
+%
+\and \Rule{Q:Set}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ B, R \vdash \htriple{ \Gamma[x/y]; Q {+} M_u {+} M_e(y) }{ x \gets y }{ \Gamma; Q' } }
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
 }
 { \textstyle
-  B; R; (\Gamma[x/x {-} y], Q) \vdash x \gets x - y \dashv (\Gamma, Q')
-}
-%
-\and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
-%
-\and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q')
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
-}
-{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+  B, R \vdash \htriple{ \Gamma[x/x {-} y]; Q {+} M_u {+} M_e(x {-} y) }{ x \gets x - y }{ \Gamma; Q' }
+}
 %
@@
 }
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
-  \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
-}
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
-{ }
-{ B; R; (\Gamma, Q) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ B, R \vdash \htriple
+  { \Gamma_f[\Vargs / \vec x] {\land} \Gamma_{\text{Loc}}; Q {+} c {+} M_f }
+  { r \gets f(\vec x) }
+  { \Gamma_f'[\Vret/r] {\land} \Gamma_{\text{Loc}}; Q' {+} c {-} M_r }
+}
+%
+\and \Rule{Q:Extend}
 { \fenv f = (\vec y, S_f)
-\\ Q_f \ge 0
-\\ Q'_f \ge 0
-\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
+\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{ \Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y] }{ S_f }{ \Gamma'; Q' }
 }
@@
 \and \Rule{Q:Weak}
-{  \!\! \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2
 \\ \Gamma'_2 \models \Gamma'_1
@@
 }
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
-%
-\and \Rule{Relax}
+{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
+%
+\\ \Rule{Relax}
 { \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
 \\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\ \forall i.\, p_i, r_i \in \Qplusz
 \\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
@@
 { Q' \succeq_\Gamma Q }
+%\vspace{-.3cm}
 \end{mathpar}
-\caption{Inference rules of the quantitative analysis.}
+\caption{Selected inference rules of the quantitative analysis.}
 \label{fig:auto}
 \end{figure*}
-}{}
-
-\jan{QUENTIN FIX ABOVE FIGURE}
 

### 2015-07-28 paper/main.tex
finish importing pldi fixes

--- 2015-07-27 paper/main.tex
@@
 \usepackage{array} %for stupid tables
+\usepackage{hhline}
 
@@
 
-\titlebanner{\iffull{Extended Version}{}}%{Draft -- Do not distribute}        % These are ignored unless
-\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.
+\titlebanner{\iffull{Extended version}{}}%{Draft -- Do not distribute}        % These are ignored unless
+\preprintfooter{\iffull{YALEU/DCS/TR-1505\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.
 
@@
 to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
-assertions are only used in the rule for the assignment
+$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
+logical assertions from potential functions with semicolons. Note that the
+logical assertions are only used in the rule for the assignment
 \code{x=x+K}.
@@
 
-\iffull{
 \begin{figure*}
@@
 %
-\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
+\Rule{Q:Skip}
 { }
-{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
+{ B, R \vdash \htriple {\Gamma; Q}{\code{skip}}{\Gamma; Q} }
 %
@@
 { }
-{ (\Gamma, Q_B); R; (\Gamma, Q_B) \vdash \code{break} \dashv (\Gamma', Q') }
+{ (\Gamma; Q_B), R \vdash \htriple{\Gamma; Q_B}{\code{break}}{\Gamma'; Q'} }
 %
 \and
-\Rule[leftskip=.3cm,rightskip=.4cm]{Q:Tick}
+\Rule{Q:Tick}
 { n < 0 \implies Q \ge 0 }
-{ B; R; (\Gamma, Q) \vdash \code{tick}(n) \dashv (\Gamma, Q{-}n) }
+{ B, R \vdash \htriple{\Gamma; Q}{\code{tick}(n)}{\Gamma; Q{-}n} }
 %
@@
 }
-{ B; (\Gamma_R,Q_R); (\Gamma, Q) \vdash \code{return}~x \dashv (\Gamma',Q') }
+{ B, (\Gamma_R; Q_R) \vdash \htriple{\Gamma; Q}{\code{return}~x}{\Gamma'; Q'} }
 %
@@
 }
-{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
-%
-\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
+{ B, R \vdash \htriple{\Gamma[x/y]; Q}{x \gets y}{\Gamma; Q'} }
+%
+\and \Rule{Q:Loop}
 { Q \ge Q'
-\\ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q)
-}
-{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
+\\ (\Gamma'; Q'), R \vdash \htriple{\Gamma; Q}{ S }{\Gamma;  Q}
+}
+{ B, R \vdash \htriple{\Gamma; Q}{\code{loop}~S}{\Gamma'; Q'} }
 %
@@
 { \textstyle
-  B; R; (\Gamma[x/x {+} y], Q) \vdash x \gets x + y \dashv (\Gamma, Q')
+  B, R \vdash \htriple{\Gamma[x/x {+} y]; Q}{x \gets x + y}{\Gamma; Q'}
 }
@@
 { \textstyle
-  B; R; (\Gamma[x/x {-} y], Q) \vdash x \gets x - y \dashv (\Gamma, Q')
+  B, R \vdash \htriple{\Gamma[x/x {-} y]; Q}{x \gets x - y}{\Gamma; Q'}
 }
@@
 \and \Rule{Q:If}
-{ B; R; (\Gamma \land e, Q) \vdash S_1 \dashv (\Gamma',Q')
-\\\\ B; R; (\Gamma \land \neg e, Q) \vdash S_2 \dashv (\Gamma',Q')
-}
-{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
+{ B, R \vdash \htriple{\Gamma \land e; Q}{S_1}{\Gamma'; Q'}
+\\\\ B, R \vdash \htriple{\Gamma \land \neg e; Q}{S_2}{\Gamma'; Q'}
+}
+{ B, R \vdash \htriple{\Gamma; Q}{\code{if}(e)~S_1~\code{else}~S_2}{\Gamma'; Q'} }
 %
 \and\Rule{Q:Seq}
-{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q')
-\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
-}
-{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
+{  B, R \vdash \htriple{\Gamma; Q}{S_1}{\Gamma'; Q'}
+\\\\ B, R \vdash \htriple{\Gamma'; Q'}{S_2}{\Gamma''; Q''}
+}
+{ B, R \vdash \htriple{\Gamma; Q}{S_1;S_2}{\Gamma''; Q''} }
 %
 \and\Rule{Q:Call}
-{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
+{ (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
 \\ \text{Loc} = \text{Locals}(Q) \!
@@
 }
-{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
-  \vdash r \gets f(\vec x) \dashv
-  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
-}
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Assert}
+{ B, R \vdash \htriple {\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}; Q {+} c}
+  {r \gets f(\vec x)}
+  {\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}; Q' {+} c}
+}
+%
+\and \Rule{Q:Assert}
 { }
-{ B; R; (\Gamma, Q) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
-%
-\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
+{ B, R \vdash \htriple{\Gamma; Q}{\code{assert}~e}{\Gamma \land e; Q} }
+%
+\and \Rule{Q:Extend}
 { \fenv f = (\vec y, S_f)
@@
 \\ Q'_f \ge 0
-\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
-}
-{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
+\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{\Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y]}{S_f}{\Gamma'; Q'}
+}
+{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
 %
 \and \Rule{Q:Weak}
-{  \!\! \Gamma_1 \models \Gamma_2
+{   \Gamma_1 \models \Gamma_2
 \\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
+\\ B, R \vdash \htriple{\Gamma_2;Q_2}{S}{\Gamma'_2; Q'_2}
 \\ \Gamma'_2 \models \Gamma'_1
@@
 }
-{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
+{ B, R \vdash \htriple{\Gamma_1; Q_1}{S}{\Gamma'_1; Q'_1} }
 %
@@
 \end{figure*}
-}{}
-
-\jan{QUENTIN FIX ABOVE FIGURE}
 
@@
 rules of the automatic amortized analysis.
+
+\paraskip
+\paragraph{Cost Aware Clight.}
+
+We present the rules for a subset of Clight.
+Clight is the first intermediate
+language of the CompCert compiler~\cite{CompcertCACM}.  It
+is a subset of C with a unified looping construct and
+side-effect free expressions.  We reuse most of CompCert's
+syntax but instrument the semantics with a resource counter
+(an arbitrary rational number) that can be modified by
+\code{tick()} function calls.
+More details on this cost-aware semantics are provided in \pref{sec:soundness}.
+
+In the rules, assignments are restricted to the form $x \gets
+y$ or $x \gets x \pm y$.  In the implementation, a Clight
+program is converted into this form prior to analysis without changing
+the resource cost.
+Non-linear operations such as $x \gets z*y$ or $x \gets
+a[y]$ are handled by assigning $0$ to coefficients
+like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
+This sound treatment ensures that no further loop bounds
+depend on the result of the non-linear operation.
 
@@
 % Function contexts are treated exactly as in the logic.  % We call
-% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
-% % $(\Gamma', Q')$.
+% % precondition the pair $(\Gamma; Q)$ and postcondition the pair
+% % $(\Gamma'; Q')$.
 % The correspondence between judgements of the
@@
 
-We describe the automatic amortized analysis for a subset of
-expressions of Clight.  Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}, it
-is a cleaner C with a unified looping construct and
-side-effect free expressions.
-Assignments are restricted to the form $x \gets
-y$ or $x \gets x \pm y$.  In the implementation, a Clight
-program is converted into this form prior to analysis without changing
-the resource cost.  This is achieved by using a series of \emph{cost-free
-  assignments} that do not result in additional cost in the
-semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-a[y]$ are handled by assigning $0$ to coefficients
-like $q_{xa}$ and $q_{ax}$ that depend on $x$ after the assignment.
+% We describe the automatic amortized analysis for a subset of
+% expressions of Clight.  Clight is the first intermediate
+% language of the CompCert compiler~\cite{CompcertCACM}, it
+% is a cleaner C with a unified looping construct and
+% side-effect free expressions.
+% Assignments are restricted to the form $x \gets
+% y$ or $x \gets x \pm y$.  In the implementation, a Clight
+% program is converted into this form prior to analysis without changing
+% the resource cost.  This is achieved by using a series of \emph{cost-free
+%   assignments} that do not result in additional cost in the
+% semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
+% a[y]$ are handled by assigning $0$ to coefficients
+% like $q_{xa}$ and $q_{ax}$ that depend on $x$ after the assignment.
+% This sound treatment ensures that no further loop bounds
+% depend on the result of the non-linear operation.
 
@@
 rule.  This rule reuses its precondition as postcondition, it is
-explained by two observations: first, no resource is consumed by the skip
-operation, thus no potential has to be used to pay for the evaluation;
-second, the program state is not changed by
-the execution of a skip statement, thus all the potential available
+explained by two observations: First, no resource is consumed by the skip
+operation, thus no potential has to be used to pay for the evaluation.
+Second, the program state is not changed by
+the execution of a skip statement.  Thus all potential available
 before the execution of the skip statement is still available after.
@@
 described later.}{}
-
-% Most of the rules correspond exactly to the respective rules of the
-% quantitative Hoare logic.
-% \iffull{
-%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
-%   as precondition.  This is sound since evaluating a $\code{skip}$
-%   statement never costs any resources and does not change the program
-%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
-%   quantitative annotation of the precondition is at least $M_t(n)$ to
-%   pay for the resource consumption.  Since the program state is also
-%   unchanged in this case, the logical part $\Gamma$ of the
-%   precondition can soundly be reused in the postcondition.  In the
-%   rule {\sc Q:Break}, where the linear control flow is broken, the
-%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
-%   taken from the break part of the judgement similarly as it is done
-%   in the quantitative logic.
-% }
-% %
-% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
-%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
-%   rules can be found in the extended version.
-% }
-% %
-% The rule {\sc Q:Return} exhibits a slight
-% difference to the quantitative logic.  While we have a function $\Z
-% \to \Assn$ to represent return values in the logic, we assume that a
-% special variable $\ret$ is part of the index set of $Q_R$.  The
-% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
-% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
-% $q_i'=q_i$ otherwise.
 
@@
 change on $q_{y0}$.
+
+The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
+loop body, one needs to find an invariant potential $Q$ that will pay for
+the iterations.  At each iteration, some resources are spent by the loop
+body.  Since the  loop can only be exited with a \code{break} statement,
+the postcondition $\{ \Gamma'; Q' \}$ for the statement
+$\code{loop}~S$ is used as break postcondition in the derivation for $S$.
+
+Another interesting rule is {\sc Q:Call}.  It needs to account for the
+changes to the stack caused by the function call, the arguments/return
+value passing, and the preservation of local variables.  We can sum up
+the main ideas of the rule as follows.
+\itemskip
+\begin{itemize}
+\item The potential in the pre- and postcondition of the function
+  specification is equalized to its matching potential in the callee's
+  pre- and postcondition.
+\itemskipIn
+\item
+  The potential of intervals $|\inter x y|$ is preserved
+  across a function call
+  if $x$ and $y$ are local\iffull{ variables.}{.}
+\item
+  The unknown potentials after the call (e.g.
+  $|\inter x g|$, with $x$ local and $g$ global)
+  are set to zero in the postcondition.
+\end{itemize}
+\itemskip
+If $x$ and $y$ are local variables and $f(x,y)$ is
+called, {\sc Q:Call} splits the potential of
+$|\inter x y|$ in two parts.  One part to perform the
+computation in the function $f$ and one part to keep for
+later use after the function call.  This splitting is
+realized by the equations $Q = P {+} S$ and $Q' =
+P' {+} S'$.  Arguments in the function precondition
+$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
+of names different from all program variables.  This
+prevents name conflicts and ensures that
+the substitution $[\Vargs/\vec x]$ is meaningful.
+Symmetrically, we use the unique name $\Vret$ to represent
+the return value in the function's postcondition
+$(\Gamma'_f; Q'_f)$.
+
+The rule {\sc Q:Weak} is not syntax directed. In the implementation
+we apply {\sc Q:Weak} before loops and between the two statements of a
+sequential composition. We could integrate weakening into every
+syntax directed rule but this simple heuristic helps to make the analysis
+efficient.  The high-level idea of {\sc Q:Weak} is the following: If
+we have a sound judgement, then it is sound to add more potential to the
+precondition and remove potential from the postcondition.  The concept of
+\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
+that is defined in the rule {\sc Relax}. This rule
+also deals with the important task of transferring constant potential
+(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
+from the logical context that the interval size $|\inter x y| \geq \ell$ is
+larger than a constant $\ell$ then we can turn the potential $q_{xy} {\cdot}
+|\inter x y|$ form the interval into the constant potential $\ell{\cdot}q_{xy}$
+and guarantee that we do not gain potential.  Conversely, if $|\inter
+x y| \leq u$ for a constant $u$ then we can transfer constant potential
+$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
+without gaining potential.
 
@@
 
-\iffull{
-Another interesting rule is {\sc Q:Call}.  It needs to account for the
-changes to the stack caused by the function call, the arguments/return
-value passing, and the preservation of local variables.  We can sum up
-the main ideas of the rule as follows.
-\itemskip
-\begin{itemize}
-\item The potential in the pre- and postcondition of the function
-  specification is equalized to its matching potential in the callee's
-  pre- and postcondition.
-\itemskipIn
-\item
-  $\!$The potential of intervals $|\inter x y|$ is preserved
-  if $x$ and $y$ are local\iffull{ variables.}{.}
-\item
-  The unknown potentials after the call (e.g.
-  $|\inter x g|$, with $x$ local and $g$ global)
-  are set to zero in the postcondition.
-\end{itemize}
-\itemskip
-If $x$ and $y$ are local variables and $f(x,y)$ is
-called, {\sc Q:Call} splits the potential of
-$|\inter x y|$ in two parts, one part to perform the
-computation in the function $f$ and one part to keep for
-later use after the function call.  This splitting is
-realized by the equations $Q = P {+} S$ and $Q' =
-P' {+} S'$.  Arguments in the function precondition
-$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
-of names different from all program variables.  This
-prevents name conflicts from happening and ensures that
-the substitution $[\Vargs/\vec x]$ is meaningful.
-Symmetrically, we use the unique name $\Vret$ to represent
-the return value in the function's postcondition
-$(\Gamma'_f, Q'_f)$.
-}{}
-
-\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition. We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}\iffull{ that
-also deals with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq l$ is
-larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}
-
 \sectskip
@@
     tick($-M_\code{bsearch}$);
-  } else return l;
+  } return l;
 }
@@
 $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
-logical variables is selected so that:
-$$
+logical variables is selected to satisfy the proposition:
+\begin{equation} \tag{*}
 \forall n\, \sigma\, \sigma'.\,
@@
 Generally, we observed that because the instrumented program
-is structurally same as the original one, it is sufficient to prove
-that the added assertions never fail to show that they have identical
-resource consumption.  This can be piggybacked on standard
-static-analysis tools.
+is structurally same as the original one, it is enough to prove
+that the added assertions never fail in order to show the two programs
+satisfy the proposition (*).  This can usually be piggybacked
+on standard static-analysis tools.
 
@@
 We define quantitative Hoare triples
-as $B;R\vdash\htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
+as $\iffull{\Delta;\,}{}B;R\vdash_L \htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
 maps from program states to an element of $\Q\cup\{\infty\}$ that represents
@@
 
-The translation of an assertion $(\Gamma, Q)$ in the automatic
+The translation of an assertion $(\Gamma; Q)$ in the automatic
 analysis is defined by
 $$
-  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
+  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
 $$
@@
 translate the assumptions in function contexts of the automatic
-analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=$
+analysis.  We define $\tr(\Gamma_f; Q_f, \Gamma'_f; Q'_f) :=$
 %
@@
 (
-  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]), \\
-  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
+  \lambda \vec v \, \state .\, \tr(\Gamma_f; Q_f)(\state[\Vargs {\mapsto} \vec v]), \\
+  \lambda v \, \state .\, \tr(\Gamma'_f; Q'_f)(\state[\Vret {\mapsto} v])
   )
@@
 %
-\iffull{These definitions let}{This lets} us translate the judgement
-$J := B; R; P \vdash S \dashv P'$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
+These definitions let us translate the judgement
+$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
 \iffull{$$
-  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
+  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash_L
     \htriple{\tr(P)}{S}{\tr(P')}.
@@
 decremented down to 100 and in the second loop $i$ is decremented further
-down to $-1$.  However, between the loops we assign \code{i=i+k+50}.
+down to $-1$.  However, between the loops we assign \lstinline{i=i+k+50}.
 So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
@@
 
-Since we have a formal cost semantics, we can run our examples with
-this semantics for different inputs and measure the cost to compare it
+With our formal cost semantics, we can run our examples
+for different inputs and measure the cost to compare it
 to our derived bound.  \pref{fig:3d} shows such a comparison for
@@
 \sectskip
+\section{Limitations}
+\label{sec:limitations}
+\aftersectskip
+
+Our implementation does not currently support
+all of Clight.  Programs with function pointers, goto statements,
+continue statements, and pointers to stack-allocated variables
+cannot be analyzed automatically.  While these limitations concern the
+current implementation, our technique is in principle capable to handle
+them.
+
+For the sake of simplicity, the automated system described here is
+restricted to finding only linear bounds.  However, the amortized analysis
+technique was shown to work with polynomial bounds~\cite{HoffmannAH11};
+we leave this extension of our system as future work.
+
+Even certain linear programs cannot be analyzed automatically by
+$\toolname$, it is usually the case for programs that rely on heap
+invariants (like nul-terminated C strings), for programs in which
+resource usage depends on the result of non-linear operations (like
+\code{\%} or \code{*}) in a non-trivial way, or for programs whose
+termination can only be proved by complex path-sensitive reasoning.
+
+\sectskip
 \section{Related Work}
@@
 analysis for C that can derive parametric bounds for loops and
-recursive functions fully automatically.
-%
-In this article, we also use a more general quantitative Hoare
-logic that is parametric in the resource of interest.
+recursive functions fully automatically.  We
+use a more general quantitative Hoare
+logic that is parametric over the resource of interest.
 
@@
 
-$\max(0, m{-}y) + \max(0, n{-}x)$ &
+$\max(0, n{-}x) + \max(0, m{-}y)$ &
 
@@
 \caption{
-  Experimental evaluation comparing the bounds generated KoAT, Rank,
-  LOOPUS, SPEED, and our automatic amortized analysis on several
+  Comparison of the bounds generated by KoAT, Rank,
+  LOOPUS, SPEED, and our tool $\toolname$ on several
   challenging linear examples. Results for KoAT and SPEED were extracted

### 2015-09-06 cfg/main.tex
cfg rules

--- 2015-07-28 paper/main.tex
@@
-% Comment for anonymization
-% Add preprint option for the extended version
-% Also look for "SHORT" comments
- \def\fullversion{}
-
-\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}
-
-% Structure:
-%
-%    Abstract                              1c
-% 1) Intro                                 3c
-% 2) The Potential Method                  1c
-% 3) Compositional Resource-Bound Analysis 4c
-%    - Gulwani/ examples
-% 4) Derivation Rules                      2c
-% 5) Automatic Inference via LP Solving    2c
-% 6) Logical State and User Interaction    2c
-% 7) Soundness Proof                       2c
-%    1) Cost Semantics for Clight
-%    2) Quantitative Hoare Logic
-% 8) Experimental Evaluation               2c
-% 9) Related Work                          1c
-% a) Conclusion                            1c
-
-\newcommand{\iffull}[2]{\ifx\fullversion\undefined{#2}\else{#1}\fi}
-\newcommand{\ifshort}[2]{\ifx\fullversion\undefined{#1}\else{#2}\fi}
-
-\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
-\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
-\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
-\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
-\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
-\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
-\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}
-
-%SHORT
-\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
-\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
-\captionsetup[figure]{format=myformat}
-\captionsetup[table]{format=myformat}
-\renewcommand{\captionlabelfont}{\bf}
-%SHORT
+\documentclass[letterpaper,11pt]{article}
+
+% \documentclass[preprint]{sigplanconf}
+\usepackage[numbers]{natbib}
 
@@
 \usepackage{mathpartir} %for typing rules
+%\usepackage{graphicx} %to include png images
 \usepackage{microtype} %better micro typing
-\usepackage{stmaryrd} %for llbracket
-\usepackage{mathabx} % for boxes
-\usepackage{graphicx} %to include png images
-\usepackage{xcolor} %for colors
-\usepackage{url}
-\usepackage{enumitem}
-\usepackage{array} %for stupid tables
-\usepackage{hhline}
-
-%----------------------------------------------------
+\usepackage[standard,amsmath,thmmarks]{ntheorem}
+\usepackage{mathabx} %for blackdiamond
 
@@
 \newcommand{\ppref}[1]{\vpageref[]{#1}}
-\newrefformat{fig}{Figure~\ref{#1}}
-\newrefformat{app}{Appendix~\ref{#1}}
-\newrefformat{tab}{Table~\ref{#1}}
-\newrefformat{cha}{Challenge~\ref{#1}}
-\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}
-
-%----------------------------------------------------
-
-\usepackage{amsthm}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}{Proposition}
-
-%----------------------------------------------------
 
 \input{definitions}
+
+%------------Draft---------------------------------------
+\usepackage{eso-pic}
+\usepackage{xcolor}
+\AddToShipoutPicture{\setlength{\unitlength}{1in}\put(4.25,10.0){\makebox[0pt][c]{\normalfont\textsf{\textbf{\large \color{gray} Draft -- \today}}}}}
+%--------------------------------------------------------
+
+\newcommand{\jan}[1]{{\color{blue}{[Jan: #1]}}}
+
+\title{Automated Amortized Resource Bound Analysis \\ for Control Flow Graphs}
+
+\author{Quentin Carbonneaux \and Jan Hoffmann}
+
 \begin{document}
 
-\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
-
-%SHORT
-\ifx\fullversion\undefined
-\abovedisplayskip=3pt
-\belowdisplayskip=3pt
-\abovedisplayshortskip=3pt
-\belowdisplayshortskip=3pt
-\else
-\fi
-%SHORT
-
-\conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
-\copyrightyear{2014}
-\copyrightdata{XXX-X-XXXX-XXXX-X/XX/XX}
-
-%----------------------------------------------------
-
-\newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}
-
-\titlebanner{\iffull{Extended version}{}}%{Draft -- Do not distribute}        % These are ignored unless
-\preprintfooter{\iffull{YALEU/DCS/TR-1505\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.
-
-\title{Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}}
-% Certified Resource Bounds for C
-% Compositional Certified Resource Bounds for C
-% Compositional Quantitative Resource Analysis of C Programs
-% Compositional Resource-Bound Certification for C Programs
-
-\authorinfo
-{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
-{Yale University \vspace{-1cm}}
-
 \maketitle
 
-\begin{abstract}
-  This paper presents a new approach for automatically deriving
-  worst-case resource bounds for C programs.
-  %
-  The described technique combines ideas from amortized analysis and
-  abstract interpretation in a unified framework to address four
-  challenges for state-of-the-art techniques: compositionality, user
-  interaction, generation of proof certificates, and scalability.
-  %
-  \emph{Compositionality} is achieved by incorporating the potential
-  method of amortized analysis.  It enables the derivation of global
-  whole-program bounds with local derivation rules by naturally
-  tracking size changes of variables in sequenced loops and function
-  calls.  The resource consumption of functions is described abstractly
-  and a function call can be analyzed without access to the function
-  body.
-  %\
-  \emph{User interaction} is supported with a new mechanism that
-  clearly separates qualitative and quantitative verification. A user
-  can guide the analysis to derive complex non-linear bounds by using
-  auxiliary variables and assertions.  The assertions are
-  separately proved using established qualitative techniques such as
-  abstract interpretation or Hoare logic.
-  %
-  \emph{Proof certificates} are automatically generated from the local
-  derivation rules. A soundness proof of the derivation system with
-  respect to a formal cost semantics guarantees the validity of the
-  certificates.
-  %
-  \emph{Scalability} is attained by an efficient reduction of
-  bound inference to a linear optimization problem that can be solved
-  by off-the-shelf LP solvers.
-  %
-  The analysis framework is implemented in the publicly-available
-  tool $\toolname$. An experimental evaluation demonstrates the
-  advantages of the new technique with a comparison of $\toolname$ with
-  existing tools on challenging micro benchmarks and the analysis of
-  more than $\totalLoc$ lines of C code from the cBench benchmark suite.
-
-% Contrary to most available tools for
-% C, ours has a compositional design that provides the
-% programmer with bounds expressed in terms of meaningful
-% variables.  The compositionality allows to handle uniformly
-% nested sequenced loops and recursive sequenced function
-% calls.  The automatic analysis is implemented
-% as a tool that leverages state-of-the-art SMT solvers
-% LP solvers to automatically infer resource bounds on
-% a wide range of C programs.  It is the first automatic
-% amortized analysis that can derive bounds that depend
-% on negative integers and differences between numbers.
-% To prove its soundness and give a concrete shape to proof
-% certificates, we developed a quantitative Hoare logic
-% that we proved sound using the Coq proof assistant.  This
-% logic allows manual proofs of arbitrarily complex resource
-% bounds on C programs and interaction with the automation.
-% To prove the scalability and precision of this new tool,
-% we run it on multiple code snippets from the literature,
-% reproduce results from previous work, and analyze a
-% consequent part of a modern benchmark suite.
-
-%   The goal of quantitative resource analysis is to provide developers
-%   with quantitative information about the runtime behavior of software
-%   at development time.  Recent years have seen tremendous progress in
-%   automatically deriving worst-case resource bounds, yet many
-%   challenges in specifying, interactively deriving, formally
-%   certifying, and composing resource bounds remain unsolved.
-%
-%   This paper describes a novel quantitative resource analysis
-%   framework that tackles these challenges for C programs.  The
-%   analysis framework consists of three parts: First, a parametric cost
-%   semantics formalizes the resource consumption of CompCert Clight
-%   programs.  Second, a quantitative Hoare logic enables users to
-%   interactively develop resource bounds in the Coq Proof Assistant.
-%   The logic is proved sound with respect to the cost semantics, and
-%   shallow embedding enables a derived bound to be any function that is
-%   definable in Coq.  Third, an automatic amortized resource analysis
-%   for Clight computes derivations in the quantitative Hoare logic.
-%   It is the first automatic amortized analysis that can derive bounds
-%   that depend on negative integers or differences between numbers,
-%   which is crucial to handle typical systems code.  Both, the
-%   quantitative logic and the automatic amortized analysis are
-%   naturally compositional and can be combined to semi-automatically
-%   derive global resource bounds.
-%   %
-%   An experimental evaluation demonstrates the practicality of the
-%   analysis framework.  The expressivity of the logic is shown by
-%   manually deriving customized bounds that are tailored to specific
-%   algorithms.  A comparison of the automatic amortized analysis with
-%   other automatic tools on 30 challenging loop and recursion patterns
-%   from the literature and open-source software shows that the bounds
-%   derived by the automatic amortized analysis are often more precise.
-\end{abstract}
-
-% \category{D.2.4}{Software Engineering}
-% {Software/Program Verification}
-% \category{F.3.1}{Logics and Meanings of Programs}
-% {Specifying and Verifying and Reasoning about Programs}
-
-% \terms Verification, Reliability
-
-% \keywords Formal Verification, Compiler Construction, Program Logics,
-% Stack-Space Bounds, Quantitative Verification
-
-\sectskip
-\section{Introduction}
-\label{sec:intro}
-\aftersectskip
-
-In software engineering and software verification,
-we often would like to have static information
-about the quantitative behavior of programs.
-For example, stack and heap-space bounds
-are important to ensure the reliability of
-safety-critical systems~\cite{Regehr05}.
-Static energy usage information is critical
-for autonomous systems and has applications in
-cloud computing~\cite{CohenZSL12,CarrollH10}.
-Worst-case time bounds can help create
-constant-time implementations that prevent
-side-channel attacks~\cite{KasperS09,BartheBCLP14}.
-Loop and recursion-depth bounds are used to
-ensure the accuracy of programs that are executed
-on unreliable hardware~\cite{CarbinMR13} and
-complexity bounds are needed to verify cryptographic
-protocols~\cite{BartheGB09}.  In general, quantitative
-resource information % at design time
-can provide useful
-feedback for developers.
-
-% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW
-
-Available techniques for {\em automatically} deriving worst-case
-resource bounds fall into two categories.  Techniques in the first
-category derive impressive bounds for numerical imperative programs,
-but are not compositional.  This is problematic if one needs to derive
-global whole-program bounds.
-%%%%%%
-Techniques in the second category derive
-tight whole-program bounds for programs with regular loop or
-recursion patterns that decrease the size of an individual variable or
-data structure. They are highly compositional, scale for large
-programs, and work directly on the syntax.  However,
-they do not support multivariate interval-based resource bounds
-(e.g., $x-y$) which are common in C programs. Indeed,
-it has been a long-time open problem to develop compositional
-resource analysis techniques that can work for typical imperative code
-with non-regular iteration patterns, signed integers, mutation, and non-linear
-control flow. % As a result, they are not suitable for
-% typical imperative code.
-
-Tools in the first category include
-SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
-PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
-and LOOPUS~\cite{SinnZV14}.
-They lack compositionality in at least two ways.
-First, they all base their analysis on some form of
-\emph{ranking function} or \emph{counter instrumentation}
-that is linked to a local analysis.  As a result, loop bounds are
-arithmetic expressions that depend on the values of
-variables just before the loop.
-This makes it hard to give a
-resource bound on a sequence of loops and function calls in
-terms of the input parameters of a function.
-Second, while all popular imperative programming languages
-provide a function or procedure abstraction, available
-tools are not able to abstract resource behavior;
-instead, they have to inline the procedure body to perform their
-analysis.
-
-Tools in the second category originate form the {\em potential method}
-of amortized analysis and type systems for functional
-programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
-definitions of object-oriented programs~\cite{Jost06} and
-data-structure predicates of separation logic~\cite{Atkey10} can play
-the role of the type system in imperative programs.
-However, a major weakness of existing potential-based techniques
-is that they can only associate \emph{potential}
-with individual program variables or data structures.
-For C programs, this
-fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
-$y-i$ decreases, but not $|i|$. %  Mutation is yet another problem.
-% While functional programs relate multiple fixed inputs (the typing
-% context) with one output (the expression typed), imperative programs
-% can change the value of variables in scope, so the potential of these
-% variables has to be updated after mutations to ensure soundness.
-
-% these
-% techniques fail for numerical imperative programs because they lack
-% fine-grained data-structure information to hook the analysis on.
-% Another issue is that existing techniques associate \emph{potential}
-% directly with single program variables.
-
-A general problem with existing tools (in both categories) is user interaction. When a tool
-fails to find a resource bound for a program, there is no possibility
-for sound user interaction to guide the tool during bound
-derivation.
-For example, there is no concept of manual proofs of
-resource bounds; and no framework can support composition
-of manually derived bounds with automatically inferred bounds.
-
-% Finally, the tools often operate on an abstract version of the and use
-% complex external tools from term rewriting and abstract
-% interpretation. This makes it hard to generate a proof certificate for
-% the bounds or even to prove the soundness of the tools with respect to
-% a formal cost semantics.
-
-This paper presents a new compositional framework for automatically
-deriving resource bounds on C programs.  This new
-approach is an attempt to unify the two aforementioned
-categories: It solves the compositionality issues of techniques
-for numerical imperative code by adapting amortized-analysis--based
-techniques from the functional world.
-%
-Our automated analysis is able to infer
-resource bounds on C programs with mutually-recursive functions and integer loops.
-%
-The resource behavior of functions can be summarized
-in a functional specification that can be used at every
-call site without accessing the function body.
-%
-To our knowledge this is the first technique based on
-amortized analysis that is able to derive bounds that depend on negative numbers
-and differences of variables.  It is also the first
-resource analysis technique for C that deals naturally with
-recursive functions and sequenced loops, and can handle resources that may
-become available during execution (e.g., when freeing
-memory).  Compared to more classical
-approaches based on ranking functions, our tool inherits
-the benefits of amortized reasoning.  Using only one
-simple mechanism, it handles:
-\itemskip
-\begin{itemize}
-\item interactions between \emph{sequential loops or
-  function calls} through size changes of variables,
-\itemskipIn
-\item \emph{nested loops} that influence each other
-  with the same set of modified variables,
-\itemskipIn
-\item and \emph{amortized bounds} as found, for example, in
-  the Knuth-Morris-Pratt algorithm for string search.
-\end{itemize}
-\itemskip
-The main innovations that make amortized analysis work
-on imperative languages are to base the analysis on a
-Hoare-like logic and to track multivariate quantities instead
-of program variables.  This leads to precise bounds
-expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
-intervals.
-%
-A distinctive feature of our analysis system is that
-it reduces linear bound inference  to a linear optimization problem
-that can be solved by off-the-shelf LP solvers.  This enables the
-efficient inference of global bounds for larger programs.
-%
-Moreover, our local inference rules automatically generate
-proof certificates that can be easily checked in linear time.
-
-The use of the potential method of amortized analysis makes user
-interaction possible in different ways. For one thing, we can directly
-combine the new automatic analysis with manually derived bounds in a
-previously-developed quantitative Hoare logic~\cite{veristack14} (see
-\pref{sec:soundness}).  For another thing, we describe a new mechanism that
-allows the separation of quantitative and qualitative verification
-(see \pref{sec:anno}).  Using this mechanism, the user can guide the
-analysis by using auxiliary variables and logical assertions that can
-be verified by existing qualitative tools such as Hoare logic or
-abstract interpretation.  In this way, we can benefit from existing
-automation techniques and provide a middle-ground between fully
-automatic and fully manual verification for bound derivation. This
-enables the semi-automatic inference of non-linear bounds, such as
-polynomial, logarithmic, and exponential bounds.
-
-We have implemented the analysis system in the tool
-\toolname and experimentally evaluated its effectiveness by
-analyzing system code and examples from the literature.
-\toolname has automatically derived global resource bounds for
-more than $\totalLoc$ lines of C code from the cBench benchmark suite.
-%
-\iffull{\pref{app:cat}}{The extended version
-  of this article~\cite{anon_extended}}{} contains more than $30$
-challenging loop and recursion patterns that we collected from open
-source software and the literature.  Our analysis can find
-asymptotically tight bounds for all but one of these patterns, and in
-most cases the derived constant factors are tight.  To compare $\toolname$
-with existing techniques, we tested our examples
-with tools such as KoAT~\cite{BrockschmidtEFFG14},
-Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
-show that the bounds that we derive are often more precise than those
-derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
-uses amortization techniques, is able to achieve a similar precision.
-
-Examples from cBench and micro benchmarks demonstrate the practicality and
-expressiveness of the user guided bound inference.  For example, we
-derive a logarithmic bound for a binary search function\iffull{, a bound that
-depends on the contents of an array to describe the exact cost of a
-function that finds a maximal element in an array, }{ }and a bound
-that amortizes the cost of $k$\iffull{ successive }{ }increments to a binary
-counter (see \pref{sec:anno}).
-
-In summary, we make the following \emph{contributions}.
-\itemskip
-\begin{itemize}
-\item We develop the first automatic amortized analysis for C
-  programs. It is naturally compositional, tracks size changes of
-  variables to derive global bounds, can handle mutually-recursive
-  functions, generates resource abstractions for functions, derives
-  proof certificates, and handles resources that may
-  become available during execution.
-\itemskipIn
-\item We show how to automatically reduce the inference of \emph{linear} resource
-    bounds to efficient LP solving.
-\itemskipIn
-\item We describe a new method of harnessing existing qualitative
-  verification techniques to guide the automatic amortized analysis
-  to derive \emph{non-linear} resource bounds with LP solving.
-\itemskipIn
-\item We prove the soundness of the analysis with respect to a
-  parametric cost semantics for C programs. The cost model can be
-  further customized with function calls (\code{tick(n)}) that
-  indicate resource usage. % This mechanism is also covered by the
-  % soundness proof.
-\itemskipIn
-\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
-\itemskipIn
-\item We present experiments with $\toolname$ on more
-  than $\totalLoc$ lines of C code. A detailed comparison shows that our
-  prototype is the only tool that can derive global bounds for larger
-  C programs while being as powerful as existing tools when deriving
-  linear local bounds for tricky loop and recursion patterns.
-\end{itemize}
-\itemskip
-%
-\iffull{}
-{\footnotesize \sl
-[To the reviewers: The auxiliary material contains an anonymized
-version of the extended article, and the source code of $\toolname$
-and all examples.] }
-
-\sectskip
-\section{The Potential Method}
-\aftersectskip
-
-The idea that underlies the design of our framework is amortized
-analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes
-on a starting state $\state$ and consumes $n$ resource units of
-some user-defined quantity.  We denote that by writing $(S, \state)
-\Downarrow_n \state'$ where $\state'$ is the program state after
-the execution.  The basic idea of amortized analysis is to define a
-\emph{potential function} $\Phi$ that maps program states to non-negative
-numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a program
-state such that $(S, \state) \Downarrow_n \state'$.  Then
-$\Phi(\state)$ is a valid resource bound.
-
-To obtain a compositional reasoning we also have to take into account the
-state resulting from a program's execution.  We thus use two potential
-functions, one that applies before the execution, and one that applies
-after.  The two functions must respect the relation $\Phi(\state)
-\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
-that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
-must provide enough \emph{potential} for both, paying for the resource
-cost of the computation and paying for the potential $\Phi'(\state')$ on
-the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
-\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
-\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
-This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
-Note that the initial potential function $\Phi$ provides an upper bound
-on the resource consumption of the whole program.  What we have observed
-is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
-$$
-\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
-n + \Phi'(\state') \; ,
-$$
-then we get the following familiar looking rule
-$$
-\RuleNolabel
-{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
-{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
-$$
-%
-This rule already shows a departure from classical techniques that are
-based on ranking functions.  Reasoning with two potential functions
-promotes compositional reasoning by focusing on the sequencing of
-programs.  In the previous rule, $\Phi$ gives a bound for
-$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
-was derived on $S_1$ only.
-%
-Similarly, other language constructs lead to rules for the potential
-functions that look very similar to Hoare logic or effect system
-rules.  These rules enable reasoning about resource usage
-in a flexible and compositional way, which, as a side effect, produces
-a certificate for the derived resource bound.
-
-% \subsection{Automating  Amortized Analysis}
-
-% To make amortized reasoning feasible on large code bases, we would
-% like to automatically infer the potential functions.  The first step
-% towards this goal is to reduce our search space by fixing the shape of
-% potential functions $\Phi$.  In this paper, we focus on a form that
-% can express tight bounds for many typical programs and allows
-% inference view \emph{linear programming}:
-% %
-% $$
-% \Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}}
-%   q_{(x,y)} \cdot |[\state(x),\state(y)]|
-% $$
-% where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  For now, we assume
-% that a program state is a simple map from program variables to integers
-% and we treat constants as program variables (that cannot be changed
-% by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
-% the sum.  % We then develop an inference rule for each syntactic construct
-% % that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
-% % the previous section.
-
-The derivation of a resource bound using potential functions is best
-explained by example.  If we use a cost model that assigns cost $n$ to
-the function call \code{tick(n)} and cost $0$ to all other operations
-then the cost of the following example can be bounded by
-$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
-  % the signs of the input variables in the examples.}
-\begin{lstlisting}[basicstyle=\tt\small]
-while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
-\end{lstlisting}
-To derive this bound, we start with the initial potential $\Phi_0 =
-|[x,y]|$, which
-we also use as the loop invariant.  For the loop body we have (like in
-Hoare logic) to derive a triple
-$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
-if we utilize the fact that $x<y$ at the beginning of the loop body.
-The reasoning then works as follows. We start with the potential
-$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
-we denote the updated version of $x$ after the assignment by $x'$ then
-the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
-after the assignment \code{x=x+1} holds.  This means that we have the
-potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
-\code{tick(1)} consumes one resource unit, we end up with potential
-$|[x,y]|$ after the loop body and have established the loop invariant
-again.
-
-\begin{figure}[t]
-  \centering
-\vspace{-1.5ex}
-\begin{lstlisting}[mathescape]
-${\color{blue}%
-  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
-while (x+K<=y) {
-  ${\color{blue}%
-  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
-  x=x+K;
-  ${\color{blue}%
-  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| \} }$
-  tick(T);
-  ${\color{blue}%
-  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
-}
-${\color{blue}%
- \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
-\end{lstlisting}
-  \caption{Derivation of a tight bound on the number of
-    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
-    $T>0$ are not program variables but denote concrete constants.
-    % The reasoning works uniformly for all such $K$ and
-    % $T$.
-  }
-  \label{fig:ex1}
-\end{figure}
-
-\pref{fig:ex1} shows a derivation of the bound
-$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
-version of Example~1 in which we increment $x$ by a constant $K>0$ and
-consume $T>0$ resources in each iteration.  The reasoning is similar
-to the one of Example~1 except that we obtain the potential
-$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
-logical assertions from potential functions with semicolons. Note that the
-logical assertions are only used in the rule for the assignment
-\code{x=x+K}.
-
-To the best of our knowledge, no other implemented tool for C is
-currently capable of deriving a tight bound on the cost of such a
-loop.  For $T=1$ (many systems focus on the number of loop iterations
-without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
-10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
-$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
-translate the program into a term-rewriting system by
-hand. We will show in the following sections that the
-potential method makes automatic bound derivation straightforward.
-
-% To automate the reasoning, we first introduce an unknown rational
-% variable for each factor in the potential functions.  We then use our
-% inference rules (see \pref{sec:AAA}) to emit linear constraints on
-% these variables that enforce that variable assignments that respect
-% the constraints correspond to sound potential annotations.  For
-% instance if $K = 1$ in \pref{fig:ex1} then we would have the
-% annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
-% |[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
-% p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
-% are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
-% $p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
-% q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
-% the coefficients of the initial potential function. LP solvers are
-% among the most mature constraint solving tools. As a result, the bound
-% derivation works very efficiently in practice.
-
-% \iffull{It might be surprising that we}{We} only track simple linear relations in
-% the logical context $\Gamma$.  While it would of course be possible to keep
-% track of more sophisticated assertions, this simple form is
-% sufficient for the examples we considered and we can efficiently
-% decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
-% assignment rule.  Nevertheless, this logical part of our analysis
-% system is orthogonal to the quantitative part and can be easily
-% extended if necessary.
-
-The concept of a potential function is a generalization of the
-concept of a ranking function.  A potential function can be used like
-a ranking function if we add the statement
-\code{tick(1)} to every back edge of the program (loops and function
-calls).  However, a potential function is more flexible.  For example,
-we can use a potential function to prove that Example~2 does not
-consume any resources.
-\begin{lstlisting}[basicstyle=\tt\small]
-#\!\!\!\!\!#while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
-
-#\!\!\!\!\!#while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
-\end{lstlisting}
-Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
-In both cases, we reason exactly like in the first version of the
-while loop to prove the bound.  Of course, such loops with different
-tick annotations can be seamlessly combined in a larger program.
-
-\iffull{In general, the potential based approach addresses several challenges
-in resource bound analysis. This includes amortized bounds in which
-individual loop iterations have different resource consumption,
-sequential loops or function calls that require the tracking of size
-changes of variables, and nested loops that interact by modifying the
-same set of variables.}{}
-
-\newlength{\progwidth}
-
-\begin{figure*}
- \setlength{\progwidth}{.22\linewidth}
-  \centering
-\vspace{-.3cm}
-\hspace{-0.4cm}
-  \begin{minipage}[b]{.18\linewidth}
-    \begin{center}
-   \begin{lstlisting}[]
-while (n>x) {
-  ${\color{blue}%
-  \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
-  if (m>y)
-    ${\color{blue}%
-    \{ m{>}y; \, |[x, n]| {+} |[y, m]| \} }$
-    y=y+1;
-    ${\color{blue}%
-    \{ \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
-  else
-    ${\color{blue}%
-    \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
-    x=x+1;
-    ${\color{blue}%
-    \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
-  ${\color{blue}%
-  \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
-  tick(1);
-} ${\color{blue}%
-  \{ \cdot; \, |[x, n]| {+} |[y, m]| \} }$
-   \end{lstlisting}
-\vspace{-2.5ex}
-$|[x, n]| + |[y, m]|$
-\\[.4\baselineskip]
-      {\bf speed\_1}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-
-while (x<n) {
-  ${\color{blue}%
-  \{ x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
-  if (z>x)
-    ${\color{blue}%
-    \{ x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
-    x=x+1;
-    ${\color{blue}%
-    \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
-  else
-    ${\color{blue}%
-    \{ z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
-    z=z+1;
-    ${\color{blue}%
-    \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
-  ${\color{blue}%
-  \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
-  tick(1);
-} ${\color{blue}%
-  \{ \cdot; \, |[x, n]| {+} |[z, n]| \} }$
-   \end{lstlisting}
-\vspace{-2.5ex}
-$|[x, n]| + |[z, n]|$
-\\[.4\baselineskip]
-      {\bf speed\_2}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (z-y>0) {
-  ${\color{blue}%
-  \{ y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-  y=y+1;
-  ${\color{blue}%
-  \{ \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-  tick(3);
-  ${\color{blue}%
-  \{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-}
-${\color{blue}%
-\{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-while (y>9) {
-  ${\color{blue}%
-  \{ y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-  y=y-10;
-  ${\color{blue}%
-  \{ \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-  tick(1);
-} ${\color{blue}%
-  \{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
-   \end{lstlisting}
-\vspace{-2.5ex}
-$3.1|[y,z]| + 0.1|[0,y]|$
-\\[.4\baselineskip]
-      {\bf t08a}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (n<0) {
-  ${\color{blue}%
-  \{ n{<}0; \,  P(n,y) \} }$
-  n=n+1;
-  ${\color{blue}%
-  \{ \cdot; \,  59 {+} P(n,y) \} }$
-  y=y+1000;
-  ${\color{blue}%
-  \{ \cdot; \, 9 {+} P(n,y) \} }$
-  while (y>=100 && *){
-    ${\color{blue}%
-    \{ y{>}99; \, 9 {+} P(n,y) \} }$
-    y=y-100;
-    ${\color{blue}%
-    \{ \cdot; \, 14 {+} P(n,y) \} }$
-    tick(5);
-  } ${\color{blue}%
-  \{ \cdot; \, 9 {+} P(n,y) \} }$
-  tick(9);
-} ${\color{blue}%
-  \{ \cdot; \, P(n,y) \} }$
-   \end{lstlisting}
-\vspace{-2.5ex}
-$59|[n,0]| {+} 0.05|[0,y]|$
-\\[.4\baselineskip]
-      {\bf t27}
-    \end{center}
-  \end{minipage}
-\vspace{1ex}
-\caption{Derivations of bounds on the number of ticks for challenging
-  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
-  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
-  \emph{t08a} contains \emph{sequential loops} so that the iterations
-  of the second loop depend on the first, and \emph{t27} contains
-  interacting \emph{nested loops}. \iffull{In the potential functions, we only
-  mention the non-zero terms and in the logical context $\Gamma$ we
-  only mention assertions that we use.}{} In
-  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
-  0.05|[0,y]|$.}
-  \label{fig:ex_list}
-\end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\sectskip
-\section{Compositional Resource-Bound Analysis}
-\label{sec:AAA}
-\aftersectskip
-
-In this section we describe the high-level design of the automatic
-amortized analysis that we implemented in $\toolname$. Examples explain
-and motivate our design decisions.
-
-\paraskip
-\paragraph{Linear Potential Functions.}
+\section{Control Flow Graphs}
+
+\paragraph{Straight Line Statements}
+
+Straight line statements (also just statements) are inductively
+defined by the following grammar.
+%
+\begin{align*}
+S &:= \code{tick}(n) \mid \code{assert}~E
+\mid \code{skip}
+\mid x \gets E
+\mid x \gets f(x^*)
+\mid S;S
+\end{align*}
+%
+Expressions $E$ are left abstract in our presentation.  For our
+analysis framework, it is only important that they are side-effect
+free. We write $\statements$ for the set of statements.
+
+\paragraph{Control Flow Graphs}
+
+A control flow graph (CFG) is a set a graph $G=(V,E,\labels)$ with
+statement-labeled nodes. Here, $V$ is a set of nodes,
+$E \subseteq V \times V$ is a set of edges, and
+$\labels : V \to \statements$ is a mapping from nodes to statements.
+
+\section{Operational Semantics of CFGs}
+
+\jan{Here we probably need some non-deterministic choice on the
+  edges. We have to watch out that this is consistent
+  with the assertions that we insert in the blocks. We might have to
+  put assertion on the edges to avoid accounting cost for a jump that
+  might not happen.}
+
+\section{Potential Annotations for CFGs}
+
+\subsection{Linear Potential Functions.}
 
@@
 |\inter {\state(x)} {\state(y)}|$.  Using these notations we can
-rewrite the above equality as\iffull{
-$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$}{
-$\Phi(\state) = \sum_{i \in I} q_i f_i(\state)$.}%
+rewrite the above equality as 
+$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$
+%
 We often write $xy$ to denote the
-index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
-the linear-algebra sense) for potential functions.}{}  This allows us to
+index $(x,y)$.  This allows us to
 uniquely represent any linear potential function
@@
 is, a family of rational numbers where only
-a finite number of elements are not zero.  \iffull{As a notable difference from
-previous works based on amortized-analysis, we do not require all the coefficients
-of our potential function to be positive.  In practice, we will see using
-a semantic argument that despite this additional freedom, potential functions
-always remain positive.}{}
+a finite number of elements are not zero.
 
@@
 
-\paraskip
-\paragraph{Abstract Program State.}
-
-In addition to the quantitative annotations, our automatic amortized
-analysis needs to maintain a minimal abstract state to justify certain
-operations on quantitative annotations.  For example when analyzing
-the code $x \gets x + y$, it is helpful to know the sign of $y$ to
-determine which intervals will increase or decrease.  The knowledge
-needed by our rules can be inferred by local reasoning (i.e. in
-basic blocks without recursion and loops) within usual theories
-(e.g. Presburger arithmetic or bit vectors).
-
-The abstract program state is represented as \emph{logical contexts} in the
-derivation system used by our automated tool.
-Our implementation finds these logical contexts using abstract
-interpretation with the domain of linear inequalities.  We observed that
-the rules of the analysis often require only minimal local knowledge.
-This means that it is not necessary for us to compute precise loop
-invariants and only a rough fixpoint (e.g. keeping only inequalities on
-variables unchanged by the loop) is sufficient to obtain good bounds.
-
 \begin{figure*}
-  \centering
-\ifshort{\vspace{-.3cm}
-\hspace{-0.4cm}}{}
-  \begin{minipage}[b]{7cm}
-    \begin{center}
-   \begin{lstlisting}[]
-void c_down (int x,int y) {
-  if (x>y) {tick(1); c_up(x-1,y);}
-}
-void c_up (int x, int y) {
-  if (y+1<x) {tick(1); c_down(x,y+2);}
-}
-   \end{lstlisting}
-\vspace{-2.7ex}
-$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
-$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
-\\[.4\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{5cm}
-    \begin{center}
-   \begin{lstlisting}
-for (; l>=8; l-=8)
-  /* process one block */
-  tick(N);
-for (; l>0; l--)
-  /* save leftovers */
-  tick(1);
-   \end{lstlisting}
-\vspace{-2.8ex}
-$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
-$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
-\\[.4\baselineskip]
-      {\bf t61}
-    \end{center}
-  \end{minipage}
-%
-\hfill
-%
-  \begin{minipage}[b]{5.5cm}
-    \begin{center}
-\begin{lstlisting}
-for (;;) {
-  do { l++; tick(1); }
-    while (l<h && *);
-  do { h--; tick(1); }
-    while (h>l && *);
-  if (h<=l) break;
-  tick(1); /* swap elems. */ }
-\end{lstlisting}
-$2+3|[l,h]|$
-\\[.4\baselineskip]
-      {\bf t62}
-    \end{center}
-  \end{minipage}
-\vspace{1ex}
-\caption{Example \emph{t39} shows two mutually-recursive functions
-  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
-  demonstrate the unique compositionality of our system. In \emph{t61},
-  $N\geq 0$ is a fixed but arbitrary constant. % The derived constant factors
-  %are tight.
-  }
-  \label{fig:ex_list_2}
-\end{figure*}
-
-\paraskip
-\paragraph{Challenging Loops.}
-
-One might think that our set of potential functions is too simplistic
-to be able to express and prove bounds for realistic
-programs. Nevertheless, we can handle challenging example programs
-without special tricks or techniques.  Examples \emph{speed\_1} and
-\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
-demonstrate that our method can handle \emph{tricky iteration
-  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
-bounds as our analysis but requires heuristics for its counter
-instrumentation.  These loops can also be handled with inference of
-\emph{disjunctive invariants}, but in the abstract interpretation
-community, these invariants are known to be notoriously difficult to
-generate.
-%
-In Example \emph{speed\_1} we have one loop that first increments
-variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
-derive the tight bound $|[x, n]| + |[y, m]|$.
-%
-Example \emph{speed\_2} is even trickier, and we found it hard to
-find a bound manually.  However, using potential transfer reasoning as
-in amortized analysis, it is easy to prove the tight bound
-$|[x, n]| + |[z, n]|$.
-
-\paraskip
-\paragraph{Nested and Sequenced Loops.}
-
-Example \emph{t08a} in Figure~\ref{fig:ex_list} shows the ability of the analysis to discover
-interaction between \emph{sequenced loops} through size change of
-variables.  We accurately track the size change of $y$ in the first
-loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
-Furthermore, \emph{t08a} shows again that we do not handle the
-constants $1$ or $0$ in any special way.  In all examples we could
-replace $0$ and $1$ with other constants like in the second loop and
-still derive a tight bound.  \iffull{The only information, that the analyzer
-needs is $y \geq c$ before assigning $y = y - c$.}{}
-%
-Example \emph{t27} in Figure~\ref{fig:ex_list}
-shows how amortization can be used to handle
-\emph{interacting nested loops}.  In the outer loop we increment the
-variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
-increment the variable $y$ by $1000$.  Then we non-deterministically
-(expressed by \code{*}) execute an inner loop that decrements $y$ by
-$100$ until $y<100$.  The analysis discovers that only the first
-execution of the inner loop depends on the initial value of $y$.  We
-again derive tight constant factors.
-%
-
-\paraskip
-\paragraph{Mutually Recursive Functions.}
-
-As mentioned, the analysis also handles advanced control flow like
-\code{break} and \code{return} statements, and mutual recursion.
-Example \emph{t39} in \pref{fig:ex_list_2} contains two mutually-recursive functions with their
-automatically derived tick bounds.  The function \code{c\_down}
-decrements its first argument $x$ until it reaches the second argument
-$y$. It then recursively calls the function \code{c\_up},
-which is dual to \code{c\_down}.  Here, we
-count up $y$ by $2$ and call \code{c\_down}.  $\toolname$ is
-the only available system that computes a tight
-bound.  \iffull{The analysis amounts to computing the meeting
-point of two trains that approach each other with different speeds.}{}
-% This is comparable to the
-% determination of the point at which to trains meet if they approach
-% each other with different speeds.
-
-\paraskip
-\paragraph{Compositionality.}
-
-With two concrete examples from open-source projects  we demonstrate
-that the compositionality of our method is indeed crucial in practice.
-
-Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
-of block-based cryptographic primitives: Data of arbitrary length is
-consumed in blocks and the leftover is stored in a buffer for future use
-when more data is available.  It is present in all the block encryption
-routines of PGP and also used in performance critical
-code to unroll a loop. For example we found it in a bit manipulating
-function of the libtiff library and a CRC computation routine of MAD,
-an MPEG decoder.  This looping pattern is handled particularly well
-by our method. If $N \geq 8$, $\toolname$ infers the bound
-$\frac{N}{8} |[0,l]|$, but if $N<8$, it infers $7\frac{8-N}{8} +
-\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
-in the second loop (\code{tick(1)}) are random choices and $\toolname$
-would also derive tight bound for other values.
-
-To understand the resource bound for the case $N<8$, first note that the cost of
-the second loop is $|[0,l]|$.  After the first loop, we still have
-$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
-to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
-we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
-loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
-potential units instead.
-%
-This level of precision and compositionality is only achieved by
-our novel analysis, no other available tool derives the aforementioned
- tight bounds.
-
-Example \emph{t62} (\pref{fig:ex_list_2})
-is the inner loop of a quick sort implementation in cBench.  More
-precisely, it is the partitioning part of the algorithm.  This partition
-loop has linear complexity, and feeding it to our analysis gives the
-worst-case bound $2+3|[l,h]|$.  This bound is not optimal but it can be
-refined by rewriting the program. % to let our logical inference
-%discover more invariants.
-%With the potential
-%method in mind, it is clear that the three loops get their
-%potential from the same interval $[l,h]$ regardless of their nesting.
-To understand the bound, we can reason as follows.
-%
-If $h \geq l$ initially, the cost of the loop is 2.
-Otherwise, the cost of each round (at
-most 3) can be payed using the potential of $[l,h]$ by the first
-increment to $l$ because we know that $l<h$.  The two inner loops
-can also use $[l,h]$ to pay for their inner costs.
-%
-KoAT fails to
-find a bound and LOOPUS derives the quadratic bound
-$(h-l-1)^2$.  Following the classical technique, these tools try to find one
-ranking function for each loop and combine them multiplicatively
-or additively.
-
-\iffull{We only show a small selections of the programs that we can
-  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
-  extended version~\cite{anon_extended}} is a list of more than $30$
-classes of challenging programs that we can automatically analyze.
-\pref{sec:exper} contains a more detailed comparison with other
-tools\iffull{ for automatic bound derivation.}{.}
-
-\begin{figure*}
+\small
+%\vspace{-.4cm}
+\def \MathparLineskip {\lineskip=0.35cm}
 \begin{mathpar}
-%
 \Rule{Q:Skip}
 { }
-{ B, R \vdash \htriple {\Gamma; Q}{\code{skip}}{\Gamma; Q} }
-%
-\and
-\Rule{Q:Break}
-{ }
-{ (\Gamma; Q_B), R \vdash \htriple{\Gamma; Q_B}{\code{break}}{\Gamma'; Q'} }
-%
-\and
-\Rule{Q:Tick}
-{ n < 0 \implies Q \ge 0 }
-{ B, R \vdash \htriple{\Gamma; Q}{\code{tick}(n)}{\Gamma; Q{-}n} }
-%
-\and
-\Rule{Q:Return}
-{ P = Q_R[\Vret/x]
-\\ \Gamma = \Gamma_R[\Vret/x]
-\\ \forall i \in \dom{P} . \, p_i = q_i
-}
-{ B, (\Gamma_R; Q_R) \vdash \htriple{\Gamma; Q}{\code{return}~x}{\Gamma'; Q'} }
-%
-\and \Rule{Q:Update}
-{ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
-}
-{ B, R \vdash \htriple{\Gamma[x/y]; Q}{x \gets y}{\Gamma; Q'} }
-%
-\and \Rule{Q:Loop}
-{ Q \ge Q'
-\\ (\Gamma'; Q'), R \vdash \htriple{\Gamma; Q}{ S }{\Gamma;  Q}
-}
-{ B, R \vdash \htriple{\Gamma; Q}{\code{loop}~S}{\Gamma'; Q'} }
-%
-\and \Rule{Q:Inc}
-{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{ux}, -q_{xu})
-\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{xu}, -q_{ux})
-}
-{ \textstyle
-  B, R \vdash \htriple{\Gamma[x/x {+} y]; Q}{x \gets x + y}{\Gamma; Q'}
-}
-%
-\and \Rule{Q:Dec}
-{  \textstyle q'_{0y} = q_{0y} - \sum_u \max(q_{xu}, -q_{ux})
-\\\\ \textstyle q'_{y0} = q_{y0} - \sum_u \max(q_{ux}, -q_{xu})
-}
-{ \textstyle
-  B, R \vdash \htriple{\Gamma[x/x {-} y]; Q}{x \gets x - y}{\Gamma; Q'}
-}
-%
-\and \Rule{Q:If}
-{ B, R \vdash \htriple{\Gamma \land e; Q}{S_1}{\Gamma'; Q'}
-\\\\ B, R \vdash \htriple{\Gamma \land \neg e; Q}{S_2}{\Gamma'; Q'}
-}
-{ B, R \vdash \htriple{\Gamma; Q}{\code{if}(e)~S_1~\code{else}~S_2}{\Gamma'; Q'} }
-%
-\and\Rule{Q:Seq}
-{  B, R \vdash \htriple{\Gamma; Q}{S_1}{\Gamma'; Q'}
-\\\\ B, R \vdash \htriple{\Gamma'; Q'}{S_2}{\Gamma''; Q''}
-}
-{ B, R \vdash \htriple{\Gamma; Q}{S_1;S_2}{\Gamma''; Q''} }
-%
-\and\Rule{Q:Call}
-{ (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
-\\ \text{Loc} = \text{Locals}(Q) \!
-\\ \forall i \neq j . \, x_i \neq x_j \!
-\\ c \in \Qplusz \!
-\\ Q = P + S \!
-\\ Q' = P' + S
-\\ U = Q_f[\Vargs / \vec x]
-\\ U' = Q'_f[\Vret/r]
-\\ \forall i \in \dom{U} . \, p_i = u_i
-\\ \forall i \in \dom{U'} .\, p'_i = u'_i
-\\ \forall i \not\in \dom{U'} .\, p'_i = 0
-\\ \forall i \not\in \text{Loc}
-  .\, s_i = 0
-}
-{ B, R \vdash \htriple {\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}; Q {+} c}
-  {r \gets f(\vec x)}
-  {\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}; Q' {+} c}
-}
+{ \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } }
 %
@@
 { }
-{ B, R \vdash \htriple{\Gamma; Q}{\code{assert}~e}{\Gamma \land e; Q} }
-%
-\and \Rule{Q:Extend}
-{ \fenv f = (\vec y, S_f)
-\\ Q_f \ge 0
-\\ Q'_f \ge 0
-\\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{\Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y]}{S_f}{\Gamma'; Q'}
-}
-{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
+{ \htriple{ \Gamma; Q {+} M_a }{ \code{assert}~e }{ \Gamma {\land} e; Q } }
+%
+\and \Rule{Q:Tick}
+{ }
+{ \htriple{ \Gamma; Q {+} M_t(n) }{ \code{tick}(n) }{ \Gamma; Q } }
+%
+\and \Rule{Q:Seq}
+{  \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
+\\ \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
+}
+{ \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
+%
+\\ \Rule{Q:IncP}
+{ \Gamma \models y \ge 0 \!
+\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
+\\\\ \textstyle q'_{0y} = q_{0y}
+      + \sum_{u \in \mathcal U} q_{xu}
+      - \sum_{v \not\in \mathcal U} q_{vx} \!
+}
+{ \textstyle
+  \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
+}
+%
+\and \Rule{Q:Inc}
+{ M = M_u + M_e(x {\pm} y)
+\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
+\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
+}
+{ \textstyle
+  \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
+}
+%
+\and \Rule{Q:Set}
+{ q'_{xy}, q'_{yx} \in \Qplusz
+\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
+}
+{ \htriple{ \Gamma[x/y]; Q {+} M_u {+} M_e(y) }{ x \gets y }{ \Gamma; Q' } }
+%
+\and \Rule{Q:DecP}
+{ \Gamma \models y \ge 0
+\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
+\\\\ \textstyle q'_{y0} = q_{y0}
+      + \sum_{u \in \mathcal U} q_{ux}
+      - \sum_{v \not\in \mathcal U} q_{xv}
+}
+{ \textstyle
+  \htriple{ \Gamma[x/x {-} y]; Q {+} M_u {+} M_e(x {-} y) }{ x \gets x - y }{ \Gamma; Q' }
+}
+% %
+% \and\Rule{Q:Call}
+% { (\Gamma_f; Q_f,  \Gamma_f'; Q_f') \in \Delta(f) \!
+% \\ \text{Loc} = \text{Locals}(Q) \!
+% \\ \forall i \neq j . \, x_i \neq x_j \!
+% \\ c \in \Qplusz \!
+% \\ Q = P + S \!
+% \\ Q' = P' + S
+% \\ U = Q_f[\Vargs / \vec x]
+% \\ U' = Q'_f[\Vret/r]
+% \\ \forall i \in \dom{U} . \, p_i = u_i
+% \\ \forall i \in \dom{U'} .\, p'_i = u'_i
+% \\ \forall i \not\in \dom{U'} .\, p'_i = 0
+% \\ \forall i \not\in \text{Loc}
+%   .\, s_i = 0
+% }
+% { B, R \vdash \htriple
+%   { \Gamma_f[\Vargs / \vec x] {\land} \Gamma_{\text{Loc}}; Q {+} c {+} M_f }
+%   { r \gets f(\vec x) }
+%   { \Gamma_f'[\Vret/r] {\land} \Gamma_{\text{Loc}}; Q' {+} c {-} M_r }
+% }
+%
+% \and \Rule{Q:Extend}
+% { \fenv f = (\vec y, S_f)
+% \\\\ B, (\Gamma_f'; Q'_f) \vdash \htriple{ \Gamma_f[\Vargs/\vec y]; Q_f[\Vargs/ \vec y] }{ S_f }{ \Gamma'; Q' }
+% }
+% { (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }
 %
 \and \Rule{Q:Weak}
-{   \Gamma_1 \models \Gamma_2
-\\ Q_1 \succeq_{\Gamma_1} Q_2
-\\ B, R \vdash \htriple{\Gamma_2;Q_2}{S}{\Gamma'_2; Q'_2}
+{ \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
+\\ \!\! \Gamma_1 \models \Gamma_2
+\\\\ Q_1 \succeq_{\Gamma_1} Q_2
 \\ \Gamma'_2 \models \Gamma'_1
@@
 }
-{ B, R \vdash \htriple{\Gamma_1; Q_1}{S}{\Gamma'_1; Q'_1} }
-%
-\and \Rule{Relax}
+{ \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
+%
+\\ \Rule{Relax}
 { \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
 \\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
-\\ \forall i.\, p_i, r_i \in \Qplusz
 \\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
@@
 { Q' \succeq_\Gamma Q }
+%\vspace{-.3cm}
 \end{mathpar}
-\caption{Inference rules of the quantitative analysis.}
+\caption{Inference rules of the straight-line statements.}
 \label{fig:auto}
@@
 
-\sectskip
-\section{Derivation System}
-
-In the following we describe the local and compositional derivation
-rules of the automatic amortized analysis.
-
-\paraskip
-\paragraph{Cost Aware Clight.}
-
-We present the rules for a subset of Clight.
-Clight is the first intermediate
-language of the CompCert compiler~\cite{CompcertCACM}.  It
-is a subset of C with a unified looping construct and
-side-effect free expressions.  We reuse most of CompCert's
-syntax but instrument the semantics with a resource counter
-(an arbitrary rational number) that can be modified by
-\code{tick()} function calls.
-More details on this cost-aware semantics are provided in \pref{sec:soundness}.
+\subsection{Analyzing Straight Line Statements}
 
@@
 
-\paraskip
 \paragraph{Judgements.}
@@
 $$
-  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
     \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
@@
 %
-The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
+The judgement
 can be seen as a quantitative Hoare triple.  All assertions are split into
@@
 
-The judgement is a bit more involved since we have to take into
-account the early exit statements \code{break} and \code{return}.
-This is similar to classical Hoare triples in the presence of non-linear
-control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
-that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
-% of looping statements by the rule {\sc Q:Loop} and used before
-% \code{break} statements by the rule {\sc Q:Break}.
-Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
-returning from a function call with \code{return}. \iffull{(This becomes clear
-in the rules.)}{}
-
-% These judgements correspond to the logic judgements $\Delta; B; R
-% \vdash \htriple P S Q$ where each assertion splits in two parts, a
-% logical part $\Gamma$ and a quantitative part $Q$.  That means that
-% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
-% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
-% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
-% simple Hoare triple.  We leave out the function context $\Delta$ in
-% the presentation to avoid cluttering notations.  Instead we assume a
-% fixed function context $\Delta$ that is implicit on all judgements.
-% Function contexts are treated exactly as in the logic.  % We call
-% % precondition the pair $(\Gamma; Q)$ and postcondition the pair
-% % $(\Gamma'; Q')$.
-% The correspondence between judgements of the
-% automation and the ones of the quantitative logic is made formal by
-% our soundness proof of the automatic amortized analysis at the end of
-% this section.
-
 As a convention, if $Q$ and $Q'$ are quantitative annotations
@@
 such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
-%
-% THE FOLLOWING IS NO LONGER TRUE
-% In all the rules, we have the implicit side condition that all rational
-% coefficients are non-negative.
+
+In all the rules, we have the implicit side condition that all rational
+coefficients are non-negative.
 %
@@
 
-% We describe the automatic amortized analysis for a subset of
-% expressions of Clight.  Clight is the first intermediate
-% language of the CompCert compiler~\cite{CompcertCACM}, it
-% is a cleaner C with a unified looping construct and
-% side-effect free expressions.
-% Assignments are restricted to the form $x \gets
-% y$ or $x \gets x \pm y$.  In the implementation, a Clight
-% program is converted into this form prior to analysis without changing
-% the resource cost.  This is achieved by using a series of \emph{cost-free
-%   assignments} that do not result in additional cost in the
-% semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
-% a[y]$ are handled by assigning $0$ to coefficients
-% like $q_{xa}$ and $q_{ax}$ that depend on $x$ after the assignment.
-% This sound treatment ensures that no further loop bounds
-% depend on the result of the non-linear operation.
-
-\paraskip
-\paragraph{Function Specifications.}
-
-During the analysis, function specifications are quadruples
-$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
-$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
-are instantiated by appropriate variables on call sites.  A distinctive
-feature of our analysis is that it respects the function abstraction:
-when deriving a function specification it generates a set of
-constraints and the above quadruple; once done, the constraint
-set can readily be reused for every call site and the function need
-not be analyzed multiple times.
-%
-Therefore, the derivation rules are parametric in a function context $\Delta$
-that we leave implicit in the rules presented here.
-% More details can be found in the extended version.
-
-\paraskip
-\paragraph{Derivation Rules.}
-
-The rules of our derivation system must serve two purposes.  They must
-\emph{attach} potential to certain program variable intervals and use
-this potential, when it is allowed, to \emph{pay} for resource consuming
-operations.  These two purposes are illustrated on the {\sc Q:Skip}
-rule.  This rule reuses its precondition as postcondition, it is
-explained by two observations: First, no resource is consumed by the skip
-operation, thus no potential has to be used to pay for the evaluation.
-Second, the program state is not changed by
-the execution of a skip statement.  Thus all potential available
-before the execution of the skip statement is still available after.
-\iffull{
-These two observations would also justify a rule with the side-condition
-$Q \ge Q'$, because some potential is always safely lost.  For the sake
-of orthogonality, it is not how we present the rule and this form of
-weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
-described later.}{}
-
-The rules {\sc Q:Inc} and {\sc Q:Dec} describe how the potential is
-distributed after a size change of a variable.
-The rule {\sc Q:Inc} is for incements $x \gets x+y$ and the symmetric
-rule {\sc Q:Dec} is for decrements $x \gets x-y$.
-The two rules are equivalent in the case where $y=0$.
-
-The two previous rules are essentially the same, so we only focus on
-the explanation of {\sc Q:Inc}. In this rule, the program updates a
-variable $x$ with $x+y$.  Since $x$ is changed, the
-quantitative annotation must be updated to reflect the change of the
-program state.  We write $x'$ for the value of $x$ after the assignment.
-Since $x$ is the only variable changed, only intervals of the form
-$\inter u x$ and $\inter x u$ will be resized.  Consider $S(x) = q_{xu}
-|\inter x u| + q_{ux} |\inter u x|$, during the increment, $S$ will
-increase by at most $y \cdot \max(q_{ux}, -q_{xu})$.
-This means that we must decrease the potential $q_{0y}$ of $|\inter 0 y|$
-by $\max(q_{ux}, -q_{xu})$ to pay for the change on
-$|\inter x u|$ and $|\inter u x|$.  A symmetric argument explains the
-change on $q_{y0}$.
-
-The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
-loop body, one needs to find an invariant potential $Q$ that will pay for
-the iterations.  At each iteration, some resources are spent by the loop
-body.  Since the  loop can only be exited with a \code{break} statement,
-the postcondition $\{ \Gamma'; Q' \}$ for the statement
-$\code{loop}~S$ is used as break postcondition in the derivation for $S$.
-
-Another interesting rule is {\sc Q:Call}.  It needs to account for the
-changes to the stack caused by the function call, the arguments/return
-value passing, and the preservation of local variables.  We can sum up
-the main ideas of the rule as follows.
-\itemskip
-\begin{itemize}
-\item The potential in the pre- and postcondition of the function
-  specification is equalized to its matching potential in the callee's
-  pre- and postcondition.
-\itemskipIn
-\item
-  The potential of intervals $|\inter x y|$ is preserved
-  across a function call
-  if $x$ and $y$ are local\iffull{ variables.}{.}
-\item
-  The unknown potentials after the call (e.g.
-  $|\inter x g|$, with $x$ local and $g$ global)
-  are set to zero in the postcondition.
-\end{itemize}
-\itemskip
-If $x$ and $y$ are local variables and $f(x,y)$ is
-called, {\sc Q:Call} splits the potential of
-$|\inter x y|$ in two parts.  One part to perform the
-computation in the function $f$ and one part to keep for
-later use after the function call.  This splitting is
-realized by the equations $Q = P {+} S$ and $Q' =
-P' {+} S'$.  Arguments in the function precondition
-$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
-of names different from all program variables.  This
-prevents name conflicts and ensures that
-the substitution $[\Vargs/\vec x]$ is meaningful.
-Symmetrically, we use the unique name $\Vret$ to represent
-the return value in the function's postcondition
-$(\Gamma'_f; Q'_f)$.
-
-The rule {\sc Q:Weak} is not syntax directed. In the implementation
-we apply {\sc Q:Weak} before loops and between the two statements of a
-sequential composition. We could integrate weakening into every
-syntax directed rule but this simple heuristic helps to make the analysis
-efficient.  The high-level idea of {\sc Q:Weak} is the following: If
-we have a sound judgement, then it is sound to add more potential to the
-precondition and remove potential from the postcondition.  The concept of
-\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
-that is defined in the rule {\sc Relax}. This rule
-also deals with the important task of transferring constant potential
-(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
-from the logical context that the interval size $|\inter x y| \geq \ell$ is
-larger than a constant $\ell$ then we can turn the potential $q_{xy} {\cdot}
-|\inter x y|$ form the interval into the constant potential $\ell{\cdot}q_{xy}$
-and guarantee that we do not gain potential.  Conversely, if $|\inter
-x y| \leq u$ for a constant $u$ then we can transfer constant potential
-$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
-without gaining potential.
-
-\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x \geq 10)~(x=x - 10; tick(5))~else~break}}}
-\begin{figure*}[t]
-\vspace{-.5cm}
-\centering
-\footnotesize
-\begin{mathpar}
-\inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
-{ \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
-  { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1cm]
-    { \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
-      { \inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
-        { \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
-          {
-          }
-          { (x {<} 10; B^\text{de}) \vdash
-	    \htriple{ x {\geq} 10; Q^\text{de} }{ \code{x = x - 10} }{ \cdot; P^\text{de} }
-          }
-        }
-        { (x {<} 10; B^\text{we}) \vdash
-	  \htriple{ x {\geq} 10; Q^\text{we} }{ \code{x = x - 10} }{ \cdot; P^\text{we} }
-        }
-        \and
-        \inferrule*[right={\scriptsize (Q:Tick)},leftskip=0cm,rightskip=0cm]
-        {
-        }
-        { (x {<} 10; B^\text{ti}) \vdash
-	  \htriple{ \cdot; Q^\text{ti} }{ \code{tick(5)} }{ \cdot; P^\text{ti} }
-        }
-      }
-      { (x {<} 10; B^\text{sq}) \vdash
-	\htriple{ x {\geq} 10; Q^\text{sq} }{ \code{x=x - 10; tick(5)} }{ \cdot; P^\text{sq} }
-      }
-    }
-    { (x {<} 10; B^\text{if}) \vdash
-      \htriple{ x {\geq} 10; Q^\text{if} }{ \code{x=x - 10; tick(5)} }{ \cdot; P^\text{if} }
-    }
-    \and
-    \inferrule*[right={\scriptsize (Q:Weak)},leftskip=4.5cm,rightskip=4cm]
-    { \inferrule*[right={\scriptsize (Q:Break)},leftskip=0cm,rightskip=0cm]
-      { }
-      { (x {<} 10; B^\text{br}) \vdash
-	\htriple{ x {<} 10; Q^\text{br} }{ \code{break} }{ \bot; P^\text{br} }
-      }
-    }
-    { (x {<} 10; B^\text{el}) \vdash
-      \htriple{ x {<} 10; Q^\text{el} }{ \code{break} }{ \cdot; P^\text{el} }
-    }
-  }
-  { \hspace{3cm} (x {<} 10; B^\text{lo}) \vdash
-    \htriple{ \cdot; Q^\text{lo} }{ \loopbody }{ \cdot; P^\text{lo} } \hspace{3cm}
-  }
-}
-{ (\cdot; B) \vdash
-  \htriple{ \cdot; Q) }{ \code{loop}~{\loopbody} }{ x {<} 10; P }
-}
-\vspace{-.3cm}
-\end{mathpar}
-\flushleft
-\begin{tabular}{l}Constraints:\end{tabular}
-\vspace{-.2cm}
-$$
-\begin{array}{lll}
-P {=} B^\text{lo} \land Q {=} Q^\text{lo} {=} P^\text{lo}
-& B^\text{el} {=} B^\text{if} {=} B^\text{lo}
-    \land Q^\text{el} {=} Q^\text{if} {=} Q^\text{lo}
-    \land P^\text{el} {=} P^\text{if} {=} P^\text{lo}
-& B^\text{el} {=} B^\text{br} \land Q^\text{el} \succeq_{(x{<}10)} Q^\text{br}
-    \land P^\text{br} \succeq_{(\cdot)} P^\text{el}
-\\
-B^\text{br} {=} Q^\text{br}
-& B^\text{if} {=} B^\text{sq} \land Q^\text{if} \succeq_{(x{<}10)} Q^\text{sq}
-    \land P^\text{sq} \succeq_{(\cdot)} P^\text{if}
-& B^\text{sq} {=} B^\text{we} {=} B^\text{ti} \land Q^\text{sq} {=} Q^\text{we}
-    \land P^\text{we} {=} Q^\text{ti} \land P^\text{ti} {=} P^\text{sq}
-\\
-Q^\text{ti} {=} P^\text{ti} + 5
-& B^\text{we} {=} B^\text{de} \land Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}
-    \land P^\text{de} \succeq_{(\cdot)} P^\text{we}
-& p^\text{de}_{0,10} {=} q^\text{de}_{0,10} + q^\text{de}_{0,x}
-    \land p^\text{de}_{0} {=} q^\text{de}_{0}
-    \land \forall (\alpha, \beta) \neq (0,10)  . \,
-              p^\text{de}_{\alpha,\beta} {=} q^\text{de}_{\alpha,\beta}
-\end{array}
-$$
-\begin{tabular}{l@{\hspace{5em}}l}
-  Linear Objective Function: $
-    1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
-  $
-& Constant Objective Function: $
-    1{\cdot}q_0 + 11{\cdot}q_{0,10}
-  $
-\end{tabular}
-\vspace{.1cm}
-\caption{An example derivation as produced $\toolname$. The constraints are resolved by an off-the-shelf LP solver.}
-\label{fig:derivation}
-\end{figure*}
-
-\sectskip
-\section{Automatic Inference via LP Solving}
-\label{sec:lp}
-\aftersectskip
-
-We separate the search of a derivation in two steps.  As a first
-step we go through the functions of the program and apply inductively
-the derivation rules of the automatic amortized analysis.  This is
-done in a bottom-up way for each strongly connected component (SCC) of the
-call graph.  During this
-process our tool uses symbolic names for the rational coefficients
-$q_i$ in the rules.  Each time a linear constraint must be
-satisfied by these coefficients, it is recorded in a global list for the SCC using
-the symbolic names.  We reuse the constraint list for
-every call from outside the SCC.
-
-We then feed the collected constraints to an
-off-the-shelf LP solver (currently CLP~\cite{clp}).
-If the solver successfully finds a solution, we know that a derivation
-exists and extract the values for the initial
-$Q$ from the solver to get a resource bound for the program.  To get
-a full derivation, we extract the complete solution from the
-solver and apply it to the symbolic names $q_i$ of the coefficients
-in the derivation.  If the LP solver fails to find a solution, an
-error is reported.
-
-\pref{fig:derivation} contains an example derivation as produced
-by $\toolname$.  The upper case letters (with optional
-superscript) such as $Q^\text{de}$ are families of
-variables that are later part of the constraint system that is passed
-to the LP solver.  For example $Q^\text{de}$ stands for the potential
-function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
-q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
-q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
-variables such as $q^\text{de}_{x,10}$ are yet unknown and later
-instantiated by the LP solver.
-
-In general, the weakening rule can be applied after every syntax directed
-rule.  However, it can be left out in practice at some places to
-increase the efficiency of the tool.
-The weakening operation $\succeq_{\Gamma}$ is defined by the rule
-\textsc{Relax}. It is parameterized by a logical context that is
-used to gather information on interval sizes.  For example,
-{%\small
-  \begin{align*}
-  P^\text{de} \succeq_{(\cdot)} P^\text{we} &\equiv
- p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10} \\
-&\land  p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10} \\
-&\land  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;.
-  \end{align*}
-}
-The other rules are syntax directed and applied inductively.  For example,
-the outermost expression is a loop, so we use the rule \text{Q:Loop}
-at the root of the derivation tree.  At this point, we do not know yet
-whether a loop invariant exists.  But we produce the constraints
-$Q^\text{lo} = P^\text{lo}$\iffull{ which is short for the following
-constraint set.
-$$
-\begin{array}{lll}
-q^\text{lo}_0 = p^\text{lo}_0 &
-q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
-q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
-q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
-q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
-q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
-\end{array}
-$$}{.}
-%
-These constraints express the fact that the potential functions before and
-after the loop body are equal and thus constitute an invariant.
-
-After the constraint generation, the LP solver is provided with an
-objective function to be minimized.  We wish to
-minimize the initial potential, which is a resource bound on the
-whole program.  Here it is given by $Q$.  Moreover, we would like to
-express that minimization of linear potential such as
-$q_{10,x} |[10,x]|$
-takes priority over minimization of constant potential such as
-$q_{0,10} |[0,10]|$.
-
-To get a tight bound, we use modern LP solvers that allow constraint
-solving and minimization at the same time:  First we consider our initial
-constraint set as given in \pref{fig:derivation} and ask the
-solver to find a solution that satisfies the constraints and minimizes
-the linear expression
-\iffull{
-$$
-1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x} \; .
-$$}
-{$
-1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
-$.}
-%
-The penalties given to certain factors are used
-to prioritize certain intervals.  For example, a bound with
-$[10,x]$ will be preferred to another with $[0,x]$ because $|[10,x]|
-\le |[0,x]|$.
-%
-The LP solver now returns a solution of the constraint set and an
-objective value\iffull{, that is, a mapping from variables to floating-point
-numbers and the value of the objective function with this
-instantiation}{}.  The solver also memorizes the optimization path that
-led to the optimal solution.  In this case, the objective value would
-be 5000
-since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
-We now add the constraint
-\iffull{
-$$
-1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
- \le 5000
-$$}{
-$
-1{\cdot}q_{x,0} + 10000{\cdot}q_{0,x} + 11{\cdot}q_{x,10} + 9990{\cdot}q_{10,x}
- \le 5000
-$
-}
-to our constraint set and ask the solver to optimize the objective
-function
-\iffull{
-$$
-q_0 + 11{\cdot}q_{0,10} \; .
-$$}
-{$
-q_0 + 11{\cdot}q_{0,10}$.}
-This happens in almost no time in practice.
-The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise. Thus
-the derived bound is $0.5 |[0,x]|$.
-
-A notable advantage of the LP-based approach compared to
-SMT-solver--based techniques is that a satisfying assignment is
-a proof certificate instead of a counter example.  To provide
-high-assurance bounds, this certificate can be checked
-in linear time by a simple validator.
-
-% SSS SPACING
-%\vspace{-.24cm}
-
-\iffull{The constraints that we generate have
-a particularly simple form that is known as \emph{network problem}.
-Such problems can be solved in linear time in practice.}{}
-
-\sectskip
-\section{Logical State and User Interaction}
-\label{sec:anno}
-\aftersectskip
-
-\begin{figure}[t]
-\begin{minipage}[b]{0.8\linewidth}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
-while (k > 0) {
-  x=0;
-  while (x < N && a[x] == 1) {
-    @assert(na > 0);@
-    a[x]=0; @na--;@
-    tick(1); x++; }
-  if (x < N) { a[x]=1; @na++;@ tick(1); }
-  k--;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\end{minipage}
-\caption{Assisted bound derivation using logical state.  We write
-  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$.
-  The derived bound is
-  $2|[0,k]| + |[0,na]|$.
-  }
-\label{fig:xmplincaux}
-\end{figure}
-
-\begin{figure}
-\vspace{-.3cm}
-\begin{lstlisting}[numbers=left]
-${\color{blue}%
-  \mbox{logical state invariant }
-  \{ lg > \log_2(h - l) \} }$
-bsearch(x,l,h@,lg@) {
-  if (h-l > 1) {
-    @assert(lg > 0);@
-    m = l + (h-l)/2;
-    @lg--;@ if (a[m]>x) h=m; else l=m;
-    tick($M_\code{bsearch}$);
-    l = bsearch(x,l,h@,lg@);
-    tick($-M_\code{bsearch}$);
-  } return l;
-}
-\end{lstlisting}
-\vspace{-.2cm}
-\caption{Assisted bound derivation using logical state.
-  % We model the stack consumption of the program using tick statements.
-  % Before a function call to \code{bsearch}, a tick statement pays for
-  % the space used by stack frame of the function.  The space is recoverd
-  % after the call.
-  We write $\log_2(x)$ for the integer part of
-  logarithm of $x$ in base 2.
-  The semi-automatically derived bound is $|[0,lg]|$.
-  }
-\label{fig:xmplbsaux}
-\end{figure}
-
-\ifdefined\fullversion
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}
-  \mbox{logical state invariant }
-  \{ nm = A(0) \} }$
-i=1; m=a[0];
-while (i < N) {
-  if (a[i] > m) {
-    @assert(nm > 0); nm--;@
-    m = a[i], tick(1);
-  } i++, tick(1);
-}
-\end{lstlisting}
-\caption{Example of assisted bound derivation using logical state.
-  We write $A(i)$ for
-  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
-  The bound derived is $|[0,N]| + |[0,nm]|$.
-  }
-% \label{fig:xmplmax2}
-\end{figure}
-\fi
-
-While complete automation is desirable, it is not always possible
-since the problem of bound derivation is undecidable.  In this section
-we present a new technique to derive complex resource bounds
-semi-automatically by leveraging our automation. Our goal is to
-develop an interface between bound derivation and
-established qualitative verification techniques.
-
-When the resource bound of a program depends on the contents of the
-heap, or is non-linear (e.g. logarithmic, exponential), we
-introduce a \emph{logical state} using \emph{auxiliary variables}.
-Auxiliary variables guide $\toolname$ during bound derivation
-but they do not change the behavior of the program.
-
-More precisely, the technique consists of the following steps. First,
-a program $P$ that fails to be analyzed automatically is enriched by
-auxiliary variables $\vec x$ to form a program
-$P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
-logical variables is selected to satisfy the proposition:
-\begin{equation} \tag{*}
-\forall n\, \sigma\, \sigma'.\,
-  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
-    {\implies}
-  \exists n' {\le} n.\,
-    (\sigma, P) \Downarrow_{n'} \sigma'.
-\end{equation}
-Since the annotated program and the original one are usually
-syntactically close, the proof of this result goes
-by simple induction on the resource-aware evaluation judgement.
-Third, using existing automation tools, a bound $B(\vec x)$ for
-$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
-$\vec X$, gives the final resource bound for the program $P$.
-
-This idea is illustrated by the program in \pref{fig:xmplincaux}.
-The parts of the code in blue are annotations that were added
-to the original program text.  The top-level loop increments
-a binary counter $k$ times.  A naive analysis of
-the algorithm yields the quadratic bound $k \cdot N$.
-However, the algorithm is in fact linear and its cost is bounded
-by $2k + \#_1(a)$ where $\#_1(a)$ denotes
-the number of one entries in the array $a$.  Since this
-number depends on the heap contents, no tool available
-for C is able to derive the linear bound.  However, it can be inferred
-by our automated tool if a logical variable $na$ is introduced.
-This logical variable is a reification of the
-number $\#_1(a)$ in the program.  For example, on line 6 of the example
-we are setting \lstinline{a[x]} to 0 and because of the condition we
-know that this array entry was 1.  To reflect this change on
-$\#_1(a)$, the logical variable $na$ is decremented.
-Similarly, on line 8, an array entry which was 0 becomes 1, so
-$na$ is incremented.  To complete the step 2 of the systematic
-procedure described above, we must show that the extra assertion
-\lstinline{na > 0} on line 5 is always true.  It
-is done by proving inductively that $na = \#_1(a)$
-and remarking that since \lstinline{a[x] == 1} is true, we must have
-$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails.
-
-Another simple example is given in \pref{fig:xmplbsaux} where
-a logarithmic bound on the stack consumption of a binary search
-program is proved using logical variable annotations.  Once again,
-annotations are in blue in the program text.  In this example,
-to ease the proof of equivalence between the annotated program
-and the original one, we use the inequality $lg >
-\log_2(h-l)$ as an invariant.  This technique allows a simpler
-proof because, when working with integer arithmetic, it is not always
-the case that $\log_2(x-x/2) = \log_2(x)-1$.
-
-Generally, we observed that because the instrumented program
-is structurally same as the original one, it is enough to prove
-that the added assertions never fail in order to show the two programs
-satisfy the proposition (*).  This can usually be piggybacked
-on standard static-analysis tools.
-
-% On the other hand, it might be the case that even with annotations, a
-% resource bound cannot be found automatically.  With existing tools, this
-% situation requires a full rewrite of the program.  However, in our
-% framework, it is possible to write a manual proof and combine it with
-% bounds derived by the automatic analysis.  This interaction is made
-% formal using our Quantitative Hoare Logic.  This resource aware program
-% logic is an extension of the one presented in~\cite{veristack14} and
-% was originally designed to prove the soundness of our automatic
-% analysis.  To make it suitable for this end, we made it independent
-% of the resource used (e.g. stack usage, heap usage, clock ticks) and
-% implemented a smoother handling of the auxiliary state through a shallow
-% embedding in Coq.  Despite being first a theoretical tool, the logic can
-% be used directly as in~\cite{veristack14}.  Its judgements are inspired
-% by typical Hoare logics and have the shape
-% $$
-% \Delta \vdash \htriple P S Q
-% $$
-% where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
-% are quantitative assertions, i.e. functions from the program states
-% to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
-% of the logic are described in further details in the extended version of
-% this paper~\cite{anon_extended}.  Because our automated tool produces
-% proofs of the resource bounds in the Quantitative Hoare Logic, it can
-% be used as an assistant when deriving a complex bound by hand in the
-% logic.  The interaction is guaranteed to be sound by the soundness
-% theorem of our Quantitative Hoare Logic.
-
-\sectskip
-\section{Soundness Proof}
-\label{sec:soundness}
-\aftersectskip
-
-The soundness of the analysis builds on a new cost semantics for Clight
-and an extended quantitative logic.  Using these two tools, the soundness
-of the automatic analysis described in \pref{sec:AAA}
-is proved by a translation morphism to the logic.
-
-\iffull{}{The main parts of the soundness proof are formalized with
-  Coq and available for download. The full definitions of the cost
-  semantics and the quantitative Hoare logic, and more details on the
-  soundness proof can be found in the extended version of this
-  article.}
-
-\paraskip
-\paragraph{Cost Semantics for Clight.}
-To base the soundness proof on a formal ground, we start by defining a
-new cost-aware operational semantics for Clight.  Clight's
-operational semantics is based on small-step transitions and
-continuations.  Expressions---which do not have side effects---are
-evaluated in a big-step fashion.
-
-A program state $\state = (\theta, \gamma)$ is composed of two maps
-from variable names to integers. The first map, $\theta : \text{Locals}
-\to \Z$, assigns integers to local variables of a function, and the
-second map, $\gamma : \text{Globals} \to \Z$, gives values to global
-variables of the program.  In this article, we assume that all values are
-integers but in the implementation we support all data types of Clight.
-The evaluation function \evalE{\cdot}{} maps an expression $e \in E$
-to a value $\evalE{e}{\state} \in \Z$ in the program state $\state$.
-% For simplicity, we assume that local variables and global variables
-% are disjoint.
-We write $\state(x)$ to obtain the value of $x$ in program state
-$\state$.
-% \iffull{Such a lookup is defined as
-% $$
-% (\theta, \gamma)(x) =
-% \left\{
-% \begin{array}{lr}
-% \theta(x) & \mbox{ if } x \in \dom \theta \\
-% \gamma(x) & \mbox{ if } x \in \dom \gamma
-% \end{array}
-% \right..
-% $$}{}
-Similarly, we write $\state[x \mapsto v]$ for the state based on $\state$
-where the value of $x$ is updated to $v$.
-
-The small-step semantics is standard, except that it tracks
-the resource consumption of a program.  The semantics is parametric in
-the resource of interest for the user of our system.  We achieve this
-independence by using a tick instruction that allows the programmer
-to specify how much and when resource is consumed.  If the resource
-usage is bound to the programming language (e.g. stack usage, number
-loop iterations and function calls), the tick statements can
-be inserted automatically by a pre-processing phase.  Resources can be
-released by using a negative tick.
-Two sample rules of the semantics for update and tick follow.
-
-\begin{mathpar}
-\Rule{U}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\rightarrow\\\\
-(\state', \code{skip}, \cont, \cost)
-}
-\and
-\Rule[leftskip=0.3cm,rightskip=0.2cm]{T}
-{ }
-{ (\state, \code{tick}(n), \cont, \cost)
-\rightarrow\\\\
-(\state, \code{skip}, \cont, \cost {-} n)
-}
-\end{mathpar}
-The rules have as implicit side condition that $\cost$ is
-non-negative.  This makes it possible to detect a resource crash as a
-stuck configuration where $\cost<0$.
-
-% \begin{figure*}
-% \small
-% \begin{mathpar}
-% %
-% \Rule{L:Update}
-% { }
-% { \Delta; B; R \vdash
-% \htriple
-%   { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-% }
-% %
-% \and
-% \Rule{L:Assert}
-% { }
-% { \Delta; B; R \vdash
-% \htriple
-%   { \code{istrue}~\sem{e}_\state \implies Q + M_a }
-%   { \code{assert}~e }
-%   { Q }
-% }
-% %
-% \and
-% \Rule{L:Tick}
-% { }
-% { \Delta; B; R \vdash
-% \htriple
-%   { Q + M_t(n) }
-%   { \code{tick}(n) }
-%   { Q }
-% }
-% %
-% \and
-% \Rule{L:Loop}
-% { \Delta; Q; R \vdash
-% \htriple
-%   { I }{ S }{ I + M_l }
-% }
-% { \Delta; B; R \vdash
-% \htriple
-%   { I }{ \code{loop}~S }{ Q }
-% }
-% \and \Rule{L:Seq}
-% { \Delta; B; R \vdash
-% \htriple{ P }{ S_1 }{ Q' + M_s }
-% \\
-% \!\!  \Delta; B; R \vdash
-% \htriple{ Q' }{ S_2 }{ Q }
-% }
-% { \Delta; B; R \vdash
-% \htriple{ P }{ S_1;S_2 }{ Q }
-% }
-% %
-% \and
-% \Rule{L:Weaken}
-% {
-%   \!\! B' \models B \!\! \\
-%   \!\! \forall v.\, (R'\,v \models R\,v) \!\! \\\\
-%   \!\! P \models P' \!\! \\
-%   \!\! \Delta; B'; R' \vdash
-%   \htriple{ P' }{ S }{ Q' } \!\! \\
-%   \!\! Q' \models Q \!\!
-% }
-% { \Delta; B; R \vdash
-% \htriple{ P }{ S }{ Q }
-% }
-% %
-% \and
-% \Rule{L:Frame}
-% { \Delta; B; R \vdash
-% \htriple{ P }{ S }{ Q } \\
-%   x \in \Qplusz
-% }
-% { \!\! \Delta; B + x; R + x \vdash
-% \htriple{ P + x }{ S }{ Q + x } \!\!
-% }
-% %
-% \end{mathpar}
-% \caption{Selected rules of the Quantitative Hoare Logic}
-% \label{fig:logic}
-% \end{figure*}
-
-\paraskip
-\paragraph{Quantitative Hoare Logic.}
-To prove the soundness of \toolname we found it useful to go through
-an intermediate step using a quantitative Hoare logic.  This logic is
-at the same time a convenient semantic tool and a clean way to
-interface manual proofs with our automation.  We base it on a logic
-for stack usage~\cite{veristack14}, add support for arbitrary
-resources, and simplify the handling of auxiliary state.
-
-We define quantitative Hoare triples
-as $\iffull{\Delta;\,}{}B;R\vdash_L \htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
-maps from program states to an element of $\Q\cup\{\infty\}$ that represents
-an amount of resources available.  The assertions $B$ and $R$ are
-postconditions for the case in which the block $S$ exits by a \code{break}
-or \code{return} statement.  Additionally, $R$ depends on the return
-value of the current function.  The meaning of the triple $\htriple Q
-S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
-the empty continuation, $Q'(\state) \ge 0$ for all state $\sigma$, and at least $Q(\state)$ resource units available
-then the evaluation does not run out of resources and there are at
-least $Q'(\state')$ resources left if the evaluation terminates in state
-$\state'$.  The logic rules are similar to the ones in previous
-work.
-
-Finally, we define a strong compositional continuation-based soundness for
-triples and prove the validity of all the rules in Coq.  \ifshort{The full
-version of this paper \cite{anon_extended}, provides explanations for
-the rules and a thorough overview of our soundness proof.}{Details about
-the logic and its soundness proof are provided in the \pref{app:logic}.}
-
-\paragraph{The Soundness Theorem.}
-We use the quantitative logic as the target of a translation
-function for the automatic derivation system.  This reveals two orthogonal
-aspects of the proof: on one side, it relies on amortized reasoning (the
-quantitative logic rules), and on the other side, it uses combinatorial
-properties of our linear potential functions (the automatic analysis
-rules).
-
-Technically, we define a translation function $\tr$ such that if
-a judgement $J$ in the automatic analysis is derivable, $\tr(J)$
-is derivable in the quantitative logic.  By using $\tr$ to translate
-derivations of the automatic analysis to derivations in the quantitative
-logic we can directly obtain a certified resource bound for the analyzed
-program.
-
-The translation of an assertion $(\Gamma; Q)$ in the automatic
-analysis is defined by
-$$
-  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
-$$
-where we write $\Phi_Q$ for the unique linear potential function
-defined by the quantitative annotation $Q$.  The logical context $\Gamma$
-is implicitly lifted to a quantitative assertion by mapping a state
-$\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
-\iffull{We also need to
-translate the assumptions in function contexts of the automatic
-analysis.  We define $\tr(\Gamma_f; Q_f, \Gamma'_f; Q'_f) :=$
-%
-\begin{multline*}
-\forall z \, \vec v \, v . \,
-(
-  \lambda \vec v \, \state .\, \tr(\Gamma_f; Q_f)(\state[\Vargs {\mapsto} \vec v]), \\
-  \lambda v \, \state .\, \tr(\Gamma'_f; Q'_f)(\state[\Vret {\mapsto} v])
-  )
-\end{multline*}}{}%
-%
-These definitions let us translate the judgement
-$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
-\iffull{$$
-  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash_L
-    \htriple{\tr(P)}{S}{\tr(P')}.
-$$}{%
-$$
-  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
-$$}%
-The soundness of the automatic analysis can now be stated
-formally with the following theorem.
-%
-\begin{theorem}[Soundness of the automatic analysis]
-  If $J$ is a judgement derived by the automatic analysis,
-  then $\tr(J)$ is a quantitative Hoare triple derivable
-  in the quantitative logic.
-\end{theorem}
-%
-\noindent
-The proof of this theorem is constructive and maps each rule
-of the automatic analysis directly to its counterpart in the
-quantitative logic.  The most tricky parts are the translations of the
-rules for increments and decrements and the rule {\sc Q:Weak} for
-weakening because they make essential use of the algebraic properties
-of our linear potential functions.
-\iffull{ For instance we prove the following lemma to show the soundness
-of the translation of {\sc Q:Weak}.
-
-\begin{lemma}[Relax]
-  If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
-  $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
-\end{lemma}
-\begin{proof}
-We observe the following inequations.
-\begin{align*}
-\Phi_{Q'}(\state) &= \sum_i q'_i f_i(\state) \\
-&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
-  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(\state)  \\ &
-  \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(\state)
-  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
- &\ge \Phi_Q(\state)
-  + \sum_{i\in\mathcal U} (u_i - f_i(\state)) p_i
-  + \sum_{i\in\mathcal L} (f_i(\state) - l_i) p_i \\
- &\ge \Phi_Q(\state).
-\end{align*}
-Since $\state \models \Gamma$, using the transitivity of $\models$ and
-the definition of $\mathcal U$ and $\mathcal L$ we get
-$\forall i\in\mathcal L.\, l_i \le f_i(\state)$ and
-$\forall i\in\mathcal U.\, f_i(\state)\le u_i$.
-These two inequalities justify the last step.
-\end{proof}
-%
-\noindent
-We prove similar lemmas for the rules that deal with assignment.  With
-the computational content of the proof, we have an effective algorithm
-to construct certificates for upper bounds derived automatically.
-}{}
-
-\begin{figure*}
-\centering
-\vspace{-.5cm}
-\begin{tabular}{r|c|c|c|c|c}
-%
-& t09 & t19 & t30 & t15 & t13\\
-%
-\hline
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-i=1; j=0;
-while (j<x) {
-  j++;
-  if (i>=4)
-    i=1, tick(40);
-  else i++;
-  tick(1); }
-\end{lstlisting}
-&
-\hspace{-.5cm}\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (i>100) {
-  i--; tick(1);
-} i += k+50;
-while (i>=0) {
-  i--; tick(1);
-}\end{lstlisting}\hspace{-.5cm}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (x>0) {
-  x--;
-  t=x, x=y, y=t;
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-assert(y>=0);
-while (x > y) {
-  x -= y+1;
-  for (z=y; z>0; z--)
-    tick(1);
-  tick(1);
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[basicstyle=\tt\footnotesize]
-while (x>0) {
-  x--;
-  if (*) y++;
-  else
-    while (y>0)
-      y--, tick(1);
-  tick(1); }
-\end{lstlisting}
-%
-\\
-\hline
-%
-%
-%
-\toolname
-  & {\small $11|[0,x]|$ }
-  & {\small $\!\!\!\!\!50 {+} |\inter {-1} i| {+} |\inter 0 k|\!\!\!\!\!\!$ }
-  & {\small $|\inter 0 x| {+} |\inter 0 y|$}
-  & {\small $|\inter 0 x|$}
-  & {\small $2 |\inter 0 x| {+} |\inter 0 y|$}
-\\
-%
-Rank
-  & {\small $23{\cdot}x - 14$ }
-  & {\small $54+k+i$ }
-  & {\small ---}
-  & {\small $2+2x-y$ }
-  & {\small $0.5{\cdot}y^2 {+} yx \ldots$}
-           % +0.5{\cdot}x^2\ldots$} % + 1{-}0.5{\cdot}y){-}0.5{\cdot}x $}
-\\
-%
-{\hspace{-.2cm}LOOPUS}
-&
-{\small
-  $41\max(x,0)$
-}
-&
-{\small
-  $\!\!\!\!$\begin{tabular}{c}
-    $\max(i{-}100, 0)$ \\
-    $+ \max(k {+} i {+} 51, 0)$
-  \end{tabular}$\!\!\!\!$
-}
-&
-{\small ---}% $1 + \max(x, y)$ (???)
-&
-{\small ---}
-&
-{\small \begin{tabular}{c}
-           $2\max(x, 0)$ \\
-           $+ \max(y, 0)$
-        \end{tabular}}
-\end{tabular}
-\vspace{.01cm}
-\caption{ Comparison of resource bounds derived by different tools on
-  several examples with linear bounds.  \iffull{The output of Rank has been manually
-  simplified to fit the table.}{}  }
-\label{fig:compar}
-\end{figure*}
-
-\sectskip
-\section{Experimental Evaluation}
-\label{sec:exper}
-\aftersectskip
-
-We have experimentally evaluated the practicality of our automatic
-amortized analysis with more than 30 challenging loop and recursion
-patterns from open-source code and the
-literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
-examples \iffull{together with the derived bounds is given in
-  \pref{app:cat}.}{is given in the extended
-  version~\cite{anon_extended}.}
-
-\pref{fig:compar} shows five representative loop patterns from the
-evaluation.
-Example \emph{t09} is a loop that performs an expensive operation every
-4 steps.  $\toolname$ is the only tool able to amortize this cost over
-the input parameter $x$.
-Example \emph{t19}
-demonstrates the compositionality of the analysis.  The program consists
-of two loops that decrement a variable $i$.  In the first loop, $i$ is
-decremented down to 100 and in the second loop $i$ is decremented further
-down to $-1$.  However, between the loops we assign \lstinline{i=i+k+50}.
-So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
-Our analysis finds this tight bound because our amortized analysis
-naturally takes into account the relation between the two loops.
-Example \emph{t30} decrements both input variables $x$ and $y$ down to
-zero in an unconventional way.  In the loop body, first $x$ is decremented
-by one, then the values of the variables $x$ and $y$ are switched using
-the local variable $t$ as a buffer.  Our analysis infers the tight
-bound $|[0,x]|+|[0,y]|$.  Sometimes we need some assumptions on the
-inputs in order to derive a bound.  Example \emph{t15} is such a case.
-We assume here that the input variable $y$ is non-negative and write
-\lstinline{assert(y>=0)}.  The assignment \lstinline{x -= y+1} in the
-loop is split in \lstinline{x--} and \lstinline{x -= y}. If we enter the
-loop then we know that $x>0$, so we can obtain constant potential from
-\lstinline{x--}.  Then we know that $x\geq y\geq 0$, as a consequence
-we can share the potential of $|[0,x]|$ between $|[0,x]|$ and $|[0,y]|$
-after \lstinline{x -= y}.  % In this way, we derive a tight linear bound.
-
-Example \emph{t13} shows how amortization can be used to find linear
-bounds for nested loops.  The outer loop is iterated $|[0,x]|$ times.
-In the conditional, we either (the branching condition is arbitrary)
-increment the variable $y$ or we execute an inner loop in which $y$ is
-counted back to $0$.  $\toolname$ computes a tight bound.
-\iffull{Again, the constants $0$ and $1$ in the
-  inner loop can as well be replace by something more interesting, say
-  $9$ and $10$ like in Example \emph{t08}.  Then we still obtain a
-  tight linear bound.}{}
-%
-\ifdefined\fullversion
-\begin{figure}[t]
-  \centering
-    \begin{minipage}[b]{.9\linewidth}
-    \begin{center}
-   \begin{lstlisting}
-int srch(
-  int t[], int n,  /* haystack */
-  int p[], int m,  /* needle */
-  int b[]
-) { int i=0, j=0, k=-1;
-    while (i < n) {
-      while (j >= 0 && t[i]!=p[j]) {
-        k = b[j];
-        assert(k > 0 && k <= j + 1);
-        j -= k; tick(1)
-      }
-      i++, j++;
-      if (j == m) break;
-      tick(1);
-    }
-    return i;
-  }
-   \end{lstlisting}
-\vspace{-1.2cm}
-\hspace{3.8cm}
-\footnotesize
-\begin{tabular}{|rl|}
-\hline
-\toolname & $1 {+} 2|\inter 0 n|$ \\
-Rank & $O(n^2)$ \\
-LOOPUS & $2 {+} 3\max(n, 0)$\\
-\hline
-\end{tabular}
-    \end{center}
-\vspace{-0.1cm}
-  \end{minipage}
-  \caption{The Knuth-Morris-Pratt algorithm for string search.}
-  \label{fig:KMP}
-\end{figure}
-
-Finally, \pref{fig:KMP} contains the search function of the
-Knuth-Morris-Pratt algorithm for string search.  Our automatic
-amortized analysis finds the tight linear bound $1+2|[0,n]|$.  We need
-to assert that the elements $b[j]$ of the failure table $b$ are in the
-interval $[1,j+1]$.  This is guaranteed by
-construction of the table in the initialization procedure of the
-algorithm, which we can also analyze automatically.  We need the
-assertion since we do not infer any logical assertions on the contents
-of the heap.  Rank derives a complex quadratic bound and LOOPUS
-derives a linear bound.
-\else
-The extended version also contains a discussion of the automatic
-bound derivation for the Knuth-Morris-Pratt algorithm for string
-search. $\toolname$ finds the tight linear bound $1+2|[0,n]|$.
-\fi
-
-To compare our tool with existing work, we focused on loop bounds and
-use tick statements to counts the number of back edges (i.e., number
-of loop iterations) that are followed in the execution of the program
-because most other tools only bound this specific cost. In
-\pref{fig:compar}, we show the bounds we derived ($\toolname$) together with
-the bounds derived by LOOPUS~\cite{SinnZV14} and
-Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
-have not been able to obtain this tool.
-KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
-cannot operate on C code and the examples would need to be manually
-translated into a term-rewriting system to be analyzed by these tools.  For
-Rank it is not completely clear how the computed bound relates to the C
-program since the computed bound is for transitions in an automaton
-that is derived from the C code.  For instance, the bound $2+y-x$ that
-is derived for \emph{t08} only applies to the first loop in the
-program.
-
-\begin{table}[t]
-\small
-\centering
-\begin{tabular}{r|c|c|c|c|c}
-& KoAT & Rank & LOOPUS & SPEED & \toolname \\
-\hline
-\#bounds & 9 & 24 & 20 & 14 & 32
-\\
-\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
-\\
-\#best bounds & 0 & 0 & 11 & 14 & 29
-\\
-\#tested & 14 & 33 & 33 & 14 & 33
-\end{tabular}
-\vspace{.01cm}
-\caption{Comparison of $\toolname$ with other
-  automatic tools.}
-\label{tab:compar}
-\end{table}
-
-\begin{table}[t]
-\centering
-\small
-\begin{tabular}{r|c|c|c}
-Function & LoC & Bound & Time (s) \\
-\hline
-\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-\code{MD5Final} & 195 & 141 & 0.22 \\
-\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-% \code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
-\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
-\hline
-\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-\end{tabular}
-\vspace{.1cm}
-\caption{Derived bounds for functions from cBench.}
-\label{tab:cbench}
-\end{table}
-
-\pref{tab:compar} summarizes the results of our experiments presented
-in \pref{tab:eval}.  It shows
-for each tool the number of derived bounds (\#bounds), the number of
-asymptotically tight bounds (\#lin. bounds), the number of bounds with
-the best constant factors in comparison with the other tools (\#best
-bounds), and the number of examples that we were able to test with the
-tool (\#tested).  Since we were not able to run the experiments for KoAT
-and SPEED, we simply used the bounds that have been reported by the
-authors of the respective tools.  The results show that our automatic
-amortized analysis outperforms the existing tools on our example
-programs.  However, this experimental evaluation has to be taken with
-a grain of salt.  Existing tools complement $\toolname$
-since they can derive polynomial bounds and support
-more features of C.  We were particularly impressed by LOOPUS which is
-very robust, works on large C files, and derives very precise bounds.
-\iffull{We did not include the running times of the tools in the table since
-all tested tools work very efficiently and need less then a second
-on every tested example.}{}
-
-\pref{tab:cbench} contains a compilation of the results of our
-experiments with the cBench benchmark suite. It shows a representative
-list of automatically derived function bounds. In total we analyzed
-more than $\totalLoc$ lines of code.  In the LoC column
-we not only count the lines of the analyzed function but
-also the ones of all the function it calls. \iffull{The line count is
-computed on pre-processed C with all unnecessary declarations removed.}{}
-We analyzed the functions by giving a cost 1 to all
-the back-edges in the control flow (loops, and function calls).
-The bounds for the functions \code{ycc\_rgb\_conv} and \code{uv\_decode}
-have been inferred with user interaction as described in \pref{sec:anno}.
-The most challenging functions for $\toolname$ have unrolled loops
-where many variables are assigned.  This stresses our analysis because
-the number of LP variables has a quadratic growth in program variables.
-Even on these stressful examples, the analysis could finish in less
-than 2 seconds.  \iffull{These times can be reduced by adding
-manual annotations to indicate $\toolname$ that certain variables do
-not have to be tracked.  We could also envision to implement heuristics.}{}
-%
-For example, the \code{sha\_update} function
-is composed of one loop calling two helper functions that in turn
-have 6 and 1 inner loops.  In the analysis of the SHA algorithm,
-the compositionality of our analysis is essential to get a tight bound
-since loops on the same index are sequenced 4 and 2 times without
-resetting it.  All other tools derive much larger constant factors.
-
-With our formal cost semantics, we can run our examples
-for different inputs and measure the cost to compare it
-to our derived bound.  \pref{fig:3d} shows such a comparison for
-Example \emph{t08}, a variant of \emph{t08a} from \pref{sec:AAA}.  One
-can see that the derived constant factors are the best possible if the
-input variable $x$ is non-negative.  \iffull{If $x$ is negative then
-  the bound is only slightly off.}{}
-
-% \begin{table}[t]
-% \small
-% \centering
-% \begin{tabular}{r|c|c|c}
-% Function & LoC & Bound & Time (s) \\
-% \hline
-% \code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
-% \code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
-% \code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
-% \code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
-% \code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
-% \code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
-% \code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
-% \code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
-% \code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
-% \code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
-% \hline
-% \code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
-% \code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
-% \end{tabular}
-% \vspace{.1cm}
-% \caption{Derived bounds for functions from cBench.}
-% \label{tab:cbench}
-% \end{table}
-
-\begin{figure}[t]
-\center
-% \vspace{-.2cm}
-\includegraphics[width=.9\linewidth]{fig/bound3d}
-\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
-  (blue lines) and the measured runtime cost (red crosses) for Example
-  \emph{t08}. For $x\ge 0$ the bound is tight.}
-\label{fig:3d}
-\end{figure}
-
-\sectskip
-\section{Limitations}
-\label{sec:limitations}
-\aftersectskip
-
-Our implementation does not currently support
-all of Clight.  Programs with function pointers, goto statements,
-continue statements, and pointers to stack-allocated variables
-cannot be analyzed automatically.  While these limitations concern the
-current implementation, our technique is in principle capable to handle
-them.
-
-For the sake of simplicity, the automated system described here is
-restricted to finding only linear bounds.  However, the amortized analysis
-technique was shown to work with polynomial bounds~\cite{HoffmannAH11};
-we leave this extension of our system as future work.
-
-Even certain linear programs cannot be analyzed automatically by
-$\toolname$, it is usually the case for programs that rely on heap
-invariants (like nul-terminated C strings), for programs in which
-resource usage depends on the result of non-linear operations (like
-\code{\%} or \code{*}) in a non-trivial way, or for programs whose
-termination can only be proved by complex path-sensitive reasoning.
-
-\sectskip
-\section{Related Work}
-\label{sec:related}
-\aftersectskip
-
-Our work has been inspired by type-based amortized resource analysis
-for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  Here,
-we present the first automatic amortized resource analysis for
-C. None of the existing techniques can handle the example programs we
-describe in this work.
-%
-The automatic analysis of realistic C programs is enabled by two
-major improvements over previous work.  First, we extended the
-analysis system to associate potential with not just individual
-program variables but also multivariate intervals
-and, more generally, auxiliary variables.  In this way, we solved the
-long-standing open problem of extending automatic amortized resource
-analysis to compute bounds for programs that loop on (possibly
-negative) integers without decreasing one individual number in each
-iteration.  Second, for the first time, we have combined an automatic
-amortized analysis with a system for interactively deriving bounds. In
-particular, recent systems~\cite{HoffmannS13} that deal with integers
-and arrays cannot derive bounds that depend on values in mutable
-locations, possibly negative integers, or on differences between
-integers.
-
-A recent project~\cite{veristack14} has implemented and verified a
-quantitative logic to reason about stack-space usage, and modified the
-verified CompCert C compiler to translate C level bound to x86 stack
-bounds.  This quantitative logic is also based on the potential method
-but has very rudimentary support for automation. It is not based on
-efficient LP solving and cannot automatically derive symbolic
-bounds.  In contrast, our main contribution is an automatic amortized
-analysis for C that can derive parametric bounds for loops and
-recursive functions fully automatically.  We
-use a more general quantitative Hoare
-logic that is parametric over the resource of interest.
-
-\iffull{
-In the development of our quantitative Hoare logic we have drawn
-inspiration from mechanically verified Hoare logics.
-Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
-logics in Isabelle/HOL has been helpful to understand the interaction
-of auxiliary variables with the consequence rule.  Appel's separation
-logic for CompCert Clight~\cite{AppelLogic} has been a blueprint for
-the general structure of the quantitative logic.  \iffull{Since we do not deal
-with memory safety, our logic is much simpler and it would be possible
-to integrate it with Appel's logic.}{}  The continuation passing style
-that we use in the quantitative logic is not only used by
-Appel~\cite{AppelLogic} but also in Hoare logics for low-level
-code~\cite{NiS06,JensenBK13}.
-
-There exist quantitative logics that are integrated into separation
-logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
-quantitative logic.  However, the purpose of these logics is slightly
-different since they focus on the verification of bounds that depend
-on the shape of heap data structures and they are not implemented for
-C.  Also closely related to our logic is a VDM-style logic for
-reasoning about resource usage of an abstract fragment of JVM byte
-code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
-Hoare-style, does not target C code, and is not designed for
-interactive bound development but to produce certificates for bounds
-derived for high-level functional programs.
-}{}
-
-There exist many tools that can automatically derive loop and
-recursion bounds for imperative programs such as
-SPEED~\cite{GulwaniMC09,GulwaniZ10}, KoAT~\cite{BrockschmidtEFFG14},
-PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
-ABC~\cite{BlancHHK10} and LOOPUS~\cite{Zuleger11,SinnZV14}.  These
-tools are based on abstract interpretation--based invariant generation
-and/or term rewriting techniques, and they derive impressive results
-on realistic software.  The importance of amortization to derive tight
-bounds is well known in the resource analysis
-community~\cite{AlonsoG12,Moser14,SinnZV14}.  Currently, the only
-other available tools that can be directly applied to C code are Rank
-and LOOPUS.  As demonstrated, $\toolname$ is more compositional than
-the aforementioned tools. Our technique, is the only one that can
-generate resource specifications for functions, deal with resources
-like memory that might become available, generate proof certificates
-for the bounds, and  support user guidance that separates
-qualitative and quantitative reasoning.
-
-There are techniques~\cite{Braberman08} that can compute the memory
-requirements of object oriented programs with region-based garbage
-collection.  \iffull{These systems infer invariants and use external tools
-that count the number of integer points in the corresponding polytopes
-to obtain bonds.  The described technique can handle loops but not
-recursive or composed functions.}{ These systems can handle loops but not
-recursive or composed functions.}
-%
-We are only aware of two verified quantitative analysis systems.
-Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
-automatically verify previously inferred loop invariants, size
-relations, and ranking functions for Java Card programs.  However,
-they do not have a formal cost semantics and do not prove the bounds
-correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
-have verified a loop bound analysis for CompCert's RTL intermediate
-language.  However, this automatic bound analysis does not
-compute symbolic bounds. \iffull{Furthermore, there is no way to
-  interactively derive bounds or to deal with resources like memory
-  usage.}
-
-\sectskip
-\section{Conclusion}
-\label{sec:concl}
-\aftersectskip
-
-We have developed a novel analysis framework for compositional and
-certified worst-case resource bound analysis for C programs.  The
-framework combines ideas from existing abstract interpretation--based
-techniques with the potential method of amortized analysis.  It is
-implemented in the publicly available tool $\toolname$. To the best of
-our knowledge, $\toolname$ is the first tool for C programs
-that automatically reduces the derivation of symbolic bounds to LP
-solving.
-
-We have demonstrated that our approach improves the state-of-the-art
-in resource bound analysis for C programs in three ways. First, our
-technique is naturally compositional, tracks size changes of
-variables, and can abstractly specify the resource cost of functions
-(\pref{sec:AAA}). Second, it is easily combined with established
-qualitative verification to guide semi-automatic bound derivation
-(\pref{sec:anno}). Third, we have shown that the local inference rules
-of the derivation system automatically produce easily checkable
-certificates for the derived bounds (\pref{sec:soundness}).
-%
-Our system is the first amortized resource analysis for C programs.
-It addresses the long-standing open problem of extending automatic
-amortized resource analysis to compute bounds for programs that loop
-on signed integers and to deal with non-linear control flow\ifshort{.}{ that is
-introduced by break and return statements.}
-
-This work is the starting point for several projects that we plan to
-investigate in the future, such as the extension to
-concurrency, better integration of low-level features like
-memory caches, and the extension of the automatic analysis to
-multivariate resource polynomials~\cite{HoffmannAH11}.
-
-% \acks
-
-% This research is based on work supported in part by NSF grants 1319671
-% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
-% Any opinions, findings, and conclusions contained in this document are
-% those of the authors and do not reflect the views of these agencies.
-
-\ifx\fullversion\undefined{}\else{
-%\clearpage
-\appendix
-
-\newpage
-
-\section{Syntax and Semantics}
-\label{app:sem}
-
-We implemented our cost semantics and the quantitative Hoare logic in
-Coq for \emph{CompCert Clight}.  Clight is the most abstract
-intermediate language used by CompCert.  Mainly, it is a subset of C
-in which loops can only be exited with a \code{break} statement and
-expressions are free of side effects.
-
-\paragraph{Syntax.}
-
-We describe our system for a subset of Clight that is sufficient to
-discuss the general ideas.  This subset is given by the following
-grammar.
-%
-\begin{align*}
-S &:= \code{assert}~E
-\mid \code{skip}
-\mid \code{break}
-\mid \code{return}~x
-\mid x \gets E
-\mid x \gets f(x^*)
-\\
-& \mid \code{loop}~S
-\mid \code{if}(E)~S~\code{else}~S
-\mid S;S
-\mid \code{tick}(n)
-\end{align*}
-%
-Expressions $E$ are left abstract in our presentation.  For our
-analysis framework, it is only important that they are side-effect
-free.
-%
-The most notable difference to full Clight is that we can only assign
-to variables and thus do not consider operations that update the heap
-or the stack through a pointer.
-Moreover, function arguments and return values are assumed to be
-variables.  This is only for simplifying the presentation; in the
-implementation we can deal with heap updates and general function
-calls and returns.  However, we have not implemented our framework for function
-pointers, \code{goto} statements, \code{continue} statements, and
-stack variables that have their address taken.
-
-We include the built-in primitive $\code{assert}~e$ that terminates
-the program if the argument $e$ evaluates to false and has no effect
-otherwise.  This is useful to express assumptions on the inputs of a
-program for the automatic analysis.  We also add the built-in function
-\code{tick(n)} that can be called with a constant rational number $n$
-as a flexible way to model resource consumption or release (if $n$ is
-negative).
-
-\begin{figure*}[t!]
-\begin{mathpar}
-%
-\Rule{S:Assert}
-{ \code{istrue}~\sem{e}_\state }
-{ (\state, \code{assert}~e, \cont, \cost)
-\smallstep
-(\state, \code{skip}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkSeq}
-{ }
-{ (\state, \code{break}, \Kseq S \cont, \cost)
-\smallstep
-(\state, \code{break}, \cont, \cost)
-}
-%
-\and
-\Rule{S:BrkLoop}
-{ }
-{ (\state, \code{break}, \Kloop S \cont, \cost)
-\smallstep
-(\state, \code{skip}, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetSeq}
-{ }
-{ (\state, \code{return}~x, \Kseq S \cont, \cost)
-\smallstep
-(\state, \code{return}~x, \cont, \cost)
-}
-\and
-%
-\and
-\Rule{S:RetLoop}
-{ }
-{ (\state, \code{return}~x, \Kloop S \cont, \cost)
-\smallstep
-(\state, \code{return}~x, \cont, \cost)
-}
-%
-\and
-\Rule{S:RetCall}
-{ \state = (\_, \gamma) \\
-  \state' = (\theta, \gamma)[r \mapsto \state(x)]
-}
-{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
-\smallstep
-(\state', \code{skip}, \cont, \cost)
-}
-%
-\and
-\Rule{S:Update}
-{ \state' = \state[x \mapsto \sem{e}_\state] }
-{ (\state, x \gets e, \cont, \cost)
-\smallstep
-(\state', \code{skip}, \cont, \cost))
-}
-%
-\and
-\Rule{S:Call}
-{ \fenv f = (\vec x, S_f) \\
-  \state = (\theta, \gamma) \\
-  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
-}
-{ (\state, r \gets f(\vec y), \cont, \cost)
-\smallstep
-(\state', S_f, \Kcall r \theta \cont, \cost)
-}
-%
-\and
-\Rule{S:Loop}
-{ }
-{ (\state, \code{loop}~S, \cont, \cost)
-\smallstep
-(\state, S, \Kloop S \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipLoop}
-{ }
-{ (\state, \code{skip}, \Kloop S \cont, \cost)
-\smallstep
-(\state, \code{loop}~S, \cont, \cost)
-}
-%
-\and
-\Rule{S:IfTrue}
-{ \code{istrue}~\sem{e}_\sigma
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep
-(\state, S_1, \cont, \cost)
-}
-%
-\and
-\Rule{S:IfFalse}
-{ \code{isfalse}~\sem{e}_\sigma
-}
-{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
-\smallstep
-(\state, S_2, \cont, \cost)
-}
-%
-\and
-\Rule{S:Seq}
-{ }
-{ (\state, S_1; S_2, \cont, \cost)
-\smallstep
-(\state, S_1, \Kseq {S_2} \cont, \cost)
-}
-%
-\and
-\Rule{S:SkipSeq}
-{ }
-{ (\state, \code{skip}, \Kseq S \cont, \cost)
-\smallstep
-(\state, S, \cont, \cost)
-}
-%
-\and
-\Rule{S:Tick}
-{ }
-{ (\state, \code{tick}(n), \cont, \cost)
-\smallstep
-(\state, \code{skip}, \cont, \cost - n)
-}
-
-\end{mathpar}
-\caption{Rules of the operational semantics of statements.}
-\label{fig:opsem}
-\end{figure*}
-
-\paragraph{Semantics.}
-
-\pref{fig:opsem} contains the reduction rules of the semantics.  The rules
-define a rewrite system for program configurations of the form $(\state,
-S, \cont, \cost)$, where $\state$ is the program state, $S$ is the
-statement being executed, $\cont$ is a continuation that describes what
-remains to be done after the execution of $S$, and $\cost \in \Q$ is
-the amount of resources available for further execution.  All the rules
-have the implicit side condition that the resource quantity available
-\emph{before} the step is non-negative.  This means that we allow
-$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side transition
-relation $\smallstep^*$ to indicate that the execution ran out of
-resources.  However, every execution that reaches such a state is stuck.
-
-A continuation $\cont$ represents the context of the execution of a
-given statement.  A continuation can be the empty continuation
-$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
-\cont$, a loop continuation $\Kloop S \cont$, or the continuation
-$\Kcall r \theta \cont$ of a function call.
-$$
-\cont := \Kstop
-\mid \Kseq S \cont
-\mid \Kloop S \cont
-\mid \Kcall r \theta \cont
-$$
-%
-During the execution we assume a fixed function context $\fenv$ that
-maps function names to a list of variables and the statement that
-defines the function body.  It is used in the rule \textsc{S:Call}.
-
-The intuitive meaning of an evaluation $(\state,
-S, \cont, \cost) \smallstep^* (\state', S', \cont', \cost')$ is the
-following.  If the statement $S$ is executed in program state
-$\state$, with continuation $\cont$, and with $\cost$ resources
-available then---after a finite number of steps---the evaluation will
-reach the new machine state $(\state', S', \cont',\cost')$ and there
-are $\cost'$ resources available.  If $\cost' \ge 0$ then the
-execution did not run out of resources and the resource consumption up
-to this point is $\cost - \cost'$.  If this difference is negative
-then resource became available during the execution.  If however
-$\cost' < 0$ then the execution ran out of resources and is stuck.
-The cost of the execution is then $\cost \ge 0$.
-
-\pref{lem:sem1} summarizes the main properties of the resource
-counter of the semantics.  It can be proved by induction on the number
-of steps.
-\begin{lemma}
-\label{lem:sem1}
-\itemskip
+\subsection{Analyzing Control Flow Graphs}
+
+Let $G = (V,E,\labels)$ be a CFG. A potential annotation $\mathcal{P}$
+for $G$ is mapping from notes $v \in V$ to pairs
+$\left<(\Gamma,Q);(\Gamma',Q')\right>$ of quantitative assertions so
+that the following holds.
+%
 \begin{enumerate}
-  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
-    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
-    \smallstep^k (\state', S', \cont', \cost' {+} n)$.
-\itemskipIn
-  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
-    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep^k (\state', S',
-    \cont', d')$ then $\cost - \cost' = d - d'$.
+\item  For every edge $(v_1,v_2) \in E$: 
+  If $\mathcal{P}(v_1) = \left<(\Gamma_1,Q_1);(\Gamma'_1,Q'_1)\right>$ and 
+   $\mathcal{P}(v_2) = \left<(\Gamma_2,Q_2);(\Gamma'_2,Q'_2)\right>$ then
+   $\Gamma'_1 \models \Gamma_2$ and $Q_1' \geq Q_2$.
+ \item For every node $v \in V$: If $\labels(v) = S$ and
+   $\mathcal{P}(v) = \left<(\Gamma,Q);(\Gamma',Q')\right>$ then
+  $\htriple{(\Gamma,Q)}{S}{(\Gamma',Q')}$.
 \end{enumerate}
-\itemskip
-\end{lemma}
-
-% OOPS this is no longer true... This is an argument agains a presentation
-% with ticks only. XXX
-%
-% If we assume that we have a \emph{positive resource metric}, that is,
-% a metric that assigns a positive cost to each evaluation step then we
-% can link bounds on the resource consumption under the metric to
-% termination.  This is formalized by \pref{lem:sem_term}.
-%
-% \begin{lemma}
-%  \label{lem:sem_term}
-%  Let $M$ be a positive metric and assume that there exists $b \in
-%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep^*
-%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
-%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
-%  \smallstep^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
-%  and $\state''$ implies $n \le B$.
-% \end{lemma}
-
-\section{Quantitative Hoare Logic}
-\label{app:logic}
-
-\begin{figure*}
-\begin{mathpar}
-%
-\Rule{L:Skip}
-{ }
-{ \Delta; B; R \ldash
-\htriple
-  { Q }{ \code{skip} }{ Q }
-}
-%
-\and
-\Rule{L:Break}
-{ }
-{ \Delta; B; R \ldash
-\htriple
-  { B }{ \code{break} }{ Q }
-}
-\and
-\Rule{L:Tick}
-{ n < 0 \implies Q \ge 0}
-{ \Delta; B; R \ldash
-\htriple
-  { Q }
-  { \code{tick}(n) }
-  { Q - n }
-}
-%
-%
-\and
-\Rule{L:Return}
-{ }
-{ \Delta; B; R \ldash
-\htriple
-  { R\,(\state(x)) }{ \code{return}~x }{ Q }
-}
-%
-\and
-\Rule{L:Assert}
-{ }
-{ \Delta; B; R \ldash
-\htriple
-  { \code{istrue}~\sem{e}_\state \implies Q }
-  { \code{assert}~e }
-  { Q }
-}
-%
-\and
-\Rule{L:Update}
-{ }
-{ \Delta; B; R \ldash
-\htriple
-  { \lambda \state.\, Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
-}
-%
-\and
-\Rule{L:Loop}
-{ I \ge Q
-\\ \Delta; Q; R \ldash
-   \htriple { I }{ S }{ I }
-}
-{ \Delta; B; R \ldash
-\htriple
-  { I }{ \code{loop}~S }{ Q }
-}
-%
-\and
-\Rule{L:Seq}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S_1 }{ Q' }
-\\\\
-  \Delta; B; R \ldash
-\htriple{ Q' }{ S_2 }{ Q }
-}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S_1;S_2 }{ Q }
-}
-%
-\and
-\Rule{L:If}
-{% P \ge 0
-%\\
-  \Delta; B; R \ldash
-  \htriple
-  { \code{istrue}~\sem{e}_\state + P }
-  { S_1 }{ Q }
-\\
-  \Delta; B; R \ldash
-\htriple
-  { \code{isfalse}~\sem{e}_\state + P }
-  { S_2 }{ Q }
-}
-{ \Delta; B; R \ldash
-\htriple
-  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
-}
-%
-\and
-\Rule{L:Call}
-{
-  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
-  P \ge P_f \, y \, (\state(\vec x)) + A \\
-  \forall v.\, (Q_f \, y \, v + A \ge \lambda \state.\, Q \, \state[r \mapsto v])
-}
-{ \Delta; B; R \ldash
-\htriple
-  { P }
-  { r \gets f(\vec x) }
-  { Q }
-}
-%
-\and
-\Rule{L:Extend}
-{  \Delta \cup \Delta'; B; R \ldash
-  \htriple{ P }{ S }{ Q }
-\\
-  P_f \ge 0
-\\\\
-  \forall f\,P_f\,Q_f.\,
-  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
-  \rightarrow
-  \forall y\,\vec v.\,
-  (\Delta \cup \Delta'; \bot; Q_f\,y \ldash
-    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
-}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Weaken}
-{
-  \Delta; B'; R' \ldash \htriple{ P' }{ S }{ Q' } \\\\
-  P \ge P' \!\! \\
-  \!\! Q' \ge Q \!\! \\
-  \!\! B' \ge B \!\! \\
-  \!\! \forall v.\, (R'\,v \ge R\,v)
-}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S }{ Q }
-}
-%
-\and
-\Rule{L:Frame}
-{ \Delta; B; R \ldash
-\htriple{ P }{ S }{ Q } \\
-  x \in \Qplusz
-}
-{ \!\! \Delta; B + x; R + x \ldash
-\htriple{ P + x }{ S }{ Q + x } \!\!
-}
-%
-\end{mathpar}
-\caption{Rules of the Quantitative Hoare Logic}
-\label{fig:logicapp}
-\end{figure*}
-
-In this section we describe a simplified version of the quantitative
-Hoare logic that we use in Coq to interactively prove resource bounds.
-%
-We generalize classic Hoare logic to express not only classical
-boolean-valued assertions but also assertions that talk about the
-future resource usage.  Instead of the usual assertions $P : \State
-\to \mathit{bool}$ of Hoare logic we use assertions
-$$
-P : \State \to \Q \cup \{ \infty \} \; .
-$$
-This can be understood as a refinement of boolean assertions where
-$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
-We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
-$\lambda \state . \, \infty$.  We sometimes call assertions
-\emph{potential functions}.  To use Coq's support for propositional
-reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
-in the implementation.  For a given $\state \in \State$, such an
-assertion can be seen as a set $B \subseteq \Q$.  However, we find
-the presentation in this article easier to read.
-
-Due to \code{break} and \code{return} statements of Clight, there are
-different possible ways to exit a block of code.  We also have to keep
-track of the resource specifications of functions.  To account for
-this in the logic, our quantitative Hoare triples have the form
-$$
-\Delta; B; R \ldash
-\htriple
-  { Q }
-  { S }
-  { Q' } \; .
-$$
-The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
-two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
-Hoare logic and the intuitive meaning is as follows.  If $S$ is
-executed with starting state $\state$, with the empty continuation
-$\Kstop$, $Q' \ge 0$, and at least $Q(\state)$ resources available then the
-evaluation does not run out of resources and there are at least
-$Q'(\state')$ resources left if the evaluation terminates in state
-$\state'$.  The assertion $B : \Assn$ provides the
-postcondition for the case in which the code block $S$ is exited by a
-\code{break} statement.  So if the execution is terminated in state
-$\state'$ with a \code{break} then $B(\state')$ resources are
-available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
-case in which the code block $S$ is exited by a $\code{return}~x$
-statement.  The integer argument of $R$ is the return value.
-Finally, the function context of judgements that we write $\Delta$ is
-a mapping from function names to specifications of the form
-$$
-  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
-$$
-The assertion $P_f \, z \, \vec v$ is the precondition of the function
-$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
-both parameterized by an arbitrary logical variable $z$ (which can be a
-tuple) that relates the function arguments with the return value.  The
-precondition also depends on $\vec v$, the values of the arguments
-at the function invocation.  Similarly, the postcondition depends on
-the return value $v$ of the function.  The use of logical variables to
-express relations between different states of an execution is a
-standard technique of Hoare logic.
-%
-To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
-the local variables on the stack, that is, $\forall z \, \vec v
-\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
-(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.
-
-For two assertions $P,Q : \Assn$, we write $P \ge Q$ to if for all
-program state $\state$, $P(\state) \ge Q(\state)$.
-
-\paragraph{Rules of the Quantitative Logic.}
-
-\pref{fig:logicapp} shows the inference rules of the quantitative logic.
-The rules are slightly simplified in comparison to the implemented
-rules in Coq.  The main difference is that the presented version does
-not formalize the heap operations.
-
-In the rule {\sc L:Skip}, we do not have to account for any resource
-consumption.  As a result, the precondition $Q$ can be any (potential)
-function and we only have to make sure that we do not end up with more
-potential.  Since the execution of \code{skip} leaves the program
-state unchanged, we can simply use the precondition as postcondition.
-The potential functions $B$ for the \code{break} and $R$ for the
-\code{return} part of the postcondition are not reachable and can
-therefore be arbitrary.
-
-The {\sc L:Tick} rule accounts for the cost $n$ of the tick statement.
-This rule is the only one to account directly for a cost on in the
-assertions because calling the \code{tick} function is the only way to
-change the amount of resource available.
-The rule also has a side condition that ensures that the potential
-remains positive during all the program execution.  It is an
-essential semantic property of this system.
-
-In the rule {\sc L:Assert}, we use the notation
-$\code{istrue}~\sem{e}_\state \implies Q$ to express that we
-require potential $Q$ in the precondition if $e$ evaluates to
-$\mathit{true}$ in the current program state.  If $e$ evaluates to
-$\mathit{false}$ then the potential in the precondition can be any
-non-negative number since the program will be terminated.
-
-In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
-be arbitrary since it is unreachable.  Instead, we have to justify the
-potential functions $B$ and $R$ that hold after a \code{return} and a
-\code{break}, respectively.  In {\sc L:Break}, we require to have
-potential $B$ in the precondition to pay for the potential after the
-\code{break}.  In {\sc L:Return} we only require to have potential
-$R$ in the precondition to pay for the potential after the
-\code{return}.
-
-The rule {\sc L:Update} is the standard assignment rule of Hoare
-logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
-the precondition we ensure that $Q$ evaluates to the same number as in
-the postcondition.
-
-The rule {\sc L:Seq} rule is crucial to understand how the
-quantitative Hoare logic works.  To account for early exits of
-statements, we must ensure in the \code{break} part $B$ of $S_1$'s
-judgement that the \code{break} part $B$ of of $S_1;S_2$ holds. The
-same is true for the return part $R$ of the judgements for $S_1$ and
-$S_2$.  The interaction between the actual pre- and postconditions is
-analogous to standard Hoare logic.
-
-In the rule {\sc L:Loop}, the \code{break} part of the
-loop body $S$ becomes the postcondition of the loop statement. We use
-an arbitrary $B$ as the \code{break} part of the judgement for
-$\code{loop}\, S$ since its operational semantics ensures that it can
-only terminate with a \code{skip} or a \code{return}.  The
-precondition $I$ of the loop is the loop invariant.  That is why we
-require to have potential $I$ available in the precondition of the
-loop body $S$.
-
-The {\sc L:If} is similar to the rule for the conditional in classic
-Hoare logic.  In the preconditions of the judgments for the two
-branches $S_1$ and $S_2$ we lift the boolean assertions
-$\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
-quantitative assertions.
-
-The pre- and postcondition $P_f$ and $Q_f$ used in the {\sc L:Call} rule
-are taken from the function context $\Delta$.  The assertions in
-the context are parametric with respect to both the values of the
-function arguments and the return value. This allows us to specify a
-bound for a function whose resource consumption depends on its
-arguments.  The arguments are instantiated by the call rule
-using the result of the evaluation of the argument variables in the
-current state.  Recall that we require that $P_f$ and $Q_f$ do
-  not depend on the local variables on the call stack, that is,
-  $\forall z \, \vec v \, \theta \, \theta' \, \gamma . \, P_f \, z \,
-  \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
-  (\theta',\gamma)$. To transfer potential that depends on local
-variables of the callee from the precondition $P$ to the postcondition
-$Q$, we use an assertion $A : \Assn$ that is independent of global
-variables, that is, $\forall \theta \, \gamma \, \gamma' . \, A
-(\theta,\gamma) = A(\theta,\gamma')$.  It is still possible to express
-relations between global and local variables using logical
-variables (see the following paragraph for details).
-
-Finally, we describe the rules which are not syntax directed.  There
-are two weakening rules available in the quantitative Hoare logic.
-The framing rule {\sc L:Frame} is designed to weaken a statement by
-stating that if $S$ needs $P$ resources to run and leaves $Q$
-resources available after its execution, then it can very well run
-with $P + c$ resources and return $Q + c$ resources.  The consequence
-{\sc L:Conseq} rule is directly imported from classical Hoare logic
-except that instead of using the logical implication $\Rightarrow$ we
-use the quantitative $\ge$.  This rule indeed weakens the statement
-since it requires more resource to run the statement and yields less
-than what has been proved to be available after its termination.
-
-\paragraph{Logical Variables and Shallow Embedding.}
-
-When specifying a program, it is often necessary to relate certain
-invariants in the pre- and postcondition.  A standard solution of Hoare
-logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
-additional variables (also called auxiliary state in the literature)
-are constant across the derivation.  For example, if $Z$ is such
-a logical variable, we can specify the function $\code{double}()$
-which doubles the global variable $x$ as
-$$
-  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
-$$
-%
-When formalizing Hoare logics in a proof assistant one can either
-fully specify the syntax and semantics of assertions and hence get a
-deep embedding, or use the assertions of the host theory to get a
-shallow embedding.  Because of its flexibility, we used the latter
-approach in our development.  This choice makes it possible to have
-logical variables almost for free: we can simply use the variable and
-binding mechanisms of Coq, our host theory.  When a logical variable
-is needed we introduce it using a universal quantifier of Coq before
-using the logic to derive triples.  For example, the Coq theorem for
-the above example would look as follows.
-%
-\begin{lstlisting}
-Theorem dbl_triple: forall Z,
-  qtriple ($\lambda\sigma$.$\sigma$(x)=Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x)=2*Z).
-\end{lstlisting}
-%
-However, this trick alone is often not sufficient when working with a
-recursive function $f$.  In that case we apply the {\sc L:Extend} rule
-of the logic.  First we add a specification $\forall z\,\vec
-v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
-$\Delta$.  Then we proceed to prove the function body $S_f$ with this
-induction hypothesis.  In this process it can be the case that we have
-to use the induction hypothesis with a different value of a logical
-variable (e.g., because the values of the arguments in the recursive
-call differ from the values of the arguments of the callee).  To cope
-with this problem, assumptions in the function context $\Delta$ are
-universally quantified over logical variables.  The {\sc L:Extend}
-rule uses the host proof assistant to require that the triple on $f$'s
-body is proved for every possible logical variable $y$.
-
-\paragraph{Using the Quantitative Logic.}
-
-In the following we demonstrate the use of the logic with two example
-derivations.
-
-In the example in~\pref{fig:xmplmax} we derive a precise runtime
-bound on a program that searches a maximal element in an array.
-Because of the location of the tick statements, the resource cost
-is closely related to the number of times the test \code{a[i] > m}
-is true during the execution.
-%
-If we define
-$$
- A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
-$$
-where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
-the number of ``maximum candidates'' in the array \code{a} seen
-by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
-So any automated tool would at best derive the linear bound $2 \cdot
-N$ for that program.  But with the expressivity of our logic it is
-possible to use the previous set cardinal directly and precisely tie
-the bound to the initial contents of the array.
-%
-The non-trivial part of this derivation is finding the loop invariant
-$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
-When the condition \code{a[i] > m} is true, we know that we
-encountered a ``maximum so far'' because $m$ is a maximal element of
-\code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
-potential unit to pay for the tick in the conditional.  In the other case, no
-maximum so far is encountered so $A(i) = A(i+1)$.
-As a side remark, if $K \ge 0$, it is possible to
-have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
-of the same program.  It can be done either by adapting the proof or by
-applying the {\sc L:Frame} rule one top of the triple already derived.
-More generally, without function calls, the {\sc L:Frame}
-rule is admissible in our system.
-
-The example in~\pref{fig:xmplbs} shows a use case for logical
-variables. On this example, we model the stack consumption:
-we account a constant cost for a function call ($M_f>0$) that
-is returned after the call ($M_r = -M_f <0$).
-We are interested in showing that a binary search function
-$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
-variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
-  + Z \cdot M_\code{bsearch}}{}{Z \cdot M_\code{bsearch}}$ to express that the stack required by the
-function is returned after the call.
-%
-The critical step in the proof is the application of the
-{\sc L:Call} rule to the recursive call.  At this point
-the context $\Delta$ contains the specification
-$
-  \forall y \, (x,l,h) \, \_.
-  ( (\lambda y \, (\_, l, h) .\, (y = \log_2(h {-} l)) {+} y {\cdot} M_\code{bsearch})
-  , (\lambda y \, \_ .\, y {\cdot} M_\code{bsearch})
-  )
-$.
-%
-Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
-- 1$.  The rest of the proof is here to make sure that the first tick
-can be paid for using regular algebra on the $\log_2$ function.
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue} \{ A(0) + N \} }$
-i=1; tick(1); m=a[0]; tick(1);
-while (i < N) {
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
-  if (a[i] > m)
-    m=a[i], tick(1);
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
-  i=i+1;
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) + 1 \} }$
-  tick(1);
-  ${\color{blue}%
-    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
-} ${\color{blue} \{ 0 \} }$
-\end{lstlisting}
-\caption{Example derivation where we wrote $A(i)$
-  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
-  we use ticks in this example to count the number of assignments performed
-  by the program.
-  }
-\label{fig:xmplmax}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-${\color{blue}%
-  P(Z) := \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
-bsearch(x,l,h) {
-  if (h-l > 1) {
-    ${\color{blue}%
-      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
-    m = h + (h-l)/2;
-    ${\color{blue}%
-      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
-    if (a[m]>x) h=m; else l=m;
-    ${\color{blue}%
-      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
-    tick($M_\code{bsearch}$);
-    ${\color{blue}\{ P(Z-1) \}}$ l = bsearch(x,l,h); ${\color{blue}\{ Q(Z-1) \}}$
-    tick($-M_\code{bsearch}$);
-  } ${\color{blue}%
-    \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
-  return l;
-} ${\color{blue} Q(Z) := \{ Z \cdot M_\code{bsearch} \} }$
-\end{lstlisting}
-\caption{Example derivation of a stack usage bound for a binary
-  search program.  The stack usage is modeled by adding a cost $M_\code{bsearch}$ before
-  the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
-  frame size of the function $\code{bsearch}$.  $Z$ is a logical variable.
-  }
-\label{fig:xmplbs}
-\end{figure}
-
-\paragraph{Soundness of Quantitative Hoare Triples.}
-
-We already gave an intuition of the meaning of judgements
-derived in the logic.  We start by defining an indexed
-predicate $\cdot \downarrow_n$ that defines a \emph{resource-safe}
-program configuration:
-\begin{multline*}
-  (\state, S, K, c) \downarrow_n \, := \\
-  \forall m \le n.\, \forall c'.\,
-    (\state, S, K, c) \smallstep^m
-    (\_, \_, \_, c') \implies 0 \le c'.
-\end{multline*}
-This predicate means that a program configuration can execute
-for a certain number of steps without running out of resources.
-Note that nothing is said about the program safety, i.e. a stuck
-configuration $C$ will satisfy $\forall n.\, C\downarrow_n$. This
-is because our logic does not prove any safety or correctness
-theorems but only focuses on resource usage.
-We can now be used to define the resource safety
-of a configuration with respect to a quantitative assertion.
-\begin{multline*}
-  \safe n P S \cont \, := \\
-  \forall \state\, c.\, P(\state) \le c \implies
-    0 \le c \wedge (\state, S, K, c) \downarrow_n.
-\end{multline*}
-Once again, we use an index in the definition of the predicate, it
-is used to perform proofs by induction for the function-call and
-loop rules of the logic.
-%
-An interesting and essential property of the previous predicate is
-that $\safe {n+1} P S \cont \implies \safe n P S \cont$.  To prove
-this weakening property, we use the quantification over all numbers
-smaller than $n$ in the definition of $\cdot \downarrow_n$.
-
-The resource safety of a continuation $\cont$ is defined using three
-assertions, one for each of the possible outcomes of a program
-statement.  We define it as follows.
-\begin{align*}
-\safeK n B R Q \cont & :=  \safe n B {\code{break}} \cont \\
- \land& \forall x. \safe n {\lambda \state.R\,(\state(x))} {\code{return}~x} \cont \\
- \land& \safe n Q {\code{skip}} \cont
-\end{align*}
-%
-We can now define the \emph{semantic validity} of a judgement $B; R
-\ldash \htriple P S Q$ of the quantitative logic without function
-context as $\valid n B R P S Q :=$
-\begin{multline*}
-\forall m \le n.\, \forall x \geq 0 .\, \forall \cont . \\
-  \safeK m {B{+}x} {R{+}x} {Q{+}x} \cont \\
-  \implies \safe m {P+x} S \cont .
-\end{multline*}
-Note how the validity of a triple embeds the frame rule of
-our logic. This refinement is necessary to have a stronger
-induction hypothesis available during the proof.
-%
-We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
-S Q$ implies $\valid n B R P S Q$.
-
-Using the semantic validity of triples we define the validity
-of a function context $\Delta$, written $\validC n \Delta$, as
-\begin{align*}
-  &\forall f .\, \Delta(f) =
-    \forall z \, \vec v \, v .\, (P_f \, z \, \vec v, Q_f \, z \, v)
-    \implies \\
-  &\qquad \forall z \, \vec v .\,
-  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
-\end{align*}
-where $S_f$ is the body of the function $f$. A full
-judgement that mentions a non-empty function context
-$\Delta$ is in fact a guarded statement: it makes
-assumptions on some functions' behavior.  The predicate
-$\validC n \Delta$ gives the precise meaning of the
-assumptions made.  It is also step-indexed to prove the
-soundness of the {\sc L:Extend} rule by induction.
-%
-We are now able to state the soundness of the quantitative logic.
-%
-\begin{theorem}[Soundness]
-  \label{thm:sound}
-  If $\Delta; B; R \ldash \htriple P S Q$ is derivable then
-  $
-    \forall n .\, \validC n \Delta
-      {\implies} \valid {n {+} 1} B R P S Q.
-  $
-\end{theorem}
-%
-\noindent
-The difference $\delta = 1$ between the index in the triple
-validity and the one in context validity arises from
-the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
-{\sc L:Call}, the language semantics makes one step and
-proceeds with the function body, so we must have
-$\delta \le 1$ to use the assumptions in $\Delta$.
-For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
-is a valid context for $n$ steps.  The induction hypothesis
-in that case says that if $\Delta \cup \Delta'$ is valid
-for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
-So if we want to solve this goal by induction, it is
-necessary that $\delta \ge 1$.  These two constraints force
-$\delta$ to be exactly one in the theorem statement.
-
-Assume that $S$ is a complete program and $\Delta$ is empty.  By
-expanding the definitions we see that $\Delta$ is valid
-for every $n$ and that $\Kstop$ is safe for every $n$ if both
-$Q$, $R$, and $B$ are non-negative on all possible program
-states. So the \pref{thm:sound} becomes
-$$
-\Delta; B; R \ldash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
-$$
-This means that from any starting state $\state$, $P(\state)$
-provides enough resources for any run of the program $S$.  This
-setting is actually the main use case of the previous theorem
-which is stated as a stronger result to allow a proof by
-induction.
-
-\section{Catalog of Automatically Analyzed Programs}
-\label{app:cat}
-
-In this appendix we provide a non-exhaustive catalog of classes of
-programs that can be automatically analyzed by our system.  For
-simplicity, and to compare our analysis with existing tools,
-the examples are automatically instrumented with tick statements
-to count the number of loop iterations and function calls.
-This is the only resource that can be measured by tools like
-KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
-we use tick statements explicitely to discuss features such as
-resource restitution.
-
-We assume that free variables in the code snippets are the inputs of
-the program. Some of the examples contain constants on which the
-computed bound depends.  These constants are randomly chosen to
-present an example but the analysis works for other constants as well.
-Note however that it is sometimes crucial that constants are positive
-(or negative) or that other relations hold.
-
-\pref{tab:eval} contains the details of our comparative evaluation.
-
-\begin{figure*}
-\setlength{\progwidth}{.24\linewidth}
-  \centering
-
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (y-x>0) {
-    x = x+1;
-  }
-  while (x>2) {
-    x=x-3;
-  }
-   \end{lstlisting}
-
-$1.33|[x,y]| + 0.33|[0,x]|$
-\\[.7\baselineskip]
-      {\bf t08}
-    \end{center}
-  \end{minipage}
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x-y>0) {
-    if (*)
-      y=y+1;
-    else
-      x=x-1;
-  }
-   \end{lstlisting}
-
-$|[y,x]|$
-\\[.7\baselineskip]
-      {\bf t10}
-    \end{center}
-  \end{minipage}
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x>0) {
-    x=x-1;
-    if (*)
-      y=y+1;
-    else {
-      while (y>0)
-        y=y-1;
-    }
-  }
-   \end{lstlisting}
-
-$2|[0,x]| + |[0,y]|$
-\\[.7\baselineskip]
-      {\bf t13}
-    \end{center}
-  \end{minipage}
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-while (n<0) {
-  n=n+1;
-  y=y+1000;
-  while (y>=100 && *){
-    y=y-100;
-  }
-}
-   \end{lstlisting}
-
-$11|[n,0]| + 0.01|[0,y]|$
-\\[.7\baselineskip]
-      {\bf t27}
-    \end{center}
-  \end{minipage}
-   \caption{Amortization and Compositionality (a).}
-  \label{fig:cat1a}
-\end{figure*}
-
-\begin{figure*}
- \setlength{\progwidth}{.18\linewidth}
-  \centering
-
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x>0) {
-    x=x-1;
-    y=y+2;
-  }
-  while (y>0) {
-    y=y-1;
-  }
-  while (y>0) {
-    y=y+1;
-  }
-   \end{lstlisting}
-
-$1 + 3|[0,x]| + |[0,y]|$
-\\[.7\baselineskip]
-      {\bf t07}
-    \end{center}
-  \end{minipage}%
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x>y) {
-    x=x-1;
-    x=x+1000;
-    y=y+1000;
-  }
-  while (y>0) {
-    y=y-1;
-  }
-  while (x<0) {
-    x=x+1;
-  }
-   \end{lstlisting}
-
-$1002|[y,x]|+|[x,0]|+|[0,y]|$
-\\[.7\baselineskip]
-      {\bf t28}
-    \end{center}
-  \end{minipage}%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  assert (y>=0);
-  while (x-y>0) {
-    x=x-1;
-    x=x-y;
-    z=y;
-    while (z>0) {
-      z=z-1;
-    }
-  }
-   \end{lstlisting}
-
-$|[0,x]|$
-\\[.7\baselineskip]
-      {\bf t15}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  assert (y>=0);
-  while (x-y>0) {
-    x=x-1;
-    x=x-y;
-    z=y;
-    z=z+y;
-    z=z+100;
-    while (z>0) {
-      z=z-1;
-    }
-  }
-   \end{lstlisting}
-
-$101|[y,x]|$
-\\[.7\baselineskip]
-      {\bf t16}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  i=1;
-  j=0;
-  while (j<x) {
-    j++;
-    if (i>=4) {
-      i=1;
-      tick(40);
-    } else
-       i++;
-    tick(1);
-  }
-   \end{lstlisting}
-
-$11|[0,x]|$
-\\[.7\baselineskip]
-      {\bf t09}
-    \end{center}
-  \end{minipage}
-
-   \caption{Amortization and Compositionality (b).}
-  \label{fig:cat1b}
-\end{figure*}
-
-\begin{figure*}
- \setlength{\progwidth}{.24\linewidth}
-  \centering
-
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (i>100) {
-    i--;
-  }
-  i=i+k+50;
-  while (i>=0) {
-    i--;
-  }
-   \end{lstlisting}
-
-$50 + |[-1,i]| + |[0,k]|$
-\\[.7\baselineskip]
-      {\bf t19}
-    \end{center}
-  \end{minipage}%
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x<y) {
-    x=x+1;
-  }
-  while (y<x) {
-    y=y+1;
-  }
-   \end{lstlisting}
-
-$|[x,y]|+|[y,x]|$
-\\[.7\baselineskip]
-      {\bf t20}
-    \end{center}
-  \end{minipage}%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x>0) {
-    x=x-1;
-    t=x;
-    x=y;
-    y=t;
-  }
-   \end{lstlisting}
-
-$|[0,x]|+|[0,y]|$
-\\[.7\baselineskip]
-      {\bf t30}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  flag=1;
-  while (flag>0) {
-    if (n>0 && *) {
-      n=n-1;
-      flag=1;
-    } else
-      flag=0;
-  }
-   \end{lstlisting}
-
-$1 + |[0, n]|$
-\\[.7\baselineskip]
-      {\bf t47}
-    \end{center}
-  \end{minipage}
-
-   \caption{Amortization and Compositionality (c).}
-  \label{fig:cat1c}
-\end{figure*}
-
-\paragraph{Amortization and Compositionality}
-
-Figures~\ref{fig:cat1a}, \ref{fig:cat1b}, and~\ref{fig:cat1c} show
-code snippets that need amortization and compositionality to obtain a
-whole program bound.
-
-Example \emph{t07} demonstrates two different features of the
-analysis.  For one thing it shows that we can precisely track size
-changes inside loops.  In the first loop, we increment $y$ by $2$ in
-each of the $|[0,x]|$ iterations.  An in the second loop, we decrement
-$y$.  For another thing it shows that we automatically recognize dead
-code if we find conflicting assertions on a branching path: After the
-second loop we know $y \leq 0$ and as a result can assign arbitrary
-potential inside the third loop where we know that $y>0$.  As a
-result, we obtain a tight bound.
-
-Example \emph{t08} shows the ability of the analysis to handle
-negative and non-negative numbers.  Note that there are no
-restrictions on the signs of $y$ and $x$.  We also see again that we
-accurately track the size change of $x$ in the first loop.
-Furthermore, \emph{t08} shows that we do not handle the constants $1$
-or $0$ in any special way.  In all examples you could replace $0$ and
-$1$ with other constants like we did in the second loop and still
-derive a tight bound.  The only information, that the analyzer needs
-is $x \geq c$ before assigning $x = x - c$.
-
-In Example \emph{t10} we also do not restrict the inputs $x$ and $y$.
-They can be negative, positive, or zero.  The star {\tt *} in the
-conditional, stands for an arbitrary assertion.  In each branch of the
-conditional we can obtain the constant potential $1$ since the interval
-size $|[y,x]|$ is decreasing.
-
-Example \emph{t13} shows how amortization can be used to handle tricky
-nested loops.  The outer loop is iterated $|[0,x]|$ times.  In the
-conditional, we either (the branching condition is again arbitrary)
-increment the variable $y$ or we execute an inner loop in which $y$ is
-counted back to $0$.  The analysis computes a tight linear bound for
-this program.  Again, the constants $0$ and $1$ in the inner loop can
-as well be replace by something more interesting, say $9$ and $10$
-like in Example \emph{t08}.  Then we still obtain a tight linear
-bound.
-
-Example \emph{t27} is similar to Example \emph{t13}.  Instead of
-decrementing the variable $x$ in the outer loop we this time increment
-the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
-we increment the variable $y$ by $1000$.  We then execute an inner
-loop that increments $y$ by $100$ until $y=0$.  The analysis can
-derive that only the first execution of the inner loop depends on the
-initial value of $y$.  We again derive a tight bound.
-
-Example \emph{t28} is particularly interesting.  In the first loop we
-decrement the size $|[y,x]|$.  However, we also shift the interval
-$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
-that this does not change the size of the interval and computes the
-tight loop bound $|[y,x]|$.  The additional two loops are in the
-program to show that the size tracking in the first loop works
-accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
-in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
-the worst case (if $x$ and $y$ are negative).
-
-Sometimes we need some assumptions on the inputs in order to derive a
-bound.  Example \emph{t15} is such a case.  We assume here that the
-input variable $y$ is non-negative and write \code{assert(y>=0)}.  The
-semantic of \code{assert} is that it has no effect if the assertion is
-true and that the program is terminated without further cost
-otherwise.  If we enter the loop then we know that $x>0$ and we can
-obtain constant potential from the assignment \code{x=x-1}.  After the
-assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
-can share the potential $|[0,x]|$ before the assignment \code{x=x-y}
-between $|[0,x]|$ and $|[0,y]|$ after the assignment.  \iffull{In this way, we
-derive a tight linear bound.}{}
-
-Example \emph{t16} is an extension of Example \emph{t15}. We again
-assume that $y$ is non-negative and use the same mechanism to iterate
-the outer loop as in \emph{t15}.  In the inner loop, we also count the
-variable $z$ down to zero and perform $|[0,z]|$ iterations.  However,
-instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
-computes the linear bound $101|[0,x]|$.  The
-assignment of potential to the size interval $|[0,x]|$ in instead of
-$|[y,x]|$ is a random choice of the LP solver.
-
-Example \emph{t09} shows how amortized reasoning handles periodically
-expensive operations.  The top-level loop is iterated $x$ times in
-a regular fashion, but every 4 iterations of this loop an expensive
-operation modeled by \code{tick(40)} is performed.  Taking the worst
-case of the loop body would yield the pessimistic bound $41|[0,x]|$,
-however our tool is able to amortize the expensive operation and derives
-instead the tight bound $11|[0,x]|$.
-
-Example \emph{t19} consists of two loops that decrement a variable
-$i$.  In the first loop, $i$ is decremented down to 100 and in the
-second loop $i$ is increment further down to $-1$.  However, between
-the loops we assign $i=i+k+50$.  So in total the program performs $50
-+ |[-1,i]| + |[0,k]|$ iterations.  Our analysis finds this tight bound
-because our amortized analysis naturally takes into account the
-relation between the two loops.  Techniques that do not use
-amortization derive a more conservative bound such as $50 + |[-1,i]| +
-|[0,k]| + |[100,i]|$.
-
-Example \emph{t20} shows how we can handle programs in which bounds
-contain absolute values like $|x-y|$.  The first loop body is only
-executed if $x<y$ and the second loop body is only executed if $y<x$.
-The analyzer finds a tight bound.
-
-At first sight, Example \emph{t30} appears to be a simple loop that
-decrements the variable $x$ down to zero.  However, a closer look
-reveals that the loop actually decrements both input variables $x$ and
-$y$ down to zero before terminating.  In the loop body, first $x$ is
-decremented by one.  Then the values of the variables $x$ and $y$ are
-switched using the local variable $t$ as a buffer.  Our analysis
-infers the tight bound $|[0,x]|+|[0,y]|$.
-
-Example \emph{t47} demonstrates how we can use integers as Booleans to
-amortize the cost of loops that depend on boolean flags.  The outer
-loop is executed as long as the variable flag is ``true'', that is,
-\code{flag>0}.  Inside the loop, there is a conditional that either
-(if $n>0$) decrements $n$ and assigns \code{flag=1}, or (if $n\geq0$)
-leaves n unchanged and assigns \code{flag=0}.  The analyzer computes
-the tight bound $1 + |[0, n]|$.  The potential in the loop invariant
-is $|[0,\mathit{flag}]| + |[0, n]|$.  In the \emph{then} branch of
-the conditional, we use the potential $|[0, n]|$ and the fact that
-$n>0$.  In the \emph{else} branch, we use the potential
-$|[0,\mathit{flag}]|$ and the fact that $\mathit{flag}=1$.
-
-\begin{figure*}
- \setlength{\progwidth}{.24\linewidth}
-  \centering
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (n>x) {
-    if (m>y)
-      y = y+1;
-    else
-      x = x+1;
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[y, m]|$
-\\[.7\baselineskip]
-      {\bf fig2\_1}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x<n) {
-    if (z>x)
-      x=x+1;
-    else
-      z=z+1;
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[z, n]|$
-\\[.7\baselineskip]
-      {\bf fig2\_2}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (x<n) {
-    while (y<m) {
-      if (*) break;
-      y=y+1;
-    }
-    x=x+1;
-  }
-   \end{lstlisting}
-
-$|[x, n]| + |[y, m]|$
-\\[.7\baselineskip]
-      {\bf nested\_multiple}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  x=0;
-  while (x<n) {
-    x=x+1;
-    while (x<n) {
-      if (*) break;
-      x=x+1;
-    }
-  }
-   \end{lstlisting}
-
-$|[0, n]|$
-\\[.7\baselineskip]
-      {\bf nested\_single}
-    \end{center}
-  \end{minipage}
-
-   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (a).}
-  \label{fig:cat2a}
-\end{figure*}
-
-\begin{figure*}
- \setlength{\progwidth}{.24\linewidth}
-  \centering
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  x=0;
-  while (x<n) {
-    if (*) break;
-    x=x+1;
-  }
-  while (x<n)
-    x=x+1;
-   \end{lstlisting}
-
-$|[0,n]|$
-\\[.7\baselineskip]
-      {\bf sequential\_single}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  x=0; y=0;
-  while (x<n) {
-    if (y<m)
-      y=y+1;
-    else
-      x=x+1;
-  }
-   \end{lstlisting}
-$|[0, m]| + |[0, n]|$
-\\[.7\baselineskip]
-      {\bf simple\_multiple}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  x=0;
-  while (x<n) {
-    if (*)
-      x=x+1;
-    else
-      x=x+1;
-  }
-   \end{lstlisting}
-
-$|[0,n]|$
-\\[.7\baselineskip]
-      {\bf simple\_single}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  x=0; y=0;
-  while (*) {
-    if (x<N) {
-      x=x+1; y=y+1;
-    } else if (y<M ) {
-      x=x+1; y=y+1;
-    } else
-      break;
-  }
-   \end{lstlisting}
-
-$|[0, M]| + |[0, N]|$
-\\[.7\baselineskip]
-      {\bf simple\_single\_2}
-    \end{center}
-  \end{minipage}
-
-   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b).}
-  \label{fig:cat2b}
-\end{figure*}
-
-\begin{figure*}
- \setlength{\progwidth}{.24\linewidth}
-  \centering
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  assert n>0;
-  assert m>0;
-  va = n; vb = 0;
-  while (va>0 && *) {
-    if (vb<m) {
-      vb=vb+1;
-      va=va-1;
-    } else {
-      vb=vb-1;
-      vb=0;
-    }
-  }
-   \end{lstlisting}
-$1 + 2|[0, n]|$
-\\[.7\baselineskip]
-      {\bf fig4\_2}
-    \end{center}
-  \end{minipage}
-%
-%
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  assert (0<m);
-  i = n;
-  while (i>0 && *) {
-    if (i<m)
-      i=i-1;
-    else
-      i=i-m;
-  }
-   \end{lstlisting}
-$|[0, n]|$
-\\[.7\baselineskip]
-      {\bf fig4\_4}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  assert (0 < m < n);
-  i=m;
-  while (0<i<n) {
-    if (dir==fwd) i++;
-    else i--;
-  }
-   \end{lstlisting}
-$---$
-\\[.7\baselineskip]
-      {\bf fig4\_5}
-    \end{center}
-  \end{minipage}
-
-   \caption{Examples from~\cite{GulwaniJK09}.}
-  \label{fig:cat2c}
-\end{figure*}
-
-\begin{figure*}
- \setlength{\progwidth}{.24\linewidth}
-  \centering
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  i=0;
-  while (i<n) {
-    j=i+1;
-    while (j<n) {
-      if (*) {
-        tick(1);
-        j=j-1; n=n-1;
-      }
-      j=j+1;
-    }
-    i=i+1;
-  }
-   \end{lstlisting}
-$|[0, n]| \text{ ticks}$
-\\[.7\baselineskip]
-      {\bf ex1}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (n>0 && m>0) {
-    n--; m--;
-    while (nondet()) {
-      n--; m++;
-    };
-    tick(1);
-  }
-   \end{lstlisting}
-$|[0,n]|$ ticks
-\\[.7\baselineskip]
-      {\bf ex2}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  while (n>0) {
-    t = x;
-    n=n-1;
-    while (n>0) {
-      if (*) break;
-      n=n-1;
-    }
-  }
-   \end{lstlisting}
-$|[0, n]|$
-\\[.7\baselineskip]
-      {\bf ex3}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-  flag=1;
-  while (flag>0) {
-    flag=0;
-    while (n>0 && *) {
-      n=n-1;
-      flag=1;
-    }
-  }
-   \end{lstlisting}
-$1 + 2|[0, n]|$
-\\[.7\baselineskip]
-      {\bf ex4}
-    \end{center}
-  \end{minipage}
-
-   \caption{Examples from~\cite{GulwaniZ10}.}
-  \label{fig:cat3a}
-\end{figure*}
-
-\paragraph{From  the Literature}
-
-Our analyzer can derive almost all linear bounds for programs that
-have been described as challenges in the literature on bound
-generation.  We found only one program with a linear bound for
-which our analyzer could not find a tight bound:  Example
-(\emph{fig4\_5}) from~~\cite{GulwaniJK09} requires \emph{path-sensitive
-  reasoning} to derive a bound.
-
-Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
-al~\cite{GulwaniMC09}.  They are both handled by the SPEED tool but
-require inference of a \emph{disjunctive invariant}.  In the abstract
-interpretation community, these invariants are known to be notoriously
-difficult to handle.
-%
-In Example \emph{fig2\_1} we have one loop that first increments
-variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
-derive the tight bound $|[x, n]| + |[y, m]|$.
-%
-Example \emph{fig2\_2} is more tricky and trying to understand how it
-works may be challenging.  However, with the amortized analysis in
-mind, using the potential transfer reasoning, it is almost trivial to
-prove a bound.  While the SPEED tool has to find a fairly
-involved invariant for the loop, our tool is simply reasoning locally
-and works without any clever tricks. We obtain the tight bound $|[x,
-n]| + |[z, n]|$.
-
-Example \emph{nested\_multiple} is similar to Example \emph{fig2\_1}.
-Instead of incrementing variable $y$ in the outer loop, $y$ is here
-potentially incremented multiple times in each iteration of the outer
-loop.  The idea of Example \emph{nested\_single} is similar.  However,
-instead of incrementing variable $y$ in the inner loop, we increment
-$x$, the counter variable of the outer loop. Our analyzer derives a
-tight bound for both programs.  Note that a star \emph{*} in a
-branching condition denotes an arbitrary boolean condition that might
-of course change while iterating (non-deterministic choice).
-
-Example \emph{sequential\_single} is like Example
-\emph{nested\_single}.  The only difference is that the inner loop of
-\emph{nested\_single} is now evaluated after the outer loop.  Example
-\emph{simple\_multiple} is a variant of Example \emph{fig2\_1} and
-\emph{simple\_single} is a simple variant of \emph{nested\_single}.
-We derive tight bounds for all aforementioned programs.
-
-Example \emph{simple\_single\_2} uses conditionals and a \code{break}
-statement to control loop iterations.  If $x<N$ then variables $x$ and
-$y$ are incremented.  Otherwise, if $y<M$ then the same increment is
-executed.  If $y\geq M$ and $x\geq N$ then the loop is terminated with
-a break.  Our tool computes the bound $|[0, M]| + |[0, N]|$.
-This bound is tight in the sense that there are inputs (such as $M =
--100$ and $N = 100$) for which the bound precisely describes the
-execution cost.  However, SPEED can compute the more precise bound
-$\max(N,M)$.  We currently cannot express this bound in our system.
-
-Example \emph{fig4\_2} from~\cite{GulwaniJK09} is quite involved.
-Amortized reasoning helps to understand how we derive the bound $1 +
-2|[0, n]|$.  We start with potential $1 + 2|[0, n]|$ and use the fact
-that $vb=0$ to establish the potential $1 + 2|[0, n]| + |[0,vb]|$ that
-serves as a loop invariant.  In the \emph{if branch} of the
-conditional, we use the constant potential $1$ of the invariant to pay
-for the potential of $|[0,vb]|$.  Since we also know that $|[0, n]|>0$
-we obtain constant potential $2$ that we use to pay for the loop
-iteration ($1$ unit) and to establish the loop invariant again ($1$
-unit).  In the \emph{else} branch, we use the potential $|[0,vb]|$
-and the fact $vb>0$ to obtain $1$ potential units to pay for the
-loop iteration.
-
-In Example \emph{fig4\_4} it is essential that $m$ is positive.  That
-ensures that we can obtain constant potential for the interval size
-$|[0,i]|$ in the \emph{else} branch of the conditional since $|[0,i]|$
-decreases.  Example \emph{fig4\_5} is an examples that we can not
-handle automatically.  The execution is bounded because the boolean
-value of the test \code{dir==fwd} does not change during the iteration
-of the loop.  As a result, the variable $i$ is either counted down to
-$0$ or up to $n$.  Our tool cannot handle Example \emph{fig4\_5}
-because we don't do path sensitive reasoning.  Note however that it
-would be more efficient to move the test \code{dir==fwd} outside of
-the loop (this would be also done by an optimizing compiler).  The
-resulting program can then be analyzed by our tool.
-
-Example \emph{ex1} from~\cite{GulwaniZ10} specifically focuses on the
-code in the \emph{if} statement.  So we
-insert \code{tick(1)} inside the if statement to derive a bound on
-the number of times the code in the if statement is executed.  Note
-that we cannot derive a bound for the whole program since the outer
-loop is executed a quadratic number of times.  Nevertheless it is
-straightforward to derive a bound on the number of ticks using the
-amortized approach: In the if statement we know that $n>0$ and assign
-\code{n=n-1}.  So we can use the potential of the interval size $|[0,n]|$
-to pay for the tick.
-
-Similar to Example \emph{ex1}, Example \emph{ex2}
-form~\cite{GulwaniZ10} focuses on the number of iterations of the
-\emph{outer loop}.  While the whole program is not terminating, the
-number of iterations of the outer loop is bounded by $|[0,n]|$.
-Finally, Example \emph{ex2} is similar to example
-\emph{nested\_single}, and \emph{ex3} is a variant of Example
-\emph{t47}.
-
-\begin{figure*}
- \setlength{\progwidth}{.28\linewidth}
-  \centering
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-void count_down (int x) {
-  int a = x;
-  if (a>0) {
-    a = a-1;
-    count_down(a);
-  }
-}
-int copy (int x, int y) {
-  if (x>0) {
-    x = x-1;
-    y = y+1;
-    y=copy(x,y);
-  };
-  return y;
-}
-void start (int x,int y) {
-  y = copy (x,y);
-  count_down(y);
-}
-   \end{lstlisting}
-
-$3 + 2|[0, x]| + |[0, y]|$
-\\[.7\baselineskip]
-      {\bf t37}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-void count_down (int x,int y)
-{ int a = x;
-  if (a>y) {
-    a = a-1;
-    count_up(a,y);
-  }
-}
-
-void count_up (int x, int y)
-{ int a = y;
-  if (a+1<x) {
-    a = a+2;
-    count_down(x,a);
-  }
-}
-
-void start (int y, int z) {
-  count_down(y,z);
-}
-   \end{lstlisting}
-
-$1.33 + 0.67 |[z,y]|$
-\\[.7\baselineskip]
-      {\bf t39}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-void produce () {
-  while (x>0) {
-    tick(-1); x=x-1; y=y+1;
-  }
-}
-void consume () {
-  while (y>0) {
-    y=y-1; x=x+1; tick(1);
-  }
-}
-void start (int y, int z) {
-  consume(); produce(); consume();
-}
-   \end{lstlisting}
-
-$|[0, y]|$ ticks
-\\[.7\baselineskip]
-      {\bf t46}
-    \end{center}
-  \end{minipage}
-
-   \caption{Programs with (recursive) functions}
-  \label{fig:cat3}
-\end{figure*}
-
-\begin{figure*}
- \setlength{\progwidth}{.32\linewidth}
-  \centering
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-int srch(
-  int t[], int n,  /* haystack */
-  int p[], int m,  /* needle */
-  int b[]
-) {
-  int i=0, j=0, k=-1;
-
-  while (i < n) {
-    while (j >= 0
-        && t[i]!=p[j]) {
-      k = b[j];
-      assert(k > 0);
-      assert(k <= j + 1);
-      j -= k;
-    }
-    i++, j++;
-    if (j == m)
-      break;
-  }
-  return i;
-}
-   \end{lstlisting}
-
-$1 + 2|\inter 0 n|$
-\\[.7\baselineskip]
-      {\bf Knuth-Morris-Pratt}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-int gcd(int x, int y) {
-  if (x <= 0)
-    return y;
-  if (y <= 0)
-    return x;
-
-  for (;;) {
-    if (x>y)      x -= y;
-    else if (y>x) y -= x;
-    else return x;
-  }
-}
-   \end{lstlisting}
-
-$|\inter 0 x| + |\inter 0 y|$
-\\[.7\baselineskip]
-      {\bf Greatest Common Divisor}
-    \end{center}
-  \end{minipage}
-%
-%
-  \begin{minipage}[b]{\progwidth}
-    \begin{center}
-   \begin{lstlisting}
-void qsort(
-  int a[], int lo, int hi
-) {
-  int m1, m2, n;
-
-  if (hi - lo < 1) return;
-
-  n = nondet(); /* partition */
-  assert( n > 0 );
-  assert( lo + n <= hi );
-
-  m1 = n + lo;
-  m2 = m1 - 1;
-
-  qsort(a, m1, hi);
-  qsort(a, lo, m2);
-}
-
-void start(int a[], int len) {
-  qsort(a, 0, len);
-}
-   \end{lstlisting}
-
-$1 + 2 |\inter 0 {\code{len}}|$
-\\[.7\baselineskip]
-      {\bf Quick Sort}
-    \end{center}
-  \end{minipage}
-
-   \caption{Well-Known Algorithms}
-  \label{fig:wknown}
-\end{figure*}
-
-\paragraph{Recursive Functions}
-
-Our approach can naturally deal with mutually-recursive functions.
-The recursion patterns can be exactly the same that are used in
-iterations of loops.  In the following, we present three simple
-examples that illustrate the analysis of functions.
-
-Example \emph{t37} illustrates that the analyzer is able to perform
-inter-procedural size tracking.  The function \code{copy} adds the
-argument $x$ to the argument $y$ if $x$ is positive.  However, this
-addition is done in steps of $1$ in each recursive call.  The function
-\code{count\_down} recursively decrements its argument down to $0$.
-The derived bound $3 + 2|[0, x]| +2|[0, y]|$ is for the function
-\code{main} in which we first add $x$ to $y$ using the function
-\code{copy} and then count down the variable $y$ using the function
-\code{count\_down}.  The derived bound is tight.
-
-Example \emph{t39} uses mutual recursion.  The function
-\code{count\_down} is similar to the function with the same name in
-Example \emph{t37}.  However, we do not count down to $0$ but to a
-variable $y$ that is passed as an argument and we call the function
-\code{count\_up} afterwards.  The function \code{count\_up} is dual to
-\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
-\code{count\_down}.  For the function \code{main}, which calls
-\code{count\_down(y,z)}, the analyzer computes the tight bound $1.33 +
-0.67 |[z,y]|$.
-
-Example \emph{t46} shows a program that uses and returns resources.
-Again, we use the function \code{tick} to
-describe the resource usage.  The function \code{produce} produces
-$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
-receives one resource unit.  Similarly, the function \code{consume}
-consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
-$|[0,y]|$ for the function \code{main}.  This is only possible since
-amortized analysis naturally tracks the size changes to the variables
-$x$ and $y$, and the interaction between \code{consume} and
-\code{produce}.
-
-\begin{table*}
-\footnotesize
-
-\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
-\newcommand{\file}[2][c]{%
-  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
-\renewcommand{\max}[0]{{\rm mx}}
-\renewcommand{\min}[0]{{\rm mn}}
-
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\centering
-\begin{tabular}{r|C{1.1cm}c|C{1.8cm}c|C{1.9cm}c|C{1.8cm}|C{2.65cm}}
-
-File &
-\multicolumn{2}{c|}{KoAT} &
-\multicolumn{2}{c|}{Rank} &
-\multicolumn{2}{c|}{LOOPUS} &
-SPEED &
-\toolname
-\\
-
-%--------------------------------------------
-\hhline{=|==|==|==|=|=} \texttt{gcd.c} &
-
-? &
-&
-
-% $((((1+1)+1)+((-5+(2{\cdot}y))+(2{\cdot}x)))+(-3+(2{\cdot}y)))+1$ &
-$(((2{+}1)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{kmp.c} &
-
-? &
-&
-
-% $((((1+1)+(n+(-1{\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+((n+(n{\cdot}j))+((-1+(-1{\cdot}j)){\cdot}i)))+1$ &
-$(((2{+}(n{+} \dots$ &
-$O(n^2)$ &
-
-% $\max(n, 0) + \max(0, (1 + \max(n, 0))) + \max(0, (1 + \max(n, 0)))$ &
-$\max(n, 0) \dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \texttt{qsort.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1 {+} 2 |\inter 0 {\code{len}}|$
-\\
-
-%--------------------------------------------
-% \hline \file{\dir{speed\_pldi09}{fig1.c}} &
-%
-% ? &
-% &
-%
-% --- &
-% &
-%
-% $2 \max(n, 0)$ &
-% $O(n)$ &
-%
-% God damn it &
-%
-% $1 + 2 |\inter 0 n|$
-% \\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_2.c}} &
-
---- &
-&
-
-% $((((1+1)+n)+(n+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}n)\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + n$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_4.c}} &
-
---- &
-&
-
-% $((((1+1)+(-1+m))+((1+n)+(-1{\cdot}m)))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\frac n m + m$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi09}{fig4\_5.c}} &
-
-$28d + 7g + 27$ &
-$O(n)$ &
-
-% $((((1+1)+(-1+n))+m)+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, n-m)$ &
-
----
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex1.c}} &
-
---- &
-&
-
---- &
-&
-
---- &
-&
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex3.c}} &
-
---- &
-&
-
-% $((((1+1)+(-1+n))+(-1+n))+1)+1$ &
-$(((2{+}(-1\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_pldi10}{ex4.c}} &
-
-$110 a + 33$ &
-$O(n)$ &
-
---- &
-&
-
---- &
-&
-
-$n + 1$ &
-
-$1 {+} 2 |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{fig2\_1.c}} &
-
-$9a + 9b + \dots$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{+}((-y\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, m{-}y)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{fig2\_2.c}} &
-
-$6a + 9b + 3c + 5$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((-1+(-1{\cdot}z))+(-1{\cdot}x))+(2{\cdot}n)))+1$ &
-$((2{-}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (x {+} 1 {+} -z)) + \max(0, (n - x)))) + \max(0, (n - x))$ &
-$\max(0, (x + 1 {-}z)\dots$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, n{-}z)$ &
-
-$|\inter x n| {+} |\inter z n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{nstd\_multiple.c}} &
-
---- &
-&
-
-% $(((1+1)+((-1{\cdot}x)+n))+(((y{\cdot}x)+((-1{\cdot}y){\cdot}n))+(((-1{\cdot}x)+n){\cdot}m)))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n^2)$ &
-
-$\max(0, m {-} y) + \max(0, n {-} x)$ &
-$O(n)$ &
-
-$\max(0, n{-}x) + \max(0, m{-}y)$ &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{nstd\_single.c}} &
-
-$48 b + 16$ &
-$O(n)$ &
-
-% $((((1+1)+((-1+(-1{\cdot}x))+n))+((-1+(-1{\cdot}x))+n))+1)+1$ &
-$(((1{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(0, \! n {-} 1)\dots$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{sqntl\_single.c}} &
-
-$21 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{}-x{+}n\dots$ &
-$O(n)$ &
-
-$2{\cdot}\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_multiple.c}} &
-
-$9 c + 10 d + 7$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 m| {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_single2.c}} &
-
-$20 d + 12 c + 17$ &
-$O(n)$ &
-
---- &
-&
-
-$\max(n, 0) + \max(m, 0)$ &
-$O(n)$ &
-
-$n + m$ &
-
-$|\inter 0 n| {+} |\inter 0 m|$
-\\
-
-%--------------------------------------------
-\hline \file{\dir{speed\_popl10}{smpl\_single.c}} &
-
-$4 b + 6$ &
-$O(n)$ &
-
-% $(((1+1)+((-1{\cdot}x)+n))+((-1{\cdot}x)+n))+1$ &
-$((2{-}x{+}n\dots$ &
-$O(n)$ &
-
-$\max(n, 0)$ &
-$O(n)$ &
-
-$n$ &
-
-$|\inter 0 n|$
-\\
-
-%--------------------------------------------
-\hline \file{\texttt{t07.c}} &
-
-? &
-&
-
-$2+x$ &
-$O(n)$ &
-
-% $\max(x, 0) + \max(0, (y +  2 {\cdot} \max(x, 0))) + \max(0, ( 2 {\cdot} \max(x, 0) + \max(y, 0)))$ &
-$\max(x, 0)\dots$ &
-$O(n)$ &
-
-? &
-
-$1 {+} 3 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t08.c} &
-
-? &
-&
-
-% $(((1+1)+-1)+(z+(-1{\cdot}y)))+1$ &
-$((2{+}z{-}y\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y + -2)) + \max(0, (z - y)))) + \max(0, (z - y))$ &
-$\max(0, \! y{-}2)\dots$ &
-$O(n)$ &
-
-? &
-
-$1.33 |\inter y z| {+} 0.33 |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t10.c} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+x))+((-1{\cdot}y)+x))+1$ &
-$((2{-}y{+}x\dots$ &
-$O(n)$ &
-
-$\max(0, x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t11.c} &
-
-? &
-&
-
-% $(((1+1)+((-1{\cdot}y)+m))+((-1{\cdot}x)+n))+1$ &
-$((2{-}y{+}m\dots$ &
-$O(n)$ &
-
-$\max(0, n {-} x)) + \max(0, m {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x n| {+} |\inter y m|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t13.c} &
-
-? &
-&
-
-% $((((1+1)+-1)+(((((-1/2){\cdot}y)+((1/2){\cdot}(y{\cdot}y)))+(((-1/2)+y){\cdot}x))+((1/2){\cdot}(x{\cdot}x))))+-1)+1$ &
-$(((1{+}y^2/2\dots$ &
-$O(n^2)$ &
-
-$2{\cdot}\max(x, 0) + \max(y, 0)$ &
-$O(n)$ &
-
-? &
-
-$2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t15.c} &
-
-? &
-&
-
-% $(((1+1)+(-1+x))+((-1{\cdot}y)+x))+1$ &
-$((1{+}x\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$|\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t16.c} &
-
-? &
-&
-
-% $(((1+1)+((-1+(-99{\cdot}y))+(101{\cdot}x)))+((-1{\cdot}y)+x))+1$ &
-$((-99{\cdot}y\dots$ &
-$O(n)$ &
-
---- &
-&
-
-? &
-
-$101 |\inter 0 x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t19.c} &
-
-? &
-&
-
-% $(((1+1)+(151+k))+(-100+i))+1$ &
-$((153{+}k\dots$ &
-$O(n)$ &
-
-$\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$ &
-$O(n)$ &
-
-? &
-
-$50 {+} |\inter {-1} i| {+} |\inter 0 k|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t20.c} &
-
-? &
-&
-
-% $((1+1)+((-1{\cdot}y)+x))+1+((1+1)+(y+(-1{\cdot}x)))+1$ &
-$(2{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, (\max(0, (y - x)) + \max(0, (x - y)))) + \max(0, (y - x))$ &
-$2 {\cdot} \max(0, \! y {-} x) \! + \max(0, \! x {-} y)$ &
-$O(n)$ &
-
-? &
-
-$|\inter x y| {+} |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t27.c} &
-
-? &
-&
-
---- &
-&
-
-% $\max(0, (1000 +  1000 {\cdot} \max(0, -n) + \max(0, (y + -99)))) + \max(0, -n)$ &
-$10^3 \max(0, \discretionary{}{}{} -n) \! \dots$ &
-$O(n)$ &
-
-? &
-
-$0.01|\inter n y| {+} 11 |\inter n 0|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t28.c} &
-
-? &
-&
-
-% $(((1+1)+-1)+((-1{\cdot}y)+x))+1$ &
-$((1{-}y{+}x\dots$ &
-$O(n)$ &
-
-% $\max(0, ( 1000 {\cdot} \max(0, (x - y)) + \max(y, 0))) + \max(0, (x - y)) + \max(0, -x)$ &
-$10^3 \,\max(0, x - y)\dots$ &
-$O(n)$ &
-
-? &
-
-$|\inter x 0| {+} |\inter 0 y| \discretionary{}{}{} {+} 1002 |\inter y x|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t30.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t37.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$3 {+} 2 |\inter 0 x| {+} |\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t39.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$1.33 {+} 0.67 |\inter z y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t46.c} &
-
-? &
-&
-
---- &
-&
-
---- &
-&
-
-? &
-
-$|\inter 0 y|$
-\\
-
-%--------------------------------------------
-\hline \texttt{t47.c} &
-
-? &
-&
-
-$4+n$ &
-$O(n)$ &
-
-$1 + \max(n, 0)$ &
-$O(n)$ &
-
-? &
-
-$1 {+} |\inter 0 n|$
-\\
-
-%--------------------------------------------
-\end{tabular}
-\caption{
-  Comparison of the bounds generated by KoAT, Rank,
-  LOOPUS, SPEED, and our tool $\toolname$ on several
-  challenging linear examples. Results for KoAT and SPEED were extracted
-  from previous publications because KoAT cannot take C programs as
-  input in its current version and SPEED is not available.
-  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
-  marked with --- indicate that the tool failed to produce a result. We write
-  $\max(a,b)$ for the maximum of $a$ and $b$.
-}
-\label{tab:eval}
-\end{table*}
-
-\paragraph{Well-Known Algorithms}
-
-\pref{fig:wknown} shows well-known algorithms that can be automatically
-analyzed in our framework.  Example \emph{Knuth-Morris-Pratt} shows
-the search function of the Knuth-Morris-Pratt algorithm for string
-search.  To derive a bound for this algorithm we have add an assertion
-that indicates the bounds of the values that are stored in the array
-\code{b}.  Using this information we derive a tight linear bound.
-
-Example \emph{Greatest Common Divisor} is the usual implementation of
-the GCD algorithm.  We automatically derive a linear bound.  Note that
-the two tests at the beginning that check if the inputs are positive
-are essential.  Finally, Example \emph{Quick Sort} shows a skeleton of
-the quick sort sorting algorithm.  Since we can only derive linear
-bounds we left out the inner loop that swaps the array elements and
-determines the position \code{n+lo} of the pivot.  We only assert that
-\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
-
-}\fi
-
-\bibliographystyle{abbrvnat}
-\iffull{
+
+\bibliographystyle{plain}
 \bibliography{lit}
-}
-{
-\sectskip
-\bibliography{lit}
-}
 

### 2015-09-07 cfg/main.tex
some typos

@@
 
-A control flow graph (CFG) is a set a graph $G=(V,E,\labels)$ with
+A control flow graph (CFG) is a graph $G=(V,E,\labels)$ with
 statement-labeled nodes. Here, $V$ is a set of nodes,
@@
 with starting state $\state$, the assertions in $\Gamma$
-hold, $Q' {\ge} 0$, and at least $Q(\state)$ resources are available then the
+hold, and at least $Q(\state)$ resources are available then the
 evaluation does not run out of resources and, if the execution
@@
 Let $G = (V,E,\labels)$ be a CFG. A potential annotation $\mathcal{P}$
-for $G$ is mapping from notes $v \in V$ to pairs
+for $G$ is mapping from nodes $v \in V$ to pairs
 $\left<(\Gamma,Q);(\Gamma',Q')\right>$ of quantitative assertions so
